### 2020-01-01 probabilisticProgrammingMartingales.tex
Initial commit
Includes first draft of proof for AST with ranking function in a
language without score.

\documentclass{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{hyperref}
\usepackage{verbatim}
\newcommand{\Y}{\textsf{Y}}
\newcommand{\sample}{\textsf{sample}}
\DeclareMathOperator{\red}{red}
\begin{document}
Following your suggestion, I will be providing a criterion for termination of programs in PPCF\cite{ppcf} based on ranking supermartingales. As it's more convenient for this proof, a sampling-based semantics will be used instead of the original distributional semantics. I assume some roughly applicable equivalence is proven somewhere, but it doesn't seem that hard anyway.

\section{Ranking functions}
Given a probabilistic program (i.e. a term $M$), in order to construct a supermartingale to prove its a.s. termination, a function to assign values to each reachable program state is necessary. Define a ranking function on $M$ to be a measurable function $f:Rch(M) \to \mathbb{R}$ such that
\begin{itemize}
    \item $f(N) > 0$ for all $N$ not in normal form
    \item $f(N) = 0$ for all $N$ in normal form
    \item $f(E[\Y N]) \geq 1 + f(E[N(\Y N)])$
    \item $f(E[\sample]) \geq \int_0^1 f(E[\underline{x}]) \, \mathrm{d}x$
    \item $f(E[R]) \geq f(E[N])$ for any other redex $R$, where $R \to N$
\end{itemize}

\section{Sampling semantics}
Let $ I = [0,1] \subset \mathbb{R} $, and let $S = I^{\mathbb{N}}$, with the $\sigma$-algebra and measure given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element. The maps $\pi_h:S \to I, \quad \pi_t:S \to S$ popping the first element are then measurable.

The one-step reduction is given by the function $\red : \Lambda \times S \to \Lambda \times S$ where
\begin{equation}
\red(M,s) = \left\{
    \begin{array}{ll}
        (E[N],s) & \text{if } M = E[R], R \to N \text{ and } R \neq \sample \\
        (\pi_h(s),\pi_t(s)) & \text{if } M =  \sample \\
        (M,s) & \text{if } M \text{ normal form}
    \end{array} \right .
\end{equation}
The limit $\red^\infty$ can then be defined as a partial function as $\lim_{n \to \infty} \red^n(M,s)$ whenever that sequence becomes constant by reaching a normal form.

\begin{comment} I'm not sure whether this section will be needed any more.
\section{Recursive steps}
Define a reduction relation on PPCF terms that excludes the $\Y$ combinator (``nonrecursive reduction''), i.e. $E[M] \dashrightarrow E[N] \text{ if } M \to N$ for any redex $M$ not of the form $\Y M'$.

This reduction relation is strongly normalizing because of the simple type system, so for any term $M$ there is some bound $b$ such that nonrecursive reduction of $M$ must terminate in at most $b$ steps. Define $\red_{nr} : \Lambda \times S \to \Lambda \times S$ like $\red(M,s)$, but using nonrecursive reduction instead of reduction, and let $\red_Y(M,s) = \red^{b_M}_{nr}(M,s)$. The sequence $(r_\Lambda(M,s,n), r_S(M,s,n)) = r(M,s,n) = \left(\red_Y \circ \red \right)^n(\red_Y(M,s))$ is then a subsequence of $\left( \red^n(M,s) \right)_n$ consisting of only terms which are either normal form or of the form $E[\Y N]$ for some $N$.
\end{comment}

\section{Supermartingales}
Given a term $M$ and a ranking function $f$ for it, define random variables on the probability space $S$ (where $s$ is a random variable) by
\begin{align*}
(M_n,s_n) & = \red^n(M,s) \\
y_0 & = 0 \\
y_{n+1} & = \text{min} \{ k>y_n | M_k \text{ normal form or of the form } E[\Y N] \} \\
M'_n & = M_{y_n} \\
X_n & = f(M'_n)
\end{align*}
and define a filtration $\mathcal{F}_n = \sigma(M_k, k \leq n)$ (i.e. all the samples used up to step $n$).

The expectation of $f(M_{n+1})$ given $\mathcal{F}_n$ is trivially less than or equal to $f(M_n)$ in the cases that $M_n \neq E[\sample]$, and in the case of $\sample$,
\begin{align*}
& \mathbb{E}[f(M_{n+1}) | \mathcal{F}_n] \\
= & \mathbb{E}[f(M_{n+1}) | M_n = E[\sample],\, \mathcal{F}_n] \\
= & \mathbb{E}[f(E[\pi_h(s_n)]) | \mathcal{F}_n] \\
= & \int_0^1 f(E[\underline x]) \, \mathrm{d} x \qquad & \text{as }s_n\text{ is independent of } \mathcal{F}_n \\
\leq & f(E[\sample]) \qquad & \text{by assumption on } f \\
= & f(M_n),
\end{align*}
therefore the values of the ranking function $f(M_n)$ are a supermartingale with respect to $\mathcal{F}_n$.

Given $M'_n$, there is some finite bound on the number of reduction steps that can take place from $M'_n$ without a $\Y$-reduction step, because of the type system, therefore $y_{n+1}$ is (conditional on $\mathcal{F}_{y_n+1}$) a bounded stopping time, therefore $\mathbb{E}[f(M_{y_{n+1}}) | \mathcal{F}_{y_n+1}] \leq f(M_{y_n+1})$. If $f(M_{y_n}) > 0$, then $M_{y_n} = E[\Y N]$ for some $E, N$, therefore $M_{y_n+1} = N (\Y N)$ and $f(M_{y_n+1}) \leq f(M_{y_n}) - 1$, therefore if $X_n > 0$, $\mathbb{E}[X_{n+1} | \mathcal{F}_{y_n+1}] \leq X_n - 1$.

The expectation of $X_n$ therefore must tend to 0 as $n \to \infty$, therefore $X_n$ almost surely is less than or equal to 1 eventually, at which point $M'_{n+1}$ must be normal form, therefore $M$ terminates almost surely.

\begin{thebibliography}{9}
\bibitem{ppcf} Thomas Ehrhard, Michele Pagani, and Christine Tasson. Measurable cones and stable, measurable functions: a model for probabilistic higher-order programming. \emph{PACMPL}, 2(POPL):59:1–59:28, 2018. doi: 10.1145/3158147. URL \href{https://doi.org/10.1145/3158147}{https://doi.org/10.1145/3158147}.
\end{thebibliography}
\end{document}

### 2020-01-20 probabilisticProgrammingMartingales.tex
Some files added. Some comments and questions.

@@
 \documentclass{article}
+
+%% BEGIN {Luke's Macros}
+%%
+\newif\ifdraft
+\drafttrue %% To hide all comments and highlighting, just comment this line out
+
+\input{Style/comment}
+
+\usepackage[square]{natbib}
+\setcitestyle{aysep={}}
+
+\definecolor{oxblue}{RGB}{0,33,71}
+\definecolor{oxgold}{HTML}{a0630a}
+\usepackage[final,
+bookmarks,
+bookmarksopen,
+colorlinks,
+final,
+linkcolor=oxred,
+citecolor=oxgold,
+pdfstartview=FitH ]{hyperref}
+%%
+%% END {Luke's Macros}
+
 \usepackage{amsmath}
 \usepackage{amsfonts}
-\usepackage{hyperref}
 \usepackage{verbatim}
@@
 \DeclareMathOperator{\red}{red}
+
 \begin{document}
-Following your suggestion, I will be providing a criterion for termination of programs in PPCF\cite{ppcf} based on ranking supermartingales.
-As it's more convenient for this proof, a sampling-based semantics will be used instead of the original distributional semantics.
+
+\lo{\LaTeX.
+For references, I recommend using the package {\tt natbib} with {\tt apalike} bibliography style (author-year is more reader-friendly), and placing the references in a separate {\tt .bib} file.
+I simply copy-and-paste the bib data from DBLP e.g.~\url{https://dblp.uni-trier.de/pers/hd/o/Ong:C==H=_Luke}.}
+
+\changed[akr]{[Email 22 Dec: I wrote a proof of termination based on ranking supermartingales (a slight strengthening of what you suggested), which is attached (I haven't made a Git repo yet because I don't know where you want it hosted).
+
+I thought some more about termination in the case of a language with a score function. 
+It isn't even clear what termination means in that case.
+None of the papers I've found include both unbounded scoring and a definition of termination.
+\lo{Can you explain your concern? I don't see any problem. See, e.g., the operational semantics in \citep{MakOP20b}; see also \citep{MakOP20a}.} 
+You said that the assumption of a.s. termination was often used. 
+Do you have any examples where it's used in a context with unbounded scoring, as going with whatever definition that uses is probably most useful?
+\lo{See \citep[Sec.~4, pp.~8 \& 9]{Wagner20} -- I am not sure if this addresses your question.}
+
+The case of bounded scoring seems simpler definitionally at least, but still seems like a bad fit for supermartingales, as the martingale property is local in a way scoring isn't. 
+I still need to think more about to what extent it can be made to make sense, though.]}
+
+Following your suggestion, I will be providing a criterion for termination of programs in PPCF \citep{DBLP:journals/pacmpl/EhrhardPT18} based on ranking supermartingales. 
+As it's more convenient for this proof, a sampling-based semantics \lo{You need to provide reference(s) for sampling-based semantics.} will be used instead of the original distributional semantics. 
 I assume some roughly applicable equivalence is proven somewhere, but it doesn't seem that hard anyway.
+\lo{Why not provide a proof as an appendix?}
 
 \section{Ranking functions}
-Given a probabilistic program (i.e.
-a term $M$), in order to construct a supermartingale to prove its a.s.
-termination, a function to assign values to each reachable program state is necessary.
-Define a ranking function on $M$ to be a measurable function $f:Rch(M) \to \mathbb{R}$ such that
+\lo{You should provide the definition of the PPCF: syntax and typing rules.}
+
+Given a probabilistic program (i.e.~a term $M$), in order to construct a supermartingale to prove its a.s.~termination, 
+\lo{\LaTeX.
+Place a space character right after ``.'' whenever the occurrence does not end a sentence (e.g.~{\tt a.s.\~{}termination} and {\tt i.e.\~{}a term}) so that \LaTeX\ does not allocate more white space than it should.}
+a function to assign values to each reachable program state is necessary. Define a ranking function on $M$ to be a measurable function $f:Rch(M) \to \mathbb{R}$ such that
 \begin{itemize}
@@
 Let $ I = [0,1] \subset \mathbb{R} $, and let $S = I^{\mathbb{N}}$, with the $\sigma$-algebra and measure given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
-The maps $\pi_h:S \to I, \quad \pi_t:S \to S$ popping the first element are then measurable.
+The maps $\pi_h:S \to I, \; \pi_t:S \to S$ popping the first element are then measurable.
+
+\lo{Is the induced $\sigma$-algebra the Borel algebra?
+The basis elements of the topology of $S$ have the form $\prod_{i \in I} U_i \times \mathbb{R}^\omega$ where $I$ ranges over finite sets and each $U_i$ is open in $\mathbb{R}$.}
 
@@
 y_0 & = 0 \\
-y_{n+1} & = \text{min} \{ k>y_n | M_k \text{ normal form or of the form } E[\Y N] \} \\
+%y_{n+1} & = \text{min} \{ k>y_n | M_k \text{ normal form or of the form } E[\Y N] \} \\
+y_{n+1} & = \min \{ k \mid k>y_n, M_k \text{ normal form or of the form } E[\Y N] \}\\
 M'_n & = M_{y_n} \\
@@
 \end{align*}
+\lo{I have made a trivial change to the third clause above.}
 and define a filtration $\mathcal{F}_n = \sigma(M_k, k \leq n)$ (i.e. all the samples used up to step $n$).
+
+\lo{So all the random variables just defined are constant functions. E.g.~$M_n$ is a constant function from $S$ to the measurable space of PPCF terms.}
+
+\lo{\LaTeX.
+Use {\tt mid} rather than ``{\tt |}'' as separator in set comprehension.
+Reason: {\tt mid} is a binary operator (and so \LaTeX\ adds white space to both side); ``{\tt |}'' is just a symbol.}
 
@@
 \begin{align*}
-& \mathbb{E}[f(M_{n+1}) | \mathcal{F}_n] \\
-= & \mathbb{E}[f(M_{n+1}) | M_n = E[\sample],\, \mathcal{F}_n] \\
-= & \mathbb{E}[f(E[\pi_h(s_n)]) | \mathcal{F}_n] \\
+& \mathbb{E}[f(M_{n+1}) \mid \mathcal{F}_n] \\
+= & \mathbb{E}[f(M_{n+1}) \mid M_n = E[\sample],\, \mathcal{F}_n] \\
+= & \mathbb{E}[f(E[\pi_h(s_n)]) \mid \mathcal{F}_n] \\
 = & \int_0^1 f(E[\underline x]) \, \mathrm{d} x \qquad & \text{as }s_n\text{ is independent of } \mathcal{F}_n \\
@@
 
-Given $M'_n$, there is some finite bound on the number of reduction steps that can take place from $M'_n$ without a $\Y$-reduction step, because of the type system, therefore $y_{n+1}$ is (conditional on $\mathcal{F}_{y_n+1}$) a bounded stopping time, therefore $\mathbb{E}[f(M_{y_{n+1}}) | \mathcal{F}_{y_n+1}] \leq f(M_{y_n+1})$.
-If $f(M_{y_n}) > 0$, then $M_{y_n} = E[\Y N]$ for some $E, N$, therefore $M_{y_n+1} = N (\Y N)$ and $f(M_{y_n+1}) \leq f(M_{y_n}) - 1$, therefore if $X_n > 0$, $\mathbb{E}[X_{n+1} | \mathcal{F}_{y_n+1}] \leq X_n - 1$.
+Given $M'_n$, there is some finite bound on the number of reduction steps that can take place from $M'_n$ without a $\Y$-reduction step, because of the type system, therefore $y_{n+1}$ is (conditional on $\mathcal{F}_{y_n+1}$) a bounded stopping time, therefore $\mathbb{E}[f(M_{y_{n+1}}) \mid \mathcal{F}_{y_n+1}] \leq f(M_{y_n+1})$.
+If $f(M_{y_n}) > 0$, then $M_{y_n} = E[\Y N]$ for some $E, N$, therefore $M_{y_n+1} = N (\Y N)$ and $f(M_{y_n+1}) \leq f(M_{y_n}) - 1$, therefore if $X_n > 0$, $\mathbb{E}[X_{n+1} \mid \mathcal{F}_{y_n+1}] \leq X_n - 1$.
 
@@
 
+\lo{You should frame the result you have just proved, precisely and formally, using a \LaTeX\ {\tt lemma} or {\tt theorem} environment.}
+
+\lo{Further directions - suggestions to come.}
+
+\bibliographystyle{apalike}
+\bibliography{references}
+
+\iffalse
 \begin{thebibliography}{9}
@@
 \end{thebibliography}
+\fi
+
 \end{document}

### 2020-01-22 probabilisticProgrammingMartingales.tex
Added some thoughts on further directions / next steps.

@@
 \lo{Is the induced $\sigma$-algebra the Borel algebra?
-The basis elements of the topology of $S$ have the form $\prod_{i \in I} U_i \times \mathbb{R}^\omega$ where $I$ ranges over finite sets and each $U_i$ is open in $\mathbb{R}$.}
+The basis elements of the topology of $S$ have the form $\big(\prod_{i \in I} U_i\big) \times \mathbb{R}^\omega$ where $I$ ranges over finite sets and each $U_i$ is open in $\mathbb{R}$.}
 
@@
 \end{equation}
+\lo{Your definition $\red$ is incomplete: it is missing the definition of $R \to N$.}
+
 The limit $\red^\infty$ can then be defined as a partial function as $\lim_{n \to \infty} \red^n(M,s)$ whenever that sequence becomes constant by reaching a normal form.
 
-\begin{comment} I'm not sure whether this section will be needed any more.
-\section{Recursive steps}
+%\begin{comment} 
+\akr{\section{Recursive steps}
+[I'm not sure whether this section will be needed any more.]
+
 Define a reduction relation on PPCF terms that excludes the $\Y$ combinator (``nonrecursive reduction''), i.e.
@@
 The sequence $(r_\Lambda(M,s,n), r_S(M,s,n)) = r(M,s,n) = \left(\red_Y \circ \red \right)^n(\red_Y(M,s))$ is then a subsequence of $\left( \red^n(M,s) \right)_n$ consisting of only terms which are either normal form or of the form $E[\Y N]$ for some $N$.
-\end{comment}
+}
+%\end{comment}
 
@@
 
-\lo{Further directions - suggestions to come.}
+\lo{\textbf{Further directions}
+
+\smallskip
+
+An obvious next step is to extend the result to the $\mathsf{score}$ construct.
+
+The following are highly topical, and could form the basis of an interesting and novel DPhil thesis.
+\begin{enumerate}
+\item Devise methods for proving (positive) a.s.~termination.
+
+- For example, develop a type system satisfying the property: if a term is typable then it is (positively) a.s.~terminating.
+See~\cite{DBLP:conf/ppdp/BreuvartL18,DBLP:conf/esop/LagoG17}.
+
+- Another approach is to develop algorithms that synthesise ranking supermartingales, following, for example, \cite{DBLP:journals/pacmpl/AgrawalC018}.
+
+\item Develop principles (e.g.~in the form of ``proof rules'') for reasoning about (positively) a.s.~termination, in the style of \cite{DBLP:journals/pacmpl/McIverMKK18}.
+
+\item Design algorithms that synthesise probabilistic invariants (\emph{qua} martingales), \`a la \cite{SchreuderO19}; see also \cite{DBLP:journals/pacmpl/HarkKGK20}.
+
+\end{enumerate}}
 

### 2020-01-24 probabilisticProgrammingMartingales.tex
Some thoughts on extension to score, and a.e. termination

@@
 final,
-linkcolor=oxred,
+linkcolor=red,
 citecolor=oxgold,
@@
 
-\lo{\LaTeX.
+\lo{23 Jan:
+
+\paragraph{Some thoughts about extension to $\mathsf{score}$}
+
+- I still think that the operational semantics (reduction semantics) of SPCF (i.e.~with $\mathsf{score}$) is not problematic.
+However the notion of a.e.~termination is interesting and would benefit from further thought.
+
+- I would propose that you use the language SPCF defined in \citep{MakOP20b}.
+They use a big-step reduction semantics.
+I would sugget that, by contrast, you develop a small-step reduction semantics for it (by a mild adaptation of your $\red : \Lambda \times S \to \Lambda \times S$).
+
+- You will find that your example 
+\[
+f \, n = n \oplus_{\frac{1}{2}} (\mathsf{score} \, 3 \, ; \, f \, (n+1))
+\]
+is not definable in SPCF, but a variant (in which the branching probabilities are drawn from the entropy sequence) is.
+OK, we will need to assume that their $\mathbb{S} := \mathbb{R}^\omega$ is replaced by your $S$.
+
+- Consider the notion of a.e.~termination defined in \citep{MakOP20b}\footnote{\url{https://drive.google.com/open?id=1JzxV0lCkZu4d7szQOzkcKspqqo8u2BP0}} (towards the end of \S 3): a program $s$ is a.e.~terminating if the set of entropy sequences where $s$ does not evaluate to a value has measure zero.
+The intuition is that scoring is about model building, and does not by itself ``affect termination''.
+}
+
+\bigskip
+
+\lo{22 Jan:
+\LaTeX.
 For references, I recommend using the package {\tt natbib} with {\tt apalike} bibliography style (author-year is more reader-friendly), and placing the references in a separate {\tt .bib} file.
@@
 
-\lo{\textbf{Further directions}
-
-\smallskip
+\lo{22 Jan: 
+\paragraph{Further directions}
 

### 2020-03-09 probabilisticProgrammingMartingales.tex
Tidy various things and respond to comments
Improve formatting. Replace end of proof with something a bit clearer.
Add an actual introduction and some references.

@@
 \usepackage{amsfonts}
+\usepackage{amsthm}
 \usepackage{verbatim}
-\newcommand{\Y}{\textsf{Y}}
-\newcommand{\sample}{\textsf{sample}}
+\newcommand{\tY}{\textsf{Y}}
+\newcommand{\tif}[3]{\textsf{if }#1\textsf{ then }#2\textsf{ else }#3}
+\newcommand{\tsample}{\textsf{sample}}
+\newcommand{\tscore}{\textsf{score}}
 \DeclareMathOperator{\red}{red}
+
+\newtheorem{theorem}{Theorem}
 
@@
 
-\section{Ranking functions}
-\lo{You should provide the definition of the PPCF: syntax and typing rules.}
+Various theorems about probabilistic programs rely on the assumption that the program terminates almost surely.
+Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \citep{DBLP:journals/pacmpl/McIverMKK18}.
+This paper's contribution is to extend this method to a higher-order setting.
 
-Given a probabilistic program (i.e.~a term $M$), in order to construct a supermartingale to prove its a.s.~termination, 
-\lo{\LaTeX.
-Place a space character right after ``.'' whenever the occurrence does not end a sentence (e.g.~{\tt a.s.\~{}termination} and {\tt i.e.\~{}a term}) so that \LaTeX\ does not allocate more white space than it should.}
-a function to assign values to each reachable program state is necessary. Define a ranking function on $M$ to be a measurable function $f:Rch(M) \to \mathbb{R}$ such that
-\begin{itemize}
-    \item $f(N) > 0$ for all $N$ not in normal form
-    \item $f(N) = 0$ for all $N$ in normal form
-    \item $f(E[\Y N]) \geq 1 + f(E[N(\Y N)])$
-    \item $f(E[\sample]) \geq \int_0^1 f(E[\underline{x}]) \, \mathrm{d}x$
-    \item $f(E[R]) \geq f(E[N])$ for any other redex $R$, where $R \to N$
-\end{itemize}
+\section{Language definition}
+The language SPCF is a simply-typed lambda calculus with sampling of real numbers from $[0,1]$ and unbounded scoring, following \cite{MakOP20b}.
+Types and terms are defined as follows, where $r$ is a real, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ and $\Gamma$ is an environment
+\begin{align*}
+  & \text{types } A, B ::= \textsf{R}  \mid  A \to B \\
+  & \text{values } v ::= x  \mid  \lambda x.s  \mid  \underline{r} \\
+  & \text{terms } s, t ::= v  \mid  t_1 t_2  \mid  \underline{f}(s_1,\dots ,s_n)  \mid  \tY s  \mid  \tif{s < 0}{t_1}{t_2}  \mid  \tsample  \mid  \tscore(s)
+\end{align*}
+\begin{align*}
+  \frac{}{\Gamma ; x:A \vdash x:A} \qquad
+  \frac{\Gamma ; x:A \vdash s : B}{\Gamma \vdash \lambda x.s : A \to B} \qquad
+  \frac{}{\underline{r} : \textsf{R}} \qquad
+  \frac{\Gamma \vdash s:A \to B \quad \Gamma \vdash t : A}{\Gamma \vdash s t : B} \\ \\
+  \frac{\Gamma \vdash s_1:\textsf{R} \dots \Gamma \vdash s_n:\textsf{R}}{\Gamma \vdash \underline{f}(s_1,\dots,s_n) : \textsf{R}} \ f : \mathbb{R}^n \to \mathbb{R} \qquad
+  \frac{\Gamma \vdash s : A \to A}{\Gamma \vdash \tY s : A} \\ \\
+  \frac{\Gamma \vdash c : \textsf{R} \quad \Gamma \vdash s_1 : A \quad \Gamma \vdash s_2 : A}{\Gamma \vdash \tif{c < 0}{s_1}{s_2} : A} \qquad
+  \frac{}{\Gamma \vdash \tsample : \textsf{R}} \qquad
+  \frac{\Gamma \vdash s : \textsf{R}}{\Gamma \vdash \tscore (s) : \textsf{R}}
+\end{align*}
 
-\section{Sampling semantics}
-Let $ I = [0,1] \subset \mathbb{R} $, and let $S = I^{\mathbb{N}}$, with the $\sigma$-algebra and measure given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
+To define the reduction relation, let environments be of the form:
+\begin{align*}
+  E ::= & \, \cdot \mid E t \mid v E \mid \underline{f}(r_1,\dots ,r_{k-1}, E, s_{k+1}, \dots, s_n) \\ & \mid \tY E \mid \tif{E<0}{s_1}{s_2} \mid \tscore (E)
+\end{align*}
+then a term reduces if it is formed by substituting a redex in an environment i.e.
+\begin{align*}
+  E[(\lambda x.s) t] & \to E[s[t/x]] \\
+  E[\underline f (\underline r_1, \dots , \underline r_n)] & \to E[\underline{f(r_1,\dots,r_n)}] \\
+  E[\tY v] & \to E[v (\tY v)] \\
+  E[\tif{\underline r < 0}{s_1}{s_2}] & \to E[s_1] \text{ where }r < 0 \\
+  E[\tif{\underline r < 0}{s_1}{s_2}] & \to E[s_2] \text{ where }r \geq 0 \\
+  E[\tsample] & \to E[\underline r] \text{ where } r \in [0,1] \\
+  E[\tscore(\underline r)] & \to E[\underline r]
+\end{align*}
+
+\subsection{Sampling semantics}
+This version of the reduction relation allows $\tsample$ to reduce to any number in $[0,1]$.
+To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples.
+Let $ I = [0,1] \subset \mathbb{R} $, and let $S = I^{\mathbb{N}}$, with the Borel $\sigma$-algebra and the measure given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
 The maps $\pi_h:S \to I, \; \pi_t:S \to S$ popping the first element are then measurable.
-
-\lo{Is the induced $\sigma$-algebra the Borel algebra?
-The basis elements of the topology of $S$ have the form $\big(\prod_{i \in I} U_i\big) \times \mathbb{R}^\omega$ where $I$ ranges over finite sets and each $U_i$ is open in $\mathbb{R}$.}
 
@@
     \begin{array}{ll}
-        (E[N],s) & \text{if } M = E[R], R \to N \text{ and } R \neq \sample \\
-        (\pi_h(s),\pi_t(s)) & \text{if } M =  \sample \\
+        (E[N],s) & \text{if } M = E[R], R \to N \text{ and } R \neq \tsample \\
+        (E[\underline{\pi_h(s)}],\pi_t(s)) & \text{if } M = E[\tsample] \\
         (M,s) & \text{if } M \text{ normal form}
@@
 \end{equation}
-\lo{Your definition $\red$ is incomplete: it is missing the definition of $R \to N$.}
 
 The limit $\red^\infty$ can then be defined as a partial function as $\lim_{n \to \infty} \red^n(M,s)$ whenever that sequence becomes constant by reaching a normal form.
+A term $M$ terminates for a sample sequence $s$ if the limit $\red^\infty(M,s)$ is defined, and it terminates almost surely if $\mu(\{s \mid M \text{ terminates at } s\}) = 1$.
 
-%\begin{comment} 
-\akr{\section{Recursive steps}
-[I'm not sure whether this section will be needed any more.]
+\paragraph{}
+This definition of almost sure termination is equivalent to that given in \citep{MakOP20b}, although the program semantics is stated in a slightly different way.
+In particular, the argument to $\tscore$ is not relevant to termination (except for the possibility that its argument's evaluation wouldn't terminate).
 
-Define a reduction relation on PPCF terms that excludes the $\Y$ combinator (``nonrecursive reduction''), i.e.
-$E[M] \dashrightarrow E[N] \text{ if } M \to N$ for any redex $M$ not of the form $\Y M'$.
-
-This reduction relation is strongly normalizing because of the simple type system, so for any term $M$ there is some bound $b$ such that nonrecursive reduction of $M$ must terminate in at most $b$ steps.
-Define $\red_{nr} : \Lambda \times S \to \Lambda \times S$ like $\red(M,s)$, but using nonrecursive reduction instead of reduction, and let $\red_Y(M,s) = \red^{b_M}_{nr}(M,s)$.
-The sequence $(r_\Lambda(M,s,n), r_S(M,s,n)) = r(M,s,n) = \left(\red_Y \circ \red \right)^n(\red_Y(M,s))$ is then a subsequence of $\left( \red^n(M,s) \right)_n$ consisting of only terms which are either normal form or of the form $E[\Y N]$ for some $N$.
-}
-%\end{comment}
+\section{Ranking functions}
+Given a probabilistic program (i.e.~a term $M$), in order to construct a supermartingale to prove its a.s.~termination, a function to assign values to each reachable program state is necessary.
+Define a ranking function on $M$ to be a measurable function $f:Rch(M) \to \mathbb{R}$ such that
+\begin{itemize}
+    \item $f(N) > 0$ for all $N$ not in normal form
+    \item $f(N) = 0$ for all $N$ in normal form
+    \item $f(E[\tY N]) \geq 1 + f(E[N(\tY N)])$
+    \item $f(E[\tsample]) \geq \int_0^1 f(E[\underline{x}]) \, \mathrm{d}x$
+    \item $f(E[R]) \geq f(E[N])$ for any other redex $R$, where $R \to N$
+\end{itemize}
 
 \section{Supermartingales}
+
+\begin{theorem}
+  If an SPCF term has a ranking function, it terminates almost surely.
+\end{theorem}
+\begin{proof}
 Given a term $M$ and a ranking function $f$ for it, define random variables on the probability space $S$ (where $s$ is a random variable) by
@@
 y_0 & = 0 \\
-%y_{n+1} & = \text{min} \{ k>y_n | M_k \text{ normal form or of the form } E[\Y N] \} \\
-y_{n+1} & = \min \{ k \mid k>y_n, M_k \text{ normal form or of the form } E[\Y N] \}\\
+y_{n+1} & = \min \{ k \mid k>y_n, M_k \text{ normal form or of the form } E[\tY N] \}\\
 M'_n & = M_{y_n} \\
@@
 \end{align*}
-\lo{I have made a trivial change to the third clause above.}
 and define a filtration $\mathcal{F}_n = \sigma(M_k, k \leq n)$ (i.e. all the samples used up to step $n$).
 
-\lo{So all the random variables just defined are constant functions. E.g.~$M_n$ is a constant function from $S$ to the measurable space of PPCF terms.}
+\lo{So all the random variables just defined are constant functions.
+E.g.~$M_n$ is a constant function from $S$ to the measurable space of PPCF terms.
+\akr{I think we discussed this and it was resolved.}}
 
-\lo{\LaTeX.
-Use {\tt mid} rather than ``{\tt |}'' as separator in set comprehension.
-Reason: {\tt mid} is a binary operator (and so \LaTeX\ adds white space to both side); ``{\tt |}'' is just a symbol.}
-
-The expectation of $f(M_{n+1})$ given $\mathcal{F}_n$ is trivially less than or equal to $f(M_n)$ in the cases that $M_n \neq E[\sample]$, and in the case of $\sample$,
+The expectation of $f(M_{n+1})$ given $\mathcal{F}_n$ is trivially less than or equal to $f(M_n)$ in the cases that $M_n \neq E[\tsample]$, and in the case of $\tsample$,
 \begin{align*}
 & \mathbb{E}[f(M_{n+1}) \mid \mathcal{F}_n] \\
-= & \mathbb{E}[f(M_{n+1}) \mid M_n = E[\sample],\, \mathcal{F}_n] \\
+= & \mathbb{E}[f(M_{n+1}) \mid M_n = E[\tsample],\, \mathcal{F}_n] \\
 = & \mathbb{E}[f(E[\pi_h(s_n)]) \mid \mathcal{F}_n] \\
 = & \int_0^1 f(E[\underline x]) \, \mathrm{d} x \qquad & \text{as }s_n\text{ is independent of } \mathcal{F}_n \\
-\leq & f(E[\sample]) \qquad & \text{by assumption on } f \\
+\leq & f(E[\tsample]) \qquad & \text{by assumption on } f \\
 = & f(M_n),
@@
 
-Given $M'_n$, there is some finite bound on the number of reduction steps that can take place from $M'_n$ without a $\Y$-reduction step, because of the type system, therefore $y_{n+1}$ is (conditional on $\mathcal{F}_{y_n+1}$) a bounded stopping time, therefore $\mathbb{E}[f(M_{y_{n+1}}) \mid \mathcal{F}_{y_n+1}] \leq f(M_{y_n+1})$.
-If $f(M_{y_n}) > 0$, then $M_{y_n} = E[\Y N]$ for some $E, N$, therefore $M_{y_n+1} = N (\Y N)$ and $f(M_{y_n+1}) \leq f(M_{y_n}) - 1$, therefore if $X_n > 0$, $\mathbb{E}[X_{n+1} \mid \mathcal{F}_{y_n+1}] \leq X_n - 1$.
+Given $M'_n$, there is some finite bound on the number of reduction steps that can take place from $M'_n$ without a $\tY$-reduction step, because of the type system, therefore $y_{n+1}$ is (conditional on $\mathcal{F}_{y_n+1}$) a bounded stopping time, therefore $\mathbb{E}[f(M_{y_{n+1}}) \mid \mathcal{F}_{y_n+1}] \leq f(M_{y_n+1})$.
+If $f(M_{y_n}) > 0$, then $M_{y_n} = E[\tY N]$ for some $E, N$, therefore $M_{y_n+1} = N (\tY N)$ and $f(M_{y_n+1}) \leq f(M_{y_n}) - 1$, therefore if $X_n > 0$, $\mathbb{E}[X_{n+1} \mid \mathcal{F}_{y_n+1}] \leq X_n - 1$.
 
-The expectation of $X_n$ therefore must tend to 0 as $n \to \infty$, therefore $X_n$ almost surely is less than or equal to 1 eventually, at which point $M'_{n+1}$ must be normal form, therefore $M$ terminates almost surely.
-
-\lo{You should frame the result you have just proved, precisely and formally, using a \LaTeX\ {\tt lemma} or {\tt theorem} environment.}
+Overall, this implies that $\mathbb{E}[X_{n+1}] - \mathbb{E}[X_{n}] \leq -\mathbb P[X_n > 0]$ therefore as $\mathbb{E}[X_n]$ is bounded below, $\mathbb P[X_n > 0]$ must tend to 0 as $n \to \infty$.
+$X_n = 0 \Rightarrow X_{n+1} = 0$ therefore $\mathbb P[X_n > 0 \text{ for infinitely many values of n}] = \mathbb P[\forall n: X_n > 0] \leq \inf_n \mathbb P[X_n > 0] = 0$, therefore $M$ terminates almost surely.
+\end{proof}
 

### 2020-03-25 probabilisticProgrammingMartingales.tex
Start writing about the construction of ranking functions
Loosen the definition of ranking function slightly so as to be more
convenient. Alter Thm 1 to match.

@@
 \begin{itemize}
-    \item $f(N) > 0$ for all $N$ not in normal form
-    \item $f(N) = 0$ for all $N$ in normal form
+    \item $f(N) \geq 0$ for all $N$
     \item $f(E[\tY N]) \geq 1 + f(E[N(\tY N)])$
     \item $f(E[\tsample]) \geq \int_0^1 f(E[\underline{x}]) \, \mathrm{d}x$
-    \item $f(E[R]) \geq f(E[N])$ for any other redex $R$, where $R \to N$
+    \item $f(E[R]) \geq f(E[N])$ for any other redex $R$, where $R \to N$.
 \end{itemize}
+
+Any term for which a ranking function exists is called ``rankable''.
 
@@
 
-\begin{theorem}
-  If an SPCF term has a ranking function, it terminates almost surely.
+\begin{theorem} \label{rankable implies ast}
+  If an SPCF term is rankable, it terminates almost surely.
 \end{theorem}
@@
 and define a filtration $\mathcal{F}_n = \sigma(M_k, k \leq n)$ (i.e. all the samples used up to step $n$).
-
-\lo{So all the random variables just defined are constant functions.
-E.g.~$M_n$ is a constant function from $S$ to the measurable space of PPCF terms.
-\akr{I think we discussed this and it was resolved.}}
 
@@
 Given $M'_n$, there is some finite bound on the number of reduction steps that can take place from $M'_n$ without a $\tY$-reduction step, because of the type system, therefore $y_{n+1}$ is (conditional on $\mathcal{F}_{y_n+1}$) a bounded stopping time, therefore $\mathbb{E}[f(M_{y_{n+1}}) \mid \mathcal{F}_{y_n+1}] \leq f(M_{y_n+1})$.
-If $f(M_{y_n}) > 0$, then $M_{y_n} = E[\tY N]$ for some $E, N$, therefore $M_{y_n+1} = N (\tY N)$ and $f(M_{y_n+1}) \leq f(M_{y_n}) - 1$, therefore if $X_n > 0$, $\mathbb{E}[X_{n+1} \mid \mathcal{F}_{y_n+1}] \leq X_n - 1$.
+If $M_{y_n}$ isn't already in normal form, then $M_{y_n} = E[\tY N]$ for some $E, N$, therefore $M_{y_n+1} = N (\tY N)$ and $f(M_{y_n+1}) \leq f(M_{y_n}) - 1$, therefore if $M_{y_n}$ isn't normal form, $\mathbb{E}[X_{n+1} \mid \mathcal{F}_{y_n+1}] \leq X_n - 1$.
 
-Overall, this implies that $\mathbb{E}[X_{n+1}] - \mathbb{E}[X_{n}] \leq -\mathbb P[X_n > 0]$ therefore as $\mathbb{E}[X_n]$ is bounded below, $\mathbb P[X_n > 0]$ must tend to 0 as $n \to \infty$.
-$X_n = 0 \Rightarrow X_{n+1} = 0$ therefore $\mathbb P[X_n > 0 \text{ for infinitely many values of n}] = \mathbb P[\forall n: X_n > 0] \leq \inf_n \mathbb P[X_n > 0] = 0$, therefore $M$ terminates almost surely.
+Overall, this implies that $\mathbb{E}[X_{n+1}] - \mathbb{E}[X_{n}] \leq -\mathbb P[M_{y_n} \text{not n.f.}]$ therefore as $\mathbb{E}[X_n]$ is bounded below, $\mathbb P[M_{y_n} \text{not n.f.}]$ must tend to 0 as $n \to \infty$.
+$M_{y_n} \text{n.f.} \Rightarrow M_{y_{n+1}} \text{n.f.}$ therefore $\mathbb P[M_{y_n} \text{not n.f.~for infinitely many values of n}] = \mathbb P[\forall n: M_{y_n} \text{not n.f.}] \leq \inf_n \mathbb P[M_{y_n} \text{not n.f.}] = 0$, therefore $M$ terminates almost surely.
 \end{proof}
+
+\section{Constructing Ranking Functions}
+Although rankability implies almost sure termination, the converse does not hold in general. For example,
+\begin{equation}
+\tif{-\tsample < 0}{\underline{0}}{\tY \lambda x. x}
+\end{equation}
+terminates in 3 steps with probability 1, but isn't rankable because $\tY \lambda x.
+x$ is reachable, although that has probability 0.
+Not only is this counterexample a.s.t, it's positively almost surely terminating i.e.~the expected time to termination is finite.
+
+A ranking function can be constructed under the stronger assumptions that, for every $N$ reachable from $M$, the expected number of $\tY$-reduction steps from $N$ to a normal form is finite.
+In particular, the expected number of $\tY$-reduction steps from each reachable term is a ranking function.
+
+\akr{I don't know whether you'll see this before I actually write the other bit, but I'm planning on giving a more practical way of constructing ranking functions later using a similar technique, which is why this theorem seems relevant.}
+\begin{theorem} \label{minimal}
+Given a term $M$, the function $f:Rch(M) \to \mathbb R$ given by $f(N) = \mathbb E [\text{the number of }\tY\text{-reduction steps from }N\text{ to a normal form}]$, if it exists, is the least of all possible ranking functions of $M$.
+\end{theorem}
+\begin{proof}
+Let $f$ be the candidate least ranking function defined above, and suppose $g$ is another ranking function such that $f(N) > g(N)$ for some $N \in Rch(M)$.
+The difference $g - f$ is then a supermartingale (with the same setup as in the proof of Theorem \ref{rankable implies ast}.
+The restrictions of $f$ and $g$ to $Rch(N)$ have the same properties assumed of $f$ and $g$, so assume w.l.o.g.~that $N=M$.
+
+\akr{TODO: finish this proof. It shouldn't be complicated.}
+\end{proof}
+
+\paragraph{}
+Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and Theorem \ref{minimal} makes even stronger assumptions than almost sure termination, so it isn't useful for proving it.
+Take, for example, the term
+\begin{align*}
+&(\tY \lambda f, n: \\
+&\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
+&) \underline{0}
+\end{align*}
+which generates a geometric distribution.
+    Despite its simplicity, its $Rch$ contains all the terms
+\begin{itemize}
+    \item $(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{i}$
+    \item $(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{i}$
+    \item $(\lambda n: \tif{\tsample - 0.5 < 0}{n}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{n}{f (m+1)}) (m+1)}) \underline{i}$
+    \item $\tif{\tsample - 0.5 < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
+    \item $\tif{\underline r - 0.5 < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
+    \item $\tif{\underline{r-0.5} < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
+    \item $\underline{i}$
+    \item $(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)$.
+\end{itemize}
+\akr{TODO: make this into a diagram with arrows.}
+Even in this simple case, defining a ranking function explicitly is awkward because of the number of cases, although in most cases, because the value need only be greater than or equal to that of the next term in sequence, it suffices to take the ranking function as taking only 3 distinct values.
 

### 2020-03-26 probabilisticProgrammingMartingales.tex
minor

--- 2020-03-09 probabilisticProgrammingMartingales.tex
@@
 - For example, develop a type system satisfying the property: if a term is typable then it is (positively) a.s.~terminating.
-See~\cite{DBLP:conf/ppdp/BreuvartL18,DBLP:conf/esop/LagoG17}.
+See~\citep{DBLP:conf/ppdp/BreuvartL18,DBLP:conf/esop/LagoG17}.
 
-- Another approach is to develop algorithms that synthesise ranking supermartingales, following, for example, \cite{DBLP:journals/pacmpl/AgrawalC018}.
+- Another approach is to develop algorithms that synthesise ranking supermartingales, following, for example, \citep{DBLP:journals/pacmpl/AgrawalC018}.
 
-\item Develop principles (e.g.~in the form of ``proof rules'') for reasoning about (positively) a.s.~termination, in the style of \cite{DBLP:journals/pacmpl/McIverMKK18}.
+\item Develop principles (e.g.~in the form of ``proof rules'') for reasoning about (positively) a.s.~termination, in the style of \citep{DBLP:journals/pacmpl/McIverMKK18}.
 
-\item Design algorithms that synthesise probabilistic invariants (\emph{qua} martingales), \`a la \cite{SchreuderO19}; see also \cite{DBLP:journals/pacmpl/HarkKGK20}.
+\item Design algorithms that synthesise probabilistic invariants (\emph{qua} martingales), \`a la \cite{SchreuderO19}; see also \citep{DBLP:journals/pacmpl/HarkKGK20}.
 

### 2020-03-27 probabilisticProgrammingMartingales.tex
Add alternative semantics
Draft description of a different version of sampling semantics more
compatible with variable reduction order.

--- 2020-03-25 probabilisticProgrammingMartingales.tex
@@
 
+\subsection{Alternative Semantics}
+\akr{This version of the semantics is much more convenient for allowing multiple different reduction orders.
+I don't currently have a proof of the equivalence of this with the more usual semantics.
+I expect that that would be somewhat complicated, but not terribly difficult.
+I don't know whether a semantics like this has already been defined elsewhere.
+I just thought I'd write it out so you can see it and in case I use it later.
+If I do end up actually using it I'll write it up nicer, but hopefully this is at least enough for you to understand what I mean.}
+
+Given a term $M$, let $L(M)$ be the set of occurrences in $M$ of variables, $\tsample$ and $\tY$, and let $F(M)$ be the free magma on $L(M)$.
+\akr{Only the elements with no \tsample{}s except in the rightmost position are used.
+I'm not sure whether making this explicit in the definition of $F(M)$ makes things more or less clear.} Let $I = [0,1]$ and $S = I^{F(M)}$, with the Borel $\sigma$-algebra and the product measure.
+
+For any reduction \akr{including reductions not included in the definition above, i.e.~out of order reductions} $M \to N$, define a corresponding map $F(M \to N) : F(N) \to F(M)$ induced (as a magma morphism) by the obvious injection along with the replacements
+\begin{align*}
+F(E[(\lambda x. F[x_i])G[y]] \to E[F(G[y]/x)[G[y_i]]]) (x_i \cdot y) & = y_i\\
+F(E[\tY_i F[y]] \to E[F[y_0] (\tY F[y])]) (y_0) & = \tY_i \cdot y
+\end{align*}
+where $y$ can be any of the things in $L(N)$.
+
+The one-step reduction is given by the function $\red : \Lambda \times S \to \Lambda \times S$ where
+\begin{equation}
+\red(M,s) = \left\{
+    \begin{array}{ll}
+        (E[N],s \circ F(E[R] \to E[N])) & \text{if } M = E[R], R \to N \text{ and } R \neq \tsample \\
+        (E[\underline{s(\tsample_i)}],s\mid_{F(E[\underline{s(\tsample_i)}])}) & \text{if } M = E[\tsample_i] \\
+        (M,s) & \text{if } M \text{ normal form}
+    \end{array} \right .
+\end{equation}
+
+Essentially the entropy value here contains a pre-selected random value for each possible \tsample-reduction, in a way independent of reduction order, taking into account the possible duplication of \tsample{}s in $\beta$- and \tY-reduction steps.
+
 \section{Ranking functions}

### 2020-03-28 probabilisticProgrammingMartingales.tex
Fix F(M -> N) conversion formula

--- 2020-03-27 probabilisticProgrammingMartingales.tex
@@
 I just thought I'd write it out so you can see it and in case I use it later.
-If I do end up actually using it I'll write it up nicer, but hopefully this is at least enough for you to understand what I mean.}
+If I do end up actually using it I'll write it up nicer, but hopefully this is at least enough for you to understand what I mean.
+
+Having thought a bit more about this, I realise it may not be quite right. I'll see if I can fix it later.}
 
@@
 \begin{align*}
-F(E[(\lambda x. F[x_i])G[y]] \to E[F(G[y]/x)[G[y_i]]]) (x_i \cdot y) & = y_i\\
-F(E[\tY_i F[y]] \to E[F[y_0] (\tY F[y])]) (y_0) & = \tY_i \cdot y
+F(E[(\lambda x. F[x_i])G[y]] \to E[F(G[y]/x)[G[y_i]]]) y_i & = (x_i \cdot y)\\
+F(E[\tY_i F[y]] \to E[F[y] (\tY F[y_0])]) (y_0) & = \tY_i \cdot y
 \end{align*}

### 2020-04-01 probabilisticProgrammingMartingales.tex
Comments, questions and corrections.

@@
 \lo{Can you explain your concern? I don't see any problem. See, e.g., the operational semantics in \citep{MakOP20b}; see also \citep{MakOP20a}.} 
-You said that the assumption of a.s. termination was often used. 
+You said that the assumption of a.s.~termination was often used. 
 Do you have any examples where it's used in a context with unbounded scoring, as going with whatever definition that uses is probably most useful?
@@
 
-To define the reduction relation, let environments be of the form:
+\lo{What are the primitive functions $f$? All measurable functions?}
+
+To define the reduction relation, let environments \lo{They are called \emph{evaluation contexts}. Environments mean something else.} be of the form:
 \begin{align*}
@@
 
+\lo{- I assume this is a CBV language (judging from the (evaluation context) clause $v \, E$). So the first rule should be
+\[E[(\lambda x.s) v] \to E[s[v/x]]\]
+
+- In a CBV language, the $\tY$-combinator should only be applied to terms of type $(A \to B) \to (A \to B)$. 
+Your rule $E[\tY v] \to E[v (\tY v)]$ will diverge ($E[v (v (v \cdots ))]$).
+The redex rule for the Y-combinator should be
+\[
+\tY (\lambda y.s) \to \lambda z.s[\tY{(\lambda y.s)}/y]\,z \quad \hbox{($z$ is fresh)}
+\]
+
+}
 \subsection{Sampling semantics}
@@
 To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples.
-Let $ I = [0,1] \subset \mathbb{R} $, and let $S = I^{\mathbb{N}}$, with the Borel $\sigma$-algebra and the measure given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
+Let $ I = [0,1] \subset \mathbb{R} $, and let $S = I^{\mathbb{N}}$, with the Borel $\sigma$-algebra and the measure, \changed[lo]{denoted $\mu$}, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
 The maps $\pi_h:S \to I, \; \pi_t:S \to S$ popping the first element are then measurable.
+\lo{What does a measurable subset of $S$ look like? Give a ``concrete'' description.}
+\lo{Clarify the measure $\mu$: give a ``formula'' for computing $\mu(A)$, for $A \in \Sigma_S$. Is $\mu$ a probability measure?}
 
@@
 
-The limit $\red^\infty$ can then be defined as a partial function as $\lim_{n \to \infty} \red^n(M,s)$ whenever that sequence becomes constant by reaching a normal form.
+The limit $\red^\infty$ can then be defined as a partial function as $\lim_{n \to \infty} \red^n(M,s)$ whenever that sequence becomes constant by reaching a normal form. \lo{Define $\red^n(M,s)$.} 
 A term $M$ terminates for a sample sequence $s$ if the limit $\red^\infty(M,s)$ is defined, and it terminates almost surely if $\mu(\{s \mid M \text{ terminates at } s\}) = 1$.
+\lo{Why is $\{s \mid M \text{ terminates at } s\}$ a measurable subset of $S$?}
+\lo{Why is your definition of a.s.~termination reasonable? 
+For example, is it the case that for all $M$, $\mu(\{s \mid M \text{ terminates at } s\}) \leq 1$?}
+
+\lo{Important: Give examples (and non-examples) of a.s.~terminating terms (such as the one you wrote on my whiteboard).}
 
@@
 Having thought a bit more about this, I realise it may not be quite right. I'll see if I can fix it later.}
+\lo{I'll read this after you fix it ;-)}
 
@@
 Given a probabilistic program (i.e.~a term $M$), in order to construct a supermartingale to prove its a.s.~termination, a function to assign values to each reachable program state is necessary.
-Define a ranking function on $M$ to be a measurable function $f:Rch(M) \to \mathbb{R}$ such that
+Define a \emph{ranking function on $M$} to be a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that
 \begin{itemize}
     \item $f(N) \geq 0$ for all $N$
-    \item $f(E[\tY N]) \geq 1 + f(E[N(\tY N)])$
+    \item $f(E[\tY N]) \geq 1 + f(E[N(\tY N)])$ \lo{Modify this in view of the revised \tY-redex rule.}
     \item $f(E[\tsample]) \geq \int_0^1 f(E[\underline{x}]) \, \mathrm{d}x$
+
+    \lo{I assume you mean $\int_{I} \mathsf{Leb} (\mathrm{d} x) \, f(E[\underline{x}])$.}
+
+    \lo{For the integral to be well-defined, we need $x \mapsto E[\underline x]$ to be a measurable function $\mathbb{R} \to \mathit{Rch}(M)$.}
+
+    \lo{Define the measurable space $\mathit{Rch}(M)$, and the measurable space of SPCF terms.}
+
     \item $f(E[R]) \geq f(E[N])$ for any other redex $R$, where $R \to N$.
@@
 Any term for which a ranking function exists is called ``rankable''.
+
+\lo{Ongoing}
 
@@
 Overall, this implies that $\mathbb{E}[X_{n+1}] - \mathbb{E}[X_{n}] \leq -\mathbb P[M_{y_n} \text{not n.f.}]$ therefore as $\mathbb{E}[X_n]$ is bounded below, $\mathbb P[M_{y_n} \text{not n.f.}]$ must tend to 0 as $n \to \infty$.
-$M_{y_n} \text{n.f.} \Rightarrow M_{y_{n+1}} \text{n.f.}$ therefore $\mathbb P[M_{y_n} \text{not n.f.~for infinitely many values of n}] = \mathbb P[\forall n: M_{y_n} \text{not n.f.}] \leq \inf_n \mathbb P[M_{y_n} \text{not n.f.}] = 0$, therefore $M$ terminates almost surely.
+$M_{y_n} \text{n.f.} \Rightarrow M_{y_{n+1}} \text{n.f.}$ therefore
+\begin{align*}
+\mathbb P[M_{y_n} \text{not n.f.~for infinitely many values of n}]  
+& = \mathbb P[\forall n: M_{y_n} \text{not n.f.}] \\
+& \leq \inf_n \mathbb P[M_{y_n} \text{not n.f.}] \\
+& = 0
+\end{align*} 
+therefore $M$ terminates almost surely.
 \end{proof}

### 2020-04-02 probabilisticProgrammingMartingales.tex
minor

@@
 
-\lo{Ongoing}
-
 \section{Supermartingales}
@@
 \begin{align*}
-\mathbb P[M_{y_n} \text{not n.f.~for infinitely many values of n}]  
+\mathbb P[M_{y_n} \text{not n.f.~for infinitely many values of $n$}]  
 & = \mathbb P[\forall n: M_{y_n} \text{not n.f.}] \\

### 2020-04-08 probabilisticProgrammingMartingales.tex
minor

@@
 
-    \lo{For the integral to be well-defined, we need $x \mapsto E[\underline x]$ to be a measurable function $\mathbb{R} \to \mathit{Rch}(M)$.}
+    \lo{For the integral to be well-defined, we need $x \mapsto E[\underline x]$ to be a measurable function $\mathbb{R} \to \mathit{Rch}(M)$ (I am sure it is, assuming an appropriate definition of SPCF terms as a measurable space.).}
 

### 2020-04-13 probabilisticProgrammingMartingales.tex
Respond to feedback
Change the Y-reduction rule. Make some things about the measures
and σ-algebras more explicit.

--- 2020-04-01 probabilisticProgrammingMartingales.tex
@@
 
-\lo{23 Jan:
-
-\paragraph{Some thoughts about extension to $\mathsf{score}$}
-
-- I still think that the operational semantics (reduction semantics) of SPCF (i.e.~with $\mathsf{score}$) is not problematic.
-However the notion of a.e.~termination is interesting and would benefit from further thought.
-
-- I would propose that you use the language SPCF defined in \citep{MakOP20b}.
-They use a big-step reduction semantics.
-I would sugget that, by contrast, you develop a small-step reduction semantics for it (by a mild adaptation of your $\red : \Lambda \times S \to \Lambda \times S$).
-
-- You will find that your example 
-\[
-f \, n = n \oplus_{\frac{1}{2}} (\mathsf{score} \, 3 \, ; \, f \, (n+1))
-\]
-is not definable in SPCF, but a variant (in which the branching probabilities are drawn from the entropy sequence) is.
-OK, we will need to assume that their $\mathbb{S} := \mathbb{R}^\omega$ is replaced by your $S$.
-
-- Consider the notion of a.e.~termination defined in \citep{MakOP20b}\footnote{\url{https://drive.google.com/open?id=1JzxV0lCkZu4d7szQOzkcKspqqo8u2BP0}} (towards the end of \S 3): a program $s$ is a.e.~terminating if the set of entropy sequences where $s$ does not evaluate to a value has measure zero.
-The intuition is that scoring is about model building, and does not by itself ``affect termination''.
-}
-
-\bigskip
-
-\lo{22 Jan:
-\LaTeX.
-For references, I recommend using the package {\tt natbib} with {\tt apalike} bibliography style (author-year is more reader-friendly), and placing the references in a separate {\tt .bib} file.
-I simply copy-and-paste the bib data from DBLP e.g.~\url{https://dblp.uni-trier.de/pers/hd/o/Ong:C==H=_Luke}.}
-
-\changed[akr]{[Email 22 Dec: I wrote a proof of termination based on ranking supermartingales (a slight strengthening of what you suggested), which is attached (I haven't made a Git repo yet because I don't know where you want it hosted).
-
-I thought some more about termination in the case of a language with a score function. 
-It isn't even clear what termination means in that case.
-None of the papers I've found include both unbounded scoring and a definition of termination.
-\lo{Can you explain your concern? I don't see any problem. See, e.g., the operational semantics in \citep{MakOP20b}; see also \citep{MakOP20a}.} 
-You said that the assumption of a.s.~termination was often used. 
-Do you have any examples where it's used in a context with unbounded scoring, as going with whatever definition that uses is probably most useful?
-\lo{See \citep[Sec.~4, pp.~8 \& 9]{Wagner20} -- I am not sure if this addresses your question.}
-
-The case of bounded scoring seems simpler definitionally at least, but still seems like a bad fit for supermartingales, as the martingale property is local in a way scoring isn't. 
-I still need to think more about to what extent it can be made to make sense, though.]}
-
-Following your suggestion, I will be providing a criterion for termination of programs in PPCF \citep{DBLP:journals/pacmpl/EhrhardPT18} based on ranking supermartingales. 
+\akr{Following your suggestion, I will be providing a criterion for termination of programs in PPCF \citep{DBLP:journals/pacmpl/EhrhardPT18} based on ranking supermartingales. 
 As it's more convenient for this proof, a sampling-based semantics \lo{You need to provide reference(s) for sampling-based semantics.} will be used instead of the original distributional semantics. 
-I assume some roughly applicable equivalence is proven somewhere, but it doesn't seem that hard anyway.
+I assume some roughly applicable equivalence is proven somewhere, but it doesn't seem that hard anyway.}
 \lo{Why not provide a proof as an appendix?}
@@
 The language SPCF is a simply-typed lambda calculus with sampling of real numbers from $[0,1]$ and unbounded scoring, following \cite{MakOP20b}.
-Types and terms are defined as follows, where $r$ is a real, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ and $\Gamma$ is an environment
+Types and terms are defined as follows, where $r$ is a real, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function and $\Gamma$ is an environment
 \begin{align*}
@@
   \frac{\Gamma \vdash s_1:\textsf{R} \dots \Gamma \vdash s_n:\textsf{R}}{\Gamma \vdash \underline{f}(s_1,\dots,s_n) : \textsf{R}} \ f : \mathbb{R}^n \to \mathbb{R} \qquad
-  \frac{\Gamma \vdash s : A \to A}{\Gamma \vdash \tY s : A} \\ \\
+  \frac{\Gamma \vdash s : (A \to B) \to (A \to B)}{\Gamma \vdash \tY s : (A \to B)} \\ \\
   \frac{\Gamma \vdash c : \textsf{R} \quad \Gamma \vdash s_1 : A \quad \Gamma \vdash s_2 : A}{\Gamma \vdash \tif{c < 0}{s_1}{s_2} : A} \qquad
@@
 
-\lo{What are the primitive functions $f$? All measurable functions?}
-
-To define the reduction relation, let environments \lo{They are called \emph{evaluation contexts}. Environments mean something else.} be of the form:
+To define the reduction relation, let evaluation contexts be of the form:
 \begin{align*}
@@
 \end{align*}
-then a term reduces if it is formed by substituting a redex in an environment i.e.
+then a term reduces if it is formed by substituting a redex in a context i.e.
 \begin{align*}
-  E[(\lambda x.s) t] & \to E[s[t/x]] \\
+  E[(\lambda x.s) v] & \to E[s[v/x]] \\
   E[\underline f (\underline r_1, \dots , \underline r_n)] & \to E[\underline{f(r_1,\dots,r_n)}] \\
-  E[\tY v] & \to E[v (\tY v)] \\
+  E[\tY \lambda x. s] & \to E[\lambda z. s[(\tY \lambda x. s)/x] z] \text{ where $z$ is not free in $s$}\\
   E[\tif{\underline r < 0}{s_1}{s_2}] & \to E[s_1] \text{ where }r < 0 \\
@@
 
-\lo{- I assume this is a CBV language (judging from the (evaluation context) clause $v \, E$). So the first rule should be
-\[E[(\lambda x.s) v] \to E[s[v/x]]\]
-
-- In a CBV language, the $\tY$-combinator should only be applied to terms of type $(A \to B) \to (A \to B)$. 
-Your rule $E[\tY v] \to E[v (\tY v)]$ will diverge ($E[v (v (v \cdots ))]$).
-The redex rule for the Y-combinator should be
-\[
-\tY (\lambda y.s) \to \lambda z.s[\tY{(\lambda y.s)}/y]\,z \quad \hbox{($z$ is fresh)}
-\]
-
-}
 \subsection{Sampling semantics}
@@
 To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples.
-Let $ I = [0,1] \subset \mathbb{R} $, and let $S = I^{\mathbb{N}}$, with the Borel $\sigma$-algebra and the measure, \changed[lo]{denoted $\mu$}, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
+Let $ I = [0,1] \subset \mathbb{R} $, and let $S = I^{\mathbb{N}}$, with the Borel $\sigma$-algebra and the probability measure, denoted $\mu$, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
+Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and $\mu (\prod_{i=0}^\infty X_i) = \prod_{i=0}^\infty \mu_{leb}(X_i)$.
 The maps $\pi_h:S \to I, \; \pi_t:S \to S$ popping the first element are then measurable.
-\lo{What does a measurable subset of $S$ look like? Give a ``concrete'' description.}
-\lo{Clarify the measure $\mu$: give a ``formula'' for computing $\mu(A)$, for $A \in \Sigma_S$. Is $\mu$ a probability measure?}
+
+The $\sigma$-algebra and measure on $\Lambda$ are defined by considering it as a disjoint union of equivalence classes under replacing all the real constants by a placeholder, where the measure on each class is that of $\mathbb{R}^n$, where $n$ is the number of real constants.
 
@@
 
-The limit $\red^\infty$ can then be defined as a partial function as $\lim_{n \to \infty} \red^n(M,s)$ whenever that sequence becomes constant by reaching a normal form. \lo{Define $\red^n(M,s)$.} 
-A term $M$ terminates for a sample sequence $s$ if the limit $\red^\infty(M,s)$ is defined, and it terminates almost surely if $\mu(\{s \mid M \text{ terminates at } s\}) = 1$.
-\lo{Why is $\{s \mid M \text{ terminates at } s\}$ a measurable subset of $S$?}
-\lo{Why is your definition of a.s.~termination reasonable? 
-For example, is it the case that for all $M$, $\mu(\{s \mid M \text{ terminates at } s\}) \leq 1$?}
+The result after $n$ steps is then simply $\red^n(M,s) = \overbrace{\red(...\red(}^n M,s)...)$, and the limit $\red^\infty$ can then be defined as a partial function as $\lim_{n \to \infty} \red^n(M,s)$ whenever that sequence becomes constant by reaching a normal form.
+A term $M$ terminates for a sample sequence $s$ if the limit $\red^\infty(M,s)$ is defined.
 
-\lo{Important: Give examples (and non-examples) of a.s.~terminating terms (such as the one you wrote on my whiteboard).}
+The reduction function is measurable, and the set of normal forms is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid M \text{ terminates at } s \}$ is measurable.
+A term $M$ is said to terminate almost surely if $\mu(\{s \mid M \text{ terminates at } s\}) = 1$.
+
+For example, the term $(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \underline{0}$, which generates a geometric distribution, terminates on the set $S \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas $\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x.
+x) \underline 0}$, which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.
 
@@
 Given a probabilistic program (i.e.~a term $M$), in order to construct a supermartingale to prove its a.s.~termination, a function to assign values to each reachable program state is necessary.
-Define a \emph{ranking function on $M$} to be a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that
+Let $Rch(M) = \{x \in \Lambda \mid \exists (y_i) : M \to y_0 \to \dots \to y_n \to x\}$, with the $\sigma$-algebra induced as a subset of $\Lambda$, then define a \emph{ranking function on $M$} to be a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that
 \begin{itemize}
     \item $f(N) \geq 0$ for all $N$
-    \item $f(E[\tY N]) \geq 1 + f(E[N(\tY N)])$ \lo{Modify this in view of the revised \tY-redex rule.}
+    \item $f(E[\tY \lambda x. N] \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z]) \text{ where $z$ is not free in $N$}$
     \item $f(E[\tsample]) \geq \int_0^1 f(E[\underline{x}]) \, \mathrm{d}x$
@@
 
-    \lo{For the integral to be well-defined, we need $x \mapsto E[\underline x]$ to be a measurable function $\mathbb{R} \to \mathit{Rch}(M)$.}
-
-    \lo{Define the measurable space $\mathit{Rch}(M)$, and the measurable space of SPCF terms.}
+    \akr{Is the meaning of the integral unclear in the context?
+I'm not sure whether $Rch(M)$ is a measurable subset of $\Lambda$, so it might not have a natural measure, but I'm pretty sure this doesn't actually matter, and the $\sigma$-algebra at least is well-defined.
+$Rch(M)$ could fail to be measurable if there are measurable functions $\mathbb R^n \to \mathbb R$ with non-measurable ranges.
+I don't know whether these exist.
+They do for the Lebesgue $\sigma$-algebra, but I'm using the Borel ones.}
 
@@
 \begin{equation}
-\tif{-\tsample < 0}{\underline{0}}{\tY \lambda x. x}
+\tif{-\tsample < 0}{\underline{0}}{(\tY \lambda x. x)x}
 \end{equation}
-terminates in 3 steps with probability 1, but isn't rankable because $\tY \lambda x.
-x$ is reachable, although that has probability 0.
+terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x.
+x)x$ is reachable, although that has probability 0.
 Not only is this counterexample a.s.t, it's positively almost surely terminating i.e.~the expected time to termination is finite.

### 2020-04-13 probabilisticProgrammingMartingales.tex

@@
 They do for the Lebesgue $\sigma$-algebra, but I'm using the Borel ones.}
+    \lo{For the integral to be well-defined, we need $x \mapsto E[\underline x]$ to be a measurable function $\mathbb{R} \to \mathit{Rch}(M)$ (I am sure it is, assuming an appropriate definition of SPCF terms as a measurable space.).}
 
@@
 Any term for which a ranking function exists is called ``rankable''.
-
-\lo{Ongoing}
 

### 2020-04-15 probabilisticProgrammingMartingales.tex
Measurability of Rch(M) and related.

@@
 
-The one-step reduction is given by the function $\red : \Lambda \times S \to \Lambda \times S$ where
+The one-step reduction, \changed[lo]{$\to$}, is given by the function $\red : \Lambda \times S \to \Lambda \times S$ where
 \begin{equation}
@@
 
-For example, the term $(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \underline{0}$, which generates a geometric distribution, terminates on the set $S \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas $\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x.
+For example, the term $(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0}$, which generates a geometric distribution, terminates on the set $S \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas $\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x.
 x) \underline 0}$, which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.
@@
 In particular, the argument to $\tscore$ is not relevant to termination (except for the possibility that its argument's evaluation wouldn't terminate).
+
+\lo{Your operational semantics does not maintain a record of the current weight of the reduction.
+Note that a.s.~termination does depend on $\tscore$: see \cite[\S 4.3]{DBLP:journals/corr/abs-2004-03924}\footnote{\url{https://arxiv.org/abs/2004.03924}}.}
 
@@
 
+    \lo{I agree that it does not actually matter whether $\mathit{Rch}(M)$ is a measurable subset of $\Lambda$ (the set of SPCF terms).}
+
+    \lo{It is straightforward to see that $E[\cdot] : x \mapsto E[\underline x]$ is a measurable function $\mathbb{R} \to \mathit{Rch}(M)$, using the ``skeletal term'' $\sigma$-algebra on $\mathit{Rch}(M)$, and so, $x \mapsto f(E[\underline{x}])$ is measurable.}
+
     \lo{I assume you mean $\int_{I} \mathsf{Leb} (\mathrm{d} x) \, f(E[\underline{x}])$.} 
+    \akr{Is the meaning of the integral unclear in the context?} 
+    \lo{It it -- I was being pedantic. What you write is a Riemann integral, but it coincides with Lebesgue in this case.}
 
-    \akr{Is the meaning of the integral unclear in the context?
-I'm not sure whether $Rch(M)$ is a measurable subset of $\Lambda$, so it might not have a natural measure, but I'm pretty sure this doesn't actually matter, and the $\sigma$-algebra at least is well-defined.
+    \akr{I'm not sure whether $Rch(M)$ is a measurable subset of $\Lambda$, so it might not have a natural measure, but I'm pretty sure this doesn't actually matter, and the $\sigma$-algebra at least is well-defined.
 $Rch(M)$ could fail to be measurable if there are measurable functions $\mathbb R^n \to \mathbb R$ with non-measurable ranges.

### 2020-04-15 probabilisticProgrammingMartingales.tex
Some comments and questions.

@@
 \lo{Your operational semantics does not maintain a record of the current weight of the reduction.
-Note that a.s.~termination does depend on $\tscore$: see \cite[\S 4.3]{DBLP:journals/corr/abs-2004-03924}\footnote{\url{https://arxiv.org/abs/2004.03924}}.}
+Note that a.s.~termination does depend on $\tscore$: see \cite[\S 4.3]{DBLP:journals/corr/abs-2004-03924}\footnote{\url{https://arxiv.org/abs/2004.03924}}.
+I think it important to take the behaviour of $\tscore$ into account, perhaps as a future task.}
 
@@
 \end{equation}
+
+\lo{By ``normal form'' in the side condition above, you mean value? (Normal form can be confusing.)}
 
@@
 
-Given $M'_n$, there is some finite bound on the number of reduction steps that can take place from $M'_n$ without a $\tY$-reduction step, because of the type system, therefore $y_{n+1}$ is (conditional on $\mathcal{F}_{y_n+1}$) a bounded stopping time, therefore $\mathbb{E}[f(M_{y_{n+1}}) \mid \mathcal{F}_{y_n+1}] \leq f(M_{y_n+1})$.
+Given $M'_n$, there is some finite bound on the number of reduction steps that can take place from $M'_n$ without a $\tY$-reduction step, because of the type system \lo{Why?}, therefore $y_{n+1}$ is (conditional on $\mathcal{F}_{y_n+1}$) a bounded stopping time, therefore $\mathbb{E}[f(M_{y_{n+1}}) \mid \mathcal{F}_{y_n+1}] \leq f(M_{y_n+1})$.
 If $M_{y_n}$ isn't already in normal form, then $M_{y_n} = E[\tY N]$ for some $E, N$, therefore $M_{y_n+1} = N (\tY N)$ and $f(M_{y_n+1}) \leq f(M_{y_n}) - 1$, therefore if $M_{y_n}$ isn't normal form, $\mathbb{E}[X_{n+1} \mid \mathcal{F}_{y_n+1}] \leq X_n - 1$.
 
-Overall, this implies that $\mathbb{E}[X_{n+1}] - \mathbb{E}[X_{n}] \leq -\mathbb P[M_{y_n} \text{not n.f.}]$ therefore as $\mathbb{E}[X_n]$ is bounded below, $\mathbb P[M_{y_n} \text{not n.f.}]$ must tend to 0 as $n \to \infty$.
+Overall, this implies \lo{Why?} that $\mathbb{E}[X_{n+1}] - \mathbb{E}[X_{n}] \leq -\mathbb P[M_{y_n} \text{not n.f.}]$ therefore as $\mathbb{E}[X_n]$ is bounded below, $\mathbb P[M_{y_n} \text{not n.f.}]$ must tend to 0 as $n \to \infty$.
 $M_{y_n} \text{n.f.} \Rightarrow M_{y_{n+1}} \text{n.f.}$ therefore

### 2020-04-15 probabilisticProgrammingMartingales.tex
Tidying up comments

--- 2020-04-15 probabilisticProgrammingMartingales.tex
@@
 
-The one-step reduction, \changed[lo]{$\to$}, is given by the function $\red : \Lambda \times S \to \Lambda \times S$ where
+The one-step reduction is given by the function $\red : \Lambda \times S \to \Lambda \times S$ where
 \begin{equation}
@@
 \end{equation}
+\lo{Replace (globally) ``normal form'' by ``value''.}
 
@@
 \lo{Your operational semantics does not maintain a record of the current weight of the reduction.
-Note that a.s.~termination does depend on $\tscore$: see \cite[\S 4.3]{DBLP:journals/corr/abs-2004-03924}\footnote{\url{https://arxiv.org/abs/2004.03924}}.}
+A.s.~termination does depend on $\tscore$: see \cite[\S 4.3]{DBLP:journals/corr/abs-2004-03924}\footnote{\url{https://arxiv.org/abs/2004.03924}}.
+I think it important to take the behaviour of $\tscore$ into account;
+you should do it as a future task.}
 
@@
 They do for the Lebesgue $\sigma$-algebra, but I'm using the Borel ones.}
-    \lo{For the integral to be well-defined, we need $x \mapsto E[\underline x]$ to be a measurable function $\mathbb{R} \to \mathit{Rch}(M)$ (I am sure it is, assuming an appropriate definition of SPCF terms as a measurable space.).}
+    \lo{It is worth inserting the preceding (including the example of a measurable function with non-measurable range) as a remark (or as a ``commenting.sty'' note).}
+    \lo{For the integral to be well-defined, we need $x \mapsto E[\underline x]$ to be a measurable function $\mathbb{R} \to \mathit{Rch}(M)$.}
 
@@
 Given $M'_n$, there is some finite bound on the number of reduction steps that can take place from $M'_n$ without a $\tY$-reduction step, because of the type system, therefore $y_{n+1}$ is (conditional on $\mathcal{F}_{y_n+1}$) a bounded stopping time, therefore $\mathbb{E}[f(M_{y_{n+1}}) \mid \mathcal{F}_{y_n+1}] \leq f(M_{y_n+1})$.
-If $M_{y_n}$ isn't already in normal form, then $M_{y_n} = E[\tY N]$ for some $E, N$, therefore $M_{y_n+1} = N (\tY N)$ and $f(M_{y_n+1}) \leq f(M_{y_n}) - 1$, therefore if $M_{y_n}$ isn't normal form, $\mathbb{E}[X_{n+1} \mid \mathcal{F}_{y_n+1}] \leq X_n - 1$.
+If $M_{y_n}$ isn't already in normal form, then \changed[lo]{$M_{y_n} = E[\tY \lambda x.N]$ for some $E, N$, therefore $M_{y_n+1} = E[\lambda z.
+N[(\tY \lambda x.
+N)/x] z]$} and $f(M_{y_n+1}) \leq f(M_{y_n}) - 1$, therefore if $M_{y_n}$ isn't normal form, $\mathbb{E}[X_{n+1} \mid \mathcal{F}_{y_n+1}] \leq X_n - 1$.
 
-Overall, this implies that $\mathbb{E}[X_{n+1}] - \mathbb{E}[X_{n}] \leq -\mathbb P[M_{y_n} \text{not n.f.}]$ therefore as $\mathbb{E}[X_n]$ is bounded below, $\mathbb P[M_{y_n} \text{not n.f.}]$ must tend to 0 as $n \to \infty$.
+Overall, this implies \lo{This implication is still not clear to me.} that $\mathbb{E}[X_{n+1}] - \mathbb{E}[X_{n}] \leq -\mathbb P[M_{y_n} \text{not n.f.}]$ therefore as $\mathbb{E}[X_n]$ is bounded below, $\mathbb P[M_{y_n} \text{not n.f.}]$ must tend to 0 as $n \to \infty$.
 $M_{y_n} \text{n.f.} \Rightarrow M_{y_{n+1}} \text{n.f.}$ therefore

### 2020-04-20 probabilisticProgrammingMartingales.tex
Consider reduction to values instead of normal forms.
Require terms to be closed for reductions to be defined, and use
"value" insead of "normal form". Also a few little edits.

@@
 The language SPCF is a simply-typed lambda calculus with sampling of real numbers from $[0,1]$ and unbounded scoring, following \cite{MakOP20b}.
-Types and terms are defined as follows, where $r$ is a real, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function and $\Gamma$ is an environment
+Types and terms are defined as follows, where $r$ is a real, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function and $\Gamma$ is an environment:
 \begin{align*}
   & \text{types } A, B ::= \textsf{R}  \mid  A \to B \\
-  & \text{values } v ::= x  \mid  \lambda x.s  \mid  \underline{r} \\
-  & \text{terms } s, t ::= v  \mid  t_1 t_2  \mid  \underline{f}(s_1,\dots ,s_n)  \mid  \tY s  \mid  \tif{s < 0}{t_1}{t_2}  \mid  \tsample  \mid  \tscore(s)
+  & \text{values } v ::= \lambda x.s  \mid  \underline{r} \\
+  & \text{terms } s, t ::= v  \mid  x  \mid  t_1 t_2  \mid  \underline{f}(s_1,\dots ,s_n)  \mid  \tY s  \mid  \tif{s < 0}{t_1}{t_2}  \mid  \tsample  \mid  \tscore(s)
 \end{align*}
@@
 
+The set of all terms is denoted $\Lambda$, and the set of closed terms is denoted $\Lambda^0$.
+
+\paragraph{}
 To define the reduction relation, let evaluation contexts be of the form:
@@
   E[\tsample] & \to E[\underline r] \text{ where } r \in [0,1] \\
-  E[\tscore(\underline r)] & \to E[\underline r]
+  E[\tscore(\underline r)] & \to E[\underline r].
 \end{align*}
+
+Every closed well-typed term either is a value or reduces to another term.
 
@@
 
-The one-step reduction is given by the function $\red : \Lambda \times S \to \Lambda \times S$ where
+The one-step reduction is given by the function $\red : \Lambda^0 \times S \to \Lambda^0 \times S$ where
 \begin{equation}
@@
         (E[\underline{\pi_h(s)}],\pi_t(s)) & \text{if } M = E[\tsample] \\
-        (M,s) & \text{if } M \text{ normal form}
+        (M,s) & \text{if } M \text{ is a value}
     \end{array} \right .
 \end{equation}
-\lo{Replace (globally) ``normal form'' by ``value''.}
 
-The result after $n$ steps is then simply $\red^n(M,s) = \overbrace{\red(...\red(}^n M,s)...)$, and the limit $\red^\infty$ can then be defined as a partial function as $\lim_{n \to \infty} \red^n(M,s)$ whenever that sequence becomes constant by reaching a normal form.
+The result after $n$ steps is then simply $\red^n(M,s) = \overbrace{\red(...\red(}^n M,s)...)$, and the limit $\red^\infty$ can then be defined as a partial function as $\lim_{n \to \infty} \red^n(M,s)$ whenever that sequence becomes constant by reaching a value.
 A term $M$ terminates for a sample sequence $s$ if the limit $\red^\infty(M,s)$ is defined.
 
-The reduction function is measurable, and the set of normal forms is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid M \text{ terminates at } s \}$ is measurable.
+The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid M \text{ terminates at } s \}$ is measurable.
 A term $M$ is said to terminate almost surely if $\mu(\{s \mid M \text{ terminates at } s\}) = 1$.
@@
 x) \underline 0}$, which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.
+
+\paragraph{}
+Note that the requirement that $M$ be closed is not important for the proofs of the theorems, but is simply required to ensure that $M$ reaches a value, rather than a general normal form like $xy$.
 
@@
         (E[\underline{s(\tsample_i)}],s\mid_{F(E[\underline{s(\tsample_i)}])}) & \text{if } M = E[\tsample_i] \\
-        (M,s) & \text{if } M \text{ normal form}
+        (M,s) & \text{if } M \text{ is a value}
     \end{array} \right .
@@
 \section{Ranking functions}
-Given a probabilistic program (i.e.~a term $M$), in order to construct a supermartingale to prove its a.s.~termination, a function to assign values to each reachable program state is necessary.
+Given a probabilistic program (i.e.~a closed term $M$), in order to construct a supermartingale to prove its a.s.~termination, a function to assign values to each reachable program state is necessary.
 Let $Rch(M) = \{x \in \Lambda \mid \exists (y_i) : M \to y_0 \to \dots \to y_n \to x\}$, with the $\sigma$-algebra induced as a subset of $\Lambda$, then define a \emph{ranking function on $M$} to be a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that
@@
     \item $f(E[\tY \lambda x. N] \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z]) \text{ where $z$ is not free in $N$}$
-    \item $f(E[\tsample]) \geq \int_0^1 f(E[\underline{x}]) \, \mathrm{d}x$
-
-    \lo{I agree that it does not actually matter whether $\mathit{Rch}(M)$ is a measurable subset of $\Lambda$ (the set of SPCF terms).}
-
-    \lo{It is straightforward to see that $E[\cdot] : x \mapsto E[\underline x]$ is a measurable function $\mathbb{R} \to \mathit{Rch}(M)$, using the ``skeletal term'' $\sigma$-algebra on $\mathit{Rch}(M)$, and so, $x \mapsto f(E[\underline{x}])$ is measurable.}
-
-    \lo{I assume you mean $\int_{I} \mathsf{Leb} (\mathrm{d} x) \, f(E[\underline{x}])$.} 
-    \akr{Is the meaning of the integral unclear in the context?} 
-    \lo{It it -- I was being pedantic. What you write is a Riemann integral, but it coincides with Lebesgue in this case.}
-
+    \item $f(E[\tsample]) \geq \int_I f(E[\underline{x}]) \, \mu_{leb}(\mathrm{d}x)$
     \akr{I'm not sure whether $Rch(M)$ is a measurable subset of $\Lambda$, so it might not have a natural measure, but I'm pretty sure this doesn't actually matter, and the $\sigma$-algebra at least is well-defined.
@@
     \lo{It is worth inserting the preceding (including the example of a measurable function with non-measurable range) as a remark (or as a ``commenting.sty'' note).}
-    \lo{For the integral to be well-defined, we need $x \mapsto E[\underline x]$ to be a measurable function $\mathbb{R} \to \mathit{Rch}(M)$.}
+    \akr{I'd like to include such a remark, but the only reference I've been able to find for the fact that these functions exist is a Stack Exchange question\footnote{\url{https://math.stackexchange.com/questions/1717282/construction-of-a-borel-measurable-function-mapping-borel-set-to-non-borel-set?rq=1}} which implies that the construction is difficult but doesn't actually say what it is (so I probably won't be able to find it on my own).}
 
@@
 
-Any term for which a ranking function exists is called ``rankable''.
+Any closed term for which a ranking function exists is called ``rankable''.
 
@@
 \begin{theorem} \label{rankable implies ast}
-  If an SPCF term is rankable, it terminates almost surely.
+  If a closed SPCF term is rankable, it terminates almost surely.
 \end{theorem}
 \begin{proof}
-Given a term $M$ and a ranking function $f$ for it, define random variables on the probability space $S$ (where $s$ is a random variable) by
+Given a closed term $M$ and a ranking function $f$ for it, define random variables on the probability space $S$ (where $s$ is a random variable) by
 \begin{align*}
@@
 y_0 & = 0 \\
-y_{n+1} & = \min \{ k \mid k>y_n, M_k \text{ normal form or of the form } E[\tY N] \}\\
+y_{n+1} & = \min \{ k \mid k>y_n, M_k \text{ a value or of the form } E[\tY N] \}\\
 M'_n & = M_{y_n} \\
@@
 Given $M'_n$, there is some finite bound on the number of reduction steps that can take place from $M'_n$ without a $\tY$-reduction step, because of the type system, therefore $y_{n+1}$ is (conditional on $\mathcal{F}_{y_n+1}$) a bounded stopping time, therefore $\mathbb{E}[f(M_{y_{n+1}}) \mid \mathcal{F}_{y_n+1}] \leq f(M_{y_n+1})$.
-If $M_{y_n}$ isn't already in normal form, then \changed[lo]{$M_{y_n} = E[\tY \lambda x.N]$ for some $E, N$, therefore $M_{y_n+1} = E[\lambda z.
+If $M_{y_n}$ isn't already a value, then \changed[lo]{$M_{y_n} = E[\tY \lambda x.N]$ for some $E, N$, therefore $M_{y_n+1} = E[\lambda z.
 N[(\tY \lambda x.
-N)/x] z]$} and $f(M_{y_n+1}) \leq f(M_{y_n}) - 1$, therefore if $M_{y_n}$ isn't normal form, $\mathbb{E}[X_{n+1} \mid \mathcal{F}_{y_n+1}] \leq X_n - 1$.
+N)/x] z]$} and $f(M_{y_n+1}) \leq f(M_{y_n}) - 1$, therefore if $M_{y_n}$ isn't a value, $\mathbb{E}[X_{n+1} \mid \mathcal{F}_{y_n+1}] \leq X_n - 1$.
 
@@
 \begin{equation}
-\tif{-\tsample < 0}{\underline{0}}{(\tY \lambda x. x)x}
+\tif{-\tsample < 0}{\underline{0}}{(\tY \lambda x. x)\underline 0}
 \end{equation}
 terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x.
-x)x$ is reachable, although that has probability 0.
+x) \underline 0$ is reachable, although that has probability 0.
 Not only is this counterexample a.s.t, it's positively almost surely terminating i.e.~the expected time to termination is finite.
 
-A ranking function can be constructed under the stronger assumptions that, for every $N$ reachable from $M$, the expected number of $\tY$-reduction steps from $N$ to a normal form is finite.
+A ranking function can be constructed under the stronger assumptions that, for every $N$ reachable from $M$, the expected number of $\tY$-reduction steps from $N$ to a value is finite.
 In particular, the expected number of $\tY$-reduction steps from each reachable term is a ranking function.
@@
 \begin{theorem} \label{minimal}
-Given a term $M$, the function $f:Rch(M) \to \mathbb R$ given by $f(N) = \mathbb E [\text{the number of }\tY\text{-reduction steps from }N\text{ to a normal form}]$, if it exists, is the least of all possible ranking functions of $M$.
+Given a closed term $M$, the function $f:Rch(M) \to \mathbb R$ given by $f(N) = \mathbb E [\text{the number of }\tY\text{-reduction steps from }N\text{ to a value}]$, if it exists, is the least of all possible ranking functions of $M$.
 \end{theorem}
@@
 Let $f$ be the candidate least ranking function defined above, and suppose $g$ is another ranking function such that $f(N) > g(N)$ for some $N \in Rch(M)$.
-The difference $g - f$ is then a supermartingale (with the same setup as in the proof of Theorem \ref{rankable implies ast}.
+The difference $g - f$ is then a supermartingale (with the same setup as in the proof of Theorem \ref{rankable implies ast}).
 The restrictions of $f$ and $g$ to $Rch(N)$ have the same properties assumed of $f$ and $g$, so assume w.l.o.g.~that $N=M$.
@@
     \item $(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{i}$
-    \item $(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{i}$
+    \item $(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{i}$
     \item $(\lambda n: \tif{\tsample - 0.5 < 0}{n}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{n}{f (m+1)}) (m+1)}) \underline{i}$

### 2020-04-20 probabilisticProgrammingMartingales.tex
Add partial ranking functions
Define partial ranking functions, and show that they also suffice
for almost sure termination.

@@
 \usepackage{amsthm}
+\usepackage{amssymb}
 \usepackage{verbatim}
@@
 \DeclareMathOperator{\red}{red}
+\DeclareMathOperator{\next}{next}
 
@@
 \begin{itemize}
-    \item $(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{i}$
+    \item $(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{0}$
+    \item $(\lambda z.(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) z) \underline{i}$
     \item $(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{i}$
@@
     \item $\underline{i}$
-    \item $(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)$.
+    \item $(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)$
+    \item $(\lambda z.(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) z) (\underline{i} + 1)$.
 \end{itemize}
@@
 Even in this simple case, defining a ranking function explicitly is awkward because of the number of cases, although in most cases, because the value need only be greater than or equal to that of the next term in sequence, it suffices to take the ranking function as taking only 3 distinct values.
+
+The definition of rankability is also inconvenient for syntactic sugar.
+It could be useful, for example, to define $M \oplus_p N = \tif{\tsample - p < 0} M N$, where $M \oplus_p N$ reduces to $M$ or $N$, depending on the first value of $s$, with probability $p$ resp.~$(1-p)$.
+Technically though, it reduces first to $\tif{\underline r - p < 0} M N$ for $r \in I$, so those terms all need values of the ranking function too.
+
+\paragraph{}
+In both of these cases, there are only some values of the rankng function that are semantically important.
+Define a \emph{partial ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \nrightarrow \mathbb R$ such that
+\begin{itemize}
+    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
+    \item $f$ is defined at $M$.
+    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will almost certainly eventually reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
+\end{itemize}
+A partial ranking function that is total is just a ranking function.
+Providing a partial ranking function is essentially part way between providing a ranking function and directly proving almost sure termination.
+
+\begin{theorem} \label{partial implies rankable}
+Every partial ranking function is a restriction of a ranking function.
+\end{theorem}
+\begin{proof}
+Take a closed term $M$ and partial ranking function $f$ on $M$. Define $f_1 : Rch(M) \nrightarrow \mathbb R$ by
+\begin{align*}
+    f_1(N) & = f(N) \text{ whenever $f(N)$ is defined,} \\
+    f_1(V) & = 0 \text{ for values $V$ not in the domain of $f$.}
+\end{align*}
+
+Define $(\next(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) = \left | \{m < n \mid \red^M(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$, then define $f_2(N) = \int_S f_1(\next(N,s)) + g(N,s) \, \mu(\mathrm d s)$.
+The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction).
+\end{proof}
+
+As a corollary, any term which admits a partial ranking function terminates almost surely.
+
+\akr{TODO: Show how this makes it easier to deal with the examples given.}
 

### 2020-04-24 probabilisticProgrammingMartingales.tex
Various improvements
Expand the proofs of theorems 1 and 3. Give an example of a rankable
but not PAST term. Fix the definition of partial ranking fuctions to
exclude terms which reach a non-terminating term in a null but nonempty
set. Give examples of how to use partial ranking functions.

@@
 
+\akr{This is quite a pedantic point but technically the proof wouldn't be correct without it.} In the following proof, conditional expectations (conditioning on $\mathcal F_n$) will be used.
+Conditional expectations are in general only defined up to a null set, but in this case there is a natural measure on each element of $\mathcal F_n$ (as they're always isomorphic to $S$), and this is the version of the conditional expectation that will be used.
+
 The expectation of $f(M_{n+1})$ given $\mathcal{F}_n$ is trivially less than or equal to $f(M_n)$ in the cases that $M_n \neq E[\tsample]$, and in the case of $\tsample$,
@@
 
-Given $M'_n$, there is some finite bound on the number of reduction steps that can take place from $M'_n$ without a $\tY$-reduction step, because of the type system, therefore $y_{n+1}$ is (conditional on $\mathcal{F}_{y_n+1}$) a bounded stopping time, therefore $\mathbb{E}[f(M_{y_{n+1}}) \mid \mathcal{F}_{y_n+1}] \leq f(M_{y_n+1})$.
-If $M_{y_n}$ isn't already a value, then \changed[lo]{$M_{y_n} = E[\tY \lambda x.N]$ for some $E, N$, therefore $M_{y_n+1} = E[\lambda z.
+Given $M'_n$, the reduction relation $\to$, \emph{excluding} $\tY$-reduction steps, is strongly normalising because of the type system, therefore there is some finite bound on the number of reduction steps that can take place from $M'_n$ without a $\tY$-reduction step, therefore $y_{n+1}$ is (conditional on $\mathcal{F}_{y_n+1}$) a bounded stopping time, therefore $\mathbb{E}[f(M_{y_{n+1}}) \mid \mathcal{F}_{y_n+1}] \leq f(M_{y_n+1})$.
+For $n>0$, if $M_{y_n}$ isn't already a value, then $M_{y_n} = E[\tY \lambda x.N]$ for some $E, N$, therefore $M_{y_n+1} = E[\lambda z.
 N[(\tY \lambda x.
-N)/x] z]$} and $f(M_{y_n+1}) \leq f(M_{y_n}) - 1$, therefore if $M_{y_n}$ isn't a value, $\mathbb{E}[X_{n+1} \mid \mathcal{F}_{y_n+1}] \leq X_n - 1$.
+N)/x] z]$ and $f(M_{y_n+1}) \leq f(M_{y_n}) - 1$, therefore if $M_{y_n}$ isn't a value, $\mathbb{E}[X_{n+1} \mid \mathcal{F}_{y_n+1}] \leq X_n - 1$.
 
-Overall, this implies \lo{This implication is still not clear to me.} that $\mathbb{E}[X_{n+1}] - \mathbb{E}[X_{n}] \leq -\mathbb P[M_{y_n} \text{not n.f.}]$ therefore as $\mathbb{E}[X_n]$ is bounded below, $\mathbb P[M_{y_n} \text{not n.f.}]$ must tend to 0 as $n \to \infty$.
-$M_{y_n} \text{n.f.} \Rightarrow M_{y_{n+1}} \text{n.f.}$ therefore
+In the other case that $X_n$ is a value, $X_{n+1} = X_n$, and $X_n = \mathbb E[X_n \mid \mathcal F_{y_n+1}]$ therefore in either case $\mathbb E[X_{n+1} \mid \mathcal F_{y_n+1}] - E[X_n \mid \mathcal F_{y_n+1}] \leq \mathbb -P[M_{y_n} \text{ not a value} \mid \mathcal F_{y_n+1}]$ (as the probability is 0 or 1).
+Marginalising over $\mathcal F_{y_n+1}$, this implies that $\mathbb{E}[X_{n+1}] - \mathbb{E}[X_{n}] \leq -\mathbb P[M_{y_n} \text{ not a value}]$ therefore as $\mathbb{E}[X_n]$ is bounded below, $\mathbb P[M_{y_n} \text{ not a value}]$ must tend to 0 as $n \to \infty$.
+$M_{y_n} \text{a value} \Rightarrow M_{y_{n+1}} \text{a value}$ therefore
 \begin{align*}
-\mathbb P[M_{y_n} \text{not n.f.~for infinitely many values of $n$}]  
-& = \mathbb P[\forall n: M_{y_n} \text{not n.f.}] \\
-& \leq \inf_n \mathbb P[M_{y_n} \text{not n.f.}] \\
+\mathbb P[M_{y_n} \text{ not a value for infinitely many values of $n$}]  
+& = \mathbb P[\forall n: M_{y_n} \text{ not a value}] \\
+& \leq \inf_n \mathbb P[M_{y_n} \text{ not a value}] \\
 & = 0
@@
 In particular, the expected number of $\tY$-reduction steps from each reachable term is a ranking function.
+Note that a finite number of expected $\tY$-reduction steps does not necessarily imply a finite number of expected total reduction steps, for example $(\tY \lambda f n.
+\tif{\tsample - 0.5 < 0}{n \underline 0}{f (\lambda x.
+n (n x))}) (\lambda x.
+x+1)$ terminates with only 2 $\tY$-reductions on average, but applies the increment function $2^n$ times with probability $2^{-1-n}$ for $n \geq 0$, which diverges.
 
-\akr{I don't know whether you'll see this before I actually write the other bit, but I'm planning on giving a more practical way of constructing ranking functions later using a similar technique, which is why this theorem seems relevant.}
 \begin{theorem} \label{minimal}
@@
     \item $\tif{\underline r - 0.5 < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
-    \item $\tif{\underline{r-0.5} < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
+    \item $\tif{\underline{r - 0.5} < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
     \item $\underline{i}$
@@
     \item $f$ is defined at $M$.
-    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will almost certainly eventually reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
+    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually \akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and ast, but that doesn't actually work because of terms reachable with 0 probability.
+Of course, astness doesn't imply rankability, so any definition of partial ranking function that includes all Y-past terms wouldn't always extend to a total ranking function.} reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
 \end{itemize}
@@
 
-Define $(\next(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) = \left | \{m < n \mid \red^M(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$, then define $f_2(N) = \int_S f_1(\next(N,s)) + g(N,s) \, \mu(\mathrm d s)$.
+Define $(\next(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) = \left | \{m < n \mid \red^M(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
+$\next$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on partial ranking functions.
+Define $f_2(N) = \int_S f_1(\next(N,s)) + g(N,s) \, \mu(\mathrm d s)$.
 The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction).
@@
 
-\akr{TODO: Show how this makes it easier to deal with the examples given.}
+\subsection{Examples}
+Let $M \oplus_p N = \tif{\tsample - p < 0} M N$, for $p \in (0,1]$, then there are the pseudo-reduction relations
+\begin{align*}
+E[M \oplus_p N] \to^3 & E[M] & \\
+E[M \oplus_p N] \to^3 & E[N] & \\
+\red^3(E[M \oplus_p N], s) = & \left\{
+    \begin{array}{ll}
+        (E[M],\pi_t(s)) & \text{if } \pi_h(s) < p \\
+        (E[N],\pi_t(s)) & \text{if } \pi_h(s) \geq p. \\
+    \end{array} \right .
+\end{align*}
+A partial ranking function could be defined with respect to this shortcut reduction simply by replacing $\to$ and $\red$ in the definition of a partial ranking function by a version that goes straight from $N \oplus_p O$ to $N$ or $O$.
+Such a pseudo-partial ranking function would then be a partial function from a subset of $Rch(M)$, so it could also be considered as a partial function from all of $Rch(M)$, and it would in fact also be an actual partial ranking function.
+It is therefore possible to prove rankability directly using the shortcutted reductions.
+
+A similar procedure would work for other forms of syntactic sugar.
+If a closed term $N$ eventually reduces to one of a set of other terms $\{N_i \mid i \in I\}$ with certain probabilities, a partial ranking function defined with respect to a reduction sequence that skips straight from $N$ to $N_i$ is also a valid partial ranking function for the original reduction function, and therefore its existence implies almost sure termination.
+There is a caveat, however, that $\tY$-reduction steps skipped over in the shortcut still need to be counted for the expected number of $\tY$-reduction steps.
+
+\paragraph{}
+With this abbreviation, the geometric distribution example from earlier can be written as $(\tY f, n.
+n \oplus_{0.5} f(n+1)) \underline 0$.
+It is then easy to see that the following is a partial ranking function:
+\begin{align*}
+(\tY f, n. n \oplus_{0.5} f(n+1)) N \mapsto 2 \\
+\underline i \oplus_{0.5} (\tY f, n. n \oplus_{0.5} f(n+1)) (\underline i + 1) \mapsto 1 \\
+\underline i \mapsto 0
+\end{align*}
 

### 2020-04-24 probabilisticProgrammingMartingales.tex
Finish the proof of Theorem 2 (on the least ranking function)

@@
 Let $f$ be the candidate least ranking function defined above, and suppose $g$ is another ranking function such that $f(N) > g(N)$ for some $N \in Rch(M)$.
-The difference $g - f$ is then a supermartingale (with the same setup as in the proof of Theorem \ref{rankable implies ast}).
 The restrictions of $f$ and $g$ to $Rch(N)$ have the same properties assumed of $f$ and $g$, so assume w.l.o.g.~that $N=M$.
+The difference $g - f$ is then a supermartingale (with the same setup as in the proof of Theorem \ref{rankable implies ast}) therefore $\forall n.\ \mathbb E[g(M_n)] \leq \mathbb E[f(M_n)] + g(M)-f(M)$.
 
-\akr{TODO: finish this proof. It shouldn't be complicated.}
+$\mathbb E[f(M_n)] = \sum_{k=n}^\infty \mathbb P[M_n = E[\tY N] \text{ for some }E, N] \to 0 \text{ as } n \to \infty$ therefore as $g(M) - f(M) < 0$, eventually $\mathbb E[g(M_n)] < 0$, which is impossible, therefore $g \geq f$ as required.
+
+In order for $f$ to be the least ranking function of $M$, it also has to actually be a ranking function itself.
+Each of the conditions on a ranking function is easily verified from the definition of $f$.
 \end{proof}

### 2020-04-29 probabilisticProgrammingMartingales.tex
minor

@@
 \begin{align*}
-  E ::= & \, \cdot \mid E t \mid v E \mid \underline{f}(r_1,\dots ,r_{k-1}, E, s_{k+1}, \dots, s_n) \\ & \mid \tY E \mid \tif{E<0}{s_1}{s_2} \mid \tscore (E)
+  E ::= & \, [\,] \mid E t \mid v E \mid \underline{f}(r_1,\dots ,r_{k-1}, E, s_{k+1}, \dots, s_n) \\ & \mid \tY E \mid \tif{E<0}{s_1}{s_2} \mid \tscore (E)
 \end{align*}
@@
 \end{align*}
+\lo{The $\tscore$ rule should have a side condition: if $r > 0$.}
 
@@
 \end{equation}
+\lo{Both $\to$ and $\red$ are called reduction relation.}
 
@@
 
-For example, the term $(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0}$, which generates a geometric distribution, terminates on the set $S \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas $\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x.
-x) \underline 0}$, which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.
+For example, the term 
+\[
+(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0},
+\] 
+which generates a geometric distribution, terminates on the set $S \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas 
+\[
+\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \underline 0},
+\] 
+which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.
 
 \paragraph{}
-Note that the requirement that $M$ be closed is not important for the proofs of the theorems, but is simply required to ensure that $M$ reaches a value, rather than a general normal form like $xy$.
+Note that the requirement that $M$ be closed is not important for the proofs of the theorems, but is simply required to ensure that $M$ reaches a value, rather than a general normal form like $x \, y$.
 

### 2020-04-29 probabilisticProgrammingMartingales.tex
minor
Some issues in the proof of Theorem 1.

@@
     \item $f(N) \geq 0$ for all $N$
-    \item $f(E[\tY \lambda x. N] \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z]) \text{ where $z$ is not free in $N$}$
+    \item $f(E[\tY \lambda x. N]) \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z]) \text{ where $z$ is not free in $N$}$
     \item $f(E[\tsample]) \geq \int_I f(E[\underline{x}]) \, \mu_{leb}(\mathrm{d}x)$
@@
 The expectation of $f(M_{n+1})$ given $\mathcal{F}_n$ is trivially less than or equal to $f(M_n)$ in the cases that $M_n \neq E[\tsample]$, and in the case of $\tsample$,
+\lo{This case analysis is problematic. The r.v.~$M_n$ is a measurable function of type $S \to \Lambda^0$ defined by $M_n(s) :=  N$ where $\red^n(M, s) = (N, s_n)$. 
+In general, it does not make sense to say ``$M_n = E[\tsample]$'' without referencing a trace (i.e.~an element of of $S$).
+Take $M = \tif{\tsample - 0.5 < 0}{\tsample}{1}$.
+Then $M_3([0.1, \cdots]) = \tsample$ but $M_3([0.6, \cdots]) = 1$.
+The case analysis should take place ``inside the conditional expectation''.}
 \begin{align*}
@@
 \end{align*}
+\lo{Problem: Some of the above terms (conditional expectation and $f(M_n)$) are random variables (i.e.~functions) but the integral is a real number.}
+
 therefore the values of the ranking function $f(M_n)$ are a supermartingale with respect to $\mathcal{F}_n$.
@@
 Define a \emph{partial ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \nrightarrow \mathbb R$ such that
+\lo{NOTATION. Common notation for partial function $\rightharpoonup$}
 \begin{itemize}

### 2020-04-30 probabilisticProgrammingMartingales.tex
Reworked proof of Theorem 1.

@@
 pdfstartview=FitH ]{hyperref}
+
+\usepackage{amsmath} % cleveref must be loaded afer amsmath
+\usepackage{cleveref}
+\Crefname{theorem}{Theorem}{Theorems}
+\Crefname{corollary}{Corollary}{Corollary}
+\crefname{proposition}{Proposition}{Propositions}
+\Crefname{claim}{Claim}{Claims}
+\Crefname{definition}{Definition}{Definitions}
+\Crefname{fact}{Fact}{Facts}
+\Crefname{conj}{Conjecture}{Conjectures}
+\Crefname{example}{Example}{Examples}
+\Crefname{remark}{Remark}{Remarks}
+\Crefname{convention}{Convention}{Conventions}
+\Crefname{lemma}{Lemma}{Lemmas}
+\Crefname{assumption}{Assumption}{Assumptions}
+\Crefname{section}{Section}{Sections}
+\Crefname{appendix}{Appendix}{Appendices}
+\Crefname{figure}{Figure}{Figures}
+%\Crefname{equation}{Eq.}{Equations}
+
+% \addtotheorempostheadhook[theorem]{\crefalias{thmlisti}{theorem}}
+% \addtotheorempostheadhook[lemma]{\crefalias{thmlisti}{lemma}}
+% \addtotheorempostheadhook[proposition]{\crefalias{thmlisti}{proposition}}
+% \addtotheorempostheadhook[corollary]{\crefalias{thmlisti}{corollary}}
+% \addtotheorempostheadhook[claim]{\crefalias{thmlisti}{claim}}
+% \addtotheorempostheadhook[fact]{\crefalias{thmlisti}{fact}}
+% \addtotheorempostheadhook[example]{\crefalias{thmlisti}{example}}
+% \addtotheorempostheadhook[definition]{\crefalias{thmlisti}{definition}}
+% \addtotheorempostheadhook[remark]{\crefalias{thmlisti}{remark}}
+% \addtotheorempostheadhook[assumption]{\crefalias{thmlisti}{assumption}}
+
+\newcommand{\Real}{\mathbb{R}}
+\newcommand\expect[1]{\mathbb{E}[#1]}
+\newcommand\set[1]{\{#1\}}
+
 %%
@@
 
-\usepackage{amsmath}
 \usepackage{amsfonts}
@@
 \newtheorem{theorem}{Theorem}
+\newtheorem{lemma}{Lemma}
+\newtheorem{definition}{Definition}
 
@@
 The $\sigma$-algebra and measure on $\Lambda$ are defined by considering it as a disjoint union of equivalence classes under replacing all the real constants by a placeholder, where the measure on each class is that of $\mathbb{R}^n$, where $n$ is the number of real constants.
+\changed[lo]{Precisely, following \cite{DBLP:conf/icfp/BorgstromLGS16},
+we view $\Lambda$ as $\bigcup_{m\in\omega} \big(\mathsf{Sk}_m \times \Real^m \big)$,
+where $\mathsf{Sk}_m$ is the set of SPCF terms with exactly $m$ place-holders (a.k.a.~\emph{skeleton terms}) for numerals.
+Thus identified, we give $\Lambda$ the countable disjoint union topology of the product topology of the discrete topology on $\mathsf{Sk}_m$ and the standard topology on $\Real^m$.
+Note that the connected components of $\Lambda$ have the form $\{M\} \times \Real^m$, with $M$ ranging over $\mathsf{Sk}_m$, and $m$ over $\omega$. 
+We fix the Borel algebra of this topology to be the $\sigma$-algebra on $\Lambda$.}
 
@@
 Given a probabilistic program (i.e.~a closed term $M$), in order to construct a supermartingale to prove its a.s.~termination, a function to assign values to each reachable program state is necessary.
-Let $Rch(M) = \{x \in \Lambda \mid \exists (y_i) : M \to y_0 \to \dots \to y_n \to x\}$, with the $\sigma$-algebra induced as a subset of $\Lambda$, then define a \emph{ranking function on $M$} to be a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that
+
+Let $Rch(M) := \{x \in \Lambda \mid \exists (y_i) : M \to y_0 \to \dots \to y_n \to x\}$, with the $\sigma$-algebra induced as a subset of $\Lambda$.
+
+\begin{definition}\rm
+A \emph{ranking function on $M$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that
 \begin{itemize}
@@
 
-    \item $f(E[R]) \geq f(E[N])$ for any other redex $R$, where $R \to N$.
+    \item $f(E[R]) \geq f(E[R'])$ for any other redex $R$, where \changed[lo]{$R \to R'$}. \lo{Metavariables: $M$ and $N$ for terms and programs; $R$ / $R'$ for redex / contractum.}
 \end{itemize}
-
 Any closed term for which a ranking function exists is called ``rankable''.
+\end{definition}
 
@@
 
+\lo{A brief explanatory note on supermartingales by way of motivation is needed here.}
+
+\lo{Supermartingales is a fitting title of this section.
+However the details of the supermartingle construction are suppressed in the proof.
+I propose that you reformulate the theorem as follows.}
+
 \begin{theorem} \label{rankable implies ast}
-  If a closed SPCF term is rankable, it terminates almost surely.
+\changed[lo]{If a closed SPCF term is rankable by $f$, then $(f(M_n))_{n \in \omega}$ is a supermartingale adapted to the filtration $(\mathcal{F}_n)_{n \in \omega}$ where $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.}
 \end{theorem}
-\begin{proof}
+
+%\begin{proof}
 Given a closed term $M$ and a ranking function $f$ for it, define random variables on the probability space $S$ (where $s$ is a random variable) by
@@
 Conditional expectations are in general only defined up to a null set, but in this case there is a natural measure on each element of $\mathcal F_n$ (as they're always isomorphic to $S$), and this is the version of the conditional expectation that will be used.
+\lo{I find this a little cryptic.
+Perhaps I am missing something, but the definition of conditional expectation $\expect{f(M_{n+1}) \mid \mathcal{F}_n}$ is standard.
+The stochastic process $(M_n)_{n \in \omega}$ is trivially adapted to the natural filtration $(\mathcal{F}_n)_{n \in \omega}$.}
 
@@
 \lo{Problem: Some of the above terms (conditional expectation and $f(M_n)$) are random variables (i.e.~functions) but the integral is a real number.}
+therefore the values of the ranking function $f(M_n)$ are a supermartingale with respect to $\mathcal{F}_n$.
 
-therefore the values of the ranking function $f(M_n)$ are a supermartingale with respect to $\mathcal{F}_n$.
+\lo{Take $M \in \Lambda^0$ (closed SPCF terms).
+Define, for each $n \in \omega$, the random variable $M_n : S \to \Lambda^0$ by $M_n(s) := \pi_0 (\red^n(M, s))$.
+
+Given $M \in \Lambda^0$, say that $M$ is \emph{type-1} (respectively 2, 3 and 4) if $M$ has the shape $E[\tY \lambda x.
+N]$ (respectively $E[\tsample]$, $E[R]$ where $R$ is any other redex, and of a value).
+By abuse of language, we refer to the three respective types of redexes as type-$1$, 2 and 3.
+
+Henceforth fix $M \in \Lambda^0$.
+For $i \in \{1, 2, 3\}$ and $n \in \omega$, define function $f_{i, n+1}(M) : S \to \Real$ by
+\[
+f_{i, n+1}(M)(s) :=
+\begin{cases}
+f(M_{n+1}(s)) & \textrm{if $M_n(s)$ is type-$i$}\\
+0 & \textrm{o/w}
+\end{cases}
+\]
+
+\begin{lemma}
+\label{lem:inde}
+For each $i$ and $n$, $f_{i, n+1}(M)$ is %not just $\mathcal{F}_{n+1}$-measurable but also 
+$\mathcal{F}_{n}$-measurable.
+\end{lemma}
+
+\begin{proof} First fix notation
+\begin{align*}
+\sigma(M_{n+1}) &= \sigma\big(\set{M_{n+1}^{-1}(U_{\beta_k})
+\mid \beta_k\in \mathsf{Sk}_k, U_{\beta_k} \in \mathcal{B}(\Real^k), k \geq 0}\big)\\
+\sigma(M_{n}) &= \sigma\big(\set{M_{n}^{-1}(U_{\alpha_j})
+\mid \alpha_j\in \mathsf{Sk}_j, U_{\alpha_j} \in \mathcal{B}(\Real^j), j \geq 0}\big);
+\end{align*}
+and for $\beta_k \in \mathsf{Sk}_i$, we write $\beta_k[\dagger]$ to mean the instantiation of $\beta_k$ by a $k$-vector of numerals $\dagger$; $\alpha_j[\dagger']$ has the same meaning.
+Now for each $\beta_k[\dagger] = E[R']$ where $R'$ is the contractum of a type-$i$ redex $R$, there exist $\alpha_j \in \mathsf{Sk}_j$ and $j$-vector $\dagger'$ such that $\alpha_j[\dagger'] = E[R]$ of type $i$.
+Moreover, for each $U_{\beta_k}$, there exists $U_{\alpha_j}$ such that $M_{n+1}^{-1}(U_{\beta_k}) = M_{n}^{-1}(U_{\alpha_j}) \in \sigma(M_n)$.
+\end{proof}
+
+Next we clearly have 
+\(
+f(M_{n+1}) = %\sum_{i=1}^3 f_{i, n+1}(M).
+f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M),
+\)
+$\mu$-almost-surely.
+(Unless otherwise stated, all equations and inequations between random variables are assumed to hold only $\mu$-a.s.)
+Therefore 
+\begin{align}
+& \expect{f(M_{n+1}) \mid \mathcal{F}_n}
+\nonumber \\ 
+& = \expect{f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M)  \mid \mathcal{F}_n} 
+\nonumber \\
+& = \expect{f_{1, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{2, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{3, n+1}(M) \mid \mathcal{F}_n} 
+\label{eqn:linear cond ex} \\
+& = f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M)
+\label{eqn:inde cond exp} \\
+%& = f_{1, n+1}(M) + \int_I f() + f_{3, n+1}(M) \label{eqn:inde cond exp} \\
+& \leq f(M_{n})[T_1] + f(M_{n})[T_2] + f(M_{n})[T_3]
+\label{eqn:ts} \\
+& = f(M_n). 
+\label{eqn:partition} 
+\end{align}
+\Cref{eqn:linear cond ex} follows from the linearity of conditional expectation. 
+\Cref{eqn:inde cond exp} is justified because $f_{i, n+1}(M)$ is $\mathcal{F}_n$-measurable (\Cref{lem:inde}).
+Inequation~(\ref{eqn:ts}) follows from $f_{i, n+1}(M) \leq f(M_n)[T_i]$ for each $i$. 
+Here we use the notation: given function $g : A \to \Real$ and $U \subseteq A$, $g[U] : A \to \Real$ denotes the function
+\[
+g[U](x) :=
+\begin{cases}
+g(x) & \textrm{$x \in U$}\\
+0 & \textrm{o/w}
+\end{cases}
+\]
+and write $T_i = \{s \in S \mid \textrm{$M_n(s)$ is type-$i$}\}$.
+To see $f_{2, n+1}(M) \leq f(M_n)[T_2]$,
+take $s \in T_2$. 
+Then $f_{2, n+1}(M)(s) = f(E[\underline{a}])$ and $M_n = E[\tsample]$, for some $a \in [0, 1]$ and evaluation context $E$.
+Hence 
+\[
+f_{2, n+1}(M)(s) \leq \int_I f(E[\underline{r}]) \, \mu_{leb}(\textrm{d} r)
+\leq f(E[\tsample]) = f(M_n)[T_2](s).
+\]
+Finally, \Cref{eqn:partition} holds because $\{T_1, T_2, T_3, T_4\}$ is a partition of $S$.
+}
 
@@
 therefore $M$ terminates almost surely.
-\end{proof}
+%\end{proof}
 

### 2020-05-01 probabilisticProgrammingMartingales.tex
Proof of Theorem 1: ongoing

@@
 \newcommand\set[1]{\{#1\}}
+\newcommand\dif{\mathrm{d}}
 
@@
 %\begin{proof}
+
+\lo{We also need to show $f(M_n)$ is integrable.}
+
 Given a closed term $M$ and a ranking function $f$ for it, define random variables on the probability space $S$ (where $s$ is a random variable) by
@@
 and write $T_i = \{s \in S \mid \textrm{$M_n(s)$ is type-$i$}\}$.
-To see $f_{2, n+1}(M) \leq f(M_n)[T_2]$,
-take $s \in T_2$. 
-Then $f_{2, n+1}(M)(s) = f(E[\underline{a}])$ and $M_n = E[\tsample]$, for some $a \in [0, 1]$ and evaluation context $E$.
-Hence 
-\[
-f_{2, n+1}(M)(s) \leq \int_I f(E[\underline{r}]) \, \mu_{leb}(\textrm{d} r)
-\leq f(E[\tsample]) = f(M_n)[T_2](s).
-\]
+[** To show $\expect{f(M_{n+1}) \mid \mathcal{F}_n} \leq f(M_n)$ a.s., it suffices to show
+\(\int_A \expect{f(M_{n+1}) \mid \mathcal{F}_n} \dif \mu \leq \int_A f(M_n) \dif \mu \), for all $A \in \mathcal{F}_n$. **]
+% To see $f_{2, n+1}(M) \leq f(M_n)[T_2]$, take $s \in T_2$. Then $f_{2, n+1}(M)(s) = f(E[\underline{a}])$ and $M_n = E[\tsample]$, for some $a \in [0, 1]$ and evaluation context $E$. Hence 
+% \[
+% f_{2, n+1}(M)(s) \leq \int_I f(E[\underline{r}]) \, \mu_{leb}(\textrm{d} r)
+% \leq f(E[\tsample]) = f(M_n)[T_2](s).
+% \]
 Finally, \Cref{eqn:partition} holds because $\{T_1, T_2, T_3, T_4\}$ is a partition of $S$.

### 2020-05-06 probabilisticProgrammingMartingales.tex
Reorganised proof of Theorem 1 and related developments.

--- 2020-04-30 probabilisticProgrammingMartingales.tex
@@
 \newcommand\set[1]{\{#1\}}
+\newcommand\dif{\mathrm{d}}
+\newcommand\calF{\mathcal{F}}
+\newcommand\calI{\mathcal{I}}
+\newcommand\Leb{\mathrm{Leb}}
+\newcommand\ndraw[2]{\#\mathrm{draw}_{#1}(#2)}
 
@@
 \newtheorem{theorem}{Theorem}
+\newtheorem{corollary}{Corollary}
 \newtheorem{lemma}{Lemma}
@@
 
-\begin{theorem} \label{rankable implies ast}
+\begin{theorem} 
+\label{thm:rankable gives supermartingale}
 \changed[lo]{If a closed SPCF term is rankable by $f$, then $(f(M_n))_{n \in \omega}$ is a supermartingale adapted to the filtration $(\mathcal{F}_n)_{n \in \omega}$ where $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.}
@@
 
-\lo{Take $M \in \Lambda^0$ (closed SPCF terms).
-Define, for each $n \in \omega$, the random variable $M_n : S \to \Lambda^0$ by $M_n(s) := \pi_0 (\red^n(M, s))$.
-
-Given $M \in \Lambda^0$, say that $M$ is \emph{type-1} (respectively 2, 3 and 4) if $M$ has the shape $E[\tY \lambda x.
-N]$ (respectively $E[\tsample]$, $E[R]$ where $R$ is any other redex, and of a value).
-By abuse of language, we refer to the three respective types of redexes as type-$1$, 2 and 3.
-
-Henceforth fix $M \in \Lambda^0$.
-For $i \in \{1, 2, 3\}$ and $n \in \omega$, define function $f_{i, n+1}(M) : S \to \Real$ by
-\[
-f_{i, n+1}(M)(s) :=
-\begin{cases}
-f(M_{n+1}(s)) & \textrm{if $M_n(s)$ is type-$i$}\\
-0 & \textrm{o/w}
-\end{cases}
-\]
-
-\begin{lemma}
-\label{lem:inde}
-For each $i$ and $n$, $f_{i, n+1}(M)$ is %not just $\mathcal{F}_{n+1}$-measurable but also 
-$\mathcal{F}_{n}$-measurable.
-\end{lemma}
-
-\begin{proof} First fix notation
-\begin{align*}
-\sigma(M_{n+1}) &= \sigma\big(\set{M_{n+1}^{-1}(U_{\beta_k})
-\mid \beta_k\in \mathsf{Sk}_k, U_{\beta_k} \in \mathcal{B}(\Real^k), k \geq 0}\big)\\
-\sigma(M_{n}) &= \sigma\big(\set{M_{n}^{-1}(U_{\alpha_j})
-\mid \alpha_j\in \mathsf{Sk}_j, U_{\alpha_j} \in \mathcal{B}(\Real^j), j \geq 0}\big);
-\end{align*}
-and for $\beta_k \in \mathsf{Sk}_i$, we write $\beta_k[\dagger]$ to mean the instantiation of $\beta_k$ by a $k$-vector of numerals $\dagger$; $\alpha_j[\dagger']$ has the same meaning.
-Now for each $\beta_k[\dagger] = E[R']$ where $R'$ is the contractum of a type-$i$ redex $R$, there exist $\alpha_j \in \mathsf{Sk}_j$ and $j$-vector $\dagger'$ such that $\alpha_j[\dagger'] = E[R]$ of type $i$.
-Moreover, for each $U_{\beta_k}$, there exists $U_{\alpha_j}$ such that $M_{n+1}^{-1}(U_{\beta_k}) = M_{n}^{-1}(U_{\alpha_j}) \in \sigma(M_n)$.
-\end{proof}
-
-Next we clearly have 
-\(
-f(M_{n+1}) = %\sum_{i=1}^3 f_{i, n+1}(M).
-f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M),
-\)
-$\mu$-almost-surely.
-(Unless otherwise stated, all equations and inequations between random variables are assumed to hold only $\mu$-a.s.)
-Therefore 
-\begin{align}
-& \expect{f(M_{n+1}) \mid \mathcal{F}_n}
-\nonumber \\ 
-& = \expect{f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M)  \mid \mathcal{F}_n} 
-\nonumber \\
-& = \expect{f_{1, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{2, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{3, n+1}(M) \mid \mathcal{F}_n} 
-\label{eqn:linear cond ex} \\
-& = f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M)
-\label{eqn:inde cond exp} \\
-%& = f_{1, n+1}(M) + \int_I f() + f_{3, n+1}(M) \label{eqn:inde cond exp} \\
-& \leq f(M_{n})[T_1] + f(M_{n})[T_2] + f(M_{n})[T_3]
-\label{eqn:ts} \\
-& = f(M_n). 
-\label{eqn:partition} 
-\end{align}
-\Cref{eqn:linear cond ex} follows from the linearity of conditional expectation. 
-\Cref{eqn:inde cond exp} is justified because $f_{i, n+1}(M)$ is $\mathcal{F}_n$-measurable (\Cref{lem:inde}).
-Inequation~(\ref{eqn:ts}) follows from $f_{i, n+1}(M) \leq f(M_n)[T_i]$ for each $i$. 
-Here we use the notation: given function $g : A \to \Real$ and $U \subseteq A$, $g[U] : A \to \Real$ denotes the function
-\[
-g[U](x) :=
-\begin{cases}
-g(x) & \textrm{$x \in U$}\\
-0 & \textrm{o/w}
-\end{cases}
-\]
-and write $T_i = \{s \in S \mid \textrm{$M_n(s)$ is type-$i$}\}$.
-To see $f_{2, n+1}(M) \leq f(M_n)[T_2]$,
-take $s \in T_2$. 
-Then $f_{2, n+1}(M)(s) = f(E[\underline{a}])$ and $M_n = E[\tsample]$, for some $a \in [0, 1]$ and evaluation context $E$.
-Hence 
-\[
-f_{2, n+1}(M)(s) \leq \int_I f(E[\underline{r}]) \, \mu_{leb}(\textrm{d} r)
-\leq f(E[\tsample]) = f(M_n)[T_2](s).
-\]
-Finally, \Cref{eqn:partition} holds because $\{T_1, T_2, T_3, T_4\}$ is a partition of $S$.
-}
-
 Given $M'_n$, the reduction relation $\to$, \emph{excluding} $\tY$-reduction steps, is strongly normalising because of the type system, therefore there is some finite bound on the number of reduction steps that can take place from $M'_n$ without a $\tY$-reduction step, therefore $y_{n+1}$ is (conditional on $\mathcal{F}_{y_n+1}$) a bounded stopping time, therefore $\mathbb{E}[f(M_{y_{n+1}}) \mid \mathcal{F}_{y_n+1}] \leq f(M_{y_n+1})$.
@@
 
+\input{ranking-supermartingales}
+
 \section{Constructing Ranking Functions}
@@
 \end{equation}
-terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x.
-x) \underline 0$ is reachable, although that has probability 0.
+terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x. x) \underline 0$ is reachable, although that has probability 0. 
 Not only is this counterexample a.s.t, it's positively almost surely terminating i.e.~the expected time to termination is finite.

### 2020-05-06 probabilisticProgrammingMartingales.tex
More progress towards Church-Rosser semantics
Add more detail on motivation for alternative semantics, prepare
to define the reduction relation itself and start proving the
necessary properties of it.

@@
 pdfstartview=FitH ]{hyperref}
-
-\usepackage{amsmath} % cleveref must be loaded afer amsmath
-\usepackage{cleveref}
-\Crefname{theorem}{Theorem}{Theorems}
-\Crefname{corollary}{Corollary}{Corollary}
-\crefname{proposition}{Proposition}{Propositions}
-\Crefname{claim}{Claim}{Claims}
-\Crefname{definition}{Definition}{Definitions}
-\Crefname{fact}{Fact}{Facts}
-\Crefname{conj}{Conjecture}{Conjectures}
-\Crefname{example}{Example}{Examples}
-\Crefname{remark}{Remark}{Remarks}
-\Crefname{convention}{Convention}{Conventions}
-\Crefname{lemma}{Lemma}{Lemmas}
-\Crefname{assumption}{Assumption}{Assumptions}
-\Crefname{section}{Section}{Sections}
-\Crefname{appendix}{Appendix}{Appendices}
-\Crefname{figure}{Figure}{Figures}
-%\Crefname{equation}{Eq.}{Equations}
-
-% \addtotheorempostheadhook[theorem]{\crefalias{thmlisti}{theorem}}
-% \addtotheorempostheadhook[lemma]{\crefalias{thmlisti}{lemma}}
-% \addtotheorempostheadhook[proposition]{\crefalias{thmlisti}{proposition}}
-% \addtotheorempostheadhook[corollary]{\crefalias{thmlisti}{corollary}}
-% \addtotheorempostheadhook[claim]{\crefalias{thmlisti}{claim}}
-% \addtotheorempostheadhook[fact]{\crefalias{thmlisti}{fact}}
-% \addtotheorempostheadhook[example]{\crefalias{thmlisti}{example}}
-% \addtotheorempostheadhook[definition]{\crefalias{thmlisti}{definition}}
-% \addtotheorempostheadhook[remark]{\crefalias{thmlisti}{remark}}
-% \addtotheorempostheadhook[assumption]{\crefalias{thmlisti}{assumption}}
-
-\newcommand{\Real}{\mathbb{R}}
-\newcommand\expect[1]{\mathbb{E}[#1]}
-\newcommand\set[1]{\{#1\}}
-\newcommand\dif{\mathrm{d}}
-\newcommand\calF{\mathcal{F}}
-\newcommand\calI{\mathcal{I}}
-\newcommand\Leb{\mathrm{Leb}}
-\newcommand\ndraw[2]{\#\mathrm{draw}_{#1}(#2)}
-
 %%
@@
 
+\usepackage{amsmath}
 \usepackage{amsfonts}
@@
 \newtheorem{theorem}{Theorem}
-\newtheorem{corollary}{Corollary}
 \newtheorem{lemma}{Lemma}
-\newtheorem{definition}{Definition}
 
@@
 \begin{align*}
-  E ::= & \, [\,] \mid E t \mid v E \mid \underline{f}(r_1,\dots ,r_{k-1}, E, s_{k+1}, \dots, s_n) \\ & \mid \tY E \mid \tif{E<0}{s_1}{s_2} \mid \tscore (E)
+  E ::= & \, \cdot \mid E t \mid v E \mid \underline{f}(r_1,\dots ,r_{k-1}, E, s_{k+1}, \dots, s_n) \\ & \mid \tY E \mid \tif{E<0}{s_1}{s_2} \mid \tscore (E)
 \end{align*}
@@
 \end{align*}
-\lo{The $\tscore$ rule should have a side condition: if $r > 0$.}
 
@@
 The $\sigma$-algebra and measure on $\Lambda$ are defined by considering it as a disjoint union of equivalence classes under replacing all the real constants by a placeholder, where the measure on each class is that of $\mathbb{R}^n$, where $n$ is the number of real constants.
-\changed[lo]{Precisely, following \cite{DBLP:conf/icfp/BorgstromLGS16},
-we view $\Lambda$ as $\bigcup_{m\in\omega} \big(\mathsf{Sk}_m \times \Real^m \big)$,
-where $\mathsf{Sk}_m$ is the set of SPCF terms with exactly $m$ place-holders (a.k.a.~\emph{skeleton terms}) for numerals.
-Thus identified, we give $\Lambda$ the countable disjoint union topology of the product topology of the discrete topology on $\mathsf{Sk}_m$ and the standard topology on $\Real^m$.
-Note that the connected components of $\Lambda$ have the form $\{M\} \times \Real^m$, with $M$ ranging over $\mathsf{Sk}_m$, and $m$ over $\omega$. 
-We fix the Borel algebra of this topology to be the $\sigma$-algebra on $\Lambda$.}
 
@@
 \end{equation}
-\lo{Both $\to$ and $\red$ are called reduction relation.}
 
@@
 
-For example, the term 
-\[
-(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0},
-\] 
-which generates a geometric distribution, terminates on the set $S \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas 
-\[
-\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \underline 0},
-\] 
-which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.
+For example, the term $(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0}$, which generates a geometric distribution, terminates on the set $S \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas $\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x.
+x) \underline 0}$, which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.
 
 \paragraph{}
-Note that the requirement that $M$ be closed is not important for the proofs of the theorems, but is simply required to ensure that $M$ reaches a value, rather than a general normal form like $x \, y$.
+Note that the requirement that $M$ be closed is not important for the proofs of the theorems, but is simply required to ensure that $M$ reaches a value, rather than a general normal form like $xy$.
 
@@
 \lo{I'll read this after you fix it ;-)}
+
+When proving almost sure termination in this way /akr{I guess I should move this section somewhere later.}, it is necessary to consider which terms a given term may reduce to.
+Sometimes however, the reduction that the programmer has in mind may not be strictly the call-by-value order defined so far, or considering an alternative reduction order may be simpler or more intuitive.
+
+Non-probabilistis lambda calculi generally have the Church-Rosser property, that if a term $A$ reduces to both $B_1$ and $B_2$, there is some $C$ with reduction sequences $B_1 \to^* C$ and $B_2 \to^* C$, so the reduction order mostly doesn't matter.
+In the probabilistic case, this may not be true, because $\beta$-reduction can duplicate $\tsample$s, so the outputs of the copies of the sample may be identical or independent, depending on whether the sample is taken before or after $\beta$-reduction.
+There are, however, some restricted variations on the reduction order that do not have this problem.
+
+\paragraph{}
+Even with this restriction, a sampling semantics in the style of the one already defined would not be entirely Church-Rosser, as, for example, $red^3(\tsample - \tsample, (1,0,\dots))$ would be either $1$ or $-1$ depending on the order of evaluation of the $\tsample$s, as that determines which sample from the pre-selected sequence is used for each one.
+To fix this, rather than pre-selecting samples according to the order they'll be drawn in, select them according to the position in the term where they'll be used instead.
+
+A position \akr{I'm not really sure how to phrase the citation, but this is adapded from \cite{Kennaway96infinitarylambda}.} is a finite sequence of steps into a term, defined inductively as
+\begin{align*}
+P ::= \cdot \mid \lambda ; P \mid @_1 ; P \mid @_2 ; P \mid \underline f_i ; P \mid \tY ; P \mid \textsf{if}_1 ; P \mid \textsf{if}_2 ; P \mid \textsf{if}_3 ; P \mid \tscore ; P.
+\end{align*}
+The subterm at a position is defined as
+\begin{align*}
+M \mid \cdot & = M \\
+\lambda x. M \mid \lambda ; P & = M \mid P \\
+M_1 M_2 \mid @_i ; P & = M_i \mid P \quad \text{for } i = 1,2 \\
+\underline f(M_1,\dots,M_n) \mid \underline f_i ; P & = M_i \mid P \quad \text{for }i \leq n \\
+\tY M \mid \tY ; P & = M \mid P \\
+\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; P & = M_i \mid P \quad \text{for } i = 1,2,3 \\
+\tscore(M) \mid \tscore ; P & = M \mid P
+\end{align*}
+so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g.
+$xy \mid \lambda ; \cdot$ is undefined).
+A position such that $M\mid P$ does exist is said to occur in $M$.
+Substitution at a given position, $M[N/P]$, is defined similarly.
+
+Two subterms $N_1$ and $N_2$ of a term $M$, corresponding to positions $P_1$ and $P_2$, can overlap in a few different ways.
+If $P_1$ is an initial segment of $P_2$, then $N_2$ is also a subterm of $N_1$, which is written as $P_1 \leq P_2$.
+If neither $P_1 \leq P_2$ nor $P_1 \geq P_2$, the positions are said to be disjoint.
+This is mostly relevant in that in any substitution at some position, the subterms at all other positions are unaffected.
+
+With this notation, a more general reduction relation can be defined. \akr{It would almost work to have one of the neater versions of $\tY$-reduction here.}
+\begin{align*}
+  \text{if } M \mid P = (\lambda x.s) v,\ & M \to M[s[v/x]/P] \\
+  \text{if } M \mid P = \underline f (\underline r_1, \dots , \underline r_n),\ & M \to M[\underline{f(r_1,\dots,r_n)}/P] \\
+  \text{if } M \mid P = \tY \lambda x. s,\ & M \to M[\lambda z. s[(\tY \lambda x. s)/x] z/P] \text{ where $z$ is not free in $s$}\\
+  \text{if } M \mid P = \tif{\underline r < 0}{s_1}{s_2},\ & M \to M[s_1/P] \text{ where }r < 0 \\
+  \text{if } M \mid P = \tif{\underline r < 0}{s_1}{s_2},\ & M \to M[s_2/P] \text{ where }r \geq 0 \\
+  \text{if } M \mid P = \tsample & \text{ and $\lambda$ does not occur after $@_2$ or $\tY$ in $P$},\\ & M \to M[\underline r/P] \text{ where } r \in [0,1] \\
+  \text{if } M \mid P = \tscore(\underline r),\ & M \to M[\underline r/P].
+\end{align*}
+
+\paragraph{}
+Labelling the pre-chosen samples by the positions in the term would also not work because in some cases, a $\tsample$ will be duplicated before being reduced, for example, in $(\lambda x.
+\underline x 0 \underline + x \underline 0) (\lambda y.
+\tsample)$, both of the $\tsample$ redexes that eventually occur originate at $@_2 ; \lambda$.
+It is therefore necessary to consider possible positions that may occur in other terms reachable from the original term.
+This is itself inadequate because some of the positions in different reachable terms need to be considered the same.
+
+Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a reduction sequence starting at $M$, and the second of which is a position in the final term of the reduction sequence.
+Reduction sequences are used rather than reachable terms because if the same term is reached twice, different samples may be needed, but they will be discussed as though they were terms, by abuse of notation.
+The relation $\sim$ is defined as the union of the symmetric relations $\sim_p$ and $\sim_c$ where
+\begin{enumerate}
+    \item If $N$ reduces to $O$ with the redex at position $P$, and $Q$ is a position in $N$ where $Q \not \geq P$, then $(N,Q) \sim_p (O,Q)$.
+    \item If $N$ $\beta$-reduces to $O$ at position $P$, $Q$ is a position in $N | P;@_1;\lambda$ and $N \mid P;@_1;\lambda;Q$ is not the variable involved in the reduction, $(N,P;@_1;\lambda;Q) \sim_p (O, P;Q)$
+    \item If $N$ $\textsf{if}$-reduces to $O$ at position $P$, with the first resp.
+second branch being taken, and $P;\textsf{if}_i;Q$ occurs in $N$ (where $i = 2$ resp.
+$3$), $(N,P;\textsf{if}_i;Q) \sim_p (O,P;Q)$
+    \item If $N \mid P = (\lambda x.
+A) B$, $A \mid Q = x$, $B \mid R = C$ is a redex reducing to $C'$ and $S$ exists in $C'$, then letting $O_1$ be $N$ reduced first at $P$ then at $P;Q;R$ and letting $O_2$ be $N$ reduced first at $P;@_2;R$ then at $P$, $(O_1,P;Q;R;S) \sim_c (O_2,P;Q;R;S)$
+    \item If $N \mid P = (\tY \lambda x.
+A)$, $A \mid Q = C$ is a redex reducing to $C'$, $R$ exists in $C'$ and $C' \mid R \neq x$, then letting $O_1$ be $N$ reduced first at $P$ then $P;\lambda;@_1;Q$ and $O_2$ reduced at $P;\tY;lambda;Q$ then $P$, $(O_1, P;\lambda;@_1;Q;R) \sim_c (O_2, P;\lambda;@_1;Q;R)$
+    \item If $N \mid P = (\tY \lambda x.
+A)$, $A \mid Q = C$ is a redex reducing to $C'$, $R$ exists in $C'$ and $A \mid S = x$, then letting $O_1$ be $N$ reduced first at $P$ then $P;\lambda;@_1;S;\tY;\lambda;Q$ and $O_2$ reduced at $P;\tY;\lambda;Q$ then $P$, $(O_1, P;\lambda;@_1;S;\tY;\lambda;Q;R) \sim_c (O_2, P;\lambda;@_1;S;\tY;\lambda;Q;R)$
+\end{enumerate}
+
+The reflexive transitive closure $\sim^*$ of this relation is used to define the set of potential positions $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable terms.
+If $(N,P) \sim^* (O,Q)$, then $N \mid P$ and $O \mid Q$ both have the same shape (i.e.~they're either both variables, both applications, both $\tsample$s ect.), therefore it's well-defined to talk of the set of potential positions where there is a $\tsample$, $L_s(M)$.
+The new sample space is then defined as $I^{L_s(M)}$, with the Borel $\sigma$-algebra and product measure.
+
+\paragraph{}
+Before defining the new version of the reduction relation $\red$, a few lemmas about the properties of $\sim$ are necessary for it to be well-defined.
+\begin{lemma}
+If $(N,P) \sim^* (N,P')$, then $P = P'$.
+\end{lemma}
+\begin{proof}
+Suppose contrariwise that $(N,P) \sim^* (N,P')$ where $P \neq P'$.
+Pick a path $(N,P) \sim (N_1,P_1) \sim \dots \sim (N_n,P_n) \sim (N,P')$ and assume without loss of generality that the sequence is the shortest for any such counterexample, then $\forall i:N_i \neq N$.
+
+The set of paths through $Rch(M)$ (which we're implicitly identifying with $Rch(M)$ itself) has a tree structure which restricts the possibilities for $N_i$.
+The only terms that $~$ relates directly are parents and children, for $\sim_p$, and first cousins for $\sim_c$.
+For any term $O$ that's the result of a reduction $O' \to O$ with the redex at position $P$, the positions in $O$ can be classified as old positions (those that correspond uniquely to structure carried over from $O'$, and which are related to positions in $O'$ by $\sim_p$), and new positions (those that either may have been duplicated in the reduction from $O'$, or aren't directly carried over).
+New positions are never related to positions in $O'$ by $\sim_p$, and old positions are never related to anything by $\sim_c$.
+Furthermore, if $(O_1,Q_1) \sim (O_2,Q_2), (O_2,Q_2')$, then $Q_2 = Q_2'$, by an exhaustive check of the cases of the positions relative to the positions of the redexes, therefore taking this together with the minimality of the counterexample, none of the $N_i$ are equal to each-other.
+
+Now simply take cases for how the sequence $N_i$ moves around the tree structure of $Rch(M)$.
+If it starts by going to a cousin of $N$, it is then on a new position within whatever term it reached therefore it can only proceed to the child of the new node or another cousin (as new positions are not related to anywhere in the parent term), at which point it would have the same problem, so it must eventually descend the tree.
+Having done so, it is unable to turn back (as all the $N_i$ are disjoint) or take any sideways steps (as it must not be on an old position), therefore it can never return to $N$, which is a contradiction, therefore $N_1$ is not a cousin of $N$.
+Symmetrically, neither is $N_n$.
+Similarly, if either $N_1$ or $N_n$ is a child of $N$, the sequence is stuck descending the tree and can't return to $N$.
+Both $N_1$ and $N_n$ are therefore the parent of $N$, therefore $n = 1$, but in that case $P = P'$, which is a contradiction, therefore the assumed couterexample could not exist.
+\end{proof}
+
+At each reduction step $M \to N$, the sample space must be restricted to $I^{L_s(N)}$, which requires a map $L_s(N) \to L_s(M)$.
+The injection $L_0(N) \to L_0(M)$ is trivial to define by appending $M \to N$ to each path, but in order for this to extend to the quotient, the following lemma is needed.
+\begin{lemma}
+If $M \to N$, the relation $\sim_N^*$ defined on $L_0(N)$ is equal to the restriction of $\sim_M^*$ from $L_0(M)$ to $L_0(N)$.
+\end{lemma}
+\begin{proof}
+If $(O_1,P_1) \sim_N (O_2,P_2)$, they're related by $\sim_M$ for exactly the same reason.
+Conversely, consider $(O_1,P_1) \sim_M (O_2,P_2)$ where both $O_1$ and $O_2$ are reachable from $N$ (i.e.~ considering them as paths, they start with the reduction $M \to N$).
+If they're related by $\sim_{p,M}$, they're related by $\sim_N$ for the same reason.
+In the case of $\sim_c$, the same proof can only be used for $\sim_N$ if the initial term in these cases ($N'$ here, because of the naming overlap) is reachable from $N$.
+Given that $O_1$ and $O_2$ are actually reduction sequences, $N'$ is their longest common prefix and they're both reachable from $N$, $N'$ must also be reachable from $N$ therefore $(O_1,P_1) \sim_N (O_2,P_2)$.
+
+The only case left to check is that $(O_1,P_1) \sim_M (O_1',P_1') \sim_M^* (O_2',P_2') \sim_M (O_2,P_2)$, where $O_1$ and $O_2$ are reachable from $N$ but $O_1'$ and $O_2'$.
+If the first relation in this sequence is $\sim_{p,M}$, then $O_1 = N$ and $O_1' = M$.
+Similarly for the last relation,
+\end{proof}
+Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function in this case, and the entropy value does not change, so it is denoted ``$\to_s$'' instead of ``$\red$'', and it relates $Rch(M)$ to itself.
+\begin{align*}
+& N \to_s O \text{ if $N \to O$ and the redex is not $\tsample$.} \\
+& N \to_s N[\underline{s(N,P)}/P] \text{ if $N \mid P = \tsample$ and $\lambda$ does not occur after $@_2$ of $\tY$ in $P$}
+\end{align*}
+
+\akr{Obsolete but keeping for now just in case:
+Given a closed term $M$, let $L_v(M), L_Y(M)$ and $L_s(M)$ be the sets of positions such that $M \mid P$ is a variable, $\tY N$ for some N, and $\tsample$, respectively.
+Although $L_s(M)$ is all the positions where sampling could occur immediately in M, in order to define values for all the samples that could possibly be taken in the course of the reduction of $M$, it is necessary to consider possible rearrangements that may occur later.
+The set of potential positions $F_x(M)$ (where $x$ stands for $v, Y$ or $s$) is defined inductively by
+\begin{align*}
+\frac{a \in L_x(M)}{a \in F_x(M)} \qquad
+\frac{a \in F_v(M), b \in F_x(M)}{(ab) \in F_x(M)} \qquad
+\frac{a \in F_Y(M)}{(Y_\eta a) \in F_v(M)} \qquad \\
+\frac{a \in F_Y(M), b \in F_x(M)}{(Y_oab) \in F_x(M)} \qquad
+\frac{}{((ab)(ac)) = (a(bc))} \qquad
+\frac{}{((Y_oab)(Y_oac)) = (Y_oa(bc))} \qquad
+\end{align*}
 
@@
 
-Essentially the entropy value here contains a pre-selected random value for each possible \tsample-reduction, in a way independent of reduction order, taking into account the possible duplication of \tsample{}s in $\beta$- and \tY-reduction steps.
+Essentially the entropy value here contains a pre-selected random value for each possible \tsample-reduction, in a way independent of reduction order, taking into account the possible duplication of \tsample{}s in $\beta$- and \tY-reduction steps.}
 
@@
 Given a probabilistic program (i.e.~a closed term $M$), in order to construct a supermartingale to prove its a.s.~termination, a function to assign values to each reachable program state is necessary.
-
-Let $Rch(M) := \{x \in \Lambda \mid \exists (y_i) : M \to y_0 \to \dots \to y_n \to x\}$, with the $\sigma$-algebra induced as a subset of $\Lambda$.
-
-\begin{definition}\rm
-A \emph{ranking function on $M$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that
+Let $Rch(M) = \{x \in \Lambda \mid \exists (y_i) : M \to y_0 \to \dots \to y_n \to x\}$, with the $\sigma$-algebra induced as a subset of $\Lambda$, then define a \emph{ranking function on $M$} to be a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that
 \begin{itemize}
     \item $f(N) \geq 0$ for all $N$
-    \item $f(E[\tY \lambda x. N]) \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z]) \text{ where $z$ is not free in $N$}$
+    \item $f(E[\tY \lambda x. N] \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z]) \text{ where $z$ is not free in $N$}$
     \item $f(E[\tsample]) \geq \int_I f(E[\underline{x}]) \, \mu_{leb}(\mathrm{d}x)$
@@
 
-    \item $f(E[R]) \geq f(E[R'])$ for any other redex $R$, where \changed[lo]{$R \to R'$}. \lo{Metavariables: $M$ and $N$ for terms and programs; $R$ / $R'$ for redex / contractum.}
+    \item $f(E[R]) \geq f(E[N])$ for any other redex $R$, where $R \to N$.
 \end{itemize}
+
 Any closed term for which a ranking function exists is called ``rankable''.
-\end{definition}
 
@@
 
-\lo{A brief explanatory note on supermartingales by way of motivation is needed here.}
-
-\lo{Supermartingales is a fitting title of this section.
-However the details of the supermartingle construction are suppressed in the proof.
-I propose that you reformulate the theorem as follows.}
-
-\begin{theorem} 
-\label{thm:rankable gives supermartingale}
-\changed[lo]{If a closed SPCF term is rankable by $f$, then $(f(M_n))_{n \in \omega}$ is a supermartingale adapted to the filtration $(\mathcal{F}_n)_{n \in \omega}$ where $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.}
+\begin{theorem} \label{rankable implies ast}
+  If a closed SPCF term is rankable, it terminates almost surely.
 \end{theorem}
-
-%\begin{proof}
-
-\lo{We also need to show $f(M_n)$ is integrable.}
-
+\begin{proof}
 Given a closed term $M$ and a ranking function $f$ for it, define random variables on the probability space $S$ (where $s$ is a random variable) by
@@
 Conditional expectations are in general only defined up to a null set, but in this case there is a natural measure on each element of $\mathcal F_n$ (as they're always isomorphic to $S$), and this is the version of the conditional expectation that will be used.
-\lo{I find this a little cryptic.
-Perhaps I am missing something, but the definition of conditional expectation $\expect{f(M_{n+1}) \mid \mathcal{F}_n}$ is standard.
-The stochastic process $(M_n)_{n \in \omega}$ is trivially adapted to the natural filtration $(\mathcal{F}_n)_{n \in \omega}$.}
 
 The expectation of $f(M_{n+1})$ given $\mathcal{F}_n$ is trivially less than or equal to $f(M_n)$ in the cases that $M_n \neq E[\tsample]$, and in the case of $\tsample$,
-\lo{This case analysis is problematic. The r.v.~$M_n$ is a measurable function of type $S \to \Lambda^0$ defined by $M_n(s) :=  N$ where $\red^n(M, s) = (N, s_n)$. 
-In general, it does not make sense to say ``$M_n = E[\tsample]$'' without referencing a trace (i.e.~an element of of $S$).
-Take $M = \tif{\tsample - 0.5 < 0}{\tsample}{1}$.
-Then $M_3([0.1, \cdots]) = \tsample$ but $M_3([0.6, \cdots]) = 1$.
-The case analysis should take place ``inside the conditional expectation''.}
 \begin{align*}
@@
 \end{align*}
-\lo{Problem: Some of the above terms (conditional expectation and $f(M_n)$) are random variables (i.e.~functions) but the integral is a real number.}
 therefore the values of the ranking function $f(M_n)$ are a supermartingale with respect to $\mathcal{F}_n$.
@@
 therefore $M$ terminates almost surely.
-%\end{proof}
-
-\input{ranking-supermartingales}
+\end{proof}
 
@@
 \end{equation}
-terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x. x) \underline 0$ is reachable, although that has probability 0. 
+terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x.
+x) \underline 0$ is reachable, although that has probability 0.
 Not only is this counterexample a.s.t, it's positively almost surely terminating i.e.~the expected time to termination is finite.
@@
 Define a \emph{partial ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \nrightarrow \mathbb R$ such that
-\lo{NOTATION. Common notation for partial function $\rightharpoonup$}
 \begin{itemize}

### 2020-05-09 probabilisticProgrammingMartingales.tex
Further work on "reorganised proof".

--- 2020-05-06 probabilisticProgrammingMartingales.tex
@@
 
+%\usepackage[bbgreekl]{mathbbol} %for \bblambda
 %%
@@
 
+\usepackage{amsthm}
+\theoremstyle{definition}
+\newtheorem{definition}{Definition}
+\newtheorem{example}{Example}
+
+\theoremstyle{lemma}
+\newtheorem{lemma}{Lemma}
 \newtheorem{theorem}{Theorem}
 \newtheorem{corollary}{Corollary}
-\newtheorem{lemma}{Lemma}
-\newtheorem{definition}{Definition}
 
+\theoremstyle{remark}
+\newtheorem*{remark}{Remark}
 \begin{document}
@@
 \end{align*}
+\changed[lo]{We write $\to^\ast$ for the reflexive, transitive closure of $\to$.}
 \lo{The $\tscore$ rule should have a side condition: if $r > 0$.}
@@
 To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples.
-Let $ I = [0,1] \subset \mathbb{R} $, and let $S = I^{\mathbb{N}}$, with the Borel $\sigma$-algebra and the probability measure, denoted $\mu$, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
-Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and $\mu (\prod_{i=0}^\infty X_i) = \prod_{i=0}^\infty \mu_{leb}(X_i)$.
+Let $ I = [0,1] \subset \mathbb{R} $, and let $S = I^{\mathbb{N}}$, with the \changed[lo]{Borel $\sigma$-algebra $\calF$} and the probability measure, denoted $\mu$, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
+Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and $\mu (\prod_{i=0}^\infty X_i) = \prod_{i=0}^\infty \Leb(X_i)$.
 The maps $\pi_h:S \to I, \; \pi_t:S \to S$ popping the first element are then measurable.
+\changed[lo]{Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(S, \calF, \mu)$ the \emph{entropy space}.}
 
@@
 A term $M$ is said to terminate almost surely if $\mu(\{s \mid M \text{ terminates at } s\}) = 1$.
+
+\lo{Alternatively, define the \emph{runtime of $M$} to be the random variable 
+\[
+T_M(s) := 
+\begin{cases}
+\min \set{n \mid \pi_0(\red^n(M, s)) \textrm{ is a value}} & \hbox{if $\red^\infty(M,s)$ is defined}\\
+\infty & \hbox{otherwise}
+\end{cases}
+\]
+Equivalently, we say that $M$ is \emph{almost-surely terminating} (AST) if $T_M < \infty$ a.s.; 
+and $M$ is \emph{positively almost-surely terminating} (PAST) if $\expect{T_M} < \infty$.}
 
@@
 
-Let $Rch(M) := \{x \in \Lambda \mid \exists (y_i) : M \to y_0 \to \dots \to y_n \to x\}$, with the $\sigma$-algebra induced as a subset of $\Lambda$.
+Let %$Rch(M) := \{x \in \Lambda \mid \exists (y_i) : M \to y_0 \to \dots \to y_n \to x\}$, 
+\changed[lo]{$Rch(M) := \{N \in \Lambda \mid M \to^\ast N\}$},
+with the $\sigma$-algebra induced as a subset of $\Lambda$.
 
 \begin{definition}\rm
-A \emph{ranking function on $M$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that
-\begin{itemize}
-    \item $f(N) \geq 0$ for all $N$
+A \emph{ranking function on $M$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
+\begin{enumerate}
     \item $f(E[\tY \lambda x. N]) \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z]) \text{ where $z$ is not free in $N$}$
-    \item $f(E[\tsample]) \geq \int_I f(E[\underline{x}]) \, \mu_{leb}(\mathrm{d}x)$
+    \item $f(E[\tsample]) \geq \int_I f(E[\underline{x}]) \, \Leb(\mathrm{d}x)$
     \akr{I'm not sure whether $Rch(M)$ is a measurable subset of $\Lambda$, so it might not have a natural measure, but I'm pretty sure this doesn't actually matter, and the $\sigma$-algebra at least is well-defined.
@@
     \item $f(E[R]) \geq f(E[R'])$ for any other redex $R$, where \changed[lo]{$R \to R'$}. \lo{Metavariables: $M$ and $N$ for terms and programs; $R$ / $R'$ for redex / contractum.}
-\end{itemize}
-Any closed term for which a ranking function exists is called ``rankable''.
+\end{enumerate}
+We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that $f(N) = 0$ iff $N$ is a value, and for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.
+
+Any closed term for which a ranking (respectively, strict ranking) function exists is called \emph{rankable} (respectively, \emph{strictly rankable}).
 \end{definition}
@@
 \section{Constructing Ranking Functions}
-Although rankability implies almost sure termination, the converse does not hold in general. For example,
+Although rankability implies almost-sure termination, the converse does not hold in general. For example,
 \begin{equation}
@@
 In particular, the expected number of $\tY$-reduction steps from each reachable term is a ranking function.
-Note that a finite number of expected $\tY$-reduction steps does not necessarily imply a finite number of expected total reduction steps, for example $(\tY \lambda f n.
-\tif{\tsample - 0.5 < 0}{n \underline 0}{f (\lambda x.
-n (n x))}) (\lambda x.
-x+1)$ terminates with only 2 $\tY$-reductions on average, but applies the increment function $2^n$ times with probability $2^{-1-n}$ for $n \geq 0$, which diverges.
+Note that a finite number of expected $\tY$-reduction steps does not necessarily imply a finite number of expected total reduction steps.
+
+\begin{example}
+\label{ex:tY finite does not imply t finite}
+The term
+\[
+M = (\tY \lambda f n. \tif{\tsample - 0.5 < 0}{n \underline 0}{f (\lambda x. n (n x))}) (\lambda x. x+1)
+\] 
+terminates with only 2 $\tY$-reductions on average i.e.~$\expect{T_M^\tY} = 2$, but applies the increment function $2^n$ times with probability $2^{-1-n}$ for $n \geq 0$, which diverges i.e.~$\expect{T_M} = \infty$.
+\end{example}
 

### 2020-05-09 ranking-supermartingales.tex
Further work on "reorganised proof".

%!TEX root = probabilisticProgrammingMartingales.tex
\subsection*{Reorganised proof}

Fix a probability space $(\Omega, \calF, \mathbb{P})$. 
We have in mind $\Omega = S$, and $\mathbb{P} = \mu$.

\begin{definition}\rm
A sequence of random variables $(Y_n)_{n \in \omega}$ adapted to a filtration $(\calF_n)_{n \in \omega}$ is a \emph{supermartingale} if for all $n \in \omega$, $\expect{|Y_n|} < \infty$, and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$.
\lo{The preceding means: $\forall A \in \calF_n$, $\int_A \dif \mu \, Y_{n+1} \leq \int_A \dif \mu \,Y_n$.}
For $\epsilon > 0$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \, \mathbf{1}_{\set{Y_n > 0}}$.
\citep{DBLP:conf/popl/FioritiH15,DBLP:conf/popl/ChatterjeeFNH16}
\end{definition}

Intuitively $Y_n$ is the rank of the program after $n$ steps of computation.
(Assume that the rank of a term is 0 iff it is a value.) In a $\epsilon$-ranking supermartingale, each computation step causes a strict decrease in rank, provided the term being reduced is not a value.

Let $S$ and $T$ be stopping times adapted to $(\calF_n)_{n \in \omega}$.
Recall the $\sigma$-algebra (consisting of measurable subsets ``priori to $T$'')
\[
\calF_T := \set{A \in \calF \mid \forall i \in \omega \, . \, A \cap \set{T \leq i} \in \calF_i}
\]
and if $S \leq T$, then $\calF_S \subseteq \calF_{T}$.

The following is an iterated version of Doob's well-known Optional Sampling Theorem (see, e.g., \cite[\S 6.7]{AshDD00}).
\begin{theorem}[Optional Sampling]
\label{thm:optional sampling}
Let $(X_n)_{n\in \omega}$ be a supermartingale, and $(T_n)_{n \in \omega}$ a sequence of increasing stopping times, then $(X_{T_n})_{n \in \omega}$ is a supermartingale adapted to $(\calF_{T_n})_{n \in \omega}$ if one of the following conditions holds:
\begin{enumerate}
\item each $T_n$ is bounded i.e.~$T_n < c_n$ where $c_n$ is a constant
\item $(X_n)_{n\in \omega}$ is uniformly integrable.
\end{enumerate}
\end{theorem}

\begin{theorem}[{\citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}, \citep[Prop 1]{DBLP:conf/popl/ChatterjeeFNH16}}]
\label{thm:rank-PAST}
Let $(Y_n)_{n \in \omega}$ be a $\epsilon$-ranking supermartingale, and set $T(s) := \min \set{n \mid Y_n(s) = 0}$.

Then $T < \infty$ almost surely, and $\expect{T} \leq \dfrac{\expect{Y_0}}{\epsilon}$.
\end{theorem}

A main result is the following theorem.
\begin{theorem} 
\label{thm:rankable and strict rankable}
If a closed SPCF term is rankable (respectively, strictly rankable) by $f$, then $(f(M_n))_{n \in \omega}$ is a 
supermartingale (respectively, ranking supermartingale) adapted to the filtration $(\mathcal{F}_n)_{n \in \omega}$ where $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
\end{theorem}

\paragraph{Proof of \Cref{thm:rankable and strict rankable}.}
\emph{Notation}. Given an $\omega$-sequence (e.g.~$s \in S$) and $m \in \omega$, we write $s_{\leq m} \in I^m$ to mean the $m$-long prefix of $s$.

Fix a closed SPCF term $M \in \Lambda^0$, and a ranking function $f$ for it, satisfying $f(N) = 0$ iff $N$ is a value.
Let $n \in \omega$, define the following random variables on the probability space $(S, \calF, \mu)$:
\begin{align*}
M_n(s) &:= \pi_0 (\red^n(M, s))\\
\ndraw{n}{s} &:= l \textrm{ where $\pi_1 (\red^n(M, s)) = \underbrace{\pi_t( \cdots (\pi_t}_l(s))$}
\end{align*}
and the filtration $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
The $\calF_n$-measurability of $M_n$ (and hence $\#\mathrm{draw}_n$) follow from \citep{DBLP:conf/icfp/BorgstromLGS16};
$\#\mathrm{draw}_n$ is a stopping time (bounded by $n$).
Take $s \in A \in \calF_n$ with $\ndraw{n}{s} = l$.
For any $s'\in S$, if $s_{\leq l} = s_{\leq l}'$ then $s' \in A$.
It follows that ${\set{s_{\leq l}} \cdot I^\omega} \subseteq A$.
\lo{NOTE. To fix notation:
\(
\sigma(M_{n}) = \sigma\big(\set{M_{n}^{-1}(\alpha_j, U_{\alpha_j})
\mid \alpha_j\in \mathsf{Sk}_j, U_{\alpha_j} \in \mathcal{B}(\Real^j), j \geq 0}\big)
\)
}
%Take $M \in \Lambda^0$ (closed SPCF terms). Define, for each $n \in \omega$, the random variable $M_n : S \to \Lambda^0$ by $M_n(s) := \pi_0 (\red^n(M, s))$.

We say that a given SPCF term is \emph{type-1} (respectively 2, 3 and 4) if it has the shape $E[\tY (\lambda x. N)]$ (respectively $E[\tsample]$, $E[R]$ where $R$ is any other redex, and of a value).
Define $\mathbf{T}_i := 
\set{s \mid M_n(s) \hbox{ is type-$i$}}$.
It is straightforward to see that each $\mathbf{T}_i \in \calF_n$, and $\set{\mathbf{T}_1, \mathbf{T}_2, \mathbf{T}_3, \mathbf{T}_4}$ is a partition of $S$.
%By abuse of language, we refer to the three respective types of redexes as type-$1$, 2 and 3.
Hence it suffices to prove the following lemma.

\iffalse
For $i \in \{1, 2, 3\}$ and $n \in \omega$, define function $f_{i, n+1}(M) : S \to \Real$ by
\[
f_{i, n+1}(M)(s) :=
\begin{cases}
f(M_{n+1}(s)) & \hbox{if }s \in \mathbf{T}_i := 
\set{s \mid M_n(s) \textrm{ is type-$i$}}\\
0 & \textrm{otherwise}
\end{cases}
\]
\begin{lemma}
\label{lem:inde}
For each $i$ and $n$, $f_{i, n+1}(M)$ is %not just $\mathcal{F}_{n+1}$-measurable but also 
$\mathcal{F}_{n}$-measurable.
\end{lemma}

\begin{proof} First fix notation
\begin{align*}
\sigma(M_{n+1}) &= \sigma\big(\set{M_{n+1}^{-1}(U_{\beta_k})
\mid \beta_k\in \mathsf{Sk}_k, U_{\beta_k} \in \mathcal{B}(\Real^k), k \geq 0}\big)\\
\sigma(M_{n}) &= \sigma\big(\set{M_{n}^{-1}(U_{\alpha_j})
\mid \alpha_j\in \mathsf{Sk}_j, U_{\alpha_j} \in \mathcal{B}(\Real^j), j \geq 0}\big);
\end{align*}
and for $\beta_k \in \mathsf{Sk}_i$, we write $\beta_k[\dagger]$ to mean the instantiation of $\beta_k$ by a $k$-vector of numerals $\dagger$; $\alpha_j[\dagger']$ has the same meaning.
Now for each $\beta_k[\dagger] = E[R']$ where $R'$ is the contractum of a type-$i$ redex $R$, there exist $\alpha_j \in \mathsf{Sk}_j$ and $j$-vector $\dagger'$ such that $\alpha_j[\dagger'] = E[R]$ of type $i$.
Moreover, for each $U_{\beta_k}$, there exists $U_{\alpha_j}$ such that $M_{n+1}^{-1}(U_{\beta_k}) = M_{n}^{-1}(U_{\alpha_j}) \in \sigma(M_n)$.
\end{proof}

Plainly
\(
f(M_{n+1}) = %\sum_{i=1}^3 f_{i, n+1}(M).
f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M),
\)
$\mu$-almost-surely.
%(Unless otherwise stated, all equations and inequations between random variables are assumed to hold only $\mu$-a.s.)
Therefore 
\[
\expect{f(M_{n+1}) \mid \mathcal{F}_n} \nonumber \\
= 
\expect{f_{1, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{2, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{3, n+1}(M) \mid \mathcal{F}_n} 
\]
\begin{align}
& \expect{f(M_{n+1}) \mid \mathcal{F}_n} \nonumber \\
& = \expect{f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M)  \mid \mathcal{F}_n} 
\nonumber \\
& = \expect{f_{1, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{2, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{3, n+1}(M) \mid \mathcal{F}_n} 
\label{eqn:linear cond ex} 
& = f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M)
\label{eqn:inde cond exp}
& = f_{1, n+1}(M) + \int_I f() + f_{3, n+1}(M) \label{eqn:inde cond exp} \\
\end{align}
\Cref{eqn:linear cond ex} follows from the linearity of conditional expectation. 
%\Cref{eqn:inde cond exp} is justified because $f_{i, n+1}(M)$ is $\mathcal{F}_n$-measurable (\Cref{lem:inde}).

It remains to show: for all $A \in \calF_n$
\begin{align}
& \int_A \dif \mu \, \expect{f(M_{n+1}) \mid \mathcal{F}_n} \nonumber \\
& \leq  \int_A \mu (\dif s) \, \big(f(M_{n})[s \in \mathbf{T}_1] + f(M_{n})[s \in \mathbf{T}_2] + f(M_{n})[s \in \mathbf{T}_3]\big)
\label{eqn:ts} \\
& = \int_A \dif \mu \, f(M_n). 
\label{eqn:partition} 
\end{align}
where $[s \in \mathbf{T}_i]$ is the Iverson bracket.

Here we use the notation: given function $g : S \to \Real$ and $U \subseteq S$, $g[U] : S \to \Real$ denotes the function
\[
g[U](x) :=
\begin{cases}
g(x) & \textrm{$x \in U$}\\
0 & \textrm{otherwise}
\end{cases}
\]

The inequality (\ref{eqn:ts}) follows immediately from the lemma below; 
and (\ref{eqn:partition}) holds because $\{\mathbf{T}_1, \mathbf{T}_2, \mathbf{T}_3, \mathbf{T}_4\}$ is a partition of $S$.
\fi

\begin{lemma}
For all $i \in \set{1, 2, 3, 4}$ and $A \in \calF_n$
\[
\int_A \mu(\dif s) \, f_{n+1}(M)[s \in \mathbf{T}_i] \leq \int_A \mu(\dif s) \, f(M_n)[s \in \mathbf{T}_i].
\] 
where $[s \in \mathbf{T}_i]$ is the Iverson bracket.
\end{lemma}

\begin{proof}
We show the non-trivial case of $i = 2$.
First we express 
\begin{equation}
f_{n+1}(M)[s \in \mathbf{T}_2] = \sum_{i \in \calI} 
f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])[s \in U_i]
\label{eqn:f 2 n+1}
\end{equation}
where 
\begin{itemize}
\item $\calI$ is a countable indexing set
\item $E_i[\cdot][\tsample] \in \mathsf{Sk}_{j_i}$, and $\sigma_i : S \to \Real^{j_i}$, and $\rho(s) = \pi_h(\pi_1(\red^n(M, s))) \in \Real$ 
\item $\set{U_i}_{i \in \calI}$ is a partition of $\mathbf{T}_2$, 
where each $U_i$ is determined by a skeletal term $E_i[\cdot][\tsample]$ and a number (of draws) $l_i \leq n$;
precisely 
\[
U_i := \#\mathrm{draw}_n^{-1}(l_i) \cap M_n^{-1}(E_i[\cdot][\tsample], \Real^{j_i}) \in \calF_n.
\]
\lo{I.e.~$(E_i[\underline{\sigma_i(s)}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s)$ iff $s \in U_i$.}
\end{itemize}

Observe that if $s \in U_i$ then $\set{s_{\leq l_i}} \cdot I^\omega \subseteq U_i$;
in fact $(U_i)_{\leq l_i} \cdot I^\omega = U_i$.
%Moreover, for any $A \in \calF_n$, if $s \in A$ then $\set{s_{\leq l}} \cdot I^\omega \subseteq A$.
This means that for any measurable $g : S \to \Real_{\geq 0}$, if $g(s)$ only depends on the $(l_i+1)$-long prefix of $s$, then, writing $g' : I^{l_i+1} \to \Real_{\geq 0}$ where $g(s) = g'(s_{\leq l_i+1})$ 
\begin{equation}
\int_{A \cap U_i}  \mu(\dif s) \, g(s) = 
\int_{(A \cap U_i)_{\leq l_i+1}} \Leb_{l_i+1}(\dif t) \, g'(t)
\label{eqn:truncate}
\end{equation}
(For simplicity, we will write $g' = g$ in the following.)
%where $A_{\leq m} := \set{s_{\leq m} \mid s \in A}$.

Take $s \in U_i$, and set $l = l_i$.
Now $\sigma_i(s)$ depends on $s_{\leq l}$, and $\rho(s)$ depends on $s_{\leq l +1}$.
Let $u$ range over $(U_i)_{\leq l}$.
Then, it follows from the definition of $f$ that
\[
\int_I \Leb(\dif r) \, f(E_i[\underline{\sigma_i(u)}][\underline{r}])) \leq f(E_i[\underline{\sigma_i(u)}][\tsample])).
\]
Integrating both sides, we get
\begin{align*}
& \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, \int_I \textrm{Leb}(\dif r) \, f(E_i[\underline{\sigma_i(u)}][\underline{r}])) \\
\leq & \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, f(E_i[\underline{\sigma_i(u)}][\tsample]))
\end{align*}
Since $\Leb_{l+1}$ is the (unique) product measure satisfying $\Leb_{l+1}(V \times B) = \Leb_l(V) \cdot \Leb(B)$, and $(U_i)_{\leq l_i} \cdot I^\omega = U_i$, we have
%it follows from the preceding observation that
\begin{align*}
& \int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\underline{\sigma_i(u')}][\underline{\rho(u')}])) \\
& \leq 
\int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\underline{\sigma_i(u')}][\tsample]))
\end{align*}
and so we have
\begin{equation}
\int_{A \cap U_i} \mu(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])) \\
\leq 
\int_{A \cap U_i} \mu(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\tsample])).
\label{eqn:a u ui}
\end{equation}

Finally, integrating both sides of (\ref{eqn:f 2 n+1}), we have
\begin{align*}
\int_A \mu(\dif s) \, f_{n+1}(M)[s \in \mathbf{T}_i] 
&= 
\int_A \mu(\dif s) \, \sum_{i \in \calI} f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])[s \in U_i] \\
&= 
\sum_{i \in \calI} \int_{A \cap U_i} \mu(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])\\
&\leq 
\sum_{i \in \calI} \int_{A \cap U_i} \mu(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\tsample])
\quad \hbox{$\because$ (\ref{eqn:a u ui})}\\
&= 
\int_{A} \mu(\dif s) \, \sum_{i \in \calI} \, f(E_i[\underline{\sigma_i(s)}][\tsample])[s \in U_i]\\
&= \int_{A} \mu(\dif s) f_{n}(M)[s \in \mathbf{T}_2]
%\quad \hbox{$\because$ (\ref{eqn:f 2 n+1})}
\end{align*}

\end{proof}

\iffalse
[** To see $f_{2, n+1}(M) \leq f(M_n)[T_2]$, take $s \in T_2$. Then $f_{2, n+1}(M)(s) = f(E[\underline{a}])$ and $M_n = E[\tsample]$, for some $a \in [0, 1]$ and evaluation context $E$. Hence 
\[
f_{2, n+1}(M)(s) \leq \int_I f(E[\underline{r}]) \, \mu_{leb}(\textrm{d} r)
\leq f(E[\tsample]) = f(M_n)[T_2](s).
\]
**]
\fi
This concludes the proof of \Cref{thm:rankable gives supermartingale}. \hfill \qed

\medskip
As before fix a $M \in \Lambda^0$.
Define random variables on $(S, \calF, \mu)$:
\begin{align*}
T_0(s) & := 0 \\
T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value or of form } E[\tY (\lambda x. N)] \}\\
X_n(s) & := f(M_n(s)) \\
Y_n(s) & := X_{T_n}(s) \\
T(s) &:= \min \set{n \mid Y_n(s) = 0}
\end{align*}
\lo{Define the \emph{$\tY$-runtime of $M$} to be the random variable $T^{\tY}_M$: 
\begin{itemize}
\item if $T_M(s) = n < \infty$ then $T^{\tY}_M(s) :={}$ number of $\tY$-reduction steps in the $n$ steps of reduction to $\red^n(M, s)$
\item if $T_M(s) = \infty$ then $T^{\tY}_M(s) := \infty$.
\end{itemize}
Observe that $T = T^{\tY}_M$.}
Note that 
\begin{enumerate}
\item $T_M < \infty$ a.s.~iff $T_M^{\tY} < \infty$ a.s.~(because $T_M(s) < \infty$ iff $T_M^{\tY}(s) < \infty$)
\item $\expect{T_M} <{} \infty$ implies $\expect{T_M^{\tY}} < \infty$ (because $T_M^{\tY} \leq T_M$), but the converse is false (\Cref{ex:tY finite does not imply t finite}).
\end{enumerate}
%Note that by definition $T_M^{\tY} \leq T_M$.

Observe that $T_0, T_1, T_2, \cdots$ are an increasing sequence of stopping times adapted to $(\calF_n)_{n \in \omega}$, and each $T_i$ is bounded.
Thanks to \Cref{thm:rankable and strict rankable}, 
$(X_n)_{n \in \omega}$ is a supermartingale;
moreover, it follows from \Cref{thm:optional sampling} that $(Y_n)_{n \in \omega}$ is a ($1$-ranking) supermartingale.
Therefore, by \Cref{thm:rank-PAST}, $\expect{T^{\tY}_M} < \infty$.

\begin{corollary}
If a closed SPCF term $M$ is rankable, then $\expect{T^{\tY}_M} < \infty$, and so, $T^{\tY}_M < \infty$ a.s.~and $M$ is AST.

If a closed SPCF term $M$ is strictly rankable, then $\expect{T_M} < \infty$ i.e.~$M$ is PAST.
\end{corollary}

### 2020-05-09 ranking-supermartingales.tex
minor

@@
 and the filtration $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
-The $\calF_n$-measurability of $M_n$ (and hence $\#\mathrm{draw}_n$) follow from \citep{DBLP:conf/icfp/BorgstromLGS16};
+The $\calF_n$-measurability of $M_n$ (and hence of $\#\mathrm{draw}_n$) follows from \citep{DBLP:conf/icfp/BorgstromLGS16};
 $\#\mathrm{draw}_n$ is a stopping time (bounded by $n$).
@@
 \[
-\int_A \mu(\dif s) \, f_{n+1}(M)[s \in \mathbf{T}_i] \leq \int_A \mu(\dif s) \, f(M_n)[s \in \mathbf{T}_i].
+\int_A \mu(\dif s) \, f_{n+1}(M)[s \in \mathbf{T}_i] \leq \int_A \mu(\dif s) \, f(M_n)[s \in \mathbf{T}_i]
 \] 
@@
 \]
-Integrating both sides, we get
+Take $A \in \calF_n$, and integrating both sides, we get
 \begin{align*}
@@
 \end{align*}
-and so we have
+and so 
 \begin{equation}
@@
 \medskip
-As before fix a $M \in \Lambda^0$.
+As before, fix a $M \in \Lambda^0$.
 Define random variables on $(S, \calF, \mu)$:

### 2020-05-09 probabilisticProgrammingMartingales.tex
minor

@@
 
-\begin{theorem} \label{minimal}
+\begin{theorem} \label{thm:minimal}
 Given a closed term $M$, the function $f:Rch(M) \to \mathbb R$ given by $f(N) = \mathbb E [\text{the number of }\tY\text{-reduction steps from }N\text{ to a value}]$, if it exists, is the least of all possible ranking functions of $M$.
@@
 \paragraph{}
-Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and Theorem \ref{minimal} makes even stronger assumptions than almost sure termination, so it isn't useful for proving it.
+Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost sure termination, so it isn't useful for proving it.
 Take, for example, the term

### 2020-05-09 ranking-supermartingales.tex
minor

@@
 \lo{The preceding means: $\forall A \in \calF_n$, $\int_A \dif \mu \, Y_{n+1} \leq \int_A \dif \mu \,Y_n$.}
-For $\epsilon > 0$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \, \mathbf{1}_{\set{Y_n > 0}}$.
+For $\epsilon > 0$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{Y_n > 0}}$.
 \citep{DBLP:conf/popl/FioritiH15,DBLP:conf/popl/ChatterjeeFNH16}
@@
 \end{theorem}
+
+\begin{proof}[Proof Sketch]
+First establish by induction on $n$: 
+\[
+\forall n \in \omega \, . \, \expect{Y_n} \leq \expect{Y_0} - \epsilon \cdot \big(\sum_{i=0}^{n-1} \mathbb{P}[Y_i > 0]\big).
+\]
+It follows that $(\sum_{i=0}^{\infty} \mathbb{P}[Y_i > 0]\big)$ converges; hence $\lim_{n \to \infty} \mathbb{P}[Y_n > 0] = 0$ and so $\mathbb{P}[T < \infty] = 1$.
+Then observe: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=1} \mathbb{P}(T \geq i)$.
+\end{proof}
+
+\medskip
 
@@
 
-Observe that $T_0, T_1, T_2, \cdots$ are an increasing sequence of stopping times adapted to $(\calF_n)_{n \in \omega}$, and each $T_i$ is bounded.
+\begin{corollary}
+\begin{enumerate}
+\item If a closed SPCF term $M$ is rankable, then $T^{\tY}_M < \infty$ a.s.~(equivalently $M$ is AST), and $\expect{T^{\tY}_M} < \infty$.
+
+\item If a closed SPCF term $M$ is strictly rankable, then $\expect{T_M} < \infty$ i.e.~$M$ is PAST.
+\end{enumerate}
+\end{corollary}
+
+\begin{proof}
+For 1, observe that $T_0, T_1, T_2, \cdots$ are an increasing sequence of stopping times adapted to $(\calF_n)_{n \in \omega}$, and each $T_i$ is bounded.
 Thanks to \Cref{thm:rankable and strict rankable}, 
 $(X_n)_{n \in \omega}$ is a supermartingale;
-moreover, it follows from \Cref{thm:optional sampling} that $(Y_n)_{n \in \omega}$ is a ($1$-ranking) supermartingale.
-Therefore, by \Cref{thm:rank-PAST}, $\expect{T^{\tY}_M} < \infty$.
+it then follows from \Cref{thm:optional sampling} that $(Y_n)_{n \in \omega}$ is a ($1$-ranking) supermartingale.
+Therefore, by \Cref{thm:rank-PAST}, $T^{\tY}_M < \infty$ a.s.~and $\expect{T^{\tY}_M} < \infty$.
 
-\begin{corollary}
-If a closed SPCF term $M$ is rankable, then $\expect{T^{\tY}_M} < \infty$, and so, $T^{\tY}_M < \infty$ a.s.~and $M$ is AST.
+Statement 2 follows immediately from \Cref{thm:rank-PAST}.
+\end{proof}
 
-If a closed SPCF term $M$ is strictly rankable, then $\expect{T_M} < \infty$ i.e.~$M$ is PAST.
-\end{corollary}
+Thus the method of ranking function is sound for proving (positive) a.s.~termination of SPCF programs.
+It is in fact also complete in a sense: if $\expect{T^\tY_N} < \infty$ for all $N \in \mathit{Rch}(M)$ then $M$ is rankable (\Cref{thm:minimal}).

### 2020-05-10 probabilisticProgrammingMartingales.tex
Minor: added each f(M_n) is integrable.

@@
 \begin{equation}
-\tif{-\tsample < 0}{\underline{0}}{(\tY \lambda x. x)\underline 0}
+\tif{{-}\tsample < 0}{\underline{0}}{(\tY \lambda x. x) \; \underline 0}
 \end{equation}
@@
 \[
-M = (\tY \lambda f n. \tif{\tsample - 0.5 < 0}{n \underline 0}{f (\lambda x. n (n x))}) (\lambda x. x+1)
+M = \big(\tY \lambda f n. \tif{\tsample - 0.5 < 0}{n \, \underline 0}{f (\lambda x. n (n x))}\big) (\lambda x. x+1)
 \] 
@@
 The restrictions of $f$ and $g$ to $Rch(N)$ have the same properties assumed of $f$ and $g$, so assume w.l.o.g.~that $N=M$.
-The difference $g - f$ is then a supermartingale (with the same setup as in the proof of Theorem \ref{rankable implies ast}) therefore $\forall n.\ \mathbb E[g(M_n)] \leq \mathbb E[f(M_n)] + g(M)-f(M)$.
+The difference $g - f$ is then a supermartingale (with the same setup as in the proof of \Cref{thm:rankable gives supermartingale}
+%Theorem \ref{rankable implies ast}) 
+therefore $\forall n.\ \mathbb E[g(M_n)] \leq \mathbb E[f(M_n)] + g(M)-f(M)$.
 
@@
 
-\paragraph{}
+\subsection{Partial ranking functions}
 Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost sure termination, so it isn't useful for proving it.
@@
 \paragraph{}
-In both of these cases, there are only some values of the rankng function that are semantically important.
-Define a \emph{partial ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \nrightarrow \mathbb R$ such that
-\lo{NOTATION. Common notation for partial function $\rightharpoonup$}
+In both of these cases, there are only some values of the rankng function that are semantically important. Define a \emph{partial ranking function} on a closed term $M$ to be a partial function 
+$f : Rch(M) \rightharpoonup \mathbb R$ 
+%$f : Rch(M) \nrightarrow \mathbb R$ 
+such that
+%\lo{NOTATION. Common notation for partial function $\rightharpoonup$}
 \begin{itemize}
@@
 \begin{proof}
-Take a closed term $M$ and partial ranking function $f$ on $M$. Define $f_1 : Rch(M) \nrightarrow \mathbb R$ by
+Take a closed term $M$ and partial ranking function $f$ on $M$. Define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
 \begin{align*}

### 2020-05-10 ranking-supermartingales.tex
Minor: added each f(M_n) is integrable.

@@
 
-Let $S$ and $T$ be stopping times adapted to $(\calF_n)_{n \in \omega}$.
+Let $T$ and $T'$ be stopping times adapted to $(\calF_n)_{n \in \omega}$.
 Recall the $\sigma$-algebra (consisting of measurable subsets ``priori to $T$'')
@@
 \]
-and if $S \leq T$, then $\calF_S \subseteq \calF_{T}$.
+and if $T \leq T'$, then $\calF_{T} \subseteq \calF_{T'}$.
 
@@
 \[
-\forall n \in \omega \, . \, \expect{Y_n} \leq \expect{Y_0} - \epsilon \cdot \big(\sum_{i=0}^{n-1} \mathbb{P}[Y_i > 0]\big).
+\forall n \in \omega \, . \, \expect{Y_n} \leq \expect{Y_0} - \epsilon \cdot \Big(\sum_{i=0}^{n-1} \mathbb{P}[Y_i > 0]\Big).
 \]
-It follows that $(\sum_{i=0}^{\infty} \mathbb{P}[Y_i > 0]\big)$ converges; hence $\lim_{n \to \infty} \mathbb{P}[Y_n > 0] = 0$ and so $\mathbb{P}[T < \infty] = 1$.
-Then observe: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=1} \mathbb{P}(T \geq i)$.
+It follows that $\sum_{i=0}^{\infty} \mathbb{P}[Y_i > 0]$ converges; hence $\lim_{n \to \infty} \mathbb{P}[Y_n > 0] = 0$ and so $\mathbb{P}[T < \infty] = 1$.
+Then it suffices to prove: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=1} \mathbb{P}[T \geq i]$.
 \end{proof}
@@
 %Take $M \in \Lambda^0$ (closed SPCF terms). Define, for each $n \in \omega$, the random variable $M_n : S \to \Lambda^0$ by $M_n(s) := \pi_0 (\red^n(M, s))$.
+
+First each $f(M_n)$ is integrable. Since $0 \leq f(N) \leq f(M)$ for all $N \in \mathit{Rch}(M)$, we have
+\[
+\int_{S} \mu(\dif s) \, |f(M_n)(s)| 
+=
+\int_{S} \mu(\dif s) \, f(M_n)(s)
+\leq
+\int_{S} \mu(\dif s) \, f(M)
+=
+f(M)
+\]
 
@@
 
-Thus the method of ranking function is sound for proving (positive) a.s.~termination of SPCF programs.
+Thus the method of (strict) ranking function is sound for proving (positive) a.s.~termination of SPCF programs.
 It is in fact also complete in a sense: if $\expect{T^\tY_N} < \infty$ for all $N \in \mathit{Rch}(M)$ then $M$ is rankable (\Cref{thm:minimal}).

### 2020-05-11 probabilisticProgrammingMartingales.tex

--- 2020-05-09 probabilisticProgrammingMartingales.tex
@@
 
-%\usepackage[bbgreekl]{mathbbol} %for \bblambda
 %%
@@
 
-\usepackage{amsthm}
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}
-\newtheorem{example}{Example}
-
-\theoremstyle{lemma}
-\newtheorem{lemma}{Lemma}
 \newtheorem{theorem}{Theorem}
 \newtheorem{corollary}{Corollary}
+\newtheorem{lemma}{Lemma}
+\newtheorem{definition}{Definition}
 
-\theoremstyle{remark}
-\newtheorem*{remark}{Remark}
 \begin{document}
@@
 \end{align*}
-\changed[lo]{We write $\to^\ast$ for the reflexive, transitive closure of $\to$.}
 \lo{The $\tscore$ rule should have a side condition: if $r > 0$.}
@@
 To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples.
-Let $ I = [0,1] \subset \mathbb{R} $, and let $S = I^{\mathbb{N}}$, with the \changed[lo]{Borel $\sigma$-algebra $\calF$} and the probability measure, denoted $\mu$, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
-Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and $\mu (\prod_{i=0}^\infty X_i) = \prod_{i=0}^\infty \Leb(X_i)$.
+Let $ I = [0,1] \subset \mathbb{R} $, and let $S = I^{\mathbb{N}}$, with the Borel $\sigma$-algebra and the probability measure, denoted $\mu$, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
+Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and $\mu (\prod_{i=0}^\infty X_i) = \prod_{i=0}^\infty \mu_{leb}(X_i)$.
 The maps $\pi_h:S \to I, \; \pi_t:S \to S$ popping the first element are then measurable.
-\changed[lo]{Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(S, \calF, \mu)$ the \emph{entropy space}.}
 
@@
 A term $M$ is said to terminate almost surely if $\mu(\{s \mid M \text{ terminates at } s\}) = 1$.
-
-\lo{Alternatively, define the \emph{runtime of $M$} to be the random variable 
-\[
-T_M(s) := 
-\begin{cases}
-\min \set{n \mid \pi_0(\red^n(M, s)) \textrm{ is a value}} & \hbox{if $\red^\infty(M,s)$ is defined}\\
-\infty & \hbox{otherwise}
-\end{cases}
-\]
-Equivalently, we say that $M$ is \emph{almost-surely terminating} (AST) if $T_M < \infty$ a.s.; 
-and $M$ is \emph{positively almost-surely terminating} (PAST) if $\expect{T_M} < \infty$.}
 
@@
 
-Given a term $M$, let $L(M)$ be the set of occurrences in $M$ of variables, $\tsample$ and $\tY$, and let $F(M)$ be the free magma on $L(M)$.
-\akr{Only the elements with no \tsample{}s except in the rightmost position are used.
-I'm not sure whether making this explicit in the definition of $F(M)$ makes things more or less clear.} Let $I = [0,1]$ and $S = I^{F(M)}$, with the Borel $\sigma$-algebra and the product measure.
+When proving almost sure termination in this way /akr{I guess I should move this section somewhere later.}, it is necessary to consider which terms a given term may reduce to.
+Sometimes however, the reduction that the programmer has in mind may not be strictly the call-by-value order defined so far, or considering an alternative reduction order may be simpler or more intuitive.
 
-For any reduction \akr{including reductions not included in the definition above, i.e.~out of order reductions} $M \to N$, define a corresponding map $F(M \to N) : F(N) \to F(M)$ induced (as a magma morphism) by the obvious injection along with the replacements
+Non-probabilistis lambda calculi generally have the Church-Rosser property, that if a term $A$ reduces to both $B_1$ and $B_2$, there is some $C$ with reduction sequences $B_1 \to^* C$ and $B_2 \to^* C$, so the reduction order mostly doesn't matter.
+In the probabilistic case, this may not be true, because $\beta$-reduction can duplicate $\tsample$s, so the outputs of the copies of the sample may be identical or independent, depending on whether the sample is taken before or after $\beta$-reduction.
+There are, however, some restricted variations on the reduction order that do not have this problem.
+
+\paragraph{}
+Even with this restriction, a sampling semantics in the style of the one already defined would not be entirely Church-Rosser, as, for example, $red^3(\tsample - \tsample, (1,0,\dots))$ would be either $1$ or $-1$ depending on the order of evaluation of the $\tsample$s, as that determines which sample from the pre-selected sequence is used for each one.
+To fix this, rather than pre-selecting samples according to the order they'll be drawn in, select them according to the position in the term where they'll be used instead.
+
+A position \akr{I'm not really sure how to phrase the citation, but this is adapded from \cite{Kennaway96infinitarylambda}.} is a finite sequence of steps into a term, defined inductively as
 \begin{align*}
-F(E[(\lambda x. F[x_i])G[y]] \to E[F(G[y]/x)[G[y_i]]]) y_i & = (x_i \cdot y)\\
-F(E[\tY_i F[y]] \to E[F[y] (\tY F[y_0])]) (y_0) & = \tY_i \cdot y
+P ::= \cdot \mid \lambda ; P \mid @_1 ; P \mid @_2 ; P \mid \underline f_i ; P \mid \tY ; P \mid \textsf{if}_1 ; P \mid \textsf{if}_2 ; P \mid \textsf{if}_3 ; P \mid \tscore ; P.
 \end{align*}
-where $y$ can be any of the things in $L(N)$.
+The subterm at a position is defined as
+\begin{align*}
+M \mid \cdot & = M \\
+\lambda x. M \mid \lambda ; P & = M \mid P \\
+M_1 M_2 \mid @_i ; P & = M_i \mid P \quad \text{for } i = 1,2 \\
+\underline f(M_1,\dots,M_n) \mid \underline f_i ; P & = M_i \mid P \quad \text{for }i \leq n \\
+\tY M \mid \tY ; P & = M \mid P \\
+\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; P & = M_i \mid P \quad \text{for } i = 1,2,3 \\
+\tscore(M) \mid \tscore ; P & = M \mid P
+\end{align*}
+so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g.
+$xy \mid \lambda ; \cdot$ is undefined).
+A position such that $M\mid P$ does exist is said to occur in $M$.
+Substitution at a given position, $M[N/P]$, is defined similarly.
 
-The one-step reduction is given by the function $\red : \Lambda^0 \times S \to \Lambda^0 \times S$ where
-\begin{equation}
-\red(M,s) = \left\{
-    \begin{array}{ll}
-        (E[N],s \circ F(E[R] \to E[N])) & \text{if } M = E[R], R \to N \text{ and } R \neq \tsample \\
-        (E[\underline{s(\tsample_i)}],s\mid_{F(E[\underline{s(\tsample_i)}])}) & \text{if } M = E[\tsample_i] \\
-        (M,s) & \text{if } M \text{ is a value}
-    \end{array} \right .
-\end{equation}
+Two subterms $N_1$ and $N_2$ of a term $M$, corresponding to positions $P_1$ and $P_2$, can overlap in a few different ways.
+If $P_1$ is an initial segment of $P_2$, then $N_2$ is also a subterm of $N_1$, which is written as $P_1 \leq P_2$.
+If neither $P_1 \leq P_2$ nor $P_1 \geq P_2$, the positions are said to be disjoint.
+This is mostly relevant in that in any substitution at some position, the subterms at all other positions are unaffected.
 
-Essentially the entropy value here contains a pre-selected random value for each possible \tsample-reduction, in a way independent of reduction order, taking into account the possible duplication of \tsample{}s in $\beta$- and \tY-reduction steps.
+With this notation, a more general reduction relation can be defined. \akr{It would almost work to have one of the neater versions of $\tY$-reduction here.}
+\begin{align*}
+  \text{if } M \mid P = (\lambda x.s) v,\ & M \to M[s[v/x]/P] \\
+  \text{if } M \mid P = \underline f (\underline r_1, \dots , \underline r_n),\ & M \to M[\underline{f(r_1,\dots,r_n)}/P] \\
+  \text{if } M \mid P = \tY \lambda x. s,\ & M \to M[\lambda z. s[(\tY \lambda x. s)/x] z/P] \text{ where $z$ is not free in $s$}\\
+  \text{if } M \mid P = \tif{\underline r < 0}{s_1}{s_2},\ & M \to M[s_1/P] \text{ where }r < 0 \\
+  \text{if } M \mid P = \tif{\underline r < 0}{s_1}{s_2},\ & M \to M[s_2/P] \text{ where }r \geq 0 \\
+  \text{if } M \mid P = \tsample & \text{ and $\lambda$ does not occur after $@_2$ or $\tY$ in $P$},\\ & M \to M[\underline r/P] \text{ where } r \in [0,1] \\
+  \text{if } M \mid P = \tscore(\underline r),\ & M \to M[\underline r/P].
+\end{align*}
+
+\paragraph{}
+Labelling the pre-chosen samples by the positions in the term would also not work because in some cases, a $\tsample$ will be duplicated before being reduced, for example, in $(\lambda x.
+\underline x 0 \underline + x \underline 0) (\lambda y.
+\tsample)$, both of the $\tsample$ redexes that eventually occur originate at $@_2 ; \lambda$.
+It is therefore necessary to consider possible positions that may occur in other terms reachable from the original term.
+This is itself inadequate because some of the positions in different reachable terms need to be considered the same.
+
+Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a reduction sequence starting at $M$, and the second of which is a position in the final term of the reduction sequence.
+Reduction sequences are used rather than reachable terms because if the same term is reached twice, different samples may be needed, but they will be discussed as though they were terms, by abuse of notation.
+The relation $\sim$ is defined as the union of the symmetric relations $\sim_p$ and $\sim_c$ where
+\begin{enumerate}
+    \item If $N$ reduces to $O$ with the redex at position $P$, and $Q$ is a position in $N$ where $Q \not \geq P$, then $(N,Q) \sim_p (O,Q)$.
+    \item If $N$ $\beta$-reduces to $O$ at position $P$, $Q$ is a position in $N | P;@_1;\lambda$ and $N \mid P;@_1;\lambda;Q$ is not the variable involved in the reduction, $(N,P;@_1;\lambda;Q) \sim_p (O, P;Q)$
+    \item If $N$ $\textsf{if}$-reduces to $O$ at position $P$, with the first resp.
+second branch being taken, and $P;\textsf{if}_i;Q$ occurs in $N$ (where $i = 2$ resp.
+$3$), $(N,P;\textsf{if}_i;Q) \sim_p (O,P;Q)$
+    \item If $N \mid P = (\lambda x.
+A) B$, $A \mid Q = x$, $B \mid R = C$ is a redex reducing to $C'$ and $S$ exists in $C'$, then letting $O_1$ be $N$ reduced first at $P$ then at $P;Q;R$ and letting $O_2$ be $N$ reduced first at $P;@_2;R$ then at $P$, $(O_1,P;Q;R;S) \sim_c (O_2,P;Q;R;S)$
+    \item If $N \mid P = (\tY \lambda x.
+A)$, $A \mid Q = C$ is a redex reducing to $C'$, $R$ exists in $C'$ and $C' \mid R \neq x$, then letting $O_1$ be $N$ reduced first at $P$ then $P;\lambda;@_1;Q$ and $O_2$ reduced at $P;\tY;\lambda;Q$ then $P$, $(O_1, P;\lambda;@_1;Q;R) \sim_c (O_2, P;\lambda;@_1;Q;R)$
+    \item If $N \mid P = (\tY \lambda x.
+A)$, $A \mid Q = C$ is a redex reducing to $C'$, $R$ exists in $C'$ and $A \mid S = x$, then letting $O_1$ be $N$ reduced first at $P$ then $P;\lambda;@_1;S;\tY;\lambda;Q$ and $O_2$ reduced at $P;\tY;\lambda;Q$ then $P$, $(O_1, P;\lambda;@_1;S;\tY;\lambda;Q;R) \sim_c (O_2, P;\lambda;@_1;S;\tY;\lambda;Q;R)$
+\end{enumerate}
+
+The reflexive transitive closure $\sim^*$ of this relation is used to define the set of potential positions $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable terms.
+If $(N,P) \sim^* (O,Q)$, then $N \mid P$ and $O \mid Q$ both have the same shape (i.e.~they're either both variables, both applications, both $\tsample$s ect.), therefore it's well-defined to talk of the set of potential positions where there is a $\tsample$, $L_s(M)$.
+The new sample space is then defined as $I^{L_s(M)}$, with the Borel $\sigma$-algebra and product measure.
+
+\paragraph{}
+Before defining the new version of the reduction relation $\red$, a few lemmas about the properties of $\sim$ are necessary for it to be well-defined.
+Considering the set of paths through $Rch(M)$ (which we're implicitly ideitifying with $Rch(M)$ itself) as a tree rooted at (the trivial path at) $M$, $\sim_p$ only relates positions in child and parent terms, and $\sim_c$ only relates first cousins.
+Furthermore, by a case analysis on the forms of positions, the possible paths $N_1 \sim N_2 \sim \dots \sim N_n$ can be strongly restricted.
+
+In any case, a position can be related to at most one position in each child term.
+Letting $P$ be the position of the redex and $Q$ the original term, this is by rule 1 if $P$ and $Q$ are disjoint, rule 2 if $P;@_1 \leq Q$ and rule 3 if $P;if_i \leq Q$.
+As $M$ itself has no cousins or parents, this is the only way positions in $M$ are related to others.
+
+For any other term $N$ and position $Q$ in it, take the parent $N'$ and the position $P$ of the reduction $N' \to N$.
+The positions that $(N,Q)$ are related to other than those in the children of $N$ are as follows.
+If $Q \not \geq P$, only case 1 applies.
+If $Q \geq P$ and $(N,Q) \sim_p (N',Q')$ by cases 2 or 3, $Q'$ is also $\geq P$ therefore if $(N,Q)$ is also related to something by $\sim_c$, $(N',Q')$ is not related to anything in its parent.
+Furthermore, letting $Q = P;R$, $Q' = P;@_1;\lambda;R$ or $P;\textsf{if}_i;R$ depending on what sort of reduction $N' \to N$ was ($\beta$, $\textsf{if}$-true or $\textsf{if}$-false respectively) therefore in any case, $(N,Q)$ is related to at most one position in $N'$.
+
+Considering now $\sim_c$, let $N'' \to N' \to N$, where the outermost of the two reductions takes place at $P$.
+Cases 5 and 6 can be distinguished from case 4 by whether $P;\tY$ exists in $N''$, and cases 5 and 6 can be distinguished by whether $N'' | R = x$ for some $R < P;\tY;\lambda;S$ where $x$ is the variable bound at $N'' | P;\tY$ and $Q = P;\lambda;@_1;S$.
+In all of the cases 4, 5 and 6, whether $Q$ is $O_1$ or $O_2$ can be distinguished by which of the reductions $N'' \to N'$, $N' \to N$ takes place at $P$.
+In all 6 cases, the term and position $(N,Q)$ is related to can be reconstructed from $Q$ and the positions of the reductions $N'' \to N' \to N$, therefore each $(N,Q)$ pair is related to at most one other by $\sim_c$.
+
+In summary, all equivalence classes under $\sim^*$ are of the shapes \akr{This would be far easier to show with a few pictures.
+I just don't yet know how to make them in Latex and I'm prioritising getting the text done.}.
+
+\paragraph{}
+
+At each reduction step $M \to N$, the sample space must be restricted to $I^{L_s(N)}$, which requires a map $L_s(N) \to L_s(M)$.
+The injection $L_0(N) \to L_0(M)$ is trivial to define by appending $M \to N$ to each path, but in order for this to extend to the quotient, the following lemma is needed.
+\begin{lemma}
+If $M \to N$, the relation $\sim_N^*$ defined on $L_0(N)$ is equal to the restriction of $\sim_M^*$ from $L_0(M)$ to $L_0(N)$.
+\end{lemma}
+\begin{proof}
+For any equivalence class under $\sim_M^*$ that includes some descendant of $N$, consider the most recent common ancestor of the terms in the equivalence class $O$.
+If $O \in Rch(N)$, so is that whole equivalence class therefore $\sim_M^*$ and $\sim_N^*$ match there.
+If $O \not \in Rch(N)$ but some of its descendants are, it must be the case that $O = M$.
+If $M$ itself has a position in the class, it it a line of $\sim_p$s, and the intersection of it with $Rch(N)$ is an equivalence class of $\sim_N^*$.
+Otherwise, there are two children $O_1, O_2$ of $M$ such that the equivalence class lies in $Rch(O_1) \cup Rch(O_2)$ and at least one of $O_1$ and $O_2$ is $N$.
+In either case, the restriction of the equivalence class to $Rch(N)$ is an equivalence class of $\sim_N^*$ therefore, in total, the restriction of $\sim_M^*$ to $Rch(N)$ is a subset of $\sim_N^*$.
+
+The converse, that $\sim_N^*$ is a subset of $\sim_M^*$ follows directly from the fact that all of the rules in the definition of $\sim$ apply at least as much in $Rch(M)$ as in $Rch(N)$ therefore $\sim_N$ is a subset of $\sim_M$.
+\end{proof}
+The injection $L_s(N) \to L_s(M)$ thus defined is denoted $i(N \to M)$.
+
+Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function in this case, so it is denoted ``$\Rightarrow$'' instead of ``$\red$'', and it relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
+\begin{align*}
+& (M,s) \Rightarrow (N,s \circ i(M \to N)) \text{ if either} \\
+& \qquad \text{$M \to N$ and the redex is not $\tsample$, or} \\
+& \qquad \text{$M \mid P = \tsample$, $N = M[\underline{s(M,P)}/P]$ and $\lambda$ does not occur after $@_2$ of $\tY$ in $P$}
+\end{align*}
 
@@
 
-Let %$Rch(M) := \{x \in \Lambda \mid \exists (y_i) : M \to y_0 \to \dots \to y_n \to x\}$, 
-\changed[lo]{$Rch(M) := \{N \in \Lambda \mid M \to^\ast N\}$},
-with the $\sigma$-algebra induced as a subset of $\Lambda$.
+Let $Rch(M) := \{x \in \Lambda \mid \exists (y_i) : M \to y_0 \to \dots \to y_n \to x\}$, with the $\sigma$-algebra induced as a subset of $\Lambda$.
 
 \begin{definition}\rm
-A \emph{ranking function on $M$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
-\begin{enumerate}
+A \emph{ranking function on $M$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that
+\begin{itemize}
+    \item $f(N) \geq 0$ for all $N$
     \item $f(E[\tY \lambda x. N]) \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z]) \text{ where $z$ is not free in $N$}$
-    \item $f(E[\tsample]) \geq \int_I f(E[\underline{x}]) \, \Leb(\mathrm{d}x)$
+    \item $f(E[\tsample]) \geq \int_I f(E[\underline{x}]) \, \mu_{leb}(\mathrm{d}x)$
     \akr{I'm not sure whether $Rch(M)$ is a measurable subset of $\Lambda$, so it might not have a natural measure, but I'm pretty sure this doesn't actually matter, and the $\sigma$-algebra at least is well-defined.
@@
     \item $f(E[R]) \geq f(E[R'])$ for any other redex $R$, where \changed[lo]{$R \to R'$}. \lo{Metavariables: $M$ and $N$ for terms and programs; $R$ / $R'$ for redex / contractum.}
-\end{enumerate}
-We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that $f(N) = 0$ iff $N$ is a value, and for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.
-
-Any closed term for which a ranking (respectively, strict ranking) function exists is called \emph{rankable} (respectively, \emph{strictly rankable}).
+\end{itemize}
+Any closed term for which a ranking function exists is called ``rankable''.
 \end{definition}
@@
 \section{Constructing Ranking Functions}
-Although rankability implies almost-sure termination, the converse does not hold in general. For example,
+Although rankability implies almost sure termination, the converse does not hold in general. For example,
 \begin{equation}
@@
 In particular, the expected number of $\tY$-reduction steps from each reachable term is a ranking function.
-Note that a finite number of expected $\tY$-reduction steps does not necessarily imply a finite number of expected total reduction steps.
-
-\begin{example}
-\label{ex:tY finite does not imply t finite}
-The term
-\[
-M = (\tY \lambda f n. \tif{\tsample - 0.5 < 0}{n \underline 0}{f (\lambda x. n (n x))}) (\lambda x. x+1)
-\] 
-terminates with only 2 $\tY$-reductions on average i.e.~$\expect{T_M^\tY} = 2$, but applies the increment function $2^n$ times with probability $2^{-1-n}$ for $n \geq 0$, which diverges i.e.~$\expect{T_M} = \infty$.
-\end{example}
+Note that a finite number of expected $\tY$-reduction steps does not necessarily imply a finite number of expected total reduction steps, for example $(\tY \lambda f n.
+\tif{\tsample - 0.5 < 0}{n \underline 0}{f (\lambda x.
+n (n x))}) (\lambda x.
+x+1)$ terminates with only 2 $\tY$-reductions on average, but applies the increment function $2^n$ times with probability $2^{-1-n}$ for $n \geq 0$, which diverges.
 

### 2020-05-11 ranking-supermartingales.tex

--- 2020-05-09 ranking-supermartingales.tex
@@
 
-A main result is the following theorem.
-\begin{theorem} 
-\label{thm:rankable and strict rankable}
-If a closed SPCF term is rankable (respectively, strictly rankable) by $f$, then $(f(M_n))_{n \in \omega}$ is a 
-supermartingale (respectively, ranking supermartingale) adapted to the filtration $(\mathcal{F}_n)_{n \in \omega}$ where $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
-\end{theorem}
-
-\paragraph{Proof of \Cref{thm:rankable and strict rankable}.}
-\emph{Notation}. Given an $\omega$-sequence (e.g.~$s \in S$) and $m \in \omega$, we write $s_{\leq m} \in I^m$ to mean the $m$-long prefix of $s$.
+\paragraph{Proof of \Cref{thm:rankable gives supermartingale}.}
 
@@
 M_n(s) &:= \pi_0 (\red^n(M, s))\\
-\ndraw{n}{s} &:= l \textrm{ where $\pi_1 (\red^n(M, s)) = \underbrace{\pi_t( \cdots (\pi_t}_l(s))$}
+\#\textrm{draws}_n(s) &:= l \textrm{ where $\pi_1 (\red^n(M, s)) = \underbrace{\pi_t( \cdots (\pi_t}_l(s))$}\\
+T_0(s) & := 0 \\
+T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value or of form } E[\tY (\lambda x. N)] \}\\
+X_n(s) & := f(M_n(s)) \\
+Y_n(s) & := X_{T_n}(s)
 \end{align*}
-and the filtration $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
-The $\calF_n$-measurability of $M_n$ (and hence of $\#\mathrm{draw}_n$) follows from \citep{DBLP:conf/icfp/BorgstromLGS16};
-$\#\mathrm{draw}_n$ is a stopping time (bounded by $n$).
-Take $s \in A \in \calF_n$ with $\ndraw{n}{s} = l$.
-For any $s'\in S$, if $s_{\leq l} = s_{\leq l}'$ then $s' \in A$.
-It follows that ${\set{s_{\leq l}} \cdot I^\omega} \subseteq A$.
-\lo{NOTE. To fix notation:
-\(
-\sigma(M_{n}) = \sigma\big(\set{M_{n}^{-1}(\alpha_j, U_{\alpha_j})
-\mid \alpha_j\in \mathsf{Sk}_j, U_{\alpha_j} \in \mathcal{B}(\Real^j), j \geq 0}\big)
-\)
-}
+and define a filtration $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
+
 %Take $M \in \Lambda^0$ (closed SPCF terms). Define, for each $n \in \omega$, the random variable $M_n : S \to \Lambda^0$ by $M_n(s) := \pi_0 (\red^n(M, s))$.
@@
 We say that a given SPCF term is \emph{type-1} (respectively 2, 3 and 4) if it has the shape $E[\tY (\lambda x. N)]$ (respectively $E[\tsample]$, $E[R]$ where $R$ is any other redex, and of a value).
-Define $\mathbf{T}_i := 
-\set{s \mid M_n(s) \hbox{ is type-$i$}}$.
-It is straightforward to see that each $\mathbf{T}_i \in \calF_n$, and $\set{\mathbf{T}_1, \mathbf{T}_2, \mathbf{T}_3, \mathbf{T}_4}$ is a partition of $S$.
 %By abuse of language, we refer to the three respective types of redexes as type-$1$, 2 and 3.
-Hence it suffices to prove the following lemma.
 
-\iffalse
+%Henceforth fix $M \in \Lambda^0$.
 For $i \in \{1, 2, 3\}$ and $n \in \omega$, define function $f_{i, n+1}(M) : S \to \Real$ by
@@
 \]
+\iffalse
 \begin{lemma}
@@
 \end{proof}
+\fi
 
@@
 \]
+\iffalse
 \begin{align}
@@
 %\Cref{eqn:inde cond exp} is justified because $f_{i, n+1}(M)$ is $\mathcal{F}_n$-measurable (\Cref{lem:inde}).
+\fi
 
@@
 
+\iffalse
 Here we use the notation: given function $g : S \to \Real$ and $U \subseteq S$, $g[U] : S \to \Real$ denotes the function
@@
 \]
+\fi
 
@@
 and (\ref{eqn:partition}) holds because $\{\mathbf{T}_1, \mathbf{T}_2, \mathbf{T}_3, \mathbf{T}_4\}$ is a partition of $S$.
-\fi
 
 \begin{lemma}
-For all $i \in \set{1, 2, 3, 4}$ and $A \in \calF_n$
+For all $i \in \set{1, 2, 3}$ and $A \in \calF_n$
 \[
-\int_A \mu(\dif s) \, f_{n+1}(M)[s \in \mathbf{T}_i] \leq \int_A \mu(\dif s) \, f(M_n)[s \in \mathbf{T}_i]
+\int_A \dif \mu \, f_{i, n+1}(M) \leq \int_A \mu(\dif s) \, f(M_n)[s \in \mathbf{T}_i].
 \] 
-where $[s \in \mathbf{T}_i]$ is the Iverson bracket.
 \end{lemma}
@@
 \begin{equation}
-f_{n+1}(M)[s \in \mathbf{T}_2] = \sum_{i \in \calI} 
-f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])[s \in U_i]
+f_{2, n+1}(M) = \sum_{i \in \calI} 
+\lambda s . f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])[s \in U_i]
 \label{eqn:f 2 n+1}
@@
 \item $\calI$ is a countable indexing set
-\item $E_i[\cdot][\tsample] \in \mathsf{Sk}_{j_i}$, and $\sigma_i : S \to \Real^{j_i}$, and $\rho(s) = \pi_h(\pi_1(\red^n(M, s))) \in \Real$ 
-\item $\set{U_i}_{i \in \calI}$ is a partition of $\mathbf{T}_2$, 
-where each $U_i$ is determined by a skeletal term $E_i[\cdot][\tsample]$ and a number (of draws) $l_i \leq n$;
-precisely 
-\[
-U_i := \#\mathrm{draw}_n^{-1}(l_i) \cap M_n^{-1}(E_i[\cdot][\tsample], \Real^{j_i}) \in \calF_n.
-\]
-\lo{I.e.~$(E_i[\underline{\sigma_i(s)}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s)$ iff $s \in U_i$.}
+\item $E_i \in \bigcup_{j \in \omega}\mathsf{Sk}_j$, and $\sigma_i : S \to \bigcup_{j \in \omega}\Real^j$ is a random variable
+\item $\rho(s) = \pi_h(\pi_1(\red^n(M, s))) \in \Real$ 
+\item $\set{U_i}_{i \in \calI}$ is a partition of $\mathbf{T}_2$ 
+\item Each $U_i$ is determined by a skeletal term, namely, $E_i$, and the number of draws (i.e.~$\ndraw{n}{s}$) consumed in $n$ steps of computation of $M$, in the sense that for all $s, s' \in U_i$, 
+%$\pi_0(\red^n(M, s))$ and $\pi_0(\red^n(M, s))$ 
+$M_n(s)$ and $M_n(s')$ have the same underlying skeleton $E_i$ and $\ndraw{n}{s} = \ndraw{n}{s'}$.
+Notice that $E_i[\underline{\sigma_i(s)}][{\tsample}] = M_{n}(s)$ for $s \in U_i$.
 \end{itemize}
 
-Observe that if $s \in U_i$ then $\set{s_{\leq l_i}} \cdot I^\omega \subseteq U_i$;
-in fact $(U_i)_{\leq l_i} \cdot I^\omega = U_i$.
-%Moreover, for any $A \in \calF_n$, if $s \in A$ then $\set{s_{\leq l}} \cdot I^\omega \subseteq A$.
-This means that for any measurable $g : S \to \Real_{\geq 0}$, if $g(s)$ only depends on the $(l_i+1)$-long prefix of $s$, then, writing $g' : I^{l_i+1} \to \Real_{\geq 0}$ where $g(s) = g'(s_{\leq l_i+1})$
+\emph{Notation}. Given an $\omega$-sequence (e.g.~$s \in S$) and $m \in \omega$, we write $s_{\leq m}$ to mean the $m$-long prefix of $s$.
+
+Observe that if $s \in U_i$ and $\ndraw{n}{s} = l$ then $s_{\leq l} = s_{\leq l}'$ implies $M_n(s) = M_n(s')$.
+It follows that $\set{s_{\leq l}} \cdot I^\omega \subseteq U_i$ (where $\cdot$ means concatenation of sequences).
+Moreover, for any $A \in \calF_n$, if $s \in A$ then $\set{s_{\leq l}} \cdot I^\omega \subseteq A$.
+This means that for any measurable $g$, if $g(s)$ only depends on the $(l+1)$-long prefix of $s$ (i.e.~$g(s) = g'(s_{\leq l+1})$), then 
 \begin{equation}
 \int_{A \cap U_i}  \mu(\dif s) \, g(s) = 
-\int_{(A \cap U_i)_{\leq l_i+1}} \Leb_{l_i+1}(\dif t) \, g'(t)
+\int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif t) \, g'(t)
 \label{eqn:truncate}
 \end{equation}
-(For simplicity, we will write $g' = g$ in the following.)
-%where $A_{\leq m} := \set{s_{\leq m} \mid s \in A}$.
+where $A_{m} := \set{s_{\leq m} \mid s \in A}$, and $\Leb_n$ is the Lebesgue measure on $I^n$.
 
-Take $s \in U_i$, and set $l = l_i$.
-Now $\sigma_i(s)$ depends on $s_{\leq l}$, and $\rho(s)$ depends on $s_{\leq l +1}$.
+Take $s \in U_i$ with $\ndraw{n}{s} = l$.
+Now $\sigma_i(s)$ depends on $s_{\leq l}$, and $\rho(s)$ depends on $s_{\leq l+1}$.
 Let $u$ range over $(U_i)_{\leq l}$.
@@
 \]
-Take $A \in \calF_n$, and integrating both sides, we get
+Integrating both sides, we get
 \begin{align*}
@@
 \end{align*}
-Since $\Leb_{l+1}$ is the (unique) product measure satisfying $\Leb_{l+1}(V \times B) = \Leb_l(V) \cdot \Leb(B)$, and $(U_i)_{\leq l_i} \cdot I^\omega = U_i$, we have
-%it follows from the preceding observation that
+It follows from the preceding observation that
 \begin{align*}
@@
 \end{align*}
-and so 
+and so we have
 \begin{equation}
@@
 \begin{align*}
-\int_A \mu(\dif s) \, f_{n+1}(M)[s \in \mathbf{T}_i] 
+\int_A \mu(\dif s) \, f_{2, n+1}(M)(s) 
 &= 
@@
 \medskip
-As before, fix a $M \in \Lambda^0$.
-Define random variables on $(S, \calF, \mu)$:
-\begin{align*}
-T_0(s) & := 0 \\
-T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value or of form } E[\tY (\lambda x. N)] \}\\
-X_n(s) & := f(M_n(s)) \\
-Y_n(s) & := X_{T_n}(s) \\
-T(s) &:= \min \set{n \mid Y_n(s) = 0}
-\end{align*}
-\lo{Define the \emph{$\tY$-runtime of $M$} to be the random variable $T^{\tY}_M$: 
-\begin{itemize}
-\item if $T_M(s) = n < \infty$ then $T^{\tY}_M(s) :={}$ number of $\tY$-reduction steps in the $n$ steps of reduction to $\red^n(M, s)$
-\item if $T_M(s) = \infty$ then $T^{\tY}_M(s) := \infty$.
-\end{itemize}
-Observe that $T = T^{\tY}_M$.}
-Note that 
-\begin{enumerate}
-\item $T_M < \infty$ a.s.~iff $T_M^{\tY} < \infty$ a.s.~(because $T_M(s) < \infty$ iff $T_M^{\tY}(s) < \infty$)
-\item $\expect{T_M} <{} \infty$ implies $\expect{T_M^{\tY}} < \infty$ (because $T_M^{\tY} \leq T_M$), but the converse is false (\Cref{ex:tY finite does not imply t finite}).
-\end{enumerate}
-%Note that by definition $T_M^{\tY} \leq T_M$.
 
 Observe that $T_0, T_1, T_2, \cdots$ are an increasing sequence of stopping times adapted to $(\calF_n)_{n \in \omega}$, and each $T_i$ is bounded.
-Thanks to \Cref{thm:rankable and strict rankable}, 
-$(X_n)_{n \in \omega}$ is a supermartingale;
-moreover, it follows from \Cref{thm:optional sampling} that $(Y_n)_{n \in \omega}$ is a ($1$-ranking) supermartingale.
-Therefore, by \Cref{thm:rank-PAST}, $\expect{T^{\tY}_M} < \infty$.
+As $(X_n)_{n \in \omega}$ is a supermartingale (\Cref{thm:rankable gives supermartingale}), it follows from \Cref{thm:optional sampling} that $(Y_n)_{n \in \omega}$ is a supermartingale (in fact, a $1$-ranking one).
+Therefore, by \Cref{thm:rank-PAST}, $M$ is PAST.
+
+Thus, to summarise
 
 \begin{corollary}
-If a closed SPCF term $M$ is rankable, then $\expect{T^{\tY}_M} < \infty$, and so, $T^{\tY}_M < \infty$ a.s.~and $M$ is AST.
-
-If a closed SPCF term $M$ is strictly rankable, then $\expect{T_M} < \infty$ i.e.~$M$ is PAST.
+If a closed SPCF term is rankable, then it is positively almost-surely terminating.
 \end{corollary}

### 2020-05-12 probabilisticProgrammingMartingales.tex

@@
 \section{Constructing Ranking Functions}
-Although rankability implies almost sure termination, the converse does not hold in general. For example,
+Although rankability implies almost-sure termination, the converse does not hold in general. For example,
 \begin{equation}
-\tif{-\tsample < 0}{\underline{0}}{(\tY \lambda x. x)\underline 0}
+\tif{{-}\tsample < 0}{\underline{0}}{(\tY \lambda x. x) \; \underline 0}
 \end{equation}
@@
 In particular, the expected number of $\tY$-reduction steps from each reachable term is a ranking function.
-Note that a finite number of expected $\tY$-reduction steps does not necessarily imply a finite number of expected total reduction steps, for example $(\tY \lambda f n.
-\tif{\tsample - 0.5 < 0}{n \underline 0}{f (\lambda x.
-n (n x))}) (\lambda x.
-x+1)$ terminates with only 2 $\tY$-reductions on average, but applies the increment function $2^n$ times with probability $2^{-1-n}$ for $n \geq 0$, which diverges.
+Note that a finite number of expected $\tY$-reduction steps does not necessarily imply a finite number of expected total reduction steps.
 
-\begin{theorem} \label{minimal}
+\begin{example}
+\label{ex:tY finite does not imply t finite}
+The term
+\[
+M = \big(\tY \lambda f n. \tif{\tsample - 0.5 < 0}{n \, \underline 0}{f (\lambda x. n (n x))}\big) (\lambda x. x+1)
+\] 
+terminates with only 2 $\tY$-reductions on average i.e.~$\expect{T_M^\tY} = 2$, but applies the increment function $2^n$ times with probability $2^{-1-n}$ for $n \geq 0$, which diverges i.e.~$\expect{T_M} = \infty$.
+\end{example}
+
+\begin{theorem} \label{thm:minimal}
 Given a closed term $M$, the function $f:Rch(M) \to \mathbb R$ given by $f(N) = \mathbb E [\text{the number of }\tY\text{-reduction steps from }N\text{ to a value}]$, if it exists, is the least of all possible ranking functions of $M$.
@@
 
-\paragraph{}
-Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and Theorem \ref{minimal} makes even stronger assumptions than almost sure termination, so it isn't useful for proving it.
+\subsection{Partial ranking functions}
+Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost sure termination, so it isn't useful for proving it.
 Take, for example, the term

### 2020-05-12 ranking-supermartingales.tex

--- 2020-05-10 ranking-supermartingales.tex
@@
 
-First each $f(M_n)$ is integrable. Since $0 \leq f(N) \leq f(M)$ for all $N \in \mathit{Rch}(M)$, we have
+\lo{We need to show that each $f(M_n)$ is integrable. LO claimed earlier that $0 \leq f(N) \leq f(M)$ for all $N \in \mathit{Rch}(M)$.
+This is of course false: take $f$ with $v \, \tsample \mapsto 1$ and $v \, \underline{r} \mapsto 2\, r$ for some value $v$. 
+However, it is true that
 \[
@@
 \leq
-\int_{S} \mu(\dif s) \, f(M)
+\int_{S} \mu(\dif s) \, f(M_0)(s)
 =
@@
 \]
+the inequality above follows from \Cref{lem:key rankable}.
+}
 
@@
 \begin{lemma}
+\label{lem:key rankable}
 For all $i \in \set{1, 2, 3, 4}$ and $A \in \calF_n$

### 2020-05-13 probabilisticProgrammingMartingales.tex
ongoing

@@
 
-When proving almost sure termination in this way /akr{I guess I should move this section somewhere later.}, it is necessary to consider which terms a given term may reduce to.
+When proving almost sure termination in this way \akr{I guess I should move this section somewhere later.}, it is necessary to consider which terms a given term may reduce to.
 Sometimes however, the reduction that the programmer has in mind may not be strictly the call-by-value order defined so far, or considering an alternative reduction order may be simpler or more intuitive.
 
-Non-probabilistis lambda calculi generally have the Church-Rosser property, that if a term $A$ reduces to both $B_1$ and $B_2$, there is some $C$ with reduction sequences $B_1 \to^* C$ and $B_2 \to^* C$, so the reduction order mostly doesn't matter.
+Non-probabilistic lambda calculi generally have the Church-Rosser property, that if a term $A$ reduces to both $B_1$ and $B_2$, there is some $C$ with reduction sequences $B_1 \to^* C$ and $B_2 \to^* C$, so the reduction order mostly doesn't matter.
 In the probabilistic case, this may not be true, because $\beta$-reduction can duplicate $\tsample$s, so the outputs of the copies of the sample may be identical or independent, depending on whether the sample is taken before or after $\beta$-reduction.
@@
 \paragraph{}
-Even with this restriction, a sampling semantics in the style of the one already defined would not be entirely Church-Rosser, as, for example, $red^3(\tsample - \tsample, (1,0,\dots))$ would be either $1$ or $-1$ depending on the order of evaluation of the $\tsample$s, as that determines which sample from the pre-selected sequence is used for each one.
+Even with this restriction, a sampling semantics in the style of the one already defined would not be entirely Church-Rosser, as, for example, $\red^3(\tsample - \tsample, (1,0,\dots))$ would be either $1$ or $-1$ depending on the order of evaluation of the $\tsample$s, as that determines which sample from the pre-selected sequence is used for each one.
 To fix this, rather than pre-selecting samples according to the order they'll be drawn in, select them according to the position in the term where they'll be used instead.
 
-A position \akr{I'm not really sure how to phrase the citation, but this is adapded from \cite{Kennaway96infinitarylambda}.} is a finite sequence of steps into a term, defined inductively as
+A position \akr{I'm not really sure how to phrase the citation, but this is adapted from \cite{Kennaway96infinitarylambda}.} is a finite sequence of steps into a term, defined inductively as
 \begin{align*}
@@
 so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g.
-$xy \mid \lambda ; \cdot$ is undefined).
+$x \, y \mid \lambda ; \cdot$ is undefined).
 A position such that $M\mid P$ does exist is said to occur in $M$.
@@
 
-With this notation, a more general reduction relation can be defined. \akr{It would almost work to have one of the neater versions of $\tY$-reduction here.}
+With this notation, a more general reduction relation $\to$ can be defined. \akr{It would almost work to have one of the neater versions of $\tY$-reduction here.}
+\lo{On first reading, I struggled to parse the following clauses.}
+\changed[lo]{The binary relation $\to$ is defined by the following rules, each is conditional on a position $P$ that occurs in the term $M$ in question.}
 \begin{align*}
@@
 \end{align*}
+\lo{The definition above is clear. But as a general rule, it is good to illustrate definitions -- especially if they are genuinely new -- with an example or two.}
 
@@
 Labelling the pre-chosen samples by the positions in the term would also not work because in some cases, a $\tsample$ will be duplicated before being reduced, for example, in $(\lambda x.
-\underline x 0 \underline + x \underline 0) (\lambda y.
+{\underline x \, 0} \, \mathbin{\underline{+}} \, {x \, \underline 0}) (\lambda y.
 \tsample)$, both of the $\tsample$ redexes that eventually occur originate at $@_2 ; \lambda$.
@@
 Reduction sequences are used rather than reachable terms because if the same term is reached twice, different samples may be needed, but they will be discussed as though they were terms, by abuse of notation.
-The relation $\sim$ is defined as the union of the symmetric relations $\sim_p$ and $\sim_c$ where
+\lo{Readers would benefit greatly from some examples of an $L_0(M)$-pair, and relations $\sim_p$ and $\sim_c$, here, before the formal definitions.
+Explain the underlying intuition.} The relation $\sim$ is defined as the union of the symmetric relations $\sim_p$ and $\sim_c$ where
 \begin{enumerate}
     \item If $N$ reduces to $O$ with the redex at position $P$, and $Q$ is a position in $N$ where $Q \not \geq P$, then $(N,Q) \sim_p (O,Q)$.
-    \item If $N$ $\beta$-reduces to $O$ at position $P$, $Q$ is a position in $N | P;@_1;\lambda$ and $N \mid P;@_1;\lambda;Q$ is not the variable involved in the reduction, $(N,P;@_1;\lambda;Q) \sim_p (O, P;Q)$
+    \item If $N$ $\beta$-reduces to $O$ at position $P$, $Q$ is a position in $N \mid P;@_1;\lambda$ and $N \mid P;@_1;\lambda;Q$ is not the variable involved in the reduction, $(N,P;@_1;\lambda;Q) \sim_p (O, P;Q)$
     \item If $N$ $\textsf{if}$-reduces to $O$ at position $P$, with the first resp.
@@
     \item If $N \mid P = (\tY \lambda x.
-A)$, $A \mid Q = C$ is a redex reducing to $C'$, $R$ exists in $C'$ and $C' \mid R \neq x$, then letting $O_1$ be $N$ reduced first at $P$ then $P;\lambda;@_1;Q$ and $O_2$ reduced at $P;\tY;\lambda;Q$ then $P$, $(O_1, P;\lambda;@_1;Q;R) \sim_c (O_2, P;\lambda;@_1;Q;R)$
+A)$, $A \mid Q = C$ is a redex reducing to $C'$, $R$ exists in $C'$ and $C' \mid R \neq x$, then letting $O_1$ be $N$ reduced first at $P$ then $P;\lambda;@_1;Q$ and $O_2$ reduced at $P;\tY;\lambda;Q$ then \changed[lo]{at} $P$, $(O_1, P;\lambda;@_1;Q;R) \sim_c (O_2, P;\lambda;@_1;Q;R)$
     \item If $N \mid P = (\tY \lambda x.
-A)$, $A \mid Q = C$ is a redex reducing to $C'$, $R$ exists in $C'$ and $A \mid S = x$, then letting $O_1$ be $N$ reduced first at $P$ then $P;\lambda;@_1;S;\tY;\lambda;Q$ and $O_2$ reduced at $P;\tY;\lambda;Q$ then $P$, $(O_1, P;\lambda;@_1;S;\tY;\lambda;Q;R) \sim_c (O_2, P;\lambda;@_1;S;\tY;\lambda;Q;R)$
+A)$, $A \mid Q = C$ is a redex reducing to $C'$, $R$ exists in $C'$ and $A \mid S = x$, then letting $O_1$ be $N$ reduced first at $P$ then $P;\lambda;@_1;S;\tY;\lambda;Q$ and $O_2$ reduced at $P;\tY;\lambda;Q$ then \changed[lo]{at} $P$,
+    \[
+    (O_1, P;\lambda;@_1;S;\tY;\lambda;Q;R) \sim_c (O_2, P;\lambda;@_1;S;\tY;\lambda;Q;R)
+    \]
 \end{enumerate}
@@
 In summary, all equivalence classes under $\sim^*$ are of the shapes \akr{This would be far easier to show with a few pictures.
-I just don't yet know how to make them in Latex and I'm prioritising getting the text done.}.
+I just don't yet know how to make them in Latex and I'm prioritising getting the text done.}
+\lo{I can suggest possible \LaTeX\ packages for drawing. Can you draw them and upload a scan?}
 

### 2020-05-13 probabilisticProgrammingMartingales.tex
question

@@
 \newcommand\ndraw[2]{\#\mathrm{draw}_{#1}(#2)}
+
+\makeatletter
+\newcommand{\dotDelta}{{\vphantom{\Delta}\mathpalette\d@tD@lta\relax}}
+\newcommand{\d@tD@lta}[2]{%
+  \ooalign{\hidewidth$\m@th#1\mkern-1mu\cdot$\hidewidth\cr$\m@th#1\Delta$\cr}%
+}
 
@@
 Two subterms $N_1$ and $N_2$ of a term $M$, corresponding to positions $P_1$ and $P_2$, can overlap in a few different ways.
-If $P_1$ is an initial segment of $P_2$, then $N_2$ is also a subterm of $N_1$, which is written as $P_1 \leq P_2$.
+If $P_1$ is an initial segment \lo{(i.e.~prefix)?} of $P_2$, then $N_2$ is also a subterm of $N_1$, which is written as $P_1 \leq P_2$.
 If neither $P_1 \leq P_2$ nor $P_1 \geq P_2$, the positions are said to be disjoint.
@@
 \begin{enumerate}
-    \item If $N$ reduces to $O$ with the redex at position $P$, and $Q$ is a position in $N$ where $Q \not \geq P$, then $(N,Q) \sim_p (O,Q)$.
+    \item If $N$ reduces to $O$ with the redex at position $P$\lo{Do you mean: $M \to^\ast N \to O = N[\dotDelta / P]$ for some redex contraction $\Delta \to \dotDelta$}, and $Q$ is a position in $N$ 
+    where $Q \not \geq P$, then $(N,Q) \sim_p (O,Q)$.
+    \lo{This implies that $Q$ is a position in $O$, but $O \mid Q$ is not guaranteed to exist.
+E.g.~$N = (\lambda x.x) 1$ and $O = 1$.
+Now $N \mid Q$ exists, where $Q = @_1 ; \lambda$, but $O \mid Q$ does not exist.}
+
+    \lo{I find this confusing. Can you express it formally?}
+
     \item If $N$ $\beta$-reduces to $O$ at position $P$, $Q$ is a position in $N \mid P;@_1;\lambda$ and $N \mid P;@_1;\lambda;Q$ is not the variable involved in the reduction, $(N,P;@_1;\lambda;Q) \sim_p (O, P;Q)$
@@
 Before defining the new version of the reduction relation $\red$, a few lemmas about the properties of $\sim$ are necessary for it to be well-defined.
-Considering the set of paths through $Rch(M)$ (which we're implicitly ideitifying with $Rch(M)$ itself) as a tree rooted at (the trivial path at) $M$, $\sim_p$ only relates positions in child and parent terms, and $\sim_c$ only relates first cousins.
+Considering the set of paths through $Rch(M)$ (which we're implicitly identifying with $Rch(M)$ itself) as a tree rooted at (the trivial path at) $M$, $\sim_p$ only relates positions in child and parent terms, and $\sim_c$ only relates first cousins.
 Furthermore, by a case analysis on the forms of positions, the possible paths $N_1 \sim N_2 \sim \dots \sim N_n$ can be strongly restricted.

### 2020-05-13 probabilisticProgrammingMartingales.tex
Added diagram.

@@
 \input{Style/comment}
+
+\usepackage{graphicx}
 
@@
 \lo{On first reading, I struggled to parse the following clauses.}
-\changed[lo]{The binary relation $\to$ is defined by the following rules, each is conditional on a position $P$ that occurs in the term $M$ in question.}
+\changed[lo]{The binary relation $\to$ is defined by the following rules, each is conditional on a redex occuring at position $P$ in the term $M$ in question.}
 \begin{align*}
@@
 Now $N \mid Q$ exists, where $Q = @_1 ; \lambda$, but $O \mid Q$ does not exist.}
+    \lo{Ah! I overlooked the condition $Q \not\geq P =
+    {\cdot}$.}
 
@@
 \end{enumerate}
+
+\begin{figure}
+\includegraphics[height=0.9\textheight]{akr-diagram.pdf}
+\caption{\label{fig:adk-diagram} Some examples of $\sim_p$- and $\sim_c$-rules}
+\end{figure}
+
+\begin{example}
+See \Cref{fig:adk-diagram} for examples of $\sim_p$- and $\sim_c$-rules.
+\end{example}
 

### 2020-05-18 probabilisticProgrammingMartingales.tex
A more succinct version of the 10? (schematic) terms in Rch of geometric distribution example

@@
   \frac{}{\underline{r} : \textsf{R}} \qquad
-  \frac{\Gamma \vdash s:A \to B \quad \Gamma \vdash t : A}{\Gamma \vdash s t : B} \\ \\
+  \frac{\Gamma \vdash s:A \to B \quad \Gamma \vdash t : A}{\Gamma \vdash s \, t : B} \\ \\
   \frac{\Gamma \vdash s_1:\textsf{R} \dots \Gamma \vdash s_n:\textsf{R}}{\Gamma \vdash \underline{f}(s_1,\dots,s_n) : \textsf{R}} \ f : \mathbb{R}^n \to \mathbb{R} \qquad
@@
 Two subterms $N_1$ and $N_2$ of a term $M$, corresponding to positions $P_1$ and $P_2$, can overlap in a few different ways.
-If $P_1$ is an initial segment \lo{(i.e.~prefix)?} of $P_2$, then $N_2$ is also a subterm of $N_1$, which is written as $P_1 \leq P_2$.
+If $P_1$ is an initial segment (i.e.~prefix) of $P_2$, then $N_2$ is also a subterm of $N_1$, which is written as $P_1 \leq P_2$.
 If neither $P_1 \leq P_2$ nor $P_1 \geq P_2$, the positions are said to be disjoint.
@@
 With this notation, a more general reduction relation $\to$ can be defined. \akr{It would almost work to have one of the neater versions of $\tY$-reduction here.}
-\lo{On first reading, I struggled to parse the following clauses.}
-\changed[lo]{The binary relation $\to$ is defined by the following rules, each is conditional on a redex occuring at position $P$ in the term $M$ in question.}
+%\lo{On first reading, I struggled to parse the following clauses.}
+\changed[lo]{The binary relation $\to$ is defined by the following rules, each is conditional on a redex occuring at position $P$ in the term $M$.}
 \begin{align*}
@@
 Reduction sequences are used rather than reachable terms because if the same term is reached twice, different samples may be needed, but they will be discussed as though they were terms, by abuse of notation.
-\lo{Readers would benefit greatly from some examples of an $L_0(M)$-pair, and relations $\sim_p$ and $\sim_c$, here, before the formal definitions.
-Explain the underlying intuition.} The relation $\sim$ is defined as the union of the symmetric relations $\sim_p$ and $\sim_c$ where
+\lo{Readers would benefit greatly from some examples of an $L_0(M)$-pair, and the relations $\sim_p$ and $\sim_c$, here, before the formal definitions.
+Explain the underlying intuition.} The relation $\sim$ is defined as the union of the symmetric relations $\sim_p$ \changed[lo]{($p$ for parent-child relationship)} and $\sim_c$ \changed[lo]{($c$ for cousin relationship)} where
+\lo{As defined, $\sim_p$ is not a symmetric relation.}
 \begin{enumerate}
@@
     where $Q \not \geq P$, then $(N,Q) \sim_p (O,Q)$.
-    \lo{This implies that $Q$ is a position in $O$, but $O \mid Q$ is not guaranteed to exist.
+    %\lo{This implies that $Q$ is a position in $O$, but $O \mid Q$ is not guaranteed to exist.
 E.g.~$N = (\lambda x.x) 1$ and $O = 1$.
-Now $N \mid Q$ exists, where $Q = @_1 ; \lambda$, but $O \mid Q$ does not exist.}
-    \lo{Ah! I overlooked the condition $Q \not\geq P =
-    {\cdot}$.}
-
-    \lo{I find this confusing. Can you express it formally?}
+Now $N \mid Q$ exists, where $Q = @_1 ; \lambda$, but $O \mid Q$ does not exist.} \lo{Ah!
+I overlooked the condition $Q \not\geq P = {\cdot}$.}
 
     \item If $N$ $\beta$-reduces to $O$ at position $P$, $Q$ is a position in $N \mid P;@_1;\lambda$ and $N \mid P;@_1;\lambda;Q$ is not the variable involved in the reduction, $(N,P;@_1;\lambda;Q) \sim_p (O, P;Q)$
+
     \item If $N$ $\textsf{if}$-reduces to $O$ at position $P$, with the first resp.
@@
 $3$), $(N,P;\textsf{if}_i;Q) \sim_p (O,P;Q)$
+
     \item If $N \mid P = (\lambda x.
 A) B$, $A \mid Q = x$, $B \mid R = C$ is a redex reducing to $C'$ and $S$ exists in $C'$, then letting $O_1$ be $N$ reduced first at $P$ then at $P;Q;R$ and letting $O_2$ be $N$ reduced first at $P;@_2;R$ then at $P$, $(O_1,P;Q;R;S) \sim_c (O_2,P;Q;R;S)$
+
     \item If $N \mid P = (\tY \lambda x.
 A)$, $A \mid Q = C$ is a redex reducing to $C'$, $R$ exists in $C'$ and $C' \mid R \neq x$, then letting $O_1$ be $N$ reduced first at $P$ then $P;\lambda;@_1;Q$ and $O_2$ reduced at $P;\tY;\lambda;Q$ then \changed[lo]{at} $P$, $(O_1, P;\lambda;@_1;Q;R) \sim_c (O_2, P;\lambda;@_1;Q;R)$
+
     \item If $N \mid P = (\tY \lambda x.
@@
 \begin{example}
-See \Cref{fig:adk-diagram} for examples of $\sim_p$- and $\sim_c$-rules.
+See \Cref{fig:adk-diagram} for a graphical depiction of $\sim_p$- and $\sim_c$-rules.
 \end{example}
@@
 I just don't yet know how to make them in Latex and I'm prioritising getting the text done.}
-\lo{I can suggest possible \LaTeX\ packages for drawing. Can you draw them and upload a scan?}
+\lo{PGF/TikZ is the standard \LaTeX\ package for generating vector graphics.
+The manual \url{https://ctan.org/pkg/pgf?lang=en} is a 1318-page tome!
+There are numerous tutorials / guides on the Internet.
+E.g.~\url{https://www.overleaf.com/learn/latex/TikZ_package}}
 
@@
     Despite its simplicity, its $Rch$ contains all the terms
-\begin{itemize}
+\begin{enumerate}
     \item $(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{0}$
@@
     \item $(\lambda z.(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) z) (\underline{i} + 1)$.
-\end{itemize}
+\end{enumerate}
 \akr{TODO: make this into a diagram with arrows.}
+
+\newpage
+
+\lo{Here is a more readable version.}
+
+Writing $\Theta = \lambda f, n : \tif{\tsample - 0.5 < 0}{n}{f (n+1)}$, we have
+\begin{enumerate}
+    \item $(\tY \, \Theta) \; \underline{0}$
+
+    \item $\big(\lambda z.\Theta \; (\tY \, \Theta) \; z \big) \; \underline{i}$
+
+    \item $\Theta \; (\tY \, \Theta) \; \underline{i}$
+
+    \item \changed[akr]{$(\lambda n: \tif{\tsample - 0.5 < 0}{n}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{n}{f (m+1)}) (m+1)}) \; \underline{i}$}
+
+    \lo{I think the preceding should be: $(\lambda n: \tif{\tsample - 0.5 < 0}{n}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (n+1)}) \; \underline{i}$, and so, more succinctly:
+    \[
+    \big(
+    \lambda n: 
+    \tif{\tsample - 0.5 < 0}{n}{(\tY \, \Theta) \; (n+1)} 
+    \big) 
+    \; \underline{i}
+    \]}
+
+    \item $\tif{\tsample - 0.5 < 0}{\underline{i}}{
+    (\tY \, \Theta) \, (\underline{i}+1)}$
+
+    \item $\tif{\underline r - 0.5 < 0}{\underline{i}}{(\tY \, \Theta) (\underline{i}+1)}$
+
+    \item $\tif{\underline{r - 0.5} < 0}{\underline{i}}{(\tY \, \Theta) (\underline{i}+1)}$
+
+    \item $\underline{i}$
+
+    \item $(\tY \, \Theta) \; (\underline{i}+1)$
+
+    \item $\big(\lambda z. \Theta \; (\tY \, \Theta) \;  z\big) \; (\underline{i} + 1)$.
+\end{enumerate}
+
 Even in this simple case, defining a ranking function explicitly is awkward because of the number of cases, although in most cases, because the value need only be greater than or equal to that of the next term in sequence, it suffices to take the ranking function as taking only 3 distinct values.

### 2020-05-18 ranking-supermartingales.tex
ongoing
A more succinct version of the 10? (schematic) terms in Rch of geometric distribution example

@@
 \lo{We need to show that each $f(M_n)$ is integrable. LO claimed earlier that $0 \leq f(N) \leq f(M)$ for all $N \in \mathit{Rch}(M)$.
-This is of course false: take $f$ with $v \, \tsample \mapsto 1$ and $v \, \underline{r} \mapsto 2\, r$ for some value $v$. 
+This is of false: take $f$ with $v \, \tsample \mapsto 1$ and $v \, \underline{r} \mapsto 2\, r$ for some value $v$. 
 However, it is true that
@@
 
-We say that a given SPCF term is \emph{type-1} (respectively 2, 3 and 4) if it has the shape $E[\tY (\lambda x. N)]$ (respectively $E[\tsample]$, $E[R]$ where $R$ is any other redex, and of a value).
+We say that a given SPCF term is \emph{type-1} (respectively 2, 3 and 4) if it has the shape $E[\tY \lambda x. N]$ (respectively $E[\tsample]$, $E[R]$ where $R$ is any other redex, and of a value).
 Define $\mathbf{T}_i := 
@@
 T_0(s) & := 0 \\
-T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value or of form } E[\tY (\lambda x. N)] \}\\
+T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value or of form } E[\tY \lambda x. N] \}\\
 X_n(s) & := f(M_n(s)) \\

### 2020-05-18 probabilisticProgrammingMartingales.tex
Questions about "partial ranking function is a restriction of a ranking function".

@@
 
-\newpage
-
 \lo{Here is a more readable version.}
@@
 
-    \item $\big(\lambda z. \Theta \; (\tY \, \Theta) \;  z\big) \; (\underline{i} + 1)$.
+    \item $\big(\lambda z. \Theta \; (\tY \, \Theta) \;  z\big) \; (\underline{i} + 1)$
+
+    \item $\Big(\lambda z. \big(
+    \lambda n: 
+    \tif{\tsample - 0.5 < 0}{n}{(\tY \, \Theta) \; (n+1)} 
+    \big) \, z \Big)
+    \; \underline{i}$
 \end{enumerate}
 
+\lo{So we have:
+\[
+\begin{array}{ll}
+1 \to 11 \to 4 \to 5 \to 6 \to 7 \to 8\\
+7 \to 9 \to 11 \to \cdots
+\end{array}
+\]
+You have 2, 3 and 10 because your $\tY$-redex rule is
+\[
+\tY \lambda x. s \to \lambda z. (\lambda x.s)\, (\tY \lambda x. s) \, z
+\]
+which is not the official definition.}
+
 Even in this simple case, defining a ranking function explicitly is awkward because of the number of cases, although in most cases, because the value need only be greater than or equal to that of the next term in sequence, it suffices to take the ranking function as taking only 3 distinct values.
+\lo{This is only true for partial ranking function.}
 
@@
 
-Define $(\next(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) = \left | \{m < n \mid \red^M(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
-$\next$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on partial ranking functions.
+Define $(\next(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) = \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
+The function $\next$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on partial ranking functions.
 Define $f_2(N) = \int_S f_1(\next(N,s)) + g(N,s) \, \mu(\mathrm d s)$.
@@
 
+\lo{Call the above partial function $h$.
+I assume that $h: (\tY f, n. n \oplus_{0.5} f(n+1)) N \mapsto 2$ for all $N$ (such that $(\tY f, n. n \oplus_{0.5} f(n+1)) N$ is reachable from $M$), 
+and the second and third clauses hold for all $i \geq 0$.
+
+Maybe I have misread / misunderstood the preceding, but I don't see why $h$ is a restriction of a ranking function. 
+Suppose $h$ is a restriction of $H$.
+Consider the reduction sequence (writing $\Xi = (\tY f, n. n \oplus_{0.5} f(n+1))$):
+\[
+\Xi \, 0 
+\xrightarrow{\tY} 
+0 \oplus_{0.5} \Xi \, 1 
+\to 
+\Xi \, 1
+\xrightarrow{\tY} 
+1 \oplus_{0.5} \Xi \, 2
+\to
+\Xi \, 2
+\xrightarrow{\tY} 
+2 \oplus_{0.5} \Xi \, 3
+\]
+Since $H(\Xi \, 0) = 2$, we must have 
+$H(\Xi \, 1) \leq 1$ and $H(\Xi \, 2) \leq 0$, contradicting the restriction assumption.
+}
+
+\bibliographystyle{apalike}
+\bibliography{references}
+
 \lo{22 Jan: 
@@
 
-\bibliographystyle{apalike}
-\bibliography{references}
-
 \iffalse

### 2020-05-26 probabilisticProgrammingMartingales.tex
Add explanatory comment.
Fix the definition of \sim, again.
New definition I'm pretty sure is correct this time (although that
was the case the previous 4 or so times too), but the accompanying
proof hasn't yet been updated.

@@
 \end{align*}
+In each of these cases, $M \mid P$ is the redex, and reduction takes place at $P$.
 \lo{The definition above is clear. But as a general rule, it is good to illustrate definitions -- especially if they are genuinely new -- with an example or two.}
@@
 \begin{enumerate}
-    \item If $N$ reduces to $O$ with the redex at position $P$\lo{Do you mean: $M \to^\ast N \to O = N[\dotDelta / P]$ for some redex contraction $\Delta \to \dotDelta$}, and $Q$ is a position in $N$ 
-    where $Q \not \geq P$, then $(N,Q) \sim_p (O,Q)$.
-    %\lo{This implies that $Q$ is a position in $O$, but $O \mid Q$ is not guaranteed to exist.
-E.g.~$N = (\lambda x.x) 1$ and $O = 1$.
-Now $N \mid Q$ exists, where $Q = @_1 ; \lambda$, but $O \mid Q$ does not exist.} \lo{Ah!
-I overlooked the condition $Q \not\geq P = {\cdot}$.}
+    \item If $N$ reduces to $O$ with the redex at position $P$\lo{Do you mean: $M \to^\ast N \to O = N[\dotDelta / P]$ for some redex contraction $\Delta \to \dotDelta$}, and $Q$ is a position in $N$ disjoint from $P$, then $(N,Q) \sim_p (O,Q)$.
 
@@
 
-    \item If $N \mid P = (\lambda x.
-A) B$, $A \mid Q = x$, $B \mid R = C$ is a redex reducing to $C'$ and $S$ exists in $C'$, then letting $O_1$ be $N$ reduced first at $P$ then at $P;Q;R$ and letting $O_2$ be $N$ reduced first at $P;@_2;R$ then at $P$, $(O_1,P;Q;R;S) \sim_c (O_2,P;Q;R;S)$
+    \item If $N$, $O_1$ and $O_2$ match any of the following cases:
+    \begin{enumerate}
+        \item $N$ contains redexes at positions $P_1$ resp.~$P_2$, $O_1$ is $N$ reduced first at $P_1$
 
-    \item If $N \mid P = (\tY \lambda x.
-A)$, $A \mid Q = C$ is a redex reducing to $C'$, $R$ exists in $C'$ and $C' \mid R \neq x$, then letting $O_1$ be $N$ reduced first at $P$ then $P;\lambda;@_1;Q$ and $O_2$ reduced at $P;\tY;\lambda;Q$ then \changed[lo]{at} $P$, $(O_1, P;\lambda;@_1;Q;R) \sim_c (O_2, P;\lambda;@_1;Q;R)$
+        \item $N \mid P = \tif{\underline r < 0}{N_1}{N_2}$, where $r < 0$ (or, respectively, $r \geq 0$), $(N_2 \text{ resp.
+} N_1) \mid Q$ is a redex, and $O_1$ is $N$ reduced at $P$ and $O_2$ is $N$ reduced first at $P;(\textsf{if}_3 \text{ resp.
+} \textsf{if}_2);Q$ then at $P$
 
-    \item If $N \mid P = (\tY \lambda x.
-A)$, $A \mid Q = C$ is a redex reducing to $C'$, $R$ exists in $C'$ and $A \mid S = x$, then letting $O_1$ be $N$ reduced first at $P$ then $P;\lambda;@_1;S;\tY;\lambda;Q$ and $O_2$ reduced at $P;\tY;\lambda;Q$ then \changed[lo]{at} $P$,
-    \[
-    (O_1, P;\lambda;@_1;S;\tY;\lambda;Q;R) \sim_c (O_2, P;\lambda;@_1;S;\tY;\lambda;Q;R)
-    \]
+        \item $N \mid P = \tif{\underline r < 0}{N_1}{N_2}$, where $r < 0$ (or, respectively, $r \geq 0$), $(N_1 \text{ resp.
+} N_2) \mid Q$ is a redex, and $O_1$ is $N$ reduced first at $P$ then at $P;Q$ and $O_2$ is $N$ reduced first at $P;(\textsf{if}_2 \text{ resp.
+} \textsf{if}_3);Q$ then at $P$
+
+        \item $N \mid P = (\lambda x. A) B$, there is a redex in $A$ at position $Q$, $O_1$ is $N$ reduced first at $P$ then at $P;Q$, and $O_2$ is $N$ reduced first at $P;@_1;\lambda;Q$ then at $P$
+
+        \item $N \mid P = (\lambda x.
+A) B$, $B \mid Q$ is a redex, $(R_i)_i$ is a list of all the positions in $A$ where $A \mid R = x$, $O_1$ is $N$ reduced first at $P ; @_2 ; Q$ then at $P$, and $O_2$ is $N$ reduced first at $P$ then at $P;R_i;Q$ for each $i$ in order.
+
+        \item $N \mid P = \tY (\lambda x.
+A)$, $A$ reduced at $Q$ is $A'$, $(R_i)_i$ is a list of all the positions where $A' \mid R  = x$, $O_1$ is $N$ reduced first at $P;\tY;\lambda;Q$ then at $P$, and $O_2$ is $N$ reduced first at $P$ then at $P;\lambda;@_1;Q$ then at $P;\lambda;@_1;R_i;Y;\lambda;Q$ for each $i$ in order.
+    \end{enumerate}
+    and $O_1$ and $O_2$ are equal as terms, but with different reduction sequences (which merely requires that in cases a, c and d, if the redexes of unspecified type are $\tsample$, corresponding $\tsample$s reduce to the same numbers), $O_1'$ and $O_2'$ are the results of applying some reduction sequence to each of $O_1$ and $O_2$ (the same reductions in each case, which is always possible because they're equal terms), and $S$ is a position in $O_1'$ (or equivalently $O_2'$), then $(O_1',S) \sim_c (O_2',S)$.
 \end{enumerate}
@@
 \includegraphics[height=0.9\textheight]{akr-diagram.pdf}
-\caption{\label{fig:adk-diagram} Some examples of $\sim_p$- and $\sim_c$-rules}
+\caption{\label{fig:adk-diagram} Some examples of the old version of the $\sim_p$- and $\sim_c$-rules}
 \end{figure}
@@
 }
+\akr{
+In this case, $H(0 \oplus_{0.5} \Xi 1) \leq 1$, but that term reduces to either $0$ or $\Xi 1$ therefore the actual condition is that $0.5 H(0) + 0.5 H(\Xi 1) \leq 1$, which is possible.
+}
 

### 2020-05-27 probabilisticProgrammingMartingales.tex
Added new section on antitone-convex ranking function; title and table of contents

--- 2020-05-18 probabilisticProgrammingMartingales.tex
@@
 \input{Style/comment}
-
+\usepackage{calculation}
 \usepackage{graphicx}
@@
 \newcommand{\Real}{\mathbb{R}}
+\newcommand{\nnReal}{\mathbb{R}_{{\geq}0}}
+\newcommand{\pReal}{\mathbb{R}_{{>}0}}
+
 \newcommand\expect[1]{\mathbb{E}[#1]}
@@
 \DeclareMathOperator{\red}{red}
-\DeclareMathOperator{\next}{next}
+\DeclareMathOperator{\nnext}{next}
 
@@
 \newtheorem*{remark}{Remark}
+
+\title{Supermartingales and Termination Analysis of Statistical PCF}
+
 \begin{document}
 
+\maketitle
+
+\lo{Tentative title}
+
+\begin{abstract}
+\lo{TODO}
+\end{abstract}
+
+\iffalse
 \akr{Following your suggestion, I will be providing a criterion for termination of programs in PPCF \citep{DBLP:journals/pacmpl/EhrhardPT18} based on ranking supermartingales. 
@@
 \lo{Why not provide a proof as an appendix?}
+\fi
+
+\tableofcontents
 
@@
 
-\section{Language definition}
+\section{Statistical PCF}
+
+\subsection{Syntax of SPCF}
+
 The language SPCF is a simply-typed lambda calculus with sampling of real numbers from $[0,1]$ and unbounded scoring, following \cite{MakOP20b}.
@@
 
-\subsection{Alternative Semantics}
+\section{Alternative Semantics}
 \akr{This version of the semantics is much more convenient for allowing multiple different reduction orders.
@@
 \tif{{-}\tsample < 0}{\underline{0}}{(\tY \lambda x. x) \; \underline 0}
+\label{ex:0 probability reachable}
 \end{equation}
 terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x. x) \underline 0$ is reachable, although that has probability 0. 
-Not only is this counterexample a.s.t, it's positively almost surely terminating i.e.~the expected time to termination is finite.
+Not only is this counterexample AST, it's positively almost surely terminating i.e.~the expected time to termination is finite.
 
@@
 Even in this simple case, defining a ranking function explicitly is awkward because of the number of cases, although in most cases, because the value need only be greater than or equal to that of the next term in sequence, it suffices to take the ranking function as taking only 3 distinct values.
-\lo{This is only true for partial ranking function.}
+(We will explain why later.)
 
@@
 
-Define $(\next(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) = \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
-The function $\next$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on partial ranking functions.
-Define $f_2(N) = \int_S f_1(\next(N,s)) + g(N,s) \, \mu(\mathrm d s)$.
+Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) = \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
+The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on partial ranking functions.
+Define $f_2(N) = \int_S f_1(\nnext(N,s)) + g(N,s) \, \mu(\mathrm d s)$.
 The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction).
@@
 }
+\akr{
+In this case, $H(0 \oplus_{0.5} \Xi 1) \leq 1$, but that term reduces to either $0$ or $\Xi 1$ therefore the actual condition is that $0.5 H(0) + 0.5 H(\Xi 1) \leq 1$, which is possible.
+}
+\lo{
+  OK. I see it now.
+}
+
+\section{Antitone-convex ranking functions}
+
+\begin{example}
+\label{ex:ac-ranking}
+\begin{enumerate}
+\item 1D unbiased random walk
+\[
+M_1 = 
+\big(\tY \, f \, n \, . \, 
+\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}\big) \, 1
+\]
+
+\item 1D biased (away from 0) random walk
+\[
+M_2 = 
+\big(\tY \, f \, n \, . \, 
+\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/3} f \, (n + 1)}\big) \, 10
+\]
+
+\item 1D biased (towards 0) random walk
+\[
+M_3 = 
+\big(\tY \, f \, n \, . \, 
+\tif{n = 0}{0}{f \, (n - 1) \oplus_{2/3} f \, (n + 2)}\big) \, 10
+\]
+
+%While {x > 0} do {x := x-1 \oplus_2/3 x:= x+1}
+
+%M2 = While {x > 0} do {x := x-1 \oplus_2/3 x:= x+2}
+\end{enumerate}
+\end{example}
+
+\paragraph{Problem.} Find a generalised notion of ranking function 
+%--- call it \emph{antitone-convex rankable} --- 
+so that $M_1$ becomes rankable, and then prove it sound i.e.~rankable in this generalised sense implies AST.
+
+\paragraph{Idea.}
+
+The definition of antitone-convex ranking function $f$ (say) for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
+we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) and convex function $\epsilon : \pReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
+\[
+f(E[R’]) \leq f(E[R]) - \epsilon(f(E[R])) 
+\]
+where $R \to R’$ is the $\tY$-redex rule.
+
+\lo{An aside: This CBV $\tY$-rule seems cleaner:
+\[
+\big(\tY f^{A \to B} \, x^A \, . \, \theta^B \big) \, v \to \theta[(\tY f \, x \, . \, \theta) / f, v / x].
+\]
+We assume $\tY f^{A \to B} \, x^A \, . \, \theta^B$ is a value.}
+
+\begin{definition}
+We say that a supermartingale $(Y_n)_{n \in \omega}$ adapted to filtration $(\calF_n)_{n \in \omega}$ is \emph{antitone-convex strict} if $Y_n \geq 0$, and there exists an antitone and convex measurable function $\epsilon : \nnReal \to \pReal$ satisfying
+\[
+\expect{Y_{n+1} \mid \calF_n} \leq Y_n - (\epsilon \circ Y_n) \cdot 1_{\set{Y_n > 0}}
+\]
+%We say that the supermartingale is \emph{antitone-convex strict} if $\epsilon$ is further assumed to be convex.
+\end{definition}
+
+\begin{theorem}
+\label{thm:a-c strict}
+Let $(Y_n)_{n \in \omega}$ be a antitone-convex strict supermartingale via function $\epsilon : \nnReal \to \pReal$, and define stopping time $T(s) := \min \set{n \mid Y_n(s) = 0}$. 
+Then 
+\begin{enumerate}
+\item $T < \infty$ almost surely, and 
+\item $\expect{T} \leq \dfrac{\expect{Y_0}}{\epsilon_0}$ where $\epsilon_0 := \epsilon(\expect{Y_0})$.
+\end{enumerate}
+\end{theorem}
+
+\begin{proof}
+\emph{Claim}: for all $n \in \omega$,
+\(
+\expect{Y_n} \leq \expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n-1} \mathbb{P}[Y_i > 0]\Big).
+\)
+We prove the claim by induction on $n$.
+The base case is trivial.
+For the inductive case, we have
+\iffalse
+\everymath{\displaystyle}
+\begin{calculation}
+\int_A\nu(x,\dif y)\;\frac{\dif \nu'}{\dif\mu}(x,y)  
+  \step{decomposition of $\nu = \kact{\mu }{ f}$} 
+\int_A\mu(x,\dif y)\;f(x,y)\cdot \frac{\dif \nu'}{\dif\mu}(x,y)
+  \step{$f=1$ outside $\infty[\nu]$} 
+\int_A\mu(x,\dif y)\;\frac{\dif \nu'}{\dif\mu}(x,y)
+  \step{RN Theorem for $\sigma$-finite measures}
+\int_A\nu'(x,\dif y)
+  \step{definition of integral}
+\nu'(x,A).
+\end{calculation}
+\fi
+{%\everymath{displaystyle}
+\begin{calculation}
+\expect{Y_{n+1}}
+  \step[=]{definition of conditional expectation}
+\expect{\expect{Y_{n+1} \mid \calF_n}}
+  \step[\leq]{linearity of expectation}
+\expect{Y_n} - \int \epsilon \circ Y_n \cdot 1_{\set{Y_n > 0}} \, \dif \mu
+\; = \;
+\expect{Y_n} - \int_{\set{Y_n > 0}} \epsilon \circ Y_n \, \dif \mu
+  \step[\leq]{convexity of $\epsilon$, and Jensen's inequality}
+\expect{Y_n} - \epsilon \big(\int_{\set{Y_n > 0}} Y_n \, \dif \mu\big)
+  \step[\leq]{by assumption $Y_n \geq 0$}
+\expect{Y_n} - \epsilon (\expect{Y_n})
+  \step[\leq]{$\mathbb{P}[Y_n > 0] \leq 1$}
+\expect{Y_n} - \epsilon (\expect{Y_n}) \cdot \mathbb{P}[Y_n > 0]
+  \step[\leq]{$\epsilon$ is antitone, and $\epsilon_0 = \expect{Y_0} \geq \expect{Y_n}$}
+\expect{Y_n} - \epsilon_0 \cdot \mathbb{P}[Y_n > 0]
+  \step[\leq]{induction hypothesis}
+%\expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n-1} \mathbb{P}[Y_i > 0]\Big) - \epsilon_0 \cdot \mathbb{P}[Y_n > 0]
+%\step[\leq]{}
+\expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n} \mathbb{P}[Y_i > 0]\Big)
+\end{calculation}
+}
+
+\iffalse
+\begin{align*}
+\expect{Y_{n+1}}
+&=
+\expect{\expect{Y_{n+1} \mid \calF_n}}\\
+&\leq
+\expect{Y_n} - \expect{\epsilon \circ Y_n} \cdot \mathbb{P}[Y_n > 0]\\
+&\leq
+\expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n-1} \mathbb{P}[Y_i > 0]\Big)
+- \expect{\epsilon \circ Y_n} \cdot \mathbb{P}[Y_n > 0]\\
+&\leq
+\expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n} \mathbb{P}[Y_i > 0]\Big)
+\end{align*}
+\fi
+
+It follows from the claim that $\sum_{i=0}^{\infty} \mathbb{P}[Y_i > 0]$ converges; hence $\lim_{n \to \infty} \mathbb{P}[Y_n > 0] = 0$ and so $\mathbb{P}[T < \infty] = 1$.
+
+For 2, it suffices to prove: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=1} \mathbb{P}[T \geq i]$.
+\end{proof}
+
+\paragraph{Results to prove:}
+\begin{enumerate}
+\item (AST Soundness) If a closed SPCF term $M$ is antitone-convex rankable, then $T^{\tY}_M < \infty$ a.s.~(equivalently $M$ is AST), and $\expect{T^{\tY}_M} < \infty$.
+
+\item A partial ranking function theorem for antitone-convex ranking functions.
+\end{enumerate}
+
+\paragraph{\Cref{ex:ac-ranking} revisited}
+
+Programs $M_1$ and $M_3$ are AST.
+To prove the existence of antitone-convex ranking functions, we appeal \lo{(hopefully)} to a partial ranking function theorem for antitone-convex ranking functions.
+
+\lo{NOTE. Suppose a function $f : E \to \Real$ from a convex subset $E \subseteq \Real^n$ has continuous second-order partial derivatives. Then it is convex iff its Hessian is positive semidefinite.}
+
+For $M_1$, consider the function $g_1(n) := \frac{2^n - 1}{2^{n-1}}$.
+Note that $(g_1 (n))_{n \geq 0}$ satisfies the recurrence relation
+\[
+z_0 = 0, \qquad
+z_n > 1/2 \cdot z_{n-1} - 1/2 \cdot z_{n+1} \quad (n \geq 1)
+\]
+and set the antitone and convex function $\epsilon_1$ to be $x \mapsto g_1(x) - 1/2 \cdot g_1(x-1) - 1/2 \cdot g_1(x+1) = \frac{1}{2^{x+1}}$.
+
+For $M_3$, we use the function $g_3(n) := \frac{3^n - 1}{3^{n-1}}$, and set $\epsilon_3$ to be $x \mapsto g_3(x) - 2/3 \cdot g_3(x-1) - 1/3 \cdot g_3(x+2) = 
+\frac{28}{3^{x+2}}$.
+
+\subparagraph{Questions.} 
+\begin{enumerate}
+\item Can we remove the convexity assumption?
+\item Complete method for AST?
+
+\lo{Example~\ref{ex:0 probability reachable} seems somewhat artificial: its validity depends on $I$ being the closed interval. 
+There is no reason why we can't define $I$ to be the open interval $(0, 1)$, in which case, $Rch(M)$ would consist of all terms that are reachable from $M$ with non-zero probability.}
+\end{enumerate}
 

### 2020-05-27 ranking-supermartingales.tex

@@
 
-Intuitively $Y_n$ is the rank of the program after $n$ steps of computation.
-(Assume that the rank of a term is 0 iff it is a value.) In a $\epsilon$-ranking supermartingale, each computation step causes a strict decrease in rank, provided the term being reduced is not a value.
+Intuitively $Y_n$ gives the rank of the program after $n$ steps of computation.
+(Assume that the rank of a term is 0 iff it is a value.) In a $\epsilon$-ranking supermartingale, each computation step causes a strict decrease in rank of at least $\epsilon$, provided the term in question is not a value.
 
@@
 \label{thm:rank-PAST}
-Let $(Y_n)_{n \in \omega}$ be a $\epsilon$-ranking supermartingale, and set $T(s) := \min \set{n \mid Y_n(s) = 0}$.
+Let $(Y_n)_{n \in \omega}$ be a $\epsilon$-ranking supermartingale, and define the r.v.~$T$ by $T(s) := \min \set{n \mid Y_n(s) = 0}$.
 
@@
 
+\iffalse
 \lo{We need to show that each $f(M_n)$ is integrable. LO claimed earlier that $0 \leq f(N) \leq f(M)$ for all $N \in \mathit{Rch}(M)$.
@@
 }
+\fi
 
@@
 
+As an immediate corollary of \Cref{lem:key rankable}, each $f(M_n)$ is integrable:
+\[
+\int_{S} \mu(\dif s) \, |f(M_n)(s)| 
+=
+\int_{S} \mu(\dif s) \, f(M_n)(s)
+\leq
+\int_{S} \mu(\dif s) \, f(M_0)(s)
+=
+f(M).
+\]
+
 \iffalse
@@
 
-\begin{corollary}
+\begin{theorem}[Soundness]
 \begin{enumerate}
@@
 \end{enumerate}
-\end{corollary}
+\end{theorem}
 

### 2020-05-27 probabilisticProgrammingMartingales.tex
Comment about Example (5) removed

--- 2020-05-27 probabilisticProgrammingMartingales.tex
@@
 \expect{Y_n} - \epsilon \big(\int_{\set{Y_n > 0}} Y_n \, \dif \mu\big)
-  \step[\leq]{by assumption $Y_n \geq 0$}
+  \step[=]{by assumption $Y_n \geq 0$}
 \expect{Y_n} - \epsilon (\expect{Y_n})
@@
 \]
-and set the antitone and convex function $\epsilon_1$ to be $x \mapsto g_1(x) - 1/2 \cdot g_1(x-1) - 1/2 \cdot g_1(x+1) = \frac{1}{2^{x+1}}$.
+and set the antitone and convex function $\epsilon_1$ to be 
+\[
+x \mapsto \big(g_1(x) - 1/2 \cdot g_1(x-1) - 1/2 \cdot g_1(x+1)\big) = \frac{1}{2^{x+1}}.
+\]
 
-For $M_3$, we use the function $g_3(n) := \frac{3^n - 1}{3^{n-1}}$, and set $\epsilon_3$ to be $x \mapsto g_3(x) - 2/3 \cdot g_3(x-1) - 1/3 \cdot g_3(x+2) = 
-\frac{28}{3^{x+2}}$.
+For $M_3$, we use the function $g_3(n) := \frac{3^n - 1}{3^{n-1}}$, and set $\epsilon_3$ to be 
+\[
+x \mapsto \big(g_3(x) - 2/3 \cdot g_3(x-1) - 1/3 \cdot g_3(x+2) \big) = 
+\frac{28}{3^{x+2}}.
+\]
 
@@
 
-\lo{Example~\ref{ex:0 probability reachable} seems somewhat artificial: its validity depends on $I$ being the closed interval. 
+%\lo{Example~\ref{ex:0 probability reachable} seems somewhat artificial: its validity depends on $I$ being the closed interval.
 There is no reason why we can't define $I$ to be the open interval $(0, 1)$, in which case, $Rch(M)$ would consist of all terms that are reachable from $M$ with non-zero probability.}

### 2020-05-28 probabilisticProgrammingMartingales.tex
Weaken the statement and correct the proof for antitone strict supermartingales.

@@
 \begin{definition}
-We say that a supermartingale $(Y_n)_{n \in \omega}$ adapted to filtration $(\calF_n)_{n \in \omega}$ is \emph{antitone-convex strict} if $Y_n \geq 0$, and there exists an antitone and convex measurable function $\epsilon : \nnReal \to \pReal$ satisfying
+We say that a supermartingale $(Y_n)_{n \in \omega}$ and a stopping time $T$ \akr{Separating the stopping time from $Y_n = 0$ allows for a neater definition of the ranking function of a term} adapted to filtration $(\calF_n)_{n \in \omega}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone measurable function $\epsilon : \nnReal \to \pReal$ satisfying
 \[
-\expect{Y_{n+1} \mid \calF_n} \leq Y_n - (\epsilon \circ Y_n) \cdot 1_{\set{Y_n > 0}}
+\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot 1_{\set{T > n}}
 \]
-%We say that the supermartingale is \emph{antitone-convex strict} if $\epsilon$ is further assumed to be convex.
 \end{definition}
@@
 \label{thm:a-c strict}
-Let $(Y_n)_{n \in \omega}$ be a antitone-convex strict supermartingale via function $\epsilon : \nnReal \to \pReal$, and define stopping time $T(s) := \min \set{n \mid Y_n(s) = 0}$. 
-Then 
-\begin{enumerate}
-\item $T < \infty$ almost surely, and 
-\item $\expect{T} \leq \dfrac{\expect{Y_0}}{\epsilon_0}$ where $\epsilon_0 := \epsilon(\expect{Y_0})$.
-\end{enumerate}
+Let $(Y_n)_{n \in \omega}$ be an antitone strict supermartingale via function $\epsilon : \nnReal \to \pReal$, and stopping time $T$. 
+Then $T < \infty$ almost surely.
+\akr{It's not actually true that $\expect{T} \leq \dfrac{\expect{Y_0}}{\epsilon_0}$ where $\epsilon_0 := \epsilon(\expect{Y_0})$.
+This is fortunate because it means that a rankability theorem based on antitone strict supermartingales actually has a chance of being stronger than the existing one (which was pretty much complete for $\tY$-PAST terms already).
+In fact, for any almost surely finite stopping time, there is a corresponding antitone strict supermartingale.
+This seems promising for the method being conditionally complete (for terms where all reachable terms are AST).}
 \end{theorem}
@@
 \begin{proof}
-\emph{Claim}: for all $n \in \omega$,
+First, as $(Y_n)$ is a supermartingale, $\expect{Y_n} \leq \expect{Y_0}$, therefore
+\begin{calculation}
+\expect{Y_n \mid T > n}
+  \step[=]{}
+\frac{\mathbb P[T > n] \expect{Y_n \mid T > n} + \mathbb P[T \leq n] \expect{Y_n \mid T \leq n} - \mathbb P[T \leq n] \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
+  \step[=]{}
+\frac{\expect{Y_n} - \mathbb P[T \leq n] \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
+  \step[\leq]{$Y_n \geq 0$ always}
+\frac{\expect{Y_n}}{\mathbb P[T > n]}
+  \step[\leq]{$(Y_n)_n$ is a supermartingale}
+\frac{\expect{Y_0}}{\mathbb P[T > n]}
+\end{calculation}
+
+\emph{Claim}: for all $0 < x \leq 1$, let $B_x = \left \lceil \frac{\expect{Y_0}+1}{x \epsilon(\expect{Y_0} x^{-1})} \right \rceil$, then
 \(
-\expect{Y_n} \leq \expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n-1} \mathbb{P}[Y_i > 0]\Big).
+\mathbb P[T > B_x] \leq x.
 \)
-We prove the claim by induction on $n$.
-The base case is trivial.
-For the inductive case, we have
-\iffalse
-\everymath{\displaystyle}
+As the convex hull of $\epsilon$ (the greatest convex function less than or equal to it) satisfies all the conditions assumed of $\epsilon$, in addition to being convex, assume wlog that $\epsilon$ is convex.
+Suppose for a contradiction that $\mathbb P[T > B_x] > x$, then for every $n \leq B_x$
 \begin{calculation}
-\int_A\nu(x,\dif y)\;\frac{\dif \nu'}{\dif\mu}(x,y)  
-  \step{decomposition of $\nu = \kact{\mu }{ f}$} 
-\int_A\mu(x,\dif y)\;f(x,y)\cdot \frac{\dif \nu'}{\dif\mu}(x,y)
-  \step{$f=1$ outside $\infty[\nu]$} 
-\int_A\mu(x,\dif y)\;\frac{\dif \nu'}{\dif\mu}(x,y)
-  \step{RN Theorem for $\sigma$-finite measures}
-\int_A\nu'(x,\dif y)
-  \step{definition of integral}
-\nu'(x,A).
+\expect{Y_n-Y_{n+1}}
+  \step[=]{definition and linearity of conditional expectation}
+\expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
+  \step[\geq]{antitone strict assumption}
+\expect{\epsilon(Y_n) \cdot 1_{\set{T > n}}}
+  \step[=]{definition of conditional expectation for a discrete variable}
+\mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
+  \step[\geq]{Jensen's inequality}
+\mathbb P[T > n]\ \epsilon(\expect{Y_n \mid T > n})
+  \step[\geq]{proved earlier}
+\mathbb P[T > n]\ \epsilon(\frac{\expect{Y_0}}{\mathbb P[T > n]})
+  \step[>]{assumption, $\mathbb P[T > n] \geq \mathbb P[T > B_x] > x$}
+x\ \epsilon(\expect{Y_0} x^{-1})
 \end{calculation}
-\fi
-{%\everymath{displaystyle}
-\begin{calculation}
-\expect{Y_{n+1}}
-  \step[=]{definition of conditional expectation}
-\expect{\expect{Y_{n+1} \mid \calF_n}}
-  \step[\leq]{linearity of expectation}
-\expect{Y_n} - \int \epsilon \circ Y_n \cdot 1_{\set{Y_n > 0}} \, \dif \mu
-\; = \;
-\expect{Y_n} - \int_{\set{Y_n > 0}} \epsilon \circ Y_n \, \dif \mu
-  \step[\leq]{convexity of $\epsilon$, and Jensen's inequality}
-\expect{Y_n} - \epsilon \big(\int_{\set{Y_n > 0}} Y_n \, \dif \mu\big)
-  \step[=]{by assumption $Y_n \geq 0$}
-\expect{Y_n} - \epsilon (\expect{Y_n})
-  \step[\leq]{$\mathbb{P}[Y_n > 0] \leq 1$}
-\expect{Y_n} - \epsilon (\expect{Y_n}) \cdot \mathbb{P}[Y_n > 0]
-  \step[\leq]{$\epsilon$ is antitone, and $\epsilon_0 = \expect{Y_0} \geq \expect{Y_n}$}
-\expect{Y_n} - \epsilon_0 \cdot \mathbb{P}[Y_n > 0]
-  \step[\leq]{induction hypothesis}
-%\expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n-1} \mathbb{P}[Y_i > 0]\Big) - \epsilon_0 \cdot \mathbb{P}[Y_n > 0]
-%\step[\leq]{}
-\expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n} \mathbb{P}[Y_i > 0]\Big)
-\end{calculation}
-}
+therefore, by a telescoping sum, $\expect{Y_{B_x}} = \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0 - B_x x \epsilon(\expect{Y_0} x^{-1})} \leq -1 < 0$, which is a contradiction, therefore the claim must be true, therefore $P[T > n]$ tends to 0 as $n$ tends to infinity, therefore $T < \infty$ almost surely.
 
-\iffalse
-\begin{align*}
-\expect{Y_{n+1}}
-&=
-\expect{\expect{Y_{n+1} \mid \calF_n}}\\
-&\leq
-\expect{Y_n} - \expect{\epsilon \circ Y_n} \cdot \mathbb{P}[Y_n > 0]\\
-&\leq
-\expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n-1} \mathbb{P}[Y_i > 0]\Big)
-- \expect{\epsilon \circ Y_n} \cdot \mathbb{P}[Y_n > 0]\\
-&\leq
-\expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n} \mathbb{P}[Y_i > 0]\Big)
-\end{align*}
-\fi
-
-It follows from the claim that $\sum_{i=0}^{\infty} \mathbb{P}[Y_i > 0]$ converges; hence $\lim_{n \to \infty} \mathbb{P}[Y_n > 0] = 0$ and so $\mathbb{P}[T < \infty] = 1$.
-
-For 2, it suffices to prove: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=1} \mathbb{P}[T \geq i]$.
 \end{proof}

### 2020-05-28 probabilisticProgrammingMartingales.tex
cosmetic changes to proof of antitone strict supermartingale theorem

@@
 \begin{definition}
-We say that a supermartingale $(Y_n)_{n \in \omega}$ and a stopping time $T$ \akr{Separating the stopping time from $Y_n = 0$ allows for a neater definition of the ranking function of a term} adapted to filtration $(\calF_n)_{n \in \omega}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone measurable function $\epsilon : \nnReal \to \pReal$ satisfying
+We say that a supermartingale $(Y_n)_{n \in \omega}$ and a stopping time $T$ \akr{Separating the stopping time from $Y_n = 0$ allows for a neater definition of the ranking function of a term} \lo{Nice!} adapted to filtration $(\calF_n)_{n \in \omega}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone measurable function $\epsilon : \nnReal \to \pReal$ satisfying
 \[
@@
 
+\lo{Given a (super)martingale $(Y_n)_{n \in \omega}$ and an a.s.~finite stopping time $T$ (meaning: for some $n \in \omega$, $T \leq n$ a.s.) both adapted to the same filtration, there exist supermartingale $(Y_n')_{n \in \omega}$ and antitone function $\epsilon$ such that $(Y_n')_{n \in \omega}$ is an antitone strict supermartingale w.r.t.~$(\epsilon, T)$.}
+
 \begin{proof}
 First, as $(Y_n)$ is a supermartingale, $\expect{Y_n} \leq \expect{Y_0}$, therefore
+
 \begin{calculation}
@@
   \step[=]{}
-\frac{\mathbb P[T > n] \expect{Y_n \mid T > n} + \mathbb P[T \leq n] \expect{Y_n \mid T \leq n} - \mathbb P[T \leq n] \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
+\dfrac{\mathbb P[T > n] \expect{Y_n \mid T > n} + \mathbb P[T \leq n] \expect{Y_n \mid T \leq n} - \mathbb P[T \leq n] \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
   \step[=]{}
-\frac{\expect{Y_n} - \mathbb P[T \leq n] \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
+\dfrac{\expect{Y_n} - \mathbb P[T \leq n] \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
   \step[\leq]{$Y_n \geq 0$ always}
-\frac{\expect{Y_n}}{\mathbb P[T > n]}
+\dfrac{\expect{Y_n}}{\mathbb P[T > n]}
   \step[\leq]{$(Y_n)_n$ is a supermartingale}
-\frac{\expect{Y_0}}{\mathbb P[T > n]}
+\dfrac{\expect{Y_0}}{\mathbb P[T > n]}
 \end{calculation}
 
-\emph{Claim}: for all $0 < x \leq 1$, let $B_x = \left \lceil \frac{\expect{Y_0}+1}{x \epsilon(\expect{Y_0} x^{-1})} \right \rceil$, then
+\emph{Claim}: for all $0 < x \leq 1$, let $B_x = \left \lceil \dfrac{\expect{Y_0}+1}{x \; \epsilon(\expect{Y_0} \, x^{-1})} \right \rceil$, then
 \(
@@
 Suppose for a contradiction that $\mathbb P[T > B_x] > x$, then for every $n \leq B_x$
+\lo{Change of justification of the second equality below: event, not discrete r.v. 
+Recall: expectation conditioning on r.v.~(resp.~event) is a r.v.~(resp.~number).}
 \begin{calculation}
@@
 \expect{\epsilon(Y_n) \cdot 1_{\set{T > n}}}
-  \step[=]{definition of conditional expectation for a discrete variable}
+%  \step[=]{definition of conditional expectation for a discrete variable}
+\step[=]{definition of expectation conditioning on an \emph{event}}
 \mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
@@
   \step[\geq]{proved earlier}
-\mathbb P[T > n]\ \epsilon(\frac{\expect{Y_0}}{\mathbb P[T > n]})
+\mathbb P[T > n]\ \epsilon\Big(\dfrac{\expect{Y_0}}{\mathbb P[T > n]}\Big)
   \step[>]{assumption, $\mathbb P[T > n] \geq \mathbb P[T > B_x] > x$}
-x\ \epsilon(\expect{Y_0} x^{-1})
+x \; \epsilon\big(\expect{Y_0} \, x^{-1}\big).
 \end{calculation}
-therefore, by a telescoping sum, $\expect{Y_{B_x}} = \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0 - B_x x \epsilon(\expect{Y_0} x^{-1})} \leq -1 < 0$, which is a contradiction, therefore the claim must be true, therefore $P[T > n]$ tends to 0 as $n$ tends to infinity, therefore $T < \infty$ almost surely.
-
+Therefore, by a telescoping sum
+\[
+\expect{Y_{B_x}} = \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1}) \leq -1 < 0 
+\]
+%\lo{$\expect{Y_0 - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})}$}
+which is a contradiction, therefore the claim must be true, therefore $P[T > n]$ tends to 0 as $n$ tends to infinity, therefore $T < \infty$ almost surely.
 \end{proof}

### 2020-05-30 probabilisticProgrammingMartingales.tex
Examples (random walk) to illustrate the use of (partial) antitone ranking functions

@@
 \paragraph{}
-In both of these cases, there are only some values of the rankng function that are semantically important. Define a \emph{partial ranking function} on a closed term $M$ to be a partial function 
+In both of these cases, there are only some values of the ranking function that are semantically important. Define a \emph{partial ranking function} on a closed term $M$ to be a partial function 
 $f : Rch(M) \rightharpoonup \mathbb R$ 
@@
 
-\section{Antitone-convex ranking functions}
+\section{Antitone ranking functions}
 
@@
 \big(\tY \, f \, n \, . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}\big) \, 1
+\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}\big) \, 10
 \]
@@
 Programs $M_1$ and $M_3$ are AST.
-To prove the existence of antitone-convex ranking functions, we appeal \lo{(hopefully)} to a partial ranking function theorem for antitone-convex ranking functions.
+To construct antitone ranking functions, we appeal \lo{(hopefully)} to a partial ranking function theorem for antitone ranking functions.
+\iffalse
+\lo{NOTE. Suppose a function $f : E \to \Real$ from a convex subset $E \subseteq \Real^n$ has continuous second-order partial derivatives. Then it is convex iff its Hessian is positive semidefinite.}
+\fi
+\iffalse
+For $M_1$, consider the function $g_1(n) := \frac{2^n - 1}{2^{n-1}}$.
+\fi
 
-\lo{NOTE. Suppose a function $f : E \to \Real$ from a convex subset $E \subseteq \Real^n$ has continuous second-order partial derivatives. Then it is convex iff its Hessian is positive semidefinite.}
+For $M_1$, define the function $g_1 : \nnReal \to [0, 2)$ by $g_1(x) = 2 - \frac{1}{2^{x-1}}$.
+Note that $(g_1 (n))_{n \in \omega}$ satisfies the recurrence relation
+\[
+\left\{
+\begin{array}{rll}
+z_0 &=& 0\\
+n \geq 1, \quad z_n &>& \dfrac{1}{2} \, \big( z_{n-1} + z_{n+1}\big).
+\end{array}
+\right.
+\]
+Set the antitone function $\epsilon_1 (y) :=  \frac{2-y}{4}$.
+Then
+\begin{equation}
+\epsilon_1(g_1(x)) = g_1(x) - \dfrac{1}{2} \, \big(g_1(x-1) + g_1(x+1)\big) = \dfrac{1}{2^{x+1}}
+\label{eqn:epsilon g M1}
+\end{equation}
+(Since $g_1$ is bijective, with inverse $g_1^{-1}(y) = 1 - \log (2-y)$, we can obtain $\epsilon_1(y)$ from \Cref{eqn:epsilon g M1}.) 
+%We have $1/2 \cdot \big(g_1(x-1) + g_1(x+1)\big) = 2 - \frac{5}{2^{x+1}}$.
 
-For $M_1$, consider the function $g_1(n) := \frac{2^n - 1}{2^{n-1}}$.
-Note that $(g_1 (n))_{n \geq 0}$ satisfies the recurrence relation
+Using shorthand 
+$\Theta_1 = \tY \, f \, n \, . \, 
+\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
+we can define a partial antitone ranking function $f_1 : \mathit{Rch}(M_1) \to \nnReal$ with antitone function $\epsilon_1$ defined as follows:
+\begin{align*}
+{\Theta_1} \, n 
+&\mapsto 
+g_1(n)
+\\
+\tif{n = 0}{0}{\Theta_1 \, (n - 1) \oplus_{1/2} \Theta_1 \, (n + 1)}
+&\mapsto
+\dfrac{1}{2} \, \big(g_1(n-1) + g_1(n+1)\big)
+\\
+0 &\mapsto 0
+\end{align*}
+
+For $M_3$, we use the function $g_3(x) := 3 - \frac{1}{3^{x-1}}$ and the antitone function $\epsilon_3(y) := \frac{84-28y}{27}$. 
+Then
 \[
-z_0 = 0, \qquad
-z_n > 1/2 \cdot z_{n-1} - 1/2 \cdot z_{n+1} \quad (n \geq 1)
-\]
-and set the antitone and convex function $\epsilon_1$ to be 
-\[
-x \mapsto \big(g_1(x) - 1/2 \cdot g_1(x-1) - 1/2 \cdot g_1(x+1)\big) = \frac{1}{2^{x+1}}.
-\]
-
-For $M_3$, we use the function $g_3(n) := \frac{3^n - 1}{3^{n-1}}$, and set $\epsilon_3$ to be 
-\[
-x \mapsto \big(g_3(x) - 2/3 \cdot g_3(x-1) - 1/3 \cdot g_3(x+2) \big) = 
+\epsilon_3(g_3(x)) = g_3(x) - \dfrac{1}{3}\big(2 \cdot g_3(x-1) + \cdot g_3(x+2) \big) = 
 \frac{28}{3^{x+2}}.
 \]
+Following $M_1$ above, we can define a partial antitone ranking function $f_3 : \mathit{Rch}(M_3) \to \nnReal$ with $\epsilon_3$ as the antitone function.
+
+\begin{example}[Continuous random walk]\label{ex:raven complex}
+In SPCF we can construct a function whose argument changes by a random amount at each recursive call:
+\[
+\underbrace{\big
+(\tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample))} \big)}_{\Theta} \, 10
+\]
+We can construct a partial antitone ranking function $f$ as follows:
+\begin{align*}
+\Theta \, l 
+&\mapsto 
+l
+\\
+\tif{l \leq 0}{0}{\Theta \, (l - \tsample)}
+&\mapsto
+l - 1/2
+\\
+0 &\mapsto 0
+\end{align*}
+with the constant antitone function $\epsilon(y) = 1/2$.
+
+For a more complex example, consider the following ``continuous random walk''
+\[
+\underbrace{\big
+(\tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample) \oplus_{1/2} f(x - \tsample)} \big)}_{\Xi} \, 10
+\]
+Let $g(x) := \sqrt{2} - \frac{1}{2^{x-\frac{1}{2}}}$ and let $\epsilon$ be the function specified by
+\[
+\epsilon(g(x)) = g(x) - \dfrac{1}{2}\Big(g(x - \dfrac{1}{2}) + \cdot g(x + \dfrac{1}{2}) \Big) = 
+\dfrac{3 - 2\sqrt{2}}{2^{x+1}}.
+\]
+We define a partial antitone ranking function as follows:
+\begin{align*}
+\Xi \, l 
+&\mapsto 
+g(l)
+\\
+\tif{l \leq 0}{0}{\Xi \, (l - \tsample) \oplus_{1/2} \Xi \, (l + \tsample)}
+&\mapsto
+\frac{1}{2}\big(g(l - \frac{1}{2}) + \cdot g(l + \frac{1}{2}) \big)
+\\
+r \leq 0, \quad r &\mapsto 0
+\end{align*}
+\end{example}
 

### 2020-06-01 probabilisticProgrammingMartingales.tex
Added 1. Proof of finiteness of Y-stopping times T_0, T_1, T_2, etc. 2. Relativised completeness conjecture

@@
 The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction).
+\lo{TODO. The preceding sentence deserves a proof.}
 \end{proof}
@@
 This seems promising for the method being conditionally complete (for terms where all reachable terms are AST).}
+\lo{See \Cref{sec:Relativised completeness}.}
 \end{theorem}
-
-\lo{Given a (super)martingale $(Y_n)_{n \in \omega}$ and an a.s.~finite stopping time $T$ (meaning: for some $n \in \omega$, $T \leq n$ a.s.) both adapted to the same filtration, there exist supermartingale $(Y_n')_{n \in \omega}$ and antitone function $\epsilon$ such that $(Y_n')_{n \in \omega}$ is an antitone strict supermartingale w.r.t.~$(\epsilon, T)$.}
 
@@
 \begin{enumerate}
-\item (AST Soundness) If a closed SPCF term $M$ is antitone-convex rankable, then $T^{\tY}_M < \infty$ a.s.~(equivalently $M$ is AST), and $\expect{T^{\tY}_M} < \infty$.
+\item (AST Soundness) If a closed SPCF term $M$ is antitone-convex rankable, then $T^{\tY}_M < \infty$ a.s.~(equivalently $M$ is AST), and $\expect{T^{\tY}_M} < \infty$. \lo{This should be fine.}
 
-\item A partial ranking function theorem for antitone-convex ranking functions.
+\item A partial ranking function theorem for antitone-convex ranking functions. \lo{TODO}
 \end{enumerate}
@@
 
-\subparagraph{Questions.} 
-\begin{enumerate}
-\item Can we remove the convexity assumption?
-\item Complete method for AST?
+\input{y-stopping-time}
 
-%\lo{Example~\ref{ex:0 probability reachable} seems somewhat artificial: its validity depends on $I$ being the closed interval.
-There is no reason why we can't define $I$ to be the open interval $(0, 1)$, in which case, $Rch(M)$ would consist of all terms that are reachable from $M$ with non-zero probability.}
-\end{enumerate}
+\input{relativised-completeness}
 

### 2020-06-01 relativised-completeness.tex
Added 1. Proof of finiteness of Y-stopping times T_0, T_1, T_2, etc. 2. Relativised completeness conjecture

% !TEX root = ./probabilisticProgrammingMartingales.tex

\section{Relativised completeness}
\label{sec:Relativised completeness}

If every term reachable from a closed SPCF program is AST then the program is also AST, but not conversely 
(because a diverging term may be reachable with probability 0 from an AST term).

We aim to prove the following conjecture:

\begin{quote}
\emph{Given a closed SPCF term $M$, if every term reachable from $M$ is AST then $M$ admits an antitone ranking function.}
\end{quote}

\begin{enumerate}

\item Given a (super)martingale $(Y_n)_{n \in \omega}$ and an a.s.~finite stopping time $T$ (meaning: for some $n \in \omega$, $T \leq n$ a.s.) both adapted to the same filtration, there exist supermartingale $(Y_n')_{n \in \omega}$ and antitone function $\epsilon$ such that $(Y_n')_{n \in \omega}$ is an antitone strict supermartingale w.r.t.~$(\epsilon, T)$.
\lo{Is this what you had in mind?}

\item Let $(Y_n')_{n \in \omega}$ be an antitone strict supermartingale w.r.t.~$(\epsilon, T)$. 
Then $M$ admits an antitone ranking function.
\end{enumerate}

For 2, note that for every $N \in \mathit{Rch}(M)$, there exists $n \geq 0$ and $s \in S$ such that $\red^n(M, s) = (N, -)$.
Define a function $f : \mathit{Rch}(M) \to \nnReal$ by $f(N) := Y_n'(s)$.

\lo{Ongoing.}

### 2020-06-01 y-stopping-time.tex
Added 1. Proof of finiteness of Y-stopping times T_0, T_1, T_2, etc. 2. Relativised completeness conjecture

% !TEX root=./probabilisticProgrammingMartingales.tex

\newcommand\transform[1]{{\ulcorner{#1}\urcorner}}

\section{$\tY$ stopping times are finite}

As before, fix a $M \in \Lambda^0$.
Recall the random variables on $(S, \calF, \mu)$:
\begin{align*}
T_0(s) & := 0 \\
T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value or of form } E[\tY \lambda x. N] \}
\end{align*}

\begin{lemma}
$T_0, T_1, T_2, \cdots$ are an increasing sequence of stopping times adapted to $(\calF_n)_{n \in \omega}$, and each $T_i$ is bounded.
\end{lemma}

We first show that $T_1$ is bounded i.e.~$T_1 \leq n$, for some $n \in \omega$.

A first proof idea is to construct a reduction argument to the strong normalisation of the simply-typed lambda-calculus.
 There is a classical transform of conditionals into the pure lambda-calculus.
However, each evaluation of $\tsample$ (almost surely) returns a different number, which complicates such a transform.

Given a closed SPCF term $M$, we transform it to a term $\transform{M}$ of the nondeterministic simply-typed lambda-calculus as follows.
(We may assume that the nondeterministic calculus is generated from a base type $\iota$ with a $\iota$-type constant symbol $r$, and a function symbol $\bot : A$  for each function type $A$.)
\iffalse
\begin{enumerate}
\item replace every $\tsample$ by $r$

\item replace each base-type subterm $f \, M_1 \cdots M_n$ by $(\lambda x_1 \cdots x_n . r) \, \transform{M_1} \cdots \transform{M_n}$ where $f$ is a primitive function

\item replace every subterm $\tif{B}{M_1}{M_2}$ by
$(\lambda x . \transform{M_1} + \transform{M_2}) \, \transform{B}$, for a fresh $x$

\item replace every $\tY \, f \, x . N$ by $\lambda x . \transform{N}[\bot / f]$

\item replace every $\tscore (N)$ by $\transform{N}$ (we set aside the requirement that $\tscore$ is undefined on negative numbers)
\end{enumerate}
\lo{The above rules can be defined formally.}
\fi
\begin{align*}
\transform{\tsample} &:= r\\
\transform{y} &:= y \\
\transform{\lambda y . N} &:= \lambda y . \transform{N}\\
\transform{M_1 \, M_2} &:= \transform{M_1} \, \transform{M_2} \\
n \geq 0, \; \transform{f \, M_1 \cdots M_n} &:= (\lambda z_1 \cdots z_n . r) \, \transform{M_1} \cdots \transform{M_n} \\
\transform{\tif{B}{M_1}{M_2}} &:= \big(\lambda z . (\transform{M_1} + \transform{M_2})\big) \, \transform{B} \\
\transform{\tY \, f \, y . N} &:= \lambda y . \transform{N}[\bot / f]\\
\transform{\tscore (N)} &:= \transform{N}
\end{align*}
In the above, variables $z, z_1, \cdots, z_n$ are assumed to be fresh; $f$ ranges over primitive functions and numerals.
In transforming $\tscore (N)$, we set aside the requirement that $\tscore$ is undefined on negative numbers.

The idea is that $\transform{M}$ captures the initial, non-recursive operational behaviour of $M$, so that $\transform{M}$ simulates the reduction of $M$ until the latter reaches a value, or a term of the form $E[\tY \lambda x. N]$.

It is straightforward to see that for every trace $s \in S$, there is a reduction sequence of $\transform{M}$ that simulates (an initial subsequene of) the reduction of $M$ under $s$;
and the simulation can be made lockstep.

Finally thanks to the following
\begin{theorem}[de Groote]
\label{thm:de groote}
The simply-typed non-deterministic lambda-calculus is strongly normalising. \qed
\end{theorem}
we have:

\begin{enumerate}
\item Every $\transform{M}$-reduction terminates on reaching $r$, or a term of the shape $E[\bot N_1 \cdots N_n]$.

\item Further there is a finite bound (say $l$) on the length of such $\transform{M}$-reduction sequences; and $l$ bounds the stopping time $T_1$.
\end{enumerate}

\lo{Note: \cite{DBLP:conf/lfcs/Groote94} considered a call-by-name version of a simply-typed nondeterministic lambda-calculus, and proved it strongly normalising, using a reducibility argument \`a la Tait.
One should check that it extends to the call-by-value calculus.
}

\iffalse
@inproceedings{DBLP:conf/lfcs/Groote94,
  author    = {Philippe de Groote},
  editor    = {Anil Nerode and
               Yuri V. Matiyasevich},
  title     = {Strong Normalization in a Non-Deterministic Typed Lambda-Calculus},
  booktitle = {Logical Foundations of Computer Science, Third International Symposium,
               LFCS'94, St. Petersburg, Russia, July 11-14, 1994, Proceedings},
  series    = {Lecture Notes in Computer Science},
  volume    = {813},
  pages     = {142--152},
  publisher = {Springer},
  year      = {1994},
  url       = {https://doi.org/10.1007/3-540-58140-5\_15},
  doi       = {10.1007/3-540-58140-5\_15},
  timestamp = {Tue, 14 May 2019 10:00:54 +0200},
  biburl    = {https://dblp.org/rec/conf/lfcs/Groote94.bib},
  bibsource = {dblp computer science bibliography, https://dblp.org}
}
\fi

### 2020-06-05 relativised-completeness.tex
Add incomplete proof of completeness of assms.

@@
 
-\item Given a (super)martingale $(Y_n)_{n \in \omega}$ and an a.s.~finite stopping time $T$ (meaning: for some $n \in \omega$, $T \leq n$ a.s.) both adapted to the same filtration, there exist supermartingale $(Y_n')_{n \in \omega}$ and antitone function $\epsilon$ such that $(Y_n')_{n \in \omega}$ is an antitone strict supermartingale w.r.t.~$(\epsilon, T)$.
+\item Given an a.s.~finite stopping time $T$ (meaning: $T < \infty$ a.s.) adapted to a filtration, there exist supermartingale $(Y_n)_{n \in \omega}$ adapted to the same filtration and antitone function $\epsilon$ such that $(Y_n)_{n \in \omega}$ is an antitone strict supermartingale w.r.t.~$(\epsilon, T)$.
 \lo{Is this what you had in mind?}
@@
 
+For 1, consider the following:
+\begin{proof}
+If $T$ is bounded by $b$ a.s.~(for $n$ constant), the statement is trivially true by taking $\epsilon$ to be constantly 1, and $Y_n = b - n$.
+Otherwise, let $(T_n)_{n \in \omega}$ be defined recursively such that $T_0 = 0$, $T_{n+1} > T_n$ and $\mathbb P[T < T_{n+1}] \leq \frac 1 4 \mathbb P[T < T_n]$.
+We will then define a supermartingale $(Y_n)_{n \in \omega}$ and nonrandom sequence $(Y'_n)_{n \in \omega}$ such that $Y_n = 0$ iff $T < n$, $Y_n = Y'_n$ iff $T \geq n$, and $Y'_n \geq 2^k$ for $n \geq T_k$.
+
+$\epsilon$ is defined piecewise and recursively in such a way as to force all the necessary constraints to hold:
+\begin{align*}
+    \text{for }x \in [0,1),\ &\epsilon(x) = 1 \\
+    \text{for }x \in [2^k,2^{k+1}),\ &\epsilon(x) = \max \left (\epsilon(2^{k-1}), \frac{2^k}{T_{k+1}-T_k} \right ).
+\end{align*}
+This ensures that
+\begin{itemize}
+\item $\epsilon$ is weakly decreasing.
+\item $\epsilon$ is bounded below by 0.
+\item For $x \geq 2^k$, $\epsilon(x) \leq \frac{2^k}{T_{k+1}-T_k}$.
+\end{itemize}
+
+$Y'_n$ is then defined recursively by
+\begin{align*}
+    Y'_0 = 2,
+    Y'_{n+1} = (Y'_n - \epsilon(Y'_n)) \frac{\mathbb P[T > n]}{\mathbb P[T > n+1]}.
+\end{align*}
+
+The fact $Y'_{T_k} \geq 2^{k+1}$ is proven by induction on $k$.
+For base case, $2 \geq 2$, as required.
+For the inductive case $k+1$, we first do another induction to prove that $Y'_n \geq 2^k$ for all $T_k \leq n \leq T_{k+1}$.
+The base case of this inner induction follows from the outer induction hypothesis a fortiori.
+take the greatest $k$ such that $n > T_k$.
+By the induction hypothesis, $Y'_{T_k} \geq 2^k$.
+\begin{align*}
+    Y'_n & = Y'_{T_k} \frac{\mathbb P[T > T_k]}{\mathbb P[T > n]} - \sum_{m=T_k}^{n-1} \epsilon(Y'_m) \frac{\mathbb P[T > m+1]}{\mathbb P[T > n]} \\
+    & \geq \frac{\mathbb P[T > T_k]}{\mathbb P[T > n]} (Y'_{T_k} - \sum_{m=T_k}^{n-1} \epsilon(Y'_m)) \\
+    & \geq \frac{\mathbb P[T > T_k]}{\mathbb P[T > n]} (Y'_{T_k} - \sum_{m=T_k}^{n-1} \frac{2^k}{T_{k+1}-T_k}) \\
+    & = \frac{\mathbb P[T > T_k]}{\mathbb P[T > n]} (Y'_{T_k} - \frac{2^k (n - T_k)}{T_{k+1}-T_k}) \\
+    & \geq \frac{\mathbb P[T > T_k]}{\mathbb P[T > n]} (Y'_{T_k} - 2^k) \\
+    & \geq \frac{\mathbb P[T > T_k]}{\mathbb P[T > n]} (2^{k+1} - 2^k) \\
+    & = \frac{\mathbb P[T > T_k]}{\mathbb P[T > n]} 2^k \\
+    & \geq 2^k
+\end{align*}
+as required. Substituting $n = T_{k+1}$ and using the fact that $\frac{\mathbb P[T > T_k]}{\mathbb P[T > T_{k+1}]}$, the same reasoning gives $Y'_{T_{k+1}} \geq 2^{k+2}$, as required.
+
+Although the stronger induction hypothesis was necessary for the induction to work, the only reason this is needed is to prove that $Y'_n > 0$ for all n, which implies that $Y_n \geq 0$.
+The other condition for $(Y_n)$ to be an antitone-strict supermartingale is \akr{The inference from the third to the fourth line is actually incorrect if $\mathcal F$ is not coarsest filtration to which $T$ is adapted.
+This error may require a substantial change to the proof to correct, but I thought you'd want to see it anyway rather than waiting for me to fix it.} \\
+\begin{align*}
+    & \expect{Y_{n+1} \mid \mathcal F_n} \\
+    = & \expect{Y'_{n+1} 1_{T \geq n+1} \mid \mathcal F_n} \\
+    = & Y'_{n+1} \expect{1_{T \geq n+1} \mid \mathcal F_n} \\ 
+    = & Y'_{n+1} 1_{T \geq n} \frac{\mathbb P[T > n+1]}{\mathbb P[T > n]} \\
+    = & (Y'_n - \epsilon(Y'_n)) 1_{T \geq n} \\
+    = & Y_n - \epsilon(Y'_n) 1_{T \geq n}
+\end{align*}
+therefore $(Y_n)_n$ is an antitone strict supermartingale with respect to $(T,\epsilon)$.
+\end{proof}
 \lo{Ongoing.}

### 2020-06-05 probabilisticProgrammingMartingales.tex
More examples

@@
 \paragraph{}
-In both of these cases, there are only some values of the ranking function that are semantically important. Define a \emph{partial ranking function} on a closed term $M$ to be a partial function 
-$f : Rch(M) \rightharpoonup \mathbb R$ 
-%$f : Rch(M) \nrightarrow \mathbb R$ 
-such that
-%\lo{NOTATION. Common notation for partial function $\rightharpoonup$}
+In both of these cases, there are only some values of the ranking function that are semantically important.
+Define a \emph{partial ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \rightharpoonup \mathbb R$ such that
 \begin{itemize}
@@
 \begin{enumerate}
-\item (AST Soundness) If a closed SPCF term $M$ is antitone-convex rankable, then $T^{\tY}_M < \infty$ a.s.~(equivalently $M$ is AST), and $\expect{T^{\tY}_M} < \infty$. \lo{This should be fine.}
+\item (AST Soundness) If a closed SPCF term $M$ is antitone rankable, then $T^{\tY}_M < \infty$ a.s.~(equivalently $M$ is AST). \lo{This should be fine.}
 
@@
 \underbrace{\big
-(\tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample) \oplus_{1/2} f(x - \tsample)} \big)}_{\Xi} \, 10
+(\tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample) \oplus_{1/2} f(x + \tsample)} \big)}_{\Xi} \, 10
 \]
@@
 \end{align*}
+\end{example}
+
+\begin{example}[Fair-in-the-limit random walk]
+\label{ex:Fair-in-the-limit random walk}\citep[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}
+\[
+\underbrace{\big
+(\tY \, f \, x . 
+\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{n}{2n+1}} f(x + 1)} \big)}_{\Xi} 
+\, 10
+\]
+To construct an antitone ranking function, we solve the recurrence relation:
+\[
+\left\{
+\begin{array}{rll}
+z_0 &=& 0\\
+n \geq 1, \quad z_n &>& \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
+\end{array}
+\right.
+\]
+It turns out the solution to Example $M_1$, $z_n = \frac{2^n - 1}{2^{n-1}}$, is nearly a solution in that
+
+For $M_1$, define the function $g_1 : \nnReal \to [0, 2)$ by $g_1(x) = 2 - \frac{1}{2^{x-1}}$.
+Note that $(g_1 (n))_{n \in \omega}$ satisfies the recurrence relation
+Set the antitone function $\epsilon_1 (y) :=  \frac{2-y}{4}$.
+Then
+\begin{equation}
+\epsilon_1(g_1(x)) = g_1(x) - \dfrac{1}{2} \, \big(g_1(x-1) + g_1(x+1)\big) = \dfrac{1}{2^{x+1}}
+\label{eqn:epsilon g M1}
+\end{equation}
+
 \end{example}

### 2020-06-06 probabilisticProgrammingMartingales.tex
Some comments and questions about the incomplete completeness proof.

@@
 (\tY \, f \, x . 
-\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{n}{2n+1}} f(x + 1)} \big)}_{\Xi} 
+\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)}_{\Xi} 
 \, 10
@@
 \]
-It turns out the solution to Example $M_1$, $z_n = \frac{2^n - 1}{2^{n-1}}$, is nearly a solution in that
-
-For $M_1$, define the function $g_1 : \nnReal \to [0, 2)$ by $g_1(x) = 2 - \frac{1}{2^{x-1}}$.
-Note that $(g_1 (n))_{n \in \omega}$ satisfies the recurrence relation
-Set the antitone function $\epsilon_1 (y) :=  \frac{2-y}{4}$.
+It turns out the solution to Example $M_1$, $z_n = \frac{2^n - 1}{2^{n-1}}$, is nearly a solution (for $n = 1$, we have equality instead).
+
+Let $g(x) = \frac{2^n - 1}{2^{n-1}}$.
 Then 
 \begin{equation}
-\epsilon_1(g_1(x)) = g_1(x) - \dfrac{1}{2} \, \big(g_1(x-1) + g_1(x+1)\big) = \dfrac{1}{2^{x+1}}
-\label{eqn:epsilon g M1}
+\epsilon(g(x)) = 
+g(x) - \dfrac{x}{2x + 1} \, g(x-1) - \dfrac{x+1}{2x + 1} \, g(x+1) 
+= \dfrac{x-1}{2^{x} \, (2x + 1)}
+\label{eqn:in the limit}
 \end{equation}
 
+\lo{ongoing}
+
+\end{example}
+
+\begin{example}[Escaping spline]
+\label{ex:escaping spline}\citep[\S 5.4]{DBLP:journals/pacmpl/McIverMKK18}
+\[
+\underbrace{\big
+(\tY \, f \, x . 
+\tif{x \leq 0}{0}{0 \oplus_{\frac{1}{x+1}} f(x + 1)} \big)}_{\Xi} 
+\, 10
+\]
 \end{example}

### 2020-06-06 relativised-completeness.tex
Some comments and questions about the incomplete completeness proof.

@@
 \item Given an a.s.~finite stopping time $T$ (meaning: $T < \infty$ a.s.) adapted to a filtration, there exist supermartingale $(Y_n)_{n \in \omega}$ adapted to the same filtration and antitone function $\epsilon$ such that $(Y_n)_{n \in \omega}$ is an antitone strict supermartingale w.r.t.~$(\epsilon, T)$.
-\lo{Is this what you had in mind?}
 
@@
 Otherwise, let $(T_n)_{n \in \omega}$ be defined recursively such that $T_0 = 0$, $T_{n+1} > T_n$ and $\mathbb P[T < T_{n+1}] \leq \frac 1 4 \mathbb P[T < T_n]$.
-We will then define a supermartingale $(Y_n)_{n \in \omega}$ and nonrandom sequence $(Y'_n)_{n \in \omega}$ such that $Y_n = 0$ iff $T < n$, $Y_n = Y'_n$ iff $T \geq n$, and $Y'_n \geq 2^k$ for $n \geq T_k$.
+\lo{Since $T$ is a stopping time, we can assume that $T_n \in \mathbb N$.
+Problem with $T_1$: we require $T_1 > 0$ which may contradict the requirement $\mathbb P[T < T_{1}] \leq \frac 1 4 \mathbb P[T < 0] = 0$ for certain $T$, such as those that satisfy $\mathbb P[T < 1] > 0$.
+We could define $T_1$ as the least number satisfying $\mathbb P[T < T_1] \geq \frac{1}{4}$.}
+We will then define a supermartingale $(Y_n)_{n \in \omega}$ and nonrandom sequence $(Y'_n)_{n \in \omega}$ such that $Y_n = 0$ iff $T < n$ \changed[lo]{a.s.}, $Y_n = Y'_n$ iff $T \geq n$ \changed[lo]{a.s.}, and $Y'_n \geq 2^k$ for $n \geq T_k$.
 
-$\epsilon$ is defined piecewise and recursively in such a way as to force all the necessary constraints to hold:
+\lo{$(T_n)_{n \in \omega}$ and $(Y'_n)_{n \in \omega}$ are $\omega$-sequences of numbers. 
+There is a convention that random variables are denoted by uppercase letters. 
+I suggest we write them as $(t_n)_{n \in \omega}$ and $(y'_n)_{n \in \omega}$ respectively:
+in this way, we emphasise that they are sequences of nunbers, and avoid any possible confusion with r.v.}
+
+The antitone function \lo{It is good practice not to begin a sentence with a maths expression / symbol.} $\epsilon$ is defined piecewise and recursively in such a way as to force all the necessary constraints to hold:
 \begin{align*}
@@
 \item For $x \geq 2^k$, $\epsilon(x) \leq \frac{2^k}{T_{k+1}-T_k}$. 
+\lo{This would imply $\frac{2^k}{T_{k+1}-T_k} \geq \epsilon(2^{k-1})$. Why?}
 \end{itemize}
@@
 \begin{align*}
-    Y'_0 = 2,
+    Y'_0 = 2, \quad
     Y'_{n+1} = (Y'_n - \epsilon(Y'_n)) \frac{\mathbb P[T > n]}{\mathbb P[T > n+1]}.

### 2020-06-06 relativised-completeness.tex
Remark / question: Claim 1 seems too general for the conjecture

@@
 
-\item Given an a.s.~finite stopping time $T$ (meaning: $T < \infty$ a.s.) adapted to a filtration, there exist supermartingale $(Y_n)_{n \in \omega}$ adapted to the same filtration and antitone function $\epsilon$ such that $(Y_n)_{n \in \omega}$ is an antitone strict supermartingale w.r.t.~$(\epsilon, T)$.
+\item[(1)] Given an a.s.~finite stopping time $T$ (meaning: $T < \infty$ a.s.) adapted to a filtration, there exist supermartingale $(Y_n)_{n \in \omega}$ adapted to the same filtration and antitone function $\epsilon$ such that $(Y_n)_{n \in \omega}$ is an antitone strict supermartingale w.r.t.~$(\epsilon, T)$.
 
-\item Let $(Y_n')_{n \in \omega}$ be an antitone strict supermartingale w.r.t.~$(\epsilon, T)$. 
+\item[(2)] Let $(Y_n)_{n \in \omega}$ be an antitone strict supermartingale w.r.t.~$(\epsilon, T)$. 
 Then $M$ admits an antitone ranking function.
@@
 
-For 2, note that for every $N \in \mathit{Rch}(M)$, there exists $n \geq 0$ and $s \in S$ such that $\red^n(M, s) = (N, -)$.
+\lo{Is (1) too general for the conjecture?
+Claim (1) assumes an arbitrary a.s.~finite stopping time $T$. This seems too general for the conjecture, which assumes a ``strongly'' AST $M \in \Lambda^0$.
+Regardless of the validity of (1), in order to prove the conjecture, I think we would need a more specific version of (1) that assumes a stopping time about terminating runs of $M$ --- such as the stopping time of definition (\ref{eq:y stopping time}) which only counts $\tY$-reduction steps --- so that the antitone-strict supermartingale that is constructed has the necessary structure to yield a ranking function for $M$.}
+
+\iffalse
+For (2), note that for every $N \in \mathit{Rch}(M)$, there exists $n \geq 0$ and $s \in S$ such that $\red^n(M, s) = (N, -)$.
 Define a function $f : \mathit{Rch}(M) \to \nnReal$ by $f(N) := Y_n'(s)$.
+\fi
 

### 2020-06-09 relativised-completeness.tex
Fix minor errors in incomplete completeness proof.

@@
 If $T$ is bounded by $b$ a.s.~(for $n$ constant), the statement is trivially true by taking $\epsilon$ to be constantly 1, and $Y_n = b - n$.
-Otherwise, let $(T_n)_{n \in \omega}$ be defined recursively such that $T_0 = 0$, $T_{n+1} > T_n$ and $\mathbb P[T < T_{n+1}] \leq \frac 1 4 \mathbb P[T < T_n]$.
+Otherwise, let $(T_n)_{n \in \omega}$ be defined recursively such that $T_0 = 0$, $T_{n+1} > T_n$ and $\mathbb P[T > T_{n+1}] \leq \frac 1 4 \mathbb P[T > T_n]$.
 \lo{Since $T$ is a stopping time, we can assume that $T_n \in \mathbb N$.
 Problem with $T_1$: we require $T_1 > 0$ which may contradict the requirement $\mathbb P[T < T_{1}] \leq \frac 1 4 \mathbb P[T < 0] = 0$ for certain $T$, such as those that satisfy $\mathbb P[T < 1] > 0$.
-We could define $T_1$ as the least number satisfying $\mathbb P[T < T_1] \geq \frac{1}{4}$.}
+We could define $T_1$ as the least number satisfying $\mathbb P[T < T_1] \geq \frac{1}{4}$.} \akr{Sorry, I just wrote it wrong.
+I assume this ($\mathbb P[T > T_n]$ instead of $\mathbb P[T < T_n]$) resolves your concern.}
 We will then define a supermartingale $(Y_n)_{n \in \omega}$ and nonrandom sequence $(Y'_n)_{n \in \omega}$ such that $Y_n = 0$ iff $T < n$ \changed[lo]{a.s.}, $Y_n = Y'_n$ iff $T \geq n$ \changed[lo]{a.s.}, and $Y'_n \geq 2^k$ for $n \geq T_k$.
@@
     \text{for }x \in [0,1),\ &\epsilon(x) = 1 \\
-    \text{for }x \in [2^k,2^{k+1}),\ &\epsilon(x) = \max \left (\epsilon(2^{k-1}), \frac{2^k}{T_{k+1}-T_k} \right ).
+    \text{for }x \in [2^k,2^{k+1}),\ &\epsilon(x) = \min \left (\epsilon(2^{k-1}), \frac{2^k}{T_{k+1}-T_k} \right ).
 \end{align*}
@@
 \item For $x \geq 2^k$, $\epsilon(x) \leq \frac{2^k}{T_{k+1}-T_k}$. 
-\lo{This would imply $\frac{2^k}{T_{k+1}-T_k} \geq \epsilon(2^{k-1})$. Why?}
+\lo{This would imply $\frac{2^k}{T_{k+1}-T_k} \geq \epsilon(2^{k-1})$. Why?} \akr{Corrected: it should have been min, not max.}
 \end{itemize}
@@
 \end{align*}
-as required. Substituting $n = T_{k+1}$ and using the fact that $\frac{\mathbb P[T > T_k]}{\mathbb P[T > T_{k+1}]}$, the same reasoning gives $Y'_{T_{k+1}} \geq 2^{k+2}$, as required.
+as required. Substituting $n = T_{k+1}$ and using the fact that $\frac{\mathbb P[T > T_k]}{\mathbb P[T > T_{k+1}]} > 4$, the same reasoning gives $Y'_{T_{k+1}} \geq 2^{k+2}$, as required.
 

### 2020-06-09 relativised-completeness.tex
Comment

@@
 I assume this ($\mathbb P[T > T_n]$ instead of $\mathbb P[T < T_n]$) resolves your concern.}
-We will then define a supermartingale $(Y_n)_{n \in \omega}$ and nonrandom sequence $(Y'_n)_{n \in \omega}$ such that $Y_n = 0$ iff $T < n$ \changed[lo]{a.s.}, $Y_n = Y'_n$ iff $T \geq n$ \changed[lo]{a.s.}, and $Y'_n \geq 2^k$ for $n \geq T_k$.
+We will then define a supermartingale $(Y_n)_{n \in \omega}$ and nonrandom sequence $(Y'_n)_{n \in \omega}$ such that $Y_n = 0$ iff $T < n$ \changed[lo]{a.s.} \akr{I see no reason that this can't just be true everywhere, as this is part of the definition of $Y_n$.
+Likewise for the other ``a.s.'' you added.}, $Y_n = Y'_n$ iff $T \geq n$ \changed[lo]{a.s.}, and $Y'_n \geq 2^k$ for $n \geq T_k$.
 

### 2020-06-12 probabilisticProgrammingMartingales.tex
Summarize the intended results about the alternative semantics.

@@
 The new sample space is then defined as $I^{L_s(M)}$, with the Borel $\sigma$-algebra and product measure.
+
+\paragraph{}
+The proofs that follow are based on a previous version of $\sim$.
+I intend to update them after making more progress on the antitone-strict supermartingale completeness result, but for now, here's a summary of the lemmas and theorems I intend to prove about it:
+\begin{itemize}
+    \item If $M \to N$, the relation $\sim_N^*$ defined on $Rch(N)$ is equal to the restriction of $\sim_M^*$ to $Rch(N)$
+    \item Given the definition below of $\Rightarrow$, which is still applicable to this version of $\sim$, and which is the equivalent of $\red$ in this version of the semantics, $\Rightarrow$ is Church-Rosser.
+    \item If $M \to N$ at $P$, no position in any term in $Rch(N)$ is related to $(M,P)$ by $\sim_M^*$, and therefore each sample is used at most once in any reduction sequence.
+    \item The subprobability distribution of values obtained by reducing a term in the usual call-by-value semantics is equal to that obtained by any reduction strategy on $\Rightarrow$ (i.e.~a sub-relation of $\Rightarrow$ that is a function) (with some allowance made for the fact that some values are not normal forms of $\Rightarrow$, so the reduction strategy might overshoot the value that $\red^\infty$ reaches.
+I'm not quite sure what form this will take.)
+    \item Similarly, the probability of termination is equal in $\red^\infty$ and any reduction strategy on $\Rightarrow$, so almost sure termination can be equivalently defined based on $\Rightarrow$.
+    \item For any closed term $N$, if there is any reduction strategy for $N$ and ranking function corresponding to that reduction strategy, $N$ is AST.
+    \item Corresponding sparse and antitone-strict ranking function theorems.
+\end{itemize}
 

### 2020-06-17 akr-thots-completeness.tex
Added file

% !TEX root = ./probabilisticProgrammingMartingales.tex
\newcommand\tvdist[1]{{|\!|}#1{|\!|}_{\textsc{tv}}}

\akr{I have some incomplete thoughts in the direction of proving completeness (both in the general case and for terms). For each non-null event measurable with respect to some step in the filtration (or reachable term in the case of proving termination of terms), there is a weakly decreasing function $\mathbb{N} \to \nnReal$ with limit 0 representing the conditional probability of having not yet terminated after $n$ more steps. 
These functions can be linearly combined and have a partial order, and form a sort of generalised supermartingale. It is then necessary to map these to real numbers, which can be done by taking the limit with respect to some ultrafilter on the natural numbers of the ratios between them, and picking some arbitrary starting point (e.g.~setting the value for the initial term to 1). The problem with this is that some terms could get assigned the value 0, so maybe it's necessary to take equivalence classes of terms whose values have ratios in the open interval $(0,\infty)$ or something.}

\begin{lemma}
If $M$ is strongly AST (i.e.~every reachable term from $N$ is AST) then for each $N \in \mathit{Rch}(M)$, there is a non-increasing function $\rho_N : \mathbb{N} \to \nnReal$ such that $\lim_{n \to \infty} \rho_N(n) = 0$.
\end{lemma}

Take $N \in \mathit{Rch}(M)$.
Assume $N$ is not a value.
Define $\rho_N : \mathbb{N} \to \nnReal$ by setting
\[
\rho_N(n) := \mu(\{s \mid \pi_0(\red^n(N, s)) \textrm{~non-value}\})
\]
I.e.~$\rho_N(n)$ is the probability that $N$ is not yet terminated after $n$ more reduction steps.
\lo{E.g.~writing $I^{m+1} := I^m \, I$ where $I^1 := I$, we have $\rho_{I^m}(n) = 1$ if $n < m - 1$, and 0 otherwise.}
Let $|\rho_N| := \sum_{n \in \omega} \rho_N(n)$.

\begin{conjecture}
$|\rho_N|$ is finite.
\end{conjecture}

Define $\widehat{\rho}_N (n):= \frac{\rho_N(n)}{|\rho_N|}$.
Assuming the conjecture, $\widehat{\rho}_N$ is a probability mass function (or subprobability mass function if we set $\widehat{\rho_N}$ to be 0 in case $N$ is a value).

\lo{Define r.v.~$Z_n(s) := \widehat{\rho}_N$ where $\red^n(M, s) = (N, {-})$. 
Then $(Z_n)_{n \in \omega}$ is a (generalised) supermartingale.
I am not sure if we need this.}

We want to map functions $\widehat{\rho}_N$ where $N$ ranges over $\mathit{Rch}(M)$ to real numbers.

The \emph{total variational distance} between probability measures $\mu, \nu$ on the same measurable space $(\Omega, \calF)$, $\tvdist{\mu - \nu} := \sup_{A \in \calF} |\mu(A) - \nu(A)|$.
The definition is presented as the supremum over all measurable subsets of the state space $\Omega$, which is not a convenient way to calculate the distance.
Fortunately there are two useful alternative characterisations:
\begin{enumerate}
\item as a simple sum over the state space: $\tvdist{\mu - \nu} = \frac{1}{2} \, \sum_{x \in \Omega} |\mu(x) - \nu(x)|$
\item via \emph{coupling}, a different probabilistic interpretation, i.e., $\tvdist{\mu - \nu}$ measures how close to identical we can force two random variables to be, such that one realises $\mu$ and the other realises $\nu$
\[
\tvdist{\mu - \nu} = \inf \{\mathbb{P}(X \not= Y) \mid \hbox{$(X, Y)$ is a coupling of $\mu$ and $\nu$}\}
\]
\end{enumerate}

\lo{NOTE. \emph{Coupling} is a powerful method and proof technique in probability theory based on the simple idea of comparing random variables with each other. 
It has been applied to prove limit theorems, derive inequalities, and obtain approximations, useful in a host of situations, ranging across random walks, Markov chains, percolation, interacting particle systems, and diffusions.
A easily accessible reference is \citep{denHollander12}; see also \citep [Ch.~4 \& 5]{LevinP17}.}

A candidate for ranking function: for $N \in \mathit{Rch}(M)$, $r(N) := 1 - \tvdist{\widehat{\rho}_M - \widehat{\rho}_N}$. 
Thus, intuitively, the ``further'' $N$ is from $M$, the closer the value $r(N)$ is to 0.

\lo{Problem: We want to show that if $\red(N, s) = (N', -)$ by contracting a $\tY$-redex, then $r(N) > r(N')$.
However we have $|\rho_{N'}| < |\rho_{N}|$ and $\rho_{N'}(n) \leq \rho_{N}(n)$.}

\lo{When proving that a candidate function is a ranking function, try making use of the sparse ranking function theorem.}

### 2020-06-17 probabilisticProgrammingMartingales.tex
Thots on completeness proof: an alternative to ultrafilter on natural numbers.
Added file

@@
 \newtheorem{corollary}{Corollary}
+\newtheorem{conjecture}{Conjecture}
 
@@
 \end{align*}
-The subterm at a position is defined as
+The \changed[lo]{\emph{subterm of $M$ at a position $P$}, denoted $M \mid P$,} is defined as
 \begin{align*}
@@
 Two subterms $N_1$ and $N_2$ of a term $M$, corresponding to positions $P_1$ and $P_2$, can overlap in a few different ways.
-If $P_1$ is an initial segment (i.e.~prefix) of $P_2$, then $N_2$ is also a subterm of $N_1$, which is written as $P_1 \leq P_2$.
+If $P_1$ is an initial segment (i.e.~prefix) of $P_2$ (written as $P_1 \leq P_2$), then $N_2$ is also a subterm of $N_1$.
 If neither $P_1 \leq P_2$ nor $P_1 \geq P_2$, the positions are said to be disjoint.
@@
 The reflexive transitive closure $\sim^*$ of this relation is used to define the set of potential positions $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable terms.
-If $(N,P) \sim^* (O,Q)$, then $N \mid P$ and $O \mid Q$ both have the same shape (i.e.~they're either both variables, both applications, both $\tsample$s ect.), therefore it's well-defined to talk of the set of potential positions where there is a $\tsample$, $L_s(M)$.
+If $(N,P) \sim^* (O,Q)$, then $N \mid P$ and $O \mid Q$ both have the same shape (i.e.~they're either both variables, both applications, both $\tsample$s etc.), therefore it's well-defined to talk of the set of potential positions where there is a $\tsample$, $L_s(M)$.
 The new sample space is then defined as $I^{L_s(M)}$, with the Borel $\sigma$-algebra and product measure.
@@
 \begin{itemize}
-    \item If $M \to N$, the relation $\sim_N^*$ defined on $Rch(N)$ is equal to the restriction of $\sim_M^*$ to $Rch(N)$
+    \item If $M \to N$, the relation $\sim_N^*$ defined on $Rch(N)$ is equal to the restriction of $\sim_M^*$ to $Rch(N)$ \lo{Define $\sim_N^*$.}
     \item Given the definition below of $\Rightarrow$, which is still applicable to this version of $\sim$, and which is the equivalent of $\red$ in this version of the semantics, $\Rightarrow$ is Church-Rosser.

### 2020-07-15 probabilisticProgrammingMartingales.tex
Add more proof for the alternative semantics stuff.

@@
     \begin{enumerate}
-        \item $N$ contains redexes at positions $P_1$ resp.~$P_2$, $O_1$ is $N$ reduced first at $P_1$
+        \item $N$ contains redexes at disjoint positions $P_1$ and $P_2$, $O_1$ is $N$ reduced first at $P_1$ then $P_2$ and $O_2$ is $N$ reduced first at $P_2$ then at $P_1$.
 
@@
         \item $N \mid P = (\lambda x.
-A) B$, $B \mid Q$ is a redex, $(R_i)_i$ is a list of all the positions in $A$ where $A \mid R = x$, $O_1$ is $N$ reduced first at $P ; @_2 ; Q$ then at $P$, and $O_2$ is $N$ reduced first at $P$ then at $P;R_i;Q$ for each $i$ in order.
+A) B$, $B \mid Q$ is a redex, $(R_i)_i$ is a list of all the positions in $A$ where $A \mid R = x$, ordered from left to right, $O_1$ is $N$ reduced first at $P ; @_2 ; Q$ then at $P$, and $O_2$ is $N$ reduced first at $P$ then at $P;R_i;Q$ for each $i$ in order.
 
         \item $N \mid P = \tY (\lambda x.
-A)$, $A$ reduced at $Q$ is $A'$, $(R_i)_i$ is a list of all the positions where $A' \mid R  = x$, $O_1$ is $N$ reduced first at $P;\tY;\lambda;Q$ then at $P$, and $O_2$ is $N$ reduced first at $P$ then at $P;\lambda;@_1;Q$ then at $P;\lambda;@_1;R_i;Y;\lambda;Q$ for each $i$ in order.
+A)$, $A$ reduced at $Q$ is $A'$, $(R_i)_i$ is a list of all the positions where $A' \mid R  = x$, ordered from left to right, $O_1$ is $N$ reduced first at $P;\tY;\lambda;Q$ then at $P$, and $O_2$ is $N$ reduced first at $P$ then at $P;\lambda;@_1;Q$ then at $P;\lambda;@_1;R_i;Y;\lambda;Q$ for each $i$ in order.
     \end{enumerate}
@@
 \end{itemize}
+
+\paragraph{}
+Before defining the new version of the reduction relation $\red$, a few lemmas about the properties of $\sim$ are necessary for it to be well-defined.
+
+\begin{lemma}
+If $(N_1,P_1) \sim^* (N_2,P_2)$, there is some term $O$, position $Q$ in $O$, and descendants $O_1$ and $O_2$ of $O$ such that $(N_1,P_1) \sim_c^* (O_1,P_1) \sim_p^* (O,Q) \sim_p^* (O_2,P_2) \sim_c^* (N_2,P_2)$.
+\end{lemma}
+\begin{proof}
+The proof will proceed by successively modifying the sequence of $\sim$-steps from $(N_1,P_1)$ to $(N_2,P_2)$ to more restricted forms.
+If we call the length of a reduction sequence from M (here being identified with its final term) its level, where longer reduction sequences are lower levels, then the relation $(N_1,P_1) \sim^* (N_2,P_2)$ implies that there exists some sequence of $\sim_p$ steps to higher or lower levels, and $\sim_c$ steps horizontally, in some arbitrary order.
+The first step of the simplification is to find a form where all the $\sim_p$ steps up happen before all the $\sim_p$ steps down.
+This proceeds by induction on the level of the innermost incorrectly-ordered pair of $\sim_p$ steps.
+
+Suppose there is, in the $\sim$-sequence, some pair of steps $(A_1,R_1) \sim_p (A'_1, R'_1)$ and $(A'_2,R'_2) \sim_p (A_2, R_2)$, occurring in this order, where $A_1 \to A'_1$ and $A_2 \to A'_2$, and furthermore the levels of $A'_1$ and $A'_2$ are the lowest of any such pair (and therefore equal, call it $n$), and there is no such $A'_2$ closer to $A'_1$ in the sequence.
+There cannot be any more $\sim_p$ steps between $A'_1$ and $A'_2$, therefore $(A'_1, R'_1) \sim_c^* (A'_2, R'_2)$.
+We aim to show that there is some sequence of $\sim$ steps from $(A_1,R_1)$ to $(A_2,R_2)$, none of whose intermediate terms pass to level $n$ or below, by analysis of the possible sequences of $\sim_c$ steps.
+
+If $A'_1 = A'_2$, as each is descended from a unique parent term, it must also be the case that $A_1 = A_2$ and $R_1 = R_2$, therefore $(A_1,R_1) \sim^* (A_2,R_2)$ with no $\sim_p$ steps to level $n$ trivially.
+
+Otherwise, consider the first $\sim_c$ step between $(A'_1, R'_1)$ and $(A'_2, R'_2)$, $(A'_1, R'_1) \sim_c (D, R'_1)$.
+If the reduction sequence $O_1 \to O'_1$ involved in the definition of $\sim_c$ in this case is not trivial (i.e.~it has at least one step), then it includes as its final step the reduction $A_1 \to A'_1$ therefore there is some equivalent $A''_1$ of $A_1$ descended from $O_2$ instead, and $(A_1, R_1) \sim_c (A''_1, R_1) \sim_p (D,R'_1)$.
+There is then a similar but strictly shorter sequence of $\sim$-steps from $(A''_1, R_1)$ to $(A_2, R_2)$, so assuming there is a solution for that one inductively, it can be composed with $(A_1, R_1) \sim_c (A''_1, R_1)$ to form a suitable sequence from $(A_1,R_1)$ to $(A_2,R_2)$.
+
+In the other case, that $O_1 = O'_1$, the reduction step from $A_1$ to $A'_1$ is the last step in the reduction sequence of $N$ to either $O_1$ or $O_2$ in one of the cases 4.a to 4.f defining $\sim_c$, which restricts the cases that are possible according to the form of the reduction $A_1 \to A'_1$.
+If $A'_1 = O_1$ resp.~$O_2$, by following the reductions $N \to A_1$ backwards then those $N \to O_2$ resp.~$O_1$ forwards, a path by which $(A_1,R_1) \sim_p^* (O_2,R'_1)$ can be constructed except in certain cases listed below, which will be dealth with separately, then as before the required relation between $(A_1,R_1)$ and $(A_2,R_2)$ can be constructed by induction.
+The cases where this doesn't work are:
+\akr{There are so many variables in scope here it's hard to avoid name clashes.
+I'll need to re-name some things later to make them more unique, but for now I'll just awkwardly avoid using some of them.}
+\begin{itemize}
+    \item Case a, when $A'_1 = O_1$ and $(O_2,R'_1)$ does not have a parent due to that position being newly generated by the last reduction, or vice-versa (swapping $O_1$ and $O_2$)
+    \item Case c, when $A'_1 = O_2$ and $R'_1 \geq P;Q$, being a position generated by the reduction $B \to B'$
+    \item Case d, when $A'_1 = O_2$ and $R'_1 \geq P;Q$, being a position generates by the reduction $R \to R'$
+    \item Case e, when $A'_1 = O_1$ and $R'_1 \geq P;R_i;Q$, where $i$ is not the last in the list, being a position generated by the reduction $B \to B'$.
+    \item Case e, when $A'_1 = O_1$ and $P;R_i;Q \geq R'_1 \geq P$, where $i$ is not the last in the list
+    \item Case f, when $A'_1 = O_1$ and $R'_1 \geq P$.
+\end{itemize}
+What all of these cases have in common is that in the branch other than the one leading to $A'_1$ (i.e.~ $N \to^* D$), $(D, R'_1)$ is newly generated by the latest reduction and is not related to anything above by $\sim_p$.
+It is therefore impossible for $D$ to be equal to $A'_2$.
+This property (that the relevant location has no parent) is preserved by any $\sim_c$ step that has its $O_1$ and $O_2$ at a strictly higher level than its $O'_1$ and $O'_2$, so consider the next $\sim_c$ step that does occur at level $n$.
+
+\akr{I want to then say that the mentioned $\sim_c$ must be the first one in reverse, because it has to be one of the cases mentioned above where the position is new, and none of the cases overlap with each other, but that doesn't quite work as planned because the intermediate $\sim_c$s at a level 1 higher could interfere with the non-overlappingness.}
+\end{proof}
 

### 2020-07-15 probabilisticProgrammingMartingales.tex
ongoing

--- 2020-06-17 probabilisticProgrammingMartingales.tex
@@
 \begin{enumerate}
-    \item If $N$ reduces to $O$ with the redex at position $P$\lo{Do you mean: $M \to^\ast N \to O = N[\dotDelta / P]$ for some redex contraction $\Delta \to \dotDelta$}, and $Q$ is a position in $N$ disjoint from $P$, then $(N,Q) \sim_p (O,Q)$.
+    \item If $N$ reduces to $O$ with the redex at position $P$ \changed[lo]{(i.e.~$M \to^\ast N \to O = N[\dotDelta / P]$ with $O \mid P = \Delta$ for some redex contraction $\Delta \to \dotDelta$)}, and $Q$ is a position in $N$ disjoint from $P$, then $(N,Q) \sim_p (O,Q)$.
 

### 2020-07-22 probabilisticProgrammingMartingales.tex
Start writing later parts of the alternative semantics stuff.

--- 2020-07-15 probabilisticProgrammingMartingales.tex
@@
 \paragraph{}
-The proofs that follow are based on a previous version of $\sim$.
-I intend to update them after making more progress on the antitone-strict supermartingale completeness result, but for now, here's a summary of the lemmas and theorems I intend to prove about it:
+For now, here's a summary of the lemmas and theorems I intend to prove about this alternative semantics:
 \begin{itemize}
@@
 
-\paragraph{}
-Before defining the new version of the reduction relation $\red$, a few lemmas about the properties of $\sim$ are necessary for it to be well-defined.
-Considering the set of paths through $Rch(M)$ (which we're implicitly identifying with $Rch(M)$ itself) as a tree rooted at (the trivial path at) $M$, $\sim_p$ only relates positions in child and parent terms, and $\sim_c$ only relates first cousins.
-Furthermore, by a case analysis on the forms of positions, the possible paths $N_1 \sim N_2 \sim \dots \sim N_n$ can be strongly restricted.
-
-In any case, a position can be related to at most one position in each child term.
-Letting $P$ be the position of the redex and $Q$ the original term, this is by rule 1 if $P$ and $Q$ are disjoint, rule 2 if $P;@_1 \leq Q$ and rule 3 if $P;if_i \leq Q$.
-As $M$ itself has no cousins or parents, this is the only way positions in $M$ are related to others.
-
-For any other term $N$ and position $Q$ in it, take the parent $N'$ and the position $P$ of the reduction $N' \to N$.
-The positions that $(N,Q)$ are related to other than those in the children of $N$ are as follows.
-If $Q \not \geq P$, only case 1 applies.
-If $Q \geq P$ and $(N,Q) \sim_p (N',Q')$ by cases 2 or 3, $Q'$ is also $\geq P$ therefore if $(N,Q)$ is also related to something by $\sim_c$, $(N',Q')$ is not related to anything in its parent.
-Furthermore, letting $Q = P;R$, $Q' = P;@_1;\lambda;R$ or $P;\textsf{if}_i;R$ depending on what sort of reduction $N' \to N$ was ($\beta$, $\textsf{if}$-true or $\textsf{if}$-false respectively) therefore in any case, $(N,Q)$ is related to at most one position in $N'$.
-
-Considering now $\sim_c$, let $N'' \to N' \to N$, where the outermost of the two reductions takes place at $P$.
-Cases 5 and 6 can be distinguished from case 4 by whether $P;\tY$ exists in $N''$, and cases 5 and 6 can be distinguished by whether $N'' | R = x$ for some $R < P;\tY;\lambda;S$ where $x$ is the variable bound at $N'' | P;\tY$ and $Q = P;\lambda;@_1;S$.
-In all of the cases 4, 5 and 6, whether $Q$ is $O_1$ or $O_2$ can be distinguished by which of the reductions $N'' \to N'$, $N' \to N$ takes place at $P$.
-In all 6 cases, the term and position $(N,Q)$ is related to can be reconstructed from $Q$ and the positions of the reductions $N'' \to N' \to N$, therefore each $(N,Q)$ pair is related to at most one other by $\sim_c$.
-
-In summary, all equivalence classes under $\sim^*$ are of the shapes \akr{This would be far easier to show with a few pictures.
-I just don't yet know how to make them in Latex and I'm prioritising getting the text done.}
-\lo{PGF/TikZ is the standard \LaTeX\ package for generating vector graphics.
-The manual \url{https://ctan.org/pkg/pgf?lang=en} is a 1318-page tome!
-There are numerous tutorials / guides on the Internet.
-E.g.~\url{https://www.overleaf.com/learn/latex/TikZ_package}}
-
-\paragraph{}
-
-At each reduction step $M \to N$, the sample space must be restricted to $I^{L_s(N)}$, which requires a map $L_s(N) \to L_s(M)$.
-The injection $L_0(N) \to L_0(M)$ is trivial to define by appending $M \to N$ to each path, but in order for this to extend to the quotient, the following lemma is needed.
+\akr{As this proof is takng a long time to sort out, I'm just going to assume for now some of the results, and proceed from there.
+The results I'm assuming are that if $M \to N$, the $\sim$ relation on $Rch(N)$ is equal to the restriction of the $\sim$ relation on $Rch(M)$, restricted to $Rch(N)$.}
+
 \begin{lemma}
-If $M \to N$, the relation $\sim_N^*$ defined on $L_0(N)$ is equal to the restriction of $\sim_M^*$ from $L_0(M)$ to $L_0(N)$.
+The relation $\sim$ is defined on $L_0(M)$ with reference to a particular starting term $M$, so different versions, $\sim_M$ and $\sim_N$, can be defined starting at different terms.
+If $M \to N$, then $\sim^*_N$ is equal to the restriction of $\sim^*_M$ to $L_0(N)$.
 \end{lemma}
 \begin{proof}
-For any equivalence class under $\sim_M^*$ that includes some descendant of $N$, consider the most recent common ancestor of the terms in the equivalence class $O$.
-If $O \in Rch(N)$, so is that whole equivalence class therefore $\sim_M^*$ and $\sim_N^*$ match there.
-If $O \not \in Rch(N)$ but some of its descendants are, it must be the case that $O = M$.
-If $M$ itself has a position in the class, it it a line of $\sim_p$s, and the intersection of it with $Rch(N)$ is an equivalence class of $\sim_N^*$.
-Otherwise, there are two children $O_1, O_2$ of $M$ such that the equivalence class lies in $Rch(O_1) \cup Rch(O_2)$ and at least one of $O_1$ and $O_2$ is $N$.
-In either case, the restriction of the equivalence class to $Rch(N)$ is an equivalence class of $\sim_N^*$ therefore, in total, the restriction of $\sim_M^*$ to $Rch(N)$ is a subset of $\sim_N^*$.
-
-The converse, that $\sim_N^*$ is a subset of $\sim_M^*$ follows directly from the fact that all of the rules in the definition of $\sim$ apply at least as much in $Rch(M)$ as in $Rch(N)$ therefore $\sim_N$ is a subset of $\sim_M$.
 \end{proof}
-The injection $L_s(N) \to L_s(M)$ thus defined is denoted $i(N \to M)$.
-
-Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function in this case, so it is denoted ``$\Rightarrow$'' instead of ``$\red$'', and it relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
+
+At each reduction step $M \to N$, the sample space must be restricted from $I^{L_s(M)}$ to $I^{L_s(N)}$.
+The injection $L_0(N) \to L_0(M)$ is trivial to define by appending $M \to N$ to each path, and using the previous lemma, this induces a corresponding injection on the quotient, $L(N) \to L(M)$.
+The corresponding map $L_s(N) \to L_s(M)$ is then denoted $i(M \to N)$.
+
+Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function, so it is denoted ``$\Rightarrow$'' instead of ``$\red$'', and it relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
 \begin{align*}
@@
 \end{align*}
+
+\paragraph{}
+\begin{lemma}
+The relation $\Rightarrow$ is Church-Rosser.
+\end{lemma}
+\begin{proof}
+Suppose that $(M,s) \Rightarrow (M_1,s_1)$ and also $(M,s) \Rightarrow (M_2,s_2)$, then it is required to prove that there is some $(M',s')$ such that both $(M_1,s_1) \Rightarrow^* (M',s')$ and $(M_2,s_2) \Rightarrow^* (M',s')$.
+Let the positions of the redexes in $M \to M_1$ and $M \to M_2$ be $P_1$ and $P_2$ respectively.
+
+First consider the case that $P_1$ and $P_2$ are disjoint.
+Let $M_1 = M[X_1/P_1]$ and similarly for $X_2$, then let $M' = M[X_1/P_1][X_2/P_2]$.
+As the positions are disjoint, the substitutions commute and both $M_1$ and $M_2$ reduce (with $\to$) to $M'$.
+Let $s' = s \circ i(M \to M_1) \circ(M_1 \to M')$.
+The injection $i(M \to M_1) \circ i(M_1 \to M')$ consists of appending $M \to M_1 \to M'$ to each reduction sequence, but by case 4.a of $\sim$, this is equivalent to appending $M \to M_2 \to M'$, which is $i(M \to M_2) \circ i(M_2 \to M')$.
+In the case that the reduction $M \to M_2$ isn't a $\tsample$-reduction, this is enough to establish that $(M_1, s_1) \Rightarrow (M', s')$, and vice-versa.
+If it is $\tsample$ though, in order for it to be the case that $(M_1, s_1) \Rightarrow (M', s')$, it is additionally necessary that $X_2$, the result of the reduction at $P_2$, be $s_1(M_1, P_2)$, which follows from the fact that $(M, P_2) \sim (M_1, P_2)$ by case 1.
+The case that the reduction $M \to M_1$ is a $\tsample$-reduction is similar.
+
+The case that $P_1 = P_2$ is trivial, because there is at most one possible $\Rightarrow$ reduction at any given position, therefore $(M_1, s_1) = (M_2, s_2)$ already.
+
+The remaining case is that $P_1 < P_2$ or $P_1 > P_2$.
+Assume without loss of generality that $P_1 < P_2$.
+For each possible case of what type of redex $M | P_1$ is, and $P_2$'s position within it, there is a corresponding case of $\sim_c$, and similarly to the case where $P_1$ and $P_2$ are disjoint, the term $O_1 = O_2$ from the definition of $\sim_c$ is a suitable value of $M'$.
+The term $M | P_1$ can't be $\tsample$, because it has strict subterms, but the case that $M | P_2 = \tsample$ is still somewhat more complicated.
+$P_2$ can't be within $P_1 ; @_2$ or $P_1 ; Y$ (cases e and f of $\sim_c$) because those are not valid positions to reduce a $\tsample$, and the case that $P_2$ is within the branch of an if statement that is deleted by the reduction at $P_1$ (case b) doesn't actually present a problem because there is no corresponding reduction to $M \to M_2$ in the other branch, which leaves cases c and d, that $P_2$ is in the $\tif$ branch that isn't deleted, and that $P_2 < P_1 ; @_1 ; \lambda$.
+The values that the $\tsample$s take in these cases match because $(M,P) \sim_p (M_1, P;Q)$ by cases 3 and 2 of $\sim_p$ respectively.
+\end{proof}
+
+\begin{lemma}
+If $M \to N$, with the redex at position $P$, then no position in any term reachable from $N$ is related by $\sim$ to $(M,P)$.
+\end{lemma}
+\begin{proof}
+\end{proof}
 

### 2020-07-28 probabilisticProgrammingMartingales.tex
Continue the outline of the alternative semantics stuff.
I really need a better name than "alternative semantics".

--- 2020-07-22 probabilisticProgrammingMartingales.tex
@@
 \DeclareMathOperator{\nnext}{next}
+\DeclareMathOperator{\cbv}{cbv}
 
@@
 \begin{proof}
+\end{proof}
+
+\paragraph{}
+The reduction relation $\Rightarrow$ is nondeterministic, so it admits multiple possible reduction strategies.
+A reduction strategy starting from a closed term $M$ is a partial function $f$ from $Rch(M)$ to positions, such that for any reachable term $N$ where $f$ is defined, $f(N)$ is a position of a valid redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
+Using a reduction strategy, a subset of $\Rightarrow$ that isn't nondeterministic, $\Rightarrow_f$, can be defined by $(N,s) \Rightarrow_f (N',s')$ iff $(N,s) \Rightarrow (N',s')$ and $N$ reduces to $N'$ with the redex at $f(N)$.
+
+The usual call-by-value semantics can be implemented as one of these reduction strategies, given by (with $V$ a value and $T$ a term that isn't a value and $M$ a general term)
+\begin{align*}
+\cbv(T M) & = @_1 ; \cbv(T) \\
+\cbv(V T) & = @_1 ; \cbv(T) \\
+\cbv(\underline f(V_1, \dots, V_{k-1}, T, M_{k+1}, \dots, M_n)) & = \underline f_k ; \cbv(T) \\
+\cbv(\tY T) & = \tY ; \cbv(T) \\
+\cbv(\tif{T < 0}{M_1}{M_2}) & = \textsf{if}_1 ; \cbv(T) \\
+\cbv(\tscore(T)) & = \tscore ; \cbv(T) \\
+\cbv(V) & \text{ is undefined} \\
+\cbv(T) & = \cdot \text{ otherwise}
+\end{align*}
+(this last case covers redexes at the root position).
+
+A closed term $M$ terminates with a given reduction strategy $f$ and samples $s$ if there is some natural number $n$ such that $(M,s) \Rightarrow_f^n (N,s')$ where $f$ gives no reduction at $N$.
+The term terminates almost surely with respect to $f$ if it terminates with $f$ for almost all $s$.
+
+\begin{theorem} \label{AstEquivalence}
+A closed term $M$ is AST with respect to $\cbv$ iff it is AST.
+\end{theorem}
+\begin{proof}
+\end{proof}
+
+\begin{theorem} \label{CbvIsTerminatingest}
+If $M$ terminates with some reduction strategy $f$ and samples $s$, it terminates with $\cbv$ and $s$.
+\end{theorem}
+\begin{proof}
+\end{proof}
+
+\begin{corollary}
+If $M$ is AST with respect to any reduction strategy, it is AST.
+\end{corollary}
+\begin{proof}
+Suppose $M$ is AST with respect to $f$.
+Let the set of samples with which it terminates with this reduction strategy be $X$.
+By \ref{CbvIsTerminatingest}, $M$ also terminates with $\cbv$ and every element of $X$, and $X$ has measure 1, by assumption, therefore $M$ is AST with respect to $\cbv$ therefore by \ref{AstEquivalence} it is AST.
 \end{proof}

### 2020-08-05 probabilisticProgrammingMartingales.tex
An example. Some comments. Ongoing.

--- 2020-07-22 probabilisticProgrammingMartingales.tex
@@
 \newcommand{\tY}{\textsf{Y}}
-\newcommand{\tif}[3]{\textsf{if }#1\textsf{ then }#2\textsf{ else }#3}
+%\newcommand{\tif}[3]{\textsf{if }#1\textsf{ then }#2\textsf{ else }#3}
+\newcommand{\tif}[3]{\textsf{if}\,(#1, #2, #3)} % space-saving concrete syntax
 \newcommand{\tsample}{\textsf{sample}}
@@
 \end{align*}
-so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g.
-$x \, y \mid \lambda ; \cdot$ is undefined).
+so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda ; \cdot$ is undefined). 
 A position such that $M\mid P$ does exist is said to occur in $M$. 
-Substitution at a given position, $M[N/P]$, is defined similarly.
+\changed[lo]{Substitution of $N$ at position $P$ in $M$, written $M[N/P]$, is defined similarly.
+Take, for example, $M = \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})$ and $P =\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2$, then 
+\[
+M[\tsample / P] = \lambda x \, y. y \, (\tif{x < 0}{y \, \tsample}{\underline 3}).
+\]}
 
@@
 If neither $P_1 \leq P_2$ nor $P_1 \geq P_2$, the positions are said to be disjoint.
-This is mostly relevant in that in any substitution at some position, the subterms at all other positions are unaffected.
+This is mostly relevant in that in any substitution at some position, the subterms at all other positions are unaffected. \lo{The preceding sentence is unclear. Typo? Mostly $\mapsto$ most?}
 
@@
 %\lo{On first reading, I struggled to parse the following clauses.}
-\changed[lo]{The binary relation $\to$ is defined by the following rules, each is conditional on a redex occuring at position $P$ in the term $M$.}
+\changed[lo]{The binary relation $\to$ is defined by the following rules, each is conditional on a redex occurring at position $P$ in the term $M$.}
 \begin{align*}
@@
 In each of these cases, $M \mid P$ is the redex, and reduction takes place at $P$.
-\lo{The definition above is clear. But as a general rule, it is good to illustrate definitions -- especially if they are genuinely new -- with an example or two.}
+%\lo{The definition above is clear. But as a general rule, it is good to illustrate definitions -- especially if they are genuinely new -- with an example or two.}
 
@@
 & \qquad \text{$M \to N$ and the redex is not $\tsample$, or} \\
-& \qquad \text{$M \mid P = \tsample$, $N = M[\underline{s(M,P)}/P]$ and $\lambda$ does not occur after $@_2$ of $\tY$ in $P$}
+& \qquad \text{$M \mid P = \tsample$, $N = M[\underline{s(M,P)}/P]$ and $\lambda$ does not occur after $@_2$ or $\tY$ in $P$}
 \end{align*}

### 2020-08-06 probabilisticProgrammingMartingales.tex
Various, mainly minor

@@
 The language SPCF is a simply-typed lambda calculus with sampling of real numbers from $[0,1]$ and unbounded scoring, following \cite{MakOP20b}.
-Types and terms are defined as follows, where $r$ is a real, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function and $\Gamma$ is an environment:
+Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
 \begin{align*}
@@
   & \text{values } v ::= \lambda x.s  \mid  \underline{r} \\
-  & \text{terms } s, t ::= v  \mid  x  \mid  t_1 t_2  \mid  \underline{f}(s_1,\dots ,s_n)  \mid  \tY s  \mid  \tif{s < 0}{t_1}{t_2}  \mid  \tsample  \mid  \tscore(s)
+  & \text{terms } s, t ::= v  \mid  x  \mid  t_1 \, t_2  \mid  \underline{f}(s_1,\dots ,s_n)  \mid  \tY s  \mid  \tif{s < 0}{t_1}{t_2}  \mid  \tsample  \mid  \tscore(s)
 \end{align*}
@@
 \paragraph{}
-To define the reduction relation, let evaluation contexts be of the form:
-\begin{align*}
-  E ::= & \, [\,] \mid E t \mid v E \mid \underline{f}(r_1,\dots ,r_{k-1}, E, s_{k+1}, \dots, s_n) \\ & \mid \tY E \mid \tif{E<0}{s_1}{s_2} \mid \tscore (E)
+To define the reduction relation, let \emph{evaluation contexts} be of the form:
+\begin{align*}
+  E ::= & \, [\,] \mid E \, t \mid v \, E \mid \underline{f}(r_1,\dots ,r_{k-1}, E, s_{k+1}, \dots, s_n) \\ & \mid \tY E \mid \tif{E<0}{s_1}{s_2} \mid \tscore (E)
 \end{align*}
@@
 Every closed well-typed term either is a value or reduces to another term.
+
+\lo{NOTATION. A minor point: metavariables for terms are $s, t$, etc.~up to this point, but $M, N$, etc.~here after all.}
 
@@
 
-\akr{As this proof is takng a long time to sort out, I'm just going to assume for now some of the results, and proceed from there.
+\akr{As this proof is taking a long time to sort out, I'm just going to assume for now some of the results, and proceed from there.
 The results I'm assuming are that if $M \to N$, the $\sim$ relation on $Rch(N)$ is equal to the restriction of the $\sim$ relation on $Rch(M)$, restricted to $Rch(N)$.}

### 2020-08-25 probabilisticProgrammingMartingales.tex
Write out more proofs and sketches for alternative semantics stuff.

--- 2020-08-05 probabilisticProgrammingMartingales.tex
@@
 \usepackage{verbatim}
+\usepackage{enumitem}
 \newcommand{\tY}{\textsf{Y}}
-%\newcommand{\tif}[3]{\textsf{if }#1\textsf{ then }#2\textsf{ else }#3}
-\newcommand{\tif}[3]{\textsf{if}\,(#1, #2, #3)} % space-saving concrete syntax
+\newcommand{\tif}[3]{\textsf{if }#1\textsf{ then }#2\textsf{ else }#3}
 \newcommand{\tsample}{\textsf{sample}}
 \newcommand{\tscore}{\textsf{score}}
+\newcommand{\skeletonPlaceholder}{\textsf{X}} % I'm not sure what the best notation here is.
 \DeclareMathOperator{\red}{red}
@@
 \end{align*}
-so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda ; \cdot$ is undefined). 
+so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g.
+$x \, y \mid \lambda ; \cdot$ is undefined).
 A position such that $M\mid P$ does exist is said to occur in $M$.
-\changed[lo]{Substitution of $N$ at position $P$ in $M$, written $M[N/P]$, is defined similarly.
-Take, for example, $M = \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})$ and $P =\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2$, then 
-\[
-M[\tsample / P] = \lambda x \, y. y \, (\tif{x < 0}{y \, \tsample}{\underline 3}).
-\]}
+Substitution at a given position, $M[N/P]$, is defined similarly.
 
@@
 If neither $P_1 \leq P_2$ nor $P_1 \geq P_2$, the positions are said to be disjoint.
-This is mostly relevant in that in any substitution at some position, the subterms at all other positions are unaffected. \lo{The preceding sentence is unclear. Typo? Mostly $\mapsto$ most?}
+This is mostly relevant in that in any substitution at some position, the subterms at all other positions are unaffected.
 
@@
 %\lo{On first reading, I struggled to parse the following clauses.}
-\changed[lo]{The binary relation $\to$ is defined by the following rules, each is conditional on a redex occurring at position $P$ in the term $M$.}
+\changed[lo]{The binary relation $\to$ is defined by the following rules, each is conditional on a redex occuring at position $P$ in the term $M$.}
 \begin{align*}
@@
 In each of these cases, $M \mid P$ is the redex, and reduction takes place at $P$.
-%\lo{The definition above is clear. But as a general rule, it is good to illustrate definitions -- especially if they are genuinely new -- with an example or two.}
+\lo{The definition above is clear. But as a general rule, it is good to illustrate definitions -- especially if they are genuinely new -- with an example or two.}
 
@@
 It is therefore necessary to consider possible positions that may occur in other terms reachable from the original term.
-This is itself inadequate because some of the positions in different reachable terms need to be considered the same.
-
-Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a reduction sequence starting at $M$, and the second of which is a position in the final term of the reduction sequence.
-Reduction sequences are used rather than reachable terms because if the same term is reached twice, different samples may be needed, but they will be discussed as though they were terms, by abuse of notation.
+This is itself inadequate because some of the positions in different reachable terms need to be considered the same, and the number of reachable terms is in general uncountable, which leads to measure-theoretic issues.
+
+Define a skeleton, to be a term but instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term has a skeleton and each skeleton can be converted to a term given the appropriate number of real numbers to substitute in.
+The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder}{A}{B}$ reducing nondeterministically to both $A$ and $M$.
+
+Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a reduction sequence of skeletons starting at $M$, and the second of which is a position in the final skeleton of the reduction sequence.
+\akr{I've yet to go through and replace references to terms with references to skeletons, but it should be reasonably obvious where the substitution should take place.
+This is potentially even more confusing though because I'm now implicitly identifying terms, skeletons, reduction sequences and skeletal reduction sequences.} Reduction sequences are used rather than reachable terms because if the same term is reached twice, different samples may be needed, but they will be discussed as though they were terms, by abuse of notation.
 \lo{Readers would benefit greatly from some examples of an $L_0(M)$-pair, and the relations $\sim_p$ and $\sim_c$, here, before the formal definitions.
@@
 \begin{enumerate}
-    \item If $N$ reduces to $O$ with the redex at position $P$ \changed[lo]{(i.e.~$M \to^\ast N \to O = N[\dotDelta / P]$ with $O \mid P = \Delta$ for some redex contraction $\Delta \to \dotDelta$)}, and $Q$ is a position in $N$ disjoint from $P$, then $(N,Q) \sim_p (O,Q)$.
+    \item If $N$ reduces to $O$ with the redex at position $P$\lo{Do you mean: $M \to^\ast N \to O = N[\dotDelta / P]$ for some redex contraction $\Delta \to \dotDelta$}, and $Q$ is a position in $N$ disjoint from $P$, then $(N,Q) \sim_p (O,Q)$.
 
@@
     \end{enumerate}
-    and $O_1$ and $O_2$ are equal as terms, but with different reduction sequences (which merely requires that in cases a, c and d, if the redexes of unspecified type are $\tsample$, corresponding $\tsample$s reduce to the same numbers), $O_1'$ and $O_2'$ are the results of applying some reduction sequence to each of $O_1$ and $O_2$ (the same reductions in each case, which is always possible because they're equal terms), and $S$ is a position in $O_1'$ (or equivalently $O_2'$), then $(O_1',S) \sim_c (O_2',S)$.
+    (in which case $O_1$ and $O_2$ are equal as terms, but with different reduction sequences), $O_1'$ and $O_2'$ are the results of applying some reduction sequence to each of $O_1$ and $O_2$ (the same reductions in each case, which is always possible because they're equal terms), and $S$ is a position in $O_1'$ (or equivalently $O_2'$), then $(O_1',S) \sim_c (O_2',S)$.
 \end{enumerate}
@@
 
-\begin{lemma}
-If $(N_1,P_1) \sim^* (N_2,P_2)$, there is some term $O$, position $Q$ in $O$, and descendants $O_1$ and $O_2$ of $O$ such that $(N_1,P_1) \sim_c^* (O_1,P_1) \sim_p^* (O,Q) \sim_p^* (O_2,P_2) \sim_c^* (N_2,P_2)$.
+\begin{lemma} \label{downAcrossUp}
+If $(N_1,P_1) \sim^* (N_2,P_2)$, there is some descendants $N'_1, N'_2$ of $N_1$ resp.~$N_2$, and a position $P'$ in both of them such that $(N_1,P_1) \sim_p^* (N'_1, P') \sim_c^* (N'_2, P') \sim_p^* (N_2, P_2)$.
 \end{lemma}
 \begin{proof}
-The proof will proceed by successively modifying the sequence of $\sim$-steps from $(N_1,P_1)$ to $(N_2,P_2)$ to more restricted forms.
-If we call the length of a reduction sequence from M (here being identified with its final term) its level, where longer reduction sequences are lower levels, then the relation $(N_1,P_1) \sim^* (N_2,P_2)$ implies that there exists some sequence of $\sim_p$ steps to higher or lower levels, and $\sim_c$ steps horizontally, in some arbitrary order.
-The first step of the simplification is to find a form where all the $\sim_p$ steps up happen before all the $\sim_p$ steps down.
-This proceeds by induction on the level of the innermost incorrectly-ordered pair of $\sim_p$ steps.
-
-Suppose there is, in the $\sim$-sequence, some pair of steps $(A_1,R_1) \sim_p (A'_1, R'_1)$ and $(A'_2,R'_2) \sim_p (A_2, R_2)$, occurring in this order, where $A_1 \to A'_1$ and $A_2 \to A'_2$, and furthermore the levels of $A'_1$ and $A'_2$ are the lowest of any such pair (and therefore equal, call it $n$), and there is no such $A'_2$ closer to $A'_1$ in the sequence.
-There cannot be any more $\sim_p$ steps between $A'_1$ and $A'_2$, therefore $(A'_1, R'_1) \sim_c^* (A'_2, R'_2)$.
-We aim to show that there is some sequence of $\sim$ steps from $(A_1,R_1)$ to $(A_2,R_2)$, none of whose intermediate terms pass to level $n$ or below, by analysis of the possible sequences of $\sim_c$ steps.
-
-If $A'_1 = A'_2$, as each is descended from a unique parent term, it must also be the case that $A_1 = A_2$ and $R_1 = R_2$, therefore $(A_1,R_1) \sim^* (A_2,R_2)$ with no $\sim_p$ steps to level $n$ trivially.
-
-Otherwise, consider the first $\sim_c$ step between $(A'_1, R'_1)$ and $(A'_2, R'_2)$, $(A'_1, R'_1) \sim_c (D, R'_1)$.
-If the reduction sequence $O_1 \to O'_1$ involved in the definition of $\sim_c$ in this case is not trivial (i.e.~it has at least one step), then it includes as its final step the reduction $A_1 \to A'_1$ therefore there is some equivalent $A''_1$ of $A_1$ descended from $O_2$ instead, and $(A_1, R_1) \sim_c (A''_1, R_1) \sim_p (D,R'_1)$.
-There is then a similar but strictly shorter sequence of $\sim$-steps from $(A''_1, R_1)$ to $(A_2, R_2)$, so assuming there is a solution for that one inductively, it can be composed with $(A_1, R_1) \sim_c (A''_1, R_1)$ to form a suitable sequence from $(A_1,R_1)$ to $(A_2,R_2)$.
-
-In the other case, that $O_1 = O'_1$, the reduction step from $A_1$ to $A'_1$ is the last step in the reduction sequence of $N$ to either $O_1$ or $O_2$ in one of the cases 4.a to 4.f defining $\sim_c$, which restricts the cases that are possible according to the form of the reduction $A_1 \to A'_1$.
-If $A'_1 = O_1$ resp.~$O_2$, by following the reductions $N \to A_1$ backwards then those $N \to O_2$ resp.~$O_1$ forwards, a path by which $(A_1,R_1) \sim_p^* (O_2,R'_1)$ can be constructed except in certain cases listed below, which will be dealth with separately, then as before the required relation between $(A_1,R_1)$ and $(A_2,R_2)$ can be constructed by induction.
-The cases where this doesn't work are:
-\akr{There are so many variables in scope here it's hard to avoid name clashes.
-I'll need to re-name some things later to make them more unique, but for now I'll just awkwardly avoid using some of them.}
-\begin{itemize}
-    \item Case a, when $A'_1 = O_1$ and $(O_2,R'_1)$ does not have a parent due to that position being newly generated by the last reduction, or vice-versa (swapping $O_1$ and $O_2$)
-    \item Case c, when $A'_1 = O_2$ and $R'_1 \geq P;Q$, being a position generated by the reduction $B \to B'$
-    \item Case d, when $A'_1 = O_2$ and $R'_1 \geq P;Q$, being a position generates by the reduction $R \to R'$
-    \item Case e, when $A'_1 = O_1$ and $R'_1 \geq P;R_i;Q$, where $i$ is not the last in the list, being a position generated by the reduction $B \to B'$.
-    \item Case e, when $A'_1 = O_1$ and $P;R_i;Q \geq R'_1 \geq P$, where $i$ is not the last in the list
-    \item Case f, when $A'_1 = O_1$ and $R'_1 \geq P$.
-\end{itemize}
-What all of these cases have in common is that in the branch other than the one leading to $A'_1$ (i.e.~ $N \to^* D$), $(D, R'_1)$ is newly generated by the latest reduction and is not related to anything above by $\sim_p$.
-It is therefore impossible for $D$ to be equal to $A'_2$.
-This property (that the relevant location has no parent) is preserved by any $\sim_c$ step that has its $O_1$ and $O_2$ at a strictly higher level than its $O'_1$ and $O'_2$, so consider the next $\sim_c$ step that does occur at level $n$.
-
-\akr{I want to then say that the mentioned $\sim_c$ must be the first one in reverse, because it has to be one of the cases mentioned above where the position is new, and none of the cases overlap with each other, but that doesn't quite work as planned because the intermediate $\sim_c$s at a level 1 higher could interfere with the non-overlappingness.}
-\end{proof}
-
-\akr{As this proof is takng a long time to sort out, I'm just going to assume for now some of the results, and proceed from there.
-The results I'm assuming are that if $M \to N$, the $\sim$ relation on $Rch(N)$ is equal to the restriction of the $\sim$ relation on $Rch(M)$, restricted to $Rch(N)$.}
+The $\sim$ relation can be split into $\sim_\downarrow \cup \sim_\uparrow$, where $(A,P_A) \sim_\downarrow (B,P_B)$ if either $A \to B$ and $(A,P_A) \sim_p (B,P_B)$ or $(A,P_A) \sim_c (B,P_B)$, and similarly $(A,P_A) \sim_\uparrow (B,P_B)$ if either $B \to A$ and $(A,P_A) \sim_p (B,P_B)$ or $(A,P_A) \sim_c (B,P_B)$.
+The sequence of $\sim$ relations linking $N_1$ to $N_2$ can then be expressed as an alternating sequence of $\sim_\downarrow^*$ and $\sim_\uparrow^*$s, starting with $\sim_\downarrow^*$ and ending with $\sim_\uparrow^*$, and the proof proceeds by induction on the number of these alternations.
+
+In the base case, that $(N_1,P_1) \sim_\downarrow^* X \sim_\uparrow^* (N_2,P_2)$, \dots
+
+\akr{Actually proving that this process terminates will be a pain although I'm confident it will work, but the basic idea is that the sequence of $\sim$ steps can be altered by repeatedly replacing the subsequence $\sim_c, \sim_{p,\downarrow}$ with $\sim_{p,\downarrow}, \sim_c$, the subsequence $\sim_{p,\uparrow}, \sim_c$ with $\sim_c, \sim_{p,\uparrow}$, and the subsequence $\sim_{p,\uparrow}, \sim_{p,\downarrow}$ with $\sim_{p,\downarrow}^*, \sim_c, \sim_{p,\uparrow}^*$, until everything is in the correct order.}
+\end{proof}
+
+\begin{lemma} \label{canonicalCousins}
+For any reduction sequence $X$, there is a unique reduction sequence $X_{cbv}$ that is related to $X$ by $\sim_c$ (for each position in the final term in $X$) and is in call-by-value order, and if $X \sim_c^* Y$, then $X_{cbv} = Y_{cbv}$.
+\end{lemma}
+\begin{proof}
+Note that although the reductions that occur in $X_{cbv}$ are in call-by-value order, it may not be the actual call-by-value reduction sequence because some redexes may remain un-reduced even when other reductions that should happen afterwards occur.
+It is only required that the reductions that do occur occur in the correct order.
+
+Proof sketch:
+Starting at $X$, take the last pair of adjacent reductions which aren't in CBV order, and switch them (with $\sim_c$). Repeat this process until CBV order is reached.
+
+If $Y \sim_c X \sim_c^* X_{cbv}$, and the reductions involved in $Y \sim_c X$ are not the last pair of reductions in $Y$ that aren't in CBV order, the order of the $\sim_c$s can be rearranged so that the sequence $Y \sim_c^* X_{cbv}$ is the canonical sequence given above, therefore $Y_{cbv} = X_{cbv}$.
+
+If Y is in CBV order and $Y \sim_c^* X$, then $Y_{cbv} = X_{cbv}$ but also $Y_{cbv} = Y$ therefore $X_{cbv}$ is unique.
+\end{proof}
 
@@
 \begin{proof}
+$\sim_N^*$ is trivially a subset of $\sim_M^*$ because $\sim_N$ is a subset of $\sim_M$.
+
+In the other direction, suppose $(N_1, P_1) \sim_M^* (N_2, P_2)$ where both $N_1$ and $N_2$ are descendants of $N$.
+By Lemma \ref{downAcrossUp}, take $(N_1, P_1) \sim_{p,M}^* (N'_1, P') \sim_{c,M}^* (N'_2, P') \sim_{p,M}^* (N_2, P_2)$.
+The $\sim_p^*$ steps remain within $Rch(N)$, and $\sim_p$ does not depend on the history of the reduction sequences, therefore $(N_1, P_1) \sim_{p,N}^* (N'_1, P') \sim_{c,M}^* (N'_2, P') \sim_{p,N}^* (N_2, P_2)$.
+
+Let $X$ be the call-by-value reduction sequence related to both $N'_1$ and $N'_2$ by $\sim_{c,M}$ given by Lemma \ref{canonicalCousins}.
+As $M \to N$ is the first reduction in the sequence $N'_1$, it is the last to be affected by the $\sim_{c,M}$ sequence $N'_1 \sim_{c,M}^* X$ given by Lemma \ref{canonicalCousins} therefore it can be split into $N'_1 \sim_{c,M}^* Y_1 \sim_{c,M}^* X$ where $Y_1$ is in CBV order except possibly for its first reduction, which is still $M -> N$.
+Let the position of the reduction $M \to N$ be $Q$.
+\akr{TODO: fill in the details that are missing here.
+It shouldn't be too hard, it's just a bit sketchy at present.} The rearrangement of the reduction sequence $Y_1 \sim_{c,M} X$ consists of moving the reduction at $Q$ down past the other reductions in $Y_1$, possibly duplicating or deleting it in the process, but not affecting the positions or the correct order for any of the other reductions.
+The reductions derived from $M \to N$ can be identified as follows:
+
+\begin{definition}
+A position in some (reduction sequence of) term(s) is \emph{derived from} another if it is related by the reflexive transitive closure of $\rightsquigarrow$, where $(V,R) \rightsquigarrow (W,T)$ iff $V \to W$ and one of the following cases holds:
+\begin{itemize}
+    \item $(V,R) \sim_p (W,T)$
+    \item $R = T$, $V \to W$ at $A$, and $A > R$
+    \item $V | A = (\lambda x.U) Z, U | B = x, R = A;@_2;C$ and $T = A;B;C$
+    \item $V | A = \tY (\lambda x.U), R = A;\tY;\lambda;B$ and $T = A;\lambda;@_1;B$
+    \item $V | A = \tY (\lambda x.U), U | B = x, R = A;C$ and $T = A;\lambda;@_1;B;C$
+\end{itemize}
+\end{definition}
+
+Crucially, the reductions in $X$ can be partitioned into 2 sets: those at positions derived from $(M,Q)$, and those at positions equal to the reductions in $Y_1$ (in the same order as they occur in $Y_1$.
+Using the same construction for $N'_2$ shows that the positions of the reductions in $Y_2$ are also the positions of the reductions in $X$ other than those derived from $(M,Q)$, therefore $Y_1 = Y_2$ therefore $N'_1 \sim_{c,M} Y_1 \sim_{c,M} N'_2$ and every reduction sequence in this sequence starts with $M \to N$ therefore $N'_1 \sim_{c,N} N'_2$.
+
+Combining this with the $\sim_{p,N}^*$s at the beginning and end then yields the desired result that $(N_1,P_1) \sim_N^* (N_2, P_2)$, therefore the restriction of $\sim_M^*$ to $\sim_N$'s domain ($Rch(N)$) is a subset of $\sim_N^*$ therefore the two versions of $\sim$ match on this domain, as desired.
 \end{proof}
@@
 & \qquad \text{$M \to N$ and the redex is not $\tsample$, or} \\
-& \qquad \text{$M \mid P = \tsample$, $N = M[\underline{s(M,P)}/P]$ and $\lambda$ does not occur after $@_2$ or $\tY$ in $P$}
+& \qquad \text{$M \mid P = \tsample$, $N = M[\underline{s(M,P)}/P]$ and $\lambda$ does not occur after $@_2$ of $\tY$ in $P$}
 \end{align*}
@@
 
-\begin{lemma}
-If $M \to N$, with the redex at position $P$, then no position in any term reachable from $N$ is related by $\sim$ to $(M,P)$.
+\begin{lemma} \label{simpDoesn'tMix}
+If $A$ is some descendant of $M$ and $(A,D) \sim^* (M,E)$, then $(A,D) \sim_p^* (M,E)$, with the length of the reduction sequences decreasing by one each step from $A$ to $M$.
 \end{lemma}
 \begin{proof}
+This is a simple induction on $\sim^*$.
+In the base case, $(A,D) = (M,E)$ therefore $(A,D) \sim_p^* (M,E)$ trivially.
+Otherwise, suppose that $(M,E) \sim_p^* (B,F) \sim (A,D)$.
+Either $(B,F) \sim_p (A,D)$ or $(B,F) \sim_c (A,D)$.
+In the first case, either $B \to A$, in which case the result follows directly, or $A \to B$, in which case the fact that each (reduction sequence of) term(s) has only one parent implies that a $(M,E) \sim_p^* (A,D)$ directly as a subsequence of the path to $(B,F)$.
+
+In the $\sim_c$ case, consider the definition of $\sim_c$.
+Either $O'_1 = A$ and $O'_2 = B$ or vice-versa.
+As $M \to^* N \to^* O_1 \to^* B$ and $(M,E) \sim_p^* (B,F)$, and $\sim_p$ only relates positions in a term and its parent, there are some positions $G, H$ such that $(M,E) \sim_p^* (N,G) \sim_p^* (O_1, H) \sim_p^* (B,F)$.
+It follows that $(O_2, H) \sim_p^* (A,D)$ by following the same path, therefore it suffices to provide the only missing portion of the path from $M$ to $A$, i.e.~to prove that $(N,G) \sim_p^* (O_2,H)$ given $(N,G) \sim_p^* (O_1,H)$ (or vice-versa).
+
+If $G$ is disjoint from all the positions of reduction from $N$ to $O_1$ and $O_2$ (and consequently $G = H$), this follows from case 1 of $\sim_p$.
+Otherwise, this can be proved by taking cases from the definition of $\sim_c$.
+This is rather long, but all of the cases are similar.
+The general idea is that the reductions from $N$ to $O_1$ correspond to the reductions from $N$ to $O_2$, so that if a position is related by $\sim_p$ across that reduction, it is related in the other branch for the same reason.
+Case d, where $B = O'_1$ rather than $O'_2$, is given here in more detail as an illustrative example:
+
+Let $I$ be $N$ reduced at $P$, and $J$ be $N$ reduced at $P;@_1;\lambda;Q$, so that $N \to I \to O_1$ and $N \to J \to O_2$.
+All of the reduction positions are $\geq P$, and $G$ is not disjoint from all of them, therefore $G$ is not disjoint from $P$.
+Let $K$ be the position such that $(N,G) \sim_p (I,K) \sim_p (O_1,H)$.
+The fact that $(N,G) \sim_p (I,K)$ implies that $G > P;@_1;\lambda$.
+Let $G = P;@_1;\lambda;L$ and $H = P;L'$.
+If $L$ is disjoint from $Q$, then $(N,G) \sim_p (J,G) \sim_p (O_2,P;L') = (O_2;H)$ by cases 1 and 2 of $\sim_p$.
+In the other case, that $L$ isn't disjoint from $Q$, $L > Q$ because none of the positions $<= P;@_1;\lambda;Q$ in $N$ are related to any position in $I$ by $\sim_p$.
+As $(I,P;L) \sim_p (O_1,P;L')$ (with the redex at $P;Q$), for exactly the same reason $(N,P;@_1;\lambda;L) \sim_p (J,P;@_1;\lambda;L')$ (with the redex at $P;@_1;\lambda;Q$).
+Because $(N,P;@_1;\lambda;L) \sim_p (J,P;@_1;\lambda;L')$, $N|P;@_1;\lambda;L = J|P;@_1;\lambda;L'$, and $N|P;@_1;\lambda;L \neq $ the variable of $N|P;@_1$, therefore $J|P;@_1;\lambda;L'$ is also not the variable therefore $(J,P;@_1;\lambda;L') \sim_p (O_2, P;L') = (O_2, H)$ by case 2 of $\sim_p$.
+Combining these results, $(N,G) \sim_p (O_2;H)$ as desired.
+
+\end{proof}
+
+\begin{lemma} \label{redexDestroyed}
+If $M \to N$, with the redex at position $P$, then no position in any term reachable from $N$ is related by $\sim^*$ to $(M,P)$.
+\end{lemma}
+\begin{proof}
+Suppose on the contrary that $(M,P) \sim^* (A,D)$, where $N \to^* A$, then by Lemma \ref{simpDoesn'tMix}, $(M,P) \sim_p^* (A,D)$.
+$M \neq A$ therefore $(M,P) \sim_p$ some position in $N$, but in all the cases of the definition of $\sim_p$, no position in the child term is related to the position of the redex.
 \end{proof}
@@
 Using a reduction strategy, a subset of $\Rightarrow$ that isn't nondeterministic, $\Rightarrow_f$, can be defined by $(N,s) \Rightarrow_f (N',s')$ iff $(N,s) \Rightarrow (N',s')$ and $N$ reduces to $N'$ with the redex at $f(N)$.
+
+Essentially what \ref{redexDestroyed} demonstrates is that the samples taken during any reduction sequence are independent of each other. This is made more precise in the following lemma.
+
+\begin{lemma}
+For any initial term $M$, reduction strategy $f$ on $M$, natural number $n$, measurable set $T$ of terms and measurable set $S$ of samples, 
+\end{lemma}
+\begin{proof}
+\end{proof}
 
@@
 \begin{proof}
+For both $\red$ and $\Rightarrow_{\cbv}$, the probability measure on the sample space can be used to define a measure on the reduction sequences of each finite length.
+The distributions are equal for each length, by induction, as follows.
+
+The base case, length 0, is trivial, because in both cases the distribution has probability 1 on the sequence containing only $M$.
+
+For the inductive case, let the distribution of terms after $n$ steps be $D_n$.
+This can be split into a random and a deterministic part, $D_n^r + D_n^d$, where $D_n^r$ has weight only on those terms of the form $E[\tsample]$, and $D_n^d$ only has weight on the terms not of that form.
 \end{proof}

### 2020-08-25 probabilisticProgrammingMartingales.tex
minor

--- 2020-08-06 probabilisticProgrammingMartingales.tex
@@
 
+\paragraph{Reduction relation $\Rightarrow$.}
+
 At each reduction step $M \to N$, the sample space must be restricted from $I^{L_s(M)}$ to $I^{L_s(N)}$.
@@
 \end{proof}
+
+\akr{For the standard CBV semantics, this example is indeed not a problem. 
+In the alternative semantics, defining 
+\begin{align*}
+A[x] &= \tif{\tif{x>0}{I}{I} \, \tsample - 0.5 > 0}{0}{\Omega}\\
+B &= \tif{\tsample - 0.5 > 0}{0}{\Omega}
+\end{align*}
+The set of samples where it terminates is 
+\[\begin{array}{l}
+\bigcup_{r \in [0,1]} \{s \in I^{L_s(A[\tsample])} \mid s([A[\tsample]], \mathsf{if}_1;\underline{ -
+}_1;@_1;\mathsf{if}_1) = r, \\
+\qquad \qquad \hfill \qquad \qquad s([A[\tsample], A[r], \ldots, B, \mathsf{if}_1) > 0.5 \}.
+\end{array}\] 
+This is a rather unwieldy expression, but the crucial part is that $r$ occurs twice in the conditions on $s$: once as the value a sample must take, and once in the location of a sample. 
+It is this set that's unmeasurable.}
 

### 2020-08-26 probabilisticProgrammingMartingales.tex
minor

@@
 
-Non-probabilistic lambda calculi generally have the Church-Rosser property, that if a term $A$ reduces to both $B_1$ and $B_2$, there is some $C$ with reduction sequences $B_1 \to^* C$ and $B_2 \to^* C$, so the reduction order mostly doesn't matter.
+Non-probabilistic lambda calculi generally have the Church-Rosser property, that if a term $A$ reduces to both $B_1$ and $B_2$, there is some $C$ 
+%with reduction sequences $B_1 \to^* C$ and $B_2 \to^* C$, 
+to which both $B_1$ and $B_2$ reduce,
+so the reduction order mostly doesn't matter.
 In the probabilistic case, this may not be true, because $\beta$-reduction can duplicate $\tsample$s, so the outputs of the copies of the sample may be identical or independent, depending on whether the sample is taken before or after $\beta$-reduction.
@@
 \changed[lo]{Substitution of $N$ at position $P$ in $M$, written $M[N/P]$, is defined similarly.
-Take, for example, $M = \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})$ and $P =\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2$, then 
-\[
+For example 
+\begin{align*}
+M &= \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})\\
+P &=\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2
+\end{align*} 
+then 
+\(
 M[\tsample / P] = \lambda x \, y. y \, (\tif{x < 0}{y \, \tsample}{\underline 3}).
-\]}
+\)}
 
@@
 %\lo{On first reading, I struggled to parse the following clauses.}
+\begin{definition}
+\label{def:more general red}
 \changed[lo]{The binary relation $\to$ is defined by the following rules, each is conditional on a redex occurring at position $P$ in the term $M$.}
@@
 In each of these cases, $M \mid P$ is the redex, and reduction takes place at $P$.
-%\lo{The definition above is clear. But as a general rule, it is good to illustrate definitions -- especially if they are genuinely new -- with an example or two.}
+\end{definition}
 
@@
 Labelling the pre-chosen samples by the positions in the term would also not work because in some cases, a $\tsample$ will be duplicated before being reduced, for example, in $(\lambda x.
-{\underline x \, 0} \, \mathbin{\underline{+}} \, {x \, \underline 0}) (\lambda y.
+x \, {\underline 0} \, \mathbin{\underline{+}} \, {x \, \underline 0}) (\lambda y.
 \tsample)$, both of the $\tsample$ redexes that eventually occur originate at $@_2 ; \lambda$.
@@
 
-Define a skeleton, to be a term but instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term has a skeleton and each skeleton can be converted to a term given the appropriate number of real numbers to substitute in.
-The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder}{A}{B}$ reducing nondeterministically to both $A$ and $M$.
+Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term has a skeleton and each skeleton can be converted to a term given the appropriate number of real numbers to substitute in.
+The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder}{A}{B}$ reducing nondeterministically to both $A$ and $B$.
 
@@
 
-The reflexive transitive closure $\sim^*$ of this relation is used to define the set of potential positions $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable terms.
+The reflexive transitive closure $\sim^*$ of this relation is used to define the set of \emph{potential positions} $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable terms.
 If $(N,P) \sim^* (O,Q)$, then $N \mid P$ and $O \mid Q$ both have the same shape (i.e.~they're either both variables, both applications, both $\tsample$s etc.), therefore it's well-defined to talk of the set of potential positions where there is a $\tsample$, $L_s(M)$.
+\lo{Thus $L_s(M) := \{[(X, P)]_{\sim^\ast_M} : X \mid P = \tsample\}$.}
 The new sample space is then defined as $I^{L_s(M)}$, with the Borel $\sigma$-algebra and product measure.
@@
 \end{lemma}
+
+\lo{To be clear, should write out the definition of $X \sim_c Y$.
+Let the final term of the reduction sequences $X$ and $Y$ be $\hat{X}$ and $\hat{Y}$ respectively. 
+Write $X \sim_c Y$ to mean for every position $P$ that occurs in $\hat X$, $(\hat X, P) \sim_c (\hat Y, P)$.
+}
+\lo{As defined, $\sim_c$ is not symmetric.}
+
 \begin{proof}
@@
 It is only required that the reductions that do occur occur in the correct order.
+\lo{This paragraph (especially the last sentence) is ambiguous.
+What does it mean to say that that a reductions that occur in a reduction sequences are \emph{in CBV order}?}
+\lo{So this is permitted?
+\[
+(II)(I (I \bar I)) \to
+(II)(I\bar I) \to
+(II)\bar I 
+\]
+but not this: 
+\(
+(II)(I (I \bar I)) \to
+(II)(I \bar I) \to
+(II)\bar I 
+\)?}
 
@@
 In the other direction, suppose $(N_1, P_1) \sim_M^* (N_2, P_2)$ where both $N_1$ and $N_2$ are descendants of $N$.
-By Lemma \ref{downAcrossUp}, take $(N_1, P_1) \sim_{p,M}^* (N'_1, P') \sim_{c,M}^* (N'_2, P') \sim_{p,M}^* (N_2, P_2)$.
+By Lemma~\ref{downAcrossUp}, take $(N_1, P_1) \sim_{p,M}^* (N'_1, P') \sim_{c,M}^* (N'_2, P') \sim_{p,M}^* (N_2, P_2)$.
 The $\sim_p^*$ steps remain within $Rch(N)$, and $\sim_p$ does not depend on the history of the reduction sequences, therefore $(N_1, P_1) \sim_{p,N}^* (N'_1, P') \sim_{c,M}^* (N'_2, P') \sim_{p,N}^* (N_2, P_2)$.
 
-Let $X$ be the call-by-value reduction sequence related to both $N'_1$ and $N'_2$ by $\sim_{c,M}$ given by Lemma \ref{canonicalCousins}.
+Let $X$ be the call-by-value reduction sequence related to both $N'_1$ and $N'_2$ by $\sim_{c,M}$ given by Lemma~\ref{canonicalCousins}. 
 As $M \to N$ is the first reduction in the sequence $N'_1$, it is the last to be affected by the $\sim_{c,M}$ sequence $N'_1 \sim_{c,M}^* X$ given by Lemma \ref{canonicalCousins} therefore it can be split into $N'_1 \sim_{c,M}^* Y_1 \sim_{c,M}^* X$ where $Y_1$ is in CBV order except possibly for its first reduction, which is still $M -> N$.
 Let the position of the reduction $M \to N$ be $Q$.
-\akr{TODO: fill in the details that are missing here.
-It shouldn't be too hard, it's just a bit sketchy at present.} The rearrangement of the reduction sequence $Y_1 \sim_{c,M} X$ consists of moving the reduction at $Q$ down past the other reductions in $Y_1$, possibly duplicating or deleting it in the process, but not affecting the positions or the correct order for any of the other reductions.
+\akr{TODO: fill in the details that are missing here. It shouldn't be too hard, it's just a bit sketchy at present.} 
+The rearrangement of the reduction sequence $Y_1 \sim_{c,M} X$ consists of moving the reduction at $Q$ down past the other reductions in $Y_1$, possibly duplicating or deleting it in the process, but not affecting the positions or the correct order for any of the other reductions.
 The reductions derived from $M \to N$ can be identified as follows:
@@
 \begin{definition}
-A position in some (reduction sequence of) term(s) is \emph{derived from} another if it is related by the reflexive transitive closure of $\rightsquigarrow$, where $(V,R) \rightsquigarrow (W,T)$ iff $V \to W$ and one of the following cases holds:
+A position in some (reduction sequence of) term(s) is \emph{derived from} another if it is related by the reflexive transitive closure of $\rightsquigarrow$, where $(V,R) \rightsquigarrow (W,T)$ just if $V \to W$ and one of the following cases holds:
 \begin{itemize}
@@
 The term $M | P_1$ can't be $\tsample$, because it has strict subterms, but the case that $M | P_2 = \tsample$ is still somewhat more complicated.
-$P_2$ can't be within $P_1 ; @_2$ or $P_1 ; Y$ (cases e and f of $\sim_c$) because those are not valid positions to reduce a $\tsample$, and the case that $P_2$ is within the branch of an if statement that is deleted by the reduction at $P_1$ (case b) doesn't actually present a problem because there is no corresponding reduction to $M \to M_2$ in the other branch, which leaves cases c and d, that $P_2$ is in the $\tif$ branch that isn't deleted, and that $P_2 < P_1 ; @_1 ; \lambda$.
+$P_2$ can't be within $P_1 ; @_2$ or $P_1 ; Y$ (cases e and f of $\sim_c$) because those are not valid positions to reduce a $\tsample$, and the case that $P_2$ is within the branch of an if statement that is deleted by the reduction at $P_1$ (case b) doesn't actually present a problem because there is no corresponding reduction to $M \to M_2$ in the other branch, which leaves cases c and d, that $P_2$ is in the $\tif{\cdot}{\cdot}{\cdot}$ branch that isn't deleted, and that $P_2 < P_1 ; @_1 ; \lambda$.
 The values that the $\tsample$s take in these cases match because $(M,P) \sim_p (M_1, P;Q)$ by cases 3 and 2 of $\sim_p$ respectively.
@@
 A reduction strategy starting from a closed term $M$ is a partial function $f$ from $Rch(M)$ to positions, such that for any reachable term $N$ where $f$ is defined, $f(N)$ is a position of a valid redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
+\lo{Redexes are defined in Definition~\ref{def:more general red}. What is an invalid redex?}
 Using a reduction strategy, a subset of $\Rightarrow$ that isn't nondeterministic, $\Rightarrow_f$, can be defined by $(N,s) \Rightarrow_f (N',s')$ iff $(N,s) \Rightarrow (N',s')$ and $N$ reduces to $N'$ with the redex at $f(N)$.
 
-Essentially what \ref{redexDestroyed} demonstrates is that the samples taken during any reduction sequence are independent of each other. This is made more precise in the following lemma.
+Essentially what Lemma~\ref{redexDestroyed} demonstrates is that the samples taken during any reduction sequence are independent of each other. This is made more precise in the following lemma.
 
@@
 \cbv(T M) & = @_1 ; \cbv(T) \\
-\cbv(V T) & = @_1 ; \cbv(T) \\
+\cbv(V T) & = {\color{red} @_2} ; \cbv(T) \\
 \cbv(\underline f(V_1, \dots, V_{k-1}, T, M_{k+1}, \dots, M_n)) & = \underline f_k ; \cbv(T) \\

### 2020-08-26 probabilisticProgrammingMartingales.tex
minor

@@
 
-A position \akr{I'm not really sure how to phrase the citation, but this is adapted from \cite{Kennaway96infinitarylambda}.} is a finite sequence of steps into a term, defined inductively as
+A \emph{position} 
+\akr{I'm not really sure how to phrase the citation, but this is adapted from \cite{Kennaway96infinitarylambda}.} 
+\lo{The idea is common enough. I think no citation necessary.}
+is a finite sequence of steps into a term, defined inductively as
 \begin{align*}
@@
 \end{align*}
-The \changed[lo]{\emph{subterm of $M$ at a position $P$}, denoted $M \mid P$,} is defined as
+The \emph{subterm of $M$ at a position $P$}, denoted $M \mid P$, is defined as
 \begin{align*}
@@
 A position such that $M\mid P$ does exist is said to occur in $M$. 
-\changed[lo]{Substitution of $N$ at position $P$ in $M$, written $M[N/P]$, is defined similarly.
+\changed[lo]{\emph{Substitution} of $N$ at position $P$ in $M$, written $M[N/P]$, is defined similarly.
 For example 
@@
 \begin{itemize}
-    \item If $M \to N$, the relation $\sim_N^*$ defined on $Rch(N)$ is equal to the restriction of $\sim_M^*$ to $Rch(N)$ \lo{Define $\sim_N^*$.}
+    \item If $M \to N$, the relation $\sim_N^*$ defined on $Rch(N)$ is equal to the restriction of $\sim_M^*$ to $Rch(N)$ %\lo{Define $\sim_N^*$.}
     \item Given the definition below of $\Rightarrow$, which is still applicable to this version of $\sim$, and which is the equivalent of $\red$ in this version of the semantics, $\Rightarrow$ is Church-Rosser.
@@
 What does it mean to say that that a reductions that occur in a reduction sequences are \emph{in CBV order}?}
+\iffalse
 \lo{So this is permitted?
@@
 \)?}
+\fi
 
@@
 
-\akr{For the standard CBV semantics, this example is indeed not a problem. 
+\lo{Give an example or two of a reduction strategy (for $\Rightarrow$) that is not $\cbv$.}
+
+\akr{NOTE. For the standard CBV semantics, this example is indeed not a problem. 
 In the alternative semantics, defining 

### 2020-08-29 probabilisticProgrammingMartingales.tex
Finish some proofs relating to independence of samples.

--- 2020-08-26 probabilisticProgrammingMartingales.tex
@@
 
+If $(X,P) \sim_c (Y,P)$ then the final skeletons in $X$ and $Y$ are equal, and $(X,Q) \sim_c (Y,Q)$, therefore $\sim_c$ can be considered to apply to reduction sequences even without positions, $X \sim_c Y$ iff $(X,\cdot) \sim_c (Y,\cdot)$.
+
 \begin{lemma} \label{canonicalCousins}
@@
 
-\lo{To be clear, should write out the definition of $X \sim_c Y$.
-Let the final term of the reduction sequences $X$ and $Y$ be $\hat{X}$ and $\hat{Y}$ respectively. 
-Write $X \sim_c Y$ to mean for every position $P$ that occurs in $\hat X$, $(\hat X, P) \sim_c (\hat Y, P)$.
-}
 \lo{As defined, $\sim_c$ is not symmetric.}
@@
 The reduction relation $\Rightarrow$ is nondeterministic, so it admits multiple possible reduction strategies.
-A reduction strategy starting from a closed term $M$ is a partial function $f$ from $Rch(M)$ to positions, such that for any reachable term $N$ where $f$ is defined, $f(N)$ is a position of a valid redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
-\lo{Redexes are defined in Definition~\ref{def:more general red}. What is an invalid redex?}
+A reduction strategy starting from a closed term $M$ is a partial function $f$ from $Rch(M)$ to positions, such that for any reachable term $N$ where $f$ is defined, $f(N)$ is a position of a redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
 Using a reduction strategy, a subset of $\Rightarrow$ that isn't nondeterministic, $\Rightarrow_f$, can be defined by $(N,s) \Rightarrow_f (N',s')$ iff $(N,s) \Rightarrow (N',s')$ and $N$ reduces to $N'$ with the redex at $f(N)$.
 
-Essentially what Lemma~\ref{redexDestroyed} demonstrates is that the samples taken during any reduction sequence are independent of each other. This is made more precise in the following lemma.
+Essentially what Lemma~\ref{redexDestroyed} demonstrates is that the samples taken during any reduction sequence are independent of each other. This is made more precise in the following lemmas.
 
 \begin{lemma}
-For any initial term $M$, reduction strategy $f$ on $M$, natural number $n$, measurable set $T$ of terms and measurable set $S$ of samples, 
+For any skeletons $M \to N$, with the redex at position $P$, and measurable set of samples $S \subset I^{L_s(N)}$, $\mu(S) = \mu(\{s \in I^{L_s(M)} \mid s \circ i(M \to N) \in S\})$, and furthermore, if $M | P = \tsample$, for any $S \subset I \times I^{L_s(N)}$, $\mu(S) = \mu(\{s \in I^{L_s(M)} \mid (s(M,P), s \circ i(M \to N)) \in S\})$.
 \end{lemma}
 \begin{proof}
+If the result holds for all sets $S$ of the form $\{s \in I^{L_s(N)} \mid \forall j : s(j) \in x_i\}$, where $(x_j)_{j \in J}$ is a family of measurable subsets of $I$ indexed by some finite set $J \subset L_s(N)$ of positions, it also holds for all other $S$ by taking limits and disjoint unions.
+
+Take such a $(x_j)_{j \in J}$, where $x_j = x_k$ and define $K = i(M \to N)[J]$.
+Because $i(M \to N)$ is injective, it defines a bijection between $J$ and $K$.
+We can then calculate the measure $\mu(\{s \in I^{L_s(M)} \mid s \circ i(M \to N) \in S\}) = \mu(\{s \in I^{L_s(M)} \mid \forall k : s(k) \in x_{i(M \to N)^{-1}(k)}\}) = \prod_{k \in K} \mu_I(x_{i(M \to N)^{-1}(k)}) = \prod_{j \in J} \mu_I(x_j) = \mu(S)$.
+
+In the case that $M|P = \tsample$, we can similarly consider only those sets $S$ of the form $x_\tsample \times \{s \in I^{L_s(N)} \mid \forall j : s(j) \in x_i\}$.
+The position $P$ in $M$ is not related to any position in $L_s(N)$, therefore $(M,P) \not \in K$.
+Again, we can calculate the measure $\mu(\{s \in I^{L_s(M)} \mid (s(M,P), s \circ i(M \to N)) \in S\}) = \mu(\{s \in I^{L_s(M)} \mid s(M,P) \in x_\tsample, \forall k : s(k) \in x_{i(M \to N)^{-1}(k)}\}) = \mu_I(x_\tsample) \prod_{k \in K} \mu_I(x_{i(M \to N)^{-1}(k)}) = \mu_I(x_\tsample) \prod_{j \in J} \mu_I(x_j) = \mu(S)$.
+\end{proof}
+
+\begin{lemma}
+For any initial term $M$, reduction strategy $f$ on $M$, natural number $n$, skeleton $N$ with $k$ holes, measurable set $T \subset \mathbb R ^ k$ and measurable set $S$ of samples in $I^{L_s(N)}$, $\mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')\}) = \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in I^{L_s(N)} : (M,s) \Rightarrow_f^n (N[r], s')\}) \mu(S)$
+\end{lemma}
+\begin{proof}
+Suppose, to begin with, that $n = 0$.
+Either $\exists r \in T : N[r] = M$, in which case both sides of the equation are $\mu(S)$, or there is no such $r$, in which case both sides of the equation are 0.
+
+For $n > 0$, suppose for induction that the lemma is true for $n - 1$, for all $N, T$ and $S$.
+If $\exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')$, the $r$ and $s'$ are necessarily unique, therefore we may assume that $T$ is the product of $k$ measurable setsets of $\mathbb R$, $(T_j)_{0 \leq j < k}$, as all of the other cases follow by taking unions and limits of these rectangles.
+Let\begin{align*}
+R_s & = \{\text{the skeleton of } O \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) = \tsample, O \to N[r]\} \\
+R_d & = \{\text{the skeleton of } O \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) \neq \tsample, O \to N[r]\}.
+\end{align*}
+Together, $R_s$ and $R_d$ contain all of the skeletons of terms that can reduce to $N$, and they're disjoint.
+For each of these skeletons, there is a map $g_O$ from the reals in $O$ to the reals in $N$, i.e.
+$g_O(r) = r'$ if $O[r] \to N[r']$ for $O \in R_d$, and $g_O(s, r) = r'$ if $O[r] \to N[r']$ with $s$ as the value the sample takes in the reduction.
+All of these functions are measurable, as they are just rearrangements of vector components, or in the case that $O|f(O) = \underline x(\skeletonPlaceholder, \dots, \skeletonPlaceholder)$, a combination fo the measurable function $x$ with a rearrangement of vector components.
+In the case that $O \in R_s$, the function $g_O$ simply inserts the sample into the vector of reals at a certain index, call it $h_O$.
+\begin{align*}
+& \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')\}) \\
+= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in S : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) + \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in S : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \\
+= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}: g_O(r,s'(O,f(O))) \in T, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) + \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, s' \circ i(O[r] \to N[g_O(r)]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \\
+= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)}: s'(O,f(O)) \in T_{h(O)}, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) + \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, s' \circ i(O[r] \to N[g_O(r)]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \\
+= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(\{s' \in I^{L_s(O)} \mid s'(O,f(O)) \in T_{h(O)}, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S\}) + \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu({s' \in I^{L_s(O)} \mid s' \circ i(O[r] \to N[g_O(r)]) \in S}) \\
+= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(T_{h(O)}) \mu(S) + \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
+= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : s'(O,f(O)) \in T_{h(O)}, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) + \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
+= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \mathbb R^{k-1}, s' \in I^{L_s(O)} : g_O(r,s'(O,f(O))) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) + \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
+= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S) + \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S) \\
+= & \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in I^{L_s(N)} : (M,s) \Rightarrow_f^n (N[r], s')\}) \mu(S)
+\end{align*}
 \end{proof}

### 2020-08-30 probabilisticProgrammingMartingales.tex
Terms -> skeletons: some changes and an example. Def. of I^{L_s(M)}

@@
 \usepackage{enumitem}
-\newcommand{\tY}{\textsf{Y}}
+\newcommand{\tY}{\mathsf{Y}}
 %\newcommand{\tif}[3]{\textsf{if }#1\textsf{ then }#2\textsf{ else }#3}
-\newcommand{\tif}[3]{\textsf{if}\,(#1, #2, #3)} % space-saving concrete syntax
-\newcommand{\tsample}{\textsf{sample}}
-\newcommand{\tscore}{\textsf{score}}
-\newcommand{\skeletonPlaceholder}{\textsf{X}} % I'm not sure what the best notation here is.
+\newcommand{\tif}[3]{\mathsf{if}(#1, #2, #3)} % space-saving concrete syntax
+\newcommand{\tsample}{\mathsf{sample}}
+\newcommand{\tscore}{\mathsf{score}}
+\newcommand{\skeletonPlaceholder}{\mathsf{X}} % I'm not sure what the best notation here is.
 \DeclareMathOperator{\red}{red}
@@
 \subsection{Sampling semantics}
+\label{sec:sampling semantics}
 This version of the reduction relation allows $\tsample$ to reduce to any number in $[0,1]$.
@@
 A \emph{position} 
-\akr{I'm not really sure how to phrase the citation, but this is adapted from \cite{Kennaway96infinitarylambda}.} 
-\lo{The idea is common enough. I think no citation necessary.}
+%\akr{I'm not really sure how to phrase the citation, but this is adapted from \cite{Kennaway96infinitarylambda}.} 
+%\lo{The idea is common enough. I think no citation necessary.}
 is a finite sequence of steps into a term, defined inductively as
@@
 so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda ; \cdot$ is undefined). 
-A position such that $M\mid P$ does exist is said to occur in $M$. 
-\changed[lo]{\emph{Substitution} of $N$ at position $P$ in $M$, written $M[N/P]$, is defined similarly.
-For example 
-\begin{align*}
-M &= \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})\\
-P &=\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2
-\end{align*} 
+A position such that $M\mid P$ does exist is said to \emph{occur} in $M$. 
+%\changed[lo]{
+\emph{Substitution} of $N$ at position $P$ in $M$, written $M[N/P]$, is defined similarly.
+For example, let 
+\[
+M = \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})
+\qquad
+P =\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2
+\]
 then 
@@
 M[\tsample / P] = \lambda x \, y. y \, (\tif{x < 0}{y \, \tsample}{\underline 3}).
-\)}
+\)
+%}
 
@@
 If neither $P_1 \leq P_2$ nor $P_1 \geq P_2$, the positions are said to be disjoint.
-This is mostly relevant in that in any substitution at some position, the subterms at all other positions are unaffected. \lo{The preceding sentence is unclear. Typo? Mostly $\mapsto$ most?}
+This is mostly relevant in that in any substitution at some position, the subterms at all other positions are unaffected. 
+%\lo{The preceding sentence is unclear. Typo? Mostly $\mapsto$ most?}
 
@@
 \label{def:more general red}
-\changed[lo]{The binary relation $\to$ is defined by the following rules, each is conditional on a redex occurring at position $P$ in the term $M$.}
+The binary relation $\to$ is defined by the following rules, each is conditional on a redex occurring at position $P$ in the term $M$:
 \begin{align*}
@@
 \end{align*}
-In each of these cases, $M \mid P$ is the redex, and reduction takes place at $P$.
+In each of these cases, $M \mid P$ is the \emph{redex}, and reduction takes place at $P$.
 \end{definition}
@@
 It is therefore necessary to consider possible positions that may occur in other terms reachable from the original term. 
-This is itself inadequate because some of the positions in different reachable terms need to be considered the same, and the number of reachable terms is in general uncountable, which leads to measure-theoretic issues.
-
-Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term has a skeleton and each skeleton can be converted to a term given the appropriate number of real numbers to substitute in.
-The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder}{A}{B}$ reducing nondeterministically to both $A$ and $B$.
-
-Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a reduction sequence of skeletons starting at $M$, and the second of which is a position in the final skeleton of the reduction sequence.
+Even this is itself inadequate because some of the positions in different reachable terms need to be considered the same, and the number of reachable terms is in general uncountable, which leads to measure-theoretic issues.
+
+Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has placeholders $\skeletonPlaceholder$, so that each term has a skeleton and each skeleton can be converted to a term given the appropriate number of real numbers to substitute in.
+\lo{Skeleton is defined in Sec.~\ref{sec:sampling semantics}.}
+The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, with \changed[lo]{$\tif{\skeletonPlaceholder < 0}{A}{B}$} reducing nondeterministically to both $A$ and $B$, \changed[lo]{and $M \to M[\skeletonPlaceholder' / P]$ for a freshly generated placeholder $\skeletonPlaceholder'$, in case $M \mid P = \tsample$ and $\lambda$ does not occur after $@_2$ or $\tY$ in $P$.
+(By abuse of notation, we use the same metavariables $M, N$, etc.~for skeletons as for terms.)}
+\changed[lo]{For example, we have 
+\(
+(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder_1}) \, \tsample
+\to
+(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder_1}) \, \skeletonPlaceholder_2
+\to
+\tif{\skeletonPlaceholder_2 < 0}{\skeletonPlaceholder_2}{\skeletonPlaceholder_1}
+\to
+\skeletonPlaceholder_i
+\) where $i$ is 1 or 2.}
+
+Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a \changed[lo]{$\to$-reduction} sequence of skeletons starting at \changed[lo]{the skeleton of} $M$, and the second of which is a position in the final skeleton of the reduction sequence.
 \akr{I've yet to go through and replace references to terms with references to skeletons, but it should be reasonably obvious where the substitution should take place. 
-This is potentially even more confusing though because I'm now implicitly identifying terms, skeletons, reduction sequences and skeletal reduction sequences.} Reduction sequences are used rather than reachable terms because if the same term is reached twice, different samples may be needed, but they will be discussed as though they were terms, by abuse of notation.
-\lo{Readers would benefit greatly from some examples of an $L_0(M)$-pair, and the relations $\sim_p$ and $\sim_c$, here, before the formal definitions.
-Explain the underlying intuition.} The relation $\sim$ is defined as the union of the symmetric relations $\sim_p$ \changed[lo]{($p$ for parent-child relationship)} and $\sim_c$ \changed[lo]{($c$ for cousin relationship)} where
+This is potentially even more confusing though because I'm now implicitly identifying terms, skeletons, reduction sequences and skeletal reduction sequences.} 
+Reduction sequences are used rather than reachable \changed[lo]{skeletons} because if the same \changed[lo]{skeleton} is reached twice, different samples may be needed, \lo{Give a simple example illustrating this.} but they will be discussed as though they were terms, by abuse of notation.
+
+\lo{Readers would benefit greatly from some examples of an $L_0(M)$-pair, and the relations $\sim_p$ and $\sim_c$, here, before the formal definitions. Explain the underlying intuition.}
+
+\begin{definition}
+The relation $\sim$ is defined as the union of the symmetric relations $\sim_p$ \changed[lo]{($p$ for parent-child)} and $\sim_c$ \changed[lo]{($c$ for cousin)} where
 \lo{As defined, $\sim_p$ is not a symmetric relation.}
@@
 \end{enumerate}
+See \Cref{fig:adk-diagram} for a graphical depiction of $\sim_p$- and $\sim_c$-rules.
+\end{definition}
 
@@
 
-\begin{example}
-See \Cref{fig:adk-diagram} for a graphical depiction of $\sim_p$- and $\sim_c$-rules.
-\end{example}
-
-The reflexive transitive closure $\sim^*$ of this relation is used to define the set of \emph{potential positions} $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable terms.
+\lo{TODO: Typeset Fig.~\ref{fig:adk-diagram} in vector graphics using PGF/TikZ.}
+
+The reflexive transitive closure $\sim^*$ of this relation is used to define the set of \emph{potential positions} $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable \changed[lo]{skeletons}.
 If $(N,P) \sim^* (O,Q)$, then $N \mid P$ and $O \mid Q$ both have the same shape (i.e.~they're either both variables, both applications, both $\tsample$s etc.), therefore it's well-defined to talk of the set of potential positions where there is a $\tsample$, $L_s(M)$.
-\lo{Thus $L_s(M) := \{[(X, P)]_{\sim^\ast_M} : X \mid P = \tsample\}$.}
+\changed[lo]{Formally $L_s(M) := \{[(X, P)]_{\sim^\ast_M} : X \mid P = \tsample\}$.}
 The new sample space is then defined as $I^{L_s(M)}$, with the Borel $\sigma$-algebra and product measure.
+\changed[lo]{Since $I^{L_s(M)}$ is a countable product, the measure space is well-defined \citep[Cor.~2.7.3]{AshDD00}.}
+
+\iffalse
+\changed[lo]{Since $L_s(M)$ is a countable set, there is a unique probability measure $\mu$ on the infinite product $\prod_{i \in L_s(M)} \Sigma_I$ of the Borel $\sigma$-algebras $\Sigma_I$ satisfying
+\[
+\mu\set{\omega \in \Omega \mid \omega_1 \in A_1, \cdots, \omega_n \in A_n} = \prod_{i=1}^n \Leb(A_i)
+\]
+where $\Leb$ is the Lebesgue measure on the measurable space $(I, \Sigma_I)$.}
+\fi
 
 \paragraph{}
-For now, here's a summary of the lemmas and theorems I intend to prove about this alternative semantics:
+\akr{For now, here's a summary of the lemmas and theorems I intend to prove about this alternative semantics:
 \begin{itemize}
@@
     \item Corresponding sparse and antitone-strict ranking function theorems.
-\end{itemize}
+\end{itemize}}
 
@@
 \begin{lemma}
+\label{lem:sim-M-N}
 The relation $\sim$ is defined on $L_0(M)$ with reference to a particular starting term $M$, so different versions, $\sim_M$ and $\sim_N$, can be defined starting at different terms.
@@
 \end{lemma}
+
 \begin{proof}
@@
 
-\begin{definition}
+%\begin{definition}
 A position in some (reduction sequence of) term(s) is \emph{derived from} another if it is related by the reflexive transitive closure of $\rightsquigarrow$, where $(V,R) \rightsquigarrow (W,T)$ just if $V \to W$ and one of the following cases holds:
@@
 \end{itemize}
-\end{definition}
+%\end{definition}
 
@@
 At each reduction step $M \to N$, the sample space must be restricted from $I^{L_s(M)}$ to $I^{L_s(N)}$. 
-The injection $L_0(N) \to L_0(M)$ is trivial to define by appending $M \to N$ to each path, and using the previous lemma, this induces a corresponding injection on the quotient, $L(N) \to L(M)$.
+The injection $L_0(N) \to L_0(M)$ is trivial to define by appending $M \to N$ to each path, and using Lemma~\ref{lem:sim-M-N}, this induces a corresponding injection on the quotient, $L(N) \to L(M)$. 
 The corresponding map $L_s(N) \to L_s(M)$ is then denoted $i(M \to N)$.
@@
 Combining these results, $(N,G) \sim_p (O_2;H)$ as desired.
-
 \end{proof}
 
-\begin{lemma} \label{redexDestroyed}
+\begin{lemma} \label{lem:redexDestroyed}
 If $M \to N$, with the redex at position $P$, then no position in any term reachable from $N$ is related by $\sim^*$ to $(M,P)$.
@@
 The reduction relation $\Rightarrow$ is nondeterministic, so it admits multiple possible reduction strategies. 
-A reduction strategy starting from a closed term $M$ is a partial function $f$ from $Rch(M)$ to positions, such that for any reachable term $N$ where $f$ is defined, $f(N)$ is a position of a redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
-Using a reduction strategy, a subset of $\Rightarrow$ that isn't nondeterministic, $\Rightarrow_f$, can be defined by $(N,s) \Rightarrow_f (N',s')$ iff $(N,s) \Rightarrow (N',s')$ and $N$ reduces to $N'$ with the redex at $f(N)$.
-
-Essentially what Lemma~\ref{redexDestroyed} demonstrates is that the samples taken during any reduction sequence are independent of each other. This is made more precise in the following lemmas.
+A \emph{reduction strategy} starting from a closed term $M$ is a partial function $f$ from $Rch(M)$ to positions, such that for any reachable term $N$ where $f$ is defined, $f(N)$ is a position of a redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
+Using a reduction strategy $f$, a subset of $\Rightarrow$ that isn't nondeterministic, $\Rightarrow_f$, can be defined by $(N,s) \Rightarrow_f (N',s')$ just if $(N,s) \Rightarrow (N',s')$ and $N$ reduces to $N'$ with the redex at $f(N)$.
+
+Essentially what Lemma~\ref{lem:redexDestroyed} demonstrates is that the samples taken during any reduction sequence are independent of each other. This is made more precise in the following lemmas.
 
@@
 \cbv(T M) & = @_1 ; \cbv(T) \\
-\cbv(V T) & = {\color{red} @_2} ; \cbv(T) \\
+\cbv(V T) & = @_2 ; \cbv(T) \\
 \cbv(\underline f(V_1, \dots, V_{k-1}, T, M_{k+1}, \dots, M_n)) & = \underline f_k ; \cbv(T) \\
@@
 
-\begin{theorem} \label{AstEquivalence}
+\begin{theorem} \label{thm:AstEquivalence}
 A closed term $M$ is AST with respect to $\cbv$ iff it is AST.
@@
 
-\begin{theorem} \label{CbvIsTerminatingest}
+\begin{theorem} \label{thm:CbvIsTerminatingest}
 If $M$ terminates with some reduction strategy $f$ and samples $s$, it terminates with $\cbv$ and $s$.
@@
 Let the set of samples with which it terminates with this reduction strategy be $X$.
-By \ref{CbvIsTerminatingest}, $M$ also terminates with $\cbv$ and every element of $X$, and $X$ has measure 1, by assumption, therefore $M$ is AST with respect to $\cbv$ therefore by \ref{AstEquivalence} it is AST.
+By Theorem~\ref{thm:CbvIsTerminatingest}, $M$ also terminates with $\cbv$ and every element of $X$, and $X$ has measure 1, by assumption, therefore $M$ is AST with respect to $\cbv$ therefore by Theorem~\ref{thm:AstEquivalence} it is AST.
 \end{proof}

### 2020-08-31 probabilisticProgrammingMartingales.tex
Clarify some things relating to skeletons.

--- 2020-08-29 probabilisticProgrammingMartingales.tex
@@
 The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder}{A}{B}$ reducing nondeterministically to both $A$ and $B$.
-
-Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a reduction sequence of skeletons starting at $M$, and the second of which is a position in the final skeleton of the reduction sequence.
+If $S$ is a skeleton containing $n$ placeholders and $r \in \mathbb R^n$, $S[r]$ is the term corresponding to $S$ with the numbers from $r$ substituted for $S$'s placeholders, and conversely, if $T$ is a term, $Sk(T)$ is its skeleton.
+
+Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a reduction sequence of skeletons starting at $Sk(M)$, and the second of which is a position in the final skeleton of the reduction sequence.
 \akr{I've yet to go through and replace references to terms with references to skeletons, but it should be reasonably obvious where the substitution should take place.
@@
 At each reduction step $M \to N$, the sample space must be restricted from $I^{L_s(M)}$ to $I^{L_s(N)}$.
-The injection $L_0(N) \to L_0(M)$ is trivial to define by appending $M \to N$ to each path, and using the previous lemma, this induces a corresponding injection on the quotient, $L(N) \to L(M)$.
+The injection $L_0(N) \to L_0(M)$ is trivial to define by appending $Sk(M) \to Sk(N)$ to each path, and using the previous lemma, this induces a corresponding injection on the quotient, $L(N) \to L(M)$.
 The corresponding map $L_s(N) \to L_s(M)$ is then denoted $i(M \to N)$.
@@
 & \qquad \text{$M \to N$ and the redex is not $\tsample$, or} \\
-& \qquad \text{$M \mid P = \tsample$, $N = M[\underline{s(M,P)}/P]$ and $\lambda$ does not occur after $@_2$ or $\tY$ in $P$}
+& \qquad \text{$M \mid P = \tsample$, $N = M[\underline{s(Sk(M),P)}/P]$ and $\lambda$ does not occur after $@_2$ or $\tY$ in $P$}
 \end{align*}

### 2020-08-31 probabilisticProgrammingMartingales.tex
Fix merge conflict, oops.

--- 2020-08-30 probabilisticProgrammingMartingales.tex
@@
 
-Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has placeholders $\skeletonPlaceholder$, so that each term has a skeleton and each skeleton can be converted to a term given the appropriate number of real numbers to substitute in.
+Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term has a skeleton $Sk(M)$ and each skeleton can be converted to a term $S[r]$ given a vector of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$.
 \lo{Skeleton is defined in Sec.~\ref{sec:sampling semantics}.}
-The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, with \changed[lo]{$\tif{\skeletonPlaceholder < 0}{A}{B}$} reducing nondeterministically to both $A$ and $B$, \changed[lo]{and $M \to M[\skeletonPlaceholder' / P]$ for a freshly generated placeholder $\skeletonPlaceholder'$, in case $M \mid P = \tsample$ and $\lambda$ does not occur after $@_2$ or $\tY$ in $P$.
+The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, \changed[lo]{and $M \to M[\skeletonPlaceholder / P]$, in case $M \mid P = \tsample$ and $\lambda$ does not occur after $@_2$ or $\tY$ in $P$.
 (By abuse of notation, we use the same metavariables $M, N$, etc.~for skeletons as for terms.)}
-\changed[lo]{For example, we have 
+For example, we have 
 \(
-(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder_1}) \, \tsample
+(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \tsample
 \to
-(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder_1}) \, \skeletonPlaceholder_2
+(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \skeletonPlaceholder
 \to
-\tif{\skeletonPlaceholder_2 < 0}{\skeletonPlaceholder_2}{\skeletonPlaceholder_1}
+\tif{\skeletonPlaceholder < 0}{\skeletonPlaceholder}{\skeletonPlaceholder}
 \to
-\skeletonPlaceholder_i
-\) where $i$ is 1 or 2.}
-
-Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a \changed[lo]{$\to$-reduction} sequence of skeletons starting at \changed[lo]{the skeleton of} $M$, and the second of which is a position in the final skeleton of the reduction sequence.
+\skeletonPlaceholder
+\)
+
+Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a $\to$-reduction sequence of skeletons starting at $Sk(M)$, and the second of which is a position in the final skeleton of the reduction sequence.
 \akr{I've yet to go through and replace references to terms with references to skeletons, but it should be reasonably obvious where the substitution should take place. 
@@
 At each reduction step $M \to N$, the sample space must be restricted from $I^{L_s(M)}$ to $I^{L_s(N)}$. 
-The injection $L_0(N) \to L_0(M)$ is trivial to define by appending $M \to N$ to each path, and using Lemma~\ref{lem:sim-M-N}, this induces a corresponding injection on the quotient, $L(N) \to L(M)$. 
+The injection $L_0(N) \to L_0(M)$ is trivial to define by appending $Sk(M) \to Sk(N)$ to each path, and using Lemma~\ref{lem:sim-M-N}, this induces a corresponding injection on the quotient, $L(N) \to L(M)$.
 The corresponding map $L_s(N) \to L_s(M)$ is then denoted $i(M \to N)$.

### 2020-08-31 probabilisticProgrammingMartingales.tex
minor

--- 2020-08-30 probabilisticProgrammingMartingales.tex
@@
 The reflexive transitive closure $\sim^*$ of this relation is used to define the set of \emph{potential positions} $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable \changed[lo]{skeletons}.
-If $(N,P) \sim^* (O,Q)$, then $N \mid P$ and $O \mid Q$ both have the same shape (i.e.~they're either both variables, both applications, both $\tsample$s etc.), therefore it's well-defined to talk of the set of potential positions where there is a $\tsample$, $L_s(M)$.
+If $(N,P) \sim^* (O,Q)$, then $N \mid P$ and $O \mid Q$ both have the same shape (i.e.~they're either \changed[lo]{both placeholders}, both variables, both applications, both $\tsample$s etc.), therefore it's well-defined to talk of the set of potential positions where there is a $\tsample$, $L_s(M)$.
 \changed[lo]{Formally $L_s(M) := \{[(X, P)]_{\sim^\ast_M} : X \mid P = \tsample\}$.}
@@
     \item $R = T$, $V \to W$ at $A$, and $A > R$
-    \item $V | A = (\lambda x.U) Z, U | B = x, R = A;@_2;C$ and $T = A;B;C$
-    \item $V | A = \tY (\lambda x.U), R = A;\tY;\lambda;B$ and $T = A;\lambda;@_1;B$
-    \item $V | A = \tY (\lambda x.U), U | B = x, R = A;C$ and $T = A;\lambda;@_1;B;C$
+    \item $V \mid A = (\lambda x.U) Z, U | B = x, R = A;@_2;C$ and $T = A;B;C$
+    \item $V \mid A = \tY (\lambda x.U), R = A;\tY;\lambda;B$ and $T = A;\lambda;@_1;B$
+    \item $V \mid A = \tY (\lambda x.U), U | B = x, R = A;C$ and $T = A;\lambda;@_1;B;C$
 \end{itemize}
@@
 At each reduction step $M \to N$, the sample space must be restricted from $I^{L_s(M)}$ to $I^{L_s(N)}$. 
-The injection $L_0(N) \to L_0(M)$ is trivial to define by appending $M \to N$ to each path, and using Lemma~\ref{lem:sim-M-N}, this induces a corresponding injection on the quotient, $L(N) \to L(M)$. 
+The injection $L_0(N) \to L_0(M)$ is trivial to define by \changed[lo]{prepending} $M \to N$ to each path, and using Lemma~\ref{lem:sim-M-N}, this induces a corresponding injection on the quotient, $L(N) \to L(M)$.
 The corresponding map $L_s(N) \to L_s(M)$ is then denoted $i(M \to N)$.
@@
 As the positions are disjoint, the substitutions commute and both $M_1$ and $M_2$ reduce (with $\to$) to $M'$.
-Let $s' = s \circ i(M \to M_1) \circ(M_1 \to M')$.
-The injection $i(M \to M_1) \circ i(M_1 \to M')$ consists of appending $M \to M_1 \to M'$ to each reduction sequence, but by case 4.a of $\sim$, this is equivalent to appending $M \to M_2 \to M'$, which is $i(M \to M_2) \circ i(M_2 \to M')$.
+Let $s' = s \circ i(M \to M_1) \circ i(M_1 \to M')$. 
+The injection $i(M \to M_1) \circ i(M_1 \to M')$ consists of \changed[lo]{prepending} $M \to M_1 \to M'$ to each reduction sequence, but by case 4.a of $\sim$, this is equivalent to \changed[lo]{prepending} $M \to M_2 \to M'$, which is $i(M \to M_2) \circ i(M_2 \to M')$.
 In the case that the reduction $M \to M_2$ isn't a $\tsample$-reduction, this is enough to establish that $(M_1, s_1) \Rightarrow (M', s')$, and vice-versa. 
@@
 For each possible case of what type of redex $M | P_1$ is, and $P_2$'s position within it, there is a corresponding case of $\sim_c$, and similarly to the case where $P_1$ and $P_2$ are disjoint, the term $O_1 = O_2$ from the definition of $\sim_c$ is a suitable value of $M'$.
-The term $M | P_1$ can't be $\tsample$, because it has strict subterms, but the case that $M | P_2 = \tsample$ is still somewhat more complicated.
-$P_2$ can't be within $P_1 ; @_2$ or $P_1 ; Y$ (cases e and f of $\sim_c$) because those are not valid positions to reduce a $\tsample$, and the case that $P_2$ is within the branch of an if statement that is deleted by the reduction at $P_1$ (case b) doesn't actually present a problem because there is no corresponding reduction to $M \to M_2$ in the other branch, which leaves cases c and d, that $P_2$ is in the $\tif{\cdot}{\cdot}{\cdot}$ branch that isn't deleted, and that $P_2 < P_1 ; @_1 ; \lambda$.
+The term $M \mid P_1$ can't be $\tsample$, because it has strict subterms, but the case that $M \mid P_2 = \tsample$ is still somewhat more complicated. 
+$P_2$ can't be within $P_1 ; @_2$ or $P_1 ; Y$ (cases e and f of $\sim_c$) because those are not valid positions to reduce a $\tsample$, 
+and the case that $P_2$ is within the branch of an if statement that is deleted by the reduction at $P_1$ (case b) doesn't actually present a problem because there is no corresponding reduction to $M \to M_2$ in the other branch,
+which leaves cases c and d, that $P_2$ is in the $\tif{\cdot}{\cdot}{\cdot}$ branch that isn't deleted, and that $P_2 < P_1 ; @_1 ; \lambda$.  
 The values that the $\tsample$s take in these cases match because $(M,P) \sim_p (M_1, P;Q)$ by cases 3 and 2 of $\sim_p$ respectively.

### 2020-08-31 probabilisticProgrammingMartingales.tex
Write the proof of thm:AstEquivalence

--- 2020-08-31 probabilisticProgrammingMartingales.tex
@@
 
-\begin{lemma}
+\begin{lemma} \label{lem:independentSamples}
 For any initial term $M$, reduction strategy $f$ on $M$, natural number $n$, skeleton $N$ with $k$ holes, measurable set $T \subset \mathbb R ^ k$ and measurable set $S$ of samples in $I^{L_s(N)}$, $\mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')\}) = \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in I^{L_s(N)} : (M,s) \Rightarrow_f^n (N[r], s')\}) \mu(S)$
@@
 \begin{proof}
+\akr{I'm not sure whether this proof is sufficiently detailed. It would be pretty easy to fill it in more, just very verbose.}
+
 For both $\red$ and $\Rightarrow_{\cbv}$, the probability measure on the sample space can be used to define a measure on the reduction sequences of each finite length.
@@
 
-For the inductive case, let the distribution of terms after $n$ steps be $D_n$.
-This can be split into a random and a deterministic part, $D_n^r + D_n^d$, where $D_n^r$ has weight only on those terms of the form $E[\tsample]$, and $D_n^d$ only has weight on the terms not of that form.
+For the inductive case, suppose that the distributions of reduction sequences of length $n$ are equal.
+\akr{TODO: mention the case that it has already terminated by step $n$, or is about to.
+The two versions of the semantics handle it a little differently.}
+
+For any term $N$ which is not a value, there is a unique environment $E$ and redex $R$ such that $N = E[R]$.
+The position of the hole in $E$ is equal to $\cbv(N)$, so that $N|\cbv(N) = R$, as the cases in the definition of $\cbv$ match the cases in the definition of environments.
+If $R \new \tsample$, let $R'$ be the result of reducing $R$ (which is equal in both versions of the semantics), then $E[R'] = N[R' / \cbv(N)]$ therefore $\red(N,s) = (E[R'],s)$ and $(N,s) \Rightarrow_{\cbv} (E[R'],s)$.
+The distribution of next terms, conditional on the previous term, in the case that that previous term reduces deterministically, is therefore equal for $\red$ and $\Rightarrow_{\cbv}$.
+
+In the other case that $R = \tsample$, the first (term) part of $\red(N,s)$ is $E[\underline{\pi_h(s)}]$.
+If $k$ is the number of samples already taken in this reduction sequence, this is equivalent to $E[\underline{s_0(k)}]$, where $s_0$ is the initial sample.
+For any sufficiently small neighbourhood of $N$ (i.e.~containing only reduction sequences with the same skeletons), the probability of reaching this neighbourhood is independent of $s(k)$, because it depends only on the samples taken so far, and the samples are independent.
+The distribution of $s(k)$ conditional on reaching $N$ is therefore the uniform distribution on $I$.
+Similarly for the other semantics, if the distribution of remaining samples after $n$ steps is independent of the term by Lemma \ref{lem:independentSamples}, therefore the distribution of $s(Sk(N),\cbv(N))$ conditional on $N$ is the uniform distribution on $I$.
+In either case, the distribution of the next term conditional on the previous term, in the case that it reduces randomly, is equal to the image under $r \mapsto E[\underline r]$ of the uniform distribution on $I$.
+
+In any case, the distribution of reduction sequences after $n+1$ steps, conditional on the reduction sequence after $n$ steps, as defined by $\Rightarrow_\cbv$ and by $\red$, is equal, any by the inductive hypothesis the distributions after $n$ steps are equal, therefore by integrating over the reduction sequence up to step $n$, the distributions on reduction sequences up to step $n+1$ are equal.
+
+The distributions on reduction sequences of any finite length as defined by $\Rightarrow_{\cbv}$ and $\red$ are therefore equal, therefore so is the probability of having reached a value after $n$ steps, therefore so is its limit, the probability of termination, therefore the probability of termination is 1 iff the probability of termination with respect to $\cbv$ is 1.
 \end{proof}

### 2020-09-03 probabilisticProgrammingMartingales.tex
Fix typos

--- 2020-08-31 probabilisticProgrammingMartingales.tex
@@
 The position of the hole in $E$ is equal to $\cbv(N)$, so that $N|\cbv(N) = R$, as the cases in the definition of $\cbv$ match the cases in the definition of environments.
-If $R \new \tsample$, let $R'$ be the result of reducing $R$ (which is equal in both versions of the semantics), then $E[R'] = N[R' / \cbv(N)]$ therefore $\red(N,s) = (E[R'],s)$ and $(N,s) \Rightarrow_{\cbv} (E[R'],s)$.
+If $R \neq \tsample$, let $R'$ be the result of reducing $R$ (which is equal in both versions of the semantics), then $E[R'] = N[R' / \cbv(N)]$ therefore $\red(N,s) = (E[R'],s)$ and $(N,s) \Rightarrow_{\cbv} (E[R'],s)$.
 The distribution of next terms, conditional on the previous term, in the case that that previous term reduces deterministically, is therefore equal for $\red$ and $\Rightarrow_{\cbv}$.
@@
 
-In any case, the distribution of reduction sequences after $n+1$ steps, conditional on the reduction sequence after $n$ steps, as defined by $\Rightarrow_\cbv$ and by $\red$, is equal, any by the inductive hypothesis the distributions after $n$ steps are equal, therefore by integrating over the reduction sequence up to step $n$, the distributions on reduction sequences up to step $n+1$ are equal.
+In any case, the distribution of reduction sequences after $n+1$ steps, conditional on the reduction sequence after $n$ steps, as defined by $\Rightarrow_{\cbv}$ and by $\red$, is equal, any by the inductive hypothesis the distributions after $n$ steps are equal, therefore by integrating over the reduction sequence up to step $n$, the distributions on reduction sequences up to step $n+1$ are equal.
 

### 2020-09-06 probabilisticProgrammingMartingales.tex
minor

@@
 The position of the hole in $E$ is equal to $\cbv(N)$, so that $N|\cbv(N) = R$, as the cases in the definition of $\cbv$ match the cases in the definition of environments.
-If $R \neq \tsample$, let $R'$ be the result of reducing $R$ (which is equal in both versions of the semantics), then $E[R'] = N[R' / \cbv(N)]$ therefore $\red(N,s) = (E[R'],s)$ and $(N,s) \Rightarrow_{\cbv} (E[R'],s)$.
+If $R \not= \tsample$, let $R'$ be the result of reducing $R$ (which is equal in both versions of the semantics), then $E[R'] = N[R' / \cbv(N)]$ therefore $\red(N,s) = (E[R'],s)$ and $(N,s) \Rightarrow_{\cbv} (E[R'],s)$.
 The distribution of next terms, conditional on the previous term, in the case that that previous term reduces deterministically, is therefore equal for $\red$ and $\Rightarrow_{\cbv}$.

### 2020-09-09 probabilisticProgrammingMartingales.tex
Minor: some terms -> skeleton; Proof of Theorem 1

--- 2020-09-03 probabilisticProgrammingMartingales.tex
@@
 
-Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term has a skeleton $Sk(M)$ and each skeleton can be converted to a term $S[r]$ given a vector of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$.
-\lo{Skeleton is defined in Sec.~\ref{sec:sampling semantics}.}
-The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, \changed[lo]{and $M \to M[\skeletonPlaceholder / P]$, in case $M \mid P = \tsample$ and $\lambda$ does not occur after $@_2$ or $\tY$ in $P$.
-(By abuse of notation, we use the same metavariables $M, N$, etc.~for skeletons as for terms.)}
+Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has \changed[lo]{(possibly more than one occurrence of the same)} placeholder $\skeletonPlaceholder$, so that each term has a skeleton $S = Sk(M)$ and each skeleton can be converted to a term $S[\overline r]$ given a $\overline r$ vector of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$.
+\lo{Need to resolve skeleton here with earlier definition in Sec.~\ref{sec:sampling semantics}.}
+The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, with \changed[lo]{$\tif{\skeletonPlaceholder < 0}{A_1}{A_2} \to A_i$ nondeterministically; and $M \to M[\skeletonPlaceholder / P]$, in case $M \mid P = \tsample$ and $\lambda$ does not occur after $@_2$ or $\tY$ in $P$.
+(By abuse of notation, we use the same metavariables $M, N$, etc.~for skeletons as for terms.
+It should be clear from the context which is meant.)}
 For example, we have 
@@
 This is potentially even more confusing though because I'm now implicitly identifying terms, skeletons, reduction sequences and skeletal reduction sequences.} 
-Reduction sequences are used rather than reachable \changed[lo]{skeletons} because if the same \changed[lo]{skeleton} is reached twice, different samples may be needed, \lo{Give a simple example illustrating this.} but they will be discussed as though they were terms, by abuse of notation.
+Reduction sequences of skeletons are used rather than reachable \changed[lo]{skeletons} because if the same \changed[lo]{skeleton} is reached twice, different samples may be needed, \lo{Give a simple example illustrating this.} but they will be discussed as though they were \changed[lo]{skeletons}, by abuse of notation.
 
@@
 
-The reflexive transitive closure $\sim^*$ of this relation is used to define the set of \emph{potential positions} $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable \changed[lo]{skeletons}.
-If $(N,P) \sim^* (O,Q)$, then $N \mid P$ and $O \mid Q$ both have the same shape (i.e.~they're either \changed[lo]{both placeholders}, both variables, both applications, both $\tsample$s etc.), therefore it's well-defined to talk of the set of potential positions where there is a $\tsample$, $L_s(M)$.
+The reflexive transitive closure of this relation $\sim$, written $\sim^*_M$, is used to define the set of \emph{potential positions} $L(M) = L_0(M) / \sim^*_M$, and each equivalence class can be considered as the same position as it may occur across multiple reachable \changed[lo]{skeletons}.
+If $(N,P) \sim^*_M (O,Q)$, then $N \mid P$ and $O \mid Q$ both have the same shape (i.e.~they're either \changed[lo]{both placeholders}, both variables, both applications, both $\tsample$s etc.), therefore it's well-defined to talk of the set of potential positions where there is a $\tsample$, $L_s(M)$.
 \changed[lo]{Formally $L_s(M) := \{[(X, P)]_{\sim^\ast_M} : X \mid P = \tsample\}$.}
 The new sample space is then defined as $I^{L_s(M)}$, with the Borel $\sigma$-algebra and product measure.
-\changed[lo]{Since $I^{L_s(M)}$ is a countable product, the measure space is well-defined \citep[Cor.~2.7.3]{AshDD00}.}
+\changed[lo]{Since $I^{L_s(M)}$ is a countable product, the measure space is well-defined (see e.g.~\citep[Cor.~2.7.3]{AshDD00}).}
+\changed[lo]{(To avoid notational clutter, we will often elide the subscript from $\sim^\ast_M$ whenever the starting skeleton $M$ of the reduction sequences is clear form the context.)}
 
@@
 \begin{lemma} \label{downAcrossUp}
-If $(N_1,P_1) \sim^* (N_2,P_2)$, there is some descendants $N'_1, N'_2$ of $N_1$ resp.~$N_2$, and a position $P'$ in both of them such that $(N_1,P_1) \sim_p^* (N'_1, P') \sim_c^* (N'_2, P') \sim_p^* (N_2, P_2)$.
+If $(N_1,P_1) \sim^* (N_2,P_2)$, there are some descendants $N'_1, N'_2$ of $N_1$ resp.~$N_2$, and a position $P'$ in both of them such that $(N_1,P_1) \sim_p^* (N'_1, P') \sim_c^* (N'_2, P') \sim_p^* (N_2, P_2)$.
 \end{lemma}
@@
 \end{align*}
+\lo{$M$ and $N$ above are skeleton. So we should have $N = Sk\big(M[\underline{s(Sk(M),P)}/P]\big)$.}
 
@@
 The reduction relation $\Rightarrow$ is nondeterministic, so it admits multiple possible reduction strategies. 
-A \emph{reduction strategy} starting from a closed term $M$ is a partial function $f$ from $Rch(M)$ to positions, such that for any reachable term $N$ where $f$ is defined, $f(N)$ is a position of a redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
+A \emph{reduction strategy} starting from a closed skeleton $M$ is a partial function $f$ from $Rch(M)$ to positions, such that for any reachable skeleton $N$ where $f$ is defined, $f(N)$ is a position of a redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
 Using a reduction strategy $f$, a subset of $\Rightarrow$ that isn't nondeterministic, $\Rightarrow_f$, can be defined by $(N,s) \Rightarrow_f (N',s')$ just if $(N,s) \Rightarrow (N',s')$ and $N$ reduces to $N'$ with the redex at $f(N)$.
@@
 \begin{theorem} \label{thm:AstEquivalence}
-A closed term $M$ is AST with respect to $\cbv$ iff it is AST.
+A closed skeleton $M$ is AST with respect to $\cbv$ iff it is AST.
 \end{theorem}
@@
 
-For both $\red$ and $\Rightarrow_{\cbv}$, the probability measure on the sample space can be used to define a measure on the reduction sequences of each finite length.
+For both $\red$ \lo{We have defined the reduction relation $\to$ on skeletons, but not (yet) $\red$.} and $\Rightarrow_{\cbv}$, the probability measure on the sample space can be used to define a measure on the reduction sequences of each finite length.
 The distributions are equal for each length, by induction, as follows.
@@
 
-For any term $N$ which is not a value, there is a unique environment $E$ and redex $R$ such that $N = E[R]$.
-The position of the hole in $E$ is equal to $\cbv(N)$, so that $N|\cbv(N) = R$, as the cases in the definition of $\cbv$ match the cases in the definition of environments.
+For any skeleton $N$ which is not a value, there is a unique \changed[lo]{context} $E$ and redex $R$ such that $N = E[R]$. 
+The position of the hole in $E$ is equal to $\cbv(N)$, so that $N \mid \cbv(N) = R$, as the cases in the definition of $\cbv$ match the cases in the definition of \changed[lo]{context}. 
 If $R \neq \tsample$, let $R'$ be the result of reducing $R$ (which is equal in both versions of the semantics), then $E[R'] = N[R' / \cbv(N)]$ therefore $\red(N,s) = (E[R'],s)$ and $(N,s) \Rightarrow_{\cbv} (E[R'],s)$.
-The distribution of next terms, conditional on the previous term, in the case that that previous term reduces deterministically, is therefore equal for $\red$ and $\Rightarrow_{\cbv}$.
-
-In the other case that $R = \tsample$, the first (term) part of $\red(N,s)$ is $E[\underline{\pi_h(s)}]$.
+The distribution of next skeletons, conditional on the previous skeleton, in the case that that previous skeleton reduces deterministically, is therefore equal for $\red$ and $\Rightarrow_{\cbv}$.
+
+In the other case that $R = \tsample$, the first (skeleton) part of $\red(N,s)$ is $E[\underline{\pi_h(s)}]$. \lo{This should be $Sk(E[\underline{\pi_h(s)}])$.} 
 If $k$ is the number of samples already taken in this reduction sequence, this is equivalent to $E[\underline{s_0(k)}]$, where $s_0$ is the initial sample.
+\lo{Should introduce a name to replace sample here.}
 For any sufficiently small neighbourhood of $N$ (i.e.~containing only reduction sequences with the same skeletons), the probability of reaching this neighbourhood is independent of $s(k)$, because it depends only on the samples taken so far, and the samples are independent.
 The distribution of $s(k)$ conditional on reaching $N$ is therefore the uniform distribution on $I$. 
-Similarly for the other semantics, if the distribution of remaining samples after $n$ steps is independent of the term by Lemma \ref{lem:independentSamples}, therefore the distribution of $s(Sk(N),\cbv(N))$ conditional on $N$ is the uniform distribution on $I$.
-In either case, the distribution of the next term conditional on the previous term, in the case that it reduces randomly, is equal to the image under $r \mapsto E[\underline r]$ of the uniform distribution on $I$.
-
-In any case, the distribution of reduction sequences after $n+1$ steps, conditional on the reduction sequence after $n$ steps, as defined by $\Rightarrow_{\cbv}$ and by $\red$, is equal, any by the inductive hypothesis the distributions after $n$ steps are equal, therefore by integrating over the reduction sequence up to step $n$, the distributions on reduction sequences up to step $n+1$ are equal.
-
-The distributions on reduction sequences of any finite length as defined by $\Rightarrow_{\cbv}$ and $\red$ are therefore equal, therefore so is the probability of having reached a value after $n$ steps, therefore so is its limit, the probability of termination, therefore the probability of termination is 1 iff the probability of termination with respect to $\cbv$ is 1.
+Similarly for the other semantics, if the distribution of remaining samples after $n$ steps is independent of the skeleton by Lemma \ref{lem:independentSamples}, therefore the distribution of $s(Sk(N),\cbv(N))$ conditional on $N$ is the uniform distribution on $I$.
+In either case, the distribution of the next skeleton conditional on the previous skeleton, in the case that it reduces randomly, is equal to the image under $r \mapsto E[\underline r]$ of the uniform distribution on $I$.
+
+In any case, the distribution of reduction sequences after $n+1$ steps, conditional on the reduction sequence after $n$ steps, as defined by $\Rightarrow_{\cbv}$ and by $\red$, is equal; 
+and by the inductive hypothesis the distributions after $n$ steps are equal, therefore by integrating over the reduction sequence up to step $n$, the distributions on reduction sequences up to step $n+1$ are equal.
+
+The distributions on reduction sequences of any finite length as defined by $\Rightarrow_{\cbv}$ and $\red$ are therefore equal; 
+therefore so is the probability of having reached a value after $n$ steps; 
+therefore so is its limit, the probability of termination; 
+therefore the probability of termination is 1 iff the probability of termination with respect to $\cbv$ is 1.
 \end{proof}

### 2020-09-09 probabilisticProgrammingMartingales.tex
Added a comment on the "dependent-type style" representation of elements of \biguplus_{M \in \Lambda_0} I^{L_s(M)}

--- 2020-09-03 probabilisticProgrammingMartingales.tex
@@
 Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function, so it is denoted ``$\Rightarrow$'' instead of ``$\red$'', and it relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
+\changed[lo]{We write an element of $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ as $(M', s)$ where the term $M' \in \Lambda^0$ identifies the component of the disjoint set, and $s \in  I^{L_s(M')}$.}
 \begin{align*}
@@
 Let $s' = s \circ i(M \to M_1) \circ i(M_1 \to M')$. 
-The injection $i(M \to M_1) \circ i(M_1 \to M')$ consists of \changed[lo]{prepending} $M \to M_1 \to M'$ to each reduction sequence, but by case 4.a of $\sim$, this is equivalent to \changed[lo]{prepending} $M \to M_2 \to M'$, which is $i(M \to M_2) \circ i(M_2 \to M')$.
+The injection $i(M \to M_1) \circ i(M_1 \to M')$ consists of prepending $M \to M_1 \to M'$ to each reduction sequence, but by case 4.a of $\sim$, this is equivalent to prepending $M \to M_2 \to M'$, which is $i(M \to M_2) \circ i(M_2 \to M')$.
 In the case that the reduction $M \to M_2$ isn't a $\tsample$-reduction, this is enough to establish that $(M_1, s_1) \Rightarrow (M', s')$, and vice-versa. 

### 2020-09-15 probabilisticProgrammingMartingales.tex
Add diagram to illustrate \sim_c using TikZ
The diagram currently looks really messy, but it's substantially complete.
It now requires lualatex to compile, because of the graph layouts

--- 2020-09-06 probabilisticProgrammingMartingales.tex
@@
 \usepackage{graphicx}
-
+\usepackage{tikz}
+\usetikzlibrary{positioning, graphs, quotes, graphdrawing, shapes.geometric}
+\usegdlibrary{layered, circular}
 \usepackage[square]{natbib}
@@
 \begin{figure}
-\includegraphics[height=0.9\textheight]{akr-diagram.pdf}
-\caption{\label{fig:adk-diagram} Some examples of the old version of the $\sim_p$- and $\sim_c$-rules}
+\include{sim-diagram}
+\caption{\label{fig:adk-diagram} Some examples of the $\sim_p$- and $\sim_c$-rules}
 \end{figure}
@@
 The position of the hole in $E$ is equal to $\cbv(N)$, so that $N|\cbv(N) = R$, as the cases in the definition of $\cbv$ match the cases in the definition of environments.
-If $R \not= \tsample$, let $R'$ be the result of reducing $R$ (which is equal in both versions of the semantics), then $E[R'] = N[R' / \cbv(N)]$ therefore $\red(N,s) = (E[R'],s)$ and $(N,s) \Rightarrow_{\cbv} (E[R'],s)$.
+If $R \neq \tsample$, let $R'$ be the result of reducing $R$ (which is equal in both versions of the semantics), then $E[R'] = N[R' / \cbv(N)]$ therefore $\red(N,s) = (E[R'],s)$ and $(N,s) \Rightarrow_{\cbv} (E[R'],s)$.
 The distribution of next terms, conditional on the previous term, in the case that that previous term reduces deterministically, is therefore equal for $\red$ and $\Rightarrow_{\cbv}$.

### 2020-09-15 probabilisticProgrammingMartingales.tex
minor

--- 2020-09-09 probabilisticProgrammingMartingales.tex
@@
 Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function, so it is denoted ``$\Rightarrow$'' instead of ``$\red$'', and it relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
-\changed[lo]{We write an element of $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ as $(M', s)$ where the term $M' \in \Lambda^0$ identifies the component of the disjoint set, and $s \in  I^{L_s(M')}$.}
+\changed[lo]{We write an element of $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ as $(M', s)$ where the \emph{term} $M' \in \Lambda^0$ identifies the component of the disjoint set, and $s \in  I^{L_s(M')}$.}
 \begin{align*}

### 2020-09-17 probabilisticProgrammingMartingales.tex
Fix the \sim_c diagram
It's still ugly, but it at least no longer has any overlaps or
extends beyond the pages.

--- 2020-09-09 probabilisticProgrammingMartingales.tex
@@
 \end{enumerate}
-See \Cref{fig:adk-diagram} for a graphical depiction of $\sim_p$- and $\sim_c$-rules.
+The $\sim_c$-rules are illustrated in the following diagrams.
 \end{definition}
 
-\begin{figure}
-\includegraphics[height=0.9\textheight]{akr-diagram.pdf}
-\caption{\label{fig:adk-diagram} Some examples of the old version of the $\sim_p$- and $\sim_c$-rules}
-\end{figure}
+\include{sim-diagram}
 

### 2020-09-17 probabilisticProgrammingMartingales.tex
minor

@@
 \end{enumerate}
-The $\sim_c$-rules are illustrated in the following diagrams.
+See \Cref{fig:adk-diagram} for a graphical depiction of $\sim_p$- and $\sim_c$-rules.
 \end{definition}
 
+\begin{figure}
 \include{sim-diagram}
+\caption{\label{fig:adk-diagram} Some examples of the $\sim_p$- and $\sim_c$-rules}
+\end{figure}
 
@@
 Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function, so it is denoted ``$\Rightarrow$'' instead of ``$\red$'', and it relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
-\changed[lo]{We write an element of $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ as $(M', s)$ where the term $M' \in \Lambda^0$ identifies the component of the disjoint set, and $s \in  I^{L_s(M')}$.}
+\changed[lo]{We write an element of $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ as $(M', s)$ where the \emph{term} $M' \in \Lambda^0$ and $s \in  I^{L_s(M')}$.}
 \begin{align*}

### 2020-09-28 probabilisticProgrammingMartingales.tex
Start writing out the full proof of the canonicalCousins lemma.

--- 2020-09-17 probabilisticProgrammingMartingales.tex
@@
 \begin{lemma} \label{canonicalCousins}
-For any reduction sequence $X$, there is a unique reduction sequence $X_{cbv}$ that is related to $X$ by $\sim_c$ (for each position in the final term in $X$) and is in call-by-value order, and if $X \sim_c^* Y$, then $X_{cbv} = Y_{cbv}$.
+For any reduction sequence $X$, there is a unique reduction sequence $X_{cbv}$ that is related to $X$ by $\sim_c$ and is in call-by-value order, and if $X \sim_c^* Y$, then $X_{cbv} = Y_{cbv}$.
 \end{lemma}
@@
 \begin{proof}
+A reduction sequence is in call-by-value order if for every pair of reductions in the sequence at positions $\alpha$ and $\beta$, where the reduction at $\alpha$ happens first, either $\alpha \leq \beta$ or $\alpha$ and $\beta$ are disjoint, with $\alpha$ to the left of $\beta$ (i.e.~the first elements of the sequences $\alpha, \beta$ where they differ are $@_1$ and $@_2$, $\underscore f_i$ and $\underscore f_j$ for $i < j$, or $\textsf{if}_i$ and $\textsf{if}_j$ for $i < j$, in that order.
 Note that although the reductions that occur in $X_{cbv}$ are in call-by-value order, it may not be the actual call-by-value reduction sequence because some redexes may remain un-reduced even when other reductions that should happen afterwards occur.
@@
 
-Proof sketch:
-Starting at $X$, take the last pair of adjacent reductions which aren't in CBV order, and switch them (with $\sim_c$). Repeat this process until CBV order is reached.
-
+Define $X_n$ recursively so that $X_0 = X$ and if $X_n$ is not in CBV order, take the last pair of reductions in $X_n$ which aren't in CBV order.
+They are adjacent in the sequence, as the CBV order is a total order.
+Let the positions of these reductions be $\alpha$ and $\beta$ respecively.
+If they are disjoint, they can be considered as one of the branches of $\sim_c$ case a.
+Otherwise, $\beta < \alpha$, and the sub-skeleton at position $\beta$ at the appropriate point in the reduction sequence is a redex with non-trivial subterms, therefore it's of one of the forms $\tif{\skeletonPlaceholder>0}{A}{B}$, $(\lambda x.A) B$ or $\tY \lambda x.
+A$, therefore the reductions $\alpha$ and $\beta$ form the $O_2$ branch of one of the cases b, c, d, e or f of $\sim_c$.
+In either case, define $X_{n+1}$ as equal to $X_n$ except taking the other (left) branch, so that the order of the reductions at $\alpha$ and $\beta$ are switched (although the equivalent(s) of the reduction at $\alpha$ may not actually be at that position).
+
+This sequence of swaps rearranges the reductions in $X$ rather like insertion-sort (but in some cases duplicating or deleting the new element), and eventually reaches some $X_n$ that is in CBV order.
+This is defined to be $X_{cbv}$.
+
+\paragraph{}
 If $Y \sim_c X \sim_c^* X_{cbv}$, and the reductions involved in $Y \sim_c X$ are not the last pair of reductions in $Y$ that aren't in CBV order, the order of the $\sim_c$s can be rearranged so that the sequence $Y \sim_c^* X_{cbv}$ is the canonical sequence given above, therefore $Y_{cbv} = X_{cbv}$.
-
+To be more precise about this rearrangement, consider the sequence of $\sim_c$ steps from $Y$ to $X_{cbv}$.
+The first step takes some sub-sequence of the reduction sequence $Y$ and swaps the order of those reductions.
+If $X = X_{cbv}$, then $X = Y_1 = Y_{cbv}$ already.
+Otherwise, consider the cases according to whether the image of this subsequence in $X$ occurs earlier than the next pair of reductions to be swapped (i.e.~the last pair of reductions in $X$ that occur not in CBV order).
+If not, then either the step $Y \sim_c X$ changes this subsequence into CBV order, or out of it.
+If it is into, then it was the last out-of-order pair in $Y$ therefore the whole sequence $Y \sim_c^* X_{cbv}$ is already in the canonical order and $Y_{cbv} = X_{cbv}$.
+If it switches the pair to the wrong order, then the result of that switch is the last out-of-order pair in $X$ therefore $Y = X_1$, and the sequence $X_1 \sim_c^* X_{cbv}$ is just a suffix of the sequence $X \sim_c^* X_{cbv}$ therefore $(X_1)_{cbv} = X_{cbv}$ and again $Y_{cbv} = X_{cbv}$.
+
+The remaining cases are when the subsequence of reduction steps involved in $Y \sim_c X$ occur before the last out-of-order pair in $X$, but possible overlapping.
+If there is no overlap, then the $\sim_c$ steps do not interfere with each other at all, and can simply be performed in the other order.
+This gives a different sequence of $\sim_c$ steps $Y \sim_c Y_1 \sim_c X_1 \sim_c^* X_{cbv}$.
+As $Y_{cbv} = (Y_1)_{cbv}$, we can proceed by induction in this case and use the fact that $(Y_1)_{cbv} = (X_1)_{cbv} = X_{cbv}$ to acheive the result.
+
+In the other case, there is some overlap between the regions involved in $Y \sim_c X$ and $X \sim_c X_1$, but the last reduction in $X \sim_c X_1$ is later than the last reduction in $Y \sim_c X$.
+There are only 2 reductions in $X$ involved in $X \sim_c X_1$, therefore the region involved in $Y \sim_c X$ ends on the first of the reductions involved in $X \sim_c X_1$.
+Now consider the specific sequence of positions of the reductions in $X$ involved in either of these steps.
+Let the last of these positions be $\alpha$, the first $\gamma$, and the second-last, third-last and so on $\beta_1, \beta_2, \dots$ respectively, and let the position of the first relevant reduction in $Y$ be $\delta$.
+Then $X \sim_c X_1$ swaps $\beta_1$ and $\alpha$, and $\alpha$ comes before $\beta_1$ in CBV order (i.e.~$\alpha < \beta_1$ or $\alpha$ is left of $\beta_1$), and the step $Y \sim_c X$ results in $\gamma, \beta_k, \dots, \beta_1$.
+In each of the cases below, it is required to show that the canonical sequence $(X_i)_i$ eventually reaches some term which is the same as one in $(Y_j)_j$, and from that point on, they match, therefore they reach the same end result: $X_{cbv} = Y_{cbv}$.
+
+\begin{itemize}
+    \item $\gamma, \beta_k, \dots, \beta_1$ has length 2, $\alpha < \beta_1$ and $\gamma$ is right of $\alpha$: In this case, the sequence $(X_i)_i$ starts by swapping $\alpha$ and $\beta_1$ by one of the cases b to f in the definition of $\sim_c$, then swapping $\gamma$ down past $\alpha$ by case a, then down past all the images of $\beta_1$, also by case a.
+Eventually it reaches some $X_i$ where the relevant sub-reduction-sequence is $\alpha$ then whatever $\beta_1$ was turned into in $X \sim_c X_1$ then $\gamma$, before moving on to swapping pairs earlier in the reduction sequence.
+The sequence $(Y_j)_j$ starts by swapping $\gamma$ down past $\alpha$ by case a, then swapping $\alpha$ and $\beta_1$ as above.
+The subskeleton at $\alpha$ at the relevant point in the reduction sequence if the same as in the $(X_i)_i$ case because $\gamma$ and $\alpha$ are disjoint therefore it's unaffected by the reduction at $\gamma$, therefore the swap of $\alpha$ and $\beta_1$ results in the same sub-reduction-sequence as in the $(X_i)_i$ case, therefore $Y_2 = X_i$ (with $i$ taking the value mentioned earlier where the rearrangement of this sub-reduction-sequence is done), therefore $X_{cbv} = Y_{cbv}$.
+    \item $\gamma, \beta_k, \dots, \beta_1$ has length 2 and $\alpha < \beta_1$ as before, but now $\gamma$ is left of $\alpha$: In this case, $(Y_j)_j$ starts by swapping $\beta_1$ and $\gamma$, therefore $Y_1 = X$ already.
+    \item $\gamma, \beta_k, \dots, \beta_1$ has length 2 and $\alpha < \beta_1$ as before, but now $\gamma < \alpha$: Again, $(Y_j)_j$ starts by swapping $\beta_1$ and $\gamma$ therefore $Y_1 = X$.
+    \item $\gamma, \beta_k, \dots, \beta_1$ has length 1 (so that $\gamma = \beta_1$): In this case, $Y \sim_c X$ either by case b or by case e (where the lambda has no occurrences of its variable), therefore $\gamma < \delta$ (the position of the reduction which is essentially deleted in $Y \sim_c X$).
+$\alpha$ is either outside or left of $\gamma$, therefore it bears the same relation to $\delta$.
+The sequence $(X_i)_i$ starts by swapping $\gamma$ below $\alpha$, and $(Y_j)_j$ starts by swapping $\gamma$ then $\delta$ below $\alpha$, then what happens next depends on the shape of the term at $\alpha$, or equivalently, by which case of $\sim_c$ are $\gamma$ and $\delta$ swapped below $\alpha$.
+If it's cases a, c or d, each of $\gamma$ and $\delta$ corresponds to one reduction in $Y_2$, and the relative position of these is the same as the relative position of $\gamma$ and $\delta$ therefore the next step of $(X_i)_i$ involves these two reductions and deletes the one equivalent to $\delta$, so $Y_3 = X_1$.
+If it's case b or e, in $X_2$ there are $n$ reductions corresponding to $\delta$ and $n$ reductions corresponding to $\gamma$, and they are all disjoint except that for each image of $\delta$, there's one image of $\gamma$ that it's inside, in a position that is deleted by the reduction at that image of $\gamma$.
+The next $\frac{n(n+1)} 2$ steps in $(X_i)_i$ are then each image of $\delta$ in turn being swapped (by case a) past the images of $\gamma$ until they reach the one they match, at which point they get deleted.
+In this case therefore, $Y_{n(n+1)/2+2} = X_1$.
+Case f, corresponding to $\tY$-reduction, is more complicated because the images of $\gamma$ and $\delta$ are not all disjoint.
+For each of $\gamma$ and $\delta$, there is one image of it that may contain other such positions.
+Let these be $\gamma_o$ and $\delta_o$.
+The sequence $(Y_j)_j$ then proceeds by moving all of the images of $\delta$ which are right of $\gamma_o$ to their corresponding $\gamma$-image and deleting them, then moving all the images of $\delta$ which are after $\delta_o$ but which are inside $\gamma_o$ (one at a time) past all the images of $\gamma$ left of $\gamma_o$ by case a, then either deleting them against $\gamma_o$ by case b or e as appropriate, or moving them past it by case c or d, then possibly past some of the images of $\gamma$ inside $\gamma_o$, then deleting them once they reach the matching image of $\gamma$.
+Then $\delta_o$ itself is moved to $\gamma_o$ then deleted, then the remaining images of $\delta$ are similarly moved and deleted.
+The end result of this process is that all the images of $\delta$ are deleted and all the images of $\gamma$ are unaffected, therefore for some $j$, $Y_j = X_1$.
+    \item \akr{There are several more cases to go, and that last case is more typical of how complex they'll be.
+I'd guess around 5 more, but I haven't worked them all out precisely.
+Also, I might find a neater way to deal with case f, or at least not have to repeat that argument, because the case where both reductions are $\tY$-reductions should include all the complexity of any other case involving case f.}
+\end{itemize}
+
+\paragraph{}
 If Y is in CBV order and $Y \sim_c^* X$, then $Y_{cbv} = X_{cbv}$ but also $Y_{cbv} = Y$ therefore $X_{cbv}$ is unique.

### 2020-09-29 probabilisticProgrammingMartingales.tex
Continue proof of canonicalCousins lemma.

--- 2020-09-28 probabilisticProgrammingMartingales.tex
@@
 \begin{proof}
-A reduction sequence is in call-by-value order if for every pair of reductions in the sequence at positions $\alpha$ and $\beta$, where the reduction at $\alpha$ happens first, either $\alpha \leq \beta$ or $\alpha$ and $\beta$ are disjoint, with $\alpha$ to the left of $\beta$ (i.e.~the first elements of the sequences $\alpha, \beta$ where they differ are $@_1$ and $@_2$, $\underscore f_i$ and $\underscore f_j$ for $i < j$, or $\textsf{if}_i$ and $\textsf{if}_j$ for $i < j$, in that order.
+A reduction sequence is in call-by-value order if for every pair of reductions in the sequence at positions $\alpha$ and $\beta$, where the reduction at $\alpha$ happens first, either $\alpha \leq \beta$ or $\alpha$ and $\beta$ are disjoint, with $\alpha$ to the left of $\beta$ (i.e.~the first elements of the sequences $\alpha, \beta$ where they differ are $@_1$ and $@_2$, $\underline f_i$ and $\underline f_j$ for $i < j$, or $\textsf{if}_i$ and $\textsf{if}_j$ for $i < j$, in that order.
 Note that although the reductions that occur in $X_{cbv}$ are in call-by-value order, it may not be the actual call-by-value reduction sequence because some redexes may remain un-reduced even when other reductions that should happen afterwards occur.
@@
 
-The remaining cases are when the subsequence of reduction steps involved in $Y \sim_c X$ occur before the last out-of-order pair in $X$, but possible overlapping.
+The remaining cases are when the subsequence of reduction steps involved in $Y \sim_c X$ occur before the last out-of-order pair in $X$, but possibly overlapping.
 If there is no overlap, then the $\sim_c$ steps do not interfere with each other at all, and can simply be performed in the other order.
@@
 Let the last of these positions be $\alpha$, the first $\gamma$, and the second-last, third-last and so on $\beta_1, \beta_2, \dots$ respectively, and let the position of the first relevant reduction in $Y$ be $\delta$.
-Then $X \sim_c X_1$ swaps $\beta_1$ and $\alpha$, and $\alpha$ comes before $\beta_1$ in CBV order (i.e.~$\alpha < \beta_1$ or $\alpha$ is left of $\beta_1$), and the step $Y \sim_c X$ results in $\gamma, \beta_k, \dots, \beta_1$.
+Then $X \sim_c X_1$ swaps $\beta_1$ and $\alpha$, and $\alpha$ comes before $\beta_1$ in CBV order (i.e.~$\alpha < \beta_1$ or $\alpha$ is left of $\beta_1$), and the step $Y \sim_c X$ either swaps $\delta$ and $\gamma$ resulting in $\gamma, \beta_k, \dots, \beta_1$, or swaps $\delta$ with the next (or several next) reduction (s) resulting in $\gamma, \beta_1$, depending on the direction of the $\sim_c$ step.
 In each of the cases below, it is required to show that the canonical sequence $(X_i)_i$ eventually reaches some term which is the same as one in $(Y_j)_j$, and from that point on, they match, therefore they reach the same end result: $X_{cbv} = Y_{cbv}$.
+Not all of the cases below are disjoint, but they do cover all of the remaining possibilities.
 
@@
 The sequence $(Y_j)_j$ starts by swapping $\gamma$ down past $\alpha$ by case a, then swapping $\alpha$ and $\beta_1$ as above.
-The subskeleton at $\alpha$ at the relevant point in the reduction sequence if the same as in the $(X_i)_i$ case because $\gamma$ and $\alpha$ are disjoint therefore it's unaffected by the reduction at $\gamma$, therefore the swap of $\alpha$ and $\beta_1$ results in the same sub-reduction-sequence as in the $(X_i)_i$ case, therefore $Y_2 = X_i$ (with $i$ taking the value mentioned earlier where the rearrangement of this sub-reduction-sequence is done), therefore $X_{cbv} = Y_{cbv}$.
+The subskeleton at $\alpha$ at the relevant point in the reduction sequence is the same as in the $(X_i)_i$ case because $\gamma$ and $\alpha$ are disjoint therefore it's unaffected by the reduction at $\gamma$, therefore the swap of $\alpha$ and $\beta_1$ results in the same sub-reduction-sequence as in the $(X_i)_i$ case, therefore $Y_2 = X_i$ (with $i$ taking the value mentioned earlier where the rearrangement of this sub-reduction-sequence is done), therefore $X_{cbv} = Y_{cbv}$.
     \item $\gamma, \beta_k, \dots, \beta_1$ has length 2 and $\alpha < \beta_1$ as before, but now $\gamma$ is left of $\alpha$: In this case, $(Y_j)_j$ starts by swapping $\beta_1$ and $\gamma$, therefore $Y_1 = X$ already.
@@
 The end result of this process is that all the images of $\delta$ are deleted and all the images of $\gamma$ are unaffected, therefore for some $j$, $Y_j = X_1$.
+    \item $\alpha$ is disjoint from $\gamma$ and all the $\beta_i$: In this case, $\alpha$ is left of all the $\beta_i$, and if it isn't also left of $\gamma$, then $Y_1 = X$ and we're done.
+Otherwise, $(X_i)_i$ proceeds by swapping each of the $\beta_i$ in turn then $\gamma$ below $\alpha$ by case a, and $(Y_j)_j$ swaps everything up to $\delta$ to below $\alpha$ by case a, then performs the same swap as in $Y \sim_c X$, but after $\alpha$ in the reduction sequence, yielding $\alpha, \gamma, \beta_k, \dots, \beta_1$, which matches $X_{k+1}$.
+    \item $\alpha > \gamma$: In this case, $Y_1 = X$.
     \item \akr{There are several more cases to go, and that last case is more typical of how complex they'll be.

### 2020-09-29 probabilisticProgrammingMartingales.tex
Add comments to give more intuition about \sim_c

--- 2020-09-29 probabilisticProgrammingMartingales.tex
@@
 \changed[lo]{Since $I^{L_s(M)}$ is a countable product, the measure space is well-defined \citep[Cor.~2.7.3]{AshDD00}.}
+
+\paragraph{}
+Although the six cases in the definition of $\sim_c$ are defined separately, there are some commonalities worth noting.
+Each case (except a, which is symmetrical) has a right branch and a left branch (the branches of case b are unfortunately displayed the wrong way around in the diagram).
+The right branch has reduction steps at two positions, $\beta$ then $\alpha$, where $\beta > \alpha$ (i.e.~$\beta$ is inside $\alpha$), and in the left branch, these order of these reductions are swapped, so the reductions are at $\alpha$ then $\beta_1, \dots, \beta_n$.
+The position $\alpha$ of one of the reductions is unchanged, and the other reduction may have multiple (in cases e and f) or 0 (in cases b and e) images in the left branch.
+The images $(\beta_i)$ are all still inside (greater than) $\alpha$, and although they are not generally equal to $\beta$, the subskeletons at these positions have a similar shape to the subskeleton (initially) at position $\beta$, so that the reductions at these positions are the same type of reduction (e.g.~both $\beta$-reduction or both $\textsf{if}$-reduction).
+Case a is also similar, except that the relevant positions are all disjoint rather than contained in one another, and either branch could be considered the right branch.
 

### 2020-09-30 probabilisticProgrammingMartingales.tex
minor

@@
 
-\lo{TODO: Typeset Fig.~\ref{fig:adk-diagram} in vector graphics using PGF/TikZ.}
+%\lo{TODO: Typeset Fig.~\ref{fig:adk-diagram} in vector graphics using PGF/TikZ.}
 
@@
 \paragraph{}
-If Y is in CBV order and $Y \sim_c^* X$, then $Y_{cbv} = X_{cbv}$ but also $Y_{cbv} = Y$ therefore $X_{cbv}$ is unique.
+If $Y$ is in CBV order and $Y \sim_c^* X$, then $Y_{cbv} = X_{cbv}$ but also $Y_{cbv} = Y$ therefore $X_{cbv}$ is unique.
 \end{proof}

### 2020-10-02 probabilisticProgrammingMartingales.tex
Re-organise the cases in canonicalCousins to be more logical.

--- 2020-09-29 probabilisticProgrammingMartingales.tex
@@
 
-\paragraph{}
-Although the six cases in the definition of $\sim_c$ are defined separately, there are some commonalities worth noting.
-Each case (except a, which is symmetrical) has a right branch and a left branch (the branches of case b are unfortunately displayed the wrong way around in the diagram).
-The right branch has reduction steps at two positions, $\beta$ then $\alpha$, where $\beta > \alpha$ (i.e.~$\beta$ is inside $\alpha$), and in the left branch, these order of these reductions are swapped, so the reductions are at $\alpha$ then $\beta_1, \dots, \beta_n$.
-The position $\alpha$ of one of the reductions is unchanged, and the other reduction may have multiple (in cases e and f) or 0 (in cases b and e) images in the left branch.
-The images $(\beta_i)$ are all still inside (greater than) $\alpha$, and although they are not generally equal to $\beta$, the subskeletons at these positions have a similar shape to the subskeleton (initially) at position $\beta$, so that the reductions at these positions are the same type of reduction (e.g.~both $\beta$-reduction or both $\textsf{if}$-reduction).
-Case a is also similar, except that the relevant positions are all disjoint rather than contained in one another, and either branch could be considered the right branch.
-
 \iffalse
@@
 Now consider the specific sequence of positions of the reductions in $X$ involved in either of these steps.
-Let the last of these positions be $\alpha$, the first $\gamma$, and the second-last, third-last and so on $\beta_1, \beta_2, \dots$ respectively, and let the position of the first relevant reduction in $Y$ be $\delta$.
-Then $X \sim_c X_1$ swaps $\beta_1$ and $\alpha$, and $\alpha$ comes before $\beta_1$ in CBV order (i.e.~$\alpha < \beta_1$ or $\alpha$ is left of $\beta_1$), and the step $Y \sim_c X$ either swaps $\delta$ and $\gamma$ resulting in $\gamma, \beta_k, \dots, \beta_1$, or swaps $\delta$ with the next (or several next) reduction (s) resulting in $\gamma, \beta_1$, depending on the direction of the $\sim_c$ step.
+Let the last of these positions be $\alpha$, the first $\gamma$, and the second-last, third-last and so on $\beta_1, \beta_2, \dots$ respectively, and let the positions of the redexes in $Y$ involved in $Y \sim_c X$ be $\beta$ then $(\gamma_i)$ in order..
+Then $X \sim_c X_1$ swaps $\beta_1$ and $\alpha$, and $\alpha$ comes before $\beta_1$ in CBV order (i.e.~$\alpha < \beta_1$ or $\alpha$ is left of $\beta_1$), and the step $Y \sim_c X$ either proceeds forwards towards CBV order, swapping $\beta$ below $\gamma_1 = \gamma$ to form $\gamma$ then $(\beta_i)$, or proceeds backwards, taking $Y$ even farther from CBV order, swapping the $(\gamma_i)$ to above $\beta$, forming $\gamma$ then $\beta_1 = \beta$.
 In each of the cases below, it is required to show that the canonical sequence $(X_i)_i$ eventually reaches some term which is the same as one in $(Y_j)_j$, and from that point on, they match, therefore they reach the same end result: $X_{cbv} = Y_{cbv}$.
-Not all of the cases below are disjoint, but they do cover all of the remaining possibilities.
 
 \begin{itemize}
+\item $Y \sim_c X$ in the forwards direction (so that $X$ is closer to CBV order than $Y$ is), and $\alpha$ comes after $\gamma$ in CBV order: In this case, the $\gamma$ and $\alpha$ reduction steps are already in the correct order in $Y$, therefore $\beta$ and $\gamma$ are already the last out-of-order pair of reductions in $Y$, $Y_1 = X$, and $Y \sim_c X \sim_c^* X_{cbv}$ is already the canonical sequence from $Y$ therefore $Y_{cbv} = X_{cbv}$.
+\item $Y \sim_c X$ is forwards and $\alpha$ is disjoint from $\gamma$ and $\beta$: In this case, the sequence $(X_i)_i$ starts by swapping $\alpha$ and $\beta_1$ by case a, then because $\gamma$ is disjoint from $\alpha$ and comes after is in CBV order, $\gamma$ is right of $\alpha$, therefore either all the remaining $\beta_i$s, are $\geq \gamma$ (if $\beta$ and $\gamma$ aren't disjoint) or there are no remaining $\beta_i$s (if $\beta$ and $\gamma$ are disjoint, in which case they swap by case a and there is only $\beta_1$).
+The sequence $(X_i)_i$ therefore proceeds to swap all the remaining $\beta_i$s then $\gamma$, in order, below $\alpha$, resulting in $X_i$ for some $i$ having as its relevant sub-reduction-sequence $\alpha, \gamma, \beta_k, \dots, \beta_1$.
+The other sequence $(Y_j)_j$ starts by swapping $\gamma$ then $beta$ below $\alpha$, then swapping $\gamma$ and $\beta$.
+Because $\alpha$ is disjoint from $\gamma$, the reduction there does not affect the sub-skeleton at position $\gamma$, therefore this last $\sim_c$ step proceeds identically to how it did in $Y \sim_c X$, resulting in $\gamma, \beta_k, \dots, \beta_i$.
+Overall, this results in $Y_3$ having the relevant sub-reduction-sequence $\alpha, \gamma, \beta_k, \dots, \beta_1$, therefore $Y_3 = X_i$ for the aforementioned value of $i$.
+\item $Y \sim_c X$ is backwards and $\alpha$ is disjoint from $\beta$ and $\gamma$: In this case, $(X_i)_i$ proceeds by swapping $\beta$ then $\gamma$ below $\alpha$ by case a, then reversing $\gamma$ and $\beta$ back to $\beta, \gamma_k, \dots, \gamma_i$ by the same case as in $Y \sim_c X$ but in the other direction, and $(Y_j)_j$ proceeds by swapping each $\gamma_i$ in turn and then $\beta$ below $\alpha$ by case a, so both sequences at some point reach the same result, with $\alpha, \beta, \gamma_k, \dots, \gamma_1$ as the relevant sub-reduction-sequence.
+\item $Y \sim_c X$ forwards, $\alpha \leq \gamma$ and $\alpha$ (and therefore also $\gamma$) is disjoint from $\beta$: In this case, $(X_i)_i$ proceeds by swapping $\beta_1 = \beta$ below alpha by case a, then swapping $\gamma$ below $\alpha$ by one of the other cases, and $(Y_j)_j$ proceeds by swapping $\gamma$ and $\alpha$, then swapping $\beta$ below $\alpha$ then all of the images of $\gamma$ (which are $\geq \alpha$ and therefore left of $\beta$).
+In both cases, the subterm at $\alpha$ when $\alpha$ and $\gamma$ are swapped is unaffected by the reduction at $\beta$, therefore it produces the same result in both cases, therefore the overall sequence in both cases is $\alpha$ then the images of $\gamma$ then $\beta$.
+\item $Y \sim_c X$ backwards, $\alpha \leq \beta$ and both $\alpha$ and $\beta$ are left of $\gamma$: This case is quite similar to the previous one.
+In this case, $(X_i)_i$ proceeds by swapping $\beta$ below $\alpha$ by one of the cases b to f, then swapping $\gamma$ below $\alpha$ then each of the images of $\beta$ in turn by case a, and $(Y_j)_j$ proceeds by swapping $\gamma$ below $\alpha$ by case a, then swapping $\beta$ below $\alpha$.
+As before, the results of the swapping of $\alpha$ and $\beta$ are the same in both cases, therefore both sequences reach the same reduction sequence containing $\alpha$ then the images of $\beta$ then $\gamma$.
+  \item Both $\gamma$ and $\beta$ are $> \alpha$ but they're disjoint from each other, and either the skeleton at position $\alpha$ is of the form $\tif{\skeletonPlaceholder > 0}{A}{B}$ or it is of the form $A B$ and either both $\gamma$ and $\beta$ are $> \alpha;@_1$ or they are both $> \alpha;@_2$: In this case, $(X_i)_i$ proceeds by swapping $\beta$ then $\gamma$ below $\alpha$, in each case possibly forming 0 or multiple images.
+The only way the number of images of each of these positions can be different is if the subskeleton at $\alpha$ is an $\textsf{if}$, and there are 0 of one of them and 1 of the other, in which case they're trivially in the correct order already.
+Otherwise, all of the images of $\beta$ and $\gamma$ are disjoint from one another, therefore none of the positions, or their relative order, change when they are swapped, therefore the next part of the sequence $(X_i)_i$ is just insertion-sort running on the images of $\beta$ and $\gamma$ with $\sim_c$ case a swaps until they're in the correct order.
+The sequence $(Y_j)_j$ starts by swapping $\gamma$ then $\beta$ below $\alpha$, then as before, sorting the images into the correct order by case a swaps.
+The images originally formed of $\gamma$ and $\beta$ are the same in both cases, therefore the final order is the same, so $X_i = Y_j$ for some $i$ and $j$.
   \item $\gamma, \beta_k, \dots, \beta_1$ has length 2, $\alpha < \beta_1$ and $\gamma$ is right of $\alpha$: In this case, the sequence $(X_i)_i$ starts by swapping $\alpha$ and $\beta_1$ by one of the cases b to f in the definition of $\sim_c$, then swapping $\gamma$ down past $\alpha$ by case a, then down past all the images of $\beta_1$, also by case a.
@@
 The subskeleton at $\alpha$ at the relevant point in the reduction sequence is the same as in the $(X_i)_i$ case because $\gamma$ and $\alpha$ are disjoint therefore it's unaffected by the reduction at $\gamma$, therefore the swap of $\alpha$ and $\beta_1$ results in the same sub-reduction-sequence as in the $(X_i)_i$ case, therefore $Y_2 = X_i$ (with $i$ taking the value mentioned earlier where the rearrangement of this sub-reduction-sequence is done), therefore $X_{cbv} = Y_{cbv}$.
+    \item \akr{The cases below this point aren't wrong, they're just not such a neat way to split up the possible cases so I probably won't be using them.}
     \item $\gamma, \beta_k, \dots, \beta_1$ has length 2 and $\alpha < \beta_1$ as before, but now $\gamma$ is left of $\alpha$: In this case, $(Y_j)_j$ starts by swapping $\beta_1$ and $\gamma$, therefore $Y_1 = X$ already.

### 2020-10-02 probabilisticProgrammingMartingales.tex
Minor correction

@@
 The position $\alpha$ of one of the reductions is unchanged, and the other reduction may have multiple (in cases e and f) or 0 (in cases b and e) images in the left branch.
-The images $(\beta_i)$ are all still inside (greater than) $\alpha$, and although they are not generally equal to $\beta$, the subskeletons at these positions have a similar shape to the subskeleton (initially) at position $\beta$, so that the reductions at these positions are the same type of reduction (e.g.~both $\beta$-reduction or both $\textsf{if}$-reduction).
+The images $(\beta_i)$ are all still inside (greater than or equal to) $\alpha$, and although they are not generally equal to $\beta$, the subskeletons at these positions have a similar shape to the subskeleton (initially) at position $\beta$, so that the reductions at these positions are the same type of reduction (e.g.~both $\beta$-reduction or both $\textsf{if}$-reduction).
 Case a is also similar, except that the relevant positions are all disjoint rather than contained in one another, and either branch could be considered the right branch.

### 2020-10-06 probabilisticProgrammingMartingales.tex
Continue proving canonicalCousins.
Write more cases, merge the forwards and backwards Y \sim_c X cases,
tweak the definition of \sim_c case f to make this proof work a bit
neater.

@@
         \item $N \mid P = \tY (\lambda x.
-A)$, $A$ reduced at $Q$ is $A'$, $(R_i)_i$ is a list of all the positions where $A' \mid R  = x$, ordered from left to right, $O_1$ is $N$ reduced first at $P;\tY;\lambda;Q$ then at $P$, and $O_2$ is $N$ reduced first at $P$ then at $P;\lambda;@_1;Q$ then at $P;\lambda;@_1;R_i;Y;\lambda;Q$ for each $i$ in order.
+A)$, $A$ reduced at $Q$ is $A'$, $(R_i)_i$ is a list of all the positions where $A' \mid R  = x$, ordered from left to right, $O_1$ is $N$ reduced first at $P;\tY;\lambda;Q$ then at $P$, and $O_2$ is $N$ reduced first at $P$ then at $P;\lambda;@_1;R_i;Y;\lambda;Q$ for each $i$ in order where $R_i$ is left of $Q$ then at $P;\lambda;@_1;Q$ then at $P;\lambda;@_1;R_i;Y;\lambda;Q$ for the remaining values of $i$.
     \end{enumerate}
@@
 \begin{itemize}
-\item $Y \sim_c X$ in the forwards direction (so that $X$ is closer to CBV order than $Y$ is), and $\alpha$ comes after $\gamma$ in CBV order: In this case, the $\gamma$ and $\alpha$ reduction steps are already in the correct order in $Y$, therefore $\beta$ and $\gamma$ are already the last out-of-order pair of reductions in $Y$, $Y_1 = X$, and $Y \sim_c X \sim_c^* X_{cbv}$ is already the canonical sequence from $Y$ therefore $Y_{cbv} = X_{cbv}$.
-\item $Y \sim_c X$ is forwards and $\alpha$ is disjoint from $\gamma$ and $\beta$: In this case, the sequence $(X_i)_i$ starts by swapping $\alpha$ and $\beta_1$ by case a, then because $\gamma$ is disjoint from $\alpha$ and comes after is in CBV order, $\gamma$ is right of $\alpha$, therefore either all the remaining $\beta_i$s, are $\geq \gamma$ (if $\beta$ and $\gamma$ aren't disjoint) or there are no remaining $\beta_i$s (if $\beta$ and $\gamma$ are disjoint, in which case they swap by case a and there is only $\beta_1$).
+  \item $Y \sim_c X$ in the backwards direction (so that $Y$ is closer to CBV order than $X$ is): By swapping the roles of $X$ and $Y$, and $\beta$ and $\gamma$, this is equivalent to the $Y \sim_c X$ forwards case.
+The assumption that $\alpha$ is before $\beta_1$ in CBV order (so that $\beta_1, \alpha$ is the pair to be swapped in $X \sim_c X_1$) maps to the assumption that $\alpha$ is before $\gamma_1$ (so that the result doesn't just follow trivially by $Y_1$ being equal to $X$) and vice-versa.
+Then one of the cases below establishes that $X_i = Y_j$ for some $i$ and $j$, therefore after swapping back, $Y_i = X_j$.
+\item $Y \sim_c X$ forwards and $\alpha$ comes after $\gamma$ in CBV order: In this case, the $\gamma$ and $\alpha$ reduction steps are already in the correct order in $Y$, therefore $\beta$ and $\gamma$ are already the last out-of-order pair of reductions in $Y$, $Y_1 = X$, and $Y \sim_c X \sim_c^* X_{cbv}$ is already the canonical sequence from $Y$ therefore $Y_{cbv} = X_{cbv}$.
+\item $Y \sim_c X$ forwards and $\alpha$ is disjoint from $\gamma$ and $\beta$: In this case, the sequence $(X_i)_i$ starts by swapping $\alpha$ and $\beta_1$ by case a, then because $\gamma$ is disjoint from $\alpha$ and comes after is in CBV order, $\gamma$ is right of $\alpha$, therefore either all the remaining $\beta_i$s, are $\geq \gamma$ (if $\beta$ and $\gamma$ aren't disjoint) or there are no remaining $\beta_i$s (if $\beta$ and $\gamma$ are disjoint, in which case they swap by case a and there is only $\beta_1$).
 The sequence $(X_i)_i$ therefore proceeds to swap all the remaining $\beta_i$s then $\gamma$, in order, below $\alpha$, resulting in $X_i$ for some $i$ having as its relevant sub-reduction-sequence $\alpha, \gamma, \beta_k, \dots, \beta_1$.
@@
 Overall, this results in $Y_3$ having the relevant sub-reduction-sequence $\alpha, \gamma, \beta_k, \dots, \beta_1$, therefore $Y_3 = X_i$ for the aforementioned value of $i$.
-\item $Y \sim_c X$ is backwards and $\alpha$ is disjoint from $\beta$ and $\gamma$: In this case, $(X_i)_i$ proceeds by swapping $\beta$ then $\gamma$ below $\alpha$ by case a, then reversing $\gamma$ and $\beta$ back to $\beta, \gamma_k, \dots, \gamma_i$ by the same case as in $Y \sim_c X$ but in the other direction, and $(Y_j)_j$ proceeds by swapping each $\gamma_i$ in turn and then $\beta$ below $\alpha$ by case a, so both sequences at some point reach the same result, with $\alpha, \beta, \gamma_k, \dots, \gamma_1$ as the relevant sub-reduction-sequence.
 \item $Y \sim_c X$ forwards, $\alpha \leq \gamma$ and $\alpha$ (and therefore also $\gamma$) is disjoint from $\beta$: In this case, $(X_i)_i$ proceeds by swapping $\beta_1 = \beta$ below alpha by case a, then swapping $\gamma$ below $\alpha$ by one of the other cases, and $(Y_j)_j$ proceeds by swapping $\gamma$ and $\alpha$, then swapping $\beta$ below $\alpha$ then all of the images of $\gamma$ (which are $\geq \alpha$ and therefore left of $\beta$).
 In both cases, the subterm at $\alpha$ when $\alpha$ and $\gamma$ are swapped is unaffected by the reduction at $\beta$, therefore it produces the same result in both cases, therefore the overall sequence in both cases is $\alpha$ then the images of $\gamma$ then $\beta$.
-\item $Y \sim_c X$ backwards, $\alpha \leq \beta$ and both $\alpha$ and $\beta$ are left of $\gamma$: This case is quite similar to the previous one.
-In this case, $(X_i)_i$ proceeds by swapping $\beta$ below $\alpha$ by one of the cases b to f, then swapping $\gamma$ below $\alpha$ then each of the images of $\beta$ in turn by case a, and $(Y_j)_j$ proceeds by swapping $\gamma$ below $\alpha$ by case a, then swapping $\beta$ below $\alpha$.
-As before, the results of the swapping of $\alpha$ and $\beta$ are the same in both cases, therefore both sequences reach the same reduction sequence containing $\alpha$ then the images of $\beta$ then $\gamma$.
 \item Both $\gamma$ and $\beta$ are $> \alpha$ but they're disjoint from each other, and either the skeleton at position $\alpha$ is of the form $\tif{\skeletonPlaceholder > 0}{A}{B}$ or it is of the form $A B$ and either both $\gamma$ and $\beta$ are $> \alpha;@_1$ or they are both $> \alpha;@_2$: In this case, $(X_i)_i$ proceeds by swapping $\beta$ then $\gamma$ below $\alpha$, in each case possibly forming 0 or multiple images.
@@
 The images originally formed of $\gamma$ and $\beta$ are the same in both cases, therefore the final order is the same, so $X_i = Y_j$ for some $i$ and $j$.
-  \item $\gamma, \beta_k, \dots, \beta_1$ has length 2, $\alpha < \beta_1$ and $\gamma$ is right of $\alpha$: In this case, the sequence $(X_i)_i$ starts by swapping $\alpha$ and $\beta_1$ by one of the cases b to f in the definition of $\sim_c$, then swapping $\gamma$ down past $\alpha$ by case a, then down past all the images of $\beta_1$, also by case a.
-Eventually it reaches some $X_i$ where the relevant sub-reduction-sequence is $\alpha$ then whatever $\beta_1$ was turned into in $X \sim_c X_1$ then $\gamma$, before moving on to swapping pairs earlier in the reduction sequence.
-The sequence $(Y_j)_j$ starts by swapping $\gamma$ down past $\alpha$ by case a, then swapping $\alpha$ and $\beta_1$ as above.
-The subskeleton at $\alpha$ at the relevant point in the reduction sequence is the same as in the $(X_i)_i$ case because $\gamma$ and $\alpha$ are disjoint therefore it's unaffected by the reduction at $\gamma$, therefore the swap of $\alpha$ and $\beta_1$ results in the same sub-reduction-sequence as in the $(X_i)_i$ case, therefore $Y_2 = X_i$ (with $i$ taking the value mentioned earlier where the rearrangement of this sub-reduction-sequence is done), therefore $X_{cbv} = Y_{cbv}$.
-    \item \akr{The cases below this point aren't wrong, they're just not such a neat way to split up the possible cases so I probably won't be using them.}
-    \item $\gamma, \beta_k, \dots, \beta_1$ has length 2 and $\alpha < \beta_1$ as before, but now $\gamma$ is left of $\alpha$: In this case, $(Y_j)_j$ starts by swapping $\beta_1$ and $\gamma$, therefore $Y_1 = X$ already.
-    \item $\gamma, \beta_k, \dots, \beta_1$ has length 2 and $\alpha < \beta_1$ as before, but now $\gamma < \alpha$: Again, $(Y_j)_j$ starts by swapping $\beta_1$ and $\gamma$ therefore $Y_1 = X$.
-    \item $\gamma, \beta_k, \dots, \beta_1$ has length 1 (so that $\gamma = \beta_1$): In this case, $Y \sim_c X$ either by case b or by case e (where the lambda has no occurrences of its variable), therefore $\gamma < \delta$ (the position of the reduction which is essentially deleted in $Y \sim_c X$).
-$\alpha$ is either outside or left of $\gamma$, therefore it bears the same relation to $\delta$.
-The sequence $(X_i)_i$ starts by swapping $\gamma$ below $\alpha$, and $(Y_j)_j$ starts by swapping $\gamma$ then $\delta$ below $\alpha$, then what happens next depends on the shape of the term at $\alpha$, or equivalently, by which case of $\sim_c$ are $\gamma$ and $\delta$ swapped below $\alpha$.
-If it's cases a, c or d, each of $\gamma$ and $\delta$ corresponds to one reduction in $Y_2$, and the relative position of these is the same as the relative position of $\gamma$ and $\delta$ therefore the next step of $(X_i)_i$ involves these two reductions and deletes the one equivalent to $\delta$, so $Y_3 = X_1$.
-If it's case b or e, in $X_2$ there are $n$ reductions corresponding to $\delta$ and $n$ reductions corresponding to $\gamma$, and they are all disjoint except that for each image of $\delta$, there's one image of $\gamma$ that it's inside, in a position that is deleted by the reduction at that image of $\gamma$.
-The next $\frac{n(n+1)} 2$ steps in $(X_i)_i$ are then each image of $\delta$ in turn being swapped (by case a) past the images of $\gamma$ until they reach the one they match, at which point they get deleted.
-In this case therefore, $Y_{n(n+1)/2+2} = X_1$.
-Case f, corresponding to $\tY$-reduction, is more complicated because the images of $\gamma$ and $\delta$ are not all disjoint.
-For each of $\gamma$ and $\delta$, there is one image of it that may contain other such positions.
-Let these be $\gamma_o$ and $\delta_o$.
-The sequence $(Y_j)_j$ then proceeds by moving all of the images of $\delta$ which are right of $\gamma_o$ to their corresponding $\gamma$-image and deleting them, then moving all the images of $\delta$ which are after $\delta_o$ but which are inside $\gamma_o$ (one at a time) past all the images of $\gamma$ left of $\gamma_o$ by case a, then either deleting them against $\gamma_o$ by case b or e as appropriate, or moving them past it by case c or d, then possibly past some of the images of $\gamma$ inside $\gamma_o$, then deleting them once they reach the matching image of $\gamma$.
-Then $\delta_o$ itself is moved to $\gamma_o$ then deleted, then the remaining images of $\delta$ are similarly moved and deleted.
-The end result of this process is that all the images of $\delta$ are deleted and all the images of $\gamma$ are unaffected, therefore for some $j$, $Y_j = X_1$.
-    \item $\alpha$ is disjoint from $\gamma$ and all the $\beta_i$: In this case, $\alpha$ is left of all the $\beta_i$, and if it isn't also left of $\gamma$, then $Y_1 = X$ and we're done.
-Otherwise, $(X_i)_i$ proceeds by swapping each of the $\beta_i$ in turn then $\gamma$ below $\alpha$ by case a, and $(Y_j)_j$ swaps everything up to $\delta$ to below $\alpha$ by case a, then performs the same swap as in $Y \sim_c X$, but after $\alpha$ in the reduction sequence, yielding $\alpha, \gamma, \beta_k, \dots, \beta_1$, which matches $X_{k+1}$.
-    \item $\alpha > \gamma$: In this case, $Y_1 = X$.
-    \item \akr{There are several more cases to go, and that last case is more typical of how complex they'll be.
-I'd guess around 5 more, but I haven't worked them all out precisely.
-Also, I might find a neater way to deal with case f, or at least not have to repeat that argument, because the case where both reductions are $\tY$-reductions should include all the complexity of any other case involving case f.}
+\item $Y \sim_c X$ forwards, $\alpha;@_1 < \gamma$, and $\alpha;@_2 < \beta$: Let $x$ be the variable involved in the $\beta$-reduction at $\alpha$, so that the sub-skeleton at $\alpha$ is $(\lambda x.
+A) B$ for some $A, B$, where $\gamma = \alpha;@_1;\lambda;\gamma'$ and $A \to A'$ at $\gamma'$, and $\beta = \alpha;@_2;\beta'$, and $B \to B'$ at $\beta'$.
+In this case, $(X_i)_i$ starts by swapping $\beta$ below $\alpha$ by case e, producing one image of $\beta$ for each $x$ in $A'$, then swapping $\gamma$ below $\alpha$ by case d, producing a single image $\alpha;\gamma'$.
+For each of the instances of $x$ in $A$ left of $\gamma'$, the reduction at $\alpha;\gamma'$ is then swapped below the corresponding image of $\beta$.
+The sequence $(Y_j)_j$ starts by swapping $\gamma$ below $\alpha$ then $\beta$ below $\alpha$, but in this case the images of $\beta$ produced may be different.
+In this case the swap happens earlier in the reduction sequence than $\gamma$, so that the body of the lambda is still $A$ rather than $A'$, and the reduction at $\gamma$ may rearrange or change the number of instances of $x$.
+Next, each of the images of $\beta$ to the right of $\alpha;\gamma'$ are swapped with it by case a.
+At this point, the next few reductions before $\alpha;\gamma'$ will in general be images of $\beta$ at positions inside $\alpha;\gamma'$.
+The subskeleton at position $\alpha;\gamma'$ before these reductions is $(A|\gamma')[B/x]$, then by the images of $\beta$ this reduces to $(A|\gamma')[B'/x]$, then it reduces at its root position ($\alpha;gamma'$ in the overall skeleton) to $(A'|\gamma')[B'/x]$.
+For each position of an $x$ in $(A|\gamma')$, there are 0 or more corresponding positions of $x$ in $(A'|\gamma')$, depending on what type of reduction $A \to A'$ is.
+Because $B$ and $B'$ cannot contain any instances of the variable involved in the reduction $A \to A'$ (if it's a $\beta$-reduction or $\tY$-reduction), each instance of $B$ in $(A|\gamma')[B/x]$ has the same set of images in $(A'|\gamma')[B/x]$ as the corresponding $x$ in $(A|\gamma')$.
+When these images of $\beta$ that overlap with the image of $\gamma$ are swapped below it, each of them has its own images, one at each position of $B$ in $(A'|\gamma')[B/x]$ corresponding to its original copy of $B$ in $(A|\gamma')[B/x]$.
+After all of them are swapped below $\alpha;\gamma'$ (and rearranged among themselved by case a), there is therefore one reduction at each copy of $B$ in $(A'|\gamma')[B/x]$, i.e.~one at position $\beta'$ relative to each $x$ in $(A'|\gamma')$.
+They are therefore the same as the images of $\beta$ in $X_1$ that overlap with $\alpha;\gamma'$, because those are also one reduction at position $\beta'$ relative to each $x$ in $(A'|\gamma)$.
+Both $(X_i)_i$ and $(Y_j)_j$ therefore eventually reach a point where the relevant sub-reduction-sequence is $\alpha$, the images of $\beta$ to the left of $\alpha;\gamma'$, $\alpha;\gamma'$ itself, the aforementioned images of $\beta$ that overlap with $\alpha;\gamma'$, then all the images of $\beta$ to the right of $\alpha;\gamma'$.
+\item $Y \sim_c X$ forwards, $\gamma$ and $\beta$ are disjoint and both $> \alpha$, and the reduction at $\alpha$ is a $\tY$-reduction: This is similar to the previous case, but more complicated, so it will need some definitions to explain properly what's going on.
+Let the term at $\alpha$ before any of the reductions be $\tY \lambda x.
+A$, let $\gamma = \alpha;\tY;\lambda;\gamma'$, let $\beta = \alpha;\tY;\lambda;\beta'$, let $A | \beta' = B \to B'$ with the reduction at $\cdot$, let $A | \gamma' = C \to C'$ with the reduction at $\cdot$, let the positions in $A$ where $x$ occurs that are left of $\gamma'$, between $\gamma'$ and $\beta'$ (but still disjoint from both), and right of $\beta'$ respectively be $(\delta^l_i)$, $(\delta^m_i)$ and $(\delta^r_i)$, let the positions in $B$, $C$, $B'$ and $C'$ respectively where $x$ occurs be $(\delta^B_i)$, $(\delta^C_i)$, $(\delta^{B'}_i)$ and $(\delta^{C'}_i)$ (so that all the positions in $A$ where $x$ occurs, in left-to-right order, are $(\delta^l_i)_i, (\delta^C_i)_i, (\delta^m_i)_i, (\delta^B_i)_i, (\delta^r_i)_i$).
+The sequence $(X_i)_i$ starts by swapping $\beta$ then $\gamma$ below $\alpha$.
+With the shorthand that $\gamma'(\epsilon) = \alpha;\lambda;@_1;\epsilon;\tY;\lambda;\gamma'$, $\gamma'_0 = \alpha;\lambda;@_1;\gamma'$, and similarly for $\beta'$, the relevant portion of $X_2$ is then $\alpha,\ (\gamma'(\delta^l_i))_i,\ \gamma'_0,\ (\gamma'(\delta^{C'}_i))_i,\ (\gamma'(\delta^m_i))_i,\ (\gamma'(\delta^B_i))_i,\ (\gamma'(\delta^r_i))_i,\ (\beta'(\delta^l_i))_i,\ (\beta'(\delta^{C'}_i))_i,\ (\beta'(\delta^m_i))_i,\ \beta'_0,\ (\beta'(\delta^{B'}_i))_i,\ (\beta'(\delta^r_i))_i$.
+Next in $(X_k)_k$, for each $i$, $\gamma'(\delta^r_i)$ are swapped past all the images of $\beta$ until it is immediately before $\beta'(\delta^r_i)$ by case a, then for each $i$, $\gamma'(\delta^B_i)$ is swapped past all the images of $\beta$ until $\beta'_0$ by case a, then swapped with $\beta'_0$ by one of the other cases depending on what type of reduction $B \to B'$ is and the relative positions.
+Even if it's case d or f, all of the images of $\gamma'(\delta^B_i)$ are disjoint from each other (and from the positions of the other reductions after $\beta'_0$) because the redex, $C$, doesn't contain any instances of the variable involved in the reduction $B \to B'$.
+Expanding the definitions, this is swapping $\alpha;\lambda;@_1;\delta^B_i;\tY;\lambda;gamma'$ with $\alpha;\lambda;@_1;\beta'$ at a point in the reduction sequence where the subskeleton at $\alpha;\lambda;@_1;\beta'$ is $B$ with a mixture of $\tY \lambda x.
+A$ and $\tY \lambda x.
+A[C'/\gamma']$ substituted for its occurrences of $x$, and one of these occurrences of $x$ is at $\delta^B_i$.
+For each $\delta^B_j$, there are some corresponding $\delta^{B'}_k$, where the images of the $x$ at $\delta^B_j$ end up after the reduction $B \to B'$.
+These cover all the $\delta^{B'}_k$, and uniquely, and for each $\delta^B_i$, the images of $\zeta;\delta^B_i;\theta$ after swapping it with $\zeta;\beta'$ are precisely $(\zeta;\delta^{B'}_k;\theta)$ for those same values of $k$, therefore after swapping all of the $(\gamma'(\delta^B_i))_i$ down past $\beta'_0$, their images are $(\gamma'(\delta^{B'}_i))_i$.
+After swapping with $\beta'_0$, each of these images than swaps by case a to take its place among the $(\beta'(\delta^{B'}_i))_i$.
+Next up in $(X_i)_i$, the $(\gamma'(\delta^m_i))_i$s and the $(\gamma'(\delta^{C'}_i))_i$s swap down past some images of $\beta$ they're disjoint from to take their place before their matching image of $\beta$, then $\gamma'_0$ swaps past the $(\beta'(\delta^l_i))_i$, stopping immediately before $\beta'(\delta^{C'}_1)$, then the $(\gamma'(\delta^l_i))_i$ and the $(\beta'(\delta^l_i))_i$ mix.
+The end result of the rearrangement of (this subsequence of) the reduction sequence is therefore $\alpha,\ (\gamma'(\delta^l_i), \beta'(\delta^l_i))_i,\ \gamma'_0,\ (\gamma'(\delta^{C'}_i), \beta'(\delta^{C'}_i))_i,\ (\gamma'(\delta^m_i), \beta'(\delta^m_i))_i,\ \beta'_0,\ (\gamma'(\delta^{B'}_i), \beta'(\delta^{B'}_i))_i,\ (\gamma'(\delta^r_i), \beta'(\delta^r_i))_i$.
+
+  The other sequence, $(Y_j)_j$, proceeds similarly, but with the images of $\gamma$ starting out after the images of $\beta$.
+It swaps all the images of $\beta$ that are right of $\gamma_0$ to the appropriate places by case a, then swaps all the $(\beta'(\delta^C_i)_i$ first to then past $\gamma'_0$, resulting in $(\beta'(\delta^{C'}_i))_i$ for the same reason as with swapping those images of $\gamma$ that overlapped with $\beta'_0$ past it in the $(X_i)_i$ case above, then finally swapping the $(\beta'(\delta^l_i))_i$s and the $(\gamma'(\delta^l_i))_i$s into the correct order.
+As required, this produces the same result as in $(X_i)_i$.
+  \item $Y \to X$ forwards, and $\alpha < \gamma < \beta$: \akr{This is the final case, although I may have to split it up depending on whether the reduction at $\alpha$ is a $tY$-reduction.}
 \end{itemize}
@@
 \paragraph{}
-If $Y$ is in CBV order and $Y \sim_c^* X$, then $Y_{cbv} = X_{cbv}$ but also $Y_{cbv} = Y$ therefore $X_{cbv}$ is unique.
+If $Y$ is in CBV order and $Y \sim_c^* X$, then $Y_{cbv} = X_{cbv}$ but also $Y_{cbv} = Y$ therefore $X_{cbv}$ is the unique reduction sequence in call-by-value order that's related to $X$ by $\sim_c^*$.
 \end{proof}

### 2020-10-07 probabilisticProgrammingMartingales.tex
Finished proof of canonicalCousing lemma!

@@
 As required, this produces the same result as in $(X_i)_i$.
-  \item $Y \to X$ forwards, and $\alpha < \gamma < \beta$: \akr{This is the final case, although I may have to split it up depending on whether the reduction at $\alpha$ is a $tY$-reduction.}
+  \item $Y \to X$ forwards, and $\alpha < \gamma < \beta$: Let $\gamma'$ be such that $\gamma = \alpha;\textsf{if}_i;\gamma'$, $\alpha;@_1;\lambda;\gamma'$, $\alpha;@_2;\gamma'$ or $\alpha;\tY;\lambda;\gamma'$, so that $\gamma'$ is the freely varying later part of $\gamma$ as in the definition of $\sim_c$, whichever case applies to swapping $\gamma$ and $\alpha$.
+Similarly, let $\beta'$ be the freely varying part of $\beta$ relative to $\gamma$.
+Let the images of $\gamma$ after swapping it with $\alpha$ (where this swap occurs later in the reduction sequence than $\beta$) then be $(\alpha;\delta_i;\gamma')_i$, and the images of $\beta$ after swapping it with $\gamma$ be $(\gamma;\epsilon_i;\beta')_i$.
+Depending on the skeleton at $\alpha$, and $\gamma$'s position within it, $(\delta_i)_i$ may be empty (case b), a singleton containing $\cdot$ (cases c and d), all the positions of the relevant variable in the lambda in the skeleton at $\alpha$ (case e), or $\lambda;@_1$ and $\lambda;@_1;\zeta;\tY;\lambda$ for each position $\zeta$ of the relevant variable (case f), but in all cases (except a, which is excluded because none of $\alpha, \gamma$ and $\beta$ are disjoint) the set of images has this general structure.
+Furthermore, the positions $(\delta_i)_i$ are determined only by the general position of $\gamma$ within $\alpha$ (the part excluded from $\gamma'$), and the positions of the relevant variable within the skeleton at $\alpha$.
+They do not depend on $\alpha$ or $\gamma$, so that if both initial positions were moved somewhere else the relative positions of the images of $\gamma$ would be unaffected, and similarly if some position within $\gamma$ were swapped with $\alpha$, the relative positions of its images would be the same (except to the extent that, in case f, the positions of the relevant variable are taken after the inner reduction takes place, so that some of those may still change).
+Because there are so many of them and they don't actually affect the multiset of positions where reductions take place, I will be ignoring swaps by case a in the proof of this case, and just assuming that all the final positions end up in the correct order.
+
+  The final set of reduction positions that I will be proving both $(X_i)_i$ and $(Y_j)_j$ reach is $\alpha, (\alpha;\delta_i;\gamma', (\alpha;\delta_i;\gamma';\epsilon_j;\beta')_j)_i$, where as usual the sequences are expanded out to the full list.
+For the sequence $(X_i)_i$, first the images of $\beta$ in $X_0$ are $(\gamma,\epsilon_i,\beta')$.
+For each of these in turn, it is swapped with $\alpha$.
+By the fact mentioned above that sub-positions are mapped to the same set of images except in case f, the images of $\gamma,\epsilon_j,\beta'$ after this swap are $(\alpha;\delta_i^j;\gamma',\epsilon_j,\beta')_i$, where $(\delta_i^j)_i$ is the same as $(\delta_i)_i$ except that in the case where this is not the first image of $\beta$ swapped below $\alpha$ and the reduction at $\alpha$ is a $\tY$-reduction (and therefore swaps with it proceed by case f), the positions of the relevant variable are taken after the reductions at $\gamma$ and $\gamma;\epsilon_k;\beta'$ for $k \leq j$ (this implies that the first of these, $\delta_i^n = \delta_i$).
+All the images $\alpha;\delta_i^j;\gamma';\epsilon_j;\beta'$ where $\delta_i^j > \lambda;@_1;\gamma';\epsilon_{j+1};\beta'$ are then swapped with $\alpha;\lambda;@_1;\gamma';\epsilon_{j+1};\beta'$ (which is $\alpha;\delta_i^{j+1};\gamma';\epsilon_{j+1};\beta'$ for some $i$).
+As in the case where $\alpha;\tY < \gamma, \beta$ and $\gamma$ and $\beta$ are disjoint, the effects of these swaps may duplicate or delete the reductions in such a way that all the remaining images of $\gamma;\epsilon_j;\beta'$ are $(\alpha;\delta_i^{j+1};\gamma';\epsilon_j;\beta'$.
+This is repeated with $\delta_i^{j+2}$ and so on until these images are all in the correct order, at which point they are $(\alpha;\delta_i;\gamma';\epsilon_j;\beta')_i$.
+After all of these are done, $\gamma$ is swapped with $\alpha$, yielding $(\alpha;\delta_i^{-1};\gamma')_i$.
+As with the images of $\beta$, the subset of these which are inside $\alpha;\lambda;@_1;\gamma';\epsilon_j;\beta'$ for each $j$ in turn are swapped with it, until they are all in the correct order and the images left of $\gamma$ are $(\alpha;\delta_i;\gamma')_i$.
+At this point, $X_i$ has the desired value.
+
+  The sequence $(Y_j)_j$ proceeds similarly.
+First $\gamma$ swaps with $\alpha$, yielding $\alpha$ and $(\alpha;\delta_i;\gamma')_i$, then $\beta$ swaps with $\alpha$, yielding one image of $\beta$ for each $\delta_i$, except that, in case f, those images at positions inside $\alpha;\lambda;@_1;\gamma'$ may differ because they are earlier in the reduction sequence than any of the images of $\gamma$.
+Each of the images of $\beta$ in turn it moved to the corresponding image of $\gamma$, except that those overlapping with $\alpha;\lambda;@_1;\gamma'$ may be duplicated or deleted on the way.
+The images of $\beta$ after just this process (which does not actually correspond to any term in $(Y_j)_j$, but it's sufficiently independent that the order doesn't matter that much) are $(\alpha;\delta_i;\gamma';\beta'')$ (where $\gamma;\beta'' = \beta$).
+The images of these for each $i$, after swapping with $\alpha;\delta_i;\gamma'$, is $(\alpha;\delta_i;\gamma';\epsilon_j;\beta')_j$, because the reductions produced by a swap are unaffected by its overall location, and the skeleton at $\alpha;\delta_i;\gamma'$ at the relevant point in the sequence is equal to the skeleton at $\gamma$ initially, therefore this is equivalent to immediately swapping $\beta$ and $\gamma$ as in $Y \sim_c X$ (except that in the case where the reduction at $\alpha$ is a $\tY$-reduction, the skeleton at $\alpha;\lambda;@_1;\gamma'$ is not actually equal to the skeleton initially at $\gamma$ because something was substituted in for the variable bound at $\alpha;\tY$, but this doesn't affect the variable involved in the reduction at $\gamma$, so this doesn't actually matter).
+Therefore also in this case, eventually the set of reductions in the relevant portion of $Y_j$ is $\alpha, (\alpha;\delta_i;\gamma', (\alpha;\delta_i;\gamma';\epsilon_j;\beta')_j)_i$, therefore it is equal to some $X_i$.
 \end{itemize}
+In summary, if the swap in $Y \sim_c X$ doesn't overlap with the swap $X \sim_c X_0$, the sequence of swaps can be rearranged until it does, and in any other case, there is some $Y_j$ that's equal to some $X_i$, therefore these sequences reach the same end point and if $Y \sim_c X$, then $Y_{cbv} = X_{cbv}$, and by chaining these together, if $Y \sim_c^* X$, $Y_{cbv} = X_{cbv}$.
 
 \paragraph{}
-If $Y$ is in CBV order and $Y \sim_c^* X$, then $Y_{cbv} = X_{cbv}$ but also $Y_{cbv} = Y$ therefore $X_{cbv}$ is the unique reduction sequence in call-by-value order that's related to $X$ by $\sim_c^*$.
+If $Y$ is in CBV order and $Y \sim_c^* X$, then $Y_{cbv} = X_{cbv}$ but also $Y_{cbv} = Y$ because the sequence $(Y_i)_i$ terminates immediately therefore $X_{cbv}$ is the unique reduction sequence in call-by-value order that's related to $X$ by $\sim_c^*$.
 \end{proof}

### 2020-10-08 probabilisticProgrammingMartingales.tex
Minor fix

@@
 If $(X,P) \sim_c (Y,P)$ then the final skeletons in $X$ and $Y$ are equal, and $(X,Q) \sim_c (Y,Q)$, therefore $\sim_c$ can be considered to apply to reduction sequences even without positions, $X \sim_c Y$ iff $(X,\cdot) \sim_c (Y,\cdot)$.
+The structure of $\sim_c$ can be seen more clearly by choosing a canonical example from each equivalence class.
+For this, the member of the class that's in call-by-value order is used.
+A reduction sequence is in call-by-value order if for every pair of reductions in the sequence at positions $\alpha$ and $\beta$, where the reduction at $\alpha$ happens first, either $\alpha \leq \beta$ or $\alpha$ and $\beta$ are disjoint, with $\alpha$ to the left of $\beta$ (i.e.~the first elements of the sequences $\alpha, \beta$ where they differ are $@_1$ and $@_2$, $\underline f_i$ and $\underline f_j$ for $i < j$, or $\textsf{if}_i$ and $\textsf{if}_j$ for $i < j$, in that order).
+Note that although the reductions that occur in the sequence are in call-by-value order, it may not be the actual call-by-value reduction sequence starting from that point because some redexes may remain un-reduced even when other reductions that should happen afterwards occur.
+It is only required that the reductions that do occur occur in the correct order.
 
@@
 \begin{proof}
-A reduction sequence is in call-by-value order if for every pair of reductions in the sequence at positions $\alpha$ and $\beta$, where the reduction at $\alpha$ happens first, either $\alpha \leq \beta$ or $\alpha$ and $\beta$ are disjoint, with $\alpha$ to the left of $\beta$ (i.e.~the first elements of the sequences $\alpha, \beta$ where they differ are $@_1$ and $@_2$, $\underline f_i$ and $\underline f_j$ for $i < j$, or $\textsf{if}_i$ and $\textsf{if}_j$ for $i < j$, in that order.
-Note that although the reductions that occur in $X_{cbv}$ are in call-by-value order, it may not be the actual call-by-value reduction sequence because some redexes may remain un-reduced even when other reductions that should happen afterwards occur.
-It is only required that the reductions that do occur occur in the correct order.
+
 \lo{This paragraph (especially the last sentence) is ambiguous.

### 2020-10-09 probabilisticProgrammingMartingales.tex
Rename variables, check "skeleton" vs. "term"
Various minor changes, but the main parts are ensuring that term and
skeleton variables are consistently uppercase, and position variables
are consistently lowercase greek. Also, check that it refers to skeletons
rather than terms in the appropriate places after the redefinition of
potential positions, and various other minor fixes.

@@
   & \text{types } A, B ::= \textsf{R}  \mid  A \to B \\
-  & \text{values } v ::= \lambda x.s  \mid  \underline{r} \\
-  & \text{terms } s, t ::= v  \mid  x  \mid  t_1 \, t_2  \mid  \underline{f}(s_1,\dots ,s_n)  \mid  \tY s  \mid  \tif{s < 0}{t_1}{t_2}  \mid  \tsample  \mid  \tscore(s)
+  & \text{values } V ::= \lambda x.M  \mid  \underline{r} \\
+  & \text{terms } M, N ::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY M  \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample  \mid  \tscore(M)
 \end{align*}
@@
   \frac{}{\Gamma ; x:A \vdash x:A} \qquad
-  \frac{\Gamma ; x:A \vdash s : B}{\Gamma \vdash \lambda x.s : A \to B} \qquad
+  \frac{\Gamma ; x:A \vdash M : B}{\Gamma \vdash \lambda x.M : A \to B} \qquad
   \frac{}{\underline{r} : \textsf{R}} \qquad
-  \frac{\Gamma \vdash s:A \to B \quad \Gamma \vdash t : A}{\Gamma \vdash s \, t : B} \\ \\
-  \frac{\Gamma \vdash s_1:\textsf{R} \dots \Gamma \vdash s_n:\textsf{R}}{\Gamma \vdash \underline{f}(s_1,\dots,s_n) : \textsf{R}} \ f : \mathbb{R}^n \to \mathbb{R} \qquad
-  \frac{\Gamma \vdash s : (A \to B) \to (A \to B)}{\Gamma \vdash \tY s : (A \to B)} \\ \\
-  \frac{\Gamma \vdash c : \textsf{R} \quad \Gamma \vdash s_1 : A \quad \Gamma \vdash s_2 : A}{\Gamma \vdash \tif{c < 0}{s_1}{s_2} : A} \qquad
+  \frac{\Gamma \vdash M:A \to B \quad \Gamma \vdash N : A}{\Gamma \vdash M \, N : B} \\ \\
+  \frac{\Gamma \vdash M_1:\textsf{R} \dots \Gamma \vdash M_n:\textsf{R}}{\Gamma \vdash \underline{f}(M_1,\dots,M_n) : \textsf{R}} \ f : \mathbb{R}^n \to \mathbb{R} \qquad
+  \frac{\Gamma \vdash M : (A \to B) \to (A \to B)}{\Gamma \vdash \tY M : (A \to B)} \\ \\
+  \frac{\Gamma \vdash M : \textsf{R} \quad \Gamma \vdash N_1 : A \quad \Gamma \vdash N_2 : A}{\Gamma \vdash \tif{M < 0}{N_1}{N_2} : A} \qquad
   \frac{}{\Gamma \vdash \tsample : \textsf{R}} \qquad
@@
 
-The set of all terms is denoted $\Lambda$, and the set of closed terms is denoted $\Lambda^0$.
+Terms are identified up to $\alpha$-equivalence, as usual. The set of all terms is denoted $\Lambda$, and the set of closed terms is denoted $\Lambda^0$.
 
@@
 \begin{align*}
-  E ::= & \, [\,] \mid E \, t \mid v \, E \mid \underline{f}(r_1,\dots ,r_{k-1}, E, s_{k+1}, \dots, s_n) \\ & \mid \tY E \mid \tif{E<0}{s_1}{s_2} \mid \tscore (E)
+  E ::= & \, [\,] \mid E \, M \mid V \, E \mid \underline{f}(\underline r_1,\dots ,\underline r_{k-1}, E, M_{k+1}, \dots, M_n) \\ & \mid \tY E \mid \tif{E<0}{M_1}{M_2} \mid \tscore (E)
 \end{align*}
@@
 \begin{align*}
-  E[(\lambda x.s) v] & \to E[s[v/x]] \\
+  E[(\lambda x.M) N] & \to E[M[N/x]] \\
   E[\underline f (\underline r_1, \dots , \underline r_n)] & \to E[\underline{f(r_1,\dots,r_n)}] \\
-  E[\tY \lambda x. s] & \to E[\lambda z. s[(\tY \lambda x. s)/x] z] \text{ where $z$ is not free in $s$}\\
-  E[\tif{\underline r < 0}{s_1}{s_2}] & \to E[s_1] \text{ where }r < 0 \\
-  E[\tif{\underline r < 0}{s_1}{s_2}] & \to E[s_2] \text{ where }r \geq 0 \\
+  E[\tY \lambda x. M] & \to E[\lambda z. M[(\tY \lambda x. M)/x] z] \text{ where $z$ is not free in $s$}\\
+  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_1] \text{ where }r < 0 \\
+  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_2] \text{ where }r \geq 0 \\
   E[\tsample] & \to E[\underline r] \text{ where } r \in [0,1] \\
@@
 Every closed well-typed term either is a value or reduces to another term.
-
-\lo{NOTATION. A minor point: metavariables for terms are $s, t$, etc.~up to this point, but $M, N$, etc.~here after all.}
 
@@
 \begin{align*}
-P ::= \cdot \mid \lambda ; P \mid @_1 ; P \mid @_2 ; P \mid \underline f_i ; P \mid \tY ; P \mid \textsf{if}_1 ; P \mid \textsf{if}_2 ; P \mid \textsf{if}_3 ; P \mid \tscore ; P.
-\end{align*}
-The \emph{subterm of $M$ at a position $P$}, denoted $M \mid P$, is defined as
+\alpha ::= \cdot \mid \lambda ; \alpha \mid @_1 ; \alpha \mid @_2 ; \alpha \mid \underline f_i ; \alpha \mid \tY ; \alpha \mid \textsf{if}_1 ; \alpha \mid \textsf{if}_2 ; \alpha \mid \textsf{if}_3 ; \alpha \mid \tscore ; \alpha.
+\end{align*}
+The \emph{subterm of $M$ at a position $\alpha$}, denoted $M \mid \alpha$, is defined as
 \begin{align*}
 M \mid \cdot & = M \\
-\lambda x. M \mid \lambda ; P & = M \mid P \\
-M_1 M_2 \mid @_i ; P & = M_i \mid P \quad \text{for } i = 1,2 \\
-\underline f(M_1,\dots,M_n) \mid \underline f_i ; P & = M_i \mid P \quad \text{for }i \leq n \\
-\tY M \mid \tY ; P & = M \mid P \\
-\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; P & = M_i \mid P \quad \text{for } i = 1,2,3 \\
-\tscore(M) \mid \tscore ; P & = M \mid P
-\end{align*}
-so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda ; \cdot$ is undefined). 
-A position such that $M\mid P$ does exist is said to \emph{occur} in $M$. 
+\lambda x. M \mid \lambda ; \alpha & = M \mid \alpha \\
+M_1 M_2 \mid @_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2 \\
+\underline f(M_1,\dots,M_n) \mid \underline f_i ; \alpha & = M_i \mid \alpha \quad \text{for }i \leq n \\
+\tY M \mid \tY ; \alpha & = M \mid \alpha \\
+\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2,3 \\
+\tscore(M) \mid \tscore ; \alpha & = M \mid \alpha
+\end{align*}
+so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda$ is undefined). 
+A position such that $M\mid \alpha$ does exist is said to \emph{occur} in $M$. 
 %\changed[lo]{
-\emph{Substitution} of $N$ at position $P$ in $M$, written $M[N/P]$, is defined similarly.
+\emph{Substitution} of $N$ at position $\alpha$ in $M$, written $M[N/\alpha]$, is defined similarly.
 For example, let 
@@
 \qquad
-P =\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2
+\alpha =\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2
 \]
@@
 \(
-M[\tsample / P] = \lambda x \, y. y \, (\tif{x < 0}{y \, \tsample}{\underline 3}).
+M[\tsample / \alpha] = \lambda x \, y. y \, (\tif{x < 0}{y \, \tsample}{\underline 3}).
 \)
@@
 
-Two subterms $N_1$ and $N_2$ of a term $M$, corresponding to positions $P_1$ and $P_2$, can overlap in a few different ways. 
-If $P_1$ is an initial segment (i.e.~prefix) of $P_2$ (written as $P_1 \leq P_2$), then $N_2$ is also a subterm of $N_1$.
-If neither $P_1 \leq P_2$ nor $P_1 \geq P_2$, the positions are said to be disjoint.
+Two subterms $N_1$ and $N_2$ of a term $M$, corresponding to positions $\alpha_1$ and $\alpha_2$, can overlap in a few different ways. 
+If $\alpha_1$ is an initial segment (i.e.~prefix) of $\alpha_2$ (written as $\alpha_1 \leq \alpha_2$), then $N_2$ is also a subterm of $N_1$.
+If neither $\alpha_1 \leq \alpha_2$ nor $\alpha_1 \geq \alpha_2$, the positions are said to be disjoint.
 This is mostly relevant in that in any substitution at some position, the subterms at all other positions are unaffected. 
@@
 \label{def:more general red}
-The binary relation $\to$ is defined by the following rules, each is conditional on a redex occurring at position $P$ in the term $M$:
-\begin{align*}
-  \text{if } M \mid P = (\lambda x.s) v,\ & M \to M[s[v/x]/P] \\
-  \text{if } M \mid P = \underline f (\underline r_1, \dots , \underline r_n),\ & M \to M[\underline{f(r_1,\dots,r_n)}/P] \\
-  \text{if } M \mid P = \tY \lambda x. s,\ & M \to M[\lambda z. s[(\tY \lambda x. s)/x] z/P] \text{ where $z$ is not free in $s$}\\
-  \text{if } M \mid P = \tif{\underline r < 0}{s_1}{s_2},\ & M \to M[s_1/P] \text{ where }r < 0 \\
-  \text{if } M \mid P = \tif{\underline r < 0}{s_1}{s_2},\ & M \to M[s_2/P] \text{ where }r \geq 0 \\
-  \text{if } M \mid P = \tsample & \text{ and $\lambda$ does not occur after $@_2$ or $\tY$ in $P$},\\ & M \to M[\underline r/P] \text{ where } r \in [0,1] \\
-  \text{if } M \mid P = \tscore(\underline r),\ & M \to M[\underline r/P].
-\end{align*}
-In each of these cases, $M \mid P$ is the \emph{redex}, and reduction takes place at $P$.
+The binary relation $\to$ is defined by the following rules, each is conditional on a redex occurring at position $\alpha$ in the term $M$:
+\begin{align*}
+  \text{if } M \mid \alpha = (\lambda x.N) V,\ & M \to M[N[V/x]/\alpha] \\
+  \text{if } M \mid \alpha = \underline f (\underline r_1, \dots , \underline r_n),\ & M \to M[\underline{f(r_1,\dots,r_n)}/\alpha] \\
+  \text{if } M \mid \alpha = \tY \lambda x. N,\ & M \to M[\lambda z. N[(\tY \lambda x. N)/x] z/\alpha] \text{ where $z$ is not free in $N$}\\
+  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_1/\alpha] \text{ where }r < 0 \\
+  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_2/\alpha] \text{ where }r \geq 0 \\
+  \text{if } M \mid \alpha = \tsample & \text{ and $\lambda$ does not occur after $@_2$ or $\tY$ in $\alpha$},\\ & M \to M[\underline r/\alpha] \text{ where } r \in [0,1] \\
+  \text{if } M \mid \alpha = \tscore(\underline r),\ & M \to M[\underline r/\alpha].
+\end{align*}
+In each of these cases, $M \mid \alpha$ is the \emph{redex}, and reduction takes place at $\alpha$.
 \end{definition}
@@
 \lo{Skeleton is defined in Sec.~\ref{sec:sampling semantics}.}
-The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, \changed[lo]{and $M \to M[\skeletonPlaceholder / P]$, in case $M \mid P = \tsample$ and $\lambda$ does not occur after $@_2$ or $\tY$ in $P$.
-(By abuse of notation, we use the same metavariables $M, N$, etc.~for skeletons as for terms.)}
+The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, and $\tsample$ reducing to $\skeletonPlaceholder$.
 For example, we have 
@@
 Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a $\to$-reduction sequence of skeletons starting at $Sk(M)$, and the second of which is a position in the final skeleton of the reduction sequence.
-\akr{I've yet to go through and replace references to terms with references to skeletons, but it should be reasonably obvious where the substitution should take place. 
-This is potentially even more confusing though because I'm now implicitly identifying terms, skeletons, reduction sequences and skeletal reduction sequences.} 
-Reduction sequences are used rather than reachable \changed[lo]{skeletons} because if the same \changed[lo]{skeleton} is reached twice, different samples may be needed, \lo{Give a simple example illustrating this.} but they will be discussed as though they were terms, by abuse of notation.
+Reduction sequences are used rather than reachable skeletons because if the same skeleton is reached twice, different samples may be needed, \lo{Give a simple example illustrating this.} but the reduction sequences of skeletons will be discussed as though they were just skeletons, identifying them with their final skeletons.
+With this abuse of notation, a reduction sequence $N$ (actually $N_1 \to \dots \to N_n = N$) may be said to reduce to a reduction sequence $O$, where the reduction sequence implicitly associated with the final skeleton $O$ is $N_1 \to \dots \to N_n \to O$.
 
@@
 \begin{enumerate}
-    \item If $N$ reduces to $O$ with the redex at position $P$ \changed[lo]{(i.e.~$M \to^\ast N \to O = N[\dotDelta / P]$ with $O \mid P = \Delta$ for some redex contraction $\Delta \to \dotDelta$)}, and $Q$ is a position in $N$ disjoint from $P$, then $(N,Q) \sim_p (O,Q)$.
-
-    \item If $N$ $\beta$-reduces to $O$ at position $P$, $Q$ is a position in $N \mid P;@_1;\lambda$ and $N \mid P;@_1;\lambda;Q$ is not the variable involved in the reduction, $(N,P;@_1;\lambda;Q) \sim_p (O, P;Q)$
-
-    \item If $N$ $\textsf{if}$-reduces to $O$ at position $P$, with the first resp.
-second branch being taken, and $P;\textsf{if}_i;Q$ occurs in $N$ (where $i = 2$ resp.
-$3$), $(N,P;\textsf{if}_i;Q) \sim_p (O,P;Q)$
+    \item If $N$ reduces to $O$ with the redex at position $\alpha$, and $\beta$ is a position in $N$ disjoint from $\alpha$, then $(N,\beta) \sim_p (O,\beta)$.
+
+    \item If $N$ $\beta$-reduces to $O$ at position $\alpha$, $\beta$ is a position in $N \mid \alpha;@_1;\lambda$ and $N \mid \alpha;@_1;\lambda;\beta$ is not the variable involved in the reduction, $(N,\alpha;@_1;\lambda;\beta) \sim_p (O, \alpha;\beta)$
+
+    \item If $N$ $\textsf{if}$-reduces to $O$ at position $\alpha$, with the first resp.
+second branch being taken, and $\alpha;\textsf{if}_i;\beta$ occurs in $N$ (where $i = 2$ resp.
+$3$), $(N,\alpha;\textsf{if}_i;\beta) \sim_p (O,\alpha;\beta)$
 
@@
     \begin{enumerate}
-        \item $N$ contains redexes at disjoint positions $P_1$ and $P_2$, $O_1$ is $N$ reduced first at $P_1$ then $P_2$ and $O_2$ is $N$ reduced first at $P_2$ then at $P_1$.
-
-        \item $N \mid P = \tif{\underline r < 0}{N_1}{N_2}$, where $r < 0$ (or, respectively, $r \geq 0$), $(N_2 \text{ resp.
-} N_1) \mid Q$ is a redex, and $O_1$ is $N$ reduced at $P$ and $O_2$ is $N$ reduced first at $P;(\textsf{if}_3 \text{ resp.
-} \textsf{if}_2);Q$ then at $P$
-
-        \item $N \mid P = \tif{\underline r < 0}{N_1}{N_2}$, where $r < 0$ (or, respectively, $r \geq 0$), $(N_1 \text{ resp.
-} N_2) \mid Q$ is a redex, and $O_1$ is $N$ reduced first at $P$ then at $P;Q$ and $O_2$ is $N$ reduced first at $P;(\textsf{if}_2 \text{ resp.
-} \textsf{if}_3);Q$ then at $P$
-
-        \item $N \mid P = (\lambda x. A) B$, there is a redex in $A$ at position $Q$, $O_1$ is $N$ reduced first at $P$ then at $P;Q$, and $O_2$ is $N$ reduced first at $P;@_1;\lambda;Q$ then at $P$
-
-        \item $N \mid P = (\lambda x.
-A) B$, $B \mid Q$ is a redex, $(R_i)_i$ is a list of all the positions in $A$ where $A \mid R = x$, ordered from left to right, $O_1$ is $N$ reduced first at $P ; @_2 ; Q$ then at $P$, and $O_2$ is $N$ reduced first at $P$ then at $P;R_i;Q$ for each $i$ in order.
-
-        \item $N \mid P = \tY (\lambda x.
-A)$, $A$ reduced at $Q$ is $A'$, $(R_i)_i$ is a list of all the positions where $A' \mid R  = x$, ordered from left to right, $O_1$ is $N$ reduced first at $P;\tY;\lambda;Q$ then at $P$, and $O_2$ is $N$ reduced first at $P$ then at $P;\lambda;@_1;R_i;Y;\lambda;Q$ for each $i$ in order where $R_i$ is left of $Q$ then at $P;\lambda;@_1;Q$ then at $P;\lambda;@_1;R_i;Y;\lambda;Q$ for the remaining values of $i$.
+        \item $N$ contains redexes at disjoint positions $\alpha_1$ and $\alpha_2$, $O_1$ is $N$ reduced first at $\alpha_1$ then $\alpha_2$ and $O_2$ is $N$ reduced first at $\alpha_2$ then at $\alpha_1$.
+
+        \item $N \mid \alpha = \tif{\underline r < 0}{N_1}{N_2}$, where $r < 0$ (or, respectively, $r \geq 0$), $(N_2 \text{ resp.
+} N_1) \mid \beta$ is a redex, and $O_1$ is $N$ reduced at $\alpha$ and $O_2$ is $N$ reduced first at $\alpha;(\textsf{if}_3 \text{ resp.
+} \textsf{if}_2);\beta$ then at $\alpha$
+
+        \item $N \mid \alpha = \tif{\underline r < 0}{N_1}{N_2}$, where $r < 0$ (or, respectively, $r \geq 0$), $(N_1 \text{ resp.
+} N_2) \mid \beta$ is a redex, and $O_1$ is $N$ reduced first at $\alpha$ then at $\alpha;\beta$ and $O_2$ is $N$ reduced first at $\alpha;(\textsf{if}_2 \text{ resp.
+} \textsf{if}_3);\beta$ then at $\alpha$
+
+        \item $N \mid \alpha = (\lambda x.
+A) B$, there is a redex in $A$ at position $\beta$, $O_1$ is $N$ reduced first at $\alpha$ then at $\alpha;\beta$, and $O_2$ is $N$ reduced first at $\alpha;@_1;\lambda;\beta$ then at $\alpha$
+
+        \item $N \mid \alpha = (\lambda x.
+A) B$, $B \mid \beta$ is a redex, $(\gamma_i)_i$ is a list of all the positions in $A$ where $A \mid \gamma = x$, ordered from left to right, $O_1$ is $N$ reduced first at $\alpha ; @_2 ; \beta$ then at $\alpha$, and $O_2$ is $N$ reduced first at $\alpha$ then at $\alpha;\gamma_i;\beta$ for each $i$ in order.
+
+        \item $N \mid \alpha = \tY (\lambda x.
+A)$, $A$ reduced at $\beta$ is $A'$, $(\gamma_i)_i$ is a list of all the positions where $A' \mid \gamma  = x$, ordered from left to right, $O_1$ is $N$ reduced first at $\alpha;\tY;\lambda;\beta$ then at $\alpha$, and $O_2$ is $N$ reduced first at $\alpha$ then at $\alpha;\lambda;@_1;\gamma_i;Y;\lambda;\beta$ for each $i$ in order where $\gamma_i$ is left of $\beta$ then at $\alpha;\lambda;@_1;\beta$ then at $\alpha;\lambda;@_1;\gamma_i;Y;\lambda;\beta$ for the remaining values of $i$.
     \end{enumerate}
-    (in which case $O_1$ and $O_2$ are equal as terms, but with different reduction sequences), $O_1'$ and $O_2'$ are the results of applying some reduction sequence to each of $O_1$ and $O_2$ (the same reductions in each case, which is always possible because they're equal terms), and $S$ is a position in $O_1'$ (or equivalently $O_2'$), then $(O_1',S) \sim_c (O_2',S)$.
+    (in which case $O_1$ and $O_2$ are equal as skeletons, but with different reduction sequences), $O_1'$ and $O_2'$ are the results of applying some reduction sequence to each of $O_1$ and $O_2$ (the same reductions in each case, which is always possible because they're equal terms), and $\delta$ is a position in $O_1'$ (or equivalently $O_2'$), then $(O_1',\delta) \sim_c (O_2',\delta)$.
 \end{enumerate}
@@
 
-%\lo{TODO: Typeset Fig.~\ref{fig:adk-diagram} in vector graphics using PGF/TikZ.}
-
-The reflexive transitive closure $\sim^*$ of this relation is used to define the set of \emph{potential positions} $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable \changed[lo]{skeletons}.
-If $(N,P) \sim^* (O,Q)$, then $N \mid P$ and $O \mid Q$ both have the same shape (i.e.~they're either \changed[lo]{both placeholders}, both variables, both applications, both $\tsample$s etc.), therefore it's well-defined to talk of the set of potential positions where there is a $\tsample$, $L_s(M)$.
-\changed[lo]{Formally $L_s(M) := \{[(X, P)]_{\sim^\ast_M} : X \mid P = \tsample\}$.}
+The reflexive transitive closure $\sim^*$ of this relation is used to define the set of \emph{potential positions} $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable skeletons.
+If $(N,\alpha) \sim^* (O,\beta)$, then $N \mid \alpha$ and $O \mid \beta$ both have the same shape (i.e.~they're either both the placeholder $\skeletonPlaceholder$, both variables, both applications, both $\tsample$s etc.), therefore it's well-defined to talk of the set of potential positions where there is a $\tsample$, $L_s(M)$.
+Formally $L_s(M) := \{[(X, \alpha)]_{\sim^\ast_M} : X \mid \alpha = \tsample\}$.
 The new sample space is then defined as $I^{L_s(M)}$, with the Borel $\sigma$-algebra and product measure.
@@
     \item Given the definition below of $\Rightarrow$, which is still applicable to this version of $\sim$, and which is the equivalent of $\red$ in this version of the semantics, $\Rightarrow$ is Church-Rosser.
-    \item If $M \to N$ at $P$, no position in any term in $Rch(N)$ is related to $(M,P)$ by $\sim_M^*$, and therefore each sample is used at most once in any reduction sequence.
+    \item If $M \to N$ at $\alpha$, no position in any term in $Rch(N)$ is related to $(M,\alpha)$ by $\sim_M^*$, and therefore each sample is used at most once in any reduction sequence.
     \item The subprobability distribution of values obtained by reducing a term in the usual call-by-value semantics is equal to that obtained by any reduction strategy on $\Rightarrow$ (i.e.~a sub-relation of $\Rightarrow$ that is a function) (with some allowance made for the fact that some values are not normal forms of $\Rightarrow$, so the reduction strategy might overshoot the value that $\red^\infty$ reaches.
@@
 \begin{lemma} \label{downAcrossUp}
-If $(N_1,P_1) \sim^* (N_2,P_2)$, there is some descendants $N'_1, N'_2$ of $N_1$ resp.~$N_2$, and a position $P'$ in both of them such that $(N_1,P_1) \sim_p^* (N'_1, P') \sim_c^* (N'_2, P') \sim_p^* (N_2, P_2)$.
+If $(N_1,\alpha_1) \sim^* (N_2,\alpha_2)$, there is some descendants $N'_1, N'_2$ of $N_1$ resp.~$N_2$, and a position $\alpha'$ in both of them such that $(N_1,\alpha_1) \sim_p^* (N'_1, \alpha') \sim_c^* (N'_2, \alpha') \sim_p^* (N_2, \alpha_2)$.
 \end{lemma}
 \begin{proof}
-The $\sim$ relation can be split into $\sim_\downarrow \cup \sim_\uparrow$, where $(A,P_A) \sim_\downarrow (B,P_B)$ if either $A \to B$ and $(A,P_A) \sim_p (B,P_B)$ or $(A,P_A) \sim_c (B,P_B)$, and similarly $(A,P_A) \sim_\uparrow (B,P_B)$ if either $B \to A$ and $(A,P_A) \sim_p (B,P_B)$ or $(A,P_A) \sim_c (B,P_B)$.
+The $\sim$ relation can be split into $\sim_\downarrow \cup \sim_\uparrow$, where $(A,\alpha) \sim_\downarrow (B,\beta)$ if either $A \to B$ and $(A,\alpha) \sim_p (B,\beta)$ or $(A,\alpha) \sim_c (B,\beta)$, and similarly $(A,\alpha) \sim_\uparrow (B,\beta)$ if either $B \to A$ and $(A,\alpha) \sim_p (B,\beta)$ or $(A,\alpha) \sim_c (B,\beta)$.
 The sequence of $\sim$ relations linking $N_1$ to $N_2$ can then be expressed as an alternating sequence of $\sim_\downarrow^*$ and $\sim_\uparrow^*$s, starting with $\sim_\downarrow^*$ and ending with $\sim_\uparrow^*$, and the proof proceeds by induction on the number of these alternations.
 
-In the base case, that $(N_1,P_1) \sim_\downarrow^* X \sim_\uparrow^* (N_2,P_2)$, \dots
+In the base case, that $(N_1,\alpha_1) \sim_\downarrow^* X \sim_\uparrow^* (N_2,\alpha_2)$, \dots
 
@@
 
-If $(X,P) \sim_c (Y,P)$ then the final skeletons in $X$ and $Y$ are equal, and $(X,Q) \sim_c (Y,Q)$, therefore $\sim_c$ can be considered to apply to reduction sequences even without positions, $X \sim_c Y$ iff $(X,\cdot) \sim_c (Y,\cdot)$.
+If $(X,\alpha) \sim_c (Y,\alpha)$ then the final skeletons in $X$ and $Y$ are equal, and $(X,\beta) \sim_c (Y,\beta)$ for any $\beta$ that occurs in this skeleton, therefore $\sim_c$ can be considered to apply to reduction sequences even without positions, $X \sim_c Y$ iff $(X,\cdot) \sim_c (Y,\cdot)$.
 The structure of $\sim_c$ can be seen more clearly by choosing a canonical example from each equivalence class.
@@
 \begin{proof}
-
-\lo{This paragraph (especially the last sentence) is ambiguous.
-What does it mean to say that that a reductions that occur in a reduction sequences are \emph{in CBV order}?}
-\iffalse
-\lo{So this is permitted?
-\[
-(II)(I (I \bar I)) \to
-(II)(I\bar I) \to
-(II)\bar I 
-\]
-but not this: 
-\(
-(II)(I (I \bar I)) \to
-(II)(I \bar I) \to
-(II)\bar I 
-\)?}
-\fi
 
@@
 
-In the other direction, suppose $(N_1, P_1) \sim_M^* (N_2, P_2)$ where both $N_1$ and $N_2$ are descendants of $N$.
-By Lemma~\ref{downAcrossUp}, take $(N_1, P_1) \sim_{p,M}^* (N'_1, P') \sim_{c,M}^* (N'_2, P') \sim_{p,M}^* (N_2, P_2)$.
-The $\sim_p^*$ steps remain within $Rch(N)$, and $\sim_p$ does not depend on the history of the reduction sequences, therefore $(N_1, P_1) \sim_{p,N}^* (N'_1, P') \sim_{c,M}^* (N'_2, P') \sim_{p,N}^* (N_2, P_2)$.
+In the other direction, suppose $(N_1, \alpha_1) \sim_M^* (N_2, \alpha_2)$ where both $N_1$ and $N_2$ are descendants of $N$.
+By Lemma~\ref{downAcrossUp}, take $(N_1, \alpha_1) \sim_{p,M}^* (N'_1, \alpha') \sim_{c,M}^* (N'_2, \alpha') \sim_{p,M}^* (N_2, \alpha_2)$.
+The $\sim_p^*$ steps remain within $Rch(N)$, and $\sim_p$ does not depend on the history of the reduction sequences, therefore $(N_1, \alpha_1) \sim_{p,N}^* (N'_1, \alpha') \sim_{c,M}^* (N'_2, \alpha') \sim_{p,N}^* (N_2, \alpha_2)$.
 
 Let $X$ be the call-by-value reduction sequence related to both $N'_1$ and $N'_2$ by $\sim_{c,M}$ given by Lemma~\ref{canonicalCousins}. 
-As $M \to N$ is the first reduction in the sequence $N'_1$, it is the last to be affected by the $\sim_{c,M}$ sequence $N'_1 \sim_{c,M}^* X$ given by Lemma \ref{canonicalCousins} therefore it can be split into $N'_1 \sim_{c,M}^* Y_1 \sim_{c,M}^* X$ where $Y_1$ is in CBV order except possibly for its first reduction, which is still $M -> N$.
-Let the position of the reduction $M \to N$ be $Q$.
+As $M \to N$ is the first reduction in the sequence $N'_1$, it is the last to be affected by the $\sim_{c,M}$ sequence $N'_1 \sim_{c,M}^* X$ given by Lemma \ref{canonicalCousins} therefore it can be split into $N'_1 \sim_{c,M}^* Y_1 \sim_{c,M}^* X$ where $Y_1$ is in CBV order except possibly for its first reduction, which is still $M \to N$.
+Let the position of the reduction $M \to N$ be $\beta$.
 \akr{TODO: fill in the details that are missing here. It shouldn't be too hard, it's just a bit sketchy at present.} 
-The rearrangement of the reduction sequence $Y_1 \sim_{c,M} X$ consists of moving the reduction at $Q$ down past the other reductions in $Y_1$, possibly duplicating or deleting it in the process, but not affecting the positions or the correct order for any of the other reductions.
+The rearrangement of the reduction sequence $Y_1 \sim_{c,M} X$ consists of moving the reduction at $\beta$ down past the other reductions in $Y_1$, possibly duplicating or deleting it in the process, but not affecting the positions or the correct order for any of the other reductions.
 The reductions derived from $M \to N$ can be identified as follows:
@@
 %\begin{definition}
-A position in some (reduction sequence of) term(s) is \emph{derived from} another if it is related by the reflexive transitive closure of $\rightsquigarrow$, where $(V,R) \rightsquigarrow (W,T)$ just if $V \to W$ and one of the following cases holds:
+A position in some (reduction sequence of) term(s) is \emph{derived from} another if it is related by the reflexive transitive closure of $\rightsquigarrow$, where $(V,\gamma) \rightsquigarrow (W,\delta)$ just if $V \to W$ and one of the following cases holds:
 \begin{itemize}
-    \item $(V,R) \sim_p (W,T)$
-    \item $R = T$, $V \to W$ at $A$, and $A > R$
-    \item $V \mid A = (\lambda x.U) Z, U | B = x, R = A;@_2;C$ and $T = A;B;C$
-    \item $V \mid A = \tY (\lambda x.U), R = A;\tY;\lambda;B$ and $T = A;\lambda;@_1;B$
-    \item $V \mid A = \tY (\lambda x.U), U | B = x, R = A;C$ and $T = A;\lambda;@_1;B;C$
+    \item $(V,\gamma) \sim_p (W,\delta)$
+    \item $\gamma = \delta$, $V \to W$ at $\epsilon$, and $\epsilon > \gamma$
+    \item $V \mid \epsilon = (\lambda x.U) Z, U | \zeta = x, \gamma = \epsilon;@_2;\theta$ and $\delta = \epsilon;\zeta;\theta$
+    \item $V \mid \epsilon = \tY (\lambda x.U), \gamma = \epsilon;\tY;\lambda;\zeta$ and $\delta = \epsilon;\lambda;@_1;\zeta$
+    \item $V \mid \epsilon = \tY (\lambda x.U), U | \zeta = x, \gamma = \epsilon;\theta$ and $\delta = \epsilon;\lambda;@_1;\zeta;\theta$
 \end{itemize}
@@
 
-Crucially, the reductions in $X$ can be partitioned into 2 sets: those at positions derived from $(M,Q)$, and those at positions equal to the reductions in $Y_1$ (in the same order as they occur in $Y_1$.
-Using the same construction for $N'_2$ shows that the positions of the reductions in $Y_2$ are also the positions of the reductions in $X$ other than those derived from $(M,Q)$, therefore $Y_1 = Y_2$ therefore $N'_1 \sim_{c,M} Y_1 \sim_{c,M} N'_2$ and every reduction sequence in this sequence starts with $M \to N$ therefore $N'_1 \sim_{c,N} N'_2$.
-
-Combining this with the $\sim_{p,N}^*$s at the beginning and end then yields the desired result that $(N_1,P_1) \sim_N^* (N_2, P_2)$, therefore the restriction of $\sim_M^*$ to $\sim_N$'s domain ($Rch(N)$) is a subset of $\sim_N^*$ therefore the two versions of $\sim$ match on this domain, as desired.
+Crucially, the reductions in $X$ can be partitioned into 2 sets: those at positions derived from $(M,\beta)$, and those at positions equal to the reductions in $Y_1$ (in the same order as they occur in $Y_1$).
+Using the same construction for $N'_2$ shows that the positions of the reductions in $Y_2$ are also the positions of the reductions in $X$ other than those derived from $(M,\beta)$, therefore $Y_1 = Y_2$ therefore $N'_1 \sim_{c,M} Y_1 \sim_{c,M} N'_2$ and every reduction sequence in this sequence starts with $M \to N$ therefore $N'_1 \sim_{c,N} N'_2$.
+
+Combining this with the $\sim_{p,N}^*$s at the beginning and end then yields the desired result that $(N_1,\alpha_1) \sim_N^* (N_2, \alpha_2)$, therefore the restriction of $\sim_M^*$ to $\sim_N$'s domain ($Rch(N)$) is a subset of $\sim_N^*$ therefore the two versions of $\sim$ match on this domain, as desired.
 \end{proof}
@@
 & \qquad \text{$M \to N$ and the redex is not $\tsample$, or} \\
-& \qquad \text{$M \mid P = \tsample$, $N = M[\underline{s(Sk(M),P)}/P]$ and $\lambda$ does not occur after $@_2$ or $\tY$ in $P$}
+& \qquad \text{$M \mid \alpha = \tsample$, $N = M[\underline{s(Sk(M),\alpha)}/\alpha]$ and $\lambda$ does not occur after $@_2$ or $\tY$ in $\alpha$}
 \end{align*}
@@
 Suppose that $(M,s) \Rightarrow (M_1,s_1)$ and also $(M,s) \Rightarrow (M_2,s_2)$, then it is required to prove that there is some $(M',s')$ such that both $(M_1,s_1) \Rightarrow^* (M',s')$ and $(M_2,s_2) \Rightarrow^* (M',s')$.
-Let the positions of the redexes in $M \to M_1$ and $M \to M_2$ be $P_1$ and $P_2$ respectively.
-
-First consider the case that $P_1$ and $P_2$ are disjoint.
-Let $M_1 = M[X_1/P_1]$ and similarly for $X_2$, then let $M' = M[X_1/P_1][X_2/P_2]$.
+Let the positions of the redexes in $M \to M_1$ and $M \to M_2$ be $\alpha_1$ and $\alpha_2$ respectively.
+
+First consider the case that $\alpha_1$ and $\alpha_2$ are disjoint.
+Let $M_1 = M[X_1/\alpha_1]$ and similarly for $X_2$, then let $M' = M[X_1/\alpha_1][X_2/\alpha_2]$.
 As the positions are disjoint, the substitutions commute and both $M_1$ and $M_2$ reduce (with $\to$) to $M'$.
 Let $s' = s \circ i(M \to M_1) \circ i(M_1 \to M')$. 
-The injection $i(M \to M_1) \circ i(M_1 \to M')$ consists of prepending $M \to M_1 \to M'$ to each reduction sequence, but by case 4.a of $\sim$, this is equivalent to prepending $M \to M_2 \to M'$, which is $i(M \to M_2) \circ i(M_2 \to M')$.
+The injection $i(M \to M_1) \circ i(M_1 \to M')$ consists of prepending $M \to M_1 \to M'$ to each reduction sequence, but by case a of $\sim_c$, this is equivalent to prepending $M \to M_2 \to M'$, which is $i(M \to M_2) \circ i(M_2 \to M')$.
 In the case that the reduction $M \to M_2$ isn't a $\tsample$-reduction, this is enough to establish that $(M_1, s_1) \Rightarrow (M', s')$, and vice-versa. 
-If it is $\tsample$ though, in order for it to be the case that $(M_1, s_1) \Rightarrow (M', s')$, it is additionally necessary that $X_2$, the result of the reduction at $P_2$, be $s_1(M_1, P_2)$, which follows from the fact that $(M, P_2) \sim (M_1, P_2)$ by case 1.
+If it is $\tsample$ though, in order for it to be the case that $(M_1, s_1) \Rightarrow (M', s')$, it is additionally necessary that $X_2$, the result of the reduction at $\alpha_2$, be $s_1(M_1, \alpha_2)$, which follows from the fact that $(M, \alpha_2) \sim (M_1, \alpha_2)$ by case 1 of $\sim_p$.
 The case that the reduction $M \to M_1$ is a $\tsample$-reduction is similar.
 
-The case that $P_1 = P_2$ is trivial, because there is at most one possible $\Rightarrow$ reduction at any given position, therefore $(M_1, s_1) = (M_2, s_2)$ already.
-
-The remaining case is that $P_1 < P_2$ or $P_1 > P_2$. 
-Assume without loss of generality that $P_1 < P_2$. 
-For each possible case of what type of redex $M | P_1$ is, and $P_2$'s position within it, there is a corresponding case of $\sim_c$, and similarly to the case where $P_1$ and $P_2$ are disjoint, the term $O_1 = O_2$ from the definition of $\sim_c$ is a suitable value of $M'$.
-The term $M \mid P_1$ can't be $\tsample$, because it has strict subterms, but the case that $M \mid P_2 = \tsample$ is still somewhat more complicated. 
-$P_2$ can't be within $P_1 ; @_2$ or $P_1 ; Y$ (cases e and f of $\sim_c$) because those are not valid positions to reduce a $\tsample$, 
-and the case that $P_2$ is within the branch of an if statement that is deleted by the reduction at $P_1$ (case b) doesn't actually present a problem because there is no corresponding reduction to $M \to M_2$ in the other branch,
-which leaves cases c and d, that $P_2$ is in the $\tif{\cdot}{\cdot}{\cdot}$ branch that isn't deleted, and that $P_2 < P_1 ; @_1 ; \lambda$.  
-The values that the $\tsample$s take in these cases match because $(M,P) \sim_p (M_1, P;Q)$ by cases 3 and 2 of $\sim_p$ respectively.
+The case that $\alpha_1 = \alpha_2$ is trivial, because there is at most one possible $\Rightarrow$ reduction at any given position, therefore $(M_1, s_1) = (M_2, s_2)$ already.
+
+The remaining case is that $\alpha_1 < \alpha_2$ or $\alpha_1 > \alpha_2$. 
+Assume without loss of generality that $\alpha_1 < \alpha_2$. 
+For each possible case of what type of redex $M | \alpha_1$ is, and $\alpha_2$'s position within it, there is a corresponding case of $\sim_c$, and similarly to the case where $\alpha_1$ and $\alpha_2$ are disjoint, the term $O_1 = O_2$ from the definition of $\sim_c$ is a suitable value of $M'$.
+The term $M \mid \alpha_1$ can't be $\tsample$, because it has strict subterms, but the case that $M \mid \alpha_2 = \tsample$ is still somewhat more complicated. 
+$\alpha_2$ can't be within $\alpha_1 ; @_2$ or $\alpha_1 ; Y$ (cases e and f of $\sim_c$) because those are not valid positions to reduce a $\tsample$, 
+and the case that $\alpha_2$ is within the branch of an if statement that is deleted by the reduction at $\alpha_1$ (case b) doesn't actually present a problem because there is no corresponding reduction to $M \to M_2$ in the other branch,
+which leaves cases c and d, that $\alpha_2$ is in the $\tif{\cdot}{\cdot}{\cdot}$ branch that isn't deleted, and that $\alpha_2 < \alpha_1 ; @_1 ; \lambda$.  
+The values that the $\tsample$s take in these cases match because $(M,\alpha) \sim_p (M_1, \alpha;\beta)$ by cases 3 and 2 of $\sim_p$ respectively.
 \end{proof}
@@
 \begin{lemma} \label{simpDoesn'tMix}
-If $A$ is some descendant of $M$ and $(A,D) \sim^* (M,E)$, then $(A,D) \sim_p^* (M,E)$, with the length of the reduction sequences decreasing by one each step from $A$ to $M$.
+If $A$ is some descendant of $M$ and $(A,\gamma) \sim^* (M,\delta)$, then $(A,\gamma) \sim_p^* (M,\delta)$, with the length of the reduction sequences decreasing by one each step from $A$ to $M$.
 \end{lemma}
@@
 This is a simple induction on $\sim^*$.
-In the base case, $(A,D) = (M,E)$ therefore $(A,D) \sim_p^* (M,E)$ trivially.
-Otherwise, suppose that $(M,E) \sim_p^* (B,F) \sim (A,D)$.
-Either $(B,F) \sim_p (A,D)$ or $(B,F) \sim_c (A,D)$.
-In the first case, either $B \to A$, in which case the result follows directly, or $A \to B$, in which case the fact that each (reduction sequence of) term(s) has only one parent implies that a $(M,E) \sim_p^* (A,D)$ directly as a subsequence of the path to $(B,F)$.
+In the base case, $(A,\gamma) = (M,\delta)$ therefore $(A,\gamma) \sim_p^* (M,\delta)$ trivially.
+Otherwise, suppose that $(M,\delta) \sim_p^* (B,\epsilon) \sim (A,\gamma)$.
+Either $(B,\epsilon) \sim_p (A,\gamma)$ or $(B,\epsilon) \sim_c (A,\gamma)$.
+In the first case, either $B \to A$, in which case the result follows directly, or $A \to B$, in which case the fact that each (reduction sequence of) skeleton(s) has only one parent implies that a $(M,\delta) \sim_p^* (A,\gamma)$ directly as a subsequence of the path to $(B,\epsilon)$.
 
@@
 Either $O'_1 = A$ and $O'_2 = B$ or vice-versa.
-As $M \to^* N \to^* O_1 \to^* B$ and $(M,E) \sim_p^* (B,F)$, and $\sim_p$ only relates positions in a term and its parent, there are some positions $G, H$ such that $(M,E) \sim_p^* (N,G) \sim_p^* (O_1, H) \sim_p^* (B,F)$.
-It follows that $(O_2, H) \sim_p^* (A,D)$ by following the same path, therefore it suffices to provide the only missing portion of the path from $M$ to $A$, i.e.~to prove that $(N,G) \sim_p^* (O_2,H)$ given $(N,G) \sim_p^* (O_1,H)$ (or vice-versa).
-
-If $G$ is disjoint from all the positions of reduction from $N$ to $O_1$ and $O_2$ (and consequently $G = H$), this follows from case 1 of $\sim_p$.
+As $M \to^* N \to^* O_1 \to^* B$ and $(M,\delta) \sim_p^* (B,\epsilon)$, and $\sim_p$ only relates positions in a term and its parent, there are some positions $\zeta, \theta$ such that $(M,\delta) \sim_p^* (N,\zeta) \sim_p^* (O_1, \theta) \sim_p^* (B,\epsilon)$.
+It follows that $(O_2, \theta) \sim_p^* (A,\gamma)$ by following the same path, therefore it suffices to provide the only missing portion of the path from $M$ to $A$, i.e.~to prove that $(N,\zeta) \sim_p^* (O_2,\theta)$ given $(N,\zeta) \sim_p^* (O_1,\theta)$ (or vice-versa).
+
+If $\zeta$ is disjoint from all the positions of reduction from $N$ to $O_1$ and $O_2$ (and consequently $\zeta = \theta$), this follows from case 1 of $\sim_p$.
 Otherwise, this can be proved by taking cases from the definition of $\sim_c$.
@@
 
-Let $I$ be $N$ reduced at $P$, and $J$ be $N$ reduced at $P;@_1;\lambda;Q$, so that $N \to I \to O_1$ and $N \to J \to O_2$.
-All of the reduction positions are $\geq P$, and $G$ is not disjoint from all of them, therefore $G$ is not disjoint from $P$.
-Let $K$ be the position such that $(N,G) \sim_p (I,K) \sim_p (O_1,H)$.
-The fact that $(N,G) \sim_p (I,K)$ implies that $G > P;@_1;\lambda$.
-Let $G = P;@_1;\lambda;L$ and $H = P;L'$.
-If $L$ is disjoint from $Q$, then $(N,G) \sim_p (J,G) \sim_p (O_2,P;L') = (O_2;H)$ by cases 1 and 2 of $\sim_p$.
-In the other case, that $L$ isn't disjoint from $Q$, $L > Q$ because none of the positions $<= P;@_1;\lambda;Q$ in $N$ are related to any position in $I$ by $\sim_p$.
-As $(I,P;L) \sim_p (O_1,P;L')$ (with the redex at $P;Q$), for exactly the same reason $(N,P;@_1;\lambda;L) \sim_p (J,P;@_1;\lambda;L')$ (with the redex at $P;@_1;\lambda;Q$).
-Because $(N,P;@_1;\lambda;L) \sim_p (J,P;@_1;\lambda;L')$, $N|P;@_1;\lambda;L = J|P;@_1;\lambda;L'$, and $N|P;@_1;\lambda;L \neq $ the variable of $N|P;@_1$, therefore $J|P;@_1;\lambda;L'$ is also not the variable therefore $(J,P;@_1;\lambda;L') \sim_p (O_2, P;L') = (O_2, H)$ by case 2 of $\sim_p$.
-Combining these results, $(N,G) \sim_p (O_2;H)$ as desired.
+Let $I$ be $N$ reduced at $\alpha$, and $J$ be $N$ reduced at $\alpha;@_1;\lambda;\beta$, so that $N \to I \to O_1$ and $N \to J \to O_2$.
+All of the reduction positions are $\geq \alpha$, and $\zeta$ is not disjoint from all of them, therefore $\zeta$ is not disjoint from $\alpha$.
+Let $\iota$ be the position such that $(N,\zeta) \sim_p (I,\iota) \sim_p (O_1,\theta)$.
+The fact that $(N,\zeta) \sim_p (I,\iota)$ implies that $\zeta > \alpha;@_1;\lambda$.
+Let $\zeta = \alpha;@_1;\lambda;\kappa$ and $\theta = \alpha;\kappa'$.
+If $\kappa$ is disjoint from $\beta$, then $(N,\zeta) \sim_p (J,\zeta) \sim_p (O_2,\alpha;\kappa') = (O_2;\theta)$ by cases 1 and 2 of $\sim_p$.
+In the other case, that $\kappa$ isn't disjoint from $\beta$, $\kappa > \beta$ because none of the positions $<= \alpha;@_1;\lambda;\beta$ in $N$ are related to any position in $I$ by $\sim_p$.
+As $(I,\alpha;\kappa) \sim_p (O_1,\alpha;\kappa')$ (with the redex at $\alpha;\beta$), for exactly the same reason $(N,\alpha;@_1;\lambda;\kappa) \sim_p (J,\alpha;@_1;\lambda;\kappa')$ (with the redex at $\alpha;@_1;\lambda;\beta$).
+Because $(N,\alpha;@_1;\lambda;\kappa) \sim_p (J,\alpha;@_1;\lambda;\kappa')$, $N|\alpha;@_1;\lambda;\kappa = J|\alpha;@_1;\lambda;\kappa'$, and $N|\alpha;@_1;\lambda;\kappa \neq $ the variable of $N|\alpha;@_1$, therefore $J|\alpha;@_1;\lambda;\kappa'$ is also not the variable therefore $(J,\alpha;@_1;\lambda;\kappa') \sim_p (O_2, \alpha;\kappa') = (O_2, \theta)$ by case 2 of $\sim_p$.
+Combining these results, $(N,\zeta) \sim_p^* (O_2;\theta)$ as desired.
 \end{proof}
@@
 \begin{lemma} \label{lem:redexDestroyed}
-If $M \to N$, with the redex at position $P$, then no position in any term reachable from $N$ is related by $\sim^*$ to $(M,P)$.
+If $M \to N$, with the redex at position $\alpha$, then no position in any term reachable from $N$ is related by $\sim^*$ to $(M,\alpha)$.
 \end{lemma}
 \begin{proof}
-Suppose on the contrary that $(M,P) \sim^* (A,D)$, where $N \to^* A$, then by Lemma \ref{simpDoesn'tMix}, $(M,P) \sim_p^* (A,D)$.
-$M \neq A$ therefore $(M,P) \sim_p$ some position in $N$, but in all the cases of the definition of $\sim_p$, no position in the child term is related to the position of the redex.
+Suppose on the contrary that $(M,\alpha) \sim^* (A,\beta)$, where $N \to^* A$, then by Lemma \ref{simpDoesn'tMix}, $(M,\alpha) \sim_p^* (A,\beta)$.
+$M \neq A$ therefore $(M,\alpha) \sim_p$ some position in $N$, but in all the cases of the definition of $\sim_p$, no position in the child term is related to the position of the redex.
 \end{proof}
@@
 \begin{lemma}
-For any skeletons $M \to N$, with the redex at position $P$, and measurable set of samples $S \subset I^{L_s(N)}$, $\mu(S) = \mu(\{s \in I^{L_s(M)} \mid s \circ i(M \to N) \in S\})$, and furthermore, if $M | P = \tsample$, for any $S \subset I \times I^{L_s(N)}$, $\mu(S) = \mu(\{s \in I^{L_s(M)} \mid (s(M,P), s \circ i(M \to N)) \in S\})$.
+For any skeletons $M \to N$, with the redex at position $\alpha$, and measurable set of samples $S \subset I^{L_s(N)}$, $\mu(S) = \mu(\{s \in I^{L_s(M)} \mid s \circ i(M \to N) \in S\})$, and furthermore, if $M | \alpha = \tsample$, for any $S \subset I \times I^{L_s(N)}$, $\mu(S) = \mu(\{s \in I^{L_s(M)} \mid (s(M,\alpha), s \circ i(M \to N)) \in S\})$.
 \end{lemma}
@@
 
-In the case that $M|P = \tsample$, we can similarly consider only those sets $S$ of the form $x_\tsample \times \{s \in I^{L_s(N)} \mid \forall j : s(j) \in x_i\}$.
-The position $P$ in $M$ is not related to any position in $L_s(N)$, therefore $(M,P) \not \in K$.
-Again, we can calculate the measure $\mu(\{s \in I^{L_s(M)} \mid (s(M,P), s \circ i(M \to N)) \in S\}) = \mu(\{s \in I^{L_s(M)} \mid s(M,P) \in x_\tsample, \forall k : s(k) \in x_{i(M \to N)^{-1}(k)}\}) = \mu_I(x_\tsample) \prod_{k \in K} \mu_I(x_{i(M \to N)^{-1}(k)}) = \mu_I(x_\tsample) \prod_{j \in J} \mu_I(x_j) = \mu(S)$.
+In the case that $M|\alpha = \tsample$, we can similarly consider only those sets $S$ of the form $x_\tsample \times \{s \in I^{L_s(N)} \mid \forall j : s(j) \in x_i\}$.
+The position $\alpha$ in $M$ is not related to any position in $L_s(N)$, therefore $(M,\alpha) \not \in K$.
+Again, we can calculate the measure $\mu(\{s \in I^{L_s(M)} \mid (s(M,\alpha), s \circ i(M \to N)) \in S\}) = \mu(\{s \in I^{L_s(M)} \mid s(M,\alpha) \in x_\tsample, \forall k : s(k) \in x_{i(M \to N)^{-1}(k)}\}) = \mu_I(x_\tsample) \prod_{k \in K} \mu_I(x_{i(M \to N)^{-1}(k)}) = \mu_I(x_\tsample) \prod_{j \in J} \mu_I(x_j) = \mu(S)$.
 \end{proof}

### 2020-10-13 probabilisticProgrammingMartingales.tex
Prove Theorem 2, thm:CbvIsTerminatingest.
Because of the restrictions on when sample reductions can occur,
this isn't actually following the same logic as the standard theorem
that lazy evaluation terminates if anything does, but it was pretty
simple anyway.

@@
 \paragraph{}
-\begin{lemma}
+\begin{lemma} \label{churchRosser}
 The relation $\Rightarrow$ is Church-Rosser.
@@
 \begin{proof}
+  Suppose $M$ terminates with the reduction strategy $f$ and samples $s$.
+Let $(M,s) = (M_0,s_0) \Rightarrow_f \dots \Rightarrow_f (M_n,s_n)$, where $M_n$ is a value.
+By Lemma \ref{canonicalCousins}, the reduction sequence $(Sk(M_i))_i$ is related by $\sim_c^*$ to some skeletal reduction sequence $X$ which is in call-by-value order.
+For each skeletal reduction sequence $X_j$ in the sequence $(Sk(M_i))_i \sim_c \dots \sim_c X$, it is possible to define a reduction sequence of terms such that $(M,s) = (N_{j,0},s_{j,0}) \Rightarrow \dots \Rightarrow (N_{j,n_j},s_{j,n_j})$ and $Sk(N_{j,k}) = X_{j,k}$, by at each step applying the reduction at the same place as in $X_j$, essentially just filling in the reals in the skeletal reduction sequence to make it a reduction sequence of terms (it will be shown shortly that the correct branch is taken in each $\textsf{if}$-reduction).
+Assume for induction on $j$ that $(N_{j,n_j},s_{j,n_j}) = (M_n,s_n)$.
+To prove that this holds true for $j+1$ as well, it suffices to show that the corresponding term and samples immediately after the reductions involved in this $\sim_c$ step match, as from that point on, the $\Rightarrow$-sequences will match, as they start from the same point and have all the same reduction positions.
+Letting the term and remaining samples immediately before the reductions involved in this $\sim_c$ step be $(N,t)$ (which is the same in both $N_j$ and $N_{j+1}$ for the same reason again, they start at the same point and the sequences of reduction positions up to this point are the same), and the terms and samples after these reductions be $(O_1,t_1)$ and $(O_2,t_2)$, so that these match the $N$, $O_1$ and $O_2$ in the definition of $\sim_c$, $t_1 = t_2$ because both of them are, by Lemma \ref{lem:sim-M-N}, equal to the restriction of $t$ to the subset of $Rch(N)$ that starts with the reduction sequences to $O_1$ and $O_2$ respectively, and the potential positions in those subsets of $Rch(N)$ are identified with each other by $\sim$ because of the same case of $\sim_c$ that means $N_j$ and $N_{j+1}$ are related.
+If none of the relevant reductions are $\tsample$-reductions, $O_1 = O_2$ trivially.
+If there are samples taken (which must be the arbitrary reduction at $\beta$), it is also required that the same samples are taken in both branches.
+Similarly to Lemma \ref{churchRosser}, the positions of these $\tsample$-reductions in each branch are related by $\sim_p^*$, therefore they correspond to the same sample in $t$, therefore $O_1 = O_2$.
+The fact that the terms are equal also implies that the same branch of an $\textsf{if}$ statements will be taken.
+
+  \paragraph{}
+  There is therefore a reduction sequence $(M,s) = (N_0,s_0') \Rightarrow \dots \Rightarrow (N_m, s_m')$ which ends in a value ($N_m = M_n$) and is in CBV order.
+It doesn't immediately follow that this is the $\Rightarrow_{\cbv}$ reduction sequence starting from $(M,s)$ because the fact that $(N_i)_i$ is in CBV order only means that the reductions that do take place in this reduction sequence are in the correct order, not that the reduction sequence finishes once it reaches a value, or that there are no other reductions that would come earlier in CBV order that don't occur at all in $(N_i)_i$.
+Suppose that at some point $j$ in the reduction sequence, the reduction that occurs is not the next reduction in CBV order, but something else, and that the reduction that should be next is at position $\alpha = \cbv(N_i)$, and the position of the actual reduction $N_j \to N_{j+1}$ is $\beta$.
+Suppose further that $N_j$ is not a value.
+Either $\alpha = \cdot$, the root position, or the reduction at the root position is dependent in some way on the reduction at $\alpha$ (e.g.
+$\alpha = \textsf{if}_1$, or $\alpha = \underline{+};@_1$).
+In any case, all the reductions in the rest of the reduction sequence are after $\alpha$ in CBV order, therefore none of them are $\leq \alpha$, and there continues to be an unreduced redex at $\alpha$, therefore the reduction at the root position never occurs, therefore $N_m$ is not a value, which is a contradiction, therefore at the point (if any) where $(N_i)_i$ departs from the $\Rightarrow_{\cbv}$ reduction sequence starting at $(M,s)$, it has already reached a value, therefore $(M,s) \Rightarrow_{\cbv} \dots$ does reach a value eventually, i.e.~$M$ terminates with the reduction strategy $\cbv$ and samples $s$.
 \end{proof}

### 2020-10-27 probabilisticProgrammingMartingales.tex
Prove Lemma 1 (downAcrossUp).
The proof turned out to be more complicated than expected, because
it's non-trivial to show that the sequence of rearrangements required
actually terminates. It was possible to adapt a proof of the Church-
Rosser theorem to prove termination.

@@
 \begin{proof}
-The $\sim$ relation can be split into $\sim_\downarrow \cup \sim_\uparrow$, where $(A,\alpha) \sim_\downarrow (B,\beta)$ if either $A \to B$ and $(A,\alpha) \sim_p (B,\beta)$ or $(A,\alpha) \sim_c (B,\beta)$, and similarly $(A,\alpha) \sim_\uparrow (B,\beta)$ if either $B \to A$ and $(A,\alpha) \sim_p (B,\beta)$ or $(A,\alpha) \sim_c (B,\beta)$.
-The sequence of $\sim$ relations linking $N_1$ to $N_2$ can then be expressed as an alternating sequence of $\sim_\downarrow^*$ and $\sim_\uparrow^*$s, starting with $\sim_\downarrow^*$ and ending with $\sim_\uparrow^*$, and the proof proceeds by induction on the number of these alternations.
-
-In the base case, that $(N_1,\alpha_1) \sim_\downarrow^* X \sim_\uparrow^* (N_2,\alpha_2)$, \dots
-
-\akr{Actually proving that this process terminates will be a pain although I'm confident it will work, but the basic idea is that the sequence of $\sim$ steps can be altered by repeatedly replacing the subsequence $\sim_c, \sim_{p,\downarrow}$ with $\sim_{p,\downarrow}, \sim_c$, the subsequence $\sim_{p,\uparrow}, \sim_c$ with $\sim_c, \sim_{p,\uparrow}$, and the subsequence $\sim_{p,\uparrow}, \sim_{p,\downarrow}$ with $\sim_{p,\downarrow}^*, \sim_c, \sim_{p,\uparrow}^*$, until everything is in the correct order.}
+  The $\sim_p$ relation can be split into $\sim_\downarrow \cup \sim_\uparrow$, where $(A,\alpha) \sim_\downarrow (B,\beta)$ if $A \to B$ and $(A,\alpha) \sim_p (B,\beta)$, and similarly $(A,\alpha) \sim_\uparrow (B,\beta)$ if $B \to A$ and $(A,\alpha) \sim_p (B,\beta)$.
+At each stage of this proof, it will be assumed that the $\sim_c$ steps and the $\sim_p$ steps in the sequence from $(N_1,\alpha_1)$ to $(N_2,\alpha_2)$ are rearranged such that there is never a $\sim_c$ immediately before a $\sim_\downarrow$, and there is never a $\sim_c$ immediately after a $\sim_\uparrow$.
+This rearrangement is always possible, because if there is some subsequence $(A,\alpha) \sim_c (B,\alpha) \sim_\downarrow (C,\beta)$, then there is an alternate path $(A,\alpha) \sim_\downarrow (B',\beta) \sim_c (C,\beta)$, where the reduction and the $\sim_\downarrow$ from $A$ to $B'$ are the same as those from $B$ to $C$, and the $\sim_c$ step is the same, but with the reduction sequences $O_1 \to^* O_1'$ and $O_2 \to^* O_2'$ in the definition of $\sim_c$ extended by the reduction $B \to C$.
+With these rearrangements assumed, it is sufficient to prove that the $\sim_p$ steps can be rearranged so that all of the $\sim_\downarrow$ steps come before all of the $\sim_\uparrow$ steps (possibly introducing some $\sim_c$ steps in the process).
+
+  The rearrangement to put all of the $\sim_\downarrow$s before all of the $\sim_\uparrow$s proceeds by induction on the number of $\sim_\downarrow$s that occur after the first occurrence of $\sim_\uparrow$.
+If it's 0, all of the $\sim_p$s are already in the correct order, and we're done.
+Otherwise, take the subsequence from the first $\sim_\uparrow$ to the first $\sim_\downarrow$ after that.
+This must be rearranged to some $\sim_\downarrow$s followed by some $\sim_\uparrow$s, then once that's done, the number of $\sim_\downarrow$s after the first $\sim_\uparrow$ in the overall sequence will have decreased by 1.
+
+  Let this subsequence be $A \sim_\uparrow^n B \sim_\downarrow C$.
+Suppose for induction that $A \sim_\uparrow^k (D,\delta) \sim_\downarrow^* (E,\epsilon) \sim_\uparrow^* C$ for some $D, \delta, E, \epsilon$, and $0 \leq k \leq n$, where in the reduction sequence $D \to^* E$, for any pair of reductions at positions which aren't disjoint, the reduction at the innermost (greater) position occurs first.
+Reduction sequences with this property will be called ``parallel'', as it is analagous to parallel reduction as in \akr{TODO: find this reference, for the proof of the Church-Rosser theorem by Tait and Martin-L\"of}.
+This induction is in reverse, with $k$ decreasing from $n$ to $0$.
+If $k = n$, simply set $(D,\delta) = B$ and $(E,\epsilon) = C$.
+If $k = 0$, then $A$ is related to $B$ by a sequence of $\sim_\downarrow$s then $\sim_\uparrow$s, as desired.
+In the intermediate steps, $k$ must be decreased by one, so a subsequence of one $\sim_\uparrow$ followed by a parallel sequence of $\sim_\downarrow$s must be replaced by a parallel sequence of $\sim_\downarrow$s followed by some $\sim_\uparrow$s.
+
+  For any individual $\sim_\uparrow$ then $\sim_\downarrow$ pair, let $(A,\alpha) \sim_\uparrow (B,\beta) \sim_\downarrow (C,\gamma)$.
+The skeleton $B$ reduces to both $A$ and $C$.
+If these are the same reduction, then $(A,\alpha) = (C,\gamma)$ and this subsequence may be removed.
+If they are different, then the way they overlap corresponds to one of the cases in the definition of $\sim_c$, either case a if the reduction positions are disjoint, or one of the cases b--f if one of the reduction positions is inside the other.
+In any case, there is a reduction sequence from each of $A$ and $C$ to some common skeleton $D$ ($O_1$ and $O_2$ in the definition of $\sim_c$).
+For the same reason that $(B,\beta) \sim_\downarrow (C,\gamma)$, there is some $\delta$ such that $(A,\alpha) \sim_\downarrow^* (D,\delta)$, and similarly for the same reason that $(B,\beta) \sim_\downarrow (A,\alpha)$, for the same $\delta$, $(C,\gamma) \sim_\downarrow^* (D,\delta)$.
+There are a lot of cases to check for this statement, but all of them are rather simple, and similar to each other.
+There is therefore an alternative $\sim^*$ sequence from $(A,\alpha)$ to $(C,\gamma)$, of the form $(A,\alpha) \sim_\downarrow^* (D_1,\delta) \sim_c (D_2,\delta) \sim_\uparrow^* (C,\gamma)$, where $D_1$ and $D_2$ are the alternative reduction sequences leading to the same skeleton $D$.
+There are only multiple $\sim_\downarrow$ steps in the result if the position of the reduction $B \to C$ is inside the position of the reduction $B \to A$, and similarly, there are only multiple $\sim_\uparrow$ steps in the result if the position of the reduction $B \to A$ is inside the position of the reduction $B \to C$.
+If there are multiple $\sim_\downarrow$ steps, they are not necessarily parallel, but in the only case where they aren't (case f of $sim_c$), the subsequence of $\sim_\downarrow$ steps with the reductions at $\alpha;\lambda;@_1;\beta$ then $\alpha;\lambda;@_1;\gamma_i;\tY;\lambda;\beta$ for each $i$ where $\gamma_i \geq \beta$ (in the notation of the definition of $\sim_c$) can be replaced by $\sim_\downarrow$s with the reductions at $\alpha;\lambda;@_1;\gamma'_i;\tY;\lambda;\beta$ then $\alpha;\lambda;@_1;\beta$, then another $\sim_c$, where $(\gamma'_i)$ is the list of positions in $A$ (\emph{before} the reduction at $\beta$ to $A'$, unlike $(\gamma_i)$) where $A | \gamma'_i = x$ and $\gamma'_i > \beta$.
+
+  The effect of this rearrangement is (ignoring the $\sim_c$s) to swap the order of the $\sim_\uparrow$ and the $\sim_\downarrow$, except that if the positions of one of these reductions is inside the other, that inner $\sim_p$ may be duplicated.
+If the positions are disjoint, both of them are unchanged (corresponding to case a of $\sim_c$), and if one is inside the other, then the outer position is unchanged and the other resultant positions, although they aren't equal to the original inner position, are still inside the outer position.
+Taking the sub-sequence of one $\sim_\uparrow$ followed by a parallel sequence of $\sim_\downarrow$s, this rearrangement can be repeatedly applied to move the $\sim_\uparrow$ further along the sequence.
+If at some point a $\sim_\uparrow$ and $\sim_\downarrow$ match (by relating the same (skeleton, position) pairs in opposite directions), they may be removed and the process may stop early.
+The $\sim_\uparrow$ may be duplicated if it passes a $\sim_\downarrow$ whose position is outside of the $\sim_\uparrow$'s position, but by the assumption that the sequence of $\sim_\downarrow$s is initially parallel, no position inside of this occurs later in the sequence, therefore all of the resultant $\sim_\uparrow$s pass the remaining $\sim_\downarrow$s without changing or duplicating them.
+The $\sim_\uparrow$s may be further duplicated, but the number of steps left in this process is bounded by the product, for all of the remaining $\sim_\downarrow$s, of $2 + \text{the number of occurrences of the variable relevant to the reduction}$, because for each $\sim_\downarrow$, that is more than the number of duplicates it could produce for any $\sim_\uparrow$ that passes it.
+It is also possible (earlier) for some of the $\sim_\downarrow$s to be duplicated, but the only case where this process would not terminate, that both the $\sim_\downarrow$s and the $\sim_\uparrow$s continue being duplicated forever so that the number of switches left to do never decreases, is prevented by the parallelness condition, as all of the duplications of $\sim_\downarrow$s occur before all of the duplications of $\sim_\uparrow$s.
+
+  It still remains to be shown that the sequence of $\sim_\downarrow$s left at the end of this can be made parallel.
+The changes that may have occurred since the previous version of this sequence (which was known to be parrallel already), are that some of the $\sim_\downarrow$s may have been duplicated by passing the $\sim_\uparrow$, and if that $\sim_\uparrow$ matches one of the $\sim_\downarrow$s, that $\sim_\downarrow$ is removed.
+The position of the $\sim_\uparrow$'s reduction is the same for all of these duplications, because it is not changed until later in the sequence, when the $\sim_\downarrow$s with positions outside of that may occur.
+The resultant positions after a duplication are in the same position relative to each other, and therefore the order of the remaining $\sim_\downarrow$s is automatically parallel, except that (letting the position of the $\sim_\uparrow$'s reduction be $\alpha$) if there are $\sim_\downarrow$s with reductions at positions $\alpha;@_1;\lambda;\beta$ and $\alpha;@_2;\gamma$, or $\alpha;\tY;\lambda;\beta$ and $\alpha;\tY;\lambda;\gamma$, some of the resultant $\sim_\downarrow$s' positions may be inside each other where they were originally not.
+Similarly to the case where a $\tY$ reduction and a reduction at a position inside it are swapped though, the order can be fixed by swapping some of the $\sim_\downarrow$ steps with each other.
+In the first case, the reduction at position $\alpha;@_1;\lambda;\beta$ ends up at $\alpha;\beta$, and the reduction at position $\alpha;@_2;\gamma$ ends up at the positions $\alpha;\delta_i;\gamma$, where $(\delta_i)$ is the positions of the variable relevant to the $\sim_\uparrow$'s reduction inside its lambda.
+The positions of the variable in the lambda may be different before the reduction at $\beta$, but the positions of the redexes corresponding to the original redex at $\alpha;@_2;\gamma$ are changed similarly.
+Rather than reducing at $\alpha;\beta$ then at $\alpha;\delta_i;\gamma$, it is therefore possible to reach the same point by reducing at $\alpha;\delta'_i;\gamma$ then at $\alpha;\beta$, where $(\delta'_i)$ is the positions of the relevant variable within the lambda before the reduction at $\alpha;@_1;\lambda;\beta$ (and again, a $\sim_c$ of some sort must also be introduced).
+The other case, that the $\sim_\uparrow$'s reduction is a $\tY$-reduction, is similar.
+Each reduction which is duplicated has one image at $\alpha;\lambda;@_1;$its original relative position, and one for each occurrence of the variable.
+Those corresponding to variable occurrences may need to be moved to before those at $\lambda;@_1$.
+
+  In summary, to reach the desired order of $\sim_p$ steps, each $\sim_\downarrow$ may have to be moved past some $\sim_\uparrow$ steps, and it may get duplicated in the process, but the resultant sequence of $\sim_\downarrow$s can all be put in parallel order, and it has been shown that moving a $\sim_\uparrow$ past a sequence of $\sim_\downarrow$s is always possible if they're in parallel order, therefore the overall process terminates and reaches a point where all of the $\sim_\downarrow$s precede all of the $\sim_\uparrow$s, and all of the $\sim_c$s are in between them.
+All of these rearrangements leave the end points of the sequence unchanged, therefore this is an alternative sequence of $\sim$ steps between $(N_1,\alpha_1)$ and $(N_2,\alpha_2)$ of the form $(N_1,\alpha_1) \sim_\downarrow^* (N_1', \alpha') \sim_c^* (N_2', \alpha') \sim_\uparrow^* (N_2,\alpha_2)$.
 \end{proof}

### 2020-10-30 probabilisticProgrammingMartingales.tex
Fix the proof of Church-Rosserness
I hadn't noticed previously that weak confluence does not in general
imply confluence, but the same argument used in Lemma \ref{downAcrossUp}
applies here too, to prove that this terminates.

@@
 \begin{proof}
-Suppose that $(M,s) \Rightarrow (M_1,s_1)$ and also $(M,s) \Rightarrow (M_2,s_2)$, then it is required to prove that there is some $(M',s')$ such that both $(M_1,s_1) \Rightarrow^* (M',s')$ and $(M_2,s_2) \Rightarrow^* (M',s')$.
+Suppose that $(M,s) \Rightarrow^* (M_1,s_1)$ and also $(M,s) \Rightarrow^* (M_2,s_2)$, then it is required to prove that there is some $(M',s')$ such that both $(M_1,s_1) \Rightarrow^* (M',s')$ and $(M_2,s_2) \Rightarrow^* (M',s')$.
+First consider the special case where $(M,s) \Rightarrow (M_1,s_1)$ and $(M,s) \Rightarrow (M_2,s_2)$, with only a single step in each case.
 Let the positions of the redexes in $M \to M_1$ and $M \to M_2$ be $\alpha_1$ and $\alpha_2$ respectively.
@@
 The injection $i(M \to M_1) \circ i(M_1 \to M')$ consists of prepending $M \to M_1 \to M'$ to each reduction sequence, but by case a of $\sim_c$, this is equivalent to prepending $M \to M_2 \to M'$, which is $i(M \to M_2) \circ i(M_2 \to M')$.
-In the case that the reduction $M \to M_2$ isn't a $\tsample$-reduction, this is enough to establish that $(M_1, s_1) \Rightarrow (M', s')$, and vice-versa. 
+In the case that the reduction $M \to M_2$ isn't a $\tsample$-reduction, this is enough to establish that $(M_1, s_1) \Rightarrow (M', s')$ (and similarly if the redex of $M \to M_1$ isn't $\tsample$, $(M_2,s_2) \Rightarrow (M',s')$).
 If it is $\tsample$ though, in order for it to be the case that $(M_1, s_1) \Rightarrow (M', s')$, it is additionally necessary that $X_2$, the result of the reduction at $\alpha_2$, be $s_1(M_1, \alpha_2)$, which follows from the fact that $(M, \alpha_2) \sim (M_1, \alpha_2)$ by case 1 of $\sim_p$.
@@
 The values that the $\tsample$s take in these cases match because $(M,\alpha) \sim_p (M_1, \alpha;\beta)$ by cases 3 and 2 of $\sim_p$ respectively.
+
+\paragraph{}
+In the case that $(M,s) \Rightarrow^* (M_1,s_1)$ or $(M,s) \Rightarrow^* (M_2,s_2)$ by multiple steps, it is possible to repeatedly replace a pair of the form $A \Leftarrow B \Rightarrow C$ by $A \Rightarrow^* D \Leftarrow^* C$ by the construction above, but it is not immediate that this process terminates, because each $\Rightarrow$ or $\Leftarrow$ may be replaced by multiple, so the sequence of $\Rightarrow$s and $\Leftarrow$s from $(M_1,s_1)$ to $(M_2,s_2)$ may get longer at some steps.
+However, termination can be proved by noting that the structure of this process is identical to the process of swapping $\sim_\uparrow$ and $\sim_\downarrow$ steps in Lemma \ref{downAcrossUp}.
+To be more precise, in this case there is a sequence of $\Leftarrow$ and $\Rightarrow$ steps, each of which has an associated reduction position and initial and final terms, and if a $\Leftarrow$ immediately precedes a $\Rightarrow$, they may be swapped to produce some number of $\Rightarrow$s, followed by some number of $\Leftarrow$s.
+In the case of Lemma \ref{downAcrossUp}, there is a sequence of $\sim_\uparrow$ and $\sim_\downarrow$ steps, each of which has an associated reduction position and initial and final skeletons, and if a $\sim_\uparrow$ immediately precedes a $\sim_\downarrow$, they may be swapped to produce some number of $\sim_\downarrow$s followed by some number of $\sim_\uparrow$s.
+In both cases, the number and reduction positions of the resultant steps are determined by the case of $\sim_c$ that matches the way the initial reduction positions overlap, and the initial skeleton (or the skeleton of the initial term).
+The same argument that the process in Lemma \ref{downAcrossUp} terminated is therefore applicable here too.
+At every stage, the sequence of reductions that results from one of the initial $\Rightarrow$s is a parallel sequence, and swapping a parallel sequence of $\Rightarrow$s with a $\Leftarrow$ always terminates, therefore each $\Rightarrow$ in turn can be moved past all of the $\Leftarrow$s, and the process as a whole will terminate in a state where all of the $\Rightarrow$s precede all of the $\Leftarrow$s, i.e.~a pair of reduction sequences $(M_1,s_1) \Rightarrow^* (M',s') \Leftarrow^* (M_2,s_2)$.
 \end{proof}

### 2020-11-24 probabilisticProgrammingMartingales.tex
Minor fixes

@@
 The sequence $(X_i)_i$ therefore proceeds to swap all the remaining $\beta_i$s then $\gamma$, in order, below $\alpha$, resulting in $X_i$ for some $i$ having as its relevant sub-reduction-sequence $\alpha, \gamma, \beta_k, \dots, \beta_1$.
-The other sequence $(Y_j)_j$ starts by swapping $\gamma$ then $beta$ below $\alpha$, then swapping $\gamma$ and $\beta$.
+The other sequence $(Y_j)_j$ starts by swapping $\gamma$ then $\beta$ below $\alpha$, then swapping $\gamma$ and $\beta$.
 Because $\alpha$ is disjoint from $\gamma$, the reduction there does not affect the sub-skeleton at position $\gamma$, therefore this last $\sim_c$ step proceeds identically to how it did in $Y \sim_c X$, resulting in $\gamma, \beta_k, \dots, \beta_i$.
 Overall, this results in $Y_3$ having the relevant sub-reduction-sequence $\alpha, \gamma, \beta_k, \dots, \beta_1$, therefore $Y_3 = X_i$ for the aforementioned value of $i$.
-\item $Y \sim_c X$ forwards, $\alpha \leq \gamma$ and $\alpha$ (and therefore also $\gamma$) is disjoint from $\beta$: In this case, $(X_i)_i$ proceeds by swapping $\beta_1 = \beta$ below alpha by case a, then swapping $\gamma$ below $\alpha$ by one of the other cases, and $(Y_j)_j$ proceeds by swapping $\gamma$ and $\alpha$, then swapping $\beta$ below $\alpha$ then all of the images of $\gamma$ (which are $\geq \alpha$ and therefore left of $\beta$).
+\item $Y \sim_c X$ forwards, $\alpha \leq \gamma$ and $\alpha$ (and therefore also $\gamma$) is disjoint from $\beta$: In this case, $(X_i)_i$ proceeds by swapping $\beta_1 = \beta$ below $\alpha$ by case a, then swapping $\gamma$ below $\alpha$ by one of the other cases, and $(Y_j)_j$ proceeds by swapping $\gamma$ and $\alpha$, then swapping $\beta$ below $\alpha$ then all of the images of $\gamma$ (which are $\geq \alpha$ and therefore left of $\beta$).
 In both cases, the subterm at $\alpha$ when $\alpha$ and $\gamma$ are swapped is unaffected by the reduction at $\beta$, therefore it produces the same result in both cases, therefore the overall sequence in both cases is $\alpha$ then the images of $\gamma$ then $\beta$.
@@
 At this point, the next few reductions before $\alpha;\gamma'$ will in general be images of $\beta$ at positions inside $\alpha;\gamma'$.
-The subskeleton at position $\alpha;\gamma'$ before these reductions is $(A|\gamma')[B/x]$, then by the images of $\beta$ this reduces to $(A|\gamma')[B'/x]$, then it reduces at its root position ($\alpha;gamma'$ in the overall skeleton) to $(A'|\gamma')[B'/x]$.
+The subskeleton at position $\alpha;\gamma'$ before these reductions is $(A|\gamma')[B/x]$, then by the images of $\beta$ this reduces to $(A|\gamma')[B'/x]$, then it reduces at its root position ($\alpha;\gamma'$ in the overall skeleton) to $(A'|\gamma')[B'/x]$.
 For each position of an $x$ in $(A|\gamma')$, there are 0 or more corresponding positions of $x$ in $(A'|\gamma')$, depending on what type of reduction $A \to A'$ is.
@@
 Even if it's case d or f, all of the images of $\gamma'(\delta^B_i)$ are disjoint from each other (and from the positions of the other reductions after $\beta'_0$) because the redex, $C$, doesn't contain any instances of the variable involved in the reduction $B \to B'$.
-Expanding the definitions, this is swapping $\alpha;\lambda;@_1;\delta^B_i;\tY;\lambda;gamma'$ with $\alpha;\lambda;@_1;\beta'$ at a point in the reduction sequence where the subskeleton at $\alpha;\lambda;@_1;\beta'$ is $B$ with a mixture of $\tY \lambda x.
+Expanding the definitions, this is swapping $\alpha;\lambda;@_1;\delta^B_i;\tY;\lambda;\gamma'$ with $\alpha;\lambda;@_1;\beta'$ at a point in the reduction sequence where the subskeleton at $\alpha;\lambda;@_1;\beta'$ is $B$ with a mixture of $\tY \lambda x.
 A$ and $\tY \lambda x.

### 2020-11-24 ranking-supermartingales.tex
Remark / question: Claim 1 seems too general for the conjecture
Minor fixes

@@
 M_n(s) &:= \pi_0 (\red^n(M, s))\\
-\ndraw{n}{s} &:= l \textrm{ where $\pi_1 (\red^n(M, s)) = \underbrace{\pi_t( \cdots (\pi_t}_l(s))$}
+\ndraw{n}{s} &:= |\{k < n \mid \exists E: M_k(s) = E[\tsample]\}|,
 \end{align*}
-and the filtration $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
+ so that $\pi_1 (\red^n(M, s)) = \underbrace{\pi_t( \cdots (\pi_t}_{\ndraw n s}(s))$, and the filtration $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
 The $\calF_n$-measurability of $M_n$ (and hence of $\#\mathrm{draw}_n$) follows from \citep{DBLP:conf/icfp/BorgstromLGS16};
 $\#\mathrm{draw}_n$ is a stopping time (bounded by $n$).
+\akr{$\#\mathrm{draw}_n$ is not a stopping time.
+It doesn't look like you actually use this claim anyway, so it should just be fine to remove, but did you mean something different?}
 Take $s \in A \in \calF_n$ with $\ndraw{n}{s} = l$.
@@
 Define random variables on $(S, \calF, \mu)$:
-\begin{align*}
-T_0(s) & := 0 \\
-T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value or of form } E[\tY \lambda x. N] \}\\
-X_n(s) & := f(M_n(s)) \\
-Y_n(s) & := X_{T_n}(s) \\
-T(s) &:= \min \set{n \mid Y_n(s) = 0}
-\end{align*}
+\begin{align}
+T_0(s) & := 0 \nonumber \\
+T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value or of form } E[\tY \lambda x. N] \} \nonumber \\
+X_n(s) & := f(M_n(s)) \nonumber \\
+Y_n(s) & := X_{T_n}(s) \nonumber \\
+T(s) &:= \min \set{n \mid Y_n(s) = 0} \label{eq:y stopping time}
+\end{align}
 \lo{Define the \emph{$\tY$-runtime of $M$} to be the random variable $T^{\tY}_M$: 

### 2020-12-01 probabilisticProgrammingMartingales.tex
Rearrange sections, various minor corrections.

@@
 
-\section{Alternative Semantics}
-\akr{This version of the semantics is much more convenient for allowing multiple different reduction orders.
-I don't currently have a proof of the equivalence of this with the more usual semantics.
-I expect that that would be somewhat complicated, but not terribly difficult.
-I don't know whether a semantics like this has already been defined elsewhere.
-I just thought I'd write it out so you can see it and in case I use it later.
-If I do end up actually using it I'll write it up nicer, but hopefully this is at least enough for you to understand what I mean.
-
-Having thought a bit more about this, I realise it may not be quite right. I'll see if I can fix it later.}
-\lo{I'll read this after you fix it ;-)}
-
-When proving almost sure termination in this way \akr{I guess I should move this section somewhere later.}, it is necessary to consider which terms a given term may reduce to.
+\section{Supermartingales}
+One approach to proving that a term terminates almost surely is to find some variant that is bounded below and, on average, decreases sufficiently quickly that it must eventually reach 0, similarly to the approach taken in \citep{DBLP:journals/pacmpl/McIverMKK18} for imperative programs.
+
+These variants are defined as functions from reachable terms (i.e.~possible states of the program's execution) to real numbers.
+Specifically, let the set of reachable terms from a given closed starting term $M$ be $Rch(M) = \{N \in \Lambda \mid M \to^* N \}$, with the $\sigma$-algebra induced as a subset of $\Lambda$.
+
+\begin{definition}\rm
+A \emph{ranking function on $M$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
+\begin{enumerate}
+    \item $f(E[\tY \lambda x. N]) \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z) \text{ where $z$ is not free in $N$}$
+    \item $f(E[\tsample]) \geq \int_I f(E[\underline{x}]) \, \Leb(\mathrm{d}x)$
+
+    \item $f(E[R]) \geq f(E[R'])$ for any other redex $R$, where $R \to R'$.
+\end{enumerate}
+We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that $f(N) = 0$ iff $N$ is a value, and for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.
+
+Any closed term for which a ranking (respectively, strict ranking) function exists is called \emph{rankable} (respectively, \emph{strictly rankable}).
+It will be demonstrated later that for any rankable term $M$, if $(M_n)_n$ is the reduction sequence starting from $M$, then $(f(M_n))_n$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales.
+\end{definition}
+
+\input{ranking-supermartingales}
+
+\section{Constructing Ranking Functions}
+Although rankability implies almost-sure termination, the converse does not hold in general. For example,
+\begin{equation}
+\tif{{-}\tsample < 0}{\underline{0}}{(\tY \lambda x. x) \; \underline 0}
+\label{ex:0 probability reachable}
+\end{equation}
+terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x. x) \underline 0$ is reachable, although that has probability 0. 
+Not only is this counterexample AST, it's positively almost surely terminating i.e.~the expected time to termination is finite.
+
+A ranking function can be constructed under the stronger assumptions that, for every $N$ reachable from $M$, the expected number of $\tY$-reduction steps from $N$ to a value is finite.
+In particular, the expected number of $\tY$-reduction steps from each reachable term is a ranking function.
+Note that a finite number of expected $\tY$-reduction steps does not necessarily imply a finite number of expected total reduction steps.
+
+\begin{example}
+\label{ex:tY finite does not imply t finite}
+The term
+\[
+M = \big(\tY \lambda f n. \tif{\tsample - 0.5 < 0}{n \, \underline 0}{f (\lambda x. n (n x))}\big) (\lambda x. x+1)
+\] 
+terminates with only 2 $\tY$-reductions on average i.e.~$\expect{T_M^\tY} = 2$, but applies the increment function $2^n$ times with probability $2^{-1-n}$ for $n \geq 0$, which diverges i.e.~$\expect{T_M} = \infty$.
+\end{example}
+
+\begin{theorem} \label{thm:minimal}
+Given a closed term $M$, the function $f:Rch(M) \to \mathbb R$ given by $f(N) = \mathbb E [\text{the number of }\tY\text{-reduction steps from }N\text{ to a value}]$, if it exists, is the least of all possible ranking functions of $M$.
+\end{theorem}
+\begin{proof}
+Let $f$ be the candidate least ranking function defined above, and suppose $g$ is another ranking function such that $f(N) > g(N)$ for some $N \in Rch(M)$.
+The restrictions of $f$ and $g$ to $Rch(N)$ have the same properties assumed of $f$ and $g$, so assume w.l.o.g.~that $N=M$.
+The difference $g - f$ is then a supermartingale (with the same setup as in the proof of \Cref{thm:rankable gives supermartingale}
+%Theorem \ref{rankable implies ast}) 
+therefore $\forall n.\ \mathbb E[g(M_n)] \leq \mathbb E[f(M_n)] + g(M)-f(M)$.
+
+$\mathbb E[f(M_n)] = \sum_{k=n}^\infty \mathbb P[M_n = E[\tY N] \text{ for some }E, N] \to 0 \text{ as } n \to \infty$ therefore as $g(M) - f(M) < 0$, eventually $\mathbb E[g(M_n)] < 0$, which is impossible, therefore $g \geq f$ as required.
+
+In order for $f$ to be the least ranking function of $M$, it also has to actually be a ranking function itself.
+Each of the conditions on a ranking function is easily verified from the definition of $f$.
+\end{proof}
+
+\subsection{Partial ranking functions}
+Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost sure termination, so it isn't useful for proving it.
+Take, for example, the term
+\begin{align*}
+&(\tY \lambda f, n: \\
+&\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
+&) \underline{0}
+\end{align*}
+which generates a geometric distribution.
+    Despite its simplicity, its $Rch$ contains all the terms
+\begin{enumerate}
+    \item $(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{0}$
+    \item $(\lambda z.(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) z) \underline{i}$
+    \item $(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{i}$
+    \item $(\lambda n: \tif{\tsample - 0.5 < 0}{n}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{n}{f (m+1)}) (m+1)}) \underline{i}$
+    \item $\tif{\tsample - 0.5 < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
+    \item $\tif{\underline r - 0.5 < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
+    \item $\tif{\underline{r - 0.5} < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
+    \item $\underline{i}$
+    \item $(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)$
+    \item $(\lambda z.(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) z) (\underline{i} + 1)$.
+\end{enumerate}
+\akr{TODO: make this into a diagram with arrows.}
+
+\lo{Here is a more readable version.}
+
+Writing $\Theta = \lambda f, n : \tif{\tsample - 0.5 < 0}{n}{f (n+1)}$, we have
+\begin{enumerate}
+    \item $(\tY \, \Theta) \; \underline{0}$
+
+    \item $\big(\lambda z.\Theta \; (\tY \, \Theta) \; z \big) \; \underline{i}$
+
+    \item $\Theta \; (\tY \, \Theta) \; \underline{i}$
+
+    \item \changed[akr]{$(\lambda n: \tif{\tsample - 0.5 < 0}{n}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{n}{f (m+1)}) (m+1)}) \; \underline{i}$}
+
+    \lo{I think the preceding should be: $(\lambda n: \tif{\tsample - 0.5 < 0}{n}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (n+1)}) \; \underline{i}$, and so, more succinctly:
+    \[
+    \big(
+    \lambda n: 
+    \tif{\tsample - 0.5 < 0}{n}{(\tY \, \Theta) \; (n+1)} 
+    \big) 
+    \; \underline{i}
+    \]}
+
+    \item $\tif{\tsample - 0.5 < 0}{\underline{i}}{
+    (\tY \, \Theta) \, (\underline{i}+1)}$
+
+    \item $\tif{\underline r - 0.5 < 0}{\underline{i}}{(\tY \, \Theta) (\underline{i}+1)}$
+
+    \item $\tif{\underline{r - 0.5} < 0}{\underline{i}}{(\tY \, \Theta) (\underline{i}+1)}$
+
+    \item $\underline{i}$
+
+    \item $(\tY \, \Theta) \; (\underline{i}+1)$
+
+    \item $\big(\lambda z. \Theta \; (\tY \, \Theta) \;  z\big) \; (\underline{i} + 1)$
+
+    \item $\Big(\lambda z. \big(
+    \lambda n: 
+    \tif{\tsample - 0.5 < 0}{n}{(\tY \, \Theta) \; (n+1)} 
+    \big) \, z \Big)
+    \; \underline{i}$
+\end{enumerate}
+
+\lo{So we have:
+\[
+\begin{array}{ll}
+1 \to 11 \to 4 \to 5 \to 6 \to 7 \to 8\\
+7 \to 9 \to 11 \to \cdots
+\end{array}
+\]
+You have 2, 3 and 10 because your $\tY$-redex rule is
+\[
+\tY \lambda x. s \to \lambda z. (\lambda x.s)\, (\tY \lambda x. s) \, z
+\]
+which is not the official definition.}
+
+Even in this simple case, defining a ranking function explicitly is awkward because of the number of cases, although in most cases, because the value need only be greater than or equal to that of the next term in sequence, it suffices to take the ranking function as taking only 3 distinct values.
+(We will explain why later.)
+
+The definition of rankability is also inconvenient for syntactic sugar.
+It could be useful, for example, to define $M \oplus_p N = \tif{\tsample - p < 0} M N$, where $M \oplus_p N$ reduces to $M$ or $N$, depending on the first value of $s$, with probability $p$ resp.~$(1-p)$.
+Technically though, it reduces first to $\tif{\underline r - p < 0} M N$ for $r \in I$, so those terms all need values of the ranking function too.
+
+\paragraph{}
+In both of these cases, there are only some values of the ranking function that are semantically important.
+Define a \emph{partial ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \rightharpoonup \mathbb R$ such that
+\begin{itemize}
+    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
+    \item $f$ is defined at $M$.
+    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually \akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and ast, but that doesn't actually work because of terms reachable with 0 probability.
+Of course, astness doesn't imply rankability, so any definition of partial ranking function that includes all Y-past terms wouldn't always extend to a total ranking function.} reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
+\end{itemize}
+A partial ranking function that is total is just a ranking function.
+Providing a partial ranking function is essentially part way between providing a ranking function and directly proving almost sure termination.
+
+\begin{theorem} \label{partial implies rankable}
+Every partial ranking function is a restriction of a ranking function.
+\end{theorem}
+\begin{proof}
+Take a closed term $M$ and partial ranking function $f$ on $M$. Define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
+\begin{align*}
+    f_1(N) & = f(N) \text{ whenever $f(N)$ is defined,} \\
+    f_1(V) & = 0 \text{ for values $V$ not in the domain of $f$.}
+\end{align*}
+
+Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) = \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
+The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on partial ranking functions.
+Define $f_2(N) = \int_S f_1(\nnext(N,s)) + g(N,s) \, \mu(\mathrm d s)$.
+The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction).
+\lo{TODO. The preceding sentence deserves a proof.}
+\end{proof}
+
+As a corollary, any term which admits a partial ranking function terminates almost surely.
+
+\subsection{Examples}
+Let $M \oplus_p N = \tif{\tsample - p < 0} M N$, for $p \in (0,1]$, then there are the pseudo-reduction relations
+\begin{align*}
+E[M \oplus_p N] \to^3 & E[M] & \\
+E[M \oplus_p N] \to^3 & E[N] & \\
+\red^3(E[M \oplus_p N], s) = & \left\{
+    \begin{array}{ll}
+        (E[M],\pi_t(s)) & \text{if } \pi_h(s) < p \\
+        (E[N],\pi_t(s)) & \text{if } \pi_h(s) \geq p. \\
+    \end{array} \right .
+\end{align*}
+A partial ranking function could be defined with respect to this shortcut reduction simply by replacing $\to$ and $\red$ in the definition of a partial ranking function by a version that goes straight from $N \oplus_p O$ to $N$ or $O$.
+Such a pseudo-partial ranking function would then be a partial function from a subset of $Rch(M)$, so it could also be considered as a partial function from all of $Rch(M)$, and it would in fact also be an actual partial ranking function.
+It is therefore possible to prove rankability directly using the shortcutted reductions.
+
+A similar procedure would work for other forms of syntactic sugar.
+If a closed term $N$ eventually reduces to one of a set of other terms $\{N_i \mid i \in I\}$ with certain probabilities, a partial ranking function defined with respect to a reduction sequence that skips straight from $N$ to $N_i$ is also a valid partial ranking function for the original reduction function, and therefore its existence implies almost sure termination.
+There is a caveat, however, that $\tY$-reduction steps skipped over in the shortcut still need to be counted for the expected number of $\tY$-reduction steps.
+
+\paragraph{}
+With this abbreviation, the geometric distribution example from earlier can be written as $(\tY f, n.
+n \oplus_{0.5} f(n+1)) \underline 0$.
+It is then easy to see that the following is a partial ranking function:
+\begin{align*}
+(\tY f, n. n \oplus_{0.5} f(n+1)) N \mapsto 2 \\
+\underline i \oplus_{0.5} (\tY f, n. n \oplus_{0.5} f(n+1)) (\underline i + 1) \mapsto 1 \\
+\underline i \mapsto 0
+\end{align*}
+
+\lo{Call the above partial function $h$.
+I assume that $h: (\tY f, n. n \oplus_{0.5} f(n+1)) N \mapsto 2$ for all $N$ (such that $(\tY f, n. n \oplus_{0.5} f(n+1)) N$ is reachable from $M$), 
+and the second and third clauses hold for all $i \geq 0$.
+
+Maybe I have misread / misunderstood the preceding, but I don't see why $h$ is a restriction of a ranking function. 
+Suppose $h$ is a restriction of $H$.
+Consider the reduction sequence (writing $\Xi = (\tY f, n. n \oplus_{0.5} f(n+1))$):
+\[
+\Xi \, 0 
+\xrightarrow{\tY} 
+0 \oplus_{0.5} \Xi \, 1 
+\to 
+\Xi \, 1
+\xrightarrow{\tY} 
+1 \oplus_{0.5} \Xi \, 2
+\to
+\Xi \, 2
+\xrightarrow{\tY} 
+2 \oplus_{0.5} \Xi \, 3
+\]
+Since $H(\Xi \, 0) = 2$, we must have 
+$H(\Xi \, 1) \leq 1$ and $H(\Xi \, 2) \leq 0$, contradicting the restriction assumption.
+}
+\akr{
+In this case, $H(0 \oplus_{0.5} \Xi 1) \leq 1$, but that term reduces to either $0$ or $\Xi 1$ therefore the actual condition is that $0.5 H(0) + 0.5 H(\Xi 1) \leq 1$, which is possible.
+}
+\lo{
+  OK. I see it now.
+}
+
+\section{Antitone ranking functions}
+
+\begin{example}
+\label{ex:ac-ranking}
+\begin{enumerate}
+\item 1D unbiased random walk
+\[
+M_1 = 
+\big(\tY \, f \, n \, . \, 
+\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}\big) \, 10
+\]
+
+\item 1D biased (away from 0) random walk
+\[
+M_2 = 
+\big(\tY \, f \, n \, . \, 
+\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/3} f \, (n + 1)}\big) \, 10
+\]
+
+\item 1D biased (towards 0) random walk
+\[
+M_3 = 
+\big(\tY \, f \, n \, . \, 
+\tif{n = 0}{0}{f \, (n - 1) \oplus_{2/3} f \, (n + 2)}\big) \, 10
+\]
+
+%While {x > 0} do {x := x-1 \oplus_2/3 x:= x+1}
+
+%M2 = While {x > 0} do {x := x-1 \oplus_2/3 x:= x+2}
+\end{enumerate}
+\end{example}
+
+\paragraph{Problem.} Find a generalised notion of ranking function 
+%--- call it \emph{antitone-convex rankable} --- 
+so that $M_1$ becomes rankable, and then prove it sound i.e.~rankable in this generalised sense implies AST.
+
+\paragraph{Idea.}
+
+The definition of antitone-convex ranking function $f$ (say) for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
+we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) and convex function $\epsilon : \pReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
+\[
+f(E[R’]) \leq f(E[R]) - \epsilon(f(E[R])) 
+\]
+where $R \to R’$ is the $\tY$-redex rule.
+
+\lo{An aside: This CBV $\tY$-rule seems cleaner:
+\[
+\big(\tY f^{A \to B} \, x^A \, . \, \theta^B \big) \, v \to \theta[(\tY f \, x \, . \, \theta) / f, v / x].
+\]
+We assume $\tY f^{A \to B} \, x^A \, . \, \theta^B$ is a value.}
+
+\begin{definition}
+We say that a supermartingale $(Y_n)_{n \in \omega}$ and a stopping time $T$ \akr{Separating the stopping time from $Y_n = 0$ allows for a neater definition of the ranking function of a term} \lo{Nice!} adapted to filtration $(\calF_n)_{n \in \omega}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone measurable function $\epsilon : \nnReal \to \pReal$ satisfying
+\[
+\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot 1_{\set{T > n}}
+\]
+\end{definition}
+
+\begin{theorem}
+\label{thm:a-c strict}
+Let $(Y_n)_{n \in \omega}$ be an antitone strict supermartingale via function $\epsilon : \nnReal \to \pReal$, and stopping time $T$. 
+Then $T < \infty$ almost surely.
+\akr{It's not actually true that $\expect{T} \leq \dfrac{\expect{Y_0}}{\epsilon_0}$ where $\epsilon_0 := \epsilon(\expect{Y_0})$.
+This is fortunate because it means that a rankability theorem based on antitone strict supermartingales actually has a chance of being stronger than the existing one (which was pretty much complete for $\tY$-PAST terms already).
+In fact, for any almost surely finite stopping time, there is a corresponding antitone strict supermartingale.
+This seems promising for the method being conditionally complete (for terms where all reachable terms are AST).}
+\lo{See \Cref{sec:Relativised completeness}.}
+\end{theorem}
+
+\begin{proof}
+First, as $(Y_n)$ is a supermartingale, $\expect{Y_n} \leq \expect{Y_0}$, therefore
+
+\begin{calculation}
+\expect{Y_n \mid T > n}
+  \step[=]{}
+\dfrac{\mathbb P[T > n] \expect{Y_n \mid T > n} + \mathbb P[T \leq n] \expect{Y_n \mid T \leq n} - \mathbb P[T \leq n] \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
+  \step[=]{}
+\dfrac{\expect{Y_n} - \mathbb P[T \leq n] \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
+  \step[\leq]{$Y_n \geq 0$ always}
+\dfrac{\expect{Y_n}}{\mathbb P[T > n]}
+  \step[\leq]{$(Y_n)_n$ is a supermartingale}
+\dfrac{\expect{Y_0}}{\mathbb P[T > n]}
+\end{calculation}
+
+\emph{Claim}: for all $0 < x \leq 1$, let $B_x = \left \lceil \dfrac{\expect{Y_0}+1}{x \; \epsilon(\expect{Y_0} \, x^{-1})} \right \rceil$, then
+\(
+\mathbb P[T > B_x] \leq x.
+\)
+As the convex hull of $\epsilon$ (the greatest convex function less than or equal to it) satisfies all the conditions assumed of $\epsilon$, in addition to being convex, assume wlog that $\epsilon$ is convex.
+Suppose for a contradiction that $\mathbb P[T > B_x] > x$, then for every $n \leq B_x$
+\lo{Change of justification of the second equality below: event, not discrete r.v. 
+Recall: expectation conditioning on r.v.~(resp.~event) is a r.v.~(resp.~number).}
+\begin{calculation}
+\expect{Y_n-Y_{n+1}}
+  \step[=]{definition and linearity of conditional expectation}
+\expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
+  \step[\geq]{antitone strict assumption}
+\expect{\epsilon(Y_n) \cdot 1_{\set{T > n}}}
+%  \step[=]{definition of conditional expectation for a discrete variable}
+\step[=]{definition of expectation conditioning on an \emph{event}}
+\mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
+  \step[\geq]{Jensen's inequality}
+\mathbb P[T > n]\ \epsilon(\expect{Y_n \mid T > n})
+  \step[\geq]{proved earlier}
+\mathbb P[T > n]\ \epsilon\Big(\dfrac{\expect{Y_0}}{\mathbb P[T > n]}\Big)
+  \step[>]{assumption, $\mathbb P[T > n] \geq \mathbb P[T > B_x] > x$}
+x \; \epsilon\big(\expect{Y_0} \, x^{-1}\big).
+\end{calculation}
+Therefore, by a telescoping sum
+\[
+\expect{Y_{B_x}} = \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1}) \leq -1 < 0 
+\]
+%\lo{$\expect{Y_0 - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})}$}
+which is a contradiction, therefore the claim must be true, therefore $P[T > n]$ tends to 0 as $n$ tends to infinity, therefore $T < \infty$ almost surely.
+\end{proof}
+
+\paragraph{Results to prove:}
+\begin{enumerate}
+\item (AST Soundness) If a closed SPCF term $M$ is antitone rankable, then $T^{\tY}_M < \infty$ a.s.~(equivalently $M$ is AST). \lo{This should be fine.}
+
+\item A partial ranking function theorem for antitone-convex ranking functions. \lo{TODO}
+\end{enumerate}
+
+\paragraph{\Cref{ex:ac-ranking} revisited}
+
+Programs $M_1$ and $M_3$ are AST.
+To construct antitone ranking functions, we appeal \lo{(hopefully)} to a partial ranking function theorem for antitone ranking functions.
+\iffalse
+\lo{NOTE. Suppose a function $f : E \to \Real$ from a convex subset $E \subseteq \Real^n$ has continuous second-order partial derivatives. Then it is convex iff its Hessian is positive semidefinite.}
+\fi
+\iffalse
+For $M_1$, consider the function $g_1(n) := \frac{2^n - 1}{2^{n-1}}$.
+\fi
+
+For $M_1$, define the function $g_1 : \nnReal \to [0, 2)$ by $g_1(x) = 2 - \frac{1}{2^{x-1}}$.
+Note that $(g_1 (n))_{n \in \omega}$ satisfies the recurrence relation
+\[
+\left\{
+\begin{array}{rll}
+z_0 &=& 0\\
+n \geq 1, \quad z_n &>& \dfrac{1}{2} \, \big( z_{n-1} + z_{n+1}\big).
+\end{array}
+\right.
+\]
+Set the antitone function $\epsilon_1 (y) :=  \frac{2-y}{4}$.
+Then
+\begin{equation}
+\epsilon_1(g_1(x)) = g_1(x) - \dfrac{1}{2} \, \big(g_1(x-1) + g_1(x+1)\big) = \dfrac{1}{2^{x+1}}
+\label{eqn:epsilon g M1}
+\end{equation}
+(Since $g_1$ is bijective, with inverse $g_1^{-1}(y) = 1 - \log (2-y)$, we can obtain $\epsilon_1(y)$ from \Cref{eqn:epsilon g M1}.) 
+%We have $1/2 \cdot \big(g_1(x-1) + g_1(x+1)\big) = 2 - \frac{5}{2^{x+1}}$.
+
+Using shorthand 
+$\Theta_1 = \tY \, f \, n \, . \, 
+\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
+we can define a partial antitone ranking function $f_1 : \mathit{Rch}(M_1) \to \nnReal$ with antitone function $\epsilon_1$ defined as follows:
+\begin{align*}
+{\Theta_1} \, n 
+&\mapsto 
+g_1(n)
+\\
+\tif{n = 0}{0}{\Theta_1 \, (n - 1) \oplus_{1/2} \Theta_1 \, (n + 1)}
+&\mapsto
+\dfrac{1}{2} \, \big(g_1(n-1) + g_1(n+1)\big)
+\\
+0 &\mapsto 0
+\end{align*}
+
+For $M_3$, we use the function $g_3(x) := 3 - \frac{1}{3^{x-1}}$ and the antitone function $\epsilon_3(y) := \frac{84-28y}{27}$. 
+Then
+\[
+\epsilon_3(g_3(x)) = g_3(x) - \dfrac{1}{3}\big(2 \cdot g_3(x-1) + \cdot g_3(x+2) \big) = 
+\frac{28}{3^{x+2}}.
+\]
+Following $M_1$ above, we can define a partial antitone ranking function $f_3 : \mathit{Rch}(M_3) \to \nnReal$ with $\epsilon_3$ as the antitone function.
+
+\begin{example}[Continuous random walk]\label{ex:raven complex}
+In SPCF we can construct a function whose argument changes by a random amount at each recursive call:
+\[
+\underbrace{\big
+(\tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample))} \big)}_{\Theta} \, 10
+\]
+We can construct a partial antitone ranking function $f$ as follows:
+\begin{align*}
+\Theta \, l 
+&\mapsto 
+l
+\\
+\tif{l \leq 0}{0}{\Theta \, (l - \tsample)}
+&\mapsto
+l - 1/2
+\\
+0 &\mapsto 0
+\end{align*}
+with the constant antitone function $\epsilon(y) = 1/2$.
+
+For a more complex example, consider the following ``continuous random walk''
+\[
+\underbrace{\big
+(\tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample) \oplus_{1/2} f(x + \tsample)} \big)}_{\Xi} \, 10
+\]
+Let $g(x) := \sqrt{2} - \frac{1}{2^{x-\frac{1}{2}}}$ and let $\epsilon$ be the function specified by
+\[
+\epsilon(g(x)) = g(x) - \dfrac{1}{2}\Big(g(x - \dfrac{1}{2}) + \cdot g(x + \dfrac{1}{2}) \Big) = 
+\dfrac{3 - 2\sqrt{2}}{2^{x+1}}.
+\]
+We define a partial antitone ranking function as follows:
+\begin{align*}
+\Xi \, l 
+&\mapsto 
+g(l)
+\\
+\tif{l \leq 0}{0}{\Xi \, (l - \tsample) \oplus_{1/2} \Xi \, (l + \tsample)}
+&\mapsto
+\frac{1}{2}\big(g(l - \frac{1}{2}) + \cdot g(l + \frac{1}{2}) \big)
+\\
+r \leq 0, \quad r &\mapsto 0
+\end{align*}
+\end{example}
+
+\begin{example}[Fair-in-the-limit random walk]
+\label{ex:Fair-in-the-limit random walk}\citep[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}
+\[
+\underbrace{\big
+(\tY \, f \, x . 
+\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)}_{\Xi} 
+\, 10
+\]
+To construct an antitone ranking function, we solve the recurrence relation:
+\[
+\left\{
+\begin{array}{rll}
+z_0 &=& 0\\
+n \geq 1, \quad z_n &>& \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
+\end{array}
+\right.
+\]
+It turns out the solution to Example $M_1$, $z_n = \frac{2^n - 1}{2^{n-1}}$, is nearly a solution (for $n = 1$, we have equality instead).
+
+Let $g(x) = \frac{2^n - 1}{2^{n-1}}$.
+Then 
+\begin{equation}
+\epsilon(g(x)) = 
+g(x) - \dfrac{x}{2x + 1} \, g(x-1) - \dfrac{x+1}{2x + 1} \, g(x+1) 
+= \dfrac{x-1}{2^{x} \, (2x + 1)}
+\label{eqn:in the limit}
+\end{equation}
+
+\lo{ongoing}
+
+\end{example}
+
+\begin{example}[Escaping spline]
+\label{ex:escaping spline}\citep[\S 5.4]{DBLP:journals/pacmpl/McIverMKK18}
+\[
+\underbrace{\big
+(\tY \, f \, x . 
+\tif{x \leq 0}{0}{0 \oplus_{\frac{1}{x+1}} f(x + 1)} \big)}_{\Xi} 
+\, 10
+\]
+\end{example}
+
+\input{y-stopping-time}
+
+\input{relativised-completeness}
+
+\section{Confluent Trace Semantics}
+When proving almost sure termination in this way, it is necessary to consider which terms a given term may reduce to.
 Sometimes however, the reduction that the programmer has in mind may not be strictly the call-by-value order defined so far, or considering an alternative reduction order may be simpler or more intuitive.
@@
 
-\section{Ranking functions}
-Given a probabilistic program (i.e.~a closed term $M$), in order to construct a supermartingale to prove its a.s.~termination, a function to assign values to each reachable program state is necessary.
-
-Let %$Rch(M) := \{x \in \Lambda \mid \exists (y_i) : M \to y_0 \to \dots \to y_n \to x\}$, 
-\changed[lo]{$Rch(M) := \{N \in \Lambda \mid M \to^\ast N\}$},
-with the $\sigma$-algebra induced as a subset of $\Lambda$.
-
-\begin{definition}\rm
-A \emph{ranking function on $M$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
-\begin{enumerate}
-    \item $f(E[\tY \lambda x. N]) \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z]) \text{ where $z$ is not free in $N$}$
-    \item $f(E[\tsample]) \geq \int_I f(E[\underline{x}]) \, \Leb(\mathrm{d}x)$
-    \akr{I'm not sure whether $Rch(M)$ is a measurable subset of $\Lambda$, so it might not have a natural measure, but I'm pretty sure this doesn't actually matter, and the $\sigma$-algebra at least is well-defined.
-$Rch(M)$ could fail to be measurable if there are measurable functions $\mathbb R^n \to \mathbb R$ with non-measurable ranges.
-I don't know whether these exist.
-They do for the Lebesgue $\sigma$-algebra, but I'm using the Borel ones.}
-    \lo{It is worth inserting the preceding (including the example of a measurable function with non-measurable range) as a remark (or as a ``commenting.sty'' note).}
-    \akr{I'd like to include such a remark, but the only reference I've been able to find for the fact that these functions exist is a Stack Exchange question\footnote{\url{https://math.stackexchange.com/questions/1717282/construction-of-a-borel-measurable-function-mapping-borel-set-to-non-borel-set?rq=1}} which implies that the construction is difficult but doesn't actually say what it is (so I probably won't be able to find it on my own).}
-
-    \item $f(E[R]) \geq f(E[R'])$ for any other redex $R$, where \changed[lo]{$R \to R'$}. \lo{Metavariables: $M$ and $N$ for terms and programs; $R$ / $R'$ for redex / contractum.}
-\end{enumerate}
-We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that $f(N) = 0$ iff $N$ is a value, and for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.
-
-Any closed term for which a ranking (respectively, strict ranking) function exists is called \emph{rankable} (respectively, \emph{strictly rankable}).
-\end{definition}
-
-\section{Supermartingales}
-
-\lo{A brief explanatory note on supermartingales by way of motivation is needed here.}
-
-\lo{Supermartingales is a fitting title of this section.
-However the details of the supermartingle construction are suppressed in the proof.
-I propose that you reformulate the theorem as follows.}
-
-\begin{theorem} 
-\label{thm:rankable gives supermartingale}
-\changed[lo]{If a closed SPCF term is rankable by $f$, then $(f(M_n))_{n \in \omega}$ is a supermartingale adapted to the filtration $(\mathcal{F}_n)_{n \in \omega}$ where $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.}
-\end{theorem}
-
-%\begin{proof}
-
-\lo{We also need to show $f(M_n)$ is integrable.}
-
-Given a closed term $M$ and a ranking function $f$ for it, define random variables on the probability space $S$ (where $s$ is a random variable) by
-\begin{align*}
-(M_n,s_n) & = \red^n(M,s) \\
-y_0 & = 0 \\
-y_{n+1} & = \min \{ k \mid k>y_n, M_k \text{ a value or of the form } E[\tY N] \}\\
-M'_n & = M_{y_n} \\
-X_n & = f(M'_n)
-\end{align*}
-and define a filtration $\mathcal{F}_n = \sigma(M_k, k \leq n)$ (i.e. all the samples used up to step $n$).
-
-\akr{This is quite a pedantic point but technically the proof wouldn't be correct without it.} In the following proof, conditional expectations (conditioning on $\mathcal F_n$) will be used.
-Conditional expectations are in general only defined up to a null set, but in this case there is a natural measure on each element of $\mathcal F_n$ (as they're always isomorphic to $S$), and this is the version of the conditional expectation that will be used.
-\lo{I find this a little cryptic.
-Perhaps I am missing something, but the definition of conditional expectation $\expect{f(M_{n+1}) \mid \mathcal{F}_n}$ is standard.
-The stochastic process $(M_n)_{n \in \omega}$ is trivially adapted to the natural filtration $(\mathcal{F}_n)_{n \in \omega}$.}
-
-The expectation of $f(M_{n+1})$ given $\mathcal{F}_n$ is trivially less than or equal to $f(M_n)$ in the cases that $M_n \neq E[\tsample]$, and in the case of $\tsample$,
-\lo{This case analysis is problematic. The r.v.~$M_n$ is a measurable function of type $S \to \Lambda^0$ defined by $M_n(s) :=  N$ where $\red^n(M, s) = (N, s_n)$. 
-In general, it does not make sense to say ``$M_n = E[\tsample]$'' without referencing a trace (i.e.~an element of of $S$).
-Take $M = \tif{\tsample - 0.5 < 0}{\tsample}{1}$.
-Then $M_3([0.1, \cdots]) = \tsample$ but $M_3([0.6, \cdots]) = 1$.
-The case analysis should take place ``inside the conditional expectation''.}
-\begin{align*}
-& \mathbb{E}[f(M_{n+1}) \mid \mathcal{F}_n] \\
-= & \mathbb{E}[f(M_{n+1}) \mid M_n = E[\tsample],\, \mathcal{F}_n] \\
-= & \mathbb{E}[f(E[\pi_h(s_n)]) \mid \mathcal{F}_n] \\
-= & \int_0^1 f(E[\underline x]) \, \mathrm{d} x \qquad & \text{as }s_n\text{ is independent of } \mathcal{F}_n \\
-\leq & f(E[\tsample]) \qquad & \text{by assumption on } f \\
-= & f(M_n),
-\end{align*}
-\lo{Problem: Some of the above terms (conditional expectation and $f(M_n)$) are random variables (i.e.~functions) but the integral is a real number.}
-therefore the values of the ranking function $f(M_n)$ are a supermartingale with respect to $\mathcal{F}_n$.
-
-Given $M'_n$, the reduction relation $\to$, \emph{excluding} $\tY$-reduction steps, is strongly normalising because of the type system, therefore there is some finite bound on the number of reduction steps that can take place from $M'_n$ without a $\tY$-reduction step, therefore $y_{n+1}$ is (conditional on $\mathcal{F}_{y_n+1}$) a bounded stopping time, therefore $\mathbb{E}[f(M_{y_{n+1}}) \mid \mathcal{F}_{y_n+1}] \leq f(M_{y_n+1})$.
-For $n>0$, if $M_{y_n}$ isn't already a value, then $M_{y_n} = E[\tY \lambda x.N]$ for some $E, N$, therefore $M_{y_n+1} = E[\lambda z.
-N[(\tY \lambda x.
-N)/x] z]$ and $f(M_{y_n+1}) \leq f(M_{y_n}) - 1$, therefore if $M_{y_n}$ isn't a value, $\mathbb{E}[X_{n+1} \mid \mathcal{F}_{y_n+1}] \leq X_n - 1$.
-
-In the other case that $X_n$ is a value, $X_{n+1} = X_n$, and $X_n = \mathbb E[X_n \mid \mathcal F_{y_n+1}]$ therefore in either case $\mathbb E[X_{n+1} \mid \mathcal F_{y_n+1}] - E[X_n \mid \mathcal F_{y_n+1}] \leq \mathbb -P[M_{y_n} \text{ not a value} \mid \mathcal F_{y_n+1}]$ (as the probability is 0 or 1).
-Marginalising over $\mathcal F_{y_n+1}$, this implies that $\mathbb{E}[X_{n+1}] - \mathbb{E}[X_{n}] \leq -\mathbb P[M_{y_n} \text{ not a value}]$ therefore as $\mathbb{E}[X_n]$ is bounded below, $\mathbb P[M_{y_n} \text{ not a value}]$ must tend to 0 as $n \to \infty$.
-$M_{y_n} \text{a value} \Rightarrow M_{y_{n+1}} \text{a value}$ therefore
-\begin{align*}
-\mathbb P[M_{y_n} \text{ not a value for infinitely many values of $n$}]  
-& = \mathbb P[\forall n: M_{y_n} \text{ not a value}] \\
-& \leq \inf_n \mathbb P[M_{y_n} \text{ not a value}] \\
-& = 0
-\end{align*} 
-therefore $M$ terminates almost surely.
-%\end{proof}
-
-\input{ranking-supermartingales}
-
-\section{Constructing Ranking Functions}
-Although rankability implies almost-sure termination, the converse does not hold in general. For example,
-\begin{equation}
-\tif{{-}\tsample < 0}{\underline{0}}{(\tY \lambda x. x) \; \underline 0}
-\label{ex:0 probability reachable}
-\end{equation}
-terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x. x) \underline 0$ is reachable, although that has probability 0. 
-Not only is this counterexample AST, it's positively almost surely terminating i.e.~the expected time to termination is finite.
-
-A ranking function can be constructed under the stronger assumptions that, for every $N$ reachable from $M$, the expected number of $\tY$-reduction steps from $N$ to a value is finite.
-In particular, the expected number of $\tY$-reduction steps from each reachable term is a ranking function.
-Note that a finite number of expected $\tY$-reduction steps does not necessarily imply a finite number of expected total reduction steps.
-
-\begin{example}
-\label{ex:tY finite does not imply t finite}
-The term
-\[
-M = \big(\tY \lambda f n. \tif{\tsample - 0.5 < 0}{n \, \underline 0}{f (\lambda x. n (n x))}\big) (\lambda x. x+1)
-\] 
-terminates with only 2 $\tY$-reductions on average i.e.~$\expect{T_M^\tY} = 2$, but applies the increment function $2^n$ times with probability $2^{-1-n}$ for $n \geq 0$, which diverges i.e.~$\expect{T_M} = \infty$.
-\end{example}
-
-\begin{theorem} \label{thm:minimal}
-Given a closed term $M$, the function $f:Rch(M) \to \mathbb R$ given by $f(N) = \mathbb E [\text{the number of }\tY\text{-reduction steps from }N\text{ to a value}]$, if it exists, is the least of all possible ranking functions of $M$.
-\end{theorem}
-\begin{proof}
-Let $f$ be the candidate least ranking function defined above, and suppose $g$ is another ranking function such that $f(N) > g(N)$ for some $N \in Rch(M)$.
-The restrictions of $f$ and $g$ to $Rch(N)$ have the same properties assumed of $f$ and $g$, so assume w.l.o.g.~that $N=M$.
-The difference $g - f$ is then a supermartingale (with the same setup as in the proof of \Cref{thm:rankable gives supermartingale}
-%Theorem \ref{rankable implies ast}) 
-therefore $\forall n.\ \mathbb E[g(M_n)] \leq \mathbb E[f(M_n)] + g(M)-f(M)$.
-
-$\mathbb E[f(M_n)] = \sum_{k=n}^\infty \mathbb P[M_n = E[\tY N] \text{ for some }E, N] \to 0 \text{ as } n \to \infty$ therefore as $g(M) - f(M) < 0$, eventually $\mathbb E[g(M_n)] < 0$, which is impossible, therefore $g \geq f$ as required.
-
-In order for $f$ to be the least ranking function of $M$, it also has to actually be a ranking function itself.
-Each of the conditions on a ranking function is easily verified from the definition of $f$.
-\end{proof}
-
-\subsection{Partial ranking functions}
-Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost sure termination, so it isn't useful for proving it.
-Take, for example, the term
-\begin{align*}
-&(\tY \lambda f, n: \\
-&\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
-&) \underline{0}
-\end{align*}
-which generates a geometric distribution.
-    Despite its simplicity, its $Rch$ contains all the terms
-\begin{enumerate}
-    \item $(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{0}$
-    \item $(\lambda z.(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) z) \underline{i}$
-    \item $(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{i}$
-    \item $(\lambda n: \tif{\tsample - 0.5 < 0}{n}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{n}{f (m+1)}) (m+1)}) \underline{i}$
-    \item $\tif{\tsample - 0.5 < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
-    \item $\tif{\underline r - 0.5 < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
-    \item $\tif{\underline{r - 0.5} < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
-    \item $\underline{i}$
-    \item $(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)$
-    \item $(\lambda z.(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) z) (\underline{i} + 1)$.
-\end{enumerate}
-\akr{TODO: make this into a diagram with arrows.}
-
-\lo{Here is a more readable version.}
-
-Writing $\Theta = \lambda f, n : \tif{\tsample - 0.5 < 0}{n}{f (n+1)}$, we have
-\begin{enumerate}
-    \item $(\tY \, \Theta) \; \underline{0}$
-
-    \item $\big(\lambda z.\Theta \; (\tY \, \Theta) \; z \big) \; \underline{i}$
-
-    \item $\Theta \; (\tY \, \Theta) \; \underline{i}$
-
-    \item \changed[akr]{$(\lambda n: \tif{\tsample - 0.5 < 0}{n}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{n}{f (m+1)}) (m+1)}) \; \underline{i}$}
-
-    \lo{I think the preceding should be: $(\lambda n: \tif{\tsample - 0.5 < 0}{n}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (n+1)}) \; \underline{i}$, and so, more succinctly:
-    \[
-    \big(
-    \lambda n: 
-    \tif{\tsample - 0.5 < 0}{n}{(\tY \, \Theta) \; (n+1)} 
-    \big) 
-    \; \underline{i}
-    \]}
-
-    \item $\tif{\tsample - 0.5 < 0}{\underline{i}}{
-    (\tY \, \Theta) \, (\underline{i}+1)}$
-
-    \item $\tif{\underline r - 0.5 < 0}{\underline{i}}{(\tY \, \Theta) (\underline{i}+1)}$
-
-    \item $\tif{\underline{r - 0.5} < 0}{\underline{i}}{(\tY \, \Theta) (\underline{i}+1)}$
-
-    \item $\underline{i}$
-
-    \item $(\tY \, \Theta) \; (\underline{i}+1)$
-
-    \item $\big(\lambda z. \Theta \; (\tY \, \Theta) \;  z\big) \; (\underline{i} + 1)$
-
-    \item $\Big(\lambda z. \big(
-    \lambda n: 
-    \tif{\tsample - 0.5 < 0}{n}{(\tY \, \Theta) \; (n+1)} 
-    \big) \, z \Big)
-    \; \underline{i}$
-\end{enumerate}
-
-\lo{So we have:
-\[
-\begin{array}{ll}
-1 \to 11 \to 4 \to 5 \to 6 \to 7 \to 8\\
-7 \to 9 \to 11 \to \cdots
-\end{array}
-\]
-You have 2, 3 and 10 because your $\tY$-redex rule is
-\[
-\tY \lambda x. s \to \lambda z. (\lambda x.s)\, (\tY \lambda x. s) \, z
-\]
-which is not the official definition.}
-
-Even in this simple case, defining a ranking function explicitly is awkward because of the number of cases, although in most cases, because the value need only be greater than or equal to that of the next term in sequence, it suffices to take the ranking function as taking only 3 distinct values.
-(We will explain why later.)
-
-The definition of rankability is also inconvenient for syntactic sugar.
-It could be useful, for example, to define $M \oplus_p N = \tif{\tsample - p < 0} M N$, where $M \oplus_p N$ reduces to $M$ or $N$, depending on the first value of $s$, with probability $p$ resp.~$(1-p)$.
-Technically though, it reduces first to $\tif{\underline r - p < 0} M N$ for $r \in I$, so those terms all need values of the ranking function too.
-
-\paragraph{}
-In both of these cases, there are only some values of the ranking function that are semantically important.
-Define a \emph{partial ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \rightharpoonup \mathbb R$ such that
-\begin{itemize}
-    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
-    \item $f$ is defined at $M$.
-    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually \akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and ast, but that doesn't actually work because of terms reachable with 0 probability.
-Of course, astness doesn't imply rankability, so any definition of partial ranking function that includes all Y-past terms wouldn't always extend to a total ranking function.} reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
-\end{itemize}
-A partial ranking function that is total is just a ranking function.
-Providing a partial ranking function is essentially part way between providing a ranking function and directly proving almost sure termination.
-
-\begin{theorem} \label{partial implies rankable}
-Every partial ranking function is a restriction of a ranking function.
-\end{theorem}
-\begin{proof}
-Take a closed term $M$ and partial ranking function $f$ on $M$. Define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
-\begin{align*}
-    f_1(N) & = f(N) \text{ whenever $f(N)$ is defined,} \\
-    f_1(V) & = 0 \text{ for values $V$ not in the domain of $f$.}
-\end{align*}
-
-Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) = \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
-The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on partial ranking functions.
-Define $f_2(N) = \int_S f_1(\nnext(N,s)) + g(N,s) \, \mu(\mathrm d s)$.
-The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction).
-\lo{TODO. The preceding sentence deserves a proof.}
-\end{proof}
-
-As a corollary, any term which admits a partial ranking function terminates almost surely.
-
-\subsection{Examples}
-Let $M \oplus_p N = \tif{\tsample - p < 0} M N$, for $p \in (0,1]$, then there are the pseudo-reduction relations
-\begin{align*}
-E[M \oplus_p N] \to^3 & E[M] & \\
-E[M \oplus_p N] \to^3 & E[N] & \\
-\red^3(E[M \oplus_p N], s) = & \left\{
-    \begin{array}{ll}
-        (E[M],\pi_t(s)) & \text{if } \pi_h(s) < p \\
-        (E[N],\pi_t(s)) & \text{if } \pi_h(s) \geq p. \\
-    \end{array} \right .
-\end{align*}
-A partial ranking function could be defined with respect to this shortcut reduction simply by replacing $\to$ and $\red$ in the definition of a partial ranking function by a version that goes straight from $N \oplus_p O$ to $N$ or $O$.
-Such a pseudo-partial ranking function would then be a partial function from a subset of $Rch(M)$, so it could also be considered as a partial function from all of $Rch(M)$, and it would in fact also be an actual partial ranking function.
-It is therefore possible to prove rankability directly using the shortcutted reductions.
-
-A similar procedure would work for other forms of syntactic sugar.
-If a closed term $N$ eventually reduces to one of a set of other terms $\{N_i \mid i \in I\}$ with certain probabilities, a partial ranking function defined with respect to a reduction sequence that skips straight from $N$ to $N_i$ is also a valid partial ranking function for the original reduction function, and therefore its existence implies almost sure termination.
-There is a caveat, however, that $\tY$-reduction steps skipped over in the shortcut still need to be counted for the expected number of $\tY$-reduction steps.
-
-\paragraph{}
-With this abbreviation, the geometric distribution example from earlier can be written as $(\tY f, n.
-n \oplus_{0.5} f(n+1)) \underline 0$.
-It is then easy to see that the following is a partial ranking function:
-\begin{align*}
-(\tY f, n. n \oplus_{0.5} f(n+1)) N \mapsto 2 \\
-\underline i \oplus_{0.5} (\tY f, n. n \oplus_{0.5} f(n+1)) (\underline i + 1) \mapsto 1 \\
-\underline i \mapsto 0
-\end{align*}
-
-\lo{Call the above partial function $h$.
-I assume that $h: (\tY f, n. n \oplus_{0.5} f(n+1)) N \mapsto 2$ for all $N$ (such that $(\tY f, n. n \oplus_{0.5} f(n+1)) N$ is reachable from $M$), 
-and the second and third clauses hold for all $i \geq 0$.
-
-Maybe I have misread / misunderstood the preceding, but I don't see why $h$ is a restriction of a ranking function. 
-Suppose $h$ is a restriction of $H$.
-Consider the reduction sequence (writing $\Xi = (\tY f, n. n \oplus_{0.5} f(n+1))$):
-\[
-\Xi \, 0 
-\xrightarrow{\tY} 
-0 \oplus_{0.5} \Xi \, 1 
-\to 
-\Xi \, 1
-\xrightarrow{\tY} 
-1 \oplus_{0.5} \Xi \, 2
-\to
-\Xi \, 2
-\xrightarrow{\tY} 
-2 \oplus_{0.5} \Xi \, 3
-\]
-Since $H(\Xi \, 0) = 2$, we must have 
-$H(\Xi \, 1) \leq 1$ and $H(\Xi \, 2) \leq 0$, contradicting the restriction assumption.
-}
-\akr{
-In this case, $H(0 \oplus_{0.5} \Xi 1) \leq 1$, but that term reduces to either $0$ or $\Xi 1$ therefore the actual condition is that $0.5 H(0) + 0.5 H(\Xi 1) \leq 1$, which is possible.
-}
-\lo{
-  OK. I see it now.
-}
-
-\section{Antitone ranking functions}
-
-\begin{example}
-\label{ex:ac-ranking}
-\begin{enumerate}
-\item 1D unbiased random walk
-\[
-M_1 = 
-\big(\tY \, f \, n \, . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}\big) \, 10
-\]
-
-\item 1D biased (away from 0) random walk
-\[
-M_2 = 
-\big(\tY \, f \, n \, . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/3} f \, (n + 1)}\big) \, 10
-\]
-
-\item 1D biased (towards 0) random walk
-\[
-M_3 = 
-\big(\tY \, f \, n \, . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{2/3} f \, (n + 2)}\big) \, 10
-\]
-
-%While {x > 0} do {x := x-1 \oplus_2/3 x:= x+1}
-
-%M2 = While {x > 0} do {x := x-1 \oplus_2/3 x:= x+2}
-\end{enumerate}
-\end{example}
-
-\paragraph{Problem.} Find a generalised notion of ranking function 
-%--- call it \emph{antitone-convex rankable} --- 
-so that $M_1$ becomes rankable, and then prove it sound i.e.~rankable in this generalised sense implies AST.
-
-\paragraph{Idea.}
-
-The definition of antitone-convex ranking function $f$ (say) for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
-we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) and convex function $\epsilon : \pReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
-\[
-f(E[R’]) \leq f(E[R]) - \epsilon(f(E[R])) 
-\]
-where $R \to R’$ is the $\tY$-redex rule.
-
-\lo{An aside: This CBV $\tY$-rule seems cleaner:
-\[
-\big(\tY f^{A \to B} \, x^A \, . \, \theta^B \big) \, v \to \theta[(\tY f \, x \, . \, \theta) / f, v / x].
-\]
-We assume $\tY f^{A \to B} \, x^A \, . \, \theta^B$ is a value.}
-
-\begin{definition}
-We say that a supermartingale $(Y_n)_{n \in \omega}$ and a stopping time $T$ \akr{Separating the stopping time from $Y_n = 0$ allows for a neater definition of the ranking function of a term} \lo{Nice!} adapted to filtration $(\calF_n)_{n \in \omega}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone measurable function $\epsilon : \nnReal \to \pReal$ satisfying
-\[
-\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot 1_{\set{T > n}}
-\]
-\end{definition}
-
-\begin{theorem}
-\label{thm:a-c strict}
-Let $(Y_n)_{n \in \omega}$ be an antitone strict supermartingale via function $\epsilon : \nnReal \to \pReal$, and stopping time $T$. 
-Then $T < \infty$ almost surely.
-\akr{It's not actually true that $\expect{T} \leq \dfrac{\expect{Y_0}}{\epsilon_0}$ where $\epsilon_0 := \epsilon(\expect{Y_0})$.
-This is fortunate because it means that a rankability theorem based on antitone strict supermartingales actually has a chance of being stronger than the existing one (which was pretty much complete for $\tY$-PAST terms already).
-In fact, for any almost surely finite stopping time, there is a corresponding antitone strict supermartingale.
-This seems promising for the method being conditionally complete (for terms where all reachable terms are AST).}
-\lo{See \Cref{sec:Relativised completeness}.}
-\end{theorem}
-
-\begin{proof}
-First, as $(Y_n)$ is a supermartingale, $\expect{Y_n} \leq \expect{Y_0}$, therefore
-
-\begin{calculation}
-\expect{Y_n \mid T > n}
-  \step[=]{}
-\dfrac{\mathbb P[T > n] \expect{Y_n \mid T > n} + \mathbb P[T \leq n] \expect{Y_n \mid T \leq n} - \mathbb P[T \leq n] \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
-  \step[=]{}
-\dfrac{\expect{Y_n} - \mathbb P[T \leq n] \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
-  \step[\leq]{$Y_n \geq 0$ always}
-\dfrac{\expect{Y_n}}{\mathbb P[T > n]}
-  \step[\leq]{$(Y_n)_n$ is a supermartingale}
-\dfrac{\expect{Y_0}}{\mathbb P[T > n]}
-\end{calculation}
-
-\emph{Claim}: for all $0 < x \leq 1$, let $B_x = \left \lceil \dfrac{\expect{Y_0}+1}{x \; \epsilon(\expect{Y_0} \, x^{-1})} \right \rceil$, then
-\(
-\mathbb P[T > B_x] \leq x.
-\)
-As the convex hull of $\epsilon$ (the greatest convex function less than or equal to it) satisfies all the conditions assumed of $\epsilon$, in addition to being convex, assume wlog that $\epsilon$ is convex.
-Suppose for a contradiction that $\mathbb P[T > B_x] > x$, then for every $n \leq B_x$
-\lo{Change of justification of the second equality below: event, not discrete r.v. 
-Recall: expectation conditioning on r.v.~(resp.~event) is a r.v.~(resp.~number).}
-\begin{calculation}
-\expect{Y_n-Y_{n+1}}
-  \step[=]{definition and linearity of conditional expectation}
-\expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
-  \step[\geq]{antitone strict assumption}
-\expect{\epsilon(Y_n) \cdot 1_{\set{T > n}}}
-%  \step[=]{definition of conditional expectation for a discrete variable}
-\step[=]{definition of expectation conditioning on an \emph{event}}
-\mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
-  \step[\geq]{Jensen's inequality}
-\mathbb P[T > n]\ \epsilon(\expect{Y_n \mid T > n})
-  \step[\geq]{proved earlier}
-\mathbb P[T > n]\ \epsilon\Big(\dfrac{\expect{Y_0}}{\mathbb P[T > n]}\Big)
-  \step[>]{assumption, $\mathbb P[T > n] \geq \mathbb P[T > B_x] > x$}
-x \; \epsilon\big(\expect{Y_0} \, x^{-1}\big).
-\end{calculation}
-Therefore, by a telescoping sum
-\[
-\expect{Y_{B_x}} = \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1}) \leq -1 < 0 
-\]
-%\lo{$\expect{Y_0 - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})}$}
-which is a contradiction, therefore the claim must be true, therefore $P[T > n]$ tends to 0 as $n$ tends to infinity, therefore $T < \infty$ almost surely.
-\end{proof}
-
-\paragraph{Results to prove:}
-\begin{enumerate}
-\item (AST Soundness) If a closed SPCF term $M$ is antitone rankable, then $T^{\tY}_M < \infty$ a.s.~(equivalently $M$ is AST). \lo{This should be fine.}
-
-\item A partial ranking function theorem for antitone-convex ranking functions. \lo{TODO}
-\end{enumerate}
-
-\paragraph{\Cref{ex:ac-ranking} revisited}
-
-Programs $M_1$ and $M_3$ are AST.
-To construct antitone ranking functions, we appeal \lo{(hopefully)} to a partial ranking function theorem for antitone ranking functions.
-\iffalse
-\lo{NOTE. Suppose a function $f : E \to \Real$ from a convex subset $E \subseteq \Real^n$ has continuous second-order partial derivatives. Then it is convex iff its Hessian is positive semidefinite.}
-\fi
-\iffalse
-For $M_1$, consider the function $g_1(n) := \frac{2^n - 1}{2^{n-1}}$.
-\fi
-
-For $M_1$, define the function $g_1 : \nnReal \to [0, 2)$ by $g_1(x) = 2 - \frac{1}{2^{x-1}}$.
-Note that $(g_1 (n))_{n \in \omega}$ satisfies the recurrence relation
-\[
-\left\{
-\begin{array}{rll}
-z_0 &=& 0\\
-n \geq 1, \quad z_n &>& \dfrac{1}{2} \, \big( z_{n-1} + z_{n+1}\big).
-\end{array}
-\right.
-\]
-Set the antitone function $\epsilon_1 (y) :=  \frac{2-y}{4}$.
-Then
-\begin{equation}
-\epsilon_1(g_1(x)) = g_1(x) - \dfrac{1}{2} \, \big(g_1(x-1) + g_1(x+1)\big) = \dfrac{1}{2^{x+1}}
-\label{eqn:epsilon g M1}
-\end{equation}
-(Since $g_1$ is bijective, with inverse $g_1^{-1}(y) = 1 - \log (2-y)$, we can obtain $\epsilon_1(y)$ from \Cref{eqn:epsilon g M1}.) 
-%We have $1/2 \cdot \big(g_1(x-1) + g_1(x+1)\big) = 2 - \frac{5}{2^{x+1}}$.
-
-Using shorthand 
-$\Theta_1 = \tY \, f \, n \, . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
-we can define a partial antitone ranking function $f_1 : \mathit{Rch}(M_1) \to \nnReal$ with antitone function $\epsilon_1$ defined as follows:
-\begin{align*}
-{\Theta_1} \, n 
-&\mapsto 
-g_1(n)
-\\
-\tif{n = 0}{0}{\Theta_1 \, (n - 1) \oplus_{1/2} \Theta_1 \, (n + 1)}
-&\mapsto
-\dfrac{1}{2} \, \big(g_1(n-1) + g_1(n+1)\big)
-\\
-0 &\mapsto 0
-\end{align*}
-
-For $M_3$, we use the function $g_3(x) := 3 - \frac{1}{3^{x-1}}$ and the antitone function $\epsilon_3(y) := \frac{84-28y}{27}$. 
-Then
-\[
-\epsilon_3(g_3(x)) = g_3(x) - \dfrac{1}{3}\big(2 \cdot g_3(x-1) + \cdot g_3(x+2) \big) = 
-\frac{28}{3^{x+2}}.
-\]
-Following $M_1$ above, we can define a partial antitone ranking function $f_3 : \mathit{Rch}(M_3) \to \nnReal$ with $\epsilon_3$ as the antitone function.
-
-\begin{example}[Continuous random walk]\label{ex:raven complex}
-In SPCF we can construct a function whose argument changes by a random amount at each recursive call:
-\[
-\underbrace{\big
-(\tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample))} \big)}_{\Theta} \, 10
-\]
-We can construct a partial antitone ranking function $f$ as follows:
-\begin{align*}
-\Theta \, l 
-&\mapsto 
-l
-\\
-\tif{l \leq 0}{0}{\Theta \, (l - \tsample)}
-&\mapsto
-l - 1/2
-\\
-0 &\mapsto 0
-\end{align*}
-with the constant antitone function $\epsilon(y) = 1/2$.
-
-For a more complex example, consider the following ``continuous random walk''
-\[
-\underbrace{\big
-(\tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample) \oplus_{1/2} f(x + \tsample)} \big)}_{\Xi} \, 10
-\]
-Let $g(x) := \sqrt{2} - \frac{1}{2^{x-\frac{1}{2}}}$ and let $\epsilon$ be the function specified by
-\[
-\epsilon(g(x)) = g(x) - \dfrac{1}{2}\Big(g(x - \dfrac{1}{2}) + \cdot g(x + \dfrac{1}{2}) \Big) = 
-\dfrac{3 - 2\sqrt{2}}{2^{x+1}}.
-\]
-We define a partial antitone ranking function as follows:
-\begin{align*}
-\Xi \, l 
-&\mapsto 
-g(l)
-\\
-\tif{l \leq 0}{0}{\Xi \, (l - \tsample) \oplus_{1/2} \Xi \, (l + \tsample)}
-&\mapsto
-\frac{1}{2}\big(g(l - \frac{1}{2}) + \cdot g(l + \frac{1}{2}) \big)
-\\
-r \leq 0, \quad r &\mapsto 0
-\end{align*}
-\end{example}
-
-\begin{example}[Fair-in-the-limit random walk]
-\label{ex:Fair-in-the-limit random walk}\citep[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}
-\[
-\underbrace{\big
-(\tY \, f \, x . 
-\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)}_{\Xi} 
-\, 10
-\]
-To construct an antitone ranking function, we solve the recurrence relation:
-\[
-\left\{
-\begin{array}{rll}
-z_0 &=& 0\\
-n \geq 1, \quad z_n &>& \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
-\end{array}
-\right.
-\]
-It turns out the solution to Example $M_1$, $z_n = \frac{2^n - 1}{2^{n-1}}$, is nearly a solution (for $n = 1$, we have equality instead).
-
-Let $g(x) = \frac{2^n - 1}{2^{n-1}}$.
-Then 
-\begin{equation}
-\epsilon(g(x)) = 
-g(x) - \dfrac{x}{2x + 1} \, g(x-1) - \dfrac{x+1}{2x + 1} \, g(x+1) 
-= \dfrac{x-1}{2^{x} \, (2x + 1)}
-\label{eqn:in the limit}
-\end{equation}
-
-\lo{ongoing}
-
-\end{example}
-
-\begin{example}[Escaping spline]
-\label{ex:escaping spline}\citep[\S 5.4]{DBLP:journals/pacmpl/McIverMKK18}
-\[
-\underbrace{\big
-(\tY \, f \, x . 
-\tif{x \leq 0}{0}{0 \oplus_{\frac{1}{x+1}} f(x + 1)} \big)}_{\Xi} 
-\, 10
-\]
-\end{example}
-
-\input{y-stopping-time}
-
-\input{relativised-completeness}
-
 \bibliographystyle{apalike}

### 2020-12-01 ranking-supermartingales.tex
Rearrange sections, various minor corrections.

@@
 %!TEX root = probabilisticProgrammingMartingales.tex
-\subsection*{Reorganised proof}
+\paragraph{}
 
@@
 \lo{The preceding means: $\forall A \in \calF_n$, $\int_A \dif \mu \, Y_{n+1} \leq \int_A \dif \mu \,Y_n$.}
-For $\epsilon > 0$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{Y_n > 0}}$.
+For a stopping time $T$ and $\epsilon > 0$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.
 \citep{DBLP:conf/popl/FioritiH15,DBLP:conf/popl/ChatterjeeFNH16}
@@
 
-Intuitively $Y_n$ gives the rank of the program after $n$ steps of computation.
-(Assume that the rank of a term is 0 iff it is a value.) In a $\epsilon$-ranking supermartingale, each computation step causes a strict decrease in rank of at least $\epsilon$, provided the term in question is not a value.
+Intuitively $Y_n$ gives the rank of the program after $n$ steps of computation, and $T$ is the time at which it reaches a value (which may be infinite if it fails to terminate).
+In a $\epsilon$-ranking supermartingale, each computation step causes a strict decrease in rank of at least $\epsilon$, provided the term in question is not a value.
 
@@
 \label{thm:rank-PAST}
-Let $(Y_n)_{n \in \omega}$ be a $\epsilon$-ranking supermartingale, and define the r.v.~$T$ by $T(s) := \min \set{n \mid Y_n(s) = 0}$.
-
-Then $T < \infty$ almost surely, and $\expect{T} \leq \dfrac{\expect{Y_0}}{\epsilon}$.
+Let $(Y_n)_{n \in \omega}$ be a $\epsilon$-ranking supermartingale with respect to the stopping time $T$, then $T < \infty$ almost surely, and $\expect{T} \leq \dfrac{\expect{Y_0}}{\epsilon}$.
 \end{theorem}
@@
 \[
-\forall n \in \omega \, . \, \expect{Y_n} \leq \expect{Y_0} - \epsilon \cdot \Big(\sum_{i=0}^{n-1} \mathbb{P}[Y_i > 0]\Big).
+\forall n \in \omega \, . \, \expect{Y_n} \leq \expect{Y_0} - \epsilon \cdot \Big(\sum_{i=0}^{n-1} \mathbb{P}[T > i]\Big).
 \]
-It follows that $\sum_{i=0}^{\infty} \mathbb{P}[Y_i > 0]$ converges; hence $\lim_{n \to \infty} \mathbb{P}[Y_n > 0] = 0$ and so $\mathbb{P}[T < \infty] = 1$.
-Then it suffices to prove: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=1} \mathbb{P}[T \geq i]$.
+It follows that $\sum_{i=0}^{\infty} \mathbb{P}[T > i]$ converges; hence $\lim_{n \to \infty} \mathbb{P}[T > i] = 0$ and so $\mathbb{P}[T < \infty] = 1$.
+Then it suffices to prove: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=0} \mathbb{P}[T > i]$.
 \end{proof}
@@
 \begin{align}
-T_0(s) & := 0 \nonumber \\
+T_{-1}(s) & := -1 \nonumber \\
 T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value or of form } E[\tY \lambda x. N] \} \nonumber \\
@@
 Y_n(s) & := X_{T_n}(s) \nonumber \\
-T(s) &:= \min \set{n \mid Y_n(s) = 0} \label{eq:y stopping time}
+T_M^\tY(s) &:= \min \set{n \mid M_{T_n(s)}(s) \textrm{ is a value}} \label{eq:y stopping time}
 \end{align}
-\lo{Define the \emph{$\tY$-runtime of $M$} to be the random variable $T^{\tY}_M$: 
-\begin{itemize}
-\item if $T_M(s) = n < \infty$ then $T^{\tY}_M(s) :={}$ number of $\tY$-reduction steps in the $n$ steps of reduction to $\red^n(M, s)$
-\item if $T_M(s) = \infty$ then $T^{\tY}_M(s) := \infty$.
-\end{itemize}
-Observe that $T = T^{\tY}_M$.}
-Note that 
-\begin{enumerate}
-\item $T_M < \infty$ a.s.~iff $T_M^{\tY} < \infty$ a.s.~(because $T_M(s) < \infty$ iff $T_M^{\tY}(s) < \infty$)
-\item $\expect{T_M} <{} \infty$ implies $\expect{T_M^{\tY}} < \infty$ (because $T_M^{\tY} \leq T_M$), but the converse is false (\Cref{ex:tY finite does not imply t finite}).
-\end{enumerate}
-%Note that by definition $T_M^{\tY} \leq T_M$.
+The random variable $T_M^\tY$ is known as the \emph{$\tY$-runtime of $M$}, and it can equivalently be defined as the number of $\tY$-reduction steps in the reduction sequence of $M$.
+Note that, as the type system ensures that the reduction relation excluding $\tY$-reduction is strongly normalising, only finitely many reductions can occur in a row without one of them being a $\tY$-reduction, therefore $T_M^\tY < \infty$ iff $M$ is AST.
 
@@
 $(X_n)_{n \in \omega}$ is a supermartingale;
-it then follows from \Cref{thm:optional sampling} that $(Y_n)_{n \in \omega}$ is a ($1$-ranking) supermartingale.
+it then follows from \Cref{thm:optional sampling} that $(Y_n)_{n \in \omega}$ is a ($1$-ranking) supermartingale with respect to the filtration $(\calF_{T_n})_{n \in \omega}$ and the stopping time $T^{\tY}_M$.
 Therefore, by \Cref{thm:rank-PAST}, $T^{\tY}_M < \infty$ a.s.~and $\expect{T^{\tY}_M} < \infty$.

### 2020-12-07 probabilisticProgrammingMartingales.tex
Add the antitone ranking theorem and its partial variant.

@@
 
-\paragraph{Results to prove:}
-\begin{enumerate}
-\item (AST Soundness) If a closed SPCF term $M$ is antitone rankable, then $T^{\tY}_M < \infty$ a.s.~(equivalently $M$ is AST). \lo{This should be fine.}
-
-\item A partial ranking function theorem for antitone-convex ranking functions. \lo{TODO}
-\end{enumerate}
+\begin{theorem}[Antitone ranking function soundness]
+If a closed SPCF term $M$ is antitone rankable, then $T_M^{\tY} < \infty$ a.s.~(equivalently, $M$ is AST).
+\end{theorem}
+\begin{proof}
+As in Theorem \ref{thm:basicRankingSoundness}, take the set $S$ of traces as a probability space, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$.
+As before, $(Y_n)_{n \in \omega}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$, therefore $T_M^{\tY}$ is almost surely finite.
+\end{proof}
+
+\paragraph{}
+As before, constructing antitone ranking fuctions completely is not necessary, and there is a corresponding notion of an antitone partial ranking function.
+
+Define an \emph{antitone partial ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$
+\begin{itemize}
+\item $f(N) \geq 0$ for all $N$ where $f$ is defined.
+\item $f$ is defined at $M$.
+\item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
+\end{itemize}
+
+\begin{theorem}
+  Every antitone partial ranking function is a restriction of an antitone ranking function.
+\end{theorem}
+\begin{proof}
+  Take a closed term $M$ and an antitone partial ranking function $f$ on $M$, with a corresponding antitone function $\epsilon$.
+Assume wlog that $\epsilon$ is convex (as if it isn't, we can just take its convex hull instead).
+As in Theorem \ref{partial implies rankable}, define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
+  \begin{align*}
+    f_1(N) &= f(N) \text{ whenever $f(N)$ is defined},\\
+    f_1(V) &= 0 \text{ for values $V$ not in the domain of $f$.}
+  \end{align*}
+
+  Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) = \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
+  The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on antitone partial ranking functions.
+Define $f_2(N) = \int_S f_1(\nnext(N,s)) + \epsilon(f_1(\nnext(N,s))) g(N,s) \, \mu(\mathrm d s)$.
+For any term $N$ where $f$ is defined, $f_2(N) = f(N)$, and the value that $f_2$ would have at $N$ if $f$ were not defined at $N$ is $\leq f(N)$, by the third condition on antitone partial ranking functions.
+In order to show that $f_2$ is an antitone partial ranking function, it therefore suffices to show that the value that $f_2$ would have had at each term if $f$ were not defined at that term is at least the expectation of $f_2$ after one reduction step (plus $\epsilon(f_2(N))$ if the reduction step is a $tY$-reduction).
+For any term $N$ which is not of the form $E[R]$ for some $\tY$-redex $R$, this is trivial.
+If $R$ is a $tY$-redex, then $\epsilon(f_2(N)) \leq \int_S \epsilon(f_1(\nnext(N,s)))\, \mu(\mathrm d s)$ by the convexity of $\epsilon$, because $n$ is bounded.
+Therefore, the (total) function $f_2$, which agrees with $f$ on $f$'s domain, is an antitone ranking function on $M$ (with the same function $\epsilon$ if it's convex).
+\end{proof}
+
+As a corollary, any term which admits an antitone partial ranking function terminates almost surely.
 
@@
 Programs $M_1$ and $M_3$ are AST.
-To construct antitone ranking functions, we appeal \lo{(hopefully)} to a partial ranking function theorem for antitone ranking functions.
+To construct antitone ranking functions, we appeal to a partial ranking function theorem for antitone ranking functions.
 \iffalse

### 2020-12-07 probabilisticProgrammingMartingales.tex
Some comments. Ongoing.

--- 2020-12-01 probabilisticProgrammingMartingales.tex
@@
 If $\alpha_1$ is an initial segment (i.e.~prefix) of $\alpha_2$ (written as $\alpha_1 \leq \alpha_2$), then $N_2$ is also a subterm of $N_1$.
-If neither $\alpha_1 \leq \alpha_2$ nor $\alpha_1 \geq \alpha_2$, the positions are said to be disjoint.
+If neither $\alpha_1 \leq \alpha_2$ nor $\alpha_1 \geq \alpha_2$, the positions are said to be \emph{disjoint}.
 This is mostly relevant in that in any substitution at some position, the subterms at all other positions are unaffected. 
-%\lo{The preceding sentence is unclear. Typo? Mostly $\mapsto$ most?}
+\lo{The preceding sentence is unclear. Typo?}
 
@@
 
-Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term has a skeleton $Sk(M)$ and each skeleton can be converted to a term $S[r]$ given a vector of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$.
+Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each \changed[lo]{term $M$} has a skeleton $Sk(M)$, and each \changed[lo]{skeleton $S$} can be converted to a term $S[r]$ given a \changed[lo]{vector $r$} of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$.
+\lo{Do we actually use the notation $S[r]$ in what follows?}
 \lo{Skeleton is defined in Sec.~\ref{sec:sampling semantics}.}

### 2020-12-11 probabilisticProgrammingMartingales.tex
Fix some examples.

--- 2020-12-07 probabilisticProgrammingMartingales.tex
@@
 
+Note that it is essential in this proof that $\epsilon$ be defined for all $\mathbb R_{\geq 0}$, not just the values that $(Y_n)_{n \in \omega}$ actually takes (or at least if $(Y_n)_{n \in \omega}$ is uniformly bounded, $\epsilon$ must be positive at the supremum as well as the actual values).
+Consider, for example, a random variable $X$, uniformly distributed in the interval $[-1,1]$, $T = $ the least $n$ such that $2^{-n} < X$, $Y_k = 4-2^{-k}$ if $k < T$, or $0$ otherwise, and $\epsilon(x) = \frac{4-x}{4}$.
+In this case, $(Y_n)_{n \in \omega}$ either tends to $4$ as $n \to \infty$, or drops to 0 at some point, with an exponentially decreasing probability.
+It is an antitone strict supermartingale except that $\epsilon$ goes to $0$ at $4$, which is larger than any value $(Y_n)_{n \in \omega}$ can actually take, and $T = \infty$ with probability $\frac 1 2$.
+
 \begin{theorem}[Antitone ranking function soundness]
@@
 To construct antitone ranking functions, we appeal to a partial ranking function theorem for antitone ranking functions.
-\iffalse
-\lo{NOTE. Suppose a function $f : E \to \Real$ from a convex subset $E \subseteq \Real^n$ has continuous second-order partial derivatives. Then it is convex iff its Hessian is positive semidefinite.}
-\fi
-\iffalse
-For $M_1$, consider the function $g_1(n) := \frac{2^n - 1}{2^{n-1}}$.
-\fi
-
-For $M_1$, define the function $g_1 : \nnReal \to [0, 2)$ by $g_1(x) = 2 - \frac{1}{2^{x-1}}$.
-Note that $(g_1 (n))_{n \in \omega}$ satisfies the recurrence relation
-\[
-\left\{
-\begin{array}{rll}
-z_0 &=& 0\\
-n \geq 1, \quad z_n &>& \dfrac{1}{2} \, \big( z_{n-1} + z_{n+1}\big).
-\end{array}
-\right.
-\]
-Set the antitone function $\epsilon_1 (y) :=  \frac{2-y}{4}$.
-Then
-\begin{equation}
-\epsilon_1(g_1(x)) = g_1(x) - \dfrac{1}{2} \, \big(g_1(x-1) + g_1(x+1)\big) = \dfrac{1}{2^{x+1}}
-\label{eqn:epsilon g M1}
-\end{equation}
-(Since $g_1$ is bijective, with inverse $g_1^{-1}(y) = 1 - \log (2-y)$, we can obtain $\epsilon_1(y)$ from \Cref{eqn:epsilon g M1}.) 
-%We have $1/2 \cdot \big(g_1(x-1) + g_1(x+1)\big) = 2 - \frac{5}{2^{x+1}}$.
-
+
+For $M_1$, define the function $g_1 : \nnReal \to \nnReal$ by $g_1(x) = \log(x+1) + 1$.
 Using shorthand 
-$\Theta_1 = \tY \, f \, n \, . \, 
+$\Theta_1 = \tY \, \lambda f n . \, 
 \tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
-we can define a partial antitone ranking function $f_1 : \mathit{Rch}(M_1) \to \nnReal$ with antitone function $\epsilon_1$ defined as follows:
+we can define an antitone partial ranking function $f_1 : \mathit{Rch}(M_1) \rightharpoonup \nnReal$ with antitone function $\epsilon_1$ defined as follows (for $n \in \mathbb N$):
 \begin{align*}
@@
 \\
-\tif{n = 0}{0}{\Theta_1 \, (n - 1) \oplus_{1/2} \Theta_1 \, (n + 1)}
-&\mapsto
-\dfrac{1}{2} \, \big(g_1(n-1) + g_1(n+1)\big)
-\\
-0 &\mapsto 0
-\end{align*}
-
-For $M_3$, we use the function $g_3(x) := 3 - \frac{1}{3^{x-1}}$ and the antitone function $\epsilon_3(y) := \frac{84-28y}{27}$. 
-Then
-\[
-\epsilon_3(g_3(x)) = g_3(x) - \dfrac{1}{3}\big(2 \cdot g_3(x-1) + \cdot g_3(x+2) \big) = 
-\frac{28}{3^{x+2}}.
-\]
-Following $M_1$ above, we can define a partial antitone ranking function $f_3 : \mathit{Rch}(M_3) \to \nnReal$ with $\epsilon_3$ as the antitone function.
+0 &\mapsto 0.
+\end{align*}
+For $n \geq 1$, $\Theta_1 \, n$ reduces in several steps to either $\Theta_1 \, (n-1)$ or $\Theta_1 \, (n+1)$, each with probability $1/2$, with one $\tY$-reduction.
+\begin{align*}
+  & g_1(n) - \frac{g_1(n-1) + g_1(n+1)} 2 \\
+  = & \log(\frac{n+1}{\sqrt{n(n+2)}}) \\
+  = & \frac 1 2 \log(\frac{(n+1)^2}{n(n+2)}) \\
+  = & \frac 1 2 \log(1 + \frac 1 {n(n+2)}) \\
+  > & \frac 1 {4n(n+2)} \\
+  > & \frac 1 2 \frac 1 {5(n+1)^2} + \frac 1 2 \frac 1 {5(n+3)^2}
+\end{align*}
+and $\Theta_1 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_1$ of $1$, therefore by setting $\epsilon_1(x) = \frac 1 {5(e^{x-1}+1)^2}$, the condition on how much $g_1$ must decrease is met.
+It is also defined at $M_1 = \Theta_1 \, 10$, and is non-negative, therefore it is an antitone partial ranking function, and $M_1$ is AST.
 

### 2020-12-12 ranking-supermartingales.tex
minor

@@
 For a stopping time $T$ and $\epsilon > 0$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.
-\citep{DBLP:conf/popl/FioritiH15,DBLP:conf/popl/ChatterjeeFNH16}
 \end{definition}
+
+\lo{N.B.~The original definition, due to \citep{DBLP:conf/popl/FioritiH15}, does not mention stopping time explicitly:
+$(Y_n)_{n \in \omega}$ is a \emph{ranking} supermartingale if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot 1_{\set{Y_n > 0}}$.}
 
@@
 Let $T$ and $T'$ be stopping times adapted to $(\calF_n)_{n \in \omega}$.
-Recall the $\sigma$-algebra (consisting of measurable subsets ``priori to $T$'')
+Recall the $\sigma$-algebra (consisting of measurable subsets ``prior to $T$'')
 \[

### 2020-12-15 probabilisticProgrammingMartingales.tex
Fix examples of partial antitone ranking functions.

--- 2020-12-11 probabilisticProgrammingMartingales.tex
@@
 \begin{enumerate}
+\item 1D biased (towards 0) random walk
+\[
+M_1 = 
+\big(\tY \, \lambda f n . \, 
+\tif{n = 0}{0}{f \, (n - 1) \oplus_{3/2} f \, (n + 1)}\big) \, 10
+\]
+
 \item 1D unbiased random walk
 \[
-M_1 = 
-\big(\tY \, f \, n \, . \, 
+M_2 = 
+\big(\tY \, \lambda f n . \, 
 \tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}\big) \, 10
@@
 \[
-M_2 = 
-\big(\tY \, f \, n \, . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/3} f \, (n + 1)}\big) \, 10
-\]
-
-\item 1D biased (towards 0) random walk
-\[
 M_3 = 
-\big(\tY \, f \, n \, . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{2/3} f \, (n + 2)}\big) \, 10
+\big(\tY \, \lambda f n . \, 
+\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/3} f \, (n + 2)}\big) \, 10
 \]
@@
 
-\paragraph{Problem.} Find a generalised notion of ranking function 
-%--- call it \emph{antitone-convex rankable} --- 
-so that $M_1$ becomes rankable, and then prove it sound i.e.~rankable in this generalised sense implies AST.
+The term $M_1$ is rankable, terminating in 31 $\tY$-steps on average, and $M_3$ only has a $1/1024$ chance of terminating, but in between, $M_2$ is AST, but isn't $\tY$-PAST, therefore it isn't rankable and Theorem \ref{thm:RankingSoundness} is insufficient to prove its termination.
+We therefore want to find a generalised notion of ranking function so that $M_2$ becomes rankable, and then prove it sound i.e.~rankable in this generalised sense implies AST.
 
@@
 The definition of antitone-convex ranking function $f$ (say) for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
-we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) and convex function $\epsilon : \pReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
+we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) function $\epsilon : \nnReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
 \[
@@
 \begin{definition}
-We say that a supermartingale $(Y_n)_{n \in \omega}$ and a stopping time $T$ \akr{Separating the stopping time from $Y_n = 0$ allows for a neater definition of the ranking function of a term} \lo{Nice!} adapted to filtration $(\calF_n)_{n \in \omega}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone measurable function $\epsilon : \nnReal \to \pReal$ satisfying
+We say that a supermartingale $(Y_n)_{n \in \omega}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \in \omega}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
 \[
@@
 Then $T < \infty$ almost surely.
-\akr{It's not actually true that $\expect{T} \leq \dfrac{\expect{Y_0}}{\epsilon_0}$ where $\epsilon_0 := \epsilon(\expect{Y_0})$.
-This is fortunate because it means that a rankability theorem based on antitone strict supermartingales actually has a chance of being stronger than the existing one (which was pretty much complete for $\tY$-PAST terms already).
-In fact, for any almost surely finite stopping time, there is a corresponding antitone strict supermartingale.
-This seems promising for the method being conditionally complete (for terms where all reachable terms are AST).}
-\lo{See \Cref{sec:Relativised completeness}.}
 \end{theorem}
@@
 
-Programs $M_1$ and $M_3$ are AST.
-To construct antitone ranking functions, we appeal to a partial ranking function theorem for antitone ranking functions.
-
-For $M_1$, define the function $g_1 : \nnReal \to \nnReal$ by $g_1(x) = \log(x+1) + 1$.
+Programs $M_1$ and $M_2$ are AST.
+For $M_1$, the partial ranking function $\big(\tY \, \lambda f n .
+\, \tif{n=0}{0}{f \, (n-1) \oplus_{3/2} f \, (n+1)}\big) \, \underline x \mapsto 3x + 1$ suffices to prove its termination (and could equivalently be considered an antitone partial ranking function with the constant antitone function $\epsilon_1(x) = 1$).
+
+For $M_2$, define the function $g_1 : \nnReal \to \nnReal$ by $g_1(x) = \log(x+1) + 1$.
 Using shorthand 
-$\Theta_1 = \tY \, \lambda f n . \, 
+$\Theta_2 = \tY \, \lambda f n . \, 
 \tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
-we can define an antitone partial ranking function $f_1 : \mathit{Rch}(M_1) \rightharpoonup \nnReal$ with antitone function $\epsilon_1$ defined as follows (for $n \in \mathbb N$):
-\begin{align*}
-{\Theta_1} \, n 
+we can define an antitone partial ranking function $f_2 : \mathit{Rch}(M_2) \rightharpoonup \nnReal$ with antitone function $\epsilon_2$ defined as follows (for $n \in \mathbb N$):
+\begin{align*}
+{\Theta_2} \, n 
 &\mapsto 
-g_1(n)
+g_2(n)
 \\
@@
 \end{align*}
-For $n \geq 1$, $\Theta_1 \, n$ reduces in several steps to either $\Theta_1 \, (n-1)$ or $\Theta_1 \, (n+1)$, each with probability $1/2$, with one $\tY$-reduction.
-\begin{align*}
-  & g_1(n) - \frac{g_1(n-1) + g_1(n+1)} 2 \\
+For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$, each with probability $1/2$, with one $\tY$-reduction.
+\begin{align*}
+  & g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 \\
   = & \log(\frac{n+1}{\sqrt{n(n+2)}}) \\
@@
 \end{align*}
-and $\Theta_1 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_1$ of $1$, therefore by setting $\epsilon_1(x) = \frac 1 {5(e^{x-1}+1)^2}$, the condition on how much $g_1$ must decrease is met.
-It is also defined at $M_1 = \Theta_1 \, 10$, and is non-negative, therefore it is an antitone partial ranking function, and $M_1$ is AST.
+and $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$, therefore by setting $\epsilon_2(x) = \frac 1 {5(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met.
+It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone partial ranking function, and $M_2$ is AST.
 
@@
 \underbrace{\big
-(\tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample))} \big)}_{\Theta} \, 10
-\]
-We can construct a partial antitone ranking function $f$ as follows:
+(\tY \, \lambda f x . \tif{x \leq 0}{0}{f(x - \tsample))} \big)}_{\Theta} \, 10
+\]
+We can construct a partial ranking function $f$ as follows:
 \begin{align*}
@@
 &\mapsto 
-l
+l + 2
 \\
@@
 &\mapsto
-l - 1/2
+l + 1
 \\
-0 &\mapsto 0
-\end{align*}
-with the constant antitone function $\epsilon(y) = 1/2$.
-
-For a more complex example, consider the following ``continuous random walk''
+0 &\mapsto 0.
+\end{align*}
+
+For a more complex (not $\tY$-PAST) example, consider the following ``continuous random walk''
 \[
 \underbrace{\big
-(\tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample) \oplus_{1/2} f(x + \tsample)} \big)}_{\Xi} \, 10
-\]
-Let $g(x) := \sqrt{2} - \frac{1}{2^{x-\frac{1}{2}}}$ and let $\epsilon$ be the function specified by
-\[
-\epsilon(g(x)) = g(x) - \dfrac{1}{2}\Big(g(x - \dfrac{1}{2}) + \cdot g(x + \dfrac{1}{2}) \Big) = 
-\dfrac{3 - 2\sqrt{2}}{2^{x+1}}.
-\]
+(\tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample + 1/2)} \big)}_{\Xi} \, 10
+\]
+Let $g(x) := 2 + log(x + 1)$ again and let $\epsilon$ be the function specified by
+\[
+\epsilon(g(x-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y.
+\]
+The limit of this as $x \to \infty$ and $g(x+1/2) \to \infty$ is 0, and $\frac {\mathrm d}{\mathrm dx} \epsilon(g(x+1/2)) = \frac 1 {x+1} + \log(1 - \frac 1 {x + 1/2}) < 0$, and $g$ is monotonic increasing therefore $\epsilon$ is antitone and bounded below by 0.
 We define a partial antitone ranking function as follows:
@@
 \\
-\tif{l \leq 0}{0}{\Xi \, (l - \tsample) \oplus_{1/2} \Xi \, (l + \tsample)}
-&\mapsto
-\frac{1}{2}\big(g(l - \frac{1}{2}) + \cdot g(l + \frac{1}{2}) \big)
-\\
-r \leq 0, \quad r &\mapsto 0
+0 &\mapsto 0
 \end{align*}
 \end{example}
+The value of $g$ after one $\tY$-reduction step is at least $g(l-1/2)$ therefore the expectation of $\epsilon$ after one $tY$-reduction step is at most $\epsilon(g(l-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y$, therefore in the case where $l > 0$, $g$ decreases by the required amount, and in the case where $l \leq 0$, $g(\Xi \, l) \geq 2 + log(1/2) > \log(\frac{3 \sqrt 3} 2 - 1 = \epsilon(g(0-1/2))$ as well, therefore this is a valid partial antitone ranking function and the term is AST.
 
@@
 \]
-It turns out the solution to Example $M_1$, $z_n = \frac{2^n - 1}{2^{n-1}}$, is nearly a solution (for $n = 1$, we have equality instead).
-
-Let $g(x) = \frac{2^n - 1}{2^{n-1}}$.
-Then 
-\begin{equation}
-\epsilon(g(x)) = 
-g(x) - \dfrac{x}{2x + 1} \, g(x-1) - \dfrac{x+1}{2x + 1} \, g(x+1) 
-= \dfrac{x-1}{2^{x} \, (2x + 1)}
-\label{eqn:in the limit}
-\end{equation}
-
-\lo{ongoing}
+For $n > 2$, $z_n = \log(n-1)$ works.
+The expected decrease is $\frac 1 2(\log(1+\frac 1 {(n-1)^2-1}) - \frac 1 {2n+1} \log(1 + \frac 2 {n-2}))$, and using the fact that $\frac x {x+1} \leq \log (1+x) \leq x$ for $x > 0$, this is at least $\frac 1 2(\frac 1 {(n-1)^2} - \frac 1 {2n + 1} \frac 2 {n-2}) = \frac {n^2}{2(n-1)^2(2n+1)(n-2)}$, which (again for $n > 2$) is positive and antitone.
+For $n = 0, 1, 2$, we take $\epsilon$ to be 9/40 (the same as its value at 3), then set $z_2 = 2 \log 2 - \log 3 - 1, z_1 = 3 \log 2 - 2 \log 3 - 3, z_0 = 4 \log 2 - 3 \log 3 - 6$.
+Some of those values are negative, but it's still bounded below, so by adding a constant offset it can be corrected, and the term is AST.
 

### 2020-12-18 probabilisticProgrammingMartingales.tex
Sec. 2

--- 2020-12-12 probabilisticProgrammingMartingales.tex
@@
 \paragraph{}
-This definition of almost sure termination is equivalent to that given in \citep{MakOP20b}, although the program semantics is stated in a slightly different way.
+This definition of almost-sure termination is equivalent to that given in \citep{MakOP20b}, although the program semantics is stated in a slightly different way.
 In particular, the argument to $\tscore$ is not relevant to termination (except for the possibility that its argument's evaluation wouldn't terminate).
 
+\iffalse
 \lo{Your operational semantics does not maintain a record of the current weight of the reduction.
@@
 you should do it as a future task.}
+\fi
 
@@
 These variants are defined as functions from reachable terms (i.e.~possible states of the program's execution) to real numbers.
-Specifically, let the set of reachable terms from a given closed starting term $M$ be $Rch(M) = \{N \in \Lambda \mid M \to^* N \}$, with the $\sigma$-algebra induced as a subset of $\Lambda$.
+Specifically, let the set of reachable terms from a given closed starting term $M$ be $Rch(M) := \{N \in \Lambda \mid M \to^* N \}$, with the $\sigma$-algebra induced as a subset of $\Lambda$.
 
@@
 Any closed term for which a ranking (respectively, strict ranking) function exists is called \emph{rankable} (respectively, \emph{strictly rankable}). 
-It will be demonstrated later that for any rankable term $M$, if $(M_n)_n$ is the reduction sequence starting from $M$, then $(f(M_n))_n$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales.
+\changed[lo]{For example, the term $(\tY \lambda x.x) \, \underline 0$ is not rankable.}
+It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \in \omega}$ is the reduction sequence starting from $M$, then $(f(M_n))_{n\in\omega}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales.
 \end{definition}
@@
 \begin{equation}
-\tif{{-}\tsample < 0}{\underline{0}}{(\tY \lambda x. x) \; \underline 0}
+\tif{{-}\tsample < 0}{\underline{0}}{(\tY \lambda x. x) \, \underline 0}
 \label{ex:0 probability reachable}
 \end{equation}
-terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x. x) \underline 0$ is reachable, although that has probability 0. 
-Not only is this counterexample AST, it's positively almost surely terminating i.e.~the expected time to termination is finite.
+terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x. x) \, \underline 0$ is reachable, although that has probability 0. 
+Not only is this counterexample AST, it's PAST. %positively almost surely terminating i.e.~the expected time to termination is finite.
 
@@
 The restrictions of $f$ and $g$ to $Rch(N)$ have the same properties assumed of $f$ and $g$, so assume w.l.o.g.~that $N=M$.
-The difference $g - f$ is then a supermartingale (with the same setup as in the proof of \Cref{thm:rankable gives supermartingale}
+The difference $g - f$ is then a supermartingale (with the same setup as in the proof of %\Cref{thm:rankable gives supermartingale}
 %Theorem \ref{rankable implies ast}) 
-therefore $\forall n.\ \mathbb E[g(M_n)] \leq \mathbb E[f(M_n)] + g(M)-f(M)$.
-
-$\mathbb E[f(M_n)] = \sum_{k=n}^\infty \mathbb P[M_n = E[\tY N] \text{ for some }E, N] \to 0 \text{ as } n \to \infty$ therefore as $g(M) - f(M) < 0$, eventually $\mathbb E[g(M_n)] < 0$, which is impossible, therefore $g \geq f$ as required.
+\Cref{thm:rankable and strict rankable});
+therefore $\mathbb E[g(M_n)] \leq \mathbb E[f(M_n)] + g(M)-f(M)$, for all $n$.
+
+We have $\mathbb E[f(M_n)] = \sum_{k=n}^\infty \mathbb P[M_n = E[\tY N] \text{ for some }E, N] \to 0 \text{ as } n \to \infty$; 
+therefore as $g(M) - f(M) < 0$, eventually $\mathbb E[g(M_n)] < 0$, which is impossible. 
+It follows that $g \geq f$ as required.
 

### 2020-12-18 ranking-supermartingales.tex
Sec. 2

--- 2020-12-01 ranking-supermartingales.tex
@@
 \lo{The preceding means: $\forall A \in \calF_n$, $\int_A \dif \mu \, Y_{n+1} \leq \int_A \dif \mu \,Y_n$.}
-For a stopping time $T$ and $\epsilon > 0$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.
-\citep{DBLP:conf/popl/FioritiH15,DBLP:conf/popl/ChatterjeeFNH16}
+%For a stopping time $T$ and $\epsilon > 0$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.
+\changed[lo]{For $\epsilon > 0$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$ where $T$ (a stopping time) is defined as
+\[T(s) := \min \set{n \mid Y_n(s) = 0}.\]}
+A \emph{ranking supermartingale} is a $\epsilon$-ranking supermartingale for some $\epsilon > 0$.
 \end{definition}
+
+\lo{N.B.~The original definition, due to \citep{DBLP:conf/popl/FioritiH15}, does not mention stopping time explicitly:
+they call a supermartingale $(Y_n)_{n \in \omega}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot 1_{\set{Y_n > 0}}$.}
 
@@
 \paragraph{Proof of \Cref{thm:rankable and strict rankable}.}
-\emph{Notation}. Given an $\omega$-sequence (e.g.~$s \in S$) and $m \in \omega$, we write $s_{\leq m} \in I^m$ to mean the $m$-long prefix of $s$.
+\emph{Notation}. Given an $\omega$-sequence (e.g.~$s \in S$) and $m \in \omega$, we write $s_{\leq m} \in I^m$ to mean the prefix of $s$ of length $m$.
 
@@
 \end{align*}
- so that $\pi_1 (\red^n(M, s)) = \underbrace{\pi_t( \cdots (\pi_t}_{\ndraw n s}(s))$, and the filtration $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
-The $\calF_n$-measurability of $M_n$ (and hence of $\#\mathrm{draw}_n$) follows from \citep{DBLP:conf/icfp/BorgstromLGS16};
-$\#\mathrm{draw}_n$ is a stopping time (bounded by $n$).
+ so that $\pi_1 (\red^n(M, s)) = \overbrace{\pi_t( \cdots (\pi_t}^{\ndraw n s}(s))$, and the filtration $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
+The $\calF_n$-measurability of $M_n$ (and hence of $\#\mathrm{draw}_n$) follows from \citep{DBLP:conf/icfp/BorgstromLGS16}.
+%$\#\mathrm{draw}_n$ is a stopping time (bounded by $n$). 
 \akr{$\#\mathrm{draw}_n$ is not a stopping time.
-It doesn't look like you actually use this claim anyway, so it should just be fine to remove, but did you mean something different?}
+It doesn't look like you actually use this claim anyway, so it should just be fine to remove, but did you mean something different?} \lo{I agree, and I don't actually use this claim.}
 Take $s \in A \in \calF_n$ with $\ndraw{n}{s} = l$.
@@
 
-We say that a given SPCF term is \emph{type-1} (respectively 2, 3 and 4) if it has the shape $E[\tY \lambda x. N]$ (respectively $E[\tsample]$, $E[R]$ where $R$ is any other redex, and of a value).
-Define $\mathbf{T}_i := 
-\set{s \mid M_n(s) \hbox{ is type-$i$}}$.
+We say that a given SPCF term is: %\emph{type-1} (respectively 2, 3 and 4) if it has the shape $E[\tY \lambda x. N]$ (respectively $E[\tsample]$, $E[R]$ where $R$ is any other redex, and of a value).
+\emph{type-1} if it has the shape $E[\tY \lambda x.
+N]$, \emph{type-2} if it has the shape $E[\tsample]$; \emph{type-3} if it has the shape $E[R]$ where $R$ is any other redex, and \emph{type-4} if it is a value.
+Henceforth fix an $n \geq 0$, and define $\mathbf{T}_i := \set{s \mid M_n(s) \hbox{ is type-$i$}}$.
 It is straightforward to see that each $\mathbf{T}_i \in \calF_n$, and $\set{\mathbf{T}_1, \mathbf{T}_2, \mathbf{T}_3, \mathbf{T}_4}$ is a partition of $S$.
@@
 \[
-\int_A \mu(\dif s) \, f_{n+1}(M)[s \in \mathbf{T}_i] \leq \int_A \mu(\dif s) \, f(M_n)[s \in \mathbf{T}_i]
+\int_A \mu(\dif s) \, f(M_{n+1})[s \in \mathbf{T}_i] \leq \int_A \mu(\dif s) \, f(M_n)[s \in \mathbf{T}_i]
+\; \footnote{where (the Iverson bracket) $[P] := 1$ if the statement $P$ hold, and 0 otherwise.}
 \] 
-where $[s \in \mathbf{T}_i]$ is the Iverson bracket.
 \end{lemma}
@@
 \begin{equation}
-f_{n+1}(M)[s \in \mathbf{T}_2] = \sum_{i \in \calI} 
+f(M_{n+1})[s \in \mathbf{T}_2] = \sum_{i \in \calI} 
 f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])[s \in U_i]
@@
 \item $\calI$ is a countable indexing set
-\item $E_i[\cdot][\tsample] \in \mathsf{Sk}_{j_i}$, and $\sigma_i : S \to \Real^{j_i}$, and $\rho(s) = \pi_h(\pi_1(\red^n(M, s))) \in \Real$ 
+\item $E_i[\cdot][\tsample] \in \mathsf{Sk}_{j_i}$, and $\sigma_i : S \to \Real^{j_i}$, and $\rho(s) := \pi_h(\pi_1(\red^n(M, s))) = \pi_h(\pi_t^{l_i}(s)) \in \Real$ 
 \item $\set{U_i}_{i \in \calI}$ is a partition of $\mathbf{T}_2$, 
-where each $U_i$ is determined by a skeletal term $E_i[\cdot][\tsample]$ and a number (of draws) $l_i \leq n$;
-precisely 
+where each $U_i$ is determined by a skeletal term $E_i[\cdot][\tsample]$ and a number (of draws) $l_i \leq n$, satisfying
 \[
+s \in U_i
+\iff
+(E_i[\underline{r}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s),
+\;
+\hbox{for some }r \in \Real^{j_i}
+\]
+(We use the (measurable) function $\sigma_i : S \to \Real^{j_i}$ to skolemise the existentially quantified $r$.)
+
+Equivalently
+\(
 U_i := \#\mathrm{draw}_n^{-1}(l_i) \cap M_n^{-1}(E_i[\cdot][\tsample], \Real^{j_i}) \in \calF_n.
-\]
-\lo{I.e.~$(E_i[\underline{\sigma_i(s)}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s)$ iff $s \in U_i$.}
+\)
+%\lo{I.e.~$(E_i[\underline{\sigma_i(s)}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s)$ iff $s \in U_i$.}
 \end{itemize}
@@
 %Moreover, for any $A \in \calF_n$, if $s \in A$ then $\set{s_{\leq l}} \cdot I^\omega \subseteq A$.
-This means that for any measurable $g : S \to \Real_{\geq 0}$, if $g(s)$ only depends on the $(l_i+1)$-long prefix of $s$, then, writing $g' : I^{l_i+1} \to \Real_{\geq 0}$ where $g(s) = g'(s_{\leq l_i+1})$
+This means that for any measurable $g : S \to \Real_{\geq 0}$, if $g(s)$ only depends on the prefix of $s$ of length $(l_i+1)$, then, writing $\widehat{g} : I^{l_i+1} \to \Real_{\geq 0}$ where $g(s) = \widehat{g}(s_{\leq l_i+1})$, we have: for any $A \in \calF_n$
 \begin{equation}
 \int_{A \cap U_i}  \mu(\dif s) \, g(s) = 
-\int_{(A \cap U_i)_{\leq l_i+1}} \Leb_{l_i+1}(\dif t) \, g'(t)
+\int_{(A \cap U_i)_{\leq l_i+1}} \Leb_{l_i+1}(\dif t) \, \widehat{g}(t)
 \label{eqn:truncate}
 \end{equation}
-(For simplicity, we will write $g' = g$ in the following.)
+%(For simplicity, we will write $g' = g$ in the following.)
 %where $A_{\leq m} := \set{s_{\leq m} \mid s \in A}$.
@@
 Take $s \in U_i$, and set $l = l_i$.
-Now $\sigma_i(s)$ depends on $s_{\leq l}$, and $\rho(s)$ depends on $s_{\leq l +1}$.
-Let $u$ range over $(U_i)_{\leq l}$.
-Then, it follows from the definition of $f$ that
+Plainly $\sigma_i(s)$ depends on $s_{\leq l}$, and $\rho(s)$ depends on $s_{\leq l +1}$.
+Take $u \in (U_i)_{\leq l}$.
+It then follows from the definition of ranking function that
 \[
-\int_I \Leb(\dif r) \, f(E_i[\underline{\sigma_i(u)}][\underline{r}])) \leq f(E_i[\underline{\sigma_i(u)}][\tsample])).
+\int_I \Leb(\dif r) \, f(E_i[\underline{\widehat{\sigma_i}(u)}][\underline{r}])) \leq f(E_i[\underline{\widehat{\sigma_i}(u)}][\tsample])).
 \]
@@
 \begin{align*}
-& \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, \int_I \textrm{Leb}(\dif r) \, f(E_i[\underline{\sigma_i(u)}][\underline{r}])) \\
-\leq & \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, f(E_i[\underline{\sigma_i(u)}][\tsample]))
+& \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, \int_I \textrm{Leb}(\dif r) \, f(E_i[\underline{\widehat{\sigma_i}(u)}][\underline{r}])) \\
+& \leq \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, f(E_i[\underline{\widehat{\sigma_i}(u)}][\tsample])).
 \end{align*}
@@
 \begin{align*}
-& \int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\underline{\sigma_i(u')}][\underline{\rho(u')}])) \\
+& \int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\underline{\widehat{\sigma_i}(u')}][\underline{\widehat{\rho}(u')}])) \\
 & \leq 
-\int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\underline{\sigma_i(u')}][\tsample]))
+\int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\underline{\widehat{\sigma_i}(u')}][\tsample]))
 \end{align*}
-and so 
+and so, by (\ref{eqn:truncate})
 \begin{equation}
@@
 
-Finally, integrating both sides of (\ref{eqn:f 2 n+1}), we have
+Now, integrating both sides of (\ref{eqn:f 2 n+1}), we have
 \begin{align*}
@@
 \fi
-This concludes the proof of \Cref{thm:rankable gives supermartingale}. \hfill \qed
+This concludes the proof of %\Cref{thm:rankable gives supermartingale}. 
+\Cref{thm:rankable and strict rankable}. \hfill \qed
 
@@
 As before, fix a $M \in \Lambda^0$.
-Define random variables on $(S, \calF, \mu)$:
+Define random variables on the probability space $(S, \calF, \mu)$:
 \begin{align}
@@
 The random variable $T_M^\tY$ is known as the \emph{$\tY$-runtime of $M$}, and it can equivalently be defined as the number of $\tY$-reduction steps in the reduction sequence of $M$.
-Note that, as the type system ensures that the reduction relation excluding $\tY$-reduction is strongly normalising, only finitely many reductions can occur in a row without one of them being a $\tY$-reduction, therefore $T_M^\tY < \infty$ iff $M$ is AST.
+Note that, as the type system ensures that the reduction relation excluding $\tY$-reduction is strongly normalising, only finitely many reductions can occur in a row without one of them being a $\tY$-reduction, therefore $T_M^\tY < \infty$ \changed[lo]{a.s.}~iff $M$ is AST.
 

### 2020-12-18 probabilisticProgrammingMartingales.tex
Finish antitone ranking examples and discuss completeness a bit.

--- 2020-12-15 probabilisticProgrammingMartingales.tex
@@
 \paragraph{\Cref{ex:ac-ranking} revisited}
+\akr{All of these ranking functions are slightly incorrect, because $\Xi \underline n$ reduces to something like $\Xi (\underline n + 1)$ then $(\lambda z.
+\text{something}[\Xi/f]) (\underline n + 1)$, rather than reducing the argument first.
+This could be fixed by appealing to the confluent semantics version of the antitone ranking theorem, but it would be nice to have at least some examples here illustrating how antitone ranking functions work, rather than actually saving all these examples until the end.
+
+Even though the difference is quite trivial, this was actually the original motivating example for the confluent trace semantics.}
 
@@
 \]
-Let $g(x) := 2 + log(x + 1)$ again and let $\epsilon$ be the function specified by
+Let $g(x) := 2 + \log(x + 1)$ again and let $\epsilon$ be the function specified by
 \[
@@
 \end{example}
-The value of $g$ after one $\tY$-reduction step is at least $g(l-1/2)$ therefore the expectation of $\epsilon$ after one $tY$-reduction step is at most $\epsilon(g(l-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y$, therefore in the case where $l > 0$, $g$ decreases by the required amount, and in the case where $l \leq 0$, $g(\Xi \, l) \geq 2 + log(1/2) > \log(\frac{3 \sqrt 3} 2 - 1 = \epsilon(g(0-1/2))$ as well, therefore this is a valid partial antitone ranking function and the term is AST.
+The value of $g$ after one $\tY$-reduction step is at least $g(l-1/2)$ therefore the expectation of $\epsilon$ after one $tY$-reduction step is at most $\epsilon(g(l-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y$, therefore in the case where $l > 0$, $g$ decreases by the required amount, and in the case where $l \leq 0$, $g(\Xi \, l) \geq 2 + \log(1/2) > \log(\frac{3 \sqrt 3} 2) - 1 = \epsilon(g(0-1/2))$ as well, therefore this is a valid partial antitone ranking function and the term is AST.
 
@@
 \underbrace{\big
-(\tY \, f \, x . 
-\tif{x \leq 0}{0}{0 \oplus_{\frac{1}{x+1}} f(x + 1)} \big)}_{\Xi} 
+(\tY \, f \, x . \,
+0 \oplus_{\frac{1}{x+1}} f(x + 1) \big)}_{\Xi} 
 \, 10
 \]
+
+In this case, the fact that the ranking function must decrease at each $\tY$-step (in an antitone partial ranking function) by the expected value of $\epsilon$ not at the current term, but at the next term where the ranking function is defined, is a little harder to deal with, because the variable $x$ can change all the way to $0$ in one step, therefore simply adding a small offset doesn't suffice to compensate for this fact.
+
+Consider the candidate ranking function $\Xi n \mapsto n + 1$.
+For each $n$, $\Xi n$ reduces to either $0$ (with probability $\frac 1 {n + 1}$) or $\Xi (n + 1)$, therefore the expected value of the ranking function is $\frac{n(n+2)}{n+1} = n + 1 - \frac 1 {(n+1)^2}$, and the required decrease is $\frac{\epsilon(0) + n \epsilon(n + 2)}{n + 1}$, therefore eventually, the expected decrease isn't enough, whatever the value of $\epsilon(0)$ is.
+
+If instead, we take the ranking function to be defined after the $\tY$-reduction but before the $\tsample$-reduction as well, this can be resolved.
+Letting $\Theta[n] = \underline 0 \oplus_{\frac 1 n} \Xi n$:
+\begin{align*}
+\Theta[n] \mapsto n \\
+\Xi 10 \mapsto 12.
+\end{align*}
+For each $n$, $\Theta[n]$ reduces to either $0$ or $\Theta[n+1]$, with a $tY$-reduction only in the latter case, and the condition that this is an antitone partial ranking function is that $n \geq \frac{(n-1)(n+1)}{n} + \frac{n-1}{n} \epsilon(n+1)$ and $12 \geq 11 + \epsilon(11)$.
+These are satisfied by setting $\epsilon(x) = \min(1, 1/x)$, therefore this term is AST.
 \end{example}
 
+\paragraph{Completeness}
+It seems very likely that this method is complete in the case that every reachable term is AST, but we have been unable to actually prove this.
+It is certainly at least capable of proving the termination of terms which terminate arbitrarily slowly (in the sense that there is no similar limitation to Theorem \ref{thm:rankable implies termination}, which can only prove termination of $\tY$-PAST terms).
+
+The following theorem does not prove completeness, but is suggestive in that direction:
+\begin{theorem}
+For any stopping time $T$ which is almost surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
+\end{theorem}
+\akr{TODO: Move proof of this theorem here, and edit it so that it's actually proving this, rather than trying to prove something stronger.}
 \input{y-stopping-time}

### 2020-12-18 probabilisticProgrammingMartingales.tex

--- 2020-12-15 probabilisticProgrammingMartingales.tex
@@
 \subsection{Partial ranking functions}
-Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost sure termination, so it isn't useful for proving it.
+Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost-sure termination, so it isn't useful for proving it.
 Take, for example, the term
@@
 &\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
-&) \underline{0}
+&) \, \underline{0}
 \end{align*}

### 2020-12-18 ranking-supermartingales.tex
Finish antitone ranking examples and discuss completeness a bit.

@@
 %For a stopping time $T$ and $\epsilon > 0$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.
-\changed[lo]{For $\epsilon > 0$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$ where $T$ (a stopping time) is defined as
-\[T(s) := \min \set{n \mid Y_n(s) = 0}.\]}
+\changed[lo]{For $\epsilon > 0$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$ where
+\(T(s) := \min \set{n \mid Y_n(s) = 0}\).
+It is straightforward to see that $T$ is a stopping time w.r.t.~$(\calF_n)_{n \in \omega}$.}
 A \emph{ranking supermartingale} is a $\epsilon$-ranking supermartingale for some $\epsilon > 0$.
@@
 \begin{proof}
-For 1, observe that $T_0, T_1, T_2, \cdots$ are an increasing sequence of stopping times adapted to $(\calF_n)_{n \in \omega}$, and each $T_i$ is bounded.
+For 1, observe that $T_0, T_1, T_2, \cdots$ are an increasing sequence of stopping times adapted to $(\calF_n)_{n \in \omega}$, and each $T_i$ is bounded.\lo{Prove it.}
 Thanks to \Cref{thm:rankable and strict rankable}, 

### 2020-12-18 probabilisticProgrammingMartingales.tex
Sec. 3: ongoing

@@
 
-We have $\mathbb E[f(M_n)] = \sum_{k=n}^\infty \mathbb P[M_n = E[\tY N] \text{ for some }E, N] \to 0 \text{ as } n \to \infty$; 
+We have $\mathbb E[f(M_n)] = \sum_{k=n}^\infty \mathbb P[M_k = E[\tY N] \text{ for some }E, N] \to 0 \text{ as } n \to \infty$; 
 therefore as $g(M) - f(M) < 0$, eventually $\mathbb E[g(M_n)] < 0$, which is impossible. 
@@
 Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost-sure termination, so it isn't useful for proving it.
-Take, for example, the term
+
+\begin{example}[Geometric distribution]
+Consider the term
 \begin{align*}
@@
 which is not the official definition.}
+\end{example}
 
@@
 With this abbreviation, the geometric distribution example from earlier can be written as $(\tY f, n.
-n \oplus_{0.5} f(n+1)) \underline 0$.
+n \oplus_{0.5} f(n+1)) \, \underline 0$.
 It is then easy to see that the following is a partial ranking function:

### 2020-12-18 ranking-supermartingales.tex
Sec. 3: ongoing

@@
 %For a stopping time $T$ and $\epsilon > 0$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.
-\changed[lo]{For $\epsilon > 0$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$ where
-\(T(s) := \min \set{n \mid Y_n(s) = 0}\).
-It is straightforward to see that $T$ is a stopping time w.r.t.~$(\calF_n)_{n \in \omega}$.}
+\changed[lo]{For $\epsilon > 0$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$ where the random variable $T$ is defined as
+\(T : s \mapsto \min \set{n \mid Y_n(s) = 0}\).
+(It is straightforward to see that $T$ is a stopping time w.r.t.~$(\calF_n)_{n \in \omega}$.)}
 A \emph{ranking supermartingale} is a $\epsilon$-ranking supermartingale for some $\epsilon > 0$.
@@
 
-\begin{theorem}[Soundness]
+\begin{theorem}[Soundness] \label{thm:rankable implies termination}
 \begin{enumerate}

### 2020-12-20 probabilisticProgrammingMartingales.tex
Sec. 2: def. ranking supermartingale, and Lemma 1.

@@
 
+\lo{NOTATION. To be consistent, we write $\lambda x. M$ rather than $\lambda x : M$.}
+
 \begin{example}[Geometric distribution]
@@
 The definition of rankability is also inconvenient for syntactic sugar.
-It could be useful, for example, to define $M \oplus_p N = \tif{\tsample - p < 0} M N$, where $M \oplus_p N$ reduces to $M$ or $N$, depending on the first value of $s$, with probability $p$ resp.~$(1-p)$.
-Technically though, it reduces first to $\tif{\underline r - p < 0} M N$ for $r \in I$, so those terms all need values of the ranking function too.
+It could be useful, for example, to define $M \oplus_p N = \tif{\tsample - p < 0} M N$, where $M \oplus_p N$ reduces to $M$ or $N$, depending on the first value of $s$\lo{Need a name for $s$.}, with probability $p$ resp.~$(1-p)$.
+Technically though, it reduces first to $\tif{\underline r - p < 0} M N$ for all $r \in I$, so those terms all need values of the ranking function too.
 
@@
 In both of these cases, there are only some values of the ranking function that are semantically important. 
+\begin{definition}
 Define a \emph{partial ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \rightharpoonup \mathbb R$ such that
@@
 \end{itemize}
+\end{definition}
 A partial ranking function that is total is just a ranking function.
@@
 \begin{align*}
-    f_1(N) & = f(N) \text{ whenever $f(N)$ is defined,} \\
-    f_1(V) & = 0 \text{ for values $V$ not in the domain of $f$.}
-\end{align*}
-
-Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) = \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
+    f_1(N) & := f(N) \text{ whenever $f(N)$ is defined,} \\
+    f_1(V) & := 0 \text{ for values $V$ not in the domain of $f$.}
+\end{align*}
+
+Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) := \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
 The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on partial ranking functions.
@@
 
-\begin{example}
+\begin{example}[Random walk]
 \label{ex:ac-ranking}
@@
 
-The term $M_1$ is rankable, terminating in 31 $\tY$-steps on average, and $M_3$ only has a $1/1024$ chance of terminating, but in between, $M_2$ is AST, but isn't $\tY$-PAST, therefore it isn't rankable and Theorem \ref{thm:RankingSoundness} is insufficient to prove its termination.
+The term $M_1$ is rankable, terminating in 31 $\tY$-steps on average, and $M_3$ only has a $1/1024$ chance of terminating, but in between, $M_2$ is AST, but isn't $\tY$-PAST, therefore it isn't rankable and \Cref{thm:rankable implies termination} is insufficient to prove its termination.
 We therefore want to find a generalised notion of ranking function so that $M_2$ becomes rankable, and then prove it sound i.e.~rankable in this generalised sense implies AST.
@@
 \end{definition}
+
+\lo{The preceding definition does not place any constraint on the stopping time $T$ (other than that it is w.r.t.~the filtration $(\calF_n)_{n \in \omega}$).
+However for the definition to make sense, $T$ must be the stopping time $T(s) := \min \set{n \mid Y_n(s) = 0}$. So I propose the following.}
 
@@
 \begin{proof}
-First, as $(Y_n)$ is a supermartingale, $\expect{Y_n} \leq \expect{Y_0}$, therefore
+First, as $(Y_n)$ is a supermartingale, $\expect{Y_n} \leq \expect{Y_0}$. 
+Therefore
 
@@
 \expect{Y_n \mid T > n}
-  \step[=]{}
-\dfrac{\mathbb P[T > n] \expect{Y_n \mid T > n} + \mathbb P[T \leq n] \expect{Y_n \mid T \leq n} - \mathbb P[T \leq n] \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
-  \step[=]{}
-\dfrac{\expect{Y_n} - \mathbb P[T \leq n] \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
+  \step[=]{rearranging terms}
+\dfrac{\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
++ 
+\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
+- 
+\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
+  \step[=]{definition of conditional expectation}
+\dfrac{\expect{Y_n} - \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
   \step[\leq]{$Y_n \geq 0$ always}
@@
 
-\emph{Claim}: for all $0 < x \leq 1$, let $B_x = \left \lceil \dfrac{\expect{Y_0}+1}{x \; \epsilon(\expect{Y_0} \, x^{-1})} \right \rceil$, then
+\emph{Claim}: For all $0 < x \leq 1$, 
 \(
-\mathbb P[T > B_x] \leq x.
+\mathbb P[T > B_x] \leq x,
 \)
+where $B_x = \left \lceil \dfrac{\expect{Y_0}+1}{x \; \epsilon(\expect{Y_0} \, x^{-1})} \right \rceil$.
+
+\smallskip
+
 As the convex hull of $\epsilon$ (the greatest convex function less than or equal to it) satisfies all the conditions assumed of $\epsilon$, in addition to being convex, assume wlog that $\epsilon$ is convex.
-Suppose for a contradiction that $\mathbb P[T > B_x] > x$, then for every $n \leq B_x$
-\lo{Change of justification of the second equality below: event, not discrete r.v. 
-Recall: expectation conditioning on r.v.~(resp.~event) is a r.v.~(resp.~number).}
+
+Assume for a contradiction that $\mathbb P[T > B_x] > x$.
+Then, take $n \leq B_x$. 
+We
+%\lo{Change of justification of the second equality below: event, not discrete r.v. Recall: expectation conditioning on r.v.~(resp.~event) is a r.v.~(resp.~number).}
 \begin{calculation}
 \expect{Y_n-Y_{n+1}}
-  \step[=]{definition and linearity of conditional expectation}
+  \step[=]{$\calF_n \subseteq \calF_{n+1}$, definition and linearity of conditional expectation}
 \expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
@@
 \begin{proof}
-As in Theorem \ref{thm:basicRankingSoundness}, take the set $S$ of traces as a probability space, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$.
-As before, $(Y_n)_{n \in \omega}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$, therefore $T_M^{\tY}$ is almost surely finite.
+As in %Theorem \ref{thm:basicRankingSoundness}, 
+\Cref{thm:rankable implies termination},
+take the set $S$ of traces as a probability space, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$.
+As before, $(Y_n)_{n \in \omega}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$.
+Therefore, it follows from \Cref{thm:a-c strict} that $T_M^{\tY}$ is almost-surely finite.
 \end{proof}
@@
 \paragraph{\Cref{ex:ac-ranking} revisited}
-\akr{All of these ranking functions are slightly incorrect, because $\Xi \underline n$ reduces to something like $\Xi (\underline n + 1)$ then $(\lambda z.
+\akr{All of these ranking functions are slightly incorrect, because $\Xi \, \underline n$ reduces to something like $\Xi \, (\underline n + 1)$ then $(\lambda z.
 \text{something}[\Xi/f]) (\underline n + 1)$, rather than reducing the argument first.
@@
 
-Programs $M_1$ and $M_2$ are AST.
-For $M_1$, the partial ranking function $\big(\tY \, \lambda f n .
-\, \tif{n=0}{0}{f \, (n-1) \oplus_{3/2} f \, (n+1)}\big) \, \underline x \mapsto 3x + 1$ suffices to prove its termination (and could equivalently be considered an antitone partial ranking function with the constant antitone function $\epsilon_1(x) = 1$).
+Programs $M_1$ and $M_2$ are AST. For $M_1$, the partial ranking function 
+\[
+\big(\tY \, \lambda f n . \, \tif{n=0}{0}{f \, (n-1) \oplus_{3/2} f \, (n+1)}\big) \, \underline x \mapsto 3x + 1
+\] 
+suffices to prove its termination (and could equivalently be considered an antitone partial ranking function with the constant antitone function $\epsilon_1(x) = 1$).
 
@@
 \tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
-we can define an antitone partial ranking function $f_2 : \mathit{Rch}(M_2) \rightharpoonup \nnReal$ with antitone function $\epsilon_2$ defined as follows (for $n \in \mathbb N$):
+we can define an antitone partial ranking function $f_2 : \mathit{Rch}(M_2) \rightharpoonup \nnReal$ 
+%with antitone function $\epsilon_2$ defined as follows (for $n \in \mathbb N$):
+defined as: for $n \in \mathbb N$
 \begin{align*}
@@
 For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$, each with probability $1/2$, with one $\tY$-reduction.
+Now
 \begin{align*}
   & g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 \\
-  = & \log(\frac{n+1}{\sqrt{n(n+2)}}) \\
-  = & \frac 1 2 \log(\frac{(n+1)^2}{n(n+2)}) \\
-  = & \frac 1 2 \log(1 + \frac 1 {n(n+2)}) \\
+  = & \log \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) \\
+  = & \frac 1 2 \log\left(\frac{(n+1)^2}{n(n+2)}\right) \\
+  = & \frac 1 2 \log\left(1 + \frac 1 {n(n+2)}\right) \\
   > & \frac 1 {4n(n+2)} \\
@@
 \end{align*}
-and $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$, therefore by setting $\epsilon_2(x) = \frac 1 {5(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met.
+and $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
+Therefore by setting $\epsilon_2(x) = \frac 1 {5(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met.
 It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone partial ranking function, and $M_2$ is AST.
@@
 \begin{align*}
-\Theta \, l 
+\Theta \, \underline l 
 &\mapsto 
@@
 \]
-Let $g(x) := 2 + \log(x + 1)$ again and let $\epsilon$ be the function specified by
+Let $g(x) := 2 + \log(x + 1)$, and let $\epsilon$ be the function specified by
 \[
@@
 \]
-The limit of this as $x \to \infty$ and $g(x+1/2) \to \infty$ is 0, and $\frac {\mathrm d}{\mathrm dx} \epsilon(g(x+1/2)) = \frac 1 {x+1} + \log(1 - \frac 1 {x + 1/2}) < 0$, and $g$ is monotonic increasing therefore $\epsilon$ is antitone and bounded below by 0.
+The limit of this as $x \to \infty$ and $g(x+1/2) \to \infty$ is 0, and $\frac {\mathrm d}{\mathrm dx} \epsilon(g(x+1/2)) = \frac 1 {x+1} + \log(1 - \frac 1 {x + 1/2}) < 0$, and $g$ is monotonic increasing;
+therefore $\epsilon$ is antitone and bounded below by 0.
 We define a partial antitone ranking function as follows:
@@
 \end{example}
-The value of $g$ after one $\tY$-reduction step is at least $g(l-1/2)$ therefore the expectation of $\epsilon$ after one $tY$-reduction step is at most $\epsilon(g(l-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y$, therefore in the case where $l > 0$, $g$ decreases by the required amount, and in the case where $l \leq 0$, $g(\Xi \, l) \geq 2 + \log(1/2) > \log(\frac{3 \sqrt 3} 2) - 1 = \epsilon(g(0-1/2))$ as well, therefore this is a valid partial antitone ranking function and the term is AST.
+The value of $g$ after one $\tY$-reduction step is at least $g(l-1/2)$, therefore the expectation of $\epsilon$ after one $\tY$-reduction step is at most $\epsilon(g(l-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y$.
+Thus 
+\begin{itemize}
+\item in case $l > 0$, $g$ decreases by the required amount
+\item in case $l \leq 0$, 
+\(
+g(\Xi \, l) \geq 2 + \log(1/2) > \log(\frac{3 \sqrt 3} 2) - 1 = \epsilon(g(0-1/2))
+\) 
+as well.
+\end{itemize}
+Hence this is a valid partial antitone ranking function and the term is AST.
 
@@
 
-Consider the candidate ranking function $\Xi n \mapsto n + 1$.
-For each $n$, $\Xi n$ reduces to either $0$ (with probability $\frac 1 {n + 1}$) or $\Xi (n + 1)$, therefore the expected value of the ranking function is $\frac{n(n+2)}{n+1} = n + 1 - \frac 1 {(n+1)^2}$, and the required decrease is $\frac{\epsilon(0) + n \epsilon(n + 2)}{n + 1}$, therefore eventually, the expected decrease isn't enough, whatever the value of $\epsilon(0)$ is.
+Consider the candidate ranking function $\Xi \, n \mapsto n + 1$.
+For each $n$, $\Xi \, n$ reduces to either $0$ (with probability $\frac 1 {n + 1}$) or $\Xi \, (n + 1)$, therefore the expected value of the ranking function is $\frac{n(n+2)}{n+1} = n + 1 - \frac 1 {(n+1)^2}$, and the required decrease is $\frac{\epsilon(0) + n \epsilon(n + 2)}{n + 1}$, therefore eventually, the expected decrease isn't enough, whatever the value of $\epsilon(0)$ is.
 

### 2020-12-20 ranking-supermartingales.tex
Sec. 2: def. ranking supermartingale, and Lemma 1.

@@
 \lo{The preceding means: $\forall A \in \calF_n$, $\int_A \dif \mu \, Y_{n+1} \leq \int_A \dif \mu \,Y_n$.}
-%For a stopping time $T$ and $\epsilon > 0$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.
-\changed[lo]{For $\epsilon > 0$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$ where the random variable $T$ is defined as
-\(T : s \mapsto \min \set{n \mid Y_n(s) = 0}\).
-(It is straightforward to see that $T$ is a stopping time w.r.t.~$(\calF_n)_{n \in \omega}$.)}
+For $\epsilon > 0$ and a stopping time $T$ w.r.t.~$(\calF_n)_{n \in \omega}$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.
 A \emph{ranking supermartingale} is a $\epsilon$-ranking supermartingale for some $\epsilon > 0$.
@@
 
-\lo{N.B.~The original definition, due to \citep{DBLP:conf/popl/FioritiH15}, does not mention stopping time explicitly:
-they call a supermartingale $(Y_n)_{n \in \omega}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot 1_{\set{Y_n > 0}}$.}
+\begin{remark}
+\changed[lo]{The original definition of ranking supermartingale, due to \citep{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, does not mention stopping time explicitly:
+Fioriti and Hermanns call a supermartingale $(Y_n)_{n \in \omega}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot 1_{\set{Y_n > 0}}$.}
+\end{remark}
 
@@
 
-\begin{theorem}[{\citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}, \citep[Prop 1]{DBLP:conf/popl/ChatterjeeFNH16}}]
-\label{thm:rank-PAST}
+\begin{lemma}[{\citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}}]
+% \citep[Prop 1]{DBLP:conf/popl/ChatterjeeFNH16}}
+\label{lem:rank-PAST}
 Let $(Y_n)_{n \in \omega}$ be a $\epsilon$-ranking supermartingale with respect to the stopping time $T$, then $T < \infty$ almost surely, and $\expect{T} \leq \dfrac{\expect{Y_0}}{\epsilon}$.
-\end{theorem}
+\end{lemma}
 
@@
 \]
-It follows that $\sum_{i=0}^{\infty} \mathbb{P}[T > i]$ converges; hence $\lim_{n \to \infty} \mathbb{P}[T > i] = 0$ and so $\mathbb{P}[T < \infty] = 1$.
+It follows that $\sum_{i=0}^{\infty} \mathbb{P}[T > i]$ converges; hence $\lim_{i \to \infty} \mathbb{P}[T > i] = 0$ and so $\mathbb{P}[T < \infty] = 1$.
 Then it suffices to prove: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=0} \mathbb{P}[T > i]$.
 \end{proof}
+
+\lo{\Cref{lem:rank-PAST} is a slight extension of \citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}. In the latter, the stopping time $T$ is assumed to be $s \mapsto \min\set{n \mid Y_n(s) = 0}$.}
 
@@
 \begin{enumerate}
-\item If a closed SPCF term $M$ is rankable, then $T^{\tY}_M < \infty$ a.s.~(equivalently $M$ is AST), and $\expect{T^{\tY}_M} < \infty$.
+\item If a closed SPCF term $M$ is rankable, then $T^{\tY}_M < \infty$ a.s.~(equivalently $M$ is AST), and $\expect{T^{\tY}_M} < \infty$ i.e.~$M$ is $\tY$-positively almost-sure terminating ($\tY$-PAST).
 
-\item If a closed SPCF term $M$ is strictly rankable, then $\expect{T_M} < \infty$ i.e.~$M$ is PAST.
+\item If a closed SPCF term $M$ is strictly rankable, then $\expect{T_M} < \infty$ i.e.~$M$ is positively almost-sure terminating (PAST).
 \end{enumerate}
@@
 it then follows from \Cref{thm:optional sampling} that $(Y_n)_{n \in \omega}$ is a ($1$-ranking) supermartingale with respect to the filtration $(\calF_{T_n})_{n \in \omega}$ and the stopping time $T^{\tY}_M$.
-Therefore, by \Cref{thm:rank-PAST}, $T^{\tY}_M < \infty$ a.s.~and $\expect{T^{\tY}_M} < \infty$.
+Therefore, by \Cref{lem:rank-PAST}, $T^{\tY}_M < \infty$ a.s.~and $\expect{T^{\tY}_M} < \infty$.
 
-Statement 2 follows immediately from \Cref{thm:rank-PAST}.
+Statement 2 follows immediately from \Cref{lem:rank-PAST}.
 \end{proof}

### 2020-12-21 probabilisticProgrammingMartingales.tex
Question about the example that illustrates that the antitone function must be everywhere defined (Example 4).

@@
 
-\lo{The preceding definition does not place any constraint on the stopping time $T$ (other than that it is w.r.t.~the filtration $(\calF_n)_{n \in \omega}$).
-However for the definition to make sense, $T$ must be the stopping time $T(s) := \min \set{n \mid Y_n(s) = 0}$. So I propose the following.}
+%\lo{The preceding definition does not place any constraint on the stopping time $T$ (other than that it is w.r.t.~the filtration $(\calF_n)_{n \in \omega}$).
+However for the definition to make sense, $T$ must be the stopping time $T(s) := \min \set{n \mid Y_n(s) = 0}$.}
 
@@
 Then, take $n \leq B_x$. 
-We
+We have
 %\lo{Change of justification of the second equality below: event, not discrete r.v. Recall: expectation conditioning on r.v.~(resp.~event) is a r.v.~(resp.~number).}
@@
 Note that it is essential in this proof that $\epsilon$ be defined for all $\mathbb R_{\geq 0}$, not just the values that $(Y_n)_{n \in \omega}$ actually takes (or at least if $(Y_n)_{n \in \omega}$ is uniformly bounded, $\epsilon$ must be positive at the supremum as well as the actual values).
-Consider, for example, a random variable $X$, uniformly distributed in the interval $[-1,1]$, $T = $ the least $n$ such that $2^{-n} < X$, $Y_k = 4-2^{-k}$ if $k < T$, or $0$ otherwise, and $\epsilon(x) = \frac{4-x}{4}$.
+
+\begin{example}
+Consider a random variable $X$, uniformly distributed in the interval $[-1,1]$, $T := \min \set{n \mid 2^{-n} < X}$, 
+\[
+Y_k := 
+\begin{cases}
+4-2^{-k} & \hbox{if $k < T$}\\
+0 & \hbox{otherwise}
+\end{cases} 
+\]
+and $\epsilon(x) = \frac{4-x}{4}$. 
 In this case, $(Y_n)_{n \in \omega}$ either tends to $4$ as $n \to \infty$, or drops to 0 at some point, with an exponentially decreasing probability.
-It is an antitone strict supermartingale except that $\epsilon$ goes to $0$ at $4$, which is larger than any value $(Y_n)_{n \in \omega}$ can actually take, and $T = \infty$ with probability $\frac 1 2$.
+It is an antitone strict supermartingale \changed[akr]{except that $\epsilon$ goes to $0$ at $4$, which is larger than any value $(Y_n)_{n \in \omega}$ can actually take}, and $T = \infty$ with probability $\frac 1 2$.
+\lo{@Andrew: I refer to the preceding highlighted claim. Did you mean: for all $n, s$, $\epsilon(Y_n(s)) > Y_n(s)$?}
+\end{example}
 
@@
 For any term $N$ which is not of the form $E[R]$ for some $\tY$-redex $R$, this is trivial.
-If $R$ is a $tY$-redex, then $\epsilon(f_2(N)) \leq \int_S \epsilon(f_1(\nnext(N,s)))\, \mu(\mathrm d s)$ by the convexity of $\epsilon$, because $n$ is bounded.
+If $R$ is a $\tY$-redex, then $\epsilon(f_2(N)) \leq \int_S \epsilon(f_1(\nnext(N,s)))\, \mu(\mathrm d s)$ by the convexity of $\epsilon$, because $n$ is bounded.
 Therefore, the (total) function $f_2$, which agrees with $f$ on $f$'s domain, is an antitone ranking function on $M$ (with the same function $\epsilon$ if it's convex).
@@
 If instead, we take the ranking function to be defined after the $\tY$-reduction but before the $\tsample$-reduction as well, this can be resolved.
-Letting $\Theta[n] = \underline 0 \oplus_{\frac 1 n} \Xi n$:
-\begin{align*}
-\Theta[n] \mapsto n \\
-\Xi 10 \mapsto 12.
-\end{align*}
-For each $n$, $\Theta[n]$ reduces to either $0$ or $\Theta[n+1]$, with a $tY$-reduction only in the latter case, and the condition that this is an antitone partial ranking function is that $n \geq \frac{(n-1)(n+1)}{n} + \frac{n-1}{n} \epsilon(n+1)$ and $12 \geq 11 + \epsilon(11)$.
+Letting $\Theta[n] = \underline 0 \oplus_{\frac 1 n} \Xi \, n$:
+\begin{align*}
+\Theta[n] &\mapsto n \\
+\Xi \, 10 & \mapsto 12.
+\end{align*}
+For each $n$, $\Theta[n]$ reduces to either $0$ or $\Theta[n+1]$, with a $\tY$-reduction only in the latter case, and the condition that this is an antitone partial ranking function is that $n \geq \frac{(n-1)(n+1)}{n} + \frac{n-1}{n} \epsilon(n+1)$ and $12 \geq 11 + \epsilon(11)$.
 These are satisfied by setting $\epsilon(x) = \min(1, 1/x)$, therefore this term is AST.

### 2020-12-21 probabilisticProgrammingMartingales.tex
Respond to comments

@@
 \end{align*}
-
-\lo{Call the above partial function $h$.
-I assume that $h: (\tY f, n. n \oplus_{0.5} f(n+1)) N \mapsto 2$ for all $N$ (such that $(\tY f, n. n \oplus_{0.5} f(n+1)) N$ is reachable from $M$), 
-and the second and third clauses hold for all $i \geq 0$.
-
-Maybe I have misread / misunderstood the preceding, but I don't see why $h$ is a restriction of a ranking function. 
-Suppose $h$ is a restriction of $H$.
-Consider the reduction sequence (writing $\Xi = (\tY f, n. n \oplus_{0.5} f(n+1))$):
-\[
-\Xi \, 0 
-\xrightarrow{\tY} 
-0 \oplus_{0.5} \Xi \, 1 
-\to 
-\Xi \, 1
-\xrightarrow{\tY} 
-1 \oplus_{0.5} \Xi \, 2
-\to
-\Xi \, 2
-\xrightarrow{\tY} 
-2 \oplus_{0.5} \Xi \, 3
-\]
-Since $H(\Xi \, 0) = 2$, we must have 
-$H(\Xi \, 1) \leq 1$ and $H(\Xi \, 2) \leq 0$, contradicting the restriction assumption.
-}
-\akr{
-In this case, $H(0 \oplus_{0.5} \Xi 1) \leq 1$, but that term reduces to either $0$ or $\Xi 1$ therefore the actual condition is that $0.5 H(0) + 0.5 H(\Xi 1) \leq 1$, which is possible.
-}
-\lo{
-  OK. I see it now.
-}
 
@@
 In this case, $(Y_n)_{n \in \omega}$ either tends to $4$ as $n \to \infty$, or drops to 0 at some point, with an exponentially decreasing probability.
-It is an antitone strict supermartingale \changed[akr]{except that $\epsilon$ goes to $0$ at $4$, which is larger than any value $(Y_n)_{n \in \omega}$ can actually take}, and $T = \infty$ with probability $\frac 1 2$.
-\lo{@Andrew: I refer to the preceding highlighted claim. Did you mean: for all $n, s$, $\epsilon(Y_n(s)) > Y_n(s)$?}
+It is an antitone strict supermartingale except that $\epsilon$ goes to $0$ at $4$, and $T = \infty$ with probability $\frac 1 2$ even though $4$ is larger than any value $(Y_n)_{n \in \omega}$ can actually take and $\epsilon(Y_n)$ never actually reaches $0$.
+\lo{@Andrew: I refer to the preceding highlighted claim. Did you mean: for all $n, s$, $\epsilon(Y_n(s)) > Y_n(s)$?} \akr{No. This phrasing should be clearer.}
 \end{example}
@@
 If neither $\alpha_1 \leq \alpha_2$ nor $\alpha_1 \geq \alpha_2$, the positions are said to be \emph{disjoint}.
-This is mostly relevant in that in any substitution at some position, the subterms at all other positions are unaffected. 
-\lo{The preceding sentence is unclear. Typo?}
+The notion of disjointness is mostly relevant in that if $\alpha_1$ and $\alpha_2$ are disjoint, performing a substitution at $\alpha_1$ will leave the subterm at $\alpha_2$ inaffected.
+\lo{The preceding sentence is unclear. Typo?} \akr{It wasn't a typo, but I've made it a little more explicit now. Is this clearer?}
 
@@
 
-Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each \changed[lo]{term $M$} has a skeleton $Sk(M)$, and each \changed[lo]{skeleton $S$} can be converted to a term $S[r]$ given a \changed[lo]{vector $r$} of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$.
-\lo{Do we actually use the notation $S[r]$ in what follows?}
+Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $Sk(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$.
+\lo{Do we actually use the notation $S[r]$ in what follows?} \akr{Yes, in Lemma \ref{lem:independentSamples} for example.}
 \lo{Skeleton is defined in Sec.~\ref{sec:sampling semantics}.}
@@
 Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a $\to$-reduction sequence of skeletons starting at $Sk(M)$, and the second of which is a position in the final skeleton of the reduction sequence.
-Reduction sequences are used rather than reachable skeletons because if the same skeleton is reached twice, different samples may be needed, \lo{Give a simple example illustrating this.} but the reduction sequences of skeletons will be discussed as though they were just skeletons, identifying them with their final skeletons.
+As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the usual linear sampling semantics, there will be (with one more caveat introduced later) a sample from $I$ pre-selected for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.
+
+Reduction sequences are used rather than reachable skeletons because if the same skeleton is reached twice, different samples may be needed.
+For example, consider the term $M = \tY (\lambda f x.\ \tif{\tsample - \underline{0.5} < 0}{f \, x}{x}) \, \underline 0$, which reduces after a few steps to $N = \tif{\tsample - \underline{0.5}  < 0}{M}{\underline 0}$.
+If the pre-selected sample for $(N,\textsf{if}_1;\underline{-}_1)$ is less than 0.5, $N$ reduces back to $M$, then $N$ again, then the same sample is used the next time, therefore it's an infinite loop, whereas if samples are labelled by reduction sequences, the samples for $M \to \dots \to N$ are independent from the samples for $M \to \dots \to N \to M \to \dots \to N$, and so on.
+
+The reduction sequences of skeletons will often be discussed as though they were just skeletons, identifying them with their final skeletons.
 With this abuse of notation, a reduction sequence $N$ (actually $N_1 \to \dots \to N_n = N$) may be said to reduce to a reduction sequence $O$, where the reduction sequence implicitly associated with the final skeleton $O$ is $N_1 \to \dots \to N_n \to O$.
 
-\lo{Readers would benefit greatly from some examples of an $L_0(M)$-pair, and the relations $\sim_p$ and $\sim_c$, here, before the formal definitions. Explain the underlying intuition.}
+\paragraph{}
+This is still not quite sufficient because sometimes the same samples must be used at corresponding positions in different reduction sequences.
+For example, the term $M = \tsample + \tsample$ has the reachable skeletons $N_1 = \skeletonPlaceholder + \tsample, N_2 = \tsample + \skeletonPlaceholder, O = \skeletonPlaceholder + \skeletonPlaceholder$ and $\skeletonPlaceholder$, with reductions $M \to N_1 \to O \to \skeletonPlaceholder$ and $M \to N_2 \to O \to \skeletonPlaceholder$.
+In the reduction $M \to N_1$, the sample labelled $(M, \underline{+}_2)$ is used, and in the reduction $N_2 \to O$, the sample labelled $(M \to N_2, \underline{+}_2)$ is used.
+Each of these samples becomes the value of the second numeral in $O$ in their respective reduction sequences, therefore in order for Church-Rosserness to be attained, they must be the same.
+Which elements of $L_0(M)$ must match can be described by the relation $\sim^*$:
 
 \begin{definition}
-The relation $\sim$ is defined as the union of the symmetric relations $\sim_p$ \changed[lo]{($p$ for parent-child)} and $\sim_c$ \changed[lo]{($c$ for cousin)} where
-\lo{As defined, $\sim_p$ is not a symmetric relation.}
+The relation $\sim$ is defined as the union of the minimal symmetric relations $\sim_p$ (``$p$'' for parent-child) and $\sim_c$ (``$c$'' for cousin) satisfying
 \begin{enumerate}
@@
 Each case (except a, which is symmetrical) has a right branch and a left branch (the branches of case b are unfortunately displayed the wrong way around in the diagram).
-The right branch has reduction steps at two positions, $\beta$ then $\alpha$, where $\beta > \alpha$ (i.e.~$\beta$ is inside $\alpha$), and in the left branch, these order of these reductions are swapped, so the reductions are at $\alpha$ then $\beta_1, \dots, \beta_n$.
+The right branch has reduction steps at two positions, $\beta$ then $\alpha$, where $\beta > \alpha$ (i.e.~$\beta$ is inside $\alpha$), and in the left branch, the order of these reductions are swapped, so the reductions are at $\alpha$ then $\beta_1, \dots, \beta_n$.
 The position $\alpha$ of one of the reductions is unchanged, and the other reduction may have multiple (in cases e and f) or 0 (in cases b and e) images in the left branch.

### 2020-12-21 ranking-supermartingales.tex
Question about the example that illustrates that the antitone function must be everywhere defined (Example 4).

@@
 
-\lo{\Cref{lem:rank-PAST} is a slight extension of \citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}. In the latter, the stopping time $T$ is assumed to be $s \mapsto \min\set{n \mid Y_n(s) = 0}$.}
+\lo{\Cref{lem:rank-PAST} is a slight extension of \citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}.
+In op.~cit., the stopping time $T$ is assumed to be the random variable $s \mapsto \min\set{n \mid Y_n(s) = 0}$.}
 
@@
 The random variable $T_M^\tY$ is known as the \emph{$\tY$-runtime of $M$}, and it can equivalently be defined as the number of $\tY$-reduction steps in the reduction sequence of $M$.
-Note that, as the type system ensures that the reduction relation excluding $\tY$-reduction is strongly normalising, only finitely many reductions can occur in a row without one of them being a $\tY$-reduction, therefore $T_M^\tY < \infty$ \changed[lo]{a.s.}~iff $M$ is AST.
+Note that, as the type system ensures that the reduction relation excluding $\tY$-reduction is strongly normalising, only finitely many reductions can occur in a row without one of them being a $\tY$-reduction, therefore $T_M^\tY < \infty$ a.s.~iff $M$ is AST.
 

### 2020-12-21 probabilisticProgrammingMartingales.tex
ongoing
Add ranking function theorems for the confluent semantics.

@@
 
-\begin{theorem} \label{partial implies rankable}
+\begin{theorem} \label{thm:partial implies rankable}
 Every partial ranking function is a restriction of a ranking function.
@@
 
-\begin{theorem}[Antitone ranking function soundness]
+\begin{theorem}[Antitone ranking function soundness] \label{thm:antitone rankable implies termination}
 If a closed SPCF term $M$ is antitone rankable, then $T_M^{\tY} < \infty$ a.s.~(equivalently, $M$ is AST).
@@
 
-\begin{theorem}
+\begin{theorem}  \label{thm:antitone partial implies rankable}
   Every antitone partial ranking function is a restriction of an antitone ranking function.
@@
 Assume wlog that $\epsilon$ is convex (as if it isn't, we can just take its convex hull instead).
-As in Theorem \ref{partial implies rankable}, define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
+As in Theorem \ref{thm:partial implies rankable}, define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
   \begin{align*}
@@
 \end{theorem}
-\akr{TODO: Move proof of this theorem here, and edit it so that it's actually proving this, rather than trying to prove something stronger.}
+
+\input{relativised-completeness}
+
 \input{y-stopping-time}
-
-\input{relativised-completeness}
 
@@
 \fi
-
-\paragraph{}
-\akr{For now, here's a summary of the lemmas and theorems I intend to prove about this alternative semantics:
-\begin{itemize}
-    \item If $M \to N$, the relation $\sim_N^*$ defined on $Rch(N)$ is equal to the restriction of $\sim_M^*$ to $Rch(N)$ %\lo{Define $\sim_N^*$.}
-    \item Given the definition below of $\Rightarrow$, which is still applicable to this version of $\sim$, and which is the equivalent of $\red$ in this version of the semantics, $\Rightarrow$ is Church-Rosser.
-    \item If $M \to N$ at $\alpha$, no position in any term in $Rch(N)$ is related to $(M,\alpha)$ by $\sim_M^*$, and therefore each sample is used at most once in any reduction sequence.
-    \item The subprobability distribution of values obtained by reducing a term in the usual call-by-value semantics is equal to that obtained by any reduction strategy on $\Rightarrow$ (i.e.~a sub-relation of $\Rightarrow$ that is a function) (with some allowance made for the fact that some values are not normal forms of $\Rightarrow$, so the reduction strategy might overshoot the value that $\red^\infty$ reaches.
-I'm not quite sure what form this will take.)
-    \item Similarly, the probability of termination is equal in $\red^\infty$ and any reduction strategy on $\Rightarrow$, so almost sure termination can be equivalently defined based on $\Rightarrow$.
-    \item For any closed term $N$, if there is any reduction strategy for $N$ and ranking function corresponding to that reduction strategy, $N$ is AST.
-    \item Corresponding sparse and antitone-strict ranking function theorems.
-\end{itemize}}
 
@@
 
+All of the theorems on the termination of rankable terms therefore extend to other reduction strategies too.
+The proofs of Theorems \ref{thm:rankable implies termination}, \ref{thm:partial implies rankable}, \ref{thm:antitone rankable implies termination} and \ref{thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies.
+They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because Theorem \ref{thm:de groote} applies equally to any reduction strategy).
+For a reduction strategy $r$ and term $M$, just substitute $N[r(N)]$ being a $\tY$-redex for $N$ being of the form $E[\tY \lambda x.
+O]$, and $r(N)$ being undefined for $N$ being a value.
+
+The domain of definition of the ranking functions also needs to be changed from the reachable terms $Rch(M)$ to the \emph{reachable terms with respect to the reduction strategy $r$},
+\[
+Rch_r(M) = \{N \mid \exists n, N_0, \dots, N_n: N_0 = M, N_n = N, N_i \to N_{i+1} \text{ at } r(N_i)\}.
+\]
+
+To be more explicit, the modified forms of the theorems are:
+\begin{definition}\rm
+A \emph{ranking function on $M$ with respect to a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
+\begin{itemize}
+    \item $f(N) \geq 1+ f(N')$ if $N[r(N)]$ is a $\tY$-redex and $N$ reduces to $N'$ at $r(N)$
+    \item $f(N) \geq \int_I f(N[\underline{x}/r(N)]) \, \Leb(\mathrm{d}x)$ if $N[r(N)] = \tsample$
+
+    \item $f(N) \geq f(N')$ if $r(N)$ is any other redex, where $N \to N'$ at $r(N)$.
+\end{itemize}
+Any closed term for which a ranking function with respect to a reduction strategy $r$ exists is called \emph{rankable} with respect to $r$. 
+\end{definition}
+
+\begin{definition}
+A \emph{partial ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : Rch_r(M) \rightharpoonup \mathbb R$ such that
+\begin{itemize}
+    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
+    \item $f$ is defined at $M$.
+    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ at the positions specified by $r$ will eventually reach some $O$ such that either $r(O)$ isn't defined or $f(O)$ is, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
+\end{itemize}
+\end{definition}
+
+\begin{definition}\rm
+An \emph{antitone ranking function on $M$ with respect to a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ such that
+\begin{itemize}
+    \item $f(N) \geq \epsilon(f(N)) + f(N')$ if $N[r(N)]$ is a $\tY$-redex and $N$ reduces to $N'$ at $r(N)$
+    \item $f(N) \geq \int_I f(N[\underline{x}/r(N)]) \, \Leb(\mathrm{d}x)$ if $N[r(N)] = \tsample$
+
+    \item $f(N) \geq f(N')$ if $r(N)$ is any other redex, where $N \to N'$ at $r(N)$.
+\end{itemize}
+Any closed term for which an antitone ranking function with respect to a reduction strategy $r$ exists is called \emph{antitone rankable} with respect to $r$. 
+\end{definition}
+
+\begin{definition}
+An \emph{antitone partial ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : Rch_r(M) \rightharpoonup \mathbb R$ such that there exists some antitone function $\epsilon : \nnReal \to \pReal$ such that
+\begin{itemize}
+    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
+    \item $f$ is defined at $M$.
+    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ at the positions specified by $r$ will eventually reach some $O$ such that either $r(O)$ isn't defined or $f(O)$ is, and $f(N) \geq \mathbb E[f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
+\end{itemize}
+\end{definition}
+
+\begin{theorem}
+If a closed SPCF term $M$ has a ranking function, partial ranking function, antitone ranking function or antitone partial ranking function with respect to a reduction strategy $r$, then $M$ is AST with respect to $r$, and AST.
+\end{theorem}
 \lo{Give an example or two of a reduction strategy (for $\Rightarrow$) that is not $\cbv$.}

### 2020-12-21 relativised-completeness.tex
Thots on completeness proof: an alternative to ultrafilter on natural numbers.
Add ranking function theorems for the confluent semantics.

--- 2020-06-06 relativised-completeness.tex
@@
 
-\section{Relativised completeness}
-\label{sec:Relativised completeness}
-
-If every term reachable from a closed SPCF program is AST then the program is also AST, but not conversely 
-(because a diverging term may be reachable with probability 0 from an AST term).
-
-We aim to prove the following conjecture:
-
-\begin{quote}
-\emph{Given a closed SPCF term $M$, if every term reachable from $M$ is AST then $M$ admits an antitone ranking function.}
-\end{quote}
-
-\begin{enumerate}
-
-\item[(1)] Given an a.s.~finite stopping time $T$ (meaning: $T < \infty$ a.s.) adapted to a filtration, there exist supermartingale $(Y_n)_{n \in \omega}$ adapted to the same filtration and antitone function $\epsilon$ such that $(Y_n)_{n \in \omega}$ is an antitone strict supermartingale w.r.t.~$(\epsilon, T)$.
-
-\item[(2)] Let $(Y_n)_{n \in \omega}$ be an antitone strict supermartingale w.r.t.~$(\epsilon, T)$. 
-Then $M$ admits an antitone ranking function.
-\end{enumerate}
-
-\lo{Is (1) too general for the conjecture?
-Claim (1) assumes an arbitrary a.s.~finite stopping time $T$. This seems too general for the conjecture, which assumes a ``strongly'' AST $M \in \Lambda^0$.
-Regardless of the validity of (1), in order to prove the conjecture, I think we would need a more specific version of (1) that assumes a stopping time about terminating runs of $M$ --- such as the stopping time of definition (\ref{eq:y stopping time}) which only counts $\tY$-reduction steps --- so that the antitone-strict supermartingale that is constructed has the necessary structure to yield a ranking function for $M$.}
-
-\iffalse
-For (2), note that for every $N \in \mathit{Rch}(M)$, there exists $n \geq 0$ and $s \in S$ such that $\red^n(M, s) = (N, -)$.
-Define a function $f : \mathit{Rch}(M) \to \nnReal$ by $f(N) := Y_n'(s)$.
-\fi
-
-For 1, consider the following:
 \begin{proof}
 If $T$ is bounded by $b$ a.s.~(for $n$ constant), the statement is trivially true by taking $\epsilon$ to be constantly 1, and $Y_n = b - n$.
-Otherwise, let $(T_n)_{n \in \omega}$ be defined recursively such that $T_0 = 0$, $T_{n+1} > T_n$ and $\mathbb P[T < T_{n+1}] \leq \frac 1 4 \mathbb P[T < T_n]$.
-\lo{Since $T$ is a stopping time, we can assume that $T_n \in \mathbb N$.
-Problem with $T_1$: we require $T_1 > 0$ which may contradict the requirement $\mathbb P[T < T_{1}] \leq \frac 1 4 \mathbb P[T < 0] = 0$ for certain $T$, such as those that satisfy $\mathbb P[T < 1] > 0$.
-We could define $T_1$ as the least number satisfying $\mathbb P[T < T_1] \geq \frac{1}{4}$.}
-We will then define a supermartingale $(Y_n)_{n \in \omega}$ and nonrandom sequence $(Y'_n)_{n \in \omega}$ such that $Y_n = 0$ iff $T < n$ \changed[lo]{a.s.}, $Y_n = Y'_n$ iff $T \geq n$ \changed[lo]{a.s.}, and $Y'_n \geq 2^k$ for $n \geq T_k$.
+Otherwise, let $(t_n)_{n \in \omega}$ be defined recursively such that $t_0 = 0$, $t_{n+1} > t_n$ and $\mathbb P[T > t_{n+1}] \leq \frac 1 4 \mathbb P[T > t_n]$.
+We will then define a supermartingale $(Y_n)_{n \in \omega}$ and nonrandom sequence $(y_n)_{n \in \omega}$ such that $Y_n = 0$ iff $T < n$, $Y_n = y_n$ iff $T \geq n$, and $y_n \geq 2^k$ for $n \geq t_k$.
 
-\lo{$(T_n)_{n \in \omega}$ and $(Y'_n)_{n \in \omega}$ are $\omega$-sequences of numbers. 
-There is a convention that random variables are denoted by uppercase letters. 
-I suggest we write them as $(t_n)_{n \in \omega}$ and $(y'_n)_{n \in \omega}$ respectively:
-in this way, we emphasise that they are sequences of nunbers, and avoid any possible confusion with r.v.}
-
-The antitone function \lo{It is good practice not to begin a sentence with a maths expression / symbol.} $\epsilon$ is defined piecewise and recursively in such a way as to force all the necessary constraints to hold:
+The antitone function $\epsilon$ is defined piecewise and recursively in such a way as to force all the necessary constraints to hold:
 \begin{align*}
     \text{for }x \in [0,1),\ &\epsilon(x) = 1 \\
-    \text{for }x \in [2^k,2^{k+1}),\ &\epsilon(x) = \max \left (\epsilon(2^{k-1}), \frac{2^k}{T_{k+1}-T_k} \right ).
+    \text{for }x \in [2^k,2^{k+1}),\ &\epsilon(x) = \min \left (\epsilon(2^{k-1}), \frac{2^k}{t_{k+1}-t_k} \right ).
 \end{align*}
@@
 \item $\epsilon$ is bounded below by 0.
-\item For $x \geq 2^k$, $\epsilon(x) \leq \frac{2^k}{T_{k+1}-T_k}$. 
-\lo{This would imply $\frac{2^k}{T_{k+1}-T_k} \geq \epsilon(2^{k-1})$. Why?}
+\item For $x \geq 2^k$, $\epsilon(x) \leq \frac{2^k}{t_{k+1}-t_k}$. 
+\lo{This would imply $\frac{2^k}{t_{k+1}-t_k} \geq \epsilon(2^{k-1})$. Why?} \akr{Corrected: it should have been min, not max.}
 \end{itemize}
 
-$Y'_n$ is then defined recursively by
+$y_n$ is then defined recursively by
 \begin{align*}
-    Y'_0 = 2, \quad
-    Y'_{n+1} = (Y'_n - \epsilon(Y'_n)) \frac{\mathbb P[T > n]}{\mathbb P[T > n+1]}.
+    y_0 = 2, \quad
+    y_{n+1} = (y_n - \epsilon(y_n)) \frac{\mathbb P[T > n]}{\mathbb P[T > n+1]}.
 \end{align*}
 
-The fact $Y'_{T_k} \geq 2^{k+1}$ is proven by induction on $k$.
+The fact $y_{t_k} \geq 2^{k+1}$ is proven by induction on $k$.
 For base case, $2 \geq 2$, as required.
-For the inductive case $k+1$, we first do another induction to prove that $Y'_n \geq 2^k$ for all $T_k \leq n \leq T_{k+1}$.
+For the inductive case $k+1$, we first do another induction to prove that $y_n \geq 2^k$ for all $t_k \leq n \leq t_{k+1}$.
 The base case of this inner induction follows from the outer induction hypothesis a fortiori.
-take the greatest $k$ such that $n > T_k$.
-By the induction hypothesis, $Y'_{T_k} \geq 2^k$.
+Take the greatest $k$ such that $n > t_k$.
+By the induction hypothesis, $y_{t_k} \geq 2^k$.
 \begin{align*}
-    Y'_n & = Y'_{T_k} \frac{\mathbb P[T > T_k]}{\mathbb P[T > n]} - \sum_{m=T_k}^{n-1} \epsilon(Y'_m) \frac{\mathbb P[T > m+1]}{\mathbb P[T > n]} \\
-    & \geq \frac{\mathbb P[T > T_k]}{\mathbb P[T > n]} (Y'_{T_k} - \sum_{m=T_k}^{n-1} \epsilon(Y'_m)) \\
-    & \geq \frac{\mathbb P[T > T_k]}{\mathbb P[T > n]} (Y'_{T_k} - \sum_{m=T_k}^{n-1} \frac{2^k}{T_{k+1}-T_k}) \\
-    & = \frac{\mathbb P[T > T_k]}{\mathbb P[T > n]} (Y'_{T_k} - \frac{2^k (n - T_k)}{T_{k+1}-T_k}) \\
-    & \geq \frac{\mathbb P[T > T_k]}{\mathbb P[T > n]} (Y'_{T_k} - 2^k) \\
-    & \geq \frac{\mathbb P[T > T_k]}{\mathbb P[T > n]} (2^{k+1} - 2^k) \\
-    & = \frac{\mathbb P[T > T_k]}{\mathbb P[T > n]} 2^k \\
+    y_n & = y_{t_k} \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} - \sum_{m=t_k}^{n-1} \epsilon(y_m) \frac{\mathbb P[T > m+1]}{\mathbb P[T > n]} \\
+    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \sum_{m=t_k}^{n-1} \epsilon(y_m)) \\
+    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \sum_{m=t_k}^{n-1} \frac{2^k}{t_{k+1}-t_k}) \\
+    & = \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \frac{2^k (n - t_k)}{t_{k+1}-t_k}) \\
+    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - 2^k) \\
+    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (2^{k+1} - 2^k) \\
+    & = \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} 2^k \\
     & \geq 2^k
 \end{align*}
-as required. Substituting $n = T_{k+1}$ and using the fact that $\frac{\mathbb P[T > T_k]}{\mathbb P[T > T_{k+1}]}$, the same reasoning gives $Y'_{T_{k+1}} \geq 2^{k+2}$, as required.
+as required. Substituting $n = t_{k+1}$ and using the fact that $\frac{\mathbb P[T > t_k]}{\mathbb P[T > t_{k+1}]} > 4$, the same reasoning gives $y_{t_{k+1}} \geq 2^{k+2}$, as required.
 
-Although the stronger induction hypothesis was necessary for the induction to work, the only reason this is needed is to prove that $Y'_n > 0$ for all n, which implies that $Y_n \geq 0$.
-The other condition for $(Y_n)$ to be an antitone-strict supermartingale is \akr{The inference from the third to the fourth line is actually incorrect if $\mathcal F$ is not coarsest filtration to which $T$ is adapted.
-This error may require a substantial change to the proof to correct, but I thought you'd want to see it anyway rather than waiting for me to fix it.} \\
+Although the stronger induction hypothesis was necessary for the induction to work, the only reason this is needed is to prove that $y_n > 0$ for all n, which implies that $Y_n \geq 0$.
+The other condition for $(Y_n)$ to be an antitone-strict supermartingale is
 \begin{align*}
     & \expect{Y_{n+1} \mid \mathcal F_n} \\
-    = & \expect{Y'_{n+1} 1_{T \geq n+1} \mid \mathcal F_n} \\
-    = & Y'_{n+1} \expect{1_{T \geq n+1} \mid \mathcal F_n} \\ 
-    = & Y'_{n+1} 1_{T \geq n} \frac{\mathbb P[T > n+1]}{\mathbb P[T > n]} \\
-    = & (Y'_n - \epsilon(Y'_n)) 1_{T \geq n} \\
-    = & Y_n - \epsilon(Y'_n) 1_{T \geq n}
+    = & \expect{y_{n+1} 1_{T \geq n+1} \mid \mathcal F_n} \\
+    = & y_{n+1} \expect{1_{T \geq n+1} \mid \mathcal F_n} \\ 
+    = & y_{n+1} 1_{T \geq n} \frac{\mathbb P[T > n+1]}{\mathbb P[T > n]} \\
+    = & (y_n - \epsilon(y_n)) 1_{T \geq n} \\
+    = & Y_n - \epsilon(y_n) 1_{T \geq n}
 \end{align*}
@@
 \end{proof}
-\lo{Ongoing.}
+
+The assumption that $(\mathcal F_n)_{n \in \omega}$ is the coarsest filtration to which $T$ is adapted is used in the equality between the third and fourth lines here.
+This condition is the main reason that this proof does not extend directly to a completeness result for antitone ranking functions.

### 2020-12-21 probabilisticProgrammingMartingales.tex
Fix geometric distribution example

@@
 
-\lo{NOTATION. To be consistent, we write $\lambda x. M$ rather than $\lambda x : M$.}
-
 \begin{example}[Geometric distribution]
@@
 \begin{align*}
-&(\tY \lambda f, n: \\
+&(\tY \lambda f, n. \\
 &\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
@@
 \end{align*}
-which generates a geometric distribution.
+which generates a geometric distribution (let $\Theta = \lambda f, n : \tif{\tsample \underline{- 0.5} < 0}{n}{f (n\underline{+ 1})}$, then the term is $(\tY \Theta)\ \underline 0$).
     Despite its simplicity, its $Rch$ contains all the terms
-\begin{enumerate}
-    \item $(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{0}$
-    \item $(\lambda z.(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) z) \underline{i}$
-    \item $(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{i}$
-    \item $(\lambda n: \tif{\tsample - 0.5 < 0}{n}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{n}{f (m+1)}) (m+1)}) \underline{i}$
-    \item $\tif{\tsample - 0.5 < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
-    \item $\tif{\underline r - 0.5 < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
-    \item $\tif{\underline{r - 0.5} < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
-    \item $\underline{i}$
-    \item $(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)$
-    \item $(\lambda z.(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) z) (\underline{i} + 1)$.
-\end{enumerate}
-\akr{TODO: make this into a diagram with arrows.}
-
-\lo{Here is a more readable version.}
-
-Writing $\Theta = \lambda f, n : \tif{\tsample - 0.5 < 0}{n}{f (n+1)}$, we have
-\begin{enumerate}
-    \item $(\tY \, \Theta) \; \underline{0}$
-
-    \item $\big(\lambda z.\Theta \; (\tY \, \Theta) \; z \big) \; \underline{i}$
-
-    \item $\Theta \; (\tY \, \Theta) \; \underline{i}$
-
-    \item \changed[akr]{$(\lambda n: \tif{\tsample - 0.5 < 0}{n}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{n}{f (m+1)}) (m+1)}) \; \underline{i}$}
-
-    \lo{I think the preceding should be: $(\lambda n: \tif{\tsample - 0.5 < 0}{n}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (n+1)}) \; \underline{i}$, and so, more succinctly:
-    \[
-    \big(
-    \lambda n: 
-    \tif{\tsample - 0.5 < 0}{n}{(\tY \, \Theta) \; (n+1)} 
-    \big) 
-    \; \underline{i}
-    \]}
-
-    \item $\tif{\tsample - 0.5 < 0}{\underline{i}}{
-    (\tY \, \Theta) \, (\underline{i}+1)}$
-
-    \item $\tif{\underline r - 0.5 < 0}{\underline{i}}{(\tY \, \Theta) (\underline{i}+1)}$
-
-    \item $\tif{\underline{r - 0.5} < 0}{\underline{i}}{(\tY \, \Theta) (\underline{i}+1)}$
-
-    \item $\underline{i}$
-
-    \item $(\tY \, \Theta) \; (\underline{i}+1)$
-
-    \item $\big(\lambda z. \Theta \; (\tY \, \Theta) \;  z\big) \; (\underline{i} + 1)$
-
-    \item $\Big(\lambda z. \big(
-    \lambda n: 
-    \tif{\tsample - 0.5 < 0}{n}{(\tY \, \Theta) \; (n+1)} 
-    \big) \, z \Big)
-    \; \underline{i}$
-\end{enumerate}
-
-\lo{So we have:
-\[
-\begin{array}{ll}
-1 \to 11 \to 4 \to 5 \to 6 \to 7 \to 8\\
-7 \to 9 \to 11 \to \cdots
-\end{array}
-\]
-You have 2, 3 and 10 because your $\tY$-redex rule is
-\[
-\tY \lambda x. s \to \lambda z. (\lambda x.s)\, (\tY \lambda x. s) \, z
-\]
-which is not the official definition.}
+
+\begin{tikzpicture}[node distance=0.5em, anchor=west]
+    \node (1) at (0,-0) {$\tY\ \Theta\ \underline 0$};
+    \node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\ \Theta\ (n \underline{+ 1})}) z)\ \underline i$};
+    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\ \Theta\ (n \underline{+ 1})})\ \underline i$};
+    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY\ \Theta\ (\underline i\ \underline{+ 1})}$};
+    \node (5) at (0,-4) {$\tif{\underline r\ \underline{- 0.5} < 0}{\underline i}{\tY\ \Theta\ (\underline i\ \underline{+ 1})}$};
+    \node (6) at (0,-5) {$\tif{\underline r < 0}{\underline i}{\tY\ \Theta\ (\underline i\ \underline{+ 1})}$};
+    \node (7) at (2,-6) [text width=] {$\underline i$};
+    \node (8) at (0,-7) {$\tY\ \Theta\ (\underline i\ \underline{+ 1})$};
+    \node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\ \Theta\ (n \underline{+ 1})}) z)\ (\underline i\ \underline{+ 1})$};
+
+    \draw [->] (1.south west) to [bend right=45] (2.north west);
+    \draw [->] (2.south west) to [bend right=45] (3.north west);
+    \draw [->] (3.south west) to [bend right=45] (4.north west);
+    \draw [->] (4.south west) to [bend right=45] (5.north west);
+    \draw [->] (5.south west) to [bend right=45] (6.north west);
+    \draw [->] (6.south west) to [bend right=45] (8.north west);
+    \draw [->] (8.south west) to [bend right=45] (9.north west);
+
+    \draw [->] (9.west) to [bend left=45] (2.west);
+    \draw [->] (6) to (7);
+\end{tikzpicture}
+
 \end{example}
 
-Even in this simple case, defining a ranking function explicitly is awkward because of the number of cases, although in most cases, because the value need only be greater than or equal to that of the next term in sequence, it suffices to take the ranking function as taking only 3 distinct values.
+Even in this simple case, defining a ranking function explicitly is awkward because of the number of cases, although in most cases, because the value need only be greater than or equal to that of the next term in sequence, it suffices to take the ranking function as having the same value as the next term, so that overall it takes only 3 distinct values.
 (We will explain why later.)
@@
 \paragraph{}
-With this abbreviation, the geometric distribution example from earlier can be written as $(\tY f, n.
-n \oplus_{0.5} f(n+1)) \, \underline 0$.
+With this abbreviation, the geometric distribution example from earlier can be written as $(\tY \lambda f n.\ n \oplus_{0.5} f(n+1)) \, \underline 0$.
 It is then easy to see that the following is a partial ranking function:
 \begin{align*}
-(\tY f, n. n \oplus_{0.5} f(n+1)) N \mapsto 2 \\
-\underline i \oplus_{0.5} (\tY f, n. n \oplus_{0.5} f(n+1)) (\underline i + 1) \mapsto 1 \\
-\underline i \mapsto 0
-\end{align*}
-
+(\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\ N & \mapsto 2 \\
+\underline i \oplus_{0.5} (\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\ (\underline i + 1) & \mapsto 1 \\
+\underline i & \mapsto 0.
+\end{align*}
+
+In fact, even the partial function $\tY \Theta N \mapsto 2$ alone is a partial ranking function for this term.
 \section{Antitone ranking functions}

### 2020-12-21 probabilisticProgrammingMartingales.tex
Fix excessively wide lines

@@
 The sequence $(X_i)_i$ starts by swapping $\beta$ then $\gamma$ below $\alpha$.
-With the shorthand that $\gamma'(\epsilon) = \alpha;\lambda;@_1;\epsilon;\tY;\lambda;\gamma'$, $\gamma'_0 = \alpha;\lambda;@_1;\gamma'$, and similarly for $\beta'$, the relevant portion of $X_2$ is then $\alpha,\ (\gamma'(\delta^l_i))_i,\ \gamma'_0,\ (\gamma'(\delta^{C'}_i))_i,\ (\gamma'(\delta^m_i))_i,\ (\gamma'(\delta^B_i))_i,\ (\gamma'(\delta^r_i))_i,\ (\beta'(\delta^l_i))_i,\ (\beta'(\delta^{C'}_i))_i,\ (\beta'(\delta^m_i))_i,\ \beta'_0,\ (\beta'(\delta^{B'}_i))_i,\ (\beta'(\delta^r_i))_i$.
+With the shorthand that $\gamma'(\epsilon) = \alpha;\lambda;@_1;\epsilon;\tY;\lambda;\gamma'$, $\gamma'_0 = \alpha;\lambda;@_1;\gamma'$, and similarly for $\beta'$, the relevant portion of $X_2$ is then $\alpha$, $(\gamma'(\delta^l_i))_i$, $\gamma'_0$, $(\gamma'(\delta^{C'}_i))_i$, $(\gamma'(\delta^m_i))_i$, $(\gamma'(\delta^B_i))_i$, $(\gamma'(\delta^r_i))_i$, $(\beta'(\delta^l_i))_i$, $(\beta'(\delta^{C'}_i))_i$, $(\beta'(\delta^m_i))_i$, $\beta'_0$, $(\beta'(\delta^{B'}_i))_i$, $(\beta'(\delta^r_i))_i$.
 Next in $(X_k)_k$, for each $i$, $\gamma'(\delta^r_i)$ are swapped past all the images of $\beta$ until it is immediately before $\beta'(\delta^r_i)$ by case a, then for each $i$, $\gamma'(\delta^B_i)$ is swapped past all the images of $\beta$ until $\beta'_0$ by case a, then swapped with $\beta'_0$ by one of the other cases depending on what type of reduction $B \to B'$ is and the relative positions.
@@
 Next up in $(X_i)_i$, the $(\gamma'(\delta^m_i))_i$s and the $(\gamma'(\delta^{C'}_i))_i$s swap down past some images of $\beta$ they're disjoint from to take their place before their matching image of $\beta$, then $\gamma'_0$ swaps past the $(\beta'(\delta^l_i))_i$, stopping immediately before $\beta'(\delta^{C'}_1)$, then the $(\gamma'(\delta^l_i))_i$ and the $(\beta'(\delta^l_i))_i$ mix.
-The end result of the rearrangement of (this subsequence of) the reduction sequence is therefore $\alpha,\ (\gamma'(\delta^l_i), \beta'(\delta^l_i))_i,\ \gamma'_0,\ (\gamma'(\delta^{C'}_i), \beta'(\delta^{C'}_i))_i,\ (\gamma'(\delta^m_i), \beta'(\delta^m_i))_i,\ \beta'_0,\ (\gamma'(\delta^{B'}_i), \beta'(\delta^{B'}_i))_i,\ (\gamma'(\delta^r_i), \beta'(\delta^r_i))_i$.
+The end result of the rearrangement of (this subsequence of) the reduction sequence is therefore $\alpha$, $(\gamma'(\delta^l_i), \beta'(\delta^l_i))_i$, $\gamma'_0$, $(\gamma'(\delta^{C'}_i), \beta'(\delta^{C'}_i))_i$, $(\gamma'(\delta^m_i), \beta'(\delta^m_i))_i$, $\beta'_0$, $(\gamma'(\delta^{B'}_i), \beta'(\delta^{B'}_i))_i$, $(\gamma'(\delta^r_i), \beta'(\delta^r_i))_i$.
 
@@
 & \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')\}) \\
-= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in S : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) + \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in S : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}: g_O(r,s'(O,f(O))) \in T, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) + \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, s' \circ i(O[r] \to N[g_O(r)]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)}: s'(O,f(O)) \in T_{h(O)}, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) + \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, s' \circ i(O[r] \to N[g_O(r)]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(\{s' \in I^{L_s(O)} \mid s'(O,f(O)) \in T_{h(O)}, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S\}) + \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu({s' \in I^{L_s(O)} \mid s' \circ i(O[r] \to N[g_O(r)]) \in S}) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(T_{h(O)}) \mu(S) + \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : s'(O,f(O)) \in T_{h(O)}, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) + \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \mathbb R^{k-1}, s' \in I^{L_s(O)} : g_O(r,s'(O,f(O))) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) + \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
-= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S) + \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S) \\
+= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in S :\\
+    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\})\\
+    + & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in S :\\
+    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \\
+\end{align*}
+\begin{align*}
+= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}:\\
+    & g_O(r,s'(O,f(O))) \in T, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
+     + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}:\\
+     & g_O(r) \in T, s' \circ i(O[r] \to N[g_O(r)]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \\
+= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)}:\\
+    & s'(O,f(O)) \in T_{h(O)}, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
+    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}:\\
+    & g_O(r) \in T, s' \circ i(O[r] \to N[g_O(r)]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \\
+= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
+    & \mu(\{s' \in I^{L_s(O)} \mid s'(O,f(O)) \in T_{h(O)}, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S\})\\
+    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
+    & \mu({s' \in I^{L_s(O)} \mid s' \circ i(O[r] \to N[g_O(r)]) \in S}) \\
+= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(T_{h(O)}) \mu(S)\\
+    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
+= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : s'(O,f(O)) \in T_{h(O)}, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S)\\
+    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
+= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \mathbb R^{k-1}, s' \in I^{L_s(O)} : g_O(r,s'(O,f(O))) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S)\\
+    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
+= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} :\\
+    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S)\\
+    + & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} :\\
+    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S) \\
 = & \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in I^{L_s(N)} : (M,s) \Rightarrow_f^n (N[r], s')\}) \mu(S)

### 2020-12-21 probabilisticProgrammingMartingales.tex
Example 4. Slight reformulation: defining T and Y_k directly in terms of a measurable function on a probability space.

--- 2020-12-21 probabilisticProgrammingMartingales.tex
@@
 \begin{example}
-Consider a random variable $X$, uniformly distributed in the interval $[-1,1]$, $T := \min \set{n \mid 2^{-n} < X}$, 
+\akr{Consider a random variable $X$, uniformly distributed in the interval $[-1,1]$, $T := \min \set{n \mid 2^{-n} < X}$, 
 \[
@@
 \]
-and $\epsilon(x) = \frac{4-x}{4}$. 
+} \lo{The following reformulation defines $T$ and $Y_k$ directly as a measurable function on a probability space (rather than via a r.v.~$X$).}
+
+Take the probability space $(\Omega, \calF, \mathbb{P})$ where $\Omega$ is the closed interval of reals $[-1, 1]$, $\calF$ the Borel $\sigma$-algebra, and $\mathbb{P}$ the corresponding Lebesgue probability measure.
+Let $\omega \in [-1, 1]$. Define random variables $T$ and $(Y_k)_{k \in \omega}$:
+\begin{align*}
+T(\omega) & := 
+\begin{cases}
+\min\set{n \in \mathbb{N} \mid \omega > 2^{-n}} & \hbox{if $\omega \in (0, 1]$}\\
+\infty & \hbox{otherwise}
+\end{cases} 
+\\
+Y_k(\omega) & := 
+\begin{cases}
+4-2^{-k} & \hbox{if $\omega \in [-1, 2^{-k}]$ (equivalently $k < T(\omega)$)}\\
+0 & \hbox{otherwise}
+\end{cases} 
+\end{align*}
+\lo{N.B.~$\set{T = n} = (\frac{1}{2^{n}}, \frac{1}{2^{n-1}}]$, and so, $\set{T > n} = [-1, \frac{1}{2^n}]$.}
+Plainly, $T$ is a stopping time, and $(Y_n)_{n \in \omega}$ is a supermartingale, adapted to the filtration $(\calF_n)_{n \in \omega}$ where $\calF_n = \calF$ for all $n$. 
 In this case, $(Y_n)_{n \in \omega}$ either tends to $4$ as $n \to \infty$, or drops to 0 at some point, with an exponentially decreasing probability. 
-It is an antitone strict supermartingale except that $\epsilon$ goes to $0$ at $4$, and $T = \infty$ with probability $\frac 1 2$ even though $4$ is larger than any value $(Y_n)_{n \in \omega}$ can actually take and $\epsilon(Y_n)$ never actually reaches $0$.
-\lo{@Andrew: I refer to the preceding highlighted claim. Did you mean: for all $n, s$, $\epsilon(Y_n(s)) > Y_n(s)$?} \akr{No. This phrasing should be clearer.}
+With respect to the antitone function $\epsilon(x) = \frac{4-x}{4}$ and stopping time $T$, we have that $(Y_n)_{n \in \omega}$ is an antitone strict supermartingale, except that $\epsilon(4) = 0$, and $T = \infty$ with probability $\frac 1 2$, even though $4$ is larger than any value $(Y_n)_{n \in \omega}$ can actually take, and $\epsilon(Y_n)$ never actually reaches $0$.
+
+%\lo{@Andrew: I refer to the preceding highlighted claim. Did you mean: for all $n, s$, $\epsilon(Y_n(s)) > Y_n(s)$?} \akr{No. This phrasing should be clearer.}
 \end{example}
@@
 \underbrace{\big
-(\tY \, \lambda f x . \tif{x \leq 0}{0}{f(x - \tsample))} \big)}_{\Theta} \, 10
+(\tY \, \lambda f x . \tif{x \leq 0}{0}{f(x - \tsample)} \big)}_{\Theta} \, 10
 \]

### 2020-12-23 probabilisticProgrammingMartingales.tex
minor

@@
 \begin{theorem}
-For any stopping time $T$ which is almost surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
+For any stopping time $T$ which is almost-surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
 \end{theorem}

### 2020-12-23 relativised-completeness.tex
minor

@@
 
-$y_n$ is then defined recursively by
+The sequence $(y_n)_{n \in \omega}$ is then defined recursively by
 \begin{align*}
@@
 
-Although the stronger induction hypothesis was necessary for the induction to work, the only reason this is needed is to prove that $y_n > 0$ for all n, which implies that $Y_n \geq 0$.
+Although the stronger induction hypothesis was necessary for the induction to work, the only reason this is needed is to prove that $y_n > 0$ for all $n$, which implies that $Y_n \geq 0$. 
 The other condition for $(Y_n)$ to be an antitone-strict supermartingale is
@@
     & \expect{Y_{n+1} \mid \mathcal F_n} \\
-    = & \expect{y_{n+1} 1_{T \geq n+1} \mid \mathcal F_n} \\
-    = & y_{n+1} \expect{1_{T \geq n+1} \mid \mathcal F_n} \\ 
-    = & y_{n+1} 1_{T \geq n} \frac{\mathbb P[T > n+1]}{\mathbb P[T > n]} \\
-    = & (y_n - \epsilon(y_n)) 1_{T \geq n} \\
-    = & Y_n - \epsilon(y_n) 1_{T \geq n}
+    = & \expect{y_{n+1} \, 1_{\set{T \geq n+1}} \mid \mathcal F_n} \\
+    = & y_{n+1} \, \expect{1_{\set{T \geq n+1}} \mid \mathcal F_n} \\ 
+    = & y_{n+1} \, 1_{\set{T \geq n}} \frac{\mathbb P[T > n+1]}{\mathbb P[T > n]} \\
+    = & (y_n - \epsilon(y_n)) \, 1_{\set{T \geq n}} \\
+    = & Y_n - \epsilon(y_n) \, 1_{\set{T \geq n}}
 \end{align*}

### 2020-12-26 LICS21/SPCF.tex
Created version for LICS21 submission in subdirectory LICS21

--- 2020-12-23 probabilisticProgrammingMartingales.tex
@@
-\documentclass{article}
-
-%% BEGIN {Luke's Macros}
-%%
-\newif\ifdraft
-\drafttrue %% To hide all comments and highlighting, just comment this line out
-
-\input{Style/comment}
-\usepackage{calculation}
-\usepackage{graphicx}
-\usepackage{tikz}
-\usetikzlibrary{positioning, graphs, quotes, graphdrawing, shapes.geometric}
-\usegdlibrary{layered, circular}
-\usepackage[square]{natbib}
-\setcitestyle{aysep={}}
-
-\definecolor{oxblue}{RGB}{0,33,71}
-\definecolor{oxgold}{HTML}{a0630a}
-\usepackage[final,
-bookmarks,
-bookmarksopen,
-colorlinks,
-final,
-linkcolor=red,
-citecolor=oxgold,
-pdfstartview=FitH ]{hyperref}
-
-\usepackage{amsmath} % cleveref must be loaded afer amsmath
-\usepackage{cleveref}
-\Crefname{theorem}{Theorem}{Theorems}
-\Crefname{corollary}{Corollary}{Corollary}
-\crefname{proposition}{Proposition}{Propositions}
-\Crefname{claim}{Claim}{Claims}
-\Crefname{definition}{Definition}{Definitions}
-\Crefname{fact}{Fact}{Facts}
-\Crefname{conj}{Conjecture}{Conjectures}
-\Crefname{example}{Example}{Examples}
-\Crefname{remark}{Remark}{Remarks}
-\Crefname{convention}{Convention}{Conventions}
-\Crefname{lemma}{Lemma}{Lemmas}
-\Crefname{assumption}{Assumption}{Assumptions}
-\Crefname{section}{Section}{Sections}
-\Crefname{appendix}{Appendix}{Appendices}
-\Crefname{figure}{Figure}{Figures}
-%\Crefname{equation}{Eq.}{Equations}
-
-% \addtotheorempostheadhook[theorem]{\crefalias{thmlisti}{theorem}}
-% \addtotheorempostheadhook[lemma]{\crefalias{thmlisti}{lemma}}
-% \addtotheorempostheadhook[proposition]{\crefalias{thmlisti}{proposition}}
-% \addtotheorempostheadhook[corollary]{\crefalias{thmlisti}{corollary}}
-% \addtotheorempostheadhook[claim]{\crefalias{thmlisti}{claim}}
-% \addtotheorempostheadhook[fact]{\crefalias{thmlisti}{fact}}
-% \addtotheorempostheadhook[example]{\crefalias{thmlisti}{example}}
-% \addtotheorempostheadhook[definition]{\crefalias{thmlisti}{definition}}
-% \addtotheorempostheadhook[remark]{\crefalias{thmlisti}{remark}}
-% \addtotheorempostheadhook[assumption]{\crefalias{thmlisti}{assumption}}
-
-\newcommand{\Real}{\mathbb{R}}
-\newcommand{\nnReal}{\mathbb{R}_{{\geq}0}}
-\newcommand{\pReal}{\mathbb{R}_{{>}0}}
-
-\newcommand\expect[1]{\mathbb{E}[#1]}
-\newcommand\set[1]{\{#1\}}
-\newcommand\dif{\mathrm{d}}
-\newcommand\calF{\mathcal{F}}
-\newcommand\calI{\mathcal{I}}
-\newcommand\Leb{\mathrm{Leb}}
-\newcommand\ndraw[2]{\#\mathrm{draw}_{#1}(#2)}
-
-\makeatletter
-\newcommand{\dotDelta}{{\vphantom{\Delta}\mathpalette\d@tD@lta\relax}}
-\newcommand{\d@tD@lta}[2]{%
-  \ooalign{\hidewidth$\m@th#1\mkern-1mu\cdot$\hidewidth\cr$\m@th#1\Delta$\cr}%
-}
-
-%\usepackage[bbgreekl]{mathbbol} %for \bblambda
-%%
-%% END {Luke's Macros}
-
-\usepackage{amsfonts}
-\usepackage{amsthm}
-\usepackage{amssymb}
-\usepackage{verbatim}
-\usepackage{enumitem}
-\newcommand{\tY}{\mathsf{Y}}
-%\newcommand{\tif}[3]{\textsf{if }#1\textsf{ then }#2\textsf{ else }#3}
-\newcommand{\tif}[3]{\mathsf{if}(#1, #2, #3)} % space-saving concrete syntax
-\newcommand{\tsample}{\mathsf{sample}}
-\newcommand{\tscore}{\mathsf{score}}
-\newcommand{\skeletonPlaceholder}{\mathsf{X}} % I'm not sure what the best notation here is.
-\DeclareMathOperator{\red}{red}
-\DeclareMathOperator{\nnext}{next}
-\DeclareMathOperator{\cbv}{cbv}
-
-\usepackage{amsthm}
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}
-\newtheorem{example}{Example}
-
-\theoremstyle{lemma}
-\newtheorem{lemma}{Lemma}
-\newtheorem{theorem}{Theorem}
-\newtheorem{corollary}{Corollary}
-\newtheorem{conjecture}{Conjecture}
-
-\theoremstyle{remark}
-\newtheorem*{remark}{Remark}
-
-\title{Supermartingales and Termination Analysis of Statistical PCF}
-
-\begin{document}
-
-\maketitle
-
-\lo{Tentative title}
-
-\begin{abstract}
-\lo{TODO}
-\end{abstract}
-
-\iffalse
-\akr{Following your suggestion, I will be providing a criterion for termination of programs in PPCF \citep{DBLP:journals/pacmpl/EhrhardPT18} based on ranking supermartingales. 
-As it's more convenient for this proof, a sampling-based semantics \lo{You need to provide reference(s) for sampling-based semantics.} will be used instead of the original distributional semantics. 
-I assume some roughly applicable equivalence is proven somewhere, but it doesn't seem that hard anyway.}
-\lo{Why not provide a proof as an appendix?}
-\fi
-
-\tableofcontents
-
-Various theorems about probabilistic programs rely on the assumption that the program terminates almost surely.
-Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \citep{DBLP:journals/pacmpl/McIverMKK18}.
-This paper's contribution is to extend this method to a higher-order setting.
-
 \section{Statistical PCF}
+\label{sec:SPCF}
 
@@
 
-The language SPCF is a simply-typed lambda calculus with sampling of real numbers from $[0,1]$ and unbounded scoring, following \cite{MakOP20b}.
+The language SPCF is a simply-typed lambda calculus with sampling of real numbers from $[0,1]$ and unbounded scoring, following \cite{MakOPW21}.
 Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
@@
 The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid M \text{ terminates at } s \}$ is measurable.
-A term $M$ is said to terminate almost surely if $\mu(\{s \mid M \text{ terminates at } s\}) = 1$.
+We say that a term $M$ is \emph{almost-surely terminating} (AST) just if $\mu(\{s \mid M \text{ terminates at } s\}) = 1$.
 
@@
 \[
-\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \underline 0},
+\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \underline 0}
 \] 
@@
 \paragraph{}
-This definition of almost-sure termination is equivalent to that given in \citep{MakOP20b}, although the program semantics is stated in a slightly different way.
+This definition of almost-sure termination is equivalent to that given in \citep{MakOPW21}, although the program semantics is stated in a slightly different way.
 In particular, the argument to $\tscore$ is not relevant to termination (except for the possibility that its argument's evaluation wouldn't terminate).
@@
 \lo{Your operational semantics does not maintain a record of the current weight of the reduction.
-A.s.~termination does depend on $\tscore$: see \cite[\S 4.3]{DBLP:journals/corr/abs-2004-03924}\footnote{\url{https://arxiv.org/abs/2004.03924}}.
+A.s.~termination does depend on $\tscore$: see \cite[\S 4.3]{MakOPW21}\footnote{\url{https://arxiv.org/abs/2004.03924}}.
 I think it important to take the behaviour of $\tscore$ into account;
@@
 \fi
-
-\section{Supermartingales}
-One approach to proving that a term terminates almost surely is to find some variant that is bounded below and, on average, decreases sufficiently quickly that it must eventually reach 0, similarly to the approach taken in \citep{DBLP:journals/pacmpl/McIverMKK18} for imperative programs.
-
-These variants are defined as functions from reachable terms (i.e.~possible states of the program's execution) to real numbers.
-Specifically, let the set of reachable terms from a given closed starting term $M$ be $Rch(M) := \{N \in \Lambda \mid M \to^* N \}$, with the $\sigma$-algebra induced as a subset of $\Lambda$.
-
-\begin{definition}\rm
-A \emph{ranking function on $M$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
-\begin{enumerate}
-    \item $f(E[\tY \lambda x. N]) \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z) \text{ where $z$ is not free in $N$}$
-    \item $f(E[\tsample]) \geq \int_I f(E[\underline{x}]) \, \Leb(\mathrm{d}x)$
-
-    \item $f(E[R]) \geq f(E[R'])$ for any other redex $R$, where $R \to R'$.
-\end{enumerate}
-We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that $f(N) = 0$ iff $N$ is a value, and for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.
-
-Any closed term for which a ranking (respectively, strict ranking) function exists is called \emph{rankable} (respectively, \emph{strictly rankable}). 
-\changed[lo]{For example, the term $(\tY \lambda x.x) \, \underline 0$ is not rankable.}
-It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \in \omega}$ is the reduction sequence starting from $M$, then $(f(M_n))_{n\in\omega}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales.
-\end{definition}
-
-\input{ranking-supermartingales}
-
-\section{Constructing Ranking Functions}
-Although rankability implies almost-sure termination, the converse does not hold in general. For example,
-\begin{equation}
-\tif{{-}\tsample < 0}{\underline{0}}{(\tY \lambda x. x) \, \underline 0}
-\label{ex:0 probability reachable}
-\end{equation}
-terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x. x) \, \underline 0$ is reachable, although that has probability 0. 
-Not only is this counterexample AST, it's PAST. %positively almost surely terminating i.e.~the expected time to termination is finite.
-
-A ranking function can be constructed under the stronger assumptions that, for every $N$ reachable from $M$, the expected number of $\tY$-reduction steps from $N$ to a value is finite.
-In particular, the expected number of $\tY$-reduction steps from each reachable term is a ranking function.
-Note that a finite number of expected $\tY$-reduction steps does not necessarily imply a finite number of expected total reduction steps.
-
-\begin{example}
-\label{ex:tY finite does not imply t finite}
-The term
-\[
-M = \big(\tY \lambda f n. \tif{\tsample - 0.5 < 0}{n \, \underline 0}{f (\lambda x. n (n x))}\big) (\lambda x. x+1)
-\] 
-terminates with only 2 $\tY$-reductions on average i.e.~$\expect{T_M^\tY} = 2$, but applies the increment function $2^n$ times with probability $2^{-1-n}$ for $n \geq 0$, which diverges i.e.~$\expect{T_M} = \infty$.
-\end{example}
-
-\begin{theorem} \label{thm:minimal}
-Given a closed term $M$, the function $f:Rch(M) \to \mathbb R$ given by $f(N) = \mathbb E [\text{the number of }\tY\text{-reduction steps from }N\text{ to a value}]$, if it exists, is the least of all possible ranking functions of $M$.
-\end{theorem}
-\begin{proof}
-Let $f$ be the candidate least ranking function defined above, and suppose $g$ is another ranking function such that $f(N) > g(N)$ for some $N \in Rch(M)$.
-The restrictions of $f$ and $g$ to $Rch(N)$ have the same properties assumed of $f$ and $g$, so assume w.l.o.g.~that $N=M$.
-The difference $g - f$ is then a supermartingale (with the same setup as in the proof of %\Cref{thm:rankable gives supermartingale}
-%Theorem \ref{rankable implies ast}) 
-\Cref{thm:rankable and strict rankable});
-therefore $\mathbb E[g(M_n)] \leq \mathbb E[f(M_n)] + g(M)-f(M)$, for all $n$.
-
-We have $\mathbb E[f(M_n)] = \sum_{k=n}^\infty \mathbb P[M_k = E[\tY N] \text{ for some }E, N] \to 0 \text{ as } n \to \infty$; 
-therefore as $g(M) - f(M) < 0$, eventually $\mathbb E[g(M_n)] < 0$, which is impossible. 
-It follows that $g \geq f$ as required.
-
-In order for $f$ to be the least ranking function of $M$, it also has to actually be a ranking function itself.
-Each of the conditions on a ranking function is easily verified from the definition of $f$.
-\end{proof}
-
-\subsection{Partial ranking functions}
-Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost-sure termination, so it isn't useful for proving it.
-
-\begin{example}[Geometric distribution]
-Consider the term
-\begin{align*}
-&(\tY \lambda f, n. \\
-&\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
-&) \, \underline{0}
-\end{align*}
-which generates a geometric distribution (let $\Theta = \lambda f, n : \tif{\tsample \underline{- 0.5} < 0}{n}{f (n\underline{+ 1})}$, then the term is $(\tY \Theta)\ \underline 0$).
-    Despite its simplicity, its $Rch$ contains all the terms
-
-\begin{tikzpicture}[node distance=0.5em, anchor=west]
-    \node (1) at (0,-0) {$\tY\ \Theta\ \underline 0$};
-    \node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\ \Theta\ (n \underline{+ 1})}) z)\ \underline i$};
-    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\ \Theta\ (n \underline{+ 1})})\ \underline i$};
-    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY\ \Theta\ (\underline i\ \underline{+ 1})}$};
-    \node (5) at (0,-4) {$\tif{\underline r\ \underline{- 0.5} < 0}{\underline i}{\tY\ \Theta\ (\underline i\ \underline{+ 1})}$};
-    \node (6) at (0,-5) {$\tif{\underline r < 0}{\underline i}{\tY\ \Theta\ (\underline i\ \underline{+ 1})}$};
-    \node (7) at (2,-6) [text width=] {$\underline i$};
-    \node (8) at (0,-7) {$\tY\ \Theta\ (\underline i\ \underline{+ 1})$};
-    \node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\ \Theta\ (n \underline{+ 1})}) z)\ (\underline i\ \underline{+ 1})$};
-
-    \draw [->] (1.south west) to [bend right=45] (2.north west);
-    \draw [->] (2.south west) to [bend right=45] (3.north west);
-    \draw [->] (3.south west) to [bend right=45] (4.north west);
-    \draw [->] (4.south west) to [bend right=45] (5.north west);
-    \draw [->] (5.south west) to [bend right=45] (6.north west);
-    \draw [->] (6.south west) to [bend right=45] (8.north west);
-    \draw [->] (8.south west) to [bend right=45] (9.north west);
-
-    \draw [->] (9.west) to [bend left=45] (2.west);
-    \draw [->] (6) to (7);
-\end{tikzpicture}
-
-\end{example}
-
-Even in this simple case, defining a ranking function explicitly is awkward because of the number of cases, although in most cases, because the value need only be greater than or equal to that of the next term in sequence, it suffices to take the ranking function as having the same value as the next term, so that overall it takes only 3 distinct values.
-(We will explain why later.)
-
-The definition of rankability is also inconvenient for syntactic sugar.
-It could be useful, for example, to define $M \oplus_p N = \tif{\tsample - p < 0} M N$, where $M \oplus_p N$ reduces to $M$ or $N$, depending on the first value of $s$\lo{Need a name for $s$.}, with probability $p$ resp.~$(1-p)$.
-Technically though, it reduces first to $\tif{\underline r - p < 0} M N$ for all $r \in I$, so those terms all need values of the ranking function too.
-
-\paragraph{}
-In both of these cases, there are only some values of the ranking function that are semantically important. 
-\begin{definition}
-Define a \emph{partial ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \rightharpoonup \mathbb R$ such that
-\begin{itemize}
-    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
-    \item $f$ is defined at $M$.
-    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually \akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and ast, but that doesn't actually work because of terms reachable with 0 probability.
-Of course, astness doesn't imply rankability, so any definition of partial ranking function that includes all Y-past terms wouldn't always extend to a total ranking function.} reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
-\end{itemize}
-\end{definition}
-A partial ranking function that is total is just a ranking function.
-Providing a partial ranking function is essentially part way between providing a ranking function and directly proving almost sure termination.
-
-\begin{theorem} \label{thm:partial implies rankable}
-Every partial ranking function is a restriction of a ranking function.
-\end{theorem}
-\begin{proof}
-Take a closed term $M$ and partial ranking function $f$ on $M$. Define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
-\begin{align*}
-    f_1(N) & := f(N) \text{ whenever $f(N)$ is defined,} \\
-    f_1(V) & := 0 \text{ for values $V$ not in the domain of $f$.}
-\end{align*}
-
-Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) := \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
-The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on partial ranking functions.
-Define $f_2(N) = \int_S f_1(\nnext(N,s)) + g(N,s) \, \mu(\mathrm d s)$.
-The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction).
-\lo{TODO. The preceding sentence deserves a proof.}
-\end{proof}
-
-As a corollary, any term which admits a partial ranking function terminates almost surely.
-
-\subsection{Examples}
-Let $M \oplus_p N = \tif{\tsample - p < 0} M N$, for $p \in (0,1]$, then there are the pseudo-reduction relations
-\begin{align*}
-E[M \oplus_p N] \to^3 & E[M] & \\
-E[M \oplus_p N] \to^3 & E[N] & \\
-\red^3(E[M \oplus_p N], s) = & \left\{
-    \begin{array}{ll}
-        (E[M],\pi_t(s)) & \text{if } \pi_h(s) < p \\
-        (E[N],\pi_t(s)) & \text{if } \pi_h(s) \geq p. \\
-    \end{array} \right .
-\end{align*}
-A partial ranking function could be defined with respect to this shortcut reduction simply by replacing $\to$ and $\red$ in the definition of a partial ranking function by a version that goes straight from $N \oplus_p O$ to $N$ or $O$.
-Such a pseudo-partial ranking function would then be a partial function from a subset of $Rch(M)$, so it could also be considered as a partial function from all of $Rch(M)$, and it would in fact also be an actual partial ranking function.
-It is therefore possible to prove rankability directly using the shortcutted reductions.
-
-A similar procedure would work for other forms of syntactic sugar.
-If a closed term $N$ eventually reduces to one of a set of other terms $\{N_i \mid i \in I\}$ with certain probabilities, a partial ranking function defined with respect to a reduction sequence that skips straight from $N$ to $N_i$ is also a valid partial ranking function for the original reduction function, and therefore its existence implies almost sure termination.
-There is a caveat, however, that $\tY$-reduction steps skipped over in the shortcut still need to be counted for the expected number of $\tY$-reduction steps.
-
-\paragraph{}
-With this abbreviation, the geometric distribution example from earlier can be written as $(\tY \lambda f n.\ n \oplus_{0.5} f(n+1)) \, \underline 0$.
-It is then easy to see that the following is a partial ranking function:
-\begin{align*}
-(\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\ N & \mapsto 2 \\
-\underline i \oplus_{0.5} (\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\ (\underline i + 1) & \mapsto 1 \\
-\underline i & \mapsto 0.
-\end{align*}
-
-In fact, even the partial function $\tY \Theta N \mapsto 2$ alone is a partial ranking function for this term.
-\section{Antitone ranking functions}
-
-\begin{example}[Random walk]
-\label{ex:ac-ranking}
-\begin{enumerate}
-\item 1D biased (towards 0) random walk
-\[
-M_1 = 
-\big(\tY \, \lambda f n . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{3/2} f \, (n + 1)}\big) \, 10
-\]
-
-\item 1D unbiased random walk
-\[
-M_2 = 
-\big(\tY \, \lambda f n . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}\big) \, 10
-\]
-
-\item 1D biased (away from 0) random walk
-\[
-M_3 = 
-\big(\tY \, \lambda f n . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/3} f \, (n + 2)}\big) \, 10
-\]
-
-%While {x > 0} do {x := x-1 \oplus_2/3 x:= x+1}
-
-%M2 = While {x > 0} do {x := x-1 \oplus_2/3 x:= x+2}
-\end{enumerate}
-\end{example}
-
-The term $M_1$ is rankable, terminating in 31 $\tY$-steps on average, and $M_3$ only has a $1/1024$ chance of terminating, but in between, $M_2$ is AST, but isn't $\tY$-PAST, therefore it isn't rankable and \Cref{thm:rankable implies termination} is insufficient to prove its termination.
-We therefore want to find a generalised notion of ranking function so that $M_2$ becomes rankable, and then prove it sound i.e.~rankable in this generalised sense implies AST.
-
-\paragraph{Idea.}
-
-The definition of antitone-convex ranking function $f$ (say) for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
-we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) function $\epsilon : \nnReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
-\[
-f(E[R’]) \leq f(E[R]) - \epsilon(f(E[R])) 
-\]
-where $R \to R’$ is the $\tY$-redex rule.
-
-\lo{An aside: This CBV $\tY$-rule seems cleaner:
-\[
-\big(\tY f^{A \to B} \, x^A \, . \, \theta^B \big) \, v \to \theta[(\tY f \, x \, . \, \theta) / f, v / x].
-\]
-We assume $\tY f^{A \to B} \, x^A \, . \, \theta^B$ is a value.}
-
-\begin{definition}
-We say that a supermartingale $(Y_n)_{n \in \omega}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \in \omega}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
-\[
-\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot 1_{\set{T > n}}
-\]
-\end{definition}
-
-%\lo{The preceding definition does not place any constraint on the stopping time $T$ (other than that it is w.r.t.~the filtration $(\calF_n)_{n \in \omega}$).
-However for the definition to make sense, $T$ must be the stopping time $T(s) := \min \set{n \mid Y_n(s) = 0}$.}
-
-\begin{theorem}
-\label{thm:a-c strict}
-Let $(Y_n)_{n \in \omega}$ be an antitone strict supermartingale via function $\epsilon : \nnReal \to \pReal$, and stopping time $T$. 
-Then $T < \infty$ almost surely.
-\end{theorem}
-
-\begin{proof}
-First, as $(Y_n)$ is a supermartingale, $\expect{Y_n} \leq \expect{Y_0}$. 
-Therefore
-
-\begin{calculation}
-\expect{Y_n \mid T > n}
-  \step[=]{rearranging terms}
-\dfrac{\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
-+ 
-\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
-- 
-\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
-  \step[=]{definition of conditional expectation}
-\dfrac{\expect{Y_n} - \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
-  \step[\leq]{$Y_n \geq 0$ always}
-\dfrac{\expect{Y_n}}{\mathbb P[T > n]}
-  \step[\leq]{$(Y_n)_n$ is a supermartingale}
-\dfrac{\expect{Y_0}}{\mathbb P[T > n]}
-\end{calculation}
-
-\emph{Claim}: For all $0 < x \leq 1$, 
-\(
-\mathbb P[T > B_x] \leq x,
-\)
-where $B_x = \left \lceil \dfrac{\expect{Y_0}+1}{x \; \epsilon(\expect{Y_0} \, x^{-1})} \right \rceil$.
-
-\smallskip
-
-As the convex hull of $\epsilon$ (the greatest convex function less than or equal to it) satisfies all the conditions assumed of $\epsilon$, in addition to being convex, assume wlog that $\epsilon$ is convex.
-
-Assume for a contradiction that $\mathbb P[T > B_x] > x$.
-Then, take $n \leq B_x$. 
-We have
-%\lo{Change of justification of the second equality below: event, not discrete r.v. Recall: expectation conditioning on r.v.~(resp.~event) is a r.v.~(resp.~number).}
-\begin{calculation}
-\expect{Y_n-Y_{n+1}}
-  \step[=]{$\calF_n \subseteq \calF_{n+1}$, definition and linearity of conditional expectation}
-\expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
-  \step[\geq]{antitone strict assumption}
-\expect{\epsilon(Y_n) \cdot 1_{\set{T > n}}}
-%  \step[=]{definition of conditional expectation for a discrete variable}
-\step[=]{definition of expectation conditioning on an \emph{event}}
-\mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
-  \step[\geq]{Jensen's inequality}
-\mathbb P[T > n]\ \epsilon(\expect{Y_n \mid T > n})
-  \step[\geq]{proved earlier}
-\mathbb P[T > n]\ \epsilon\Big(\dfrac{\expect{Y_0}}{\mathbb P[T > n]}\Big)
-  \step[>]{assumption, $\mathbb P[T > n] \geq \mathbb P[T > B_x] > x$}
-x \; \epsilon\big(\expect{Y_0} \, x^{-1}\big).
-\end{calculation}
-Therefore, by a telescoping sum
-\[
-\expect{Y_{B_x}} = \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1}) \leq -1 < 0 
-\]
-%\lo{$\expect{Y_0 - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})}$}
-which is a contradiction, therefore the claim must be true, therefore $P[T > n]$ tends to 0 as $n$ tends to infinity, therefore $T < \infty$ almost surely.
-\end{proof}
-
-Note that it is essential in this proof that $\epsilon$ be defined for all $\mathbb R_{\geq 0}$, not just the values that $(Y_n)_{n \in \omega}$ actually takes (or at least if $(Y_n)_{n \in \omega}$ is uniformly bounded, $\epsilon$ must be positive at the supremum as well as the actual values).
-
-\begin{example}
-\akr{Consider a random variable $X$, uniformly distributed in the interval $[-1,1]$, $T := \min \set{n \mid 2^{-n} < X}$, 
-\[
-Y_k := 
-\begin{cases}
-4-2^{-k} & \hbox{if $k < T$}\\
-0 & \hbox{otherwise}
-\end{cases} 
-\]
-} \lo{The following reformulation defines $T$ and $Y_k$ directly as a measurable function on a probability space (rather than via a r.v.~$X$).}
-
-Take the probability space $(\Omega, \calF, \mathbb{P})$ where $\Omega$ is the closed interval of reals $[-1, 1]$, $\calF$ the Borel $\sigma$-algebra, and $\mathbb{P}$ the corresponding Lebesgue probability measure.
-Let $\omega \in [-1, 1]$. Define random variables $T$ and $(Y_k)_{k \in \omega}$:
-\begin{align*}
-T(\omega) & := 
-\begin{cases}
-\min\set{n \in \mathbb{N} \mid \omega > 2^{-n}} & \hbox{if $\omega \in (0, 1]$}\\
-\infty & \hbox{otherwise}
-\end{cases} 
-\\
-Y_k(\omega) & := 
-\begin{cases}
-4-2^{-k} & \hbox{if $\omega \in [-1, 2^{-k}]$ (equivalently $k < T(\omega)$)}\\
-0 & \hbox{otherwise}
-\end{cases} 
-\end{align*}
-\lo{N.B.~$\set{T = n} = (\frac{1}{2^{n}}, \frac{1}{2^{n-1}}]$, and so, $\set{T > n} = [-1, \frac{1}{2^n}]$.}
-Plainly, $T$ is a stopping time, and $(Y_n)_{n \in \omega}$ is a supermartingale, adapted to the filtration $(\calF_n)_{n \in \omega}$ where $\calF_n = \calF$ for all $n$. 
-In this case, $(Y_n)_{n \in \omega}$ either tends to $4$ as $n \to \infty$, or drops to 0 at some point, with an exponentially decreasing probability. 
-With respect to the antitone function $\epsilon(x) = \frac{4-x}{4}$ and stopping time $T$, we have that $(Y_n)_{n \in \omega}$ is an antitone strict supermartingale, except that $\epsilon(4) = 0$, and $T = \infty$ with probability $\frac 1 2$, even though $4$ is larger than any value $(Y_n)_{n \in \omega}$ can actually take, and $\epsilon(Y_n)$ never actually reaches $0$.
-
-%\lo{@Andrew: I refer to the preceding highlighted claim. Did you mean: for all $n, s$, $\epsilon(Y_n(s)) > Y_n(s)$?} \akr{No. This phrasing should be clearer.}
-\end{example}
-
-\begin{theorem}[Antitone ranking function soundness] \label{thm:antitone rankable implies termination}
-If a closed SPCF term $M$ is antitone rankable, then $T_M^{\tY} < \infty$ a.s.~(equivalently, $M$ is AST).
-\end{theorem}
-\begin{proof}
-As in %Theorem \ref{thm:basicRankingSoundness}, 
-\Cref{thm:rankable implies termination},
-take the set $S$ of traces as a probability space, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$.
-As before, $(Y_n)_{n \in \omega}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$.
-Therefore, it follows from \Cref{thm:a-c strict} that $T_M^{\tY}$ is almost-surely finite.
-\end{proof}
-
-\paragraph{}
-As before, constructing antitone ranking fuctions completely is not necessary, and there is a corresponding notion of an antitone partial ranking function.
-
-Define an \emph{antitone partial ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$
-\begin{itemize}
-\item $f(N) \geq 0$ for all $N$ where $f$ is defined.
-\item $f$ is defined at $M$.
-\item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
-\end{itemize}
-
-\begin{theorem}  \label{thm:antitone partial implies rankable}
-  Every antitone partial ranking function is a restriction of an antitone ranking function.
-\end{theorem}
-\begin{proof}
-  Take a closed term $M$ and an antitone partial ranking function $f$ on $M$, with a corresponding antitone function $\epsilon$.
-Assume wlog that $\epsilon$ is convex (as if it isn't, we can just take its convex hull instead).
-As in Theorem \ref{thm:partial implies rankable}, define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
-  \begin{align*}
-    f_1(N) &= f(N) \text{ whenever $f(N)$ is defined},\\
-    f_1(V) &= 0 \text{ for values $V$ not in the domain of $f$.}
-  \end{align*}
-
-  Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) = \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
-  The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on antitone partial ranking functions.
-Define $f_2(N) = \int_S f_1(\nnext(N,s)) + \epsilon(f_1(\nnext(N,s))) g(N,s) \, \mu(\mathrm d s)$.
-For any term $N$ where $f$ is defined, $f_2(N) = f(N)$, and the value that $f_2$ would have at $N$ if $f$ were not defined at $N$ is $\leq f(N)$, by the third condition on antitone partial ranking functions.
-In order to show that $f_2$ is an antitone partial ranking function, it therefore suffices to show that the value that $f_2$ would have had at each term if $f$ were not defined at that term is at least the expectation of $f_2$ after one reduction step (plus $\epsilon(f_2(N))$ if the reduction step is a $tY$-reduction).
-For any term $N$ which is not of the form $E[R]$ for some $\tY$-redex $R$, this is trivial.
-If $R$ is a $\tY$-redex, then $\epsilon(f_2(N)) \leq \int_S \epsilon(f_1(\nnext(N,s)))\, \mu(\mathrm d s)$ by the convexity of $\epsilon$, because $n$ is bounded.
-Therefore, the (total) function $f_2$, which agrees with $f$ on $f$'s domain, is an antitone ranking function on $M$ (with the same function $\epsilon$ if it's convex).
-\end{proof}
-
-As a corollary, any term which admits an antitone partial ranking function terminates almost surely.
-
-\paragraph{\Cref{ex:ac-ranking} revisited}
-\akr{All of these ranking functions are slightly incorrect, because $\Xi \, \underline n$ reduces to something like $\Xi \, (\underline n + 1)$ then $(\lambda z.
-\text{something}[\Xi/f]) (\underline n + 1)$, rather than reducing the argument first.
-This could be fixed by appealing to the confluent semantics version of the antitone ranking theorem, but it would be nice to have at least some examples here illustrating how antitone ranking functions work, rather than actually saving all these examples until the end.
-
-Even though the difference is quite trivial, this was actually the original motivating example for the confluent trace semantics.}
-
-Programs $M_1$ and $M_2$ are AST. For $M_1$, the partial ranking function 
-\[
-\big(\tY \, \lambda f n . \, \tif{n=0}{0}{f \, (n-1) \oplus_{3/2} f \, (n+1)}\big) \, \underline x \mapsto 3x + 1
-\] 
-suffices to prove its termination (and could equivalently be considered an antitone partial ranking function with the constant antitone function $\epsilon_1(x) = 1$).
-
-For $M_2$, define the function $g_1 : \nnReal \to \nnReal$ by $g_1(x) = \log(x+1) + 1$.
-Using shorthand 
-$\Theta_2 = \tY \, \lambda f n . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
-we can define an antitone partial ranking function $f_2 : \mathit{Rch}(M_2) \rightharpoonup \nnReal$ 
-%with antitone function $\epsilon_2$ defined as follows (for $n \in \mathbb N$):
-defined as: for $n \in \mathbb N$
-\begin{align*}
-{\Theta_2} \, n 
-&\mapsto 
-g_2(n)
-\\
-0 &\mapsto 0.
-\end{align*}
-For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$, each with probability $1/2$, with one $\tY$-reduction.
-Now
-\begin{align*}
-  & g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 \\
-  = & \log \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) \\
-  = & \frac 1 2 \log\left(\frac{(n+1)^2}{n(n+2)}\right) \\
-  = & \frac 1 2 \log\left(1 + \frac 1 {n(n+2)}\right) \\
-  > & \frac 1 {4n(n+2)} \\
-  > & \frac 1 2 \frac 1 {5(n+1)^2} + \frac 1 2 \frac 1 {5(n+3)^2}
-\end{align*}
-and $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
-Therefore by setting $\epsilon_2(x) = \frac 1 {5(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met.
-It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone partial ranking function, and $M_2$ is AST.
-
-\begin{example}[Continuous random walk]\label{ex:raven complex}
-In SPCF we can construct a function whose argument changes by a random amount at each recursive call:
-\[
-\underbrace{\big
-(\tY \, \lambda f x . \tif{x \leq 0}{0}{f(x - \tsample)} \big)}_{\Theta} \, 10
-\]
-We can construct a partial ranking function $f$ as follows:
-\begin{align*}
-\Theta \, \underline l 
-&\mapsto 
-l + 2
-\\
-\tif{l \leq 0}{0}{\Theta \, (l - \tsample)}
-&\mapsto
-l + 1
-\\
-0 &\mapsto 0.
-\end{align*}
-
-For a more complex (not $\tY$-PAST) example, consider the following ``continuous random walk''
-\[
-\underbrace{\big
-(\tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample + 1/2)} \big)}_{\Xi} \, 10
-\]
-Let $g(x) := 2 + \log(x + 1)$, and let $\epsilon$ be the function specified by
-\[
-\epsilon(g(x-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y.
-\]
-The limit of this as $x \to \infty$ and $g(x+1/2) \to \infty$ is 0, and $\frac {\mathrm d}{\mathrm dx} \epsilon(g(x+1/2)) = \frac 1 {x+1} + \log(1 - \frac 1 {x + 1/2}) < 0$, and $g$ is monotonic increasing;
-therefore $\epsilon$ is antitone and bounded below by 0.
-We define a partial antitone ranking function as follows:
-\begin{align*}
-\Xi \, l 
-&\mapsto 
-g(l)
-\\
-0 &\mapsto 0
-\end{align*}
-\end{example}
-The value of $g$ after one $\tY$-reduction step is at least $g(l-1/2)$, therefore the expectation of $\epsilon$ after one $\tY$-reduction step is at most $\epsilon(g(l-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y$.
-Thus 
-\begin{itemize}
-\item in case $l > 0$, $g$ decreases by the required amount
-\item in case $l \leq 0$, 
-\(
-g(\Xi \, l) \geq 2 + \log(1/2) > \log(\frac{3 \sqrt 3} 2) - 1 = \epsilon(g(0-1/2))
-\) 
-as well.
-\end{itemize}
-Hence this is a valid partial antitone ranking function and the term is AST.
-
-\begin{example}[Fair-in-the-limit random walk]
-\label{ex:Fair-in-the-limit random walk}\citep[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}
-\[
-\underbrace{\big
-(\tY \, f \, x . 
-\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)}_{\Xi} 
-\, 10
-\]
-To construct an antitone ranking function, we solve the recurrence relation:
-\[
-\left\{
-\begin{array}{rll}
-z_0 &=& 0\\
-n \geq 1, \quad z_n &>& \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
-\end{array}
-\right.
-\]
-For $n > 2$, $z_n = \log(n-1)$ works.
-The expected decrease is $\frac 1 2(\log(1+\frac 1 {(n-1)^2-1}) - \frac 1 {2n+1} \log(1 + \frac 2 {n-2}))$, and using the fact that $\frac x {x+1} \leq \log (1+x) \leq x$ for $x > 0$, this is at least $\frac 1 2(\frac 1 {(n-1)^2} - \frac 1 {2n + 1} \frac 2 {n-2}) = \frac {n^2}{2(n-1)^2(2n+1)(n-2)}$, which (again for $n > 2$) is positive and antitone.
-For $n = 0, 1, 2$, we take $\epsilon$ to be 9/40 (the same as its value at 3), then set $z_2 = 2 \log 2 - \log 3 - 1, z_1 = 3 \log 2 - 2 \log 3 - 3, z_0 = 4 \log 2 - 3 \log 3 - 6$.
-Some of those values are negative, but it's still bounded below, so by adding a constant offset it can be corrected, and the term is AST.
-
-\end{example}
-
-\begin{example}[Escaping spline]
-\label{ex:escaping spline}\citep[\S 5.4]{DBLP:journals/pacmpl/McIverMKK18}
-\[
-\underbrace{\big
-(\tY \, f \, x . \,
-0 \oplus_{\frac{1}{x+1}} f(x + 1) \big)}_{\Xi} 
-\, 10
-\]
-
-In this case, the fact that the ranking function must decrease at each $\tY$-step (in an antitone partial ranking function) by the expected value of $\epsilon$ not at the current term, but at the next term where the ranking function is defined, is a little harder to deal with, because the variable $x$ can change all the way to $0$ in one step, therefore simply adding a small offset doesn't suffice to compensate for this fact.
-
-Consider the candidate ranking function $\Xi \, n \mapsto n + 1$.
-For each $n$, $\Xi \, n$ reduces to either $0$ (with probability $\frac 1 {n + 1}$) or $\Xi \, (n + 1)$, therefore the expected value of the ranking function is $\frac{n(n+2)}{n+1} = n + 1 - \frac 1 {(n+1)^2}$, and the required decrease is $\frac{\epsilon(0) + n \epsilon(n + 2)}{n + 1}$, therefore eventually, the expected decrease isn't enough, whatever the value of $\epsilon(0)$ is.
-
-If instead, we take the ranking function to be defined after the $\tY$-reduction but before the $\tsample$-reduction as well, this can be resolved.
-Letting $\Theta[n] = \underline 0 \oplus_{\frac 1 n} \Xi \, n$:
-\begin{align*}
-\Theta[n] &\mapsto n \\
-\Xi \, 10 & \mapsto 12.
-\end{align*}
-For each $n$, $\Theta[n]$ reduces to either $0$ or $\Theta[n+1]$, with a $\tY$-reduction only in the latter case, and the condition that this is an antitone partial ranking function is that $n \geq \frac{(n-1)(n+1)}{n} + \frac{n-1}{n} \epsilon(n+1)$ and $12 \geq 11 + \epsilon(11)$.
-These are satisfied by setting $\epsilon(x) = \min(1, 1/x)$, therefore this term is AST.
-\end{example}
-
-\paragraph{Completeness}
-It seems very likely that this method is complete in the case that every reachable term is AST, but we have been unable to actually prove this.
-It is certainly at least capable of proving the termination of terms which terminate arbitrarily slowly (in the sense that there is no similar limitation to Theorem \ref{thm:rankable implies termination}, which can only prove termination of $\tY$-PAST terms).
-
-The following theorem does not prove completeness, but is suggestive in that direction:
-\begin{theorem}
-For any stopping time $T$ which is almost-surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
-\end{theorem}
-
-\input{relativised-completeness}
-
-\input{y-stopping-time}
-
-\section{Confluent Trace Semantics}
-When proving almost sure termination in this way, it is necessary to consider which terms a given term may reduce to.
-Sometimes however, the reduction that the programmer has in mind may not be strictly the call-by-value order defined so far, or considering an alternative reduction order may be simpler or more intuitive.
-
-Non-probabilistic lambda calculi generally have the Church-Rosser property, that if a term $A$ reduces to both $B_1$ and $B_2$, there is some $C$ 
-%with reduction sequences $B_1 \to^* C$ and $B_2 \to^* C$, 
-to which both $B_1$ and $B_2$ reduce,
-so the reduction order mostly doesn't matter. 
-In the probabilistic case, this may not be true, because $\beta$-reduction can duplicate $\tsample$s, so the outputs of the copies of the sample may be identical or independent, depending on whether the sample is taken before or after $\beta$-reduction.
-There are, however, some restricted variations on the reduction order that do not have this problem.
-
-\paragraph{}
-Even with this restriction, a sampling semantics in the style of the one already defined would not be entirely Church-Rosser, as, for example, $\red^3(\tsample - \tsample, (1,0,\dots))$ would be either $1$ or $-1$ depending on the order of evaluation of the $\tsample$s, as that determines which sample from the pre-selected sequence is used for each one.
-To fix this, rather than pre-selecting samples according to the order they'll be drawn in, select them according to the position in the term where they'll be used instead.
-
-A \emph{position} 
-%\akr{I'm not really sure how to phrase the citation, but this is adapted from \cite{Kennaway96infinitarylambda}.} 
-%\lo{The idea is common enough. I think no citation necessary.}
-is a finite sequence of steps into a term, defined inductively as
-\begin{align*}
-\alpha ::= \cdot \mid \lambda ; \alpha \mid @_1 ; \alpha \mid @_2 ; \alpha \mid \underline f_i ; \alpha \mid \tY ; \alpha \mid \textsf{if}_1 ; \alpha \mid \textsf{if}_2 ; \alpha \mid \textsf{if}_3 ; \alpha \mid \tscore ; \alpha.
-\end{align*}
-The \emph{subterm of $M$ at a position $\alpha$}, denoted $M \mid \alpha$, is defined as
-\begin{align*}
-M \mid \cdot & = M \\
-\lambda x. M \mid \lambda ; \alpha & = M \mid \alpha \\
-M_1 M_2 \mid @_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2 \\
-\underline f(M_1,\dots,M_n) \mid \underline f_i ; \alpha & = M_i \mid \alpha \quad \text{for }i \leq n \\
-\tY M \mid \tY ; \alpha & = M \mid \alpha \\
-\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2,3 \\
-\tscore(M) \mid \tscore ; \alpha & = M \mid \alpha
-\end{align*}
-so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda$ is undefined). 
-A position such that $M\mid \alpha$ does exist is said to \emph{occur} in $M$. 
-%\changed[lo]{
-\emph{Substitution} of $N$ at position $\alpha$ in $M$, written $M[N/\alpha]$, is defined similarly.
-For example, let 
-\[
-M = \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})
-\qquad
-\alpha =\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2
-\]
-then 
-\(
-M[\tsample / \alpha] = \lambda x \, y. y \, (\tif{x < 0}{y \, \tsample}{\underline 3}).
-\)
-%}
-
-Two subterms $N_1$ and $N_2$ of a term $M$, corresponding to positions $\alpha_1$ and $\alpha_2$, can overlap in a few different ways. 
-If $\alpha_1$ is an initial segment (i.e.~prefix) of $\alpha_2$ (written as $\alpha_1 \leq \alpha_2$), then $N_2$ is also a subterm of $N_1$.
-If neither $\alpha_1 \leq \alpha_2$ nor $\alpha_1 \geq \alpha_2$, the positions are said to be \emph{disjoint}.
-The notion of disjointness is mostly relevant in that if $\alpha_1$ and $\alpha_2$ are disjoint, performing a substitution at $\alpha_1$ will leave the subterm at $\alpha_2$ inaffected.
-\lo{The preceding sentence is unclear. Typo?} \akr{It wasn't a typo, but I've made it a little more explicit now. Is this clearer?}
-
-With this notation, a more general reduction relation $\to$ can be defined. \akr{It would almost work to have one of the neater versions of $\tY$-reduction here.}
-%\lo{On first reading, I struggled to parse the following clauses.}
-\begin{definition}
-\label{def:more general red}
-The binary relation $\to$ is defined by the following rules, each is conditional on a redex occurring at position $\alpha$ in the term $M$:
-\begin{align*}
-  \text{if } M \mid \alpha = (\lambda x.N) V,\ & M \to M[N[V/x]/\alpha] \\
-  \text{if } M \mid \alpha = \underline f (\underline r_1, \dots , \underline r_n),\ & M \to M[\underline{f(r_1,\dots,r_n)}/\alpha] \\
-  \text{if } M \mid \alpha = \tY \lambda x. N,\ & M \to M[\lambda z. N[(\tY \lambda x. N)/x] z/\alpha] \text{ where $z$ is not free in $N$}\\
-  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_1/\alpha] \text{ where }r < 0 \\
-  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_2/\alpha] \text{ where }r \geq 0 \\
-  \text{if } M \mid \alpha = \tsample & \text{ and $\lambda$ does not occur after $@_2$ or $\tY$ in $\alpha$},\\ & M \to M[\underline r/\alpha] \text{ where } r \in [0,1] \\
-  \text{if } M \mid \alpha = \tscore(\underline r),\ & M \to M[\underline r/\alpha].
-\end{align*}
-In each of these cases, $M \mid \alpha$ is the \emph{redex}, and reduction takes place at $\alpha$.
-\end{definition}
-
-\paragraph{}
-Labelling the pre-chosen samples by the positions in the term would also not work because in some cases, a $\tsample$ will be duplicated before being reduced, for example, in $(\lambda x.
-x \, {\underline 0} \, \mathbin{\underline{+}} \, {x \, \underline 0}) (\lambda y.
-\tsample)$, both of the $\tsample$ redexes that eventually occur originate at $@_2 ; \lambda$.
-It is therefore necessary to consider possible positions that may occur in other terms reachable from the original term. 
-Even this is itself inadequate because some of the positions in different reachable terms need to be considered the same, and the number of reachable terms is in general uncountable, which leads to measure-theoretic issues.
-
-Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $Sk(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$.
-\lo{Do we actually use the notation $S[r]$ in what follows?} \akr{Yes, in Lemma \ref{lem:independentSamples} for example.}
-\lo{Skeleton is defined in Sec.~\ref{sec:sampling semantics}.}
-The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, and $\tsample$ reducing to $\skeletonPlaceholder$.
-For example, we have 
-\(
-(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \tsample
-\to
-(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \skeletonPlaceholder
-\to
-\tif{\skeletonPlaceholder < 0}{\skeletonPlaceholder}{\skeletonPlaceholder}
-\to
-\skeletonPlaceholder
-\)
-
-Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a $\to$-reduction sequence of skeletons starting at $Sk(M)$, and the second of which is a position in the final skeleton of the reduction sequence.
-As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the usual linear sampling semantics, there will be (with one more caveat introduced later) a sample from $I$ pre-selected for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.
-
-Reduction sequences are used rather than reachable skeletons because if the same skeleton is reached twice, different samples may be needed.
-For example, consider the term $M = \tY (\lambda f x.\ \tif{\tsample - \underline{0.5} < 0}{f \, x}{x}) \, \underline 0$, which reduces after a few steps to $N = \tif{\tsample - \underline{0.5}  < 0}{M}{\underline 0}$.
-If the pre-selected sample for $(N,\textsf{if}_1;\underline{-}_1)$ is less than 0.5, $N$ reduces back to $M$, then $N$ again, then the same sample is used the next time, therefore it's an infinite loop, whereas if samples are labelled by reduction sequences, the samples for $M \to \dots \to N$ are independent from the samples for $M \to \dots \to N \to M \to \dots \to N$, and so on.
-
-The reduction sequences of skeletons will often be discussed as though they were just skeletons, identifying them with their final skeletons.
-With this abuse of notation, a reduction sequence $N$ (actually $N_1 \to \dots \to N_n = N$) may be said to reduce to a reduction sequence $O$, where the reduction sequence implicitly associated with the final skeleton $O$ is $N_1 \to \dots \to N_n \to O$.
-
-\paragraph{}
-This is still not quite sufficient because sometimes the same samples must be used at corresponding positions in different reduction sequences.
-For example, the term $M = \tsample + \tsample$ has the reachable skeletons $N_1 = \skeletonPlaceholder + \tsample, N_2 = \tsample + \skeletonPlaceholder, O = \skeletonPlaceholder + \skeletonPlaceholder$ and $\skeletonPlaceholder$, with reductions $M \to N_1 \to O \to \skeletonPlaceholder$ and $M \to N_2 \to O \to \skeletonPlaceholder$.
-In the reduction $M \to N_1$, the sample labelled $(M, \underline{+}_2)$ is used, and in the reduction $N_2 \to O$, the sample labelled $(M \to N_2, \underline{+}_2)$ is used.
-Each of these samples becomes the value of the second numeral in $O$ in their respective reduction sequences, therefore in order for Church-Rosserness to be attained, they must be the same.
-Which elements of $L_0(M)$ must match can be described by the relation $\sim^*$:
-
-\begin{definition}
-The relation $\sim$ is defined as the union of the minimal symmetric relations $\sim_p$ (``$p$'' for parent-child) and $\sim_c$ (``$c$'' for cousin) satisfying
-\begin{enumerate}
-    \item If $N$ reduces to $O$ with the redex at position $\alpha$, and $\beta$ is a position in $N$ disjoint from $\alpha$, then $(N,\beta) \sim_p (O,\beta)$.
-
-    \item If $N$ $\beta$-reduces to $O$ at position $\alpha$, $\beta$ is a position in $N \mid \alpha;@_1;\lambda$ and $N \mid \alpha;@_1;\lambda;\beta$ is not the variable involved in the reduction, $(N,\alpha;@_1;\lambda;\beta) \sim_p (O, \alpha;\beta)$
-
-    \item If $N$ $\textsf{if}$-reduces to $O$ at position $\alpha$, with the first resp.
-second branch being taken, and $\alpha;\textsf{if}_i;\beta$ occurs in $N$ (where $i = 2$ resp.
-$3$), $(N,\alpha;\textsf{if}_i;\beta) \sim_p (O,\alpha;\beta)$
-
-    \item If $N$, $O_1$ and $O_2$ match any of the following cases:
-    \begin{enumerate}
-        \item $N$ contains redexes at disjoint positions $\alpha_1$ and $\alpha_2$, $O_1$ is $N$ reduced first at $\alpha_1$ then $\alpha_2$ and $O_2$ is $N$ reduced first at $\alpha_2$ then at $\alpha_1$.
-
-        \item $N \mid \alpha = \tif{\underline r < 0}{N_1}{N_2}$, where $r < 0$ (or, respectively, $r \geq 0$), $(N_2 \text{ resp.
-} N_1) \mid \beta$ is a redex, and $O_1$ is $N$ reduced at $\alpha$ and $O_2$ is $N$ reduced first at $\alpha;(\textsf{if}_3 \text{ resp.
-} \textsf{if}_2);\beta$ then at $\alpha$
-
-        \item $N \mid \alpha = \tif{\underline r < 0}{N_1}{N_2}$, where $r < 0$ (or, respectively, $r \geq 0$), $(N_1 \text{ resp.
-} N_2) \mid \beta$ is a redex, and $O_1$ is $N$ reduced first at $\alpha$ then at $\alpha;\beta$ and $O_2$ is $N$ reduced first at $\alpha;(\textsf{if}_2 \text{ resp.
-} \textsf{if}_3);\beta$ then at $\alpha$
-
-        \item $N \mid \alpha = (\lambda x.
-A) B$, there is a redex in $A$ at position $\beta$, $O_1$ is $N$ reduced first at $\alpha$ then at $\alpha;\beta$, and $O_2$ is $N$ reduced first at $\alpha;@_1;\lambda;\beta$ then at $\alpha$
-
-        \item $N \mid \alpha = (\lambda x.
-A) B$, $B \mid \beta$ is a redex, $(\gamma_i)_i$ is a list of all the positions in $A$ where $A \mid \gamma = x$, ordered from left to right, $O_1$ is $N$ reduced first at $\alpha ; @_2 ; \beta$ then at $\alpha$, and $O_2$ is $N$ reduced first at $\alpha$ then at $\alpha;\gamma_i;\beta$ for each $i$ in order.
-
-        \item $N \mid \alpha = \tY (\lambda x.
-A)$, $A$ reduced at $\beta$ is $A'$, $(\gamma_i)_i$ is a list of all the positions where $A' \mid \gamma  = x$, ordered from left to right, $O_1$ is $N$ reduced first at $\alpha;\tY;\lambda;\beta$ then at $\alpha$, and $O_2$ is $N$ reduced first at $\alpha$ then at $\alpha;\lambda;@_1;\gamma_i;Y;\lambda;\beta$ for each $i$ in order where $\gamma_i$ is left of $\beta$ then at $\alpha;\lambda;@_1;\beta$ then at $\alpha;\lambda;@_1;\gamma_i;Y;\lambda;\beta$ for the remaining values of $i$.
-    \end{enumerate}
-    (in which case $O_1$ and $O_2$ are equal as skeletons, but with different reduction sequences), $O_1'$ and $O_2'$ are the results of applying some reduction sequence to each of $O_1$ and $O_2$ (the same reductions in each case, which is always possible because they're equal terms), and $\delta$ is a position in $O_1'$ (or equivalently $O_2'$), then $(O_1',\delta) \sim_c (O_2',\delta)$.
-\end{enumerate}
-The $\sim_c$-rules are illustrated in the following diagrams.
-\end{definition}
-
-\include{sim-diagram}
-
-The reflexive transitive closure $\sim^*$ of this relation is used to define the set of \emph{potential positions} $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable skeletons.
-If $(N,\alpha) \sim^* (O,\beta)$, then $N \mid \alpha$ and $O \mid \beta$ both have the same shape (i.e.~they're either both the placeholder $\skeletonPlaceholder$, both variables, both applications, both $\tsample$s etc.), therefore it's well-defined to talk of the set of potential positions where there is a $\tsample$, $L_s(M)$.
-Formally $L_s(M) := \{[(X, \alpha)]_{\sim^\ast_M} : X \mid \alpha = \tsample\}$.
-The new sample space is then defined as $I^{L_s(M)}$, with the Borel $\sigma$-algebra and product measure.
-\changed[lo]{Since $I^{L_s(M)}$ is a countable product, the measure space is well-defined \citep[Cor.~2.7.3]{AshDD00}.}
-
-\paragraph{}
-Although the six cases in the definition of $\sim_c$ are defined separately, there are some commonalities worth noting.
-Each case (except a, which is symmetrical) has a right branch and a left branch (the branches of case b are unfortunately displayed the wrong way around in the diagram).
-The right branch has reduction steps at two positions, $\beta$ then $\alpha$, where $\beta > \alpha$ (i.e.~$\beta$ is inside $\alpha$), and in the left branch, the order of these reductions are swapped, so the reductions are at $\alpha$ then $\beta_1, \dots, \beta_n$.
-The position $\alpha$ of one of the reductions is unchanged, and the other reduction may have multiple (in cases e and f) or 0 (in cases b and e) images in the left branch.
-The images $(\beta_i)$ are all still inside (greater than or equal to) $\alpha$, and although they are not generally equal to $\beta$, the subskeletons at these positions have a similar shape to the subskeleton (initially) at position $\beta$, so that the reductions at these positions are the same type of reduction (e.g.~both $\beta$-reduction or both $\textsf{if}$-reduction).
-Case a is also similar, except that the relevant positions are all disjoint rather than contained in one another, and either branch could be considered the right branch.
-
-\iffalse
-\changed[lo]{Since $L_s(M)$ is a countable set, there is a unique probability measure $\mu$ on the infinite product $\prod_{i \in L_s(M)} \Sigma_I$ of the Borel $\sigma$-algebras $\Sigma_I$ satisfying
-\[
-\mu\set{\omega \in \Omega \mid \omega_1 \in A_1, \cdots, \omega_n \in A_n} = \prod_{i=1}^n \Leb(A_i)
-\]
-where $\Leb$ is the Lebesgue measure on the measurable space $(I, \Sigma_I)$.}
-\fi
-
-\paragraph{}
-Before defining the new version of the reduction relation $\red$, a few lemmas about the properties of $\sim$ are necessary for it to be well-defined.
-
-\begin{lemma} \label{downAcrossUp}
-If $(N_1,\alpha_1) \sim^* (N_2,\alpha_2)$, there is some descendants $N'_1, N'_2$ of $N_1$ resp.~$N_2$, and a position $\alpha'$ in both of them such that $(N_1,\alpha_1) \sim_p^* (N'_1, \alpha') \sim_c^* (N'_2, \alpha') \sim_p^* (N_2, \alpha_2)$.
-\end{lemma}
-\begin{proof}
-  The $\sim_p$ relation can be split into $\sim_\downarrow \cup \sim_\uparrow$, where $(A,\alpha) \sim_\downarrow (B,\beta)$ if $A \to B$ and $(A,\alpha) \sim_p (B,\beta)$, and similarly $(A,\alpha) \sim_\uparrow (B,\beta)$ if $B \to A$ and $(A,\alpha) \sim_p (B,\beta)$.
-At each stage of this proof, it will be assumed that the $\sim_c$ steps and the $\sim_p$ steps in the sequence from $(N_1,\alpha_1)$ to $(N_2,\alpha_2)$ are rearranged such that there is never a $\sim_c$ immediately before a $\sim_\downarrow$, and there is never a $\sim_c$ immediately after a $\sim_\uparrow$.
-This rearrangement is always possible, because if there is some subsequence $(A,\alpha) \sim_c (B,\alpha) \sim_\downarrow (C,\beta)$, then there is an alternate path $(A,\alpha) \sim_\downarrow (B',\beta) \sim_c (C,\beta)$, where the reduction and the $\sim_\downarrow$ from $A$ to $B'$ are the same as those from $B$ to $C$, and the $\sim_c$ step is the same, but with the reduction sequences $O_1 \to^* O_1'$ and $O_2 \to^* O_2'$ in the definition of $\sim_c$ extended by the reduction $B \to C$.
-With these rearrangements assumed, it is sufficient to prove that the $\sim_p$ steps can be rearranged so that all of the $\sim_\downarrow$ steps come before all of the $\sim_\uparrow$ steps (possibly introducing some $\sim_c$ steps in the process).
-
-  The rearrangement to put all of the $\sim_\downarrow$s before all of the $\sim_\uparrow$s proceeds by induction on the number of $\sim_\downarrow$s that occur after the first occurrence of $\sim_\uparrow$.
-If it's 0, all of the $\sim_p$s are already in the correct order, and we're done.
-Otherwise, take the subsequence from the first $\sim_\uparrow$ to the first $\sim_\downarrow$ after that.
-This must be rearranged to some $\sim_\downarrow$s followed by some $\sim_\uparrow$s, then once that's done, the number of $\sim_\downarrow$s after the first $\sim_\uparrow$ in the overall sequence will have decreased by 1.
-
-  Let this subsequence be $A \sim_\uparrow^n B \sim_\downarrow C$.
-Suppose for induction that $A \sim_\uparrow^k (D,\delta) \sim_\downarrow^* (E,\epsilon) \sim_\uparrow^* C$ for some $D, \delta, E, \epsilon$, and $0 \leq k \leq n$, where in the reduction sequence $D \to^* E$, for any pair of reductions at positions which aren't disjoint, the reduction at the innermost (greater) position occurs first.
-Reduction sequences with this property will be called ``parallel'', as it is analagous to parallel reduction as in \akr{TODO: find this reference, for the proof of the Church-Rosser theorem by Tait and Martin-L\"of}.
-This induction is in reverse, with $k$ decreasing from $n$ to $0$.
-If $k = n$, simply set $(D,\delta) = B$ and $(E,\epsilon) = C$.
-If $k = 0$, then $A$ is related to $B$ by a sequence of $\sim_\downarrow$s then $\sim_\uparrow$s, as desired.
-In the intermediate steps, $k$ must be decreased by one, so a subsequence of one $\sim_\uparrow$ followed by a parallel sequence of $\sim_\downarrow$s must be replaced by a parallel sequence of $\sim_\downarrow$s followed by some $\sim_\uparrow$s.
-
-  For any individual $\sim_\uparrow$ then $\sim_\downarrow$ pair, let $(A,\alpha) \sim_\uparrow (B,\beta) \sim_\downarrow (C,\gamma)$.
-The skeleton $B$ reduces to both $A$ and $C$.
-If these are the same reduction, then $(A,\alpha) = (C,\gamma)$ and this subsequence may be removed.
-If they are different, then the way they overlap corresponds to one of the cases in the definition of $\sim_c$, either case a if the reduction positions are disjoint, or one of the cases b--f if one of the reduction positions is inside the other.
-In any case, there is a reduction sequence from each of $A$ and $C$ to some common skeleton $D$ ($O_1$ and $O_2$ in the definition of $\sim_c$).
-For the same reason that $(B,\beta) \sim_\downarrow (C,\gamma)$, there is some $\delta$ such that $(A,\alpha) \sim_\downarrow^* (D,\delta)$, and similarly for the same reason that $(B,\beta) \sim_\downarrow (A,\alpha)$, for the same $\delta$, $(C,\gamma) \sim_\downarrow^* (D,\delta)$.
-There are a lot of cases to check for this statement, but all of them are rather simple, and similar to each other.
-There is therefore an alternative $\sim^*$ sequence from $(A,\alpha)$ to $(C,\gamma)$, of the form $(A,\alpha) \sim_\downarrow^* (D_1,\delta) \sim_c (D_2,\delta) \sim_\uparrow^* (C,\gamma)$, where $D_1$ and $D_2$ are the alternative reduction sequences leading to the same skeleton $D$.
-There are only multiple $\sim_\downarrow$ steps in the result if the position of the reduction $B \to C$ is inside the position of the reduction $B \to A$, and similarly, there are only multiple $\sim_\uparrow$ steps in the result if the position of the reduction $B \to A$ is inside the position of the reduction $B \to C$.
-If there are multiple $\sim_\downarrow$ steps, they are not necessarily parallel, but in the only case where they aren't (case f of $sim_c$), the subsequence of $\sim_\downarrow$ steps with the reductions at $\alpha;\lambda;@_1;\beta$ then $\alpha;\lambda;@_1;\gamma_i;\tY;\lambda;\beta$ for each $i$ where $\gamma_i \geq \beta$ (in the notation of the definition of $\sim_c$) can be replaced by $\sim_\downarrow$s with the reductions at $\alpha;\lambda;@_1;\gamma'_i;\tY;\lambda;\beta$ then $\alpha;\lambda;@_1;\beta$, then another $\sim_c$, where $(\gamma'_i)$ is the list of positions in $A$ (\emph{before} the reduction at $\beta$ to $A'$, unlike $(\gamma_i)$) where $A | \gamma'_i = x$ and $\gamma'_i > \beta$.
-
-  The effect of this rearrangement is (ignoring the $\sim_c$s) to swap the order of the $\sim_\uparrow$ and the $\sim_\downarrow$, except that if the positions of one of these reductions is inside the other, that inner $\sim_p$ may be duplicated.
-If the positions are disjoint, both of them are unchanged (corresponding to case a of $\sim_c$), and if one is inside the other, then the outer position is unchanged and the other resultant positions, although they aren't equal to the original inner position, are still inside the outer position.
-Taking the sub-sequence of one $\sim_\uparrow$ followed by a parallel sequence of $\sim_\downarrow$s, this rearrangement can be repeatedly applied to move the $\sim_\uparrow$ further along the sequence.
-If at some point a $\sim_\uparrow$ and $\sim_\downarrow$ match (by relating the same (skeleton, position) pairs in opposite directions), they may be removed and the process may stop early.
-The $\sim_\uparrow$ may be duplicated if it passes a $\sim_\downarrow$ whose position is outside of the $\sim_\uparrow$'s position, but by the assumption that the sequence of $\sim_\downarrow$s is initially parallel, no position inside of this occurs later in the sequence, therefore all of the resultant $\sim_\uparrow$s pass the remaining $\sim_\downarrow$s without changing or duplicating them.
-The $\sim_\uparrow$s may be further duplicated, but the number of steps left in this process is bounded by the product, for all of the remaining $\sim_\downarrow$s, of $2 + \text{the number of occurrences of the variable relevant to the reduction}$, because for each $\sim_\downarrow$, that is more than the number of duplicates it could produce for any $\sim_\uparrow$ that passes it.
-It is also possible (earlier) for some of the $\sim_\downarrow$s to be duplicated, but the only case where this process would not terminate, that both the $\sim_\downarrow$s and the $\sim_\uparrow$s continue being duplicated forever so that the number of switches left to do never decreases, is prevented by the parallelness condition, as all of the duplications of $\sim_\downarrow$s occur before all of the duplications of $\sim_\uparrow$s.
-
-  It still remains to be shown that the sequence of $\sim_\downarrow$s left at the end of this can be made parallel.
-The changes that may have occurred since the previous version of this sequence (which was known to be parrallel already), are that some of the $\sim_\downarrow$s may have been duplicated by passing the $\sim_\uparrow$, and if that $\sim_\uparrow$ matches one of the $\sim_\downarrow$s, that $\sim_\downarrow$ is removed.
-The position of the $\sim_\uparrow$'s reduction is the same for all of these duplications, because it is not changed until later in the sequence, when the $\sim_\downarrow$s with positions outside of that may occur.
-The resultant positions after a duplication are in the same position relative to each other, and therefore the order of the remaining $\sim_\downarrow$s is automatically parallel, except that (letting the position of the $\sim_\uparrow$'s reduction be $\alpha$) if there are $\sim_\downarrow$s with reductions at positions $\alpha;@_1;\lambda;\beta$ and $\alpha;@_2;\gamma$, or $\alpha;\tY;\lambda;\beta$ and $\alpha;\tY;\lambda;\gamma$, some of the resultant $\sim_\downarrow$s' positions may be inside each other where they were originally not.
-Similarly to the case where a $\tY$ reduction and a reduction at a position inside it are swapped though, the order can be fixed by swapping some of the $\sim_\downarrow$ steps with each other.
-In the first case, the reduction at position $\alpha;@_1;\lambda;\beta$ ends up at $\alpha;\beta$, and the reduction at position $\alpha;@_2;\gamma$ ends up at the positions $\alpha;\delta_i;\gamma$, where $(\delta_i)$ is the positions of the variable relevant to the $\sim_\uparrow$'s reduction inside its lambda.
-The positions of the variable in the lambda may be different before the reduction at $\beta$, but the positions of the redexes corresponding to the original redex at $\alpha;@_2;\gamma$ are changed similarly.
-Rather than reducing at $\alpha;\beta$ then at $\alpha;\delta_i;\gamma$, it is therefore possible to reach the same point by reducing at $\alpha;\delta'_i;\gamma$ then at $\alpha;\beta$, where $(\delta'_i)$ is the positions of the relevant variable within the lambda before the reduction at $\alpha;@_1;\lambda;\beta$ (and again, a $\sim_c$ of some sort must also be introduced).
-The other case, that the $\sim_\uparrow$'s reduction is a $\tY$-reduction, is similar.
-Each reduction which is duplicated has one image at $\alpha;\lambda;@_1;$its original relative position, and one for each occurrence of the variable.
-Those corresponding to variable occurrences may need to be moved to before those at $\lambda;@_1$.
-
-  In summary, to reach the desired order of $\sim_p$ steps, each $\sim_\downarrow$ may have to be moved past some $\sim_\uparrow$ steps, and it may get duplicated in the process, but the resultant sequence of $\sim_\downarrow$s can all be put in parallel order, and it has been shown that moving a $\sim_\uparrow$ past a sequence of $\sim_\downarrow$s is always possible if they're in parallel order, therefore the overall process terminates and reaches a point where all of the $\sim_\downarrow$s precede all of the $\sim_\uparrow$s, and all of the $\sim_c$s are in between them.
-All of these rearrangements leave the end points of the sequence unchanged, therefore this is an alternative sequence of $\sim$ steps between $(N_1,\alpha_1)$ and $(N_2,\alpha_2)$ of the form $(N_1,\alpha_1) \sim_\downarrow^* (N_1', \alpha') \sim_c^* (N_2', \alpha') \sim_\uparrow^* (N_2,\alpha_2)$.
-\end{proof}
-
-If $(X,\alpha) \sim_c (Y,\alpha)$ then the final skeletons in $X$ and $Y$ are equal, and $(X,\beta) \sim_c (Y,\beta)$ for any $\beta$ that occurs in this skeleton, therefore $\sim_c$ can be considered to apply to reduction sequences even without positions, $X \sim_c Y$ iff $(X,\cdot) \sim_c (Y,\cdot)$.
-The structure of $\sim_c$ can be seen more clearly by choosing a canonical example from each equivalence class.
-For this, the member of the class that's in call-by-value order is used.
-A reduction sequence is in call-by-value order if for every pair of reductions in the sequence at positions $\alpha$ and $\beta$, where the reduction at $\alpha$ happens first, either $\alpha \leq \beta$ or $\alpha$ and $\beta$ are disjoint, with $\alpha$ to the left of $\beta$ (i.e.~the first elements of the sequences $\alpha, \beta$ where they differ are $@_1$ and $@_2$, $\underline f_i$ and $\underline f_j$ for $i < j$, or $\textsf{if}_i$ and $\textsf{if}_j$ for $i < j$, in that order).
-Note that although the reductions that occur in the sequence are in call-by-value order, it may not be the actual call-by-value reduction sequence starting from that point because some redexes may remain un-reduced even when other reductions that should happen afterwards occur.
-It is only required that the reductions that do occur occur in the correct order.
-
-\begin{lemma} \label{canonicalCousins}
-For any reduction sequence $X$, there is a unique reduction sequence $X_{cbv}$ that is related to $X$ by $\sim_c$ and is in call-by-value order, and if $X \sim_c^* Y$, then $X_{cbv} = Y_{cbv}$.
-\end{lemma}
-
-\lo{As defined, $\sim_c$ is not symmetric.}
-
-\begin{proof}
-
-Define $X_n$ recursively so that $X_0 = X$ and if $X_n$ is not in CBV order, take the last pair of reductions in $X_n$ which aren't in CBV order.
-They are adjacent in the sequence, as the CBV order is a total order.
-Let the positions of these reductions be $\alpha$ and $\beta$ respecively.
-If they are disjoint, they can be considered as one of the branches of $\sim_c$ case a.
-Otherwise, $\beta < \alpha$, and the sub-skeleton at position $\beta$ at the appropriate point in the reduction sequence is a redex with non-trivial subterms, therefore it's of one of the forms $\tif{\skeletonPlaceholder>0}{A}{B}$, $(\lambda x.A) B$ or $\tY \lambda x.
-A$, therefore the reductions $\alpha$ and $\beta$ form the $O_2$ branch of one of the cases b, c, d, e or f of $\sim_c$.
-In either case, define $X_{n+1}$ as equal to $X_n$ except taking the other (left) branch, so that the order of the reductions at $\alpha$ and $\beta$ are switched (although the equivalent(s) of the reduction at $\alpha$ may not actually be at that position).
-
-This sequence of swaps rearranges the reductions in $X$ rather like insertion-sort (but in some cases duplicating or deleting the new element), and eventually reaches some $X_n$ that is in CBV order.
-This is defined to be $X_{cbv}$.
-
-\paragraph{}
-If $Y \sim_c X \sim_c^* X_{cbv}$, and the reductions involved in $Y \sim_c X$ are not the last pair of reductions in $Y$ that aren't in CBV order, the order of the $\sim_c$s can be rearranged so that the sequence $Y \sim_c^* X_{cbv}$ is the canonical sequence given above, therefore $Y_{cbv} = X_{cbv}$.
-To be more precise about this rearrangement, consider the sequence of $\sim_c$ steps from $Y$ to $X_{cbv}$.
-The first step takes some sub-sequence of the reduction sequence $Y$ and swaps the order of those reductions.
-If $X = X_{cbv}$, then $X = Y_1 = Y_{cbv}$ already.
-Otherwise, consider the cases according to whether the image of this subsequence in $X$ occurs earlier than the next pair of reductions to be swapped (i.e.~the last pair of reductions in $X$ that occur not in CBV order).
-If not, then either the step $Y \sim_c X$ changes this subsequence into CBV order, or out of it.
-If it is into, then it was the last out-of-order pair in $Y$ therefore the whole sequence $Y \sim_c^* X_{cbv}$ is already in the canonical order and $Y_{cbv} = X_{cbv}$.
-If it switches the pair to the wrong order, then the result of that switch is the last out-of-order pair in $X$ therefore $Y = X_1$, and the sequence $X_1 \sim_c^* X_{cbv}$ is just a suffix of the sequence $X \sim_c^* X_{cbv}$ therefore $(X_1)_{cbv} = X_{cbv}$ and again $Y_{cbv} = X_{cbv}$.
-
-The remaining cases are when the subsequence of reduction steps involved in $Y \sim_c X$ occur before the last out-of-order pair in $X$, but possibly overlapping.
-If there is no overlap, then the $\sim_c$ steps do not interfere with each other at all, and can simply be performed in the other order.
-This gives a different sequence of $\sim_c$ steps $Y \sim_c Y_1 \sim_c X_1 \sim_c^* X_{cbv}$.
-As $Y_{cbv} = (Y_1)_{cbv}$, we can proceed by induction in this case and use the fact that $(Y_1)_{cbv} = (X_1)_{cbv} = X_{cbv}$ to acheive the result.
-
-In the other case, there is some overlap between the regions involved in $Y \sim_c X$ and $X \sim_c X_1$, but the last reduction in $X \sim_c X_1$ is later than the last reduction in $Y \sim_c X$.
-There are only 2 reductions in $X$ involved in $X \sim_c X_1$, therefore the region involved in $Y \sim_c X$ ends on the first of the reductions involved in $X \sim_c X_1$.
-Now consider the specific sequence of positions of the reductions in $X$ involved in either of these steps.
-Let the last of these positions be $\alpha$, the first $\gamma$, and the second-last, third-last and so on $\beta_1, \beta_2, \dots$ respectively, and let the positions of the redexes in $Y$ involved in $Y \sim_c X$ be $\beta$ then $(\gamma_i)$ in order..
-Then $X \sim_c X_1$ swaps $\beta_1$ and $\alpha$, and $\alpha$ comes before $\beta_1$ in CBV order (i.e.~$\alpha < \beta_1$ or $\alpha$ is left of $\beta_1$), and the step $Y \sim_c X$ either proceeds forwards towards CBV order, swapping $\beta$ below $\gamma_1 = \gamma$ to form $\gamma$ then $(\beta_i)$, or proceeds backwards, taking $Y$ even farther from CBV order, swapping the $(\gamma_i)$ to above $\beta$, forming $\gamma$ then $\beta_1 = \beta$.
-In each of the cases below, it is required to show that the canonical sequence $(X_i)_i$ eventually reaches some term which is the same as one in $(Y_j)_j$, and from that point on, they match, therefore they reach the same end result: $X_{cbv} = Y_{cbv}$.
-
-\begin{itemize}
-  \item $Y \sim_c X$ in the backwards direction (so that $Y$ is closer to CBV order than $X$ is): By swapping the roles of $X$ and $Y$, and $\beta$ and $\gamma$, this is equivalent to the $Y \sim_c X$ forwards case.
-The assumption that $\alpha$ is before $\beta_1$ in CBV order (so that $\beta_1, \alpha$ is the pair to be swapped in $X \sim_c X_1$) maps to the assumption that $\alpha$ is before $\gamma_1$ (so that the result doesn't just follow trivially by $Y_1$ being equal to $X$) and vice-versa.
-Then one of the cases below establishes that $X_i = Y_j$ for some $i$ and $j$, therefore after swapping back, $Y_i = X_j$.
-\item $Y \sim_c X$ forwards and $\alpha$ comes after $\gamma$ in CBV order: In this case, the $\gamma$ and $\alpha$ reduction steps are already in the correct order in $Y$, therefore $\beta$ and $\gamma$ are already the last out-of-order pair of reductions in $Y$, $Y_1 = X$, and $Y \sim_c X \sim_c^* X_{cbv}$ is already the canonical sequence from $Y$ therefore $Y_{cbv} = X_{cbv}$.
-\item $Y \sim_c X$ forwards and $\alpha$ is disjoint from $\gamma$ and $\beta$: In this case, the sequence $(X_i)_i$ starts by swapping $\alpha$ and $\beta_1$ by case a, then because $\gamma$ is disjoint from $\alpha$ and comes after is in CBV order, $\gamma$ is right of $\alpha$, therefore either all the remaining $\beta_i$s, are $\geq \gamma$ (if $\beta$ and $\gamma$ aren't disjoint) or there are no remaining $\beta_i$s (if $\beta$ and $\gamma$ are disjoint, in which case they swap by case a and there is only $\beta_1$).
-The sequence $(X_i)_i$ therefore proceeds to swap all the remaining $\beta_i$s then $\gamma$, in order, below $\alpha$, resulting in $X_i$ for some $i$ having as its relevant sub-reduction-sequence $\alpha, \gamma, \beta_k, \dots, \beta_1$.
-The other sequence $(Y_j)_j$ starts by swapping $\gamma$ then $\beta$ below $\alpha$, then swapping $\gamma$ and $\beta$.
-Because $\alpha$ is disjoint from $\gamma$, the reduction there does not affect the sub-skeleton at position $\gamma$, therefore this last $\sim_c$ step proceeds identically to how it did in $Y \sim_c X$, resulting in $\gamma, \beta_k, \dots, \beta_i$.
-Overall, this results in $Y_3$ having the relevant sub-reduction-sequence $\alpha, \gamma, \beta_k, \dots, \beta_1$, therefore $Y_3 = X_i$ for the aforementioned value of $i$.
-\item $Y \sim_c X$ forwards, $\alpha \leq \gamma$ and $\alpha$ (and therefore also $\gamma$) is disjoint from $\beta$: In this case, $(X_i)_i$ proceeds by swapping $\beta_1 = \beta$ below $\alpha$ by case a, then swapping $\gamma$ below $\alpha$ by one of the other cases, and $(Y_j)_j$ proceeds by swapping $\gamma$ and $\alpha$, then swapping $\beta$ below $\alpha$ then all of the images of $\gamma$ (which are $\geq \alpha$ and therefore left of $\beta$).
-In both cases, the subterm at $\alpha$ when $\alpha$ and $\gamma$ are swapped is unaffected by the reduction at $\beta$, therefore it produces the same result in both cases, therefore the overall sequence in both cases is $\alpha$ then the images of $\gamma$ then $\beta$.
-\item Both $\gamma$ and $\beta$ are $> \alpha$ but they're disjoint from each other, and either the skeleton at position $\alpha$ is of the form $\tif{\skeletonPlaceholder > 0}{A}{B}$ or it is of the form $A B$ and either both $\gamma$ and $\beta$ are $> \alpha;@_1$ or they are both $> \alpha;@_2$: In this case, $(X_i)_i$ proceeds by swapping $\beta$ then $\gamma$ below $\alpha$, in each case possibly forming 0 or multiple images.
-The only way the number of images of each of these positions can be different is if the subskeleton at $\alpha$ is an $\textsf{if}$, and there are 0 of one of them and 1 of the other, in which case they're trivially in the correct order already.
-Otherwise, all of the images of $\beta$ and $\gamma$ are disjoint from one another, therefore none of the positions, or their relative order, change when they are swapped, therefore the next part of the sequence $(X_i)_i$ is just insertion-sort running on the images of $\beta$ and $\gamma$ with $\sim_c$ case a swaps until they're in the correct order.
-The sequence $(Y_j)_j$ starts by swapping $\gamma$ then $\beta$ below $\alpha$, then as before, sorting the images into the correct order by case a swaps.
-The images originally formed of $\gamma$ and $\beta$ are the same in both cases, therefore the final order is the same, so $X_i = Y_j$ for some $i$ and $j$.
-\item $Y \sim_c X$ forwards, $\alpha;@_1 < \gamma$, and $\alpha;@_2 < \beta$: Let $x$ be the variable involved in the $\beta$-reduction at $\alpha$, so that the sub-skeleton at $\alpha$ is $(\lambda x.
-A) B$ for some $A, B$, where $\gamma = \alpha;@_1;\lambda;\gamma'$ and $A \to A'$ at $\gamma'$, and $\beta = \alpha;@_2;\beta'$, and $B \to B'$ at $\beta'$.
-In this case, $(X_i)_i$ starts by swapping $\beta$ below $\alpha$ by case e, producing one image of $\beta$ for each $x$ in $A'$, then swapping $\gamma$ below $\alpha$ by case d, producing a single image $\alpha;\gamma'$.
-For each of the instances of $x$ in $A$ left of $\gamma'$, the reduction at $\alpha;\gamma'$ is then swapped below the corresponding image of $\beta$.
-The sequence $(Y_j)_j$ starts by swapping $\gamma$ below $\alpha$ then $\beta$ below $\alpha$, but in this case the images of $\beta$ produced may be different.
-In this case the swap happens earlier in the reduction sequence than $\gamma$, so that the body of the lambda is still $A$ rather than $A'$, and the reduction at $\gamma$ may rearrange or change the number of instances of $x$.
-Next, each of the images of $\beta$ to the right of $\alpha;\gamma'$ are swapped with it by case a.
-At this point, the next few reductions before $\alpha;\gamma'$ will in general be images of $\beta$ at positions inside $\alpha;\gamma'$.
-The subskeleton at position $\alpha;\gamma'$ before these reductions is $(A|\gamma')[B/x]$, then by the images of $\beta$ this reduces to $(A|\gamma')[B'/x]$, then it reduces at its root position ($\alpha;\gamma'$ in the overall skeleton) to $(A'|\gamma')[B'/x]$.
-For each position of an $x$ in $(A|\gamma')$, there are 0 or more corresponding positions of $x$ in $(A'|\gamma')$, depending on what type of reduction $A \to A'$ is.
-Because $B$ and $B'$ cannot contain any instances of the variable involved in the reduction $A \to A'$ (if it's a $\beta$-reduction or $\tY$-reduction), each instance of $B$ in $(A|\gamma')[B/x]$ has the same set of images in $(A'|\gamma')[B/x]$ as the corresponding $x$ in $(A|\gamma')$.
-When these images of $\beta$ that overlap with the image of $\gamma$ are swapped below it, each of them has its own images, one at each position of $B$ in $(A'|\gamma')[B/x]$ corresponding to its original copy of $B$ in $(A|\gamma')[B/x]$.
-After all of them are swapped below $\alpha;\gamma'$ (and rearranged among themselved by case a), there is therefore one reduction at each copy of $B$ in $(A'|\gamma')[B/x]$, i.e.~one at position $\beta'$ relative to each $x$ in $(A'|\gamma')$.
-They are therefore the same as the images of $\beta$ in $X_1$ that overlap with $\alpha;\gamma'$, because those are also one reduction at position $\beta'$ relative to each $x$ in $(A'|\gamma)$.
-Both $(X_i)_i$ and $(Y_j)_j$ therefore eventually reach a point where the relevant sub-reduction-sequence is $\alpha$, the images of $\beta$ to the left of $\alpha;\gamma'$, $\alpha;\gamma'$ itself, the aforementioned images of $\beta$ that overlap with $\alpha;\gamma'$, then all the images of $\beta$ to the right of $\alpha;\gamma'$.
-\item $Y \sim_c X$ forwards, $\gamma$ and $\beta$ are disjoint and both $> \alpha$, and the reduction at $\alpha$ is a $\tY$-reduction: This is similar to the previous case, but more complicated, so it will need some definitions to explain properly what's going on.
-Let the term at $\alpha$ before any of the reductions be $\tY \lambda x.
-A$, let $\gamma = \alpha;\tY;\lambda;\gamma'$, let $\beta = \alpha;\tY;\lambda;\beta'$, let $A | \beta' = B \to B'$ with the reduction at $\cdot$, let $A | \gamma' = C \to C'$ with the reduction at $\cdot$, let the positions in $A$ where $x$ occurs that are left of $\gamma'$, between $\gamma'$ and $\beta'$ (but still disjoint from both), and right of $\beta'$ respectively be $(\delta^l_i)$, $(\delta^m_i)$ and $(\delta^r_i)$, let the positions in $B$, $C$, $B'$ and $C'$ respectively where $x$ occurs be $(\delta^B_i)$, $(\delta^C_i)$, $(\delta^{B'}_i)$ and $(\delta^{C'}_i)$ (so that all the positions in $A$ where $x$ occurs, in left-to-right order, are $(\delta^l_i)_i, (\delta^C_i)_i, (\delta^m_i)_i, (\delta^B_i)_i, (\delta^r_i)_i$).
-The sequence $(X_i)_i$ starts by swapping $\beta$ then $\gamma$ below $\alpha$.
-With the shorthand that $\gamma'(\epsilon) = \alpha;\lambda;@_1;\epsilon;\tY;\lambda;\gamma'$, $\gamma'_0 = \alpha;\lambda;@_1;\gamma'$, and similarly for $\beta'$, the relevant portion of $X_2$ is then $\alpha$, $(\gamma'(\delta^l_i))_i$, $\gamma'_0$, $(\gamma'(\delta^{C'}_i))_i$, $(\gamma'(\delta^m_i))_i$, $(\gamma'(\delta^B_i))_i$, $(\gamma'(\delta^r_i))_i$, $(\beta'(\delta^l_i))_i$, $(\beta'(\delta^{C'}_i))_i$, $(\beta'(\delta^m_i))_i$, $\beta'_0$, $(\beta'(\delta^{B'}_i))_i$, $(\beta'(\delta^r_i))_i$.
-Next in $(X_k)_k$, for each $i$, $\gamma'(\delta^r_i)$ are swapped past all the images of $\beta$ until it is immediately before $\beta'(\delta^r_i)$ by case a, then for each $i$, $\gamma'(\delta^B_i)$ is swapped past all the images of $\beta$ until $\beta'_0$ by case a, then swapped with $\beta'_0$ by one of the other cases depending on what type of reduction $B \to B'$ is and the relative positions.
-Even if it's case d or f, all of the images of $\gamma'(\delta^B_i)$ are disjoint from each other (and from the positions of the other reductions after $\beta'_0$) because the redex, $C$, doesn't contain any instances of the variable involved in the reduction $B \to B'$.
-Expanding the definitions, this is swapping $\alpha;\lambda;@_1;\delta^B_i;\tY;\lambda;\gamma'$ with $\alpha;\lambda;@_1;\beta'$ at a point in the reduction sequence where the subskeleton at $\alpha;\lambda;@_1;\beta'$ is $B$ with a mixture of $\tY \lambda x.
-A$ and $\tY \lambda x.
-A[C'/\gamma']$ substituted for its occurrences of $x$, and one of these occurrences of $x$ is at $\delta^B_i$.
-For each $\delta^B_j$, there are some corresponding $\delta^{B'}_k$, where the images of the $x$ at $\delta^B_j$ end up after the reduction $B \to B'$.
-These cover all the $\delta^{B'}_k$, and uniquely, and for each $\delta^B_i$, the images of $\zeta;\delta^B_i;\theta$ after swapping it with $\zeta;\beta'$ are precisely $(\zeta;\delta^{B'}_k;\theta)$ for those same values of $k$, therefore after swapping all of the $(\gamma'(\delta^B_i))_i$ down past $\beta'_0$, their images are $(\gamma'(\delta^{B'}_i))_i$.
-After swapping with $\beta'_0$, each of these images than swaps by case a to take its place among the $(\beta'(\delta^{B'}_i))_i$.
-Next up in $(X_i)_i$, the $(\gamma'(\delta^m_i))_i$s and the $(\gamma'(\delta^{C'}_i))_i$s swap down past some images of $\beta$ they're disjoint from to take their place before their matching image of $\beta$, then $\gamma'_0$ swaps past the $(\beta'(\delta^l_i))_i$, stopping immediately before $\beta'(\delta^{C'}_1)$, then the $(\gamma'(\delta^l_i))_i$ and the $(\beta'(\delta^l_i))_i$ mix.
-The end result of the rearrangement of (this subsequence of) the reduction sequence is therefore $\alpha$, $(\gamma'(\delta^l_i), \beta'(\delta^l_i))_i$, $\gamma'_0$, $(\gamma'(\delta^{C'}_i), \beta'(\delta^{C'}_i))_i$, $(\gamma'(\delta^m_i), \beta'(\delta^m_i))_i$, $\beta'_0$, $(\gamma'(\delta^{B'}_i), \beta'(\delta^{B'}_i))_i$, $(\gamma'(\delta^r_i), \beta'(\delta^r_i))_i$.
-
-  The other sequence, $(Y_j)_j$, proceeds similarly, but with the images of $\gamma$ starting out after the images of $\beta$.
-It swaps all the images of $\beta$ that are right of $\gamma_0$ to the appropriate places by case a, then swaps all the $(\beta'(\delta^C_i)_i$ first to then past $\gamma'_0$, resulting in $(\beta'(\delta^{C'}_i))_i$ for the same reason as with swapping those images of $\gamma$ that overlapped with $\beta'_0$ past it in the $(X_i)_i$ case above, then finally swapping the $(\beta'(\delta^l_i))_i$s and the $(\gamma'(\delta^l_i))_i$s into the correct order.
-As required, this produces the same result as in $(X_i)_i$.
-  \item $Y \to X$ forwards, and $\alpha < \gamma < \beta$: Let $\gamma'$ be such that $\gamma = \alpha;\textsf{if}_i;\gamma'$, $\alpha;@_1;\lambda;\gamma'$, $\alpha;@_2;\gamma'$ or $\alpha;\tY;\lambda;\gamma'$, so that $\gamma'$ is the freely varying later part of $\gamma$ as in the definition of $\sim_c$, whichever case applies to swapping $\gamma$ and $\alpha$.
-Similarly, let $\beta'$ be the freely varying part of $\beta$ relative to $\gamma$.
-Let the images of $\gamma$ after swapping it with $\alpha$ (where this swap occurs later in the reduction sequence than $\beta$) then be $(\alpha;\delta_i;\gamma')_i$, and the images of $\beta$ after swapping it with $\gamma$ be $(\gamma;\epsilon_i;\beta')_i$.
-Depending on the skeleton at $\alpha$, and $\gamma$'s position within it, $(\delta_i)_i$ may be empty (case b), a singleton containing $\cdot$ (cases c and d), all the positions of the relevant variable in the lambda in the skeleton at $\alpha$ (case e), or $\lambda;@_1$ and $\lambda;@_1;\zeta;\tY;\lambda$ for each position $\zeta$ of the relevant variable (case f), but in all cases (except a, which is excluded because none of $\alpha, \gamma$ and $\beta$ are disjoint) the set of images has this general structure.
-Furthermore, the positions $(\delta_i)_i$ are determined only by the general position of $\gamma$ within $\alpha$ (the part excluded from $\gamma'$), and the positions of the relevant variable within the skeleton at $\alpha$.
-They do not depend on $\alpha$ or $\gamma$, so that if both initial positions were moved somewhere else the relative positions of the images of $\gamma$ would be unaffected, and similarly if some position within $\gamma$ were swapped with $\alpha$, the relative positions of its images would be the same (except to the extent that, in case f, the positions of the relevant variable are taken after the inner reduction takes place, so that some of those may still change).
-Because there are so many of them and they don't actually affect the multiset of positions where reductions take place, I will be ignoring swaps by case a in the proof of this case, and just assuming that all the final positions end up in the correct order.
-
-  The final set of reduction positions that I will be proving both $(X_i)_i$ and $(Y_j)_j$ reach is $\alpha, (\alpha;\delta_i;\gamma', (\alpha;\delta_i;\gamma';\epsilon_j;\beta')_j)_i$, where as usual the sequences are expanded out to the full list.
-For the sequence $(X_i)_i$, first the images of $\beta$ in $X_0$ are $(\gamma,\epsilon_i,\beta')$.
-For each of these in turn, it is swapped with $\alpha$.
-By the fact mentioned above that sub-positions are mapped to the same set of images except in case f, the images of $\gamma,\epsilon_j,\beta'$ after this swap are $(\alpha;\delta_i^j;\gamma',\epsilon_j,\beta')_i$, where $(\delta_i^j)_i$ is the same as $(\delta_i)_i$ except that in the case where this is not the first image of $\beta$ swapped below $\alpha$ and the reduction at $\alpha$ is a $\tY$-reduction (and therefore swaps with it proceed by case f), the positions of the relevant variable are taken after the reductions at $\gamma$ and $\gamma;\epsilon_k;\beta'$ for $k \leq j$ (this implies that the first of these, $\delta_i^n = \delta_i$).
-All the images $\alpha;\delta_i^j;\gamma';\epsilon_j;\beta'$ where $\delta_i^j > \lambda;@_1;\gamma';\epsilon_{j+1};\beta'$ are then swapped with $\alpha;\lambda;@_1;\gamma';\epsilon_{j+1};\beta'$ (which is $\alpha;\delta_i^{j+1};\gamma';\epsilon_{j+1};\beta'$ for some $i$).
-As in the case where $\alpha;\tY < \gamma, \beta$ and $\gamma$ and $\beta$ are disjoint, the effects of these swaps may duplicate or delete the reductions in such a way that all the remaining images of $\gamma;\epsilon_j;\beta'$ are $(\alpha;\delta_i^{j+1};\gamma';\epsilon_j;\beta'$.
-This is repeated with $\delta_i^{j+2}$ and so on until these images are all in the correct order, at which point they are $(\alpha;\delta_i;\gamma';\epsilon_j;\beta')_i$.
-After all of these are done, $\gamma$ is swapped with $\alpha$, yielding $(\alpha;\delta_i^{-1};\gamma')_i$.
-As with the images of $\beta$, the subset of these which are inside $\alpha;\lambda;@_1;\gamma';\epsilon_j;\beta'$ for each $j$ in turn are swapped with it, until they are all in the correct order and the images left of $\gamma$ are $(\alpha;\delta_i;\gamma')_i$.
-At this point, $X_i$ has the desired value.
-
-  The sequence $(Y_j)_j$ proceeds similarly.
-First $\gamma$ swaps with $\alpha$, yielding $\alpha$ and $(\alpha;\delta_i;\gamma')_i$, then $\beta$ swaps with $\alpha$, yielding one image of $\beta$ for each $\delta_i$, except that, in case f, those images at positions inside $\alpha;\lambda;@_1;\gamma'$ may differ because they are earlier in the reduction sequence than any of the images of $\gamma$.
-Each of the images of $\beta$ in turn it moved to the corresponding image of $\gamma$, except that those overlapping with $\alpha;\lambda;@_1;\gamma'$ may be duplicated or deleted on the way.
-The images of $\beta$ after just this process (which does not actually correspond to any term in $(Y_j)_j$, but it's sufficiently independent that the order doesn't matter that much) are $(\alpha;\delta_i;\gamma';\beta'')$ (where $\gamma;\beta'' = \beta$).
-The images of these for each $i$, after swapping with $\alpha;\delta_i;\gamma'$, is $(\alpha;\delta_i;\gamma';\epsilon_j;\beta')_j$, because the reductions produced by a swap are unaffected by its overall location, and the skeleton at $\alpha;\delta_i;\gamma'$ at the relevant point in the sequence is equal to the skeleton at $\gamma$ initially, therefore this is equivalent to immediately swapping $\beta$ and $\gamma$ as in $Y \sim_c X$ (except that in the case where the reduction at $\alpha$ is a $\tY$-reduction, the skeleton at $\alpha;\lambda;@_1;\gamma'$ is not actually equal to the skeleton initially at $\gamma$ because something was substituted in for the variable bound at $\alpha;\tY$, but this doesn't affect the variable involved in the reduction at $\gamma$, so this doesn't actually matter).
-Therefore also in this case, eventually the set of reductions in the relevant portion of $Y_j$ is $\alpha, (\alpha;\delta_i;\gamma', (\alpha;\delta_i;\gamma';\epsilon_j;\beta')_j)_i$, therefore it is equal to some $X_i$.
-\end{itemize}
-In summary, if the swap in $Y \sim_c X$ doesn't overlap with the swap $X \sim_c X_0$, the sequence of swaps can be rearranged until it does, and in any other case, there is some $Y_j$ that's equal to some $X_i$, therefore these sequences reach the same end point and if $Y \sim_c X$, then $Y_{cbv} = X_{cbv}$, and by chaining these together, if $Y \sim_c^* X$, $Y_{cbv} = X_{cbv}$.
-
-\paragraph{}
-If $Y$ is in CBV order and $Y \sim_c^* X$, then $Y_{cbv} = X_{cbv}$ but also $Y_{cbv} = Y$ because the sequence $(Y_i)_i$ terminates immediately therefore $X_{cbv}$ is the unique reduction sequence in call-by-value order that's related to $X$ by $\sim_c^*$.
-\end{proof}
-
-\begin{lemma}
-\label{lem:sim-M-N}
-The relation $\sim$ is defined on $L_0(M)$ with reference to a particular starting term $M$, so different versions, $\sim_M$ and $\sim_N$, can be defined starting at different terms.
-If $M \to N$, then $\sim^*_N$ is equal to the restriction of $\sim^*_M$ to $L_0(N)$.
-\end{lemma}
-
-\begin{proof}
-$\sim_N^*$ is trivially a subset of $\sim_M^*$ because $\sim_N$ is a subset of $\sim_M$.
-
-In the other direction, suppose $(N_1, \alpha_1) \sim_M^* (N_2, \alpha_2)$ where both $N_1$ and $N_2$ are descendants of $N$.
-By Lemma~\ref{downAcrossUp}, take $(N_1, \alpha_1) \sim_{p,M}^* (N'_1, \alpha') \sim_{c,M}^* (N'_2, \alpha') \sim_{p,M}^* (N_2, \alpha_2)$.
-The $\sim_p^*$ steps remain within $Rch(N)$, and $\sim_p$ does not depend on the history of the reduction sequences, therefore $(N_1, \alpha_1) \sim_{p,N}^* (N'_1, \alpha') \sim_{c,M}^* (N'_2, \alpha') \sim_{p,N}^* (N_2, \alpha_2)$.
-
-Let $X$ be the call-by-value reduction sequence related to both $N'_1$ and $N'_2$ by $\sim_{c,M}$ given by Lemma~\ref{canonicalCousins}. 
-As $M \to N$ is the first reduction in the sequence $N'_1$, it is the last to be affected by the $\sim_{c,M}$ sequence $N'_1 \sim_{c,M}^* X$ given by Lemma \ref{canonicalCousins} therefore it can be split into $N'_1 \sim_{c,M}^* Y_1 \sim_{c,M}^* X$ where $Y_1$ is in CBV order except possibly for its first reduction, which is still $M \to N$.
-Let the position of the reduction $M \to N$ be $\beta$.
-\akr{TODO: fill in the details that are missing here. It shouldn't be too hard, it's just a bit sketchy at present.} 
-The rearrangement of the reduction sequence $Y_1 \sim_{c,M} X$ consists of moving the reduction at $\beta$ down past the other reductions in $Y_1$, possibly duplicating or deleting it in the process, but not affecting the positions or the correct order for any of the other reductions.
-The reductions derived from $M \to N$ can be identified as follows:
-
-%\begin{definition}
-A position in some (reduction sequence of) term(s) is \emph{derived from} another if it is related by the reflexive transitive closure of $\rightsquigarrow$, where $(V,\gamma) \rightsquigarrow (W,\delta)$ just if $V \to W$ and one of the following cases holds:
-\begin{itemize}
-    \item $(V,\gamma) \sim_p (W,\delta)$
-    \item $\gamma = \delta$, $V \to W$ at $\epsilon$, and $\epsilon > \gamma$
-    \item $V \mid \epsilon = (\lambda x.U) Z, U | \zeta = x, \gamma = \epsilon;@_2;\theta$ and $\delta = \epsilon;\zeta;\theta$
-    \item $V \mid \epsilon = \tY (\lambda x.U), \gamma = \epsilon;\tY;\lambda;\zeta$ and $\delta = \epsilon;\lambda;@_1;\zeta$
-    \item $V \mid \epsilon = \tY (\lambda x.U), U | \zeta = x, \gamma = \epsilon;\theta$ and $\delta = \epsilon;\lambda;@_1;\zeta;\theta$
-\end{itemize}
-%\end{definition}
-
-Crucially, the reductions in $X$ can be partitioned into 2 sets: those at positions derived from $(M,\beta)$, and those at positions equal to the reductions in $Y_1$ (in the same order as they occur in $Y_1$).
-Using the same construction for $N'_2$ shows that the positions of the reductions in $Y_2$ are also the positions of the reductions in $X$ other than those derived from $(M,\beta)$, therefore $Y_1 = Y_2$ therefore $N'_1 \sim_{c,M} Y_1 \sim_{c,M} N'_2$ and every reduction sequence in this sequence starts with $M \to N$ therefore $N'_1 \sim_{c,N} N'_2$.
-
-Combining this with the $\sim_{p,N}^*$s at the beginning and end then yields the desired result that $(N_1,\alpha_1) \sim_N^* (N_2, \alpha_2)$, therefore the restriction of $\sim_M^*$ to $\sim_N$'s domain ($Rch(N)$) is a subset of $\sim_N^*$ therefore the two versions of $\sim$ match on this domain, as desired.
-\end{proof}
-
-At each reduction step $M \to N$, the sample space must be restricted from $I^{L_s(M)}$ to $I^{L_s(N)}$. 
-The injection $L_0(N) \to L_0(M)$ is trivial to define by appending $Sk(M) \to Sk(N)$ to each path, and using Lemma~\ref{lem:sim-M-N}, this induces a corresponding injection on the quotient, $L(N) \to L(M)$.
-The corresponding map $L_s(N) \to L_s(M)$ is then denoted $i(M \to N)$.
-
-Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function, so it is denoted ``$\Rightarrow$'' instead of ``$\red$'', and it relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
-\changed[lo]{We write an element of $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ as $(M', s)$ where the \emph{term} $M' \in \Lambda^0$ and $s \in  I^{L_s(M')}$.}
-\begin{align*}
-& (M,s) \Rightarrow (N,s \circ i(M \to N)) \text{ if either} \\
-& \qquad \text{$M \to N$ and the redex is not $\tsample$, or} \\
-& \qquad \text{$M \mid \alpha = \tsample$, $N = M[\underline{s(Sk(M),\alpha)}/\alpha]$ and $\lambda$ does not occur after $@_2$ or $\tY$ in $\alpha$}
-\end{align*}
-
-\paragraph{}
-\begin{lemma} \label{churchRosser}
-The relation $\Rightarrow$ is Church-Rosser.
-\end{lemma}
-\begin{proof}
-Suppose that $(M,s) \Rightarrow^* (M_1,s_1)$ and also $(M,s) \Rightarrow^* (M_2,s_2)$, then it is required to prove that there is some $(M',s')$ such that both $(M_1,s_1) \Rightarrow^* (M',s')$ and $(M_2,s_2) \Rightarrow^* (M',s')$.
-First consider the special case where $(M,s) \Rightarrow (M_1,s_1)$ and $(M,s) \Rightarrow (M_2,s_2)$, with only a single step in each case.
-Let the positions of the redexes in $M \to M_1$ and $M \to M_2$ be $\alpha_1$ and $\alpha_2$ respectively.
-
-First consider the case that $\alpha_1$ and $\alpha_2$ are disjoint.
-Let $M_1 = M[X_1/\alpha_1]$ and similarly for $X_2$, then let $M' = M[X_1/\alpha_1][X_2/\alpha_2]$.
-As the positions are disjoint, the substitutions commute and both $M_1$ and $M_2$ reduce (with $\to$) to $M'$.
-Let $s' = s \circ i(M \to M_1) \circ i(M_1 \to M')$. 
-The injection $i(M \to M_1) \circ i(M_1 \to M')$ consists of prepending $M \to M_1 \to M'$ to each reduction sequence, but by case a of $\sim_c$, this is equivalent to prepending $M \to M_2 \to M'$, which is $i(M \to M_2) \circ i(M_2 \to M')$.
-In the case that the reduction $M \to M_2$ isn't a $\tsample$-reduction, this is enough to establish that $(M_1, s_1) \Rightarrow (M', s')$ (and similarly if the redex of $M \to M_1$ isn't $\tsample$, $(M_2,s_2) \Rightarrow (M',s')$).
-If it is $\tsample$ though, in order for it to be the case that $(M_1, s_1) \Rightarrow (M', s')$, it is additionally necessary that $X_2$, the result of the reduction at $\alpha_2$, be $s_1(M_1, \alpha_2)$, which follows from the fact that $(M, \alpha_2) \sim (M_1, \alpha_2)$ by case 1 of $\sim_p$.
-The case that the reduction $M \to M_1$ is a $\tsample$-reduction is similar.
-
-The case that $\alpha_1 = \alpha_2$ is trivial, because there is at most one possible $\Rightarrow$ reduction at any given position, therefore $(M_1, s_1) = (M_2, s_2)$ already.
-
-The remaining case is that $\alpha_1 < \alpha_2$ or $\alpha_1 > \alpha_2$. 
-Assume without loss of generality that $\alpha_1 < \alpha_2$. 
-For each possible case of what type of redex $M | \alpha_1$ is, and $\alpha_2$'s position within it, there is a corresponding case of $\sim_c$, and similarly to the case where $\alpha_1$ and $\alpha_2$ are disjoint, the term $O_1 = O_2$ from the definition of $\sim_c$ is a suitable value of $M'$.
-The term $M \mid \alpha_1$ can't be $\tsample$, because it has strict subterms, but the case that $M \mid \alpha_2 = \tsample$ is still somewhat more complicated. 
-$\alpha_2$ can't be within $\alpha_1 ; @_2$ or $\alpha_1 ; Y$ (cases e and f of $\sim_c$) because those are not valid positions to reduce a $\tsample$, 
-and the case that $\alpha_2$ is within the branch of an if statement that is deleted by the reduction at $\alpha_1$ (case b) doesn't actually present a problem because there is no corresponding reduction to $M \to M_2$ in the other branch,
-which leaves cases c and d, that $\alpha_2$ is in the $\tif{\cdot}{\cdot}{\cdot}$ branch that isn't deleted, and that $\alpha_2 < \alpha_1 ; @_1 ; \lambda$.  
-The values that the $\tsample$s take in these cases match because $(M,\alpha) \sim_p (M_1, \alpha;\beta)$ by cases 3 and 2 of $\sim_p$ respectively.
-
-\paragraph{}
-In the case that $(M,s) \Rightarrow^* (M_1,s_1)$ or $(M,s) \Rightarrow^* (M_2,s_2)$ by multiple steps, it is possible to repeatedly replace a pair of the form $A \Leftarrow B \Rightarrow C$ by $A \Rightarrow^* D \Leftarrow^* C$ by the construction above, but it is not immediate that this process terminates, because each $\Rightarrow$ or $\Leftarrow$ may be replaced by multiple, so the sequence of $\Rightarrow$s and $\Leftarrow$s from $(M_1,s_1)$ to $(M_2,s_2)$ may get longer at some steps.
-However, termination can be proved by noting that the structure of this process is identical to the process of swapping $\sim_\uparrow$ and $\sim_\downarrow$ steps in Lemma \ref{downAcrossUp}.
-To be more precise, in this case there is a sequence of $\Leftarrow$ and $\Rightarrow$ steps, each of which has an associated reduction position and initial and final terms, and if a $\Leftarrow$ immediately precedes a $\Rightarrow$, they may be swapped to produce some number of $\Rightarrow$s, followed by some number of $\Leftarrow$s.
-In the case of Lemma \ref{downAcrossUp}, there is a sequence of $\sim_\uparrow$ and $\sim_\downarrow$ steps, each of which has an associated reduction position and initial and final skeletons, and if a $\sim_\uparrow$ immediately precedes a $\sim_\downarrow$, they may be swapped to produce some number of $\sim_\downarrow$s followed by some number of $\sim_\uparrow$s.
-In both cases, the number and reduction positions of the resultant steps are determined by the case of $\sim_c$ that matches the way the initial reduction positions overlap, and the initial skeleton (or the skeleton of the initial term).
-The same argument that the process in Lemma \ref{downAcrossUp} terminated is therefore applicable here too.
-At every stage, the sequence of reductions that results from one of the initial $\Rightarrow$s is a parallel sequence, and swapping a parallel sequence of $\Rightarrow$s with a $\Leftarrow$ always terminates, therefore each $\Rightarrow$ in turn can be moved past all of the $\Leftarrow$s, and the process as a whole will terminate in a state where all of the $\Rightarrow$s precede all of the $\Leftarrow$s, i.e.~a pair of reduction sequences $(M_1,s_1) \Rightarrow^* (M',s') \Leftarrow^* (M_2,s_2)$.
-\end{proof}
-
-\begin{lemma} \label{simpDoesn'tMix}
-If $A$ is some descendant of $M$ and $(A,\gamma) \sim^* (M,\delta)$, then $(A,\gamma) \sim_p^* (M,\delta)$, with the length of the reduction sequences decreasing by one each step from $A$ to $M$.
-\end{lemma}
-\begin{proof}
-This is a simple induction on $\sim^*$.
-In the base case, $(A,\gamma) = (M,\delta)$ therefore $(A,\gamma) \sim_p^* (M,\delta)$ trivially.
-Otherwise, suppose that $(M,\delta) \sim_p^* (B,\epsilon) \sim (A,\gamma)$.
-Either $(B,\epsilon) \sim_p (A,\gamma)$ or $(B,\epsilon) \sim_c (A,\gamma)$.
-In the first case, either $B \to A$, in which case the result follows directly, or $A \to B$, in which case the fact that each (reduction sequence of) skeleton(s) has only one parent implies that a $(M,\delta) \sim_p^* (A,\gamma)$ directly as a subsequence of the path to $(B,\epsilon)$.
-
-In the $\sim_c$ case, consider the definition of $\sim_c$.
-Either $O'_1 = A$ and $O'_2 = B$ or vice-versa.
-As $M \to^* N \to^* O_1 \to^* B$ and $(M,\delta) \sim_p^* (B,\epsilon)$, and $\sim_p$ only relates positions in a term and its parent, there are some positions $\zeta, \theta$ such that $(M,\delta) \sim_p^* (N,\zeta) \sim_p^* (O_1, \theta) \sim_p^* (B,\epsilon)$.
-It follows that $(O_2, \theta) \sim_p^* (A,\gamma)$ by following the same path, therefore it suffices to provide the only missing portion of the path from $M$ to $A$, i.e.~to prove that $(N,\zeta) \sim_p^* (O_2,\theta)$ given $(N,\zeta) \sim_p^* (O_1,\theta)$ (or vice-versa).
-
-If $\zeta$ is disjoint from all the positions of reduction from $N$ to $O_1$ and $O_2$ (and consequently $\zeta = \theta$), this follows from case 1 of $\sim_p$.
-Otherwise, this can be proved by taking cases from the definition of $\sim_c$.
-This is rather long, but all of the cases are similar.
-The general idea is that the reductions from $N$ to $O_1$ correspond to the reductions from $N$ to $O_2$, so that if a position is related by $\sim_p$ across that reduction, it is related in the other branch for the same reason.
-Case d, where $B = O'_1$ rather than $O'_2$, is given here in more detail as an illustrative example:
-
-Let $I$ be $N$ reduced at $\alpha$, and $J$ be $N$ reduced at $\alpha;@_1;\lambda;\beta$, so that $N \to I \to O_1$ and $N \to J \to O_2$.
-All of the reduction positions are $\geq \alpha$, and $\zeta$ is not disjoint from all of them, therefore $\zeta$ is not disjoint from $\alpha$.
-Let $\iota$ be the position such that $(N,\zeta) \sim_p (I,\iota) \sim_p (O_1,\theta)$.
-The fact that $(N,\zeta) \sim_p (I,\iota)$ implies that $\zeta > \alpha;@_1;\lambda$.
-Let $\zeta = \alpha;@_1;\lambda;\kappa$ and $\theta = \alpha;\kappa'$.
-If $\kappa$ is disjoint from $\beta$, then $(N,\zeta) \sim_p (J,\zeta) \sim_p (O_2,\alpha;\kappa') = (O_2;\theta)$ by cases 1 and 2 of $\sim_p$.
-In the other case, that $\kappa$ isn't disjoint from $\beta$, $\kappa > \beta$ because none of the positions $<= \alpha;@_1;\lambda;\beta$ in $N$ are related to any position in $I$ by $\sim_p$.
-As $(I,\alpha;\kappa) \sim_p (O_1,\alpha;\kappa')$ (with the redex at $\alpha;\beta$), for exactly the same reason $(N,\alpha;@_1;\lambda;\kappa) \sim_p (J,\alpha;@_1;\lambda;\kappa')$ (with the redex at $\alpha;@_1;\lambda;\beta$).
-Because $(N,\alpha;@_1;\lambda;\kappa) \sim_p (J,\alpha;@_1;\lambda;\kappa')$, $N|\alpha;@_1;\lambda;\kappa = J|\alpha;@_1;\lambda;\kappa'$, and $N|\alpha;@_1;\lambda;\kappa \neq $ the variable of $N|\alpha;@_1$, therefore $J|\alpha;@_1;\lambda;\kappa'$ is also not the variable therefore $(J,\alpha;@_1;\lambda;\kappa') \sim_p (O_2, \alpha;\kappa') = (O_2, \theta)$ by case 2 of $\sim_p$.
-Combining these results, $(N,\zeta) \sim_p^* (O_2;\theta)$ as desired.
-\end{proof}
-
-\begin{lemma} \label{lem:redexDestroyed}
-If $M \to N$, with the redex at position $\alpha$, then no position in any term reachable from $N$ is related by $\sim^*$ to $(M,\alpha)$.
-\end{lemma}
-\begin{proof}
-Suppose on the contrary that $(M,\alpha) \sim^* (A,\beta)$, where $N \to^* A$, then by Lemma \ref{simpDoesn'tMix}, $(M,\alpha) \sim_p^* (A,\beta)$.
-$M \neq A$ therefore $(M,\alpha) \sim_p$ some position in $N$, but in all the cases of the definition of $\sim_p$, no position in the child term is related to the position of the redex.
-\end{proof}
-
-\paragraph{}
-The reduction relation $\Rightarrow$ is nondeterministic, so it admits multiple possible reduction strategies. 
-A \emph{reduction strategy} starting from a closed term $M$ is a partial function $f$ from $Rch(M)$ to positions, such that for any reachable term $N$ where $f$ is defined, $f(N)$ is a position of a redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
-Using a reduction strategy $f$, a subset of $\Rightarrow$ that isn't nondeterministic, $\Rightarrow_f$, can be defined by $(N,s) \Rightarrow_f (N',s')$ just if $(N,s) \Rightarrow (N',s')$ and $N$ reduces to $N'$ with the redex at $f(N)$.
-
-Essentially what Lemma~\ref{lem:redexDestroyed} demonstrates is that the samples taken during any reduction sequence are independent of each other. This is made more precise in the following lemmas.
-
-\begin{lemma}
-For any skeletons $M \to N$, with the redex at position $\alpha$, and measurable set of samples $S \subset I^{L_s(N)}$, $\mu(S) = \mu(\{s \in I^{L_s(M)} \mid s \circ i(M \to N) \in S\})$, and furthermore, if $M | \alpha = \tsample$, for any $S \subset I \times I^{L_s(N)}$, $\mu(S) = \mu(\{s \in I^{L_s(M)} \mid (s(M,\alpha), s \circ i(M \to N)) \in S\})$.
-\end{lemma}
-\begin{proof}
-If the result holds for all sets $S$ of the form $\{s \in I^{L_s(N)} \mid \forall j : s(j) \in x_i\}$, where $(x_j)_{j \in J}$ is a family of measurable subsets of $I$ indexed by some finite set $J \subset L_s(N)$ of positions, it also holds for all other $S$ by taking limits and disjoint unions.
-
-Take such a $(x_j)_{j \in J}$, where $x_j = x_k$ and define $K = i(M \to N)[J]$.
-Because $i(M \to N)$ is injective, it defines a bijection between $J$ and $K$.
-We can then calculate the measure $\mu(\{s \in I^{L_s(M)} \mid s \circ i(M \to N) \in S\}) = \mu(\{s \in I^{L_s(M)} \mid \forall k : s(k) \in x_{i(M \to N)^{-1}(k)}\}) = \prod_{k \in K} \mu_I(x_{i(M \to N)^{-1}(k)}) = \prod_{j \in J} \mu_I(x_j) = \mu(S)$.
-
-In the case that $M|\alpha = \tsample$, we can similarly consider only those sets $S$ of the form $x_\tsample \times \{s \in I^{L_s(N)} \mid \forall j : s(j) \in x_i\}$.
-The position $\alpha$ in $M$ is not related to any position in $L_s(N)$, therefore $(M,\alpha) \not \in K$.
-Again, we can calculate the measure $\mu(\{s \in I^{L_s(M)} \mid (s(M,\alpha), s \circ i(M \to N)) \in S\}) = \mu(\{s \in I^{L_s(M)} \mid s(M,\alpha) \in x_\tsample, \forall k : s(k) \in x_{i(M \to N)^{-1}(k)}\}) = \mu_I(x_\tsample) \prod_{k \in K} \mu_I(x_{i(M \to N)^{-1}(k)}) = \mu_I(x_\tsample) \prod_{j \in J} \mu_I(x_j) = \mu(S)$.
-\end{proof}
-
-\begin{lemma} \label{lem:independentSamples}
-For any initial term $M$, reduction strategy $f$ on $M$, natural number $n$, skeleton $N$ with $k$ holes, measurable set $T \subset \mathbb R ^ k$ and measurable set $S$ of samples in $I^{L_s(N)}$, $\mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')\}) = \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in I^{L_s(N)} : (M,s) \Rightarrow_f^n (N[r], s')\}) \mu(S)$
-\end{lemma}
-\begin{proof}
-Suppose, to begin with, that $n = 0$.
-Either $\exists r \in T : N[r] = M$, in which case both sides of the equation are $\mu(S)$, or there is no such $r$, in which case both sides of the equation are 0.
-
-For $n > 0$, suppose for induction that the lemma is true for $n - 1$, for all $N, T$ and $S$.
-If $\exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')$, the $r$ and $s'$ are necessarily unique, therefore we may assume that $T$ is the product of $k$ measurable setsets of $\mathbb R$, $(T_j)_{0 \leq j < k}$, as all of the other cases follow by taking unions and limits of these rectangles.
-Let\begin{align*}
-R_s & = \{\text{the skeleton of } O \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) = \tsample, O \to N[r]\} \\
-R_d & = \{\text{the skeleton of } O \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) \neq \tsample, O \to N[r]\}.
-\end{align*}
-Together, $R_s$ and $R_d$ contain all of the skeletons of terms that can reduce to $N$, and they're disjoint.
-For each of these skeletons, there is a map $g_O$ from the reals in $O$ to the reals in $N$, i.e.
-$g_O(r) = r'$ if $O[r] \to N[r']$ for $O \in R_d$, and $g_O(s, r) = r'$ if $O[r] \to N[r']$ with $s$ as the value the sample takes in the reduction.
-All of these functions are measurable, as they are just rearrangements of vector components, or in the case that $O|f(O) = \underline x(\skeletonPlaceholder, \dots, \skeletonPlaceholder)$, a combination fo the measurable function $x$ with a rearrangement of vector components.
-In the case that $O \in R_s$, the function $g_O$ simply inserts the sample into the vector of reals at a certain index, call it $h_O$.
-\begin{align*}
-& \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')\}) \\
-= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in S :\\
-    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\})\\
-    + & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in S :\\
-    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \\
-\end{align*}
-\begin{align*}
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}:\\
-    & g_O(r,s'(O,f(O))) \in T, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
-     + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}:\\
-     & g_O(r) \in T, s' \circ i(O[r] \to N[g_O(r)]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)}:\\
-    & s'(O,f(O)) \in T_{h(O)}, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
-    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}:\\
-    & g_O(r) \in T, s' \circ i(O[r] \to N[g_O(r)]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
-    & \mu(\{s' \in I^{L_s(O)} \mid s'(O,f(O)) \in T_{h(O)}, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S\})\\
-    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
-    & \mu({s' \in I^{L_s(O)} \mid s' \circ i(O[r] \to N[g_O(r)]) \in S}) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(T_{h(O)}) \mu(S)\\
-    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : s'(O,f(O)) \in T_{h(O)}, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S)\\
-    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \mathbb R^{k-1}, s' \in I^{L_s(O)} : g_O(r,s'(O,f(O))) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S)\\
-    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
-= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} :\\
-    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S)\\
-    + & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} :\\
-    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S) \\
-= & \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in I^{L_s(N)} : (M,s) \Rightarrow_f^n (N[r], s')\}) \mu(S)
-\end{align*}
-\end{proof}
-
-The usual call-by-value semantics can be implemented as one of these reduction strategies, given by (with $V$ a value and $T$ a term that isn't a value and $M$ a general term)
-\begin{align*}
-\cbv(T M) & = @_1 ; \cbv(T) \\
-\cbv(V T) & = @_2 ; \cbv(T) \\
-\cbv(\underline f(V_1, \dots, V_{k-1}, T, M_{k+1}, \dots, M_n)) & = \underline f_k ; \cbv(T) \\
-\cbv(\tY T) & = \tY ; \cbv(T) \\
-\cbv(\tif{T < 0}{M_1}{M_2}) & = \textsf{if}_1 ; \cbv(T) \\
-\cbv(\tscore(T)) & = \tscore ; \cbv(T) \\
-\cbv(V) & \text{ is undefined} \\
-\cbv(T) & = \cdot \text{ otherwise}
-\end{align*}
-(this last case covers redexes at the root position).
-
-A closed term $M$ terminates with a given reduction strategy $f$ and samples $s$ if there is some natural number $n$ such that $(M,s) \Rightarrow_f^n (N,s')$ where $f$ gives no reduction at $N$.
-The term terminates almost surely with respect to $f$ if it terminates with $f$ for almost all $s$.
-
-\begin{theorem} \label{thm:AstEquivalence}
-A closed term $M$ is AST with respect to $\cbv$ iff it is AST.
-\end{theorem}
-\begin{proof}
-\akr{I'm not sure whether this proof is sufficiently detailed. It would be pretty easy to fill it in more, just very verbose.}
-
-For both $\red$ and $\Rightarrow_{\cbv}$, the probability measure on the sample space can be used to define a measure on the reduction sequences of each finite length.
-The distributions are equal for each length, by induction, as follows.
-
-The base case, length 0, is trivial, because in both cases the distribution has probability 1 on the sequence containing only $M$.
-
-For the inductive case, suppose that the distributions of reduction sequences of length $n$ are equal.
-\akr{TODO: mention the case that it has already terminated by step $n$, or is about to.
-The two versions of the semantics handle it a little differently.}
-
-For any term $N$ which is not a value, there is a unique environment $E$ and redex $R$ such that $N = E[R]$.
-The position of the hole in $E$ is equal to $\cbv(N)$, so that $N|\cbv(N) = R$, as the cases in the definition of $\cbv$ match the cases in the definition of environments.
-If $R \neq \tsample$, let $R'$ be the result of reducing $R$ (which is equal in both versions of the semantics), then $E[R'] = N[R' / \cbv(N)]$ therefore $\red(N,s) = (E[R'],s)$ and $(N,s) \Rightarrow_{\cbv} (E[R'],s)$.
-The distribution of next terms, conditional on the previous term, in the case that that previous term reduces deterministically, is therefore equal for $\red$ and $\Rightarrow_{\cbv}$.
-
-In the other case that $R = \tsample$, the first (term) part of $\red(N,s)$ is $E[\underline{\pi_h(s)}]$.
-If $k$ is the number of samples already taken in this reduction sequence, this is equivalent to $E[\underline{s_0(k)}]$, where $s_0$ is the initial sample.
-For any sufficiently small neighbourhood of $N$ (i.e.~containing only reduction sequences with the same skeletons), the probability of reaching this neighbourhood is independent of $s(k)$, because it depends only on the samples taken so far, and the samples are independent.
-The distribution of $s(k)$ conditional on reaching $N$ is therefore the uniform distribution on $I$.
-Similarly for the other semantics, if the distribution of remaining samples after $n$ steps is independent of the term by Lemma \ref{lem:independentSamples}, therefore the distribution of $s(Sk(N),\cbv(N))$ conditional on $N$ is the uniform distribution on $I$.
-In either case, the distribution of the next term conditional on the previous term, in the case that it reduces randomly, is equal to the image under $r \mapsto E[\underline r]$ of the uniform distribution on $I$.
-
-In any case, the distribution of reduction sequences after $n+1$ steps, conditional on the reduction sequence after $n$ steps, as defined by $\Rightarrow_{\cbv}$ and by $\red$, is equal, any by the inductive hypothesis the distributions after $n$ steps are equal, therefore by integrating over the reduction sequence up to step $n$, the distributions on reduction sequences up to step $n+1$ are equal.
-
-The distributions on reduction sequences of any finite length as defined by $\Rightarrow_{\cbv}$ and $\red$ are therefore equal, therefore so is the probability of having reached a value after $n$ steps, therefore so is its limit, the probability of termination, therefore the probability of termination is 1 iff the probability of termination with respect to $\cbv$ is 1.
-\end{proof}
-
-\begin{theorem} \label{thm:CbvIsTerminatingest}
-If $M$ terminates with some reduction strategy $f$ and samples $s$, it terminates with $\cbv$ and $s$.
-\end{theorem}
-\begin{proof}
-  Suppose $M$ terminates with the reduction strategy $f$ and samples $s$.
-Let $(M,s) = (M_0,s_0) \Rightarrow_f \dots \Rightarrow_f (M_n,s_n)$, where $M_n$ is a value.
-By Lemma \ref{canonicalCousins}, the reduction sequence $(Sk(M_i))_i$ is related by $\sim_c^*$ to some skeletal reduction sequence $X$ which is in call-by-value order.
-For each skeletal reduction sequence $X_j$ in the sequence $(Sk(M_i))_i \sim_c \dots \sim_c X$, it is possible to define a reduction sequence of terms such that $(M,s) = (N_{j,0},s_{j,0}) \Rightarrow \dots \Rightarrow (N_{j,n_j},s_{j,n_j})$ and $Sk(N_{j,k}) = X_{j,k}$, by at each step applying the reduction at the same place as in $X_j$, essentially just filling in the reals in the skeletal reduction sequence to make it a reduction sequence of terms (it will be shown shortly that the correct branch is taken in each $\textsf{if}$-reduction).
-Assume for induction on $j$ that $(N_{j,n_j},s_{j,n_j}) = (M_n,s_n)$.
-To prove that this holds true for $j+1$ as well, it suffices to show that the corresponding term and samples immediately after the reductions involved in this $\sim_c$ step match, as from that point on, the $\Rightarrow$-sequences will match, as they start from the same point and have all the same reduction positions.
-Letting the term and remaining samples immediately before the reductions involved in this $\sim_c$ step be $(N,t)$ (which is the same in both $N_j$ and $N_{j+1}$ for the same reason again, they start at the same point and the sequences of reduction positions up to this point are the same), and the terms and samples after these reductions be $(O_1,t_1)$ and $(O_2,t_2)$, so that these match the $N$, $O_1$ and $O_2$ in the definition of $\sim_c$, $t_1 = t_2$ because both of them are, by Lemma \ref{lem:sim-M-N}, equal to the restriction of $t$ to the subset of $Rch(N)$ that starts with the reduction sequences to $O_1$ and $O_2$ respectively, and the potential positions in those subsets of $Rch(N)$ are identified with each other by $\sim$ because of the same case of $\sim_c$ that means $N_j$ and $N_{j+1}$ are related.
-If none of the relevant reductions are $\tsample$-reductions, $O_1 = O_2$ trivially.
-If there are samples taken (which must be the arbitrary reduction at $\beta$), it is also required that the same samples are taken in both branches.
-Similarly to Lemma \ref{churchRosser}, the positions of these $\tsample$-reductions in each branch are related by $\sim_p^*$, therefore they correspond to the same sample in $t$, therefore $O_1 = O_2$.
-The fact that the terms are equal also implies that the same branch of an $\textsf{if}$ statements will be taken.
-
-  \paragraph{}
-  There is therefore a reduction sequence $(M,s) = (N_0,s_0') \Rightarrow \dots \Rightarrow (N_m, s_m')$ which ends in a value ($N_m = M_n$) and is in CBV order.
-It doesn't immediately follow that this is the $\Rightarrow_{\cbv}$ reduction sequence starting from $(M,s)$ because the fact that $(N_i)_i$ is in CBV order only means that the reductions that do take place in this reduction sequence are in the correct order, not that the reduction sequence finishes once it reaches a value, or that there are no other reductions that would come earlier in CBV order that don't occur at all in $(N_i)_i$.
-Suppose that at some point $j$ in the reduction sequence, the reduction that occurs is not the next reduction in CBV order, but something else, and that the reduction that should be next is at position $\alpha = \cbv(N_i)$, and the position of the actual reduction $N_j \to N_{j+1}$ is $\beta$.
-Suppose further that $N_j$ is not a value.
-Either $\alpha = \cdot$, the root position, or the reduction at the root position is dependent in some way on the reduction at $\alpha$ (e.g.
-$\alpha = \textsf{if}_1$, or $\alpha = \underline{+};@_1$).
-In any case, all the reductions in the rest of the reduction sequence are after $\alpha$ in CBV order, therefore none of them are $\leq \alpha$, and there continues to be an unreduced redex at $\alpha$, therefore the reduction at the root position never occurs, therefore $N_m$ is not a value, which is a contradiction, therefore at the point (if any) where $(N_i)_i$ departs from the $\Rightarrow_{\cbv}$ reduction sequence starting at $(M,s)$, it has already reached a value, therefore $(M,s) \Rightarrow_{\cbv} \dots$ does reach a value eventually, i.e.~$M$ terminates with the reduction strategy $\cbv$ and samples $s$.
-\end{proof}
-
-\begin{corollary}
-If $M$ is AST with respect to any reduction strategy, it is AST.
-\end{corollary}
-\begin{proof}
-Suppose $M$ is AST with respect to $f$.
-Let the set of samples with which it terminates with this reduction strategy be $X$.
-By Theorem~\ref{thm:CbvIsTerminatingest}, $M$ also terminates with $\cbv$ and every element of $X$, and $X$ has measure 1, by assumption, therefore $M$ is AST with respect to $\cbv$ therefore by Theorem~\ref{thm:AstEquivalence} it is AST.
-\end{proof}
-
-All of the theorems on the termination of rankable terms therefore extend to other reduction strategies too.
-The proofs of Theorems \ref{thm:rankable implies termination}, \ref{thm:partial implies rankable}, \ref{thm:antitone rankable implies termination} and \ref{thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies.
-They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because Theorem \ref{thm:de groote} applies equally to any reduction strategy).
-For a reduction strategy $r$ and term $M$, just substitute $N[r(N)]$ being a $\tY$-redex for $N$ being of the form $E[\tY \lambda x.
-O]$, and $r(N)$ being undefined for $N$ being a value.
-
-The domain of definition of the ranking functions also needs to be changed from the reachable terms $Rch(M)$ to the \emph{reachable terms with respect to the reduction strategy $r$},
-\[
-Rch_r(M) = \{N \mid \exists n, N_0, \dots, N_n: N_0 = M, N_n = N, N_i \to N_{i+1} \text{ at } r(N_i)\}.
-\]
-
-To be more explicit, the modified forms of the theorems are:
-\begin{definition}\rm
-A \emph{ranking function on $M$ with respect to a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
-\begin{itemize}
-    \item $f(N) \geq 1+ f(N')$ if $N[r(N)]$ is a $\tY$-redex and $N$ reduces to $N'$ at $r(N)$
-    \item $f(N) \geq \int_I f(N[\underline{x}/r(N)]) \, \Leb(\mathrm{d}x)$ if $N[r(N)] = \tsample$
-
-    \item $f(N) \geq f(N')$ if $r(N)$ is any other redex, where $N \to N'$ at $r(N)$.
-\end{itemize}
-Any closed term for which a ranking function with respect to a reduction strategy $r$ exists is called \emph{rankable} with respect to $r$. 
-\end{definition}
-
-\begin{definition}
-A \emph{partial ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : Rch_r(M) \rightharpoonup \mathbb R$ such that
-\begin{itemize}
-    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
-    \item $f$ is defined at $M$.
-    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ at the positions specified by $r$ will eventually reach some $O$ such that either $r(O)$ isn't defined or $f(O)$ is, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
-\end{itemize}
-\end{definition}
-
-\begin{definition}\rm
-An \emph{antitone ranking function on $M$ with respect to a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ such that
-\begin{itemize}
-    \item $f(N) \geq \epsilon(f(N)) + f(N')$ if $N[r(N)]$ is a $\tY$-redex and $N$ reduces to $N'$ at $r(N)$
-    \item $f(N) \geq \int_I f(N[\underline{x}/r(N)]) \, \Leb(\mathrm{d}x)$ if $N[r(N)] = \tsample$
-
-    \item $f(N) \geq f(N')$ if $r(N)$ is any other redex, where $N \to N'$ at $r(N)$.
-\end{itemize}
-Any closed term for which an antitone ranking function with respect to a reduction strategy $r$ exists is called \emph{antitone rankable} with respect to $r$. 
-\end{definition}
-
-\begin{definition}
-An \emph{antitone partial ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : Rch_r(M) \rightharpoonup \mathbb R$ such that there exists some antitone function $\epsilon : \nnReal \to \pReal$ such that
-\begin{itemize}
-    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
-    \item $f$ is defined at $M$.
-    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ at the positions specified by $r$ will eventually reach some $O$ such that either $r(O)$ isn't defined or $f(O)$ is, and $f(N) \geq \mathbb E[f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
-\end{itemize}
-\end{definition}
-
-\begin{theorem}
-If a closed SPCF term $M$ has a ranking function, partial ranking function, antitone ranking function or antitone partial ranking function with respect to a reduction strategy $r$, then $M$ is AST with respect to $r$, and AST.
-\end{theorem}
-\lo{Give an example or two of a reduction strategy (for $\Rightarrow$) that is not $\cbv$.}
-
-\akr{NOTE. For the standard CBV semantics, this example is indeed not a problem. 
-In the alternative semantics, defining 
-\begin{align*}
-A[x] &= \tif{\tif{x>0}{I}{I} \, \tsample - 0.5 > 0}{0}{\Omega}\\
-B &= \tif{\tsample - 0.5 > 0}{0}{\Omega}
-\end{align*}
-The set of samples where it terminates is 
-\[\begin{array}{l}
-\bigcup_{r \in [0,1]} \{s \in I^{L_s(A[\tsample])} \mid s([A[\tsample]], \mathsf{if}_1;\underline{ -
-}_1;@_1;\mathsf{if}_1) = r, \\
-\qquad \qquad \hfill \qquad \qquad s([A[\tsample], A[r], \ldots, B, \mathsf{if}_1) > 0.5 \}.
-\end{array}\] 
-This is a rather unwieldy expression, but the crucial part is that $r$ occurs twice in the conditions on $s$: once as the value a sample must take, and once in the location of a sample. 
-It is this set that's unmeasurable.}
-
-\bibliographystyle{apalike}
-\bibliography{references}
-
-\lo{22 Jan: 
-\paragraph{Further directions}
-
-An obvious next step is to extend the result to the $\mathsf{score}$ construct.
-
-The following are highly topical, and could form the basis of an interesting and novel DPhil thesis.
-\begin{enumerate}
-\item Devise methods for proving (positive) a.s.~termination.
-
-- For example, develop a type system satisfying the property: if a term is typable then it is (positively) a.s.~terminating.
-See~\citep{DBLP:conf/ppdp/BreuvartL18,DBLP:conf/esop/LagoG17}.
-
-- Another approach is to develop algorithms that synthesise ranking supermartingales, following, for example, \citep{DBLP:journals/pacmpl/AgrawalC018}.
-
-\item Develop principles (e.g.~in the form of ``proof rules'') for reasoning about (positively) a.s.~termination, in the style of \citep{DBLP:journals/pacmpl/McIverMKK18}.
-
-\item Design algorithms that synthesise probabilistic invariants (\emph{qua} martingales), \`a la \cite{SchreuderO19}; see also \citep{DBLP:journals/pacmpl/HarkKGK20}.
-
-\end{enumerate}}
-
-\iffalse
-\begin{thebibliography}{9}
-\bibitem{ppcf} Thomas Ehrhard, Michele Pagani, and Christine Tasson.
-Measurable cones and stable, measurable functions: a model for probabilistic higher-order programming.
-\emph{PACMPL}, 2(POPL):59:1–59:28, 2018.
-doi: 10.1145/3158147.
-URL \href{https://doi.org/10.1145/3158147}{https://doi.org/10.1145/3158147}.
-\end{thebibliography}
-\fi
-
-\end{document}

### 2020-12-26 LICS21/antitone.tex
Created version for LICS21 submission in subdirectory LICS21

--- 2020-12-23 probabilisticProgrammingMartingales.tex
@@
-\documentclass{article}
-
-%% BEGIN {Luke's Macros}
-%%
-\newif\ifdraft
-\drafttrue %% To hide all comments and highlighting, just comment this line out
-
-\input{Style/comment}
-\usepackage{calculation}
-\usepackage{graphicx}
-\usepackage{tikz}
-\usetikzlibrary{positioning, graphs, quotes, graphdrawing, shapes.geometric}
-\usegdlibrary{layered, circular}
-\usepackage[square]{natbib}
-\setcitestyle{aysep={}}
-
-\definecolor{oxblue}{RGB}{0,33,71}
-\definecolor{oxgold}{HTML}{a0630a}
-\usepackage[final,
-bookmarks,
-bookmarksopen,
-colorlinks,
-final,
-linkcolor=red,
-citecolor=oxgold,
-pdfstartview=FitH ]{hyperref}
-
-\usepackage{amsmath} % cleveref must be loaded afer amsmath
-\usepackage{cleveref}
-\Crefname{theorem}{Theorem}{Theorems}
-\Crefname{corollary}{Corollary}{Corollary}
-\crefname{proposition}{Proposition}{Propositions}
-\Crefname{claim}{Claim}{Claims}
-\Crefname{definition}{Definition}{Definitions}
-\Crefname{fact}{Fact}{Facts}
-\Crefname{conj}{Conjecture}{Conjectures}
-\Crefname{example}{Example}{Examples}
-\Crefname{remark}{Remark}{Remarks}
-\Crefname{convention}{Convention}{Conventions}
-\Crefname{lemma}{Lemma}{Lemmas}
-\Crefname{assumption}{Assumption}{Assumptions}
-\Crefname{section}{Section}{Sections}
-\Crefname{appendix}{Appendix}{Appendices}
-\Crefname{figure}{Figure}{Figures}
-%\Crefname{equation}{Eq.}{Equations}
-
-% \addtotheorempostheadhook[theorem]{\crefalias{thmlisti}{theorem}}
-% \addtotheorempostheadhook[lemma]{\crefalias{thmlisti}{lemma}}
-% \addtotheorempostheadhook[proposition]{\crefalias{thmlisti}{proposition}}
-% \addtotheorempostheadhook[corollary]{\crefalias{thmlisti}{corollary}}
-% \addtotheorempostheadhook[claim]{\crefalias{thmlisti}{claim}}
-% \addtotheorempostheadhook[fact]{\crefalias{thmlisti}{fact}}
-% \addtotheorempostheadhook[example]{\crefalias{thmlisti}{example}}
-% \addtotheorempostheadhook[definition]{\crefalias{thmlisti}{definition}}
-% \addtotheorempostheadhook[remark]{\crefalias{thmlisti}{remark}}
-% \addtotheorempostheadhook[assumption]{\crefalias{thmlisti}{assumption}}
-
-\newcommand{\Real}{\mathbb{R}}
-\newcommand{\nnReal}{\mathbb{R}_{{\geq}0}}
-\newcommand{\pReal}{\mathbb{R}_{{>}0}}
-
-\newcommand\expect[1]{\mathbb{E}[#1]}
-\newcommand\set[1]{\{#1\}}
-\newcommand\dif{\mathrm{d}}
-\newcommand\calF{\mathcal{F}}
-\newcommand\calI{\mathcal{I}}
-\newcommand\Leb{\mathrm{Leb}}
-\newcommand\ndraw[2]{\#\mathrm{draw}_{#1}(#2)}
-
-\makeatletter
-\newcommand{\dotDelta}{{\vphantom{\Delta}\mathpalette\d@tD@lta\relax}}
-\newcommand{\d@tD@lta}[2]{%
-  \ooalign{\hidewidth$\m@th#1\mkern-1mu\cdot$\hidewidth\cr$\m@th#1\Delta$\cr}%
-}
-
-%\usepackage[bbgreekl]{mathbbol} %for \bblambda
-%%
-%% END {Luke's Macros}
-
-\usepackage{amsfonts}
-\usepackage{amsthm}
-\usepackage{amssymb}
-\usepackage{verbatim}
-\usepackage{enumitem}
-\newcommand{\tY}{\mathsf{Y}}
-%\newcommand{\tif}[3]{\textsf{if }#1\textsf{ then }#2\textsf{ else }#3}
-\newcommand{\tif}[3]{\mathsf{if}(#1, #2, #3)} % space-saving concrete syntax
-\newcommand{\tsample}{\mathsf{sample}}
-\newcommand{\tscore}{\mathsf{score}}
-\newcommand{\skeletonPlaceholder}{\mathsf{X}} % I'm not sure what the best notation here is.
-\DeclareMathOperator{\red}{red}
-\DeclareMathOperator{\nnext}{next}
-\DeclareMathOperator{\cbv}{cbv}
-
-\usepackage{amsthm}
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}
-\newtheorem{example}{Example}
-
-\theoremstyle{lemma}
-\newtheorem{lemma}{Lemma}
-\newtheorem{theorem}{Theorem}
-\newtheorem{corollary}{Corollary}
-\newtheorem{conjecture}{Conjecture}
-
-\theoremstyle{remark}
-\newtheorem*{remark}{Remark}
-
-\title{Supermartingales and Termination Analysis of Statistical PCF}
-
-\begin{document}
-
-\maketitle
-
-\lo{Tentative title}
-
-\begin{abstract}
-\lo{TODO}
-\end{abstract}
-
-\iffalse
-\akr{Following your suggestion, I will be providing a criterion for termination of programs in PPCF \citep{DBLP:journals/pacmpl/EhrhardPT18} based on ranking supermartingales. 
-As it's more convenient for this proof, a sampling-based semantics \lo{You need to provide reference(s) for sampling-based semantics.} will be used instead of the original distributional semantics. 
-I assume some roughly applicable equivalence is proven somewhere, but it doesn't seem that hard anyway.}
-\lo{Why not provide a proof as an appendix?}
-\fi
-
-\tableofcontents
-
-Various theorems about probabilistic programs rely on the assumption that the program terminates almost surely.
-Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \citep{DBLP:journals/pacmpl/McIverMKK18}.
-This paper's contribution is to extend this method to a higher-order setting.
-
-\section{Statistical PCF}
-
-\subsection{Syntax of SPCF}
-
-The language SPCF is a simply-typed lambda calculus with sampling of real numbers from $[0,1]$ and unbounded scoring, following \cite{MakOP20b}.
-Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
-\begin{align*}
-  & \text{types } A, B ::= \textsf{R}  \mid  A \to B \\
-  & \text{values } V ::= \lambda x.M  \mid  \underline{r} \\
-  & \text{terms } M, N ::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY M  \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample  \mid  \tscore(M)
-\end{align*}
-\begin{align*}
-  \frac{}{\Gamma ; x:A \vdash x:A} \qquad
-  \frac{\Gamma ; x:A \vdash M : B}{\Gamma \vdash \lambda x.M : A \to B} \qquad
-  \frac{}{\underline{r} : \textsf{R}} \qquad
-  \frac{\Gamma \vdash M:A \to B \quad \Gamma \vdash N : A}{\Gamma \vdash M \, N : B} \\ \\
-  \frac{\Gamma \vdash M_1:\textsf{R} \dots \Gamma \vdash M_n:\textsf{R}}{\Gamma \vdash \underline{f}(M_1,\dots,M_n) : \textsf{R}} \ f : \mathbb{R}^n \to \mathbb{R} \qquad
-  \frac{\Gamma \vdash M : (A \to B) \to (A \to B)}{\Gamma \vdash \tY M : (A \to B)} \\ \\
-  \frac{\Gamma \vdash M : \textsf{R} \quad \Gamma \vdash N_1 : A \quad \Gamma \vdash N_2 : A}{\Gamma \vdash \tif{M < 0}{N_1}{N_2} : A} \qquad
-  \frac{}{\Gamma \vdash \tsample : \textsf{R}} \qquad
-  \frac{\Gamma \vdash s : \textsf{R}}{\Gamma \vdash \tscore (s) : \textsf{R}}
-\end{align*}
-
-Terms are identified up to $\alpha$-equivalence, as usual. The set of all terms is denoted $\Lambda$, and the set of closed terms is denoted $\Lambda^0$.
-
-\paragraph{}
-To define the reduction relation, let \emph{evaluation contexts} be of the form:
-\begin{align*}
-  E ::= & \, [\,] \mid E \, M \mid V \, E \mid \underline{f}(\underline r_1,\dots ,\underline r_{k-1}, E, M_{k+1}, \dots, M_n) \\ & \mid \tY E \mid \tif{E<0}{M_1}{M_2} \mid \tscore (E)
-\end{align*}
-then a term reduces if it is formed by substituting a redex in a context i.e.
-\begin{align*}
-  E[(\lambda x.M) N] & \to E[M[N/x]] \\
-  E[\underline f (\underline r_1, \dots , \underline r_n)] & \to E[\underline{f(r_1,\dots,r_n)}] \\
-  E[\tY \lambda x. M] & \to E[\lambda z. M[(\tY \lambda x. M)/x] z] \text{ where $z$ is not free in $s$}\\
-  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_1] \text{ where }r < 0 \\
-  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_2] \text{ where }r \geq 0 \\
-  E[\tsample] & \to E[\underline r] \text{ where } r \in [0,1] \\
-  E[\tscore(\underline r)] & \to E[\underline r].
-\end{align*}
-\changed[lo]{We write $\to^\ast$ for the reflexive, transitive closure of $\to$.}
-\lo{The $\tscore$ rule should have a side condition: if $r > 0$.}
-
-Every closed well-typed term either is a value or reduces to another term.
-
-\subsection{Sampling semantics}
-\label{sec:sampling semantics}
-This version of the reduction relation allows $\tsample$ to reduce to any number in $[0,1]$.
-To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples.
-Let $ I = [0,1] \subset \mathbb{R} $, and let $S = I^{\mathbb{N}}$, with the \changed[lo]{Borel $\sigma$-algebra $\calF$} and the probability measure, denoted $\mu$, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
-Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and $\mu (\prod_{i=0}^\infty X_i) = \prod_{i=0}^\infty \Leb(X_i)$.
-The maps $\pi_h:S \to I, \; \pi_t:S \to S$ popping the first element are then measurable.
-\changed[lo]{Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(S, \calF, \mu)$ the \emph{entropy space}.}
-
-The $\sigma$-algebra and measure on $\Lambda$ are defined by considering it as a disjoint union of equivalence classes under replacing all the real constants by a placeholder, where the measure on each class is that of $\mathbb{R}^n$, where $n$ is the number of real constants.
-\changed[lo]{Precisely, following \cite{DBLP:conf/icfp/BorgstromLGS16},
-we view $\Lambda$ as $\bigcup_{m\in\omega} \big(\mathsf{Sk}_m \times \Real^m \big)$,
-where $\mathsf{Sk}_m$ is the set of SPCF terms with exactly $m$ place-holders (a.k.a.~\emph{skeleton terms}) for numerals.
-Thus identified, we give $\Lambda$ the countable disjoint union topology of the product topology of the discrete topology on $\mathsf{Sk}_m$ and the standard topology on $\Real^m$.
-Note that the connected components of $\Lambda$ have the form $\{M\} \times \Real^m$, with $M$ ranging over $\mathsf{Sk}_m$, and $m$ over $\omega$. 
-We fix the Borel algebra of this topology to be the $\sigma$-algebra on $\Lambda$.}
-
-The one-step reduction is given by the function $\red : \Lambda^0 \times S \to \Lambda^0 \times S$ where
-\begin{equation}
-\red(M,s) = \left\{
-    \begin{array}{ll}
-        (E[N],s) & \text{if } M = E[R], R \to N \text{ and } R \neq \tsample \\
-        (E[\underline{\pi_h(s)}],\pi_t(s)) & \text{if } M = E[\tsample] \\
-        (M,s) & \text{if } M \text{ is a value}
-    \end{array} \right .
-\end{equation}
-\lo{Both $\to$ and $\red$ are called reduction relation.}
-
-The result after $n$ steps is then simply $\red^n(M,s) = \overbrace{\red(...\red(}^n M,s)...)$, and the limit $\red^\infty$ can then be defined as a partial function as $\lim_{n \to \infty} \red^n(M,s)$ whenever that sequence becomes constant by reaching a value.
-A term $M$ terminates for a sample sequence $s$ if the limit $\red^\infty(M,s)$ is defined.
-
-The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid M \text{ terminates at } s \}$ is measurable.
-A term $M$ is said to terminate almost surely if $\mu(\{s \mid M \text{ terminates at } s\}) = 1$.
-
-\lo{Alternatively, define the \emph{runtime of $M$} to be the random variable 
-\[
-T_M(s) := 
-\begin{cases}
-\min \set{n \mid \pi_0(\red^n(M, s)) \textrm{ is a value}} & \hbox{if $\red^\infty(M,s)$ is defined}\\
-\infty & \hbox{otherwise}
-\end{cases}
-\]
-Equivalently, we say that $M$ is \emph{almost-surely terminating} (AST) if $T_M < \infty$ a.s.; 
-and $M$ is \emph{positively almost-surely terminating} (PAST) if $\expect{T_M} < \infty$.}
-
-For example, the term 
-\[
-(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0},
-\] 
-which generates a geometric distribution, terminates on the set $S \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas 
-\[
-\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \underline 0},
-\] 
-which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.
-
-\paragraph{}
-Note that the requirement that $M$ be closed is not important for the proofs of the theorems, but is simply required to ensure that $M$ reaches a value, rather than a general normal form like $x \, y$.
-
-\paragraph{}
-This definition of almost-sure termination is equivalent to that given in \citep{MakOP20b}, although the program semantics is stated in a slightly different way.
-In particular, the argument to $\tscore$ is not relevant to termination (except for the possibility that its argument's evaluation wouldn't terminate).
-
-\iffalse
-\lo{Your operational semantics does not maintain a record of the current weight of the reduction.
-A.s.~termination does depend on $\tscore$: see \cite[\S 4.3]{DBLP:journals/corr/abs-2004-03924}\footnote{\url{https://arxiv.org/abs/2004.03924}}.
-I think it important to take the behaviour of $\tscore$ into account;
-you should do it as a future task.}
-\fi
-
-\section{Supermartingales}
-One approach to proving that a term terminates almost surely is to find some variant that is bounded below and, on average, decreases sufficiently quickly that it must eventually reach 0, similarly to the approach taken in \citep{DBLP:journals/pacmpl/McIverMKK18} for imperative programs.
-
-These variants are defined as functions from reachable terms (i.e.~possible states of the program's execution) to real numbers.
-Specifically, let the set of reachable terms from a given closed starting term $M$ be $Rch(M) := \{N \in \Lambda \mid M \to^* N \}$, with the $\sigma$-algebra induced as a subset of $\Lambda$.
-
-\begin{definition}\rm
-A \emph{ranking function on $M$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
-\begin{enumerate}
-    \item $f(E[\tY \lambda x. N]) \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z) \text{ where $z$ is not free in $N$}$
-    \item $f(E[\tsample]) \geq \int_I f(E[\underline{x}]) \, \Leb(\mathrm{d}x)$
-
-    \item $f(E[R]) \geq f(E[R'])$ for any other redex $R$, where $R \to R'$.
-\end{enumerate}
-We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that $f(N) = 0$ iff $N$ is a value, and for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.
-
-Any closed term for which a ranking (respectively, strict ranking) function exists is called \emph{rankable} (respectively, \emph{strictly rankable}). 
-\changed[lo]{For example, the term $(\tY \lambda x.x) \, \underline 0$ is not rankable.}
-It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \in \omega}$ is the reduction sequence starting from $M$, then $(f(M_n))_{n\in\omega}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales.
-\end{definition}
-
-\input{ranking-supermartingales}
-
-\section{Constructing Ranking Functions}
-Although rankability implies almost-sure termination, the converse does not hold in general. For example,
-\begin{equation}
-\tif{{-}\tsample < 0}{\underline{0}}{(\tY \lambda x. x) \, \underline 0}
-\label{ex:0 probability reachable}
-\end{equation}
-terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x. x) \, \underline 0$ is reachable, although that has probability 0. 
-Not only is this counterexample AST, it's PAST. %positively almost surely terminating i.e.~the expected time to termination is finite.
-
-A ranking function can be constructed under the stronger assumptions that, for every $N$ reachable from $M$, the expected number of $\tY$-reduction steps from $N$ to a value is finite.
-In particular, the expected number of $\tY$-reduction steps from each reachable term is a ranking function.
-Note that a finite number of expected $\tY$-reduction steps does not necessarily imply a finite number of expected total reduction steps.
-
-\begin{example}
-\label{ex:tY finite does not imply t finite}
-The term
-\[
-M = \big(\tY \lambda f n. \tif{\tsample - 0.5 < 0}{n \, \underline 0}{f (\lambda x. n (n x))}\big) (\lambda x. x+1)
-\] 
-terminates with only 2 $\tY$-reductions on average i.e.~$\expect{T_M^\tY} = 2$, but applies the increment function $2^n$ times with probability $2^{-1-n}$ for $n \geq 0$, which diverges i.e.~$\expect{T_M} = \infty$.
-\end{example}
-
-\begin{theorem} \label{thm:minimal}
-Given a closed term $M$, the function $f:Rch(M) \to \mathbb R$ given by $f(N) = \mathbb E [\text{the number of }\tY\text{-reduction steps from }N\text{ to a value}]$, if it exists, is the least of all possible ranking functions of $M$.
-\end{theorem}
-\begin{proof}
-Let $f$ be the candidate least ranking function defined above, and suppose $g$ is another ranking function such that $f(N) > g(N)$ for some $N \in Rch(M)$.
-The restrictions of $f$ and $g$ to $Rch(N)$ have the same properties assumed of $f$ and $g$, so assume w.l.o.g.~that $N=M$.
-The difference $g - f$ is then a supermartingale (with the same setup as in the proof of %\Cref{thm:rankable gives supermartingale}
-%Theorem \ref{rankable implies ast}) 
-\Cref{thm:rankable and strict rankable});
-therefore $\mathbb E[g(M_n)] \leq \mathbb E[f(M_n)] + g(M)-f(M)$, for all $n$.
-
-We have $\mathbb E[f(M_n)] = \sum_{k=n}^\infty \mathbb P[M_k = E[\tY N] \text{ for some }E, N] \to 0 \text{ as } n \to \infty$; 
-therefore as $g(M) - f(M) < 0$, eventually $\mathbb E[g(M_n)] < 0$, which is impossible. 
-It follows that $g \geq f$ as required.
-
-In order for $f$ to be the least ranking function of $M$, it also has to actually be a ranking function itself.
-Each of the conditions on a ranking function is easily verified from the definition of $f$.
-\end{proof}
-
-\subsection{Partial ranking functions}
-Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost-sure termination, so it isn't useful for proving it.
-
-\begin{example}[Geometric distribution]
-Consider the term
-\begin{align*}
-&(\tY \lambda f, n. \\
-&\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
-&) \, \underline{0}
-\end{align*}
-which generates a geometric distribution (let $\Theta = \lambda f, n : \tif{\tsample \underline{- 0.5} < 0}{n}{f (n\underline{+ 1})}$, then the term is $(\tY \Theta)\ \underline 0$).
-    Despite its simplicity, its $Rch$ contains all the terms
-
-\begin{tikzpicture}[node distance=0.5em, anchor=west]
-    \node (1) at (0,-0) {$\tY\ \Theta\ \underline 0$};
-    \node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\ \Theta\ (n \underline{+ 1})}) z)\ \underline i$};
-    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\ \Theta\ (n \underline{+ 1})})\ \underline i$};
-    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY\ \Theta\ (\underline i\ \underline{+ 1})}$};
-    \node (5) at (0,-4) {$\tif{\underline r\ \underline{- 0.5} < 0}{\underline i}{\tY\ \Theta\ (\underline i\ \underline{+ 1})}$};
-    \node (6) at (0,-5) {$\tif{\underline r < 0}{\underline i}{\tY\ \Theta\ (\underline i\ \underline{+ 1})}$};
-    \node (7) at (2,-6) [text width=] {$\underline i$};
-    \node (8) at (0,-7) {$\tY\ \Theta\ (\underline i\ \underline{+ 1})$};
-    \node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\ \Theta\ (n \underline{+ 1})}) z)\ (\underline i\ \underline{+ 1})$};
-
-    \draw [->] (1.south west) to [bend right=45] (2.north west);
-    \draw [->] (2.south west) to [bend right=45] (3.north west);
-    \draw [->] (3.south west) to [bend right=45] (4.north west);
-    \draw [->] (4.south west) to [bend right=45] (5.north west);
-    \draw [->] (5.south west) to [bend right=45] (6.north west);
-    \draw [->] (6.south west) to [bend right=45] (8.north west);
-    \draw [->] (8.south west) to [bend right=45] (9.north west);
-
-    \draw [->] (9.west) to [bend left=45] (2.west);
-    \draw [->] (6) to (7);
-\end{tikzpicture}
-
-\end{example}
-
-Even in this simple case, defining a ranking function explicitly is awkward because of the number of cases, although in most cases, because the value need only be greater than or equal to that of the next term in sequence, it suffices to take the ranking function as having the same value as the next term, so that overall it takes only 3 distinct values.
-(We will explain why later.)
-
-The definition of rankability is also inconvenient for syntactic sugar.
-It could be useful, for example, to define $M \oplus_p N = \tif{\tsample - p < 0} M N$, where $M \oplus_p N$ reduces to $M$ or $N$, depending on the first value of $s$\lo{Need a name for $s$.}, with probability $p$ resp.~$(1-p)$.
-Technically though, it reduces first to $\tif{\underline r - p < 0} M N$ for all $r \in I$, so those terms all need values of the ranking function too.
-
-\paragraph{}
-In both of these cases, there are only some values of the ranking function that are semantically important. 
-\begin{definition}
-Define a \emph{partial ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \rightharpoonup \mathbb R$ such that
-\begin{itemize}
-    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
-    \item $f$ is defined at $M$.
-    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually \akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and ast, but that doesn't actually work because of terms reachable with 0 probability.
-Of course, astness doesn't imply rankability, so any definition of partial ranking function that includes all Y-past terms wouldn't always extend to a total ranking function.} reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
-\end{itemize}
-\end{definition}
-A partial ranking function that is total is just a ranking function.
-Providing a partial ranking function is essentially part way between providing a ranking function and directly proving almost sure termination.
-
-\begin{theorem} \label{thm:partial implies rankable}
-Every partial ranking function is a restriction of a ranking function.
-\end{theorem}
-\begin{proof}
-Take a closed term $M$ and partial ranking function $f$ on $M$. Define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
-\begin{align*}
-    f_1(N) & := f(N) \text{ whenever $f(N)$ is defined,} \\
-    f_1(V) & := 0 \text{ for values $V$ not in the domain of $f$.}
-\end{align*}
-
-Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) := \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
-The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on partial ranking functions.
-Define $f_2(N) = \int_S f_1(\nnext(N,s)) + g(N,s) \, \mu(\mathrm d s)$.
-The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction).
-\lo{TODO. The preceding sentence deserves a proof.}
-\end{proof}
-
-As a corollary, any term which admits a partial ranking function terminates almost surely.
-
-\subsection{Examples}
-Let $M \oplus_p N = \tif{\tsample - p < 0} M N$, for $p \in (0,1]$, then there are the pseudo-reduction relations
-\begin{align*}
-E[M \oplus_p N] \to^3 & E[M] & \\
-E[M \oplus_p N] \to^3 & E[N] & \\
-\red^3(E[M \oplus_p N], s) = & \left\{
-    \begin{array}{ll}
-        (E[M],\pi_t(s)) & \text{if } \pi_h(s) < p \\
-        (E[N],\pi_t(s)) & \text{if } \pi_h(s) \geq p. \\
-    \end{array} \right .
-\end{align*}
-A partial ranking function could be defined with respect to this shortcut reduction simply by replacing $\to$ and $\red$ in the definition of a partial ranking function by a version that goes straight from $N \oplus_p O$ to $N$ or $O$.
-Such a pseudo-partial ranking function would then be a partial function from a subset of $Rch(M)$, so it could also be considered as a partial function from all of $Rch(M)$, and it would in fact also be an actual partial ranking function.
-It is therefore possible to prove rankability directly using the shortcutted reductions.
-
-A similar procedure would work for other forms of syntactic sugar.
-If a closed term $N$ eventually reduces to one of a set of other terms $\{N_i \mid i \in I\}$ with certain probabilities, a partial ranking function defined with respect to a reduction sequence that skips straight from $N$ to $N_i$ is also a valid partial ranking function for the original reduction function, and therefore its existence implies almost sure termination.
-There is a caveat, however, that $\tY$-reduction steps skipped over in the shortcut still need to be counted for the expected number of $\tY$-reduction steps.
-
-\paragraph{}
-With this abbreviation, the geometric distribution example from earlier can be written as $(\tY \lambda f n.\ n \oplus_{0.5} f(n+1)) \, \underline 0$.
-It is then easy to see that the following is a partial ranking function:
-\begin{align*}
-(\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\ N & \mapsto 2 \\
-\underline i \oplus_{0.5} (\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\ (\underline i + 1) & \mapsto 1 \\
-\underline i & \mapsto 0.
-\end{align*}
-
-In fact, even the partial function $\tY \Theta N \mapsto 2$ alone is a partial ranking function for this term.
 \section{Antitone ranking functions}
+\label{sec:antitone}
 
@@
 \[
-f(E[R’]) \leq f(E[R]) - \epsilon(f(E[R])) 
+f(E[R']) \leq f(E[R]) - \epsilon(f(E[R])) 
 \]
-where $R \to R’$ is the $\tY$-redex rule.
+where $R \to R'$ is the $\tY$-redex rule.
 
+\iffalse
 \lo{An aside: This CBV $\tY$-rule seems cleaner:
@@
 We assume $\tY f^{A \to B} \, x^A \, . \, \theta^B$ is a value.}
+\fi
 
@@
 With respect to the antitone function $\epsilon(x) = \frac{4-x}{4}$ and stopping time $T$, we have that $(Y_n)_{n \in \omega}$ is an antitone strict supermartingale, except that $\epsilon(4) = 0$, and $T = \infty$ with probability $\frac 1 2$, even though $4$ is larger than any value $(Y_n)_{n \in \omega}$ can actually take, and $\epsilon(Y_n)$ never actually reaches $0$.
-
 %\lo{@Andrew: I refer to the preceding highlighted claim. Did you mean: for all $n, s$, $\epsilon(Y_n(s)) > Y_n(s)$?} \akr{No. This phrasing should be clearer.}
 \end{example}
+
+\changed[lo]{Let $f$ be a ranking function on a closed SPCF term $M$.
+%Recall that $(f(M_n))_{n \in \omega}$ is a supermartingale.
+We say that $M$ is \emph{antitone rankable} by $f$ if the supermartingale $(f(M_n))_{n \in \omega}$ and the stopping time $T_M^{\tY}$, both adapted to the fitration $(\calF_n)_{n \in \omega}$ where $\calF_n := \sigma(M_1, \ldots, M_n)$, are antitone strict.}
 
@@
 \Cref{thm:rankable implies termination},
-take the set $S$ of traces as a probability space, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$.
+take the set $S$ of traces \lo{TODO: We should find a better name for traces $s \in S$.} as a probability space, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$.
 As before, $(Y_n)_{n \in \omega}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$.
@@
 \begin{align*}
-  & g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 \\
-  = & \log \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) \\
-  = & \frac 1 2 \log\left(\frac{(n+1)^2}{n(n+2)}\right) \\
-  = & \frac 1 2 \log\left(1 + \frac 1 {n(n+2)}\right) \\
-  > & \frac 1 {4n(n+2)} \\
-  > & \frac 1 2 \frac 1 {5(n+1)^2} + \frac 1 2 \frac 1 {5(n+3)^2}
+g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 
+  &=  \log \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) \\
+  &=  \frac 1 2 \log\left(\frac{(n+1)^2}{n(n+2)}\right) \\
+  &=  \frac 1 2 \log\left(1 + \frac 1 {n(n+2)}\right) \\
+  &>  \frac 1 {4n(n+2)} \\
+  &>  \frac 1 2 \frac 1 {5(n+1)^2} + \frac 1 2 \frac 1 {5(n+3)^2}
 \end{align*}
-and $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
+\changed[lo]{(The first inequality follows from the fact that for all $x \in (0, 1)$, $\log (1 + x) > x$.)}
+Moreover $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
 Therefore by setting $\epsilon_2(x) = \frac 1 {5(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met.
@@
 For $n > 2$, $z_n = \log(n-1)$ works.
-The expected decrease is $\frac 1 2(\log(1+\frac 1 {(n-1)^2-1}) - \frac 1 {2n+1} \log(1 + \frac 2 {n-2}))$, and using the fact that $\frac x {x+1} \leq \log (1+x) \leq x$ for $x > 0$, this is at least $\frac 1 2(\frac 1 {(n-1)^2} - \frac 1 {2n + 1} \frac 2 {n-2}) = \frac {n^2}{2(n-1)^2(2n+1)(n-2)}$, which (again for $n > 2$) is positive and antitone.
+The expected decrease is $\frac 1 2(\log(1+\frac 1 {(n-1)^2-1}) - \frac 1 {2n+1} \log(1 + \frac 2 {n-2}))$, and using the fact that $\frac x {x+1} \leq \log (1+x) \leq x$ for $x > 0$,\lo{The preceding inequality is incorrect: for $x \in (0, 1)$, $\log (1 + x) > x$.} this is at least $\frac 1 2(\frac 1 {(n-1)^2} - \frac 1 {2n + 1} \frac 2 {n-2}) = \frac {n^2}{2(n-1)^2(2n+1)(n-2)}$, which (again for $n > 2$) is positive and antitone.
 For $n = 0, 1, 2$, we take $\epsilon$ to be 9/40 (the same as its value at 3), then set $z_2 = 2 \log 2 - \log 3 - 1, z_1 = 3 \log 2 - 2 \log 3 - 3, z_0 = 4 \log 2 - 3 \log 3 - 6$.
@@
 
-\input{relativised-completeness}
+% !TEX root = ./probabilisticProgrammingMartingales.tex
 
-\input{y-stopping-time}
+\begin{proof}
+If $T$ is bounded by $b$ a.s.~(for $n$ constant), the statement is trivially true by taking $\epsilon$ to be constantly 1, and $Y_n = b - n$.
+Otherwise, let $(t_n)_{n \in \omega}$ be defined recursively such that $t_0 = 0$, $t_{n+1} > t_n$ and $\mathbb P[T > t_{n+1}] \leq \frac 1 4 \mathbb P[T > t_n]$.
+We will then define a supermartingale $(Y_n)_{n \in \omega}$ and nonrandom sequence $(y_n)_{n \in \omega}$ such that $Y_n = 0$ iff $T < n$, $Y_n = y_n$ iff $T \geq n$, and $y_n \geq 2^k$ for $n \geq t_k$.
 
-\section{Confluent Trace Semantics}
-When proving almost sure termination in this way, it is necessary to consider which terms a given term may reduce to.
-Sometimes however, the reduction that the programmer has in mind may not be strictly the call-by-value order defined so far, or considering an alternative reduction order may be simpler or more intuitive.
+The antitone function $\epsilon$ is defined piecewise and recursively in such a way as to force all the necessary constraints to hold:
+\begin{align*}
+    \text{for }x \in [0,1),\ &\epsilon(x) = 1 \\
+    \text{for }x \in [2^k,2^{k+1}),\ &\epsilon(x) = \min \left (\epsilon(2^{k-1}), \frac{2^k}{t_{k+1}-t_k} \right ).
+\end{align*}
+This ensures that
+\begin{itemize}
+\item $\epsilon$ is weakly decreasing. 
+\item $\epsilon$ is bounded below by 0.
+\item For $x \geq 2^k$, $\epsilon(x) \leq \frac{2^k}{t_{k+1}-t_k}$. 
+\lo{This would imply $\frac{2^k}{t_{k+1}-t_k} \geq \epsilon(2^{k-1})$. Why?} \akr{Corrected: it should have been min, not max.}
+\end{itemize}
 
-Non-probabilistic lambda calculi generally have the Church-Rosser property, that if a term $A$ reduces to both $B_1$ and $B_2$, there is some $C$ 
-%with reduction sequences $B_1 \to^* C$ and $B_2 \to^* C$, 
-to which both $B_1$ and $B_2$ reduce,
-so the reduction order mostly doesn't matter. 
-In the probabilistic case, this may not be true, because $\beta$-reduction can duplicate $\tsample$s, so the outputs of the copies of the sample may be identical or independent, depending on whether the sample is taken before or after $\beta$-reduction.
-There are, however, some restricted variations on the reduction order that do not have this problem.
+The sequence $(y_n)_{n \in \omega}$ is then defined recursively by
+\begin{align*}
+    y_0 = 2, \quad
+    y_{n+1} = (y_n - \epsilon(y_n)) \frac{\mathbb P[T > n]}{\mathbb P[T > n+1]}.
+\end{align*}
 
-\paragraph{}
-Even with this restriction, a sampling semantics in the style of the one already defined would not be entirely Church-Rosser, as, for example, $\red^3(\tsample - \tsample, (1,0,\dots))$ would be either $1$ or $-1$ depending on the order of evaluation of the $\tsample$s, as that determines which sample from the pre-selected sequence is used for each one.
-To fix this, rather than pre-selecting samples according to the order they'll be drawn in, select them according to the position in the term where they'll be used instead.
+The fact $y_{t_k} \geq 2^{k+1}$ is proven by induction on $k$.
+For base case, $2 \geq 2$, as required.
+For the inductive case $k+1$, we first do another induction to prove that $y_n \geq 2^k$ for all $t_k \leq n \leq t_{k+1}$.
+The base case of this inner induction follows from the outer induction hypothesis a fortiori.
+Take the greatest $k$ such that $n > t_k$.
+By the induction hypothesis, $y_{t_k} \geq 2^k$.
+\begin{align*}
+    y_n & = y_{t_k} \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} - \sum_{m=t_k}^{n-1} \epsilon(y_m) \frac{\mathbb P[T > m+1]}{\mathbb P[T > n]} \\
+    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \sum_{m=t_k}^{n-1} \epsilon(y_m)) \\
+    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \sum_{m=t_k}^{n-1} \frac{2^k}{t_{k+1}-t_k}) \\
+    & = \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \frac{2^k (n - t_k)}{t_{k+1}-t_k}) \\
+    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - 2^k) \\
+    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (2^{k+1} - 2^k) \\
+    & = \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} 2^k \\
+    & \geq 2^k
+\end{align*}
+as required. Substituting $n = t_{k+1}$ and using the fact that $\frac{\mathbb P[T > t_k]}{\mathbb P[T > t_{k+1}]} > 4$, the same reasoning gives $y_{t_{k+1}} \geq 2^{k+2}$, as required.
 
-A \emph{position} 
-%\akr{I'm not really sure how to phrase the citation, but this is adapted from \cite{Kennaway96infinitarylambda}.} 
-%\lo{The idea is common enough. I think no citation necessary.}
-is a finite sequence of steps into a term, defined inductively as
+Although the stronger induction hypothesis was necessary for the induction to work, the only reason this is needed is to prove that $y_n > 0$ for all $n$, which implies that $Y_n \geq 0$. 
+The other condition for $(Y_n)$ to be an antitone-strict supermartingale is
 \begin{align*}
-\alpha ::= \cdot \mid \lambda ; \alpha \mid @_1 ; \alpha \mid @_2 ; \alpha \mid \underline f_i ; \alpha \mid \tY ; \alpha \mid \textsf{if}_1 ; \alpha \mid \textsf{if}_2 ; \alpha \mid \textsf{if}_3 ; \alpha \mid \tscore ; \alpha.
+    & \expect{Y_{n+1} \mid \mathcal F_n} \\
+    = & \expect{y_{n+1} \, 1_{\set{T \geq n+1}} \mid \mathcal F_n} \\
+    = & y_{n+1} \, \expect{1_{\set{T \geq n+1}} \mid \mathcal F_n} \\ 
+    = & y_{n+1} \, 1_{\set{T \geq n}} \frac{\mathbb P[T > n+1]}{\mathbb P[T > n]} \\
+    = & (y_n - \epsilon(y_n)) \, 1_{\set{T \geq n}} \\
+    = & Y_n - \epsilon(y_n) \, 1_{\set{T \geq n}}
 \end{align*}
-The \emph{subterm of $M$ at a position $\alpha$}, denoted $M \mid \alpha$, is defined as
-\begin{align*}
-M \mid \cdot & = M \\
-\lambda x. M \mid \lambda ; \alpha & = M \mid \alpha \\
-M_1 M_2 \mid @_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2 \\
-\underline f(M_1,\dots,M_n) \mid \underline f_i ; \alpha & = M_i \mid \alpha \quad \text{for }i \leq n \\
-\tY M \mid \tY ; \alpha & = M \mid \alpha \\
-\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2,3 \\
-\tscore(M) \mid \tscore ; \alpha & = M \mid \alpha
-\end{align*}
-so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda$ is undefined). 
-A position such that $M\mid \alpha$ does exist is said to \emph{occur} in $M$. 
-%\changed[lo]{
-\emph{Substitution} of $N$ at position $\alpha$ in $M$, written $M[N/\alpha]$, is defined similarly.
-For example, let 
-\[
-M = \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})
-\qquad
-\alpha =\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2
-\]
-then 
-\(
-M[\tsample / \alpha] = \lambda x \, y. y \, (\tif{x < 0}{y \, \tsample}{\underline 3}).
-\)
-%}
-
-Two subterms $N_1$ and $N_2$ of a term $M$, corresponding to positions $\alpha_1$ and $\alpha_2$, can overlap in a few different ways. 
-If $\alpha_1$ is an initial segment (i.e.~prefix) of $\alpha_2$ (written as $\alpha_1 \leq \alpha_2$), then $N_2$ is also a subterm of $N_1$.
-If neither $\alpha_1 \leq \alpha_2$ nor $\alpha_1 \geq \alpha_2$, the positions are said to be \emph{disjoint}.
-The notion of disjointness is mostly relevant in that if $\alpha_1$ and $\alpha_2$ are disjoint, performing a substitution at $\alpha_1$ will leave the subterm at $\alpha_2$ inaffected.
-\lo{The preceding sentence is unclear. Typo?} \akr{It wasn't a typo, but I've made it a little more explicit now. Is this clearer?}
-
-With this notation, a more general reduction relation $\to$ can be defined. \akr{It would almost work to have one of the neater versions of $\tY$-reduction here.}
-%\lo{On first reading, I struggled to parse the following clauses.}
-\begin{definition}
-\label{def:more general red}
-The binary relation $\to$ is defined by the following rules, each is conditional on a redex occurring at position $\alpha$ in the term $M$:
-\begin{align*}
-  \text{if } M \mid \alpha = (\lambda x.N) V,\ & M \to M[N[V/x]/\alpha] \\
-  \text{if } M \mid \alpha = \underline f (\underline r_1, \dots , \underline r_n),\ & M \to M[\underline{f(r_1,\dots,r_n)}/\alpha] \\
-  \text{if } M \mid \alpha = \tY \lambda x. N,\ & M \to M[\lambda z. N[(\tY \lambda x. N)/x] z/\alpha] \text{ where $z$ is not free in $N$}\\
-  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_1/\alpha] \text{ where }r < 0 \\
-  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_2/\alpha] \text{ where }r \geq 0 \\
-  \text{if } M \mid \alpha = \tsample & \text{ and $\lambda$ does not occur after $@_2$ or $\tY$ in $\alpha$},\\ & M \to M[\underline r/\alpha] \text{ where } r \in [0,1] \\
-  \text{if } M \mid \alpha = \tscore(\underline r),\ & M \to M[\underline r/\alpha].
-\end{align*}
-In each of these cases, $M \mid \alpha$ is the \emph{redex}, and reduction takes place at $\alpha$.
-\end{definition}
-
-\paragraph{}
-Labelling the pre-chosen samples by the positions in the term would also not work because in some cases, a $\tsample$ will be duplicated before being reduced, for example, in $(\lambda x.
-x \, {\underline 0} \, \mathbin{\underline{+}} \, {x \, \underline 0}) (\lambda y.
-\tsample)$, both of the $\tsample$ redexes that eventually occur originate at $@_2 ; \lambda$.
-It is therefore necessary to consider possible positions that may occur in other terms reachable from the original term. 
-Even this is itself inadequate because some of the positions in different reachable terms need to be considered the same, and the number of reachable terms is in general uncountable, which leads to measure-theoretic issues.
-
-Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $Sk(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$.
-\lo{Do we actually use the notation $S[r]$ in what follows?} \akr{Yes, in Lemma \ref{lem:independentSamples} for example.}
-\lo{Skeleton is defined in Sec.~\ref{sec:sampling semantics}.}
-The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, and $\tsample$ reducing to $\skeletonPlaceholder$.
-For example, we have 
-\(
-(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \tsample
-\to
-(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \skeletonPlaceholder
-\to
-\tif{\skeletonPlaceholder < 0}{\skeletonPlaceholder}{\skeletonPlaceholder}
-\to
-\skeletonPlaceholder
-\)
-
-Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a $\to$-reduction sequence of skeletons starting at $Sk(M)$, and the second of which is a position in the final skeleton of the reduction sequence.
-As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the usual linear sampling semantics, there will be (with one more caveat introduced later) a sample from $I$ pre-selected for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.
-
-Reduction sequences are used rather than reachable skeletons because if the same skeleton is reached twice, different samples may be needed.
-For example, consider the term $M = \tY (\lambda f x.\ \tif{\tsample - \underline{0.5} < 0}{f \, x}{x}) \, \underline 0$, which reduces after a few steps to $N = \tif{\tsample - \underline{0.5}  < 0}{M}{\underline 0}$.
-If the pre-selected sample for $(N,\textsf{if}_1;\underline{-}_1)$ is less than 0.5, $N$ reduces back to $M$, then $N$ again, then the same sample is used the next time, therefore it's an infinite loop, whereas if samples are labelled by reduction sequences, the samples for $M \to \dots \to N$ are independent from the samples for $M \to \dots \to N \to M \to \dots \to N$, and so on.
-
-The reduction sequences of skeletons will often be discussed as though they were just skeletons, identifying them with their final skeletons.
-With this abuse of notation, a reduction sequence $N$ (actually $N_1 \to \dots \to N_n = N$) may be said to reduce to a reduction sequence $O$, where the reduction sequence implicitly associated with the final skeleton $O$ is $N_1 \to \dots \to N_n \to O$.
-
-\paragraph{}
-This is still not quite sufficient because sometimes the same samples must be used at corresponding positions in different reduction sequences.
-For example, the term $M = \tsample + \tsample$ has the reachable skeletons $N_1 = \skeletonPlaceholder + \tsample, N_2 = \tsample + \skeletonPlaceholder, O = \skeletonPlaceholder + \skeletonPlaceholder$ and $\skeletonPlaceholder$, with reductions $M \to N_1 \to O \to \skeletonPlaceholder$ and $M \to N_2 \to O \to \skeletonPlaceholder$.
-In the reduction $M \to N_1$, the sample labelled $(M, \underline{+}_2)$ is used, and in the reduction $N_2 \to O$, the sample labelled $(M \to N_2, \underline{+}_2)$ is used.
-Each of these samples becomes the value of the second numeral in $O$ in their respective reduction sequences, therefore in order for Church-Rosserness to be attained, they must be the same.
-Which elements of $L_0(M)$ must match can be described by the relation $\sim^*$:
-
-\begin{definition}
-The relation $\sim$ is defined as the union of the minimal symmetric relations $\sim_p$ (``$p$'' for parent-child) and $\sim_c$ (``$c$'' for cousin) satisfying
-\begin{enumerate}
-    \item If $N$ reduces to $O$ with the redex at position $\alpha$, and $\beta$ is a position in $N$ disjoint from $\alpha$, then $(N,\beta) \sim_p (O,\beta)$.
-
-    \item If $N$ $\beta$-reduces to $O$ at position $\alpha$, $\beta$ is a position in $N \mid \alpha;@_1;\lambda$ and $N \mid \alpha;@_1;\lambda;\beta$ is not the variable involved in the reduction, $(N,\alpha;@_1;\lambda;\beta) \sim_p (O, \alpha;\beta)$
-
-    \item If $N$ $\textsf{if}$-reduces to $O$ at position $\alpha$, with the first resp.
-second branch being taken, and $\alpha;\textsf{if}_i;\beta$ occurs in $N$ (where $i = 2$ resp.
-$3$), $(N,\alpha;\textsf{if}_i;\beta) \sim_p (O,\alpha;\beta)$
-
-    \item If $N$, $O_1$ and $O_2$ match any of the following cases:
-    \begin{enumerate}
-        \item $N$ contains redexes at disjoint positions $\alpha_1$ and $\alpha_2$, $O_1$ is $N$ reduced first at $\alpha_1$ then $\alpha_2$ and $O_2$ is $N$ reduced first at $\alpha_2$ then at $\alpha_1$.
-
-        \item $N \mid \alpha = \tif{\underline r < 0}{N_1}{N_2}$, where $r < 0$ (or, respectively, $r \geq 0$), $(N_2 \text{ resp.
-} N_1) \mid \beta$ is a redex, and $O_1$ is $N$ reduced at $\alpha$ and $O_2$ is $N$ reduced first at $\alpha;(\textsf{if}_3 \text{ resp.
-} \textsf{if}_2);\beta$ then at $\alpha$
-
-        \item $N \mid \alpha = \tif{\underline r < 0}{N_1}{N_2}$, where $r < 0$ (or, respectively, $r \geq 0$), $(N_1 \text{ resp.
-} N_2) \mid \beta$ is a redex, and $O_1$ is $N$ reduced first at $\alpha$ then at $\alpha;\beta$ and $O_2$ is $N$ reduced first at $\alpha;(\textsf{if}_2 \text{ resp.
-} \textsf{if}_3);\beta$ then at $\alpha$
-
-        \item $N \mid \alpha = (\lambda x.
-A) B$, there is a redex in $A$ at position $\beta$, $O_1$ is $N$ reduced first at $\alpha$ then at $\alpha;\beta$, and $O_2$ is $N$ reduced first at $\alpha;@_1;\lambda;\beta$ then at $\alpha$
-
-        \item $N \mid \alpha = (\lambda x.
-A) B$, $B \mid \beta$ is a redex, $(\gamma_i)_i$ is a list of all the positions in $A$ where $A \mid \gamma = x$, ordered from left to right, $O_1$ is $N$ reduced first at $\alpha ; @_2 ; \beta$ then at $\alpha$, and $O_2$ is $N$ reduced first at $\alpha$ then at $\alpha;\gamma_i;\beta$ for each $i$ in order.
-
-        \item $N \mid \alpha = \tY (\lambda x.
-A)$, $A$ reduced at $\beta$ is $A'$, $(\gamma_i)_i$ is a list of all the positions where $A' \mid \gamma  = x$, ordered from left to right, $O_1$ is $N$ reduced first at $\alpha;\tY;\lambda;\beta$ then at $\alpha$, and $O_2$ is $N$ reduced first at $\alpha$ then at $\alpha;\lambda;@_1;\gamma_i;Y;\lambda;\beta$ for each $i$ in order where $\gamma_i$ is left of $\beta$ then at $\alpha;\lambda;@_1;\beta$ then at $\alpha;\lambda;@_1;\gamma_i;Y;\lambda;\beta$ for the remaining values of $i$.
-    \end{enumerate}
-    (in which case $O_1$ and $O_2$ are equal as skeletons, but with different reduction sequences), $O_1'$ and $O_2'$ are the results of applying some reduction sequence to each of $O_1$ and $O_2$ (the same reductions in each case, which is always possible because they're equal terms), and $\delta$ is a position in $O_1'$ (or equivalently $O_2'$), then $(O_1',\delta) \sim_c (O_2',\delta)$.
-\end{enumerate}
-The $\sim_c$-rules are illustrated in the following diagrams.
-\end{definition}
-
-\include{sim-diagram}
-
-The reflexive transitive closure $\sim^*$ of this relation is used to define the set of \emph{potential positions} $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable skeletons.
-If $(N,\alpha) \sim^* (O,\beta)$, then $N \mid \alpha$ and $O \mid \beta$ both have the same shape (i.e.~they're either both the placeholder $\skeletonPlaceholder$, both variables, both applications, both $\tsample$s etc.), therefore it's well-defined to talk of the set of potential positions where there is a $\tsample$, $L_s(M)$.
-Formally $L_s(M) := \{[(X, \alpha)]_{\sim^\ast_M} : X \mid \alpha = \tsample\}$.
-The new sample space is then defined as $I^{L_s(M)}$, with the Borel $\sigma$-algebra and product measure.
-\changed[lo]{Since $I^{L_s(M)}$ is a countable product, the measure space is well-defined \citep[Cor.~2.7.3]{AshDD00}.}
-
-\paragraph{}
-Although the six cases in the definition of $\sim_c$ are defined separately, there are some commonalities worth noting.
-Each case (except a, which is symmetrical) has a right branch and a left branch (the branches of case b are unfortunately displayed the wrong way around in the diagram).
-The right branch has reduction steps at two positions, $\beta$ then $\alpha$, where $\beta > \alpha$ (i.e.~$\beta$ is inside $\alpha$), and in the left branch, the order of these reductions are swapped, so the reductions are at $\alpha$ then $\beta_1, \dots, \beta_n$.
-The position $\alpha$ of one of the reductions is unchanged, and the other reduction may have multiple (in cases e and f) or 0 (in cases b and e) images in the left branch.
-The images $(\beta_i)$ are all still inside (greater than or equal to) $\alpha$, and although they are not generally equal to $\beta$, the subskeletons at these positions have a similar shape to the subskeleton (initially) at position $\beta$, so that the reductions at these positions are the same type of reduction (e.g.~both $\beta$-reduction or both $\textsf{if}$-reduction).
-Case a is also similar, except that the relevant positions are all disjoint rather than contained in one another, and either branch could be considered the right branch.
-
-\iffalse
-\changed[lo]{Since $L_s(M)$ is a countable set, there is a unique probability measure $\mu$ on the infinite product $\prod_{i \in L_s(M)} \Sigma_I$ of the Borel $\sigma$-algebras $\Sigma_I$ satisfying
-\[
-\mu\set{\omega \in \Omega \mid \omega_1 \in A_1, \cdots, \omega_n \in A_n} = \prod_{i=1}^n \Leb(A_i)
-\]
-where $\Leb$ is the Lebesgue measure on the measurable space $(I, \Sigma_I)$.}
-\fi
-
-\paragraph{}
-Before defining the new version of the reduction relation $\red$, a few lemmas about the properties of $\sim$ are necessary for it to be well-defined.
-
-\begin{lemma} \label{downAcrossUp}
-If $(N_1,\alpha_1) \sim^* (N_2,\alpha_2)$, there is some descendants $N'_1, N'_2$ of $N_1$ resp.~$N_2$, and a position $\alpha'$ in both of them such that $(N_1,\alpha_1) \sim_p^* (N'_1, \alpha') \sim_c^* (N'_2, \alpha') \sim_p^* (N_2, \alpha_2)$.
-\end{lemma}
-\begin{proof}
-  The $\sim_p$ relation can be split into $\sim_\downarrow \cup \sim_\uparrow$, where $(A,\alpha) \sim_\downarrow (B,\beta)$ if $A \to B$ and $(A,\alpha) \sim_p (B,\beta)$, and similarly $(A,\alpha) \sim_\uparrow (B,\beta)$ if $B \to A$ and $(A,\alpha) \sim_p (B,\beta)$.
-At each stage of this proof, it will be assumed that the $\sim_c$ steps and the $\sim_p$ steps in the sequence from $(N_1,\alpha_1)$ to $(N_2,\alpha_2)$ are rearranged such that there is never a $\sim_c$ immediately before a $\sim_\downarrow$, and there is never a $\sim_c$ immediately after a $\sim_\uparrow$.
-This rearrangement is always possible, because if there is some subsequence $(A,\alpha) \sim_c (B,\alpha) \sim_\downarrow (C,\beta)$, then there is an alternate path $(A,\alpha) \sim_\downarrow (B',\beta) \sim_c (C,\beta)$, where the reduction and the $\sim_\downarrow$ from $A$ to $B'$ are the same as those from $B$ to $C$, and the $\sim_c$ step is the same, but with the reduction sequences $O_1 \to^* O_1'$ and $O_2 \to^* O_2'$ in the definition of $\sim_c$ extended by the reduction $B \to C$.
-With these rearrangements assumed, it is sufficient to prove that the $\sim_p$ steps can be rearranged so that all of the $\sim_\downarrow$ steps come before all of the $\sim_\uparrow$ steps (possibly introducing some $\sim_c$ steps in the process).
-
-  The rearrangement to put all of the $\sim_\downarrow$s before all of the $\sim_\uparrow$s proceeds by induction on the number of $\sim_\downarrow$s that occur after the first occurrence of $\sim_\uparrow$.
-If it's 0, all of the $\sim_p$s are already in the correct order, and we're done.
-Otherwise, take the subsequence from the first $\sim_\uparrow$ to the first $\sim_\downarrow$ after that.
-This must be rearranged to some $\sim_\downarrow$s followed by some $\sim_\uparrow$s, then once that's done, the number of $\sim_\downarrow$s after the first $\sim_\uparrow$ in the overall sequence will have decreased by 1.
-
-  Let this subsequence be $A \sim_\uparrow^n B \sim_\downarrow C$.
-Suppose for induction that $A \sim_\uparrow^k (D,\delta) \sim_\downarrow^* (E,\epsilon) \sim_\uparrow^* C$ for some $D, \delta, E, \epsilon$, and $0 \leq k \leq n$, where in the reduction sequence $D \to^* E$, for any pair of reductions at positions which aren't disjoint, the reduction at the innermost (greater) position occurs first.
-Reduction sequences with this property will be called ``parallel'', as it is analagous to parallel reduction as in \akr{TODO: find this reference, for the proof of the Church-Rosser theorem by Tait and Martin-L\"of}.
-This induction is in reverse, with $k$ decreasing from $n$ to $0$.
-If $k = n$, simply set $(D,\delta) = B$ and $(E,\epsilon) = C$.
-If $k = 0$, then $A$ is related to $B$ by a sequence of $\sim_\downarrow$s then $\sim_\uparrow$s, as desired.
-In the intermediate steps, $k$ must be decreased by one, so a subsequence of one $\sim_\uparrow$ followed by a parallel sequence of $\sim_\downarrow$s must be replaced by a parallel sequence of $\sim_\downarrow$s followed by some $\sim_\uparrow$s.
-
-  For any individual $\sim_\uparrow$ then $\sim_\downarrow$ pair, let $(A,\alpha) \sim_\uparrow (B,\beta) \sim_\downarrow (C,\gamma)$.
-The skeleton $B$ reduces to both $A$ and $C$.
-If these are the same reduction, then $(A,\alpha) = (C,\gamma)$ and this subsequence may be removed.
-If they are different, then the way they overlap corresponds to one of the cases in the definition of $\sim_c$, either case a if the reduction positions are disjoint, or one of the cases b--f if one of the reduction positions is inside the other.
-In any case, there is a reduction sequence from each of $A$ and $C$ to some common skeleton $D$ ($O_1$ and $O_2$ in the definition of $\sim_c$).
-For the same reason that $(B,\beta) \sim_\downarrow (C,\gamma)$, there is some $\delta$ such that $(A,\alpha) \sim_\downarrow^* (D,\delta)$, and similarly for the same reason that $(B,\beta) \sim_\downarrow (A,\alpha)$, for the same $\delta$, $(C,\gamma) \sim_\downarrow^* (D,\delta)$.
-There are a lot of cases to check for this statement, but all of them are rather simple, and similar to each other.
-There is therefore an alternative $\sim^*$ sequence from $(A,\alpha)$ to $(C,\gamma)$, of the form $(A,\alpha) \sim_\downarrow^* (D_1,\delta) \sim_c (D_2,\delta) \sim_\uparrow^* (C,\gamma)$, where $D_1$ and $D_2$ are the alternative reduction sequences leading to the same skeleton $D$.
-There are only multiple $\sim_\downarrow$ steps in the result if the position of the reduction $B \to C$ is inside the position of the reduction $B \to A$, and similarly, there are only multiple $\sim_\uparrow$ steps in the result if the position of the reduction $B \to A$ is inside the position of the reduction $B \to C$.
-If there are multiple $\sim_\downarrow$ steps, they are not necessarily parallel, but in the only case where they aren't (case f of $sim_c$), the subsequence of $\sim_\downarrow$ steps with the reductions at $\alpha;\lambda;@_1;\beta$ then $\alpha;\lambda;@_1;\gamma_i;\tY;\lambda;\beta$ for each $i$ where $\gamma_i \geq \beta$ (in the notation of the definition of $\sim_c$) can be replaced by $\sim_\downarrow$s with the reductions at $\alpha;\lambda;@_1;\gamma'_i;\tY;\lambda;\beta$ then $\alpha;\lambda;@_1;\beta$, then another $\sim_c$, where $(\gamma'_i)$ is the list of positions in $A$ (\emph{before} the reduction at $\beta$ to $A'$, unlike $(\gamma_i)$) where $A | \gamma'_i = x$ and $\gamma'_i > \beta$.
-
-  The effect of this rearrangement is (ignoring the $\sim_c$s) to swap the order of the $\sim_\uparrow$ and the $\sim_\downarrow$, except that if the positions of one of these reductions is inside the other, that inner $\sim_p$ may be duplicated.
-If the positions are disjoint, both of them are unchanged (corresponding to case a of $\sim_c$), and if one is inside the other, then the outer position is unchanged and the other resultant positions, although they aren't equal to the original inner position, are still inside the outer position.
-Taking the sub-sequence of one $\sim_\uparrow$ followed by a parallel sequence of $\sim_\downarrow$s, this rearrangement can be repeatedly applied to move the $\sim_\uparrow$ further along the sequence.
-If at some point a $\sim_\uparrow$ and $\sim_\downarrow$ match (by relating the same (skeleton, position) pairs in opposite directions), they may be removed and the process may stop early.
-The $\sim_\uparrow$ may be duplicated if it passes a $\sim_\downarrow$ whose position is outside of the $\sim_\uparrow$'s position, but by the assumption that the sequence of $\sim_\downarrow$s is initially parallel, no position inside of this occurs later in the sequence, therefore all of the resultant $\sim_\uparrow$s pass the remaining $\sim_\downarrow$s without changing or duplicating them.
-The $\sim_\uparrow$s may be further duplicated, but the number of steps left in this process is bounded by the product, for all of the remaining $\sim_\downarrow$s, of $2 + \text{the number of occurrences of the variable relevant to the reduction}$, because for each $\sim_\downarrow$, that is more than the number of duplicates it could produce for any $\sim_\uparrow$ that passes it.
-It is also possible (earlier) for some of the $\sim_\downarrow$s to be duplicated, but the only case where this process would not terminate, that both the $\sim_\downarrow$s and the $\sim_\uparrow$s continue being duplicated forever so that the number of switches left to do never decreases, is prevented by the parallelness condition, as all of the duplications of $\sim_\downarrow$s occur before all of the duplications of $\sim_\uparrow$s.
-
-  It still remains to be shown that the sequence of $\sim_\downarrow$s left at the end of this can be made parallel.
-The changes that may have occurred since the previous version of this sequence (which was known to be parrallel already), are that some of the $\sim_\downarrow$s may have been duplicated by passing the $\sim_\uparrow$, and if that $\sim_\uparrow$ matches one of the $\sim_\downarrow$s, that $\sim_\downarrow$ is removed.
-The position of the $\sim_\uparrow$'s reduction is the same for all of these duplications, because it is not changed until later in the sequence, when the $\sim_\downarrow$s with positions outside of that may occur.
-The resultant positions after a duplication are in the same position relative to each other, and therefore the order of the remaining $\sim_\downarrow$s is automatically parallel, except that (letting the position of the $\sim_\uparrow$'s reduction be $\alpha$) if there are $\sim_\downarrow$s with reductions at positions $\alpha;@_1;\lambda;\beta$ and $\alpha;@_2;\gamma$, or $\alpha;\tY;\lambda;\beta$ and $\alpha;\tY;\lambda;\gamma$, some of the resultant $\sim_\downarrow$s' positions may be inside each other where they were originally not.
-Similarly to the case where a $\tY$ reduction and a reduction at a position inside it are swapped though, the order can be fixed by swapping some of the $\sim_\downarrow$ steps with each other.
-In the first case, the reduction at position $\alpha;@_1;\lambda;\beta$ ends up at $\alpha;\beta$, and the reduction at position $\alpha;@_2;\gamma$ ends up at the positions $\alpha;\delta_i;\gamma$, where $(\delta_i)$ is the positions of the variable relevant to the $\sim_\uparrow$'s reduction inside its lambda.
-The positions of the variable in the lambda may be different before the reduction at $\beta$, but the positions of the redexes corresponding to the original redex at $\alpha;@_2;\gamma$ are changed similarly.
-Rather than reducing at $\alpha;\beta$ then at $\alpha;\delta_i;\gamma$, it is therefore possible to reach the same point by reducing at $\alpha;\delta'_i;\gamma$ then at $\alpha;\beta$, where $(\delta'_i)$ is the positions of the relevant variable within the lambda before the reduction at $\alpha;@_1;\lambda;\beta$ (and again, a $\sim_c$ of some sort must also be introduced).
-The other case, that the $\sim_\uparrow$'s reduction is a $\tY$-reduction, is similar.
-Each reduction which is duplicated has one image at $\alpha;\lambda;@_1;$its original relative position, and one for each occurrence of the variable.
-Those corresponding to variable occurrences may need to be moved to before those at $\lambda;@_1$.
-
-  In summary, to reach the desired order of $\sim_p$ steps, each $\sim_\downarrow$ may have to be moved past some $\sim_\uparrow$ steps, and it may get duplicated in the process, but the resultant sequence of $\sim_\downarrow$s can all be put in parallel order, and it has been shown that moving a $\sim_\uparrow$ past a sequence of $\sim_\downarrow$s is always possible if they're in parallel order, therefore the overall process terminates and reaches a point where all of the $\sim_\downarrow$s precede all of the $\sim_\uparrow$s, and all of the $\sim_c$s are in between them.
-All of these rearrangements leave the end points of the sequence unchanged, therefore this is an alternative sequence of $\sim$ steps between $(N_1,\alpha_1)$ and $(N_2,\alpha_2)$ of the form $(N_1,\alpha_1) \sim_\downarrow^* (N_1', \alpha') \sim_c^* (N_2', \alpha') \sim_\uparrow^* (N_2,\alpha_2)$.
+therefore $(Y_n)_n$ is an antitone strict supermartingale with respect to $(T,\epsilon)$.
 \end{proof}
 
-If $(X,\alpha) \sim_c (Y,\alpha)$ then the final skeletons in $X$ and $Y$ are equal, and $(X,\beta) \sim_c (Y,\beta)$ for any $\beta$ that occurs in this skeleton, therefore $\sim_c$ can be considered to apply to reduction sequences even without positions, $X \sim_c Y$ iff $(X,\cdot) \sim_c (Y,\cdot)$.
-The structure of $\sim_c$ can be seen more clearly by choosing a canonical example from each equivalence class.
-For this, the member of the class that's in call-by-value order is used.
-A reduction sequence is in call-by-value order if for every pair of reductions in the sequence at positions $\alpha$ and $\beta$, where the reduction at $\alpha$ happens first, either $\alpha \leq \beta$ or $\alpha$ and $\beta$ are disjoint, with $\alpha$ to the left of $\beta$ (i.e.~the first elements of the sequences $\alpha, \beta$ where they differ are $@_1$ and $@_2$, $\underline f_i$ and $\underline f_j$ for $i < j$, or $\textsf{if}_i$ and $\textsf{if}_j$ for $i < j$, in that order).
-Note that although the reductions that occur in the sequence are in call-by-value order, it may not be the actual call-by-value reduction sequence starting from that point because some redexes may remain un-reduced even when other reductions that should happen afterwards occur.
-It is only required that the reductions that do occur occur in the correct order.
-
-\begin{lemma} \label{canonicalCousins}
-For any reduction sequence $X$, there is a unique reduction sequence $X_{cbv}$ that is related to $X$ by $\sim_c$ and is in call-by-value order, and if $X \sim_c^* Y$, then $X_{cbv} = Y_{cbv}$.
-\end{lemma}
-
-\lo{As defined, $\sim_c$ is not symmetric.}
-
-\begin{proof}
-
-Define $X_n$ recursively so that $X_0 = X$ and if $X_n$ is not in CBV order, take the last pair of reductions in $X_n$ which aren't in CBV order.
-They are adjacent in the sequence, as the CBV order is a total order.
-Let the positions of these reductions be $\alpha$ and $\beta$ respecively.
-If they are disjoint, they can be considered as one of the branches of $\sim_c$ case a.
-Otherwise, $\beta < \alpha$, and the sub-skeleton at position $\beta$ at the appropriate point in the reduction sequence is a redex with non-trivial subterms, therefore it's of one of the forms $\tif{\skeletonPlaceholder>0}{A}{B}$, $(\lambda x.A) B$ or $\tY \lambda x.
-A$, therefore the reductions $\alpha$ and $\beta$ form the $O_2$ branch of one of the cases b, c, d, e or f of $\sim_c$.
-In either case, define $X_{n+1}$ as equal to $X_n$ except taking the other (left) branch, so that the order of the reductions at $\alpha$ and $\beta$ are switched (although the equivalent(s) of the reduction at $\alpha$ may not actually be at that position).
-
-This sequence of swaps rearranges the reductions in $X$ rather like insertion-sort (but in some cases duplicating or deleting the new element), and eventually reaches some $X_n$ that is in CBV order.
-This is defined to be $X_{cbv}$.
-
-\paragraph{}
-If $Y \sim_c X \sim_c^* X_{cbv}$, and the reductions involved in $Y \sim_c X$ are not the last pair of reductions in $Y$ that aren't in CBV order, the order of the $\sim_c$s can be rearranged so that the sequence $Y \sim_c^* X_{cbv}$ is the canonical sequence given above, therefore $Y_{cbv} = X_{cbv}$.
-To be more precise about this rearrangement, consider the sequence of $\sim_c$ steps from $Y$ to $X_{cbv}$.
-The first step takes some sub-sequence of the reduction sequence $Y$ and swaps the order of those reductions.
-If $X = X_{cbv}$, then $X = Y_1 = Y_{cbv}$ already.
-Otherwise, consider the cases according to whether the image of this subsequence in $X$ occurs earlier than the next pair of reductions to be swapped (i.e.~the last pair of reductions in $X$ that occur not in CBV order).
-If not, then either the step $Y \sim_c X$ changes this subsequence into CBV order, or out of it.
-If it is into, then it was the last out-of-order pair in $Y$ therefore the whole sequence $Y \sim_c^* X_{cbv}$ is already in the canonical order and $Y_{cbv} = X_{cbv}$.
-If it switches the pair to the wrong order, then the result of that switch is the last out-of-order pair in $X$ therefore $Y = X_1$, and the sequence $X_1 \sim_c^* X_{cbv}$ is just a suffix of the sequence $X \sim_c^* X_{cbv}$ therefore $(X_1)_{cbv} = X_{cbv}$ and again $Y_{cbv} = X_{cbv}$.
-
-The remaining cases are when the subsequence of reduction steps involved in $Y \sim_c X$ occur before the last out-of-order pair in $X$, but possibly overlapping.
-If there is no overlap, then the $\sim_c$ steps do not interfere with each other at all, and can simply be performed in the other order.
-This gives a different sequence of $\sim_c$ steps $Y \sim_c Y_1 \sim_c X_1 \sim_c^* X_{cbv}$.
-As $Y_{cbv} = (Y_1)_{cbv}$, we can proceed by induction in this case and use the fact that $(Y_1)_{cbv} = (X_1)_{cbv} = X_{cbv}$ to acheive the result.
-
-In the other case, there is some overlap between the regions involved in $Y \sim_c X$ and $X \sim_c X_1$, but the last reduction in $X \sim_c X_1$ is later than the last reduction in $Y \sim_c X$.
-There are only 2 reductions in $X$ involved in $X \sim_c X_1$, therefore the region involved in $Y \sim_c X$ ends on the first of the reductions involved in $X \sim_c X_1$.
-Now consider the specific sequence of positions of the reductions in $X$ involved in either of these steps.
-Let the last of these positions be $\alpha$, the first $\gamma$, and the second-last, third-last and so on $\beta_1, \beta_2, \dots$ respectively, and let the positions of the redexes in $Y$ involved in $Y \sim_c X$ be $\beta$ then $(\gamma_i)$ in order..
-Then $X \sim_c X_1$ swaps $\beta_1$ and $\alpha$, and $\alpha$ comes before $\beta_1$ in CBV order (i.e.~$\alpha < \beta_1$ or $\alpha$ is left of $\beta_1$), and the step $Y \sim_c X$ either proceeds forwards towards CBV order, swapping $\beta$ below $\gamma_1 = \gamma$ to form $\gamma$ then $(\beta_i)$, or proceeds backwards, taking $Y$ even farther from CBV order, swapping the $(\gamma_i)$ to above $\beta$, forming $\gamma$ then $\beta_1 = \beta$.
-In each of the cases below, it is required to show that the canonical sequence $(X_i)_i$ eventually reaches some term which is the same as one in $(Y_j)_j$, and from that point on, they match, therefore they reach the same end result: $X_{cbv} = Y_{cbv}$.
-
-\begin{itemize}
-  \item $Y \sim_c X$ in the backwards direction (so that $Y$ is closer to CBV order than $X$ is): By swapping the roles of $X$ and $Y$, and $\beta$ and $\gamma$, this is equivalent to the $Y \sim_c X$ forwards case.
-The assumption that $\alpha$ is before $\beta_1$ in CBV order (so that $\beta_1, \alpha$ is the pair to be swapped in $X \sim_c X_1$) maps to the assumption that $\alpha$ is before $\gamma_1$ (so that the result doesn't just follow trivially by $Y_1$ being equal to $X$) and vice-versa.
-Then one of the cases below establishes that $X_i = Y_j$ for some $i$ and $j$, therefore after swapping back, $Y_i = X_j$.
-\item $Y \sim_c X$ forwards and $\alpha$ comes after $\gamma$ in CBV order: In this case, the $\gamma$ and $\alpha$ reduction steps are already in the correct order in $Y$, therefore $\beta$ and $\gamma$ are already the last out-of-order pair of reductions in $Y$, $Y_1 = X$, and $Y \sim_c X \sim_c^* X_{cbv}$ is already the canonical sequence from $Y$ therefore $Y_{cbv} = X_{cbv}$.
-\item $Y \sim_c X$ forwards and $\alpha$ is disjoint from $\gamma$ and $\beta$: In this case, the sequence $(X_i)_i$ starts by swapping $\alpha$ and $\beta_1$ by case a, then because $\gamma$ is disjoint from $\alpha$ and comes after is in CBV order, $\gamma$ is right of $\alpha$, therefore either all the remaining $\beta_i$s, are $\geq \gamma$ (if $\beta$ and $\gamma$ aren't disjoint) or there are no remaining $\beta_i$s (if $\beta$ and $\gamma$ are disjoint, in which case they swap by case a and there is only $\beta_1$).
-The sequence $(X_i)_i$ therefore proceeds to swap all the remaining $\beta_i$s then $\gamma$, in order, below $\alpha$, resulting in $X_i$ for some $i$ having as its relevant sub-reduction-sequence $\alpha, \gamma, \beta_k, \dots, \beta_1$.
-The other sequence $(Y_j)_j$ starts by swapping $\gamma$ then $\beta$ below $\alpha$, then swapping $\gamma$ and $\beta$.
-Because $\alpha$ is disjoint from $\gamma$, the reduction there does not affect the sub-skeleton at position $\gamma$, therefore this last $\sim_c$ step proceeds identically to how it did in $Y \sim_c X$, resulting in $\gamma, \beta_k, \dots, \beta_i$.
-Overall, this results in $Y_3$ having the relevant sub-reduction-sequence $\alpha, \gamma, \beta_k, \dots, \beta_1$, therefore $Y_3 = X_i$ for the aforementioned value of $i$.
-\item $Y \sim_c X$ forwards, $\alpha \leq \gamma$ and $\alpha$ (and therefore also $\gamma$) is disjoint from $\beta$: In this case, $(X_i)_i$ proceeds by swapping $\beta_1 = \beta$ below $\alpha$ by case a, then swapping $\gamma$ below $\alpha$ by one of the other cases, and $(Y_j)_j$ proceeds by swapping $\gamma$ and $\alpha$, then swapping $\beta$ below $\alpha$ then all of the images of $\gamma$ (which are $\geq \alpha$ and therefore left of $\beta$).
-In both cases, the subterm at $\alpha$ when $\alpha$ and $\gamma$ are swapped is unaffected by the reduction at $\beta$, therefore it produces the same result in both cases, therefore the overall sequence in both cases is $\alpha$ then the images of $\gamma$ then $\beta$.
-\item Both $\gamma$ and $\beta$ are $> \alpha$ but they're disjoint from each other, and either the skeleton at position $\alpha$ is of the form $\tif{\skeletonPlaceholder > 0}{A}{B}$ or it is of the form $A B$ and either both $\gamma$ and $\beta$ are $> \alpha;@_1$ or they are both $> \alpha;@_2$: In this case, $(X_i)_i$ proceeds by swapping $\beta$ then $\gamma$ below $\alpha$, in each case possibly forming 0 or multiple images.
-The only way the number of images of each of these positions can be different is if the subskeleton at $\alpha$ is an $\textsf{if}$, and there are 0 of one of them and 1 of the other, in which case they're trivially in the correct order already.
-Otherwise, all of the images of $\beta$ and $\gamma$ are disjoint from one another, therefore none of the positions, or their relative order, change when they are swapped, therefore the next part of the sequence $(X_i)_i$ is just insertion-sort running on the images of $\beta$ and $\gamma$ with $\sim_c$ case a swaps until they're in the correct order.
-The sequence $(Y_j)_j$ starts by swapping $\gamma$ then $\beta$ below $\alpha$, then as before, sorting the images into the correct order by case a swaps.
-The images originally formed of $\gamma$ and $\beta$ are the same in both cases, therefore the final order is the same, so $X_i = Y_j$ for some $i$ and $j$.
-\item $Y \sim_c X$ forwards, $\alpha;@_1 < \gamma$, and $\alpha;@_2 < \beta$: Let $x$ be the variable involved in the $\beta$-reduction at $\alpha$, so that the sub-skeleton at $\alpha$ is $(\lambda x.
-A) B$ for some $A, B$, where $\gamma = \alpha;@_1;\lambda;\gamma'$ and $A \to A'$ at $\gamma'$, and $\beta = \alpha;@_2;\beta'$, and $B \to B'$ at $\beta'$.
-In this case, $(X_i)_i$ starts by swapping $\beta$ below $\alpha$ by case e, producing one image of $\beta$ for each $x$ in $A'$, then swapping $\gamma$ below $\alpha$ by case d, producing a single image $\alpha;\gamma'$.
-For each of the instances of $x$ in $A$ left of $\gamma'$, the reduction at $\alpha;\gamma'$ is then swapped below the corresponding image of $\beta$.
-The sequence $(Y_j)_j$ starts by swapping $\gamma$ below $\alpha$ then $\beta$ below $\alpha$, but in this case the images of $\beta$ produced may be different.
-In this case the swap happens earlier in the reduction sequence than $\gamma$, so that the body of the lambda is still $A$ rather than $A'$, and the reduction at $\gamma$ may rearrange or change the number of instances of $x$.
-Next, each of the images of $\beta$ to the right of $\alpha;\gamma'$ are swapped with it by case a.
-At this point, the next few reductions before $\alpha;\gamma'$ will in general be images of $\beta$ at positions inside $\alpha;\gamma'$.
-The subskeleton at position $\alpha;\gamma'$ before these reductions is $(A|\gamma')[B/x]$, then by the images of $\beta$ this reduces to $(A|\gamma')[B'/x]$, then it reduces at its root position ($\alpha;\gamma'$ in the overall skeleton) to $(A'|\gamma')[B'/x]$.
-For each position of an $x$ in $(A|\gamma')$, there are 0 or more corresponding positions of $x$ in $(A'|\gamma')$, depending on what type of reduction $A \to A'$ is.
-Because $B$ and $B'$ cannot contain any instances of the variable involved in the reduction $A \to A'$ (if it's a $\beta$-reduction or $\tY$-reduction), each instance of $B$ in $(A|\gamma')[B/x]$ has the same set of images in $(A'|\gamma')[B/x]$ as the corresponding $x$ in $(A|\gamma')$.
-When these images of $\beta$ that overlap with the image of $\gamma$ are swapped below it, each of them has its own images, one at each position of $B$ in $(A'|\gamma')[B/x]$ corresponding to its original copy of $B$ in $(A|\gamma')[B/x]$.
-After all of them are swapped below $\alpha;\gamma'$ (and rearranged among themselved by case a), there is therefore one reduction at each copy of $B$ in $(A'|\gamma')[B/x]$, i.e.~one at position $\beta'$ relative to each $x$ in $(A'|\gamma')$.
-They are therefore the same as the images of $\beta$ in $X_1$ that overlap with $\alpha;\gamma'$, because those are also one reduction at position $\beta'$ relative to each $x$ in $(A'|\gamma)$.
-Both $(X_i)_i$ and $(Y_j)_j$ therefore eventually reach a point where the relevant sub-reduction-sequence is $\alpha$, the images of $\beta$ to the left of $\alpha;\gamma'$, $\alpha;\gamma'$ itself, the aforementioned images of $\beta$ that overlap with $\alpha;\gamma'$, then all the images of $\beta$ to the right of $\alpha;\gamma'$.
-\item $Y \sim_c X$ forwards, $\gamma$ and $\beta$ are disjoint and both $> \alpha$, and the reduction at $\alpha$ is a $\tY$-reduction: This is similar to the previous case, but more complicated, so it will need some definitions to explain properly what's going on.
-Let the term at $\alpha$ before any of the reductions be $\tY \lambda x.
-A$, let $\gamma = \alpha;\tY;\lambda;\gamma'$, let $\beta = \alpha;\tY;\lambda;\beta'$, let $A | \beta' = B \to B'$ with the reduction at $\cdot$, let $A | \gamma' = C \to C'$ with the reduction at $\cdot$, let the positions in $A$ where $x$ occurs that are left of $\gamma'$, between $\gamma'$ and $\beta'$ (but still disjoint from both), and right of $\beta'$ respectively be $(\delta^l_i)$, $(\delta^m_i)$ and $(\delta^r_i)$, let the positions in $B$, $C$, $B'$ and $C'$ respectively where $x$ occurs be $(\delta^B_i)$, $(\delta^C_i)$, $(\delta^{B'}_i)$ and $(\delta^{C'}_i)$ (so that all the positions in $A$ where $x$ occurs, in left-to-right order, are $(\delta^l_i)_i, (\delta^C_i)_i, (\delta^m_i)_i, (\delta^B_i)_i, (\delta^r_i)_i$).
-The sequence $(X_i)_i$ starts by swapping $\beta$ then $\gamma$ below $\alpha$.
-With the shorthand that $\gamma'(\epsilon) = \alpha;\lambda;@_1;\epsilon;\tY;\lambda;\gamma'$, $\gamma'_0 = \alpha;\lambda;@_1;\gamma'$, and similarly for $\beta'$, the relevant portion of $X_2$ is then $\alpha$, $(\gamma'(\delta^l_i))_i$, $\gamma'_0$, $(\gamma'(\delta^{C'}_i))_i$, $(\gamma'(\delta^m_i))_i$, $(\gamma'(\delta^B_i))_i$, $(\gamma'(\delta^r_i))_i$, $(\beta'(\delta^l_i))_i$, $(\beta'(\delta^{C'}_i))_i$, $(\beta'(\delta^m_i))_i$, $\beta'_0$, $(\beta'(\delta^{B'}_i))_i$, $(\beta'(\delta^r_i))_i$.
-Next in $(X_k)_k$, for each $i$, $\gamma'(\delta^r_i)$ are swapped past all the images of $\beta$ until it is immediately before $\beta'(\delta^r_i)$ by case a, then for each $i$, $\gamma'(\delta^B_i)$ is swapped past all the images of $\beta$ until $\beta'_0$ by case a, then swapped with $\beta'_0$ by one of the other cases depending on what type of reduction $B \to B'$ is and the relative positions.
-Even if it's case d or f, all of the images of $\gamma'(\delta^B_i)$ are disjoint from each other (and from the positions of the other reductions after $\beta'_0$) because the redex, $C$, doesn't contain any instances of the variable involved in the reduction $B \to B'$.
-Expanding the definitions, this is swapping $\alpha;\lambda;@_1;\delta^B_i;\tY;\lambda;\gamma'$ with $\alpha;\lambda;@_1;\beta'$ at a point in the reduction sequence where the subskeleton at $\alpha;\lambda;@_1;\beta'$ is $B$ with a mixture of $\tY \lambda x.
-A$ and $\tY \lambda x.
-A[C'/\gamma']$ substituted for its occurrences of $x$, and one of these occurrences of $x$ is at $\delta^B_i$.
-For each $\delta^B_j$, there are some corresponding $\delta^{B'}_k$, where the images of the $x$ at $\delta^B_j$ end up after the reduction $B \to B'$.
-These cover all the $\delta^{B'}_k$, and uniquely, and for each $\delta^B_i$, the images of $\zeta;\delta^B_i;\theta$ after swapping it with $\zeta;\beta'$ are precisely $(\zeta;\delta^{B'}_k;\theta)$ for those same values of $k$, therefore after swapping all of the $(\gamma'(\delta^B_i))_i$ down past $\beta'_0$, their images are $(\gamma'(\delta^{B'}_i))_i$.
-After swapping with $\beta'_0$, each of these images than swaps by case a to take its place among the $(\beta'(\delta^{B'}_i))_i$.
-Next up in $(X_i)_i$, the $(\gamma'(\delta^m_i))_i$s and the $(\gamma'(\delta^{C'}_i))_i$s swap down past some images of $\beta$ they're disjoint from to take their place before their matching image of $\beta$, then $\gamma'_0$ swaps past the $(\beta'(\delta^l_i))_i$, stopping immediately before $\beta'(\delta^{C'}_1)$, then the $(\gamma'(\delta^l_i))_i$ and the $(\beta'(\delta^l_i))_i$ mix.
-The end result of the rearrangement of (this subsequence of) the reduction sequence is therefore $\alpha$, $(\gamma'(\delta^l_i), \beta'(\delta^l_i))_i$, $\gamma'_0$, $(\gamma'(\delta^{C'}_i), \beta'(\delta^{C'}_i))_i$, $(\gamma'(\delta^m_i), \beta'(\delta^m_i))_i$, $\beta'_0$, $(\gamma'(\delta^{B'}_i), \beta'(\delta^{B'}_i))_i$, $(\gamma'(\delta^r_i), \beta'(\delta^r_i))_i$.
-
-  The other sequence, $(Y_j)_j$, proceeds similarly, but with the images of $\gamma$ starting out after the images of $\beta$.
-It swaps all the images of $\beta$ that are right of $\gamma_0$ to the appropriate places by case a, then swaps all the $(\beta'(\delta^C_i)_i$ first to then past $\gamma'_0$, resulting in $(\beta'(\delta^{C'}_i))_i$ for the same reason as with swapping those images of $\gamma$ that overlapped with $\beta'_0$ past it in the $(X_i)_i$ case above, then finally swapping the $(\beta'(\delta^l_i))_i$s and the $(\gamma'(\delta^l_i))_i$s into the correct order.
-As required, this produces the same result as in $(X_i)_i$.
-  \item $Y \to X$ forwards, and $\alpha < \gamma < \beta$: Let $\gamma'$ be such that $\gamma = \alpha;\textsf{if}_i;\gamma'$, $\alpha;@_1;\lambda;\gamma'$, $\alpha;@_2;\gamma'$ or $\alpha;\tY;\lambda;\gamma'$, so that $\gamma'$ is the freely varying later part of $\gamma$ as in the definition of $\sim_c$, whichever case applies to swapping $\gamma$ and $\alpha$.
-Similarly, let $\beta'$ be the freely varying part of $\beta$ relative to $\gamma$.
-Let the images of $\gamma$ after swapping it with $\alpha$ (where this swap occurs later in the reduction sequence than $\beta$) then be $(\alpha;\delta_i;\gamma')_i$, and the images of $\beta$ after swapping it with $\gamma$ be $(\gamma;\epsilon_i;\beta')_i$.
-Depending on the skeleton at $\alpha$, and $\gamma$'s position within it, $(\delta_i)_i$ may be empty (case b), a singleton containing $\cdot$ (cases c and d), all the positions of the relevant variable in the lambda in the skeleton at $\alpha$ (case e), or $\lambda;@_1$ and $\lambda;@_1;\zeta;\tY;\lambda$ for each position $\zeta$ of the relevant variable (case f), but in all cases (except a, which is excluded because none of $\alpha, \gamma$ and $\beta$ are disjoint) the set of images has this general structure.
-Furthermore, the positions $(\delta_i)_i$ are determined only by the general position of $\gamma$ within $\alpha$ (the part excluded from $\gamma'$), and the positions of the relevant variable within the skeleton at $\alpha$.
-They do not depend on $\alpha$ or $\gamma$, so that if both initial positions were moved somewhere else the relative positions of the images of $\gamma$ would be unaffected, and similarly if some position within $\gamma$ were swapped with $\alpha$, the relative positions of its images would be the same (except to the extent that, in case f, the positions of the relevant variable are taken after the inner reduction takes place, so that some of those may still change).
-Because there are so many of them and they don't actually affect the multiset of positions where reductions take place, I will be ignoring swaps by case a in the proof of this case, and just assuming that all the final positions end up in the correct order.
-
-  The final set of reduction positions that I will be proving both $(X_i)_i$ and $(Y_j)_j$ reach is $\alpha, (\alpha;\delta_i;\gamma', (\alpha;\delta_i;\gamma';\epsilon_j;\beta')_j)_i$, where as usual the sequences are expanded out to the full list.
-For the sequence $(X_i)_i$, first the images of $\beta$ in $X_0$ are $(\gamma,\epsilon_i,\beta')$.
-For each of these in turn, it is swapped with $\alpha$.
-By the fact mentioned above that sub-positions are mapped to the same set of images except in case f, the images of $\gamma,\epsilon_j,\beta'$ after this swap are $(\alpha;\delta_i^j;\gamma',\epsilon_j,\beta')_i$, where $(\delta_i^j)_i$ is the same as $(\delta_i)_i$ except that in the case where this is not the first image of $\beta$ swapped below $\alpha$ and the reduction at $\alpha$ is a $\tY$-reduction (and therefore swaps with it proceed by case f), the positions of the relevant variable are taken after the reductions at $\gamma$ and $\gamma;\epsilon_k;\beta'$ for $k \leq j$ (this implies that the first of these, $\delta_i^n = \delta_i$).
-All the images $\alpha;\delta_i^j;\gamma';\epsilon_j;\beta'$ where $\delta_i^j > \lambda;@_1;\gamma';\epsilon_{j+1};\beta'$ are then swapped with $\alpha;\lambda;@_1;\gamma';\epsilon_{j+1};\beta'$ (which is $\alpha;\delta_i^{j+1};\gamma';\epsilon_{j+1};\beta'$ for some $i$).
-As in the case where $\alpha;\tY < \gamma, \beta$ and $\gamma$ and $\beta$ are disjoint, the effects of these swaps may duplicate or delete the reductions in such a way that all the remaining images of $\gamma;\epsilon_j;\beta'$ are $(\alpha;\delta_i^{j+1};\gamma';\epsilon_j;\beta'$.
-This is repeated with $\delta_i^{j+2}$ and so on until these images are all in the correct order, at which point they are $(\alpha;\delta_i;\gamma';\epsilon_j;\beta')_i$.
-After all of these are done, $\gamma$ is swapped with $\alpha$, yielding $(\alpha;\delta_i^{-1};\gamma')_i$.
-As with the images of $\beta$, the subset of these which are inside $\alpha;\lambda;@_1;\gamma';\epsilon_j;\beta'$ for each $j$ in turn are swapped with it, until they are all in the correct order and the images left of $\gamma$ are $(\alpha;\delta_i;\gamma')_i$.
-At this point, $X_i$ has the desired value.
-
-  The sequence $(Y_j)_j$ proceeds similarly.
-First $\gamma$ swaps with $\alpha$, yielding $\alpha$ and $(\alpha;\delta_i;\gamma')_i$, then $\beta$ swaps with $\alpha$, yielding one image of $\beta$ for each $\delta_i$, except that, in case f, those images at positions inside $\alpha;\lambda;@_1;\gamma'$ may differ because they are earlier in the reduction sequence than any of the images of $\gamma$.
-Each of the images of $\beta$ in turn it moved to the corresponding image of $\gamma$, except that those overlapping with $\alpha;\lambda;@_1;\gamma'$ may be duplicated or deleted on the way.
-The images of $\beta$ after just this process (which does not actually correspond to any term in $(Y_j)_j$, but it's sufficiently independent that the order doesn't matter that much) are $(\alpha;\delta_i;\gamma';\beta'')$ (where $\gamma;\beta'' = \beta$).
-The images of these for each $i$, after swapping with $\alpha;\delta_i;\gamma'$, is $(\alpha;\delta_i;\gamma';\epsilon_j;\beta')_j$, because the reductions produced by a swap are unaffected by its overall location, and the skeleton at $\alpha;\delta_i;\gamma'$ at the relevant point in the sequence is equal to the skeleton at $\gamma$ initially, therefore this is equivalent to immediately swapping $\beta$ and $\gamma$ as in $Y \sim_c X$ (except that in the case where the reduction at $\alpha$ is a $\tY$-reduction, the skeleton at $\alpha;\lambda;@_1;\gamma'$ is not actually equal to the skeleton initially at $\gamma$ because something was substituted in for the variable bound at $\alpha;\tY$, but this doesn't affect the variable involved in the reduction at $\gamma$, so this doesn't actually matter).
-Therefore also in this case, eventually the set of reductions in the relevant portion of $Y_j$ is $\alpha, (\alpha;\delta_i;\gamma', (\alpha;\delta_i;\gamma';\epsilon_j;\beta')_j)_i$, therefore it is equal to some $X_i$.
-\end{itemize}
-In summary, if the swap in $Y \sim_c X$ doesn't overlap with the swap $X \sim_c X_0$, the sequence of swaps can be rearranged until it does, and in any other case, there is some $Y_j$ that's equal to some $X_i$, therefore these sequences reach the same end point and if $Y \sim_c X$, then $Y_{cbv} = X_{cbv}$, and by chaining these together, if $Y \sim_c^* X$, $Y_{cbv} = X_{cbv}$.
-
-\paragraph{}
-If $Y$ is in CBV order and $Y \sim_c^* X$, then $Y_{cbv} = X_{cbv}$ but also $Y_{cbv} = Y$ because the sequence $(Y_i)_i$ terminates immediately therefore $X_{cbv}$ is the unique reduction sequence in call-by-value order that's related to $X$ by $\sim_c^*$.
-\end{proof}
-
-\begin{lemma}
-\label{lem:sim-M-N}
-The relation $\sim$ is defined on $L_0(M)$ with reference to a particular starting term $M$, so different versions, $\sim_M$ and $\sim_N$, can be defined starting at different terms.
-If $M \to N$, then $\sim^*_N$ is equal to the restriction of $\sim^*_M$ to $L_0(N)$.
-\end{lemma}
-
-\begin{proof}
-$\sim_N^*$ is trivially a subset of $\sim_M^*$ because $\sim_N$ is a subset of $\sim_M$.
-
-In the other direction, suppose $(N_1, \alpha_1) \sim_M^* (N_2, \alpha_2)$ where both $N_1$ and $N_2$ are descendants of $N$.
-By Lemma~\ref{downAcrossUp}, take $(N_1, \alpha_1) \sim_{p,M}^* (N'_1, \alpha') \sim_{c,M}^* (N'_2, \alpha') \sim_{p,M}^* (N_2, \alpha_2)$.
-The $\sim_p^*$ steps remain within $Rch(N)$, and $\sim_p$ does not depend on the history of the reduction sequences, therefore $(N_1, \alpha_1) \sim_{p,N}^* (N'_1, \alpha') \sim_{c,M}^* (N'_2, \alpha') \sim_{p,N}^* (N_2, \alpha_2)$.
-
-Let $X$ be the call-by-value reduction sequence related to both $N'_1$ and $N'_2$ by $\sim_{c,M}$ given by Lemma~\ref{canonicalCousins}. 
-As $M \to N$ is the first reduction in the sequence $N'_1$, it is the last to be affected by the $\sim_{c,M}$ sequence $N'_1 \sim_{c,M}^* X$ given by Lemma \ref{canonicalCousins} therefore it can be split into $N'_1 \sim_{c,M}^* Y_1 \sim_{c,M}^* X$ where $Y_1$ is in CBV order except possibly for its first reduction, which is still $M \to N$.
-Let the position of the reduction $M \to N$ be $\beta$.
-\akr{TODO: fill in the details that are missing here. It shouldn't be too hard, it's just a bit sketchy at present.} 
-The rearrangement of the reduction sequence $Y_1 \sim_{c,M} X$ consists of moving the reduction at $\beta$ down past the other reductions in $Y_1$, possibly duplicating or deleting it in the process, but not affecting the positions or the correct order for any of the other reductions.
-The reductions derived from $M \to N$ can be identified as follows:
-
-%\begin{definition}
-A position in some (reduction sequence of) term(s) is \emph{derived from} another if it is related by the reflexive transitive closure of $\rightsquigarrow$, where $(V,\gamma) \rightsquigarrow (W,\delta)$ just if $V \to W$ and one of the following cases holds:
-\begin{itemize}
-    \item $(V,\gamma) \sim_p (W,\delta)$
-    \item $\gamma = \delta$, $V \to W$ at $\epsilon$, and $\epsilon > \gamma$
-    \item $V \mid \epsilon = (\lambda x.U) Z, U | \zeta = x, \gamma = \epsilon;@_2;\theta$ and $\delta = \epsilon;\zeta;\theta$
-    \item $V \mid \epsilon = \tY (\lambda x.U), \gamma = \epsilon;\tY;\lambda;\zeta$ and $\delta = \epsilon;\lambda;@_1;\zeta$
-    \item $V \mid \epsilon = \tY (\lambda x.U), U | \zeta = x, \gamma = \epsilon;\theta$ and $\delta = \epsilon;\lambda;@_1;\zeta;\theta$
-\end{itemize}
-%\end{definition}
-
-Crucially, the reductions in $X$ can be partitioned into 2 sets: those at positions derived from $(M,\beta)$, and those at positions equal to the reductions in $Y_1$ (in the same order as they occur in $Y_1$).
-Using the same construction for $N'_2$ shows that the positions of the reductions in $Y_2$ are also the positions of the reductions in $X$ other than those derived from $(M,\beta)$, therefore $Y_1 = Y_2$ therefore $N'_1 \sim_{c,M} Y_1 \sim_{c,M} N'_2$ and every reduction sequence in this sequence starts with $M \to N$ therefore $N'_1 \sim_{c,N} N'_2$.
-
-Combining this with the $\sim_{p,N}^*$s at the beginning and end then yields the desired result that $(N_1,\alpha_1) \sim_N^* (N_2, \alpha_2)$, therefore the restriction of $\sim_M^*$ to $\sim_N$'s domain ($Rch(N)$) is a subset of $\sim_N^*$ therefore the two versions of $\sim$ match on this domain, as desired.
-\end{proof}
-
-At each reduction step $M \to N$, the sample space must be restricted from $I^{L_s(M)}$ to $I^{L_s(N)}$. 
-The injection $L_0(N) \to L_0(M)$ is trivial to define by appending $Sk(M) \to Sk(N)$ to each path, and using Lemma~\ref{lem:sim-M-N}, this induces a corresponding injection on the quotient, $L(N) \to L(M)$.
-The corresponding map $L_s(N) \to L_s(M)$ is then denoted $i(M \to N)$.
-
-Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function, so it is denoted ``$\Rightarrow$'' instead of ``$\red$'', and it relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
-\changed[lo]{We write an element of $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ as $(M', s)$ where the \emph{term} $M' \in \Lambda^0$ and $s \in  I^{L_s(M')}$.}
-\begin{align*}
-& (M,s) \Rightarrow (N,s \circ i(M \to N)) \text{ if either} \\
-& \qquad \text{$M \to N$ and the redex is not $\tsample$, or} \\
-& \qquad \text{$M \mid \alpha = \tsample$, $N = M[\underline{s(Sk(M),\alpha)}/\alpha]$ and $\lambda$ does not occur after $@_2$ or $\tY$ in $\alpha$}
-\end{align*}
-
-\paragraph{}
-\begin{lemma} \label{churchRosser}
-The relation $\Rightarrow$ is Church-Rosser.
-\end{lemma}
-\begin{proof}
-Suppose that $(M,s) \Rightarrow^* (M_1,s_1)$ and also $(M,s) \Rightarrow^* (M_2,s_2)$, then it is required to prove that there is some $(M',s')$ such that both $(M_1,s_1) \Rightarrow^* (M',s')$ and $(M_2,s_2) \Rightarrow^* (M',s')$.
-First consider the special case where $(M,s) \Rightarrow (M_1,s_1)$ and $(M,s) \Rightarrow (M_2,s_2)$, with only a single step in each case.
-Let the positions of the redexes in $M \to M_1$ and $M \to M_2$ be $\alpha_1$ and $\alpha_2$ respectively.
-
-First consider the case that $\alpha_1$ and $\alpha_2$ are disjoint.
-Let $M_1 = M[X_1/\alpha_1]$ and similarly for $X_2$, then let $M' = M[X_1/\alpha_1][X_2/\alpha_2]$.
-As the positions are disjoint, the substitutions commute and both $M_1$ and $M_2$ reduce (with $\to$) to $M'$.
-Let $s' = s \circ i(M \to M_1) \circ i(M_1 \to M')$. 
-The injection $i(M \to M_1) \circ i(M_1 \to M')$ consists of prepending $M \to M_1 \to M'$ to each reduction sequence, but by case a of $\sim_c$, this is equivalent to prepending $M \to M_2 \to M'$, which is $i(M \to M_2) \circ i(M_2 \to M')$.
-In the case that the reduction $M \to M_2$ isn't a $\tsample$-reduction, this is enough to establish that $(M_1, s_1) \Rightarrow (M', s')$ (and similarly if the redex of $M \to M_1$ isn't $\tsample$, $(M_2,s_2) \Rightarrow (M',s')$).
-If it is $\tsample$ though, in order for it to be the case that $(M_1, s_1) \Rightarrow (M', s')$, it is additionally necessary that $X_2$, the result of the reduction at $\alpha_2$, be $s_1(M_1, \alpha_2)$, which follows from the fact that $(M, \alpha_2) \sim (M_1, \alpha_2)$ by case 1 of $\sim_p$.
-The case that the reduction $M \to M_1$ is a $\tsample$-reduction is similar.
-
-The case that $\alpha_1 = \alpha_2$ is trivial, because there is at most one possible $\Rightarrow$ reduction at any given position, therefore $(M_1, s_1) = (M_2, s_2)$ already.
-
-The remaining case is that $\alpha_1 < \alpha_2$ or $\alpha_1 > \alpha_2$. 
-Assume without loss of generality that $\alpha_1 < \alpha_2$. 
-For each possible case of what type of redex $M | \alpha_1$ is, and $\alpha_2$'s position within it, there is a corresponding case of $\sim_c$, and similarly to the case where $\alpha_1$ and $\alpha_2$ are disjoint, the term $O_1 = O_2$ from the definition of $\sim_c$ is a suitable value of $M'$.
-The term $M \mid \alpha_1$ can't be $\tsample$, because it has strict subterms, but the case that $M \mid \alpha_2 = \tsample$ is still somewhat more complicated. 
-$\alpha_2$ can't be within $\alpha_1 ; @_2$ or $\alpha_1 ; Y$ (cases e and f of $\sim_c$) because those are not valid positions to reduce a $\tsample$, 
-and the case that $\alpha_2$ is within the branch of an if statement that is deleted by the reduction at $\alpha_1$ (case b) doesn't actually present a problem because there is no corresponding reduction to $M \to M_2$ in the other branch,
-which leaves cases c and d, that $\alpha_2$ is in the $\tif{\cdot}{\cdot}{\cdot}$ branch that isn't deleted, and that $\alpha_2 < \alpha_1 ; @_1 ; \lambda$.  
-The values that the $\tsample$s take in these cases match because $(M,\alpha) \sim_p (M_1, \alpha;\beta)$ by cases 3 and 2 of $\sim_p$ respectively.
-
-\paragraph{}
-In the case that $(M,s) \Rightarrow^* (M_1,s_1)$ or $(M,s) \Rightarrow^* (M_2,s_2)$ by multiple steps, it is possible to repeatedly replace a pair of the form $A \Leftarrow B \Rightarrow C$ by $A \Rightarrow^* D \Leftarrow^* C$ by the construction above, but it is not immediate that this process terminates, because each $\Rightarrow$ or $\Leftarrow$ may be replaced by multiple, so the sequence of $\Rightarrow$s and $\Leftarrow$s from $(M_1,s_1)$ to $(M_2,s_2)$ may get longer at some steps.
-However, termination can be proved by noting that the structure of this process is identical to the process of swapping $\sim_\uparrow$ and $\sim_\downarrow$ steps in Lemma \ref{downAcrossUp}.
-To be more precise, in this case there is a sequence of $\Leftarrow$ and $\Rightarrow$ steps, each of which has an associated reduction position and initial and final terms, and if a $\Leftarrow$ immediately precedes a $\Rightarrow$, they may be swapped to produce some number of $\Rightarrow$s, followed by some number of $\Leftarrow$s.
-In the case of Lemma \ref{downAcrossUp}, there is a sequence of $\sim_\uparrow$ and $\sim_\downarrow$ steps, each of which has an associated reduction position and initial and final skeletons, and if a $\sim_\uparrow$ immediately precedes a $\sim_\downarrow$, they may be swapped to produce some number of $\sim_\downarrow$s followed by some number of $\sim_\uparrow$s.
-In both cases, the number and reduction positions of the resultant steps are determined by the case of $\sim_c$ that matches the way the initial reduction positions overlap, and the initial skeleton (or the skeleton of the initial term).
-The same argument that the process in Lemma \ref{downAcrossUp} terminated is therefore applicable here too.
-At every stage, the sequence of reductions that results from one of the initial $\Rightarrow$s is a parallel sequence, and swapping a parallel sequence of $\Rightarrow$s with a $\Leftarrow$ always terminates, therefore each $\Rightarrow$ in turn can be moved past all of the $\Leftarrow$s, and the process as a whole will terminate in a state where all of the $\Rightarrow$s precede all of the $\Leftarrow$s, i.e.~a pair of reduction sequences $(M_1,s_1) \Rightarrow^* (M',s') \Leftarrow^* (M_2,s_2)$.
-\end{proof}
-
-\begin{lemma} \label{simpDoesn'tMix}
-If $A$ is some descendant of $M$ and $(A,\gamma) \sim^* (M,\delta)$, then $(A,\gamma) \sim_p^* (M,\delta)$, with the length of the reduction sequences decreasing by one each step from $A$ to $M$.
-\end{lemma}
-\begin{proof}
-This is a simple induction on $\sim^*$.
-In the base case, $(A,\gamma) = (M,\delta)$ therefore $(A,\gamma) \sim_p^* (M,\delta)$ trivially.
-Otherwise, suppose that $(M,\delta) \sim_p^* (B,\epsilon) \sim (A,\gamma)$.
-Either $(B,\epsilon) \sim_p (A,\gamma)$ or $(B,\epsilon) \sim_c (A,\gamma)$.
-In the first case, either $B \to A$, in which case the result follows directly, or $A \to B$, in which case the fact that each (reduction sequence of) skeleton(s) has only one parent implies that a $(M,\delta) \sim_p^* (A,\gamma)$ directly as a subsequence of the path to $(B,\epsilon)$.
-
-In the $\sim_c$ case, consider the definition of $\sim_c$.
-Either $O'_1 = A$ and $O'_2 = B$ or vice-versa.
-As $M \to^* N \to^* O_1 \to^* B$ and $(M,\delta) \sim_p^* (B,\epsilon)$, and $\sim_p$ only relates positions in a term and its parent, there are some positions $\zeta, \theta$ such that $(M,\delta) \sim_p^* (N,\zeta) \sim_p^* (O_1, \theta) \sim_p^* (B,\epsilon)$.
-It follows that $(O_2, \theta) \sim_p^* (A,\gamma)$ by following the same path, therefore it suffices to provide the only missing portion of the path from $M$ to $A$, i.e.~to prove that $(N,\zeta) \sim_p^* (O_2,\theta)$ given $(N,\zeta) \sim_p^* (O_1,\theta)$ (or vice-versa).
-
-If $\zeta$ is disjoint from all the positions of reduction from $N$ to $O_1$ and $O_2$ (and consequently $\zeta = \theta$), this follows from case 1 of $\sim_p$.
-Otherwise, this can be proved by taking cases from the definition of $\sim_c$.
-This is rather long, but all of the cases are similar.
-The general idea is that the reductions from $N$ to $O_1$ correspond to the reductions from $N$ to $O_2$, so that if a position is related by $\sim_p$ across that reduction, it is related in the other branch for the same reason.
-Case d, where $B = O'_1$ rather than $O'_2$, is given here in more detail as an illustrative example:
-
-Let $I$ be $N$ reduced at $\alpha$, and $J$ be $N$ reduced at $\alpha;@_1;\lambda;\beta$, so that $N \to I \to O_1$ and $N \to J \to O_2$.
-All of the reduction positions are $\geq \alpha$, and $\zeta$ is not disjoint from all of them, therefore $\zeta$ is not disjoint from $\alpha$.
-Let $\iota$ be the position such that $(N,\zeta) \sim_p (I,\iota) \sim_p (O_1,\theta)$.
-The fact that $(N,\zeta) \sim_p (I,\iota)$ implies that $\zeta > \alpha;@_1;\lambda$.
-Let $\zeta = \alpha;@_1;\lambda;\kappa$ and $\theta = \alpha;\kappa'$.
-If $\kappa$ is disjoint from $\beta$, then $(N,\zeta) \sim_p (J,\zeta) \sim_p (O_2,\alpha;\kappa') = (O_2;\theta)$ by cases 1 and 2 of $\sim_p$.
-In the other case, that $\kappa$ isn't disjoint from $\beta$, $\kappa > \beta$ because none of the positions $<= \alpha;@_1;\lambda;\beta$ in $N$ are related to any position in $I$ by $\sim_p$.
-As $(I,\alpha;\kappa) \sim_p (O_1,\alpha;\kappa')$ (with the redex at $\alpha;\beta$), for exactly the same reason $(N,\alpha;@_1;\lambda;\kappa) \sim_p (J,\alpha;@_1;\lambda;\kappa')$ (with the redex at $\alpha;@_1;\lambda;\beta$).
-Because $(N,\alpha;@_1;\lambda;\kappa) \sim_p (J,\alpha;@_1;\lambda;\kappa')$, $N|\alpha;@_1;\lambda;\kappa = J|\alpha;@_1;\lambda;\kappa'$, and $N|\alpha;@_1;\lambda;\kappa \neq $ the variable of $N|\alpha;@_1$, therefore $J|\alpha;@_1;\lambda;\kappa'$ is also not the variable therefore $(J,\alpha;@_1;\lambda;\kappa') \sim_p (O_2, \alpha;\kappa') = (O_2, \theta)$ by case 2 of $\sim_p$.
-Combining these results, $(N,\zeta) \sim_p^* (O_2;\theta)$ as desired.
-\end{proof}
-
-\begin{lemma} \label{lem:redexDestroyed}
-If $M \to N$, with the redex at position $\alpha$, then no position in any term reachable from $N$ is related by $\sim^*$ to $(M,\alpha)$.
-\end{lemma}
-\begin{proof}
-Suppose on the contrary that $(M,\alpha) \sim^* (A,\beta)$, where $N \to^* A$, then by Lemma \ref{simpDoesn'tMix}, $(M,\alpha) \sim_p^* (A,\beta)$.
-$M \neq A$ therefore $(M,\alpha) \sim_p$ some position in $N$, but in all the cases of the definition of $\sim_p$, no position in the child term is related to the position of the redex.
-\end{proof}
-
-\paragraph{}
-The reduction relation $\Rightarrow$ is nondeterministic, so it admits multiple possible reduction strategies. 
-A \emph{reduction strategy} starting from a closed term $M$ is a partial function $f$ from $Rch(M)$ to positions, such that for any reachable term $N$ where $f$ is defined, $f(N)$ is a position of a redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
-Using a reduction strategy $f$, a subset of $\Rightarrow$ that isn't nondeterministic, $\Rightarrow_f$, can be defined by $(N,s) \Rightarrow_f (N',s')$ just if $(N,s) \Rightarrow (N',s')$ and $N$ reduces to $N'$ with the redex at $f(N)$.
-
-Essentially what Lemma~\ref{lem:redexDestroyed} demonstrates is that the samples taken during any reduction sequence are independent of each other. This is made more precise in the following lemmas.
-
-\begin{lemma}
-For any skeletons $M \to N$, with the redex at position $\alpha$, and measurable set of samples $S \subset I^{L_s(N)}$, $\mu(S) = \mu(\{s \in I^{L_s(M)} \mid s \circ i(M \to N) \in S\})$, and furthermore, if $M | \alpha = \tsample$, for any $S \subset I \times I^{L_s(N)}$, $\mu(S) = \mu(\{s \in I^{L_s(M)} \mid (s(M,\alpha), s \circ i(M \to N)) \in S\})$.
-\end{lemma}
-\begin{proof}
-If the result holds for all sets $S$ of the form $\{s \in I^{L_s(N)} \mid \forall j : s(j) \in x_i\}$, where $(x_j)_{j \in J}$ is a family of measurable subsets of $I$ indexed by some finite set $J \subset L_s(N)$ of positions, it also holds for all other $S$ by taking limits and disjoint unions.
-
-Take such a $(x_j)_{j \in J}$, where $x_j = x_k$ and define $K = i(M \to N)[J]$.
-Because $i(M \to N)$ is injective, it defines a bijection between $J$ and $K$.
-We can then calculate the measure $\mu(\{s \in I^{L_s(M)} \mid s \circ i(M \to N) \in S\}) = \mu(\{s \in I^{L_s(M)} \mid \forall k : s(k) \in x_{i(M \to N)^{-1}(k)}\}) = \prod_{k \in K} \mu_I(x_{i(M \to N)^{-1}(k)}) = \prod_{j \in J} \mu_I(x_j) = \mu(S)$.
-
-In the case that $M|\alpha = \tsample$, we can similarly consider only those sets $S$ of the form $x_\tsample \times \{s \in I^{L_s(N)} \mid \forall j : s(j) \in x_i\}$.
-The position $\alpha$ in $M$ is not related to any position in $L_s(N)$, therefore $(M,\alpha) \not \in K$.
-Again, we can calculate the measure $\mu(\{s \in I^{L_s(M)} \mid (s(M,\alpha), s \circ i(M \to N)) \in S\}) = \mu(\{s \in I^{L_s(M)} \mid s(M,\alpha) \in x_\tsample, \forall k : s(k) \in x_{i(M \to N)^{-1}(k)}\}) = \mu_I(x_\tsample) \prod_{k \in K} \mu_I(x_{i(M \to N)^{-1}(k)}) = \mu_I(x_\tsample) \prod_{j \in J} \mu_I(x_j) = \mu(S)$.
-\end{proof}
-
-\begin{lemma} \label{lem:independentSamples}
-For any initial term $M$, reduction strategy $f$ on $M$, natural number $n$, skeleton $N$ with $k$ holes, measurable set $T \subset \mathbb R ^ k$ and measurable set $S$ of samples in $I^{L_s(N)}$, $\mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')\}) = \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in I^{L_s(N)} : (M,s) \Rightarrow_f^n (N[r], s')\}) \mu(S)$
-\end{lemma}
-\begin{proof}
-Suppose, to begin with, that $n = 0$.
-Either $\exists r \in T : N[r] = M$, in which case both sides of the equation are $\mu(S)$, or there is no such $r$, in which case both sides of the equation are 0.
-
-For $n > 0$, suppose for induction that the lemma is true for $n - 1$, for all $N, T$ and $S$.
-If $\exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')$, the $r$ and $s'$ are necessarily unique, therefore we may assume that $T$ is the product of $k$ measurable setsets of $\mathbb R$, $(T_j)_{0 \leq j < k}$, as all of the other cases follow by taking unions and limits of these rectangles.
-Let\begin{align*}
-R_s & = \{\text{the skeleton of } O \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) = \tsample, O \to N[r]\} \\
-R_d & = \{\text{the skeleton of } O \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) \neq \tsample, O \to N[r]\}.
-\end{align*}
-Together, $R_s$ and $R_d$ contain all of the skeletons of terms that can reduce to $N$, and they're disjoint.
-For each of these skeletons, there is a map $g_O$ from the reals in $O$ to the reals in $N$, i.e.
-$g_O(r) = r'$ if $O[r] \to N[r']$ for $O \in R_d$, and $g_O(s, r) = r'$ if $O[r] \to N[r']$ with $s$ as the value the sample takes in the reduction.
-All of these functions are measurable, as they are just rearrangements of vector components, or in the case that $O|f(O) = \underline x(\skeletonPlaceholder, \dots, \skeletonPlaceholder)$, a combination fo the measurable function $x$ with a rearrangement of vector components.
-In the case that $O \in R_s$, the function $g_O$ simply inserts the sample into the vector of reals at a certain index, call it $h_O$.
-\begin{align*}
-& \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')\}) \\
-= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in S :\\
-    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\})\\
-    + & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in S :\\
-    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \\
-\end{align*}
-\begin{align*}
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}:\\
-    & g_O(r,s'(O,f(O))) \in T, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
-     + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}:\\
-     & g_O(r) \in T, s' \circ i(O[r] \to N[g_O(r)]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)}:\\
-    & s'(O,f(O)) \in T_{h(O)}, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
-    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}:\\
-    & g_O(r) \in T, s' \circ i(O[r] \to N[g_O(r)]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
-    & \mu(\{s' \in I^{L_s(O)} \mid s'(O,f(O)) \in T_{h(O)}, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S\})\\
-    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
-    & \mu({s' \in I^{L_s(O)} \mid s' \circ i(O[r] \to N[g_O(r)]) \in S}) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(T_{h(O)}) \mu(S)\\
-    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : s'(O,f(O)) \in T_{h(O)}, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S)\\
-    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \mathbb R^{k-1}, s' \in I^{L_s(O)} : g_O(r,s'(O,f(O))) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S)\\
-    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
-= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} :\\
-    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S)\\
-    + & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} :\\
-    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S) \\
-= & \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in I^{L_s(N)} : (M,s) \Rightarrow_f^n (N[r], s')\}) \mu(S)
-\end{align*}
-\end{proof}
-
-The usual call-by-value semantics can be implemented as one of these reduction strategies, given by (with $V$ a value and $T$ a term that isn't a value and $M$ a general term)
-\begin{align*}
-\cbv(T M) & = @_1 ; \cbv(T) \\
-\cbv(V T) & = @_2 ; \cbv(T) \\
-\cbv(\underline f(V_1, \dots, V_{k-1}, T, M_{k+1}, \dots, M_n)) & = \underline f_k ; \cbv(T) \\
-\cbv(\tY T) & = \tY ; \cbv(T) \\
-\cbv(\tif{T < 0}{M_1}{M_2}) & = \textsf{if}_1 ; \cbv(T) \\
-\cbv(\tscore(T)) & = \tscore ; \cbv(T) \\
-\cbv(V) & \text{ is undefined} \\
-\cbv(T) & = \cdot \text{ otherwise}
-\end{align*}
-(this last case covers redexes at the root position).
-
-A closed term $M$ terminates with a given reduction strategy $f$ and samples $s$ if there is some natural number $n$ such that $(M,s) \Rightarrow_f^n (N,s')$ where $f$ gives no reduction at $N$.
-The term terminates almost surely with respect to $f$ if it terminates with $f$ for almost all $s$.
-
-\begin{theorem} \label{thm:AstEquivalence}
-A closed term $M$ is AST with respect to $\cbv$ iff it is AST.
-\end{theorem}
-\begin{proof}
-\akr{I'm not sure whether this proof is sufficiently detailed. It would be pretty easy to fill it in more, just very verbose.}
-
-For both $\red$ and $\Rightarrow_{\cbv}$, the probability measure on the sample space can be used to define a measure on the reduction sequences of each finite length.
-The distributions are equal for each length, by induction, as follows.
-
-The base case, length 0, is trivial, because in both cases the distribution has probability 1 on the sequence containing only $M$.
-
-For the inductive case, suppose that the distributions of reduction sequences of length $n$ are equal.
-\akr{TODO: mention the case that it has already terminated by step $n$, or is about to.
-The two versions of the semantics handle it a little differently.}
-
-For any term $N$ which is not a value, there is a unique environment $E$ and redex $R$ such that $N = E[R]$.
-The position of the hole in $E$ is equal to $\cbv(N)$, so that $N|\cbv(N) = R$, as the cases in the definition of $\cbv$ match the cases in the definition of environments.
-If $R \neq \tsample$, let $R'$ be the result of reducing $R$ (which is equal in both versions of the semantics), then $E[R'] = N[R' / \cbv(N)]$ therefore $\red(N,s) = (E[R'],s)$ and $(N,s) \Rightarrow_{\cbv} (E[R'],s)$.
-The distribution of next terms, conditional on the previous term, in the case that that previous term reduces deterministically, is therefore equal for $\red$ and $\Rightarrow_{\cbv}$.
-
-In the other case that $R = \tsample$, the first (term) part of $\red(N,s)$ is $E[\underline{\pi_h(s)}]$.
-If $k$ is the number of samples already taken in this reduction sequence, this is equivalent to $E[\underline{s_0(k)}]$, where $s_0$ is the initial sample.
-For any sufficiently small neighbourhood of $N$ (i.e.~containing only reduction sequences with the same skeletons), the probability of reaching this neighbourhood is independent of $s(k)$, because it depends only on the samples taken so far, and the samples are independent.
-The distribution of $s(k)$ conditional on reaching $N$ is therefore the uniform distribution on $I$.
-Similarly for the other semantics, if the distribution of remaining samples after $n$ steps is independent of the term by Lemma \ref{lem:independentSamples}, therefore the distribution of $s(Sk(N),\cbv(N))$ conditional on $N$ is the uniform distribution on $I$.
-In either case, the distribution of the next term conditional on the previous term, in the case that it reduces randomly, is equal to the image under $r \mapsto E[\underline r]$ of the uniform distribution on $I$.
-
-In any case, the distribution of reduction sequences after $n+1$ steps, conditional on the reduction sequence after $n$ steps, as defined by $\Rightarrow_{\cbv}$ and by $\red$, is equal, any by the inductive hypothesis the distributions after $n$ steps are equal, therefore by integrating over the reduction sequence up to step $n$, the distributions on reduction sequences up to step $n+1$ are equal.
-
-The distributions on reduction sequences of any finite length as defined by $\Rightarrow_{\cbv}$ and $\red$ are therefore equal, therefore so is the probability of having reached a value after $n$ steps, therefore so is its limit, the probability of termination, therefore the probability of termination is 1 iff the probability of termination with respect to $\cbv$ is 1.
-\end{proof}
-
-\begin{theorem} \label{thm:CbvIsTerminatingest}
-If $M$ terminates with some reduction strategy $f$ and samples $s$, it terminates with $\cbv$ and $s$.
-\end{theorem}
-\begin{proof}
-  Suppose $M$ terminates with the reduction strategy $f$ and samples $s$.
-Let $(M,s) = (M_0,s_0) \Rightarrow_f \dots \Rightarrow_f (M_n,s_n)$, where $M_n$ is a value.
-By Lemma \ref{canonicalCousins}, the reduction sequence $(Sk(M_i))_i$ is related by $\sim_c^*$ to some skeletal reduction sequence $X$ which is in call-by-value order.
-For each skeletal reduction sequence $X_j$ in the sequence $(Sk(M_i))_i \sim_c \dots \sim_c X$, it is possible to define a reduction sequence of terms such that $(M,s) = (N_{j,0},s_{j,0}) \Rightarrow \dots \Rightarrow (N_{j,n_j},s_{j,n_j})$ and $Sk(N_{j,k}) = X_{j,k}$, by at each step applying the reduction at the same place as in $X_j$, essentially just filling in the reals in the skeletal reduction sequence to make it a reduction sequence of terms (it will be shown shortly that the correct branch is taken in each $\textsf{if}$-reduction).
-Assume for induction on $j$ that $(N_{j,n_j},s_{j,n_j}) = (M_n,s_n)$.
-To prove that this holds true for $j+1$ as well, it suffices to show that the corresponding term and samples immediately after the reductions involved in this $\sim_c$ step match, as from that point on, the $\Rightarrow$-sequences will match, as they start from the same point and have all the same reduction positions.
-Letting the term and remaining samples immediately before the reductions involved in this $\sim_c$ step be $(N,t)$ (which is the same in both $N_j$ and $N_{j+1}$ for the same reason again, they start at the same point and the sequences of reduction positions up to this point are the same), and the terms and samples after these reductions be $(O_1,t_1)$ and $(O_2,t_2)$, so that these match the $N$, $O_1$ and $O_2$ in the definition of $\sim_c$, $t_1 = t_2$ because both of them are, by Lemma \ref{lem:sim-M-N}, equal to the restriction of $t$ to the subset of $Rch(N)$ that starts with the reduction sequences to $O_1$ and $O_2$ respectively, and the potential positions in those subsets of $Rch(N)$ are identified with each other by $\sim$ because of the same case of $\sim_c$ that means $N_j$ and $N_{j+1}$ are related.
-If none of the relevant reductions are $\tsample$-reductions, $O_1 = O_2$ trivially.
-If there are samples taken (which must be the arbitrary reduction at $\beta$), it is also required that the same samples are taken in both branches.
-Similarly to Lemma \ref{churchRosser}, the positions of these $\tsample$-reductions in each branch are related by $\sim_p^*$, therefore they correspond to the same sample in $t$, therefore $O_1 = O_2$.
-The fact that the terms are equal also implies that the same branch of an $\textsf{if}$ statements will be taken.
-
-  \paragraph{}
-  There is therefore a reduction sequence $(M,s) = (N_0,s_0') \Rightarrow \dots \Rightarrow (N_m, s_m')$ which ends in a value ($N_m = M_n$) and is in CBV order.
-It doesn't immediately follow that this is the $\Rightarrow_{\cbv}$ reduction sequence starting from $(M,s)$ because the fact that $(N_i)_i$ is in CBV order only means that the reductions that do take place in this reduction sequence are in the correct order, not that the reduction sequence finishes once it reaches a value, or that there are no other reductions that would come earlier in CBV order that don't occur at all in $(N_i)_i$.
-Suppose that at some point $j$ in the reduction sequence, the reduction that occurs is not the next reduction in CBV order, but something else, and that the reduction that should be next is at position $\alpha = \cbv(N_i)$, and the position of the actual reduction $N_j \to N_{j+1}$ is $\beta$.
-Suppose further that $N_j$ is not a value.
-Either $\alpha = \cdot$, the root position, or the reduction at the root position is dependent in some way on the reduction at $\alpha$ (e.g.
-$\alpha = \textsf{if}_1$, or $\alpha = \underline{+};@_1$).
-In any case, all the reductions in the rest of the reduction sequence are after $\alpha$ in CBV order, therefore none of them are $\leq \alpha$, and there continues to be an unreduced redex at $\alpha$, therefore the reduction at the root position never occurs, therefore $N_m$ is not a value, which is a contradiction, therefore at the point (if any) where $(N_i)_i$ departs from the $\Rightarrow_{\cbv}$ reduction sequence starting at $(M,s)$, it has already reached a value, therefore $(M,s) \Rightarrow_{\cbv} \dots$ does reach a value eventually, i.e.~$M$ terminates with the reduction strategy $\cbv$ and samples $s$.
-\end{proof}
-
-\begin{corollary}
-If $M$ is AST with respect to any reduction strategy, it is AST.
-\end{corollary}
-\begin{proof}
-Suppose $M$ is AST with respect to $f$.
-Let the set of samples with which it terminates with this reduction strategy be $X$.
-By Theorem~\ref{thm:CbvIsTerminatingest}, $M$ also terminates with $\cbv$ and every element of $X$, and $X$ has measure 1, by assumption, therefore $M$ is AST with respect to $\cbv$ therefore by Theorem~\ref{thm:AstEquivalence} it is AST.
-\end{proof}
-
-All of the theorems on the termination of rankable terms therefore extend to other reduction strategies too.
-The proofs of Theorems \ref{thm:rankable implies termination}, \ref{thm:partial implies rankable}, \ref{thm:antitone rankable implies termination} and \ref{thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies.
-They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because Theorem \ref{thm:de groote} applies equally to any reduction strategy).
-For a reduction strategy $r$ and term $M$, just substitute $N[r(N)]$ being a $\tY$-redex for $N$ being of the form $E[\tY \lambda x.
-O]$, and $r(N)$ being undefined for $N$ being a value.
-
-The domain of definition of the ranking functions also needs to be changed from the reachable terms $Rch(M)$ to the \emph{reachable terms with respect to the reduction strategy $r$},
-\[
-Rch_r(M) = \{N \mid \exists n, N_0, \dots, N_n: N_0 = M, N_n = N, N_i \to N_{i+1} \text{ at } r(N_i)\}.
-\]
-
-To be more explicit, the modified forms of the theorems are:
-\begin{definition}\rm
-A \emph{ranking function on $M$ with respect to a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
-\begin{itemize}
-    \item $f(N) \geq 1+ f(N')$ if $N[r(N)]$ is a $\tY$-redex and $N$ reduces to $N'$ at $r(N)$
-    \item $f(N) \geq \int_I f(N[\underline{x}/r(N)]) \, \Leb(\mathrm{d}x)$ if $N[r(N)] = \tsample$
-
-    \item $f(N) \geq f(N')$ if $r(N)$ is any other redex, where $N \to N'$ at $r(N)$.
-\end{itemize}
-Any closed term for which a ranking function with respect to a reduction strategy $r$ exists is called \emph{rankable} with respect to $r$. 
-\end{definition}
-
-\begin{definition}
-A \emph{partial ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : Rch_r(M) \rightharpoonup \mathbb R$ such that
-\begin{itemize}
-    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
-    \item $f$ is defined at $M$.
-    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ at the positions specified by $r$ will eventually reach some $O$ such that either $r(O)$ isn't defined or $f(O)$ is, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
-\end{itemize}
-\end{definition}
-
-\begin{definition}\rm
-An \emph{antitone ranking function on $M$ with respect to a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ such that
-\begin{itemize}
-    \item $f(N) \geq \epsilon(f(N)) + f(N')$ if $N[r(N)]$ is a $\tY$-redex and $N$ reduces to $N'$ at $r(N)$
-    \item $f(N) \geq \int_I f(N[\underline{x}/r(N)]) \, \Leb(\mathrm{d}x)$ if $N[r(N)] = \tsample$
-
-    \item $f(N) \geq f(N')$ if $r(N)$ is any other redex, where $N \to N'$ at $r(N)$.
-\end{itemize}
-Any closed term for which an antitone ranking function with respect to a reduction strategy $r$ exists is called \emph{antitone rankable} with respect to $r$. 
-\end{definition}
-
-\begin{definition}
-An \emph{antitone partial ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : Rch_r(M) \rightharpoonup \mathbb R$ such that there exists some antitone function $\epsilon : \nnReal \to \pReal$ such that
-\begin{itemize}
-    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
-    \item $f$ is defined at $M$.
-    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ at the positions specified by $r$ will eventually reach some $O$ such that either $r(O)$ isn't defined or $f(O)$ is, and $f(N) \geq \mathbb E[f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
-\end{itemize}
-\end{definition}
-
-\begin{theorem}
-If a closed SPCF term $M$ has a ranking function, partial ranking function, antitone ranking function or antitone partial ranking function with respect to a reduction strategy $r$, then $M$ is AST with respect to $r$, and AST.
-\end{theorem}
-\lo{Give an example or two of a reduction strategy (for $\Rightarrow$) that is not $\cbv$.}
-
-\akr{NOTE. For the standard CBV semantics, this example is indeed not a problem. 
-In the alternative semantics, defining 
-\begin{align*}
-A[x] &= \tif{\tif{x>0}{I}{I} \, \tsample - 0.5 > 0}{0}{\Omega}\\
-B &= \tif{\tsample - 0.5 > 0}{0}{\Omega}
-\end{align*}
-The set of samples where it terminates is 
-\[\begin{array}{l}
-\bigcup_{r \in [0,1]} \{s \in I^{L_s(A[\tsample])} \mid s([A[\tsample]], \mathsf{if}_1;\underline{ -
-}_1;@_1;\mathsf{if}_1) = r, \\
-\qquad \qquad \hfill \qquad \qquad s([A[\tsample], A[r], \ldots, B, \mathsf{if}_1) > 0.5 \}.
-\end{array}\] 
-This is a rather unwieldy expression, but the crucial part is that $r$ occurs twice in the conditions on $s$: once as the value a sample must take, and once in the location of a sample. 
-It is this set that's unmeasurable.}
-
-\bibliographystyle{apalike}
-\bibliography{references}
-
-\lo{22 Jan: 
-\paragraph{Further directions}
-
-An obvious next step is to extend the result to the $\mathsf{score}$ construct.
-
-The following are highly topical, and could form the basis of an interesting and novel DPhil thesis.
-\begin{enumerate}
-\item Devise methods for proving (positive) a.s.~termination.
-
-- For example, develop a type system satisfying the property: if a term is typable then it is (positively) a.s.~terminating.
-See~\citep{DBLP:conf/ppdp/BreuvartL18,DBLP:conf/esop/LagoG17}.
-
-- Another approach is to develop algorithms that synthesise ranking supermartingales, following, for example, \citep{DBLP:journals/pacmpl/AgrawalC018}.
-
-\item Develop principles (e.g.~in the form of ``proof rules'') for reasoning about (positively) a.s.~termination, in the style of \citep{DBLP:journals/pacmpl/McIverMKK18}.
-
-\item Design algorithms that synthesise probabilistic invariants (\emph{qua} martingales), \`a la \cite{SchreuderO19}; see also \citep{DBLP:journals/pacmpl/HarkKGK20}.
-
-\end{enumerate}}
-
-\iffalse
-\begin{thebibliography}{9}
-\bibitem{ppcf} Thomas Ehrhard, Michele Pagani, and Christine Tasson.
-Measurable cones and stable, measurable functions: a model for probabilistic higher-order programming.
-\emph{PACMPL}, 2(POPL):59:1–59:28, 2018.
-doi: 10.1145/3158147.
-URL \href{https://doi.org/10.1145/3158147}{https://doi.org/10.1145/3158147}.
-\end{thebibliography}
-\fi
-
-\end{document}
+The assumption that $(\mathcal F_n)_{n \in \omega}$ is the coarsest filtration to which $T$ is adapted is used in the equality between the third and fourth lines here.
+This condition is the main reason that this proof does not extend directly to a completeness result for antitone ranking functions.

### 2020-12-26 LICS21/confluent.tex
Created version for LICS21 submission in subdirectory LICS21

--- 2020-12-23 probabilisticProgrammingMartingales.tex
@@
-\documentclass{article}
-
-%% BEGIN {Luke's Macros}
-%%
-\newif\ifdraft
-\drafttrue %% To hide all comments and highlighting, just comment this line out
-
-\input{Style/comment}
-\usepackage{calculation}
-\usepackage{graphicx}
-\usepackage{tikz}
-\usetikzlibrary{positioning, graphs, quotes, graphdrawing, shapes.geometric}
-\usegdlibrary{layered, circular}
-\usepackage[square]{natbib}
-\setcitestyle{aysep={}}
-
-\definecolor{oxblue}{RGB}{0,33,71}
-\definecolor{oxgold}{HTML}{a0630a}
-\usepackage[final,
-bookmarks,
-bookmarksopen,
-colorlinks,
-final,
-linkcolor=red,
-citecolor=oxgold,
-pdfstartview=FitH ]{hyperref}
-
-\usepackage{amsmath} % cleveref must be loaded afer amsmath
-\usepackage{cleveref}
-\Crefname{theorem}{Theorem}{Theorems}
-\Crefname{corollary}{Corollary}{Corollary}
-\crefname{proposition}{Proposition}{Propositions}
-\Crefname{claim}{Claim}{Claims}
-\Crefname{definition}{Definition}{Definitions}
-\Crefname{fact}{Fact}{Facts}
-\Crefname{conj}{Conjecture}{Conjectures}
-\Crefname{example}{Example}{Examples}
-\Crefname{remark}{Remark}{Remarks}
-\Crefname{convention}{Convention}{Conventions}
-\Crefname{lemma}{Lemma}{Lemmas}
-\Crefname{assumption}{Assumption}{Assumptions}
-\Crefname{section}{Section}{Sections}
-\Crefname{appendix}{Appendix}{Appendices}
-\Crefname{figure}{Figure}{Figures}
-%\Crefname{equation}{Eq.}{Equations}
-
-% \addtotheorempostheadhook[theorem]{\crefalias{thmlisti}{theorem}}
-% \addtotheorempostheadhook[lemma]{\crefalias{thmlisti}{lemma}}
-% \addtotheorempostheadhook[proposition]{\crefalias{thmlisti}{proposition}}
-% \addtotheorempostheadhook[corollary]{\crefalias{thmlisti}{corollary}}
-% \addtotheorempostheadhook[claim]{\crefalias{thmlisti}{claim}}
-% \addtotheorempostheadhook[fact]{\crefalias{thmlisti}{fact}}
-% \addtotheorempostheadhook[example]{\crefalias{thmlisti}{example}}
-% \addtotheorempostheadhook[definition]{\crefalias{thmlisti}{definition}}
-% \addtotheorempostheadhook[remark]{\crefalias{thmlisti}{remark}}
-% \addtotheorempostheadhook[assumption]{\crefalias{thmlisti}{assumption}}
-
-\newcommand{\Real}{\mathbb{R}}
-\newcommand{\nnReal}{\mathbb{R}_{{\geq}0}}
-\newcommand{\pReal}{\mathbb{R}_{{>}0}}
-
-\newcommand\expect[1]{\mathbb{E}[#1]}
-\newcommand\set[1]{\{#1\}}
-\newcommand\dif{\mathrm{d}}
-\newcommand\calF{\mathcal{F}}
-\newcommand\calI{\mathcal{I}}
-\newcommand\Leb{\mathrm{Leb}}
-\newcommand\ndraw[2]{\#\mathrm{draw}_{#1}(#2)}
-
-\makeatletter
-\newcommand{\dotDelta}{{\vphantom{\Delta}\mathpalette\d@tD@lta\relax}}
-\newcommand{\d@tD@lta}[2]{%
-  \ooalign{\hidewidth$\m@th#1\mkern-1mu\cdot$\hidewidth\cr$\m@th#1\Delta$\cr}%
-}
-
-%\usepackage[bbgreekl]{mathbbol} %for \bblambda
-%%
-%% END {Luke's Macros}
-
-\usepackage{amsfonts}
-\usepackage{amsthm}
-\usepackage{amssymb}
-\usepackage{verbatim}
-\usepackage{enumitem}
-\newcommand{\tY}{\mathsf{Y}}
-%\newcommand{\tif}[3]{\textsf{if }#1\textsf{ then }#2\textsf{ else }#3}
-\newcommand{\tif}[3]{\mathsf{if}(#1, #2, #3)} % space-saving concrete syntax
-\newcommand{\tsample}{\mathsf{sample}}
-\newcommand{\tscore}{\mathsf{score}}
-\newcommand{\skeletonPlaceholder}{\mathsf{X}} % I'm not sure what the best notation here is.
-\DeclareMathOperator{\red}{red}
-\DeclareMathOperator{\nnext}{next}
-\DeclareMathOperator{\cbv}{cbv}
-
-\usepackage{amsthm}
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}
-\newtheorem{example}{Example}
-
-\theoremstyle{lemma}
-\newtheorem{lemma}{Lemma}
-\newtheorem{theorem}{Theorem}
-\newtheorem{corollary}{Corollary}
-\newtheorem{conjecture}{Conjecture}
-
-\theoremstyle{remark}
-\newtheorem*{remark}{Remark}
-
-\title{Supermartingales and Termination Analysis of Statistical PCF}
-
-\begin{document}
-
-\maketitle
-
-\lo{Tentative title}
-
-\begin{abstract}
-\lo{TODO}
-\end{abstract}
-
-\iffalse
-\akr{Following your suggestion, I will be providing a criterion for termination of programs in PPCF \citep{DBLP:journals/pacmpl/EhrhardPT18} based on ranking supermartingales. 
-As it's more convenient for this proof, a sampling-based semantics \lo{You need to provide reference(s) for sampling-based semantics.} will be used instead of the original distributional semantics. 
-I assume some roughly applicable equivalence is proven somewhere, but it doesn't seem that hard anyway.}
-\lo{Why not provide a proof as an appendix?}
-\fi
-
-\tableofcontents
-
-Various theorems about probabilistic programs rely on the assumption that the program terminates almost surely.
-Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \citep{DBLP:journals/pacmpl/McIverMKK18}.
-This paper's contribution is to extend this method to a higher-order setting.
-
-\section{Statistical PCF}
-
-\subsection{Syntax of SPCF}
-
-The language SPCF is a simply-typed lambda calculus with sampling of real numbers from $[0,1]$ and unbounded scoring, following \cite{MakOP20b}.
-Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
-\begin{align*}
-  & \text{types } A, B ::= \textsf{R}  \mid  A \to B \\
-  & \text{values } V ::= \lambda x.M  \mid  \underline{r} \\
-  & \text{terms } M, N ::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY M  \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample  \mid  \tscore(M)
-\end{align*}
-\begin{align*}
-  \frac{}{\Gamma ; x:A \vdash x:A} \qquad
-  \frac{\Gamma ; x:A \vdash M : B}{\Gamma \vdash \lambda x.M : A \to B} \qquad
-  \frac{}{\underline{r} : \textsf{R}} \qquad
-  \frac{\Gamma \vdash M:A \to B \quad \Gamma \vdash N : A}{\Gamma \vdash M \, N : B} \\ \\
-  \frac{\Gamma \vdash M_1:\textsf{R} \dots \Gamma \vdash M_n:\textsf{R}}{\Gamma \vdash \underline{f}(M_1,\dots,M_n) : \textsf{R}} \ f : \mathbb{R}^n \to \mathbb{R} \qquad
-  \frac{\Gamma \vdash M : (A \to B) \to (A \to B)}{\Gamma \vdash \tY M : (A \to B)} \\ \\
-  \frac{\Gamma \vdash M : \textsf{R} \quad \Gamma \vdash N_1 : A \quad \Gamma \vdash N_2 : A}{\Gamma \vdash \tif{M < 0}{N_1}{N_2} : A} \qquad
-  \frac{}{\Gamma \vdash \tsample : \textsf{R}} \qquad
-  \frac{\Gamma \vdash s : \textsf{R}}{\Gamma \vdash \tscore (s) : \textsf{R}}
-\end{align*}
-
-Terms are identified up to $\alpha$-equivalence, as usual. The set of all terms is denoted $\Lambda$, and the set of closed terms is denoted $\Lambda^0$.
-
-\paragraph{}
-To define the reduction relation, let \emph{evaluation contexts} be of the form:
-\begin{align*}
-  E ::= & \, [\,] \mid E \, M \mid V \, E \mid \underline{f}(\underline r_1,\dots ,\underline r_{k-1}, E, M_{k+1}, \dots, M_n) \\ & \mid \tY E \mid \tif{E<0}{M_1}{M_2} \mid \tscore (E)
-\end{align*}
-then a term reduces if it is formed by substituting a redex in a context i.e.
-\begin{align*}
-  E[(\lambda x.M) N] & \to E[M[N/x]] \\
-  E[\underline f (\underline r_1, \dots , \underline r_n)] & \to E[\underline{f(r_1,\dots,r_n)}] \\
-  E[\tY \lambda x. M] & \to E[\lambda z. M[(\tY \lambda x. M)/x] z] \text{ where $z$ is not free in $s$}\\
-  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_1] \text{ where }r < 0 \\
-  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_2] \text{ where }r \geq 0 \\
-  E[\tsample] & \to E[\underline r] \text{ where } r \in [0,1] \\
-  E[\tscore(\underline r)] & \to E[\underline r].
-\end{align*}
-\changed[lo]{We write $\to^\ast$ for the reflexive, transitive closure of $\to$.}
-\lo{The $\tscore$ rule should have a side condition: if $r > 0$.}
-
-Every closed well-typed term either is a value or reduces to another term.
-
-\subsection{Sampling semantics}
-\label{sec:sampling semantics}
-This version of the reduction relation allows $\tsample$ to reduce to any number in $[0,1]$.
-To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples.
-Let $ I = [0,1] \subset \mathbb{R} $, and let $S = I^{\mathbb{N}}$, with the \changed[lo]{Borel $\sigma$-algebra $\calF$} and the probability measure, denoted $\mu$, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
-Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and $\mu (\prod_{i=0}^\infty X_i) = \prod_{i=0}^\infty \Leb(X_i)$.
-The maps $\pi_h:S \to I, \; \pi_t:S \to S$ popping the first element are then measurable.
-\changed[lo]{Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(S, \calF, \mu)$ the \emph{entropy space}.}
-
-The $\sigma$-algebra and measure on $\Lambda$ are defined by considering it as a disjoint union of equivalence classes under replacing all the real constants by a placeholder, where the measure on each class is that of $\mathbb{R}^n$, where $n$ is the number of real constants.
-\changed[lo]{Precisely, following \cite{DBLP:conf/icfp/BorgstromLGS16},
-we view $\Lambda$ as $\bigcup_{m\in\omega} \big(\mathsf{Sk}_m \times \Real^m \big)$,
-where $\mathsf{Sk}_m$ is the set of SPCF terms with exactly $m$ place-holders (a.k.a.~\emph{skeleton terms}) for numerals.
-Thus identified, we give $\Lambda$ the countable disjoint union topology of the product topology of the discrete topology on $\mathsf{Sk}_m$ and the standard topology on $\Real^m$.
-Note that the connected components of $\Lambda$ have the form $\{M\} \times \Real^m$, with $M$ ranging over $\mathsf{Sk}_m$, and $m$ over $\omega$. 
-We fix the Borel algebra of this topology to be the $\sigma$-algebra on $\Lambda$.}
-
-The one-step reduction is given by the function $\red : \Lambda^0 \times S \to \Lambda^0 \times S$ where
-\begin{equation}
-\red(M,s) = \left\{
-    \begin{array}{ll}
-        (E[N],s) & \text{if } M = E[R], R \to N \text{ and } R \neq \tsample \\
-        (E[\underline{\pi_h(s)}],\pi_t(s)) & \text{if } M = E[\tsample] \\
-        (M,s) & \text{if } M \text{ is a value}
-    \end{array} \right .
-\end{equation}
-\lo{Both $\to$ and $\red$ are called reduction relation.}
-
-The result after $n$ steps is then simply $\red^n(M,s) = \overbrace{\red(...\red(}^n M,s)...)$, and the limit $\red^\infty$ can then be defined as a partial function as $\lim_{n \to \infty} \red^n(M,s)$ whenever that sequence becomes constant by reaching a value.
-A term $M$ terminates for a sample sequence $s$ if the limit $\red^\infty(M,s)$ is defined.
-
-The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid M \text{ terminates at } s \}$ is measurable.
-A term $M$ is said to terminate almost surely if $\mu(\{s \mid M \text{ terminates at } s\}) = 1$.
-
-\lo{Alternatively, define the \emph{runtime of $M$} to be the random variable 
-\[
-T_M(s) := 
-\begin{cases}
-\min \set{n \mid \pi_0(\red^n(M, s)) \textrm{ is a value}} & \hbox{if $\red^\infty(M,s)$ is defined}\\
-\infty & \hbox{otherwise}
-\end{cases}
-\]
-Equivalently, we say that $M$ is \emph{almost-surely terminating} (AST) if $T_M < \infty$ a.s.; 
-and $M$ is \emph{positively almost-surely terminating} (PAST) if $\expect{T_M} < \infty$.}
-
-For example, the term 
-\[
-(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0},
-\] 
-which generates a geometric distribution, terminates on the set $S \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas 
-\[
-\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \underline 0},
-\] 
-which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.
-
-\paragraph{}
-Note that the requirement that $M$ be closed is not important for the proofs of the theorems, but is simply required to ensure that $M$ reaches a value, rather than a general normal form like $x \, y$.
-
-\paragraph{}
-This definition of almost-sure termination is equivalent to that given in \citep{MakOP20b}, although the program semantics is stated in a slightly different way.
-In particular, the argument to $\tscore$ is not relevant to termination (except for the possibility that its argument's evaluation wouldn't terminate).
-
-\iffalse
-\lo{Your operational semantics does not maintain a record of the current weight of the reduction.
-A.s.~termination does depend on $\tscore$: see \cite[\S 4.3]{DBLP:journals/corr/abs-2004-03924}\footnote{\url{https://arxiv.org/abs/2004.03924}}.
-I think it important to take the behaviour of $\tscore$ into account;
-you should do it as a future task.}
-\fi
-
-\section{Supermartingales}
-One approach to proving that a term terminates almost surely is to find some variant that is bounded below and, on average, decreases sufficiently quickly that it must eventually reach 0, similarly to the approach taken in \citep{DBLP:journals/pacmpl/McIverMKK18} for imperative programs.
-
-These variants are defined as functions from reachable terms (i.e.~possible states of the program's execution) to real numbers.
-Specifically, let the set of reachable terms from a given closed starting term $M$ be $Rch(M) := \{N \in \Lambda \mid M \to^* N \}$, with the $\sigma$-algebra induced as a subset of $\Lambda$.
-
-\begin{definition}\rm
-A \emph{ranking function on $M$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
-\begin{enumerate}
-    \item $f(E[\tY \lambda x. N]) \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z) \text{ where $z$ is not free in $N$}$
-    \item $f(E[\tsample]) \geq \int_I f(E[\underline{x}]) \, \Leb(\mathrm{d}x)$
-
-    \item $f(E[R]) \geq f(E[R'])$ for any other redex $R$, where $R \to R'$.
-\end{enumerate}
-We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that $f(N) = 0$ iff $N$ is a value, and for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.
-
-Any closed term for which a ranking (respectively, strict ranking) function exists is called \emph{rankable} (respectively, \emph{strictly rankable}). 
-\changed[lo]{For example, the term $(\tY \lambda x.x) \, \underline 0$ is not rankable.}
-It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \in \omega}$ is the reduction sequence starting from $M$, then $(f(M_n))_{n\in\omega}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales.
-\end{definition}
-
-\input{ranking-supermartingales}
-
-\section{Constructing Ranking Functions}
-Although rankability implies almost-sure termination, the converse does not hold in general. For example,
-\begin{equation}
-\tif{{-}\tsample < 0}{\underline{0}}{(\tY \lambda x. x) \, \underline 0}
-\label{ex:0 probability reachable}
-\end{equation}
-terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x. x) \, \underline 0$ is reachable, although that has probability 0. 
-Not only is this counterexample AST, it's PAST. %positively almost surely terminating i.e.~the expected time to termination is finite.
-
-A ranking function can be constructed under the stronger assumptions that, for every $N$ reachable from $M$, the expected number of $\tY$-reduction steps from $N$ to a value is finite.
-In particular, the expected number of $\tY$-reduction steps from each reachable term is a ranking function.
-Note that a finite number of expected $\tY$-reduction steps does not necessarily imply a finite number of expected total reduction steps.
-
-\begin{example}
-\label{ex:tY finite does not imply t finite}
-The term
-\[
-M = \big(\tY \lambda f n. \tif{\tsample - 0.5 < 0}{n \, \underline 0}{f (\lambda x. n (n x))}\big) (\lambda x. x+1)
-\] 
-terminates with only 2 $\tY$-reductions on average i.e.~$\expect{T_M^\tY} = 2$, but applies the increment function $2^n$ times with probability $2^{-1-n}$ for $n \geq 0$, which diverges i.e.~$\expect{T_M} = \infty$.
-\end{example}
-
-\begin{theorem} \label{thm:minimal}
-Given a closed term $M$, the function $f:Rch(M) \to \mathbb R$ given by $f(N) = \mathbb E [\text{the number of }\tY\text{-reduction steps from }N\text{ to a value}]$, if it exists, is the least of all possible ranking functions of $M$.
-\end{theorem}
-\begin{proof}
-Let $f$ be the candidate least ranking function defined above, and suppose $g$ is another ranking function such that $f(N) > g(N)$ for some $N \in Rch(M)$.
-The restrictions of $f$ and $g$ to $Rch(N)$ have the same properties assumed of $f$ and $g$, so assume w.l.o.g.~that $N=M$.
-The difference $g - f$ is then a supermartingale (with the same setup as in the proof of %\Cref{thm:rankable gives supermartingale}
-%Theorem \ref{rankable implies ast}) 
-\Cref{thm:rankable and strict rankable});
-therefore $\mathbb E[g(M_n)] \leq \mathbb E[f(M_n)] + g(M)-f(M)$, for all $n$.
-
-We have $\mathbb E[f(M_n)] = \sum_{k=n}^\infty \mathbb P[M_k = E[\tY N] \text{ for some }E, N] \to 0 \text{ as } n \to \infty$; 
-therefore as $g(M) - f(M) < 0$, eventually $\mathbb E[g(M_n)] < 0$, which is impossible. 
-It follows that $g \geq f$ as required.
-
-In order for $f$ to be the least ranking function of $M$, it also has to actually be a ranking function itself.
-Each of the conditions on a ranking function is easily verified from the definition of $f$.
-\end{proof}
-
-\subsection{Partial ranking functions}
-Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost-sure termination, so it isn't useful for proving it.
-
-\begin{example}[Geometric distribution]
-Consider the term
-\begin{align*}
-&(\tY \lambda f, n. \\
-&\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
-&) \, \underline{0}
-\end{align*}
-which generates a geometric distribution (let $\Theta = \lambda f, n : \tif{\tsample \underline{- 0.5} < 0}{n}{f (n\underline{+ 1})}$, then the term is $(\tY \Theta)\ \underline 0$).
-    Despite its simplicity, its $Rch$ contains all the terms
-
-\begin{tikzpicture}[node distance=0.5em, anchor=west]
-    \node (1) at (0,-0) {$\tY\ \Theta\ \underline 0$};
-    \node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\ \Theta\ (n \underline{+ 1})}) z)\ \underline i$};
-    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\ \Theta\ (n \underline{+ 1})})\ \underline i$};
-    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY\ \Theta\ (\underline i\ \underline{+ 1})}$};
-    \node (5) at (0,-4) {$\tif{\underline r\ \underline{- 0.5} < 0}{\underline i}{\tY\ \Theta\ (\underline i\ \underline{+ 1})}$};
-    \node (6) at (0,-5) {$\tif{\underline r < 0}{\underline i}{\tY\ \Theta\ (\underline i\ \underline{+ 1})}$};
-    \node (7) at (2,-6) [text width=] {$\underline i$};
-    \node (8) at (0,-7) {$\tY\ \Theta\ (\underline i\ \underline{+ 1})$};
-    \node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\ \Theta\ (n \underline{+ 1})}) z)\ (\underline i\ \underline{+ 1})$};
-
-    \draw [->] (1.south west) to [bend right=45] (2.north west);
-    \draw [->] (2.south west) to [bend right=45] (3.north west);
-    \draw [->] (3.south west) to [bend right=45] (4.north west);
-    \draw [->] (4.south west) to [bend right=45] (5.north west);
-    \draw [->] (5.south west) to [bend right=45] (6.north west);
-    \draw [->] (6.south west) to [bend right=45] (8.north west);
-    \draw [->] (8.south west) to [bend right=45] (9.north west);
-
-    \draw [->] (9.west) to [bend left=45] (2.west);
-    \draw [->] (6) to (7);
-\end{tikzpicture}
-
-\end{example}
-
-Even in this simple case, defining a ranking function explicitly is awkward because of the number of cases, although in most cases, because the value need only be greater than or equal to that of the next term in sequence, it suffices to take the ranking function as having the same value as the next term, so that overall it takes only 3 distinct values.
-(We will explain why later.)
-
-The definition of rankability is also inconvenient for syntactic sugar.
-It could be useful, for example, to define $M \oplus_p N = \tif{\tsample - p < 0} M N$, where $M \oplus_p N$ reduces to $M$ or $N$, depending on the first value of $s$\lo{Need a name for $s$.}, with probability $p$ resp.~$(1-p)$.
-Technically though, it reduces first to $\tif{\underline r - p < 0} M N$ for all $r \in I$, so those terms all need values of the ranking function too.
-
-\paragraph{}
-In both of these cases, there are only some values of the ranking function that are semantically important. 
-\begin{definition}
-Define a \emph{partial ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \rightharpoonup \mathbb R$ such that
-\begin{itemize}
-    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
-    \item $f$ is defined at $M$.
-    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually \akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and ast, but that doesn't actually work because of terms reachable with 0 probability.
-Of course, astness doesn't imply rankability, so any definition of partial ranking function that includes all Y-past terms wouldn't always extend to a total ranking function.} reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
-\end{itemize}
-\end{definition}
-A partial ranking function that is total is just a ranking function.
-Providing a partial ranking function is essentially part way between providing a ranking function and directly proving almost sure termination.
-
-\begin{theorem} \label{thm:partial implies rankable}
-Every partial ranking function is a restriction of a ranking function.
-\end{theorem}
-\begin{proof}
-Take a closed term $M$ and partial ranking function $f$ on $M$. Define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
-\begin{align*}
-    f_1(N) & := f(N) \text{ whenever $f(N)$ is defined,} \\
-    f_1(V) & := 0 \text{ for values $V$ not in the domain of $f$.}
-\end{align*}
-
-Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) := \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
-The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on partial ranking functions.
-Define $f_2(N) = \int_S f_1(\nnext(N,s)) + g(N,s) \, \mu(\mathrm d s)$.
-The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction).
-\lo{TODO. The preceding sentence deserves a proof.}
-\end{proof}
-
-As a corollary, any term which admits a partial ranking function terminates almost surely.
-
-\subsection{Examples}
-Let $M \oplus_p N = \tif{\tsample - p < 0} M N$, for $p \in (0,1]$, then there are the pseudo-reduction relations
-\begin{align*}
-E[M \oplus_p N] \to^3 & E[M] & \\
-E[M \oplus_p N] \to^3 & E[N] & \\
-\red^3(E[M \oplus_p N], s) = & \left\{
-    \begin{array}{ll}
-        (E[M],\pi_t(s)) & \text{if } \pi_h(s) < p \\
-        (E[N],\pi_t(s)) & \text{if } \pi_h(s) \geq p. \\
-    \end{array} \right .
-\end{align*}
-A partial ranking function could be defined with respect to this shortcut reduction simply by replacing $\to$ and $\red$ in the definition of a partial ranking function by a version that goes straight from $N \oplus_p O$ to $N$ or $O$.
-Such a pseudo-partial ranking function would then be a partial function from a subset of $Rch(M)$, so it could also be considered as a partial function from all of $Rch(M)$, and it would in fact also be an actual partial ranking function.
-It is therefore possible to prove rankability directly using the shortcutted reductions.
-
-A similar procedure would work for other forms of syntactic sugar.
-If a closed term $N$ eventually reduces to one of a set of other terms $\{N_i \mid i \in I\}$ with certain probabilities, a partial ranking function defined with respect to a reduction sequence that skips straight from $N$ to $N_i$ is also a valid partial ranking function for the original reduction function, and therefore its existence implies almost sure termination.
-There is a caveat, however, that $\tY$-reduction steps skipped over in the shortcut still need to be counted for the expected number of $\tY$-reduction steps.
-
-\paragraph{}
-With this abbreviation, the geometric distribution example from earlier can be written as $(\tY \lambda f n.\ n \oplus_{0.5} f(n+1)) \, \underline 0$.
-It is then easy to see that the following is a partial ranking function:
-\begin{align*}
-(\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\ N & \mapsto 2 \\
-\underline i \oplus_{0.5} (\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\ (\underline i + 1) & \mapsto 1 \\
-\underline i & \mapsto 0.
-\end{align*}
-
-In fact, even the partial function $\tY \Theta N \mapsto 2$ alone is a partial ranking function for this term.
-\section{Antitone ranking functions}
-
-\begin{example}[Random walk]
-\label{ex:ac-ranking}
-\begin{enumerate}
-\item 1D biased (towards 0) random walk
-\[
-M_1 = 
-\big(\tY \, \lambda f n . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{3/2} f \, (n + 1)}\big) \, 10
-\]
-
-\item 1D unbiased random walk
-\[
-M_2 = 
-\big(\tY \, \lambda f n . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}\big) \, 10
-\]
-
-\item 1D biased (away from 0) random walk
-\[
-M_3 = 
-\big(\tY \, \lambda f n . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/3} f \, (n + 2)}\big) \, 10
-\]
-
-%While {x > 0} do {x := x-1 \oplus_2/3 x:= x+1}
-
-%M2 = While {x > 0} do {x := x-1 \oplus_2/3 x:= x+2}
-\end{enumerate}
-\end{example}
-
-The term $M_1$ is rankable, terminating in 31 $\tY$-steps on average, and $M_3$ only has a $1/1024$ chance of terminating, but in between, $M_2$ is AST, but isn't $\tY$-PAST, therefore it isn't rankable and \Cref{thm:rankable implies termination} is insufficient to prove its termination.
-We therefore want to find a generalised notion of ranking function so that $M_2$ becomes rankable, and then prove it sound i.e.~rankable in this generalised sense implies AST.
-
-\paragraph{Idea.}
-
-The definition of antitone-convex ranking function $f$ (say) for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
-we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) function $\epsilon : \nnReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
-\[
-f(E[R’]) \leq f(E[R]) - \epsilon(f(E[R])) 
-\]
-where $R \to R’$ is the $\tY$-redex rule.
-
-\lo{An aside: This CBV $\tY$-rule seems cleaner:
-\[
-\big(\tY f^{A \to B} \, x^A \, . \, \theta^B \big) \, v \to \theta[(\tY f \, x \, . \, \theta) / f, v / x].
-\]
-We assume $\tY f^{A \to B} \, x^A \, . \, \theta^B$ is a value.}
-
-\begin{definition}
-We say that a supermartingale $(Y_n)_{n \in \omega}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \in \omega}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
-\[
-\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot 1_{\set{T > n}}
-\]
-\end{definition}
-
-%\lo{The preceding definition does not place any constraint on the stopping time $T$ (other than that it is w.r.t.~the filtration $(\calF_n)_{n \in \omega}$).
-However for the definition to make sense, $T$ must be the stopping time $T(s) := \min \set{n \mid Y_n(s) = 0}$.}
-
-\begin{theorem}
-\label{thm:a-c strict}
-Let $(Y_n)_{n \in \omega}$ be an antitone strict supermartingale via function $\epsilon : \nnReal \to \pReal$, and stopping time $T$. 
-Then $T < \infty$ almost surely.
-\end{theorem}
-
-\begin{proof}
-First, as $(Y_n)$ is a supermartingale, $\expect{Y_n} \leq \expect{Y_0}$. 
-Therefore
-
-\begin{calculation}
-\expect{Y_n \mid T > n}
-  \step[=]{rearranging terms}
-\dfrac{\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
-+ 
-\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
-- 
-\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
-  \step[=]{definition of conditional expectation}
-\dfrac{\expect{Y_n} - \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
-  \step[\leq]{$Y_n \geq 0$ always}
-\dfrac{\expect{Y_n}}{\mathbb P[T > n]}
-  \step[\leq]{$(Y_n)_n$ is a supermartingale}
-\dfrac{\expect{Y_0}}{\mathbb P[T > n]}
-\end{calculation}
-
-\emph{Claim}: For all $0 < x \leq 1$, 
-\(
-\mathbb P[T > B_x] \leq x,
-\)
-where $B_x = \left \lceil \dfrac{\expect{Y_0}+1}{x \; \epsilon(\expect{Y_0} \, x^{-1})} \right \rceil$.
-
-\smallskip
-
-As the convex hull of $\epsilon$ (the greatest convex function less than or equal to it) satisfies all the conditions assumed of $\epsilon$, in addition to being convex, assume wlog that $\epsilon$ is convex.
-
-Assume for a contradiction that $\mathbb P[T > B_x] > x$.
-Then, take $n \leq B_x$. 
-We have
-%\lo{Change of justification of the second equality below: event, not discrete r.v. Recall: expectation conditioning on r.v.~(resp.~event) is a r.v.~(resp.~number).}
-\begin{calculation}
-\expect{Y_n-Y_{n+1}}
-  \step[=]{$\calF_n \subseteq \calF_{n+1}$, definition and linearity of conditional expectation}
-\expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
-  \step[\geq]{antitone strict assumption}
-\expect{\epsilon(Y_n) \cdot 1_{\set{T > n}}}
-%  \step[=]{definition of conditional expectation for a discrete variable}
-\step[=]{definition of expectation conditioning on an \emph{event}}
-\mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
-  \step[\geq]{Jensen's inequality}
-\mathbb P[T > n]\ \epsilon(\expect{Y_n \mid T > n})
-  \step[\geq]{proved earlier}
-\mathbb P[T > n]\ \epsilon\Big(\dfrac{\expect{Y_0}}{\mathbb P[T > n]}\Big)
-  \step[>]{assumption, $\mathbb P[T > n] \geq \mathbb P[T > B_x] > x$}
-x \; \epsilon\big(\expect{Y_0} \, x^{-1}\big).
-\end{calculation}
-Therefore, by a telescoping sum
-\[
-\expect{Y_{B_x}} = \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1}) \leq -1 < 0 
-\]
-%\lo{$\expect{Y_0 - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})}$}
-which is a contradiction, therefore the claim must be true, therefore $P[T > n]$ tends to 0 as $n$ tends to infinity, therefore $T < \infty$ almost surely.
-\end{proof}
-
-Note that it is essential in this proof that $\epsilon$ be defined for all $\mathbb R_{\geq 0}$, not just the values that $(Y_n)_{n \in \omega}$ actually takes (or at least if $(Y_n)_{n \in \omega}$ is uniformly bounded, $\epsilon$ must be positive at the supremum as well as the actual values).
-
-\begin{example}
-\akr{Consider a random variable $X$, uniformly distributed in the interval $[-1,1]$, $T := \min \set{n \mid 2^{-n} < X}$, 
-\[
-Y_k := 
-\begin{cases}
-4-2^{-k} & \hbox{if $k < T$}\\
-0 & \hbox{otherwise}
-\end{cases} 
-\]
-} \lo{The following reformulation defines $T$ and $Y_k$ directly as a measurable function on a probability space (rather than via a r.v.~$X$).}
-
-Take the probability space $(\Omega, \calF, \mathbb{P})$ where $\Omega$ is the closed interval of reals $[-1, 1]$, $\calF$ the Borel $\sigma$-algebra, and $\mathbb{P}$ the corresponding Lebesgue probability measure.
-Let $\omega \in [-1, 1]$. Define random variables $T$ and $(Y_k)_{k \in \omega}$:
-\begin{align*}
-T(\omega) & := 
-\begin{cases}
-\min\set{n \in \mathbb{N} \mid \omega > 2^{-n}} & \hbox{if $\omega \in (0, 1]$}\\
-\infty & \hbox{otherwise}
-\end{cases} 
-\\
-Y_k(\omega) & := 
-\begin{cases}
-4-2^{-k} & \hbox{if $\omega \in [-1, 2^{-k}]$ (equivalently $k < T(\omega)$)}\\
-0 & \hbox{otherwise}
-\end{cases} 
-\end{align*}
-\lo{N.B.~$\set{T = n} = (\frac{1}{2^{n}}, \frac{1}{2^{n-1}}]$, and so, $\set{T > n} = [-1, \frac{1}{2^n}]$.}
-Plainly, $T$ is a stopping time, and $(Y_n)_{n \in \omega}$ is a supermartingale, adapted to the filtration $(\calF_n)_{n \in \omega}$ where $\calF_n = \calF$ for all $n$. 
-In this case, $(Y_n)_{n \in \omega}$ either tends to $4$ as $n \to \infty$, or drops to 0 at some point, with an exponentially decreasing probability. 
-With respect to the antitone function $\epsilon(x) = \frac{4-x}{4}$ and stopping time $T$, we have that $(Y_n)_{n \in \omega}$ is an antitone strict supermartingale, except that $\epsilon(4) = 0$, and $T = \infty$ with probability $\frac 1 2$, even though $4$ is larger than any value $(Y_n)_{n \in \omega}$ can actually take, and $\epsilon(Y_n)$ never actually reaches $0$.
-
-%\lo{@Andrew: I refer to the preceding highlighted claim. Did you mean: for all $n, s$, $\epsilon(Y_n(s)) > Y_n(s)$?} \akr{No. This phrasing should be clearer.}
-\end{example}
-
-\begin{theorem}[Antitone ranking function soundness] \label{thm:antitone rankable implies termination}
-If a closed SPCF term $M$ is antitone rankable, then $T_M^{\tY} < \infty$ a.s.~(equivalently, $M$ is AST).
-\end{theorem}
-\begin{proof}
-As in %Theorem \ref{thm:basicRankingSoundness}, 
-\Cref{thm:rankable implies termination},
-take the set $S$ of traces as a probability space, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$.
-As before, $(Y_n)_{n \in \omega}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$.
-Therefore, it follows from \Cref{thm:a-c strict} that $T_M^{\tY}$ is almost-surely finite.
-\end{proof}
-
-\paragraph{}
-As before, constructing antitone ranking fuctions completely is not necessary, and there is a corresponding notion of an antitone partial ranking function.
-
-Define an \emph{antitone partial ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$
-\begin{itemize}
-\item $f(N) \geq 0$ for all $N$ where $f$ is defined.
-\item $f$ is defined at $M$.
-\item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
-\end{itemize}
-
-\begin{theorem}  \label{thm:antitone partial implies rankable}
-  Every antitone partial ranking function is a restriction of an antitone ranking function.
-\end{theorem}
-\begin{proof}
-  Take a closed term $M$ and an antitone partial ranking function $f$ on $M$, with a corresponding antitone function $\epsilon$.
-Assume wlog that $\epsilon$ is convex (as if it isn't, we can just take its convex hull instead).
-As in Theorem \ref{thm:partial implies rankable}, define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
-  \begin{align*}
-    f_1(N) &= f(N) \text{ whenever $f(N)$ is defined},\\
-    f_1(V) &= 0 \text{ for values $V$ not in the domain of $f$.}
-  \end{align*}
-
-  Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) = \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
-  The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on antitone partial ranking functions.
-Define $f_2(N) = \int_S f_1(\nnext(N,s)) + \epsilon(f_1(\nnext(N,s))) g(N,s) \, \mu(\mathrm d s)$.
-For any term $N$ where $f$ is defined, $f_2(N) = f(N)$, and the value that $f_2$ would have at $N$ if $f$ were not defined at $N$ is $\leq f(N)$, by the third condition on antitone partial ranking functions.
-In order to show that $f_2$ is an antitone partial ranking function, it therefore suffices to show that the value that $f_2$ would have had at each term if $f$ were not defined at that term is at least the expectation of $f_2$ after one reduction step (plus $\epsilon(f_2(N))$ if the reduction step is a $tY$-reduction).
-For any term $N$ which is not of the form $E[R]$ for some $\tY$-redex $R$, this is trivial.
-If $R$ is a $\tY$-redex, then $\epsilon(f_2(N)) \leq \int_S \epsilon(f_1(\nnext(N,s)))\, \mu(\mathrm d s)$ by the convexity of $\epsilon$, because $n$ is bounded.
-Therefore, the (total) function $f_2$, which agrees with $f$ on $f$'s domain, is an antitone ranking function on $M$ (with the same function $\epsilon$ if it's convex).
-\end{proof}
-
-As a corollary, any term which admits an antitone partial ranking function terminates almost surely.
-
-\paragraph{\Cref{ex:ac-ranking} revisited}
-\akr{All of these ranking functions are slightly incorrect, because $\Xi \, \underline n$ reduces to something like $\Xi \, (\underline n + 1)$ then $(\lambda z.
-\text{something}[\Xi/f]) (\underline n + 1)$, rather than reducing the argument first.
-This could be fixed by appealing to the confluent semantics version of the antitone ranking theorem, but it would be nice to have at least some examples here illustrating how antitone ranking functions work, rather than actually saving all these examples until the end.
-
-Even though the difference is quite trivial, this was actually the original motivating example for the confluent trace semantics.}
-
-Programs $M_1$ and $M_2$ are AST. For $M_1$, the partial ranking function 
-\[
-\big(\tY \, \lambda f n . \, \tif{n=0}{0}{f \, (n-1) \oplus_{3/2} f \, (n+1)}\big) \, \underline x \mapsto 3x + 1
-\] 
-suffices to prove its termination (and could equivalently be considered an antitone partial ranking function with the constant antitone function $\epsilon_1(x) = 1$).
-
-For $M_2$, define the function $g_1 : \nnReal \to \nnReal$ by $g_1(x) = \log(x+1) + 1$.
-Using shorthand 
-$\Theta_2 = \tY \, \lambda f n . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
-we can define an antitone partial ranking function $f_2 : \mathit{Rch}(M_2) \rightharpoonup \nnReal$ 
-%with antitone function $\epsilon_2$ defined as follows (for $n \in \mathbb N$):
-defined as: for $n \in \mathbb N$
-\begin{align*}
-{\Theta_2} \, n 
-&\mapsto 
-g_2(n)
-\\
-0 &\mapsto 0.
-\end{align*}
-For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$, each with probability $1/2$, with one $\tY$-reduction.
-Now
-\begin{align*}
-  & g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 \\
-  = & \log \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) \\
-  = & \frac 1 2 \log\left(\frac{(n+1)^2}{n(n+2)}\right) \\
-  = & \frac 1 2 \log\left(1 + \frac 1 {n(n+2)}\right) \\
-  > & \frac 1 {4n(n+2)} \\
-  > & \frac 1 2 \frac 1 {5(n+1)^2} + \frac 1 2 \frac 1 {5(n+3)^2}
-\end{align*}
-and $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
-Therefore by setting $\epsilon_2(x) = \frac 1 {5(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met.
-It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone partial ranking function, and $M_2$ is AST.
-
-\begin{example}[Continuous random walk]\label{ex:raven complex}
-In SPCF we can construct a function whose argument changes by a random amount at each recursive call:
-\[
-\underbrace{\big
-(\tY \, \lambda f x . \tif{x \leq 0}{0}{f(x - \tsample)} \big)}_{\Theta} \, 10
-\]
-We can construct a partial ranking function $f$ as follows:
-\begin{align*}
-\Theta \, \underline l 
-&\mapsto 
-l + 2
-\\
-\tif{l \leq 0}{0}{\Theta \, (l - \tsample)}
-&\mapsto
-l + 1
-\\
-0 &\mapsto 0.
-\end{align*}
-
-For a more complex (not $\tY$-PAST) example, consider the following ``continuous random walk''
-\[
-\underbrace{\big
-(\tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample + 1/2)} \big)}_{\Xi} \, 10
-\]
-Let $g(x) := 2 + \log(x + 1)$, and let $\epsilon$ be the function specified by
-\[
-\epsilon(g(x-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y.
-\]
-The limit of this as $x \to \infty$ and $g(x+1/2) \to \infty$ is 0, and $\frac {\mathrm d}{\mathrm dx} \epsilon(g(x+1/2)) = \frac 1 {x+1} + \log(1 - \frac 1 {x + 1/2}) < 0$, and $g$ is monotonic increasing;
-therefore $\epsilon$ is antitone and bounded below by 0.
-We define a partial antitone ranking function as follows:
-\begin{align*}
-\Xi \, l 
-&\mapsto 
-g(l)
-\\
-0 &\mapsto 0
-\end{align*}
-\end{example}
-The value of $g$ after one $\tY$-reduction step is at least $g(l-1/2)$, therefore the expectation of $\epsilon$ after one $\tY$-reduction step is at most $\epsilon(g(l-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y$.
-Thus 
-\begin{itemize}
-\item in case $l > 0$, $g$ decreases by the required amount
-\item in case $l \leq 0$, 
-\(
-g(\Xi \, l) \geq 2 + \log(1/2) > \log(\frac{3 \sqrt 3} 2) - 1 = \epsilon(g(0-1/2))
-\) 
-as well.
-\end{itemize}
-Hence this is a valid partial antitone ranking function and the term is AST.
-
-\begin{example}[Fair-in-the-limit random walk]
-\label{ex:Fair-in-the-limit random walk}\citep[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}
-\[
-\underbrace{\big
-(\tY \, f \, x . 
-\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)}_{\Xi} 
-\, 10
-\]
-To construct an antitone ranking function, we solve the recurrence relation:
-\[
-\left\{
-\begin{array}{rll}
-z_0 &=& 0\\
-n \geq 1, \quad z_n &>& \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
-\end{array}
-\right.
-\]
-For $n > 2$, $z_n = \log(n-1)$ works.
-The expected decrease is $\frac 1 2(\log(1+\frac 1 {(n-1)^2-1}) - \frac 1 {2n+1} \log(1 + \frac 2 {n-2}))$, and using the fact that $\frac x {x+1} \leq \log (1+x) \leq x$ for $x > 0$, this is at least $\frac 1 2(\frac 1 {(n-1)^2} - \frac 1 {2n + 1} \frac 2 {n-2}) = \frac {n^2}{2(n-1)^2(2n+1)(n-2)}$, which (again for $n > 2$) is positive and antitone.
-For $n = 0, 1, 2$, we take $\epsilon$ to be 9/40 (the same as its value at 3), then set $z_2 = 2 \log 2 - \log 3 - 1, z_1 = 3 \log 2 - 2 \log 3 - 3, z_0 = 4 \log 2 - 3 \log 3 - 6$.
-Some of those values are negative, but it's still bounded below, so by adding a constant offset it can be corrected, and the term is AST.
-
-\end{example}
-
-\begin{example}[Escaping spline]
-\label{ex:escaping spline}\citep[\S 5.4]{DBLP:journals/pacmpl/McIverMKK18}
-\[
-\underbrace{\big
-(\tY \, f \, x . \,
-0 \oplus_{\frac{1}{x+1}} f(x + 1) \big)}_{\Xi} 
-\, 10
-\]
-
-In this case, the fact that the ranking function must decrease at each $\tY$-step (in an antitone partial ranking function) by the expected value of $\epsilon$ not at the current term, but at the next term where the ranking function is defined, is a little harder to deal with, because the variable $x$ can change all the way to $0$ in one step, therefore simply adding a small offset doesn't suffice to compensate for this fact.
-
-Consider the candidate ranking function $\Xi \, n \mapsto n + 1$.
-For each $n$, $\Xi \, n$ reduces to either $0$ (with probability $\frac 1 {n + 1}$) or $\Xi \, (n + 1)$, therefore the expected value of the ranking function is $\frac{n(n+2)}{n+1} = n + 1 - \frac 1 {(n+1)^2}$, and the required decrease is $\frac{\epsilon(0) + n \epsilon(n + 2)}{n + 1}$, therefore eventually, the expected decrease isn't enough, whatever the value of $\epsilon(0)$ is.
-
-If instead, we take the ranking function to be defined after the $\tY$-reduction but before the $\tsample$-reduction as well, this can be resolved.
-Letting $\Theta[n] = \underline 0 \oplus_{\frac 1 n} \Xi \, n$:
-\begin{align*}
-\Theta[n] &\mapsto n \\
-\Xi \, 10 & \mapsto 12.
-\end{align*}
-For each $n$, $\Theta[n]$ reduces to either $0$ or $\Theta[n+1]$, with a $\tY$-reduction only in the latter case, and the condition that this is an antitone partial ranking function is that $n \geq \frac{(n-1)(n+1)}{n} + \frac{n-1}{n} \epsilon(n+1)$ and $12 \geq 11 + \epsilon(11)$.
-These are satisfied by setting $\epsilon(x) = \min(1, 1/x)$, therefore this term is AST.
-\end{example}
-
-\paragraph{Completeness}
-It seems very likely that this method is complete in the case that every reachable term is AST, but we have been unable to actually prove this.
-It is certainly at least capable of proving the termination of terms which terminate arbitrarily slowly (in the sense that there is no similar limitation to Theorem \ref{thm:rankable implies termination}, which can only prove termination of $\tY$-PAST terms).
-
-The following theorem does not prove completeness, but is suggestive in that direction:
-\begin{theorem}
-For any stopping time $T$ which is almost-surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
-\end{theorem}
-
-\input{relativised-completeness}
-
-\input{y-stopping-time}
-
-\section{Confluent Trace Semantics}
+\section{Confluent trace semantics}
+\label{sec:confluent}
+
 When proving almost sure termination in this way, it is necessary to consider which terms a given term may reduce to.
@@
 It is this set that's unmeasurable.}
-
-\bibliographystyle{apalike}
-\bibliography{references}
-
-\lo{22 Jan: 
-\paragraph{Further directions}
-
-An obvious next step is to extend the result to the $\mathsf{score}$ construct.
-
-The following are highly topical, and could form the basis of an interesting and novel DPhil thesis.
-\begin{enumerate}
-\item Devise methods for proving (positive) a.s.~termination.
-
-- For example, develop a type system satisfying the property: if a term is typable then it is (positively) a.s.~terminating.
-See~\citep{DBLP:conf/ppdp/BreuvartL18,DBLP:conf/esop/LagoG17}.
-
-- Another approach is to develop algorithms that synthesise ranking supermartingales, following, for example, \citep{DBLP:journals/pacmpl/AgrawalC018}.
-
-\item Develop principles (e.g.~in the form of ``proof rules'') for reasoning about (positively) a.s.~termination, in the style of \citep{DBLP:journals/pacmpl/McIverMKK18}.
-
-\item Design algorithms that synthesise probabilistic invariants (\emph{qua} martingales), \`a la \cite{SchreuderO19}; see also \citep{DBLP:journals/pacmpl/HarkKGK20}.
-
-\end{enumerate}}
-
-\iffalse
-\begin{thebibliography}{9}
-\bibitem{ppcf} Thomas Ehrhard, Michele Pagani, and Christine Tasson.
-Measurable cones and stable, measurable functions: a model for probabilistic higher-order programming.
-\emph{PACMPL}, 2(POPL):59:1–59:28, 2018.
-doi: 10.1145/3158147.
-URL \href{https://doi.org/10.1145/3158147}{https://doi.org/10.1145/3158147}.
-\end{thebibliography}
-\fi
-
-\end{document}

### 2020-12-26 LICS21/intro.tex
Created version for LICS21 submission in subdirectory LICS21

--- 2020-12-23 probabilisticProgrammingMartingales.tex
@@
-\documentclass{article}
-
-%% BEGIN {Luke's Macros}
-%%
-\newif\ifdraft
-\drafttrue %% To hide all comments and highlighting, just comment this line out
-
-\input{Style/comment}
-\usepackage{calculation}
-\usepackage{graphicx}
-\usepackage{tikz}
-\usetikzlibrary{positioning, graphs, quotes, graphdrawing, shapes.geometric}
-\usegdlibrary{layered, circular}
-\usepackage[square]{natbib}
-\setcitestyle{aysep={}}
-
-\definecolor{oxblue}{RGB}{0,33,71}
-\definecolor{oxgold}{HTML}{a0630a}
-\usepackage[final,
-bookmarks,
-bookmarksopen,
-colorlinks,
-final,
-linkcolor=red,
-citecolor=oxgold,
-pdfstartview=FitH ]{hyperref}
-
-\usepackage{amsmath} % cleveref must be loaded afer amsmath
-\usepackage{cleveref}
-\Crefname{theorem}{Theorem}{Theorems}
-\Crefname{corollary}{Corollary}{Corollary}
-\crefname{proposition}{Proposition}{Propositions}
-\Crefname{claim}{Claim}{Claims}
-\Crefname{definition}{Definition}{Definitions}
-\Crefname{fact}{Fact}{Facts}
-\Crefname{conj}{Conjecture}{Conjectures}
-\Crefname{example}{Example}{Examples}
-\Crefname{remark}{Remark}{Remarks}
-\Crefname{convention}{Convention}{Conventions}
-\Crefname{lemma}{Lemma}{Lemmas}
-\Crefname{assumption}{Assumption}{Assumptions}
-\Crefname{section}{Section}{Sections}
-\Crefname{appendix}{Appendix}{Appendices}
-\Crefname{figure}{Figure}{Figures}
-%\Crefname{equation}{Eq.}{Equations}
-
-% \addtotheorempostheadhook[theorem]{\crefalias{thmlisti}{theorem}}
-% \addtotheorempostheadhook[lemma]{\crefalias{thmlisti}{lemma}}
-% \addtotheorempostheadhook[proposition]{\crefalias{thmlisti}{proposition}}
-% \addtotheorempostheadhook[corollary]{\crefalias{thmlisti}{corollary}}
-% \addtotheorempostheadhook[claim]{\crefalias{thmlisti}{claim}}
-% \addtotheorempostheadhook[fact]{\crefalias{thmlisti}{fact}}
-% \addtotheorempostheadhook[example]{\crefalias{thmlisti}{example}}
-% \addtotheorempostheadhook[definition]{\crefalias{thmlisti}{definition}}
-% \addtotheorempostheadhook[remark]{\crefalias{thmlisti}{remark}}
-% \addtotheorempostheadhook[assumption]{\crefalias{thmlisti}{assumption}}
-
-\newcommand{\Real}{\mathbb{R}}
-\newcommand{\nnReal}{\mathbb{R}_{{\geq}0}}
-\newcommand{\pReal}{\mathbb{R}_{{>}0}}
-
-\newcommand\expect[1]{\mathbb{E}[#1]}
-\newcommand\set[1]{\{#1\}}
-\newcommand\dif{\mathrm{d}}
-\newcommand\calF{\mathcal{F}}
-\newcommand\calI{\mathcal{I}}
-\newcommand\Leb{\mathrm{Leb}}
-\newcommand\ndraw[2]{\#\mathrm{draw}_{#1}(#2)}
-
-\makeatletter
-\newcommand{\dotDelta}{{\vphantom{\Delta}\mathpalette\d@tD@lta\relax}}
-\newcommand{\d@tD@lta}[2]{%
-  \ooalign{\hidewidth$\m@th#1\mkern-1mu\cdot$\hidewidth\cr$\m@th#1\Delta$\cr}%
-}
-
-%\usepackage[bbgreekl]{mathbbol} %for \bblambda
-%%
-%% END {Luke's Macros}
-
-\usepackage{amsfonts}
-\usepackage{amsthm}
-\usepackage{amssymb}
-\usepackage{verbatim}
-\usepackage{enumitem}
-\newcommand{\tY}{\mathsf{Y}}
-%\newcommand{\tif}[3]{\textsf{if }#1\textsf{ then }#2\textsf{ else }#3}
-\newcommand{\tif}[3]{\mathsf{if}(#1, #2, #3)} % space-saving concrete syntax
-\newcommand{\tsample}{\mathsf{sample}}
-\newcommand{\tscore}{\mathsf{score}}
-\newcommand{\skeletonPlaceholder}{\mathsf{X}} % I'm not sure what the best notation here is.
-\DeclareMathOperator{\red}{red}
-\DeclareMathOperator{\nnext}{next}
-\DeclareMathOperator{\cbv}{cbv}
-
-\usepackage{amsthm}
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}
-\newtheorem{example}{Example}
-
-\theoremstyle{lemma}
-\newtheorem{lemma}{Lemma}
-\newtheorem{theorem}{Theorem}
-\newtheorem{corollary}{Corollary}
-\newtheorem{conjecture}{Conjecture}
-
-\theoremstyle{remark}
-\newtheorem*{remark}{Remark}
-
-\title{Supermartingales and Termination Analysis of Statistical PCF}
-
-\begin{document}
-
-\maketitle
-
-\lo{Tentative title}
-
-\begin{abstract}
-\lo{TODO}
-\end{abstract}
-
-\iffalse
-\akr{Following your suggestion, I will be providing a criterion for termination of programs in PPCF \citep{DBLP:journals/pacmpl/EhrhardPT18} based on ranking supermartingales. 
-As it's more convenient for this proof, a sampling-based semantics \lo{You need to provide reference(s) for sampling-based semantics.} will be used instead of the original distributional semantics. 
-I assume some roughly applicable equivalence is proven somewhere, but it doesn't seem that hard anyway.}
-\lo{Why not provide a proof as an appendix?}
-\fi
-
-\tableofcontents
+\section{Introduction}
+\label{sec:intro}
 
@@
 This paper's contribution is to extend this method to a higher-order setting.
-
-\section{Statistical PCF}
-
-\subsection{Syntax of SPCF}
-
-The language SPCF is a simply-typed lambda calculus with sampling of real numbers from $[0,1]$ and unbounded scoring, following \cite{MakOP20b}.
-Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
-\begin{align*}
-  & \text{types } A, B ::= \textsf{R}  \mid  A \to B \\
-  & \text{values } V ::= \lambda x.M  \mid  \underline{r} \\
-  & \text{terms } M, N ::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY M  \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample  \mid  \tscore(M)
-\end{align*}
-\begin{align*}
-  \frac{}{\Gamma ; x:A \vdash x:A} \qquad
-  \frac{\Gamma ; x:A \vdash M : B}{\Gamma \vdash \lambda x.M : A \to B} \qquad
-  \frac{}{\underline{r} : \textsf{R}} \qquad
-  \frac{\Gamma \vdash M:A \to B \quad \Gamma \vdash N : A}{\Gamma \vdash M \, N : B} \\ \\
-  \frac{\Gamma \vdash M_1:\textsf{R} \dots \Gamma \vdash M_n:\textsf{R}}{\Gamma \vdash \underline{f}(M_1,\dots,M_n) : \textsf{R}} \ f : \mathbb{R}^n \to \mathbb{R} \qquad
-  \frac{\Gamma \vdash M : (A \to B) \to (A \to B)}{\Gamma \vdash \tY M : (A \to B)} \\ \\
-  \frac{\Gamma \vdash M : \textsf{R} \quad \Gamma \vdash N_1 : A \quad \Gamma \vdash N_2 : A}{\Gamma \vdash \tif{M < 0}{N_1}{N_2} : A} \qquad
-  \frac{}{\Gamma \vdash \tsample : \textsf{R}} \qquad
-  \frac{\Gamma \vdash s : \textsf{R}}{\Gamma \vdash \tscore (s) : \textsf{R}}
-\end{align*}
-
-Terms are identified up to $\alpha$-equivalence, as usual. The set of all terms is denoted $\Lambda$, and the set of closed terms is denoted $\Lambda^0$.
-
-\paragraph{}
-To define the reduction relation, let \emph{evaluation contexts} be of the form:
-\begin{align*}
-  E ::= & \, [\,] \mid E \, M \mid V \, E \mid \underline{f}(\underline r_1,\dots ,\underline r_{k-1}, E, M_{k+1}, \dots, M_n) \\ & \mid \tY E \mid \tif{E<0}{M_1}{M_2} \mid \tscore (E)
-\end{align*}
-then a term reduces if it is formed by substituting a redex in a context i.e.
-\begin{align*}
-  E[(\lambda x.M) N] & \to E[M[N/x]] \\
-  E[\underline f (\underline r_1, \dots , \underline r_n)] & \to E[\underline{f(r_1,\dots,r_n)}] \\
-  E[\tY \lambda x. M] & \to E[\lambda z. M[(\tY \lambda x. M)/x] z] \text{ where $z$ is not free in $s$}\\
-  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_1] \text{ where }r < 0 \\
-  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_2] \text{ where }r \geq 0 \\
-  E[\tsample] & \to E[\underline r] \text{ where } r \in [0,1] \\
-  E[\tscore(\underline r)] & \to E[\underline r].
-\end{align*}
-\changed[lo]{We write $\to^\ast$ for the reflexive, transitive closure of $\to$.}
-\lo{The $\tscore$ rule should have a side condition: if $r > 0$.}
-
-Every closed well-typed term either is a value or reduces to another term.
-
-\subsection{Sampling semantics}
-\label{sec:sampling semantics}
-This version of the reduction relation allows $\tsample$ to reduce to any number in $[0,1]$.
-To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples.
-Let $ I = [0,1] \subset \mathbb{R} $, and let $S = I^{\mathbb{N}}$, with the \changed[lo]{Borel $\sigma$-algebra $\calF$} and the probability measure, denoted $\mu$, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
-Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and $\mu (\prod_{i=0}^\infty X_i) = \prod_{i=0}^\infty \Leb(X_i)$.
-The maps $\pi_h:S \to I, \; \pi_t:S \to S$ popping the first element are then measurable.
-\changed[lo]{Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(S, \calF, \mu)$ the \emph{entropy space}.}
-
-The $\sigma$-algebra and measure on $\Lambda$ are defined by considering it as a disjoint union of equivalence classes under replacing all the real constants by a placeholder, where the measure on each class is that of $\mathbb{R}^n$, where $n$ is the number of real constants.
-\changed[lo]{Precisely, following \cite{DBLP:conf/icfp/BorgstromLGS16},
-we view $\Lambda$ as $\bigcup_{m\in\omega} \big(\mathsf{Sk}_m \times \Real^m \big)$,
-where $\mathsf{Sk}_m$ is the set of SPCF terms with exactly $m$ place-holders (a.k.a.~\emph{skeleton terms}) for numerals.
-Thus identified, we give $\Lambda$ the countable disjoint union topology of the product topology of the discrete topology on $\mathsf{Sk}_m$ and the standard topology on $\Real^m$.
-Note that the connected components of $\Lambda$ have the form $\{M\} \times \Real^m$, with $M$ ranging over $\mathsf{Sk}_m$, and $m$ over $\omega$. 
-We fix the Borel algebra of this topology to be the $\sigma$-algebra on $\Lambda$.}
-
-The one-step reduction is given by the function $\red : \Lambda^0 \times S \to \Lambda^0 \times S$ where
-\begin{equation}
-\red(M,s) = \left\{
-    \begin{array}{ll}
-        (E[N],s) & \text{if } M = E[R], R \to N \text{ and } R \neq \tsample \\
-        (E[\underline{\pi_h(s)}],\pi_t(s)) & \text{if } M = E[\tsample] \\
-        (M,s) & \text{if } M \text{ is a value}
-    \end{array} \right .
-\end{equation}
-\lo{Both $\to$ and $\red$ are called reduction relation.}
-
-The result after $n$ steps is then simply $\red^n(M,s) = \overbrace{\red(...\red(}^n M,s)...)$, and the limit $\red^\infty$ can then be defined as a partial function as $\lim_{n \to \infty} \red^n(M,s)$ whenever that sequence becomes constant by reaching a value.
-A term $M$ terminates for a sample sequence $s$ if the limit $\red^\infty(M,s)$ is defined.
-
-The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid M \text{ terminates at } s \}$ is measurable.
-A term $M$ is said to terminate almost surely if $\mu(\{s \mid M \text{ terminates at } s\}) = 1$.
-
-\lo{Alternatively, define the \emph{runtime of $M$} to be the random variable 
-\[
-T_M(s) := 
-\begin{cases}
-\min \set{n \mid \pi_0(\red^n(M, s)) \textrm{ is a value}} & \hbox{if $\red^\infty(M,s)$ is defined}\\
-\infty & \hbox{otherwise}
-\end{cases}
-\]
-Equivalently, we say that $M$ is \emph{almost-surely terminating} (AST) if $T_M < \infty$ a.s.; 
-and $M$ is \emph{positively almost-surely terminating} (PAST) if $\expect{T_M} < \infty$.}
-
-For example, the term 
-\[
-(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0},
-\] 
-which generates a geometric distribution, terminates on the set $S \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas 
-\[
-\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \underline 0},
-\] 
-which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.
-
-\paragraph{}
-Note that the requirement that $M$ be closed is not important for the proofs of the theorems, but is simply required to ensure that $M$ reaches a value, rather than a general normal form like $x \, y$.
-
-\paragraph{}
-This definition of almost-sure termination is equivalent to that given in \citep{MakOP20b}, although the program semantics is stated in a slightly different way.
-In particular, the argument to $\tscore$ is not relevant to termination (except for the possibility that its argument's evaluation wouldn't terminate).
-
-\iffalse
-\lo{Your operational semantics does not maintain a record of the current weight of the reduction.
-A.s.~termination does depend on $\tscore$: see \cite[\S 4.3]{DBLP:journals/corr/abs-2004-03924}\footnote{\url{https://arxiv.org/abs/2004.03924}}.
-I think it important to take the behaviour of $\tscore$ into account;
-you should do it as a future task.}
-\fi
-
-\section{Supermartingales}
-One approach to proving that a term terminates almost surely is to find some variant that is bounded below and, on average, decreases sufficiently quickly that it must eventually reach 0, similarly to the approach taken in \citep{DBLP:journals/pacmpl/McIverMKK18} for imperative programs.
-
-These variants are defined as functions from reachable terms (i.e.~possible states of the program's execution) to real numbers.
-Specifically, let the set of reachable terms from a given closed starting term $M$ be $Rch(M) := \{N \in \Lambda \mid M \to^* N \}$, with the $\sigma$-algebra induced as a subset of $\Lambda$.
-
-\begin{definition}\rm
-A \emph{ranking function on $M$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
-\begin{enumerate}
-    \item $f(E[\tY \lambda x. N]) \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z) \text{ where $z$ is not free in $N$}$
-    \item $f(E[\tsample]) \geq \int_I f(E[\underline{x}]) \, \Leb(\mathrm{d}x)$
-
-    \item $f(E[R]) \geq f(E[R'])$ for any other redex $R$, where $R \to R'$.
-\end{enumerate}
-We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that $f(N) = 0$ iff $N$ is a value, and for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.
-
-Any closed term for which a ranking (respectively, strict ranking) function exists is called \emph{rankable} (respectively, \emph{strictly rankable}). 
-\changed[lo]{For example, the term $(\tY \lambda x.x) \, \underline 0$ is not rankable.}
-It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \in \omega}$ is the reduction sequence starting from $M$, then $(f(M_n))_{n\in\omega}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales.
-\end{definition}
-
-\input{ranking-supermartingales}
-
-\section{Constructing Ranking Functions}
-Although rankability implies almost-sure termination, the converse does not hold in general. For example,
-\begin{equation}
-\tif{{-}\tsample < 0}{\underline{0}}{(\tY \lambda x. x) \, \underline 0}
-\label{ex:0 probability reachable}
-\end{equation}
-terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x. x) \, \underline 0$ is reachable, although that has probability 0. 
-Not only is this counterexample AST, it's PAST. %positively almost surely terminating i.e.~the expected time to termination is finite.
-
-A ranking function can be constructed under the stronger assumptions that, for every $N$ reachable from $M$, the expected number of $\tY$-reduction steps from $N$ to a value is finite.
-In particular, the expected number of $\tY$-reduction steps from each reachable term is a ranking function.
-Note that a finite number of expected $\tY$-reduction steps does not necessarily imply a finite number of expected total reduction steps.
-
-\begin{example}
-\label{ex:tY finite does not imply t finite}
-The term
-\[
-M = \big(\tY \lambda f n. \tif{\tsample - 0.5 < 0}{n \, \underline 0}{f (\lambda x. n (n x))}\big) (\lambda x. x+1)
-\] 
-terminates with only 2 $\tY$-reductions on average i.e.~$\expect{T_M^\tY} = 2$, but applies the increment function $2^n$ times with probability $2^{-1-n}$ for $n \geq 0$, which diverges i.e.~$\expect{T_M} = \infty$.
-\end{example}
-
-\begin{theorem} \label{thm:minimal}
-Given a closed term $M$, the function $f:Rch(M) \to \mathbb R$ given by $f(N) = \mathbb E [\text{the number of }\tY\text{-reduction steps from }N\text{ to a value}]$, if it exists, is the least of all possible ranking functions of $M$.
-\end{theorem}
-\begin{proof}
-Let $f$ be the candidate least ranking function defined above, and suppose $g$ is another ranking function such that $f(N) > g(N)$ for some $N \in Rch(M)$.
-The restrictions of $f$ and $g$ to $Rch(N)$ have the same properties assumed of $f$ and $g$, so assume w.l.o.g.~that $N=M$.
-The difference $g - f$ is then a supermartingale (with the same setup as in the proof of %\Cref{thm:rankable gives supermartingale}
-%Theorem \ref{rankable implies ast}) 
-\Cref{thm:rankable and strict rankable});
-therefore $\mathbb E[g(M_n)] \leq \mathbb E[f(M_n)] + g(M)-f(M)$, for all $n$.
-
-We have $\mathbb E[f(M_n)] = \sum_{k=n}^\infty \mathbb P[M_k = E[\tY N] \text{ for some }E, N] \to 0 \text{ as } n \to \infty$; 
-therefore as $g(M) - f(M) < 0$, eventually $\mathbb E[g(M_n)] < 0$, which is impossible. 
-It follows that $g \geq f$ as required.
-
-In order for $f$ to be the least ranking function of $M$, it also has to actually be a ranking function itself.
-Each of the conditions on a ranking function is easily verified from the definition of $f$.
-\end{proof}
-
-\subsection{Partial ranking functions}
-Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost-sure termination, so it isn't useful for proving it.
-
-\begin{example}[Geometric distribution]
-Consider the term
-\begin{align*}
-&(\tY \lambda f, n. \\
-&\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
-&) \, \underline{0}
-\end{align*}
-which generates a geometric distribution (let $\Theta = \lambda f, n : \tif{\tsample \underline{- 0.5} < 0}{n}{f (n\underline{+ 1})}$, then the term is $(\tY \Theta)\ \underline 0$).
-    Despite its simplicity, its $Rch$ contains all the terms
-
-\begin{tikzpicture}[node distance=0.5em, anchor=west]
-    \node (1) at (0,-0) {$\tY\ \Theta\ \underline 0$};
-    \node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\ \Theta\ (n \underline{+ 1})}) z)\ \underline i$};
-    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\ \Theta\ (n \underline{+ 1})})\ \underline i$};
-    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY\ \Theta\ (\underline i\ \underline{+ 1})}$};
-    \node (5) at (0,-4) {$\tif{\underline r\ \underline{- 0.5} < 0}{\underline i}{\tY\ \Theta\ (\underline i\ \underline{+ 1})}$};
-    \node (6) at (0,-5) {$\tif{\underline r < 0}{\underline i}{\tY\ \Theta\ (\underline i\ \underline{+ 1})}$};
-    \node (7) at (2,-6) [text width=] {$\underline i$};
-    \node (8) at (0,-7) {$\tY\ \Theta\ (\underline i\ \underline{+ 1})$};
-    \node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\ \Theta\ (n \underline{+ 1})}) z)\ (\underline i\ \underline{+ 1})$};
-
-    \draw [->] (1.south west) to [bend right=45] (2.north west);
-    \draw [->] (2.south west) to [bend right=45] (3.north west);
-    \draw [->] (3.south west) to [bend right=45] (4.north west);
-    \draw [->] (4.south west) to [bend right=45] (5.north west);
-    \draw [->] (5.south west) to [bend right=45] (6.north west);
-    \draw [->] (6.south west) to [bend right=45] (8.north west);
-    \draw [->] (8.south west) to [bend right=45] (9.north west);
-
-    \draw [->] (9.west) to [bend left=45] (2.west);
-    \draw [->] (6) to (7);
-\end{tikzpicture}
-
-\end{example}
-
-Even in this simple case, defining a ranking function explicitly is awkward because of the number of cases, although in most cases, because the value need only be greater than or equal to that of the next term in sequence, it suffices to take the ranking function as having the same value as the next term, so that overall it takes only 3 distinct values.
-(We will explain why later.)
-
-The definition of rankability is also inconvenient for syntactic sugar.
-It could be useful, for example, to define $M \oplus_p N = \tif{\tsample - p < 0} M N$, where $M \oplus_p N$ reduces to $M$ or $N$, depending on the first value of $s$\lo{Need a name for $s$.}, with probability $p$ resp.~$(1-p)$.
-Technically though, it reduces first to $\tif{\underline r - p < 0} M N$ for all $r \in I$, so those terms all need values of the ranking function too.
-
-\paragraph{}
-In both of these cases, there are only some values of the ranking function that are semantically important. 
-\begin{definition}
-Define a \emph{partial ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \rightharpoonup \mathbb R$ such that
-\begin{itemize}
-    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
-    \item $f$ is defined at $M$.
-    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually \akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and ast, but that doesn't actually work because of terms reachable with 0 probability.
-Of course, astness doesn't imply rankability, so any definition of partial ranking function that includes all Y-past terms wouldn't always extend to a total ranking function.} reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
-\end{itemize}
-\end{definition}
-A partial ranking function that is total is just a ranking function.
-Providing a partial ranking function is essentially part way between providing a ranking function and directly proving almost sure termination.
-
-\begin{theorem} \label{thm:partial implies rankable}
-Every partial ranking function is a restriction of a ranking function.
-\end{theorem}
-\begin{proof}
-Take a closed term $M$ and partial ranking function $f$ on $M$. Define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
-\begin{align*}
-    f_1(N) & := f(N) \text{ whenever $f(N)$ is defined,} \\
-    f_1(V) & := 0 \text{ for values $V$ not in the domain of $f$.}
-\end{align*}
-
-Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) := \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
-The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on partial ranking functions.
-Define $f_2(N) = \int_S f_1(\nnext(N,s)) + g(N,s) \, \mu(\mathrm d s)$.
-The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction).
-\lo{TODO. The preceding sentence deserves a proof.}
-\end{proof}
-
-As a corollary, any term which admits a partial ranking function terminates almost surely.
-
-\subsection{Examples}
-Let $M \oplus_p N = \tif{\tsample - p < 0} M N$, for $p \in (0,1]$, then there are the pseudo-reduction relations
-\begin{align*}
-E[M \oplus_p N] \to^3 & E[M] & \\
-E[M \oplus_p N] \to^3 & E[N] & \\
-\red^3(E[M \oplus_p N], s) = & \left\{
-    \begin{array}{ll}
-        (E[M],\pi_t(s)) & \text{if } \pi_h(s) < p \\
-        (E[N],\pi_t(s)) & \text{if } \pi_h(s) \geq p. \\
-    \end{array} \right .
-\end{align*}
-A partial ranking function could be defined with respect to this shortcut reduction simply by replacing $\to$ and $\red$ in the definition of a partial ranking function by a version that goes straight from $N \oplus_p O$ to $N$ or $O$.
-Such a pseudo-partial ranking function would then be a partial function from a subset of $Rch(M)$, so it could also be considered as a partial function from all of $Rch(M)$, and it would in fact also be an actual partial ranking function.
-It is therefore possible to prove rankability directly using the shortcutted reductions.
-
-A similar procedure would work for other forms of syntactic sugar.
-If a closed term $N$ eventually reduces to one of a set of other terms $\{N_i \mid i \in I\}$ with certain probabilities, a partial ranking function defined with respect to a reduction sequence that skips straight from $N$ to $N_i$ is also a valid partial ranking function for the original reduction function, and therefore its existence implies almost sure termination.
-There is a caveat, however, that $\tY$-reduction steps skipped over in the shortcut still need to be counted for the expected number of $\tY$-reduction steps.
-
-\paragraph{}
-With this abbreviation, the geometric distribution example from earlier can be written as $(\tY \lambda f n.\ n \oplus_{0.5} f(n+1)) \, \underline 0$.
-It is then easy to see that the following is a partial ranking function:
-\begin{align*}
-(\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\ N & \mapsto 2 \\
-\underline i \oplus_{0.5} (\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\ (\underline i + 1) & \mapsto 1 \\
-\underline i & \mapsto 0.
-\end{align*}
-
-In fact, even the partial function $\tY \Theta N \mapsto 2$ alone is a partial ranking function for this term.
-\section{Antitone ranking functions}
-
-\begin{example}[Random walk]
-\label{ex:ac-ranking}
-\begin{enumerate}
-\item 1D biased (towards 0) random walk
-\[
-M_1 = 
-\big(\tY \, \lambda f n . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{3/2} f \, (n + 1)}\big) \, 10
-\]
-
-\item 1D unbiased random walk
-\[
-M_2 = 
-\big(\tY \, \lambda f n . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}\big) \, 10
-\]
-
-\item 1D biased (away from 0) random walk
-\[
-M_3 = 
-\big(\tY \, \lambda f n . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/3} f \, (n + 2)}\big) \, 10
-\]
-
-%While {x > 0} do {x := x-1 \oplus_2/3 x:= x+1}
-
-%M2 = While {x > 0} do {x := x-1 \oplus_2/3 x:= x+2}
-\end{enumerate}
-\end{example}
-
-The term $M_1$ is rankable, terminating in 31 $\tY$-steps on average, and $M_3$ only has a $1/1024$ chance of terminating, but in between, $M_2$ is AST, but isn't $\tY$-PAST, therefore it isn't rankable and \Cref{thm:rankable implies termination} is insufficient to prove its termination.
-We therefore want to find a generalised notion of ranking function so that $M_2$ becomes rankable, and then prove it sound i.e.~rankable in this generalised sense implies AST.
-
-\paragraph{Idea.}
-
-The definition of antitone-convex ranking function $f$ (say) for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
-we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) function $\epsilon : \nnReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
-\[
-f(E[R’]) \leq f(E[R]) - \epsilon(f(E[R])) 
-\]
-where $R \to R’$ is the $\tY$-redex rule.
-
-\lo{An aside: This CBV $\tY$-rule seems cleaner:
-\[
-\big(\tY f^{A \to B} \, x^A \, . \, \theta^B \big) \, v \to \theta[(\tY f \, x \, . \, \theta) / f, v / x].
-\]
-We assume $\tY f^{A \to B} \, x^A \, . \, \theta^B$ is a value.}
-
-\begin{definition}
-We say that a supermartingale $(Y_n)_{n \in \omega}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \in \omega}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
-\[
-\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot 1_{\set{T > n}}
-\]
-\end{definition}
-
-%\lo{The preceding definition does not place any constraint on the stopping time $T$ (other than that it is w.r.t.~the filtration $(\calF_n)_{n \in \omega}$).
-However for the definition to make sense, $T$ must be the stopping time $T(s) := \min \set{n \mid Y_n(s) = 0}$.}
-
-\begin{theorem}
-\label{thm:a-c strict}
-Let $(Y_n)_{n \in \omega}$ be an antitone strict supermartingale via function $\epsilon : \nnReal \to \pReal$, and stopping time $T$. 
-Then $T < \infty$ almost surely.
-\end{theorem}
-
-\begin{proof}
-First, as $(Y_n)$ is a supermartingale, $\expect{Y_n} \leq \expect{Y_0}$. 
-Therefore
-
-\begin{calculation}
-\expect{Y_n \mid T > n}
-  \step[=]{rearranging terms}
-\dfrac{\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
-+ 
-\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
-- 
-\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
-  \step[=]{definition of conditional expectation}
-\dfrac{\expect{Y_n} - \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
-  \step[\leq]{$Y_n \geq 0$ always}
-\dfrac{\expect{Y_n}}{\mathbb P[T > n]}
-  \step[\leq]{$(Y_n)_n$ is a supermartingale}
-\dfrac{\expect{Y_0}}{\mathbb P[T > n]}
-\end{calculation}
-
-\emph{Claim}: For all $0 < x \leq 1$, 
-\(
-\mathbb P[T > B_x] \leq x,
-\)
-where $B_x = \left \lceil \dfrac{\expect{Y_0}+1}{x \; \epsilon(\expect{Y_0} \, x^{-1})} \right \rceil$.
-
-\smallskip
-
-As the convex hull of $\epsilon$ (the greatest convex function less than or equal to it) satisfies all the conditions assumed of $\epsilon$, in addition to being convex, assume wlog that $\epsilon$ is convex.
-
-Assume for a contradiction that $\mathbb P[T > B_x] > x$.
-Then, take $n \leq B_x$. 
-We have
-%\lo{Change of justification of the second equality below: event, not discrete r.v. Recall: expectation conditioning on r.v.~(resp.~event) is a r.v.~(resp.~number).}
-\begin{calculation}
-\expect{Y_n-Y_{n+1}}
-  \step[=]{$\calF_n \subseteq \calF_{n+1}$, definition and linearity of conditional expectation}
-\expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
-  \step[\geq]{antitone strict assumption}
-\expect{\epsilon(Y_n) \cdot 1_{\set{T > n}}}
-%  \step[=]{definition of conditional expectation for a discrete variable}
-\step[=]{definition of expectation conditioning on an \emph{event}}
-\mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
-  \step[\geq]{Jensen's inequality}
-\mathbb P[T > n]\ \epsilon(\expect{Y_n \mid T > n})
-  \step[\geq]{proved earlier}
-\mathbb P[T > n]\ \epsilon\Big(\dfrac{\expect{Y_0}}{\mathbb P[T > n]}\Big)
-  \step[>]{assumption, $\mathbb P[T > n] \geq \mathbb P[T > B_x] > x$}
-x \; \epsilon\big(\expect{Y_0} \, x^{-1}\big).
-\end{calculation}
-Therefore, by a telescoping sum
-\[
-\expect{Y_{B_x}} = \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1}) \leq -1 < 0 
-\]
-%\lo{$\expect{Y_0 - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})}$}
-which is a contradiction, therefore the claim must be true, therefore $P[T > n]$ tends to 0 as $n$ tends to infinity, therefore $T < \infty$ almost surely.
-\end{proof}
-
-Note that it is essential in this proof that $\epsilon$ be defined for all $\mathbb R_{\geq 0}$, not just the values that $(Y_n)_{n \in \omega}$ actually takes (or at least if $(Y_n)_{n \in \omega}$ is uniformly bounded, $\epsilon$ must be positive at the supremum as well as the actual values).
-
-\begin{example}
-\akr{Consider a random variable $X$, uniformly distributed in the interval $[-1,1]$, $T := \min \set{n \mid 2^{-n} < X}$, 
-\[
-Y_k := 
-\begin{cases}
-4-2^{-k} & \hbox{if $k < T$}\\
-0 & \hbox{otherwise}
-\end{cases} 
-\]
-} \lo{The following reformulation defines $T$ and $Y_k$ directly as a measurable function on a probability space (rather than via a r.v.~$X$).}
-
-Take the probability space $(\Omega, \calF, \mathbb{P})$ where $\Omega$ is the closed interval of reals $[-1, 1]$, $\calF$ the Borel $\sigma$-algebra, and $\mathbb{P}$ the corresponding Lebesgue probability measure.
-Let $\omega \in [-1, 1]$. Define random variables $T$ and $(Y_k)_{k \in \omega}$:
-\begin{align*}
-T(\omega) & := 
-\begin{cases}
-\min\set{n \in \mathbb{N} \mid \omega > 2^{-n}} & \hbox{if $\omega \in (0, 1]$}\\
-\infty & \hbox{otherwise}
-\end{cases} 
-\\
-Y_k(\omega) & := 
-\begin{cases}
-4-2^{-k} & \hbox{if $\omega \in [-1, 2^{-k}]$ (equivalently $k < T(\omega)$)}\\
-0 & \hbox{otherwise}
-\end{cases} 
-\end{align*}
-\lo{N.B.~$\set{T = n} = (\frac{1}{2^{n}}, \frac{1}{2^{n-1}}]$, and so, $\set{T > n} = [-1, \frac{1}{2^n}]$.}
-Plainly, $T$ is a stopping time, and $(Y_n)_{n \in \omega}$ is a supermartingale, adapted to the filtration $(\calF_n)_{n \in \omega}$ where $\calF_n = \calF$ for all $n$. 
-In this case, $(Y_n)_{n \in \omega}$ either tends to $4$ as $n \to \infty$, or drops to 0 at some point, with an exponentially decreasing probability. 
-With respect to the antitone function $\epsilon(x) = \frac{4-x}{4}$ and stopping time $T$, we have that $(Y_n)_{n \in \omega}$ is an antitone strict supermartingale, except that $\epsilon(4) = 0$, and $T = \infty$ with probability $\frac 1 2$, even though $4$ is larger than any value $(Y_n)_{n \in \omega}$ can actually take, and $\epsilon(Y_n)$ never actually reaches $0$.
-
-%\lo{@Andrew: I refer to the preceding highlighted claim. Did you mean: for all $n, s$, $\epsilon(Y_n(s)) > Y_n(s)$?} \akr{No. This phrasing should be clearer.}
-\end{example}
-
-\begin{theorem}[Antitone ranking function soundness] \label{thm:antitone rankable implies termination}
-If a closed SPCF term $M$ is antitone rankable, then $T_M^{\tY} < \infty$ a.s.~(equivalently, $M$ is AST).
-\end{theorem}
-\begin{proof}
-As in %Theorem \ref{thm:basicRankingSoundness}, 
-\Cref{thm:rankable implies termination},
-take the set $S$ of traces as a probability space, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$.
-As before, $(Y_n)_{n \in \omega}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$.
-Therefore, it follows from \Cref{thm:a-c strict} that $T_M^{\tY}$ is almost-surely finite.
-\end{proof}
-
-\paragraph{}
-As before, constructing antitone ranking fuctions completely is not necessary, and there is a corresponding notion of an antitone partial ranking function.
-
-Define an \emph{antitone partial ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$
-\begin{itemize}
-\item $f(N) \geq 0$ for all $N$ where $f$ is defined.
-\item $f$ is defined at $M$.
-\item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
-\end{itemize}
-
-\begin{theorem}  \label{thm:antitone partial implies rankable}
-  Every antitone partial ranking function is a restriction of an antitone ranking function.
-\end{theorem}
-\begin{proof}
-  Take a closed term $M$ and an antitone partial ranking function $f$ on $M$, with a corresponding antitone function $\epsilon$.
-Assume wlog that $\epsilon$ is convex (as if it isn't, we can just take its convex hull instead).
-As in Theorem \ref{thm:partial implies rankable}, define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
-  \begin{align*}
-    f_1(N) &= f(N) \text{ whenever $f(N)$ is defined},\\
-    f_1(V) &= 0 \text{ for values $V$ not in the domain of $f$.}
-  \end{align*}
-
-  Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) = \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
-  The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on antitone partial ranking functions.
-Define $f_2(N) = \int_S f_1(\nnext(N,s)) + \epsilon(f_1(\nnext(N,s))) g(N,s) \, \mu(\mathrm d s)$.
-For any term $N$ where $f$ is defined, $f_2(N) = f(N)$, and the value that $f_2$ would have at $N$ if $f$ were not defined at $N$ is $\leq f(N)$, by the third condition on antitone partial ranking functions.
-In order to show that $f_2$ is an antitone partial ranking function, it therefore suffices to show that the value that $f_2$ would have had at each term if $f$ were not defined at that term is at least the expectation of $f_2$ after one reduction step (plus $\epsilon(f_2(N))$ if the reduction step is a $tY$-reduction).
-For any term $N$ which is not of the form $E[R]$ for some $\tY$-redex $R$, this is trivial.
-If $R$ is a $\tY$-redex, then $\epsilon(f_2(N)) \leq \int_S \epsilon(f_1(\nnext(N,s)))\, \mu(\mathrm d s)$ by the convexity of $\epsilon$, because $n$ is bounded.
-Therefore, the (total) function $f_2$, which agrees with $f$ on $f$'s domain, is an antitone ranking function on $M$ (with the same function $\epsilon$ if it's convex).
-\end{proof}
-
-As a corollary, any term which admits an antitone partial ranking function terminates almost surely.
-
-\paragraph{\Cref{ex:ac-ranking} revisited}
-\akr{All of these ranking functions are slightly incorrect, because $\Xi \, \underline n$ reduces to something like $\Xi \, (\underline n + 1)$ then $(\lambda z.
-\text{something}[\Xi/f]) (\underline n + 1)$, rather than reducing the argument first.
-This could be fixed by appealing to the confluent semantics version of the antitone ranking theorem, but it would be nice to have at least some examples here illustrating how antitone ranking functions work, rather than actually saving all these examples until the end.
-
-Even though the difference is quite trivial, this was actually the original motivating example for the confluent trace semantics.}
-
-Programs $M_1$ and $M_2$ are AST. For $M_1$, the partial ranking function 
-\[
-\big(\tY \, \lambda f n . \, \tif{n=0}{0}{f \, (n-1) \oplus_{3/2} f \, (n+1)}\big) \, \underline x \mapsto 3x + 1
-\] 
-suffices to prove its termination (and could equivalently be considered an antitone partial ranking function with the constant antitone function $\epsilon_1(x) = 1$).
-
-For $M_2$, define the function $g_1 : \nnReal \to \nnReal$ by $g_1(x) = \log(x+1) + 1$.
-Using shorthand 
-$\Theta_2 = \tY \, \lambda f n . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
-we can define an antitone partial ranking function $f_2 : \mathit{Rch}(M_2) \rightharpoonup \nnReal$ 
-%with antitone function $\epsilon_2$ defined as follows (for $n \in \mathbb N$):
-defined as: for $n \in \mathbb N$
-\begin{align*}
-{\Theta_2} \, n 
-&\mapsto 
-g_2(n)
-\\
-0 &\mapsto 0.
-\end{align*}
-For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$, each with probability $1/2$, with one $\tY$-reduction.
-Now
-\begin{align*}
-  & g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 \\
-  = & \log \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) \\
-  = & \frac 1 2 \log\left(\frac{(n+1)^2}{n(n+2)}\right) \\
-  = & \frac 1 2 \log\left(1 + \frac 1 {n(n+2)}\right) \\
-  > & \frac 1 {4n(n+2)} \\
-  > & \frac 1 2 \frac 1 {5(n+1)^2} + \frac 1 2 \frac 1 {5(n+3)^2}
-\end{align*}
-and $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
-Therefore by setting $\epsilon_2(x) = \frac 1 {5(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met.
-It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone partial ranking function, and $M_2$ is AST.
-
-\begin{example}[Continuous random walk]\label{ex:raven complex}
-In SPCF we can construct a function whose argument changes by a random amount at each recursive call:
-\[
-\underbrace{\big
-(\tY \, \lambda f x . \tif{x \leq 0}{0}{f(x - \tsample)} \big)}_{\Theta} \, 10
-\]
-We can construct a partial ranking function $f$ as follows:
-\begin{align*}
-\Theta \, \underline l 
-&\mapsto 
-l + 2
-\\
-\tif{l \leq 0}{0}{\Theta \, (l - \tsample)}
-&\mapsto
-l + 1
-\\
-0 &\mapsto 0.
-\end{align*}
-
-For a more complex (not $\tY$-PAST) example, consider the following ``continuous random walk''
-\[
-\underbrace{\big
-(\tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample + 1/2)} \big)}_{\Xi} \, 10
-\]
-Let $g(x) := 2 + \log(x + 1)$, and let $\epsilon$ be the function specified by
-\[
-\epsilon(g(x-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y.
-\]
-The limit of this as $x \to \infty$ and $g(x+1/2) \to \infty$ is 0, and $\frac {\mathrm d}{\mathrm dx} \epsilon(g(x+1/2)) = \frac 1 {x+1} + \log(1 - \frac 1 {x + 1/2}) < 0$, and $g$ is monotonic increasing;
-therefore $\epsilon$ is antitone and bounded below by 0.
-We define a partial antitone ranking function as follows:
-\begin{align*}
-\Xi \, l 
-&\mapsto 
-g(l)
-\\
-0 &\mapsto 0
-\end{align*}
-\end{example}
-The value of $g$ after one $\tY$-reduction step is at least $g(l-1/2)$, therefore the expectation of $\epsilon$ after one $\tY$-reduction step is at most $\epsilon(g(l-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y$.
-Thus 
-\begin{itemize}
-\item in case $l > 0$, $g$ decreases by the required amount
-\item in case $l \leq 0$, 
-\(
-g(\Xi \, l) \geq 2 + \log(1/2) > \log(\frac{3 \sqrt 3} 2) - 1 = \epsilon(g(0-1/2))
-\) 
-as well.
-\end{itemize}
-Hence this is a valid partial antitone ranking function and the term is AST.
-
-\begin{example}[Fair-in-the-limit random walk]
-\label{ex:Fair-in-the-limit random walk}\citep[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}
-\[
-\underbrace{\big
-(\tY \, f \, x . 
-\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)}_{\Xi} 
-\, 10
-\]
-To construct an antitone ranking function, we solve the recurrence relation:
-\[
-\left\{
-\begin{array}{rll}
-z_0 &=& 0\\
-n \geq 1, \quad z_n &>& \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
-\end{array}
-\right.
-\]
-For $n > 2$, $z_n = \log(n-1)$ works.
-The expected decrease is $\frac 1 2(\log(1+\frac 1 {(n-1)^2-1}) - \frac 1 {2n+1} \log(1 + \frac 2 {n-2}))$, and using the fact that $\frac x {x+1} \leq \log (1+x) \leq x$ for $x > 0$, this is at least $\frac 1 2(\frac 1 {(n-1)^2} - \frac 1 {2n + 1} \frac 2 {n-2}) = \frac {n^2}{2(n-1)^2(2n+1)(n-2)}$, which (again for $n > 2$) is positive and antitone.
-For $n = 0, 1, 2$, we take $\epsilon$ to be 9/40 (the same as its value at 3), then set $z_2 = 2 \log 2 - \log 3 - 1, z_1 = 3 \log 2 - 2 \log 3 - 3, z_0 = 4 \log 2 - 3 \log 3 - 6$.
-Some of those values are negative, but it's still bounded below, so by adding a constant offset it can be corrected, and the term is AST.
-
-\end{example}
-
-\begin{example}[Escaping spline]
-\label{ex:escaping spline}\citep[\S 5.4]{DBLP:journals/pacmpl/McIverMKK18}
-\[
-\underbrace{\big
-(\tY \, f \, x . \,
-0 \oplus_{\frac{1}{x+1}} f(x + 1) \big)}_{\Xi} 
-\, 10
-\]
-
-In this case, the fact that the ranking function must decrease at each $\tY$-step (in an antitone partial ranking function) by the expected value of $\epsilon$ not at the current term, but at the next term where the ranking function is defined, is a little harder to deal with, because the variable $x$ can change all the way to $0$ in one step, therefore simply adding a small offset doesn't suffice to compensate for this fact.
-
-Consider the candidate ranking function $\Xi \, n \mapsto n + 1$.
-For each $n$, $\Xi \, n$ reduces to either $0$ (with probability $\frac 1 {n + 1}$) or $\Xi \, (n + 1)$, therefore the expected value of the ranking function is $\frac{n(n+2)}{n+1} = n + 1 - \frac 1 {(n+1)^2}$, and the required decrease is $\frac{\epsilon(0) + n \epsilon(n + 2)}{n + 1}$, therefore eventually, the expected decrease isn't enough, whatever the value of $\epsilon(0)$ is.
-
-If instead, we take the ranking function to be defined after the $\tY$-reduction but before the $\tsample$-reduction as well, this can be resolved.
-Letting $\Theta[n] = \underline 0 \oplus_{\frac 1 n} \Xi \, n$:
-\begin{align*}
-\Theta[n] &\mapsto n \\
-\Xi \, 10 & \mapsto 12.
-\end{align*}
-For each $n$, $\Theta[n]$ reduces to either $0$ or $\Theta[n+1]$, with a $\tY$-reduction only in the latter case, and the condition that this is an antitone partial ranking function is that $n \geq \frac{(n-1)(n+1)}{n} + \frac{n-1}{n} \epsilon(n+1)$ and $12 \geq 11 + \epsilon(11)$.
-These are satisfied by setting $\epsilon(x) = \min(1, 1/x)$, therefore this term is AST.
-\end{example}
-
-\paragraph{Completeness}
-It seems very likely that this method is complete in the case that every reachable term is AST, but we have been unable to actually prove this.
-It is certainly at least capable of proving the termination of terms which terminate arbitrarily slowly (in the sense that there is no similar limitation to Theorem \ref{thm:rankable implies termination}, which can only prove termination of $\tY$-PAST terms).
-
-The following theorem does not prove completeness, but is suggestive in that direction:
-\begin{theorem}
-For any stopping time $T$ which is almost-surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
-\end{theorem}
-
-\input{relativised-completeness}
-
-\input{y-stopping-time}
-
-\section{Confluent Trace Semantics}
-When proving almost sure termination in this way, it is necessary to consider which terms a given term may reduce to.
-Sometimes however, the reduction that the programmer has in mind may not be strictly the call-by-value order defined so far, or considering an alternative reduction order may be simpler or more intuitive.
-
-Non-probabilistic lambda calculi generally have the Church-Rosser property, that if a term $A$ reduces to both $B_1$ and $B_2$, there is some $C$ 
-%with reduction sequences $B_1 \to^* C$ and $B_2 \to^* C$, 
-to which both $B_1$ and $B_2$ reduce,
-so the reduction order mostly doesn't matter. 
-In the probabilistic case, this may not be true, because $\beta$-reduction can duplicate $\tsample$s, so the outputs of the copies of the sample may be identical or independent, depending on whether the sample is taken before or after $\beta$-reduction.
-There are, however, some restricted variations on the reduction order that do not have this problem.
-
-\paragraph{}
-Even with this restriction, a sampling semantics in the style of the one already defined would not be entirely Church-Rosser, as, for example, $\red^3(\tsample - \tsample, (1,0,\dots))$ would be either $1$ or $-1$ depending on the order of evaluation of the $\tsample$s, as that determines which sample from the pre-selected sequence is used for each one.
-To fix this, rather than pre-selecting samples according to the order they'll be drawn in, select them according to the position in the term where they'll be used instead.
-
-A \emph{position} 
-%\akr{I'm not really sure how to phrase the citation, but this is adapted from \cite{Kennaway96infinitarylambda}.} 
-%\lo{The idea is common enough. I think no citation necessary.}
-is a finite sequence of steps into a term, defined inductively as
-\begin{align*}
-\alpha ::= \cdot \mid \lambda ; \alpha \mid @_1 ; \alpha \mid @_2 ; \alpha \mid \underline f_i ; \alpha \mid \tY ; \alpha \mid \textsf{if}_1 ; \alpha \mid \textsf{if}_2 ; \alpha \mid \textsf{if}_3 ; \alpha \mid \tscore ; \alpha.
-\end{align*}
-The \emph{subterm of $M$ at a position $\alpha$}, denoted $M \mid \alpha$, is defined as
-\begin{align*}
-M \mid \cdot & = M \\
-\lambda x. M \mid \lambda ; \alpha & = M \mid \alpha \\
-M_1 M_2 \mid @_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2 \\
-\underline f(M_1,\dots,M_n) \mid \underline f_i ; \alpha & = M_i \mid \alpha \quad \text{for }i \leq n \\
-\tY M \mid \tY ; \alpha & = M \mid \alpha \\
-\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2,3 \\
-\tscore(M) \mid \tscore ; \alpha & = M \mid \alpha
-\end{align*}
-so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda$ is undefined). 
-A position such that $M\mid \alpha$ does exist is said to \emph{occur} in $M$. 
-%\changed[lo]{
-\emph{Substitution} of $N$ at position $\alpha$ in $M$, written $M[N/\alpha]$, is defined similarly.
-For example, let 
-\[
-M = \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})
-\qquad
-\alpha =\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2
-\]
-then 
-\(
-M[\tsample / \alpha] = \lambda x \, y. y \, (\tif{x < 0}{y \, \tsample}{\underline 3}).
-\)
-%}
-
-Two subterms $N_1$ and $N_2$ of a term $M$, corresponding to positions $\alpha_1$ and $\alpha_2$, can overlap in a few different ways. 
-If $\alpha_1$ is an initial segment (i.e.~prefix) of $\alpha_2$ (written as $\alpha_1 \leq \alpha_2$), then $N_2$ is also a subterm of $N_1$.
-If neither $\alpha_1 \leq \alpha_2$ nor $\alpha_1 \geq \alpha_2$, the positions are said to be \emph{disjoint}.
-The notion of disjointness is mostly relevant in that if $\alpha_1$ and $\alpha_2$ are disjoint, performing a substitution at $\alpha_1$ will leave the subterm at $\alpha_2$ inaffected.
-\lo{The preceding sentence is unclear. Typo?} \akr{It wasn't a typo, but I've made it a little more explicit now. Is this clearer?}
-
-With this notation, a more general reduction relation $\to$ can be defined. \akr{It would almost work to have one of the neater versions of $\tY$-reduction here.}
-%\lo{On first reading, I struggled to parse the following clauses.}
-\begin{definition}
-\label{def:more general red}
-The binary relation $\to$ is defined by the following rules, each is conditional on a redex occurring at position $\alpha$ in the term $M$:
-\begin{align*}
-  \text{if } M \mid \alpha = (\lambda x.N) V,\ & M \to M[N[V/x]/\alpha] \\
-  \text{if } M \mid \alpha = \underline f (\underline r_1, \dots , \underline r_n),\ & M \to M[\underline{f(r_1,\dots,r_n)}/\alpha] \\
-  \text{if } M \mid \alpha = \tY \lambda x. N,\ & M \to M[\lambda z. N[(\tY \lambda x. N)/x] z/\alpha] \text{ where $z$ is not free in $N$}\\
-  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_1/\alpha] \text{ where }r < 0 \\
-  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_2/\alpha] \text{ where }r \geq 0 \\
-  \text{if } M \mid \alpha = \tsample & \text{ and $\lambda$ does not occur after $@_2$ or $\tY$ in $\alpha$},\\ & M \to M[\underline r/\alpha] \text{ where } r \in [0,1] \\
-  \text{if } M \mid \alpha = \tscore(\underline r),\ & M \to M[\underline r/\alpha].
-\end{align*}
-In each of these cases, $M \mid \alpha$ is the \emph{redex}, and reduction takes place at $\alpha$.
-\end{definition}
-
-\paragraph{}
-Labelling the pre-chosen samples by the positions in the term would also not work because in some cases, a $\tsample$ will be duplicated before being reduced, for example, in $(\lambda x.
-x \, {\underline 0} \, \mathbin{\underline{+}} \, {x \, \underline 0}) (\lambda y.
-\tsample)$, both of the $\tsample$ redexes that eventually occur originate at $@_2 ; \lambda$.
-It is therefore necessary to consider possible positions that may occur in other terms reachable from the original term. 
-Even this is itself inadequate because some of the positions in different reachable terms need to be considered the same, and the number of reachable terms is in general uncountable, which leads to measure-theoretic issues.
-
-Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $Sk(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$.
-\lo{Do we actually use the notation $S[r]$ in what follows?} \akr{Yes, in Lemma \ref{lem:independentSamples} for example.}
-\lo{Skeleton is defined in Sec.~\ref{sec:sampling semantics}.}
-The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, and $\tsample$ reducing to $\skeletonPlaceholder$.
-For example, we have 
-\(
-(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \tsample
-\to
-(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \skeletonPlaceholder
-\to
-\tif{\skeletonPlaceholder < 0}{\skeletonPlaceholder}{\skeletonPlaceholder}
-\to
-\skeletonPlaceholder
-\)
-
-Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a $\to$-reduction sequence of skeletons starting at $Sk(M)$, and the second of which is a position in the final skeleton of the reduction sequence.
-As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the usual linear sampling semantics, there will be (with one more caveat introduced later) a sample from $I$ pre-selected for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.
-
-Reduction sequences are used rather than reachable skeletons because if the same skeleton is reached twice, different samples may be needed.
-For example, consider the term $M = \tY (\lambda f x.\ \tif{\tsample - \underline{0.5} < 0}{f \, x}{x}) \, \underline 0$, which reduces after a few steps to $N = \tif{\tsample - \underline{0.5}  < 0}{M}{\underline 0}$.
-If the pre-selected sample for $(N,\textsf{if}_1;\underline{-}_1)$ is less than 0.5, $N$ reduces back to $M$, then $N$ again, then the same sample is used the next time, therefore it's an infinite loop, whereas if samples are labelled by reduction sequences, the samples for $M \to \dots \to N$ are independent from the samples for $M \to \dots \to N \to M \to \dots \to N$, and so on.
-
-The reduction sequences of skeletons will often be discussed as though they were just skeletons, identifying them with their final skeletons.
-With this abuse of notation, a reduction sequence $N$ (actually $N_1 \to \dots \to N_n = N$) may be said to reduce to a reduction sequence $O$, where the reduction sequence implicitly associated with the final skeleton $O$ is $N_1 \to \dots \to N_n \to O$.
-
-\paragraph{}
-This is still not quite sufficient because sometimes the same samples must be used at corresponding positions in different reduction sequences.
-For example, the term $M = \tsample + \tsample$ has the reachable skeletons $N_1 = \skeletonPlaceholder + \tsample, N_2 = \tsample + \skeletonPlaceholder, O = \skeletonPlaceholder + \skeletonPlaceholder$ and $\skeletonPlaceholder$, with reductions $M \to N_1 \to O \to \skeletonPlaceholder$ and $M \to N_2 \to O \to \skeletonPlaceholder$.
-In the reduction $M \to N_1$, the sample labelled $(M, \underline{+}_2)$ is used, and in the reduction $N_2 \to O$, the sample labelled $(M \to N_2, \underline{+}_2)$ is used.
-Each of these samples becomes the value of the second numeral in $O$ in their respective reduction sequences, therefore in order for Church-Rosserness to be attained, they must be the same.
-Which elements of $L_0(M)$ must match can be described by the relation $\sim^*$:
-
-\begin{definition}
-The relation $\sim$ is defined as the union of the minimal symmetric relations $\sim_p$ (``$p$'' for parent-child) and $\sim_c$ (``$c$'' for cousin) satisfying
-\begin{enumerate}
-    \item If $N$ reduces to $O$ with the redex at position $\alpha$, and $\beta$ is a position in $N$ disjoint from $\alpha$, then $(N,\beta) \sim_p (O,\beta)$.
-
-    \item If $N$ $\beta$-reduces to $O$ at position $\alpha$, $\beta$ is a position in $N \mid \alpha;@_1;\lambda$ and $N \mid \alpha;@_1;\lambda;\beta$ is not the variable involved in the reduction, $(N,\alpha;@_1;\lambda;\beta) \sim_p (O, \alpha;\beta)$
-
-    \item If $N$ $\textsf{if}$-reduces to $O$ at position $\alpha$, with the first resp.
-second branch being taken, and $\alpha;\textsf{if}_i;\beta$ occurs in $N$ (where $i = 2$ resp.
-$3$), $(N,\alpha;\textsf{if}_i;\beta) \sim_p (O,\alpha;\beta)$
-
-    \item If $N$, $O_1$ and $O_2$ match any of the following cases:
-    \begin{enumerate}
-        \item $N$ contains redexes at disjoint positions $\alpha_1$ and $\alpha_2$, $O_1$ is $N$ reduced first at $\alpha_1$ then $\alpha_2$ and $O_2$ is $N$ reduced first at $\alpha_2$ then at $\alpha_1$.
-
-        \item $N \mid \alpha = \tif{\underline r < 0}{N_1}{N_2}$, where $r < 0$ (or, respectively, $r \geq 0$), $(N_2 \text{ resp.
-} N_1) \mid \beta$ is a redex, and $O_1$ is $N$ reduced at $\alpha$ and $O_2$ is $N$ reduced first at $\alpha;(\textsf{if}_3 \text{ resp.
-} \textsf{if}_2);\beta$ then at $\alpha$
-
-        \item $N \mid \alpha = \tif{\underline r < 0}{N_1}{N_2}$, where $r < 0$ (or, respectively, $r \geq 0$), $(N_1 \text{ resp.
-} N_2) \mid \beta$ is a redex, and $O_1$ is $N$ reduced first at $\alpha$ then at $\alpha;\beta$ and $O_2$ is $N$ reduced first at $\alpha;(\textsf{if}_2 \text{ resp.
-} \textsf{if}_3);\beta$ then at $\alpha$
-
-        \item $N \mid \alpha = (\lambda x.
-A) B$, there is a redex in $A$ at position $\beta$, $O_1$ is $N$ reduced first at $\alpha$ then at $\alpha;\beta$, and $O_2$ is $N$ reduced first at $\alpha;@_1;\lambda;\beta$ then at $\alpha$
-
-        \item $N \mid \alpha = (\lambda x.
-A) B$, $B \mid \beta$ is a redex, $(\gamma_i)_i$ is a list of all the positions in $A$ where $A \mid \gamma = x$, ordered from left to right, $O_1$ is $N$ reduced first at $\alpha ; @_2 ; \beta$ then at $\alpha$, and $O_2$ is $N$ reduced first at $\alpha$ then at $\alpha;\gamma_i;\beta$ for each $i$ in order.
-
-        \item $N \mid \alpha = \tY (\lambda x.
-A)$, $A$ reduced at $\beta$ is $A'$, $(\gamma_i)_i$ is a list of all the positions where $A' \mid \gamma  = x$, ordered from left to right, $O_1$ is $N$ reduced first at $\alpha;\tY;\lambda;\beta$ then at $\alpha$, and $O_2$ is $N$ reduced first at $\alpha$ then at $\alpha;\lambda;@_1;\gamma_i;Y;\lambda;\beta$ for each $i$ in order where $\gamma_i$ is left of $\beta$ then at $\alpha;\lambda;@_1;\beta$ then at $\alpha;\lambda;@_1;\gamma_i;Y;\lambda;\beta$ for the remaining values of $i$.
-    \end{enumerate}
-    (in which case $O_1$ and $O_2$ are equal as skeletons, but with different reduction sequences), $O_1'$ and $O_2'$ are the results of applying some reduction sequence to each of $O_1$ and $O_2$ (the same reductions in each case, which is always possible because they're equal terms), and $\delta$ is a position in $O_1'$ (or equivalently $O_2'$), then $(O_1',\delta) \sim_c (O_2',\delta)$.
-\end{enumerate}
-The $\sim_c$-rules are illustrated in the following diagrams.
-\end{definition}
-
-\include{sim-diagram}
-
-The reflexive transitive closure $\sim^*$ of this relation is used to define the set of \emph{potential positions} $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable skeletons.
-If $(N,\alpha) \sim^* (O,\beta)$, then $N \mid \alpha$ and $O \mid \beta$ both have the same shape (i.e.~they're either both the placeholder $\skeletonPlaceholder$, both variables, both applications, both $\tsample$s etc.), therefore it's well-defined to talk of the set of potential positions where there is a $\tsample$, $L_s(M)$.
-Formally $L_s(M) := \{[(X, \alpha)]_{\sim^\ast_M} : X \mid \alpha = \tsample\}$.
-The new sample space is then defined as $I^{L_s(M)}$, with the Borel $\sigma$-algebra and product measure.
-\changed[lo]{Since $I^{L_s(M)}$ is a countable product, the measure space is well-defined \citep[Cor.~2.7.3]{AshDD00}.}
-
-\paragraph{}
-Although the six cases in the definition of $\sim_c$ are defined separately, there are some commonalities worth noting.
-Each case (except a, which is symmetrical) has a right branch and a left branch (the branches of case b are unfortunately displayed the wrong way around in the diagram).
-The right branch has reduction steps at two positions, $\beta$ then $\alpha$, where $\beta > \alpha$ (i.e.~$\beta$ is inside $\alpha$), and in the left branch, the order of these reductions are swapped, so the reductions are at $\alpha$ then $\beta_1, \dots, \beta_n$.
-The position $\alpha$ of one of the reductions is unchanged, and the other reduction may have multiple (in cases e and f) or 0 (in cases b and e) images in the left branch.
-The images $(\beta_i)$ are all still inside (greater than or equal to) $\alpha$, and although they are not generally equal to $\beta$, the subskeletons at these positions have a similar shape to the subskeleton (initially) at position $\beta$, so that the reductions at these positions are the same type of reduction (e.g.~both $\beta$-reduction or both $\textsf{if}$-reduction).
-Case a is also similar, except that the relevant positions are all disjoint rather than contained in one another, and either branch could be considered the right branch.
-
-\iffalse
-\changed[lo]{Since $L_s(M)$ is a countable set, there is a unique probability measure $\mu$ on the infinite product $\prod_{i \in L_s(M)} \Sigma_I$ of the Borel $\sigma$-algebras $\Sigma_I$ satisfying
-\[
-\mu\set{\omega \in \Omega \mid \omega_1 \in A_1, \cdots, \omega_n \in A_n} = \prod_{i=1}^n \Leb(A_i)
-\]
-where $\Leb$ is the Lebesgue measure on the measurable space $(I, \Sigma_I)$.}
-\fi
-
-\paragraph{}
-Before defining the new version of the reduction relation $\red$, a few lemmas about the properties of $\sim$ are necessary for it to be well-defined.
-
-\begin{lemma} \label{downAcrossUp}
-If $(N_1,\alpha_1) \sim^* (N_2,\alpha_2)$, there is some descendants $N'_1, N'_2$ of $N_1$ resp.~$N_2$, and a position $\alpha'$ in both of them such that $(N_1,\alpha_1) \sim_p^* (N'_1, \alpha') \sim_c^* (N'_2, \alpha') \sim_p^* (N_2, \alpha_2)$.
-\end{lemma}
-\begin{proof}
-  The $\sim_p$ relation can be split into $\sim_\downarrow \cup \sim_\uparrow$, where $(A,\alpha) \sim_\downarrow (B,\beta)$ if $A \to B$ and $(A,\alpha) \sim_p (B,\beta)$, and similarly $(A,\alpha) \sim_\uparrow (B,\beta)$ if $B \to A$ and $(A,\alpha) \sim_p (B,\beta)$.
-At each stage of this proof, it will be assumed that the $\sim_c$ steps and the $\sim_p$ steps in the sequence from $(N_1,\alpha_1)$ to $(N_2,\alpha_2)$ are rearranged such that there is never a $\sim_c$ immediately before a $\sim_\downarrow$, and there is never a $\sim_c$ immediately after a $\sim_\uparrow$.
-This rearrangement is always possible, because if there is some subsequence $(A,\alpha) \sim_c (B,\alpha) \sim_\downarrow (C,\beta)$, then there is an alternate path $(A,\alpha) \sim_\downarrow (B',\beta) \sim_c (C,\beta)$, where the reduction and the $\sim_\downarrow$ from $A$ to $B'$ are the same as those from $B$ to $C$, and the $\sim_c$ step is the same, but with the reduction sequences $O_1 \to^* O_1'$ and $O_2 \to^* O_2'$ in the definition of $\sim_c$ extended by the reduction $B \to C$.
-With these rearrangements assumed, it is sufficient to prove that the $\sim_p$ steps can be rearranged so that all of the $\sim_\downarrow$ steps come before all of the $\sim_\uparrow$ steps (possibly introducing some $\sim_c$ steps in the process).
-
-  The rearrangement to put all of the $\sim_\downarrow$s before all of the $\sim_\uparrow$s proceeds by induction on the number of $\sim_\downarrow$s that occur after the first occurrence of $\sim_\uparrow$.
-If it's 0, all of the $\sim_p$s are already in the correct order, and we're done.
-Otherwise, take the subsequence from the first $\sim_\uparrow$ to the first $\sim_\downarrow$ after that.
-This must be rearranged to some $\sim_\downarrow$s followed by some $\sim_\uparrow$s, then once that's done, the number of $\sim_\downarrow$s after the first $\sim_\uparrow$ in the overall sequence will have decreased by 1.
-
-  Let this subsequence be $A \sim_\uparrow^n B \sim_\downarrow C$.
-Suppose for induction that $A \sim_\uparrow^k (D,\delta) \sim_\downarrow^* (E,\epsilon) \sim_\uparrow^* C$ for some $D, \delta, E, \epsilon$, and $0 \leq k \leq n$, where in the reduction sequence $D \to^* E$, for any pair of reductions at positions which aren't disjoint, the reduction at the innermost (greater) position occurs first.
-Reduction sequences with this property will be called ``parallel'', as it is analagous to parallel reduction as in \akr{TODO: find this reference, for the proof of the Church-Rosser theorem by Tait and Martin-L\"of}.
-This induction is in reverse, with $k$ decreasing from $n$ to $0$.
-If $k = n$, simply set $(D,\delta) = B$ and $(E,\epsilon) = C$.
-If $k = 0$, then $A$ is related to $B$ by a sequence of $\sim_\downarrow$s then $\sim_\uparrow$s, as desired.
-In the intermediate steps, $k$ must be decreased by one, so a subsequence of one $\sim_\uparrow$ followed by a parallel sequence of $\sim_\downarrow$s must be replaced by a parallel sequence of $\sim_\downarrow$s followed by some $\sim_\uparrow$s.
-
-  For any individual $\sim_\uparrow$ then $\sim_\downarrow$ pair, let $(A,\alpha) \sim_\uparrow (B,\beta) \sim_\downarrow (C,\gamma)$.
-The skeleton $B$ reduces to both $A$ and $C$.
-If these are the same reduction, then $(A,\alpha) = (C,\gamma)$ and this subsequence may be removed.
-If they are different, then the way they overlap corresponds to one of the cases in the definition of $\sim_c$, either case a if the reduction positions are disjoint, or one of the cases b--f if one of the reduction positions is inside the other.
-In any case, there is a reduction sequence from each of $A$ and $C$ to some common skeleton $D$ ($O_1$ and $O_2$ in the definition of $\sim_c$).
-For the same reason that $(B,\beta) \sim_\downarrow (C,\gamma)$, there is some $\delta$ such that $(A,\alpha) \sim_\downarrow^* (D,\delta)$, and similarly for the same reason that $(B,\beta) \sim_\downarrow (A,\alpha)$, for the same $\delta$, $(C,\gamma) \sim_\downarrow^* (D,\delta)$.
-There are a lot of cases to check for this statement, but all of them are rather simple, and similar to each other.
-There is therefore an alternative $\sim^*$ sequence from $(A,\alpha)$ to $(C,\gamma)$, of the form $(A,\alpha) \sim_\downarrow^* (D_1,\delta) \sim_c (D_2,\delta) \sim_\uparrow^* (C,\gamma)$, where $D_1$ and $D_2$ are the alternative reduction sequences leading to the same skeleton $D$.
-There are only multiple $\sim_\downarrow$ steps in the result if the position of the reduction $B \to C$ is inside the position of the reduction $B \to A$, and similarly, there are only multiple $\sim_\uparrow$ steps in the result if the position of the reduction $B \to A$ is inside the position of the reduction $B \to C$.
-If there are multiple $\sim_\downarrow$ steps, they are not necessarily parallel, but in the only case where they aren't (case f of $sim_c$), the subsequence of $\sim_\downarrow$ steps with the reductions at $\alpha;\lambda;@_1;\beta$ then $\alpha;\lambda;@_1;\gamma_i;\tY;\lambda;\beta$ for each $i$ where $\gamma_i \geq \beta$ (in the notation of the definition of $\sim_c$) can be replaced by $\sim_\downarrow$s with the reductions at $\alpha;\lambda;@_1;\gamma'_i;\tY;\lambda;\beta$ then $\alpha;\lambda;@_1;\beta$, then another $\sim_c$, where $(\gamma'_i)$ is the list of positions in $A$ (\emph{before} the reduction at $\beta$ to $A'$, unlike $(\gamma_i)$) where $A | \gamma'_i = x$ and $\gamma'_i > \beta$.
-
-  The effect of this rearrangement is (ignoring the $\sim_c$s) to swap the order of the $\sim_\uparrow$ and the $\sim_\downarrow$, except that if the positions of one of these reductions is inside the other, that inner $\sim_p$ may be duplicated.
-If the positions are disjoint, both of them are unchanged (corresponding to case a of $\sim_c$), and if one is inside the other, then the outer position is unchanged and the other resultant positions, although they aren't equal to the original inner position, are still inside the outer position.
-Taking the sub-sequence of one $\sim_\uparrow$ followed by a parallel sequence of $\sim_\downarrow$s, this rearrangement can be repeatedly applied to move the $\sim_\uparrow$ further along the sequence.
-If at some point a $\sim_\uparrow$ and $\sim_\downarrow$ match (by relating the same (skeleton, position) pairs in opposite directions), they may be removed and the process may stop early.
-The $\sim_\uparrow$ may be duplicated if it passes a $\sim_\downarrow$ whose position is outside of the $\sim_\uparrow$'s position, but by the assumption that the sequence of $\sim_\downarrow$s is initially parallel, no position inside of this occurs later in the sequence, therefore all of the resultant $\sim_\uparrow$s pass the remaining $\sim_\downarrow$s without changing or duplicating them.
-The $\sim_\uparrow$s may be further duplicated, but the number of steps left in this process is bounded by the product, for all of the remaining $\sim_\downarrow$s, of $2 + \text{the number of occurrences of the variable relevant to the reduction}$, because for each $\sim_\downarrow$, that is more than the number of duplicates it could produce for any $\sim_\uparrow$ that passes it.
-It is also possible (earlier) for some of the $\sim_\downarrow$s to be duplicated, but the only case where this process would not terminate, that both the $\sim_\downarrow$s and the $\sim_\uparrow$s continue being duplicated forever so that the number of switches left to do never decreases, is prevented by the parallelness condition, as all of the duplications of $\sim_\downarrow$s occur before all of the duplications of $\sim_\uparrow$s.
-
-  It still remains to be shown that the sequence of $\sim_\downarrow$s left at the end of this can be made parallel.
-The changes that may have occurred since the previous version of this sequence (which was known to be parrallel already), are that some of the $\sim_\downarrow$s may have been duplicated by passing the $\sim_\uparrow$, and if that $\sim_\uparrow$ matches one of the $\sim_\downarrow$s, that $\sim_\downarrow$ is removed.
-The position of the $\sim_\uparrow$'s reduction is the same for all of these duplications, because it is not changed until later in the sequence, when the $\sim_\downarrow$s with positions outside of that may occur.
-The resultant positions after a duplication are in the same position relative to each other, and therefore the order of the remaining $\sim_\downarrow$s is automatically parallel, except that (letting the position of the $\sim_\uparrow$'s reduction be $\alpha$) if there are $\sim_\downarrow$s with reductions at positions $\alpha;@_1;\lambda;\beta$ and $\alpha;@_2;\gamma$, or $\alpha;\tY;\lambda;\beta$ and $\alpha;\tY;\lambda;\gamma$, some of the resultant $\sim_\downarrow$s' positions may be inside each other where they were originally not.
-Similarly to the case where a $\tY$ reduction and a reduction at a position inside it are swapped though, the order can be fixed by swapping some of the $\sim_\downarrow$ steps with each other.
-In the first case, the reduction at position $\alpha;@_1;\lambda;\beta$ ends up at $\alpha;\beta$, and the reduction at position $\alpha;@_2;\gamma$ ends up at the positions $\alpha;\delta_i;\gamma$, where $(\delta_i)$ is the positions of the variable relevant to the $\sim_\uparrow$'s reduction inside its lambda.
-The positions of the variable in the lambda may be different before the reduction at $\beta$, but the positions of the redexes corresponding to the original redex at $\alpha;@_2;\gamma$ are changed similarly.
-Rather than reducing at $\alpha;\beta$ then at $\alpha;\delta_i;\gamma$, it is therefore possible to reach the same point by reducing at $\alpha;\delta'_i;\gamma$ then at $\alpha;\beta$, where $(\delta'_i)$ is the positions of the relevant variable within the lambda before the reduction at $\alpha;@_1;\lambda;\beta$ (and again, a $\sim_c$ of some sort must also be introduced).
-The other case, that the $\sim_\uparrow$'s reduction is a $\tY$-reduction, is similar.
-Each reduction which is duplicated has one image at $\alpha;\lambda;@_1;$its original relative position, and one for each occurrence of the variable.
-Those corresponding to variable occurrences may need to be moved to before those at $\lambda;@_1$.
-
-  In summary, to reach the desired order of $\sim_p$ steps, each $\sim_\downarrow$ may have to be moved past some $\sim_\uparrow$ steps, and it may get duplicated in the process, but the resultant sequence of $\sim_\downarrow$s can all be put in parallel order, and it has been shown that moving a $\sim_\uparrow$ past a sequence of $\sim_\downarrow$s is always possible if they're in parallel order, therefore the overall process terminates and reaches a point where all of the $\sim_\downarrow$s precede all of the $\sim_\uparrow$s, and all of the $\sim_c$s are in between them.
-All of these rearrangements leave the end points of the sequence unchanged, therefore this is an alternative sequence of $\sim$ steps between $(N_1,\alpha_1)$ and $(N_2,\alpha_2)$ of the form $(N_1,\alpha_1) \sim_\downarrow^* (N_1', \alpha') \sim_c^* (N_2', \alpha') \sim_\uparrow^* (N_2,\alpha_2)$.
-\end{proof}
-
-If $(X,\alpha) \sim_c (Y,\alpha)$ then the final skeletons in $X$ and $Y$ are equal, and $(X,\beta) \sim_c (Y,\beta)$ for any $\beta$ that occurs in this skeleton, therefore $\sim_c$ can be considered to apply to reduction sequences even without positions, $X \sim_c Y$ iff $(X,\cdot) \sim_c (Y,\cdot)$.
-The structure of $\sim_c$ can be seen more clearly by choosing a canonical example from each equivalence class.
-For this, the member of the class that's in call-by-value order is used.
-A reduction sequence is in call-by-value order if for every pair of reductions in the sequence at positions $\alpha$ and $\beta$, where the reduction at $\alpha$ happens first, either $\alpha \leq \beta$ or $\alpha$ and $\beta$ are disjoint, with $\alpha$ to the left of $\beta$ (i.e.~the first elements of the sequences $\alpha, \beta$ where they differ are $@_1$ and $@_2$, $\underline f_i$ and $\underline f_j$ for $i < j$, or $\textsf{if}_i$ and $\textsf{if}_j$ for $i < j$, in that order).
-Note that although the reductions that occur in the sequence are in call-by-value order, it may not be the actual call-by-value reduction sequence starting from that point because some redexes may remain un-reduced even when other reductions that should happen afterwards occur.
-It is only required that the reductions that do occur occur in the correct order.
-
-\begin{lemma} \label{canonicalCousins}
-For any reduction sequence $X$, there is a unique reduction sequence $X_{cbv}$ that is related to $X$ by $\sim_c$ and is in call-by-value order, and if $X \sim_c^* Y$, then $X_{cbv} = Y_{cbv}$.
-\end{lemma}
-
-\lo{As defined, $\sim_c$ is not symmetric.}
-
-\begin{proof}
-
-Define $X_n$ recursively so that $X_0 = X$ and if $X_n$ is not in CBV order, take the last pair of reductions in $X_n$ which aren't in CBV order.
-They are adjacent in the sequence, as the CBV order is a total order.
-Let the positions of these reductions be $\alpha$ and $\beta$ respecively.
-If they are disjoint, they can be considered as one of the branches of $\sim_c$ case a.
-Otherwise, $\beta < \alpha$, and the sub-skeleton at position $\beta$ at the appropriate point in the reduction sequence is a redex with non-trivial subterms, therefore it's of one of the forms $\tif{\skeletonPlaceholder>0}{A}{B}$, $(\lambda x.A) B$ or $\tY \lambda x.
-A$, therefore the reductions $\alpha$ and $\beta$ form the $O_2$ branch of one of the cases b, c, d, e or f of $\sim_c$.
-In either case, define $X_{n+1}$ as equal to $X_n$ except taking the other (left) branch, so that the order of the reductions at $\alpha$ and $\beta$ are switched (although the equivalent(s) of the reduction at $\alpha$ may not actually be at that position).
-
-This sequence of swaps rearranges the reductions in $X$ rather like insertion-sort (but in some cases duplicating or deleting the new element), and eventually reaches some $X_n$ that is in CBV order.
-This is defined to be $X_{cbv}$.
-
-\paragraph{}
-If $Y \sim_c X \sim_c^* X_{cbv}$, and the reductions involved in $Y \sim_c X$ are not the last pair of reductions in $Y$ that aren't in CBV order, the order of the $\sim_c$s can be rearranged so that the sequence $Y \sim_c^* X_{cbv}$ is the canonical sequence given above, therefore $Y_{cbv} = X_{cbv}$.
-To be more precise about this rearrangement, consider the sequence of $\sim_c$ steps from $Y$ to $X_{cbv}$.
-The first step takes some sub-sequence of the reduction sequence $Y$ and swaps the order of those reductions.
-If $X = X_{cbv}$, then $X = Y_1 = Y_{cbv}$ already.
-Otherwise, consider the cases according to whether the image of this subsequence in $X$ occurs earlier than the next pair of reductions to be swapped (i.e.~the last pair of reductions in $X$ that occur not in CBV order).
-If not, then either the step $Y \sim_c X$ changes this subsequence into CBV order, or out of it.
-If it is into, then it was the last out-of-order pair in $Y$ therefore the whole sequence $Y \sim_c^* X_{cbv}$ is already in the canonical order and $Y_{cbv} = X_{cbv}$.
-If it switches the pair to the wrong order, then the result of that switch is the last out-of-order pair in $X$ therefore $Y = X_1$, and the sequence $X_1 \sim_c^* X_{cbv}$ is just a suffix of the sequence $X \sim_c^* X_{cbv}$ therefore $(X_1)_{cbv} = X_{cbv}$ and again $Y_{cbv} = X_{cbv}$.
-
-The remaining cases are when the subsequence of reduction steps involved in $Y \sim_c X$ occur before the last out-of-order pair in $X$, but possibly overlapping.
-If there is no overlap, then the $\sim_c$ steps do not interfere with each other at all, and can simply be performed in the other order.
-This gives a different sequence of $\sim_c$ steps $Y \sim_c Y_1 \sim_c X_1 \sim_c^* X_{cbv}$.
-As $Y_{cbv} = (Y_1)_{cbv}$, we can proceed by induction in this case and use the fact that $(Y_1)_{cbv} = (X_1)_{cbv} = X_{cbv}$ to acheive the result.
-
-In the other case, there is some overlap between the regions involved in $Y \sim_c X$ and $X \sim_c X_1$, but the last reduction in $X \sim_c X_1$ is later than the last reduction in $Y \sim_c X$.
-There are only 2 reductions in $X$ involved in $X \sim_c X_1$, therefore the region involved in $Y \sim_c X$ ends on the first of the reductions involved in $X \sim_c X_1$.
-Now consider the specific sequence of positions of the reductions in $X$ involved in either of these steps.
-Let the last of these positions be $\alpha$, the first $\gamma$, and the second-last, third-last and so on $\beta_1, \beta_2, \dots$ respectively, and let the positions of the redexes in $Y$ involved in $Y \sim_c X$ be $\beta$ then $(\gamma_i)$ in order..
-Then $X \sim_c X_1$ swaps $\beta_1$ and $\alpha$, and $\alpha$ comes before $\beta_1$ in CBV order (i.e.~$\alpha < \beta_1$ or $\alpha$ is left of $\beta_1$), and the step $Y \sim_c X$ either proceeds forwards towards CBV order, swapping $\beta$ below $\gamma_1 = \gamma$ to form $\gamma$ then $(\beta_i)$, or proceeds backwards, taking $Y$ even farther from CBV order, swapping the $(\gamma_i)$ to above $\beta$, forming $\gamma$ then $\beta_1 = \beta$.
-In each of the cases below, it is required to show that the canonical sequence $(X_i)_i$ eventually reaches some term which is the same as one in $(Y_j)_j$, and from that point on, they match, therefore they reach the same end result: $X_{cbv} = Y_{cbv}$.
-
-\begin{itemize}
-  \item $Y \sim_c X$ in the backwards direction (so that $Y$ is closer to CBV order than $X$ is): By swapping the roles of $X$ and $Y$, and $\beta$ and $\gamma$, this is equivalent to the $Y \sim_c X$ forwards case.
-The assumption that $\alpha$ is before $\beta_1$ in CBV order (so that $\beta_1, \alpha$ is the pair to be swapped in $X \sim_c X_1$) maps to the assumption that $\alpha$ is before $\gamma_1$ (so that the result doesn't just follow trivially by $Y_1$ being equal to $X$) and vice-versa.
-Then one of the cases below establishes that $X_i = Y_j$ for some $i$ and $j$, therefore after swapping back, $Y_i = X_j$.
-\item $Y \sim_c X$ forwards and $\alpha$ comes after $\gamma$ in CBV order: In this case, the $\gamma$ and $\alpha$ reduction steps are already in the correct order in $Y$, therefore $\beta$ and $\gamma$ are already the last out-of-order pair of reductions in $Y$, $Y_1 = X$, and $Y \sim_c X \sim_c^* X_{cbv}$ is already the canonical sequence from $Y$ therefore $Y_{cbv} = X_{cbv}$.
-\item $Y \sim_c X$ forwards and $\alpha$ is disjoint from $\gamma$ and $\beta$: In this case, the sequence $(X_i)_i$ starts by swapping $\alpha$ and $\beta_1$ by case a, then because $\gamma$ is disjoint from $\alpha$ and comes after is in CBV order, $\gamma$ is right of $\alpha$, therefore either all the remaining $\beta_i$s, are $\geq \gamma$ (if $\beta$ and $\gamma$ aren't disjoint) or there are no remaining $\beta_i$s (if $\beta$ and $\gamma$ are disjoint, in which case they swap by case a and there is only $\beta_1$).
-The sequence $(X_i)_i$ therefore proceeds to swap all the remaining $\beta_i$s then $\gamma$, in order, below $\alpha$, resulting in $X_i$ for some $i$ having as its relevant sub-reduction-sequence $\alpha, \gamma, \beta_k, \dots, \beta_1$.
-The other sequence $(Y_j)_j$ starts by swapping $\gamma$ then $\beta$ below $\alpha$, then swapping $\gamma$ and $\beta$.
-Because $\alpha$ is disjoint from $\gamma$, the reduction there does not affect the sub-skeleton at position $\gamma$, therefore this last $\sim_c$ step proceeds identically to how it did in $Y \sim_c X$, resulting in $\gamma, \beta_k, \dots, \beta_i$.
-Overall, this results in $Y_3$ having the relevant sub-reduction-sequence $\alpha, \gamma, \beta_k, \dots, \beta_1$, therefore $Y_3 = X_i$ for the aforementioned value of $i$.
-\item $Y \sim_c X$ forwards, $\alpha \leq \gamma$ and $\alpha$ (and therefore also $\gamma$) is disjoint from $\beta$: In this case, $(X_i)_i$ proceeds by swapping $\beta_1 = \beta$ below $\alpha$ by case a, then swapping $\gamma$ below $\alpha$ by one of the other cases, and $(Y_j)_j$ proceeds by swapping $\gamma$ and $\alpha$, then swapping $\beta$ below $\alpha$ then all of the images of $\gamma$ (which are $\geq \alpha$ and therefore left of $\beta$).
-In both cases, the subterm at $\alpha$ when $\alpha$ and $\gamma$ are swapped is unaffected by the reduction at $\beta$, therefore it produces the same result in both cases, therefore the overall sequence in both cases is $\alpha$ then the images of $\gamma$ then $\beta$.
-\item Both $\gamma$ and $\beta$ are $> \alpha$ but they're disjoint from each other, and either the skeleton at position $\alpha$ is of the form $\tif{\skeletonPlaceholder > 0}{A}{B}$ or it is of the form $A B$ and either both $\gamma$ and $\beta$ are $> \alpha;@_1$ or they are both $> \alpha;@_2$: In this case, $(X_i)_i$ proceeds by swapping $\beta$ then $\gamma$ below $\alpha$, in each case possibly forming 0 or multiple images.
-The only way the number of images of each of these positions can be different is if the subskeleton at $\alpha$ is an $\textsf{if}$, and there are 0 of one of them and 1 of the other, in which case they're trivially in the correct order already.
-Otherwise, all of the images of $\beta$ and $\gamma$ are disjoint from one another, therefore none of the positions, or their relative order, change when they are swapped, therefore the next part of the sequence $(X_i)_i$ is just insertion-sort running on the images of $\beta$ and $\gamma$ with $\sim_c$ case a swaps until they're in the correct order.
-The sequence $(Y_j)_j$ starts by swapping $\gamma$ then $\beta$ below $\alpha$, then as before, sorting the images into the correct order by case a swaps.
-The images originally formed of $\gamma$ and $\beta$ are the same in both cases, therefore the final order is the same, so $X_i = Y_j$ for some $i$ and $j$.
-\item $Y \sim_c X$ forwards, $\alpha;@_1 < \gamma$, and $\alpha;@_2 < \beta$: Let $x$ be the variable involved in the $\beta$-reduction at $\alpha$, so that the sub-skeleton at $\alpha$ is $(\lambda x.
-A) B$ for some $A, B$, where $\gamma = \alpha;@_1;\lambda;\gamma'$ and $A \to A'$ at $\gamma'$, and $\beta = \alpha;@_2;\beta'$, and $B \to B'$ at $\beta'$.
-In this case, $(X_i)_i$ starts by swapping $\beta$ below $\alpha$ by case e, producing one image of $\beta$ for each $x$ in $A'$, then swapping $\gamma$ below $\alpha$ by case d, producing a single image $\alpha;\gamma'$.
-For each of the instances of $x$ in $A$ left of $\gamma'$, the reduction at $\alpha;\gamma'$ is then swapped below the corresponding image of $\beta$.
-The sequence $(Y_j)_j$ starts by swapping $\gamma$ below $\alpha$ then $\beta$ below $\alpha$, but in this case the images of $\beta$ produced may be different.
-In this case the swap happens earlier in the reduction sequence than $\gamma$, so that the body of the lambda is still $A$ rather than $A'$, and the reduction at $\gamma$ may rearrange or change the number of instances of $x$.
-Next, each of the images of $\beta$ to the right of $\alpha;\gamma'$ are swapped with it by case a.
-At this point, the next few reductions before $\alpha;\gamma'$ will in general be images of $\beta$ at positions inside $\alpha;\gamma'$.
-The subskeleton at position $\alpha;\gamma'$ before these reductions is $(A|\gamma')[B/x]$, then by the images of $\beta$ this reduces to $(A|\gamma')[B'/x]$, then it reduces at its root position ($\alpha;\gamma'$ in the overall skeleton) to $(A'|\gamma')[B'/x]$.
-For each position of an $x$ in $(A|\gamma')$, there are 0 or more corresponding positions of $x$ in $(A'|\gamma')$, depending on what type of reduction $A \to A'$ is.
-Because $B$ and $B'$ cannot contain any instances of the variable involved in the reduction $A \to A'$ (if it's a $\beta$-reduction or $\tY$-reduction), each instance of $B$ in $(A|\gamma')[B/x]$ has the same set of images in $(A'|\gamma')[B/x]$ as the corresponding $x$ in $(A|\gamma')$.
-When these images of $\beta$ that overlap with the image of $\gamma$ are swapped below it, each of them has its own images, one at each position of $B$ in $(A'|\gamma')[B/x]$ corresponding to its original copy of $B$ in $(A|\gamma')[B/x]$.
-After all of them are swapped below $\alpha;\gamma'$ (and rearranged among themselved by case a), there is therefore one reduction at each copy of $B$ in $(A'|\gamma')[B/x]$, i.e.~one at position $\beta'$ relative to each $x$ in $(A'|\gamma')$.
-They are therefore the same as the images of $\beta$ in $X_1$ that overlap with $\alpha;\gamma'$, because those are also one reduction at position $\beta'$ relative to each $x$ in $(A'|\gamma)$.
-Both $(X_i)_i$ and $(Y_j)_j$ therefore eventually reach a point where the relevant sub-reduction-sequence is $\alpha$, the images of $\beta$ to the left of $\alpha;\gamma'$, $\alpha;\gamma'$ itself, the aforementioned images of $\beta$ that overlap with $\alpha;\gamma'$, then all the images of $\beta$ to the right of $\alpha;\gamma'$.
-\item $Y \sim_c X$ forwards, $\gamma$ and $\beta$ are disjoint and both $> \alpha$, and the reduction at $\alpha$ is a $\tY$-reduction: This is similar to the previous case, but more complicated, so it will need some definitions to explain properly what's going on.
-Let the term at $\alpha$ before any of the reductions be $\tY \lambda x.
-A$, let $\gamma = \alpha;\tY;\lambda;\gamma'$, let $\beta = \alpha;\tY;\lambda;\beta'$, let $A | \beta' = B \to B'$ with the reduction at $\cdot$, let $A | \gamma' = C \to C'$ with the reduction at $\cdot$, let the positions in $A$ where $x$ occurs that are left of $\gamma'$, between $\gamma'$ and $\beta'$ (but still disjoint from both), and right of $\beta'$ respectively be $(\delta^l_i)$, $(\delta^m_i)$ and $(\delta^r_i)$, let the positions in $B$, $C$, $B'$ and $C'$ respectively where $x$ occurs be $(\delta^B_i)$, $(\delta^C_i)$, $(\delta^{B'}_i)$ and $(\delta^{C'}_i)$ (so that all the positions in $A$ where $x$ occurs, in left-to-right order, are $(\delta^l_i)_i, (\delta^C_i)_i, (\delta^m_i)_i, (\delta^B_i)_i, (\delta^r_i)_i$).
-The sequence $(X_i)_i$ starts by swapping $\beta$ then $\gamma$ below $\alpha$.
-With the shorthand that $\gamma'(\epsilon) = \alpha;\lambda;@_1;\epsilon;\tY;\lambda;\gamma'$, $\gamma'_0 = \alpha;\lambda;@_1;\gamma'$, and similarly for $\beta'$, the relevant portion of $X_2$ is then $\alpha$, $(\gamma'(\delta^l_i))_i$, $\gamma'_0$, $(\gamma'(\delta^{C'}_i))_i$, $(\gamma'(\delta^m_i))_i$, $(\gamma'(\delta^B_i))_i$, $(\gamma'(\delta^r_i))_i$, $(\beta'(\delta^l_i))_i$, $(\beta'(\delta^{C'}_i))_i$, $(\beta'(\delta^m_i))_i$, $\beta'_0$, $(\beta'(\delta^{B'}_i))_i$, $(\beta'(\delta^r_i))_i$.
-Next in $(X_k)_k$, for each $i$, $\gamma'(\delta^r_i)$ are swapped past all the images of $\beta$ until it is immediately before $\beta'(\delta^r_i)$ by case a, then for each $i$, $\gamma'(\delta^B_i)$ is swapped past all the images of $\beta$ until $\beta'_0$ by case a, then swapped with $\beta'_0$ by one of the other cases depending on what type of reduction $B \to B'$ is and the relative positions.
-Even if it's case d or f, all of the images of $\gamma'(\delta^B_i)$ are disjoint from each other (and from the positions of the other reductions after $\beta'_0$) because the redex, $C$, doesn't contain any instances of the variable involved in the reduction $B \to B'$.
-Expanding the definitions, this is swapping $\alpha;\lambda;@_1;\delta^B_i;\tY;\lambda;\gamma'$ with $\alpha;\lambda;@_1;\beta'$ at a point in the reduction sequence where the subskeleton at $\alpha;\lambda;@_1;\beta'$ is $B$ with a mixture of $\tY \lambda x.
-A$ and $\tY \lambda x.
-A[C'/\gamma']$ substituted for its occurrences of $x$, and one of these occurrences of $x$ is at $\delta^B_i$.
-For each $\delta^B_j$, there are some corresponding $\delta^{B'}_k$, where the images of the $x$ at $\delta^B_j$ end up after the reduction $B \to B'$.
-These cover all the $\delta^{B'}_k$, and uniquely, and for each $\delta^B_i$, the images of $\zeta;\delta^B_i;\theta$ after swapping it with $\zeta;\beta'$ are precisely $(\zeta;\delta^{B'}_k;\theta)$ for those same values of $k$, therefore after swapping all of the $(\gamma'(\delta^B_i))_i$ down past $\beta'_0$, their images are $(\gamma'(\delta^{B'}_i))_i$.
-After swapping with $\beta'_0$, each of these images than swaps by case a to take its place among the $(\beta'(\delta^{B'}_i))_i$.
-Next up in $(X_i)_i$, the $(\gamma'(\delta^m_i))_i$s and the $(\gamma'(\delta^{C'}_i))_i$s swap down past some images of $\beta$ they're disjoint from to take their place before their matching image of $\beta$, then $\gamma'_0$ swaps past the $(\beta'(\delta^l_i))_i$, stopping immediately before $\beta'(\delta^{C'}_1)$, then the $(\gamma'(\delta^l_i))_i$ and the $(\beta'(\delta^l_i))_i$ mix.
-The end result of the rearrangement of (this subsequence of) the reduction sequence is therefore $\alpha$, $(\gamma'(\delta^l_i), \beta'(\delta^l_i))_i$, $\gamma'_0$, $(\gamma'(\delta^{C'}_i), \beta'(\delta^{C'}_i))_i$, $(\gamma'(\delta^m_i), \beta'(\delta^m_i))_i$, $\beta'_0$, $(\gamma'(\delta^{B'}_i), \beta'(\delta^{B'}_i))_i$, $(\gamma'(\delta^r_i), \beta'(\delta^r_i))_i$.
-
-  The other sequence, $(Y_j)_j$, proceeds similarly, but with the images of $\gamma$ starting out after the images of $\beta$.
-It swaps all the images of $\beta$ that are right of $\gamma_0$ to the appropriate places by case a, then swaps all the $(\beta'(\delta^C_i)_i$ first to then past $\gamma'_0$, resulting in $(\beta'(\delta^{C'}_i))_i$ for the same reason as with swapping those images of $\gamma$ that overlapped with $\beta'_0$ past it in the $(X_i)_i$ case above, then finally swapping the $(\beta'(\delta^l_i))_i$s and the $(\gamma'(\delta^l_i))_i$s into the correct order.
-As required, this produces the same result as in $(X_i)_i$.
-  \item $Y \to X$ forwards, and $\alpha < \gamma < \beta$: Let $\gamma'$ be such that $\gamma = \alpha;\textsf{if}_i;\gamma'$, $\alpha;@_1;\lambda;\gamma'$, $\alpha;@_2;\gamma'$ or $\alpha;\tY;\lambda;\gamma'$, so that $\gamma'$ is the freely varying later part of $\gamma$ as in the definition of $\sim_c$, whichever case applies to swapping $\gamma$ and $\alpha$.
-Similarly, let $\beta'$ be the freely varying part of $\beta$ relative to $\gamma$.
-Let the images of $\gamma$ after swapping it with $\alpha$ (where this swap occurs later in the reduction sequence than $\beta$) then be $(\alpha;\delta_i;\gamma')_i$, and the images of $\beta$ after swapping it with $\gamma$ be $(\gamma;\epsilon_i;\beta')_i$.
-Depending on the skeleton at $\alpha$, and $\gamma$'s position within it, $(\delta_i)_i$ may be empty (case b), a singleton containing $\cdot$ (cases c and d), all the positions of the relevant variable in the lambda in the skeleton at $\alpha$ (case e), or $\lambda;@_1$ and $\lambda;@_1;\zeta;\tY;\lambda$ for each position $\zeta$ of the relevant variable (case f), but in all cases (except a, which is excluded because none of $\alpha, \gamma$ and $\beta$ are disjoint) the set of images has this general structure.
-Furthermore, the positions $(\delta_i)_i$ are determined only by the general position of $\gamma$ within $\alpha$ (the part excluded from $\gamma'$), and the positions of the relevant variable within the skeleton at $\alpha$.
-They do not depend on $\alpha$ or $\gamma$, so that if both initial positions were moved somewhere else the relative positions of the images of $\gamma$ would be unaffected, and similarly if some position within $\gamma$ were swapped with $\alpha$, the relative positions of its images would be the same (except to the extent that, in case f, the positions of the relevant variable are taken after the inner reduction takes place, so that some of those may still change).
-Because there are so many of them and they don't actually affect the multiset of positions where reductions take place, I will be ignoring swaps by case a in the proof of this case, and just assuming that all the final positions end up in the correct order.
-
-  The final set of reduction positions that I will be proving both $(X_i)_i$ and $(Y_j)_j$ reach is $\alpha, (\alpha;\delta_i;\gamma', (\alpha;\delta_i;\gamma';\epsilon_j;\beta')_j)_i$, where as usual the sequences are expanded out to the full list.
-For the sequence $(X_i)_i$, first the images of $\beta$ in $X_0$ are $(\gamma,\epsilon_i,\beta')$.
-For each of these in turn, it is swapped with $\alpha$.
-By the fact mentioned above that sub-positions are mapped to the same set of images except in case f, the images of $\gamma,\epsilon_j,\beta'$ after this swap are $(\alpha;\delta_i^j;\gamma',\epsilon_j,\beta')_i$, where $(\delta_i^j)_i$ is the same as $(\delta_i)_i$ except that in the case where this is not the first image of $\beta$ swapped below $\alpha$ and the reduction at $\alpha$ is a $\tY$-reduction (and therefore swaps with it proceed by case f), the positions of the relevant variable are taken after the reductions at $\gamma$ and $\gamma;\epsilon_k;\beta'$ for $k \leq j$ (this implies that the first of these, $\delta_i^n = \delta_i$).
-All the images $\alpha;\delta_i^j;\gamma';\epsilon_j;\beta'$ where $\delta_i^j > \lambda;@_1;\gamma';\epsilon_{j+1};\beta'$ are then swapped with $\alpha;\lambda;@_1;\gamma';\epsilon_{j+1};\beta'$ (which is $\alpha;\delta_i^{j+1};\gamma';\epsilon_{j+1};\beta'$ for some $i$).
-As in the case where $\alpha;\tY < \gamma, \beta$ and $\gamma$ and $\beta$ are disjoint, the effects of these swaps may duplicate or delete the reductions in such a way that all the remaining images of $\gamma;\epsilon_j;\beta'$ are $(\alpha;\delta_i^{j+1};\gamma';\epsilon_j;\beta'$.
-This is repeated with $\delta_i^{j+2}$ and so on until these images are all in the correct order, at which point they are $(\alpha;\delta_i;\gamma';\epsilon_j;\beta')_i$.
-After all of these are done, $\gamma$ is swapped with $\alpha$, yielding $(\alpha;\delta_i^{-1};\gamma')_i$.
-As with the images of $\beta$, the subset of these which are inside $\alpha;\lambda;@_1;\gamma';\epsilon_j;\beta'$ for each $j$ in turn are swapped with it, until they are all in the correct order and the images left of $\gamma$ are $(\alpha;\delta_i;\gamma')_i$.
-At this point, $X_i$ has the desired value.
-
-  The sequence $(Y_j)_j$ proceeds similarly.
-First $\gamma$ swaps with $\alpha$, yielding $\alpha$ and $(\alpha;\delta_i;\gamma')_i$, then $\beta$ swaps with $\alpha$, yielding one image of $\beta$ for each $\delta_i$, except that, in case f, those images at positions inside $\alpha;\lambda;@_1;\gamma'$ may differ because they are earlier in the reduction sequence than any of the images of $\gamma$.
-Each of the images of $\beta$ in turn it moved to the corresponding image of $\gamma$, except that those overlapping with $\alpha;\lambda;@_1;\gamma'$ may be duplicated or deleted on the way.
-The images of $\beta$ after just this process (which does not actually correspond to any term in $(Y_j)_j$, but it's sufficiently independent that the order doesn't matter that much) are $(\alpha;\delta_i;\gamma';\beta'')$ (where $\gamma;\beta'' = \beta$).
-The images of these for each $i$, after swapping with $\alpha;\delta_i;\gamma'$, is $(\alpha;\delta_i;\gamma';\epsilon_j;\beta')_j$, because the reductions produced by a swap are unaffected by its overall location, and the skeleton at $\alpha;\delta_i;\gamma'$ at the relevant point in the sequence is equal to the skeleton at $\gamma$ initially, therefore this is equivalent to immediately swapping $\beta$ and $\gamma$ as in $Y \sim_c X$ (except that in the case where the reduction at $\alpha$ is a $\tY$-reduction, the skeleton at $\alpha;\lambda;@_1;\gamma'$ is not actually equal to the skeleton initially at $\gamma$ because something was substituted in for the variable bound at $\alpha;\tY$, but this doesn't affect the variable involved in the reduction at $\gamma$, so this doesn't actually matter).
-Therefore also in this case, eventually the set of reductions in the relevant portion of $Y_j$ is $\alpha, (\alpha;\delta_i;\gamma', (\alpha;\delta_i;\gamma';\epsilon_j;\beta')_j)_i$, therefore it is equal to some $X_i$.
-\end{itemize}
-In summary, if the swap in $Y \sim_c X$ doesn't overlap with the swap $X \sim_c X_0$, the sequence of swaps can be rearranged until it does, and in any other case, there is some $Y_j$ that's equal to some $X_i$, therefore these sequences reach the same end point and if $Y \sim_c X$, then $Y_{cbv} = X_{cbv}$, and by chaining these together, if $Y \sim_c^* X$, $Y_{cbv} = X_{cbv}$.
-
-\paragraph{}
-If $Y$ is in CBV order and $Y \sim_c^* X$, then $Y_{cbv} = X_{cbv}$ but also $Y_{cbv} = Y$ because the sequence $(Y_i)_i$ terminates immediately therefore $X_{cbv}$ is the unique reduction sequence in call-by-value order that's related to $X$ by $\sim_c^*$.
-\end{proof}
-
-\begin{lemma}
-\label{lem:sim-M-N}
-The relation $\sim$ is defined on $L_0(M)$ with reference to a particular starting term $M$, so different versions, $\sim_M$ and $\sim_N$, can be defined starting at different terms.
-If $M \to N$, then $\sim^*_N$ is equal to the restriction of $\sim^*_M$ to $L_0(N)$.
-\end{lemma}
-
-\begin{proof}
-$\sim_N^*$ is trivially a subset of $\sim_M^*$ because $\sim_N$ is a subset of $\sim_M$.
-
-In the other direction, suppose $(N_1, \alpha_1) \sim_M^* (N_2, \alpha_2)$ where both $N_1$ and $N_2$ are descendants of $N$.
-By Lemma~\ref{downAcrossUp}, take $(N_1, \alpha_1) \sim_{p,M}^* (N'_1, \alpha') \sim_{c,M}^* (N'_2, \alpha') \sim_{p,M}^* (N_2, \alpha_2)$.
-The $\sim_p^*$ steps remain within $Rch(N)$, and $\sim_p$ does not depend on the history of the reduction sequences, therefore $(N_1, \alpha_1) \sim_{p,N}^* (N'_1, \alpha') \sim_{c,M}^* (N'_2, \alpha') \sim_{p,N}^* (N_2, \alpha_2)$.
-
-Let $X$ be the call-by-value reduction sequence related to both $N'_1$ and $N'_2$ by $\sim_{c,M}$ given by Lemma~\ref{canonicalCousins}. 
-As $M \to N$ is the first reduction in the sequence $N'_1$, it is the last to be affected by the $\sim_{c,M}$ sequence $N'_1 \sim_{c,M}^* X$ given by Lemma \ref{canonicalCousins} therefore it can be split into $N'_1 \sim_{c,M}^* Y_1 \sim_{c,M}^* X$ where $Y_1$ is in CBV order except possibly for its first reduction, which is still $M \to N$.
-Let the position of the reduction $M \to N$ be $\beta$.
-\akr{TODO: fill in the details that are missing here. It shouldn't be too hard, it's just a bit sketchy at present.} 
-The rearrangement of the reduction sequence $Y_1 \sim_{c,M} X$ consists of moving the reduction at $\beta$ down past the other reductions in $Y_1$, possibly duplicating or deleting it in the process, but not affecting the positions or the correct order for any of the other reductions.
-The reductions derived from $M \to N$ can be identified as follows:
-
-%\begin{definition}
-A position in some (reduction sequence of) term(s) is \emph{derived from} another if it is related by the reflexive transitive closure of $\rightsquigarrow$, where $(V,\gamma) \rightsquigarrow (W,\delta)$ just if $V \to W$ and one of the following cases holds:
-\begin{itemize}
-    \item $(V,\gamma) \sim_p (W,\delta)$
-    \item $\gamma = \delta$, $V \to W$ at $\epsilon$, and $\epsilon > \gamma$
-    \item $V \mid \epsilon = (\lambda x.U) Z, U | \zeta = x, \gamma = \epsilon;@_2;\theta$ and $\delta = \epsilon;\zeta;\theta$
-    \item $V \mid \epsilon = \tY (\lambda x.U), \gamma = \epsilon;\tY;\lambda;\zeta$ and $\delta = \epsilon;\lambda;@_1;\zeta$
-    \item $V \mid \epsilon = \tY (\lambda x.U), U | \zeta = x, \gamma = \epsilon;\theta$ and $\delta = \epsilon;\lambda;@_1;\zeta;\theta$
-\end{itemize}
-%\end{definition}
-
-Crucially, the reductions in $X$ can be partitioned into 2 sets: those at positions derived from $(M,\beta)$, and those at positions equal to the reductions in $Y_1$ (in the same order as they occur in $Y_1$).
-Using the same construction for $N'_2$ shows that the positions of the reductions in $Y_2$ are also the positions of the reductions in $X$ other than those derived from $(M,\beta)$, therefore $Y_1 = Y_2$ therefore $N'_1 \sim_{c,M} Y_1 \sim_{c,M} N'_2$ and every reduction sequence in this sequence starts with $M \to N$ therefore $N'_1 \sim_{c,N} N'_2$.
-
-Combining this with the $\sim_{p,N}^*$s at the beginning and end then yields the desired result that $(N_1,\alpha_1) \sim_N^* (N_2, \alpha_2)$, therefore the restriction of $\sim_M^*$ to $\sim_N$'s domain ($Rch(N)$) is a subset of $\sim_N^*$ therefore the two versions of $\sim$ match on this domain, as desired.
-\end{proof}
-
-At each reduction step $M \to N$, the sample space must be restricted from $I^{L_s(M)}$ to $I^{L_s(N)}$. 
-The injection $L_0(N) \to L_0(M)$ is trivial to define by appending $Sk(M) \to Sk(N)$ to each path, and using Lemma~\ref{lem:sim-M-N}, this induces a corresponding injection on the quotient, $L(N) \to L(M)$.
-The corresponding map $L_s(N) \to L_s(M)$ is then denoted $i(M \to N)$.
-
-Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function, so it is denoted ``$\Rightarrow$'' instead of ``$\red$'', and it relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
-\changed[lo]{We write an element of $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ as $(M', s)$ where the \emph{term} $M' \in \Lambda^0$ and $s \in  I^{L_s(M')}$.}
-\begin{align*}
-& (M,s) \Rightarrow (N,s \circ i(M \to N)) \text{ if either} \\
-& \qquad \text{$M \to N$ and the redex is not $\tsample$, or} \\
-& \qquad \text{$M \mid \alpha = \tsample$, $N = M[\underline{s(Sk(M),\alpha)}/\alpha]$ and $\lambda$ does not occur after $@_2$ or $\tY$ in $\alpha$}
-\end{align*}
-
-\paragraph{}
-\begin{lemma} \label{churchRosser}
-The relation $\Rightarrow$ is Church-Rosser.
-\end{lemma}
-\begin{proof}
-Suppose that $(M,s) \Rightarrow^* (M_1,s_1)$ and also $(M,s) \Rightarrow^* (M_2,s_2)$, then it is required to prove that there is some $(M',s')$ such that both $(M_1,s_1) \Rightarrow^* (M',s')$ and $(M_2,s_2) \Rightarrow^* (M',s')$.
-First consider the special case where $(M,s) \Rightarrow (M_1,s_1)$ and $(M,s) \Rightarrow (M_2,s_2)$, with only a single step in each case.
-Let the positions of the redexes in $M \to M_1$ and $M \to M_2$ be $\alpha_1$ and $\alpha_2$ respectively.
-
-First consider the case that $\alpha_1$ and $\alpha_2$ are disjoint.
-Let $M_1 = M[X_1/\alpha_1]$ and similarly for $X_2$, then let $M' = M[X_1/\alpha_1][X_2/\alpha_2]$.
-As the positions are disjoint, the substitutions commute and both $M_1$ and $M_2$ reduce (with $\to$) to $M'$.
-Let $s' = s \circ i(M \to M_1) \circ i(M_1 \to M')$. 
-The injection $i(M \to M_1) \circ i(M_1 \to M')$ consists of prepending $M \to M_1 \to M'$ to each reduction sequence, but by case a of $\sim_c$, this is equivalent to prepending $M \to M_2 \to M'$, which is $i(M \to M_2) \circ i(M_2 \to M')$.
-In the case that the reduction $M \to M_2$ isn't a $\tsample$-reduction, this is enough to establish that $(M_1, s_1) \Rightarrow (M', s')$ (and similarly if the redex of $M \to M_1$ isn't $\tsample$, $(M_2,s_2) \Rightarrow (M',s')$).
-If it is $\tsample$ though, in order for it to be the case that $(M_1, s_1) \Rightarrow (M', s')$, it is additionally necessary that $X_2$, the result of the reduction at $\alpha_2$, be $s_1(M_1, \alpha_2)$, which follows from the fact that $(M, \alpha_2) \sim (M_1, \alpha_2)$ by case 1 of $\sim_p$.
-The case that the reduction $M \to M_1$ is a $\tsample$-reduction is similar.
-
-The case that $\alpha_1 = \alpha_2$ is trivial, because there is at most one possible $\Rightarrow$ reduction at any given position, therefore $(M_1, s_1) = (M_2, s_2)$ already.
-
-The remaining case is that $\alpha_1 < \alpha_2$ or $\alpha_1 > \alpha_2$. 
-Assume without loss of generality that $\alpha_1 < \alpha_2$. 
-For each possible case of what type of redex $M | \alpha_1$ is, and $\alpha_2$'s position within it, there is a corresponding case of $\sim_c$, and similarly to the case where $\alpha_1$ and $\alpha_2$ are disjoint, the term $O_1 = O_2$ from the definition of $\sim_c$ is a suitable value of $M'$.
-The term $M \mid \alpha_1$ can't be $\tsample$, because it has strict subterms, but the case that $M \mid \alpha_2 = \tsample$ is still somewhat more complicated. 
-$\alpha_2$ can't be within $\alpha_1 ; @_2$ or $\alpha_1 ; Y$ (cases e and f of $\sim_c$) because those are not valid positions to reduce a $\tsample$, 
-and the case that $\alpha_2$ is within the branch of an if statement that is deleted by the reduction at $\alpha_1$ (case b) doesn't actually present a problem because there is no corresponding reduction to $M \to M_2$ in the other branch,
-which leaves cases c and d, that $\alpha_2$ is in the $\tif{\cdot}{\cdot}{\cdot}$ branch that isn't deleted, and that $\alpha_2 < \alpha_1 ; @_1 ; \lambda$.  
-The values that the $\tsample$s take in these cases match because $(M,\alpha) \sim_p (M_1, \alpha;\beta)$ by cases 3 and 2 of $\sim_p$ respectively.
-
-\paragraph{}
-In the case that $(M,s) \Rightarrow^* (M_1,s_1)$ or $(M,s) \Rightarrow^* (M_2,s_2)$ by multiple steps, it is possible to repeatedly replace a pair of the form $A \Leftarrow B \Rightarrow C$ by $A \Rightarrow^* D \Leftarrow^* C$ by the construction above, but it is not immediate that this process terminates, because each $\Rightarrow$ or $\Leftarrow$ may be replaced by multiple, so the sequence of $\Rightarrow$s and $\Leftarrow$s from $(M_1,s_1)$ to $(M_2,s_2)$ may get longer at some steps.
-However, termination can be proved by noting that the structure of this process is identical to the process of swapping $\sim_\uparrow$ and $\sim_\downarrow$ steps in Lemma \ref{downAcrossUp}.
-To be more precise, in this case there is a sequence of $\Leftarrow$ and $\Rightarrow$ steps, each of which has an associated reduction position and initial and final terms, and if a $\Leftarrow$ immediately precedes a $\Rightarrow$, they may be swapped to produce some number of $\Rightarrow$s, followed by some number of $\Leftarrow$s.
-In the case of Lemma \ref{downAcrossUp}, there is a sequence of $\sim_\uparrow$ and $\sim_\downarrow$ steps, each of which has an associated reduction position and initial and final skeletons, and if a $\sim_\uparrow$ immediately precedes a $\sim_\downarrow$, they may be swapped to produce some number of $\sim_\downarrow$s followed by some number of $\sim_\uparrow$s.
-In both cases, the number and reduction positions of the resultant steps are determined by the case of $\sim_c$ that matches the way the initial reduction positions overlap, and the initial skeleton (or the skeleton of the initial term).
-The same argument that the process in Lemma \ref{downAcrossUp} terminated is therefore applicable here too.
-At every stage, the sequence of reductions that results from one of the initial $\Rightarrow$s is a parallel sequence, and swapping a parallel sequence of $\Rightarrow$s with a $\Leftarrow$ always terminates, therefore each $\Rightarrow$ in turn can be moved past all of the $\Leftarrow$s, and the process as a whole will terminate in a state where all of the $\Rightarrow$s precede all of the $\Leftarrow$s, i.e.~a pair of reduction sequences $(M_1,s_1) \Rightarrow^* (M',s') \Leftarrow^* (M_2,s_2)$.
-\end{proof}
-
-\begin{lemma} \label{simpDoesn'tMix}
-If $A$ is some descendant of $M$ and $(A,\gamma) \sim^* (M,\delta)$, then $(A,\gamma) \sim_p^* (M,\delta)$, with the length of the reduction sequences decreasing by one each step from $A$ to $M$.
-\end{lemma}
-\begin{proof}
-This is a simple induction on $\sim^*$.
-In the base case, $(A,\gamma) = (M,\delta)$ therefore $(A,\gamma) \sim_p^* (M,\delta)$ trivially.
-Otherwise, suppose that $(M,\delta) \sim_p^* (B,\epsilon) \sim (A,\gamma)$.
-Either $(B,\epsilon) \sim_p (A,\gamma)$ or $(B,\epsilon) \sim_c (A,\gamma)$.
-In the first case, either $B \to A$, in which case the result follows directly, or $A \to B$, in which case the fact that each (reduction sequence of) skeleton(s) has only one parent implies that a $(M,\delta) \sim_p^* (A,\gamma)$ directly as a subsequence of the path to $(B,\epsilon)$.
-
-In the $\sim_c$ case, consider the definition of $\sim_c$.
-Either $O'_1 = A$ and $O'_2 = B$ or vice-versa.
-As $M \to^* N \to^* O_1 \to^* B$ and $(M,\delta) \sim_p^* (B,\epsilon)$, and $\sim_p$ only relates positions in a term and its parent, there are some positions $\zeta, \theta$ such that $(M,\delta) \sim_p^* (N,\zeta) \sim_p^* (O_1, \theta) \sim_p^* (B,\epsilon)$.
-It follows that $(O_2, \theta) \sim_p^* (A,\gamma)$ by following the same path, therefore it suffices to provide the only missing portion of the path from $M$ to $A$, i.e.~to prove that $(N,\zeta) \sim_p^* (O_2,\theta)$ given $(N,\zeta) \sim_p^* (O_1,\theta)$ (or vice-versa).
-
-If $\zeta$ is disjoint from all the positions of reduction from $N$ to $O_1$ and $O_2$ (and consequently $\zeta = \theta$), this follows from case 1 of $\sim_p$.
-Otherwise, this can be proved by taking cases from the definition of $\sim_c$.
-This is rather long, but all of the cases are similar.
-The general idea is that the reductions from $N$ to $O_1$ correspond to the reductions from $N$ to $O_2$, so that if a position is related by $\sim_p$ across that reduction, it is related in the other branch for the same reason.
-Case d, where $B = O'_1$ rather than $O'_2$, is given here in more detail as an illustrative example:
-
-Let $I$ be $N$ reduced at $\alpha$, and $J$ be $N$ reduced at $\alpha;@_1;\lambda;\beta$, so that $N \to I \to O_1$ and $N \to J \to O_2$.
-All of the reduction positions are $\geq \alpha$, and $\zeta$ is not disjoint from all of them, therefore $\zeta$ is not disjoint from $\alpha$.
-Let $\iota$ be the position such that $(N,\zeta) \sim_p (I,\iota) \sim_p (O_1,\theta)$.
-The fact that $(N,\zeta) \sim_p (I,\iota)$ implies that $\zeta > \alpha;@_1;\lambda$.
-Let $\zeta = \alpha;@_1;\lambda;\kappa$ and $\theta = \alpha;\kappa'$.
-If $\kappa$ is disjoint from $\beta$, then $(N,\zeta) \sim_p (J,\zeta) \sim_p (O_2,\alpha;\kappa') = (O_2;\theta)$ by cases 1 and 2 of $\sim_p$.
-In the other case, that $\kappa$ isn't disjoint from $\beta$, $\kappa > \beta$ because none of the positions $<= \alpha;@_1;\lambda;\beta$ in $N$ are related to any position in $I$ by $\sim_p$.
-As $(I,\alpha;\kappa) \sim_p (O_1,\alpha;\kappa')$ (with the redex at $\alpha;\beta$), for exactly the same reason $(N,\alpha;@_1;\lambda;\kappa) \sim_p (J,\alpha;@_1;\lambda;\kappa')$ (with the redex at $\alpha;@_1;\lambda;\beta$).
-Because $(N,\alpha;@_1;\lambda;\kappa) \sim_p (J,\alpha;@_1;\lambda;\kappa')$, $N|\alpha;@_1;\lambda;\kappa = J|\alpha;@_1;\lambda;\kappa'$, and $N|\alpha;@_1;\lambda;\kappa \neq $ the variable of $N|\alpha;@_1$, therefore $J|\alpha;@_1;\lambda;\kappa'$ is also not the variable therefore $(J,\alpha;@_1;\lambda;\kappa') \sim_p (O_2, \alpha;\kappa') = (O_2, \theta)$ by case 2 of $\sim_p$.
-Combining these results, $(N,\zeta) \sim_p^* (O_2;\theta)$ as desired.
-\end{proof}
-
-\begin{lemma} \label{lem:redexDestroyed}
-If $M \to N$, with the redex at position $\alpha$, then no position in any term reachable from $N$ is related by $\sim^*$ to $(M,\alpha)$.
-\end{lemma}
-\begin{proof}
-Suppose on the contrary that $(M,\alpha) \sim^* (A,\beta)$, where $N \to^* A$, then by Lemma \ref{simpDoesn'tMix}, $(M,\alpha) \sim_p^* (A,\beta)$.
-$M \neq A$ therefore $(M,\alpha) \sim_p$ some position in $N$, but in all the cases of the definition of $\sim_p$, no position in the child term is related to the position of the redex.
-\end{proof}
-
-\paragraph{}
-The reduction relation $\Rightarrow$ is nondeterministic, so it admits multiple possible reduction strategies. 
-A \emph{reduction strategy} starting from a closed term $M$ is a partial function $f$ from $Rch(M)$ to positions, such that for any reachable term $N$ where $f$ is defined, $f(N)$ is a position of a redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
-Using a reduction strategy $f$, a subset of $\Rightarrow$ that isn't nondeterministic, $\Rightarrow_f$, can be defined by $(N,s) \Rightarrow_f (N',s')$ just if $(N,s) \Rightarrow (N',s')$ and $N$ reduces to $N'$ with the redex at $f(N)$.
-
-Essentially what Lemma~\ref{lem:redexDestroyed} demonstrates is that the samples taken during any reduction sequence are independent of each other. This is made more precise in the following lemmas.
-
-\begin{lemma}
-For any skeletons $M \to N$, with the redex at position $\alpha$, and measurable set of samples $S \subset I^{L_s(N)}$, $\mu(S) = \mu(\{s \in I^{L_s(M)} \mid s \circ i(M \to N) \in S\})$, and furthermore, if $M | \alpha = \tsample$, for any $S \subset I \times I^{L_s(N)}$, $\mu(S) = \mu(\{s \in I^{L_s(M)} \mid (s(M,\alpha), s \circ i(M \to N)) \in S\})$.
-\end{lemma}
-\begin{proof}
-If the result holds for all sets $S$ of the form $\{s \in I^{L_s(N)} \mid \forall j : s(j) \in x_i\}$, where $(x_j)_{j \in J}$ is a family of measurable subsets of $I$ indexed by some finite set $J \subset L_s(N)$ of positions, it also holds for all other $S$ by taking limits and disjoint unions.
-
-Take such a $(x_j)_{j \in J}$, where $x_j = x_k$ and define $K = i(M \to N)[J]$.
-Because $i(M \to N)$ is injective, it defines a bijection between $J$ and $K$.
-We can then calculate the measure $\mu(\{s \in I^{L_s(M)} \mid s \circ i(M \to N) \in S\}) = \mu(\{s \in I^{L_s(M)} \mid \forall k : s(k) \in x_{i(M \to N)^{-1}(k)}\}) = \prod_{k \in K} \mu_I(x_{i(M \to N)^{-1}(k)}) = \prod_{j \in J} \mu_I(x_j) = \mu(S)$.
-
-In the case that $M|\alpha = \tsample$, we can similarly consider only those sets $S$ of the form $x_\tsample \times \{s \in I^{L_s(N)} \mid \forall j : s(j) \in x_i\}$.
-The position $\alpha$ in $M$ is not related to any position in $L_s(N)$, therefore $(M,\alpha) \not \in K$.
-Again, we can calculate the measure $\mu(\{s \in I^{L_s(M)} \mid (s(M,\alpha), s \circ i(M \to N)) \in S\}) = \mu(\{s \in I^{L_s(M)} \mid s(M,\alpha) \in x_\tsample, \forall k : s(k) \in x_{i(M \to N)^{-1}(k)}\}) = \mu_I(x_\tsample) \prod_{k \in K} \mu_I(x_{i(M \to N)^{-1}(k)}) = \mu_I(x_\tsample) \prod_{j \in J} \mu_I(x_j) = \mu(S)$.
-\end{proof}
-
-\begin{lemma} \label{lem:independentSamples}
-For any initial term $M$, reduction strategy $f$ on $M$, natural number $n$, skeleton $N$ with $k$ holes, measurable set $T \subset \mathbb R ^ k$ and measurable set $S$ of samples in $I^{L_s(N)}$, $\mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')\}) = \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in I^{L_s(N)} : (M,s) \Rightarrow_f^n (N[r], s')\}) \mu(S)$
-\end{lemma}
-\begin{proof}
-Suppose, to begin with, that $n = 0$.
-Either $\exists r \in T : N[r] = M$, in which case both sides of the equation are $\mu(S)$, or there is no such $r$, in which case both sides of the equation are 0.
-
-For $n > 0$, suppose for induction that the lemma is true for $n - 1$, for all $N, T$ and $S$.
-If $\exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')$, the $r$ and $s'$ are necessarily unique, therefore we may assume that $T$ is the product of $k$ measurable setsets of $\mathbb R$, $(T_j)_{0 \leq j < k}$, as all of the other cases follow by taking unions and limits of these rectangles.
-Let\begin{align*}
-R_s & = \{\text{the skeleton of } O \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) = \tsample, O \to N[r]\} \\
-R_d & = \{\text{the skeleton of } O \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) \neq \tsample, O \to N[r]\}.
-\end{align*}
-Together, $R_s$ and $R_d$ contain all of the skeletons of terms that can reduce to $N$, and they're disjoint.
-For each of these skeletons, there is a map $g_O$ from the reals in $O$ to the reals in $N$, i.e.
-$g_O(r) = r'$ if $O[r] \to N[r']$ for $O \in R_d$, and $g_O(s, r) = r'$ if $O[r] \to N[r']$ with $s$ as the value the sample takes in the reduction.
-All of these functions are measurable, as they are just rearrangements of vector components, or in the case that $O|f(O) = \underline x(\skeletonPlaceholder, \dots, \skeletonPlaceholder)$, a combination fo the measurable function $x$ with a rearrangement of vector components.
-In the case that $O \in R_s$, the function $g_O$ simply inserts the sample into the vector of reals at a certain index, call it $h_O$.
-\begin{align*}
-& \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')\}) \\
-= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in S :\\
-    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\})\\
-    + & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in S :\\
-    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \\
-\end{align*}
-\begin{align*}
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}:\\
-    & g_O(r,s'(O,f(O))) \in T, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
-     + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}:\\
-     & g_O(r) \in T, s' \circ i(O[r] \to N[g_O(r)]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)}:\\
-    & s'(O,f(O)) \in T_{h(O)}, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
-    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}:\\
-    & g_O(r) \in T, s' \circ i(O[r] \to N[g_O(r)]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
-    & \mu(\{s' \in I^{L_s(O)} \mid s'(O,f(O)) \in T_{h(O)}, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S\})\\
-    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
-    & \mu({s' \in I^{L_s(O)} \mid s' \circ i(O[r] \to N[g_O(r)]) \in S}) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(T_{h(O)}) \mu(S)\\
-    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : s'(O,f(O)) \in T_{h(O)}, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S)\\
-    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \mathbb R^{k-1}, s' \in I^{L_s(O)} : g_O(r,s'(O,f(O))) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S)\\
-    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
-= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} :\\
-    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S)\\
-    + & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} :\\
-    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S) \\
-= & \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in I^{L_s(N)} : (M,s) \Rightarrow_f^n (N[r], s')\}) \mu(S)
-\end{align*}
-\end{proof}
-
-The usual call-by-value semantics can be implemented as one of these reduction strategies, given by (with $V$ a value and $T$ a term that isn't a value and $M$ a general term)
-\begin{align*}
-\cbv(T M) & = @_1 ; \cbv(T) \\
-\cbv(V T) & = @_2 ; \cbv(T) \\
-\cbv(\underline f(V_1, \dots, V_{k-1}, T, M_{k+1}, \dots, M_n)) & = \underline f_k ; \cbv(T) \\
-\cbv(\tY T) & = \tY ; \cbv(T) \\
-\cbv(\tif{T < 0}{M_1}{M_2}) & = \textsf{if}_1 ; \cbv(T) \\
-\cbv(\tscore(T)) & = \tscore ; \cbv(T) \\
-\cbv(V) & \text{ is undefined} \\
-\cbv(T) & = \cdot \text{ otherwise}
-\end{align*}
-(this last case covers redexes at the root position).
-
-A closed term $M$ terminates with a given reduction strategy $f$ and samples $s$ if there is some natural number $n$ such that $(M,s) \Rightarrow_f^n (N,s')$ where $f$ gives no reduction at $N$.
-The term terminates almost surely with respect to $f$ if it terminates with $f$ for almost all $s$.
-
-\begin{theorem} \label{thm:AstEquivalence}
-A closed term $M$ is AST with respect to $\cbv$ iff it is AST.
-\end{theorem}
-\begin{proof}
-\akr{I'm not sure whether this proof is sufficiently detailed. It would be pretty easy to fill it in more, just very verbose.}
-
-For both $\red$ and $\Rightarrow_{\cbv}$, the probability measure on the sample space can be used to define a measure on the reduction sequences of each finite length.
-The distributions are equal for each length, by induction, as follows.
-
-The base case, length 0, is trivial, because in both cases the distribution has probability 1 on the sequence containing only $M$.
-
-For the inductive case, suppose that the distributions of reduction sequences of length $n$ are equal.
-\akr{TODO: mention the case that it has already terminated by step $n$, or is about to.
-The two versions of the semantics handle it a little differently.}
-
-For any term $N$ which is not a value, there is a unique environment $E$ and redex $R$ such that $N = E[R]$.
-The position of the hole in $E$ is equal to $\cbv(N)$, so that $N|\cbv(N) = R$, as the cases in the definition of $\cbv$ match the cases in the definition of environments.
-If $R \neq \tsample$, let $R'$ be the result of reducing $R$ (which is equal in both versions of the semantics), then $E[R'] = N[R' / \cbv(N)]$ therefore $\red(N,s) = (E[R'],s)$ and $(N,s) \Rightarrow_{\cbv} (E[R'],s)$.
-The distribution of next terms, conditional on the previous term, in the case that that previous term reduces deterministically, is therefore equal for $\red$ and $\Rightarrow_{\cbv}$.
-
-In the other case that $R = \tsample$, the first (term) part of $\red(N,s)$ is $E[\underline{\pi_h(s)}]$.
-If $k$ is the number of samples already taken in this reduction sequence, this is equivalent to $E[\underline{s_0(k)}]$, where $s_0$ is the initial sample.
-For any sufficiently small neighbourhood of $N$ (i.e.~containing only reduction sequences with the same skeletons), the probability of reaching this neighbourhood is independent of $s(k)$, because it depends only on the samples taken so far, and the samples are independent.
-The distribution of $s(k)$ conditional on reaching $N$ is therefore the uniform distribution on $I$.
-Similarly for the other semantics, if the distribution of remaining samples after $n$ steps is independent of the term by Lemma \ref{lem:independentSamples}, therefore the distribution of $s(Sk(N),\cbv(N))$ conditional on $N$ is the uniform distribution on $I$.
-In either case, the distribution of the next term conditional on the previous term, in the case that it reduces randomly, is equal to the image under $r \mapsto E[\underline r]$ of the uniform distribution on $I$.
-
-In any case, the distribution of reduction sequences after $n+1$ steps, conditional on the reduction sequence after $n$ steps, as defined by $\Rightarrow_{\cbv}$ and by $\red$, is equal, any by the inductive hypothesis the distributions after $n$ steps are equal, therefore by integrating over the reduction sequence up to step $n$, the distributions on reduction sequences up to step $n+1$ are equal.
-
-The distributions on reduction sequences of any finite length as defined by $\Rightarrow_{\cbv}$ and $\red$ are therefore equal, therefore so is the probability of having reached a value after $n$ steps, therefore so is its limit, the probability of termination, therefore the probability of termination is 1 iff the probability of termination with respect to $\cbv$ is 1.
-\end{proof}
-
-\begin{theorem} \label{thm:CbvIsTerminatingest}
-If $M$ terminates with some reduction strategy $f$ and samples $s$, it terminates with $\cbv$ and $s$.
-\end{theorem}
-\begin{proof}
-  Suppose $M$ terminates with the reduction strategy $f$ and samples $s$.
-Let $(M,s) = (M_0,s_0) \Rightarrow_f \dots \Rightarrow_f (M_n,s_n)$, where $M_n$ is a value.
-By Lemma \ref{canonicalCousins}, the reduction sequence $(Sk(M_i))_i$ is related by $\sim_c^*$ to some skeletal reduction sequence $X$ which is in call-by-value order.
-For each skeletal reduction sequence $X_j$ in the sequence $(Sk(M_i))_i \sim_c \dots \sim_c X$, it is possible to define a reduction sequence of terms such that $(M,s) = (N_{j,0},s_{j,0}) \Rightarrow \dots \Rightarrow (N_{j,n_j},s_{j,n_j})$ and $Sk(N_{j,k}) = X_{j,k}$, by at each step applying the reduction at the same place as in $X_j$, essentially just filling in the reals in the skeletal reduction sequence to make it a reduction sequence of terms (it will be shown shortly that the correct branch is taken in each $\textsf{if}$-reduction).
-Assume for induction on $j$ that $(N_{j,n_j},s_{j,n_j}) = (M_n,s_n)$.
-To prove that this holds true for $j+1$ as well, it suffices to show that the corresponding term and samples immediately after the reductions involved in this $\sim_c$ step match, as from that point on, the $\Rightarrow$-sequences will match, as they start from the same point and have all the same reduction positions.
-Letting the term and remaining samples immediately before the reductions involved in this $\sim_c$ step be $(N,t)$ (which is the same in both $N_j$ and $N_{j+1}$ for the same reason again, they start at the same point and the sequences of reduction positions up to this point are the same), and the terms and samples after these reductions be $(O_1,t_1)$ and $(O_2,t_2)$, so that these match the $N$, $O_1$ and $O_2$ in the definition of $\sim_c$, $t_1 = t_2$ because both of them are, by Lemma \ref{lem:sim-M-N}, equal to the restriction of $t$ to the subset of $Rch(N)$ that starts with the reduction sequences to $O_1$ and $O_2$ respectively, and the potential positions in those subsets of $Rch(N)$ are identified with each other by $\sim$ because of the same case of $\sim_c$ that means $N_j$ and $N_{j+1}$ are related.
-If none of the relevant reductions are $\tsample$-reductions, $O_1 = O_2$ trivially.
-If there are samples taken (which must be the arbitrary reduction at $\beta$), it is also required that the same samples are taken in both branches.
-Similarly to Lemma \ref{churchRosser}, the positions of these $\tsample$-reductions in each branch are related by $\sim_p^*$, therefore they correspond to the same sample in $t$, therefore $O_1 = O_2$.
-The fact that the terms are equal also implies that the same branch of an $\textsf{if}$ statements will be taken.
-
-  \paragraph{}
-  There is therefore a reduction sequence $(M,s) = (N_0,s_0') \Rightarrow \dots \Rightarrow (N_m, s_m')$ which ends in a value ($N_m = M_n$) and is in CBV order.
-It doesn't immediately follow that this is the $\Rightarrow_{\cbv}$ reduction sequence starting from $(M,s)$ because the fact that $(N_i)_i$ is in CBV order only means that the reductions that do take place in this reduction sequence are in the correct order, not that the reduction sequence finishes once it reaches a value, or that there are no other reductions that would come earlier in CBV order that don't occur at all in $(N_i)_i$.
-Suppose that at some point $j$ in the reduction sequence, the reduction that occurs is not the next reduction in CBV order, but something else, and that the reduction that should be next is at position $\alpha = \cbv(N_i)$, and the position of the actual reduction $N_j \to N_{j+1}$ is $\beta$.
-Suppose further that $N_j$ is not a value.
-Either $\alpha = \cdot$, the root position, or the reduction at the root position is dependent in some way on the reduction at $\alpha$ (e.g.
-$\alpha = \textsf{if}_1$, or $\alpha = \underline{+};@_1$).
-In any case, all the reductions in the rest of the reduction sequence are after $\alpha$ in CBV order, therefore none of them are $\leq \alpha$, and there continues to be an unreduced redex at $\alpha$, therefore the reduction at the root position never occurs, therefore $N_m$ is not a value, which is a contradiction, therefore at the point (if any) where $(N_i)_i$ departs from the $\Rightarrow_{\cbv}$ reduction sequence starting at $(M,s)$, it has already reached a value, therefore $(M,s) \Rightarrow_{\cbv} \dots$ does reach a value eventually, i.e.~$M$ terminates with the reduction strategy $\cbv$ and samples $s$.
-\end{proof}
-
-\begin{corollary}
-If $M$ is AST with respect to any reduction strategy, it is AST.
-\end{corollary}
-\begin{proof}
-Suppose $M$ is AST with respect to $f$.
-Let the set of samples with which it terminates with this reduction strategy be $X$.
-By Theorem~\ref{thm:CbvIsTerminatingest}, $M$ also terminates with $\cbv$ and every element of $X$, and $X$ has measure 1, by assumption, therefore $M$ is AST with respect to $\cbv$ therefore by Theorem~\ref{thm:AstEquivalence} it is AST.
-\end{proof}
-
-All of the theorems on the termination of rankable terms therefore extend to other reduction strategies too.
-The proofs of Theorems \ref{thm:rankable implies termination}, \ref{thm:partial implies rankable}, \ref{thm:antitone rankable implies termination} and \ref{thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies.
-They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because Theorem \ref{thm:de groote} applies equally to any reduction strategy).
-For a reduction strategy $r$ and term $M$, just substitute $N[r(N)]$ being a $\tY$-redex for $N$ being of the form $E[\tY \lambda x.
-O]$, and $r(N)$ being undefined for $N$ being a value.
-
-The domain of definition of the ranking functions also needs to be changed from the reachable terms $Rch(M)$ to the \emph{reachable terms with respect to the reduction strategy $r$},
-\[
-Rch_r(M) = \{N \mid \exists n, N_0, \dots, N_n: N_0 = M, N_n = N, N_i \to N_{i+1} \text{ at } r(N_i)\}.
-\]
-
-To be more explicit, the modified forms of the theorems are:
-\begin{definition}\rm
-A \emph{ranking function on $M$ with respect to a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
-\begin{itemize}
-    \item $f(N) \geq 1+ f(N')$ if $N[r(N)]$ is a $\tY$-redex and $N$ reduces to $N'$ at $r(N)$
-    \item $f(N) \geq \int_I f(N[\underline{x}/r(N)]) \, \Leb(\mathrm{d}x)$ if $N[r(N)] = \tsample$
-
-    \item $f(N) \geq f(N')$ if $r(N)$ is any other redex, where $N \to N'$ at $r(N)$.
-\end{itemize}
-Any closed term for which a ranking function with respect to a reduction strategy $r$ exists is called \emph{rankable} with respect to $r$. 
-\end{definition}
-
-\begin{definition}
-A \emph{partial ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : Rch_r(M) \rightharpoonup \mathbb R$ such that
-\begin{itemize}
-    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
-    \item $f$ is defined at $M$.
-    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ at the positions specified by $r$ will eventually reach some $O$ such that either $r(O)$ isn't defined or $f(O)$ is, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
-\end{itemize}
-\end{definition}
-
-\begin{definition}\rm
-An \emph{antitone ranking function on $M$ with respect to a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ such that
-\begin{itemize}
-    \item $f(N) \geq \epsilon(f(N)) + f(N')$ if $N[r(N)]$ is a $\tY$-redex and $N$ reduces to $N'$ at $r(N)$
-    \item $f(N) \geq \int_I f(N[\underline{x}/r(N)]) \, \Leb(\mathrm{d}x)$ if $N[r(N)] = \tsample$
-
-    \item $f(N) \geq f(N')$ if $r(N)$ is any other redex, where $N \to N'$ at $r(N)$.
-\end{itemize}
-Any closed term for which an antitone ranking function with respect to a reduction strategy $r$ exists is called \emph{antitone rankable} with respect to $r$. 
-\end{definition}
-
-\begin{definition}
-An \emph{antitone partial ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : Rch_r(M) \rightharpoonup \mathbb R$ such that there exists some antitone function $\epsilon : \nnReal \to \pReal$ such that
-\begin{itemize}
-    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
-    \item $f$ is defined at $M$.
-    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ at the positions specified by $r$ will eventually reach some $O$ such that either $r(O)$ isn't defined or $f(O)$ is, and $f(N) \geq \mathbb E[f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
-\end{itemize}
-\end{definition}
-
-\begin{theorem}
-If a closed SPCF term $M$ has a ranking function, partial ranking function, antitone ranking function or antitone partial ranking function with respect to a reduction strategy $r$, then $M$ is AST with respect to $r$, and AST.
-\end{theorem}
-\lo{Give an example or two of a reduction strategy (for $\Rightarrow$) that is not $\cbv$.}
-
-\akr{NOTE. For the standard CBV semantics, this example is indeed not a problem. 
-In the alternative semantics, defining 
-\begin{align*}
-A[x] &= \tif{\tif{x>0}{I}{I} \, \tsample - 0.5 > 0}{0}{\Omega}\\
-B &= \tif{\tsample - 0.5 > 0}{0}{\Omega}
-\end{align*}
-The set of samples where it terminates is 
-\[\begin{array}{l}
-\bigcup_{r \in [0,1]} \{s \in I^{L_s(A[\tsample])} \mid s([A[\tsample]], \mathsf{if}_1;\underline{ -
-}_1;@_1;\mathsf{if}_1) = r, \\
-\qquad \qquad \hfill \qquad \qquad s([A[\tsample], A[r], \ldots, B, \mathsf{if}_1) > 0.5 \}.
-\end{array}\] 
-This is a rather unwieldy expression, but the crucial part is that $r$ occurs twice in the conditions on $s$: once as the value a sample must take, and once in the location of a sample. 
-It is this set that's unmeasurable.}
-
-\bibliographystyle{apalike}
-\bibliography{references}
-
-\lo{22 Jan: 
-\paragraph{Further directions}
-
-An obvious next step is to extend the result to the $\mathsf{score}$ construct.
-
-The following are highly topical, and could form the basis of an interesting and novel DPhil thesis.
-\begin{enumerate}
-\item Devise methods for proving (positive) a.s.~termination.
-
-- For example, develop a type system satisfying the property: if a term is typable then it is (positively) a.s.~terminating.
-See~\citep{DBLP:conf/ppdp/BreuvartL18,DBLP:conf/esop/LagoG17}.
-
-- Another approach is to develop algorithms that synthesise ranking supermartingales, following, for example, \citep{DBLP:journals/pacmpl/AgrawalC018}.
-
-\item Develop principles (e.g.~in the form of ``proof rules'') for reasoning about (positively) a.s.~termination, in the style of \citep{DBLP:journals/pacmpl/McIverMKK18}.
-
-\item Design algorithms that synthesise probabilistic invariants (\emph{qua} martingales), \`a la \cite{SchreuderO19}; see also \citep{DBLP:journals/pacmpl/HarkKGK20}.
-
-\end{enumerate}}
-
-\iffalse
-\begin{thebibliography}{9}
-\bibitem{ppcf} Thomas Ehrhard, Michele Pagani, and Christine Tasson.
-Measurable cones and stable, measurable functions: a model for probabilistic higher-order programming.
-\emph{PACMPL}, 2(POPL):59:1–59:28, 2018.
-doi: 10.1145/3158147.
-URL \href{https://doi.org/10.1145/3158147}{https://doi.org/10.1145/3158147}.
-\end{thebibliography}
-\fi
-
-\end{document}

### 2020-12-26 LICS21/probabilisticProgrammingMartingales.tex
Created version for LICS21 submission in subdirectory LICS21

--- 2020-12-23 probabilisticProgrammingMartingales.tex
@@
 
-\input{Style/comment}
+\input{../Style/comment}
 \usepackage{calculation}
@@
 \section{Supermartingales}
-One approach to proving that a term terminates almost surely is to find some variant that is bounded below and, on average, decreases sufficiently quickly that it must eventually reach 0, similarly to the approach taken in \citep{DBLP:journals/pacmpl/McIverMKK18} for imperative programs.
+One approach to proving that a term terminates almost surely is to find some variant that is bounded below and, on average, decreases sufficiently quickly that it must eventually reach 0, similarly to the approach taken in
+%\citep{DBLP:journals/pacmpl/McIverMKK18} 
+\cite{DBLP:conf/popl/FioritiH15} for imperative programs.
 
@@
 \[
-f(E[R’]) \leq f(E[R]) - \epsilon(f(E[R])) 
-\]
-where $R \to R’$ is the $\tY$-redex rule.
-
+f(E[R']) \leq f(E[R]) - \epsilon(f(E[R])) 
+\]
+where $R \to R'$ is the $\tY$-redex rule.
+
+\iffalse
 \lo{An aside: This CBV $\tY$-rule seems cleaner:
@@
 \begin{align*}
-  & g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 \\
-  = & \log \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) \\
-  = & \frac 1 2 \log\left(\frac{(n+1)^2}{n(n+2)}\right) \\
-  = & \frac 1 2 \log\left(1 + \frac 1 {n(n+2)}\right) \\
-  > & \frac 1 {4n(n+2)} \\
-  > & \frac 1 2 \frac 1 {5(n+1)^2} + \frac 1 2 \frac 1 {5(n+3)^2}
-\end{align*}
-and $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
+g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 
+  &=  \log \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) \\
+  &=  \frac 1 2 \log\left(\frac{(n+1)^2}{n(n+2)}\right) \\
+  &=  \frac 1 2 \log\left(1 + \frac 1 {n(n+2)}\right) \\
+  &>  \frac 1 {4n(n+2)} \\
+  &>  \frac 1 2 \frac 1 {5(n+1)^2} + \frac 1 2 \frac 1 {5(n+3)^2}
+\end{align*}
+\changed[lo]{(The first inequality follows from the fact that for all $x \in (0, 1)$, $\log (1 + x) > x$.)}
+Moreover $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
 Therefore by setting $\epsilon_2(x) = \frac 1 {5(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met.
@@
 \bibliographystyle{apalike}
-\bibliography{references}
+\bibliography{../references}
 

### 2020-12-26 LICS21/ranking-supermartingales.tex
Created version for LICS21 submission in subdirectory LICS21

--- 2020-12-21 ranking-supermartingales.tex
@@
 \begin{definition}\rm
-A sequence of random variables $(Y_n)_{n \in \omega}$ adapted to a filtration $(\calF_n)_{n \in \omega}$ is a \emph{supermartingale} if for all $n \in \omega$, $\expect{|Y_n|} < \infty$, and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$.
-\lo{The preceding means: $\forall A \in \calF_n$, $\int_A \dif \mu \, Y_{n+1} \leq \int_A \dif \mu \,Y_n$.}
-For $\epsilon > 0$ and a stopping time $T$ w.r.t.~$(\calF_n)_{n \in \omega}$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.
+\begin{enumerate}
+\item A sequence of random variables $(Y_n)_{n \in \omega}$ adapted to a filtration $(\calF_n)_{n \in \omega}$ is a \emph{supermartingale} if for all $n \in \omega$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~\changed[lo]{(i.e.~$\forall A \in \calF_n$, $\int_A \dif \mu \, Y_{n+1} \leq \int_A \dif \mu \,Y_n$).}
+\item Let $\epsilon > 0$.
+A stopping time $T$ and a supermartingale $(Y_n)_{n \in \omega}$, both adapted to filtration $(\calF_n)_{n \in \omega}$, are a \emph{$\epsilon$-ranking supermartingale} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.\footnote{For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable $s \mapsto \begin{cases}
+1 & \hbox{if $s \in A$}\\
+0 & \hbox{otherwise}
+\end{cases}$.}
 A \emph{ranking supermartingale} is a $\epsilon$-ranking supermartingale for some $\epsilon > 0$.
+\end{enumerate}
 \end{definition}
@@
 \begin{remark}
-\changed[lo]{The original definition of ranking supermartingale, due to \citep{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, does not mention stopping time explicitly:
+\changed[lo]{The original definition of ranking supermartingale, due to \citep{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, does not refer to a stopping time explicitly:
 Fioriti and Hermanns call a supermartingale $(Y_n)_{n \in \omega}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot 1_{\set{Y_n > 0}}$.}
@@
 
-The following is an iterated version of Doob's well-known Optional Sampling Theorem (see, e.g., \cite[\S 6.7]{AshDD00}).
+The following is an iterated version of Doob's well-known Optional Sampling Theorem (see, e.g., \cite[\S 6.7]{AshDD00} and \cite[Theorem 7.2]{DBLP:conf/popl/FioritiH15}).
 \begin{theorem}[Optional Sampling]
 \label{thm:optional sampling}
-Let $(X_n)_{n\in \omega}$ be a supermartingale, and $(T_n)_{n \in \omega}$ a sequence of increasing stopping times, then $(X_{T_n})_{n \in \omega}$ is a supermartingale adapted to $(\calF_{T_n})_{n \in \omega}$ if one of the following conditions holds:
+Let $(X_n)_{n\in \omega}$ be a supermartingale, and $(T_n)_{n \in \omega}$ a sequence of increasing stopping times, both adapted to filtration $(\calF_n)_{n \in \omega}$, then $(X_{T_n})_{n \in \omega}$ is a supermartingale adapted to $(\calF_{T_n})_{n \in \omega}$ if one of the following conditions holds:
 \begin{enumerate}
@@
 
-\begin{lemma}[{\citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}}]
+\begin{lemma}%[{\citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}}]
 % \citep[Prop 1]{DBLP:conf/popl/ChatterjeeFNH16}}
@@
 
-\begin{proof}[Proof Sketch]
-First establish by induction on $n$: 
+\begin{proof}
+We first prove by induction on $n$: 
 \[
@@
 It follows that $\sum_{i=0}^{\infty} \mathbb{P}[T > i]$ converges; hence $\lim_{i \to \infty} \mathbb{P}[T > i] = 0$ and so $\mathbb{P}[T < \infty] = 1$.
-Then it suffices to prove: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=0} \mathbb{P}[T > i]$.
+It then remains to observe: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=0} \mathbb{P}[T > i]$.
 \end{proof}
 
-\lo{\Cref{lem:rank-PAST} is a slight extension of \citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}.
-In op.~cit., the stopping time $T$ is assumed to be the random variable $s \mapsto \min\set{n \mid Y_n(s) = 0}$.}
+\changed[lo]{\Cref{lem:rank-PAST} is a slight extension of \citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}. 
+In Lemma 5.5 of op.~cit., the stopping time $T$ is assumed to be the random variable $s \mapsto \min\set{n \mid Y_n(s) = 0}$.}
 
@@
 
-A main result is the following theorem.
-\begin{theorem} 
+Our first result is the following theorem.
+\begin{theorem}[Rankability] 
 \label{thm:rankable and strict rankable}
@@
 
-\paragraph{Proof of \Cref{thm:rankable and strict rankable}.}
+%\subsection{Proof of \Cref{thm:rankable and strict rankable}.}
 \emph{Notation}. Given an $\omega$-sequence (e.g.~$s \in S$) and $m \in \omega$, we write $s_{\leq m} \in I^m$ to mean the prefix of $s$ of length $m$.
@@
 %$\#\mathrm{draw}_n$ is a stopping time (bounded by $n$). 
+\iffalse
 \akr{$\#\mathrm{draw}_n$ is not a stopping time.
 It doesn't look like you actually use this claim anyway, so it should just be fine to remove, but did you mean something different?} \lo{I agree, and I don't actually use this claim.}
+\fi
 Take $s \in A \in \calF_n$ with $\ndraw{n}{s} = l$.
@@
 \medskip
-As before, fix a $M \in \Lambda^0$.
+As before, fix a closed SPCF term $M \in \Lambda^0$.
 Define random variables on the probability space $(S, \calF, \mu)$:
@@
 \end{align}
-The random variable $T_M^\tY$ is known as the \emph{$\tY$-runtime of $M$}, and it can equivalently be defined as the number of $\tY$-reduction steps in the reduction sequence of $M$.
+The random variable $T_M^\tY$, which we call the \emph{$\tY$-runtime of $M$}, can equivalently be defined as the number of $\tY$-reduction steps in the reduction sequence of $M$. 
 Note that, as the type system ensures that the reduction relation excluding $\tY$-reduction is strongly normalising, only finitely many reductions can occur in a row without one of them being a $\tY$-reduction, therefore $T_M^\tY < \infty$ a.s.~iff $M$ is AST.
+
+The r.v.~$T_M^\tY$ is a stopping time adapted to the filtration $(\calF_{T_n})_{n \in \omega}$.
 
@@
 \begin{enumerate}
-\item If a closed SPCF term $M$ is rankable, then $T^{\tY}_M < \infty$ a.s.~(equivalently $M$ is AST), and $\expect{T^{\tY}_M} < \infty$ i.e.~$M$ is $\tY$-positively almost-sure terminating ($\tY$-PAST).
+\item If a closed SPCF term $M$ is rankable, then $M$ is AST (equivalently, $T^{\tY}_M < \infty$ almost surely), and $M$ is $\tY$-positively almost-sure terminating ($\tY$-PAST) i.e.~$\expect{T^{\tY}_M} < \infty$.
 
@@
 \begin{proof}
-For 1, observe that $T_0, T_1, T_2, \cdots$ are an increasing sequence of stopping times adapted to $(\calF_n)_{n \in \omega}$, and each $T_i$ is bounded.\lo{Prove it.}
-Thanks to \Cref{thm:rankable and strict rankable}, 
-$(X_n)_{n \in \omega}$ is a supermartingale;
-it then follows from \Cref{thm:optional sampling} that $(Y_n)_{n \in \omega}$ is a ($1$-ranking) supermartingale with respect to the filtration $(\calF_{T_n})_{n \in \omega}$ and the stopping time $T^{\tY}_M$.
+Let $f$ be a ranking function on $M$.
+For 1, observe that $T_0, T_1, T_2, \cdots$ are an increasing sequence of stopping times, each is bounded \lo{Prove it.} and adapted to $(\calF_n)_{n \in \omega}$.
+Then, since $(f(M_n))_{n \in \omega}$ is a supermartingale also adapted to $(\calF_n)_{n \in \omega}$ (\Cref{thm:rankable and strict rankable}),
+it follows from the Optional Sampling \Cref{thm:optional sampling} that $(Y_n)_{n \in \omega}$ where $Y_n := f(M_{T_n})$ is a supermartingale adapted to the filtration $(\calF_{T_n})_{n \in \omega}$.
+Since $T_M^\tY$ is a stopping time also adapted to $(\calF_{T_n})_{n \in \omega}$, we have that $(Y_n)_{n \in \omega}$ is a $1$-ranking supermartingale.
 Therefore, by \Cref{lem:rank-PAST}, $T^{\tY}_M < \infty$ a.s.~and $\expect{T^{\tY}_M} < \infty$.

### 2020-12-26 LICS21/ranking.tex
Created version for LICS21 submission in subdirectory LICS21

--- 2020-12-23 probabilisticProgrammingMartingales.tex
@@
-\documentclass{article}
+\section{Constructing Ranking Functions}
+\label{sec:ranking}
 
-%% BEGIN {Luke's Macros}
-%%
-\newif\ifdraft
-\drafttrue %% To hide all comments and highlighting, just comment this line out
-
-\input{Style/comment}
-\usepackage{calculation}
-\usepackage{graphicx}
-\usepackage{tikz}
-\usetikzlibrary{positioning, graphs, quotes, graphdrawing, shapes.geometric}
-\usegdlibrary{layered, circular}
-\usepackage[square]{natbib}
-\setcitestyle{aysep={}}
-
-\definecolor{oxblue}{RGB}{0,33,71}
-\definecolor{oxgold}{HTML}{a0630a}
-\usepackage[final,
-bookmarks,
-bookmarksopen,
-colorlinks,
-final,
-linkcolor=red,
-citecolor=oxgold,
-pdfstartview=FitH ]{hyperref}
-
-\usepackage{amsmath} % cleveref must be loaded afer amsmath
-\usepackage{cleveref}
-\Crefname{theorem}{Theorem}{Theorems}
-\Crefname{corollary}{Corollary}{Corollary}
-\crefname{proposition}{Proposition}{Propositions}
-\Crefname{claim}{Claim}{Claims}
-\Crefname{definition}{Definition}{Definitions}
-\Crefname{fact}{Fact}{Facts}
-\Crefname{conj}{Conjecture}{Conjectures}
-\Crefname{example}{Example}{Examples}
-\Crefname{remark}{Remark}{Remarks}
-\Crefname{convention}{Convention}{Conventions}
-\Crefname{lemma}{Lemma}{Lemmas}
-\Crefname{assumption}{Assumption}{Assumptions}
-\Crefname{section}{Section}{Sections}
-\Crefname{appendix}{Appendix}{Appendices}
-\Crefname{figure}{Figure}{Figures}
-%\Crefname{equation}{Eq.}{Equations}
-
-% \addtotheorempostheadhook[theorem]{\crefalias{thmlisti}{theorem}}
-% \addtotheorempostheadhook[lemma]{\crefalias{thmlisti}{lemma}}
-% \addtotheorempostheadhook[proposition]{\crefalias{thmlisti}{proposition}}
-% \addtotheorempostheadhook[corollary]{\crefalias{thmlisti}{corollary}}
-% \addtotheorempostheadhook[claim]{\crefalias{thmlisti}{claim}}
-% \addtotheorempostheadhook[fact]{\crefalias{thmlisti}{fact}}
-% \addtotheorempostheadhook[example]{\crefalias{thmlisti}{example}}
-% \addtotheorempostheadhook[definition]{\crefalias{thmlisti}{definition}}
-% \addtotheorempostheadhook[remark]{\crefalias{thmlisti}{remark}}
-% \addtotheorempostheadhook[assumption]{\crefalias{thmlisti}{assumption}}
-
-\newcommand{\Real}{\mathbb{R}}
-\newcommand{\nnReal}{\mathbb{R}_{{\geq}0}}
-\newcommand{\pReal}{\mathbb{R}_{{>}0}}
-
-\newcommand\expect[1]{\mathbb{E}[#1]}
-\newcommand\set[1]{\{#1\}}
-\newcommand\dif{\mathrm{d}}
-\newcommand\calF{\mathcal{F}}
-\newcommand\calI{\mathcal{I}}
-\newcommand\Leb{\mathrm{Leb}}
-\newcommand\ndraw[2]{\#\mathrm{draw}_{#1}(#2)}
-
-\makeatletter
-\newcommand{\dotDelta}{{\vphantom{\Delta}\mathpalette\d@tD@lta\relax}}
-\newcommand{\d@tD@lta}[2]{%
-  \ooalign{\hidewidth$\m@th#1\mkern-1mu\cdot$\hidewidth\cr$\m@th#1\Delta$\cr}%
-}
-
-%\usepackage[bbgreekl]{mathbbol} %for \bblambda
-%%
-%% END {Luke's Macros}
-
-\usepackage{amsfonts}
-\usepackage{amsthm}
-\usepackage{amssymb}
-\usepackage{verbatim}
-\usepackage{enumitem}
-\newcommand{\tY}{\mathsf{Y}}
-%\newcommand{\tif}[3]{\textsf{if }#1\textsf{ then }#2\textsf{ else }#3}
-\newcommand{\tif}[3]{\mathsf{if}(#1, #2, #3)} % space-saving concrete syntax
-\newcommand{\tsample}{\mathsf{sample}}
-\newcommand{\tscore}{\mathsf{score}}
-\newcommand{\skeletonPlaceholder}{\mathsf{X}} % I'm not sure what the best notation here is.
-\DeclareMathOperator{\red}{red}
-\DeclareMathOperator{\nnext}{next}
-\DeclareMathOperator{\cbv}{cbv}
-
-\usepackage{amsthm}
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}
-\newtheorem{example}{Example}
-
-\theoremstyle{lemma}
-\newtheorem{lemma}{Lemma}
-\newtheorem{theorem}{Theorem}
-\newtheorem{corollary}{Corollary}
-\newtheorem{conjecture}{Conjecture}
-
-\theoremstyle{remark}
-\newtheorem*{remark}{Remark}
-
-\title{Supermartingales and Termination Analysis of Statistical PCF}
-
-\begin{document}
-
-\maketitle
-
-\lo{Tentative title}
-
-\begin{abstract}
-\lo{TODO}
-\end{abstract}
-
-\iffalse
-\akr{Following your suggestion, I will be providing a criterion for termination of programs in PPCF \citep{DBLP:journals/pacmpl/EhrhardPT18} based on ranking supermartingales. 
-As it's more convenient for this proof, a sampling-based semantics \lo{You need to provide reference(s) for sampling-based semantics.} will be used instead of the original distributional semantics. 
-I assume some roughly applicable equivalence is proven somewhere, but it doesn't seem that hard anyway.}
-\lo{Why not provide a proof as an appendix?}
-\fi
-
-\tableofcontents
-
-Various theorems about probabilistic programs rely on the assumption that the program terminates almost surely.
-Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \citep{DBLP:journals/pacmpl/McIverMKK18}.
-This paper's contribution is to extend this method to a higher-order setting.
-
-\section{Statistical PCF}
-
-\subsection{Syntax of SPCF}
-
-The language SPCF is a simply-typed lambda calculus with sampling of real numbers from $[0,1]$ and unbounded scoring, following \cite{MakOP20b}.
-Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
-\begin{align*}
-  & \text{types } A, B ::= \textsf{R}  \mid  A \to B \\
-  & \text{values } V ::= \lambda x.M  \mid  \underline{r} \\
-  & \text{terms } M, N ::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY M  \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample  \mid  \tscore(M)
-\end{align*}
-\begin{align*}
-  \frac{}{\Gamma ; x:A \vdash x:A} \qquad
-  \frac{\Gamma ; x:A \vdash M : B}{\Gamma \vdash \lambda x.M : A \to B} \qquad
-  \frac{}{\underline{r} : \textsf{R}} \qquad
-  \frac{\Gamma \vdash M:A \to B \quad \Gamma \vdash N : A}{\Gamma \vdash M \, N : B} \\ \\
-  \frac{\Gamma \vdash M_1:\textsf{R} \dots \Gamma \vdash M_n:\textsf{R}}{\Gamma \vdash \underline{f}(M_1,\dots,M_n) : \textsf{R}} \ f : \mathbb{R}^n \to \mathbb{R} \qquad
-  \frac{\Gamma \vdash M : (A \to B) \to (A \to B)}{\Gamma \vdash \tY M : (A \to B)} \\ \\
-  \frac{\Gamma \vdash M : \textsf{R} \quad \Gamma \vdash N_1 : A \quad \Gamma \vdash N_2 : A}{\Gamma \vdash \tif{M < 0}{N_1}{N_2} : A} \qquad
-  \frac{}{\Gamma \vdash \tsample : \textsf{R}} \qquad
-  \frac{\Gamma \vdash s : \textsf{R}}{\Gamma \vdash \tscore (s) : \textsf{R}}
-\end{align*}
-
-Terms are identified up to $\alpha$-equivalence, as usual. The set of all terms is denoted $\Lambda$, and the set of closed terms is denoted $\Lambda^0$.
-
-\paragraph{}
-To define the reduction relation, let \emph{evaluation contexts} be of the form:
-\begin{align*}
-  E ::= & \, [\,] \mid E \, M \mid V \, E \mid \underline{f}(\underline r_1,\dots ,\underline r_{k-1}, E, M_{k+1}, \dots, M_n) \\ & \mid \tY E \mid \tif{E<0}{M_1}{M_2} \mid \tscore (E)
-\end{align*}
-then a term reduces if it is formed by substituting a redex in a context i.e.
-\begin{align*}
-  E[(\lambda x.M) N] & \to E[M[N/x]] \\
-  E[\underline f (\underline r_1, \dots , \underline r_n)] & \to E[\underline{f(r_1,\dots,r_n)}] \\
-  E[\tY \lambda x. M] & \to E[\lambda z. M[(\tY \lambda x. M)/x] z] \text{ where $z$ is not free in $s$}\\
-  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_1] \text{ where }r < 0 \\
-  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_2] \text{ where }r \geq 0 \\
-  E[\tsample] & \to E[\underline r] \text{ where } r \in [0,1] \\
-  E[\tscore(\underline r)] & \to E[\underline r].
-\end{align*}
-\changed[lo]{We write $\to^\ast$ for the reflexive, transitive closure of $\to$.}
-\lo{The $\tscore$ rule should have a side condition: if $r > 0$.}
-
-Every closed well-typed term either is a value or reduces to another term.
-
-\subsection{Sampling semantics}
-\label{sec:sampling semantics}
-This version of the reduction relation allows $\tsample$ to reduce to any number in $[0,1]$.
-To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples.
-Let $ I = [0,1] \subset \mathbb{R} $, and let $S = I^{\mathbb{N}}$, with the \changed[lo]{Borel $\sigma$-algebra $\calF$} and the probability measure, denoted $\mu$, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
-Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and $\mu (\prod_{i=0}^\infty X_i) = \prod_{i=0}^\infty \Leb(X_i)$.
-The maps $\pi_h:S \to I, \; \pi_t:S \to S$ popping the first element are then measurable.
-\changed[lo]{Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(S, \calF, \mu)$ the \emph{entropy space}.}
-
-The $\sigma$-algebra and measure on $\Lambda$ are defined by considering it as a disjoint union of equivalence classes under replacing all the real constants by a placeholder, where the measure on each class is that of $\mathbb{R}^n$, where $n$ is the number of real constants.
-\changed[lo]{Precisely, following \cite{DBLP:conf/icfp/BorgstromLGS16},
-we view $\Lambda$ as $\bigcup_{m\in\omega} \big(\mathsf{Sk}_m \times \Real^m \big)$,
-where $\mathsf{Sk}_m$ is the set of SPCF terms with exactly $m$ place-holders (a.k.a.~\emph{skeleton terms}) for numerals.
-Thus identified, we give $\Lambda$ the countable disjoint union topology of the product topology of the discrete topology on $\mathsf{Sk}_m$ and the standard topology on $\Real^m$.
-Note that the connected components of $\Lambda$ have the form $\{M\} \times \Real^m$, with $M$ ranging over $\mathsf{Sk}_m$, and $m$ over $\omega$. 
-We fix the Borel algebra of this topology to be the $\sigma$-algebra on $\Lambda$.}
-
-The one-step reduction is given by the function $\red : \Lambda^0 \times S \to \Lambda^0 \times S$ where
-\begin{equation}
-\red(M,s) = \left\{
-    \begin{array}{ll}
-        (E[N],s) & \text{if } M = E[R], R \to N \text{ and } R \neq \tsample \\
-        (E[\underline{\pi_h(s)}],\pi_t(s)) & \text{if } M = E[\tsample] \\
-        (M,s) & \text{if } M \text{ is a value}
-    \end{array} \right .
-\end{equation}
-\lo{Both $\to$ and $\red$ are called reduction relation.}
-
-The result after $n$ steps is then simply $\red^n(M,s) = \overbrace{\red(...\red(}^n M,s)...)$, and the limit $\red^\infty$ can then be defined as a partial function as $\lim_{n \to \infty} \red^n(M,s)$ whenever that sequence becomes constant by reaching a value.
-A term $M$ terminates for a sample sequence $s$ if the limit $\red^\infty(M,s)$ is defined.
-
-The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid M \text{ terminates at } s \}$ is measurable.
-A term $M$ is said to terminate almost surely if $\mu(\{s \mid M \text{ terminates at } s\}) = 1$.
-
-\lo{Alternatively, define the \emph{runtime of $M$} to be the random variable 
-\[
-T_M(s) := 
-\begin{cases}
-\min \set{n \mid \pi_0(\red^n(M, s)) \textrm{ is a value}} & \hbox{if $\red^\infty(M,s)$ is defined}\\
-\infty & \hbox{otherwise}
-\end{cases}
-\]
-Equivalently, we say that $M$ is \emph{almost-surely terminating} (AST) if $T_M < \infty$ a.s.; 
-and $M$ is \emph{positively almost-surely terminating} (PAST) if $\expect{T_M} < \infty$.}
-
-For example, the term 
-\[
-(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0},
-\] 
-which generates a geometric distribution, terminates on the set $S \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas 
-\[
-\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \underline 0},
-\] 
-which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.
-
-\paragraph{}
-Note that the requirement that $M$ be closed is not important for the proofs of the theorems, but is simply required to ensure that $M$ reaches a value, rather than a general normal form like $x \, y$.
-
-\paragraph{}
-This definition of almost-sure termination is equivalent to that given in \citep{MakOP20b}, although the program semantics is stated in a slightly different way.
-In particular, the argument to $\tscore$ is not relevant to termination (except for the possibility that its argument's evaluation wouldn't terminate).
-
-\iffalse
-\lo{Your operational semantics does not maintain a record of the current weight of the reduction.
-A.s.~termination does depend on $\tscore$: see \cite[\S 4.3]{DBLP:journals/corr/abs-2004-03924}\footnote{\url{https://arxiv.org/abs/2004.03924}}.
-I think it important to take the behaviour of $\tscore$ into account;
-you should do it as a future task.}
-\fi
-
-\section{Supermartingales}
-One approach to proving that a term terminates almost surely is to find some variant that is bounded below and, on average, decreases sufficiently quickly that it must eventually reach 0, similarly to the approach taken in \citep{DBLP:journals/pacmpl/McIverMKK18} for imperative programs.
-
-These variants are defined as functions from reachable terms (i.e.~possible states of the program's execution) to real numbers.
-Specifically, let the set of reachable terms from a given closed starting term $M$ be $Rch(M) := \{N \in \Lambda \mid M \to^* N \}$, with the $\sigma$-algebra induced as a subset of $\Lambda$.
-
-\begin{definition}\rm
-A \emph{ranking function on $M$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
-\begin{enumerate}
-    \item $f(E[\tY \lambda x. N]) \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z) \text{ where $z$ is not free in $N$}$
-    \item $f(E[\tsample]) \geq \int_I f(E[\underline{x}]) \, \Leb(\mathrm{d}x)$
-
-    \item $f(E[R]) \geq f(E[R'])$ for any other redex $R$, where $R \to R'$.
-\end{enumerate}
-We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that $f(N) = 0$ iff $N$ is a value, and for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.
-
-Any closed term for which a ranking (respectively, strict ranking) function exists is called \emph{rankable} (respectively, \emph{strictly rankable}). 
-\changed[lo]{For example, the term $(\tY \lambda x.x) \, \underline 0$ is not rankable.}
-It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \in \omega}$ is the reduction sequence starting from $M$, then $(f(M_n))_{n\in\omega}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales.
-\end{definition}
-
-\input{ranking-supermartingales}
-
-\section{Constructing Ranking Functions}
 Although rankability implies almost-sure termination, the converse does not hold in general. For example,
@@
 In fact, even the partial function $\tY \Theta N \mapsto 2$ alone is a partial ranking function for this term.
-\section{Antitone ranking functions}
-
-\begin{example}[Random walk]
-\label{ex:ac-ranking}
-\begin{enumerate}
-\item 1D biased (towards 0) random walk
-\[
-M_1 = 
-\big(\tY \, \lambda f n . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{3/2} f \, (n + 1)}\big) \, 10
-\]
-
-\item 1D unbiased random walk
-\[
-M_2 = 
-\big(\tY \, \lambda f n . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}\big) \, 10
-\]
-
-\item 1D biased (away from 0) random walk
-\[
-M_3 = 
-\big(\tY \, \lambda f n . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/3} f \, (n + 2)}\big) \, 10
-\]
-
-%While {x > 0} do {x := x-1 \oplus_2/3 x:= x+1}
-
-%M2 = While {x > 0} do {x := x-1 \oplus_2/3 x:= x+2}
-\end{enumerate}
-\end{example}
-
-The term $M_1$ is rankable, terminating in 31 $\tY$-steps on average, and $M_3$ only has a $1/1024$ chance of terminating, but in between, $M_2$ is AST, but isn't $\tY$-PAST, therefore it isn't rankable and \Cref{thm:rankable implies termination} is insufficient to prove its termination.
-We therefore want to find a generalised notion of ranking function so that $M_2$ becomes rankable, and then prove it sound i.e.~rankable in this generalised sense implies AST.
-
-\paragraph{Idea.}
-
-The definition of antitone-convex ranking function $f$ (say) for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
-we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) function $\epsilon : \nnReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
-\[
-f(E[R’]) \leq f(E[R]) - \epsilon(f(E[R])) 
-\]
-where $R \to R’$ is the $\tY$-redex rule.
-
-\lo{An aside: This CBV $\tY$-rule seems cleaner:
-\[
-\big(\tY f^{A \to B} \, x^A \, . \, \theta^B \big) \, v \to \theta[(\tY f \, x \, . \, \theta) / f, v / x].
-\]
-We assume $\tY f^{A \to B} \, x^A \, . \, \theta^B$ is a value.}
-
-\begin{definition}
-We say that a supermartingale $(Y_n)_{n \in \omega}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \in \omega}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
-\[
-\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot 1_{\set{T > n}}
-\]
-\end{definition}
-
-%\lo{The preceding definition does not place any constraint on the stopping time $T$ (other than that it is w.r.t.~the filtration $(\calF_n)_{n \in \omega}$).
-However for the definition to make sense, $T$ must be the stopping time $T(s) := \min \set{n \mid Y_n(s) = 0}$.}
-
-\begin{theorem}
-\label{thm:a-c strict}
-Let $(Y_n)_{n \in \omega}$ be an antitone strict supermartingale via function $\epsilon : \nnReal \to \pReal$, and stopping time $T$. 
-Then $T < \infty$ almost surely.
-\end{theorem}
-
-\begin{proof}
-First, as $(Y_n)$ is a supermartingale, $\expect{Y_n} \leq \expect{Y_0}$. 
-Therefore
-
-\begin{calculation}
-\expect{Y_n \mid T > n}
-  \step[=]{rearranging terms}
-\dfrac{\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
-+ 
-\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
-- 
-\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
-  \step[=]{definition of conditional expectation}
-\dfrac{\expect{Y_n} - \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
-  \step[\leq]{$Y_n \geq 0$ always}
-\dfrac{\expect{Y_n}}{\mathbb P[T > n]}
-  \step[\leq]{$(Y_n)_n$ is a supermartingale}
-\dfrac{\expect{Y_0}}{\mathbb P[T > n]}
-\end{calculation}
-
-\emph{Claim}: For all $0 < x \leq 1$, 
-\(
-\mathbb P[T > B_x] \leq x,
-\)
-where $B_x = \left \lceil \dfrac{\expect{Y_0}+1}{x \; \epsilon(\expect{Y_0} \, x^{-1})} \right \rceil$.
-
-\smallskip
-
-As the convex hull of $\epsilon$ (the greatest convex function less than or equal to it) satisfies all the conditions assumed of $\epsilon$, in addition to being convex, assume wlog that $\epsilon$ is convex.
-
-Assume for a contradiction that $\mathbb P[T > B_x] > x$.
-Then, take $n \leq B_x$. 
-We have
-%\lo{Change of justification of the second equality below: event, not discrete r.v. Recall: expectation conditioning on r.v.~(resp.~event) is a r.v.~(resp.~number).}
-\begin{calculation}
-\expect{Y_n-Y_{n+1}}
-  \step[=]{$\calF_n \subseteq \calF_{n+1}$, definition and linearity of conditional expectation}
-\expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
-  \step[\geq]{antitone strict assumption}
-\expect{\epsilon(Y_n) \cdot 1_{\set{T > n}}}
-%  \step[=]{definition of conditional expectation for a discrete variable}
-\step[=]{definition of expectation conditioning on an \emph{event}}
-\mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
-  \step[\geq]{Jensen's inequality}
-\mathbb P[T > n]\ \epsilon(\expect{Y_n \mid T > n})
-  \step[\geq]{proved earlier}
-\mathbb P[T > n]\ \epsilon\Big(\dfrac{\expect{Y_0}}{\mathbb P[T > n]}\Big)
-  \step[>]{assumption, $\mathbb P[T > n] \geq \mathbb P[T > B_x] > x$}
-x \; \epsilon\big(\expect{Y_0} \, x^{-1}\big).
-\end{calculation}
-Therefore, by a telescoping sum
-\[
-\expect{Y_{B_x}} = \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1}) \leq -1 < 0 
-\]
-%\lo{$\expect{Y_0 - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})}$}
-which is a contradiction, therefore the claim must be true, therefore $P[T > n]$ tends to 0 as $n$ tends to infinity, therefore $T < \infty$ almost surely.
-\end{proof}
-
-Note that it is essential in this proof that $\epsilon$ be defined for all $\mathbb R_{\geq 0}$, not just the values that $(Y_n)_{n \in \omega}$ actually takes (or at least if $(Y_n)_{n \in \omega}$ is uniformly bounded, $\epsilon$ must be positive at the supremum as well as the actual values).
-
-\begin{example}
-\akr{Consider a random variable $X$, uniformly distributed in the interval $[-1,1]$, $T := \min \set{n \mid 2^{-n} < X}$, 
-\[
-Y_k := 
-\begin{cases}
-4-2^{-k} & \hbox{if $k < T$}\\
-0 & \hbox{otherwise}
-\end{cases} 
-\]
-} \lo{The following reformulation defines $T$ and $Y_k$ directly as a measurable function on a probability space (rather than via a r.v.~$X$).}
-
-Take the probability space $(\Omega, \calF, \mathbb{P})$ where $\Omega$ is the closed interval of reals $[-1, 1]$, $\calF$ the Borel $\sigma$-algebra, and $\mathbb{P}$ the corresponding Lebesgue probability measure.
-Let $\omega \in [-1, 1]$. Define random variables $T$ and $(Y_k)_{k \in \omega}$:
-\begin{align*}
-T(\omega) & := 
-\begin{cases}
-\min\set{n \in \mathbb{N} \mid \omega > 2^{-n}} & \hbox{if $\omega \in (0, 1]$}\\
-\infty & \hbox{otherwise}
-\end{cases} 
-\\
-Y_k(\omega) & := 
-\begin{cases}
-4-2^{-k} & \hbox{if $\omega \in [-1, 2^{-k}]$ (equivalently $k < T(\omega)$)}\\
-0 & \hbox{otherwise}
-\end{cases} 
-\end{align*}
-\lo{N.B.~$\set{T = n} = (\frac{1}{2^{n}}, \frac{1}{2^{n-1}}]$, and so, $\set{T > n} = [-1, \frac{1}{2^n}]$.}
-Plainly, $T$ is a stopping time, and $(Y_n)_{n \in \omega}$ is a supermartingale, adapted to the filtration $(\calF_n)_{n \in \omega}$ where $\calF_n = \calF$ for all $n$. 
-In this case, $(Y_n)_{n \in \omega}$ either tends to $4$ as $n \to \infty$, or drops to 0 at some point, with an exponentially decreasing probability. 
-With respect to the antitone function $\epsilon(x) = \frac{4-x}{4}$ and stopping time $T$, we have that $(Y_n)_{n \in \omega}$ is an antitone strict supermartingale, except that $\epsilon(4) = 0$, and $T = \infty$ with probability $\frac 1 2$, even though $4$ is larger than any value $(Y_n)_{n \in \omega}$ can actually take, and $\epsilon(Y_n)$ never actually reaches $0$.
-
-%\lo{@Andrew: I refer to the preceding highlighted claim. Did you mean: for all $n, s$, $\epsilon(Y_n(s)) > Y_n(s)$?} \akr{No. This phrasing should be clearer.}
-\end{example}
-
-\begin{theorem}[Antitone ranking function soundness] \label{thm:antitone rankable implies termination}
-If a closed SPCF term $M$ is antitone rankable, then $T_M^{\tY} < \infty$ a.s.~(equivalently, $M$ is AST).
-\end{theorem}
-\begin{proof}
-As in %Theorem \ref{thm:basicRankingSoundness}, 
-\Cref{thm:rankable implies termination},
-take the set $S$ of traces as a probability space, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$.
-As before, $(Y_n)_{n \in \omega}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$.
-Therefore, it follows from \Cref{thm:a-c strict} that $T_M^{\tY}$ is almost-surely finite.
-\end{proof}
-
-\paragraph{}
-As before, constructing antitone ranking fuctions completely is not necessary, and there is a corresponding notion of an antitone partial ranking function.
-
-Define an \emph{antitone partial ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$
-\begin{itemize}
-\item $f(N) \geq 0$ for all $N$ where $f$ is defined.
-\item $f$ is defined at $M$.
-\item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
-\end{itemize}
-
-\begin{theorem}  \label{thm:antitone partial implies rankable}
-  Every antitone partial ranking function is a restriction of an antitone ranking function.
-\end{theorem}
-\begin{proof}
-  Take a closed term $M$ and an antitone partial ranking function $f$ on $M$, with a corresponding antitone function $\epsilon$.
-Assume wlog that $\epsilon$ is convex (as if it isn't, we can just take its convex hull instead).
-As in Theorem \ref{thm:partial implies rankable}, define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
-  \begin{align*}
-    f_1(N) &= f(N) \text{ whenever $f(N)$ is defined},\\
-    f_1(V) &= 0 \text{ for values $V$ not in the domain of $f$.}
-  \end{align*}
-
-  Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) = \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
-  The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on antitone partial ranking functions.
-Define $f_2(N) = \int_S f_1(\nnext(N,s)) + \epsilon(f_1(\nnext(N,s))) g(N,s) \, \mu(\mathrm d s)$.
-For any term $N$ where $f$ is defined, $f_2(N) = f(N)$, and the value that $f_2$ would have at $N$ if $f$ were not defined at $N$ is $\leq f(N)$, by the third condition on antitone partial ranking functions.
-In order to show that $f_2$ is an antitone partial ranking function, it therefore suffices to show that the value that $f_2$ would have had at each term if $f$ were not defined at that term is at least the expectation of $f_2$ after one reduction step (plus $\epsilon(f_2(N))$ if the reduction step is a $tY$-reduction).
-For any term $N$ which is not of the form $E[R]$ for some $\tY$-redex $R$, this is trivial.
-If $R$ is a $\tY$-redex, then $\epsilon(f_2(N)) \leq \int_S \epsilon(f_1(\nnext(N,s)))\, \mu(\mathrm d s)$ by the convexity of $\epsilon$, because $n$ is bounded.
-Therefore, the (total) function $f_2$, which agrees with $f$ on $f$'s domain, is an antitone ranking function on $M$ (with the same function $\epsilon$ if it's convex).
-\end{proof}
-
-As a corollary, any term which admits an antitone partial ranking function terminates almost surely.
-
-\paragraph{\Cref{ex:ac-ranking} revisited}
-\akr{All of these ranking functions are slightly incorrect, because $\Xi \, \underline n$ reduces to something like $\Xi \, (\underline n + 1)$ then $(\lambda z.
-\text{something}[\Xi/f]) (\underline n + 1)$, rather than reducing the argument first.
-This could be fixed by appealing to the confluent semantics version of the antitone ranking theorem, but it would be nice to have at least some examples here illustrating how antitone ranking functions work, rather than actually saving all these examples until the end.
-
-Even though the difference is quite trivial, this was actually the original motivating example for the confluent trace semantics.}
-
-Programs $M_1$ and $M_2$ are AST. For $M_1$, the partial ranking function 
-\[
-\big(\tY \, \lambda f n . \, \tif{n=0}{0}{f \, (n-1) \oplus_{3/2} f \, (n+1)}\big) \, \underline x \mapsto 3x + 1
-\] 
-suffices to prove its termination (and could equivalently be considered an antitone partial ranking function with the constant antitone function $\epsilon_1(x) = 1$).
-
-For $M_2$, define the function $g_1 : \nnReal \to \nnReal$ by $g_1(x) = \log(x+1) + 1$.
-Using shorthand 
-$\Theta_2 = \tY \, \lambda f n . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
-we can define an antitone partial ranking function $f_2 : \mathit{Rch}(M_2) \rightharpoonup \nnReal$ 
-%with antitone function $\epsilon_2$ defined as follows (for $n \in \mathbb N$):
-defined as: for $n \in \mathbb N$
-\begin{align*}
-{\Theta_2} \, n 
-&\mapsto 
-g_2(n)
-\\
-0 &\mapsto 0.
-\end{align*}
-For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$, each with probability $1/2$, with one $\tY$-reduction.
-Now
-\begin{align*}
-  & g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 \\
-  = & \log \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) \\
-  = & \frac 1 2 \log\left(\frac{(n+1)^2}{n(n+2)}\right) \\
-  = & \frac 1 2 \log\left(1 + \frac 1 {n(n+2)}\right) \\
-  > & \frac 1 {4n(n+2)} \\
-  > & \frac 1 2 \frac 1 {5(n+1)^2} + \frac 1 2 \frac 1 {5(n+3)^2}
-\end{align*}
-and $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
-Therefore by setting $\epsilon_2(x) = \frac 1 {5(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met.
-It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone partial ranking function, and $M_2$ is AST.
-
-\begin{example}[Continuous random walk]\label{ex:raven complex}
-In SPCF we can construct a function whose argument changes by a random amount at each recursive call:
-\[
-\underbrace{\big
-(\tY \, \lambda f x . \tif{x \leq 0}{0}{f(x - \tsample)} \big)}_{\Theta} \, 10
-\]
-We can construct a partial ranking function $f$ as follows:
-\begin{align*}
-\Theta \, \underline l 
-&\mapsto 
-l + 2
-\\
-\tif{l \leq 0}{0}{\Theta \, (l - \tsample)}
-&\mapsto
-l + 1
-\\
-0 &\mapsto 0.
-\end{align*}
-
-For a more complex (not $\tY$-PAST) example, consider the following ``continuous random walk''
-\[
-\underbrace{\big
-(\tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample + 1/2)} \big)}_{\Xi} \, 10
-\]
-Let $g(x) := 2 + \log(x + 1)$, and let $\epsilon$ be the function specified by
-\[
-\epsilon(g(x-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y.
-\]
-The limit of this as $x \to \infty$ and $g(x+1/2) \to \infty$ is 0, and $\frac {\mathrm d}{\mathrm dx} \epsilon(g(x+1/2)) = \frac 1 {x+1} + \log(1 - \frac 1 {x + 1/2}) < 0$, and $g$ is monotonic increasing;
-therefore $\epsilon$ is antitone and bounded below by 0.
-We define a partial antitone ranking function as follows:
-\begin{align*}
-\Xi \, l 
-&\mapsto 
-g(l)
-\\
-0 &\mapsto 0
-\end{align*}
-\end{example}
-The value of $g$ after one $\tY$-reduction step is at least $g(l-1/2)$, therefore the expectation of $\epsilon$ after one $\tY$-reduction step is at most $\epsilon(g(l-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y$.
-Thus 
-\begin{itemize}
-\item in case $l > 0$, $g$ decreases by the required amount
-\item in case $l \leq 0$, 
-\(
-g(\Xi \, l) \geq 2 + \log(1/2) > \log(\frac{3 \sqrt 3} 2) - 1 = \epsilon(g(0-1/2))
-\) 
-as well.
-\end{itemize}
-Hence this is a valid partial antitone ranking function and the term is AST.
-
-\begin{example}[Fair-in-the-limit random walk]
-\label{ex:Fair-in-the-limit random walk}\citep[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}
-\[
-\underbrace{\big
-(\tY \, f \, x . 
-\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)}_{\Xi} 
-\, 10
-\]
-To construct an antitone ranking function, we solve the recurrence relation:
-\[
-\left\{
-\begin{array}{rll}
-z_0 &=& 0\\
-n \geq 1, \quad z_n &>& \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
-\end{array}
-\right.
-\]
-For $n > 2$, $z_n = \log(n-1)$ works.
-The expected decrease is $\frac 1 2(\log(1+\frac 1 {(n-1)^2-1}) - \frac 1 {2n+1} \log(1 + \frac 2 {n-2}))$, and using the fact that $\frac x {x+1} \leq \log (1+x) \leq x$ for $x > 0$, this is at least $\frac 1 2(\frac 1 {(n-1)^2} - \frac 1 {2n + 1} \frac 2 {n-2}) = \frac {n^2}{2(n-1)^2(2n+1)(n-2)}$, which (again for $n > 2$) is positive and antitone.
-For $n = 0, 1, 2$, we take $\epsilon$ to be 9/40 (the same as its value at 3), then set $z_2 = 2 \log 2 - \log 3 - 1, z_1 = 3 \log 2 - 2 \log 3 - 3, z_0 = 4 \log 2 - 3 \log 3 - 6$.
-Some of those values are negative, but it's still bounded below, so by adding a constant offset it can be corrected, and the term is AST.
-
-\end{example}
-
-\begin{example}[Escaping spline]
-\label{ex:escaping spline}\citep[\S 5.4]{DBLP:journals/pacmpl/McIverMKK18}
-\[
-\underbrace{\big
-(\tY \, f \, x . \,
-0 \oplus_{\frac{1}{x+1}} f(x + 1) \big)}_{\Xi} 
-\, 10
-\]
-
-In this case, the fact that the ranking function must decrease at each $\tY$-step (in an antitone partial ranking function) by the expected value of $\epsilon$ not at the current term, but at the next term where the ranking function is defined, is a little harder to deal with, because the variable $x$ can change all the way to $0$ in one step, therefore simply adding a small offset doesn't suffice to compensate for this fact.
-
-Consider the candidate ranking function $\Xi \, n \mapsto n + 1$.
-For each $n$, $\Xi \, n$ reduces to either $0$ (with probability $\frac 1 {n + 1}$) or $\Xi \, (n + 1)$, therefore the expected value of the ranking function is $\frac{n(n+2)}{n+1} = n + 1 - \frac 1 {(n+1)^2}$, and the required decrease is $\frac{\epsilon(0) + n \epsilon(n + 2)}{n + 1}$, therefore eventually, the expected decrease isn't enough, whatever the value of $\epsilon(0)$ is.
-
-If instead, we take the ranking function to be defined after the $\tY$-reduction but before the $\tsample$-reduction as well, this can be resolved.
-Letting $\Theta[n] = \underline 0 \oplus_{\frac 1 n} \Xi \, n$:
-\begin{align*}
-\Theta[n] &\mapsto n \\
-\Xi \, 10 & \mapsto 12.
-\end{align*}
-For each $n$, $\Theta[n]$ reduces to either $0$ or $\Theta[n+1]$, with a $\tY$-reduction only in the latter case, and the condition that this is an antitone partial ranking function is that $n \geq \frac{(n-1)(n+1)}{n} + \frac{n-1}{n} \epsilon(n+1)$ and $12 \geq 11 + \epsilon(11)$.
-These are satisfied by setting $\epsilon(x) = \min(1, 1/x)$, therefore this term is AST.
-\end{example}
-
-\paragraph{Completeness}
-It seems very likely that this method is complete in the case that every reachable term is AST, but we have been unable to actually prove this.
-It is certainly at least capable of proving the termination of terms which terminate arbitrarily slowly (in the sense that there is no similar limitation to Theorem \ref{thm:rankable implies termination}, which can only prove termination of $\tY$-PAST terms).
-
-The following theorem does not prove completeness, but is suggestive in that direction:
-\begin{theorem}
-For any stopping time $T$ which is almost-surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
-\end{theorem}
-
-\input{relativised-completeness}
-
-\input{y-stopping-time}
-
-\section{Confluent Trace Semantics}
-When proving almost sure termination in this way, it is necessary to consider which terms a given term may reduce to.
-Sometimes however, the reduction that the programmer has in mind may not be strictly the call-by-value order defined so far, or considering an alternative reduction order may be simpler or more intuitive.
-
-Non-probabilistic lambda calculi generally have the Church-Rosser property, that if a term $A$ reduces to both $B_1$ and $B_2$, there is some $C$ 
-%with reduction sequences $B_1 \to^* C$ and $B_2 \to^* C$, 
-to which both $B_1$ and $B_2$ reduce,
-so the reduction order mostly doesn't matter. 
-In the probabilistic case, this may not be true, because $\beta$-reduction can duplicate $\tsample$s, so the outputs of the copies of the sample may be identical or independent, depending on whether the sample is taken before or after $\beta$-reduction.
-There are, however, some restricted variations on the reduction order that do not have this problem.
-
-\paragraph{}
-Even with this restriction, a sampling semantics in the style of the one already defined would not be entirely Church-Rosser, as, for example, $\red^3(\tsample - \tsample, (1,0,\dots))$ would be either $1$ or $-1$ depending on the order of evaluation of the $\tsample$s, as that determines which sample from the pre-selected sequence is used for each one.
-To fix this, rather than pre-selecting samples according to the order they'll be drawn in, select them according to the position in the term where they'll be used instead.
-
-A \emph{position} 
-%\akr{I'm not really sure how to phrase the citation, but this is adapted from \cite{Kennaway96infinitarylambda}.} 
-%\lo{The idea is common enough. I think no citation necessary.}
-is a finite sequence of steps into a term, defined inductively as
-\begin{align*}
-\alpha ::= \cdot \mid \lambda ; \alpha \mid @_1 ; \alpha \mid @_2 ; \alpha \mid \underline f_i ; \alpha \mid \tY ; \alpha \mid \textsf{if}_1 ; \alpha \mid \textsf{if}_2 ; \alpha \mid \textsf{if}_3 ; \alpha \mid \tscore ; \alpha.
-\end{align*}
-The \emph{subterm of $M$ at a position $\alpha$}, denoted $M \mid \alpha$, is defined as
-\begin{align*}
-M \mid \cdot & = M \\
-\lambda x. M \mid \lambda ; \alpha & = M \mid \alpha \\
-M_1 M_2 \mid @_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2 \\
-\underline f(M_1,\dots,M_n) \mid \underline f_i ; \alpha & = M_i \mid \alpha \quad \text{for }i \leq n \\
-\tY M \mid \tY ; \alpha & = M \mid \alpha \\
-\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2,3 \\
-\tscore(M) \mid \tscore ; \alpha & = M \mid \alpha
-\end{align*}
-so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda$ is undefined). 
-A position such that $M\mid \alpha$ does exist is said to \emph{occur} in $M$. 
-%\changed[lo]{
-\emph{Substitution} of $N$ at position $\alpha$ in $M$, written $M[N/\alpha]$, is defined similarly.
-For example, let 
-\[
-M = \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})
-\qquad
-\alpha =\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2
-\]
-then 
-\(
-M[\tsample / \alpha] = \lambda x \, y. y \, (\tif{x < 0}{y \, \tsample}{\underline 3}).
-\)
-%}
-
-Two subterms $N_1$ and $N_2$ of a term $M$, corresponding to positions $\alpha_1$ and $\alpha_2$, can overlap in a few different ways. 
-If $\alpha_1$ is an initial segment (i.e.~prefix) of $\alpha_2$ (written as $\alpha_1 \leq \alpha_2$), then $N_2$ is also a subterm of $N_1$.
-If neither $\alpha_1 \leq \alpha_2$ nor $\alpha_1 \geq \alpha_2$, the positions are said to be \emph{disjoint}.
-The notion of disjointness is mostly relevant in that if $\alpha_1$ and $\alpha_2$ are disjoint, performing a substitution at $\alpha_1$ will leave the subterm at $\alpha_2$ inaffected.
-\lo{The preceding sentence is unclear. Typo?} \akr{It wasn't a typo, but I've made it a little more explicit now. Is this clearer?}
-
-With this notation, a more general reduction relation $\to$ can be defined. \akr{It would almost work to have one of the neater versions of $\tY$-reduction here.}
-%\lo{On first reading, I struggled to parse the following clauses.}
-\begin{definition}
-\label{def:more general red}
-The binary relation $\to$ is defined by the following rules, each is conditional on a redex occurring at position $\alpha$ in the term $M$:
-\begin{align*}
-  \text{if } M \mid \alpha = (\lambda x.N) V,\ & M \to M[N[V/x]/\alpha] \\
-  \text{if } M \mid \alpha = \underline f (\underline r_1, \dots , \underline r_n),\ & M \to M[\underline{f(r_1,\dots,r_n)}/\alpha] \\
-  \text{if } M \mid \alpha = \tY \lambda x. N,\ & M \to M[\lambda z. N[(\tY \lambda x. N)/x] z/\alpha] \text{ where $z$ is not free in $N$}\\
-  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_1/\alpha] \text{ where }r < 0 \\
-  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_2/\alpha] \text{ where }r \geq 0 \\
-  \text{if } M \mid \alpha = \tsample & \text{ and $\lambda$ does not occur after $@_2$ or $\tY$ in $\alpha$},\\ & M \to M[\underline r/\alpha] \text{ where } r \in [0,1] \\
-  \text{if } M \mid \alpha = \tscore(\underline r),\ & M \to M[\underline r/\alpha].
-\end{align*}
-In each of these cases, $M \mid \alpha$ is the \emph{redex}, and reduction takes place at $\alpha$.
-\end{definition}
-
-\paragraph{}
-Labelling the pre-chosen samples by the positions in the term would also not work because in some cases, a $\tsample$ will be duplicated before being reduced, for example, in $(\lambda x.
-x \, {\underline 0} \, \mathbin{\underline{+}} \, {x \, \underline 0}) (\lambda y.
-\tsample)$, both of the $\tsample$ redexes that eventually occur originate at $@_2 ; \lambda$.
-It is therefore necessary to consider possible positions that may occur in other terms reachable from the original term. 
-Even this is itself inadequate because some of the positions in different reachable terms need to be considered the same, and the number of reachable terms is in general uncountable, which leads to measure-theoretic issues.
-
-Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $Sk(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$.
-\lo{Do we actually use the notation $S[r]$ in what follows?} \akr{Yes, in Lemma \ref{lem:independentSamples} for example.}
-\lo{Skeleton is defined in Sec.~\ref{sec:sampling semantics}.}
-The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, and $\tsample$ reducing to $\skeletonPlaceholder$.
-For example, we have 
-\(
-(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \tsample
-\to
-(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \skeletonPlaceholder
-\to
-\tif{\skeletonPlaceholder < 0}{\skeletonPlaceholder}{\skeletonPlaceholder}
-\to
-\skeletonPlaceholder
-\)
-
-Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a $\to$-reduction sequence of skeletons starting at $Sk(M)$, and the second of which is a position in the final skeleton of the reduction sequence.
-As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the usual linear sampling semantics, there will be (with one more caveat introduced later) a sample from $I$ pre-selected for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.
-
-Reduction sequences are used rather than reachable skeletons because if the same skeleton is reached twice, different samples may be needed.
-For example, consider the term $M = \tY (\lambda f x.\ \tif{\tsample - \underline{0.5} < 0}{f \, x}{x}) \, \underline 0$, which reduces after a few steps to $N = \tif{\tsample - \underline{0.5}  < 0}{M}{\underline 0}$.
-If the pre-selected sample for $(N,\textsf{if}_1;\underline{-}_1)$ is less than 0.5, $N$ reduces back to $M$, then $N$ again, then the same sample is used the next time, therefore it's an infinite loop, whereas if samples are labelled by reduction sequences, the samples for $M \to \dots \to N$ are independent from the samples for $M \to \dots \to N \to M \to \dots \to N$, and so on.
-
-The reduction sequences of skeletons will often be discussed as though they were just skeletons, identifying them with their final skeletons.
-With this abuse of notation, a reduction sequence $N$ (actually $N_1 \to \dots \to N_n = N$) may be said to reduce to a reduction sequence $O$, where the reduction sequence implicitly associated with the final skeleton $O$ is $N_1 \to \dots \to N_n \to O$.
-
-\paragraph{}
-This is still not quite sufficient because sometimes the same samples must be used at corresponding positions in different reduction sequences.
-For example, the term $M = \tsample + \tsample$ has the reachable skeletons $N_1 = \skeletonPlaceholder + \tsample, N_2 = \tsample + \skeletonPlaceholder, O = \skeletonPlaceholder + \skeletonPlaceholder$ and $\skeletonPlaceholder$, with reductions $M \to N_1 \to O \to \skeletonPlaceholder$ and $M \to N_2 \to O \to \skeletonPlaceholder$.
-In the reduction $M \to N_1$, the sample labelled $(M, \underline{+}_2)$ is used, and in the reduction $N_2 \to O$, the sample labelled $(M \to N_2, \underline{+}_2)$ is used.
-Each of these samples becomes the value of the second numeral in $O$ in their respective reduction sequences, therefore in order for Church-Rosserness to be attained, they must be the same.
-Which elements of $L_0(M)$ must match can be described by the relation $\sim^*$:
-
-\begin{definition}
-The relation $\sim$ is defined as the union of the minimal symmetric relations $\sim_p$ (``$p$'' for parent-child) and $\sim_c$ (``$c$'' for cousin) satisfying
-\begin{enumerate}
-    \item If $N$ reduces to $O$ with the redex at position $\alpha$, and $\beta$ is a position in $N$ disjoint from $\alpha$, then $(N,\beta) \sim_p (O,\beta)$.
-
-    \item If $N$ $\beta$-reduces to $O$ at position $\alpha$, $\beta$ is a position in $N \mid \alpha;@_1;\lambda$ and $N \mid \alpha;@_1;\lambda;\beta$ is not the variable involved in the reduction, $(N,\alpha;@_1;\lambda;\beta) \sim_p (O, \alpha;\beta)$
-
-    \item If $N$ $\textsf{if}$-reduces to $O$ at position $\alpha$, with the first resp.
-second branch being taken, and $\alpha;\textsf{if}_i;\beta$ occurs in $N$ (where $i = 2$ resp.
-$3$), $(N,\alpha;\textsf{if}_i;\beta) \sim_p (O,\alpha;\beta)$
-
-    \item If $N$, $O_1$ and $O_2$ match any of the following cases:
-    \begin{enumerate}
-        \item $N$ contains redexes at disjoint positions $\alpha_1$ and $\alpha_2$, $O_1$ is $N$ reduced first at $\alpha_1$ then $\alpha_2$ and $O_2$ is $N$ reduced first at $\alpha_2$ then at $\alpha_1$.
-
-        \item $N \mid \alpha = \tif{\underline r < 0}{N_1}{N_2}$, where $r < 0$ (or, respectively, $r \geq 0$), $(N_2 \text{ resp.
-} N_1) \mid \beta$ is a redex, and $O_1$ is $N$ reduced at $\alpha$ and $O_2$ is $N$ reduced first at $\alpha;(\textsf{if}_3 \text{ resp.
-} \textsf{if}_2);\beta$ then at $\alpha$
-
-        \item $N \mid \alpha = \tif{\underline r < 0}{N_1}{N_2}$, where $r < 0$ (or, respectively, $r \geq 0$), $(N_1 \text{ resp.
-} N_2) \mid \beta$ is a redex, and $O_1$ is $N$ reduced first at $\alpha$ then at $\alpha;\beta$ and $O_2$ is $N$ reduced first at $\alpha;(\textsf{if}_2 \text{ resp.
-} \textsf{if}_3);\beta$ then at $\alpha$
-
-        \item $N \mid \alpha = (\lambda x.
-A) B$, there is a redex in $A$ at position $\beta$, $O_1$ is $N$ reduced first at $\alpha$ then at $\alpha;\beta$, and $O_2$ is $N$ reduced first at $\alpha;@_1;\lambda;\beta$ then at $\alpha$
-
-        \item $N \mid \alpha = (\lambda x.
-A) B$, $B \mid \beta$ is a redex, $(\gamma_i)_i$ is a list of all the positions in $A$ where $A \mid \gamma = x$, ordered from left to right, $O_1$ is $N$ reduced first at $\alpha ; @_2 ; \beta$ then at $\alpha$, and $O_2$ is $N$ reduced first at $\alpha$ then at $\alpha;\gamma_i;\beta$ for each $i$ in order.
-
-        \item $N \mid \alpha = \tY (\lambda x.
-A)$, $A$ reduced at $\beta$ is $A'$, $(\gamma_i)_i$ is a list of all the positions where $A' \mid \gamma  = x$, ordered from left to right, $O_1$ is $N$ reduced first at $\alpha;\tY;\lambda;\beta$ then at $\alpha$, and $O_2$ is $N$ reduced first at $\alpha$ then at $\alpha;\lambda;@_1;\gamma_i;Y;\lambda;\beta$ for each $i$ in order where $\gamma_i$ is left of $\beta$ then at $\alpha;\lambda;@_1;\beta$ then at $\alpha;\lambda;@_1;\gamma_i;Y;\lambda;\beta$ for the remaining values of $i$.
-    \end{enumerate}
-    (in which case $O_1$ and $O_2$ are equal as skeletons, but with different reduction sequences), $O_1'$ and $O_2'$ are the results of applying some reduction sequence to each of $O_1$ and $O_2$ (the same reductions in each case, which is always possible because they're equal terms), and $\delta$ is a position in $O_1'$ (or equivalently $O_2'$), then $(O_1',\delta) \sim_c (O_2',\delta)$.
-\end{enumerate}
-The $\sim_c$-rules are illustrated in the following diagrams.
-\end{definition}
-
-\include{sim-diagram}
-
-The reflexive transitive closure $\sim^*$ of this relation is used to define the set of \emph{potential positions} $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable skeletons.
-If $(N,\alpha) \sim^* (O,\beta)$, then $N \mid \alpha$ and $O \mid \beta$ both have the same shape (i.e.~they're either both the placeholder $\skeletonPlaceholder$, both variables, both applications, both $\tsample$s etc.), therefore it's well-defined to talk of the set of potential positions where there is a $\tsample$, $L_s(M)$.
-Formally $L_s(M) := \{[(X, \alpha)]_{\sim^\ast_M} : X \mid \alpha = \tsample\}$.
-The new sample space is then defined as $I^{L_s(M)}$, with the Borel $\sigma$-algebra and product measure.
-\changed[lo]{Since $I^{L_s(M)}$ is a countable product, the measure space is well-defined \citep[Cor.~2.7.3]{AshDD00}.}
-
-\paragraph{}
-Although the six cases in the definition of $\sim_c$ are defined separately, there are some commonalities worth noting.
-Each case (except a, which is symmetrical) has a right branch and a left branch (the branches of case b are unfortunately displayed the wrong way around in the diagram).
-The right branch has reduction steps at two positions, $\beta$ then $\alpha$, where $\beta > \alpha$ (i.e.~$\beta$ is inside $\alpha$), and in the left branch, the order of these reductions are swapped, so the reductions are at $\alpha$ then $\beta_1, \dots, \beta_n$.
-The position $\alpha$ of one of the reductions is unchanged, and the other reduction may have multiple (in cases e and f) or 0 (in cases b and e) images in the left branch.
-The images $(\beta_i)$ are all still inside (greater than or equal to) $\alpha$, and although they are not generally equal to $\beta$, the subskeletons at these positions have a similar shape to the subskeleton (initially) at position $\beta$, so that the reductions at these positions are the same type of reduction (e.g.~both $\beta$-reduction or both $\textsf{if}$-reduction).
-Case a is also similar, except that the relevant positions are all disjoint rather than contained in one another, and either branch could be considered the right branch.
-
-\iffalse
-\changed[lo]{Since $L_s(M)$ is a countable set, there is a unique probability measure $\mu$ on the infinite product $\prod_{i \in L_s(M)} \Sigma_I$ of the Borel $\sigma$-algebras $\Sigma_I$ satisfying
-\[
-\mu\set{\omega \in \Omega \mid \omega_1 \in A_1, \cdots, \omega_n \in A_n} = \prod_{i=1}^n \Leb(A_i)
-\]
-where $\Leb$ is the Lebesgue measure on the measurable space $(I, \Sigma_I)$.}
-\fi
-
-\paragraph{}
-Before defining the new version of the reduction relation $\red$, a few lemmas about the properties of $\sim$ are necessary for it to be well-defined.
-
-\begin{lemma} \label{downAcrossUp}
-If $(N_1,\alpha_1) \sim^* (N_2,\alpha_2)$, there is some descendants $N'_1, N'_2$ of $N_1$ resp.~$N_2$, and a position $\alpha'$ in both of them such that $(N_1,\alpha_1) \sim_p^* (N'_1, \alpha') \sim_c^* (N'_2, \alpha') \sim_p^* (N_2, \alpha_2)$.
-\end{lemma}
-\begin{proof}
-  The $\sim_p$ relation can be split into $\sim_\downarrow \cup \sim_\uparrow$, where $(A,\alpha) \sim_\downarrow (B,\beta)$ if $A \to B$ and $(A,\alpha) \sim_p (B,\beta)$, and similarly $(A,\alpha) \sim_\uparrow (B,\beta)$ if $B \to A$ and $(A,\alpha) \sim_p (B,\beta)$.
-At each stage of this proof, it will be assumed that the $\sim_c$ steps and the $\sim_p$ steps in the sequence from $(N_1,\alpha_1)$ to $(N_2,\alpha_2)$ are rearranged such that there is never a $\sim_c$ immediately before a $\sim_\downarrow$, and there is never a $\sim_c$ immediately after a $\sim_\uparrow$.
-This rearrangement is always possible, because if there is some subsequence $(A,\alpha) \sim_c (B,\alpha) \sim_\downarrow (C,\beta)$, then there is an alternate path $(A,\alpha) \sim_\downarrow (B',\beta) \sim_c (C,\beta)$, where the reduction and the $\sim_\downarrow$ from $A$ to $B'$ are the same as those from $B$ to $C$, and the $\sim_c$ step is the same, but with the reduction sequences $O_1 \to^* O_1'$ and $O_2 \to^* O_2'$ in the definition of $\sim_c$ extended by the reduction $B \to C$.
-With these rearrangements assumed, it is sufficient to prove that the $\sim_p$ steps can be rearranged so that all of the $\sim_\downarrow$ steps come before all of the $\sim_\uparrow$ steps (possibly introducing some $\sim_c$ steps in the process).
-
-  The rearrangement to put all of the $\sim_\downarrow$s before all of the $\sim_\uparrow$s proceeds by induction on the number of $\sim_\downarrow$s that occur after the first occurrence of $\sim_\uparrow$.
-If it's 0, all of the $\sim_p$s are already in the correct order, and we're done.
-Otherwise, take the subsequence from the first $\sim_\uparrow$ to the first $\sim_\downarrow$ after that.
-This must be rearranged to some $\sim_\downarrow$s followed by some $\sim_\uparrow$s, then once that's done, the number of $\sim_\downarrow$s after the first $\sim_\uparrow$ in the overall sequence will have decreased by 1.
-
-  Let this subsequence be $A \sim_\uparrow^n B \sim_\downarrow C$.
-Suppose for induction that $A \sim_\uparrow^k (D,\delta) \sim_\downarrow^* (E,\epsilon) \sim_\uparrow^* C$ for some $D, \delta, E, \epsilon$, and $0 \leq k \leq n$, where in the reduction sequence $D \to^* E$, for any pair of reductions at positions which aren't disjoint, the reduction at the innermost (greater) position occurs first.
-Reduction sequences with this property will be called ``parallel'', as it is analagous to parallel reduction as in \akr{TODO: find this reference, for the proof of the Church-Rosser theorem by Tait and Martin-L\"of}.
-This induction is in reverse, with $k$ decreasing from $n$ to $0$.
-If $k = n$, simply set $(D,\delta) = B$ and $(E,\epsilon) = C$.
-If $k = 0$, then $A$ is related to $B$ by a sequence of $\sim_\downarrow$s then $\sim_\uparrow$s, as desired.
-In the intermediate steps, $k$ must be decreased by one, so a subsequence of one $\sim_\uparrow$ followed by a parallel sequence of $\sim_\downarrow$s must be replaced by a parallel sequence of $\sim_\downarrow$s followed by some $\sim_\uparrow$s.
-
-  For any individual $\sim_\uparrow$ then $\sim_\downarrow$ pair, let $(A,\alpha) \sim_\uparrow (B,\beta) \sim_\downarrow (C,\gamma)$.
-The skeleton $B$ reduces to both $A$ and $C$.
-If these are the same reduction, then $(A,\alpha) = (C,\gamma)$ and this subsequence may be removed.
-If they are different, then the way they overlap corresponds to one of the cases in the definition of $\sim_c$, either case a if the reduction positions are disjoint, or one of the cases b--f if one of the reduction positions is inside the other.
-In any case, there is a reduction sequence from each of $A$ and $C$ to some common skeleton $D$ ($O_1$ and $O_2$ in the definition of $\sim_c$).
-For the same reason that $(B,\beta) \sim_\downarrow (C,\gamma)$, there is some $\delta$ such that $(A,\alpha) \sim_\downarrow^* (D,\delta)$, and similarly for the same reason that $(B,\beta) \sim_\downarrow (A,\alpha)$, for the same $\delta$, $(C,\gamma) \sim_\downarrow^* (D,\delta)$.
-There are a lot of cases to check for this statement, but all of them are rather simple, and similar to each other.
-There is therefore an alternative $\sim^*$ sequence from $(A,\alpha)$ to $(C,\gamma)$, of the form $(A,\alpha) \sim_\downarrow^* (D_1,\delta) \sim_c (D_2,\delta) \sim_\uparrow^* (C,\gamma)$, where $D_1$ and $D_2$ are the alternative reduction sequences leading to the same skeleton $D$.
-There are only multiple $\sim_\downarrow$ steps in the result if the position of the reduction $B \to C$ is inside the position of the reduction $B \to A$, and similarly, there are only multiple $\sim_\uparrow$ steps in the result if the position of the reduction $B \to A$ is inside the position of the reduction $B \to C$.
-If there are multiple $\sim_\downarrow$ steps, they are not necessarily parallel, but in the only case where they aren't (case f of $sim_c$), the subsequence of $\sim_\downarrow$ steps with the reductions at $\alpha;\lambda;@_1;\beta$ then $\alpha;\lambda;@_1;\gamma_i;\tY;\lambda;\beta$ for each $i$ where $\gamma_i \geq \beta$ (in the notation of the definition of $\sim_c$) can be replaced by $\sim_\downarrow$s with the reductions at $\alpha;\lambda;@_1;\gamma'_i;\tY;\lambda;\beta$ then $\alpha;\lambda;@_1;\beta$, then another $\sim_c$, where $(\gamma'_i)$ is the list of positions in $A$ (\emph{before} the reduction at $\beta$ to $A'$, unlike $(\gamma_i)$) where $A | \gamma'_i = x$ and $\gamma'_i > \beta$.
-
-  The effect of this rearrangement is (ignoring the $\sim_c$s) to swap the order of the $\sim_\uparrow$ and the $\sim_\downarrow$, except that if the positions of one of these reductions is inside the other, that inner $\sim_p$ may be duplicated.
-If the positions are disjoint, both of them are unchanged (corresponding to case a of $\sim_c$), and if one is inside the other, then the outer position is unchanged and the other resultant positions, although they aren't equal to the original inner position, are still inside the outer position.
-Taking the sub-sequence of one $\sim_\uparrow$ followed by a parallel sequence of $\sim_\downarrow$s, this rearrangement can be repeatedly applied to move the $\sim_\uparrow$ further along the sequence.
-If at some point a $\sim_\uparrow$ and $\sim_\downarrow$ match (by relating the same (skeleton, position) pairs in opposite directions), they may be removed and the process may stop early.
-The $\sim_\uparrow$ may be duplicated if it passes a $\sim_\downarrow$ whose position is outside of the $\sim_\uparrow$'s position, but by the assumption that the sequence of $\sim_\downarrow$s is initially parallel, no position inside of this occurs later in the sequence, therefore all of the resultant $\sim_\uparrow$s pass the remaining $\sim_\downarrow$s without changing or duplicating them.
-The $\sim_\uparrow$s may be further duplicated, but the number of steps left in this process is bounded by the product, for all of the remaining $\sim_\downarrow$s, of $2 + \text{the number of occurrences of the variable relevant to the reduction}$, because for each $\sim_\downarrow$, that is more than the number of duplicates it could produce for any $\sim_\uparrow$ that passes it.
-It is also possible (earlier) for some of the $\sim_\downarrow$s to be duplicated, but the only case where this process would not terminate, that both the $\sim_\downarrow$s and the $\sim_\uparrow$s continue being duplicated forever so that the number of switches left to do never decreases, is prevented by the parallelness condition, as all of the duplications of $\sim_\downarrow$s occur before all of the duplications of $\sim_\uparrow$s.
-
-  It still remains to be shown that the sequence of $\sim_\downarrow$s left at the end of this can be made parallel.
-The changes that may have occurred since the previous version of this sequence (which was known to be parrallel already), are that some of the $\sim_\downarrow$s may have been duplicated by passing the $\sim_\uparrow$, and if that $\sim_\uparrow$ matches one of the $\sim_\downarrow$s, that $\sim_\downarrow$ is removed.
-The position of the $\sim_\uparrow$'s reduction is the same for all of these duplications, because it is not changed until later in the sequence, when the $\sim_\downarrow$s with positions outside of that may occur.
-The resultant positions after a duplication are in the same position relative to each other, and therefore the order of the remaining $\sim_\downarrow$s is automatically parallel, except that (letting the position of the $\sim_\uparrow$'s reduction be $\alpha$) if there are $\sim_\downarrow$s with reductions at positions $\alpha;@_1;\lambda;\beta$ and $\alpha;@_2;\gamma$, or $\alpha;\tY;\lambda;\beta$ and $\alpha;\tY;\lambda;\gamma$, some of the resultant $\sim_\downarrow$s' positions may be inside each other where they were originally not.
-Similarly to the case where a $\tY$ reduction and a reduction at a position inside it are swapped though, the order can be fixed by swapping some of the $\sim_\downarrow$ steps with each other.
-In the first case, the reduction at position $\alpha;@_1;\lambda;\beta$ ends up at $\alpha;\beta$, and the reduction at position $\alpha;@_2;\gamma$ ends up at the positions $\alpha;\delta_i;\gamma$, where $(\delta_i)$ is the positions of the variable relevant to the $\sim_\uparrow$'s reduction inside its lambda.
-The positions of the variable in the lambda may be different before the reduction at $\beta$, but the positions of the redexes corresponding to the original redex at $\alpha;@_2;\gamma$ are changed similarly.
-Rather than reducing at $\alpha;\beta$ then at $\alpha;\delta_i;\gamma$, it is therefore possible to reach the same point by reducing at $\alpha;\delta'_i;\gamma$ then at $\alpha;\beta$, where $(\delta'_i)$ is the positions of the relevant variable within the lambda before the reduction at $\alpha;@_1;\lambda;\beta$ (and again, a $\sim_c$ of some sort must also be introduced).
-The other case, that the $\sim_\uparrow$'s reduction is a $\tY$-reduction, is similar.
-Each reduction which is duplicated has one image at $\alpha;\lambda;@_1;$its original relative position, and one for each occurrence of the variable.
-Those corresponding to variable occurrences may need to be moved to before those at $\lambda;@_1$.
-
-  In summary, to reach the desired order of $\sim_p$ steps, each $\sim_\downarrow$ may have to be moved past some $\sim_\uparrow$ steps, and it may get duplicated in the process, but the resultant sequence of $\sim_\downarrow$s can all be put in parallel order, and it has been shown that moving a $\sim_\uparrow$ past a sequence of $\sim_\downarrow$s is always possible if they're in parallel order, therefore the overall process terminates and reaches a point where all of the $\sim_\downarrow$s precede all of the $\sim_\uparrow$s, and all of the $\sim_c$s are in between them.
-All of these rearrangements leave the end points of the sequence unchanged, therefore this is an alternative sequence of $\sim$ steps between $(N_1,\alpha_1)$ and $(N_2,\alpha_2)$ of the form $(N_1,\alpha_1) \sim_\downarrow^* (N_1', \alpha') \sim_c^* (N_2', \alpha') \sim_\uparrow^* (N_2,\alpha_2)$.
-\end{proof}
-
-If $(X,\alpha) \sim_c (Y,\alpha)$ then the final skeletons in $X$ and $Y$ are equal, and $(X,\beta) \sim_c (Y,\beta)$ for any $\beta$ that occurs in this skeleton, therefore $\sim_c$ can be considered to apply to reduction sequences even without positions, $X \sim_c Y$ iff $(X,\cdot) \sim_c (Y,\cdot)$.
-The structure of $\sim_c$ can be seen more clearly by choosing a canonical example from each equivalence class.
-For this, the member of the class that's in call-by-value order is used.
-A reduction sequence is in call-by-value order if for every pair of reductions in the sequence at positions $\alpha$ and $\beta$, where the reduction at $\alpha$ happens first, either $\alpha \leq \beta$ or $\alpha$ and $\beta$ are disjoint, with $\alpha$ to the left of $\beta$ (i.e.~the first elements of the sequences $\alpha, \beta$ where they differ are $@_1$ and $@_2$, $\underline f_i$ and $\underline f_j$ for $i < j$, or $\textsf{if}_i$ and $\textsf{if}_j$ for $i < j$, in that order).
-Note that although the reductions that occur in the sequence are in call-by-value order, it may not be the actual call-by-value reduction sequence starting from that point because some redexes may remain un-reduced even when other reductions that should happen afterwards occur.
-It is only required that the reductions that do occur occur in the correct order.
-
-\begin{lemma} \label{canonicalCousins}
-For any reduction sequence $X$, there is a unique reduction sequence $X_{cbv}$ that is related to $X$ by $\sim_c$ and is in call-by-value order, and if $X \sim_c^* Y$, then $X_{cbv} = Y_{cbv}$.
-\end{lemma}
-
-\lo{As defined, $\sim_c$ is not symmetric.}
-
-\begin{proof}
-
-Define $X_n$ recursively so that $X_0 = X$ and if $X_n$ is not in CBV order, take the last pair of reductions in $X_n$ which aren't in CBV order.
-They are adjacent in the sequence, as the CBV order is a total order.
-Let the positions of these reductions be $\alpha$ and $\beta$ respecively.
-If they are disjoint, they can be considered as one of the branches of $\sim_c$ case a.
-Otherwise, $\beta < \alpha$, and the sub-skeleton at position $\beta$ at the appropriate point in the reduction sequence is a redex with non-trivial subterms, therefore it's of one of the forms $\tif{\skeletonPlaceholder>0}{A}{B}$, $(\lambda x.A) B$ or $\tY \lambda x.
-A$, therefore the reductions $\alpha$ and $\beta$ form the $O_2$ branch of one of the cases b, c, d, e or f of $\sim_c$.
-In either case, define $X_{n+1}$ as equal to $X_n$ except taking the other (left) branch, so that the order of the reductions at $\alpha$ and $\beta$ are switched (although the equivalent(s) of the reduction at $\alpha$ may not actually be at that position).
-
-This sequence of swaps rearranges the reductions in $X$ rather like insertion-sort (but in some cases duplicating or deleting the new element), and eventually reaches some $X_n$ that is in CBV order.
-This is defined to be $X_{cbv}$.
-
-\paragraph{}
-If $Y \sim_c X \sim_c^* X_{cbv}$, and the reductions involved in $Y \sim_c X$ are not the last pair of reductions in $Y$ that aren't in CBV order, the order of the $\sim_c$s can be rearranged so that the sequence $Y \sim_c^* X_{cbv}$ is the canonical sequence given above, therefore $Y_{cbv} = X_{cbv}$.
-To be more precise about this rearrangement, consider the sequence of $\sim_c$ steps from $Y$ to $X_{cbv}$.
-The first step takes some sub-sequence of the reduction sequence $Y$ and swaps the order of those reductions.
-If $X = X_{cbv}$, then $X = Y_1 = Y_{cbv}$ already.
-Otherwise, consider the cases according to whether the image of this subsequence in $X$ occurs earlier than the next pair of reductions to be swapped (i.e.~the last pair of reductions in $X$ that occur not in CBV order).
-If not, then either the step $Y \sim_c X$ changes this subsequence into CBV order, or out of it.
-If it is into, then it was the last out-of-order pair in $Y$ therefore the whole sequence $Y \sim_c^* X_{cbv}$ is already in the canonical order and $Y_{cbv} = X_{cbv}$.
-If it switches the pair to the wrong order, then the result of that switch is the last out-of-order pair in $X$ therefore $Y = X_1$, and the sequence $X_1 \sim_c^* X_{cbv}$ is just a suffix of the sequence $X \sim_c^* X_{cbv}$ therefore $(X_1)_{cbv} = X_{cbv}$ and again $Y_{cbv} = X_{cbv}$.
-
-The remaining cases are when the subsequence of reduction steps involved in $Y \sim_c X$ occur before the last out-of-order pair in $X$, but possibly overlapping.
-If there is no overlap, then the $\sim_c$ steps do not interfere with each other at all, and can simply be performed in the other order.
-This gives a different sequence of $\sim_c$ steps $Y \sim_c Y_1 \sim_c X_1 \sim_c^* X_{cbv}$.
-As $Y_{cbv} = (Y_1)_{cbv}$, we can proceed by induction in this case and use the fact that $(Y_1)_{cbv} = (X_1)_{cbv} = X_{cbv}$ to acheive the result.
-
-In the other case, there is some overlap between the regions involved in $Y \sim_c X$ and $X \sim_c X_1$, but the last reduction in $X \sim_c X_1$ is later than the last reduction in $Y \sim_c X$.
-There are only 2 reductions in $X$ involved in $X \sim_c X_1$, therefore the region involved in $Y \sim_c X$ ends on the first of the reductions involved in $X \sim_c X_1$.
-Now consider the specific sequence of positions of the reductions in $X$ involved in either of these steps.
-Let the last of these positions be $\alpha$, the first $\gamma$, and the second-last, third-last and so on $\beta_1, \beta_2, \dots$ respectively, and let the positions of the redexes in $Y$ involved in $Y \sim_c X$ be $\beta$ then $(\gamma_i)$ in order..
-Then $X \sim_c X_1$ swaps $\beta_1$ and $\alpha$, and $\alpha$ comes before $\beta_1$ in CBV order (i.e.~$\alpha < \beta_1$ or $\alpha$ is left of $\beta_1$), and the step $Y \sim_c X$ either proceeds forwards towards CBV order, swapping $\beta$ below $\gamma_1 = \gamma$ to form $\gamma$ then $(\beta_i)$, or proceeds backwards, taking $Y$ even farther from CBV order, swapping the $(\gamma_i)$ to above $\beta$, forming $\gamma$ then $\beta_1 = \beta$.
-In each of the cases below, it is required to show that the canonical sequence $(X_i)_i$ eventually reaches some term which is the same as one in $(Y_j)_j$, and from that point on, they match, therefore they reach the same end result: $X_{cbv} = Y_{cbv}$.
-
-\begin{itemize}
-  \item $Y \sim_c X$ in the backwards direction (so that $Y$ is closer to CBV order than $X$ is): By swapping the roles of $X$ and $Y$, and $\beta$ and $\gamma$, this is equivalent to the $Y \sim_c X$ forwards case.
-The assumption that $\alpha$ is before $\beta_1$ in CBV order (so that $\beta_1, \alpha$ is the pair to be swapped in $X \sim_c X_1$) maps to the assumption that $\alpha$ is before $\gamma_1$ (so that the result doesn't just follow trivially by $Y_1$ being equal to $X$) and vice-versa.
-Then one of the cases below establishes that $X_i = Y_j$ for some $i$ and $j$, therefore after swapping back, $Y_i = X_j$.
-\item $Y \sim_c X$ forwards and $\alpha$ comes after $\gamma$ in CBV order: In this case, the $\gamma$ and $\alpha$ reduction steps are already in the correct order in $Y$, therefore $\beta$ and $\gamma$ are already the last out-of-order pair of reductions in $Y$, $Y_1 = X$, and $Y \sim_c X \sim_c^* X_{cbv}$ is already the canonical sequence from $Y$ therefore $Y_{cbv} = X_{cbv}$.
-\item $Y \sim_c X$ forwards and $\alpha$ is disjoint from $\gamma$ and $\beta$: In this case, the sequence $(X_i)_i$ starts by swapping $\alpha$ and $\beta_1$ by case a, then because $\gamma$ is disjoint from $\alpha$ and comes after is in CBV order, $\gamma$ is right of $\alpha$, therefore either all the remaining $\beta_i$s, are $\geq \gamma$ (if $\beta$ and $\gamma$ aren't disjoint) or there are no remaining $\beta_i$s (if $\beta$ and $\gamma$ are disjoint, in which case they swap by case a and there is only $\beta_1$).
-The sequence $(X_i)_i$ therefore proceeds to swap all the remaining $\beta_i$s then $\gamma$, in order, below $\alpha$, resulting in $X_i$ for some $i$ having as its relevant sub-reduction-sequence $\alpha, \gamma, \beta_k, \dots, \beta_1$.
-The other sequence $(Y_j)_j$ starts by swapping $\gamma$ then $\beta$ below $\alpha$, then swapping $\gamma$ and $\beta$.
-Because $\alpha$ is disjoint from $\gamma$, the reduction there does not affect the sub-skeleton at position $\gamma$, therefore this last $\sim_c$ step proceeds identically to how it did in $Y \sim_c X$, resulting in $\gamma, \beta_k, \dots, \beta_i$.
-Overall, this results in $Y_3$ having the relevant sub-reduction-sequence $\alpha, \gamma, \beta_k, \dots, \beta_1$, therefore $Y_3 = X_i$ for the aforementioned value of $i$.
-\item $Y \sim_c X$ forwards, $\alpha \leq \gamma$ and $\alpha$ (and therefore also $\gamma$) is disjoint from $\beta$: In this case, $(X_i)_i$ proceeds by swapping $\beta_1 = \beta$ below $\alpha$ by case a, then swapping $\gamma$ below $\alpha$ by one of the other cases, and $(Y_j)_j$ proceeds by swapping $\gamma$ and $\alpha$, then swapping $\beta$ below $\alpha$ then all of the images of $\gamma$ (which are $\geq \alpha$ and therefore left of $\beta$).
-In both cases, the subterm at $\alpha$ when $\alpha$ and $\gamma$ are swapped is unaffected by the reduction at $\beta$, therefore it produces the same result in both cases, therefore the overall sequence in both cases is $\alpha$ then the images of $\gamma$ then $\beta$.
-\item Both $\gamma$ and $\beta$ are $> \alpha$ but they're disjoint from each other, and either the skeleton at position $\alpha$ is of the form $\tif{\skeletonPlaceholder > 0}{A}{B}$ or it is of the form $A B$ and either both $\gamma$ and $\beta$ are $> \alpha;@_1$ or they are both $> \alpha;@_2$: In this case, $(X_i)_i$ proceeds by swapping $\beta$ then $\gamma$ below $\alpha$, in each case possibly forming 0 or multiple images.
-The only way the number of images of each of these positions can be different is if the subskeleton at $\alpha$ is an $\textsf{if}$, and there are 0 of one of them and 1 of the other, in which case they're trivially in the correct order already.
-Otherwise, all of the images of $\beta$ and $\gamma$ are disjoint from one another, therefore none of the positions, or their relative order, change when they are swapped, therefore the next part of the sequence $(X_i)_i$ is just insertion-sort running on the images of $\beta$ and $\gamma$ with $\sim_c$ case a swaps until they're in the correct order.
-The sequence $(Y_j)_j$ starts by swapping $\gamma$ then $\beta$ below $\alpha$, then as before, sorting the images into the correct order by case a swaps.
-The images originally formed of $\gamma$ and $\beta$ are the same in both cases, therefore the final order is the same, so $X_i = Y_j$ for some $i$ and $j$.
-\item $Y \sim_c X$ forwards, $\alpha;@_1 < \gamma$, and $\alpha;@_2 < \beta$: Let $x$ be the variable involved in the $\beta$-reduction at $\alpha$, so that the sub-skeleton at $\alpha$ is $(\lambda x.
-A) B$ for some $A, B$, where $\gamma = \alpha;@_1;\lambda;\gamma'$ and $A \to A'$ at $\gamma'$, and $\beta = \alpha;@_2;\beta'$, and $B \to B'$ at $\beta'$.
-In this case, $(X_i)_i$ starts by swapping $\beta$ below $\alpha$ by case e, producing one image of $\beta$ for each $x$ in $A'$, then swapping $\gamma$ below $\alpha$ by case d, producing a single image $\alpha;\gamma'$.
-For each of the instances of $x$ in $A$ left of $\gamma'$, the reduction at $\alpha;\gamma'$ is then swapped below the corresponding image of $\beta$.
-The sequence $(Y_j)_j$ starts by swapping $\gamma$ below $\alpha$ then $\beta$ below $\alpha$, but in this case the images of $\beta$ produced may be different.
-In this case the swap happens earlier in the reduction sequence than $\gamma$, so that the body of the lambda is still $A$ rather than $A'$, and the reduction at $\gamma$ may rearrange or change the number of instances of $x$.
-Next, each of the images of $\beta$ to the right of $\alpha;\gamma'$ are swapped with it by case a.
-At this point, the next few reductions before $\alpha;\gamma'$ will in general be images of $\beta$ at positions inside $\alpha;\gamma'$.
-The subskeleton at position $\alpha;\gamma'$ before these reductions is $(A|\gamma')[B/x]$, then by the images of $\beta$ this reduces to $(A|\gamma')[B'/x]$, then it reduces at its root position ($\alpha;\gamma'$ in the overall skeleton) to $(A'|\gamma')[B'/x]$.
-For each position of an $x$ in $(A|\gamma')$, there are 0 or more corresponding positions of $x$ in $(A'|\gamma')$, depending on what type of reduction $A \to A'$ is.
-Because $B$ and $B'$ cannot contain any instances of the variable involved in the reduction $A \to A'$ (if it's a $\beta$-reduction or $\tY$-reduction), each instance of $B$ in $(A|\gamma')[B/x]$ has the same set of images in $(A'|\gamma')[B/x]$ as the corresponding $x$ in $(A|\gamma')$.
-When these images of $\beta$ that overlap with the image of $\gamma$ are swapped below it, each of them has its own images, one at each position of $B$ in $(A'|\gamma')[B/x]$ corresponding to its original copy of $B$ in $(A|\gamma')[B/x]$.
-After all of them are swapped below $\alpha;\gamma'$ (and rearranged among themselved by case a), there is therefore one reduction at each copy of $B$ in $(A'|\gamma')[B/x]$, i.e.~one at position $\beta'$ relative to each $x$ in $(A'|\gamma')$.
-They are therefore the same as the images of $\beta$ in $X_1$ that overlap with $\alpha;\gamma'$, because those are also one reduction at position $\beta'$ relative to each $x$ in $(A'|\gamma)$.
-Both $(X_i)_i$ and $(Y_j)_j$ therefore eventually reach a point where the relevant sub-reduction-sequence is $\alpha$, the images of $\beta$ to the left of $\alpha;\gamma'$, $\alpha;\gamma'$ itself, the aforementioned images of $\beta$ that overlap with $\alpha;\gamma'$, then all the images of $\beta$ to the right of $\alpha;\gamma'$.
-\item $Y \sim_c X$ forwards, $\gamma$ and $\beta$ are disjoint and both $> \alpha$, and the reduction at $\alpha$ is a $\tY$-reduction: This is similar to the previous case, but more complicated, so it will need some definitions to explain properly what's going on.
-Let the term at $\alpha$ before any of the reductions be $\tY \lambda x.
-A$, let $\gamma = \alpha;\tY;\lambda;\gamma'$, let $\beta = \alpha;\tY;\lambda;\beta'$, let $A | \beta' = B \to B'$ with the reduction at $\cdot$, let $A | \gamma' = C \to C'$ with the reduction at $\cdot$, let the positions in $A$ where $x$ occurs that are left of $\gamma'$, between $\gamma'$ and $\beta'$ (but still disjoint from both), and right of $\beta'$ respectively be $(\delta^l_i)$, $(\delta^m_i)$ and $(\delta^r_i)$, let the positions in $B$, $C$, $B'$ and $C'$ respectively where $x$ occurs be $(\delta^B_i)$, $(\delta^C_i)$, $(\delta^{B'}_i)$ and $(\delta^{C'}_i)$ (so that all the positions in $A$ where $x$ occurs, in left-to-right order, are $(\delta^l_i)_i, (\delta^C_i)_i, (\delta^m_i)_i, (\delta^B_i)_i, (\delta^r_i)_i$).
-The sequence $(X_i)_i$ starts by swapping $\beta$ then $\gamma$ below $\alpha$.
-With the shorthand that $\gamma'(\epsilon) = \alpha;\lambda;@_1;\epsilon;\tY;\lambda;\gamma'$, $\gamma'_0 = \alpha;\lambda;@_1;\gamma'$, and similarly for $\beta'$, the relevant portion of $X_2$ is then $\alpha$, $(\gamma'(\delta^l_i))_i$, $\gamma'_0$, $(\gamma'(\delta^{C'}_i))_i$, $(\gamma'(\delta^m_i))_i$, $(\gamma'(\delta^B_i))_i$, $(\gamma'(\delta^r_i))_i$, $(\beta'(\delta^l_i))_i$, $(\beta'(\delta^{C'}_i))_i$, $(\beta'(\delta^m_i))_i$, $\beta'_0$, $(\beta'(\delta^{B'}_i))_i$, $(\beta'(\delta^r_i))_i$.
-Next in $(X_k)_k$, for each $i$, $\gamma'(\delta^r_i)$ are swapped past all the images of $\beta$ until it is immediately before $\beta'(\delta^r_i)$ by case a, then for each $i$, $\gamma'(\delta^B_i)$ is swapped past all the images of $\beta$ until $\beta'_0$ by case a, then swapped with $\beta'_0$ by one of the other cases depending on what type of reduction $B \to B'$ is and the relative positions.
-Even if it's case d or f, all of the images of $\gamma'(\delta^B_i)$ are disjoint from each other (and from the positions of the other reductions after $\beta'_0$) because the redex, $C$, doesn't contain any instances of the variable involved in the reduction $B \to B'$.
-Expanding the definitions, this is swapping $\alpha;\lambda;@_1;\delta^B_i;\tY;\lambda;\gamma'$ with $\alpha;\lambda;@_1;\beta'$ at a point in the reduction sequence where the subskeleton at $\alpha;\lambda;@_1;\beta'$ is $B$ with a mixture of $\tY \lambda x.
-A$ and $\tY \lambda x.
-A[C'/\gamma']$ substituted for its occurrences of $x$, and one of these occurrences of $x$ is at $\delta^B_i$.
-For each $\delta^B_j$, there are some corresponding $\delta^{B'}_k$, where the images of the $x$ at $\delta^B_j$ end up after the reduction $B \to B'$.
-These cover all the $\delta^{B'}_k$, and uniquely, and for each $\delta^B_i$, the images of $\zeta;\delta^B_i;\theta$ after swapping it with $\zeta;\beta'$ are precisely $(\zeta;\delta^{B'}_k;\theta)$ for those same values of $k$, therefore after swapping all of the $(\gamma'(\delta^B_i))_i$ down past $\beta'_0$, their images are $(\gamma'(\delta^{B'}_i))_i$.
-After swapping with $\beta'_0$, each of these images than swaps by case a to take its place among the $(\beta'(\delta^{B'}_i))_i$.
-Next up in $(X_i)_i$, the $(\gamma'(\delta^m_i))_i$s and the $(\gamma'(\delta^{C'}_i))_i$s swap down past some images of $\beta$ they're disjoint from to take their place before their matching image of $\beta$, then $\gamma'_0$ swaps past the $(\beta'(\delta^l_i))_i$, stopping immediately before $\beta'(\delta^{C'}_1)$, then the $(\gamma'(\delta^l_i))_i$ and the $(\beta'(\delta^l_i))_i$ mix.
-The end result of the rearrangement of (this subsequence of) the reduction sequence is therefore $\alpha$, $(\gamma'(\delta^l_i), \beta'(\delta^l_i))_i$, $\gamma'_0$, $(\gamma'(\delta^{C'}_i), \beta'(\delta^{C'}_i))_i$, $(\gamma'(\delta^m_i), \beta'(\delta^m_i))_i$, $\beta'_0$, $(\gamma'(\delta^{B'}_i), \beta'(\delta^{B'}_i))_i$, $(\gamma'(\delta^r_i), \beta'(\delta^r_i))_i$.
-
-  The other sequence, $(Y_j)_j$, proceeds similarly, but with the images of $\gamma$ starting out after the images of $\beta$.
-It swaps all the images of $\beta$ that are right of $\gamma_0$ to the appropriate places by case a, then swaps all the $(\beta'(\delta^C_i)_i$ first to then past $\gamma'_0$, resulting in $(\beta'(\delta^{C'}_i))_i$ for the same reason as with swapping those images of $\gamma$ that overlapped with $\beta'_0$ past it in the $(X_i)_i$ case above, then finally swapping the $(\beta'(\delta^l_i))_i$s and the $(\gamma'(\delta^l_i))_i$s into the correct order.
-As required, this produces the same result as in $(X_i)_i$.
-  \item $Y \to X$ forwards, and $\alpha < \gamma < \beta$: Let $\gamma'$ be such that $\gamma = \alpha;\textsf{if}_i;\gamma'$, $\alpha;@_1;\lambda;\gamma'$, $\alpha;@_2;\gamma'$ or $\alpha;\tY;\lambda;\gamma'$, so that $\gamma'$ is the freely varying later part of $\gamma$ as in the definition of $\sim_c$, whichever case applies to swapping $\gamma$ and $\alpha$.
-Similarly, let $\beta'$ be the freely varying part of $\beta$ relative to $\gamma$.
-Let the images of $\gamma$ after swapping it with $\alpha$ (where this swap occurs later in the reduction sequence than $\beta$) then be $(\alpha;\delta_i;\gamma')_i$, and the images of $\beta$ after swapping it with $\gamma$ be $(\gamma;\epsilon_i;\beta')_i$.
-Depending on the skeleton at $\alpha$, and $\gamma$'s position within it, $(\delta_i)_i$ may be empty (case b), a singleton containing $\cdot$ (cases c and d), all the positions of the relevant variable in the lambda in the skeleton at $\alpha$ (case e), or $\lambda;@_1$ and $\lambda;@_1;\zeta;\tY;\lambda$ for each position $\zeta$ of the relevant variable (case f), but in all cases (except a, which is excluded because none of $\alpha, \gamma$ and $\beta$ are disjoint) the set of images has this general structure.
-Furthermore, the positions $(\delta_i)_i$ are determined only by the general position of $\gamma$ within $\alpha$ (the part excluded from $\gamma'$), and the positions of the relevant variable within the skeleton at $\alpha$.
-They do not depend on $\alpha$ or $\gamma$, so that if both initial positions were moved somewhere else the relative positions of the images of $\gamma$ would be unaffected, and similarly if some position within $\gamma$ were swapped with $\alpha$, the relative positions of its images would be the same (except to the extent that, in case f, the positions of the relevant variable are taken after the inner reduction takes place, so that some of those may still change).
-Because there are so many of them and they don't actually affect the multiset of positions where reductions take place, I will be ignoring swaps by case a in the proof of this case, and just assuming that all the final positions end up in the correct order.
-
-  The final set of reduction positions that I will be proving both $(X_i)_i$ and $(Y_j)_j$ reach is $\alpha, (\alpha;\delta_i;\gamma', (\alpha;\delta_i;\gamma';\epsilon_j;\beta')_j)_i$, where as usual the sequences are expanded out to the full list.
-For the sequence $(X_i)_i$, first the images of $\beta$ in $X_0$ are $(\gamma,\epsilon_i,\beta')$.
-For each of these in turn, it is swapped with $\alpha$.
-By the fact mentioned above that sub-positions are mapped to the same set of images except in case f, the images of $\gamma,\epsilon_j,\beta'$ after this swap are $(\alpha;\delta_i^j;\gamma',\epsilon_j,\beta')_i$, where $(\delta_i^j)_i$ is the same as $(\delta_i)_i$ except that in the case where this is not the first image of $\beta$ swapped below $\alpha$ and the reduction at $\alpha$ is a $\tY$-reduction (and therefore swaps with it proceed by case f), the positions of the relevant variable are taken after the reductions at $\gamma$ and $\gamma;\epsilon_k;\beta'$ for $k \leq j$ (this implies that the first of these, $\delta_i^n = \delta_i$).
-All the images $\alpha;\delta_i^j;\gamma';\epsilon_j;\beta'$ where $\delta_i^j > \lambda;@_1;\gamma';\epsilon_{j+1};\beta'$ are then swapped with $\alpha;\lambda;@_1;\gamma';\epsilon_{j+1};\beta'$ (which is $\alpha;\delta_i^{j+1};\gamma';\epsilon_{j+1};\beta'$ for some $i$).
-As in the case where $\alpha;\tY < \gamma, \beta$ and $\gamma$ and $\beta$ are disjoint, the effects of these swaps may duplicate or delete the reductions in such a way that all the remaining images of $\gamma;\epsilon_j;\beta'$ are $(\alpha;\delta_i^{j+1};\gamma';\epsilon_j;\beta'$.
-This is repeated with $\delta_i^{j+2}$ and so on until these images are all in the correct order, at which point they are $(\alpha;\delta_i;\gamma';\epsilon_j;\beta')_i$.
-After all of these are done, $\gamma$ is swapped with $\alpha$, yielding $(\alpha;\delta_i^{-1};\gamma')_i$.
-As with the images of $\beta$, the subset of these which are inside $\alpha;\lambda;@_1;\gamma';\epsilon_j;\beta'$ for each $j$ in turn are swapped with it, until they are all in the correct order and the images left of $\gamma$ are $(\alpha;\delta_i;\gamma')_i$.
-At this point, $X_i$ has the desired value.
-
-  The sequence $(Y_j)_j$ proceeds similarly.
-First $\gamma$ swaps with $\alpha$, yielding $\alpha$ and $(\alpha;\delta_i;\gamma')_i$, then $\beta$ swaps with $\alpha$, yielding one image of $\beta$ for each $\delta_i$, except that, in case f, those images at positions inside $\alpha;\lambda;@_1;\gamma'$ may differ because they are earlier in the reduction sequence than any of the images of $\gamma$.
-Each of the images of $\beta$ in turn it moved to the corresponding image of $\gamma$, except that those overlapping with $\alpha;\lambda;@_1;\gamma'$ may be duplicated or deleted on the way.
-The images of $\beta$ after just this process (which does not actually correspond to any term in $(Y_j)_j$, but it's sufficiently independent that the order doesn't matter that much) are $(\alpha;\delta_i;\gamma';\beta'')$ (where $\gamma;\beta'' = \beta$).
-The images of these for each $i$, after swapping with $\alpha;\delta_i;\gamma'$, is $(\alpha;\delta_i;\gamma';\epsilon_j;\beta')_j$, because the reductions produced by a swap are unaffected by its overall location, and the skeleton at $\alpha;\delta_i;\gamma'$ at the relevant point in the sequence is equal to the skeleton at $\gamma$ initially, therefore this is equivalent to immediately swapping $\beta$ and $\gamma$ as in $Y \sim_c X$ (except that in the case where the reduction at $\alpha$ is a $\tY$-reduction, the skeleton at $\alpha;\lambda;@_1;\gamma'$ is not actually equal to the skeleton initially at $\gamma$ because something was substituted in for the variable bound at $\alpha;\tY$, but this doesn't affect the variable involved in the reduction at $\gamma$, so this doesn't actually matter).
-Therefore also in this case, eventually the set of reductions in the relevant portion of $Y_j$ is $\alpha, (\alpha;\delta_i;\gamma', (\alpha;\delta_i;\gamma';\epsilon_j;\beta')_j)_i$, therefore it is equal to some $X_i$.
-\end{itemize}
-In summary, if the swap in $Y \sim_c X$ doesn't overlap with the swap $X \sim_c X_0$, the sequence of swaps can be rearranged until it does, and in any other case, there is some $Y_j$ that's equal to some $X_i$, therefore these sequences reach the same end point and if $Y \sim_c X$, then $Y_{cbv} = X_{cbv}$, and by chaining these together, if $Y \sim_c^* X$, $Y_{cbv} = X_{cbv}$.
-
-\paragraph{}
-If $Y$ is in CBV order and $Y \sim_c^* X$, then $Y_{cbv} = X_{cbv}$ but also $Y_{cbv} = Y$ because the sequence $(Y_i)_i$ terminates immediately therefore $X_{cbv}$ is the unique reduction sequence in call-by-value order that's related to $X$ by $\sim_c^*$.
-\end{proof}
-
-\begin{lemma}
-\label{lem:sim-M-N}
-The relation $\sim$ is defined on $L_0(M)$ with reference to a particular starting term $M$, so different versions, $\sim_M$ and $\sim_N$, can be defined starting at different terms.
-If $M \to N$, then $\sim^*_N$ is equal to the restriction of $\sim^*_M$ to $L_0(N)$.
-\end{lemma}
-
-\begin{proof}
-$\sim_N^*$ is trivially a subset of $\sim_M^*$ because $\sim_N$ is a subset of $\sim_M$.
-
-In the other direction, suppose $(N_1, \alpha_1) \sim_M^* (N_2, \alpha_2)$ where both $N_1$ and $N_2$ are descendants of $N$.
-By Lemma~\ref{downAcrossUp}, take $(N_1, \alpha_1) \sim_{p,M}^* (N'_1, \alpha') \sim_{c,M}^* (N'_2, \alpha') \sim_{p,M}^* (N_2, \alpha_2)$.
-The $\sim_p^*$ steps remain within $Rch(N)$, and $\sim_p$ does not depend on the history of the reduction sequences, therefore $(N_1, \alpha_1) \sim_{p,N}^* (N'_1, \alpha') \sim_{c,M}^* (N'_2, \alpha') \sim_{p,N}^* (N_2, \alpha_2)$.
-
-Let $X$ be the call-by-value reduction sequence related to both $N'_1$ and $N'_2$ by $\sim_{c,M}$ given by Lemma~\ref{canonicalCousins}. 
-As $M \to N$ is the first reduction in the sequence $N'_1$, it is the last to be affected by the $\sim_{c,M}$ sequence $N'_1 \sim_{c,M}^* X$ given by Lemma \ref{canonicalCousins} therefore it can be split into $N'_1 \sim_{c,M}^* Y_1 \sim_{c,M}^* X$ where $Y_1$ is in CBV order except possibly for its first reduction, which is still $M \to N$.
-Let the position of the reduction $M \to N$ be $\beta$.
-\akr{TODO: fill in the details that are missing here. It shouldn't be too hard, it's just a bit sketchy at present.} 
-The rearrangement of the reduction sequence $Y_1 \sim_{c,M} X$ consists of moving the reduction at $\beta$ down past the other reductions in $Y_1$, possibly duplicating or deleting it in the process, but not affecting the positions or the correct order for any of the other reductions.
-The reductions derived from $M \to N$ can be identified as follows:
-
-%\begin{definition}
-A position in some (reduction sequence of) term(s) is \emph{derived from} another if it is related by the reflexive transitive closure of $\rightsquigarrow$, where $(V,\gamma) \rightsquigarrow (W,\delta)$ just if $V \to W$ and one of the following cases holds:
-\begin{itemize}
-    \item $(V,\gamma) \sim_p (W,\delta)$
-    \item $\gamma = \delta$, $V \to W$ at $\epsilon$, and $\epsilon > \gamma$
-    \item $V \mid \epsilon = (\lambda x.U) Z, U | \zeta = x, \gamma = \epsilon;@_2;\theta$ and $\delta = \epsilon;\zeta;\theta$
-    \item $V \mid \epsilon = \tY (\lambda x.U), \gamma = \epsilon;\tY;\lambda;\zeta$ and $\delta = \epsilon;\lambda;@_1;\zeta$
-    \item $V \mid \epsilon = \tY (\lambda x.U), U | \zeta = x, \gamma = \epsilon;\theta$ and $\delta = \epsilon;\lambda;@_1;\zeta;\theta$
-\end{itemize}
-%\end{definition}
-
-Crucially, the reductions in $X$ can be partitioned into 2 sets: those at positions derived from $(M,\beta)$, and those at positions equal to the reductions in $Y_1$ (in the same order as they occur in $Y_1$).
-Using the same construction for $N'_2$ shows that the positions of the reductions in $Y_2$ are also the positions of the reductions in $X$ other than those derived from $(M,\beta)$, therefore $Y_1 = Y_2$ therefore $N'_1 \sim_{c,M} Y_1 \sim_{c,M} N'_2$ and every reduction sequence in this sequence starts with $M \to N$ therefore $N'_1 \sim_{c,N} N'_2$.
-
-Combining this with the $\sim_{p,N}^*$s at the beginning and end then yields the desired result that $(N_1,\alpha_1) \sim_N^* (N_2, \alpha_2)$, therefore the restriction of $\sim_M^*$ to $\sim_N$'s domain ($Rch(N)$) is a subset of $\sim_N^*$ therefore the two versions of $\sim$ match on this domain, as desired.
-\end{proof}
-
-At each reduction step $M \to N$, the sample space must be restricted from $I^{L_s(M)}$ to $I^{L_s(N)}$. 
-The injection $L_0(N) \to L_0(M)$ is trivial to define by appending $Sk(M) \to Sk(N)$ to each path, and using Lemma~\ref{lem:sim-M-N}, this induces a corresponding injection on the quotient, $L(N) \to L(M)$.
-The corresponding map $L_s(N) \to L_s(M)$ is then denoted $i(M \to N)$.
-
-Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function, so it is denoted ``$\Rightarrow$'' instead of ``$\red$'', and it relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
-\changed[lo]{We write an element of $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ as $(M', s)$ where the \emph{term} $M' \in \Lambda^0$ and $s \in  I^{L_s(M')}$.}
-\begin{align*}
-& (M,s) \Rightarrow (N,s \circ i(M \to N)) \text{ if either} \\
-& \qquad \text{$M \to N$ and the redex is not $\tsample$, or} \\
-& \qquad \text{$M \mid \alpha = \tsample$, $N = M[\underline{s(Sk(M),\alpha)}/\alpha]$ and $\lambda$ does not occur after $@_2$ or $\tY$ in $\alpha$}
-\end{align*}
-
-\paragraph{}
-\begin{lemma} \label{churchRosser}
-The relation $\Rightarrow$ is Church-Rosser.
-\end{lemma}
-\begin{proof}
-Suppose that $(M,s) \Rightarrow^* (M_1,s_1)$ and also $(M,s) \Rightarrow^* (M_2,s_2)$, then it is required to prove that there is some $(M',s')$ such that both $(M_1,s_1) \Rightarrow^* (M',s')$ and $(M_2,s_2) \Rightarrow^* (M',s')$.
-First consider the special case where $(M,s) \Rightarrow (M_1,s_1)$ and $(M,s) \Rightarrow (M_2,s_2)$, with only a single step in each case.
-Let the positions of the redexes in $M \to M_1$ and $M \to M_2$ be $\alpha_1$ and $\alpha_2$ respectively.
-
-First consider the case that $\alpha_1$ and $\alpha_2$ are disjoint.
-Let $M_1 = M[X_1/\alpha_1]$ and similarly for $X_2$, then let $M' = M[X_1/\alpha_1][X_2/\alpha_2]$.
-As the positions are disjoint, the substitutions commute and both $M_1$ and $M_2$ reduce (with $\to$) to $M'$.
-Let $s' = s \circ i(M \to M_1) \circ i(M_1 \to M')$. 
-The injection $i(M \to M_1) \circ i(M_1 \to M')$ consists of prepending $M \to M_1 \to M'$ to each reduction sequence, but by case a of $\sim_c$, this is equivalent to prepending $M \to M_2 \to M'$, which is $i(M \to M_2) \circ i(M_2 \to M')$.
-In the case that the reduction $M \to M_2$ isn't a $\tsample$-reduction, this is enough to establish that $(M_1, s_1) \Rightarrow (M', s')$ (and similarly if the redex of $M \to M_1$ isn't $\tsample$, $(M_2,s_2) \Rightarrow (M',s')$).
-If it is $\tsample$ though, in order for it to be the case that $(M_1, s_1) \Rightarrow (M', s')$, it is additionally necessary that $X_2$, the result of the reduction at $\alpha_2$, be $s_1(M_1, \alpha_2)$, which follows from the fact that $(M, \alpha_2) \sim (M_1, \alpha_2)$ by case 1 of $\sim_p$.
-The case that the reduction $M \to M_1$ is a $\tsample$-reduction is similar.
-
-The case that $\alpha_1 = \alpha_2$ is trivial, because there is at most one possible $\Rightarrow$ reduction at any given position, therefore $(M_1, s_1) = (M_2, s_2)$ already.
-
-The remaining case is that $\alpha_1 < \alpha_2$ or $\alpha_1 > \alpha_2$. 
-Assume without loss of generality that $\alpha_1 < \alpha_2$. 
-For each possible case of what type of redex $M | \alpha_1$ is, and $\alpha_2$'s position within it, there is a corresponding case of $\sim_c$, and similarly to the case where $\alpha_1$ and $\alpha_2$ are disjoint, the term $O_1 = O_2$ from the definition of $\sim_c$ is a suitable value of $M'$.
-The term $M \mid \alpha_1$ can't be $\tsample$, because it has strict subterms, but the case that $M \mid \alpha_2 = \tsample$ is still somewhat more complicated. 
-$\alpha_2$ can't be within $\alpha_1 ; @_2$ or $\alpha_1 ; Y$ (cases e and f of $\sim_c$) because those are not valid positions to reduce a $\tsample$, 
-and the case that $\alpha_2$ is within the branch of an if statement that is deleted by the reduction at $\alpha_1$ (case b) doesn't actually present a problem because there is no corresponding reduction to $M \to M_2$ in the other branch,
-which leaves cases c and d, that $\alpha_2$ is in the $\tif{\cdot}{\cdot}{\cdot}$ branch that isn't deleted, and that $\alpha_2 < \alpha_1 ; @_1 ; \lambda$.  
-The values that the $\tsample$s take in these cases match because $(M,\alpha) \sim_p (M_1, \alpha;\beta)$ by cases 3 and 2 of $\sim_p$ respectively.
-
-\paragraph{}
-In the case that $(M,s) \Rightarrow^* (M_1,s_1)$ or $(M,s) \Rightarrow^* (M_2,s_2)$ by multiple steps, it is possible to repeatedly replace a pair of the form $A \Leftarrow B \Rightarrow C$ by $A \Rightarrow^* D \Leftarrow^* C$ by the construction above, but it is not immediate that this process terminates, because each $\Rightarrow$ or $\Leftarrow$ may be replaced by multiple, so the sequence of $\Rightarrow$s and $\Leftarrow$s from $(M_1,s_1)$ to $(M_2,s_2)$ may get longer at some steps.
-However, termination can be proved by noting that the structure of this process is identical to the process of swapping $\sim_\uparrow$ and $\sim_\downarrow$ steps in Lemma \ref{downAcrossUp}.
-To be more precise, in this case there is a sequence of $\Leftarrow$ and $\Rightarrow$ steps, each of which has an associated reduction position and initial and final terms, and if a $\Leftarrow$ immediately precedes a $\Rightarrow$, they may be swapped to produce some number of $\Rightarrow$s, followed by some number of $\Leftarrow$s.
-In the case of Lemma \ref{downAcrossUp}, there is a sequence of $\sim_\uparrow$ and $\sim_\downarrow$ steps, each of which has an associated reduction position and initial and final skeletons, and if a $\sim_\uparrow$ immediately precedes a $\sim_\downarrow$, they may be swapped to produce some number of $\sim_\downarrow$s followed by some number of $\sim_\uparrow$s.
-In both cases, the number and reduction positions of the resultant steps are determined by the case of $\sim_c$ that matches the way the initial reduction positions overlap, and the initial skeleton (or the skeleton of the initial term).
-The same argument that the process in Lemma \ref{downAcrossUp} terminated is therefore applicable here too.
-At every stage, the sequence of reductions that results from one of the initial $\Rightarrow$s is a parallel sequence, and swapping a parallel sequence of $\Rightarrow$s with a $\Leftarrow$ always terminates, therefore each $\Rightarrow$ in turn can be moved past all of the $\Leftarrow$s, and the process as a whole will terminate in a state where all of the $\Rightarrow$s precede all of the $\Leftarrow$s, i.e.~a pair of reduction sequences $(M_1,s_1) \Rightarrow^* (M',s') \Leftarrow^* (M_2,s_2)$.
-\end{proof}
-
-\begin{lemma} \label{simpDoesn'tMix}
-If $A$ is some descendant of $M$ and $(A,\gamma) \sim^* (M,\delta)$, then $(A,\gamma) \sim_p^* (M,\delta)$, with the length of the reduction sequences decreasing by one each step from $A$ to $M$.
-\end{lemma}
-\begin{proof}
-This is a simple induction on $\sim^*$.
-In the base case, $(A,\gamma) = (M,\delta)$ therefore $(A,\gamma) \sim_p^* (M,\delta)$ trivially.
-Otherwise, suppose that $(M,\delta) \sim_p^* (B,\epsilon) \sim (A,\gamma)$.
-Either $(B,\epsilon) \sim_p (A,\gamma)$ or $(B,\epsilon) \sim_c (A,\gamma)$.
-In the first case, either $B \to A$, in which case the result follows directly, or $A \to B$, in which case the fact that each (reduction sequence of) skeleton(s) has only one parent implies that a $(M,\delta) \sim_p^* (A,\gamma)$ directly as a subsequence of the path to $(B,\epsilon)$.
-
-In the $\sim_c$ case, consider the definition of $\sim_c$.
-Either $O'_1 = A$ and $O'_2 = B$ or vice-versa.
-As $M \to^* N \to^* O_1 \to^* B$ and $(M,\delta) \sim_p^* (B,\epsilon)$, and $\sim_p$ only relates positions in a term and its parent, there are some positions $\zeta, \theta$ such that $(M,\delta) \sim_p^* (N,\zeta) \sim_p^* (O_1, \theta) \sim_p^* (B,\epsilon)$.
-It follows that $(O_2, \theta) \sim_p^* (A,\gamma)$ by following the same path, therefore it suffices to provide the only missing portion of the path from $M$ to $A$, i.e.~to prove that $(N,\zeta) \sim_p^* (O_2,\theta)$ given $(N,\zeta) \sim_p^* (O_1,\theta)$ (or vice-versa).
-
-If $\zeta$ is disjoint from all the positions of reduction from $N$ to $O_1$ and $O_2$ (and consequently $\zeta = \theta$), this follows from case 1 of $\sim_p$.
-Otherwise, this can be proved by taking cases from the definition of $\sim_c$.
-This is rather long, but all of the cases are similar.
-The general idea is that the reductions from $N$ to $O_1$ correspond to the reductions from $N$ to $O_2$, so that if a position is related by $\sim_p$ across that reduction, it is related in the other branch for the same reason.
-Case d, where $B = O'_1$ rather than $O'_2$, is given here in more detail as an illustrative example:
-
-Let $I$ be $N$ reduced at $\alpha$, and $J$ be $N$ reduced at $\alpha;@_1;\lambda;\beta$, so that $N \to I \to O_1$ and $N \to J \to O_2$.
-All of the reduction positions are $\geq \alpha$, and $\zeta$ is not disjoint from all of them, therefore $\zeta$ is not disjoint from $\alpha$.
-Let $\iota$ be the position such that $(N,\zeta) \sim_p (I,\iota) \sim_p (O_1,\theta)$.
-The fact that $(N,\zeta) \sim_p (I,\iota)$ implies that $\zeta > \alpha;@_1;\lambda$.
-Let $\zeta = \alpha;@_1;\lambda;\kappa$ and $\theta = \alpha;\kappa'$.
-If $\kappa$ is disjoint from $\beta$, then $(N,\zeta) \sim_p (J,\zeta) \sim_p (O_2,\alpha;\kappa') = (O_2;\theta)$ by cases 1 and 2 of $\sim_p$.
-In the other case, that $\kappa$ isn't disjoint from $\beta$, $\kappa > \beta$ because none of the positions $<= \alpha;@_1;\lambda;\beta$ in $N$ are related to any position in $I$ by $\sim_p$.
-As $(I,\alpha;\kappa) \sim_p (O_1,\alpha;\kappa')$ (with the redex at $\alpha;\beta$), for exactly the same reason $(N,\alpha;@_1;\lambda;\kappa) \sim_p (J,\alpha;@_1;\lambda;\kappa')$ (with the redex at $\alpha;@_1;\lambda;\beta$).
-Because $(N,\alpha;@_1;\lambda;\kappa) \sim_p (J,\alpha;@_1;\lambda;\kappa')$, $N|\alpha;@_1;\lambda;\kappa = J|\alpha;@_1;\lambda;\kappa'$, and $N|\alpha;@_1;\lambda;\kappa \neq $ the variable of $N|\alpha;@_1$, therefore $J|\alpha;@_1;\lambda;\kappa'$ is also not the variable therefore $(J,\alpha;@_1;\lambda;\kappa') \sim_p (O_2, \alpha;\kappa') = (O_2, \theta)$ by case 2 of $\sim_p$.
-Combining these results, $(N,\zeta) \sim_p^* (O_2;\theta)$ as desired.
-\end{proof}
-
-\begin{lemma} \label{lem:redexDestroyed}
-If $M \to N$, with the redex at position $\alpha$, then no position in any term reachable from $N$ is related by $\sim^*$ to $(M,\alpha)$.
-\end{lemma}
-\begin{proof}
-Suppose on the contrary that $(M,\alpha) \sim^* (A,\beta)$, where $N \to^* A$, then by Lemma \ref{simpDoesn'tMix}, $(M,\alpha) \sim_p^* (A,\beta)$.
-$M \neq A$ therefore $(M,\alpha) \sim_p$ some position in $N$, but in all the cases of the definition of $\sim_p$, no position in the child term is related to the position of the redex.
-\end{proof}
-
-\paragraph{}
-The reduction relation $\Rightarrow$ is nondeterministic, so it admits multiple possible reduction strategies. 
-A \emph{reduction strategy} starting from a closed term $M$ is a partial function $f$ from $Rch(M)$ to positions, such that for any reachable term $N$ where $f$ is defined, $f(N)$ is a position of a redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
-Using a reduction strategy $f$, a subset of $\Rightarrow$ that isn't nondeterministic, $\Rightarrow_f$, can be defined by $(N,s) \Rightarrow_f (N',s')$ just if $(N,s) \Rightarrow (N',s')$ and $N$ reduces to $N'$ with the redex at $f(N)$.
-
-Essentially what Lemma~\ref{lem:redexDestroyed} demonstrates is that the samples taken during any reduction sequence are independent of each other. This is made more precise in the following lemmas.
-
-\begin{lemma}
-For any skeletons $M \to N$, with the redex at position $\alpha$, and measurable set of samples $S \subset I^{L_s(N)}$, $\mu(S) = \mu(\{s \in I^{L_s(M)} \mid s \circ i(M \to N) \in S\})$, and furthermore, if $M | \alpha = \tsample$, for any $S \subset I \times I^{L_s(N)}$, $\mu(S) = \mu(\{s \in I^{L_s(M)} \mid (s(M,\alpha), s \circ i(M \to N)) \in S\})$.
-\end{lemma}
-\begin{proof}
-If the result holds for all sets $S$ of the form $\{s \in I^{L_s(N)} \mid \forall j : s(j) \in x_i\}$, where $(x_j)_{j \in J}$ is a family of measurable subsets of $I$ indexed by some finite set $J \subset L_s(N)$ of positions, it also holds for all other $S$ by taking limits and disjoint unions.
-
-Take such a $(x_j)_{j \in J}$, where $x_j = x_k$ and define $K = i(M \to N)[J]$.
-Because $i(M \to N)$ is injective, it defines a bijection between $J$ and $K$.
-We can then calculate the measure $\mu(\{s \in I^{L_s(M)} \mid s \circ i(M \to N) \in S\}) = \mu(\{s \in I^{L_s(M)} \mid \forall k : s(k) \in x_{i(M \to N)^{-1}(k)}\}) = \prod_{k \in K} \mu_I(x_{i(M \to N)^{-1}(k)}) = \prod_{j \in J} \mu_I(x_j) = \mu(S)$.
-
-In the case that $M|\alpha = \tsample$, we can similarly consider only those sets $S$ of the form $x_\tsample \times \{s \in I^{L_s(N)} \mid \forall j : s(j) \in x_i\}$.
-The position $\alpha$ in $M$ is not related to any position in $L_s(N)$, therefore $(M,\alpha) \not \in K$.
-Again, we can calculate the measure $\mu(\{s \in I^{L_s(M)} \mid (s(M,\alpha), s \circ i(M \to N)) \in S\}) = \mu(\{s \in I^{L_s(M)} \mid s(M,\alpha) \in x_\tsample, \forall k : s(k) \in x_{i(M \to N)^{-1}(k)}\}) = \mu_I(x_\tsample) \prod_{k \in K} \mu_I(x_{i(M \to N)^{-1}(k)}) = \mu_I(x_\tsample) \prod_{j \in J} \mu_I(x_j) = \mu(S)$.
-\end{proof}
-
-\begin{lemma} \label{lem:independentSamples}
-For any initial term $M$, reduction strategy $f$ on $M$, natural number $n$, skeleton $N$ with $k$ holes, measurable set $T \subset \mathbb R ^ k$ and measurable set $S$ of samples in $I^{L_s(N)}$, $\mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')\}) = \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in I^{L_s(N)} : (M,s) \Rightarrow_f^n (N[r], s')\}) \mu(S)$
-\end{lemma}
-\begin{proof}
-Suppose, to begin with, that $n = 0$.
-Either $\exists r \in T : N[r] = M$, in which case both sides of the equation are $\mu(S)$, or there is no such $r$, in which case both sides of the equation are 0.
-
-For $n > 0$, suppose for induction that the lemma is true for $n - 1$, for all $N, T$ and $S$.
-If $\exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')$, the $r$ and $s'$ are necessarily unique, therefore we may assume that $T$ is the product of $k$ measurable setsets of $\mathbb R$, $(T_j)_{0 \leq j < k}$, as all of the other cases follow by taking unions and limits of these rectangles.
-Let\begin{align*}
-R_s & = \{\text{the skeleton of } O \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) = \tsample, O \to N[r]\} \\
-R_d & = \{\text{the skeleton of } O \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) \neq \tsample, O \to N[r]\}.
-\end{align*}
-Together, $R_s$ and $R_d$ contain all of the skeletons of terms that can reduce to $N$, and they're disjoint.
-For each of these skeletons, there is a map $g_O$ from the reals in $O$ to the reals in $N$, i.e.
-$g_O(r) = r'$ if $O[r] \to N[r']$ for $O \in R_d$, and $g_O(s, r) = r'$ if $O[r] \to N[r']$ with $s$ as the value the sample takes in the reduction.
-All of these functions are measurable, as they are just rearrangements of vector components, or in the case that $O|f(O) = \underline x(\skeletonPlaceholder, \dots, \skeletonPlaceholder)$, a combination fo the measurable function $x$ with a rearrangement of vector components.
-In the case that $O \in R_s$, the function $g_O$ simply inserts the sample into the vector of reals at a certain index, call it $h_O$.
-\begin{align*}
-& \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')\}) \\
-= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in S :\\
-    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\})\\
-    + & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in S :\\
-    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \\
-\end{align*}
-\begin{align*}
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}:\\
-    & g_O(r,s'(O,f(O))) \in T, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
-     + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}:\\
-     & g_O(r) \in T, s' \circ i(O[r] \to N[g_O(r)]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)}:\\
-    & s'(O,f(O)) \in T_{h(O)}, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
-    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}:\\
-    & g_O(r) \in T, s' \circ i(O[r] \to N[g_O(r)]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
-    & \mu(\{s' \in I^{L_s(O)} \mid s'(O,f(O)) \in T_{h(O)}, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S\})\\
-    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
-    & \mu({s' \in I^{L_s(O)} \mid s' \circ i(O[r] \to N[g_O(r)]) \in S}) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(T_{h(O)}) \mu(S)\\
-    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : s'(O,f(O)) \in T_{h(O)}, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S)\\
-    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \mathbb R^{k-1}, s' \in I^{L_s(O)} : g_O(r,s'(O,f(O))) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S)\\
-    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
-= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} :\\
-    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S)\\
-    + & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} :\\
-    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S) \\
-= & \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in I^{L_s(N)} : (M,s) \Rightarrow_f^n (N[r], s')\}) \mu(S)
-\end{align*}
-\end{proof}
-
-The usual call-by-value semantics can be implemented as one of these reduction strategies, given by (with $V$ a value and $T$ a term that isn't a value and $M$ a general term)
-\begin{align*}
-\cbv(T M) & = @_1 ; \cbv(T) \\
-\cbv(V T) & = @_2 ; \cbv(T) \\
-\cbv(\underline f(V_1, \dots, V_{k-1}, T, M_{k+1}, \dots, M_n)) & = \underline f_k ; \cbv(T) \\
-\cbv(\tY T) & = \tY ; \cbv(T) \\
-\cbv(\tif{T < 0}{M_1}{M_2}) & = \textsf{if}_1 ; \cbv(T) \\
-\cbv(\tscore(T)) & = \tscore ; \cbv(T) \\
-\cbv(V) & \text{ is undefined} \\
-\cbv(T) & = \cdot \text{ otherwise}
-\end{align*}
-(this last case covers redexes at the root position).
-
-A closed term $M$ terminates with a given reduction strategy $f$ and samples $s$ if there is some natural number $n$ such that $(M,s) \Rightarrow_f^n (N,s')$ where $f$ gives no reduction at $N$.
-The term terminates almost surely with respect to $f$ if it terminates with $f$ for almost all $s$.
-
-\begin{theorem} \label{thm:AstEquivalence}
-A closed term $M$ is AST with respect to $\cbv$ iff it is AST.
-\end{theorem}
-\begin{proof}
-\akr{I'm not sure whether this proof is sufficiently detailed. It would be pretty easy to fill it in more, just very verbose.}
-
-For both $\red$ and $\Rightarrow_{\cbv}$, the probability measure on the sample space can be used to define a measure on the reduction sequences of each finite length.
-The distributions are equal for each length, by induction, as follows.
-
-The base case, length 0, is trivial, because in both cases the distribution has probability 1 on the sequence containing only $M$.
-
-For the inductive case, suppose that the distributions of reduction sequences of length $n$ are equal.
-\akr{TODO: mention the case that it has already terminated by step $n$, or is about to.
-The two versions of the semantics handle it a little differently.}
-
-For any term $N$ which is not a value, there is a unique environment $E$ and redex $R$ such that $N = E[R]$.
-The position of the hole in $E$ is equal to $\cbv(N)$, so that $N|\cbv(N) = R$, as the cases in the definition of $\cbv$ match the cases in the definition of environments.
-If $R \neq \tsample$, let $R'$ be the result of reducing $R$ (which is equal in both versions of the semantics), then $E[R'] = N[R' / \cbv(N)]$ therefore $\red(N,s) = (E[R'],s)$ and $(N,s) \Rightarrow_{\cbv} (E[R'],s)$.
-The distribution of next terms, conditional on the previous term, in the case that that previous term reduces deterministically, is therefore equal for $\red$ and $\Rightarrow_{\cbv}$.
-
-In the other case that $R = \tsample$, the first (term) part of $\red(N,s)$ is $E[\underline{\pi_h(s)}]$.
-If $k$ is the number of samples already taken in this reduction sequence, this is equivalent to $E[\underline{s_0(k)}]$, where $s_0$ is the initial sample.
-For any sufficiently small neighbourhood of $N$ (i.e.~containing only reduction sequences with the same skeletons), the probability of reaching this neighbourhood is independent of $s(k)$, because it depends only on the samples taken so far, and the samples are independent.
-The distribution of $s(k)$ conditional on reaching $N$ is therefore the uniform distribution on $I$.
-Similarly for the other semantics, if the distribution of remaining samples after $n$ steps is independent of the term by Lemma \ref{lem:independentSamples}, therefore the distribution of $s(Sk(N),\cbv(N))$ conditional on $N$ is the uniform distribution on $I$.
-In either case, the distribution of the next term conditional on the previous term, in the case that it reduces randomly, is equal to the image under $r \mapsto E[\underline r]$ of the uniform distribution on $I$.
-
-In any case, the distribution of reduction sequences after $n+1$ steps, conditional on the reduction sequence after $n$ steps, as defined by $\Rightarrow_{\cbv}$ and by $\red$, is equal, any by the inductive hypothesis the distributions after $n$ steps are equal, therefore by integrating over the reduction sequence up to step $n$, the distributions on reduction sequences up to step $n+1$ are equal.
-
-The distributions on reduction sequences of any finite length as defined by $\Rightarrow_{\cbv}$ and $\red$ are therefore equal, therefore so is the probability of having reached a value after $n$ steps, therefore so is its limit, the probability of termination, therefore the probability of termination is 1 iff the probability of termination with respect to $\cbv$ is 1.
-\end{proof}
-
-\begin{theorem} \label{thm:CbvIsTerminatingest}
-If $M$ terminates with some reduction strategy $f$ and samples $s$, it terminates with $\cbv$ and $s$.
-\end{theorem}
-\begin{proof}
-  Suppose $M$ terminates with the reduction strategy $f$ and samples $s$.
-Let $(M,s) = (M_0,s_0) \Rightarrow_f \dots \Rightarrow_f (M_n,s_n)$, where $M_n$ is a value.
-By Lemma \ref{canonicalCousins}, the reduction sequence $(Sk(M_i))_i$ is related by $\sim_c^*$ to some skeletal reduction sequence $X$ which is in call-by-value order.
-For each skeletal reduction sequence $X_j$ in the sequence $(Sk(M_i))_i \sim_c \dots \sim_c X$, it is possible to define a reduction sequence of terms such that $(M,s) = (N_{j,0},s_{j,0}) \Rightarrow \dots \Rightarrow (N_{j,n_j},s_{j,n_j})$ and $Sk(N_{j,k}) = X_{j,k}$, by at each step applying the reduction at the same place as in $X_j$, essentially just filling in the reals in the skeletal reduction sequence to make it a reduction sequence of terms (it will be shown shortly that the correct branch is taken in each $\textsf{if}$-reduction).
-Assume for induction on $j$ that $(N_{j,n_j},s_{j,n_j}) = (M_n,s_n)$.
-To prove that this holds true for $j+1$ as well, it suffices to show that the corresponding term and samples immediately after the reductions involved in this $\sim_c$ step match, as from that point on, the $\Rightarrow$-sequences will match, as they start from the same point and have all the same reduction positions.
-Letting the term and remaining samples immediately before the reductions involved in this $\sim_c$ step be $(N,t)$ (which is the same in both $N_j$ and $N_{j+1}$ for the same reason again, they start at the same point and the sequences of reduction positions up to this point are the same), and the terms and samples after these reductions be $(O_1,t_1)$ and $(O_2,t_2)$, so that these match the $N$, $O_1$ and $O_2$ in the definition of $\sim_c$, $t_1 = t_2$ because both of them are, by Lemma \ref{lem:sim-M-N}, equal to the restriction of $t$ to the subset of $Rch(N)$ that starts with the reduction sequences to $O_1$ and $O_2$ respectively, and the potential positions in those subsets of $Rch(N)$ are identified with each other by $\sim$ because of the same case of $\sim_c$ that means $N_j$ and $N_{j+1}$ are related.
-If none of the relevant reductions are $\tsample$-reductions, $O_1 = O_2$ trivially.
-If there are samples taken (which must be the arbitrary reduction at $\beta$), it is also required that the same samples are taken in both branches.
-Similarly to Lemma \ref{churchRosser}, the positions of these $\tsample$-reductions in each branch are related by $\sim_p^*$, therefore they correspond to the same sample in $t$, therefore $O_1 = O_2$.
-The fact that the terms are equal also implies that the same branch of an $\textsf{if}$ statements will be taken.
-
-  \paragraph{}
-  There is therefore a reduction sequence $(M,s) = (N_0,s_0') \Rightarrow \dots \Rightarrow (N_m, s_m')$ which ends in a value ($N_m = M_n$) and is in CBV order.
-It doesn't immediately follow that this is the $\Rightarrow_{\cbv}$ reduction sequence starting from $(M,s)$ because the fact that $(N_i)_i$ is in CBV order only means that the reductions that do take place in this reduction sequence are in the correct order, not that the reduction sequence finishes once it reaches a value, or that there are no other reductions that would come earlier in CBV order that don't occur at all in $(N_i)_i$.
-Suppose that at some point $j$ in the reduction sequence, the reduction that occurs is not the next reduction in CBV order, but something else, and that the reduction that should be next is at position $\alpha = \cbv(N_i)$, and the position of the actual reduction $N_j \to N_{j+1}$ is $\beta$.
-Suppose further that $N_j$ is not a value.
-Either $\alpha = \cdot$, the root position, or the reduction at the root position is dependent in some way on the reduction at $\alpha$ (e.g.
-$\alpha = \textsf{if}_1$, or $\alpha = \underline{+};@_1$).
-In any case, all the reductions in the rest of the reduction sequence are after $\alpha$ in CBV order, therefore none of them are $\leq \alpha$, and there continues to be an unreduced redex at $\alpha$, therefore the reduction at the root position never occurs, therefore $N_m$ is not a value, which is a contradiction, therefore at the point (if any) where $(N_i)_i$ departs from the $\Rightarrow_{\cbv}$ reduction sequence starting at $(M,s)$, it has already reached a value, therefore $(M,s) \Rightarrow_{\cbv} \dots$ does reach a value eventually, i.e.~$M$ terminates with the reduction strategy $\cbv$ and samples $s$.
-\end{proof}
-
-\begin{corollary}
-If $M$ is AST with respect to any reduction strategy, it is AST.
-\end{corollary}
-\begin{proof}
-Suppose $M$ is AST with respect to $f$.
-Let the set of samples with which it terminates with this reduction strategy be $X$.
-By Theorem~\ref{thm:CbvIsTerminatingest}, $M$ also terminates with $\cbv$ and every element of $X$, and $X$ has measure 1, by assumption, therefore $M$ is AST with respect to $\cbv$ therefore by Theorem~\ref{thm:AstEquivalence} it is AST.
-\end{proof}
-
-All of the theorems on the termination of rankable terms therefore extend to other reduction strategies too.
-The proofs of Theorems \ref{thm:rankable implies termination}, \ref{thm:partial implies rankable}, \ref{thm:antitone rankable implies termination} and \ref{thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies.
-They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because Theorem \ref{thm:de groote} applies equally to any reduction strategy).
-For a reduction strategy $r$ and term $M$, just substitute $N[r(N)]$ being a $\tY$-redex for $N$ being of the form $E[\tY \lambda x.
-O]$, and $r(N)$ being undefined for $N$ being a value.
-
-The domain of definition of the ranking functions also needs to be changed from the reachable terms $Rch(M)$ to the \emph{reachable terms with respect to the reduction strategy $r$},
-\[
-Rch_r(M) = \{N \mid \exists n, N_0, \dots, N_n: N_0 = M, N_n = N, N_i \to N_{i+1} \text{ at } r(N_i)\}.
-\]
-
-To be more explicit, the modified forms of the theorems are:
-\begin{definition}\rm
-A \emph{ranking function on $M$ with respect to a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
-\begin{itemize}
-    \item $f(N) \geq 1+ f(N')$ if $N[r(N)]$ is a $\tY$-redex and $N$ reduces to $N'$ at $r(N)$
-    \item $f(N) \geq \int_I f(N[\underline{x}/r(N)]) \, \Leb(\mathrm{d}x)$ if $N[r(N)] = \tsample$
-
-    \item $f(N) \geq f(N')$ if $r(N)$ is any other redex, where $N \to N'$ at $r(N)$.
-\end{itemize}
-Any closed term for which a ranking function with respect to a reduction strategy $r$ exists is called \emph{rankable} with respect to $r$. 
-\end{definition}
-
-\begin{definition}
-A \emph{partial ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : Rch_r(M) \rightharpoonup \mathbb R$ such that
-\begin{itemize}
-    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
-    \item $f$ is defined at $M$.
-    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ at the positions specified by $r$ will eventually reach some $O$ such that either $r(O)$ isn't defined or $f(O)$ is, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
-\end{itemize}
-\end{definition}
-
-\begin{definition}\rm
-An \emph{antitone ranking function on $M$ with respect to a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ such that
-\begin{itemize}
-    \item $f(N) \geq \epsilon(f(N)) + f(N')$ if $N[r(N)]$ is a $\tY$-redex and $N$ reduces to $N'$ at $r(N)$
-    \item $f(N) \geq \int_I f(N[\underline{x}/r(N)]) \, \Leb(\mathrm{d}x)$ if $N[r(N)] = \tsample$
-
-    \item $f(N) \geq f(N')$ if $r(N)$ is any other redex, where $N \to N'$ at $r(N)$.
-\end{itemize}
-Any closed term for which an antitone ranking function with respect to a reduction strategy $r$ exists is called \emph{antitone rankable} with respect to $r$. 
-\end{definition}
-
-\begin{definition}
-An \emph{antitone partial ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : Rch_r(M) \rightharpoonup \mathbb R$ such that there exists some antitone function $\epsilon : \nnReal \to \pReal$ such that
-\begin{itemize}
-    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
-    \item $f$ is defined at $M$.
-    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ at the positions specified by $r$ will eventually reach some $O$ such that either $r(O)$ isn't defined or $f(O)$ is, and $f(N) \geq \mathbb E[f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
-\end{itemize}
-\end{definition}
-
-\begin{theorem}
-If a closed SPCF term $M$ has a ranking function, partial ranking function, antitone ranking function or antitone partial ranking function with respect to a reduction strategy $r$, then $M$ is AST with respect to $r$, and AST.
-\end{theorem}
-\lo{Give an example or two of a reduction strategy (for $\Rightarrow$) that is not $\cbv$.}
-
-\akr{NOTE. For the standard CBV semantics, this example is indeed not a problem. 
-In the alternative semantics, defining 
-\begin{align*}
-A[x] &= \tif{\tif{x>0}{I}{I} \, \tsample - 0.5 > 0}{0}{\Omega}\\
-B &= \tif{\tsample - 0.5 > 0}{0}{\Omega}
-\end{align*}
-The set of samples where it terminates is 
-\[\begin{array}{l}
-\bigcup_{r \in [0,1]} \{s \in I^{L_s(A[\tsample])} \mid s([A[\tsample]], \mathsf{if}_1;\underline{ -
-}_1;@_1;\mathsf{if}_1) = r, \\
-\qquad \qquad \hfill \qquad \qquad s([A[\tsample], A[r], \ldots, B, \mathsf{if}_1) > 0.5 \}.
-\end{array}\] 
-This is a rather unwieldy expression, but the crucial part is that $r$ occurs twice in the conditions on $s$: once as the value a sample must take, and once in the location of a sample. 
-It is this set that's unmeasurable.}
-
-\bibliographystyle{apalike}
-\bibliography{references}
-
-\lo{22 Jan: 
-\paragraph{Further directions}
-
-An obvious next step is to extend the result to the $\mathsf{score}$ construct.
-
-The following are highly topical, and could form the basis of an interesting and novel DPhil thesis.
-\begin{enumerate}
-\item Devise methods for proving (positive) a.s.~termination.
-
-- For example, develop a type system satisfying the property: if a term is typable then it is (positively) a.s.~terminating.
-See~\citep{DBLP:conf/ppdp/BreuvartL18,DBLP:conf/esop/LagoG17}.
-
-- Another approach is to develop algorithms that synthesise ranking supermartingales, following, for example, \citep{DBLP:journals/pacmpl/AgrawalC018}.
-
-\item Develop principles (e.g.~in the form of ``proof rules'') for reasoning about (positively) a.s.~termination, in the style of \citep{DBLP:journals/pacmpl/McIverMKK18}.
-
-\item Design algorithms that synthesise probabilistic invariants (\emph{qua} martingales), \`a la \cite{SchreuderO19}; see also \citep{DBLP:journals/pacmpl/HarkKGK20}.
-
-\end{enumerate}}
-
-\iffalse
-\begin{thebibliography}{9}
-\bibitem{ppcf} Thomas Ehrhard, Michele Pagani, and Christine Tasson.
-Measurable cones and stable, measurable functions: a model for probabilistic higher-order programming.
-\emph{PACMPL}, 2(POPL):59:1–59:28, 2018.
-doi: 10.1145/3158147.
-URL \href{https://doi.org/10.1145/3158147}{https://doi.org/10.1145/3158147}.
-\end{thebibliography}
-\fi
-
-\end{document}

### 2020-12-26 LICS21/related.tex
Created version for LICS21 submission in subdirectory LICS21

\section{Related work, conclusion and further directions}

### 2020-12-26 LICS21/supermartingales.tex
Created version for LICS21 submission in subdirectory LICS21

--- 2020-12-26 LICS21/ranking-supermartingales.tex
@@
-%!TEX root = probabilisticProgrammingMartingales.tex
+\section{Supermartingales}
+
+\section{Supermartingales}
+One approach to proving that a term terminates almost surely is to find some variant that is bounded below and, on average, decreases sufficiently quickly that it must eventually reach 0, similarly to the approach taken in
+%\citep{DBLP:journals/pacmpl/McIverMKK18} 
+\cite{DBLP:conf/popl/FioritiH15} for imperative programs.
+
+These variants are defined as functions from reachable terms (i.e.~possible states of the program's execution) to real numbers.
+Specifically, let the set of reachable terms from a given closed starting term $M$ be $Rch(M) := \{N \in \Lambda \mid M \to^* N \}$, with the $\sigma$-algebra induced as a subset of $\Lambda$.
+
+\begin{definition}\rm
+A \emph{ranking function on $M$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
+\begin{enumerate}
+    \item $f(E[\tY \lambda x. N]) \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z) \text{ where $z$ is not free in $N$}$
+    \item $f(E[\tsample]) \geq \int_I f(E[\underline{x}]) \, \Leb(\mathrm{d}x)$
+
+    \item $f(E[R]) \geq f(E[R'])$ for any other redex $R$, where $R \to R'$.
+\end{enumerate}
+We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that $f(N) = 0$ iff $N$ is a value, and for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.
+
+Any closed term for which a ranking (respectively, strict ranking) function exists is called \emph{rankable} (respectively, \emph{strictly rankable}). 
+\changed[lo]{For example, the term $(\tY \lambda x.x) \, \underline 0$ is not rankable.}
+It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \in \omega}$ is the reduction sequence starting from $M$, then $(f(M_n))_{n\in\omega}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales.
+\end{definition}
+
 \paragraph{}

### 2020-12-26 LICS21/applications.tex
Mainly LaTeX reformatting to fit 2-column style

% !TEX root = main.tex
\section{Applications}
\label{sec:applications}

### 2020-12-26 LICS21/supermartingales.tex
Mainly LaTeX reformatting to fit 2-column style

@@
+% !TEX root = main.tex
 \section{Supermartingales}
+\label{sec:supermartingales}
 
-\section{Supermartingales}
 One approach to proving that a term terminates almost surely is to find some variant that is bounded below and, on average, decreases sufficiently quickly that it must eventually reach 0, similarly to the approach taken in
@@
 and so, by (\ref{eqn:truncate})
-\begin{equation}
-\int_{A \cap U_i} \mu(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])) \\
-\leq 
+\begin{align}
+& \int_{A \cap U_i} \mu(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])) \nonumber \\
+& \leq 
 \int_{A \cap U_i} \mu(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\tsample])).
 \label{eqn:a u ui}
-\end{equation}
+\end{align}
 
@@
 \begin{align*}
-\int_A \mu(\dif s) \, f_{n+1}(M)[s \in \mathbf{T}_i] 
+& \int_A \mu(\dif s) \, f_{n+1}(M)[s \in \mathbf{T}_i] \\
 &= 
@@
 As an immediate corollary of \Cref{lem:key rankable}, each $f(M_n)$ is integrable:
-\[
+\begin{align*}
 \int_{S} \mu(\dif s) \, |f(M_n)(s)| 
-=
+&=
 \int_{S} \mu(\dif s) \, f(M_n)(s)
-\leq
+\\
+&\leq
 \int_{S} \mu(\dif s) \, f(M_0)(s)
@@
 f(M).
-\]
+\end{align*}
 
@@
 T_{-1}(s) & := -1 \nonumber \\
-T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value or of form } E[\tY \lambda x. N] \} \nonumber \\
+T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value} \nonumber\\
+& \qquad\qquad\qquad\qquad\qquad \hbox{or of form } E[\tY \lambda x. N] \} \nonumber \\
 X_n(s) & := f(M_n(s)) \nonumber \\

### 2020-12-26 LICS21/SPCF.tex
Mainly LaTeX reformatting to fit 2-column style
\appendices

@@
+% !TEX root = main.tex
 \section{Statistical PCF}
@@
 The one-step reduction is given by the function $\red : \Lambda^0 \times S \to \Lambda^0 \times S$ where
-\begin{equation}
+\begin{equation*}
 \red(M,s) = \left\{
     \begin{array}{ll}
-        (E[N],s) & \text{if } M = E[R], R \to N \text{ and } R \neq \tsample \\
+        (E[N],s) & \text{if } M = E[R], R \to N\\
+        & \qquad \text{ and } R \neq \tsample \\
         (E[\underline{\pi_h(s)}],\pi_t(s)) & \text{if } M = E[\tsample] \\
@@
     \end{array} \right .
-\end{equation}
+\end{equation*}
 \lo{Both $\to$ and $\red$ are called reduction relation.}
 
-The result after $n$ steps is then simply $\red^n(M,s) = \overbrace{\red(...\red(}^n M,s)...)$, and the limit $\red^\infty$ can then be defined as a partial function as $\lim_{n \to \infty} \red^n(M,s)$ whenever that sequence becomes constant by reaching a value.
+The result after $n$ steps is then simply 
+\[
+\red^n(M,s) = \underbrace{\red(\ldots\red(}_n M,s)\ldots)
+\] 
+and the limit $\red^\infty$ can then be defined as a partial function as $\lim_{n \to \infty} \red^n(M,s)$ whenever that sequence becomes constant by reaching a value.
 A term $M$ terminates for a sample sequence $s$ if the limit $\red^\infty(M,s)$ is defined.
@@
 
+\iffalse
 \lo{Alternatively, define the \emph{runtime of $M$} to be the random variable 
@@
 and $M$ is \emph{positively almost-surely terminating} (PAST) if $\expect{T_M} < \infty$.}
-
+\fi
 For example, the term 
@@
 
-\paragraph{}
+%\paragraph{}
 Note that the requirement that $M$ be closed is not important for the proofs of the theorems, but is simply required to ensure that $M$ reaches a value, rather than a general normal form like $x \, y$.
 
-\paragraph{}
+%\paragraph{}
 This definition of almost-sure termination is equivalent to that given in \citep{MakOPW21}, although the program semantics is stated in a slightly different way.

### 2020-12-26 LICS21/supermartingales.tex
\appendices

@@
 \begin{definition}\rm
-A \emph{ranking function on $M$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
+A \emph{ranking function on $M \in \Lambda^0$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
 \begin{enumerate}
-    \item $f(E[\tY \lambda x. N]) \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z) \text{ where $z$ is not free in $N$}$
+    \item $f(E[\tY \lambda x. N]) \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z)$ where $z$ is not free in $N$
     \item $f(E[\tsample]) \geq \int_I f(E[\underline{x}]) \, \Leb(\mathrm{d}x)$
-
     \item $f(E[R]) \geq f(E[R'])$ for any other redex $R$, where $R \to R'$.
 \end{enumerate}
-We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that $f(N) = 0$ iff $N$ is a value, and for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.
+We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that $f(N) = 0$ iff $N$ is a value; and for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.
 
 Any closed term for which a ranking (respectively, strict ranking) function exists is called \emph{rankable} (respectively, \emph{strictly rankable}). 
-\changed[lo]{For example, the term $(\tY \lambda x.x) \, \underline 0$ is not rankable.}
+For example, the term $(\tY \lambda x.x) \, \underline 0$ is not rankable.
+
 It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \in \omega}$ is the reduction sequence starting from $M$, then $(f(M_n))_{n\in\omega}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales.
@@
 
-\paragraph{}
+%\paragraph{}
 
@@
 
-The following is an iterated version of Doob's well-known Optional Sampling Theorem (see, e.g., \cite[\S 6.7]{AshDD00} and \cite[Theorem 7.2]{DBLP:conf/popl/FioritiH15}).
+The following is an iterated version of Doob's well-known Optional Sampling Theorem (see, e.g., \cite[\S 6.7]{AshDD00} and \cite[Thm.~7.2]{DBLP:conf/popl/FioritiH15}).
 \begin{theorem}[Optional Sampling]
@@
 \label{lem:rank-PAST}
-Let $(Y_n)_{n \in \omega}$ be a $\epsilon$-ranking supermartingale with respect to the stopping time $T$, then $T < \infty$ almost surely, and $\expect{T} \leq \dfrac{\expect{Y_0}}{\epsilon}$.
+Let $(Y_n)_{n \in \omega}$ be a $\epsilon$-ranking supermartingale with respect to the stopping time $T$, then $T < \infty$ almost surely, and $\expect{T} \leq \frac{\expect{Y_0}}{\epsilon}$.
 \end{lemma}
@@
 \begin{proof}
-We first prove by induction on $n$: 
+We first prove (by induction on $n$): for all $n \in \omega$
 \[
-\forall n \in \omega \, . \, \expect{Y_n} \leq \expect{Y_0} - \epsilon \cdot \Big(\sum_{i=0}^{n-1} \mathbb{P}[T > i]\Big).
+%\forall n \in \omega \, . \, 
+\expect{Y_n} \leq \expect{Y_0} - \epsilon \cdot \big(\textstyle\sum_{i=0}^{n-1} \mathbb{P}[T > i]\big).
 \]
-It follows that $\sum_{i=0}^{\infty} \mathbb{P}[T > i]$ converges; hence $\lim_{i \to \infty} \mathbb{P}[T > i] = 0$ and so $\mathbb{P}[T < \infty] = 1$.
+It follows that $\sum_{i=0}^{\infty} \mathbb{P}[T > i]$ converges; hence we have $\lim_{i \to \infty} \mathbb{P}[T > i] = 0$ and so $\mathbb{P}[T < \infty] = 1$.
 It then remains to observe: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=0} \mathbb{P}[T > i]$.
@@
 
-\changed[lo]{\Cref{lem:rank-PAST} is a slight extension of \citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}. 
-In Lemma 5.5 of op.~cit., the stopping time $T$ is assumed to be the random variable $s \mapsto \min\set{n \mid Y_n(s) = 0}$.}
+\changed[lo]{\Cref{lem:rank-PAST} is a slight extension of \citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15},
+wherein the stopping time $T$ is assumed to be the random variable $s \mapsto \min\set{n \mid Y_n(s) = 0}$.}
 
@@
 where each $U_i$ is determined by a skeletal term $E_i[\cdot][\tsample]$ and a number (of draws) $l_i \leq n$, satisfying
-\[
+\(
 s \in U_i
@@
 \;
-\hbox{for some }r \in \Real^{j_i}
+\hbox{for some }r \in \Real^{j_i}.
+\)
+(We use the (measurable) function $\sigma_i : S \to \Real^{j_i}$ to skolemise the existentially quantified $r$.)
+Equivalently
+\[
+U_i := \#\mathrm{draw}_n^{-1}(l_i) \cap M_n^{-1}(E_i[\cdot][\tsample], \Real^{j_i}) \in \calF_n.
 \]
-(We use the (measurable) function $\sigma_i : S \to \Real^{j_i}$ to skolemise the existentially quantified $r$.)
-
-Equivalently
-\(
-U_i := \#\mathrm{draw}_n^{-1}(l_i) \cap M_n^{-1}(E_i[\cdot][\tsample], \Real^{j_i}) \in \calF_n.
-\)
 %\lo{I.e.~$(E_i[\underline{\sigma_i(s)}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s)$ iff $s \in U_i$.}
@@
 It then follows from the definition of ranking function that
-\[
+\[%\textstyle
 \int_I \Leb(\dif r) \, f(E_i[\underline{\widehat{\sigma_i}(u)}][\underline{r}])) \leq f(E_i[\underline{\widehat{\sigma_i}(u)}][\tsample])).

### 2020-12-27 LICS21/SPCF.tex
Added two lemmas for proof of (rankability) Soundness Theorem. Their proofs are placed in appendix, using restatable.

@@
 \begin{align*}
-  & \text{types } A, B ::= \textsf{R}  \mid  A \to B \\
-  & \text{values } V ::= \lambda x.M  \mid  \underline{r} \\
-  & \text{terms } M, N ::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY M  \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample  \mid  \tscore(M)
+  \text{types } A, B &::= \textsf{R}  \mid  A \to B \\
+  \text{values } V &::= \lambda x.M  \mid  \underline{r} \\
+  \text{terms } M, N &::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY M \\
+   & \qquad \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample  \mid  \tscore(M)
 \end{align*}
-\begin{align*}
-  \frac{}{\Gamma ; x:A \vdash x:A} \qquad
-  \frac{\Gamma ; x:A \vdash M : B}{\Gamma \vdash \lambda x.M : A \to B} \qquad
-  \frac{}{\underline{r} : \textsf{R}} \qquad
-  \frac{\Gamma \vdash M:A \to B \quad \Gamma \vdash N : A}{\Gamma \vdash M \, N : B} \\ \\
-  \frac{\Gamma \vdash M_1:\textsf{R} \dots \Gamma \vdash M_n:\textsf{R}}{\Gamma \vdash \underline{f}(M_1,\dots,M_n) : \textsf{R}} \ f : \mathbb{R}^n \to \mathbb{R} \qquad
-  \frac{\Gamma \vdash M : (A \to B) \to (A \to B)}{\Gamma \vdash \tY M : (A \to B)} \\ \\
-  \frac{\Gamma \vdash M : \textsf{R} \quad \Gamma \vdash N_1 : A \quad \Gamma \vdash N_2 : A}{\Gamma \vdash \tif{M < 0}{N_1}{N_2} : A} \qquad
-  \frac{}{\Gamma \vdash \tsample : \textsf{R}} \qquad
-  \frac{\Gamma \vdash s : \textsf{R}}{\Gamma \vdash \tscore (s) : \textsf{R}}
-\end{align*}
+The typing rules are presented in \Cref{fig:typing rules}.
+Terms are identified up to $\alpha$-equivalence, as usual. 
+The set of all terms is denoted $\Lambda$, and the set of closed terms is denoted $\Lambda^0$.
 
-Terms are identified up to $\alpha$-equivalence, as usual. The set of all terms is denoted $\Lambda$, and the set of closed terms is denoted $\Lambda^0$.
-
-\paragraph{}
+%\paragraph{}
 To define the reduction relation, let \emph{evaluation contexts} be of the form:
@@
 then a term reduces if it is formed by substituting a redex in a context i.e.
+
+\begin{figure*}
+\begin{align*}
+  \frac{}{\Gamma ; x:A \vdash x:A} 
+  \qquad
+  \frac{\Gamma ; x:A \vdash M : B}{\Gamma \vdash \lambda x.M : A \to B} 
+  \qquad
+  \frac{}{\underline{r} : \textsf{R}} 
+  \qquad
+  \frac{\Gamma \vdash M:A \to B \quad \Gamma \vdash N : A}{\Gamma \vdash M \, N : B} 
+  \\ \\
+  \qquad
+  \frac{\Gamma \vdash M : (A \to B) \to (A \to B)}{\Gamma \vdash \tY M : (A \to B)}
+  \qquad
+  \frac{\Gamma \vdash M_1:\textsf{R} \quad \dots \quad \Gamma \vdash M_n:\textsf{R}}{\Gamma \vdash \underline{f}(M_1,\dots,M_n) : \textsf{R}} \ (f : \mathbb{R}^n \to \mathbb{R})    
+  \\ \\
+  \qquad
+  \frac{\Gamma \vdash M : \textsf{R} \quad \Gamma \vdash N_1 : A \quad \Gamma \vdash N_2 : A}{\Gamma \vdash \tif{M < 0}{N_1}{N_2} : A}
+  \qquad
+  \frac{}{\Gamma \vdash \tsample : \textsf{R}} 
+  \qquad
+  \frac{\Gamma \vdash s : \textsf{R}}{\Gamma \vdash \tscore (s) : \textsf{R}}
+\end{align*}
+\caption{Typing rules of SPCF \label{fig:typing rules}}
+\end{figure*}
+
 \begin{align*}
@@
   E[\underline f (\underline r_1, \dots , \underline r_n)] & \to E[\underline{f(r_1,\dots,r_n)}] \\
-  E[\tY \lambda x. M] & \to E[\lambda z. M[(\tY \lambda x. M)/x] z] \text{ where $z$ is not free in $s$}\\
+  E[\tY \lambda x. M] & \to E[\lambda z. M[(\tY \lambda x. M)/x] z] \\
+  & \qquad \qquad \text{ where $z$ is not free in $s$}\\
   E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_1] \text{ where }r < 0 \\
@@
 \end{align*}
-\changed[lo]{We write $\to^\ast$ for the reflexive, transitive closure of $\to$.}
+We write $\to^\ast$ for the reflexive, transitive closure of $\to$.
 \lo{The $\tscore$ rule should have a side condition: if $r > 0$.}

### 2020-12-27 LICS21/apx-y-stopping-times.tex
Created version for LICS21 submission in subdirectory LICS21
Added two lemmas for proof of (rankability) Soundness Theorem. Their proofs are placed in appendix, using restatable.

--- 2020-06-01 y-stopping-time.tex
@@
-% !TEX root=./probabilisticProgrammingMartingales.tex
+% !TEX root=./main.tex
 
@@
 
-\section{$\tY$ stopping times are finite}
-
+%\section{$\tY$ stopping times are finite}
+\iffalse
 As before, fix a $M \in \Lambda^0$.
@@
 \end{align*}
+\fi
 
-\begin{lemma}
-$T_0, T_1, T_2, \cdots$ are an increasing sequence of stopping times adapted to $(\calF_n)_{n \in \omega}$, and each $T_i$ is bounded.
-\end{lemma}
+\TnBounded*
 
@@
 \label{thm:de groote}
-The simply-typed non-deterministic lambda-calculus is strongly normalising. \qed
+The simply-typed nondeterministic lambda-calculus is strongly normalising. \qed
 \end{theorem}
@@
 \fi
+
+\TMtYStoppingTime*
+\begin{proof}
+%For each $n \in \omega$, $\set{T_M^\tY = n} \in \calF_{T_n}$.
+To see $\set{T_M^\tY = n} \in \calF_{T_n}$ for a given $n$, we need to show that $\set{T_M^\tY = n} \cap \set{T_n \leq i} \in \calF_i$, for all $i \in \omega$.
+Let $\cal V$ be the measurable subset of $\Lambda^0$ given by the countable union of $\set{\alpha_j} \times \mathbb{R}^j$ where $\alpha_j$ ranges over skeletons which are values with $j$ place-holders, and $j$ ranges over $\omega$.
+Let $\cal W$ be the measurable subset of $\Lambda^0$ given by the countable union of $\set{\beta_j} \times \mathbb{R}^j$ where $\beta_j$ ranges over skeletons which are not values, and with $j$ place-holders, and $j$ ranges over $\omega$.
+%Then $\set{T_M^\tY = n} \cap \set{T_n \leq i}$ is given by
+Then, it suffices to observe that $\set{T_M^\tY = n} \cap \set{T_n \leq i}
+= \bigcup_{l=n}^i 
+\big( M_l^{-1} ({\cal V}) \cap  
+M_{l-1}^{-1} ({\cal W}) \cap 
+\set{T_n = l} \big)
+$.
+\end{proof}

### 2020-12-27 LICS21/ranking.tex
Added two lemmas for proof of (rankability) Soundness Theorem. Their proofs are placed in appendix, using restatable.

@@
+% !TEX root = main.tex
 \section{Constructing Ranking Functions}
@@
 \begin{theorem} \label{thm:minimal}
-Given a closed term $M$, the function $f:Rch(M) \to \mathbb R$ given by $f(N) = \mathbb E [\text{the number of }\tY\text{-reduction steps from }N\text{ to a value}]$, if it exists, is the least of all possible ranking functions of $M$.
+Given a closed term $M$, the function $f:Rch(M) \to \mathbb R$ given by 
+\[
+f(N) := \mathbb E [\text{number of }\tY\text{-reduction steps from }N\text{ to a value}]
+\] 
+if it exists, is the least of all possible ranking functions of $M$.
 \end{theorem}
@@
 which generates a geometric distribution (let $\Theta = \lambda f, n : \tif{\tsample \underline{- 0.5} < 0}{n}{f (n\underline{+ 1})}$, then the term is $(\tY \Theta)\ \underline 0$).
-    Despite its simplicity, its $Rch$ contains all the terms
-
+    Despite its simplicity, its $Rch$ contains all the terms: see \Cref{fig:geometric distribution}.
+\begin{figure}
 \begin{tikzpicture}[node distance=0.5em, anchor=west]
@@
 \end{tikzpicture}
-
+\caption{Some reachable terms of $\tY\ \Theta\ \underline 0$\label{fig:geometric distribution}}
+\end{figure}
 \end{example}
@@
 
-\paragraph{}
+%\paragraph{}
 In both of these cases, there are only some values of the ranking function that are semantically important. 
@@
 
-\paragraph{}
+%\paragraph{}
 With this abbreviation, the geometric distribution example from earlier can be written as $(\tY \lambda f n.\ n \oplus_{0.5} f(n+1)) \, \underline 0$.

### 2020-12-27 LICS21/supermartingales.tex
Added two lemmas for proof of (rankability) Soundness Theorem. Their proofs are placed in appendix, using restatable.

@@
 \item Let $\epsilon > 0$.
-A stopping time $T$ and a supermartingale $(Y_n)_{n \in \omega}$, both adapted to filtration $(\calF_n)_{n \in \omega}$, are a \emph{$\epsilon$-ranking supermartingale} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.\footnote{For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable $s \mapsto \begin{cases}
-1 & \hbox{if $s \in A$}\\
-0 & \hbox{otherwise}
-\end{cases}$.}
+A stopping time $T$ and a supermartingale $(Y_n)_{n \in \omega}$, both adapted to filtration $(\calF_n)_{n \in \omega}$, are a \emph{$\epsilon$-ranking supermartingale} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.\footnote{For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable defined by ${\bf 1}_A(s) := 1$ if $s \in A$, and $0$ otherwise.}
 A \emph{ranking supermartingale} is a $\epsilon$-ranking supermartingale for some $\epsilon > 0$.
@@
 \label{thm:rankable and strict rankable}
-If a closed SPCF term is rankable (respectively, strictly rankable) by $f$, then $(f(M_n))_{n \in \omega}$ is a 
-supermartingale (respectively, ranking supermartingale) adapted to the filtration $(\mathcal{F}_n)_{n \in \omega}$ where $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
+If a closed SPCF term $M$ is rankable (respectively, strictly rankable) by $f$ satisfying $f(N) = 0$ iff $N$ is a value, then $(f(M_n))_{n \in \omega}$, where the random variable $M_n(s) := \pi_0 (\red^n(M, s))$, is a supermartingale (respectively, ranking supermartingale) adapted to the filtration $(\mathcal{F}_n)_{n \in \omega}$ where $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
 \end{theorem}
@@
 %\subsection{Proof of \Cref{thm:rankable and strict rankable}.}
-\emph{Notation}. Given an $\omega$-sequence (e.g.~$s \in S$) and $m \in \omega$, we write $s_{\leq m} \in I^m$ to mean the prefix of $s$ of length $m$.
+We shall obtain the theorem as a corollary of a technical lemma (\Cref{lem:key rankable}).
+First, some notation.
 
-Fix a closed SPCF term $M \in \Lambda^0$, and a ranking function $f$ for it, satisfying $f(N) = 0$ iff $N$ is a value.
-Let $n \in \omega$, define the following random variables on the probability space $(S, \calF, \mu)$:
+Given an $\omega$-sequence (e.g.~$s \in S$) and $m \in \omega$, we write $s_{\leq m} \in I^m$ to mean the prefix of $s$ of length $m$.
+%Fix a closed SPCF term $M \in \Lambda^0$, and a ranking function $f$ for it, satisfying $f(N) = 0$ iff $N$ is a value.
+%Let $n \in \omega$, define the following random variables on the probability space $(S, \calF, \mu)$:
+%\begin{align*}
+%M_n(s) &:= \pi_0 (\red^n(M, s))\\
+%\ndraw{n}{s} &:= |\{k < n \mid \exists E: M_k(s) = E[\tsample]\}|,
+%\end{align*}
+Fix $M \in \Lambda^0$; let $n \in \omega$, set
 \begin{align*}
-M_n(s) &:= \pi_0 (\red^n(M, s))\\
+%M_n(s) &:= \pi_0 (\red^n(M, s))\\
 \ndraw{n}{s} &:= |\{k < n \mid \exists E: M_k(s) = E[\tsample]\}|,
 \end{align*}
- so that $\pi_1 (\red^n(M, s)) = \overbrace{\pi_t( \cdots (\pi_t}^{\ndraw n s}(s))$, and the filtration $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
+so that $\pi_1 (\red^n(M, s)) = \overbrace{\pi_t( \cdots (\pi_t}^{\ndraw n s}(s))$. 
+%and the filtration $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
 The $\calF_n$-measurability of $M_n$ (and hence of $\#\mathrm{draw}_n$) follows from \citep{DBLP:conf/icfp/BorgstromLGS16}.
@@
 \medskip
-As before, fix a closed SPCF term $M \in \Lambda^0$.
+As before, fix $M \in \Lambda^0$.
 Define random variables on the probability space $(S, \calF, \mu)$:
@@
 \end{align}
+
+%The random variable $T_0(s)$ is the least $k \geq 0$ such that $M_k(s)$ is either a value or a $\tY$-redex.
+
+%$T_1(s)$ is the least $k \geq 0$ such that $M_k(s)$ is either a value or the second occurrence of a $\tY$-redex.
+
+%Thus the random variable $T_n(s)$ is the least $k \geq 0$ such that $M_k(s)$ is either a value or the $(n+1)$-occurrence of a $\tY$-redex (in the reduction sequence $M_0(s), M_1(s), \ldots$).
+%Notice that each $T_n$ is a stopping time w.r.t.~filtration $(\calF_n)_{n \in \omega}$.
+First a useful property about the r.v.~$T_0, T_1, T_2, \cdots$.
+\begin{restatable}{lemma}{TnBounded}
+\label{lem:TnBounded}
+$T_0, T_1, T_2, \cdots$ are an increasing sequence of stopping times adapted to $(\calF_n)_{n \in \omega}$, and each $T_i$ is bounded.
+\end{restatable}
+
 The random variable $T_M^\tY$, which we call the \emph{$\tY$-runtime of $M$}, can equivalently be defined as the number of $\tY$-reduction steps in the reduction sequence of $M$. 
 Note that, as the type system ensures that the reduction relation excluding $\tY$-reduction is strongly normalising, only finitely many reductions can occur in a row without one of them being a $\tY$-reduction, therefore $T_M^\tY < \infty$ a.s.~iff $M$ is AST.
-
-The r.v.~$T_M^\tY$ is a stopping time adapted to the filtration $(\calF_{T_n})_{n \in \omega}$.
+Moreover 
+\begin{restatable}{lemma}{TMtYStoppingTime}
+\label{lem:TMtY is a stopping time}
+$T_M^\tY$ is a stopping time adapted to $(\calF_{T_n})_{n \in \omega}$.
+\end{restatable}
 
@@
 Let $f$ be a ranking function on $M$.
-For 1, observe that $T_0, T_1, T_2, \cdots$ are an increasing sequence of stopping times, each is bounded \lo{Prove it.} and adapted to $(\calF_n)_{n \in \omega}$.
-Then, since $(f(M_n))_{n \in \omega}$ is a supermartingale also adapted to $(\calF_n)_{n \in \omega}$ (\Cref{thm:rankable and strict rankable}),
+For 1, since $T_0, T_1, T_2, \cdots$ are an increasing sequence of stopping times, each is bounded and adapted to $(\calF_n)_{n \in \omega}$ (\Cref{lem:TnBounded}),
+and $(f(M_n))_{n \in \omega}$ is a supermartingale also adapted to $(\calF_n)_{n \in \omega}$ (\Cref{thm:rankable and strict rankable}),
 it follows from the Optional Sampling \Cref{thm:optional sampling} that $(Y_n)_{n \in \omega}$ where $Y_n := f(M_{T_n})$ is a supermartingale adapted to the filtration $(\calF_{T_n})_{n \in \omega}$.
-Since $T_M^\tY$ is a stopping time also adapted to $(\calF_{T_n})_{n \in \omega}$, we have that $(Y_n)_{n \in \omega}$ is a $1$-ranking supermartingale.
+Since the stopping time $T_M^\tY$ is also adapted to $(\calF_{T_n})_{n \in \omega}$ (\Cref{lem:TMtY is a stopping time}), we have that $(Y_n)_{n \in \omega}$ is a $1$-ranking supermartingale.
 Therefore, by \Cref{lem:rank-PAST}, $T^{\tY}_M < \infty$ a.s.~and $\expect{T^{\tY}_M} < \infty$.

### 2020-12-27 LICS21/antitone.tex
Mainly LaTeX reformatting to fit 2-column style
\appendices
Sec. 3: Repositioning ranking supermaringale lemma; introduce two subsections.

@@
+% !TEX root = main.tex
 \section{Antitone ranking functions}
@@
 
-\paragraph{Idea.}
+\paragraph*{Idea}
 
@@
 Therefore
-
 \begin{calculation}
@@
    \step[=]{rearranging terms}
-\dfrac{\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
+% \dfrac{\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
+% + 
+% \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
+% - 
+% \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
+\dfrac{
+\begin{array}{l}
+\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
 + 
 \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
-- 
-\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
+\\
+{} - 
+\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}
+\end{array}
+}{\mathbb P[T > n]}
   \step[=]{definition of conditional expectation}
@@
 \expect{Y_n-Y_{n+1}}
-  \step[=]{$\calF_n \subseteq \calF_{n+1}$, definition and linearity of conditional expectation}
+  \step[=]{$\calF_n \subseteq \calF_{n+1}$, def.~\& linearity of cond.~expectation}
 \expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
@@
 %  \step[=]{definition of conditional expectation for a discrete variable}
-\step[=]{definition of expectation conditioning on an \emph{event}}
+\step[=]{def.~expectation conditioning on an \emph{event}}
 \mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
@@
 Therefore, by a telescoping sum
-\[
-\expect{Y_{B_x}} = \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1}) \leq -1 < 0 
-\]
+\begin{align*}
+\expect{Y_{B_x}} 
+&= \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})\\
+&\leq -1 < 0 
+\end{align*}
 %\lo{$\expect{Y_0 - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})}$}
@@
 As before, $(Y_n)_{n \in \omega}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$.
-Therefore, it follows from \Cref{thm:a-c strict} that $T_M^{\tY}$ is almost-surely finite.
+Therefore, it follows from \Cref{thm:a-c strict} that $T_M^{\tY} < \infty$ a.s.
 \end{proof}
 
-\paragraph{}
+%\paragraph{}
 As before, constructing antitone ranking fuctions completely is not necessary, and there is a corresponding notion of an antitone partial ranking function.
@@
 
-\paragraph{\Cref{ex:ac-ranking} revisited}
+\paragraph*{\Cref{ex:ac-ranking} revisited}
 \akr{All of these ranking functions are slightly incorrect, because $\Xi \, \underline n$ reduces to something like $\Xi \, (\underline n + 1)$ then $(\lambda z.
@@
 \begin{align*}
-g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 
+& g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 \\
   &=  \log \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) \\
@@
 To construct an antitone ranking function, we solve the recurrence relation:
-\[
-\left\{
-\begin{array}{rll}
-z_0 &=& 0\\
-n \geq 1, \quad z_n &>& \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
-\end{array}
-\right.
-\]
+% \[
+% \left\{
+% \begin{array}{rll}
+% z_0 &=& 0\\
+% n \geq 1, \quad z_n &>& \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
+% \end{array}
+% \right.
+% \]
+\begin{align*}
+z_0 &= 0\\
+n \geq 1, \quad z_n &> \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
+\end{align*}
 For $n > 2$, $z_n = \log(n-1)$ works.
@@
 
-\paragraph{Completeness}
+\paragraph*{Completeness}
 It seems very likely that this method is complete in the case that every reachable term is AST, but we have been unable to actually prove this.
@@
     & \expect{Y_{n+1} \mid \mathcal F_n} \\
-    = & \expect{y_{n+1} \, 1_{\set{T \geq n+1}} \mid \mathcal F_n} \\
-    = & y_{n+1} \, \expect{1_{\set{T \geq n+1}} \mid \mathcal F_n} \\ 
-    = & y_{n+1} \, 1_{\set{T \geq n}} \frac{\mathbb P[T > n+1]}{\mathbb P[T > n]} \\
-    = & (y_n - \epsilon(y_n)) \, 1_{\set{T \geq n}} \\
-    = & Y_n - \epsilon(y_n) \, 1_{\set{T \geq n}}
+    &=  \expect{y_{n+1} \, 1_{\set{T \geq n+1}} \mid \mathcal F_n} \\
+    &=  y_{n+1} \, \expect{1_{\set{T \geq n+1}} \mid \mathcal F_n} \\ 
+    &=  y_{n+1} \, 1_{\set{T \geq n}} \frac{\mathbb P[T > n+1]}{\mathbb P[T > n]} \\
+    &=  (y_n - \epsilon(y_n)) \, 1_{\set{T \geq n}} \\
+    &= Y_n - \epsilon(y_n) \, 1_{\set{T \geq n}}
 \end{align*}

### 2020-12-27 LICS21/supermartingales.tex
Sec. 3: Repositioning ranking supermaringale lemma; introduce two subsections.

@@
 
-One approach to proving that a term terminates almost surely is to find some variant that is bounded below and, on average, decreases sufficiently quickly that it must eventually reach 0, similarly to the approach taken in
+One approach to proving that a term terminates almost surely is to find some variant that is bounded below and, on average, decreases sufficiently quickly that it must eventually reach 0, similarly to the approach taken by
 %\citep{DBLP:journals/pacmpl/McIverMKK18} 
-\cite{DBLP:conf/popl/FioritiH15} for imperative programs.
+\cite{DBLP:conf/popl/FioritiH15} and others for imperative programs.
 
@@
 \end{enumerate}
+\lo{Ranking function $f$ is not required to satisfy: $f(N) = 0$ iff $N$ is a value.}
 We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that $f(N) = 0$ iff $N$ is a value; and for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.
@@
 \begin{remark}
-\changed[lo]{The original definition of ranking supermartingale, due to \citep{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, does not refer to a stopping time explicitly:
+\changed[lo]{Our notion of ranking supermartingale is a slight generalisation of the original definition, due to \citep{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, which does not refer to a stopping time explicitly:
 Fioriti and Hermanns call a supermartingale $(Y_n)_{n \in \omega}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot 1_{\set{Y_n > 0}}$.}
@@
 In a $\epsilon$-ranking supermartingale, each computation step causes a strict decrease in rank of at least $\epsilon$, provided the term in question is not a value.
+
+\begin{lemma}%[{\citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}}]
+% \citep[Prop 1]{DBLP:conf/popl/ChatterjeeFNH16}}
+\label{lem:rank-PAST}
+Let $(Y_n)_{n \in \omega}$ be a $\epsilon$-ranking supermartingale with respect to the stopping time $T$, then $T < \infty$ almost surely, and $\expect{T} \leq \frac{\expect{Y_0}}{\epsilon}$.
+\end{lemma}
+
+\begin{proof}
+We first prove (by induction on $n$): for all $n \in \omega$
+\[
+%\forall n \in \omega \, . \, 
+\expect{Y_n} \leq \expect{Y_0} - \epsilon \cdot \big(\textstyle\sum_{i=0}^{n-1} \mathbb{P}[T > i]\big).
+\]
+It follows that $\sum_{i=0}^{\infty} \mathbb{P}[T > i]$ converges; hence we have $\lim_{i \to \infty} \mathbb{P}[T > i] = 0$ and so $\mathbb{P}[T < \infty] = 1$.
+It then remains to observe: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=0} \mathbb{P}[T > i]$.
+\end{proof}
+
+\begin{remark}
+\changed[lo]{\Cref{lem:rank-PAST} is a slight extension of \citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15},
+which asserts the same result for the specific stopping time $T : s \mapsto \min\set{n \mid Y_n(s) = 0}$.}
+\end{remark}
 
@@
 
-\begin{lemma}%[{\citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}}]
-% \citep[Prop 1]{DBLP:conf/popl/ChatterjeeFNH16}}
-\label{lem:rank-PAST}
-Let $(Y_n)_{n \in \omega}$ be a $\epsilon$-ranking supermartingale with respect to the stopping time $T$, then $T < \infty$ almost surely, and $\expect{T} \leq \frac{\expect{Y_0}}{\epsilon}$.
-\end{lemma}
-
-\begin{proof}
-We first prove (by induction on $n$): for all $n \in \omega$
-\[
-%\forall n \in \omega \, . \, 
-\expect{Y_n} \leq \expect{Y_0} - \epsilon \cdot \big(\textstyle\sum_{i=0}^{n-1} \mathbb{P}[T > i]\big).
-\]
-It follows that $\sum_{i=0}^{\infty} \mathbb{P}[T > i]$ converges; hence we have $\lim_{i \to \infty} \mathbb{P}[T > i] = 0$ and so $\mathbb{P}[T < \infty] = 1$.
-It then remains to observe: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=0} \mathbb{P}[T > i]$.
-\end{proof}
-
-\changed[lo]{\Cref{lem:rank-PAST} is a slight extension of \citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15},
-wherein the stopping time $T$ is assumed to be the random variable $s \mapsto \min\set{n \mid Y_n(s) = 0}$.}
-
-\medskip
+\subsection{Ranking functions and supermartingales}
 
@@
 
-\medskip
+\subsection{Soundness of rankability}
+
 As before, fix $M \in \Lambda^0$.
@@
 
-\begin{theorem}[Soundness] \label{thm:rankable implies termination}
+\begin{theorem}[Soundness of rankability] \label{thm:rankable implies termination}
 \begin{enumerate}

### 2020-12-27 LICS21/supermartingales.tex
Sec 2: minor

@@
 \end{enumerate}
-\lo{Ranking function $f$ is not required to satisfy: $f(N) = 0$ iff $N$ is a value.}
+\lo{@Andrew: As defined, ranking function $f$ is not required to satisfy: $f(N) = 0$ iff $N$ is a value.
+But for AST analysis, we always require $f$ to satisfy this property.
+It seems cleaner to include this in the definition.}
 We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that $f(N) = 0$ iff $N$ is a value; and for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.
@@
 Fix a probability space $(\Omega, \calF, \mathbb{P})$. 
-We have in mind $\Omega = S$, and $\mathbb{P} = \mu$.
+(In subsequent applications to AST analysis, $\Omega = S$, and $\mathbb{P} = \mu$.)
 
@@
 
+%\emph{Notation.}
+\changed[lo]{For the rest of this section, fix the probability space $(S, \calF, \mu)$, and a closed SPCF term $M$.
+For $n \in \omega$, define the random variables $M_n(s) := \pi_0 (\red^n(M, s))$, and the filtration $(\mathcal{F}_n)_{n \in \omega}$ where $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.}
+
 Our first result is the following theorem.
@@
 \label{thm:rankable and strict rankable}
-If a closed SPCF term $M$ is rankable (respectively, strictly rankable) by $f$ satisfying $f(N) = 0$ iff $N$ is a value, then $(f(M_n))_{n \in \omega}$, where the random variable $M_n(s) := \pi_0 (\red^n(M, s))$, is a supermartingale (respectively, ranking supermartingale) adapted to the filtration $(\mathcal{F}_n)_{n \in \omega}$ where $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
+If a closed SPCF term $M$ is rankable (respectively, strictly rankable) by $f$ satisfying $f(N) = 0$ iff $N$ is a value, then $(f(M_n))_{n \in \omega}$ is a supermartingale (respectively, ranking supermartingale with stopping time $T_M: s \mapsto \min \set{n \mid f(M_n(s)) = 0}$) adapted to the filtration $(\mathcal{F}_n)_{n \in \omega}$.
+%where $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
 \end{theorem}
@@
 %\end{align*}
-Fix $M \in \Lambda^0$; let $n \in \omega$, set
+For $n \in \omega$, define
 \begin{align*}
@@
 We say that a given SPCF term is: %\emph{type-1} (respectively 2, 3 and 4) if it has the shape $E[\tY \lambda x. N]$ (respectively $E[\tsample]$, $E[R]$ where $R$ is any other redex, and of a value).
-\emph{type-1} if it has the shape $E[\tY \lambda x.
-N]$, \emph{type-2} if it has the shape $E[\tsample]$; \emph{type-3} if it has the shape $E[R]$ where $R$ is any other redex, and \emph{type-4} if it is a value.
+\emph{type 1} if it has the shape $E[\tY \lambda x.
+N]$, \emph{type 2} if it has the shape $E[\tsample]$; \emph{type 3} if it has the shape $E[R]$ where $R$ is any other redex, and \emph{type 4} if it is a value.
 Henceforth fix an $n \geq 0$, and define $\mathbf{T}_i := \set{s \mid M_n(s) \hbox{ is type-$i$}}$.
@@
 \] 
+\changed[lo]{Thus 
+%$\int_A \mu(\dif s) \, f(M_{n+1}) \leq \int_A \mu(\dif s) \, f(M_n)$, i.e., 
+$\mathbb{E}[f(M_{n+1}) \mid \calF_n] \leq f(M_n)$ a.s.}
 \end{lemma}
@@
 This concludes the proof of %\Cref{thm:rankable gives supermartingale}. 
-\Cref{thm:rankable and strict rankable}. \hfill \qed
+\Cref{thm:rankable and strict rankable}. %\hfill \qed
 
@@
 
-As before, fix $M \in \Lambda^0$.
+In the rest of this section we show that if $M \in \Lambda^0$ is rankable by $f$, then $M$ is AST.
+
+Let $f$ be a ranking function on $M$ satisfying $f(N) = 0$ iff $N$ is a value.
 Define random variables on the probability space $(S, \calF, \mu)$:
@@
 T_M^\tY(s) &:= \min \set{n \mid M_{T_n(s)}(s) \textrm{ is a value}} \label{eq:y stopping time} 
+%T_M(s) & := \min \set{n \mid X_n(s) = 0} \nonumber
 \end{align}
@@
 %Notice that each $T_n$ is a stopping time w.r.t.~filtration $(\calF_n)_{n \in \omega}$.
-First a useful property about the r.v.~$T_0, T_1, T_2, \cdots$.
+We first state a useful property about r.v.~$T_0, T_1, T_2, \cdots$.
 \begin{restatable}{lemma}{TnBounded}
 \label{lem:TnBounded}
-$T_0, T_1, T_2, \cdots$ are an increasing sequence of stopping times adapted to $(\calF_n)_{n \in \omega}$, and each $T_i$ is bounded.
+$(T_n)_{n \in\omega}$ is an increasing sequence of stopping times adapted to $(\calF_n)_{n \in \omega}$, and each $T_i$ is bounded.
 \end{restatable}
+
+%The random variable $T_M$ is a stopping time w.r.t.~filtration $(\calF_n)_{n \in \omega}$. 
+%Since $f(M_n(s)) = 0$ iff $M_n(s)$ is a value, $T_M$ is the \emph{runtime of $M$}.
 
@@
 Let $f$ be a ranking function on $M$.
-For 1, since $T_0, T_1, T_2, \cdots$ are an increasing sequence of stopping times, each is bounded and adapted to $(\calF_n)_{n \in \omega}$ (\Cref{lem:TnBounded}),
+For 1, since $(T_n)_{n \in\omega}$ is an increasing sequence of stopping times, each is bounded and adapted to $(\calF_n)_{n \in \omega}$ (\Cref{lem:TnBounded}),
 and $(f(M_n))_{n \in \omega}$ is a supermartingale also adapted to $(\calF_n)_{n \in \omega}$ (\Cref{thm:rankable and strict rankable}),
 it follows from the Optional Sampling \Cref{thm:optional sampling} that $(Y_n)_{n \in \omega}$ where $Y_n := f(M_{T_n})$ is a supermartingale adapted to the filtration $(\calF_{T_n})_{n \in \omega}$.
-Since the stopping time $T_M^\tY$ is also adapted to $(\calF_{T_n})_{n \in \omega}$ (\Cref{lem:TMtY is a stopping time}), we have that $(Y_n)_{n \in \omega}$ is a $1$-ranking supermartingale.
+Notice the stopping time $T_M^\tY$ is also adapted to $(\calF_{T_n})_{n \in \omega}$ (\Cref{lem:TMtY is a stopping time}); and we have that $(Y_n)_{n \in \omega}$ is a $1$-ranking supermartingale.
 Therefore, by \Cref{lem:rank-PAST}, $T^{\tY}_M < \infty$ a.s.~and $\expect{T^{\tY}_M} < \infty$.
 
-Statement 2 follows immediately from \Cref{lem:rank-PAST}.
+%For 2, we have that $(f(M_n))_{n \in \omega}$ is a ranking supermartingale (\Cref{thm:rankable and strict rankable}) w.r.t.~stopping time $T_M$. 
+%It follows immediately from \Cref{lem:rank-PAST} that $\expect{T_M} < \infty$.
+Statement 2 follows immediately from \Cref{thm:rankable and strict rankable} and \Cref{lem:rank-PAST}.
 \end{proof}

### 2020-12-28 LICS21/antitone.tex
Global replace: _{n \in \omega} -> _{n \geq 0}

@@
 \begin{definition}
-We say that a supermartingale $(Y_n)_{n \in \omega}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \in \omega}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
+We say that a supermartingale $(Y_n)_{n \geq 0}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \geq 0}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
 \[
@@
 
-%\lo{The preceding definition does not place any constraint on the stopping time $T$ (other than that it is w.r.t.~the filtration $(\calF_n)_{n \in \omega}$).
+%\lo{The preceding definition does not place any constraint on the stopping time $T$ (other than that it is w.r.t.~the filtration $(\calF_n)_{n \geq 0}$).
 However for the definition to make sense, $T$ must be the stopping time $T(s) := \min \set{n \mid Y_n(s) = 0}$.}
@@
 \label{thm:a-c strict}
-Let $(Y_n)_{n \in \omega}$ be an antitone strict supermartingale via function $\epsilon : \nnReal \to \pReal$, and stopping time $T$. 
+Let $(Y_n)_{n \geq 0}$ be an antitone strict supermartingale via function $\epsilon : \nnReal \to \pReal$, and stopping time $T$. 
 Then $T < \infty$ almost surely.
@@
 
-Note that it is essential in this proof that $\epsilon$ be defined for all $\mathbb R_{\geq 0}$, not just the values that $(Y_n)_{n \in \omega}$ actually takes (or at least if $(Y_n)_{n \in \omega}$ is uniformly bounded, $\epsilon$ must be positive at the supremum as well as the actual values).
+Note that it is essential in this proof that $\epsilon$ be defined for all $\mathbb R_{\geq 0}$, not just the values that $(Y_n)_{n \geq 0}$ actually takes (or at least if $(Y_n)_{n \geq 0}$ is uniformly bounded, $\epsilon$ must be positive at the supremum as well as the actual values).
 
@@
 \lo{N.B.~$\set{T = n} = (\frac{1}{2^{n}}, \frac{1}{2^{n-1}}]$, and so, $\set{T > n} = [-1, \frac{1}{2^n}]$.}
-Plainly, $T$ is a stopping time, and $(Y_n)_{n \in \omega}$ is a supermartingale, adapted to the filtration $(\calF_n)_{n \in \omega}$ where $\calF_n = \calF$ for all $n$. 
-In this case, $(Y_n)_{n \in \omega}$ either tends to $4$ as $n \to \infty$, or drops to 0 at some point, with an exponentially decreasing probability. 
-With respect to the antitone function $\epsilon(x) = \frac{4-x}{4}$ and stopping time $T$, we have that $(Y_n)_{n \in \omega}$ is an antitone strict supermartingale, except that $\epsilon(4) = 0$, and $T = \infty$ with probability $\frac 1 2$, even though $4$ is larger than any value $(Y_n)_{n \in \omega}$ can actually take, and $\epsilon(Y_n)$ never actually reaches $0$.
+Plainly, $T$ is a stopping time, and $(Y_n)_{n \geq 0}$ is a supermartingale, adapted to the filtration $(\calF_n)_{n \geq 0}$ where $\calF_n = \calF$ for all $n$. 
+In this case, $(Y_n)_{n \geq 0}$ either tends to $4$ as $n \to \infty$, or drops to 0 at some point, with an exponentially decreasing probability. 
+With respect to the antitone function $\epsilon(x) = \frac{4-x}{4}$ and stopping time $T$, we have that $(Y_n)_{n \geq 0}$ is an antitone strict supermartingale, except that $\epsilon(4) = 0$, and $T = \infty$ with probability $\frac 1 2$, even though $4$ is larger than any value $(Y_n)_{n \geq 0}$ can actually take, and $\epsilon(Y_n)$ never actually reaches $0$.
 %\lo{@Andrew: I refer to the preceding highlighted claim. Did you mean: for all $n, s$, $\epsilon(Y_n(s)) > Y_n(s)$?} \akr{No. This phrasing should be clearer.}
@@
 \changed[lo]{Let $f$ be a ranking function on a closed SPCF term $M$.
-%Recall that $(f(M_n))_{n \in \omega}$ is a supermartingale.
-We say that $M$ is \emph{antitone rankable} by $f$ if the supermartingale $(f(M_n))_{n \in \omega}$ and the stopping time $T_M^{\tY}$, both adapted to the fitration $(\calF_n)_{n \in \omega}$ where $\calF_n := \sigma(M_1, \ldots, M_n)$, are antitone strict.}
+%Recall that $(f(M_n))_{n \geq 0}$ is a supermartingale.
+We say that $M$ is \emph{antitone rankable} by $f$ if the supermartingale $(f(M_n))_{n \geq 0}$ and the stopping time $T_M^{\tY}$, both adapted to the fitration $(\calF_n)_{n \geq 0}$ where $\calF_n := \sigma(M_1, \ldots, M_n)$, are antitone strict.}
 
@@
 take the set $S$ of traces \lo{TODO: We should find a better name for traces $s \in S$.} as a probability space, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$.
-As before, $(Y_n)_{n \in \omega}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$.
+As before, $(Y_n)_{n \geq 0}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$.
 Therefore, it follows from \Cref{thm:a-c strict} that $T_M^{\tY} < \infty$ a.s.
@@
 If $T$ is bounded by $b$ a.s.~(for $n$ constant), the statement is trivially true by taking $\epsilon$ to be constantly 1, and $Y_n = b - n$.
-Otherwise, let $(t_n)_{n \in \omega}$ be defined recursively such that $t_0 = 0$, $t_{n+1} > t_n$ and $\mathbb P[T > t_{n+1}] \leq \frac 1 4 \mathbb P[T > t_n]$.
-We will then define a supermartingale $(Y_n)_{n \in \omega}$ and nonrandom sequence $(y_n)_{n \in \omega}$ such that $Y_n = 0$ iff $T < n$, $Y_n = y_n$ iff $T \geq n$, and $y_n \geq 2^k$ for $n \geq t_k$.
+Otherwise, let $(t_n)_{n \geq 0}$ be defined recursively such that $t_0 = 0$, $t_{n+1} > t_n$ and $\mathbb P[T > t_{n+1}] \leq \frac 1 4 \mathbb P[T > t_n]$.
+We will then define a supermartingale $(Y_n)_{n \geq 0}$ and nonrandom sequence $(y_n)_{n \geq 0}$ such that $Y_n = 0$ iff $T < n$, $Y_n = y_n$ iff $T \geq n$, and $y_n \geq 2^k$ for $n \geq t_k$.
 
@@
 
-The sequence $(y_n)_{n \in \omega}$ is then defined recursively by
+The sequence $(y_n)_{n \geq 0}$ is then defined recursively by
 \begin{align*}
@@
 
-The assumption that $(\mathcal F_n)_{n \in \omega}$ is the coarsest filtration to which $T$ is adapted is used in the equality between the third and fourth lines here.
+The assumption that $(\mathcal F_n)_{n \geq 0}$ is the coarsest filtration to which $T$ is adapted is used in the equality between the third and fourth lines here.
 This condition is the main reason that this proof does not extend directly to a completeness result for antitone ranking functions.

### 2020-12-28 LICS21/supermartingales.tex
Global replace: _{n \in \omega} -> _{n \geq 0}

@@
 For example, the term $(\tY \lambda x.x) \, \underline 0$ is not rankable.
+\end{definition}
 
-It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \in \omega}$ is the reduction sequence starting from $M$, then $(f(M_n))_{n\in\omega}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales.
-\end{definition}
+It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \geq 0}$ is the reduction sequence starting from $M$, then $(f(M_n))_{n \geq 0}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales.
+\lo{It is confusing (strictly speaking, incorrect) to say that $(M_n)_{n \geq 0}$ is a reduction sequence from $M$. Formally $M_n$ is the random variable $M_n: s \mapsto \pi_0 (\red^n(M, s))$.}
 
@@
 
-Fix a probability space $(\Omega, \calF, \mathbb{P})$. 
-(In subsequent applications to AST analysis, $\Omega = S$, and $\mathbb{P} = \mu$.)
+\changed[lo]{Fix a probability space $(\Omega, \calF, \mathbb{P})$, and a filtration $(\calF_n)_{n \geq 0}$. 
+(In subsequent applications to termination analysis, $\Omega = S$, and $\mathbb{P} = \mu$.)
+Let $T$ be a random variable that takes values in $\mathbb{N} \cup \set{\infty}$.
+We call $T$ a \emph{stopping time adapted to $(\calF_n)_{n \geq 0}$} just if $\set{T = n} \in \calF_n$, for all $n \in \omega$.}
 
@@
 \begin{enumerate}
-\item A sequence of random variables $(Y_n)_{n \in \omega}$ adapted to a filtration $(\calF_n)_{n \in \omega}$ is a \emph{supermartingale} if for all $n \in \omega$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~\changed[lo]{(i.e.~$\forall A \in \calF_n$, $\int_A \dif \mu \, Y_{n+1} \leq \int_A \dif \mu \,Y_n$).}
+\item A sequence of random variables $(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \emph{supermartingale} if for all $n \in \omega$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~\changed[lo]{(i.e.~for all $A \in \calF_n$, $\int_A \mathbb{P}(\dif \omega) \, Y_{n+1}(\omega) \leq \int_A \mathbb{P}(\dif \omega) \,Y_n(\omega)$).}
 \item Let $\epsilon > 0$.
-A stopping time $T$ and a supermartingale $(Y_n)_{n \in \omega}$, both adapted to filtration $(\calF_n)_{n \in \omega}$, are a \emph{$\epsilon$-ranking supermartingale} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.\footnote{For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable defined by ${\bf 1}_A(s) := 1$ if $s \in A$, and $0$ otherwise.}
+A stopping time $T$ and a supermartingale $(Y_n)_{n \geq 0}$, both adapted to filtration $(\calF_n)_{n \geq 0}$, are a \emph{$\epsilon$-ranking supermartingale} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.\footnote{For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable defined by ${\bf 1}_A(s) := 1$ if $s \in A$, and $0$ otherwise.}
 A \emph{ranking supermartingale} is a $\epsilon$-ranking supermartingale for some $\epsilon > 0$.
@@
 \changed[lo]{Our notion of ranking supermartingale is a slight generalisation of the original definition, due to \citep{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, which does not refer to a stopping time explicitly:
-Fioriti and Hermanns call a supermartingale $(Y_n)_{n \in \omega}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot 1_{\set{Y_n > 0}}$.}
+Fioriti and Hermanns call a supermartingale $(Y_n)_{n \geq 0}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot 1_{\set{Y_n > 0}}$.}
 \end{remark}
@@
 \label{lem:rank-PAST}
-Let $(Y_n)_{n \in \omega}$ be a $\epsilon$-ranking supermartingale with respect to the stopping time $T$, then $T < \infty$ almost surely, and $\expect{T} \leq \frac{\expect{Y_0}}{\epsilon}$.
+Let $(Y_n)_{n \geq 0}$ be a $\epsilon$-ranking supermartingale with respect to the stopping time $T$, then $T < \infty$ almost surely, and $\expect{T} \leq \frac{\expect{Y_0}}{\epsilon}$.
 \end{lemma}
@@
 
-Let $T$ and $T'$ be stopping times adapted to $(\calF_n)_{n \in \omega}$.
+Let $T$ and $T'$ be stopping times adapted to $(\calF_n)_{n \geq 0}$.
 Recall the $\sigma$-algebra (consisting of measurable subsets ``prior to $T$'')
@@
 \label{thm:optional sampling}
-Let $(X_n)_{n\in \omega}$ be a supermartingale, and $(T_n)_{n \in \omega}$ a sequence of increasing stopping times, both adapted to filtration $(\calF_n)_{n \in \omega}$, then $(X_{T_n})_{n \in \omega}$ is a supermartingale adapted to $(\calF_{T_n})_{n \in \omega}$ if one of the following conditions holds:
+Let $(X_n)_{n \geq 0}$ be a supermartingale, and $(T_n)_{n \geq 0}$ a sequence of increasing stopping times, both adapted to filtration $(\calF_n)_{n \geq 0}$, then $(X_{T_n})_{n \geq 0}$ is a supermartingale adapted to $(\calF_{T_n})_{n \geq 0}$ if one of the following conditions holds:
 \begin{enumerate}
 \item each $T_n$ is bounded i.e.~$T_n < c_n$ where $c_n$ is a constant
-\item $(X_n)_{n\in \omega}$ is uniformly integrable.
+\item $(X_n)_{n \geq 0}$ is uniformly integrable.
 \end{enumerate}
@@
 \changed[lo]{For the rest of this section, fix the probability space $(S, \calF, \mu)$, and a closed SPCF term $M$.
-For $n \in \omega$, define the random variables $M_n(s) := \pi_0 (\red^n(M, s))$, and the filtration $(\mathcal{F}_n)_{n \in \omega}$ where $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.}
+For $n \in \omega$, define the random variables $M_n(s) := \pi_0 (\red^n(M, s))$, and the filtration $(\mathcal{F}_n)_{n \geq 0}$ where $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.}
 
@@
 \label{thm:rankable and strict rankable}
-If a closed SPCF term $M$ is rankable (respectively, strictly rankable) by $f$ satisfying $f(N) = 0$ iff $N$ is a value, then $(f(M_n))_{n \in \omega}$ is a supermartingale (respectively, ranking supermartingale with stopping time $T_M: s \mapsto \min \set{n \mid f(M_n(s)) = 0}$) adapted to the filtration $(\mathcal{F}_n)_{n \in \omega}$.
+If a closed SPCF term $M$ is rankable (respectively, strictly rankable) by $f$ satisfying $f(N) = 0$ iff $N$ is a value, then $(f(M_n))_{n \geq 0}$ is a supermartingale (respectively, ranking supermartingale with stopping time $T_M: s \mapsto \min \set{n \mid f(M_n(s)) = 0}$) adapted to the filtration $(\mathcal{F}_n)_{n \geq 0}$.
 %where $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
@@
 %Thus the random variable $T_n(s)$ is the least $k \geq 0$ such that $M_k(s)$ is either a value or the $(n+1)$-occurrence of a $\tY$-redex (in the reduction sequence $M_0(s), M_1(s), \ldots$).
-%Notice that each $T_n$ is a stopping time w.r.t.~filtration $(\calF_n)_{n \in \omega}$.
+%Notice that each $T_n$ is a stopping time w.r.t.~filtration $(\calF_n)_{n \geq 0}$.
 We first state a useful property about r.v.~$T_0, T_1, T_2, \cdots$.
@@
 \label{lem:TnBounded}
-$(T_n)_{n \in\omega}$ is an increasing sequence of stopping times adapted to $(\calF_n)_{n \in \omega}$, and each $T_i$ is bounded.
+$(T_n)_{n \geq 0}$ is an increasing sequence of stopping times adapted to $(\calF_n)_{n \geq 0}$, and each $T_i$ is bounded.
 \end{restatable}
 
-%The random variable $T_M$ is a stopping time w.r.t.~filtration $(\calF_n)_{n \in \omega}$. 
+%The random variable $T_M$ is a stopping time w.r.t.~filtration $(\calF_n)_{n \geq 0}$. 
 %Since $f(M_n(s)) = 0$ iff $M_n(s)$ is a value, $T_M$ is the \emph{runtime of $M$}.
@@
 \label{lem:TMtY is a stopping time}
-$T_M^\tY$ is a stopping time adapted to $(\calF_{T_n})_{n \in \omega}$.
+$T_M^\tY$ is a stopping time adapted to $(\calF_{T_n})_{n \geq 0}$.
 \end{restatable}
@@
 Let $f$ be a ranking function on $M$.
-For 1, since $(T_n)_{n \in\omega}$ is an increasing sequence of stopping times, each is bounded and adapted to $(\calF_n)_{n \in \omega}$ (\Cref{lem:TnBounded}),
-and $(f(M_n))_{n \in \omega}$ is a supermartingale also adapted to $(\calF_n)_{n \in \omega}$ (\Cref{thm:rankable and strict rankable}),
-it follows from the Optional Sampling \Cref{thm:optional sampling} that $(Y_n)_{n \in \omega}$ where $Y_n := f(M_{T_n})$ is a supermartingale adapted to the filtration $(\calF_{T_n})_{n \in \omega}$.
-Notice the stopping time $T_M^\tY$ is also adapted to $(\calF_{T_n})_{n \in \omega}$ (\Cref{lem:TMtY is a stopping time}); and we have that $(Y_n)_{n \in \omega}$ is a $1$-ranking supermartingale.
+For 1, since $(T_n)_{n \geq 0}$ is an increasing sequence of stopping times, each is bounded and adapted to $(\calF_n)_{n \geq 0}$ (\Cref{lem:TnBounded}),
+and $(f(M_n))_{n \geq 0}$ is a supermartingale also adapted to $(\calF_n)_{n \geq 0}$ (\Cref{thm:rankable and strict rankable}),
+it follows from the Optional Sampling \Cref{thm:optional sampling} that $(Y_n)_{n \geq 0}$ where $Y_n := f(M_{T_n})$ is a supermartingale adapted to the filtration $(\calF_{T_n})_{n \geq 0}$.
+Notice the stopping time $T_M^\tY$ is also adapted to $(\calF_{T_n})_{n \geq 0}$ (\Cref{lem:TMtY is a stopping time}); and we have that $(Y_n)_{n \geq 0}$ is a $1$-ranking supermartingale.
 Therefore, by \Cref{lem:rank-PAST}, $T^{\tY}_M < \infty$ a.s.~and $\expect{T^{\tY}_M} < \infty$.
 
-%For 2, we have that $(f(M_n))_{n \in \omega}$ is a ranking supermartingale (\Cref{thm:rankable and strict rankable}) w.r.t.~stopping time $T_M$. 
+%For 2, we have that $(f(M_n))_{n \geq 0}$ is a ranking supermartingale (\Cref{thm:rankable and strict rankable}) w.r.t.~stopping time $T_M$. 
 %It follows immediately from \Cref{lem:rank-PAST} that $\expect{T_M} < \infty$.

### 2020-12-28 akr-thots-completeness.tex
Global replace: _{n \in \omega} -> _{n \geq 0}

@@
 \lo{E.g.~writing $I^{m+1} := I^m \, I$ where $I^1 := I$, we have $\rho_{I^m}(n) = 1$ if $n < m - 1$, and 0 otherwise.}
-Let $|\rho_N| := \sum_{n \in \omega} \rho_N(n)$.
+Let $|\rho_N| := \sum_{n \geq 0} \rho_N(n)$.
 
@@
 \lo{Define r.v.~$Z_n(s) := \widehat{\rho}_N$ where $\red^n(M, s) = (N, {-})$. 
-Then $(Z_n)_{n \in \omega}$ is a (generalised) supermartingale.
+Then $(Z_n)_{n \geq 0}$ is a (generalised) supermartingale.
 I am not sure if we need this.}

### 2020-12-28 probabilisticProgrammingMartingales.tex
Global replace: _{n \in \omega} -> _{n \geq 0}

@@
 \changed[lo]{For example, the term $(\tY \lambda x.x) \, \underline 0$ is not rankable.}
-It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \in \omega}$ is the reduction sequence starting from $M$, then $(f(M_n))_{n\in\omega}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales.
+It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \geq 0}$ is the reduction sequence starting from $M$, then $(f(M_n))_{n \geq 0}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales.
 \end{definition}
@@
 We assume $\tY f^{A \to B} \, x^A \, . \, \theta^B$ is a value.}
+\fi
 
 \begin{definition}
-We say that a supermartingale $(Y_n)_{n \in \omega}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \in \omega}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
+We say that a supermartingale $(Y_n)_{n \geq 0}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \geq 0}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
 \[
@@
 \lo{N.B.~$\set{T = n} = (\frac{1}{2^{n}}, \frac{1}{2^{n-1}}]$, and so, $\set{T > n} = [-1, \frac{1}{2^n}]$.}
-Plainly, $T$ is a stopping time, and $(Y_n)_{n \in \omega}$ is a supermartingale, adapted to the filtration $(\calF_n)_{n \in \omega}$ where $\calF_n = \calF$ for all $n$. 
-In this case, $(Y_n)_{n \in \omega}$ either tends to $4$ as $n \to \infty$, or drops to 0 at some point, with an exponentially decreasing probability. 
-With respect to the antitone function $\epsilon(x) = \frac{4-x}{4}$ and stopping time $T$, we have that $(Y_n)_{n \in \omega}$ is an antitone strict supermartingale, except that $\epsilon(4) = 0$, and $T = \infty$ with probability $\frac 1 2$, even though $4$ is larger than any value $(Y_n)_{n \in \omega}$ can actually take, and $\epsilon(Y_n)$ never actually reaches $0$.
-
+Plainly, $T$ is a stopping time, and $(Y_n)_{n \geq 0}$ is a supermartingale, adapted to the filtration $(\calF_n)_{n \geq 0}$ where $\calF_n = \calF$ for all $n$. 
+In this case, $(Y_n)_{n \geq 0}$ either tends to $4$ as $n \to \infty$, or drops to 0 at some point, with an exponentially decreasing probability. 
+With respect to the antitone function $\epsilon(x) = \frac{4-x}{4}$ and stopping time $T$, we have that $(Y_n)_{n \geq 0}$ is an antitone strict supermartingale, except that $\epsilon(4) = 0$, and $T = \infty$ with probability $\frac 1 2$, even though $4$ is larger than any value $(Y_n)_{n \geq 0}$ can actually take, and $\epsilon(Y_n)$ never actually reaches $0$.
 %\lo{@Andrew: I refer to the preceding highlighted claim. Did you mean: for all $n, s$, $\epsilon(Y_n(s)) > Y_n(s)$?} \akr{No. This phrasing should be clearer.}
 \end{example}
+
+\changed[lo]{Let $f$ be a ranking function on a closed SPCF term $M$.
+%Recall that $(f(M_n))_{n \geq 0}$ is a supermartingale.
+We say that $M$ is \emph{antitone rankable} by $f$ if the supermartingale $(f(M_n))_{n \geq 0}$ and the stopping time $T_M^{\tY}$, both adapted to the fitration $(\calF_n)_{n \geq 0}$ where $\calF_n := \sigma(M_1, \ldots, M_n)$, are antitone strict.}
 
@@
 \Cref{thm:rankable implies termination},
-take the set $S$ of traces as a probability space, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$.
-As before, $(Y_n)_{n \in \omega}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$.
+take the set $S$ of traces \lo{TODO: We should find a better name for traces $s \in S$.} as a probability space, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$.
+As before, $(Y_n)_{n \geq 0}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$.
 Therefore, it follows from \Cref{thm:a-c strict} that $T_M^{\tY}$ is almost-surely finite.

### 2020-12-28 ranking-supermartingales.tex
Global replace: _{n \in \omega} -> _{n \geq 0}

@@
 \begin{definition}\rm
-A sequence of random variables $(Y_n)_{n \in \omega}$ adapted to a filtration $(\calF_n)_{n \in \omega}$ is a \emph{supermartingale} if for all $n \in \omega$, $\expect{|Y_n|} < \infty$, and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$.
-\lo{The preceding means: $\forall A \in \calF_n$, $\int_A \dif \mu \, Y_{n+1} \leq \int_A \dif \mu \,Y_n$.}
-For $\epsilon > 0$ and a stopping time $T$ w.r.t.~$(\calF_n)_{n \in \omega}$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.
+\begin{enumerate}
+\item A sequence of random variables $(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \emph{supermartingale} if for all $n \in \omega$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~\changed[lo]{(i.e.~$\forall A \in \calF_n$, $\int_A \dif \mu \, Y_{n+1} \leq \int_A \dif \mu \,Y_n$).}
+\item Let $\epsilon > 0$.
+A stopping time $T$ and a supermartingale $(Y_n)_{n \geq 0}$, both adapted to filtration $(\calF_n)_{n \geq 0}$, are a \emph{$\epsilon$-ranking supermartingale} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.\footnote{For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable $s \mapsto \begin{cases}
+1 & \hbox{if $s \in A$}\\
+0 & \hbox{otherwise}
+\end{cases}$.}
 A \emph{ranking supermartingale} is a $\epsilon$-ranking supermartingale for some $\epsilon > 0$.
+\end{enumerate}
 \end{definition}
@@
 \begin{remark}
-\changed[lo]{The original definition of ranking supermartingale, due to \citep{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, does not mention stopping time explicitly:
-Fioriti and Hermanns call a supermartingale $(Y_n)_{n \in \omega}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot 1_{\set{Y_n > 0}}$.}
+\changed[lo]{The original definition of ranking supermartingale, due to \citep{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, does not refer to a stopping time explicitly:
+Fioriti and Hermanns call a supermartingale $(Y_n)_{n \geq 0}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot 1_{\set{Y_n > 0}}$.}
 \end{remark}
@@
 
-The following is an iterated version of Doob's well-known Optional Sampling Theorem (see, e.g., \cite[\S 6.7]{AshDD00}).
+The following is an iterated version of Doob's well-known Optional Sampling Theorem (see, e.g., \cite[\S 6.7]{AshDD00} and \cite[Theorem 7.2]{DBLP:conf/popl/FioritiH15}).
 \begin{theorem}[Optional Sampling]
 \label{thm:optional sampling}
-Let $(X_n)_{n\in \omega}$ be a supermartingale, and $(T_n)_{n \in \omega}$ a sequence of increasing stopping times, then $(X_{T_n})_{n \in \omega}$ is a supermartingale adapted to $(\calF_{T_n})_{n \in \omega}$ if one of the following conditions holds:
+Let $(X_n)_{n \geq 0}$ be a supermartingale, and $(T_n)_{n \geq 0}$ a sequence of increasing stopping times, both adapted to filtration $(\calF_n)_{n \geq 0}$, then $(X_{T_n})_{n \geq 0}$ is a supermartingale adapted to $(\calF_{T_n})_{n \geq 0}$ if one of the following conditions holds:
 \begin{enumerate}
 \item each $T_n$ is bounded i.e.~$T_n < c_n$ where $c_n$ is a constant
-\item $(X_n)_{n\in \omega}$ is uniformly integrable.
+\item $(X_n)_{n \geq 0}$ is uniformly integrable.
 \end{enumerate}
@@
 
-\begin{lemma}[{\citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}}]
+\begin{lemma}%[{\citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}}]
 % \citep[Prop 1]{DBLP:conf/popl/ChatterjeeFNH16}}
 \label{lem:rank-PAST}
-Let $(Y_n)_{n \in \omega}$ be a $\epsilon$-ranking supermartingale with respect to the stopping time $T$, then $T < \infty$ almost surely, and $\expect{T} \leq \dfrac{\expect{Y_0}}{\epsilon}$.
+Let $(Y_n)_{n \geq 0}$ be a $\epsilon$-ranking supermartingale with respect to the stopping time $T$, then $T < \infty$ almost surely, and $\expect{T} \leq \dfrac{\expect{Y_0}}{\epsilon}$.
 \end{lemma}
 
-\begin{proof}[Proof Sketch]
-First establish by induction on $n$: 
+\begin{proof}
+We first prove by induction on $n$: 
 \[
@@
 
-A main result is the following theorem.
-\begin{theorem} 
+Our first result is the following theorem.
+\begin{theorem}[Rankability] 
 \label{thm:rankable and strict rankable}
-If a closed SPCF term is rankable (respectively, strictly rankable) by $f$, then $(f(M_n))_{n \in \omega}$ is a 
-supermartingale (respectively, ranking supermartingale) adapted to the filtration $(\mathcal{F}_n)_{n \in \omega}$ where $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
+If a closed SPCF term is rankable (respectively, strictly rankable) by $f$, then $(f(M_n))_{n \geq 0}$ is a 
+supermartingale (respectively, ranking supermartingale) adapted to the filtration $(\mathcal{F}_n)_{n \geq 0}$ where $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
 \end{theorem}
 
-\paragraph{Proof of \Cref{thm:rankable and strict rankable}.}
+%\subsection{Proof of \Cref{thm:rankable and strict rankable}.}
 \emph{Notation}. Given an $\omega$-sequence (e.g.~$s \in S$) and $m \in \omega$, we write $s_{\leq m} \in I^m$ to mean the prefix of $s$ of length $m$.
@@
 \end{align}
-The random variable $T_M^\tY$ is known as the \emph{$\tY$-runtime of $M$}, and it can equivalently be defined as the number of $\tY$-reduction steps in the reduction sequence of $M$.
+The random variable $T_M^\tY$, which we call the \emph{$\tY$-runtime of $M$}, can equivalently be defined as the number of $\tY$-reduction steps in the reduction sequence of $M$. 
 Note that, as the type system ensures that the reduction relation excluding $\tY$-reduction is strongly normalising, only finitely many reductions can occur in a row without one of them being a $\tY$-reduction, therefore $T_M^\tY < \infty$ a.s.~iff $M$ is AST.
+
+The r.v.~$T_M^\tY$ is a stopping time adapted to the filtration $(\calF_{T_n})_{n \geq 0}$.
 
@@
 \begin{proof}
-For 1, observe that $T_0, T_1, T_2, \cdots$ are an increasing sequence of stopping times adapted to $(\calF_n)_{n \in \omega}$, and each $T_i$ is bounded.\lo{Prove it.}
-Thanks to \Cref{thm:rankable and strict rankable}, 
-$(X_n)_{n \in \omega}$ is a supermartingale;
-it then follows from \Cref{thm:optional sampling} that $(Y_n)_{n \in \omega}$ is a ($1$-ranking) supermartingale with respect to the filtration $(\calF_{T_n})_{n \in \omega}$ and the stopping time $T^{\tY}_M$.
+Let $f$ be a ranking function on $M$.
+For 1, observe that $T_0, T_1, T_2, \cdots$ are an increasing sequence of stopping times, each is bounded \lo{Prove it.} and adapted to $(\calF_n)_{n \geq 0}$.
+Then, since $(f(M_n))_{n \geq 0}$ is a supermartingale also adapted to $(\calF_n)_{n \geq 0}$ (\Cref{thm:rankable and strict rankable}),
+it follows from the Optional Sampling \Cref{thm:optional sampling} that $(Y_n)_{n \geq 0}$ where $Y_n := f(M_{T_n})$ is a supermartingale adapted to the filtration $(\calF_{T_n})_{n \geq 0}$.
+Since $T_M^\tY$ is a stopping time also adapted to $(\calF_{T_n})_{n \geq 0}$, we have that $(Y_n)_{n \geq 0}$ is a $1$-ranking supermartingale.
 Therefore, by \Cref{lem:rank-PAST}, $T^{\tY}_M < \infty$ a.s.~and $\expect{T^{\tY}_M} < \infty$.

### 2020-12-28 y-stopping-time.tex
More examples
Global replace: _{n \in \omega} -> _{n \geq 0}

@@
 \begin{lemma}
-$T_0, T_1, T_2, \cdots$ are an increasing sequence of stopping times adapted to $(\calF_n)_{n \in \omega}$, and each $T_i$ is bounded.
+$T_0, T_1, T_2, \cdots$ are an increasing sequence of stopping times adapted to $(\calF_n)_{n \geq 0}$, and each $T_i$ is bounded.
 \end{lemma}

### 2020-12-28 LICS21/supermartingales.tex
For consistency: global replace indicator function 1_{...} -> {\bf 1}_{...}

@@
 \changed[lo]{Our notion of ranking supermartingale is a slight generalisation of the original definition, due to \citep{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, which does not refer to a stopping time explicitly:
-Fioriti and Hermanns call a supermartingale $(Y_n)_{n \geq 0}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot 1_{\set{Y_n > 0}}$.}
+Fioriti and Hermanns call a supermartingale $(Y_n)_{n \geq 0}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot {\bf 1}_{\set{Y_n > 0}}$.}
 \end{remark}

### 2020-12-28 ranking-supermartingales.tex
For consistency: global replace indicator function 1_{...} -> {\bf 1}_{...}

@@
 \changed[lo]{The original definition of ranking supermartingale, due to \citep{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, does not refer to a stopping time explicitly:
-Fioriti and Hermanns call a supermartingale $(Y_n)_{n \geq 0}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot 1_{\set{Y_n > 0}}$.}
+Fioriti and Hermanns call a supermartingale $(Y_n)_{n \geq 0}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot {\bf 1}_{\set{Y_n > 0}}$.}
 \end{remark}

### 2020-12-28 LICS21/SPCF.tex
n \in \omega -> n \geq 0, and similar.

@@
 \changed[lo]{Precisely, following \cite{DBLP:conf/icfp/BorgstromLGS16},
-we view $\Lambda$ as $\bigcup_{m\in\omega} \big(\mathsf{Sk}_m \times \Real^m \big)$,
+we view $\Lambda$ as $\bigcup_{m\geq 0} \big(\mathsf{Sk}_m \times \Real^m \big)$,
 where $\mathsf{Sk}_m$ is the set of SPCF terms with exactly $m$ place-holders (a.k.a.~\emph{skeleton terms}) for numerals.
 Thus identified, we give $\Lambda$ the countable disjoint union topology of the product topology of the discrete topology on $\mathsf{Sk}_m$ and the standard topology on $\Real^m$.
-Note that the connected components of $\Lambda$ have the form $\{M\} \times \Real^m$, with $M$ ranging over $\mathsf{Sk}_m$, and $m$ over $\omega$. 
+Note that the connected components of $\Lambda$ have the form $\{M\} \times \Real^m$, with $M$ ranging over $\mathsf{Sk}_m$, and $m$ over $\mathbb{N}$. 
 We fix the Borel algebra of this topology to be the $\sigma$-algebra on $\Lambda$.}

### 2020-12-28 LICS21/supermartingales.tex
n \in \omega -> n \geq 0, and similar.

@@
 Let $T$ be a random variable that takes values in $\mathbb{N} \cup \set{\infty}$.
-We call $T$ a \emph{stopping time adapted to $(\calF_n)_{n \geq 0}$} just if $\set{T = n} \in \calF_n$, for all $n \in \omega$.}
+We call $T$ a \emph{stopping time adapted to $(\calF_n)_{n \geq 0}$} just if $\set{T = n} \in \calF_n$, for all $n \geq 0$.}
 
@@
 \begin{enumerate}
-\item A sequence of random variables $(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \emph{supermartingale} if for all $n \in \omega$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~\changed[lo]{(i.e.~for all $A \in \calF_n$, $\int_A \mathbb{P}(\dif \omega) \, Y_{n+1}(\omega) \leq \int_A \mathbb{P}(\dif \omega) \,Y_n(\omega)$).}
+\item A sequence of random variables $(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \emph{supermartingale} if for all $n \geq 0$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~\changed[lo]{(i.e.~for all $A \in \calF_n$, $\int_A \mathbb{P}(\dif \omega) \, Y_{n+1}(\omega) \leq \int_A \mathbb{P}(\dif \omega) \,Y_n(\omega)$).}
 \item Let $\epsilon > 0$.
@@
 \begin{proof}
-We first prove (by induction on $n$): for all $n \in \omega$
+We first prove (by induction on $n$): for all $n \geq 0$
 \[
-%\forall n \in \omega \, . \, 
+%\forall n \geq 0 \, . \, 
 \expect{Y_n} \leq \expect{Y_0} - \epsilon \cdot \big(\textstyle\sum_{i=0}^{n-1} \mathbb{P}[T > i]\big).
@@
 \[
-\calF_T := \set{A \in \calF \mid \forall i \in \omega \, . \, A \cap \set{T \leq i} \in \calF_i}
+\calF_T := \set{A \in \calF \mid \forall i \geq 0 \, . \, A \cap \set{T \leq i} \in \calF_i}
 \]
@@
 \changed[lo]{For the rest of this section, fix the probability space $(S, \calF, \mu)$, and a closed SPCF term $M$.
-For $n \in \omega$, define the random variables $M_n(s) := \pi_0 (\red^n(M, s))$, and the filtration $(\mathcal{F}_n)_{n \geq 0}$ where $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.}
+For $n \geq 0$, define the random variables $M_n(s) := \pi_0 (\red^n(M, s))$, and the filtration $(\mathcal{F}_n)_{n \geq 0}$ where $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.}
 
@@
 
-Given an $\omega$-sequence (e.g.~$s \in S$) and $m \in \omega$, we write $s_{\leq m} \in I^m$ to mean the prefix of $s$ of length $m$.
+Given an $\omega$-sequence (e.g.~$s \in S$) and $m \geq 0$, we write $s_{\leq m} \in I^m$ to mean the prefix of $s$ of length $m$.
 %Fix a closed SPCF term $M \in \Lambda^0$, and a ranking function $f$ for it, satisfying $f(N) = 0$ iff $N$ is a value.
-%Let $n \in \omega$, define the following random variables on the probability space $(S, \calF, \mu)$:
+%Let $n \geq 0$, define the following random variables on the probability space $(S, \calF, \mu)$:
 %\begin{align*}
@@
 %\end{align*}
-For $n \in \omega$, define
+For $n \geq 0$, define
 \begin{align*}
@@
 \iffalse
-For $i \in \{1, 2, 3\}$ and $n \in \omega$, define function $f_{i, n+1}(M) : S \to \Real$ by
+For $i \in \{1, 2, 3\}$ and $n \geq 0$, define function $f_{i, n+1}(M) : S \to \Real$ by
 \[

### 2020-12-28 LICS21/SPCF.tex
(Global change: probability space (S, \calF, \mu) -> (\entrosp, \calF, \mu_\entrosp)

@@
 
-The language SPCF is a simply-typed lambda calculus with sampling of real numbers from $[0,1]$ and unbounded scoring, following \cite{MakOPW21}. 
+The language SPCF is a simply-typed lambda calculus with sampling of real numbers from the closed interval $[0,1]$ and unbounded scoring, following \cite{MakOPW21}. 
 Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
 \begin{align*}
-  \text{types } A, B &::= \textsf{R}  \mid  A \to B \\
-  \text{values } V &::= \lambda x.M  \mid  \underline{r} \\
-  \text{terms } M, N &::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY M \\
+  \text{\em types } A, B &::= \textsf{R}  \mid  A \to B \\
+  \text{\em values } V &::= \lambda x.M  \mid  \underline{r} \\
+  \text{\em terms } M, N &::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY M \\
    & \qquad \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample  \mid  \tscore(M)
@@
 \begin{align*}
-  E[(\lambda x.M) N] & \to E[M[N/x]] \\
+  E[(\lambda x.M) V] & \to E[M[V/x]] \\
   E[\underline f (\underline r_1, \dots , \underline r_n)] & \to E[\underline{f(r_1,\dots,r_n)}] \\
@@
 To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples.
-Let $ I = [0,1] \subset \mathbb{R} $, and let $S = I^{\mathbb{N}}$, with the \changed[lo]{Borel $\sigma$-algebra $\calF$} and the probability measure, denoted $\mu$, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
-Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and $\mu (\prod_{i=0}^\infty X_i) = \prod_{i=0}^\infty \Leb(X_i)$.
-The maps $\pi_h:S \to I, \; \pi_t:S \to S$ popping the first element are then measurable.
-\changed[lo]{Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(S, \calF, \mu)$ the \emph{entropy space}.}
+Let $ I = [0,1] \subset \mathbb{R} $, and let $\entrosp = I^{\mathbb{N}}$, with the \changed[lo]{Borel $\sigma$-algebra $\calF$} and the probability measure, denoted $\mu_{\entrosp}$, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
+Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and $\mu_{\entrosp} (\prod_{i=0}^\infty X_i) = \prod_{i=0}^\infty \Leb(X_i)$.
+The maps $\pi_h:\entrosp \to I, \; \pi_t:\entrosp \to \entrosp$ popping the first element are then measurable.
+\changed[lo]{Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(\entrosp, \calF, \mu_{\entrosp})$ the \emph{entropy space}.}
 
@@
 
-The one-step reduction is given by the function $\red : \Lambda^0 \times S \to \Lambda^0 \times S$ where
+The one-step reduction is given by the function $\red : \Lambda^0 \times \entrosp \to \Lambda^0 \times \entrosp$ where
 \begin{equation*}
@@
 The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid M \text{ terminates at } s \}$ is measurable.
-We say that a term $M$ is \emph{almost-surely terminating} (AST) just if $\mu(\{s \mid M \text{ terminates at } s\}) = 1$.
+We say that a term $M$ is \emph{almost-surely terminating} (AST) just if $\mu_{\entrosp}(\{s \mid M \text{ terminates at } s\}) = 1$.
 
@@
 \] 
-which generates a geometric distribution, terminates on the set $S \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas 
+which generates a geometric distribution, terminates on the set $\entrosp \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas 
 \[

### 2020-12-28 LICS21/supermartingales.tex
(Global change: probability space (S, \calF, \mu) -> (\entrosp, \calF, \mu_\entrosp)

@@
 \changed[lo]{Fix a probability space $(\Omega, \calF, \mathbb{P})$, and a filtration $(\calF_n)_{n \geq 0}$. 
-(In subsequent applications to termination analysis, $\Omega = S$, and $\mathbb{P} = \mu$.)
+(In subsequent applications to termination analysis, $\Omega = S$, and $\mathbb{P} = \mu_{\entrosp}$.)
 Let $T$ be a random variable that takes values in $\mathbb{N} \cup \set{\infty}$.
@@
 \item Let $\epsilon > 0$. 
-A stopping time $T$ and a supermartingale $(Y_n)_{n \geq 0}$, both adapted to filtration $(\calF_n)_{n \geq 0}$, are a \emph{$\epsilon$-ranking supermartingale} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.\footnote{For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable defined by ${\bf 1}_A(s) := 1$ if $s \in A$, and $0$ otherwise.}
-A \emph{ranking supermartingale} is a $\epsilon$-ranking supermartingale for some $\epsilon > 0$.
+Given a stopping time $T$ and a supermartingale $(Y_n)_{n \geq 0}$, both adapted to filtration $(\calF_n)_{n \geq 0}$, 
+we say that $(Y_n)_{n \geq 0}$ is a \emph{$\epsilon$-ranking supermartingale w.r.t.~$T$} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.
+{(For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable defined by ${\bf 1}_A(\omega) := 1$ if $\omega \in A$, and $0$ otherwise.)} \lo{\LaTeX\ does not allow the preceding bracketed sentence to be typeset as a footnote.
+Strange!}
+A \emph{ranking supermartingale} w.r.t.~$T$ is a $\epsilon$-ranking supermartingale w.r.t.~$T$ for some $\epsilon > 0$.
 \end{enumerate}
@@
 \label{lem:rank-PAST}
-Let $(Y_n)_{n \geq 0}$ be a $\epsilon$-ranking supermartingale with respect to the stopping time $T$, then $T < \infty$ almost surely, and $\expect{T} \leq \frac{\expect{Y_0}}{\epsilon}$.
+Let $(Y_n)_{n \geq 0}$ be a $\epsilon$-ranking supermartingale w.r.t.~the stopping time $T$, then $T < \infty$ a.s., and $\expect{T} \leq \frac{\expect{Y_0}}{\epsilon}$.
 \end{lemma}
@@
 \changed[lo]{\Cref{lem:rank-PAST} is a slight extension of \citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15},
-which asserts the same result for the specific stopping time $T : s \mapsto \min\set{n \mid Y_n(s) = 0}$.}
+which asserts the same result but for the specific stopping time $T : \omega \mapsto \min\set{n \mid Y_n(\omega) = 0}$.}
 \end{remark}
@@
 \label{thm:optional sampling}
-Let $(X_n)_{n \geq 0}$ be a supermartingale, and $(T_n)_{n \geq 0}$ a sequence of increasing stopping times, both adapted to filtration $(\calF_n)_{n \geq 0}$, then $(X_{T_n})_{n \geq 0}$ is a supermartingale adapted to $(\calF_{T_n})_{n \geq 0}$ if one of the following conditions holds:
+Let $(X_n)_{n \geq 0}$ be a supermartingale, and $(T_n)_{n \geq 0}$ a sequence of increasing stopping times, all adapted to filtration $(\calF_n)_{n \geq 0}$, then $(X_{T_n})_{n \geq 0}$ is a supermartingale adapted to $(\calF_{T_n})_{n \geq 0}$ if one of the following conditions holds:
 \begin{enumerate}
@@
 %\emph{Notation.}
-\changed[lo]{For the rest of this section, fix the probability space $(S, \calF, \mu)$, and a closed SPCF term $M$.
+\changed[lo]{Henceforth, fix the probability space $(S, \calF, \mu_{\entrosp})$, and a closed SPCF term $M$.
 For $n \geq 0$, define the random variables $M_n(s) := \pi_0 (\red^n(M, s))$, and the filtration $(\mathcal{F}_n)_{n \geq 0}$ where $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.}
@@
 \label{thm:rankable and strict rankable}
-If a closed SPCF term $M$ is rankable (respectively, strictly rankable) by $f$ satisfying $f(N) = 0$ iff $N$ is a value, then $(f(M_n))_{n \geq 0}$ is a supermartingale (respectively, ranking supermartingale with stopping time $T_M: s \mapsto \min \set{n \mid f(M_n(s)) = 0}$) adapted to the filtration $(\mathcal{F}_n)_{n \geq 0}$.
+If a closed SPCF term $M$ is rankable (respectively, strictly rankable) by $f$ satisfying $f(N) = 0$ iff $N$ is a value, then $(f(M_n))_{n \geq 0}$ is a supermartingale (respectively, ranking supermartingale w.r.t.~stopping time $T_M: s \mapsto \min \set{n \mid f(M_n(s)) = 0}$) adapted to $(\mathcal{F}_n)_{n \geq 0}$.
 %where $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
@@
 %M_n(s) &:= \pi_0 (\red^n(M, s))\\
-\ndraw{n}{s} &:= |\{k < n \mid \exists E: M_k(s) = E[\tsample]\}|,
+\ndraw{n}{s} &:= |\{k < n \mid \exists E: M_k(s) = E[\tsample]\}|
 \end{align*}
-so that $\pi_1 (\red^n(M, s)) = \overbrace{\pi_t( \cdots (\pi_t}^{\ndraw n s}(s))$. 
+so that $\pi_1 (\red^n(M, s)) = %\overbrace{\pi_t( \cdots (\pi_t}^{\ndraw n s}(s))$. 
+\pi_t( \cdots (\pi_t (s))$, with $\pi_t$ applied ${\ndraw n s}$ times.
 %and the filtration $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
@@
 
-In the rest of this section we show that if $M \in \Lambda^0$ is rankable by $f$, then $M$ is AST.
+In the rest of this section we show that if $M \in \Lambda^0$ is rankable, then $M$ is AST: in other words, the method of ranking function is sound for proving a.s.~termination of SPCF programs.
 

### 2020-12-28 LICS21/supermartingales.tex
minor

@@
 \changed[lo]{Fix a probability space $(\Omega, \calF, \mathbb{P})$, and a filtration $(\calF_n)_{n \geq 0}$. 
-(In subsequent applications to termination analysis, $\Omega = S$, and $\mathbb{P} = \mu_{\entrosp}$.)
+(In subsequent applications to termination analysis, $\Omega = \entrosp$, and $\mathbb{P} = \mu_{\entrosp}$.)
 Let $T$ be a random variable that takes values in $\mathbb{N} \cup \set{\infty}$.
@@
 
-Given an $\omega$-sequence (e.g.~$s \in S$) and $m \geq 0$, we write $s_{\leq m} \in I^m$ to mean the prefix of $s$ of length $m$.
+Given an $\omega$-sequence (e.g.~$s \in \entrosp$) and $m \geq 0$, we write $s_{\leq m} \in I^m$ to mean the prefix of $s$ of length $m$.
 %Fix a closed SPCF term $M \in \Lambda^0$, and a ranking function $f$ for it, satisfying $f(N) = 0$ iff $N$ is a value.
@@
 Take $s \in A \in \calF_n$ with $\ndraw{n}{s} = l$.
-For any $s'\in S$, if $s_{\leq l} = s_{\leq l}'$ then $s' \in A$.
+For any $s'\in \entrosp$, if $s_{\leq l} = s_{\leq l}'$ then $s' \in A$.
 It follows that ${\set{s_{\leq l}} \cdot I^\omega} \subseteq A$.
@@
 Henceforth fix an $n \geq 0$, and define $\mathbf{T}_i := \set{s \mid M_n(s) \hbox{ is type-$i$}}$.
-It is straightforward to see that each $\mathbf{T}_i \in \calF_n$, and $\set{\mathbf{T}_1, \mathbf{T}_2, \mathbf{T}_3, \mathbf{T}_4}$ is a partition of $S$.
+It is straightforward to see that each $\mathbf{T}_i \in \calF_n$, and $\set{\mathbf{T}_1, \mathbf{T}_2, \mathbf{T}_3, \mathbf{T}_4}$ is a partition of $\entrosp$.
 %By abuse of language, we refer to the three respective types of redexes as type-$1$, 2 and 3.
@@
 \iffalse
-For $i \in \{1, 2, 3\}$ and $n \geq 0$, define function $f_{i, n+1}(M) : S \to \Real$ by
+For $i \in \{1, 2, 3\}$ and $n \geq 0$, define function $f_{i, n+1}(M) : \entrosp \to \Real$ by
 \[
@@
 
-Here we use the notation: given function $g : S \to \Real$ and $U \subseteq S$, $g[U] : S \to \Real$ denotes the function
+Here we use the notation: given function $g : \entrosp \to \Real$ and $U \subseteq \entrosp$, $g[U] : \entrosp \to \Real$ denotes the function
 \[
@@
 The inequality (\ref{eqn:ts}) follows immediately from the lemma below; 
-and (\ref{eqn:partition}) holds because $\{\mathbf{T}_1, \mathbf{T}_2, \mathbf{T}_3, \mathbf{T}_4\}$ is a partition of $S$.
+and (\ref{eqn:partition}) holds because $\{\mathbf{T}_1, \mathbf{T}_2, \mathbf{T}_3, \mathbf{T}_4\}$ is a partition of $\entrosp$.
 \fi
@@
 \item $\calI$ is a countable indexing set
-\item $E_i[\cdot][\tsample] \in \mathsf{Sk}_{j_i}$, and $\sigma_i : S \to \Real^{j_i}$, and $\rho(s) := \pi_h(\pi_1(\red^n(M, s))) = \pi_h(\pi_t^{l_i}(s)) \in \Real$ 
+\item $E_i[\cdot][\tsample] \in \mathsf{Sk}_{j_i}$, and $\sigma_i : \entrosp \to \Real^{j_i}$, and $\rho(s) := \pi_h(\pi_1(\red^n(M, s))) = \pi_h(\pi_t^{l_i}(s)) \in \Real$ 
 \item $\set{U_i}_{i \in \calI}$ is a partition of $\mathbf{T}_2$, 
@@
 %Moreover, for any $A \in \calF_n$, if $s \in A$ then $\set{s_{\leq l}} \cdot I^\omega \subseteq A$.
-This means that for any measurable $g : S \to \Real_{\geq 0}$, if $g(s)$ only depends on the prefix of $s$ of length $(l_i+1)$, then, writing $\widehat{g} : I^{l_i+1} \to \Real_{\geq 0}$ where $g(s) = \widehat{g}(s_{\leq l_i+1})$, we have: for any $A \in \calF_n$
+This means that for any measurable $g : \entrosp \to \Real_{\geq 0}$, if $g(s)$ only depends on the prefix of $s$ of length $(l_i+1)$, then, writing $\widehat{g} : I^{l_i+1} \to \Real_{\geq 0}$ where $g(s) = \widehat{g}(s_{\leq l_i+1})$, we have: for any $A \in \calF_n$
 \begin{equation}

### 2020-12-28 LICS21/antitone.tex
For consistency: global replace indicator function 1_{...} -> {\bf 1}_{...}
Changed S and \mu to \entrosp and \mu_\entrosp for all files except confluent.tex

@@
 \[
-\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot 1_{\set{T > n}}
+\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot {\bf 1}_{\set{T > n}}
 \]
@@
   \step[\geq]{antitone strict assumption}
-\expect{\epsilon(Y_n) \cdot 1_{\set{T > n}}}
+\expect{\epsilon(Y_n) \cdot {\bf 1}_{\set{T > n}}}
 %  \step[=]{definition of conditional expectation for a discrete variable}
@@
 \Cref{thm:rankable implies termination},
-take the set $S$ of traces \lo{TODO: We should find a better name for traces $s \in S$.} as a probability space, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$.
+take the set $\entrosp$ of traces \lo{TODO: We should find a better name for traces $s \in \entrosp$.} as a probability space, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$.
 As before, $(Y_n)_{n \geq 0}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$.
@@
   The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on antitone partial ranking functions.
-Define $f_2(N) = \int_S f_1(\nnext(N,s)) + \epsilon(f_1(\nnext(N,s))) g(N,s) \, \mu(\mathrm d s)$.
+Define $f_2(N) = \int_\entrosp f_1(\nnext(N,s)) + \epsilon(f_1(\nnext(N,s))) g(N,s) \, \mu_{\entrosp}(\mathrm d s)$.
 For any term $N$ where $f$ is defined, $f_2(N) = f(N)$, and the value that $f_2$ would have at $N$ if $f$ were not defined at $N$ is $\leq f(N)$, by the third condition on antitone partial ranking functions.
@@
 For any term $N$ which is not of the form $E[R]$ for some $\tY$-redex $R$, this is trivial.
-If $R$ is a $\tY$-redex, then $\epsilon(f_2(N)) \leq \int_S \epsilon(f_1(\nnext(N,s)))\, \mu(\mathrm d s)$ by the convexity of $\epsilon$, because $n$ is bounded.
+If $R$ is a $\tY$-redex, then $\epsilon(f_2(N)) \leq \int_\entrosp \epsilon(f_1(\nnext(N,s)))\, \mu_{\entrosp}(\mathrm d s)$ by the convexity of $\epsilon$, because $n$ is bounded.
 Therefore, the (total) function $f_2$, which agrees with $f$ on $f$'s domain, is an antitone ranking function on $M$ (with the same function $\epsilon$ if it's convex).
@@
     & \expect{Y_{n+1} \mid \mathcal F_n} \\
-    &=  \expect{y_{n+1} \, 1_{\set{T \geq n+1}} \mid \mathcal F_n} \\
-    &=  y_{n+1} \, \expect{1_{\set{T \geq n+1}} \mid \mathcal F_n} \\ 
-    &=  y_{n+1} \, 1_{\set{T \geq n}} \frac{\mathbb P[T > n+1]}{\mathbb P[T > n]} \\
-    &=  (y_n - \epsilon(y_n)) \, 1_{\set{T \geq n}} \\
-    &= Y_n - \epsilon(y_n) \, 1_{\set{T \geq n}}
+    &=  \expect{y_{n+1} \, {\bf 1}_{\set{T \geq n+1}} \mid \mathcal F_n} \\
+    &=  y_{n+1} \, \expect{{\bf 1}_{\set{T \geq n+1}} \mid \mathcal F_n} \\ 
+    &=  y_{n+1} \, {\bf 1}_{\set{T \geq n}} \frac{\mathbb P[T > n+1]}{\mathbb P[T > n]} \\
+    &=  (y_n - \epsilon(y_n)) \, {\bf 1}_{\set{T \geq n}} \\
+    &= Y_n - \epsilon(y_n) \, {\bf 1}_{\set{T \geq n}}
 \end{align*}

### 2020-12-28 LICS21/supermartingales.tex
Changed S and \mu to \entrosp and \mu_\entrosp for all files except confluent.tex

@@
 %\emph{Notation.}
-\changed[lo]{Henceforth, fix the probability space $(S, \calF, \mu_{\entrosp})$, and a closed SPCF term $M$.
+\changed[lo]{Henceforth, fix the probability space $(\entrosp, \calF, \mu_{\entrosp})$, and a closed SPCF term $M$.
 For $n \geq 0$, define the random variables $M_n(s) := \pi_0 (\red^n(M, s))$, and the filtration $(\mathcal{F}_n)_{n \geq 0}$ where $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.}
@@
 }
-%Take $M \in \Lambda^0$ (closed SPCF terms). Define, for each $n \in \omega$, the random variable $M_n : S \to \Lambda^0$ by $M_n(s) := \pi_0 (\red^n(M, s))$.
+%Take $M \in \Lambda^0$ (closed SPCF terms). Define, for each $n \in \omega$, the random variable $M_n : \entrosp \to \Lambda^0$ by $M_n(s) := \pi_0 (\red^n(M, s))$.
 
@@
 \)
-(We use the (measurable) function $\sigma_i : S \to \Real^{j_i}$ to skolemise the existentially quantified $r$.)
+(We use the (measurable) function $\sigma_i : \entrosp \to \Real^{j_i}$ to skolemise the existentially quantified $r$.)
 Equivalently
@@
 Let $f$ be a ranking function on $M$ satisfying $f(N) = 0$ iff $N$ is a value.
-Define random variables on the probability space $(S, \calF, \mu)$:
+Define random variables on the probability space $(\entrosp, \calF, \mu)$:
 \begin{align}

### 2020-12-28 LICS21/supermartingales.tex

@@
 \begin{definition}\rm
+\label{def:ranking function}
 A \emph{ranking function on $M \in \Lambda^0$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
@@
 It seems cleaner to include this in the definition.}
+\akr{The fact that the ranking function can be 0 in some cases before it reaches a value is necessary to get \Cref{thm:minimal} to work neatly.}
+\lo{OK. Let's leave \Cref{def:ranking function} as it is.}
 We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that $f(N) = 0$ iff $N$ is a value; and for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.
@@
 \changed[lo]{Henceforth, fix the probability space $(\entrosp, \calF, \mu_{\entrosp})$, and a closed SPCF term $M$.
-For $n \geq 0$, define the random variables $M_n(s) := \pi_0 (\red^n(M, s))$, and the filtration $(\mathcal{F}_n)_{n \geq 0}$ where $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.}
+For $n \geq 0$, define the random variables 
+\begin{align*}
+M_n(s) &:= \pi_0 (\red^n(M, s))\\
+T_M(s) &:= \min \set{n \mid \hbox{$M_n(s)$ is a value}}
+\end{align*} 
+and the filtration $(\mathcal{F}_n)_{n \geq 0}$ where $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
+Thus $T_M$ is the \emph{termination time} of $M$.}
 
@@
 \label{thm:rankable and strict rankable}
-If a closed SPCF term $M$ is rankable (respectively, strictly rankable) by $f$ satisfying $f(N) = 0$ iff $N$ is a value, then $(f(M_n))_{n \geq 0}$ is a supermartingale (respectively, ranking supermartingale w.r.t.~stopping time $T_M: s \mapsto \min \set{n \mid f(M_n(s)) = 0}$) adapted to $(\mathcal{F}_n)_{n \geq 0}$.
+If a closed SPCF term $M$ is rankable (respectively, strictly rankable) by $f$ 
+%satisfying $f(N) = 0$ iff $N$ is a value, 
+then $(f(M_n))_{n \geq 0}$ is a supermartingale (respectively, ranking supermartingale w.r.t.~stopping time $T_M$) adapted to $(\mathcal{F}_n)_{n \geq 0}$.
 %where $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
@@
 \[
-\int_A \mu(\dif s) \, f(M_{n+1})[s \in \mathbf{T}_i] \leq \int_A \mu(\dif s) \, f(M_n)[s \in \mathbf{T}_i]
+\int_A \mu_{\entrosp}(\dif s) \, f(M_{n+1})[s \in \mathbf{T}_i] \leq \int_A \mu_{\entrosp}(\dif s) \, f(M_n)[s \in \mathbf{T}_i]
 \; \footnote{where (the Iverson bracket) $[P] := 1$ if the statement $P$ hold, and 0 otherwise.}
@@
 \changed[lo]{Thus 
-%$\int_A \mu(\dif s) \, f(M_{n+1}) \leq \int_A \mu(\dif s) \, f(M_n)$, i.e., 
+%$\int_A \mu_{\entrosp}(\dif s) \, f(M_{n+1}) \leq \int_A \mu_{\entrosp}(\dif s) \, f(M_n)$, i.e., 
 $\mathbb{E}[f(M_{n+1}) \mid \calF_n] \leq f(M_n)$ a.s.}
@@
 \begin{equation}
-\int_{A \cap U_i}  \mu(\dif s) \, g(s) = 
+\int_{A \cap U_i}  \mu_{\entrosp}(\dif s) \, g(s) = 
 \int_{(A \cap U_i)_{\leq l_i+1}} \Leb_{l_i+1}(\dif t) \, \widehat{g}(t)
@@
 \begin{align}
-& \int_{A \cap U_i} \mu(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])) \nonumber \\
+& \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])) \nonumber \\
 & \leq 
-\int_{A \cap U_i} \mu(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\tsample])).
+\int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\tsample])).
 \label{eqn:a u ui}
@@
 \begin{align*}
-& \int_A \mu(\dif s) \, f_{n+1}(M)[s \in \mathbf{T}_i] \\
+& \int_A \mu_{\entrosp}(\dif s) \, f_{n+1}(M)[s \in \mathbf{T}_i] \\
 &= 
-\int_A \mu(\dif s) \, \sum_{i \in \calI} f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])[s \in U_i] \\
+\int_A \mu_{\entrosp}(\dif s) \, \sum_{i \in \calI} f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])[s \in U_i] \\
 &= 
-\sum_{i \in \calI} \int_{A \cap U_i} \mu(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])\\
+\sum_{i \in \calI} \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])\\
 &\leq 
-\sum_{i \in \calI} \int_{A \cap U_i} \mu(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\tsample])
+\sum_{i \in \calI} \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\tsample])
 \quad \hbox{$\because$ (\ref{eqn:a u ui})}\\
 &= 
-\int_{A} \mu(\dif s) \, \sum_{i \in \calI} \, f(E_i[\underline{\sigma_i(s)}][\tsample])[s \in U_i]\\
-&= \int_{A} \mu(\dif s) f_{n}(M)[s \in \mathbf{T}_2]
+\int_{A} \mu_{\entrosp}(\dif s) \, \sum_{i \in \calI} \, f(E_i[\underline{\sigma_i(s)}][\tsample])[s \in U_i]\\
+&= \int_{A} \mu_{\entrosp}(\dif s) f_{n}(M)[s \in \mathbf{T}_2]
 %\quad \hbox{$\because$ (\ref{eqn:f 2 n+1})}
@@
 \begin{align*}
-\int_{S} \mu(\dif s) \, |f(M_n)(s)| 
+\int_{\entrosp} \mu_{\entrosp}(\dif s) \, |f(M_n)(s)| 
 &=
-\int_{S} \mu(\dif s) \, f(M_n)(s)
+\int_{\entrosp} \mu_{\entrosp}(\dif s) \, f(M_n)(s)
 \\
 &\leq
-\int_{S} \mu(\dif s) \, f(M_0)(s)
+\int_{\entrosp} \mu_{\entrosp}(\dif s) \, f(M_0)(s)
 =
@@
 Let $f$ be a ranking function on $M$ satisfying $f(N) = 0$ iff $N$ is a value.
-Define random variables on the probability space $(\entrosp, \calF, \mu)$:
+Define random variables on the probability space $(\entrosp, \calF, \mu_{\entrosp})$:
 \begin{align}

### 2020-12-28 LICS21/antitone.tex
Minor

@@
 \end{align*}
-\changed[lo]{(The first inequality follows from the fact that for all $x \in (0, 1)$, $\log (1 + x) > x$.)}
+\changed[lo]{(The first inequality follows from the fact that for all $x \in (0, 1)$, $\log_2 (1 + x) > x$.)}
 Moreover $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
@@
 For $n > 2$, $z_n = \log(n-1)$ works.
-The expected decrease is $\frac 1 2(\log(1+\frac 1 {(n-1)^2-1}) - \frac 1 {2n+1} \log(1 + \frac 2 {n-2}))$, and using the fact that $\frac x {x+1} \leq \log (1+x) \leq x$ for $x > 0$,\lo{The preceding inequality is incorrect: for $x \in (0, 1)$, $\log (1 + x) > x$.} this is at least $\frac 1 2(\frac 1 {(n-1)^2} - \frac 1 {2n + 1} \frac 2 {n-2}) = \frac {n^2}{2(n-1)^2(2n+1)(n-2)}$, which (again for $n > 2$) is positive and antitone.
+The expected decrease is $\frac 1 2(\log(1+\frac 1 {(n-1)^2-1}) - \frac 1 {2n+1} \log(1 + \frac 2 {n-2}))$, and using the fact that $\frac x {x+1} \leq \log (1+x) \leq x$ for $x > 0$,\lo{The preceding inequality is incorrect: for $x \in (0, 1)$, $\log (1 + x) > x$.}
+\lo{OK.
+You mean $\ln$ rather than $\log_2$.} this is at least $\frac 1 2(\frac 1 {(n-1)^2} - \frac 1 {2n + 1} \frac 2 {n-2}) = \frac {n^2}{2(n-1)^2(2n+1)(n-2)}$, which (again for $n > 2$) is positive and antitone.
 For $n = 0, 1, 2$, we take $\epsilon$ to be 9/40 (the same as its value at 3), then set $z_2 = 2 \log 2 - \log 3 - 1, z_1 = 3 \log 2 - 2 \log 3 - 3, z_0 = 4 \log 2 - 3 \log 3 - 6$.

### 2020-12-28 LICS21/confluent.tex
Mainly LaTeX reformatting to fit 2-column style

@@
+% !TEX root = main.tex
 \section{Confluent trace semantics}
@@
 
-A \emph{position} 
-%\akr{I'm not really sure how to phrase the citation, but this is adapted from \cite{Kennaway96infinitarylambda}.} 
-%\lo{The idea is common enough. I think no citation necessary.}
-is a finite sequence of steps into a term, defined inductively as
+A \emph{position} is a finite sequence of steps into a term, defined inductively as
 \begin{align*}
-\alpha ::= \cdot \mid \lambda ; \alpha \mid @_1 ; \alpha \mid @_2 ; \alpha \mid \underline f_i ; \alpha \mid \tY ; \alpha \mid \textsf{if}_1 ; \alpha \mid \textsf{if}_2 ; \alpha \mid \textsf{if}_3 ; \alpha \mid \tscore ; \alpha.
+\alpha ::= \cdot \mid \lambda ; \alpha \mid @_1 ; \alpha \mid @_2 ; \alpha \mid \underline f_i ; \alpha \\
+\mid \tY ; \alpha \mid \textsf{if}_1 ; \alpha \mid \textsf{if}_2 ; \alpha \mid \textsf{if}_3 ; \alpha \mid \tscore ; \alpha.
 \end{align*}
@@
   \text{if } M \mid \alpha = \underline f (\underline r_1, \dots , \underline r_n),\ & M \to M[\underline{f(r_1,\dots,r_n)}/\alpha] \\
-  \text{if } M \mid \alpha = \tY \lambda x. N,\ & M \to M[\lambda z. N[(\tY \lambda x. N)/x] z/\alpha] \text{ where $z$ is not free in $N$}\\
+  \text{if } M \mid \alpha = \tY \lambda x. N,\ & M \to M[\lambda z. N[(\tY \lambda x. N)/x] z/\alpha]\\ \text{where }z&\text{ is not free in $N$}\\
   \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_1/\alpha] \text{ where }r < 0 \\
   \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_2/\alpha] \text{ where }r \geq 0 \\
-  \text{if } M \mid \alpha = \tsample & \text{ and $\lambda$ does not occur after $@_2$ or $\tY$ in $\alpha$},\\ & M \to M[\underline r/\alpha] \text{ where } r \in [0,1] \\
+  \text{if } M \mid \alpha = \tsample \text{ and $\lambda$ does}&\text{ not occur after $@_2$ or $\tY$ in $\alpha$},\\ & M \to M[\underline r/\alpha] \text{ where } r \in [0,1] \\
   \text{if } M \mid \alpha = \tscore(\underline r),\ & M \to M[\underline r/\alpha].
@@
 The $\sim_\uparrow$ may be duplicated if it passes a $\sim_\downarrow$ whose position is outside of the $\sim_\uparrow$'s position, but by the assumption that the sequence of $\sim_\downarrow$s is initially parallel, no position inside of this occurs later in the sequence, therefore all of the resultant $\sim_\uparrow$s pass the remaining $\sim_\downarrow$s without changing or duplicating them.
-The $\sim_\uparrow$s may be further duplicated, but the number of steps left in this process is bounded by the product, for all of the remaining $\sim_\downarrow$s, of $2 + \text{the number of occurrences of the variable relevant to the reduction}$, because for each $\sim_\downarrow$, that is more than the number of duplicates it could produce for any $\sim_\uparrow$ that passes it.
+The $\sim_\uparrow$s may be further duplicated, but the number of steps left in this process is bounded by the product, for all of the remaining $\sim_\downarrow$s, of $2 +$ the number of occurrences of the variable relevant to the reduction, because for each $\sim_\downarrow$, that is more than the number of duplicates it could produce for any $\sim_\uparrow$ that passes it.
 It is also possible (earlier) for some of the $\sim_\downarrow$s to be duplicated, but the only case where this process would not terminate, that both the $\sim_\downarrow$s and the $\sim_\uparrow$s continue being duplicated forever so that the number of switches left to do never decreases, is prevented by the parallelness condition, as all of the duplications of $\sim_\downarrow$s occur before all of the duplications of $\sim_\uparrow$s.
@@
 \begin{align*}
-& (M,s) \Rightarrow (N,s \circ i(M \to N)) \text{ if either} \\
-& \qquad \text{$M \to N$ and the redex is not $\tsample$, or} \\
-& \qquad \text{$M \mid \alpha = \tsample$, $N = M[\underline{s(Sk(M),\alpha)}/\alpha]$ and $\lambda$ does not occur after $@_2$ or $\tY$ in $\alpha$}
+& (M,s) \Rightarrow (N,s \circ i(M \to N)) \text{ if $M \to N$ at $\alpha$ and either} \\
+& \qquad \text{the redex is not $\tsample$, or} \\
+& \qquad \text{$M \mid \alpha = \tsample$ and $N = M[\underline{s(Sk(M),\alpha)}/\alpha]$}
 \end{align*}
@@
 Let\begin{align*}
-R_s & = \{\text{the skeleton of } O \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) = \tsample, O \to N[r]\} \\
-R_d & = \{\text{the skeleton of } O \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) \neq \tsample, O \to N[r]\}.
+R_s & = \{Sk(O) \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) = \tsample, O \to N[r]\} \\
+R_d & = \{Sk(O) \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) \neq \tsample, O \to N[r]\}.
 \end{align*}
@@
 In the case that $O \in R_s$, the function $g_O$ simply inserts the sample into the vector of reals at a certain index, call it $h_O$.
+\clearpage
 \begin{align*}
 & \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')\}) \\
-= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in S :\\
-    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\})\\
-    + & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in S :\\
-    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \\
-\end{align*}
-\begin{align*}
+= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in S : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\})\\
+    + & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in S : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \\
 = & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}:\\
@@
     + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
-= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} :\\
-    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S)\\
-    + & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} :\\
-    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S) \\
+= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S)\\
+    + & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S) \\
 = & \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in I^{L_s(N)} : (M,s) \Rightarrow_f^n (N[r], s')\}) \mu(S)
@@
 \end{proof}
+\clearpage
 
@@
 The domain of definition of the ranking functions also needs to be changed from the reachable terms $Rch(M)$ to the \emph{reachable terms with respect to the reduction strategy $r$},
-\[
-Rch_r(M) = \{N \mid \exists n, N_0, \dots, N_n: N_0 = M, N_n = N, N_i \to N_{i+1} \text{ at } r(N_i)\}.
-\]
+\begin{align*}
+& Rch_r(M) = \\ & \{N \mid \exists n, (N_i)_{0 \leq i \leq n}: N_0 = M, N_n = N, N_i \to N_{i+1} \text{ at } r(N_i)\}.
+\end{align*}
 

### 2020-12-28 LICS21/SPCF.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper

@@
 \end{align*}
-The typing rules are presented in \Cref{fig:typing rules}.
+The typing rules are standard (see \Cref{fig:typing rules}).
 Terms are identified up to $\alpha$-equivalence, as usual. 
@@
 then a term reduces if it is formed by substituting a redex in a context i.e.
-
-\begin{figure*}
-\begin{align*}
-  \frac{}{\Gamma ; x:A \vdash x:A} 
-  \qquad
-  \frac{\Gamma ; x:A \vdash M : B}{\Gamma \vdash \lambda x.M : A \to B} 
-  \qquad
-  \frac{}{\underline{r} : \textsf{R}} 
-  \qquad
-  \frac{\Gamma \vdash M:A \to B \quad \Gamma \vdash N : A}{\Gamma \vdash M \, N : B} 
-  \\ \\
-  \qquad
-  \frac{\Gamma \vdash M : (A \to B) \to (A \to B)}{\Gamma \vdash \tY M : (A \to B)}
-  \qquad
-  \frac{\Gamma \vdash M_1:\textsf{R} \quad \dots \quad \Gamma \vdash M_n:\textsf{R}}{\Gamma \vdash \underline{f}(M_1,\dots,M_n) : \textsf{R}} \ (f : \mathbb{R}^n \to \mathbb{R})    
-  \\ \\
-  \qquad
-  \frac{\Gamma \vdash M : \textsf{R} \quad \Gamma \vdash N_1 : A \quad \Gamma \vdash N_2 : A}{\Gamma \vdash \tif{M < 0}{N_1}{N_2} : A}
-  \qquad
-  \frac{}{\Gamma \vdash \tsample : \textsf{R}} 
-  \qquad
-  \frac{\Gamma \vdash s : \textsf{R}}{\Gamma \vdash \tscore (s) : \textsf{R}}
-\end{align*}
-\caption{Typing rules of SPCF \label{fig:typing rules}}
-\end{figure*}
-
 \begin{align*}
@@
 To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples.
-Let $ I = [0,1] \subset \mathbb{R} $, and let $\entrosp = I^{\mathbb{N}}$, with the \changed[lo]{Borel $\sigma$-algebra $\calF$} and the probability measure, denoted $\mu_{\entrosp}$, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
-Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and $\mu_{\entrosp} (\prod_{i=0}^\infty X_i) = \prod_{i=0}^\infty \Leb(X_i)$.
+Let $ I := [0,1] \subset \mathbb{R} $, and define $\entrosp := I^{\mathbb{N}}$, with the \changed[lo]{Borel $\sigma$-algebra $\calF$} and the probability measure, denoted $\mu_{\entrosp}$, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
+Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and $\mu_{\entrosp} (\prod_{i=0}^\infty X_i) := \prod_{i=0}^\infty \Leb(X_i)$.
 The maps $\pi_h:\entrosp \to I, \; \pi_t:\entrosp \to \entrosp$ popping the first element are then measurable.
@@
 \begin{equation*}
-\red(M,s) = \left\{
+\red(M,s) := \left\{
     \begin{array}{ll}

### 2020-12-28 LICS21/supermartingales-proofs.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper Moved proof of Lemma \ref{lem:rank-Past} to appendix.

--- 2020-12-27 LICS21/apx-y-stopping-times.tex
@@
 % !TEX root=./main.tex
+
+\rankPast*
+\begin{proof}%\rankPast
+We first prove (by induction on $n$): for all $n \geq 0$
+\[
+%\forall n \geq 0 \, . \, 
+\expect{Y_n} \leq \expect{Y_0} - \epsilon \cdot \big(\textstyle\sum_{i=0}^{n-1} \mathbb{P}[T > i]\big).
+\]
+It follows that $\sum_{i=0}^{\infty} \mathbb{P}[T > i]$ converges; hence we have $\lim_{i \to \infty} \mathbb{P}[T > i] = 0$ and so $\mathbb{P}[T < \infty] = 1$.
+It then remains to observe: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=0} \mathbb{P}[T > i]$.
+\end{proof}
 

### 2020-12-28 LICS21/supermartingales.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper Moved proof of Lemma \ref{lem:rank-Past} to appendix.

@@
 
-\changed[lo]{Fix a probability space $(\Omega, \calF, \mathbb{P})$, and a filtration $(\calF_n)_{n \geq 0}$. 
+\changed[lo]{Fix a probability space $(\Omega, \calF, \mathbb{P})$ and a filtration $(\calF_n)_{n \geq 0}$. 
 (In subsequent applications to termination analysis, $\Omega = \entrosp$, and $\mathbb{P} = \mu_{\entrosp}$.)
@@
 we say that $(Y_n)_{n \geq 0}$ is a \emph{$\epsilon$-ranking supermartingale w.r.t.~$T$} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.
-{(For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable defined by ${\bf 1}_A(\omega) := 1$ if $\omega \in A$, and $0$ otherwise.)} \lo{\LaTeX\ does not allow the preceding bracketed sentence to be typeset as a footnote.
-Strange!}
+\footnote{(For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable defined by ${\bf 1}_A(\omega) := 1$ if $\omega \in A$, and $0$ otherwise.)} %\lo{\LaTeX\ does not allow the preceding bracketed sentence to be typeset as a footnote.
+Strange!}\lo{Now it does.
+Stranger.}
 A \emph{ranking supermartingale} w.r.t.~$T$ is a $\epsilon$-ranking supermartingale w.r.t.~$T$ for some $\epsilon > 0$.
@@
 
-\begin{lemma}%[{\citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}}]
+\begin{restatable}{lemma}{rankPast}
+%\begin{lemma}%[{\citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}}]
 % \citep[Prop 1]{DBLP:conf/popl/ChatterjeeFNH16}}
@@
 Let $(Y_n)_{n \geq 0}$ be a $\epsilon$-ranking supermartingale w.r.t.~the stopping time $T$, then $T < \infty$ a.s., and $\expect{T} \leq \frac{\expect{Y_0}}{\epsilon}$.
-\end{lemma}
-
-\begin{proof}
-We first prove (by induction on $n$): for all $n \geq 0$
-\[
-%\forall n \geq 0 \, . \, 
-\expect{Y_n} \leq \expect{Y_0} - \epsilon \cdot \big(\textstyle\sum_{i=0}^{n-1} \mathbb{P}[T > i]\big).
-\]
-It follows that $\sum_{i=0}^{\infty} \mathbb{P}[T > i]$ converges; hence we have $\lim_{i \to \infty} \mathbb{P}[T > i] = 0$ and so $\mathbb{P}[T < \infty] = 1$.
-It then remains to observe: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=0} \mathbb{P}[T > i]$.
-\end{proof}
+%\end{lemma}
+\end{restatable}
 

### 2020-12-29 LICS21/antitone.tex
Motivation and examples for non-affine recursion, higher-order recursion, and programs with continuous distributions.

@@
 We say that a supermartingale $(Y_n)_{n \geq 0}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \geq 0}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
-\[
+\(
 \expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot {\bf 1}_{\set{T > n}}
-\]
+\).
 \end{definition}
@@
 \label{thm:a-c strict}
-Let $(Y_n)_{n \geq 0}$ be an antitone strict supermartingale via function $\epsilon : \nnReal \to \pReal$, and stopping time $T$. 
-Then $T < \infty$ almost surely.
+Let $(Y_n)_{n \geq 0}$ be an antitone strict supermartingale via $\epsilon : \nnReal \to \pReal$ and stopping time $T$. 
+Then $T < \infty$ a.s.
 \end{theorem}
@@
 \Cref{thm:rankable implies termination},
-take the set $\entrosp$ of traces \lo{TODO: We should find a better name for traces $s \in \entrosp$.} as a probability space, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$.
+take the probability space $(\entrosp, \calF, \mu_\entrosp)$, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$. 
 As before, $(Y_n)_{n \geq 0}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$.
-Therefore, it follows from \Cref{thm:a-c strict} that $T_M^{\tY} < \infty$ a.s.
+Thus, by \Cref{thm:a-c strict}, $T_M^{\tY} < \infty$ a.s.
 \end{proof}
@@
 
+\begin{example}[Unbiased random walk]
+\label{ex:unbiased random walk}
 For $M_2$, define the function $g_1 : \nnReal \to \nnReal$ by $g_1(x) = \log(x+1) + 1$.
@@
 %with antitone function $\epsilon_2$ defined as follows (for $n \in \mathbb N$):
-defined as: for $n \in \mathbb N$
-\begin{align*}
-{\Theta_2} \, n 
-&\mapsto 
-g_2(n)
-\\
-0 &\mapsto 0.
-\end{align*}
+by
+\[
+f: {\Theta_2} \, n 
+\mapsto 
+g_2(n), \qquad 
+0 \mapsto 0
+\]
+for $n \in \mathbb N$.
+% \begin{align*}
+% {\Theta_2} \, n 
+% &\mapsto 
+% g_2(n)
+% \\
+% 0 &\mapsto 0.
+% \end{align*}
 For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$, each with probability $1/2$, with one $\tY$-reduction.
@@
 It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone partial ranking function, and $M_2$ is AST.
+\end{example}
 
 \begin{example}[Continuous random walk]\label{ex:raven complex}
-In SPCF we can construct a function whose argument changes by a random amount at each recursive call:
+In SPCF we can construct a function whose argument changes by a random amount at each recursive call: $\Theta \, 10$ where
 \[
-\underbrace{\big
-(\tY \, \lambda f x . \tif{x \leq 0}{0}{f(x - \tsample)} \big)}_{\Theta} \, 10
+%\underbrace{\big
+\Theta := \tY \, \lambda f x . \tif{x \leq 0}{0}{f(x - \tsample)}
+%}_{\Theta} \, 10
 \]
@@
 
-For a more complex (not $\tY$-PAST) example, consider the following ``continuous random walk''
+For a more complex (not $\tY$-PAST) example, consider the following ``continuous random walk'': $\Xi \, 10$ where
 \[
-\underbrace{\big
-(\tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample + 1/2)} \big)}_{\Xi} \, 10
+%\underbrace{\big(
+\Xi := \tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample + 1/2)} 
+%\big)}_{\Xi} \, 10
 \]
@@
 therefore $\epsilon$ is antitone and bounded below by 0.
-We define a partial antitone ranking function as follows:
-\begin{align*}
+We define a partial antitone ranking function by:
+\[
 \Xi \, l 
-&\mapsto 
-g(l)
-\\
-0 &\mapsto 0
-\end{align*}
-\end{example}
+\mapsto 
+g(l), \qquad 
+0 \mapsto 0
+\]
+
+% \begin{align*}
+% \Xi \, l 
+% &\mapsto 
+% g(l)
+% \\
+% 0 &\mapsto 0
+% \end{align*}
 The value of $g$ after one $\tY$-reduction step is at least $g(l-1/2)$, therefore the expectation of $\epsilon$ after one $\tY$-reduction step is at most $\epsilon(g(l-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y$.
@@
 Hence this is a valid partial antitone ranking function and the term is AST.
+\end{example}
 
@@
 z_0 &= 0\\
-n \geq 1, \quad z_n &> \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
+n \geq 1, \quad z_n &> \textstyle \frac{n}{2n + 1} \, z_{n-1} + \frac{n+1}{2n + 1} \, z_{n+1}.
 \end{align*}
@@
 \begin{align*}
-    & \expect{Y_{n+1} \mid \mathcal F_n} \\
+    \expect{Y_{n+1} \mid \mathcal F_n} 
     &=  \expect{y_{n+1} \, {\bf 1}_{\set{T \geq n+1}} \mid \mathcal F_n} \\
     &=  y_{n+1} \, \expect{{\bf 1}_{\set{T \geq n+1}} \mid \mathcal F_n} \\ 
-    &=  y_{n+1} \, {\bf 1}_{\set{T \geq n}} \frac{\mathbb P[T > n+1]}{\mathbb P[T > n]} \\
+    &=  y_{n+1} \, {\bf 1}_{\set{T \geq n}} \textstyle \frac{\mathbb P[T > n+1]}{\mathbb P[T > n]} \\
     &=  (y_n - \epsilon(y_n)) \, {\bf 1}_{\set{T \geq n}} \\
@@
 This condition is the main reason that this proof does not extend directly to a completeness result for antitone ranking functions.
+
+\begin{example}[Non-affine recursion]
+\changed[akr]{Let 
+\[
+\Xi := \tY \lambda f x. x \oplus_{2/3} f (f (x + 1))
+\] 
+A partial ranking function for $\Xi \, 1$ would be $\Xi^n \, i \mapsto 3 \, n$ (for $n \geq 0$), but w.r.t.~the reduction strategy that first reduces the argument of an application to a value, then the function, then the application itself.
+
+For other values of $p > 1/2$, $n/(2p-1)$ works, and for $p = 1/2$, something much like the antitone ranking function for \Cref{ex:unbiased random walk} would work.
+
+A standard partial ranking function (i.e.~one that doesn't depend on the confluent semantics) would have a slightly more verbose definition, as the functions that accumulate would not be $\Xi$, but what $\Xi$ reduces to after one step.}
+
+\lo{This is a genuine (and simple) non-affine recursive program. 
+We can use this example to motivate the confluent semantics, though it is perhaps not the most compelling possible.
+Reason: one could use a more restrictive but equally expressive recursion \emph{operator} (as opposed to a $\tY$-\emph{combinator}, which has the status of a constant) to implement general recursion which would make it unnecessary to consider an alternative reduction strategy when applying the partial ranking function theorem.
+E.g.~define letrec-terms, $\mathsf{letrec} \, f^{A \to B} \, x^A . M$, as values, via the formation rule
+\[{\Gamma, f: A \to B, x:A  \vdash M : B} \over
+\Gamma \vdash \mathsf{letrec} \, f^{A \to B} \, x^A . M : A \to B
+\]
+with the redex rule 
+\[
+(\mathsf{letrec} \, f^{A \to B} \, x^A . M^B) \, V \to M[\mathsf{letrec} \, f \, x . M / f, V / x].
+\]}
+\end{example}
+
+\lo{
+Motivation: Non-affine recursion. This could be placed in \Cref{sec:intro}.
+}
+Many of the recent advances in the development of AST verification methods \citep{DBLP:conf/pldi/ChenH20,DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18,DBLP:conf/aplas/HuangFC18,DBLP:conf/popl/ChatterjeeNZ17,DBLP:journals/pacmpl/AgrawalC018,DBLP:conf/cav/ChatterjeeFG16,DBLP:conf/lics/OlmedoKKM16,DBLP:journals/pacmpl/Huang0CG19} are concerned with loop-based programs.
+We can view such loops as tail-recursive programs that are, in particular, \emph{affine recursive}, i.e., 
+in each evaluation (or run) of the body of the recursion, recursive calls are made from at most one call site \citep[\S 4.1]{DBLP:journals/toplas/LagoG19}.
+Termination analysis of \emph{non-affine recursive} probabilistic programs does not seem to have received much attention.
+Methods such as those presented in \citep{DBLP:journals/toplas/LagoG19} are explicitly restricted to affine programs, and are unsound otherwise.
+By contrast, many probabilistic programming languages allow for richer recursive structures \cite{DBLP:conf/pkdd/TolpinMW15,DBLP:conf/uai/GoodmanMRBT08,DBLP:journals/corr/MansinghkaSP14}.
+
+\medskip
+
+\lo{
+DRAFT. Motivation: Continuous distributions. This could be placed in \Cref{sec:intro}.
+}
+Sampling from continuous distributions is an essential feature of probabilistic programming languages.
+(See e.g.~Church \citep{DBLP:conf/uai/GoodmanMRBT08}, Stan \citep{carpenter2017stan}, Anglican \citep{DBLP:conf/pkdd/TolpinMW15}, Gen \citep{cusumano-towner2019Gen}, Pyro \citep{bingham2019Pyro}, Edward \citep{tran2016edward} and Turing \citep{ge2018Turing}.)
+To our knowledge, methods of proving AST of probabilistic computation have been developed for probabilistic programs with discrete districutions only.
+The problem of proving AST of programs with continuous distribution is new.
+
+\lo{The following examples are for our discussion at 5:30. Details to be checked.}
+
+\begin{example}[Non-affine recursion with continuous distribution]
+\label{ex:non-affine with continuous distribution}
+Consider 
+\[
+\Xi := \tY \lambda f x . \big(\mathsf{let} \; e = \tsample \; \mathsf{in} \;
+\mathsf{if}\boldsymbol{(}e \leq p, x, X\boldsymbol{)}\big)
+\]
+where
+\(
+X = \big( f^3(x+1) \oplus_e f^2(x+1) \big) \oplus_{\mathit{sig}(x)} f^2(x+1) 
+\)
+and $\mathit{sig}(x)$ is the Sigmoid function.
+
+Not only is this program non-affine recursive,
+note also the use of a random sample, $e$, as a first-class value, and the subsequent use as a probability in the binary choice $\oplus_e$.
+Such a computation cannot be modelled via discrete distributions.
+
+We can use the ranking function method (coupled with the solution of linear recurrence relations) to show that provided $p \geq 2 - \sqrt{7}$, the program $\Xi \, \underline r$ is AST for all $r \in \Real$.
+
+\lo{This example is taken from the series of 3d-printing-company running examples (Beutner-Ong 2021).
+This paper is as yet unpublished, and effectively uncitable for the time being (because of strict double-blind submission rules).}
+\end{example}
+
+\lo{I believe the ranking function method can be used to prove soundness of the Beutner-Ong counting-based method (which reduces the AST problem of first-order (possibly non-affine) recursive SPCF programs to random walk on $\mathbb{N}$), based on Theorem 5.9 and Lemma 5.10 in \emph{op.~cit.}}
+
+\iffalse
+\[\begin{array}{l}
+\mathsf{let} \; \mathit{add} \; x \; y = x + y \\
+\mathsf{letrec} \; \mathit{iter} \; f \; s \; n = 
+\tif{n \leq 0}{s}{f \; n \; (\mathit{iter} \; f \; s \; (n - 1)}
+\end{array}
+\]
+\fi
+
+\lo{@Andrew: By the way, we should use the name \emph{sparse ranking function} (theorem / method).}
+
+\lo{
+DRAFT. Motivation: higher-order recursion. This could be placed in \Cref{sec:intro} or \Cref{sec:related}.
+}
+There is an obvious source of deterministic (i.e.~non-probabilistic) higher-order functions defined by recursion, viz., \emph{higher-order recursion schemes} (HORS) (see e.g.~\citep{DBLP:conf/lics/Ong06,DBLP:conf/lics/Ong15}).
+(Incidentally the (sparse) ranking function method is just as applicable to the termination analysis of \emph{deterministic} SPCF programs.)
+Recently \cite{DBLP:conf/lics/KobayashiLG19} have extended HORS to \emph{probabilistic higher-order recursion schemes} (PHORS), which are HORS augmented with probabilistic (binary) branching $\oplus_p$.
+As HORS are in essence the $\lambda \tY$-calculus \cite{DBLP:conf/lics/Statman02} (i.e.~pure simply-typed lambda calculus with recursion, generated from a finite base type), PHORS are definable in SPCF:
+(order-$n$) PHORS is encodable as (order-$n$) CbN SPCF, but the former is strictly less expressive (because the underlying recursion schemes are not Turing complete). 
+Some interesting SPCF terms such as \Cref{ex:non-affine with continuous distribution} %\refExample{complexExample} 
+cannot be expressed as PHORS.
+A relevant result here is that the AST problem is decidable for order-1 PHORS (by reduction to the PSPACE-hard solvability of finite systems of polynomial equations with real coefficients; see \citep{DBLP:journals/jacm/EtessamiY09}), but undecidable for order-2 PHORS.
+
+\begin{example}[Higher-order recursion]
+Consider the higher-order function
+$\Xi : (\textsf{R} \to \textsf{R} \to \textsf{R}) \to \textsf{R} \to \textsf{R} \to \textsf{R}$
+recursively defined\footnote{Inspired by examples in \citep{DBLP:journals/pacmpl/BurnOR18,DBLP:conf/lics/OngW19}.} by
+\[
+\begin{array}{l}
+\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} . \\
+\quad\tif{n \leq 0}{s}{
+f \; n \; (\varphi \; f \; s \; (n - 1))
+\oplus_p
+f \; (n+1) \; (\varphi \; f \; s \; n)
+}
+\end{array}
+\]
+Let $g : \textsf{R} \to \textsf{R} \to \textsf{R}$ be a function that terminates on all arguments.
+Then $\Xi \, g$ is AST for $p \geq 1/2$.
+\end{example}

### 2020-12-29 LICS21/supermartingales.tex
Motivation and examples for non-affine recursion, higher-order recursion, and programs with continuous distributions.

@@
 we say that $(Y_n)_{n \geq 0}$ is a \emph{$\epsilon$-ranking supermartingale w.r.t.~$T$} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.
-\footnote{(For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable defined by ${\bf 1}_A(\omega) := 1$ if $\omega \in A$, and $0$ otherwise.)} %\lo{\LaTeX\ does not allow the preceding bracketed sentence to be typeset as a footnote.
+\footnote{For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable defined by ${\bf 1}_A(\omega) := 1$ if $\omega \in A$, and $0$ otherwise.} %\lo{\LaTeX\ does not allow the preceding bracketed sentence to be typeset as a footnote.
 Strange!}\lo{Now it does.
@@
 and the filtration $(\mathcal{F}_n)_{n \geq 0}$ where $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
-Thus $T_M$ is the \emph{termination time} of $M$.}
+Thus $T_M$ is the \emph{runtime} of $M$ (and $M$ is AST iff $\mu_{\entrosp}(T_M < \infty) = 1$).}
 
@@
 \end{theorem}
+\lo{A better name for this theorem: ``Ranking function'', ``Supermartingale'', ``(Ranking) supermartingale''?}
 
@@
 Let $f$ be a ranking function on $M$.
-For 1, since $(T_n)_{n \geq 0}$ is an increasing sequence of stopping times, each is bounded and adapted to $(\calF_n)_{n \geq 0}$ (\Cref{lem:TnBounded}),
+For 1, since $(T_n)_{n \geq 0}$ is an increasing sequence of stopping times, each is adapted to $(\calF_n)_{n \geq 0}$ (\Cref{lem:TnBounded}) and bounded,
 and $(f(M_n))_{n \geq 0}$ is a supermartingale also adapted to $(\calF_n)_{n \geq 0}$ (\Cref{thm:rankable and strict rankable}),
-it follows from the Optional Sampling \Cref{thm:optional sampling} that $(Y_n)_{n \geq 0}$ where $Y_n := f(M_{T_n})$ is a supermartingale adapted to the filtration $(\calF_{T_n})_{n \geq 0}$.
+it follows from the Optional Sampling \Cref{thm:optional sampling} that $(Y_n)_{n \geq 0}$ %where $Y_n := f(M_{T_n})$ 
+is a supermartingale adapted to $(\calF_{T_n})_{n \geq 0}$.
 Notice the stopping time $T_M^\tY$ is also adapted to $(\calF_{T_n})_{n \geq 0}$ (\Cref{lem:TMtY is a stopping time}); and we have that $(Y_n)_{n \geq 0}$ is a $1$-ranking supermartingale.
@@
 %It follows immediately from \Cref{lem:rank-PAST} that $\expect{T_M} < \infty$.
-Statement 2 follows immediately from \Cref{thm:rankable and strict rankable} and \Cref{lem:rank-PAST}.
+Statement 2 follows at once from \Cref{thm:rankable and strict rankable} and \Cref{lem:rank-PAST}.
 \end{proof}

### 2020-12-29 LICS21/antitone.tex
Minor: references

@@
 
+\medskip
+
+\lo{
+Motivation: Non-affine recursion. This could be placed in \Cref{sec:intro}.
+}
+Many of the recent advances in the development of AST verification methods \citep{DBLP:conf/pldi/ChenH20,DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18,DBLP:conf/aplas/HuangFC18,DBLP:conf/popl/ChatterjeeNZ17,DBLP:journals/pacmpl/AgrawalC018,DBLP:conf/cav/ChatterjeeFG16,DBLP:conf/lics/OlmedoKKM16,DBLP:journals/pacmpl/Huang0CG19} are concerned with loop-based programs.
+We can view such loops as tail-recursive programs that are, in particular, \emph{affine recursive}, i.e., 
+in each evaluation (or run) of the body of the recursion, recursive calls are made from at most one call site \citep[\S 4.1]{DBLP:journals/toplas/LagoG19}.
+(Note that whether a program is affine recursive cannot be checked by just counting textual occurrences of variables.)
+Termination analysis of \emph{non-affine recursive} probabilistic programs does not seem to have received much attention.
+Methods such as those presented in \citep{DBLP:journals/toplas/LagoG19} are explicitly restricted to affine programs, and are unsound otherwise.
+By contrast, many probabilistic programming languages allow for richer recursive structures \citep{DBLP:conf/pkdd/TolpinMW15,DBLP:conf/uai/GoodmanMRBT08,DBLP:journals/corr/MansinghkaSP14}.
+
+\medskip
+
 \begin{example}[Non-affine recursion]
@@
 
-\lo{This is a genuine (and simple) non-affine recursive program. 
+\lo{This is a genuine (and simple) \emph{non-affine recursive program}, i.e., recursive programs that can, during the evaluation of the recursive body, make multiple recursive
+calls (of a first-order function) from distinct call sites. 
 We can use this example to motivate the confluent semantics, though it is perhaps not the most compelling possible.
@@
 \lo{
-Motivation: Non-affine recursion. This could be placed in \Cref{sec:intro}.
-}
-Many of the recent advances in the development of AST verification methods \citep{DBLP:conf/pldi/ChenH20,DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18,DBLP:conf/aplas/HuangFC18,DBLP:conf/popl/ChatterjeeNZ17,DBLP:journals/pacmpl/AgrawalC018,DBLP:conf/cav/ChatterjeeFG16,DBLP:conf/lics/OlmedoKKM16,DBLP:journals/pacmpl/Huang0CG19} are concerned with loop-based programs.
-We can view such loops as tail-recursive programs that are, in particular, \emph{affine recursive}, i.e., 
-in each evaluation (or run) of the body of the recursion, recursive calls are made from at most one call site \citep[\S 4.1]{DBLP:journals/toplas/LagoG19}.
-Termination analysis of \emph{non-affine recursive} probabilistic programs does not seem to have received much attention.
-Methods such as those presented in \citep{DBLP:journals/toplas/LagoG19} are explicitly restricted to affine programs, and are unsound otherwise.
-By contrast, many probabilistic programming languages allow for richer recursive structures \cite{DBLP:conf/pkdd/TolpinMW15,DBLP:conf/uai/GoodmanMRBT08,DBLP:journals/corr/MansinghkaSP14}.
-
-\medskip
-
-\lo{
 DRAFT. Motivation: Continuous distributions. This could be placed in \Cref{sec:intro}.
@@
 (See e.g.~Church \citep{DBLP:conf/uai/GoodmanMRBT08}, Stan \citep{carpenter2017stan}, Anglican \citep{DBLP:conf/pkdd/TolpinMW15}, Gen \citep{cusumano-towner2019Gen}, Pyro \citep{bingham2019Pyro}, Edward \citep{tran2016edward} and Turing \citep{ge2018Turing}.)
-To our knowledge, methods of proving AST of probabilistic computation have been developed for probabilistic programs with discrete districutions only.
-The problem of proving AST of programs with continuous distribution is new.
+Methods of proving AST of probabilistic computation have been developed for probabilistic programs with discrete districutions (see e.g.~\citep{DBLP:journals/toplas/LagoG19,DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/lics/KobayashiLG19,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05}).
+To our knowledge, the problem of proving AST of programs with continuous distribution is new.
 
-\lo{The following examples are for our discussion at 5:30. Details to be checked.}
+\lo{The following examples are for our discussion at 5:30. Details yet to be checked.}
 

### 2020-12-29 LICS21/antitone.tex
Higher-order recursion example: question
Added apx-typing-rules.tex

@@
 }
-Sampling from continuous distributions is an essential feature of probabilistic programming languages.
+Sampling from continuous distributions is an essential feature of statistical probabilistic programming languages.
 (See e.g.~Church \citep{DBLP:conf/uai/GoodmanMRBT08}, Stan \citep{carpenter2017stan}, Anglican \citep{DBLP:conf/pkdd/TolpinMW15}, Gen \citep{cusumano-towner2019Gen}, Pyro \citep{bingham2019Pyro}, Edward \citep{tran2016edward} and Turing \citep{ge2018Turing}.)
-Methods of proving AST of probabilistic computation have been developed for probabilistic programs with discrete districutions (see e.g.~\citep{DBLP:journals/toplas/LagoG19,DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/lics/KobayashiLG19,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05}).
-To our knowledge, the problem of proving AST of programs with continuous distribution is new.
+Methods of proving AST of probabilistic computation have been developed for probabilistic programs with discrete distributions (see e.g.~\citep{DBLP:journals/toplas/LagoG19,DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/lics/KobayashiLG19,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05}).
+To our knowledge, the problem of proving AST of probabilistic programs with continuous distribution is new.
 
@@
 \]
-Let $g : \textsf{R} \to \textsf{R} \to \textsf{R}$ be a function that terminates on all arguments.
-Then $\Xi \, g$ is AST for $p \geq 1/2$.
+\lo{QUESTION: Let $g : \textsf{R} \to \textsf{R} \to \textsf{R}$ be a function that terminates on all arguments.
+Then $\Xi \, g$ is AST for $p \geq 1/2$.}
 \end{example}

### 2020-12-29 LICS21/antitone.tex
Sec. 3: Supermartingales - cutting down to 2- pp.

@@
 calls (of a first-order function) from distinct call sites. 
-We can use this example to motivate the confluent semantics, though it is perhaps not the most compelling possible.
-Reason: one could use a more restrictive but equally expressive recursion \emph{operator} (as opposed to a $\tY$-\emph{combinator}, which has the status of a constant) to implement general recursion which would make it unnecessary to consider an alternative reduction strategy when applying the partial ranking function theorem.
+We can use this example to motivate the confluent trace semantics, though it is perhaps not the most compelling possible.
+Reason: one could use a more restrictive but equally expressive recursion \emph{operator} (as opposed to a $\tY$-\emph{combinator}, which is a constant) to implement general recursion which would make it unnecessary to consider an alternative reduction strategy when applying the partial ranking function theorem.
 E.g.~define letrec-terms, $\mathsf{letrec} \, f^{A \to B} \, x^A . M$, as values, via the formation rule
@@
 \)
-and $\mathit{sig}(x)$ is the Sigmoid function.
+and $\mathit{sig}(x)$ is the sigmoid function.
 
@@
 Recently \cite{DBLP:conf/lics/KobayashiLG19} have extended HORS to \emph{probabilistic higher-order recursion schemes} (PHORS), which are HORS augmented with probabilistic (binary) branching $\oplus_p$.
-As HORS are in essence the $\lambda \tY$-calculus \cite{DBLP:conf/lics/Statman02} (i.e.~pure simply-typed lambda calculus with recursion, generated from a finite base type), PHORS are definable in SPCF:
-(order-$n$) PHORS is encodable as (order-$n$) CbN SPCF, but the former is strictly less expressive (because the underlying recursion schemes are not Turing complete). 
+As HORS are in essence the $\lambda \tY$-calculus \citep{DBLP:conf/lics/Statman02} (i.e.~pure simply-typed lambda calculus with recursion, generated from a finite base type), PHORS are definable in SPCF:
+(order-$n$) PHORS is encodable as (order-$n$) (call-by-name) SPCF, but the former is strictly less expressive (because the underlying recursion schemes are not Turing complete). 
 Some interesting SPCF terms such as \Cref{ex:non-affine with continuous distribution} %\refExample{complexExample} 

### 2020-12-29 LICS21/supermartingales-proofs.tex
Sec. 3: Supermartingales - cutting down to 2- pp.

@@
 It then remains to observe: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=0} \mathbb{P}[T > i]$.
+\end{proof}
+
+\TechnicalRankingSupermartingale*
+
+First, some notation.
+
+Given an $\omega$-sequence (e.g.~$s \in \entrosp$) and $m \geq 0$, we write $s_{\leq m} \in I^m$ to mean the prefix of $s$ of length $m$.
+For $n \geq 0$, define
+\begin{align*}
+%M_n(s) &:= \pi_0 (\red^n(M, s))\\
+\ndraw{n}{s} &:= |\{k < n \mid \exists E: M_k(s) = E[\tsample]\}|
+\end{align*}
+so that $\pi_1 (\red^n(M, s)) = %\overbrace{\pi_t( \cdots (\pi_t}^{\ndraw n s}(s))$. 
+\pi_t( \cdots (\pi_t (s))$, with $\pi_t$ applied ${\ndraw n s}$ times.
+%and the filtration $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
+The $\calF_n$-measurability of $M_n$ (and hence of $\#\mathrm{draw}_n$) follows from \citep{DBLP:conf/icfp/BorgstromLGS16}.
+%$\#\mathrm{draw}_n$ is a stopping time (bounded by $n$). 
+\iffalse
+\akr{$\#\mathrm{draw}_n$ is not a stopping time.
+It doesn't look like you actually use this claim anyway, so it should just be fine to remove, but did you mean something different?} \lo{I agree, and I don't actually use this claim.}
+\fi
+Take $s \in A \in \calF_n$ with $\ndraw{n}{s} = l$.
+For any $s'\in \entrosp$, if $s_{\leq l} = s_{\leq l}'$ then $s' \in A$.
+It follows that ${\set{s_{\leq l}} \cdot I^\omega} \subseteq A$.
+\lo{NOTE. To fix notation:
+\(
+\sigma(M_{n}) = \sigma\big(\set{M_{n}^{-1}(\alpha_j, U_{\alpha_j})
+\mid \alpha_j\in \mathsf{Sk}_j, U_{\alpha_j} \in \mathcal{B}(\Real^j), j \geq 0}\big)
+\)
+}
+%Take $M \in \Lambda^0$ (closed SPCF terms). Define, for each $n \in \omega$, the random variable $M_n : \entrosp \to \Lambda^0$ by $M_n(s) := \pi_0 (\red^n(M, s))$.
+
+\begin{proof}
+We show the non-trivial case of $i = 2$.
+First we express 
+\begin{equation}
+f(M_{n+1})[s \in \mathbf{T}_2] = \sum_{i \in \calI} 
+f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])[s \in U_i]
+\label{eqn:f 2 n+1}
+\end{equation}
+where 
+\begin{itemize}
+\item $\calI$ is a countable indexing set
+\item $E_i[\cdot][\tsample] \in \mathsf{Sk}_{j_i}$, and $\sigma_i : \entrosp \to \Real^{j_i}$, and $\rho(s) := \pi_h(\pi_1(\red^n(M, s))) = \pi_h(\pi_t^{l_i}(s)) \in \Real$ 
+\item $\set{U_i}_{i \in \calI}$ is a partition of $\mathbf{T}_2$, 
+where each $U_i$ is determined by a skeletal term $E_i[\cdot][\tsample]$ and a number (of draws) $l_i \leq n$, satisfying
+\(
+s \in U_i
+\iff
+(E_i[\underline{r}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s),
+\;
+\hbox{for some }r \in \Real^{j_i}.
+\)
+(We use the (measurable) function $\sigma_i : \entrosp \to \Real^{j_i}$ to skolemise the existentially quantified $r$.)
+Equivalently
+\[
+U_i := \#\mathrm{draw}_n^{-1}(l_i) \cap M_n^{-1}(E_i[\cdot][\tsample], \Real^{j_i}) \in \calF_n.
+\]
+%\lo{I.e.~$(E_i[\underline{\sigma_i(s)}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s)$ iff $s \in U_i$.}
+\end{itemize}
+
+Observe that if $s \in U_i$ then $\set{s_{\leq l_i}} \cdot I^\omega \subseteq U_i$;
+in fact $(U_i)_{\leq l_i} \cdot I^\omega = U_i$.
+%Moreover, for any $A \in \calF_n$, if $s \in A$ then $\set{s_{\leq l}} \cdot I^\omega \subseteq A$.
+This means that for any measurable $g : \entrosp \to \Real_{\geq 0}$, if $g(s)$ only depends on the prefix of $s$ of length $(l_i+1)$, then, writing $\widehat{g} : I^{l_i+1} \to \Real_{\geq 0}$ where $g(s) = \widehat{g}(s_{\leq l_i+1})$, we have: for any $A \in \calF_n$
+\begin{equation}
+\int_{A \cap U_i}  \mu_{\entrosp}(\dif s) \, g(s) = 
+\int_{(A \cap U_i)_{\leq l_i+1}} \Leb_{l_i+1}(\dif t) \, \widehat{g}(t)
+\label{eqn:truncate}
+\end{equation}
+%(For simplicity, we will write $g' = g$ in the following.)
+%where $A_{\leq m} := \set{s_{\leq m} \mid s \in A}$.
+
+Take $s \in U_i$, and set $l = l_i$.
+Plainly $\sigma_i(s)$ depends on $s_{\leq l}$, and $\rho(s)$ depends on $s_{\leq l +1}$.
+Take $u \in (U_i)_{\leq l}$.
+It then follows from the definition of ranking function that
+\[%\textstyle
+\int_I \Leb(\dif r) \, f(E_i[\underline{\widehat{\sigma_i}(u)}][\underline{r}])) \leq f(E_i[\underline{\widehat{\sigma_i}(u)}][\tsample])).
+\]
+Take $A \in \calF_n$, and integrating both sides, we get
+\begin{align*}
+& \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, \int_I \textrm{Leb}(\dif r) \, f(E_i[\underline{\widehat{\sigma_i}(u)}][\underline{r}])) \\
+& \leq \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, f(E_i[\underline{\widehat{\sigma_i}(u)}][\tsample])).
+\end{align*}
+Since $\Leb_{l+1}$ is the (unique) product measure satisfying $\Leb_{l+1}(V \times B) = \Leb_l(V) \cdot \Leb(B)$, and $(U_i)_{\leq l_i} \cdot I^\omega = U_i$, we have
+%it follows from the preceding observation that
+\begin{align*}
+& \int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\underline{\widehat{\sigma_i}(u')}][\underline{\widehat{\rho}(u')}])) \\
+& \leq 
+\int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\underline{\widehat{\sigma_i}(u')}][\tsample]))
+\end{align*}
+and so, by (\ref{eqn:truncate})
+\begin{align}
+& \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])) \nonumber \\
+& \leq 
+\int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\tsample])).
+\label{eqn:a u ui}
+\end{align}
+
+Now, integrating both sides of (\ref{eqn:f 2 n+1}), we have
+\begin{align*}
+& \int_A \mu_{\entrosp}(\dif s) \, f_{n+1}(M)[s \in \mathbf{T}_i] \\
+&= 
+\int_A \mu_{\entrosp}(\dif s) \, \sum_{i \in \calI} f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])[s \in U_i] \\
+&= 
+\sum_{i \in \calI} \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])\\
+&\leq 
+\sum_{i \in \calI} \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\tsample])
+\quad \hbox{$\because$ (\ref{eqn:a u ui})}\\
+&= 
+\int_{A} \mu_{\entrosp}(\dif s) \, \sum_{i \in \calI} \, f(E_i[\underline{\sigma_i(s)}][\tsample])[s \in U_i]\\
+&= \int_{A} \mu_{\entrosp}(\dif s) f_{n}(M)[s \in \mathbf{T}_2]
+%\quad \hbox{$\because$ (\ref{eqn:f 2 n+1})}
+\end{align*}
+
 \end{proof}

### 2020-12-29 LICS21/supermartingales.tex
Sec. 3: Supermartingales - cutting down to 2- pp.

@@
 \label{def:ranking function}
-A \emph{ranking function on $M \in \Lambda^0$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
+A \emph{ranking function on $M \in \Lambda^0$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$\footnote{The fact that the ranking function can be 0 in some cases before it reaches a value is necessary to get \Cref{thm:minimal} to work neatly.}, and
 \begin{enumerate}
@@
 \end{enumerate}
+\iffalse
 \lo{@Andrew: As defined, ranking function $f$ is not required to satisfy: $f(N) = 0$ iff $N$ is a value.
@@
 \lo{OK. Let's leave \Cref{def:ranking function} as it is.}
+\fi
 We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that $f(N) = 0$ iff $N$ is a value; and for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.
@@
 
-\changed[lo]{Fix a probability space $(\Omega, \calF, \mathbb{P})$ and a filtration $(\calF_n)_{n \geq 0}$. 
+Fix a probability space $(\Omega, \calF, \mathbb{P})$ and a filtration $(\calF_n)_{n \geq 0}$. 
 (In subsequent applications to termination analysis, $\Omega = \entrosp$, and $\mathbb{P} = \mu_{\entrosp}$.)
 Let $T$ be a random variable that takes values in $\mathbb{N} \cup \set{\infty}$.
-We call $T$ a \emph{stopping time adapted to $(\calF_n)_{n \geq 0}$} just if $\set{T = n} \in \calF_n$, for all $n \geq 0$.}
+We call $T$ a \emph{stopping time adapted to $(\calF_n)_{n \geq 0}$} just if $\set{T = n} \in \calF_n$, for all $n \geq 0$.
 
@@
 \begin{enumerate}
-\item A sequence of random variables $(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \emph{supermartingale} if for all $n \geq 0$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~\changed[lo]{(i.e.~for all $A \in \calF_n$, $\int_A \mathbb{P}(\dif \omega) \, Y_{n+1}(\omega) \leq \int_A \mathbb{P}(\dif \omega) \,Y_n(\omega)$).}
+\item A sequence of random variables $(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \emph{supermartingale} if for all $n \geq 0$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~(i.e.~for all $A \in \calF_n$, $\int_A \mathbb{P}(\dif \omega) \, Y_{n+1}(\omega) \leq \int_A \mathbb{P}(\dif \omega) \,Y_n(\omega)$).
 \item Let $\epsilon > 0$. 
@@
 \begin{remark}
-\changed[lo]{Our notion of ranking supermartingale is a slight generalisation of the original definition, due to \citep{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, which does not refer to a stopping time explicitly:
-Fioriti and Hermanns call a supermartingale $(Y_n)_{n \geq 0}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot {\bf 1}_{\set{Y_n > 0}}$.}
+Our notion of ranking supermartingale is a slight generalisation of the original definition, due to \citep{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, which does not refer to a stopping time explicitly:
+Fioriti and Hermanns call a supermartingale $(Y_n)_{n \geq 0}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot {\bf 1}_{\set{Y_n > 0}}$.
 \end{remark}
@@
 
-%\emph{Notation.}
-\changed[lo]{Henceforth, fix the probability space $(\entrosp, \calF, \mu_{\entrosp})$, and a closed SPCF term $M$.
+Henceforth, fix the probability space $(\entrosp, \calF, \mu_{\entrosp})$, and a closed SPCF term $M$.
 For $n \geq 0$, define the random variables 
@@
 and the filtration $(\mathcal{F}_n)_{n \geq 0}$ where $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
-Thus $T_M$ is the \emph{runtime} of $M$ (and $M$ is AST iff $\mu_{\entrosp}(T_M < \infty) = 1$).}
+Thus $T_M$ is the \emph{runtime} of $M$ (and $M$ is AST iff $\mu_{\entrosp}(T_M < \infty) = 1$).
 
@@
 \end{theorem}
-\lo{A better name for this theorem: ``Ranking function'', ``Supermartingale'', ``(Ranking) supermartingale''?}
+\lo{@Andrew: A better name for this theorem: ``Ranking function'', ``Supermartingale'', ``(Ranking) supermartingale''?}
 
@@
 We shall obtain the theorem as a corollary of a technical lemma (\Cref{lem:key rankable}).
-First, some notation.
-
-Given an $\omega$-sequence (e.g.~$s \in \entrosp$) and $m \geq 0$, we write $s_{\leq m} \in I^m$ to mean the prefix of $s$ of length $m$.
-%Fix a closed SPCF term $M \in \Lambda^0$, and a ranking function $f$ for it, satisfying $f(N) = 0$ iff $N$ is a value.
-%Let $n \geq 0$, define the following random variables on the probability space $(S, \calF, \mu)$:
-%\begin{align*}
-%M_n(s) &:= \pi_0 (\red^n(M, s))\\
-%\ndraw{n}{s} &:= |\{k < n \mid \exists E: M_k(s) = E[\tsample]\}|,
-%\end{align*}
-For $n \geq 0$, define
-\begin{align*}
-%M_n(s) &:= \pi_0 (\red^n(M, s))\\
-\ndraw{n}{s} &:= |\{k < n \mid \exists E: M_k(s) = E[\tsample]\}|
-\end{align*}
-so that $\pi_1 (\red^n(M, s)) = %\overbrace{\pi_t( \cdots (\pi_t}^{\ndraw n s}(s))$. 
-\pi_t( \cdots (\pi_t (s))$, with $\pi_t$ applied ${\ndraw n s}$ times.
-%and the filtration $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
-The $\calF_n$-measurability of $M_n$ (and hence of $\#\mathrm{draw}_n$) follows from \citep{DBLP:conf/icfp/BorgstromLGS16}.
-%$\#\mathrm{draw}_n$ is a stopping time (bounded by $n$). 
-\iffalse
-\akr{$\#\mathrm{draw}_n$ is not a stopping time.
-It doesn't look like you actually use this claim anyway, so it should just be fine to remove, but did you mean something different?} \lo{I agree, and I don't actually use this claim.}
-\fi
-Take $s \in A \in \calF_n$ with $\ndraw{n}{s} = l$.
-For any $s'\in \entrosp$, if $s_{\leq l} = s_{\leq l}'$ then $s' \in A$.
-It follows that ${\set{s_{\leq l}} \cdot I^\omega} \subseteq A$.
-\lo{NOTE. To fix notation:
-\(
-\sigma(M_{n}) = \sigma\big(\set{M_{n}^{-1}(\alpha_j, U_{\alpha_j})
-\mid \alpha_j\in \mathsf{Sk}_j, U_{\alpha_j} \in \mathcal{B}(\Real^j), j \geq 0}\big)
-\)
-}
-%Take $M \in \Lambda^0$ (closed SPCF terms). Define, for each $n \in \omega$, the random variable $M_n : \entrosp \to \Lambda^0$ by $M_n(s) := \pi_0 (\red^n(M, s))$.
 
@@
 N]$, \emph{type 2} if it has the shape $E[\tsample]$; \emph{type 3} if it has the shape $E[R]$ where $R$ is any other redex, and \emph{type 4} if it is a value.
-Henceforth fix an $n \geq 0$, and define $\mathbf{T}_i := \set{s \mid M_n(s) \hbox{ is type-$i$}}$.
+Henceforth fix an $n \geq 0$, and define $\mathbf{T}_i := \set{s \mid M_n(s) \hbox{ is type $i$}}$.
 It is straightforward to see that each $\mathbf{T}_i \in \calF_n$, and $\set{\mathbf{T}_1, \mathbf{T}_2, \mathbf{T}_3, \mathbf{T}_4}$ is a partition of $\entrosp$.
@@
 
-\begin{lemma}
+\begin{restatable}[Technical]{lemma}{TechnicalRankingSupermartingale}
 \label{lem:key rankable}
@@
 \] 
-\changed[lo]{Thus 
 %$\int_A \mu_{\entrosp}(\dif s) \, f(M_{n+1}) \leq \int_A \mu_{\entrosp}(\dif s) \, f(M_n)$, i.e., 
-$\mathbb{E}[f(M_{n+1}) \mid \calF_n] \leq f(M_n)$ a.s.}
-\end{lemma}
-
-\begin{proof}
-We show the non-trivial case of $i = 2$.
-First we express 
-\begin{equation}
-f(M_{n+1})[s \in \mathbf{T}_2] = \sum_{i \in \calI} 
-f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])[s \in U_i]
-\label{eqn:f 2 n+1}
-\end{equation}
-where 
-\begin{itemize}
-\item $\calI$ is a countable indexing set
-\item $E_i[\cdot][\tsample] \in \mathsf{Sk}_{j_i}$, and $\sigma_i : \entrosp \to \Real^{j_i}$, and $\rho(s) := \pi_h(\pi_1(\red^n(M, s))) = \pi_h(\pi_t^{l_i}(s)) \in \Real$ 
-\item $\set{U_i}_{i \in \calI}$ is a partition of $\mathbf{T}_2$, 
-where each $U_i$ is determined by a skeletal term $E_i[\cdot][\tsample]$ and a number (of draws) $l_i \leq n$, satisfying
-\(
-s \in U_i
-\iff
-(E_i[\underline{r}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s),
-\;
-\hbox{for some }r \in \Real^{j_i}.
-\)
-(We use the (measurable) function $\sigma_i : \entrosp \to \Real^{j_i}$ to skolemise the existentially quantified $r$.)
-Equivalently
-\[
-U_i := \#\mathrm{draw}_n^{-1}(l_i) \cap M_n^{-1}(E_i[\cdot][\tsample], \Real^{j_i}) \in \calF_n.
-\]
-%\lo{I.e.~$(E_i[\underline{\sigma_i(s)}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s)$ iff $s \in U_i$.}
-\end{itemize}
-
-Observe that if $s \in U_i$ then $\set{s_{\leq l_i}} \cdot I^\omega \subseteq U_i$;
-in fact $(U_i)_{\leq l_i} \cdot I^\omega = U_i$.
-%Moreover, for any $A \in \calF_n$, if $s \in A$ then $\set{s_{\leq l}} \cdot I^\omega \subseteq A$.
-This means that for any measurable $g : \entrosp \to \Real_{\geq 0}$, if $g(s)$ only depends on the prefix of $s$ of length $(l_i+1)$, then, writing $\widehat{g} : I^{l_i+1} \to \Real_{\geq 0}$ where $g(s) = \widehat{g}(s_{\leq l_i+1})$, we have: for any $A \in \calF_n$
-\begin{equation}
-\int_{A \cap U_i}  \mu_{\entrosp}(\dif s) \, g(s) = 
-\int_{(A \cap U_i)_{\leq l_i+1}} \Leb_{l_i+1}(\dif t) \, \widehat{g}(t)
-\label{eqn:truncate}
-\end{equation}
-%(For simplicity, we will write $g' = g$ in the following.)
-%where $A_{\leq m} := \set{s_{\leq m} \mid s \in A}$.
-
-Take $s \in U_i$, and set $l = l_i$.
-Plainly $\sigma_i(s)$ depends on $s_{\leq l}$, and $\rho(s)$ depends on $s_{\leq l +1}$.
-Take $u \in (U_i)_{\leq l}$.
-It then follows from the definition of ranking function that
-\[%\textstyle
-\int_I \Leb(\dif r) \, f(E_i[\underline{\widehat{\sigma_i}(u)}][\underline{r}])) \leq f(E_i[\underline{\widehat{\sigma_i}(u)}][\tsample])).
-\]
-Take $A \in \calF_n$, and integrating both sides, we get
-\begin{align*}
-& \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, \int_I \textrm{Leb}(\dif r) \, f(E_i[\underline{\widehat{\sigma_i}(u)}][\underline{r}])) \\
-& \leq \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, f(E_i[\underline{\widehat{\sigma_i}(u)}][\tsample])).
-\end{align*}
-Since $\Leb_{l+1}$ is the (unique) product measure satisfying $\Leb_{l+1}(V \times B) = \Leb_l(V) \cdot \Leb(B)$, and $(U_i)_{\leq l_i} \cdot I^\omega = U_i$, we have
-%it follows from the preceding observation that
-\begin{align*}
-& \int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\underline{\widehat{\sigma_i}(u')}][\underline{\widehat{\rho}(u')}])) \\
-& \leq 
-\int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\underline{\widehat{\sigma_i}(u')}][\tsample]))
-\end{align*}
-and so, by (\ref{eqn:truncate})
-\begin{align}
-& \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])) \nonumber \\
-& \leq 
-\int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\tsample])).
-\label{eqn:a u ui}
-\end{align}
-
-Now, integrating both sides of (\ref{eqn:f 2 n+1}), we have
-\begin{align*}
-& \int_A \mu_{\entrosp}(\dif s) \, f_{n+1}(M)[s \in \mathbf{T}_i] \\
-&= 
-\int_A \mu_{\entrosp}(\dif s) \, \sum_{i \in \calI} f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])[s \in U_i] \\
-&= 
-\sum_{i \in \calI} \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])\\
-&\leq 
-\sum_{i \in \calI} \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\tsample])
-\quad \hbox{$\because$ (\ref{eqn:a u ui})}\\
-&= 
-\int_{A} \mu_{\entrosp}(\dif s) \, \sum_{i \in \calI} \, f(E_i[\underline{\sigma_i(s)}][\tsample])[s \in U_i]\\
-&= \int_{A} \mu_{\entrosp}(\dif s) f_{n}(M)[s \in \mathbf{T}_2]
-%\quad \hbox{$\because$ (\ref{eqn:f 2 n+1})}
-\end{align*}
-
-\end{proof}
-
-As an immediate corollary of \Cref{lem:key rankable}, each $f(M_n)$ is integrable:
+Hence $\mathbb{E}[f(M_{n+1}) \mid \calF_n] \leq f(M_n)$ a.s.
+\end{restatable}
+As an immediate corollary of \Cref{lem:key rankable}, each $f(M_n)$ is integrable
+\iffalse
 \begin{align*}
@@
 
-\iffalse
 [** To see $f_{2, n+1}(M) \leq f(M_n)[T_2]$, take $s \in T_2$. Then $f_{2, n+1}(M)(s) = f(E[\underline{a}])$ and $M_n = E[\tsample]$, for some $a \in [0, 1]$ and evaluation context $E$. Hence 
@@
 \fi
-This concludes the proof of %\Cref{thm:rankable gives supermartingale}. 
-\Cref{thm:rankable and strict rankable}. %\hfill \qed
+This concludes the proof of \Cref{thm:rankable and strict rankable}.
 
@@
 Let $f$ be a ranking function on $M$.
-For 1, since $(T_n)_{n \geq 0}$ is an increasing sequence of stopping times, each is adapted to $(\calF_n)_{n \geq 0}$ (\Cref{lem:TnBounded}) and bounded,
+For 1, since $(T_n)_{n \geq 0}$ is an increasing sequence of stopping times, each is adapted to $(\calF_n)_{n \geq 0}$ and bounded (\Cref{lem:TnBounded}),
 and $(f(M_n))_{n \geq 0}$ is a supermartingale also adapted to $(\calF_n)_{n \geq 0}$ (\Cref{thm:rankable and strict rankable}),

### 2020-12-31 LICS21/ranking.tex
Sec. 3: Repositioning ranking supermaringale lemma; introduce two subsections.
Changed S and \mu to \entrosp and \mu_\entrosp for all files except confluent.tex
Sec. IV: Various minors. Revised figure for geometric distribution example.

@@
 % !TEX root = main.tex
-\section{Constructing Ranking Functions}
+\section{Constructing ranking functions}
 \label{sec:ranking}
@@
 In particular, the expected number of $\tY$-reduction steps from each reachable term is a ranking function. 
-Note that a finite number of expected $\tY$-reduction steps does not necessarily imply a finite number of expected total reduction steps.
+However, note that a finite number of expected $\tY$-reduction steps does not necessarily imply a finite number of expected total reduction steps.
 
@@
 \label{ex:tY finite does not imply t finite}
-The term
+The term $M = \Xi \, (\lambda x. x+1)$ where
 \[
-M = \big(\tY \lambda f n. \tif{\tsample - 0.5 < 0}{n \, \underline 0}{f (\lambda x. n (n x))}\big) (\lambda x. x+1)
+\Xi := \tY \lambda f n. \tif{\tsample - 0.5 < 0}{n \, \underline 0}{f (\lambda x. n (n x))}
 \] 
-terminates with only 2 $\tY$-reductions on average i.e.~$\expect{T_M^\tY} = 2$, but applies the increment function $2^n$ times with probability $2^{-1-n}$ for $n \geq 0$, which diverges i.e.~$\expect{T_M} = \infty$.
+terminates with only 2 $\tY$-reductions on average, i.e.,~$\expect{T_M^\tY} = 2$, but applies the increment function $2^n$ times with probability $2^{-n-1}$ for $n \geq 0$, which diverges, i.e.,~$\expect{T_M} = \infty$.
 \end{example}
@@
 therefore $\mathbb E[g(M_n)] \leq \mathbb E[f(M_n)] + g(M)-f(M)$, for all $n$.
-
-We have $\mathbb E[f(M_n)] = \sum_{k=n}^\infty \mathbb P[M_k = E[\tY N] \text{ for some }E, N] \to 0 \text{ as } n \to \infty$; 
+Now $\mathbb E[f(M_n)] = \sum_{k=n}^\infty \mathbb P[M_k = E[\tY N] \text{ for some }E, N] \to 0 \text{ as } n \to \infty$; 
 therefore as $g(M) - f(M) < 0$, eventually $\mathbb E[g(M_n)] < 0$, which is impossible. 
@@
 \begin{example}[Geometric distribution]
-Consider the term
-\begin{align*}
-&(\tY \lambda f, n. \\
-&\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
-&) \, \underline{0}
-\end{align*}
-which generates a geometric distribution (let $\Theta = \lambda f, n : \tif{\tsample \underline{- 0.5} < 0}{n}{f (n\underline{+ 1})}$, then the term is $(\tY \Theta)\ \underline 0$).
-    Despite its simplicity, its $Rch$ contains all the terms: see \Cref{fig:geometric distribution}.
+% Consider the term
+% \begin{align*}
+% &(\tY \lambda f, n. \\
+% &\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
+% &) \, \underline{0}
+% \end{align*}
+%\[
+%\big(\tY \lambda f, n. \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
+%\big) \, \underline{0}
+%\]
+Let
+\[
+\Theta := \lambda f, n. \tif{\tsample - 0.5 < 0}{n}{f (n+1)}. 
+\]
+The term  $(\tY \, \Theta) \, \underline 0$ generates a geometric distribution.
+Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}.
+\lo{\Cref{fig:geometric distribution} is confusing: the arrow means either one-step reduction, or its reflexive transitive closure, or ``has the same syntactic shape''.
+I think \Cref{fig:geometric distribution 1} is not confusing (arrow does mean one-step reduction; note $r'$), and it would be correct to say that $\mathit{Rch}(\tY \, \Theta \, \underline 0)$ contains all the terms in \Cref{fig:geometric distribution 1}, for each $i \geq 0$.}
 \begin{figure}
 \begin{tikzpicture}[node distance=0.5em, anchor=west]
-    \node (1) at (0,-0) {$\tY\ \Theta\ \underline 0$};
-    \node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\ \Theta\ (n \underline{+ 1})}) z)\ \underline i$};
-    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\ \Theta\ (n \underline{+ 1})})\ \underline i$};
-    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY\ \Theta\ (\underline i\ \underline{+ 1})}$};
-    \node (5) at (0,-4) {$\tif{\underline r\ \underline{- 0.5} < 0}{\underline i}{\tY\ \Theta\ (\underline i\ \underline{+ 1})}$};
-    \node (6) at (0,-5) {$\tif{\underline r < 0}{\underline i}{\tY\ \Theta\ (\underline i\ \underline{+ 1})}$};
+    \node (1) at (0,-0) {$\tY\, \Theta\, \underline 0$};
+    \node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, \underline i$};
+    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})})\, \underline i$};
+    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
+    \node (5) at (0,-4) {$\tif{\underline r\, \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
+    \node (6) at (0,-5) {$\tif{\underline r < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
     \node (7) at (2,-6) [text width=] {$\underline i$};
-    \node (8) at (0,-7) {$\tY\ \Theta\ (\underline i\ \underline{+ 1})$};
-    \node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\ \Theta\ (n \underline{+ 1})}) z)\ (\underline i\ \underline{+ 1})$};
+    \node (8) at (0,-7) {$\tY\, \Theta\, (\underline i\, \underline{+ 1})$};
+    \node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, (\underline i\, \underline{+ 1})$};
 
@@
 \end{tikzpicture}
-\caption{Some reachable terms of $\tY\ \Theta\ \underline 0$\label{fig:geometric distribution}}
+\caption{Some reachable terms of $\tY \, \Theta \, \underline 0$\label{fig:geometric distribution}}
+\end{figure}
+\begin{figure}
+\centering
+\begin{tikzpicture}[node distance=0.5em, anchor=west]
+    \node (1) at (0,-0) {$\tY\, \Theta\, \underline i$};
+    \node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, \underline i$};
+    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})})\, \underline i$};
+    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
+    \node (5) at (0,-4) {$\tif{\underline r\, \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
+    \node (6) at (0,-5) {$\tif{\underline r' < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
+    \node (7) at (2,-6) [text width=] {$\underline i$};
+    \node (8) at (0,-7) {$\tY\, \Theta\, (\underline i\, \underline{+ 1})$};
+    %\node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, (\underline i\, \underline{+ 1})$};
+
+    \draw [->] (1.south west) to [bend right=45] (2.north west);
+    \draw [->] (2.south west) to [bend right=45] (3.north west);
+    \draw [->] (3.south west) to [bend right=45] (4.north west);
+    \draw [->] (4.south west) to [bend right=45] (5.north west);
+    \draw [->] (5.south west) to [bend right=45] (6.north west);
+    \draw [->] (6.south west) to [bend right=45] (8.north west);
+    %\draw [->] (8.south west) to [bend right=45] (9.north west);
+    %\draw [->] (9.west) to [bend left=45] (2.west);
+    \draw [->] (6) to (7);
+\end{tikzpicture}
+\caption{Some reachable terms of $\tY \, \Theta \, \underline 0$\label{fig:geometric distribution 1}}
 \end{figure}
@@
 The definition of rankability is also inconvenient for syntactic sugar.
-It could be useful, for example, to define $M \oplus_p N = \tif{\tsample - p < 0} M N$, where $M \oplus_p N$ reduces to $M$ or $N$, depending on the first value of $s$\lo{Need a name for $s$.}, with probability $p$ resp.~$(1-p)$.
+It could be useful, for example, to define $M \oplus_p N = \tif{\tsample - p < 0} M N$, where $M \oplus_p N$ reduces to $M$ or $N$, depending on the first value of $s \in \entrosp$\lo{Need a name for $s$.}, with probability $p$ resp.~$(1-p)$.
 Technically though, it reduces first to $\tif{\underline r - p < 0} M N$ for all $r \in I$, so those terms all need values of the ranking function too.
@@
 The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on partial ranking functions.
-Define $f_2(N) = \int_S f_1(\nnext(N,s)) + g(N,s) \, \mu(\mathrm d s)$.
+Define $f_2(N) = \int_\entrosp f_1(\nnext(N,s)) + g(N,s) \, \mu_{\entrosp}(\mathrm d s)$.
 The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction).

### 2020-12-31 LICS21/ranking.tex
Further comments on figure about reachable terms (geometric distribution example)

@@
 The term  $(\tY \, \Theta) \, \underline 0$ generates a geometric distribution.
-Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}.
-\lo{\Cref{fig:geometric distribution} is confusing: the arrow means either one-step reduction, or its reflexive transitive closure, or ``has the same syntactic shape''.
-I think \Cref{fig:geometric distribution 1} is not confusing (arrow does mean one-step reduction; note $r'$), and it would be correct to say that $\mathit{Rch}(\tY \, \Theta \, \underline 0)$ contains all the terms in \Cref{fig:geometric distribution 1}, for each $i \geq 0$.}
+\changed[akr]{Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}.}
+\lo{\Cref{fig:geometric distribution} is confusing: the arrow means either one-step reduction, or its reflexive transitive closure, or ``has the same syntactic shape''.}
+
+\lo{Perhaps you have in mind an informal, schematic interpretation, reading a term $C[i, r]$ as ``there exists a term $C[i, r]$, for appropriate $i, r \in \mathbb{R}$''.
+One way to make this precise is to consider the graph obtained by quotienting the one-step reduction relation by an appropriate equivalence relation over terms (that specifies this scheme reading).}
+
+\lo{\Cref{fig:geometric distribution 1} is less ambitious, but it is not confusing in that arrow does mean one-step reduction (also note $r'$), and it would be correct to say (cf.~the preceding blue-highlighted phrase): ``Despite its simplicity, $\mathit{Rch}(\tY \, \Theta \, \underline 0)$ contains all the terms in \Cref{fig:geometric distribution 1}, for each $i \geq 0$''.}
+
+\lo{The graph can be simplified if we redefine general recursion using a fixpoint operator as per \Cref{eq:letrec redex}.
+But (especially in view of the vast investment in the confluent trace semantics) it is perhaps too late to change things now?}
+
 \begin{figure}
@@
 \begin{tikzpicture}[node distance=0.5em, anchor=west]
-    \node (1) at (0,-0) {$\tY\, \Theta\, \underline i$};
+    \node (1) at (0,-0) {$\tY\, \Theta\, \underline i$ \quad [provided $i = 0$]};
     \node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, \underline i$};
@@
     \node (8) at (0,-7) {$\tY\, \Theta\, (\underline i\, \underline{+ 1})$};
-    %\node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, (\underline i\, \underline{+ 1})$};
+    \node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, (\underline i\, \underline{+ 1})$};
+    \node (10) at (0,-9) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, \underline{i + 1}$};
 
@@
     \draw [->] (6.south west) to [bend right=45] (8.north west);
-    %\draw [->] (8.south west) to [bend right=45] (9.north west);
-    %\draw [->] (9.west) to [bend left=45] (2.west);
+    \draw [->] (8.south west) to [bend right=45] (9.north west);
+    \draw [->] (9.south west) to [bend right=45] (10.north west);
     \draw [->] (6) to (7);

### 2020-12-31 LICS21/ranking.tex
Geometric distribution example: Adding another graph of reachable terms using fixpoint operator (instead of Y-combinator).

@@
 
-\lo{The graph can be simplified if we redefine general recursion using a fixpoint operator as per \Cref{eq:letrec redex}.
+\lo{The graph can be simplified somewhat if we redefine general recursion using a fixpoint operator $\tY_{f, n}\Theta$ with redex rule as per \Cref{eq:letrec redex}: see \Cref{fig:geometric distribution 2}.
 But (especially in view of the vast investment in the confluent trace semantics) it is perhaps too late to change things now?}
@@
 \end{figure}
+
+\begin{figure}
+\centering
+\begin{tikzpicture}[node distance=0.5em, anchor=west]
+    \node (1) at (0,-0) {$\tY_{f, n}\Theta\, \underline i$};
+    %\node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY_{f, n}\Theta\, (n \underline{+ 1})}) z)\, \underline i$};
+    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY_{f, n}\Theta\, (n \underline{+ 1})})\, \underline i$};
+    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY_{f, n}\Theta\, (\underline i\, \underline{+ 1})}$};
+    \node (5) at (0,-4) {$\tif{\underline r\, \underline{- 0.5} < 0}{\underline i}{\tY_{f, n}\Theta\, (\underline i\, \underline{+ 1})}$};
+    \node (6) at (0,-5) {$\tif{\underline r' < 0}{\underline i}{\tY_{f, n}\Theta\, (\underline i\, \underline{+ 1})}$};
+    \node (7) at (2,-6) [text width=] {$\underline i$};
+    \node (8) at (0,-7) {$\tY_{f, n}\Theta\, (\underline i\, \underline{+ 1})$};
+    \node (9) at (0,-8) {$\tY_{f, n}\Theta\, \underline{i + 1}$};
+    %\node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY_{f, n}\Theta\, (n \underline{+ 1})}) z)\, (\underline i\, \underline{+ 1})$};
+    %\node (10) at (0,-9) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY_{f, n}\Theta\, (n \underline{+ 1})}) z)\, \underline{i + 1}$};
+
+    \draw [->] (1.south west) to [bend right=45] (3.north west);
+    %\draw [->] (2.south west) to [bend right=45] (3.north west);
+    \draw [->] (3.south west) to [bend right=45] (4.north west);
+    \draw [->] (4.south west) to [bend right=45] (5.north west);
+    \draw [->] (5.south west) to [bend right=45] (6.north west);
+    \draw [->] (6.south west) to [bend right=45] (8.north west);
+    \draw [->] (8.south west) to [bend right=45] (9.north west);
+    %\draw [->] (9.south west) to [bend right=45] (10.north west);
+    \draw [->] (6) to (7);
+\end{tikzpicture}
+\caption{Some reachable terms of $\tY_{f, n}\Theta \, \underline 0$\label{fig:geometric distribution 2}}
+\end{figure}
+
 \end{example}

### 2020-12-31 LICS21/antitone.tex
Further comments on figure about reachable terms (geometric distribution example)
Set theorem (similarly, definition, example, etc.) to number within sections.

@@
 \begin{definition}
-We say that a supermartingale $(Y_n)_{n \geq 0}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \geq 0}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
+We say that a supermartingale $(Y_n)_{n \geq 0}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \geq 0}$ are \emph{antitone strict} if for all $n \geq 0$, we have $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
 \(
@@
 \begin{example}
+\iffalse
 \akr{Consider a random variable $X$, uniformly distributed in the interval $[-1,1]$, $T := \min \set{n \mid 2^{-n} < X}$, 
@@
 } \lo{The following reformulation defines $T$ and $Y_k$ directly as a measurable function on a probability space (rather than via a r.v.~$X$).}
-
+\fi
 Take the probability space $(\Omega, \calF, \mathbb{P})$ where $\Omega$ is the closed interval of reals $[-1, 1]$, $\calF$ the Borel $\sigma$-algebra, and $\mathbb{P}$ the corresponding Lebesgue probability measure.
@@
 with the redex rule 
-\[
+\begin{equation}
 (\mathsf{letrec} \, f^{A \to B} \, x^A . M^B) \, V \to M[\mathsf{letrec} \, f \, x . M / f, V / x].
-\]}
+\label{eq:letrec redex}
+\end{equation}
+}
 \end{example}

### 2020-12-31 LICS21/SPCF.tex
Global change: the entrospy space is denoted (\entrosp, \Sigma_\entrosp, \mu_\entrosp). NOTE: \calF -> \Sigma_\entrosp

@@
 To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples.
-Let $ I := [0,1] \subset \mathbb{R} $, and define $\entrosp := I^{\mathbb{N}}$, with the \changed[lo]{Borel $\sigma$-algebra $\calF$} and the probability measure, denoted $\mu_{\entrosp}$, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
+Let $ I := [0,1] \subset \mathbb{R} $, and define $\entrosp := I^{\mathbb{N}}$, with the \changed[lo]{Borel $\sigma$-algebra $\Sigma_\entrosp$} and the probability measure, denoted $\mu_{\entrosp}$, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
 Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and $\mu_{\entrosp} (\prod_{i=0}^\infty X_i) := \prod_{i=0}^\infty \Leb(X_i)$.
 The maps $\pi_h:\entrosp \to I, \; \pi_t:\entrosp \to \entrosp$ popping the first element are then measurable.
-\changed[lo]{Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(\entrosp, \calF, \mu_{\entrosp})$ the \emph{entropy space}.}
+\changed[lo]{Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$ the \emph{entropy space}.}
 

### 2020-12-31 LICS21/antitone.tex
Global change: the entrospy space is denoted (\entrosp, \Sigma_\entrosp, \mu_\entrosp). NOTE: \calF -> \Sigma_\entrosp

@@
 \Cref{thm:rankable implies termination},
-take the probability space $(\entrosp, \calF, \mu_\entrosp)$, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$. 
+take the probability space $(\entrosp, \Sigma_\entrosp, \mu_\entrosp)$, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$. 
 As before, $(Y_n)_{n \geq 0}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$.

### 2020-12-31 LICS21/supermartingales.tex
Global change: the entrospy space is denoted (\entrosp, \Sigma_\entrosp, \mu_\entrosp). NOTE: \calF -> \Sigma_\entrosp

@@
 
-Henceforth, fix the probability space $(\entrosp, \calF, \mu_{\entrosp})$, and a closed SPCF term $M$.
+Henceforth, fix the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$, and a closed SPCF term $M$.
 For $n \geq 0$, define the random variables 
@@
 Let $f$ be a ranking function on $M$ satisfying $f(N) = 0$ iff $N$ is a value.
-Define random variables on the probability space $(\entrosp, \calF, \mu_{\entrosp})$:
+Define random variables on the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$:
 \begin{align}

### 2020-12-31 LICS21/antitone.tex
Tidying up definitions (antitone rankable, antitone strict supermartingale w.r.t. T) and antitone ranking function soundness theorem

@@
 
-\paragraph*{Idea}
+%\paragraph*{Idea}
 
-The definition of antitone-convex ranking function $f$ (say) for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
-we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) function $\epsilon : \nnReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
+\changed[lo]{
+\begin{definition}
+%The definition of antitone ranking function $f$ for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
+%we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) function $\epsilon : \nnReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
+%\[f(E[R']) \leq f(E[R]) - \epsilon(f(E[R])) \] where $R \to R'$ is the $\tY$-redex rule.
+Given a ranking function $f$ on $M \in \Lambda^0$, we say that $f$ is \emph{antitone} if there exists an antitone function\footnote{i.e.~$r < r'$ implies $\epsilon(r) \geq \epsilon(r')$} $\epsilon : \nnReal \to \pReal$ such that %the ranking function $f$ satisfies
 \[
@@
 \]
-where $R \to R'$ is the $\tY$-redex rule.
+for all $E[R'] \in \mathit{Rch}(M)$ where $R \to R'$ is the $\tY$-redex rule.
+\end{definition}
 
@@
 \begin{definition}
-We say that a supermartingale $(Y_n)_{n \geq 0}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \geq 0}$ are \emph{antitone strict} if for all $n \geq 0$, we have $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
+Given a probability space $(\Omega, \calF, \mathbb{P})$, and a supermartingale $(Y_n)_{n \geq 0}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \geq 0}$,
+we say that $(Y_n)_{n \geq 0}$ is an \emph{antitone strict supermartingale w.r.t.~$T$} if for all $n \geq 0$, we have $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
 \(
@@
 \).
+Any closed term for which an antitone ranking function exists is called \emph{antitone rankable}.
 \end{definition}
-
+}
 %\lo{The preceding definition does not place any constraint on the stopping time $T$ (other than that it is w.r.t.~the filtration $(\calF_n)_{n \geq 0}$).
@@
 \label{thm:a-c strict}
-Let $(Y_n)_{n \geq 0}$ be an antitone strict supermartingale via $\epsilon : \nnReal \to \pReal$ and stopping time $T$. 
+Let $(Y_n)_{n \geq 0}$ be an antitone strict supermartingale w.r.t.~stopping time $T$. 
 Then $T < \infty$ a.s.
@@
 \)
-where $B_x = \left \lceil \dfrac{\expect{Y_0}+1}{x \; \epsilon(\expect{Y_0} \, x^{-1})} \right \rceil$.
+where $B_x = \left \lceil \dfrac{\expect{Y_0}+1}{x \; \epsilon(\expect{Y_0} \, x^{-1})} \right \rceil$ and $\epsilon : \nnReal \to \pReal$ is the antitone function.
 
@@
 
-\changed[lo]{Let $f$ be a ranking function on a closed SPCF term $M$.
-%Recall that $(f(M_n))_{n \geq 0}$ is a supermartingale.
+%\changed[lo]{Let $f$ be a ranking function on a closed SPCF term $M$.
 We say that $M$ is \emph{antitone rankable} by $f$ if the supermartingale $(f(M_n))_{n \geq 0}$ and the stopping time $T_M^{\tY}$, both adapted to the fitration $(\calF_n)_{n \geq 0}$ where $\calF_n := \sigma(M_1, \ldots, M_n)$, are antitone strict.}
@@
 take the probability space $(\entrosp, \Sigma_\entrosp, \mu_\entrosp)$, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$. 
-As before, $(Y_n)_{n \geq 0}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$.
+Thanks to \Cref{thm:rankable and strict rankable}, $(Y_n)_{n \geq 0}$ is a supermartingale; and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale w.r.t.~stopping time $T_M^{\tY}$.
 Thus, by \Cref{thm:a-c strict}, $T_M^{\tY} < \infty$ a.s.

### 2021-01-01 LICS21/SPCF.tex
Rearrange the confluent semantics section.

--- 2020-12-28 LICS21/SPCF.tex
@@
 
-The $\sigma$-algebra and measure on $\Lambda$ are defined by considering it as a disjoint union of equivalence classes under replacing all the real constants by a placeholder, where the measure on each class is that of $\mathbb{R}^n$, where $n$ is the number of real constants.
-\changed[lo]{Precisely, following \cite{DBLP:conf/icfp/BorgstromLGS16},
-we view $\Lambda$ as $\bigcup_{m\geq 0} \big(\mathsf{Sk}_m \times \Real^m \big)$,
-where $\mathsf{Sk}_m$ is the set of SPCF terms with exactly $m$ place-holders (a.k.a.~\emph{skeleton terms}) for numerals.
-Thus identified, we give $\Lambda$ the countable disjoint union topology of the product topology of the discrete topology on $\mathsf{Sk}_m$ and the standard topology on $\Real^m$.
-Note that the connected components of $\Lambda$ have the form $\{M\} \times \Real^m$, with $M$ ranging over $\mathsf{Sk}_m$, and $m$ over $\mathbb{N}$. 
-We fix the Borel algebra of this topology to be the $\sigma$-algebra on $\Lambda$.}
+Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $Sk(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$.
+Following \cite{DBLP:conf/icfp/BorgstromLGS16}, the $\sigma$-algebra on $\Lambda$ is defined by identifying $\Lambda$ with $\bigcup_{m\geq 0} \big(\mathsf{Sk}_m \times \Real^m \big)$, where $\mathsf{Sk}_m$ is the set of skeletons containing $m$ occurrences of $\skeletonPlaceholder$.
+Thus identified, we give $\Lambda$ the countable disjoint union topology of the product topology of the discrete topology on $\mathsf{Sk}_m$ and the standard topology on $\Real^m$, and take the corresponding Borel $\sigma$-algebra.
+Note that the connected components of $\Lambda$ have the form $\{S\} \times \Real^m$, with $S$ ranging over $\mathsf{Sk}_m$, and $m$ over $\mathbb{N}$.
 

### 2021-01-01 LICS21/confluent.tex
Motivation and examples for non-affine recursion, higher-order recursion, and programs with continuous distributions.
Rearrange the confluent semantics section.

@@
 \begin{align*}
-\alpha ::= \cdot \mid \lambda ; \alpha \mid @_1 ; \alpha \mid @_2 ; \alpha \mid \underline f_i ; \alpha \\
-\mid \tY ; \alpha \mid \textsf{if}_1 ; \alpha \mid \textsf{if}_2 ; \alpha \mid \textsf{if}_3 ; \alpha \mid \tscore ; \alpha.
+\alpha ::= & \; {\cdot} \mid \lambda ; \alpha \mid @_1 ; \alpha \mid @_2 ; \alpha \mid \underline f_i ; \alpha \\
+& {} \mid \tY ; \alpha \mid \textsf{if}_1 ; \alpha \mid \textsf{if}_2 ; \alpha \mid \textsf{if}_3 ; \alpha \mid \tscore ; \alpha.
 \end{align*}
@@
 
-Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $Sk(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$.
-\lo{Do we actually use the notation $S[r]$ in what follows?} \akr{Yes, in Lemma \ref{lem:independentSamples} for example.}
-\lo{Skeleton is defined in Sec.~\ref{sec:sampling semantics}.}
 The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, and $\tsample$ reducing to $\skeletonPlaceholder$.
@@
 \skeletonPlaceholder
-\)
+\).
 
@@
 As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the usual linear sampling semantics, there will be (with one more caveat introduced later) a sample from $I$ pre-selected for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.
+
+\begin{example}[Labelling samples by terms]
+Consider the terms
+\begin{align*}
+A[x] &= \tif{\tif{x>0}{I}{I} \, \tsample - 0.5 > 0}{0}{\Omega}\\
+B &= \tif{\tsample - 0.5 > 0}{0}{\Omega}
+\end{align*}
+If terms rather than skeletons were used to label samples, the set of traces where $A[\tsample]$ terminates would be
+\[\begin{array}{l}
+\bigcup_{r \in [0,1]} \{s \in I^{L_s(A[\tsample])} \mid \\
+\qquad s([A[\tsample]], \mathsf{if}_1;\underline{-}_1;@_1;\mathsf{if}_1) = r, \\
+\qquad s([A[\tsample], A[\underline r], \ldots, B, \mathsf{if}_1) > 0.5 \}.
+\end{array}\] 
+This is a rather unwieldy expression, but the crucial part is that $r$ occurs twice in the conditions on $s$: once as the value a sample must take, and once in the location of a sample.
+This set is unmeasurable, therefore the termination probability would not even be well-defined.
+Labelling samples by skeletons instead, this problem does not occur because there are only countably many skeleton, and at each step in a reduction sequence, only finitely many could have occurred yet.
+For this particular term, $Sk(A[\underline r])$ does not depend on the value of $r$, therefore the set where it terminates becomes simply
+\begin{align*}
+&\{s \in I^{L_s(A[\tsample])} \mid \\
+&\qquad s(Sk(A[\tsample]), Sk(A[\underline 0]), \ldots, Sk(B), \mathsf{if}_1) > 0.5 \},
+\end{align*}
+which is measurable. Skeletal reduction sequences still contain just as much informationas reduction sequences of terms, however, on which reductions took place.
+\end{example}
 
@@
 \end{enumerate}
-The $\sim_c$-rules are illustrated in the following diagrams.
+The $\sim_c$-rules are illustrated in Figure \ref{sim-diagram}.
 \end{definition}
-
-\include{sim-diagram}
 
@@
 \paragraph{}
-Although the six cases in the definition of $\sim_c$ are defined separately, there are some commonalities worth noting.
-Each case (except a, which is symmetrical) has a right branch and a left branch (the branches of case b are unfortunately displayed the wrong way around in the diagram).
-The right branch has reduction steps at two positions, $\beta$ then $\alpha$, where $\beta > \alpha$ (i.e.~$\beta$ is inside $\alpha$), and in the left branch, the order of these reductions are swapped, so the reductions are at $\alpha$ then $\beta_1, \dots, \beta_n$.
-The position $\alpha$ of one of the reductions is unchanged, and the other reduction may have multiple (in cases e and f) or 0 (in cases b and e) images in the left branch.
-The images $(\beta_i)$ are all still inside (greater than or equal to) $\alpha$, and although they are not generally equal to $\beta$, the subskeletons at these positions have a similar shape to the subskeleton (initially) at position $\beta$, so that the reductions at these positions are the same type of reduction (e.g.~both $\beta$-reduction or both $\textsf{if}$-reduction).
-Case a is also similar, except that the relevant positions are all disjoint rather than contained in one another, and either branch could be considered the right branch.
+Before defining the new version of the reduction relation $\red$, the following lemma is necessary for it to be well-defined.
 
-\iffalse
-\changed[lo]{Since $L_s(M)$ is a countable set, there is a unique probability measure $\mu$ on the infinite product $\prod_{i \in L_s(M)} \Sigma_I$ of the Borel $\sigma$-algebras $\Sigma_I$ satisfying
-\[
-\mu\set{\omega \in \Omega \mid \omega_1 \in A_1, \cdots, \omega_n \in A_n} = \prod_{i=1}^n \Leb(A_i)
-\]
-where $\Leb$ is the Lebesgue measure on the measurable space $(I, \Sigma_I)$.}
-\fi
-
-\paragraph{}
-Before defining the new version of the reduction relation $\red$, a few lemmas about the properties of $\sim$ are necessary for it to be well-defined.
-
-\begin{lemma} \label{downAcrossUp}
-If $(N_1,\alpha_1) \sim^* (N_2,\alpha_2)$, there is some descendants $N'_1, N'_2$ of $N_1$ resp.~$N_2$, and a position $\alpha'$ in both of them such that $(N_1,\alpha_1) \sim_p^* (N'_1, \alpha') \sim_c^* (N'_2, \alpha') \sim_p^* (N_2, \alpha_2)$.
-\end{lemma}
-\begin{proof}
-  The $\sim_p$ relation can be split into $\sim_\downarrow \cup \sim_\uparrow$, where $(A,\alpha) \sim_\downarrow (B,\beta)$ if $A \to B$ and $(A,\alpha) \sim_p (B,\beta)$, and similarly $(A,\alpha) \sim_\uparrow (B,\beta)$ if $B \to A$ and $(A,\alpha) \sim_p (B,\beta)$.
-At each stage of this proof, it will be assumed that the $\sim_c$ steps and the $\sim_p$ steps in the sequence from $(N_1,\alpha_1)$ to $(N_2,\alpha_2)$ are rearranged such that there is never a $\sim_c$ immediately before a $\sim_\downarrow$, and there is never a $\sim_c$ immediately after a $\sim_\uparrow$.
-This rearrangement is always possible, because if there is some subsequence $(A,\alpha) \sim_c (B,\alpha) \sim_\downarrow (C,\beta)$, then there is an alternate path $(A,\alpha) \sim_\downarrow (B',\beta) \sim_c (C,\beta)$, where the reduction and the $\sim_\downarrow$ from $A$ to $B'$ are the same as those from $B$ to $C$, and the $\sim_c$ step is the same, but with the reduction sequences $O_1 \to^* O_1'$ and $O_2 \to^* O_2'$ in the definition of $\sim_c$ extended by the reduction $B \to C$.
-With these rearrangements assumed, it is sufficient to prove that the $\sim_p$ steps can be rearranged so that all of the $\sim_\downarrow$ steps come before all of the $\sim_\uparrow$ steps (possibly introducing some $\sim_c$ steps in the process).
-
-  The rearrangement to put all of the $\sim_\downarrow$s before all of the $\sim_\uparrow$s proceeds by induction on the number of $\sim_\downarrow$s that occur after the first occurrence of $\sim_\uparrow$.
-If it's 0, all of the $\sim_p$s are already in the correct order, and we're done.
-Otherwise, take the subsequence from the first $\sim_\uparrow$ to the first $\sim_\downarrow$ after that.
-This must be rearranged to some $\sim_\downarrow$s followed by some $\sim_\uparrow$s, then once that's done, the number of $\sim_\downarrow$s after the first $\sim_\uparrow$ in the overall sequence will have decreased by 1.
-
-  Let this subsequence be $A \sim_\uparrow^n B \sim_\downarrow C$.
-Suppose for induction that $A \sim_\uparrow^k (D,\delta) \sim_\downarrow^* (E,\epsilon) \sim_\uparrow^* C$ for some $D, \delta, E, \epsilon$, and $0 \leq k \leq n$, where in the reduction sequence $D \to^* E$, for any pair of reductions at positions which aren't disjoint, the reduction at the innermost (greater) position occurs first.
-Reduction sequences with this property will be called ``parallel'', as it is analagous to parallel reduction as in \akr{TODO: find this reference, for the proof of the Church-Rosser theorem by Tait and Martin-L\"of}.
-This induction is in reverse, with $k$ decreasing from $n$ to $0$.
-If $k = n$, simply set $(D,\delta) = B$ and $(E,\epsilon) = C$.
-If $k = 0$, then $A$ is related to $B$ by a sequence of $\sim_\downarrow$s then $\sim_\uparrow$s, as desired.
-In the intermediate steps, $k$ must be decreased by one, so a subsequence of one $\sim_\uparrow$ followed by a parallel sequence of $\sim_\downarrow$s must be replaced by a parallel sequence of $\sim_\downarrow$s followed by some $\sim_\uparrow$s.
-
-  For any individual $\sim_\uparrow$ then $\sim_\downarrow$ pair, let $(A,\alpha) \sim_\uparrow (B,\beta) \sim_\downarrow (C,\gamma)$.
-The skeleton $B$ reduces to both $A$ and $C$.
-If these are the same reduction, then $(A,\alpha) = (C,\gamma)$ and this subsequence may be removed.
-If they are different, then the way they overlap corresponds to one of the cases in the definition of $\sim_c$, either case a if the reduction positions are disjoint, or one of the cases b--f if one of the reduction positions is inside the other.
-In any case, there is a reduction sequence from each of $A$ and $C$ to some common skeleton $D$ ($O_1$ and $O_2$ in the definition of $\sim_c$).
-For the same reason that $(B,\beta) \sim_\downarrow (C,\gamma)$, there is some $\delta$ such that $(A,\alpha) \sim_\downarrow^* (D,\delta)$, and similarly for the same reason that $(B,\beta) \sim_\downarrow (A,\alpha)$, for the same $\delta$, $(C,\gamma) \sim_\downarrow^* (D,\delta)$.
-There are a lot of cases to check for this statement, but all of them are rather simple, and similar to each other.
-There is therefore an alternative $\sim^*$ sequence from $(A,\alpha)$ to $(C,\gamma)$, of the form $(A,\alpha) \sim_\downarrow^* (D_1,\delta) \sim_c (D_2,\delta) \sim_\uparrow^* (C,\gamma)$, where $D_1$ and $D_2$ are the alternative reduction sequences leading to the same skeleton $D$.
-There are only multiple $\sim_\downarrow$ steps in the result if the position of the reduction $B \to C$ is inside the position of the reduction $B \to A$, and similarly, there are only multiple $\sim_\uparrow$ steps in the result if the position of the reduction $B \to A$ is inside the position of the reduction $B \to C$.
-If there are multiple $\sim_\downarrow$ steps, they are not necessarily parallel, but in the only case where they aren't (case f of $sim_c$), the subsequence of $\sim_\downarrow$ steps with the reductions at $\alpha;\lambda;@_1;\beta$ then $\alpha;\lambda;@_1;\gamma_i;\tY;\lambda;\beta$ for each $i$ where $\gamma_i \geq \beta$ (in the notation of the definition of $\sim_c$) can be replaced by $\sim_\downarrow$s with the reductions at $\alpha;\lambda;@_1;\gamma'_i;\tY;\lambda;\beta$ then $\alpha;\lambda;@_1;\beta$, then another $\sim_c$, where $(\gamma'_i)$ is the list of positions in $A$ (\emph{before} the reduction at $\beta$ to $A'$, unlike $(\gamma_i)$) where $A | \gamma'_i = x$ and $\gamma'_i > \beta$.
-
-  The effect of this rearrangement is (ignoring the $\sim_c$s) to swap the order of the $\sim_\uparrow$ and the $\sim_\downarrow$, except that if the positions of one of these reductions is inside the other, that inner $\sim_p$ may be duplicated.
-If the positions are disjoint, both of them are unchanged (corresponding to case a of $\sim_c$), and if one is inside the other, then the outer position is unchanged and the other resultant positions, although they aren't equal to the original inner position, are still inside the outer position.
-Taking the sub-sequence of one $\sim_\uparrow$ followed by a parallel sequence of $\sim_\downarrow$s, this rearrangement can be repeatedly applied to move the $\sim_\uparrow$ further along the sequence.
-If at some point a $\sim_\uparrow$ and $\sim_\downarrow$ match (by relating the same (skeleton, position) pairs in opposite directions), they may be removed and the process may stop early.
-The $\sim_\uparrow$ may be duplicated if it passes a $\sim_\downarrow$ whose position is outside of the $\sim_\uparrow$'s position, but by the assumption that the sequence of $\sim_\downarrow$s is initially parallel, no position inside of this occurs later in the sequence, therefore all of the resultant $\sim_\uparrow$s pass the remaining $\sim_\downarrow$s without changing or duplicating them.
-The $\sim_\uparrow$s may be further duplicated, but the number of steps left in this process is bounded by the product, for all of the remaining $\sim_\downarrow$s, of $2 +$ the number of occurrences of the variable relevant to the reduction, because for each $\sim_\downarrow$, that is more than the number of duplicates it could produce for any $\sim_\uparrow$ that passes it.
-It is also possible (earlier) for some of the $\sim_\downarrow$s to be duplicated, but the only case where this process would not terminate, that both the $\sim_\downarrow$s and the $\sim_\uparrow$s continue being duplicated forever so that the number of switches left to do never decreases, is prevented by the parallelness condition, as all of the duplications of $\sim_\downarrow$s occur before all of the duplications of $\sim_\uparrow$s.
-
-  It still remains to be shown that the sequence of $\sim_\downarrow$s left at the end of this can be made parallel.
-The changes that may have occurred since the previous version of this sequence (which was known to be parrallel already), are that some of the $\sim_\downarrow$s may have been duplicated by passing the $\sim_\uparrow$, and if that $\sim_\uparrow$ matches one of the $\sim_\downarrow$s, that $\sim_\downarrow$ is removed.
-The position of the $\sim_\uparrow$'s reduction is the same for all of these duplications, because it is not changed until later in the sequence, when the $\sim_\downarrow$s with positions outside of that may occur.
-The resultant positions after a duplication are in the same position relative to each other, and therefore the order of the remaining $\sim_\downarrow$s is automatically parallel, except that (letting the position of the $\sim_\uparrow$'s reduction be $\alpha$) if there are $\sim_\downarrow$s with reductions at positions $\alpha;@_1;\lambda;\beta$ and $\alpha;@_2;\gamma$, or $\alpha;\tY;\lambda;\beta$ and $\alpha;\tY;\lambda;\gamma$, some of the resultant $\sim_\downarrow$s' positions may be inside each other where they were originally not.
-Similarly to the case where a $\tY$ reduction and a reduction at a position inside it are swapped though, the order can be fixed by swapping some of the $\sim_\downarrow$ steps with each other.
-In the first case, the reduction at position $\alpha;@_1;\lambda;\beta$ ends up at $\alpha;\beta$, and the reduction at position $\alpha;@_2;\gamma$ ends up at the positions $\alpha;\delta_i;\gamma$, where $(\delta_i)$ is the positions of the variable relevant to the $\sim_\uparrow$'s reduction inside its lambda.
-The positions of the variable in the lambda may be different before the reduction at $\beta$, but the positions of the redexes corresponding to the original redex at $\alpha;@_2;\gamma$ are changed similarly.
-Rather than reducing at $\alpha;\beta$ then at $\alpha;\delta_i;\gamma$, it is therefore possible to reach the same point by reducing at $\alpha;\delta'_i;\gamma$ then at $\alpha;\beta$, where $(\delta'_i)$ is the positions of the relevant variable within the lambda before the reduction at $\alpha;@_1;\lambda;\beta$ (and again, a $\sim_c$ of some sort must also be introduced).
-The other case, that the $\sim_\uparrow$'s reduction is a $\tY$-reduction, is similar.
-Each reduction which is duplicated has one image at $\alpha;\lambda;@_1;$its original relative position, and one for each occurrence of the variable.
-Those corresponding to variable occurrences may need to be moved to before those at $\lambda;@_1$.
-
-  In summary, to reach the desired order of $\sim_p$ steps, each $\sim_\downarrow$ may have to be moved past some $\sim_\uparrow$ steps, and it may get duplicated in the process, but the resultant sequence of $\sim_\downarrow$s can all be put in parallel order, and it has been shown that moving a $\sim_\uparrow$ past a sequence of $\sim_\downarrow$s is always possible if they're in parallel order, therefore the overall process terminates and reaches a point where all of the $\sim_\downarrow$s precede all of the $\sim_\uparrow$s, and all of the $\sim_c$s are in between them.
-All of these rearrangements leave the end points of the sequence unchanged, therefore this is an alternative sequence of $\sim$ steps between $(N_1,\alpha_1)$ and $(N_2,\alpha_2)$ of the form $(N_1,\alpha_1) \sim_\downarrow^* (N_1', \alpha') \sim_c^* (N_2', \alpha') \sim_\uparrow^* (N_2,\alpha_2)$.
-\end{proof}
-
-If $(X,\alpha) \sim_c (Y,\alpha)$ then the final skeletons in $X$ and $Y$ are equal, and $(X,\beta) \sim_c (Y,\beta)$ for any $\beta$ that occurs in this skeleton, therefore $\sim_c$ can be considered to apply to reduction sequences even without positions, $X \sim_c Y$ iff $(X,\cdot) \sim_c (Y,\cdot)$.
-The structure of $\sim_c$ can be seen more clearly by choosing a canonical example from each equivalence class.
-For this, the member of the class that's in call-by-value order is used.
-A reduction sequence is in call-by-value order if for every pair of reductions in the sequence at positions $\alpha$ and $\beta$, where the reduction at $\alpha$ happens first, either $\alpha \leq \beta$ or $\alpha$ and $\beta$ are disjoint, with $\alpha$ to the left of $\beta$ (i.e.~the first elements of the sequences $\alpha, \beta$ where they differ are $@_1$ and $@_2$, $\underline f_i$ and $\underline f_j$ for $i < j$, or $\textsf{if}_i$ and $\textsf{if}_j$ for $i < j$, in that order).
-Note that although the reductions that occur in the sequence are in call-by-value order, it may not be the actual call-by-value reduction sequence starting from that point because some redexes may remain un-reduced even when other reductions that should happen afterwards occur.
-It is only required that the reductions that do occur occur in the correct order.
-
-\begin{lemma} \label{canonicalCousins}
-For any reduction sequence $X$, there is a unique reduction sequence $X_{cbv}$ that is related to $X$ by $\sim_c$ and is in call-by-value order, and if $X \sim_c^* Y$, then $X_{cbv} = Y_{cbv}$.
-\end{lemma}
-
-\lo{As defined, $\sim_c$ is not symmetric.}
-
-\begin{proof}
-
-Define $X_n$ recursively so that $X_0 = X$ and if $X_n$ is not in CBV order, take the last pair of reductions in $X_n$ which aren't in CBV order.
-They are adjacent in the sequence, as the CBV order is a total order.
-Let the positions of these reductions be $\alpha$ and $\beta$ respecively.
-If they are disjoint, they can be considered as one of the branches of $\sim_c$ case a.
-Otherwise, $\beta < \alpha$, and the sub-skeleton at position $\beta$ at the appropriate point in the reduction sequence is a redex with non-trivial subterms, therefore it's of one of the forms $\tif{\skeletonPlaceholder>0}{A}{B}$, $(\lambda x.A) B$ or $\tY \lambda x.
-A$, therefore the reductions $\alpha$ and $\beta$ form the $O_2$ branch of one of the cases b, c, d, e or f of $\sim_c$.
-In either case, define $X_{n+1}$ as equal to $X_n$ except taking the other (left) branch, so that the order of the reductions at $\alpha$ and $\beta$ are switched (although the equivalent(s) of the reduction at $\alpha$ may not actually be at that position).
-
-This sequence of swaps rearranges the reductions in $X$ rather like insertion-sort (but in some cases duplicating or deleting the new element), and eventually reaches some $X_n$ that is in CBV order.
-This is defined to be $X_{cbv}$.
-
-\paragraph{}
-If $Y \sim_c X \sim_c^* X_{cbv}$, and the reductions involved in $Y \sim_c X$ are not the last pair of reductions in $Y$ that aren't in CBV order, the order of the $\sim_c$s can be rearranged so that the sequence $Y \sim_c^* X_{cbv}$ is the canonical sequence given above, therefore $Y_{cbv} = X_{cbv}$.
-To be more precise about this rearrangement, consider the sequence of $\sim_c$ steps from $Y$ to $X_{cbv}$.
-The first step takes some sub-sequence of the reduction sequence $Y$ and swaps the order of those reductions.
-If $X = X_{cbv}$, then $X = Y_1 = Y_{cbv}$ already.
-Otherwise, consider the cases according to whether the image of this subsequence in $X$ occurs earlier than the next pair of reductions to be swapped (i.e.~the last pair of reductions in $X$ that occur not in CBV order).
-If not, then either the step $Y \sim_c X$ changes this subsequence into CBV order, or out of it.
-If it is into, then it was the last out-of-order pair in $Y$ therefore the whole sequence $Y \sim_c^* X_{cbv}$ is already in the canonical order and $Y_{cbv} = X_{cbv}$.
-If it switches the pair to the wrong order, then the result of that switch is the last out-of-order pair in $X$ therefore $Y = X_1$, and the sequence $X_1 \sim_c^* X_{cbv}$ is just a suffix of the sequence $X \sim_c^* X_{cbv}$ therefore $(X_1)_{cbv} = X_{cbv}$ and again $Y_{cbv} = X_{cbv}$.
-
-The remaining cases are when the subsequence of reduction steps involved in $Y \sim_c X$ occur before the last out-of-order pair in $X$, but possibly overlapping.
-If there is no overlap, then the $\sim_c$ steps do not interfere with each other at all, and can simply be performed in the other order.
-This gives a different sequence of $\sim_c$ steps $Y \sim_c Y_1 \sim_c X_1 \sim_c^* X_{cbv}$.
-As $Y_{cbv} = (Y_1)_{cbv}$, we can proceed by induction in this case and use the fact that $(Y_1)_{cbv} = (X_1)_{cbv} = X_{cbv}$ to acheive the result.
-
-In the other case, there is some overlap between the regions involved in $Y \sim_c X$ and $X \sim_c X_1$, but the last reduction in $X \sim_c X_1$ is later than the last reduction in $Y \sim_c X$.
-There are only 2 reductions in $X$ involved in $X \sim_c X_1$, therefore the region involved in $Y \sim_c X$ ends on the first of the reductions involved in $X \sim_c X_1$.
-Now consider the specific sequence of positions of the reductions in $X$ involved in either of these steps.
-Let the last of these positions be $\alpha$, the first $\gamma$, and the second-last, third-last and so on $\beta_1, \beta_2, \dots$ respectively, and let the positions of the redexes in $Y$ involved in $Y \sim_c X$ be $\beta$ then $(\gamma_i)$ in order..
-Then $X \sim_c X_1$ swaps $\beta_1$ and $\alpha$, and $\alpha$ comes before $\beta_1$ in CBV order (i.e.~$\alpha < \beta_1$ or $\alpha$ is left of $\beta_1$), and the step $Y \sim_c X$ either proceeds forwards towards CBV order, swapping $\beta$ below $\gamma_1 = \gamma$ to form $\gamma$ then $(\beta_i)$, or proceeds backwards, taking $Y$ even farther from CBV order, swapping the $(\gamma_i)$ to above $\beta$, forming $\gamma$ then $\beta_1 = \beta$.
-In each of the cases below, it is required to show that the canonical sequence $(X_i)_i$ eventually reaches some term which is the same as one in $(Y_j)_j$, and from that point on, they match, therefore they reach the same end result: $X_{cbv} = Y_{cbv}$.
-
-\begin{itemize}
-  \item $Y \sim_c X$ in the backwards direction (so that $Y$ is closer to CBV order than $X$ is): By swapping the roles of $X$ and $Y$, and $\beta$ and $\gamma$, this is equivalent to the $Y \sim_c X$ forwards case.
-The assumption that $\alpha$ is before $\beta_1$ in CBV order (so that $\beta_1, \alpha$ is the pair to be swapped in $X \sim_c X_1$) maps to the assumption that $\alpha$ is before $\gamma_1$ (so that the result doesn't just follow trivially by $Y_1$ being equal to $X$) and vice-versa.
-Then one of the cases below establishes that $X_i = Y_j$ for some $i$ and $j$, therefore after swapping back, $Y_i = X_j$.
-\item $Y \sim_c X$ forwards and $\alpha$ comes after $\gamma$ in CBV order: In this case, the $\gamma$ and $\alpha$ reduction steps are already in the correct order in $Y$, therefore $\beta$ and $\gamma$ are already the last out-of-order pair of reductions in $Y$, $Y_1 = X$, and $Y \sim_c X \sim_c^* X_{cbv}$ is already the canonical sequence from $Y$ therefore $Y_{cbv} = X_{cbv}$.
-\item $Y \sim_c X$ forwards and $\alpha$ is disjoint from $\gamma$ and $\beta$: In this case, the sequence $(X_i)_i$ starts by swapping $\alpha$ and $\beta_1$ by case a, then because $\gamma$ is disjoint from $\alpha$ and comes after is in CBV order, $\gamma$ is right of $\alpha$, therefore either all the remaining $\beta_i$s, are $\geq \gamma$ (if $\beta$ and $\gamma$ aren't disjoint) or there are no remaining $\beta_i$s (if $\beta$ and $\gamma$ are disjoint, in which case they swap by case a and there is only $\beta_1$).
-The sequence $(X_i)_i$ therefore proceeds to swap all the remaining $\beta_i$s then $\gamma$, in order, below $\alpha$, resulting in $X_i$ for some $i$ having as its relevant sub-reduction-sequence $\alpha, \gamma, \beta_k, \dots, \beta_1$.
-The other sequence $(Y_j)_j$ starts by swapping $\gamma$ then $\beta$ below $\alpha$, then swapping $\gamma$ and $\beta$.
-Because $\alpha$ is disjoint from $\gamma$, the reduction there does not affect the sub-skeleton at position $\gamma$, therefore this last $\sim_c$ step proceeds identically to how it did in $Y \sim_c X$, resulting in $\gamma, \beta_k, \dots, \beta_i$.
-Overall, this results in $Y_3$ having the relevant sub-reduction-sequence $\alpha, \gamma, \beta_k, \dots, \beta_1$, therefore $Y_3 = X_i$ for the aforementioned value of $i$.
-\item $Y \sim_c X$ forwards, $\alpha \leq \gamma$ and $\alpha$ (and therefore also $\gamma$) is disjoint from $\beta$: In this case, $(X_i)_i$ proceeds by swapping $\beta_1 = \beta$ below $\alpha$ by case a, then swapping $\gamma$ below $\alpha$ by one of the other cases, and $(Y_j)_j$ proceeds by swapping $\gamma$ and $\alpha$, then swapping $\beta$ below $\alpha$ then all of the images of $\gamma$ (which are $\geq \alpha$ and therefore left of $\beta$).
-In both cases, the subterm at $\alpha$ when $\alpha$ and $\gamma$ are swapped is unaffected by the reduction at $\beta$, therefore it produces the same result in both cases, therefore the overall sequence in both cases is $\alpha$ then the images of $\gamma$ then $\beta$.
-\item Both $\gamma$ and $\beta$ are $> \alpha$ but they're disjoint from each other, and either the skeleton at position $\alpha$ is of the form $\tif{\skeletonPlaceholder > 0}{A}{B}$ or it is of the form $A B$ and either both $\gamma$ and $\beta$ are $> \alpha;@_1$ or they are both $> \alpha;@_2$: In this case, $(X_i)_i$ proceeds by swapping $\beta$ then $\gamma$ below $\alpha$, in each case possibly forming 0 or multiple images.
-The only way the number of images of each of these positions can be different is if the subskeleton at $\alpha$ is an $\textsf{if}$, and there are 0 of one of them and 1 of the other, in which case they're trivially in the correct order already.
-Otherwise, all of the images of $\beta$ and $\gamma$ are disjoint from one another, therefore none of the positions, or their relative order, change when they are swapped, therefore the next part of the sequence $(X_i)_i$ is just insertion-sort running on the images of $\beta$ and $\gamma$ with $\sim_c$ case a swaps until they're in the correct order.
-The sequence $(Y_j)_j$ starts by swapping $\gamma$ then $\beta$ below $\alpha$, then as before, sorting the images into the correct order by case a swaps.
-The images originally formed of $\gamma$ and $\beta$ are the same in both cases, therefore the final order is the same, so $X_i = Y_j$ for some $i$ and $j$.
-\item $Y \sim_c X$ forwards, $\alpha;@_1 < \gamma$, and $\alpha;@_2 < \beta$: Let $x$ be the variable involved in the $\beta$-reduction at $\alpha$, so that the sub-skeleton at $\alpha$ is $(\lambda x.
-A) B$ for some $A, B$, where $\gamma = \alpha;@_1;\lambda;\gamma'$ and $A \to A'$ at $\gamma'$, and $\beta = \alpha;@_2;\beta'$, and $B \to B'$ at $\beta'$.
-In this case, $(X_i)_i$ starts by swapping $\beta$ below $\alpha$ by case e, producing one image of $\beta$ for each $x$ in $A'$, then swapping $\gamma$ below $\alpha$ by case d, producing a single image $\alpha;\gamma'$.
-For each of the instances of $x$ in $A$ left of $\gamma'$, the reduction at $\alpha;\gamma'$ is then swapped below the corresponding image of $\beta$.
-The sequence $(Y_j)_j$ starts by swapping $\gamma$ below $\alpha$ then $\beta$ below $\alpha$, but in this case the images of $\beta$ produced may be different.
-In this case the swap happens earlier in the reduction sequence than $\gamma$, so that the body of the lambda is still $A$ rather than $A'$, and the reduction at $\gamma$ may rearrange or change the number of instances of $x$.
-Next, each of the images of $\beta$ to the right of $\alpha;\gamma'$ are swapped with it by case a.
-At this point, the next few reductions before $\alpha;\gamma'$ will in general be images of $\beta$ at positions inside $\alpha;\gamma'$.
-The subskeleton at position $\alpha;\gamma'$ before these reductions is $(A|\gamma')[B/x]$, then by the images of $\beta$ this reduces to $(A|\gamma')[B'/x]$, then it reduces at its root position ($\alpha;\gamma'$ in the overall skeleton) to $(A'|\gamma')[B'/x]$.
-For each position of an $x$ in $(A|\gamma')$, there are 0 or more corresponding positions of $x$ in $(A'|\gamma')$, depending on what type of reduction $A \to A'$ is.
-Because $B$ and $B'$ cannot contain any instances of the variable involved in the reduction $A \to A'$ (if it's a $\beta$-reduction or $\tY$-reduction), each instance of $B$ in $(A|\gamma')[B/x]$ has the same set of images in $(A'|\gamma')[B/x]$ as the corresponding $x$ in $(A|\gamma')$.
-When these images of $\beta$ that overlap with the image of $\gamma$ are swapped below it, each of them has its own images, one at each position of $B$ in $(A'|\gamma')[B/x]$ corresponding to its original copy of $B$ in $(A|\gamma')[B/x]$.
-After all of them are swapped below $\alpha;\gamma'$ (and rearranged among themselved by case a), there is therefore one reduction at each copy of $B$ in $(A'|\gamma')[B/x]$, i.e.~one at position $\beta'$ relative to each $x$ in $(A'|\gamma')$.
-They are therefore the same as the images of $\beta$ in $X_1$ that overlap with $\alpha;\gamma'$, because those are also one reduction at position $\beta'$ relative to each $x$ in $(A'|\gamma)$.
-Both $(X_i)_i$ and $(Y_j)_j$ therefore eventually reach a point where the relevant sub-reduction-sequence is $\alpha$, the images of $\beta$ to the left of $\alpha;\gamma'$, $\alpha;\gamma'$ itself, the aforementioned images of $\beta$ that overlap with $\alpha;\gamma'$, then all the images of $\beta$ to the right of $\alpha;\gamma'$.
-\item $Y \sim_c X$ forwards, $\gamma$ and $\beta$ are disjoint and both $> \alpha$, and the reduction at $\alpha$ is a $\tY$-reduction: This is similar to the previous case, but more complicated, so it will need some definitions to explain properly what's going on.
-Let the term at $\alpha$ before any of the reductions be $\tY \lambda x.
-A$, let $\gamma = \alpha;\tY;\lambda;\gamma'$, let $\beta = \alpha;\tY;\lambda;\beta'$, let $A | \beta' = B \to B'$ with the reduction at $\cdot$, let $A | \gamma' = C \to C'$ with the reduction at $\cdot$, let the positions in $A$ where $x$ occurs that are left of $\gamma'$, between $\gamma'$ and $\beta'$ (but still disjoint from both), and right of $\beta'$ respectively be $(\delta^l_i)$, $(\delta^m_i)$ and $(\delta^r_i)$, let the positions in $B$, $C$, $B'$ and $C'$ respectively where $x$ occurs be $(\delta^B_i)$, $(\delta^C_i)$, $(\delta^{B'}_i)$ and $(\delta^{C'}_i)$ (so that all the positions in $A$ where $x$ occurs, in left-to-right order, are $(\delta^l_i)_i, (\delta^C_i)_i, (\delta^m_i)_i, (\delta^B_i)_i, (\delta^r_i)_i$).
-The sequence $(X_i)_i$ starts by swapping $\beta$ then $\gamma$ below $\alpha$.
-With the shorthand that $\gamma'(\epsilon) = \alpha;\lambda;@_1;\epsilon;\tY;\lambda;\gamma'$, $\gamma'_0 = \alpha;\lambda;@_1;\gamma'$, and similarly for $\beta'$, the relevant portion of $X_2$ is then $\alpha$, $(\gamma'(\delta^l_i))_i$, $\gamma'_0$, $(\gamma'(\delta^{C'}_i))_i$, $(\gamma'(\delta^m_i))_i$, $(\gamma'(\delta^B_i))_i$, $(\gamma'(\delta^r_i))_i$, $(\beta'(\delta^l_i))_i$, $(\beta'(\delta^{C'}_i))_i$, $(\beta'(\delta^m_i))_i$, $\beta'_0$, $(\beta'(\delta^{B'}_i))_i$, $(\beta'(\delta^r_i))_i$.
-Next in $(X_k)_k$, for each $i$, $\gamma'(\delta^r_i)$ are swapped past all the images of $\beta$ until it is immediately before $\beta'(\delta^r_i)$ by case a, then for each $i$, $\gamma'(\delta^B_i)$ is swapped past all the images of $\beta$ until $\beta'_0$ by case a, then swapped with $\beta'_0$ by one of the other cases depending on what type of reduction $B \to B'$ is and the relative positions.
-Even if it's case d or f, all of the images of $\gamma'(\delta^B_i)$ are disjoint from each other (and from the positions of the other reductions after $\beta'_0$) because the redex, $C$, doesn't contain any instances of the variable involved in the reduction $B \to B'$.
-Expanding the definitions, this is swapping $\alpha;\lambda;@_1;\delta^B_i;\tY;\lambda;\gamma'$ with $\alpha;\lambda;@_1;\beta'$ at a point in the reduction sequence where the subskeleton at $\alpha;\lambda;@_1;\beta'$ is $B$ with a mixture of $\tY \lambda x.
-A$ and $\tY \lambda x.
-A[C'/\gamma']$ substituted for its occurrences of $x$, and one of these occurrences of $x$ is at $\delta^B_i$.
-For each $\delta^B_j$, there are some corresponding $\delta^{B'}_k$, where the images of the $x$ at $\delta^B_j$ end up after the reduction $B \to B'$.
-These cover all the $\delta^{B'}_k$, and uniquely, and for each $\delta^B_i$, the images of $\zeta;\delta^B_i;\theta$ after swapping it with $\zeta;\beta'$ are precisely $(\zeta;\delta^{B'}_k;\theta)$ for those same values of $k$, therefore after swapping all of the $(\gamma'(\delta^B_i))_i$ down past $\beta'_0$, their images are $(\gamma'(\delta^{B'}_i))_i$.
-After swapping with $\beta'_0$, each of these images than swaps by case a to take its place among the $(\beta'(\delta^{B'}_i))_i$.
-Next up in $(X_i)_i$, the $(\gamma'(\delta^m_i))_i$s and the $(\gamma'(\delta^{C'}_i))_i$s swap down past some images of $\beta$ they're disjoint from to take their place before their matching image of $\beta$, then $\gamma'_0$ swaps past the $(\beta'(\delta^l_i))_i$, stopping immediately before $\beta'(\delta^{C'}_1)$, then the $(\gamma'(\delta^l_i))_i$ and the $(\beta'(\delta^l_i))_i$ mix.
-The end result of the rearrangement of (this subsequence of) the reduction sequence is therefore $\alpha$, $(\gamma'(\delta^l_i), \beta'(\delta^l_i))_i$, $\gamma'_0$, $(\gamma'(\delta^{C'}_i), \beta'(\delta^{C'}_i))_i$, $(\gamma'(\delta^m_i), \beta'(\delta^m_i))_i$, $\beta'_0$, $(\gamma'(\delta^{B'}_i), \beta'(\delta^{B'}_i))_i$, $(\gamma'(\delta^r_i), \beta'(\delta^r_i))_i$.
-
-  The other sequence, $(Y_j)_j$, proceeds similarly, but with the images of $\gamma$ starting out after the images of $\beta$.
-It swaps all the images of $\beta$ that are right of $\gamma_0$ to the appropriate places by case a, then swaps all the $(\beta'(\delta^C_i)_i$ first to then past $\gamma'_0$, resulting in $(\beta'(\delta^{C'}_i))_i$ for the same reason as with swapping those images of $\gamma$ that overlapped with $\beta'_0$ past it in the $(X_i)_i$ case above, then finally swapping the $(\beta'(\delta^l_i))_i$s and the $(\gamma'(\delta^l_i))_i$s into the correct order.
-As required, this produces the same result as in $(X_i)_i$.
-  \item $Y \to X$ forwards, and $\alpha < \gamma < \beta$: Let $\gamma'$ be such that $\gamma = \alpha;\textsf{if}_i;\gamma'$, $\alpha;@_1;\lambda;\gamma'$, $\alpha;@_2;\gamma'$ or $\alpha;\tY;\lambda;\gamma'$, so that $\gamma'$ is the freely varying later part of $\gamma$ as in the definition of $\sim_c$, whichever case applies to swapping $\gamma$ and $\alpha$.
-Similarly, let $\beta'$ be the freely varying part of $\beta$ relative to $\gamma$.
-Let the images of $\gamma$ after swapping it with $\alpha$ (where this swap occurs later in the reduction sequence than $\beta$) then be $(\alpha;\delta_i;\gamma')_i$, and the images of $\beta$ after swapping it with $\gamma$ be $(\gamma;\epsilon_i;\beta')_i$.
-Depending on the skeleton at $\alpha$, and $\gamma$'s position within it, $(\delta_i)_i$ may be empty (case b), a singleton containing $\cdot$ (cases c and d), all the positions of the relevant variable in the lambda in the skeleton at $\alpha$ (case e), or $\lambda;@_1$ and $\lambda;@_1;\zeta;\tY;\lambda$ for each position $\zeta$ of the relevant variable (case f), but in all cases (except a, which is excluded because none of $\alpha, \gamma$ and $\beta$ are disjoint) the set of images has this general structure.
-Furthermore, the positions $(\delta_i)_i$ are determined only by the general position of $\gamma$ within $\alpha$ (the part excluded from $\gamma'$), and the positions of the relevant variable within the skeleton at $\alpha$.
-They do not depend on $\alpha$ or $\gamma$, so that if both initial positions were moved somewhere else the relative positions of the images of $\gamma$ would be unaffected, and similarly if some position within $\gamma$ were swapped with $\alpha$, the relative positions of its images would be the same (except to the extent that, in case f, the positions of the relevant variable are taken after the inner reduction takes place, so that some of those may still change).
-Because there are so many of them and they don't actually affect the multiset of positions where reductions take place, I will be ignoring swaps by case a in the proof of this case, and just assuming that all the final positions end up in the correct order.
-
-  The final set of reduction positions that I will be proving both $(X_i)_i$ and $(Y_j)_j$ reach is $\alpha, (\alpha;\delta_i;\gamma', (\alpha;\delta_i;\gamma';\epsilon_j;\beta')_j)_i$, where as usual the sequences are expanded out to the full list.
-For the sequence $(X_i)_i$, first the images of $\beta$ in $X_0$ are $(\gamma,\epsilon_i,\beta')$.
-For each of these in turn, it is swapped with $\alpha$.
-By the fact mentioned above that sub-positions are mapped to the same set of images except in case f, the images of $\gamma,\epsilon_j,\beta'$ after this swap are $(\alpha;\delta_i^j;\gamma',\epsilon_j,\beta')_i$, where $(\delta_i^j)_i$ is the same as $(\delta_i)_i$ except that in the case where this is not the first image of $\beta$ swapped below $\alpha$ and the reduction at $\alpha$ is a $\tY$-reduction (and therefore swaps with it proceed by case f), the positions of the relevant variable are taken after the reductions at $\gamma$ and $\gamma;\epsilon_k;\beta'$ for $k \leq j$ (this implies that the first of these, $\delta_i^n = \delta_i$).
-All the images $\alpha;\delta_i^j;\gamma';\epsilon_j;\beta'$ where $\delta_i^j > \lambda;@_1;\gamma';\epsilon_{j+1};\beta'$ are then swapped with $\alpha;\lambda;@_1;\gamma';\epsilon_{j+1};\beta'$ (which is $\alpha;\delta_i^{j+1};\gamma';\epsilon_{j+1};\beta'$ for some $i$).
-As in the case where $\alpha;\tY < \gamma, \beta$ and $\gamma$ and $\beta$ are disjoint, the effects of these swaps may duplicate or delete the reductions in such a way that all the remaining images of $\gamma;\epsilon_j;\beta'$ are $(\alpha;\delta_i^{j+1};\gamma';\epsilon_j;\beta'$.
-This is repeated with $\delta_i^{j+2}$ and so on until these images are all in the correct order, at which point they are $(\alpha;\delta_i;\gamma';\epsilon_j;\beta')_i$.
-After all of these are done, $\gamma$ is swapped with $\alpha$, yielding $(\alpha;\delta_i^{-1};\gamma')_i$.
-As with the images of $\beta$, the subset of these which are inside $\alpha;\lambda;@_1;\gamma';\epsilon_j;\beta'$ for each $j$ in turn are swapped with it, until they are all in the correct order and the images left of $\gamma$ are $(\alpha;\delta_i;\gamma')_i$.
-At this point, $X_i$ has the desired value.
-
-  The sequence $(Y_j)_j$ proceeds similarly.
-First $\gamma$ swaps with $\alpha$, yielding $\alpha$ and $(\alpha;\delta_i;\gamma')_i$, then $\beta$ swaps with $\alpha$, yielding one image of $\beta$ for each $\delta_i$, except that, in case f, those images at positions inside $\alpha;\lambda;@_1;\gamma'$ may differ because they are earlier in the reduction sequence than any of the images of $\gamma$.
-Each of the images of $\beta$ in turn it moved to the corresponding image of $\gamma$, except that those overlapping with $\alpha;\lambda;@_1;\gamma'$ may be duplicated or deleted on the way.
-The images of $\beta$ after just this process (which does not actually correspond to any term in $(Y_j)_j$, but it's sufficiently independent that the order doesn't matter that much) are $(\alpha;\delta_i;\gamma';\beta'')$ (where $\gamma;\beta'' = \beta$).
-The images of these for each $i$, after swapping with $\alpha;\delta_i;\gamma'$, is $(\alpha;\delta_i;\gamma';\epsilon_j;\beta')_j$, because the reductions produced by a swap are unaffected by its overall location, and the skeleton at $\alpha;\delta_i;\gamma'$ at the relevant point in the sequence is equal to the skeleton at $\gamma$ initially, therefore this is equivalent to immediately swapping $\beta$ and $\gamma$ as in $Y \sim_c X$ (except that in the case where the reduction at $\alpha$ is a $\tY$-reduction, the skeleton at $\alpha;\lambda;@_1;\gamma'$ is not actually equal to the skeleton initially at $\gamma$ because something was substituted in for the variable bound at $\alpha;\tY$, but this doesn't affect the variable involved in the reduction at $\gamma$, so this doesn't actually matter).
-Therefore also in this case, eventually the set of reductions in the relevant portion of $Y_j$ is $\alpha, (\alpha;\delta_i;\gamma', (\alpha;\delta_i;\gamma';\epsilon_j;\beta')_j)_i$, therefore it is equal to some $X_i$.
-\end{itemize}
-In summary, if the swap in $Y \sim_c X$ doesn't overlap with the swap $X \sim_c X_0$, the sequence of swaps can be rearranged until it does, and in any other case, there is some $Y_j$ that's equal to some $X_i$, therefore these sequences reach the same end point and if $Y \sim_c X$, then $Y_{cbv} = X_{cbv}$, and by chaining these together, if $Y \sim_c^* X$, $Y_{cbv} = X_{cbv}$.
-
-\paragraph{}
-If $Y$ is in CBV order and $Y \sim_c^* X$, then $Y_{cbv} = X_{cbv}$ but also $Y_{cbv} = Y$ because the sequence $(Y_i)_i$ terminates immediately therefore $X_{cbv}$ is the unique reduction sequence in call-by-value order that's related to $X$ by $\sim_c^*$.
-\end{proof}
-
-\begin{lemma}
+\begin{restatable}{lemma}{lemmaSimMN}
 \label{lem:sim-M-N}
@@
 If $M \to N$, then $\sim^*_N$ is equal to the restriction of $\sim^*_M$ to $L_0(N)$.
-\end{lemma}
-
-\begin{proof}
-$\sim_N^*$ is trivially a subset of $\sim_M^*$ because $\sim_N$ is a subset of $\sim_M$.
-
-In the other direction, suppose $(N_1, \alpha_1) \sim_M^* (N_2, \alpha_2)$ where both $N_1$ and $N_2$ are descendants of $N$.
-By Lemma~\ref{downAcrossUp}, take $(N_1, \alpha_1) \sim_{p,M}^* (N'_1, \alpha') \sim_{c,M}^* (N'_2, \alpha') \sim_{p,M}^* (N_2, \alpha_2)$.
-The $\sim_p^*$ steps remain within $Rch(N)$, and $\sim_p$ does not depend on the history of the reduction sequences, therefore $(N_1, \alpha_1) \sim_{p,N}^* (N'_1, \alpha') \sim_{c,M}^* (N'_2, \alpha') \sim_{p,N}^* (N_2, \alpha_2)$.
-
-Let $X$ be the call-by-value reduction sequence related to both $N'_1$ and $N'_2$ by $\sim_{c,M}$ given by Lemma~\ref{canonicalCousins}. 
-As $M \to N$ is the first reduction in the sequence $N'_1$, it is the last to be affected by the $\sim_{c,M}$ sequence $N'_1 \sim_{c,M}^* X$ given by Lemma \ref{canonicalCousins} therefore it can be split into $N'_1 \sim_{c,M}^* Y_1 \sim_{c,M}^* X$ where $Y_1$ is in CBV order except possibly for its first reduction, which is still $M \to N$.
-Let the position of the reduction $M \to N$ be $\beta$.
-\akr{TODO: fill in the details that are missing here. It shouldn't be too hard, it's just a bit sketchy at present.} 
-The rearrangement of the reduction sequence $Y_1 \sim_{c,M} X$ consists of moving the reduction at $\beta$ down past the other reductions in $Y_1$, possibly duplicating or deleting it in the process, but not affecting the positions or the correct order for any of the other reductions.
-The reductions derived from $M \to N$ can be identified as follows:
-
-%\begin{definition}
-A position in some (reduction sequence of) term(s) is \emph{derived from} another if it is related by the reflexive transitive closure of $\rightsquigarrow$, where $(V,\gamma) \rightsquigarrow (W,\delta)$ just if $V \to W$ and one of the following cases holds:
-\begin{itemize}
-    \item $(V,\gamma) \sim_p (W,\delta)$
-    \item $\gamma = \delta$, $V \to W$ at $\epsilon$, and $\epsilon > \gamma$
-    \item $V \mid \epsilon = (\lambda x.U) Z, U | \zeta = x, \gamma = \epsilon;@_2;\theta$ and $\delta = \epsilon;\zeta;\theta$
-    \item $V \mid \epsilon = \tY (\lambda x.U), \gamma = \epsilon;\tY;\lambda;\zeta$ and $\delta = \epsilon;\lambda;@_1;\zeta$
-    \item $V \mid \epsilon = \tY (\lambda x.U), U | \zeta = x, \gamma = \epsilon;\theta$ and $\delta = \epsilon;\lambda;@_1;\zeta;\theta$
-\end{itemize}
-%\end{definition}
-
-Crucially, the reductions in $X$ can be partitioned into 2 sets: those at positions derived from $(M,\beta)$, and those at positions equal to the reductions in $Y_1$ (in the same order as they occur in $Y_1$).
-Using the same construction for $N'_2$ shows that the positions of the reductions in $Y_2$ are also the positions of the reductions in $X$ other than those derived from $(M,\beta)$, therefore $Y_1 = Y_2$ therefore $N'_1 \sim_{c,M} Y_1 \sim_{c,M} N'_2$ and every reduction sequence in this sequence starts with $M \to N$ therefore $N'_1 \sim_{c,N} N'_2$.
-
-Combining this with the $\sim_{p,N}^*$s at the beginning and end then yields the desired result that $(N_1,\alpha_1) \sim_N^* (N_2, \alpha_2)$, therefore the restriction of $\sim_M^*$ to $\sim_N$'s domain ($Rch(N)$) is a subset of $\sim_N^*$ therefore the two versions of $\sim$ match on this domain, as desired.
-\end{proof}
+\end{restatable}
 
@@
 \paragraph{}
-\begin{lemma} \label{churchRosser}
+This reduction relation now has all of the properties required of it.
+In particular, it can be considered an extension of the usual linear trace semantics (as will be seen later in Theorem \ref{thm:AstEquivalence}), and also:
+\begin{restatable}{lemma}{churchRosser} \label{churchRosser}
 The relation $\Rightarrow$ is Church-Rosser.
-\end{lemma}
-\begin{proof}
-Suppose that $(M,s) \Rightarrow^* (M_1,s_1)$ and also $(M,s) \Rightarrow^* (M_2,s_2)$, then it is required to prove that there is some $(M',s')$ such that both $(M_1,s_1) \Rightarrow^* (M',s')$ and $(M_2,s_2) \Rightarrow^* (M',s')$.
-First consider the special case where $(M,s) \Rightarrow (M_1,s_1)$ and $(M,s) \Rightarrow (M_2,s_2)$, with only a single step in each case.
-Let the positions of the redexes in $M \to M_1$ and $M \to M_2$ be $\alpha_1$ and $\alpha_2$ respectively.
-
-First consider the case that $\alpha_1$ and $\alpha_2$ are disjoint.
-Let $M_1 = M[X_1/\alpha_1]$ and similarly for $X_2$, then let $M' = M[X_1/\alpha_1][X_2/\alpha_2]$.
-As the positions are disjoint, the substitutions commute and both $M_1$ and $M_2$ reduce (with $\to$) to $M'$.
-Let $s' = s \circ i(M \to M_1) \circ i(M_1 \to M')$. 
-The injection $i(M \to M_1) \circ i(M_1 \to M')$ consists of prepending $M \to M_1 \to M'$ to each reduction sequence, but by case a of $\sim_c$, this is equivalent to prepending $M \to M_2 \to M'$, which is $i(M \to M_2) \circ i(M_2 \to M')$.
-In the case that the reduction $M \to M_2$ isn't a $\tsample$-reduction, this is enough to establish that $(M_1, s_1) \Rightarrow (M', s')$ (and similarly if the redex of $M \to M_1$ isn't $\tsample$, $(M_2,s_2) \Rightarrow (M',s')$).
-If it is $\tsample$ though, in order for it to be the case that $(M_1, s_1) \Rightarrow (M', s')$, it is additionally necessary that $X_2$, the result of the reduction at $\alpha_2$, be $s_1(M_1, \alpha_2)$, which follows from the fact that $(M, \alpha_2) \sim (M_1, \alpha_2)$ by case 1 of $\sim_p$.
-The case that the reduction $M \to M_1$ is a $\tsample$-reduction is similar.
-
-The case that $\alpha_1 = \alpha_2$ is trivial, because there is at most one possible $\Rightarrow$ reduction at any given position, therefore $(M_1, s_1) = (M_2, s_2)$ already.
-
-The remaining case is that $\alpha_1 < \alpha_2$ or $\alpha_1 > \alpha_2$. 
-Assume without loss of generality that $\alpha_1 < \alpha_2$. 
-For each possible case of what type of redex $M | \alpha_1$ is, and $\alpha_2$'s position within it, there is a corresponding case of $\sim_c$, and similarly to the case where $\alpha_1$ and $\alpha_2$ are disjoint, the term $O_1 = O_2$ from the definition of $\sim_c$ is a suitable value of $M'$.
-The term $M \mid \alpha_1$ can't be $\tsample$, because it has strict subterms, but the case that $M \mid \alpha_2 = \tsample$ is still somewhat more complicated. 
-$\alpha_2$ can't be within $\alpha_1 ; @_2$ or $\alpha_1 ; Y$ (cases e and f of $\sim_c$) because those are not valid positions to reduce a $\tsample$, 
-and the case that $\alpha_2$ is within the branch of an if statement that is deleted by the reduction at $\alpha_1$ (case b) doesn't actually present a problem because there is no corresponding reduction to $M \to M_2$ in the other branch,
-which leaves cases c and d, that $\alpha_2$ is in the $\tif{\cdot}{\cdot}{\cdot}$ branch that isn't deleted, and that $\alpha_2 < \alpha_1 ; @_1 ; \lambda$.  
-The values that the $\tsample$s take in these cases match because $(M,\alpha) \sim_p (M_1, \alpha;\beta)$ by cases 3 and 2 of $\sim_p$ respectively.
-
-\paragraph{}
-In the case that $(M,s) \Rightarrow^* (M_1,s_1)$ or $(M,s) \Rightarrow^* (M_2,s_2)$ by multiple steps, it is possible to repeatedly replace a pair of the form $A \Leftarrow B \Rightarrow C$ by $A \Rightarrow^* D \Leftarrow^* C$ by the construction above, but it is not immediate that this process terminates, because each $\Rightarrow$ or $\Leftarrow$ may be replaced by multiple, so the sequence of $\Rightarrow$s and $\Leftarrow$s from $(M_1,s_1)$ to $(M_2,s_2)$ may get longer at some steps.
-However, termination can be proved by noting that the structure of this process is identical to the process of swapping $\sim_\uparrow$ and $\sim_\downarrow$ steps in Lemma \ref{downAcrossUp}.
-To be more precise, in this case there is a sequence of $\Leftarrow$ and $\Rightarrow$ steps, each of which has an associated reduction position and initial and final terms, and if a $\Leftarrow$ immediately precedes a $\Rightarrow$, they may be swapped to produce some number of $\Rightarrow$s, followed by some number of $\Leftarrow$s.
-In the case of Lemma \ref{downAcrossUp}, there is a sequence of $\sim_\uparrow$ and $\sim_\downarrow$ steps, each of which has an associated reduction position and initial and final skeletons, and if a $\sim_\uparrow$ immediately precedes a $\sim_\downarrow$, they may be swapped to produce some number of $\sim_\downarrow$s followed by some number of $\sim_\uparrow$s.
-In both cases, the number and reduction positions of the resultant steps are determined by the case of $\sim_c$ that matches the way the initial reduction positions overlap, and the initial skeleton (or the skeleton of the initial term).
-The same argument that the process in Lemma \ref{downAcrossUp} terminated is therefore applicable here too.
-At every stage, the sequence of reductions that results from one of the initial $\Rightarrow$s is a parallel sequence, and swapping a parallel sequence of $\Rightarrow$s with a $\Leftarrow$ always terminates, therefore each $\Rightarrow$ in turn can be moved past all of the $\Leftarrow$s, and the process as a whole will terminate in a state where all of the $\Rightarrow$s precede all of the $\Leftarrow$s, i.e.~a pair of reduction sequences $(M_1,s_1) \Rightarrow^* (M',s') \Leftarrow^* (M_2,s_2)$.
-\end{proof}
-
-\begin{lemma} \label{simpDoesn'tMix}
-If $A$ is some descendant of $M$ and $(A,\gamma) \sim^* (M,\delta)$, then $(A,\gamma) \sim_p^* (M,\delta)$, with the length of the reduction sequences decreasing by one each step from $A$ to $M$.
-\end{lemma}
-\begin{proof}
-This is a simple induction on $\sim^*$.
-In the base case, $(A,\gamma) = (M,\delta)$ therefore $(A,\gamma) \sim_p^* (M,\delta)$ trivially.
-Otherwise, suppose that $(M,\delta) \sim_p^* (B,\epsilon) \sim (A,\gamma)$.
-Either $(B,\epsilon) \sim_p (A,\gamma)$ or $(B,\epsilon) \sim_c (A,\gamma)$.
-In the first case, either $B \to A$, in which case the result follows directly, or $A \to B$, in which case the fact that each (reduction sequence of) skeleton(s) has only one parent implies that a $(M,\delta) \sim_p^* (A,\gamma)$ directly as a subsequence of the path to $(B,\epsilon)$.
-
-In the $\sim_c$ case, consider the definition of $\sim_c$.
-Either $O'_1 = A$ and $O'_2 = B$ or vice-versa.
-As $M \to^* N \to^* O_1 \to^* B$ and $(M,\delta) \sim_p^* (B,\epsilon)$, and $\sim_p$ only relates positions in a term and its parent, there are some positions $\zeta, \theta$ such that $(M,\delta) \sim_p^* (N,\zeta) \sim_p^* (O_1, \theta) \sim_p^* (B,\epsilon)$.
-It follows that $(O_2, \theta) \sim_p^* (A,\gamma)$ by following the same path, therefore it suffices to provide the only missing portion of the path from $M$ to $A$, i.e.~to prove that $(N,\zeta) \sim_p^* (O_2,\theta)$ given $(N,\zeta) \sim_p^* (O_1,\theta)$ (or vice-versa).
-
-If $\zeta$ is disjoint from all the positions of reduction from $N$ to $O_1$ and $O_2$ (and consequently $\zeta = \theta$), this follows from case 1 of $\sim_p$.
-Otherwise, this can be proved by taking cases from the definition of $\sim_c$.
-This is rather long, but all of the cases are similar.
-The general idea is that the reductions from $N$ to $O_1$ correspond to the reductions from $N$ to $O_2$, so that if a position is related by $\sim_p$ across that reduction, it is related in the other branch for the same reason.
-Case d, where $B = O'_1$ rather than $O'_2$, is given here in more detail as an illustrative example:
-
-Let $I$ be $N$ reduced at $\alpha$, and $J$ be $N$ reduced at $\alpha;@_1;\lambda;\beta$, so that $N \to I \to O_1$ and $N \to J \to O_2$.
-All of the reduction positions are $\geq \alpha$, and $\zeta$ is not disjoint from all of them, therefore $\zeta$ is not disjoint from $\alpha$.
-Let $\iota$ be the position such that $(N,\zeta) \sim_p (I,\iota) \sim_p (O_1,\theta)$.
-The fact that $(N,\zeta) \sim_p (I,\iota)$ implies that $\zeta > \alpha;@_1;\lambda$.
-Let $\zeta = \alpha;@_1;\lambda;\kappa$ and $\theta = \alpha;\kappa'$.
-If $\kappa$ is disjoint from $\beta$, then $(N,\zeta) \sim_p (J,\zeta) \sim_p (O_2,\alpha;\kappa') = (O_2;\theta)$ by cases 1 and 2 of $\sim_p$.
-In the other case, that $\kappa$ isn't disjoint from $\beta$, $\kappa > \beta$ because none of the positions $<= \alpha;@_1;\lambda;\beta$ in $N$ are related to any position in $I$ by $\sim_p$.
-As $(I,\alpha;\kappa) \sim_p (O_1,\alpha;\kappa')$ (with the redex at $\alpha;\beta$), for exactly the same reason $(N,\alpha;@_1;\lambda;\kappa) \sim_p (J,\alpha;@_1;\lambda;\kappa')$ (with the redex at $\alpha;@_1;\lambda;\beta$).
-Because $(N,\alpha;@_1;\lambda;\kappa) \sim_p (J,\alpha;@_1;\lambda;\kappa')$, $N|\alpha;@_1;\lambda;\kappa = J|\alpha;@_1;\lambda;\kappa'$, and $N|\alpha;@_1;\lambda;\kappa \neq $ the variable of $N|\alpha;@_1$, therefore $J|\alpha;@_1;\lambda;\kappa'$ is also not the variable therefore $(J,\alpha;@_1;\lambda;\kappa') \sim_p (O_2, \alpha;\kappa') = (O_2, \theta)$ by case 2 of $\sim_p$.
-Combining these results, $(N,\zeta) \sim_p^* (O_2;\theta)$ as desired.
-\end{proof}
-
-\begin{lemma} \label{lem:redexDestroyed}
-If $M \to N$, with the redex at position $\alpha$, then no position in any term reachable from $N$ is related by $\sim^*$ to $(M,\alpha)$.
-\end{lemma}
-\begin{proof}
-Suppose on the contrary that $(M,\alpha) \sim^* (A,\beta)$, where $N \to^* A$, then by Lemma \ref{simpDoesn'tMix}, $(M,\alpha) \sim_p^* (A,\beta)$.
-$M \neq A$ therefore $(M,\alpha) \sim_p$ some position in $N$, but in all the cases of the definition of $\sim_p$, no position in the child term is related to the position of the redex.
-\end{proof}
+\end{restatable}
 
@@
 Using a reduction strategy $f$, a subset of $\Rightarrow$ that isn't nondeterministic, $\Rightarrow_f$, can be defined by $(N,s) \Rightarrow_f (N',s')$ just if $(N,s) \Rightarrow (N',s')$ and $N$ reduces to $N'$ with the redex at $f(N)$.
-
-Essentially what Lemma~\ref{lem:redexDestroyed} demonstrates is that the samples taken during any reduction sequence are independent of each other. This is made more precise in the following lemmas.
-
-\begin{lemma}
-For any skeletons $M \to N$, with the redex at position $\alpha$, and measurable set of samples $S \subset I^{L_s(N)}$, $\mu(S) = \mu(\{s \in I^{L_s(M)} \mid s \circ i(M \to N) \in S\})$, and furthermore, if $M | \alpha = \tsample$, for any $S \subset I \times I^{L_s(N)}$, $\mu(S) = \mu(\{s \in I^{L_s(M)} \mid (s(M,\alpha), s \circ i(M \to N)) \in S\})$.
-\end{lemma}
-\begin{proof}
-If the result holds for all sets $S$ of the form $\{s \in I^{L_s(N)} \mid \forall j : s(j) \in x_i\}$, where $(x_j)_{j \in J}$ is a family of measurable subsets of $I$ indexed by some finite set $J \subset L_s(N)$ of positions, it also holds for all other $S$ by taking limits and disjoint unions.
-
-Take such a $(x_j)_{j \in J}$, where $x_j = x_k$ and define $K = i(M \to N)[J]$.
-Because $i(M \to N)$ is injective, it defines a bijection between $J$ and $K$.
-We can then calculate the measure $\mu(\{s \in I^{L_s(M)} \mid s \circ i(M \to N) \in S\}) = \mu(\{s \in I^{L_s(M)} \mid \forall k : s(k) \in x_{i(M \to N)^{-1}(k)}\}) = \prod_{k \in K} \mu_I(x_{i(M \to N)^{-1}(k)}) = \prod_{j \in J} \mu_I(x_j) = \mu(S)$.
-
-In the case that $M|\alpha = \tsample$, we can similarly consider only those sets $S$ of the form $x_\tsample \times \{s \in I^{L_s(N)} \mid \forall j : s(j) \in x_i\}$.
-The position $\alpha$ in $M$ is not related to any position in $L_s(N)$, therefore $(M,\alpha) \not \in K$.
-Again, we can calculate the measure $\mu(\{s \in I^{L_s(M)} \mid (s(M,\alpha), s \circ i(M \to N)) \in S\}) = \mu(\{s \in I^{L_s(M)} \mid s(M,\alpha) \in x_\tsample, \forall k : s(k) \in x_{i(M \to N)^{-1}(k)}\}) = \mu_I(x_\tsample) \prod_{k \in K} \mu_I(x_{i(M \to N)^{-1}(k)}) = \mu_I(x_\tsample) \prod_{j \in J} \mu_I(x_j) = \mu(S)$.
-\end{proof}
-
-\begin{lemma} \label{lem:independentSamples}
-For any initial term $M$, reduction strategy $f$ on $M$, natural number $n$, skeleton $N$ with $k$ holes, measurable set $T \subset \mathbb R ^ k$ and measurable set $S$ of samples in $I^{L_s(N)}$, $\mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')\}) = \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in I^{L_s(N)} : (M,s) \Rightarrow_f^n (N[r], s')\}) \mu(S)$
-\end{lemma}
-\begin{proof}
-Suppose, to begin with, that $n = 0$.
-Either $\exists r \in T : N[r] = M$, in which case both sides of the equation are $\mu(S)$, or there is no such $r$, in which case both sides of the equation are 0.
-
-For $n > 0$, suppose for induction that the lemma is true for $n - 1$, for all $N, T$ and $S$.
-If $\exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')$, the $r$ and $s'$ are necessarily unique, therefore we may assume that $T$ is the product of $k$ measurable setsets of $\mathbb R$, $(T_j)_{0 \leq j < k}$, as all of the other cases follow by taking unions and limits of these rectangles.
-Let\begin{align*}
-R_s & = \{Sk(O) \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) = \tsample, O \to N[r]\} \\
-R_d & = \{Sk(O) \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) \neq \tsample, O \to N[r]\}.
-\end{align*}
-Together, $R_s$ and $R_d$ contain all of the skeletons of terms that can reduce to $N$, and they're disjoint.
-For each of these skeletons, there is a map $g_O$ from the reals in $O$ to the reals in $N$, i.e.
-$g_O(r) = r'$ if $O[r] \to N[r']$ for $O \in R_d$, and $g_O(s, r) = r'$ if $O[r] \to N[r']$ with $s$ as the value the sample takes in the reduction.
-All of these functions are measurable, as they are just rearrangements of vector components, or in the case that $O|f(O) = \underline x(\skeletonPlaceholder, \dots, \skeletonPlaceholder)$, a combination fo the measurable function $x$ with a rearrangement of vector components.
-In the case that $O \in R_s$, the function $g_O$ simply inserts the sample into the vector of reals at a certain index, call it $h_O$.
-\clearpage
-\begin{align*}
-& \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')\}) \\
-= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in S : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\})\\
-    + & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in S : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}:\\
-    & g_O(r,s'(O,f(O))) \in T, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
-     + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}:\\
-     & g_O(r) \in T, s' \circ i(O[r] \to N[g_O(r)]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)}:\\
-    & s'(O,f(O)) \in T_{h(O)}, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
-    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}:\\
-    & g_O(r) \in T, s' \circ i(O[r] \to N[g_O(r)]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
-    & \mu(\{s' \in I^{L_s(O)} \mid s'(O,f(O)) \in T_{h(O)}, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S\})\\
-    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
-    & \mu({s' \in I^{L_s(O)} \mid s' \circ i(O[r] \to N[g_O(r)]) \in S}) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(T_{h(O)}) \mu(S)\\
-    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : s'(O,f(O)) \in T_{h(O)}, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S)\\
-    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \mathbb R^{k-1}, s' \in I^{L_s(O)} : g_O(r,s'(O,f(O))) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S)\\
-    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
-= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S)\\
-    + & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S) \\
-= & \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in I^{L_s(N)} : (M,s) \Rightarrow_f^n (N[r], s')\}) \mu(S)
-\end{align*}
-\end{proof}
-\clearpage
 
@@
 
-\begin{theorem} \label{thm:AstEquivalence}
+With these definitions, it is now possible to relate the confluent sampling semantics back to the linear sampling semantics.
+
+\begin{restatable}{theorem}{AstEquivalence} \label{thm:AstEquivalence}
 A closed term $M$ is AST with respect to $\cbv$ iff it is AST.
-\end{theorem}
-\begin{proof}
-\akr{I'm not sure whether this proof is sufficiently detailed. It would be pretty easy to fill it in more, just very verbose.}
+\end{restatable}
 
-For both $\red$ and $\Rightarrow_{\cbv}$, the probability measure on the sample space can be used to define a measure on the reduction sequences of each finite length.
-The distributions are equal for each length, by induction, as follows.
+It is also true that the distributions of terms produced by the linear semantics and by the reduction strategy $\cbv$ are the same, for much the same reason, although that is not relevant to our main results.
 
-The base case, length 0, is trivial, because in both cases the distribution has probability 1 on the sequence containing only $M$.
+\begin{restatable}{theorem}{CbvIsTerminatingest} \label{thm:CbvIsTerminatingest}
+If $M$ terminates with some reduction strategy $f$ and trace $s$, it terminates with $\cbv$ and $s$.
+\end{restatable}
 
-For the inductive case, suppose that the distributions of reduction sequences of length $n$ are equal.
-\akr{TODO: mention the case that it has already terminated by step $n$, or is about to.
-The two versions of the semantics handle it a little differently.}
-
-For any term $N$ which is not a value, there is a unique environment $E$ and redex $R$ such that $N = E[R]$.
-The position of the hole in $E$ is equal to $\cbv(N)$, so that $N|\cbv(N) = R$, as the cases in the definition of $\cbv$ match the cases in the definition of environments.
-If $R \neq \tsample$, let $R'$ be the result of reducing $R$ (which is equal in both versions of the semantics), then $E[R'] = N[R' / \cbv(N)]$ therefore $\red(N,s) = (E[R'],s)$ and $(N,s) \Rightarrow_{\cbv} (E[R'],s)$.
-The distribution of next terms, conditional on the previous term, in the case that that previous term reduces deterministically, is therefore equal for $\red$ and $\Rightarrow_{\cbv}$.
-
-In the other case that $R = \tsample$, the first (term) part of $\red(N,s)$ is $E[\underline{\pi_h(s)}]$.
-If $k$ is the number of samples already taken in this reduction sequence, this is equivalent to $E[\underline{s_0(k)}]$, where $s_0$ is the initial sample.
-For any sufficiently small neighbourhood of $N$ (i.e.~containing only reduction sequences with the same skeletons), the probability of reaching this neighbourhood is independent of $s(k)$, because it depends only on the samples taken so far, and the samples are independent.
-The distribution of $s(k)$ conditional on reaching $N$ is therefore the uniform distribution on $I$.
-Similarly for the other semantics, if the distribution of remaining samples after $n$ steps is independent of the term by Lemma \ref{lem:independentSamples}, therefore the distribution of $s(Sk(N),\cbv(N))$ conditional on $N$ is the uniform distribution on $I$.
-In either case, the distribution of the next term conditional on the previous term, in the case that it reduces randomly, is equal to the image under $r \mapsto E[\underline r]$ of the uniform distribution on $I$.
-
-In any case, the distribution of reduction sequences after $n+1$ steps, conditional on the reduction sequence after $n$ steps, as defined by $\Rightarrow_{\cbv}$ and by $\red$, is equal, any by the inductive hypothesis the distributions after $n$ steps are equal, therefore by integrating over the reduction sequence up to step $n$, the distributions on reduction sequences up to step $n+1$ are equal.
-
-The distributions on reduction sequences of any finite length as defined by $\Rightarrow_{\cbv}$ and $\red$ are therefore equal, therefore so is the probability of having reached a value after $n$ steps, therefore so is its limit, the probability of termination, therefore the probability of termination is 1 iff the probability of termination with respect to $\cbv$ is 1.
-\end{proof}
-
-\begin{theorem} \label{thm:CbvIsTerminatingest}
-If $M$ terminates with some reduction strategy $f$ and samples $s$, it terminates with $\cbv$ and $s$.
-\end{theorem}
-\begin{proof}
-  Suppose $M$ terminates with the reduction strategy $f$ and samples $s$.
-Let $(M,s) = (M_0,s_0) \Rightarrow_f \dots \Rightarrow_f (M_n,s_n)$, where $M_n$ is a value.
-By Lemma \ref{canonicalCousins}, the reduction sequence $(Sk(M_i))_i$ is related by $\sim_c^*$ to some skeletal reduction sequence $X$ which is in call-by-value order.
-For each skeletal reduction sequence $X_j$ in the sequence $(Sk(M_i))_i \sim_c \dots \sim_c X$, it is possible to define a reduction sequence of terms such that $(M,s) = (N_{j,0},s_{j,0}) \Rightarrow \dots \Rightarrow (N_{j,n_j},s_{j,n_j})$ and $Sk(N_{j,k}) = X_{j,k}$, by at each step applying the reduction at the same place as in $X_j$, essentially just filling in the reals in the skeletal reduction sequence to make it a reduction sequence of terms (it will be shown shortly that the correct branch is taken in each $\textsf{if}$-reduction).
-Assume for induction on $j$ that $(N_{j,n_j},s_{j,n_j}) = (M_n,s_n)$.
-To prove that this holds true for $j+1$ as well, it suffices to show that the corresponding term and samples immediately after the reductions involved in this $\sim_c$ step match, as from that point on, the $\Rightarrow$-sequences will match, as they start from the same point and have all the same reduction positions.
-Letting the term and remaining samples immediately before the reductions involved in this $\sim_c$ step be $(N,t)$ (which is the same in both $N_j$ and $N_{j+1}$ for the same reason again, they start at the same point and the sequences of reduction positions up to this point are the same), and the terms and samples after these reductions be $(O_1,t_1)$ and $(O_2,t_2)$, so that these match the $N$, $O_1$ and $O_2$ in the definition of $\sim_c$, $t_1 = t_2$ because both of them are, by Lemma \ref{lem:sim-M-N}, equal to the restriction of $t$ to the subset of $Rch(N)$ that starts with the reduction sequences to $O_1$ and $O_2$ respectively, and the potential positions in those subsets of $Rch(N)$ are identified with each other by $\sim$ because of the same case of $\sim_c$ that means $N_j$ and $N_{j+1}$ are related.
-If none of the relevant reductions are $\tsample$-reductions, $O_1 = O_2$ trivially.
-If there are samples taken (which must be the arbitrary reduction at $\beta$), it is also required that the same samples are taken in both branches.
-Similarly to Lemma \ref{churchRosser}, the positions of these $\tsample$-reductions in each branch are related by $\sim_p^*$, therefore they correspond to the same sample in $t$, therefore $O_1 = O_2$.
-The fact that the terms are equal also implies that the same branch of an $\textsf{if}$ statements will be taken.
-
-  \paragraph{}
-  There is therefore a reduction sequence $(M,s) = (N_0,s_0') \Rightarrow \dots \Rightarrow (N_m, s_m')$ which ends in a value ($N_m = M_n$) and is in CBV order.
-It doesn't immediately follow that this is the $\Rightarrow_{\cbv}$ reduction sequence starting from $(M,s)$ because the fact that $(N_i)_i$ is in CBV order only means that the reductions that do take place in this reduction sequence are in the correct order, not that the reduction sequence finishes once it reaches a value, or that there are no other reductions that would come earlier in CBV order that don't occur at all in $(N_i)_i$.
-Suppose that at some point $j$ in the reduction sequence, the reduction that occurs is not the next reduction in CBV order, but something else, and that the reduction that should be next is at position $\alpha = \cbv(N_i)$, and the position of the actual reduction $N_j \to N_{j+1}$ is $\beta$.
-Suppose further that $N_j$ is not a value.
-Either $\alpha = \cdot$, the root position, or the reduction at the root position is dependent in some way on the reduction at $\alpha$ (e.g.
-$\alpha = \textsf{if}_1$, or $\alpha = \underline{+};@_1$).
-In any case, all the reductions in the rest of the reduction sequence are after $\alpha$ in CBV order, therefore none of them are $\leq \alpha$, and there continues to be an unreduced redex at $\alpha$, therefore the reduction at the root position never occurs, therefore $N_m$ is not a value, which is a contradiction, therefore at the point (if any) where $(N_i)_i$ departs from the $\Rightarrow_{\cbv}$ reduction sequence starting at $(M,s)$, it has already reached a value, therefore $(M,s) \Rightarrow_{\cbv} \dots$ does reach a value eventually, i.e.~$M$ terminates with the reduction strategy $\cbv$ and samples $s$.
-\end{proof}
+It is not quite true, however, that all reduction strategies produce the same values (barring some reduction strategies terminating where others don't), because it is possible for a reduction strategy to overshoot the value that the linear semantics produces by, for example, performing reductions inside of a lambda.
+Even for the same value of the samples, the final values may be different.
+A more restricted result is true, however, that if a term terminates with some reduction strategy $f$ and samples $s$, the value that $\Rightarrow_{\cbv}$ terminates with can reduce via $\Rightarrow$ to the value that $\Rightarrow_f$ terminates with (as a corollary to the proof of Theorem \ref{thm:CbvIsTerminatingest}).
 
@@
 \lo{Give an example or two of a reduction strategy (for $\Rightarrow$) that is not $\cbv$.}
-
-\akr{NOTE. For the standard CBV semantics, this example is indeed not a problem. 
-In the alternative semantics, defining 
-\begin{align*}
-A[x] &= \tif{\tif{x>0}{I}{I} \, \tsample - 0.5 > 0}{0}{\Omega}\\
-B &= \tif{\tsample - 0.5 > 0}{0}{\Omega}
-\end{align*}
-The set of samples where it terminates is 
-\[\begin{array}{l}
-\bigcup_{r \in [0,1]} \{s \in I^{L_s(A[\tsample])} \mid s([A[\tsample]], \mathsf{if}_1;\underline{ -
-}_1;@_1;\mathsf{if}_1) = r, \\
-\qquad \qquad \hfill \qquad \qquad s([A[\tsample], A[r], \ldots, B, \mathsf{if}_1) > 0.5 \}.
-\end{array}\] 
-This is a rather unwieldy expression, but the crucial part is that $r$ occurs twice in the conditions on $s$: once as the value a sample must take, and once in the location of a sample. 
-It is this set that's unmeasurable.}

### 2021-01-01 LICS21/antitone.tex
minor

@@
 
-  Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) = \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
+  Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) := \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
   The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on antitone partial ranking functions.
-Define $f_2(N) = \int_\entrosp f_1(\nnext(N,s)) + \epsilon(f_1(\nnext(N,s))) g(N,s) \, \mu_{\entrosp}(\mathrm d s)$.
+Define $f_2(N) := \int_\entrosp f_1(\nnext(N,s)) + \epsilon(f_1(\nnext(N,s))) g(N,s) \, \mu_{\entrosp}(\mathrm d s)$.
 For any term $N$ where $f$ is defined, $f_2(N) = f(N)$, and the value that $f_2$ would have at $N$ if $f$ were not defined at $N$ is $\leq f(N)$, by the third condition on antitone partial ranking functions.
@@
 
-We can use the ranking function method (coupled with the solution of linear recurrence relations) to show that provided $p \geq 2 - \sqrt{7}$, the program $\Xi \, \underline r$ is AST for all $r \in \Real$.
+We can use the ranking function method (coupled with the solution of linear recurrence relations) to show that provided $p \geq \sqrt{7} - 2$, the program $\Xi \, \underline r$ is AST for all $r \in \Real$.
 

### 2021-01-01 LICS21/antitone.tex
Rename "partial ranking function" -> "sparse ranking function"

--- 2020-12-31 LICS21/antitone.tex
@@
 \begin{definition}
-We say that a supermartingale $(Y_n)_{n \geq 0}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \geq 0}$ are \emph{antitone strict} if for all $n \geq 0$, we have $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
+We say that a supermartingale $(Y_n)_{n \geq 0}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \geq 0}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
 \(
@@
 \begin{example}
-\iffalse
 \akr{Consider a random variable $X$, uniformly distributed in the interval $[-1,1]$, $T := \min \set{n \mid 2^{-n} < X}$, 
@@
 } \lo{The following reformulation defines $T$ and $Y_k$ directly as a measurable function on a probability space (rather than via a r.v.~$X$).}
-\fi
+
 Take the probability space $(\Omega, \calF, \mathbb{P})$ where $\Omega$ is the closed interval of reals $[-1, 1]$, $\calF$ the Borel $\sigma$-algebra, and $\mathbb{P}$ the corresponding Lebesgue probability measure.
@@
 \Cref{thm:rankable implies termination},
-take the probability space $(\entrosp, \Sigma_\entrosp, \mu_\entrosp)$, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$. 
+take the probability space $(\entrosp, \calF, \mu_\entrosp)$, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$. 
 As before, $(Y_n)_{n \geq 0}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$.
@@
 %\paragraph{}
-As before, constructing antitone ranking fuctions completely is not necessary, and there is a corresponding notion of an antitone partial ranking function.
+As before, constructing antitone ranking fuctions completely is not necessary, and there is a corresponding notion of an antitone sparse ranking function.
 
-Define an \emph{antitone partial ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$
+Define an \emph{antitone sparse ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$
 \begin{itemize}
@@
 \begin{theorem}  \label{thm:antitone partial implies rankable}
-  Every antitone partial ranking function is a restriction of an antitone ranking function.
+  Every antitone sparse ranking function is a restriction of an antitone ranking function.
 \end{theorem}
 \begin{proof}
-  Take a closed term $M$ and an antitone partial ranking function $f$ on $M$, with a corresponding antitone function $\epsilon$.
+  Take a closed term $M$ and an antitone sparse ranking function $f$ on $M$, with a corresponding antitone function $\epsilon$.
 Assume wlog that $\epsilon$ is convex (as if it isn't, we can just take its convex hull instead).
@@
   Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) = \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
-  The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on antitone partial ranking functions.
+  The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on antitone sparse ranking functions.
 Define $f_2(N) = \int_\entrosp f_1(\nnext(N,s)) + \epsilon(f_1(\nnext(N,s))) g(N,s) \, \mu_{\entrosp}(\mathrm d s)$.
-For any term $N$ where $f$ is defined, $f_2(N) = f(N)$, and the value that $f_2$ would have at $N$ if $f$ were not defined at $N$ is $\leq f(N)$, by the third condition on antitone partial ranking functions.
-In order to show that $f_2$ is an antitone partial ranking function, it therefore suffices to show that the value that $f_2$ would have had at each term if $f$ were not defined at that term is at least the expectation of $f_2$ after one reduction step (plus $\epsilon(f_2(N))$ if the reduction step is a $tY$-reduction).
+For any term $N$ where $f$ is defined, $f_2(N) = f(N)$, and the value that $f_2$ would have at $N$ if $f$ were not defined at $N$ is $\leq f(N)$, by the third condition on antitone sparse ranking functions.
+In order to show that $f_2$ is an antitone sparse ranking function, it therefore suffices to show that the value that $f_2$ would have had at each term if $f$ were not defined at that term is at least the expectation of $f_2$ after one reduction step (plus $\epsilon(f_2(N))$ if the reduction step is a $tY$-reduction).
 For any term $N$ which is not of the form $E[R]$ for some $\tY$-redex $R$, this is trivial.
@@
 
-As a corollary, any term which admits an antitone partial ranking function terminates almost surely.
+As a corollary, any term which admits an antitone sparse ranking function terminates almost surely.
 
@@
 
-Programs $M_1$ and $M_2$ are AST. For $M_1$, the partial ranking function 
+Programs $M_1$ and $M_2$ are AST. For $M_1$, the sparse ranking function 
 \[
@@
 \] 
-suffices to prove its termination (and could equivalently be considered an antitone partial ranking function with the constant antitone function $\epsilon_1(x) = 1$).
+suffices to prove its termination (and could equivalently be considered an antitone sparse ranking function with the constant antitone function $\epsilon_1(x) = 1$).
 
@@
 \tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
-we can define an antitone partial ranking function $f_2 : \mathit{Rch}(M_2) \rightharpoonup \nnReal$ 
+we can define an antitone sparse ranking function $f_2 : \mathit{Rch}(M_2) \rightharpoonup \nnReal$ 
 %with antitone function $\epsilon_2$ defined as follows (for $n \in \mathbb N$):
@@
 Therefore by setting $\epsilon_2(x) = \frac 1 {5(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met.
-It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone partial ranking function, and $M_2$ is AST.
+It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone sparse ranking function, and $M_2$ is AST.
 \end{example}
@@
 \]
-We can construct a partial ranking function $f$ as follows:
+We can construct a sparse ranking function $f$ as follows:
 \begin{align*}
@@
 therefore $\epsilon$ is antitone and bounded below by 0.
-We define a partial antitone ranking function by:
+We define an antitone sparse ranking function by:
 \[
@@
 \end{itemize}
-Hence this is a valid partial antitone ranking function and the term is AST.
+Hence this is a valid antitone sparse ranking function and the term is AST.
 \end{example}
@@
 
-In this case, the fact that the ranking function must decrease at each $\tY$-step (in an antitone partial ranking function) by the expected value of $\epsilon$ not at the current term, but at the next term where the ranking function is defined, is a little harder to deal with, because the variable $x$ can change all the way to $0$ in one step, therefore simply adding a small offset doesn't suffice to compensate for this fact.
+In this case, the fact that the ranking function must decrease at each $\tY$-step (in an antitone sparse ranking function) by the expected value of $\epsilon$ not at the current term, but at the next term where the ranking function is defined, is a little harder to deal with, because the variable $x$ can change all the way to $0$ in one step, therefore simply adding a small offset doesn't suffice to compensate for this fact.
 
@@
 \end{align*}
-For each $n$, $\Theta[n]$ reduces to either $0$ or $\Theta[n+1]$, with a $\tY$-reduction only in the latter case, and the condition that this is an antitone partial ranking function is that $n \geq \frac{(n-1)(n+1)}{n} + \frac{n-1}{n} \epsilon(n+1)$ and $12 \geq 11 + \epsilon(11)$.
+For each $n$, $\Theta[n]$ reduces to either $0$ or $\Theta[n+1]$, with a $\tY$-reduction only in the latter case, and the condition that this is an antitone sparse ranking function is that $n \geq \frac{(n-1)(n+1)}{n} + \frac{n-1}{n} \epsilon(n+1)$ and $12 \geq 11 + \epsilon(11)$.
 These are satisfied by setting $\epsilon(x) = \min(1, 1/x)$, therefore this term is AST.
@@
 \] 
-A partial ranking function for $\Xi \, 1$ would be $\Xi^n \, i \mapsto 3 \, n$ (for $n \geq 0$), but w.r.t.~the reduction strategy that first reduces the argument of an application to a value, then the function, then the application itself.
+A sparse ranking function for $\Xi \, 1$ would be $\Xi^n \, i \mapsto 3 \, n$ (for $n \geq 0$), but w.r.t.~the reduction strategy that first reduces the argument of an application to a value, then the function, then the application itself.
 
@@
 
-A standard partial ranking function (i.e.~one that doesn't depend on the confluent semantics) would have a slightly more verbose definition, as the functions that accumulate would not be $\Xi$, but what $\Xi$ reduces to after one step.}
+A standard sparse ranking function (i.e.~one that doesn't depend on the confluent semantics) would have a slightly more verbose definition, as the functions that accumulate would not be $\Xi$, but what $\Xi$ reduces to after one step.}
 
@@
 We can use this example to motivate the confluent trace semantics, though it is perhaps not the most compelling possible.
-Reason: one could use a more restrictive but equally expressive recursion \emph{operator} (as opposed to a $\tY$-\emph{combinator}, which is a constant) to implement general recursion which would make it unnecessary to consider an alternative reduction strategy when applying the partial ranking function theorem.
+Reason: one could use a more restrictive but equally expressive recursion \emph{operator} (as opposed to a $\tY$-\emph{combinator}, which is a constant) to implement general recursion which would make it unnecessary to consider an alternative reduction strategy when applying the sparse ranking function theorem.
 E.g.~define letrec-terms, $\mathsf{letrec} \, f^{A \to B} \, x^A . M$, as values, via the formation rule

### 2021-01-01 LICS21/confluent.tex
Rename "partial ranking function" -> "sparse ranking function"

@@
 \begin{definition}
-A \emph{partial ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : Rch_r(M) \rightharpoonup \mathbb R$ such that
+A \emph{sparse ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : Rch_r(M) \rightharpoonup \mathbb R$ such that
 \begin{itemize}
@@
 \begin{definition}
-An \emph{antitone partial ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : Rch_r(M) \rightharpoonup \mathbb R$ such that there exists some antitone function $\epsilon : \nnReal \to \pReal$ such that
+An \emph{antitone sparse ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : Rch_r(M) \rightharpoonup \mathbb R$ such that there exists some antitone function $\epsilon : \nnReal \to \pReal$ such that
 \begin{itemize}
@@
 \begin{theorem}
-If a closed SPCF term $M$ has a ranking function, partial ranking function, antitone ranking function or antitone partial ranking function with respect to a reduction strategy $r$, then $M$ is AST with respect to $r$, and AST.
+If a closed SPCF term $M$ has a ranking function, sparse ranking function, antitone ranking function or antitone sparse ranking function with respect to a reduction strategy $r$, then $M$ is AST with respect to $r$, and AST.
 \end{theorem}

### 2021-01-01 LICS21/confluent-proofs.tex

--- 2020-12-23 probabilisticProgrammingMartingales.tex
@@
-\documentclass{article}
-
-%% BEGIN {Luke's Macros}
-%%
-\newif\ifdraft
-\drafttrue %% To hide all comments and highlighting, just comment this line out
-
-\input{Style/comment}
-\usepackage{calculation}
-\usepackage{graphicx}
-\usepackage{tikz}
-\usetikzlibrary{positioning, graphs, quotes, graphdrawing, shapes.geometric}
-\usegdlibrary{layered, circular}
-\usepackage[square]{natbib}
-\setcitestyle{aysep={}}
-
-\definecolor{oxblue}{RGB}{0,33,71}
-\definecolor{oxgold}{HTML}{a0630a}
-\usepackage[final,
-bookmarks,
-bookmarksopen,
-colorlinks,
-final,
-linkcolor=red,
-citecolor=oxgold,
-pdfstartview=FitH ]{hyperref}
-
-\usepackage{amsmath} % cleveref must be loaded afer amsmath
-\usepackage{cleveref}
-\Crefname{theorem}{Theorem}{Theorems}
-\Crefname{corollary}{Corollary}{Corollary}
-\crefname{proposition}{Proposition}{Propositions}
-\Crefname{claim}{Claim}{Claims}
-\Crefname{definition}{Definition}{Definitions}
-\Crefname{fact}{Fact}{Facts}
-\Crefname{conj}{Conjecture}{Conjectures}
-\Crefname{example}{Example}{Examples}
-\Crefname{remark}{Remark}{Remarks}
-\Crefname{convention}{Convention}{Conventions}
-\Crefname{lemma}{Lemma}{Lemmas}
-\Crefname{assumption}{Assumption}{Assumptions}
-\Crefname{section}{Section}{Sections}
-\Crefname{appendix}{Appendix}{Appendices}
-\Crefname{figure}{Figure}{Figures}
-%\Crefname{equation}{Eq.}{Equations}
-
-% \addtotheorempostheadhook[theorem]{\crefalias{thmlisti}{theorem}}
-% \addtotheorempostheadhook[lemma]{\crefalias{thmlisti}{lemma}}
-% \addtotheorempostheadhook[proposition]{\crefalias{thmlisti}{proposition}}
-% \addtotheorempostheadhook[corollary]{\crefalias{thmlisti}{corollary}}
-% \addtotheorempostheadhook[claim]{\crefalias{thmlisti}{claim}}
-% \addtotheorempostheadhook[fact]{\crefalias{thmlisti}{fact}}
-% \addtotheorempostheadhook[example]{\crefalias{thmlisti}{example}}
-% \addtotheorempostheadhook[definition]{\crefalias{thmlisti}{definition}}
-% \addtotheorempostheadhook[remark]{\crefalias{thmlisti}{remark}}
-% \addtotheorempostheadhook[assumption]{\crefalias{thmlisti}{assumption}}
-
-\newcommand{\Real}{\mathbb{R}}
-\newcommand{\nnReal}{\mathbb{R}_{{\geq}0}}
-\newcommand{\pReal}{\mathbb{R}_{{>}0}}
-
-\newcommand\expect[1]{\mathbb{E}[#1]}
-\newcommand\set[1]{\{#1\}}
-\newcommand\dif{\mathrm{d}}
-\newcommand\calF{\mathcal{F}}
-\newcommand\calI{\mathcal{I}}
-\newcommand\Leb{\mathrm{Leb}}
-\newcommand\ndraw[2]{\#\mathrm{draw}_{#1}(#2)}
-
-\makeatletter
-\newcommand{\dotDelta}{{\vphantom{\Delta}\mathpalette\d@tD@lta\relax}}
-\newcommand{\d@tD@lta}[2]{%
-  \ooalign{\hidewidth$\m@th#1\mkern-1mu\cdot$\hidewidth\cr$\m@th#1\Delta$\cr}%
-}
-
-%\usepackage[bbgreekl]{mathbbol} %for \bblambda
-%%
-%% END {Luke's Macros}
-
-\usepackage{amsfonts}
-\usepackage{amsthm}
-\usepackage{amssymb}
-\usepackage{verbatim}
-\usepackage{enumitem}
-\newcommand{\tY}{\mathsf{Y}}
-%\newcommand{\tif}[3]{\textsf{if }#1\textsf{ then }#2\textsf{ else }#3}
-\newcommand{\tif}[3]{\mathsf{if}(#1, #2, #3)} % space-saving concrete syntax
-\newcommand{\tsample}{\mathsf{sample}}
-\newcommand{\tscore}{\mathsf{score}}
-\newcommand{\skeletonPlaceholder}{\mathsf{X}} % I'm not sure what the best notation here is.
-\DeclareMathOperator{\red}{red}
-\DeclareMathOperator{\nnext}{next}
-\DeclareMathOperator{\cbv}{cbv}
-
-\usepackage{amsthm}
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}
-\newtheorem{example}{Example}
-
-\theoremstyle{lemma}
-\newtheorem{lemma}{Lemma}
-\newtheorem{theorem}{Theorem}
-\newtheorem{corollary}{Corollary}
-\newtheorem{conjecture}{Conjecture}
-
-\theoremstyle{remark}
-\newtheorem*{remark}{Remark}
-
-\title{Supermartingales and Termination Analysis of Statistical PCF}
-
-\begin{document}
-
-\maketitle
-
-\lo{Tentative title}
-
-\begin{abstract}
-\lo{TODO}
-\end{abstract}
-
-\iffalse
-\akr{Following your suggestion, I will be providing a criterion for termination of programs in PPCF \citep{DBLP:journals/pacmpl/EhrhardPT18} based on ranking supermartingales. 
-As it's more convenient for this proof, a sampling-based semantics \lo{You need to provide reference(s) for sampling-based semantics.} will be used instead of the original distributional semantics. 
-I assume some roughly applicable equivalence is proven somewhere, but it doesn't seem that hard anyway.}
-\lo{Why not provide a proof as an appendix?}
-\fi
-
-\tableofcontents
-
-Various theorems about probabilistic programs rely on the assumption that the program terminates almost surely.
-Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \citep{DBLP:journals/pacmpl/McIverMKK18}.
-This paper's contribution is to extend this method to a higher-order setting.
-
-\section{Statistical PCF}
-
-\subsection{Syntax of SPCF}
-
-The language SPCF is a simply-typed lambda calculus with sampling of real numbers from $[0,1]$ and unbounded scoring, following \cite{MakOP20b}.
-Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
-\begin{align*}
-  & \text{types } A, B ::= \textsf{R}  \mid  A \to B \\
-  & \text{values } V ::= \lambda x.M  \mid  \underline{r} \\
-  & \text{terms } M, N ::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY M  \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample  \mid  \tscore(M)
-\end{align*}
-\begin{align*}
-  \frac{}{\Gamma ; x:A \vdash x:A} \qquad
-  \frac{\Gamma ; x:A \vdash M : B}{\Gamma \vdash \lambda x.M : A \to B} \qquad
-  \frac{}{\underline{r} : \textsf{R}} \qquad
-  \frac{\Gamma \vdash M:A \to B \quad \Gamma \vdash N : A}{\Gamma \vdash M \, N : B} \\ \\
-  \frac{\Gamma \vdash M_1:\textsf{R} \dots \Gamma \vdash M_n:\textsf{R}}{\Gamma \vdash \underline{f}(M_1,\dots,M_n) : \textsf{R}} \ f : \mathbb{R}^n \to \mathbb{R} \qquad
-  \frac{\Gamma \vdash M : (A \to B) \to (A \to B)}{\Gamma \vdash \tY M : (A \to B)} \\ \\
-  \frac{\Gamma \vdash M : \textsf{R} \quad \Gamma \vdash N_1 : A \quad \Gamma \vdash N_2 : A}{\Gamma \vdash \tif{M < 0}{N_1}{N_2} : A} \qquad
-  \frac{}{\Gamma \vdash \tsample : \textsf{R}} \qquad
-  \frac{\Gamma \vdash s : \textsf{R}}{\Gamma \vdash \tscore (s) : \textsf{R}}
-\end{align*}
-
-Terms are identified up to $\alpha$-equivalence, as usual. The set of all terms is denoted $\Lambda$, and the set of closed terms is denoted $\Lambda^0$.
-
-\paragraph{}
-To define the reduction relation, let \emph{evaluation contexts} be of the form:
-\begin{align*}
-  E ::= & \, [\,] \mid E \, M \mid V \, E \mid \underline{f}(\underline r_1,\dots ,\underline r_{k-1}, E, M_{k+1}, \dots, M_n) \\ & \mid \tY E \mid \tif{E<0}{M_1}{M_2} \mid \tscore (E)
-\end{align*}
-then a term reduces if it is formed by substituting a redex in a context i.e.
-\begin{align*}
-  E[(\lambda x.M) N] & \to E[M[N/x]] \\
-  E[\underline f (\underline r_1, \dots , \underline r_n)] & \to E[\underline{f(r_1,\dots,r_n)}] \\
-  E[\tY \lambda x. M] & \to E[\lambda z. M[(\tY \lambda x. M)/x] z] \text{ where $z$ is not free in $s$}\\
-  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_1] \text{ where }r < 0 \\
-  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_2] \text{ where }r \geq 0 \\
-  E[\tsample] & \to E[\underline r] \text{ where } r \in [0,1] \\
-  E[\tscore(\underline r)] & \to E[\underline r].
-\end{align*}
-\changed[lo]{We write $\to^\ast$ for the reflexive, transitive closure of $\to$.}
-\lo{The $\tscore$ rule should have a side condition: if $r > 0$.}
-
-Every closed well-typed term either is a value or reduces to another term.
-
-\subsection{Sampling semantics}
-\label{sec:sampling semantics}
-This version of the reduction relation allows $\tsample$ to reduce to any number in $[0,1]$.
-To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples.
-Let $ I = [0,1] \subset \mathbb{R} $, and let $S = I^{\mathbb{N}}$, with the \changed[lo]{Borel $\sigma$-algebra $\calF$} and the probability measure, denoted $\mu$, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
-Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and $\mu (\prod_{i=0}^\infty X_i) = \prod_{i=0}^\infty \Leb(X_i)$.
-The maps $\pi_h:S \to I, \; \pi_t:S \to S$ popping the first element are then measurable.
-\changed[lo]{Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(S, \calF, \mu)$ the \emph{entropy space}.}
-
-The $\sigma$-algebra and measure on $\Lambda$ are defined by considering it as a disjoint union of equivalence classes under replacing all the real constants by a placeholder, where the measure on each class is that of $\mathbb{R}^n$, where $n$ is the number of real constants.
-\changed[lo]{Precisely, following \cite{DBLP:conf/icfp/BorgstromLGS16},
-we view $\Lambda$ as $\bigcup_{m\in\omega} \big(\mathsf{Sk}_m \times \Real^m \big)$,
-where $\mathsf{Sk}_m$ is the set of SPCF terms with exactly $m$ place-holders (a.k.a.~\emph{skeleton terms}) for numerals.
-Thus identified, we give $\Lambda$ the countable disjoint union topology of the product topology of the discrete topology on $\mathsf{Sk}_m$ and the standard topology on $\Real^m$.
-Note that the connected components of $\Lambda$ have the form $\{M\} \times \Real^m$, with $M$ ranging over $\mathsf{Sk}_m$, and $m$ over $\omega$. 
-We fix the Borel algebra of this topology to be the $\sigma$-algebra on $\Lambda$.}
-
-The one-step reduction is given by the function $\red : \Lambda^0 \times S \to \Lambda^0 \times S$ where
-\begin{equation}
-\red(M,s) = \left\{
-    \begin{array}{ll}
-        (E[N],s) & \text{if } M = E[R], R \to N \text{ and } R \neq \tsample \\
-        (E[\underline{\pi_h(s)}],\pi_t(s)) & \text{if } M = E[\tsample] \\
-        (M,s) & \text{if } M \text{ is a value}
-    \end{array} \right .
-\end{equation}
-\lo{Both $\to$ and $\red$ are called reduction relation.}
-
-The result after $n$ steps is then simply $\red^n(M,s) = \overbrace{\red(...\red(}^n M,s)...)$, and the limit $\red^\infty$ can then be defined as a partial function as $\lim_{n \to \infty} \red^n(M,s)$ whenever that sequence becomes constant by reaching a value.
-A term $M$ terminates for a sample sequence $s$ if the limit $\red^\infty(M,s)$ is defined.
-
-The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid M \text{ terminates at } s \}$ is measurable.
-A term $M$ is said to terminate almost surely if $\mu(\{s \mid M \text{ terminates at } s\}) = 1$.
-
-\lo{Alternatively, define the \emph{runtime of $M$} to be the random variable 
-\[
-T_M(s) := 
-\begin{cases}
-\min \set{n \mid \pi_0(\red^n(M, s)) \textrm{ is a value}} & \hbox{if $\red^\infty(M,s)$ is defined}\\
-\infty & \hbox{otherwise}
-\end{cases}
-\]
-Equivalently, we say that $M$ is \emph{almost-surely terminating} (AST) if $T_M < \infty$ a.s.; 
-and $M$ is \emph{positively almost-surely terminating} (PAST) if $\expect{T_M} < \infty$.}
-
-For example, the term 
-\[
-(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0},
-\] 
-which generates a geometric distribution, terminates on the set $S \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas 
-\[
-\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \underline 0},
-\] 
-which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.
-
-\paragraph{}
-Note that the requirement that $M$ be closed is not important for the proofs of the theorems, but is simply required to ensure that $M$ reaches a value, rather than a general normal form like $x \, y$.
-
-\paragraph{}
-This definition of almost-sure termination is equivalent to that given in \citep{MakOP20b}, although the program semantics is stated in a slightly different way.
-In particular, the argument to $\tscore$ is not relevant to termination (except for the possibility that its argument's evaluation wouldn't terminate).
-
-\iffalse
-\lo{Your operational semantics does not maintain a record of the current weight of the reduction.
-A.s.~termination does depend on $\tscore$: see \cite[\S 4.3]{DBLP:journals/corr/abs-2004-03924}\footnote{\url{https://arxiv.org/abs/2004.03924}}.
-I think it important to take the behaviour of $\tscore$ into account;
-you should do it as a future task.}
-\fi
-
-\section{Supermartingales}
-One approach to proving that a term terminates almost surely is to find some variant that is bounded below and, on average, decreases sufficiently quickly that it must eventually reach 0, similarly to the approach taken in \citep{DBLP:journals/pacmpl/McIverMKK18} for imperative programs.
-
-These variants are defined as functions from reachable terms (i.e.~possible states of the program's execution) to real numbers.
-Specifically, let the set of reachable terms from a given closed starting term $M$ be $Rch(M) := \{N \in \Lambda \mid M \to^* N \}$, with the $\sigma$-algebra induced as a subset of $\Lambda$.
-
-\begin{definition}\rm
-A \emph{ranking function on $M$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
-\begin{enumerate}
-    \item $f(E[\tY \lambda x. N]) \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z) \text{ where $z$ is not free in $N$}$
-    \item $f(E[\tsample]) \geq \int_I f(E[\underline{x}]) \, \Leb(\mathrm{d}x)$
-
-    \item $f(E[R]) \geq f(E[R'])$ for any other redex $R$, where $R \to R'$.
-\end{enumerate}
-We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that $f(N) = 0$ iff $N$ is a value, and for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.
-
-Any closed term for which a ranking (respectively, strict ranking) function exists is called \emph{rankable} (respectively, \emph{strictly rankable}). 
-\changed[lo]{For example, the term $(\tY \lambda x.x) \, \underline 0$ is not rankable.}
-It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \in \omega}$ is the reduction sequence starting from $M$, then $(f(M_n))_{n\in\omega}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales.
-\end{definition}
-
-\input{ranking-supermartingales}
-
-\section{Constructing Ranking Functions}
-Although rankability implies almost-sure termination, the converse does not hold in general. For example,
-\begin{equation}
-\tif{{-}\tsample < 0}{\underline{0}}{(\tY \lambda x. x) \, \underline 0}
-\label{ex:0 probability reachable}
-\end{equation}
-terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x. x) \, \underline 0$ is reachable, although that has probability 0. 
-Not only is this counterexample AST, it's PAST. %positively almost surely terminating i.e.~the expected time to termination is finite.
-
-A ranking function can be constructed under the stronger assumptions that, for every $N$ reachable from $M$, the expected number of $\tY$-reduction steps from $N$ to a value is finite.
-In particular, the expected number of $\tY$-reduction steps from each reachable term is a ranking function.
-Note that a finite number of expected $\tY$-reduction steps does not necessarily imply a finite number of expected total reduction steps.
-
-\begin{example}
-\label{ex:tY finite does not imply t finite}
-The term
-\[
-M = \big(\tY \lambda f n. \tif{\tsample - 0.5 < 0}{n \, \underline 0}{f (\lambda x. n (n x))}\big) (\lambda x. x+1)
-\] 
-terminates with only 2 $\tY$-reductions on average i.e.~$\expect{T_M^\tY} = 2$, but applies the increment function $2^n$ times with probability $2^{-1-n}$ for $n \geq 0$, which diverges i.e.~$\expect{T_M} = \infty$.
-\end{example}
-
-\begin{theorem} \label{thm:minimal}
-Given a closed term $M$, the function $f:Rch(M) \to \mathbb R$ given by $f(N) = \mathbb E [\text{the number of }\tY\text{-reduction steps from }N\text{ to a value}]$, if it exists, is the least of all possible ranking functions of $M$.
-\end{theorem}
-\begin{proof}
-Let $f$ be the candidate least ranking function defined above, and suppose $g$ is another ranking function such that $f(N) > g(N)$ for some $N \in Rch(M)$.
-The restrictions of $f$ and $g$ to $Rch(N)$ have the same properties assumed of $f$ and $g$, so assume w.l.o.g.~that $N=M$.
-The difference $g - f$ is then a supermartingale (with the same setup as in the proof of %\Cref{thm:rankable gives supermartingale}
-%Theorem \ref{rankable implies ast}) 
-\Cref{thm:rankable and strict rankable});
-therefore $\mathbb E[g(M_n)] \leq \mathbb E[f(M_n)] + g(M)-f(M)$, for all $n$.
-
-We have $\mathbb E[f(M_n)] = \sum_{k=n}^\infty \mathbb P[M_k = E[\tY N] \text{ for some }E, N] \to 0 \text{ as } n \to \infty$; 
-therefore as $g(M) - f(M) < 0$, eventually $\mathbb E[g(M_n)] < 0$, which is impossible. 
-It follows that $g \geq f$ as required.
-
-In order for $f$ to be the least ranking function of $M$, it also has to actually be a ranking function itself.
-Each of the conditions on a ranking function is easily verified from the definition of $f$.
-\end{proof}
-
-\subsection{Partial ranking functions}
-Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost-sure termination, so it isn't useful for proving it.
-
-\begin{example}[Geometric distribution]
-Consider the term
-\begin{align*}
-&(\tY \lambda f, n. \\
-&\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
-&) \, \underline{0}
-\end{align*}
-which generates a geometric distribution (let $\Theta = \lambda f, n : \tif{\tsample \underline{- 0.5} < 0}{n}{f (n\underline{+ 1})}$, then the term is $(\tY \Theta)\ \underline 0$).
-    Despite its simplicity, its $Rch$ contains all the terms
-
-\begin{tikzpicture}[node distance=0.5em, anchor=west]
-    \node (1) at (0,-0) {$\tY\ \Theta\ \underline 0$};
-    \node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\ \Theta\ (n \underline{+ 1})}) z)\ \underline i$};
-    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\ \Theta\ (n \underline{+ 1})})\ \underline i$};
-    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY\ \Theta\ (\underline i\ \underline{+ 1})}$};
-    \node (5) at (0,-4) {$\tif{\underline r\ \underline{- 0.5} < 0}{\underline i}{\tY\ \Theta\ (\underline i\ \underline{+ 1})}$};
-    \node (6) at (0,-5) {$\tif{\underline r < 0}{\underline i}{\tY\ \Theta\ (\underline i\ \underline{+ 1})}$};
-    \node (7) at (2,-6) [text width=] {$\underline i$};
-    \node (8) at (0,-7) {$\tY\ \Theta\ (\underline i\ \underline{+ 1})$};
-    \node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\ \Theta\ (n \underline{+ 1})}) z)\ (\underline i\ \underline{+ 1})$};
-
-    \draw [->] (1.south west) to [bend right=45] (2.north west);
-    \draw [->] (2.south west) to [bend right=45] (3.north west);
-    \draw [->] (3.south west) to [bend right=45] (4.north west);
-    \draw [->] (4.south west) to [bend right=45] (5.north west);
-    \draw [->] (5.south west) to [bend right=45] (6.north west);
-    \draw [->] (6.south west) to [bend right=45] (8.north west);
-    \draw [->] (8.south west) to [bend right=45] (9.north west);
-
-    \draw [->] (9.west) to [bend left=45] (2.west);
-    \draw [->] (6) to (7);
-\end{tikzpicture}
-
-\end{example}
-
-Even in this simple case, defining a ranking function explicitly is awkward because of the number of cases, although in most cases, because the value need only be greater than or equal to that of the next term in sequence, it suffices to take the ranking function as having the same value as the next term, so that overall it takes only 3 distinct values.
-(We will explain why later.)
-
-The definition of rankability is also inconvenient for syntactic sugar.
-It could be useful, for example, to define $M \oplus_p N = \tif{\tsample - p < 0} M N$, where $M \oplus_p N$ reduces to $M$ or $N$, depending on the first value of $s$\lo{Need a name for $s$.}, with probability $p$ resp.~$(1-p)$.
-Technically though, it reduces first to $\tif{\underline r - p < 0} M N$ for all $r \in I$, so those terms all need values of the ranking function too.
-
-\paragraph{}
-In both of these cases, there are only some values of the ranking function that are semantically important. 
-\begin{definition}
-Define a \emph{partial ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \rightharpoonup \mathbb R$ such that
-\begin{itemize}
-    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
-    \item $f$ is defined at $M$.
-    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually \akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and ast, but that doesn't actually work because of terms reachable with 0 probability.
-Of course, astness doesn't imply rankability, so any definition of partial ranking function that includes all Y-past terms wouldn't always extend to a total ranking function.} reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
-\end{itemize}
-\end{definition}
-A partial ranking function that is total is just a ranking function.
-Providing a partial ranking function is essentially part way between providing a ranking function and directly proving almost sure termination.
-
-\begin{theorem} \label{thm:partial implies rankable}
-Every partial ranking function is a restriction of a ranking function.
-\end{theorem}
-\begin{proof}
-Take a closed term $M$ and partial ranking function $f$ on $M$. Define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
-\begin{align*}
-    f_1(N) & := f(N) \text{ whenever $f(N)$ is defined,} \\
-    f_1(V) & := 0 \text{ for values $V$ not in the domain of $f$.}
-\end{align*}
-
-Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) := \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
-The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on partial ranking functions.
-Define $f_2(N) = \int_S f_1(\nnext(N,s)) + g(N,s) \, \mu(\mathrm d s)$.
-The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction).
-\lo{TODO. The preceding sentence deserves a proof.}
-\end{proof}
-
-As a corollary, any term which admits a partial ranking function terminates almost surely.
-
-\subsection{Examples}
-Let $M \oplus_p N = \tif{\tsample - p < 0} M N$, for $p \in (0,1]$, then there are the pseudo-reduction relations
-\begin{align*}
-E[M \oplus_p N] \to^3 & E[M] & \\
-E[M \oplus_p N] \to^3 & E[N] & \\
-\red^3(E[M \oplus_p N], s) = & \left\{
-    \begin{array}{ll}
-        (E[M],\pi_t(s)) & \text{if } \pi_h(s) < p \\
-        (E[N],\pi_t(s)) & \text{if } \pi_h(s) \geq p. \\
-    \end{array} \right .
-\end{align*}
-A partial ranking function could be defined with respect to this shortcut reduction simply by replacing $\to$ and $\red$ in the definition of a partial ranking function by a version that goes straight from $N \oplus_p O$ to $N$ or $O$.
-Such a pseudo-partial ranking function would then be a partial function from a subset of $Rch(M)$, so it could also be considered as a partial function from all of $Rch(M)$, and it would in fact also be an actual partial ranking function.
-It is therefore possible to prove rankability directly using the shortcutted reductions.
-
-A similar procedure would work for other forms of syntactic sugar.
-If a closed term $N$ eventually reduces to one of a set of other terms $\{N_i \mid i \in I\}$ with certain probabilities, a partial ranking function defined with respect to a reduction sequence that skips straight from $N$ to $N_i$ is also a valid partial ranking function for the original reduction function, and therefore its existence implies almost sure termination.
-There is a caveat, however, that $\tY$-reduction steps skipped over in the shortcut still need to be counted for the expected number of $\tY$-reduction steps.
-
-\paragraph{}
-With this abbreviation, the geometric distribution example from earlier can be written as $(\tY \lambda f n.\ n \oplus_{0.5} f(n+1)) \, \underline 0$.
-It is then easy to see that the following is a partial ranking function:
-\begin{align*}
-(\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\ N & \mapsto 2 \\
-\underline i \oplus_{0.5} (\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\ (\underline i + 1) & \mapsto 1 \\
-\underline i & \mapsto 0.
-\end{align*}
-
-In fact, even the partial function $\tY \Theta N \mapsto 2$ alone is a partial ranking function for this term.
-\section{Antitone ranking functions}
-
-\begin{example}[Random walk]
-\label{ex:ac-ranking}
-\begin{enumerate}
-\item 1D biased (towards 0) random walk
-\[
-M_1 = 
-\big(\tY \, \lambda f n . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{3/2} f \, (n + 1)}\big) \, 10
-\]
-
-\item 1D unbiased random walk
-\[
-M_2 = 
-\big(\tY \, \lambda f n . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}\big) \, 10
-\]
-
-\item 1D biased (away from 0) random walk
-\[
-M_3 = 
-\big(\tY \, \lambda f n . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/3} f \, (n + 2)}\big) \, 10
-\]
-
-%While {x > 0} do {x := x-1 \oplus_2/3 x:= x+1}
-
-%M2 = While {x > 0} do {x := x-1 \oplus_2/3 x:= x+2}
-\end{enumerate}
-\end{example}
-
-The term $M_1$ is rankable, terminating in 31 $\tY$-steps on average, and $M_3$ only has a $1/1024$ chance of terminating, but in between, $M_2$ is AST, but isn't $\tY$-PAST, therefore it isn't rankable and \Cref{thm:rankable implies termination} is insufficient to prove its termination.
-We therefore want to find a generalised notion of ranking function so that $M_2$ becomes rankable, and then prove it sound i.e.~rankable in this generalised sense implies AST.
-
-\paragraph{Idea.}
-
-The definition of antitone-convex ranking function $f$ (say) for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
-we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) function $\epsilon : \nnReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
-\[
-f(E[R’]) \leq f(E[R]) - \epsilon(f(E[R])) 
-\]
-where $R \to R’$ is the $\tY$-redex rule.
-
-\lo{An aside: This CBV $\tY$-rule seems cleaner:
-\[
-\big(\tY f^{A \to B} \, x^A \, . \, \theta^B \big) \, v \to \theta[(\tY f \, x \, . \, \theta) / f, v / x].
-\]
-We assume $\tY f^{A \to B} \, x^A \, . \, \theta^B$ is a value.}
-
-\begin{definition}
-We say that a supermartingale $(Y_n)_{n \in \omega}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \in \omega}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
-\[
-\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot 1_{\set{T > n}}
-\]
-\end{definition}
-
-%\lo{The preceding definition does not place any constraint on the stopping time $T$ (other than that it is w.r.t.~the filtration $(\calF_n)_{n \in \omega}$).
-However for the definition to make sense, $T$ must be the stopping time $T(s) := \min \set{n \mid Y_n(s) = 0}$.}
-
-\begin{theorem}
-\label{thm:a-c strict}
-Let $(Y_n)_{n \in \omega}$ be an antitone strict supermartingale via function $\epsilon : \nnReal \to \pReal$, and stopping time $T$. 
-Then $T < \infty$ almost surely.
-\end{theorem}
-
-\begin{proof}
-First, as $(Y_n)$ is a supermartingale, $\expect{Y_n} \leq \expect{Y_0}$. 
-Therefore
-
-\begin{calculation}
-\expect{Y_n \mid T > n}
-  \step[=]{rearranging terms}
-\dfrac{\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
-+ 
-\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
-- 
-\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
-  \step[=]{definition of conditional expectation}
-\dfrac{\expect{Y_n} - \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
-  \step[\leq]{$Y_n \geq 0$ always}
-\dfrac{\expect{Y_n}}{\mathbb P[T > n]}
-  \step[\leq]{$(Y_n)_n$ is a supermartingale}
-\dfrac{\expect{Y_0}}{\mathbb P[T > n]}
-\end{calculation}
-
-\emph{Claim}: For all $0 < x \leq 1$, 
-\(
-\mathbb P[T > B_x] \leq x,
-\)
-where $B_x = \left \lceil \dfrac{\expect{Y_0}+1}{x \; \epsilon(\expect{Y_0} \, x^{-1})} \right \rceil$.
-
-\smallskip
-
-As the convex hull of $\epsilon$ (the greatest convex function less than or equal to it) satisfies all the conditions assumed of $\epsilon$, in addition to being convex, assume wlog that $\epsilon$ is convex.
-
-Assume for a contradiction that $\mathbb P[T > B_x] > x$.
-Then, take $n \leq B_x$. 
-We have
-%\lo{Change of justification of the second equality below: event, not discrete r.v. Recall: expectation conditioning on r.v.~(resp.~event) is a r.v.~(resp.~number).}
-\begin{calculation}
-\expect{Y_n-Y_{n+1}}
-  \step[=]{$\calF_n \subseteq \calF_{n+1}$, definition and linearity of conditional expectation}
-\expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
-  \step[\geq]{antitone strict assumption}
-\expect{\epsilon(Y_n) \cdot 1_{\set{T > n}}}
-%  \step[=]{definition of conditional expectation for a discrete variable}
-\step[=]{definition of expectation conditioning on an \emph{event}}
-\mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
-  \step[\geq]{Jensen's inequality}
-\mathbb P[T > n]\ \epsilon(\expect{Y_n \mid T > n})
-  \step[\geq]{proved earlier}
-\mathbb P[T > n]\ \epsilon\Big(\dfrac{\expect{Y_0}}{\mathbb P[T > n]}\Big)
-  \step[>]{assumption, $\mathbb P[T > n] \geq \mathbb P[T > B_x] > x$}
-x \; \epsilon\big(\expect{Y_0} \, x^{-1}\big).
-\end{calculation}
-Therefore, by a telescoping sum
-\[
-\expect{Y_{B_x}} = \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1}) \leq -1 < 0 
-\]
-%\lo{$\expect{Y_0 - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})}$}
-which is a contradiction, therefore the claim must be true, therefore $P[T > n]$ tends to 0 as $n$ tends to infinity, therefore $T < \infty$ almost surely.
-\end{proof}
-
-Note that it is essential in this proof that $\epsilon$ be defined for all $\mathbb R_{\geq 0}$, not just the values that $(Y_n)_{n \in \omega}$ actually takes (or at least if $(Y_n)_{n \in \omega}$ is uniformly bounded, $\epsilon$ must be positive at the supremum as well as the actual values).
-
-\begin{example}
-\akr{Consider a random variable $X$, uniformly distributed in the interval $[-1,1]$, $T := \min \set{n \mid 2^{-n} < X}$, 
-\[
-Y_k := 
-\begin{cases}
-4-2^{-k} & \hbox{if $k < T$}\\
-0 & \hbox{otherwise}
-\end{cases} 
-\]
-} \lo{The following reformulation defines $T$ and $Y_k$ directly as a measurable function on a probability space (rather than via a r.v.~$X$).}
-
-Take the probability space $(\Omega, \calF, \mathbb{P})$ where $\Omega$ is the closed interval of reals $[-1, 1]$, $\calF$ the Borel $\sigma$-algebra, and $\mathbb{P}$ the corresponding Lebesgue probability measure.
-Let $\omega \in [-1, 1]$. Define random variables $T$ and $(Y_k)_{k \in \omega}$:
-\begin{align*}
-T(\omega) & := 
-\begin{cases}
-\min\set{n \in \mathbb{N} \mid \omega > 2^{-n}} & \hbox{if $\omega \in (0, 1]$}\\
-\infty & \hbox{otherwise}
-\end{cases} 
-\\
-Y_k(\omega) & := 
-\begin{cases}
-4-2^{-k} & \hbox{if $\omega \in [-1, 2^{-k}]$ (equivalently $k < T(\omega)$)}\\
-0 & \hbox{otherwise}
-\end{cases} 
-\end{align*}
-\lo{N.B.~$\set{T = n} = (\frac{1}{2^{n}}, \frac{1}{2^{n-1}}]$, and so, $\set{T > n} = [-1, \frac{1}{2^n}]$.}
-Plainly, $T$ is a stopping time, and $(Y_n)_{n \in \omega}$ is a supermartingale, adapted to the filtration $(\calF_n)_{n \in \omega}$ where $\calF_n = \calF$ for all $n$. 
-In this case, $(Y_n)_{n \in \omega}$ either tends to $4$ as $n \to \infty$, or drops to 0 at some point, with an exponentially decreasing probability. 
-With respect to the antitone function $\epsilon(x) = \frac{4-x}{4}$ and stopping time $T$, we have that $(Y_n)_{n \in \omega}$ is an antitone strict supermartingale, except that $\epsilon(4) = 0$, and $T = \infty$ with probability $\frac 1 2$, even though $4$ is larger than any value $(Y_n)_{n \in \omega}$ can actually take, and $\epsilon(Y_n)$ never actually reaches $0$.
-
-%\lo{@Andrew: I refer to the preceding highlighted claim. Did you mean: for all $n, s$, $\epsilon(Y_n(s)) > Y_n(s)$?} \akr{No. This phrasing should be clearer.}
-\end{example}
-
-\begin{theorem}[Antitone ranking function soundness] \label{thm:antitone rankable implies termination}
-If a closed SPCF term $M$ is antitone rankable, then $T_M^{\tY} < \infty$ a.s.~(equivalently, $M$ is AST).
-\end{theorem}
-\begin{proof}
-As in %Theorem \ref{thm:basicRankingSoundness}, 
-\Cref{thm:rankable implies termination},
-take the set $S$ of traces as a probability space, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$.
-As before, $(Y_n)_{n \in \omega}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$.
-Therefore, it follows from \Cref{thm:a-c strict} that $T_M^{\tY}$ is almost-surely finite.
-\end{proof}
-
-\paragraph{}
-As before, constructing antitone ranking fuctions completely is not necessary, and there is a corresponding notion of an antitone partial ranking function.
-
-Define an \emph{antitone partial ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$
-\begin{itemize}
-\item $f(N) \geq 0$ for all $N$ where $f$ is defined.
-\item $f$ is defined at $M$.
-\item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
-\end{itemize}
-
-\begin{theorem}  \label{thm:antitone partial implies rankable}
-  Every antitone partial ranking function is a restriction of an antitone ranking function.
-\end{theorem}
-\begin{proof}
-  Take a closed term $M$ and an antitone partial ranking function $f$ on $M$, with a corresponding antitone function $\epsilon$.
-Assume wlog that $\epsilon$ is convex (as if it isn't, we can just take its convex hull instead).
-As in Theorem \ref{thm:partial implies rankable}, define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
-  \begin{align*}
-    f_1(N) &= f(N) \text{ whenever $f(N)$ is defined},\\
-    f_1(V) &= 0 \text{ for values $V$ not in the domain of $f$.}
-  \end{align*}
-
-  Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) = \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
-  The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on antitone partial ranking functions.
-Define $f_2(N) = \int_S f_1(\nnext(N,s)) + \epsilon(f_1(\nnext(N,s))) g(N,s) \, \mu(\mathrm d s)$.
-For any term $N$ where $f$ is defined, $f_2(N) = f(N)$, and the value that $f_2$ would have at $N$ if $f$ were not defined at $N$ is $\leq f(N)$, by the third condition on antitone partial ranking functions.
-In order to show that $f_2$ is an antitone partial ranking function, it therefore suffices to show that the value that $f_2$ would have had at each term if $f$ were not defined at that term is at least the expectation of $f_2$ after one reduction step (plus $\epsilon(f_2(N))$ if the reduction step is a $tY$-reduction).
-For any term $N$ which is not of the form $E[R]$ for some $\tY$-redex $R$, this is trivial.
-If $R$ is a $\tY$-redex, then $\epsilon(f_2(N)) \leq \int_S \epsilon(f_1(\nnext(N,s)))\, \mu(\mathrm d s)$ by the convexity of $\epsilon$, because $n$ is bounded.
-Therefore, the (total) function $f_2$, which agrees with $f$ on $f$'s domain, is an antitone ranking function on $M$ (with the same function $\epsilon$ if it's convex).
-\end{proof}
-
-As a corollary, any term which admits an antitone partial ranking function terminates almost surely.
-
-\paragraph{\Cref{ex:ac-ranking} revisited}
-\akr{All of these ranking functions are slightly incorrect, because $\Xi \, \underline n$ reduces to something like $\Xi \, (\underline n + 1)$ then $(\lambda z.
-\text{something}[\Xi/f]) (\underline n + 1)$, rather than reducing the argument first.
-This could be fixed by appealing to the confluent semantics version of the antitone ranking theorem, but it would be nice to have at least some examples here illustrating how antitone ranking functions work, rather than actually saving all these examples until the end.
-
-Even though the difference is quite trivial, this was actually the original motivating example for the confluent trace semantics.}
-
-Programs $M_1$ and $M_2$ are AST. For $M_1$, the partial ranking function 
-\[
-\big(\tY \, \lambda f n . \, \tif{n=0}{0}{f \, (n-1) \oplus_{3/2} f \, (n+1)}\big) \, \underline x \mapsto 3x + 1
-\] 
-suffices to prove its termination (and could equivalently be considered an antitone partial ranking function with the constant antitone function $\epsilon_1(x) = 1$).
-
-For $M_2$, define the function $g_1 : \nnReal \to \nnReal$ by $g_1(x) = \log(x+1) + 1$.
-Using shorthand 
-$\Theta_2 = \tY \, \lambda f n . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
-we can define an antitone partial ranking function $f_2 : \mathit{Rch}(M_2) \rightharpoonup \nnReal$ 
-%with antitone function $\epsilon_2$ defined as follows (for $n \in \mathbb N$):
-defined as: for $n \in \mathbb N$
-\begin{align*}
-{\Theta_2} \, n 
-&\mapsto 
-g_2(n)
-\\
-0 &\mapsto 0.
-\end{align*}
-For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$, each with probability $1/2$, with one $\tY$-reduction.
-Now
-\begin{align*}
-  & g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 \\
-  = & \log \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) \\
-  = & \frac 1 2 \log\left(\frac{(n+1)^2}{n(n+2)}\right) \\
-  = & \frac 1 2 \log\left(1 + \frac 1 {n(n+2)}\right) \\
-  > & \frac 1 {4n(n+2)} \\
-  > & \frac 1 2 \frac 1 {5(n+1)^2} + \frac 1 2 \frac 1 {5(n+3)^2}
-\end{align*}
-and $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
-Therefore by setting $\epsilon_2(x) = \frac 1 {5(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met.
-It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone partial ranking function, and $M_2$ is AST.
-
-\begin{example}[Continuous random walk]\label{ex:raven complex}
-In SPCF we can construct a function whose argument changes by a random amount at each recursive call:
-\[
-\underbrace{\big
-(\tY \, \lambda f x . \tif{x \leq 0}{0}{f(x - \tsample)} \big)}_{\Theta} \, 10
-\]
-We can construct a partial ranking function $f$ as follows:
-\begin{align*}
-\Theta \, \underline l 
-&\mapsto 
-l + 2
-\\
-\tif{l \leq 0}{0}{\Theta \, (l - \tsample)}
-&\mapsto
-l + 1
-\\
-0 &\mapsto 0.
-\end{align*}
-
-For a more complex (not $\tY$-PAST) example, consider the following ``continuous random walk''
-\[
-\underbrace{\big
-(\tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample + 1/2)} \big)}_{\Xi} \, 10
-\]
-Let $g(x) := 2 + \log(x + 1)$, and let $\epsilon$ be the function specified by
-\[
-\epsilon(g(x-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y.
-\]
-The limit of this as $x \to \infty$ and $g(x+1/2) \to \infty$ is 0, and $\frac {\mathrm d}{\mathrm dx} \epsilon(g(x+1/2)) = \frac 1 {x+1} + \log(1 - \frac 1 {x + 1/2}) < 0$, and $g$ is monotonic increasing;
-therefore $\epsilon$ is antitone and bounded below by 0.
-We define a partial antitone ranking function as follows:
-\begin{align*}
-\Xi \, l 
-&\mapsto 
-g(l)
-\\
-0 &\mapsto 0
-\end{align*}
-\end{example}
-The value of $g$ after one $\tY$-reduction step is at least $g(l-1/2)$, therefore the expectation of $\epsilon$ after one $\tY$-reduction step is at most $\epsilon(g(l-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y$.
-Thus 
-\begin{itemize}
-\item in case $l > 0$, $g$ decreases by the required amount
-\item in case $l \leq 0$, 
-\(
-g(\Xi \, l) \geq 2 + \log(1/2) > \log(\frac{3 \sqrt 3} 2) - 1 = \epsilon(g(0-1/2))
-\) 
-as well.
-\end{itemize}
-Hence this is a valid partial antitone ranking function and the term is AST.
-
-\begin{example}[Fair-in-the-limit random walk]
-\label{ex:Fair-in-the-limit random walk}\citep[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}
-\[
-\underbrace{\big
-(\tY \, f \, x . 
-\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)}_{\Xi} 
-\, 10
-\]
-To construct an antitone ranking function, we solve the recurrence relation:
-\[
-\left\{
-\begin{array}{rll}
-z_0 &=& 0\\
-n \geq 1, \quad z_n &>& \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
-\end{array}
-\right.
-\]
-For $n > 2$, $z_n = \log(n-1)$ works.
-The expected decrease is $\frac 1 2(\log(1+\frac 1 {(n-1)^2-1}) - \frac 1 {2n+1} \log(1 + \frac 2 {n-2}))$, and using the fact that $\frac x {x+1} \leq \log (1+x) \leq x$ for $x > 0$, this is at least $\frac 1 2(\frac 1 {(n-1)^2} - \frac 1 {2n + 1} \frac 2 {n-2}) = \frac {n^2}{2(n-1)^2(2n+1)(n-2)}$, which (again for $n > 2$) is positive and antitone.
-For $n = 0, 1, 2$, we take $\epsilon$ to be 9/40 (the same as its value at 3), then set $z_2 = 2 \log 2 - \log 3 - 1, z_1 = 3 \log 2 - 2 \log 3 - 3, z_0 = 4 \log 2 - 3 \log 3 - 6$.
-Some of those values are negative, but it's still bounded below, so by adding a constant offset it can be corrected, and the term is AST.
-
-\end{example}
-
-\begin{example}[Escaping spline]
-\label{ex:escaping spline}\citep[\S 5.4]{DBLP:journals/pacmpl/McIverMKK18}
-\[
-\underbrace{\big
-(\tY \, f \, x . \,
-0 \oplus_{\frac{1}{x+1}} f(x + 1) \big)}_{\Xi} 
-\, 10
-\]
-
-In this case, the fact that the ranking function must decrease at each $\tY$-step (in an antitone partial ranking function) by the expected value of $\epsilon$ not at the current term, but at the next term where the ranking function is defined, is a little harder to deal with, because the variable $x$ can change all the way to $0$ in one step, therefore simply adding a small offset doesn't suffice to compensate for this fact.
-
-Consider the candidate ranking function $\Xi \, n \mapsto n + 1$.
-For each $n$, $\Xi \, n$ reduces to either $0$ (with probability $\frac 1 {n + 1}$) or $\Xi \, (n + 1)$, therefore the expected value of the ranking function is $\frac{n(n+2)}{n+1} = n + 1 - \frac 1 {(n+1)^2}$, and the required decrease is $\frac{\epsilon(0) + n \epsilon(n + 2)}{n + 1}$, therefore eventually, the expected decrease isn't enough, whatever the value of $\epsilon(0)$ is.
-
-If instead, we take the ranking function to be defined after the $\tY$-reduction but before the $\tsample$-reduction as well, this can be resolved.
-Letting $\Theta[n] = \underline 0 \oplus_{\frac 1 n} \Xi \, n$:
-\begin{align*}
-\Theta[n] &\mapsto n \\
-\Xi \, 10 & \mapsto 12.
-\end{align*}
-For each $n$, $\Theta[n]$ reduces to either $0$ or $\Theta[n+1]$, with a $\tY$-reduction only in the latter case, and the condition that this is an antitone partial ranking function is that $n \geq \frac{(n-1)(n+1)}{n} + \frac{n-1}{n} \epsilon(n+1)$ and $12 \geq 11 + \epsilon(11)$.
-These are satisfied by setting $\epsilon(x) = \min(1, 1/x)$, therefore this term is AST.
-\end{example}
-
-\paragraph{Completeness}
-It seems very likely that this method is complete in the case that every reachable term is AST, but we have been unable to actually prove this.
-It is certainly at least capable of proving the termination of terms which terminate arbitrarily slowly (in the sense that there is no similar limitation to Theorem \ref{thm:rankable implies termination}, which can only prove termination of $\tY$-PAST terms).
-
-The following theorem does not prove completeness, but is suggestive in that direction:
-\begin{theorem}
-For any stopping time $T$ which is almost-surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
-\end{theorem}
-
-\input{relativised-completeness}
-
-\input{y-stopping-time}
-
-\section{Confluent Trace Semantics}
-When proving almost sure termination in this way, it is necessary to consider which terms a given term may reduce to.
-Sometimes however, the reduction that the programmer has in mind may not be strictly the call-by-value order defined so far, or considering an alternative reduction order may be simpler or more intuitive.
-
-Non-probabilistic lambda calculi generally have the Church-Rosser property, that if a term $A$ reduces to both $B_1$ and $B_2$, there is some $C$ 
-%with reduction sequences $B_1 \to^* C$ and $B_2 \to^* C$, 
-to which both $B_1$ and $B_2$ reduce,
-so the reduction order mostly doesn't matter. 
-In the probabilistic case, this may not be true, because $\beta$-reduction can duplicate $\tsample$s, so the outputs of the copies of the sample may be identical or independent, depending on whether the sample is taken before or after $\beta$-reduction.
-There are, however, some restricted variations on the reduction order that do not have this problem.
-
-\paragraph{}
-Even with this restriction, a sampling semantics in the style of the one already defined would not be entirely Church-Rosser, as, for example, $\red^3(\tsample - \tsample, (1,0,\dots))$ would be either $1$ or $-1$ depending on the order of evaluation of the $\tsample$s, as that determines which sample from the pre-selected sequence is used for each one.
-To fix this, rather than pre-selecting samples according to the order they'll be drawn in, select them according to the position in the term where they'll be used instead.
-
-A \emph{position} 
-%\akr{I'm not really sure how to phrase the citation, but this is adapted from \cite{Kennaway96infinitarylambda}.} 
-%\lo{The idea is common enough. I think no citation necessary.}
-is a finite sequence of steps into a term, defined inductively as
-\begin{align*}
-\alpha ::= \cdot \mid \lambda ; \alpha \mid @_1 ; \alpha \mid @_2 ; \alpha \mid \underline f_i ; \alpha \mid \tY ; \alpha \mid \textsf{if}_1 ; \alpha \mid \textsf{if}_2 ; \alpha \mid \textsf{if}_3 ; \alpha \mid \tscore ; \alpha.
-\end{align*}
-The \emph{subterm of $M$ at a position $\alpha$}, denoted $M \mid \alpha$, is defined as
-\begin{align*}
-M \mid \cdot & = M \\
-\lambda x. M \mid \lambda ; \alpha & = M \mid \alpha \\
-M_1 M_2 \mid @_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2 \\
-\underline f(M_1,\dots,M_n) \mid \underline f_i ; \alpha & = M_i \mid \alpha \quad \text{for }i \leq n \\
-\tY M \mid \tY ; \alpha & = M \mid \alpha \\
-\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2,3 \\
-\tscore(M) \mid \tscore ; \alpha & = M \mid \alpha
-\end{align*}
-so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda$ is undefined). 
-A position such that $M\mid \alpha$ does exist is said to \emph{occur} in $M$. 
-%\changed[lo]{
-\emph{Substitution} of $N$ at position $\alpha$ in $M$, written $M[N/\alpha]$, is defined similarly.
-For example, let 
-\[
-M = \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})
-\qquad
-\alpha =\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2
-\]
-then 
-\(
-M[\tsample / \alpha] = \lambda x \, y. y \, (\tif{x < 0}{y \, \tsample}{\underline 3}).
-\)
-%}
-
-Two subterms $N_1$ and $N_2$ of a term $M$, corresponding to positions $\alpha_1$ and $\alpha_2$, can overlap in a few different ways. 
-If $\alpha_1$ is an initial segment (i.e.~prefix) of $\alpha_2$ (written as $\alpha_1 \leq \alpha_2$), then $N_2$ is also a subterm of $N_1$.
-If neither $\alpha_1 \leq \alpha_2$ nor $\alpha_1 \geq \alpha_2$, the positions are said to be \emph{disjoint}.
-The notion of disjointness is mostly relevant in that if $\alpha_1$ and $\alpha_2$ are disjoint, performing a substitution at $\alpha_1$ will leave the subterm at $\alpha_2$ inaffected.
-\lo{The preceding sentence is unclear. Typo?} \akr{It wasn't a typo, but I've made it a little more explicit now. Is this clearer?}
-
-With this notation, a more general reduction relation $\to$ can be defined. \akr{It would almost work to have one of the neater versions of $\tY$-reduction here.}
-%\lo{On first reading, I struggled to parse the following clauses.}
-\begin{definition}
-\label{def:more general red}
-The binary relation $\to$ is defined by the following rules, each is conditional on a redex occurring at position $\alpha$ in the term $M$:
-\begin{align*}
-  \text{if } M \mid \alpha = (\lambda x.N) V,\ & M \to M[N[V/x]/\alpha] \\
-  \text{if } M \mid \alpha = \underline f (\underline r_1, \dots , \underline r_n),\ & M \to M[\underline{f(r_1,\dots,r_n)}/\alpha] \\
-  \text{if } M \mid \alpha = \tY \lambda x. N,\ & M \to M[\lambda z. N[(\tY \lambda x. N)/x] z/\alpha] \text{ where $z$ is not free in $N$}\\
-  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_1/\alpha] \text{ where }r < 0 \\
-  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_2/\alpha] \text{ where }r \geq 0 \\
-  \text{if } M \mid \alpha = \tsample & \text{ and $\lambda$ does not occur after $@_2$ or $\tY$ in $\alpha$},\\ & M \to M[\underline r/\alpha] \text{ where } r \in [0,1] \\
-  \text{if } M \mid \alpha = \tscore(\underline r),\ & M \to M[\underline r/\alpha].
-\end{align*}
-In each of these cases, $M \mid \alpha$ is the \emph{redex}, and reduction takes place at $\alpha$.
-\end{definition}
-
-\paragraph{}
-Labelling the pre-chosen samples by the positions in the term would also not work because in some cases, a $\tsample$ will be duplicated before being reduced, for example, in $(\lambda x.
-x \, {\underline 0} \, \mathbin{\underline{+}} \, {x \, \underline 0}) (\lambda y.
-\tsample)$, both of the $\tsample$ redexes that eventually occur originate at $@_2 ; \lambda$.
-It is therefore necessary to consider possible positions that may occur in other terms reachable from the original term. 
-Even this is itself inadequate because some of the positions in different reachable terms need to be considered the same, and the number of reachable terms is in general uncountable, which leads to measure-theoretic issues.
-
-Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $Sk(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$.
-\lo{Do we actually use the notation $S[r]$ in what follows?} \akr{Yes, in Lemma \ref{lem:independentSamples} for example.}
-\lo{Skeleton is defined in Sec.~\ref{sec:sampling semantics}.}
-The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, and $\tsample$ reducing to $\skeletonPlaceholder$.
-For example, we have 
-\(
-(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \tsample
-\to
-(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \skeletonPlaceholder
-\to
-\tif{\skeletonPlaceholder < 0}{\skeletonPlaceholder}{\skeletonPlaceholder}
-\to
-\skeletonPlaceholder
-\)
-
-Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a $\to$-reduction sequence of skeletons starting at $Sk(M)$, and the second of which is a position in the final skeleton of the reduction sequence.
-As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the usual linear sampling semantics, there will be (with one more caveat introduced later) a sample from $I$ pre-selected for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.
-
-Reduction sequences are used rather than reachable skeletons because if the same skeleton is reached twice, different samples may be needed.
-For example, consider the term $M = \tY (\lambda f x.\ \tif{\tsample - \underline{0.5} < 0}{f \, x}{x}) \, \underline 0$, which reduces after a few steps to $N = \tif{\tsample - \underline{0.5}  < 0}{M}{\underline 0}$.
-If the pre-selected sample for $(N,\textsf{if}_1;\underline{-}_1)$ is less than 0.5, $N$ reduces back to $M$, then $N$ again, then the same sample is used the next time, therefore it's an infinite loop, whereas if samples are labelled by reduction sequences, the samples for $M \to \dots \to N$ are independent from the samples for $M \to \dots \to N \to M \to \dots \to N$, and so on.
-
-The reduction sequences of skeletons will often be discussed as though they were just skeletons, identifying them with their final skeletons.
-With this abuse of notation, a reduction sequence $N$ (actually $N_1 \to \dots \to N_n = N$) may be said to reduce to a reduction sequence $O$, where the reduction sequence implicitly associated with the final skeleton $O$ is $N_1 \to \dots \to N_n \to O$.
-
-\paragraph{}
-This is still not quite sufficient because sometimes the same samples must be used at corresponding positions in different reduction sequences.
-For example, the term $M = \tsample + \tsample$ has the reachable skeletons $N_1 = \skeletonPlaceholder + \tsample, N_2 = \tsample + \skeletonPlaceholder, O = \skeletonPlaceholder + \skeletonPlaceholder$ and $\skeletonPlaceholder$, with reductions $M \to N_1 \to O \to \skeletonPlaceholder$ and $M \to N_2 \to O \to \skeletonPlaceholder$.
-In the reduction $M \to N_1$, the sample labelled $(M, \underline{+}_2)$ is used, and in the reduction $N_2 \to O$, the sample labelled $(M \to N_2, \underline{+}_2)$ is used.
-Each of these samples becomes the value of the second numeral in $O$ in their respective reduction sequences, therefore in order for Church-Rosserness to be attained, they must be the same.
-Which elements of $L_0(M)$ must match can be described by the relation $\sim^*$:
-
-\begin{definition}
-The relation $\sim$ is defined as the union of the minimal symmetric relations $\sim_p$ (``$p$'' for parent-child) and $\sim_c$ (``$c$'' for cousin) satisfying
-\begin{enumerate}
-    \item If $N$ reduces to $O$ with the redex at position $\alpha$, and $\beta$ is a position in $N$ disjoint from $\alpha$, then $(N,\beta) \sim_p (O,\beta)$.
-
-    \item If $N$ $\beta$-reduces to $O$ at position $\alpha$, $\beta$ is a position in $N \mid \alpha;@_1;\lambda$ and $N \mid \alpha;@_1;\lambda;\beta$ is not the variable involved in the reduction, $(N,\alpha;@_1;\lambda;\beta) \sim_p (O, \alpha;\beta)$
-
-    \item If $N$ $\textsf{if}$-reduces to $O$ at position $\alpha$, with the first resp.
-second branch being taken, and $\alpha;\textsf{if}_i;\beta$ occurs in $N$ (where $i = 2$ resp.
-$3$), $(N,\alpha;\textsf{if}_i;\beta) \sim_p (O,\alpha;\beta)$
-
-    \item If $N$, $O_1$ and $O_2$ match any of the following cases:
-    \begin{enumerate}
-        \item $N$ contains redexes at disjoint positions $\alpha_1$ and $\alpha_2$, $O_1$ is $N$ reduced first at $\alpha_1$ then $\alpha_2$ and $O_2$ is $N$ reduced first at $\alpha_2$ then at $\alpha_1$.
-
-        \item $N \mid \alpha = \tif{\underline r < 0}{N_1}{N_2}$, where $r < 0$ (or, respectively, $r \geq 0$), $(N_2 \text{ resp.
-} N_1) \mid \beta$ is a redex, and $O_1$ is $N$ reduced at $\alpha$ and $O_2$ is $N$ reduced first at $\alpha;(\textsf{if}_3 \text{ resp.
-} \textsf{if}_2);\beta$ then at $\alpha$
-
-        \item $N \mid \alpha = \tif{\underline r < 0}{N_1}{N_2}$, where $r < 0$ (or, respectively, $r \geq 0$), $(N_1 \text{ resp.
-} N_2) \mid \beta$ is a redex, and $O_1$ is $N$ reduced first at $\alpha$ then at $\alpha;\beta$ and $O_2$ is $N$ reduced first at $\alpha;(\textsf{if}_2 \text{ resp.
-} \textsf{if}_3);\beta$ then at $\alpha$
-
-        \item $N \mid \alpha = (\lambda x.
-A) B$, there is a redex in $A$ at position $\beta$, $O_1$ is $N$ reduced first at $\alpha$ then at $\alpha;\beta$, and $O_2$ is $N$ reduced first at $\alpha;@_1;\lambda;\beta$ then at $\alpha$
-
-        \item $N \mid \alpha = (\lambda x.
-A) B$, $B \mid \beta$ is a redex, $(\gamma_i)_i$ is a list of all the positions in $A$ where $A \mid \gamma = x$, ordered from left to right, $O_1$ is $N$ reduced first at $\alpha ; @_2 ; \beta$ then at $\alpha$, and $O_2$ is $N$ reduced first at $\alpha$ then at $\alpha;\gamma_i;\beta$ for each $i$ in order.
-
-        \item $N \mid \alpha = \tY (\lambda x.
-A)$, $A$ reduced at $\beta$ is $A'$, $(\gamma_i)_i$ is a list of all the positions where $A' \mid \gamma  = x$, ordered from left to right, $O_1$ is $N$ reduced first at $\alpha;\tY;\lambda;\beta$ then at $\alpha$, and $O_2$ is $N$ reduced first at $\alpha$ then at $\alpha;\lambda;@_1;\gamma_i;Y;\lambda;\beta$ for each $i$ in order where $\gamma_i$ is left of $\beta$ then at $\alpha;\lambda;@_1;\beta$ then at $\alpha;\lambda;@_1;\gamma_i;Y;\lambda;\beta$ for the remaining values of $i$.
-    \end{enumerate}
-    (in which case $O_1$ and $O_2$ are equal as skeletons, but with different reduction sequences), $O_1'$ and $O_2'$ are the results of applying some reduction sequence to each of $O_1$ and $O_2$ (the same reductions in each case, which is always possible because they're equal terms), and $\delta$ is a position in $O_1'$ (or equivalently $O_2'$), then $(O_1',\delta) \sim_c (O_2',\delta)$.
-\end{enumerate}
-The $\sim_c$-rules are illustrated in the following diagrams.
-\end{definition}
-
+% !TEX root = main.tex
+The rules defining $\sim_c$ can be more intuitively understood as diagrams.
+In the following diagrams, a circle represents a term (or skeleton), and the leaves on the trees are subterms at the positions indicated by the labels in the other nodes.
+The labels on the arrows between trees are the positions of the reductions, and dashed arrows represent some number of reductions in a row.
+If a reduction sequence includes one branch of one of these cases at some point in it, then the positions in the reduction sequence obtained by substituting in the other branch are all related by $\sim_c$ to the corresponding positions in the original reduction sequence.
 \include{sim-diagram}
-
-The reflexive transitive closure $\sim^*$ of this relation is used to define the set of \emph{potential positions} $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable skeletons.
-If $(N,\alpha) \sim^* (O,\beta)$, then $N \mid \alpha$ and $O \mid \beta$ both have the same shape (i.e.~they're either both the placeholder $\skeletonPlaceholder$, both variables, both applications, both $\tsample$s etc.), therefore it's well-defined to talk of the set of potential positions where there is a $\tsample$, $L_s(M)$.
-Formally $L_s(M) := \{[(X, \alpha)]_{\sim^\ast_M} : X \mid \alpha = \tsample\}$.
-The new sample space is then defined as $I^{L_s(M)}$, with the Borel $\sigma$-algebra and product measure.
-\changed[lo]{Since $I^{L_s(M)}$ is a countable product, the measure space is well-defined \citep[Cor.~2.7.3]{AshDD00}.}
+%The diagram is not actually in the figure environment because that doesn't work properly with it extending over several pages. The empty figure is just needed to get the label and caption.
+\begin{figure}[htb]
+\label{sim-diagram}
+\caption{Illustration of the $\sim_c$ rules.}
+\end{figure}
 
@@
 Case a is also similar, except that the relevant positions are all disjoint rather than contained in one another, and either branch could be considered the right branch.
-
-\iffalse
-\changed[lo]{Since $L_s(M)$ is a countable set, there is a unique probability measure $\mu$ on the infinite product $\prod_{i \in L_s(M)} \Sigma_I$ of the Borel $\sigma$-algebras $\Sigma_I$ satisfying
-\[
-\mu\set{\omega \in \Omega \mid \omega_1 \in A_1, \cdots, \omega_n \in A_n} = \prod_{i=1}^n \Leb(A_i)
-\]
-where $\Leb$ is the Lebesgue measure on the measurable space $(I, \Sigma_I)$.}
-\fi
-
-\paragraph{}
-Before defining the new version of the reduction relation $\red$, a few lemmas about the properties of $\sim$ are necessary for it to be well-defined.
 
@@
 \end{lemma}
-
-\lo{As defined, $\sim_c$ is not symmetric.}
 
@@
 
-\begin{lemma}
-\label{lem:sim-M-N}
-The relation $\sim$ is defined on $L_0(M)$ with reference to a particular starting term $M$, so different versions, $\sim_M$ and $\sim_N$, can be defined starting at different terms.
-If $M \to N$, then $\sim^*_N$ is equal to the restriction of $\sim^*_M$ to $L_0(N)$.
-\end{lemma}
+\lemmaSimMN*
 
@@
 
-At each reduction step $M \to N$, the sample space must be restricted from $I^{L_s(M)}$ to $I^{L_s(N)}$. 
-The injection $L_0(N) \to L_0(M)$ is trivial to define by appending $Sk(M) \to Sk(N)$ to each path, and using Lemma~\ref{lem:sim-M-N}, this induces a corresponding injection on the quotient, $L(N) \to L(M)$.
-The corresponding map $L_s(N) \to L_s(M)$ is then denoted $i(M \to N)$.
+\churchRosser*
 
-Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function, so it is denoted ``$\Rightarrow$'' instead of ``$\red$'', and it relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
-\changed[lo]{We write an element of $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ as $(M', s)$ where the \emph{term} $M' \in \Lambda^0$ and $s \in  I^{L_s(M')}$.}
-\begin{align*}
-& (M,s) \Rightarrow (N,s \circ i(M \to N)) \text{ if either} \\
-& \qquad \text{$M \to N$ and the redex is not $\tsample$, or} \\
-& \qquad \text{$M \mid \alpha = \tsample$, $N = M[\underline{s(Sk(M),\alpha)}/\alpha]$ and $\lambda$ does not occur after $@_2$ or $\tY$ in $\alpha$}
-\end{align*}
-
-\paragraph{}
-\begin{lemma} \label{churchRosser}
-The relation $\Rightarrow$ is Church-Rosser.
-\end{lemma}
 \begin{proof}
@@
 
-\paragraph{}
-The reduction relation $\Rightarrow$ is nondeterministic, so it admits multiple possible reduction strategies. 
-A \emph{reduction strategy} starting from a closed term $M$ is a partial function $f$ from $Rch(M)$ to positions, such that for any reachable term $N$ where $f$ is defined, $f(N)$ is a position of a redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
-Using a reduction strategy $f$, a subset of $\Rightarrow$ that isn't nondeterministic, $\Rightarrow_f$, can be defined by $(N,s) \Rightarrow_f (N',s')$ just if $(N,s) \Rightarrow (N',s')$ and $N$ reduces to $N'$ with the redex at $f(N)$.
-
 Essentially what Lemma~\ref{lem:redexDestroyed} demonstrates is that the samples taken during any reduction sequence are independent of each other. This is made more precise in the following lemmas.
@@
 \end{proof}
+\clearpage
 
-The usual call-by-value semantics can be implemented as one of these reduction strategies, given by (with $V$ a value and $T$ a term that isn't a value and $M$ a general term)
-\begin{align*}
-\cbv(T M) & = @_1 ; \cbv(T) \\
-\cbv(V T) & = @_2 ; \cbv(T) \\
-\cbv(\underline f(V_1, \dots, V_{k-1}, T, M_{k+1}, \dots, M_n)) & = \underline f_k ; \cbv(T) \\
-\cbv(\tY T) & = \tY ; \cbv(T) \\
-\cbv(\tif{T < 0}{M_1}{M_2}) & = \textsf{if}_1 ; \cbv(T) \\
-\cbv(\tscore(T)) & = \tscore ; \cbv(T) \\
-\cbv(V) & \text{ is undefined} \\
-\cbv(T) & = \cdot \text{ otherwise}
-\end{align*}
-(this last case covers redexes at the root position).
-
-A closed term $M$ terminates with a given reduction strategy $f$ and samples $s$ if there is some natural number $n$ such that $(M,s) \Rightarrow_f^n (N,s')$ where $f$ gives no reduction at $N$.
-The term terminates almost surely with respect to $f$ if it terminates with $f$ for almost all $s$.
-
-\begin{theorem} \label{thm:AstEquivalence}
-A closed term $M$ is AST with respect to $\cbv$ iff it is AST.
-\end{theorem}
+\AstEquivalence*
 \begin{proof}
@@
 
-\begin{theorem} \label{thm:CbvIsTerminatingest}
-If $M$ terminates with some reduction strategy $f$ and samples $s$, it terminates with $\cbv$ and $s$.
-\end{theorem}
+\CbvIsTerminatingest*
 \begin{proof}
@@
 \end{proof}
-
-\begin{corollary}
-If $M$ is AST with respect to any reduction strategy, it is AST.
-\end{corollary}
-\begin{proof}
-Suppose $M$ is AST with respect to $f$.
-Let the set of samples with which it terminates with this reduction strategy be $X$.
-By Theorem~\ref{thm:CbvIsTerminatingest}, $M$ also terminates with $\cbv$ and every element of $X$, and $X$ has measure 1, by assumption, therefore $M$ is AST with respect to $\cbv$ therefore by Theorem~\ref{thm:AstEquivalence} it is AST.
-\end{proof}
-
-All of the theorems on the termination of rankable terms therefore extend to other reduction strategies too.
-The proofs of Theorems \ref{thm:rankable implies termination}, \ref{thm:partial implies rankable}, \ref{thm:antitone rankable implies termination} and \ref{thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies.
-They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because Theorem \ref{thm:de groote} applies equally to any reduction strategy).
-For a reduction strategy $r$ and term $M$, just substitute $N[r(N)]$ being a $\tY$-redex for $N$ being of the form $E[\tY \lambda x.
-O]$, and $r(N)$ being undefined for $N$ being a value.
-
-The domain of definition of the ranking functions also needs to be changed from the reachable terms $Rch(M)$ to the \emph{reachable terms with respect to the reduction strategy $r$},
-\[
-Rch_r(M) = \{N \mid \exists n, N_0, \dots, N_n: N_0 = M, N_n = N, N_i \to N_{i+1} \text{ at } r(N_i)\}.
-\]
-
-To be more explicit, the modified forms of the theorems are:
-\begin{definition}\rm
-A \emph{ranking function on $M$ with respect to a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
-\begin{itemize}
-    \item $f(N) \geq 1+ f(N')$ if $N[r(N)]$ is a $\tY$-redex and $N$ reduces to $N'$ at $r(N)$
-    \item $f(N) \geq \int_I f(N[\underline{x}/r(N)]) \, \Leb(\mathrm{d}x)$ if $N[r(N)] = \tsample$
-
-    \item $f(N) \geq f(N')$ if $r(N)$ is any other redex, where $N \to N'$ at $r(N)$.
-\end{itemize}
-Any closed term for which a ranking function with respect to a reduction strategy $r$ exists is called \emph{rankable} with respect to $r$. 
-\end{definition}
-
-\begin{definition}
-A \emph{partial ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : Rch_r(M) \rightharpoonup \mathbb R$ such that
-\begin{itemize}
-    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
-    \item $f$ is defined at $M$.
-    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ at the positions specified by $r$ will eventually reach some $O$ such that either $r(O)$ isn't defined or $f(O)$ is, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
-\end{itemize}
-\end{definition}
-
-\begin{definition}\rm
-An \emph{antitone ranking function on $M$ with respect to a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ such that
-\begin{itemize}
-    \item $f(N) \geq \epsilon(f(N)) + f(N')$ if $N[r(N)]$ is a $\tY$-redex and $N$ reduces to $N'$ at $r(N)$
-    \item $f(N) \geq \int_I f(N[\underline{x}/r(N)]) \, \Leb(\mathrm{d}x)$ if $N[r(N)] = \tsample$
-
-    \item $f(N) \geq f(N')$ if $r(N)$ is any other redex, where $N \to N'$ at $r(N)$.
-\end{itemize}
-Any closed term for which an antitone ranking function with respect to a reduction strategy $r$ exists is called \emph{antitone rankable} with respect to $r$. 
-\end{definition}
-
-\begin{definition}
-An \emph{antitone partial ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : Rch_r(M) \rightharpoonup \mathbb R$ such that there exists some antitone function $\epsilon : \nnReal \to \pReal$ such that
-\begin{itemize}
-    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
-    \item $f$ is defined at $M$.
-    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ at the positions specified by $r$ will eventually reach some $O$ such that either $r(O)$ isn't defined or $f(O)$ is, and $f(N) \geq \mathbb E[f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
-\end{itemize}
-\end{definition}
-
-\begin{theorem}
-If a closed SPCF term $M$ has a ranking function, partial ranking function, antitone ranking function or antitone partial ranking function with respect to a reduction strategy $r$, then $M$ is AST with respect to $r$, and AST.
-\end{theorem}
-\lo{Give an example or two of a reduction strategy (for $\Rightarrow$) that is not $\cbv$.}
-
-\akr{NOTE. For the standard CBV semantics, this example is indeed not a problem. 
-In the alternative semantics, defining 
-\begin{align*}
-A[x] &= \tif{\tif{x>0}{I}{I} \, \tsample - 0.5 > 0}{0}{\Omega}\\
-B &= \tif{\tsample - 0.5 > 0}{0}{\Omega}
-\end{align*}
-The set of samples where it terminates is 
-\[\begin{array}{l}
-\bigcup_{r \in [0,1]} \{s \in I^{L_s(A[\tsample])} \mid s([A[\tsample]], \mathsf{if}_1;\underline{ -
-}_1;@_1;\mathsf{if}_1) = r, \\
-\qquad \qquad \hfill \qquad \qquad s([A[\tsample], A[r], \ldots, B, \mathsf{if}_1) > 0.5 \}.
-\end{array}\] 
-This is a rather unwieldy expression, but the crucial part is that $r$ occurs twice in the conditions on $s$: once as the value a sample must take, and once in the location of a sample. 
-It is this set that's unmeasurable.}
-
-\bibliographystyle{apalike}
-\bibliography{references}
-
-\lo{22 Jan: 
-\paragraph{Further directions}
-
-An obvious next step is to extend the result to the $\mathsf{score}$ construct.
-
-The following are highly topical, and could form the basis of an interesting and novel DPhil thesis.
-\begin{enumerate}
-\item Devise methods for proving (positive) a.s.~termination.
-
-- For example, develop a type system satisfying the property: if a term is typable then it is (positively) a.s.~terminating.
-See~\citep{DBLP:conf/ppdp/BreuvartL18,DBLP:conf/esop/LagoG17}.
-
-- Another approach is to develop algorithms that synthesise ranking supermartingales, following, for example, \citep{DBLP:journals/pacmpl/AgrawalC018}.
-
-\item Develop principles (e.g.~in the form of ``proof rules'') for reasoning about (positively) a.s.~termination, in the style of \citep{DBLP:journals/pacmpl/McIverMKK18}.
-
-\item Design algorithms that synthesise probabilistic invariants (\emph{qua} martingales), \`a la \cite{SchreuderO19}; see also \citep{DBLP:journals/pacmpl/HarkKGK20}.
-
-\end{enumerate}}
-
-\iffalse
-\begin{thebibliography}{9}
-\bibitem{ppcf} Thomas Ehrhard, Michele Pagani, and Christine Tasson.
-Measurable cones and stable, measurable functions: a model for probabilistic higher-order programming.
-\emph{PACMPL}, 2(POPL):59:1–59:28, 2018.
-doi: 10.1145/3158147.
-URL \href{https://doi.org/10.1145/3158147}{https://doi.org/10.1145/3158147}.
-\end{thebibliography}
-\fi
-
-\end{document}

### 2021-01-01 LICS21/ranking.tex

@@
 
-\subsection{Partial ranking functions}
+\subsection{Sparse ranking functions}
 Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost-sure termination, so it isn't useful for proving it.
@@
 \begin{definition}
-Define a \emph{partial ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \rightharpoonup \mathbb R$ such that
+Define a \emph{sparse ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \rightharpoonup \mathbb R$ such that
 \begin{itemize}
@@
     \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually \akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and ast, but that doesn't actually work because of terms reachable with 0 probability.
-Of course, astness doesn't imply rankability, so any definition of partial ranking function that includes all Y-past terms wouldn't always extend to a total ranking function.} reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
+Of course, astness doesn't imply rankability, so any definition of sparse ranking function that includes all Y-past terms wouldn't always extend to a total ranking function.} reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
 \end{itemize}
 \end{definition}
-A partial ranking function that is total is just a ranking function.
-Providing a partial ranking function is essentially part way between providing a ranking function and directly proving almost sure termination.
+A sparse ranking function that is total is just a ranking function.
+Providing a sparse ranking function is essentially part way between providing a ranking function and directly proving almost sure termination.
 
 \begin{theorem} \label{thm:partial implies rankable}
-Every partial ranking function is a restriction of a ranking function.
+Every sparse ranking function is a restriction of a ranking function.
 \end{theorem}
 \begin{proof}
-Take a closed term $M$ and partial ranking function $f$ on $M$. Define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
+Take a closed term $M$ and sparse ranking function $f$ on $M$. Define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
 \begin{align*}
@@
 Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) := \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
-The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on partial ranking functions.
+The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on sparse ranking functions.
 Define $f_2(N) = \int_\entrosp f_1(\nnext(N,s)) + g(N,s) \, \mu_{\entrosp}(\mathrm d s)$.
@@
 
-As a corollary, any term which admits a partial ranking function terminates almost surely.
+As a corollary, any term which admits a sparse ranking function terminates almost surely.
 
@@
 \end{align*}
-A partial ranking function could be defined with respect to this shortcut reduction simply by replacing $\to$ and $\red$ in the definition of a partial ranking function by a version that goes straight from $N \oplus_p O$ to $N$ or $O$.
-Such a pseudo-partial ranking function would then be a partial function from a subset of $Rch(M)$, so it could also be considered as a partial function from all of $Rch(M)$, and it would in fact also be an actual partial ranking function.
+A sparse ranking function could be defined with respect to this shortcut reduction simply by replacing $\to$ and $\red$ in the definition of a sparse ranking function by a version that goes straight from $N \oplus_p O$ to $N$ or $O$.
+Such a pseudo-sparse ranking function would then be a partial function from a subset of $Rch(M)$, so it could also be considered as a partial function from all of $Rch(M)$, and it would in fact also be an actual sparse ranking function.
 It is therefore possible to prove rankability directly using the shortcutted reductions.
@@
 A similar procedure would work for other forms of syntactic sugar.
-If a closed term $N$ eventually reduces to one of a set of other terms $\{N_i \mid i \in I\}$ with certain probabilities, a partial ranking function defined with respect to a reduction sequence that skips straight from $N$ to $N_i$ is also a valid partial ranking function for the original reduction function, and therefore its existence implies almost sure termination.
+If a closed term $N$ eventually reduces to one of a set of other terms $\{N_i \mid i \in I\}$ with certain probabilities, a sparse ranking function defined with respect to a reduction sequence that skips straight from $N$ to $N_i$ is also a valid sparse ranking function for the original reduction function, and therefore its existence implies almost sure termination.
 There is a caveat, however, that $\tY$-reduction steps skipped over in the shortcut still need to be counted for the expected number of $\tY$-reduction steps.
@@
 With this abbreviation, the geometric distribution example from earlier can be written as $(\tY \lambda f n.\ n \oplus_{0.5} f(n+1)) \, \underline 0$.
-It is then easy to see that the following is a partial ranking function:
+It is then easy to see that the following is a sparse ranking function:
 \begin{align*}
@@
 
-In fact, even the partial function $\tY \Theta N \mapsto 2$ alone is a partial ranking function for this term.
+In fact, even the partial function $\tY \Theta N \mapsto 2$ alone is a sparse ranking function for this term.

### 2021-01-01 LICS21/supermartingales.tex
Minor clarifications

--- 2020-12-29 LICS21/supermartingales.tex
@@
 \lo{It is confusing (strictly speaking, incorrect) to say that $(M_n)_{n \geq 0}$ is a reduction sequence from $M$. Formally $M_n$ is the random variable $M_n: s \mapsto \pi_0 (\red^n(M, s))$.}
+\akr{I guess it's implicitly identifying a sequence of functions ($(M_n)_{n \geq 0}$), and a function producing sequences (a reduction-sequence-valued random variable), but this sort of thing seems very standard in dealing with random variables (e.g.~applying a function to some random variables, which is really a sort of concatenation).}
 

### 2021-01-01 LICS21/ranking.tex

@@
 The term  $(\tY \, \Theta) \, \underline 0$ generates a geometric distribution.
-\changed[akr]{Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}.}
+\changed[akr]{Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.}
 \lo{\Cref{fig:geometric distribution} is confusing: the arrow means either one-step reduction, or its reflexive transitive closure, or ``has the same syntactic shape''.}
+
+\akr{The arrows all mean the one-step reduction $\to$, I just didn't intend for the values of $i$ and $r$ to be consistent between the terms.
+I would much rather have the diagram be complete (with all the reductions and all the terms).}
 
@@
     \node (5) at (0,-4) {$\tif{\underline r\, \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (6) at (0,-5) {$\tif{\underline r < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (7) at (2,-6) [text width=] {$\underline i$};
+    \node (6) at (0,-5) {$\tif{\underline {r - 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
+    \node (7) at (2,-6.3) [text width=] {$\underline i$};
     \node (8) at (0,-7) {$\tY\, \Theta\, (\underline i\, \underline{+ 1})$};
@@
 
-    \draw [->] (1.south west) to [bend right=45] (2.north west);
+    \draw [->] (1.south west) to [bend right=45, "$i := 0$"'] (2.north west);
     \draw [->] (2.south west) to [bend right=45] (3.north west);
@@
     \draw [->] (5.south west) to [bend right=45] (6.north west);
-    \draw [->] (6.south west) to [bend right=45] (8.north west);
+    \draw [->] (6.south west) to [bend right=45, "$r \geq 0.5$"] (8.north west);
     \draw [->] (8.south west) to [bend right=45] (9.north west);
 
-    \draw [->] (9.west) to [bend left=45] (2.west);
-    \draw [->] (6) to (7);
+    \draw [->] (9.west) to [bend left=45, "$i := i + 1$" {above,sloped}] (2.west);
+    \draw [->] (6) to ["$r < 0.5$"] (7);
 \end{tikzpicture}
-\caption{Some reachable terms of $\tY \, \Theta \, \underline 0$\label{fig:geometric distribution}}
+\caption{The reachable terms of $\tY \, \Theta \, \underline 0$\label{fig:geometric distribution}}
 \end{figure}

### 2021-01-01 LICS21/antitone.tex
Misc stuff
I don't even remember what changes this contains. Nothing major.

@@
 The following theorem does not prove completeness, but is suggestive in that direction:
-\begin{theorem}
+\begin{restatable}{theorem}{antitoneSeemsComplete}
 For any stopping time $T$ which is almost-surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
-\end{theorem}
-
-% !TEX root = ./probabilisticProgrammingMartingales.tex
-
-\begin{proof}
-If $T$ is bounded by $b$ a.s.~(for $n$ constant), the statement is trivially true by taking $\epsilon$ to be constantly 1, and $Y_n = b - n$.
-Otherwise, let $(t_n)_{n \geq 0}$ be defined recursively such that $t_0 = 0$, $t_{n+1} > t_n$ and $\mathbb P[T > t_{n+1}] \leq \frac 1 4 \mathbb P[T > t_n]$.
-We will then define a supermartingale $(Y_n)_{n \geq 0}$ and nonrandom sequence $(y_n)_{n \geq 0}$ such that $Y_n = 0$ iff $T < n$, $Y_n = y_n$ iff $T \geq n$, and $y_n \geq 2^k$ for $n \geq t_k$.
-
-The antitone function $\epsilon$ is defined piecewise and recursively in such a way as to force all the necessary constraints to hold:
-\begin{align*}
-    \text{for }x \in [0,1),\ &\epsilon(x) = 1 \\
-    \text{for }x \in [2^k,2^{k+1}),\ &\epsilon(x) = \min \left (\epsilon(2^{k-1}), \frac{2^k}{t_{k+1}-t_k} \right ).
-\end{align*}
-This ensures that
-\begin{itemize}
-\item $\epsilon$ is weakly decreasing. 
-\item $\epsilon$ is bounded below by 0.
-\item For $x \geq 2^k$, $\epsilon(x) \leq \frac{2^k}{t_{k+1}-t_k}$. 
-\lo{This would imply $\frac{2^k}{t_{k+1}-t_k} \geq \epsilon(2^{k-1})$. Why?} \akr{Corrected: it should have been min, not max.}
-\end{itemize}
-
-The sequence $(y_n)_{n \geq 0}$ is then defined recursively by
-\begin{align*}
-    y_0 = 2, \quad
-    y_{n+1} = (y_n - \epsilon(y_n)) \frac{\mathbb P[T > n]}{\mathbb P[T > n+1]}.
-\end{align*}
-
-The fact $y_{t_k} \geq 2^{k+1}$ is proven by induction on $k$.
-For base case, $2 \geq 2$, as required.
-For the inductive case $k+1$, we first do another induction to prove that $y_n \geq 2^k$ for all $t_k \leq n \leq t_{k+1}$.
-The base case of this inner induction follows from the outer induction hypothesis a fortiori.
-Take the greatest $k$ such that $n > t_k$.
-By the induction hypothesis, $y_{t_k} \geq 2^k$.
-\begin{align*}
-    y_n & = y_{t_k} \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} - \sum_{m=t_k}^{n-1} \epsilon(y_m) \frac{\mathbb P[T > m+1]}{\mathbb P[T > n]} \\
-    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \sum_{m=t_k}^{n-1} \epsilon(y_m)) \\
-    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \sum_{m=t_k}^{n-1} \frac{2^k}{t_{k+1}-t_k}) \\
-    & = \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \frac{2^k (n - t_k)}{t_{k+1}-t_k}) \\
-    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - 2^k) \\
-    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (2^{k+1} - 2^k) \\
-    & = \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} 2^k \\
-    & \geq 2^k
-\end{align*}
-as required. Substituting $n = t_{k+1}$ and using the fact that $\frac{\mathbb P[T > t_k]}{\mathbb P[T > t_{k+1}]} > 4$, the same reasoning gives $y_{t_{k+1}} \geq 2^{k+2}$, as required.
-
-Although the stronger induction hypothesis was necessary for the induction to work, the only reason this is needed is to prove that $y_n > 0$ for all $n$, which implies that $Y_n \geq 0$. 
-The other condition for $(Y_n)$ to be an antitone-strict supermartingale is
-\begin{align*}
-    \expect{Y_{n+1} \mid \mathcal F_n} 
-    &=  \expect{y_{n+1} \, {\bf 1}_{\set{T \geq n+1}} \mid \mathcal F_n} \\
-    &=  y_{n+1} \, \expect{{\bf 1}_{\set{T \geq n+1}} \mid \mathcal F_n} \\ 
-    &=  y_{n+1} \, {\bf 1}_{\set{T \geq n}} \textstyle \frac{\mathbb P[T > n+1]}{\mathbb P[T > n]} \\
-    &=  (y_n - \epsilon(y_n)) \, {\bf 1}_{\set{T \geq n}} \\
-    &= Y_n - \epsilon(y_n) \, {\bf 1}_{\set{T \geq n}}
-\end{align*}
-therefore $(Y_n)_n$ is an antitone strict supermartingale with respect to $(T,\epsilon)$.
-\end{proof}
-
-The assumption that $(\mathcal F_n)_{n \geq 0}$ is the coarsest filtration to which $T$ is adapted is used in the equality between the third and fourth lines here.
-This condition is the main reason that this proof does not extend directly to a completeness result for antitone ranking functions.
+\end{restatable}
 
@@
 Then $\Xi \, g$ is AST for $p \geq 1/2$.}
+\akr{In order to actually construct a ranking function, it will be necessary to assume a bound on the number of $\tY$-reduction steps in the reduction of $g \underline n \underline m$ too.}
 \end{example}

### 2021-01-01 LICS21/ranking.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
Remove deprecated figures and comments from geometric distribution example.
Misc stuff
I don't even remember what changes this contains. Nothing major.

@@
 The term  $(\tY \, \Theta) \, \underline 0$ generates a geometric distribution.
-\changed[akr]{Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.}
-\lo{\Cref{fig:geometric distribution} is confusing: the arrow means either one-step reduction, or its reflexive transitive closure, or ``has the same syntactic shape''.}
-
-\akr{The arrows all mean the one-step reduction $\to$, I just didn't intend for the values of $i$ and $r$ to be consistent between the terms.
-I would much rather have the diagram be complete (with all the reductions and all the terms).}
-
-\lo{Perhaps you have in mind an informal, schematic interpretation, reading a term $C[i, r]$ as ``there exists a term $C[i, r]$, for appropriate $i, r \in \mathbb{R}$''.
-One way to make this precise is to consider the graph obtained by quotienting the one-step reduction relation by an appropriate equivalence relation over terms (that specifies this scheme reading).}
-
-\lo{\Cref{fig:geometric distribution 1} is less ambitious, but it is not confusing in that arrow does mean one-step reduction (also note $r'$), and it would be correct to say (cf.~the preceding blue-highlighted phrase): ``Despite its simplicity, $\mathit{Rch}(\tY \, \Theta \, \underline 0)$ contains all the terms in \Cref{fig:geometric distribution 1}, for each $i \geq 0$''.}
-
-\lo{The graph can be simplified somewhat if we redefine general recursion using a fixpoint operator $\tY_{f, n}\Theta$ with redex rule as per \Cref{eq:letrec redex}: see \Cref{fig:geometric distribution 2}.
-But (especially in view of the vast investment in the confluent trace semantics) it is perhaps too late to change things now?}
+Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.
+The meanings of the substitutions and conditions that label some of the reductions in the diagram should be self-explanatory.
 
@@
 
-    \draw [->] (9.west) to [bend left=45, "$i := i + 1$" {above,sloped}] (2.west);
+    \draw [->] (9.west) to [bend left=45, "$i := i + 1$" {below,sloped}] (2.west);
     \draw [->] (6) to ["$r < 0.5$"] (7);
@@
 \caption{The reachable terms of $\tY \, \Theta \, \underline 0$\label{fig:geometric distribution}}
-\end{figure}
-\begin{figure}
-\centering
-\begin{tikzpicture}[node distance=0.5em, anchor=west]
-    \node (1) at (0,-0) {$\tY\, \Theta\, \underline i$ \quad [provided $i = 0$]};
-    \node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, \underline i$};
-    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})})\, \underline i$};
-    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (5) at (0,-4) {$\tif{\underline r\, \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (6) at (0,-5) {$\tif{\underline r' < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (7) at (2,-6) [text width=] {$\underline i$};
-    \node (8) at (0,-7) {$\tY\, \Theta\, (\underline i\, \underline{+ 1})$};
-    \node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, (\underline i\, \underline{+ 1})$};
-    \node (10) at (0,-9) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, \underline{i + 1}$};
-
-    \draw [->] (1.south west) to [bend right=45] (2.north west);
-    \draw [->] (2.south west) to [bend right=45] (3.north west);
-    \draw [->] (3.south west) to [bend right=45] (4.north west);
-    \draw [->] (4.south west) to [bend right=45] (5.north west);
-    \draw [->] (5.south west) to [bend right=45] (6.north west);
-    \draw [->] (6.south west) to [bend right=45] (8.north west);
-    \draw [->] (8.south west) to [bend right=45] (9.north west);
-    \draw [->] (9.south west) to [bend right=45] (10.north west);
-    \draw [->] (6) to (7);
-\end{tikzpicture}
-\caption{Some reachable terms of $\tY \, \Theta \, \underline 0$\label{fig:geometric distribution 1}}
-\end{figure}
-
-\begin{figure}
-\centering
-\begin{tikzpicture}[node distance=0.5em, anchor=west]
-    \node (1) at (0,-0) {$\tY_{f, n}\Theta\, \underline i$};
-    %\node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY_{f, n}\Theta\, (n \underline{+ 1})}) z)\, \underline i$};
-    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY_{f, n}\Theta\, (n \underline{+ 1})})\, \underline i$};
-    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY_{f, n}\Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (5) at (0,-4) {$\tif{\underline r\, \underline{- 0.5} < 0}{\underline i}{\tY_{f, n}\Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (6) at (0,-5) {$\tif{\underline r' < 0}{\underline i}{\tY_{f, n}\Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (7) at (2,-6) [text width=] {$\underline i$};
-    \node (8) at (0,-7) {$\tY_{f, n}\Theta\, (\underline i\, \underline{+ 1})$};
-    \node (9) at (0,-8) {$\tY_{f, n}\Theta\, \underline{i + 1}$};
-    %\node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY_{f, n}\Theta\, (n \underline{+ 1})}) z)\, (\underline i\, \underline{+ 1})$};
-    %\node (10) at (0,-9) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY_{f, n}\Theta\, (n \underline{+ 1})}) z)\, \underline{i + 1}$};
-
-    \draw [->] (1.south west) to [bend right=45] (3.north west);
-    %\draw [->] (2.south west) to [bend right=45] (3.north west);
-    \draw [->] (3.south west) to [bend right=45] (4.north west);
-    \draw [->] (4.south west) to [bend right=45] (5.north west);
-    \draw [->] (5.south west) to [bend right=45] (6.north west);
-    \draw [->] (6.south west) to [bend right=45] (8.north west);
-    \draw [->] (8.south west) to [bend right=45] (9.north west);
-    %\draw [->] (9.south west) to [bend right=45] (10.north west);
-    \draw [->] (6) to (7);
-\end{tikzpicture}
-\caption{Some reachable terms of $\tY_{f, n}\Theta \, \underline 0$\label{fig:geometric distribution 2}}
 \end{figure}

### 2021-01-01 LICS21/ranking.tex
Minor: geometric distribution example.

--- 2021-01-01 LICS21/ranking.tex
@@
 The term  $(\tY \, \Theta) \, \underline 0$ generates a geometric distribution.
-\changed[akr]{Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.}
+\iffalse
 \lo{\Cref{fig:geometric distribution} is confusing: the arrow means either one-step reduction, or its reflexive transitive closure, or ``has the same syntactic shape''.}
@@
 But (especially in view of the vast investment in the confluent trace semantics) it is perhaps too late to change things now?}
+\fi
+Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.
+The meanings of the substitutions and conditions that label some of the reductions in the diagram should be self-explanatory.
 
@@
 \end{figure}
+
+\iffalse
 \begin{figure}
@@
 \end{figure}
-
+\fi
 \end{example}

### 2021-01-02 LICS21/SPCF.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
# Conflicts:
#	LICS21/ranking.tex

almost-sure -> almost sure (for consistency)

@@
 The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid M \text{ terminates at } s \}$ is measurable.
-We say that a term $M$ is \emph{almost-surely terminating} (AST) just if $\mu_{\entrosp}(\{s \mid M \text{ terminates at } s\}) = 1$.
+We say that a term $M$ is \emph{almost surely terminating} (AST) just if $\mu_{\entrosp}(\{s \mid M \text{ terminates at } s\}) = 1$.
 
@@
 \]
-Equivalently, we say that $M$ is \emph{almost-surely terminating} (AST) if $T_M < \infty$ a.s.; 
-and $M$ is \emph{positively almost-surely terminating} (PAST) if $\expect{T_M} < \infty$.}
+Equivalently, we say that $M$ is \emph{almost surely terminating} (AST) if $T_M < \infty$ a.s.; 
+and $M$ is \emph{positively almost surely terminating} (PAST) if $\expect{T_M} < \infty$.}
 \fi
@@
 %\paragraph{}
-This definition of almost-sure termination is equivalent to that given in \citep{MakOPW21}, although the program semantics is stated in a slightly different way.
+This definition of almost sure termination is equivalent to that given in \citep{MakOPW21}, although the program semantics is stated in a slightly different way.
 In particular, the argument to $\tscore$ is not relevant to termination (except for the possibility that its argument's evaluation wouldn't terminate).

### 2021-01-02 LICS21/antitone.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
# Conflicts:
#	LICS21/ranking.tex

almost-sure -> almost sure (for consistency)

@@
 \begin{restatable}{theorem}{antitoneSeemsComplete}
-For any stopping time $T$ which is almost-surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
+For any stopping time $T$ which is almost surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
 \end{restatable}
@@
 Then $\Xi \, g$ is AST for $p \geq 1/2$.}
-\akr{In order to actually construct a ranking function, it will be necessary to assume a bound on the number of $\tY$-reduction steps in the reduction of $g \underline n \underline m$ too.}
+\akr{In order to actually construct a ranking function, it will be necessary to assume a bound on the number of $\tY$-reduction steps in the reduction of $g \, \underline n \, \underline m$ too.}
+
+\akr{I suspect that the term you actually had in mind for the higher-order recursion example was different from what you wrote, because you said that it terminates for $p \geq 1/2$, but it actually terminates for all $p > 0$.
+Passing $n+1$ as an argument to $f$ doesn't actually make the recursion any longer.
+Did you mean
+\[
+\begin{array}{l}
+\tY \lambda \varphi f s n. \\
+\qquad \tif{n \leq 0}{s}{ f \, n \, (\varphi \, f \, s \, (n - 1)) \oplus_p f \, n \, (\varphi \, f \, s \, (n+1))}
+\end{array}
+\]}
 \end{example}

### 2021-01-02 LICS21/confluent.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
# Conflicts:
#	LICS21/ranking.tex

almost-sure -> almost sure (for consistency)

@@
 When proving almost sure termination in this way, it is necessary to consider which terms a given term may reduce to.
-Sometimes however, the reduction that the programmer has in mind may not be strictly the call-by-value order defined so far, or considering an alternative reduction order may be simpler or more intuitive.
+Sometimes however, the reduction that the programmer has in mind may not be strictly the call-by-value order defined so far; or considering an alternative reduction order may be simpler or more intuitive.
 
@@
 If neither $\alpha_1 \leq \alpha_2$ nor $\alpha_1 \geq \alpha_2$, the positions are said to be \emph{disjoint}.
-The notion of disjointness is mostly relevant in that if $\alpha_1$ and $\alpha_2$ are disjoint, performing a substitution at $\alpha_1$ will leave the subterm at $\alpha_2$ inaffected.
-\lo{The preceding sentence is unclear. Typo?} \akr{It wasn't a typo, but I've made it a little more explicit now. Is this clearer?}
+The notion of disjointness is mostly relevant in that if $\alpha_1$ and $\alpha_2$ are disjoint, performing a substitution at $\alpha_1$ will leave the subterm at $\alpha_2$ unaffected.
+\lo{The preceding sentence is unclear. Typo?} \akr{It wasn't a typo, but I've made it a little more explicit now. Is this clearer?} \lo{Yes, it's clear now.}
 
@@
 \end{align*}
-In each of these cases, $M \mid \alpha$ is the \emph{redex}, and reduction takes place at $\alpha$.
+In each of these cases, $M \mid \alpha$ is the \emph{redex}, and the reduction takes place at $\alpha$.
 \end{definition}
@@
 \bigcup_{r \in [0,1]} \{s \in I^{L_s(A[\tsample])} \mid \\
-\qquad s([A[\tsample]], \mathsf{if}_1;\underline{-}_1;@_1;\mathsf{if}_1) = r, \\
-\qquad s([A[\tsample], A[\underline r], \ldots, B, \mathsf{if}_1) > 0.5 \}.
+\qquad\qquad\qquad\quad s([A[\tsample]], \mathsf{if}_1;\underline{-}_1;@_1;\mathsf{if}_1) = r, \\
+\qquad\qquad\qquad\quad s([A[\tsample], A[\underline r], \ldots, B, \mathsf{if}_1) > 0.5 \}.
 \end{array}\] 
@@
 which is measurable. Skeletal reduction sequences still contain just as much informationas reduction sequences of terms, however, on which reductions took place.
+\qed
 \end{example}

### 2021-01-02 LICS21/ranking.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
# Conflicts:
#	LICS21/ranking.tex

almost-sure -> almost sure (for consistency)

@@
 
-Although rankability implies almost-sure termination, the converse does not hold in general. For example,
+Although rankability implies almost sure termination, the converse does not hold in general. For example,
 \begin{equation}
@@
 \subsection{Sparse ranking functions}
-Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost-sure termination, so it isn't useful for proving it.
+Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost sure termination, so it isn't useful for proving it.
 
@@
 Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.
-The meanings of the substitutions and conditions that label some of the reductions in the diagram should be self-explanatory.
+The meanings of the substitutions \changed[lo]{(to be applied to the contractum)} and conditions that label some of the reductions in the diagram should be self-explanatory.
 

### 2021-01-02 LICS21/supermartingales-proofs.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
# Conflicts:
#	LICS21/ranking.tex

almost-sure -> almost sure (for consistency)

@@
 
-It is straightforward to see that for every trace $s \in S$, there is a reduction sequence of $\transform{M}$ that simulates (an initial subsequene of) the reduction of $M$ under $s$;
+It is straightforward to see that for every trace $s \in \entrosp$, there is a reduction sequence of $\transform{M}$ that simulates (an initial subsequene of) the reduction of $M$ under $s$;
 and the simulation can be made lockstep.

### 2021-01-02 LICS21/supermartingales.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
# Conflicts:
#	LICS21/ranking.tex

almost-sure -> almost sure (for consistency)

@@
 \)
-$\mu$-almost-surely.
+$\mu$-almost surely.
 %(Unless otherwise stated, all equations and inequations between random variables are assumed to hold only $\mu$-a.s.)
@@
 \begin{enumerate}
-\item If a closed SPCF term $M$ is rankable, then $M$ is AST (equivalently, $T^{\tY}_M < \infty$ almost surely), and $M$ is $\tY$-positively almost-sure terminating ($\tY$-PAST) i.e.~$\expect{T^{\tY}_M} < \infty$.
+\item If a closed SPCF term $M$ is rankable, then $M$ is AST (equivalently, $T^{\tY}_M < \infty$ almost surely), and $M$ is $\tY$-positively almost sure terminating ($\tY$-PAST) i.e.~$\expect{T^{\tY}_M} < \infty$.
 
-\item If a closed SPCF term $M$ is strictly rankable, then $\expect{T_M} < \infty$ i.e.~$M$ is positively almost-sure terminating (PAST).
+\item If a closed SPCF term $M$ is strictly rankable, then $\expect{T_M} < \infty$ i.e.~$M$ is positively almost sure terminating (PAST).
 \end{enumerate}

### 2021-01-02 probabilisticProgrammingMartingales.tex
For consistency: global replace indicator function 1_{...} -> {\bf 1}_{...}
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
# Conflicts:
#	LICS21/ranking.tex

almost-sure -> almost sure (for consistency)

@@
 The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid M \text{ terminates at } s \}$ is measurable.
-We say that a term $M$ is \emph{almost-surely terminating} (AST) just if $\mu(\{s \mid M \text{ terminates at } s\}) = 1$.
+We say that a term $M$ is \emph{almost surely terminating} (AST) just if $\mu(\{s \mid M \text{ terminates at } s\}) = 1$.
 
@@
 \]
-Equivalently, we say that $M$ is \emph{almost-surely terminating} (AST) if $T_M < \infty$ a.s.; 
-and $M$ is \emph{positively almost-surely terminating} (PAST) if $\expect{T_M} < \infty$.}
+Equivalently, we say that $M$ is \emph{almost surely terminating} (AST) if $T_M < \infty$ a.s.; 
+and $M$ is \emph{positively almost surely terminating} (PAST) if $\expect{T_M} < \infty$.}
 
@@
 \paragraph{}
-This definition of almost-sure termination is equivalent to that given in \citep{MakOPW21}, although the program semantics is stated in a slightly different way.
+This definition of almost sure termination is equivalent to that given in \citep{MakOPW21}, although the program semantics is stated in a slightly different way.
 In particular, the argument to $\tscore$ is not relevant to termination (except for the possibility that its argument's evaluation wouldn't terminate).
@@
 \subsection{Partial ranking functions}
-Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost-sure termination, so it isn't useful for proving it.
+Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost sure termination, so it isn't useful for proving it.
 
@@
 As before, $(Y_n)_{n \geq 0}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$.
-Therefore, it follows from \Cref{thm:a-c strict} that $T_M^{\tY}$ is almost-surely finite.
+Therefore, it follows from \Cref{thm:a-c strict} that $T_M^{\tY}$ is almost surely finite.
 \end{proof}
@@
 \begin{theorem}
-For any stopping time $T$ which is almost-surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
+For any stopping time $T$ which is almost surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
 \end{theorem}

### 2021-01-02 LICS21/SPCF.tex
Redefining \tY as a value.
More on def. of \tY and \tY-redex.

@@
 Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
+\akr{\begin{align*}
+  \text{\em types } A, B &::= \textsf{R}  \mid  A \to B \\
+  \text{\em values } V &::= \lambda x.M  \mid  \underline{r} \\
+  \text{\em terms } M, N &::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY \, M \\
+   & \qquad \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample  \mid  \tscore(M)
+\end{align*}}
+\lo{Currently $\tY$ is not a legal term.
+This is unnatural: there is no reason to so restrict $\tY$.
+Following \citep[p.~376]{DBLP:conf/fsttcs/Sieber90}, we should define $\tY$ as a value (and so a proper term); it follows that $\tY \, M$ is redundant as a clause.}
+\lo{N.B. It is also standard to regard variables as values in a CBV language.}
+
+\lo{Problem: $\tY \, \tY$ is then a ``stuck'' term: neither value nor redex.
+A possible solution is to redefine the $\tY$-redex rule as $\tY \, V \to \lambda z .
+V \, (\tY \, V) \, z$, but I expect this would affect many proofs in the confluent semantics section, and there may not be time for such a change.}
 \begin{align*}
   \text{\em types } A, B &::= \textsf{R}  \mid  A \to B \\
-  \text{\em values } V &::= \lambda x.M  \mid  \underline{r} \\
-  \text{\em terms } M, N &::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY M \\
+  \text{\em values } V &::= x \mid \lambda x.M  \mid  \underline{r} \mid \tY \\
+  \text{\em terms } M, N &::= V  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n) \\
    & \qquad \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample  \mid  \tscore(M)

### 2021-01-02 LICS21/SPCF.tex
Example VI.2 (Labelling samples by terms): typos

@@
 The maps $\pi_h:\entrosp \to I, \; \pi_t:\entrosp \to \entrosp$ popping the first element are then measurable.
-\changed[lo]{Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$ the \emph{entropy space}.}
+\changed[lo]{Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$ the \emph{entropy space}, and call elements of $\entrosp$ \emph{traces}.}
 

### 2021-01-02 LICS21/confluent.tex
Example VI.2 (Labelling samples by terms): typos

@@
 
-The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, and $\tsample$ reducing to $\skeletonPlaceholder$.
+\changed[lo]{We are thus lead to consider the reduction relation on skeletons (and positions in a skeleton), which can be extended from the definitions on terms in the obvious way,} 
+%The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, 
+with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, and $\tsample$ reducing to $\skeletonPlaceholder$.
 For example, we have 
@@
 Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a $\to$-reduction sequence of skeletons starting at $Sk(M)$, and the second of which is a position in the final skeleton of the reduction sequence.
-As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the usual linear sampling semantics, there will be (with one more caveat introduced later) a sample from $I$ pre-selected for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.
+As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the usual linear sampling semantics,\lo{I assume by linear sampling semantics, you mean the standard sampling semantics.
+If so, I would drop linear.} there will be (with one more caveat introduced later) a sample from $I$ pre-selected for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.
 
@@
 \end{array}\] 
+\lo{We have just defined $L_0(M)$, but what is $L_s(\cdot)$ (subscript $s$)? I assume it is the version of $L_0(M)$ that labels samples by term.}
+
+\lo{The occurrence of $L_s(M)$ in the set comprehension below should be $L_0(M)$?}
+
+\lo{The $\mathsf{if}_1$ in the line above (and in the set comprehension below) should be $\mathsf{if}_1 ; \underline{-}_1$.}
+
 This is a rather unwieldy expression, but the crucial part is that $r$ occurs twice in the conditions on $s$: once as the value a sample must take, and once in the location of a sample. 
@@
 \end{align*}
-which is measurable. Skeletal reduction sequences still contain just as much informationas reduction sequences of terms, however, on which reductions took place.
+which is measurable. Skeletal reduction sequences still contain just as much information as reduction sequences of terms, however, on which reductions took place.
 \qed
@@
 Reduction sequences are used rather than reachable skeletons because if the same skeleton is reached twice, different samples may be needed.
-For example, consider the term $M = \tY (\lambda f x.\ \tif{\tsample - \underline{0.5} < 0}{f \, x}{x}) \, \underline 0$, which reduces after a few steps to $N = \tif{\tsample - \underline{0.5}  < 0}{M}{\underline 0}$.
+
+\begin{example} Consider the term $M = \tY (\lambda f x.\ \tif{\tsample - \underline{0.5} < 0}{f \, x}{x}) \, \underline 0$, which reduces after a few steps to $N = \tif{\tsample - \underline{0.5}  < 0}{M}{\underline 0}$.
 If the pre-selected sample for $(N,\textsf{if}_1;\underline{-}_1)$ is less than 0.5, $N$ reduces back to $M$, then $N$ again, then the same sample is used the next time, therefore it's an infinite loop, whereas if samples are labelled by reduction sequences, the samples for $M \to \dots \to N$ are independent from the samples for $M \to \dots \to N \to M \to \dots \to N$, and so on.
+\end{example}
 

### 2021-01-02 LICS21/ranking.tex
Example VI.2 (Labelling samples by terms): typos

@@
 \fi
-Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.
-The meanings of the substitutions \changed[lo]{(to be applied to the contractum)} and conditions that label some of the reductions in the diagram should be self-explanatory.
+Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.\lo{TODO: reduce the figure to avoid overflow.} The meanings of the substitutions \changed[lo]{(to be applied to the contractum)} and conditions that label some of the reductions in the diagram should be self-explanatory.
 

### 2021-01-02 LICS21/confluent-proofs.tex
Global change: Sk -> \mathit{Sk}

@@
 Let\begin{align*}
-R_s & = \{Sk(O) \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) = \tsample, O \to N[r]\} \\
-R_d & = \{Sk(O) \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) \neq \tsample, O \to N[r]\}.
+R_s & = \{\mathit{Sk}(O) \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) = \tsample, O \to N[r]\} \\
+R_d & = \{\mathit{Sk}(O) \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) \neq \tsample, O \to N[r]\}.
 \end{align*}
@@
 The distribution of $s(k)$ conditional on reaching $N$ is therefore the uniform distribution on $I$.
-Similarly for the other semantics, if the distribution of remaining samples after $n$ steps is independent of the term by Lemma \ref{lem:independentSamples}, therefore the distribution of $s(Sk(N),\cbv(N))$ conditional on $N$ is the uniform distribution on $I$.
+Similarly for the other semantics, if the distribution of remaining samples after $n$ steps is independent of the term by Lemma \ref{lem:independentSamples}, therefore the distribution of $s(\mathit{Sk}(N),\cbv(N))$ conditional on $N$ is the uniform distribution on $I$.
 In either case, the distribution of the next term conditional on the previous term, in the case that it reduces randomly, is equal to the image under $r \mapsto E[\underline r]$ of the uniform distribution on $I$.
@@
 Let $(M,s) = (M_0,s_0) \Rightarrow_f \dots \Rightarrow_f (M_n,s_n)$, where $M_n$ is a value.
-By Lemma \ref{canonicalCousins}, the reduction sequence $(Sk(M_i))_i$ is related by $\sim_c^*$ to some skeletal reduction sequence $X$ which is in call-by-value order.
-For each skeletal reduction sequence $X_j$ in the sequence $(Sk(M_i))_i \sim_c \dots \sim_c X$, it is possible to define a reduction sequence of terms such that $(M,s) = (N_{j,0},s_{j,0}) \Rightarrow \dots \Rightarrow (N_{j,n_j},s_{j,n_j})$ and $Sk(N_{j,k}) = X_{j,k}$, by at each step applying the reduction at the same place as in $X_j$, essentially just filling in the reals in the skeletal reduction sequence to make it a reduction sequence of terms (it will be shown shortly that the correct branch is taken in each $\textsf{if}$-reduction).
+By Lemma \ref{canonicalCousins}, the reduction sequence $(\mathit{Sk}(M_i))_i$ is related by $\sim_c^*$ to some skeletal reduction sequence $X$ which is in call-by-value order.
+For each skeletal reduction sequence $X_j$ in the sequence $(\mathit{Sk}(M_i))_i \sim_c \dots \sim_c X$, it is possible to define a reduction sequence of terms such that $(M,s) = (N_{j,0},s_{j,0}) \Rightarrow \dots \Rightarrow (N_{j,n_j},s_{j,n_j})$ and $\mathit{Sk}(N_{j,k}) = X_{j,k}$, by at each step applying the reduction at the same place as in $X_j$, essentially just filling in the reals in the skeletal reduction sequence to make it a reduction sequence of terms (it will be shown shortly that the correct branch is taken in each $\textsf{if}$-reduction).
 Assume for induction on $j$ that $(N_{j,n_j},s_{j,n_j}) = (M_n,s_n)$.

### 2021-01-02 LICS21/confluent.tex
Global change: Sk -> \mathit{Sk}

@@
 
-Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a $\to$-reduction sequence of skeletons starting at $Sk(M)$, and the second of which is a position in the final skeleton of the reduction sequence.
+Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a $\to$-reduction sequence of skeletons starting at $\mathit{Sk}(M)$, and the second of which is a position in the final skeleton of the reduction sequence.
 As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the usual linear sampling semantics,\lo{I assume by linear sampling semantics, you mean the standard sampling semantics.
@@
 
+\lo{I think we should write ``$[A[\tsample], A[\underline r], \ldots, B$'' in the line above as ``$A[\tsample] \to A[\underline r] \to^\ast B$''; similarly $\mathit{Sk}(A[\tsample]) \to \mathit{Sk}(A[\underline 0]) \to^\ast \mathit{Sk}(B)$ below.}
 This is a rather unwieldy expression, but the crucial part is that $r$ occurs twice in the conditions on $s$: once as the value a sample must take, and once in the location of a sample. 
@@
 Labelling samples by skeletons instead, this problem does not occur because there are only countably many skeleton, and at each step in a reduction sequence, only finitely many could have occurred yet.
-For this particular term, $Sk(A[\underline r])$ does not depend on the value of $r$, therefore the set where it terminates becomes simply
+For this particular term, $\mathit{Sk}(A[\underline r])$ does not depend on the value of $r$, therefore the set where it terminates becomes simply
 \begin{align*}
 &\{s \in I^{L_s(A[\tsample])} \mid \\
-&\qquad s(Sk(A[\tsample]), Sk(A[\underline 0]), \ldots, Sk(B), \mathsf{if}_1) > 0.5 \},
+&\qquad s(\mathit{Sk}(A[\tsample]), \mathit{Sk}(A[\underline 0]), \ldots, \mathit{Sk}(B), \mathsf{if}_1) > 0.5 \},
 \end{align*}
@@
 At each reduction step $M \to N$, the sample space must be restricted from $I^{L_s(M)}$ to $I^{L_s(N)}$. 
-The injection $L_0(N) \to L_0(M)$ is trivial to define by appending $Sk(M) \to Sk(N)$ to each path, and using Lemma~\ref{lem:sim-M-N}, this induces a corresponding injection on the quotient, $L(N) \to L(M)$.
+The injection $L_0(N) \to L_0(M)$ is trivial to define by appending $\mathit{Sk}(M) \to \mathit{Sk}(N)$ to each path, and using Lemma~\ref{lem:sim-M-N}, this induces a corresponding injection on the quotient, $L(N) \to L(M)$.
 The corresponding map $L_s(N) \to L_s(M)$ is then denoted $i(M \to N)$.
@@
 & \qquad \text{the redex is not $\tsample$, or} \\
-& \qquad \text{$M \mid \alpha = \tsample$ and $N = M[\underline{s(Sk(M),\alpha)}/\alpha]$}
+& \qquad \text{$M \mid \alpha = \tsample$ and $N = M[\underline{s(\mathit{Sk}(M),\alpha)}/\alpha]$}
 \end{align*}

### 2021-01-02 LICS21/SPCF.tex
Reference for Tait and Martin-Löf proof. Example for \sim_c.

@@
 
-Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $Sk(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$.
+Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $\mathit{Sk}(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$.
 Following \cite{DBLP:conf/icfp/BorgstromLGS16}, the $\sigma$-algebra on $\Lambda$ is defined by identifying $\Lambda$ with $\bigcup_{m\geq 0} \big(\mathsf{Sk}_m \times \Real^m \big)$, where $\mathsf{Sk}_m$ is the set of skeletons containing $m$ occurrences of $\skeletonPlaceholder$.

### 2021-01-02 LICS21/confluent-proofs.tex
Reference for Tait and Martin-Löf proof. Example for \sim_c.

@@
 If a reduction sequence includes one branch of one of these cases at some point in it, then the positions in the reduction sequence obtained by substituting in the other branch are all related by $\sim_c$ to the corresponding positions in the original reduction sequence.
+
+\lo{TODO. Remove the white space by repositioning the sim diagrams.}
 \include{sim-diagram}
@@
 \begin{figure}[htb]
-\label{sim-diagram}
-\caption{Illustration of the $\sim_c$ rules.}
+\caption{Illustration of the $\sim_c$ rules. \label{sim-diagram}
+}
 \end{figure}
@@
 Suppose for induction that $A \sim_\uparrow^k (D,\delta) \sim_\downarrow^* (E,\epsilon) \sim_\uparrow^* C$ for some $D, \delta, E, \epsilon$, and $0 \leq k \leq n$, where in the reduction sequence $D \to^* E$, for any pair of reductions at positions which aren't disjoint, the reduction at the innermost (greater) position occurs first.
-Reduction sequences with this property will be called ``parallel'', as it is analagous to parallel reduction as in \akr{TODO: find this reference, for the proof of the Church-Rosser theorem by Tait and Martin-L\"of}.
-This induction is in reverse, with $k$ decreasing from $n$ to $0$.
+Reduction sequences with this property will be called ``parallel'', as it is analagous to \changed[lo]{the parallel reduction introduced by Tait and Martin-L\"of in their proof of the Church-Rosser theorem \citep{Bar84,Takahashi1995}.}
+  \akr{TODO: find this reference, for the proof of the Church-Rosser theorem by Tait and Martin-L\"of.}
+  \lo{Tait and Martin-L\"of's proof is folklore; I don't think there is a first-hand documentary record.} This induction is in reverse, with $k$ decreasing from $n$ to $0$.
 If $k = n$, simply set $(D,\delta) = B$ and $(E,\epsilon) = C$.

### 2021-01-02 LICS21/confluent.tex
Reference for Tait and Martin-Löf proof. Example for \sim_c.

@@
 \end{array}\] 
-\lo{We have just defined $L_0(M)$, but what is $L_s(\cdot)$ (subscript $s$)? I assume it is the version of $L_0(M)$ that labels samples by term.}
+\lo{We have just defined $L_0(M)$, but what is $L_s(\cdot)$ (subscript $s$)? I assume it is the version of $L_0(M)$ that labels samples by terms.}
 
@@
 This is still not quite sufficient because sometimes the same samples must be used at corresponding positions in different reduction sequences.
-For example, the term $M = \tsample + \tsample$ has the reachable skeletons $N_1 = \skeletonPlaceholder + \tsample, N_2 = \tsample + \skeletonPlaceholder, O = \skeletonPlaceholder + \skeletonPlaceholder$ and $\skeletonPlaceholder$, with reductions $M \to N_1 \to O \to \skeletonPlaceholder$ and $M \to N_2 \to O \to \skeletonPlaceholder$.
-In the reduction $M \to N_1$, the sample labelled $(M, \underline{+}_2)$ is used, and in the reduction $N_2 \to O$, the sample labelled $(M \to N_2, \underline{+}_2)$ is used.
+
+\begin{example} \label{ex:cousin sim} The term $M = \tsample + \tsample$ has the reachable skeletons $N_1 = \skeletonPlaceholder + \tsample, N_2 = \tsample + \skeletonPlaceholder, O = \skeletonPlaceholder + \skeletonPlaceholder$ and $\skeletonPlaceholder$, with reductions $M \to N_1 \to O \to \skeletonPlaceholder$ and $M \to N_2 \to O \to \skeletonPlaceholder$.
+In the reduction $M \to N_1$, the sample labelled $(M, \underline{+}_2)$ is used, and in the reduction $N_2 \to O$, the sample labelled %$(M \to N_2, \underline{+}_2)$
+\changed[lo]{$(M \to N_2, \underline{+}_1)$} is used. 
 Each of these samples becomes the value of the second numeral in $O$ in their respective reduction sequences, therefore in order for Church-Rosserness to be attained, they must be the same.
 Which elements of $L_0(M)$ must match can be described by the relation $\sim^*$:
+\end{example}
 
@@
 The $\sim_c$-rules are illustrated in Figure \ref{sim-diagram}.
+\lo{TODO. Use \Cref{ex:cousin sim} to give an example of $\sim_c$.}
 \end{definition}

### 2021-01-03 LICS21/confluent.tex
lead -> led; Ex. VI.2: minor

@@
 
-\changed[lo]{We are thus lead to consider the reduction relation on skeletons (and positions in a skeleton), which can be extended from the definitions on terms in the obvious way,} 
+\changed[lo]{We are thus led to consider the reduction relation on skeletons (and positions in a skeleton), which can be extended from the definitions on terms in the obvious way,} 
 %The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, 
@@
 \[\begin{array}{l}
-\bigcup_{r \in [0,1]} \{s \in I^{L_s(A[\tsample])} \mid \\
+\bigcup_{r \in [0,1]} \Big\{s \in I^{L_s(A[\tsample])} \mid \\
 \qquad\qquad\qquad\quad s([A[\tsample]], \mathsf{if}_1;\underline{-}_1;@_1;\mathsf{if}_1) = r, \\
-\qquad\qquad\qquad\quad s([A[\tsample], A[\underline r], \ldots, B, \mathsf{if}_1) > 0.5 \}.
+\qquad\qquad\qquad\quad s([A[\tsample], A[\underline r], \ldots, B, \mathsf{if}_1) > 0.5 \Big\}.
 \end{array}\] 
+\lo{There are a number of spurious square bracket in the predicate of the set comprehension above.}
+
 \lo{We have just defined $L_0(M)$, but what is $L_s(\cdot)$ (subscript $s$)? I assume it is the version of $L_0(M)$ that labels samples by terms.}

### 2021-01-03 LICS21/confluent.tex
Ex. VI.4 reviisted: typos?

@@
 
-\begin{example} \label{ex:cousin sim} The term $M = \tsample + \tsample$ has the reachable skeletons $N_1 = \skeletonPlaceholder + \tsample, N_2 = \tsample + \skeletonPlaceholder, O = \skeletonPlaceholder + \skeletonPlaceholder$ and $\skeletonPlaceholder$, with reductions $M \to N_1 \to O \to \skeletonPlaceholder$ and $M \to N_2 \to O \to \skeletonPlaceholder$.
-In the reduction $M \to N_1$, the sample labelled $(M, \underline{+}_2)$ is used, and in the reduction $N_2 \to O$, the sample labelled %$(M \to N_2, \underline{+}_2)$
-\changed[lo]{$(M \to N_2, \underline{+}_1)$} is used. 
+\begin{example} \label{ex:cousin sim} 
+The term $M = \tsample + \tsample$ has the reachable skeletons $N_1 = \skeletonPlaceholder + \tsample, N_2 = \tsample + \skeletonPlaceholder, O = \skeletonPlaceholder + \skeletonPlaceholder$ and $\skeletonPlaceholder$, with reductions $M \to N_1 \to O \to \skeletonPlaceholder$ and $M \to N_2 \to O \to \skeletonPlaceholder$.
+In the reduction $M \to N_1$, the sample labelled $(M, \underline{+}_2)$ is used, and in the reduction $N_2 \to O$, the sample labelled $(M \to N_2, \underline{+}_2)$ is used. 
 Each of these samples becomes the value of the second numeral in $O$ in their respective reduction sequences, therefore in order for Church-Rosserness to be attained, they must be the same.
 Which elements of $L_0(M)$ must match can be described by the relation $\sim^*$:
+
+\lo{I think there are typos above: see the corrected(?) version below.}[The term $M = \tsample + \tsample$ has the reachable skeletons $N_1 = \skeletonPlaceholder + \tsample, N_2 = \tsample + \skeletonPlaceholder, O = \skeletonPlaceholder + \skeletonPlaceholder$ and $\skeletonPlaceholder$, with reductions $M \to N_1 \to O \to \skeletonPlaceholder$ and $M \to N_2 \to O \to \skeletonPlaceholder$.
+In the reduction $M \to N_1$, the sample labelled \changed[lo]{$(M, \underline{+}_1)$} is used, and in the reduction $N_2 \to O$, the sample labelled %$(M \to N_2, \underline{+}_2)$
+\changed[lo]{$(M \to N_2, \underline{+}_1)$} is used. 
+Each of these samples becomes the value of the \changed[lo]{first} \lo{left?} numeral in $O$ in their respective reduction sequences, therefore in order for Church-Rosserness to be attained, they must be the same.
+Which elements of $L_0(M)$ must match can be described by the relation $\sim^*$:]
 \end{example}

### 2021-01-04 LICS21/antitone.tex
Redefining \tY as a value.
Reference for Tait and Martin-Löf proof. Example for \sim_c.
Respond to comments

@@
 However for the definition to make sense, $T$ must be the stopping time $T(s) := \min \set{n \mid Y_n(s) = 0}$.}
+\akr{The definition you gave for an antitone ranking function is incorrect, because the antitone ranking functions are a proper subset of ranking functions.
+Although we may have been referring to antitone ranking functions as a type of ranking function, they're a generalisation, so they don't actually meet the definition of a ranking function in general.}
 
@@
 Even though the difference is quite trivial, this was actually the original motivating example for the confluent trace semantics.}
-
 Programs $M_1$ and $M_2$ are AST. For $M_1$, the sparse ranking function 
@@
 
-\lo{@Andrew: By the way, we should use the name \emph{sparse ranking function} (theorem / method).}
-
 \lo{
@@
 \quad\tif{n \leq 0}{s}{
-f \; n \; (\varphi \; f \; s \; (n - 1))
+f \, n \, (\varphi \, f \, s \, (n - 1))
 \oplus_p
-f \; (n+1) \; (\varphi \; f \; s \; n)
+f \, (n+1) \, (\varphi \, f \, s \, n)
 }

### 2021-01-04 LICS21/confluent.tex
Respond to comments

--- 2021-01-02 LICS21/confluent.tex
@@
 The notion of disjointness is mostly relevant in that if $\alpha_1$ and $\alpha_2$ are disjoint, performing a substitution at $\alpha_1$ will leave the subterm at $\alpha_2$ unaffected.
-\lo{The preceding sentence is unclear. Typo?} \akr{It wasn't a typo, but I've made it a little more explicit now. Is this clearer?} \lo{Yes, it's clear now.}
 
-With this notation, a more general reduction relation $\to$ can be defined. \akr{It would almost work to have one of the neater versions of $\tY$-reduction here.}
-%\lo{On first reading, I struggled to parse the following clauses.}
+With this notation, a more general reduction relation $\to$ can be defined.
 \begin{definition}
@@
 
-\changed[lo]{We are thus lead to consider the reduction relation on skeletons (and positions in a skeleton), which can be extended from the definitions on terms in the obvious way,} 
-%The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, 
-with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, and $\tsample$ reducing to $\skeletonPlaceholder$.
+We are thus led to consider the reduction relation on skeletons (and positions in a skeleton), which can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, and $\tsample$ reducing to $\skeletonPlaceholder$.
 For example, we have 
@@
 Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a $\to$-reduction sequence of skeletons starting at $\mathit{Sk}(M)$, and the second of which is a position in the final skeleton of the reduction sequence.
-As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the usual linear sampling semantics,\lo{I assume by linear sampling semantics, you mean the standard sampling semantics.
-If so, I would drop linear.} there will be (with one more caveat introduced later) a sample from $I$ pre-selected for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.
+As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the standard sampling semantics, there will be (with one more caveat introduced later) a sample from $I$ pre-selected for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.
+
+A (skeletal) reduction sequence is assumed to contain the information on the locations of all of the redexes as well as the actual sequence of skeletons that occurs.
+For example, $(\lambda x.
+x) ((\lambda x.
+x) \skeletonPlaceholder)$ could reduce to $(\lambda x.
+x) \skeletonPlaceholder$ with the redex at either $\cdot$ or $@_2$, and these give different reduction sequences.
 
@@
 If terms rather than skeletons were used to label samples, the set of traces where $A[\tsample]$ terminates would be
-\[\begin{array}{l}
-\bigcup_{r \in [0,1]} \{s \in I^{L_s(A[\tsample])} \mid \\
-\qquad\qquad\qquad\quad s([A[\tsample]], \mathsf{if}_1;\underline{-}_1;@_1;\mathsf{if}_1) = r, \\
-\qquad\qquad\qquad\quad s([A[\tsample], A[\underline r], \ldots, B, \mathsf{if}_1) > 0.5 \}.
-\end{array}\] 
+\begin{align*}
+\bigcup_{r \in [0,1]} \Big\{s \mid\ & s(A[\tsample], \mathsf{if}_1;\underline{-}_1;@_1;\mathsf{if}_1) = r, \\
+& s(A[\tsample] \to A[\underline r] \to^* B, \mathsf{if}_1 ; \underline{-}_1) > 0.5 \Big\}.
+\end{align*}
+
 \lo{We have just defined $L_0(M)$, but what is $L_s(\cdot)$ (subscript $s$)? I assume it is the version of $L_0(M)$ that labels samples by terms.}
 
+\akr{$L_s(M)$ is the set of potential positions (quotiented by $\sim^*$) where there is a $\tsample$.
+I don't want to move this example until after $L_s(M)$ has actually been defined, but also I don't want to imply that $I^{L_0(M)}$ is the actual entropy space too explicitly, so I'll just leave the set $s$ is restricted to implicit for now.}
+
 \lo{The occurrence of $L_s(M)$ in the set comprehension below should be $L_0(M)$?}
-
-\lo{The $\mathsf{if}_1$ in the line above (and in the set comprehension below) should be $\mathsf{if}_1 ; \underline{-}_1$.}
 
@@
 Labelling samples by skeletons instead, this problem does not occur because there are only countably many skeleton, and at each step in a reduction sequence, only finitely many could have occurred yet.
+Although skeletal reduction sequences omit the information on what the results of sampling were, they still contain all the necessary information on how many, and which, reductions took place.
+
 For this particular term, $\mathit{Sk}(A[\underline r])$ does not depend on the value of $r$, therefore the set where it terminates becomes simply
 \begin{align*}
-&\{s \in I^{L_s(A[\tsample])} \mid \\
-&\qquad s(\mathit{Sk}(A[\tsample]), \mathit{Sk}(A[\underline 0]), \ldots, \mathit{Sk}(B), \mathsf{if}_1) > 0.5 \},
+\{s \mid s(\mathit{Sk}(A[\tsample]) \to \mathit{Sk}(A[\underline 0]) \to^* \mathit{Sk}(B), \mathsf{if}_1 ; \underline{-}_1) > 0.5 \},
 \end{align*}
-which is measurable. Skeletal reduction sequences still contain just as much information as reduction sequences of terms, however, on which reductions took place.
+which is measurable.
 \qed
@@
 
-\begin{example} \label{ex:cousin sim} The term $M = \tsample + \tsample$ has the reachable skeletons $N_1 = \skeletonPlaceholder + \tsample, N_2 = \tsample + \skeletonPlaceholder, O = \skeletonPlaceholder + \skeletonPlaceholder$ and $\skeletonPlaceholder$, with reductions $M \to N_1 \to O \to \skeletonPlaceholder$ and $M \to N_2 \to O \to \skeletonPlaceholder$.
-In the reduction $M \to N_1$, the sample labelled $(M, \underline{+}_2)$ is used, and in the reduction $N_2 \to O$, the sample labelled %$(M \to N_2, \underline{+}_2)$
-\changed[lo]{$(M \to N_2, \underline{+}_1)$} is used. 
-Each of these samples becomes the value of the second numeral in $O$ in their respective reduction sequences, therefore in order for Church-Rosserness to be attained, they must be the same.
+\begin{example} \label{ex:cousin sim} 
+The term $M = \tsample + \tsample$ has the reachable skeletons $N_1 = \skeletonPlaceholder + \tsample, N_2 = \tsample + \skeletonPlaceholder, O = \skeletonPlaceholder + \skeletonPlaceholder$ and $\skeletonPlaceholder$, with reductions $M \to N_1 \to O \to \skeletonPlaceholder$ and $M \to N_2 \to O \to \skeletonPlaceholder$.
+In the reduction $M \to N_1$, the sample labelled $(M, \underline{+}_1)$ is used, and in the reduction $N_2 \to O$, the sample labelled $(M \to N_2, \underline{+}_1)$ is used. 
+Each of these samples becomes the value of the first numeral in $O$ in their respective reduction sequences, therefore in order for Church-Rosserness to be attained, they must be the same.
 Which elements of $L_0(M)$ must match can be described by the relation $\sim^*$:
@@
 The $\sim_c$-rules are illustrated in Figure \ref{sim-diagram}.
-\lo{TODO. Use \Cref{ex:cousin sim} to give an example of $\sim_c$.}
 \end{definition}
+
+In \Cref{ex:cousin sim}, $(M,\underline{+}_1) \sim_p (M \to N_2,\underline{+}_1)$ by case i of $\sim_p$ (because the reduction $M \to N_2$ occurs at $\underline{+}_2$, which is disjoint from $\underline{+}_1$), and similarly, $(M,\underline{+}_2) \sim_p (M \to N_1,\underline{+}_2)$.
+
+If we extend it to have three $\tsample$s, $\sim_c$ becomes necessary as well: Let $M_{sss} = \tsample + \tsample + \tsample$ (taking the three-way addition to be a single primitive function), $M_{Xss} = \skeletonPlaceholder + \tsample + \tsample$, and so on.
+There are then reduction sequences $M_{sss} \to M_{Xss} \to M_{XXs} \to M_{XXX} \to \skeletonPlaceholder$ and $M_{sss} \to M_{sXs} \to M_{XXs} \to M_{XXX} \to \skeletonPlaceholder$.
+For the first two reductions, these reduction sequences take the same samples by $\sim_p$, case i, as in \Cref{ex:cousin sim}~.
+The next reduction uses the samples labelled by $(M_{sss} \to M_{Xss} \to M_{XXs}, \underline{+}_3)$ and $(M_{sss} \to M_{sXs} \to M_{XXs}, \underline{+}_3)$, which are related by $\sim_c$, case a, therefore when these reduction sequences reach $M_{XXX}$, they still contain all the same numbers, as desired.
 
@@
 This reduction relation now has all of the properties required of it.
-In particular, it can be considered an extension of the usual linear trace semantics (as will be seen later in Theorem \ref{thm:AstEquivalence}), and also:
+In particular, it can be considered an extension of the standard trace semantics (as will be seen later in Theorem \ref{thm:AstEquivalence}), and also:
 \begin{restatable}{lemma}{churchRosser} \label{churchRosser}
@@
 
-With these definitions, it is now possible to relate the confluent sampling semantics back to the linear sampling semantics.
+With these definitions, it is now possible to relate the confluent sampling semantics back to the standard sampling semantics.
 
@@
 
-It is also true that the distributions of terms produced by the linear semantics and by the reduction strategy $\cbv$ are the same, for much the same reason, although that is not relevant to our main results.
+It is also true that the distributions of terms produced by the standard semantics and by the reduction strategy $\cbv$ are the same, for much the same reason, although that is not relevant to our main results.
 
@@
 
-It is not quite true, however, that all reduction strategies produce the same values (barring some reduction strategies terminating where others don't), because it is possible for a reduction strategy to overshoot the value that the linear semantics produces by, for example, performing reductions inside of a lambda.
+It is not quite true, however, that all reduction strategies produce the same values (barring some reduction strategies terminating where others don't), because it is possible for a reduction strategy to overshoot the value that the standard semantics produces by, for example, performing reductions inside of a lambda.
 Even for the same value of the samples, the final values may be different.

### 2021-01-04 LICS21/ranking.tex
Respond to comments

--- 2021-01-02 LICS21/ranking.tex
@@
 \fi
-Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.
-The meanings of the substitutions \changed[lo]{(to be applied to the contractum)} and conditions that label some of the reductions in the diagram should be self-explanatory.
+Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.\lo{TODO: reduce the figure to avoid overflow.}\akr{It still overflows the gutter, but not the actual text of the other column.} The meanings of the substitutions (to be applied to the contractum) and conditions that label some of the reductions in the diagram should be self-explanatory.
 
@@
 
-    \draw [->] (9.west) to [bend left=45, "$i := i + 1$" {below,sloped}] (2.west);
+    \draw [->] (9.west) to [bend left=30, "$i := i + 1$" {below,sloped}] (2.west);
     \draw [->] (6) to ["$r < 0.5$"] (7);

### 2021-01-04 LICS21/SPCF.tex
Remove score entirely.
Remove the score construct from the language definition, and add a
note that this differs from the previous definition of SPCF and it
would be trivial to add it back in.

@@
   \text{\em terms } M, N &::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY \, M \\
-   & \qquad \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample  \mid  \tscore(M)
+   & \qquad \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample
 \end{align*}}
@@
   \text{\em terms } M, N &::= V  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n) \\
-   & \qquad \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample  \mid  \tscore(M)
+   & \qquad \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample
 \end{align*}
@@
 \begin{align*}
-  E ::= & \, [\,] \mid E \, M \mid V \, E \mid \underline{f}(\underline r_1,\dots ,\underline r_{k-1}, E, M_{k+1}, \dots, M_n) \\ & \mid \tY E \mid \tif{E<0}{M_1}{M_2} \mid \tscore (E)
+  E ::= & \, [\,] \mid E \, M \mid V \, E \mid \underline{f}(\underline r_1,\dots ,\underline r_{k-1}, E, M_{k+1}, \dots, M_n) \\ & \mid \tY E \mid \tif{E<0}{M_1}{M_2}
 \end{align*}
@@
   E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_2] \text{ where }r \geq 0 \\
-  E[\tsample] & \to E[\underline r] \text{ where } r \in [0,1] \\
-  E[\tscore(\underline r)] & \to E[\underline r].
+  E[\tsample] & \to E[\underline r] \text{ where } r \in [0,1].
 \end{align*}
 We write $\to^\ast$ for the reflexive, transitive closure of $\to$.
-\lo{The $\tscore$ rule should have a side condition: if $r > 0$.}
 
@@
 %\paragraph{}
-This definition of almost sure termination is equivalent to that given in \citep{MakOPW21}, although the program semantics is stated in a slightly different way.
-In particular, the argument to $\tscore$ is not relevant to termination (except for the possibility that its argument's evaluation wouldn't terminate).
+This definition of almost sure termination is equivalent to that given in \citep{MakOPW21} (although the program semantics is stated in a slightly different way), except for the presence of a $\tscore(r)$ construct for soft conditioning.
+The $\tscore(r)$ construct is not relevant to termination except that it fails if its argument is less than or equal to zero, which allows computations to fail after finitely many steps.
+If such a thing were added, it would merely require an additional side condition to our main theorems, that no failing term was reachable (or, in the case of the confluent semantics based results, no term where the reduction stragegy is not defined is reachable via the reduction strategy unless it's a value).
 

### 2021-01-04 LICS21/confluent-proofs.tex
Remove score entirely.
Remove the score construct from the language definition, and add a
note that this differs from the previous definition of SPCF and it
would be trivial to add it back in.

@@
 For both $\red$ and $\Rightarrow_{\cbv}$, the probability measure on the sample space can be used to define a measure on the reduction sequences of each finite length.
+In the case that the reduction sequences terminate after a finite number of steps by reaching a value, the reduction sequence defined using $\Rightarrow_{\cbv}$ should be taken to continue by repeating the last term indefinitely, as $\red$ does, even though $\Rightarrow_{\cbv}$ specifies no next step in this case.
 The distributions are equal for each length, by induction, as follows.
@@
 Suppose that at some point $j$ in the reduction sequence, the reduction that occurs is not the next reduction in CBV order, but something else, and that the reduction that should be next is at position $\alpha = \cbv(N_i)$, and the position of the actual reduction $N_j \to N_{j+1}$ is $\beta$.
-Suppose further that $N_j$ is not a value.
+Suppose further, for a contradiction, that $N_j$ is not a value.
 Either $\alpha = \cdot$, the root position, or the reduction at the root position is dependent in some way on the reduction at $\alpha$ (e.g.

### 2021-01-04 LICS21/confluent.tex
Remove score entirely.
Remove the score construct from the language definition, and add a
note that this differs from the previous definition of SPCF and it
would be trivial to add it back in.

@@
 \alpha ::= & \; {\cdot} \mid \lambda ; \alpha \mid @_1 ; \alpha \mid @_2 ; \alpha \mid \underline f_i ; \alpha \\
-& {} \mid \tY ; \alpha \mid \textsf{if}_1 ; \alpha \mid \textsf{if}_2 ; \alpha \mid \textsf{if}_3 ; \alpha \mid \tscore ; \alpha.
+& {} \mid \tY ; \alpha \mid \textsf{if}_1 ; \alpha \mid \textsf{if}_2 ; \alpha \mid \textsf{if}_3 ; \alpha.
 \end{align*}
@@
 \tY M \mid \tY ; \alpha & = M \mid \alpha \\
-\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2,3 \\
-\tscore(M) \mid \tscore ; \alpha & = M \mid \alpha
+\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2,3
 \end{align*}
@@
   \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_2/\alpha] \text{ where }r \geq 0 \\
-  \text{if } M \mid \alpha = \tsample \text{ and $\lambda$ does}&\text{ not occur after $@_2$ or $\tY$ in $\alpha$},\\ & M \to M[\underline r/\alpha] \text{ where } r \in [0,1] \\
-  \text{if } M \mid \alpha = \tscore(\underline r),\ & M \to M[\underline r/\alpha].
+  \text{if } M \mid \alpha = \tsample \text{ and $\lambda$ does}&\text{ not occur after $@_2$ or $\tY$ in $\alpha$},\\ & M \to M[\underline r/\alpha] \text{ where } r \in [0,1].
 \end{align*}
@@
 \cbv(\tif{T < 0}{M_1}{M_2}) & = \textsf{if}_1 ; \cbv(T) \\
-\cbv(\tscore(T)) & = \tscore ; \cbv(T) \\
 \cbv(V) & \text{ is undefined} \\

### 2021-01-04 LICS21/ranking-supermartingales.tex
Remove score entirely.
Remove the score construct from the language definition, and add a
note that this differs from the previous definition of SPCF and it
would be trivial to add it back in.

@@
 \label{thm:rankable and strict rankable}
-If a closed SPCF term is rankable (respectively, strictly rankable) by $f$, then $(f(M_n))_{n \in \omega}$ is a 
+If a closed SPCF term is rankable (respectively, strictly rankable) by $f$ and $Rch(M)$ contains no failing terms, then $(f(M_n))_{n \in \omega}$ is a 
 supermartingale (respectively, ranking supermartingale) adapted to the filtration $(\mathcal{F}_n)_{n \in \omega}$ where $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
@@
 N]$, \emph{type-2} if it has the shape $E[\tsample]$; \emph{type-3} if it has the shape $E[R]$ where $R$ is any other redex, and \emph{type-4} if it is a value.
+Failing terms are not included because it is assumed that no failing term is reachable from $M$.
 Henceforth fix an $n \geq 0$, and define $\mathbf{T}_i := \set{s \mid M_n(s) \hbox{ is type-$i$}}$.

### 2021-01-04 LICS21/supermartingales.tex
Remove score entirely.
Remove the score construct from the language definition, and add a
note that this differs from the previous definition of SPCF and it
would be trivial to add it back in.

@@
 \fi
-We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that $f(N) = 0$ iff $N$ is a value; and for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.
+We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.
 
@@
 If a closed SPCF term $M$ is rankable (respectively, strictly rankable) by $f$ 
-%satisfying $f(N) = 0$ iff $N$ is a value, 
 then $(f(M_n))_{n \geq 0}$ is a supermartingale (respectively, ranking supermartingale w.r.t.~stopping time $T_M$) adapted to $(\mathcal{F}_n)_{n \geq 0}$.
@@
 
-Let $f$ be a ranking function on $M$ satisfying $f(N) = 0$ iff $N$ is a value.
+Let $f$ be a ranking function on $M$.
 Define random variables on the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$:

### 2021-01-04 LICS21/SPCF.tex
Corrected def. of antitone ranking functions, and a few minors.

@@
 %\paragraph{}
+\begin{remark}
 This definition of almost sure termination is equivalent to that given in \citep{MakOPW21} (although the program semantics is stated in a slightly different way), except for the presence of a $\tscore(r)$ construct for soft conditioning.
@@
 If such a thing were added, it would merely require an additional side condition to our main theorems, that no failing term was reachable (or, in the case of the confluent semantics based results, no term where the reduction stragegy is not defined is reachable via the reduction strategy unless it's a value).
+\end{remark}
 

### 2021-01-04 LICS21/antitone.tex
Corrected def. of antitone ranking functions, and a few minors.

@@
 
-\changed[lo]{
 \begin{definition}
-%The definition of antitone ranking function $f$ for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
-%we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) function $\epsilon : \nnReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
-%\[f(E[R']) \leq f(E[R]) - \epsilon(f(E[R])) \] where $R \to R'$ is the $\tY$-redex rule.
+\changed[akr]{The definition of \emph{antitone ranking function} $f$ for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
+we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) function $\epsilon : \nnReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
+\[f(E[R']) \leq f(E[R]) - \epsilon(f(E[R])) \] 
+where $R \to R'$ is the $\tY$-redex rule.}
+Any closed term for which an antitone ranking function exists is called \emph{antitone rankable}.
+\iffalse
 Given a ranking function $f$ on $M \in \Lambda^0$, we say that $f$ is \emph{antitone} if there exists an antitone function\footnote{i.e.~$r < r'$ implies $\epsilon(r) \geq \epsilon(r')$} $\epsilon : \nnReal \to \pReal$ such that %the ranking function $f$ satisfies
@@
 for all $E[R'] \in \mathit{Rch}(M)$ where $R \to R'$ is the $\tY$-redex rule.
+\fi
 \end{definition}
+
+\changed[akr]{Note that antitone ranking functions are actually a generalisation of ranking functions, even though their name may suggest that they are a type of ranking functions.}
 
@@
 \).
-Any closed term for which an antitone ranking function exists is called \emph{antitone rankable}.
+%Any closed term for which an antitone ranking function exists is called \emph{antitone rankable}.
 \end{definition}
-}
+
 %\lo{The preceding definition does not place any constraint on the stopping time $T$ (other than that it is w.r.t.~the filtration $(\calF_n)_{n \geq 0}$).

### 2021-01-04 LICS21/confluent.tex
Corrected def. of antitone ranking functions, and a few minors.

@@
 x) ((\lambda x.
-x) \skeletonPlaceholder)$ could reduce to $(\lambda x.
-x) \skeletonPlaceholder$ with the redex at either $\cdot$ or $@_2$, and these give different reduction sequences.
+x) \, \skeletonPlaceholder)$ could reduce to $(\lambda x.
+x) \, \skeletonPlaceholder$ with the redex at either $\cdot$ or $@_2$, and these give different reduction sequences.
 
@@
 \begin{align*}
-\{s \mid s(\mathit{Sk}(A[\tsample]) \to \mathit{Sk}(A[\underline 0]) \to^* \mathit{Sk}(B), \mathsf{if}_1 ; \underline{-}_1) > 0.5 \},
+\big\{s \mid s(\mathit{Sk}(A[\tsample]) \to \mathit{Sk}(A[\underline 0]) \to^* \mathit{Sk}(B), \mathsf{if}_1 ; \underline{-}_1) > 0.5 \big\}
 \end{align*}
@@
 \end{enumerate}
-The $\sim_c$-rules are illustrated in Figure \ref{sim-diagram}.
+The $\sim_c$-rules are illustrated in \Cref{sim-diagram}.
 \end{definition}
 
+\begin{example}
 In \Cref{ex:cousin sim}, $(M,\underline{+}_1) \sim_p (M \to N_2,\underline{+}_1)$ by case i of $\sim_p$ (because the reduction $M \to N_2$ occurs at $\underline{+}_2$, which is disjoint from $\underline{+}_1$), and similarly, $(M,\underline{+}_2) \sim_p (M \to N_1,\underline{+}_2)$.
@@
 The next reduction uses the samples labelled by $(M_{sss} \to M_{Xss} \to M_{XXs}, \underline{+}_3)$ and $(M_{sss} \to M_{sXs} \to M_{XXs}, \underline{+}_3)$, which are related by $\sim_c$, case a, therefore when these reduction sequences reach $M_{XXX}$, they still contain all the same numbers, as desired.
+\end{example}
 
@@
 At each reduction step $M \to N$, the sample space must be restricted from $I^{L_s(M)}$ to $I^{L_s(N)}$. 
-The injection $L_0(N) \to L_0(M)$ is trivial to define by appending $\mathit{Sk}(M) \to \mathit{Sk}(N)$ to each path, and using Lemma~\ref{lem:sim-M-N}, this induces a corresponding injection on the quotient, $L(N) \to L(M)$.
+The injection $L_0(N) \to L_0(M)$ is trivial to define by appending $\mathit{Sk}(M) \to \mathit{Sk}(N)$ to each path, and using \Cref{lem:sim-M-N}, this induces a corresponding injection on the quotient, $L(N) \to L(M)$.
 The corresponding map $L_s(N) \to L_s(M)$ is then denoted $i(M \to N)$.
@@
 This reduction relation now has all of the properties required of it.
-In particular, it can be considered an extension of the standard trace semantics (as will be seen later in Theorem \ref{thm:AstEquivalence}), and also:
+In particular, it can be considered an extension of the standard trace semantics (as will be seen later in \Cref{thm:AstEquivalence}), and also:
 \begin{restatable}{lemma}{churchRosser} \label{churchRosser}
@@
 Even for the same value of the samples, the final values may be different.
-A more restricted result is true, however, that if a term terminates with some reduction strategy $f$ and samples $s$, the value that $\Rightarrow_{\cbv}$ terminates with can reduce via $\Rightarrow$ to the value that $\Rightarrow_f$ terminates with (as a corollary to the proof of Theorem \ref{thm:CbvIsTerminatingest}).
+A more restricted result is true, however, that if a term terminates with some reduction strategy $f$ and samples $s$, the value that $\Rightarrow_{\cbv}$ terminates with can reduce via $\Rightarrow$ to the value that $\Rightarrow_f$ terminates with (as a corollary to the proof of \Cref{thm:CbvIsTerminatingest}).
 
@@
 Let the set of samples with which it terminates with this reduction strategy be $X$.
-By Theorem~\ref{thm:CbvIsTerminatingest}, $M$ also terminates with $\cbv$ and every element of $X$, and $X$ has measure 1, by assumption, therefore $M$ is AST with respect to $\cbv$ therefore by Theorem~\ref{thm:AstEquivalence} it is AST.
+By \Cref{thm:CbvIsTerminatingest}, $M$ also terminates with $\cbv$ and every element of $X$, and $X$ has measure 1, by assumption, therefore $M$ is AST with respect to $\cbv$ therefore by \Cref{thm:AstEquivalence} it is AST.
 \end{proof}
@@
 All of the theorems on the termination of rankable terms therefore extend to other reduction strategies too.
-The proofs of Theorems \ref{thm:rankable implies termination}, \ref{thm:partial implies rankable}, \ref{thm:antitone rankable implies termination} and \ref{thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies.
-They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because Theorem \ref{thm:de groote} applies equally to any reduction strategy).
+The proofs of \Cref{thm:rankable implies termination,thm:partial implies rankable,thm:antitone rankable implies termination,thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies.
+They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because \Cref{thm:de groote} applies equally to any reduction strategy).
 For a reduction strategy $r$ and term $M$, just substitute $N[r(N)]$ being a $\tY$-redex for $N$ being of the form $E[\tY \lambda x.

### 2021-01-04 LICS21/ranking.tex
Corrected def. of antitone ranking functions, and a few minors.

@@
 \fi
-Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.\lo{TODO: reduce the figure to avoid overflow.}\akr{It still overflows the gutter, but not the actual text of the other column.} The meanings of the substitutions (to be applied to the contractum) and conditions that label some of the reductions in the diagram should be self-explanatory.
+Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.\lo{TODO: reduce the figure to avoid overflow.}\akr{It still overflows the gutter, but not the actual text of the other column.}\lo{OK.
+Thanks.} The meanings of the substitutions (to be applied to the contractum) and conditions that label some of the reductions in the diagram should be self-explanatory.
 

### 2021-01-04 LICS21/SPCF.tex
Global change: SPCF -> PPCF.

--- 2021-01-04 LICS21/SPCF.tex
@@
 % !TEX root = main.tex
-\section{Statistical PCF}
-\label{sec:SPCF}
+\section{Probabilistic PCF}
+\label{sec:PPCF}
 
-\subsection{Syntax of SPCF}
+\subsection{Syntax of PPCF}
 
-The language SPCF is a simply-typed lambda calculus with sampling of real numbers from the closed interval $[0,1]$ and unbounded scoring, following \cite{MakOPW21}. 
+The language PPCF is a simply-typed lambda calculus with sampling of real numbers from the closed interval $[0,1]$, \changed[lo]{following \cite{MakOPW21} except the score function (\Cref{rem:score})}.
 Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
@@
 %\paragraph{}
+\begin{remark}
+\label{rem:score}
 This definition of almost sure termination is equivalent to that given in \citep{MakOPW21} (although the program semantics is stated in a slightly different way), except for the presence of a $\tscore(r)$ construct for soft conditioning.

### 2021-01-04 LICS21/antitone.tex
Global change: SPCF -> PPCF.

@@
 
-%\changed[lo]{Let $f$ be a ranking function on a closed SPCF term $M$.
+%\changed[lo]{Let $f$ be a ranking function on a closed PPCF term $M$.
 We say that $M$ is \emph{antitone rankable} by $f$ if the supermartingale $(f(M_n))_{n \geq 0}$ and the stopping time $T_M^{\tY}$, both adapted to the fitration $(\calF_n)_{n \geq 0}$ where $\calF_n := \sigma(M_1, \ldots, M_n)$, are antitone strict.}
@@
 \begin{theorem}[Antitone ranking function soundness] \label{thm:antitone rankable implies termination}
-If a closed SPCF term $M$ is antitone rankable, then $T_M^{\tY} < \infty$ a.s.~(equivalently, $M$ is AST).
+If a closed PPCF term $M$ is antitone rankable, then $T_M^{\tY} < \infty$ a.s.~(equivalently, $M$ is AST).
 \end{theorem}
@@
 \begin{example}[Continuous random walk]\label{ex:raven complex}
-In SPCF we can construct a function whose argument changes by a random amount at each recursive call: $\Theta \, 10$ where
+In PPCF we can construct a function whose argument changes by a random amount at each recursive call: $\Theta \, 10$ where
 \[
@@
 
-\lo{I believe the ranking function method can be used to prove soundness of the Beutner-Ong counting-based method (which reduces the AST problem of first-order (possibly non-affine) recursive SPCF programs to random walk on $\mathbb{N}$), based on Theorem 5.9 and Lemma 5.10 in \emph{op.~cit.}}
+\lo{I believe the ranking function method can be used to prove soundness of the Beutner-Ong counting-based method (which reduces the AST problem of first-order (possibly non-affine) recursive PPCF programs to random walk on $\mathbb{N}$), based on Theorem 5.9 and Lemma 5.10 in \emph{op.~cit.}}
 
@@
 There is an obvious source of deterministic (i.e.~non-probabilistic) higher-order functions defined by recursion, viz., \emph{higher-order recursion schemes} (HORS) (see e.g.~\citep{DBLP:conf/lics/Ong06,DBLP:conf/lics/Ong15}).
-(Incidentally the (sparse) ranking function method is just as applicable to the termination analysis of \emph{deterministic} SPCF programs.)
+(Incidentally the (sparse) ranking function method is just as applicable to the termination analysis of \emph{deterministic} PPCF programs.)
 Recently \cite{DBLP:conf/lics/KobayashiLG19} have extended HORS to \emph{probabilistic higher-order recursion schemes} (PHORS), which are HORS augmented with probabilistic (binary) branching $\oplus_p$.
-As HORS are in essence the $\lambda \tY$-calculus \citep{DBLP:conf/lics/Statman02} (i.e.~pure simply-typed lambda calculus with recursion, generated from a finite base type), PHORS are definable in SPCF:
-(order-$n$) PHORS is encodable as (order-$n$) (call-by-name) SPCF, but the former is strictly less expressive (because the underlying recursion schemes are not Turing complete). 
-Some interesting SPCF terms such as \Cref{ex:non-affine with continuous distribution} %\refExample{complexExample} 
+As HORS are in essence the $\lambda \tY$-calculus \citep{DBLP:conf/lics/Statman02} (i.e.~pure simply-typed lambda calculus with recursion, generated from a finite base type), PHORS are definable in PPCF:
+(order-$n$) PHORS is encodable as (order-$n$) (call-by-name) PPCF, but the former is strictly less expressive (because the underlying recursion schemes are not Turing complete). 
+Some interesting PPCF terms such as \Cref{ex:non-affine with continuous distribution} %\refExample{complexExample} 
 cannot be expressed as PHORS.

### 2021-01-04 LICS21/confluent.tex
Global change: SPCF -> PPCF.

@@
 \begin{theorem}
-If a closed SPCF term $M$ has a ranking function, sparse ranking function, antitone ranking function or antitone sparse ranking function with respect to a reduction strategy $r$, then $M$ is AST with respect to $r$, and AST.
+If a closed PPCF term $M$ has a ranking function, sparse ranking function, antitone ranking function or antitone sparse ranking function with respect to a reduction strategy $r$, then $M$ is AST with respect to $r$, and AST.
 \end{theorem}

### 2021-01-04 LICS21/supermartingales-proofs.tex
Remove score entirely.
Remove the score construct from the language definition, and add a
note that this differs from the previous definition of SPCF and it
would be trivial to add it back in.
Global change: SPCF -> PPCF.

@@
 }
-%Take $M \in \Lambda^0$ (closed SPCF terms). Define, for each $n \in \omega$, the random variable $M_n : \entrosp \to \Lambda^0$ by $M_n(s) := \pi_0 (\red^n(M, s))$.
+%Take $M \in \Lambda^0$ (closed PPCF terms). Define, for each $n \in \omega$, the random variable $M_n : \entrosp \to \Lambda^0$ by $M_n(s) := \pi_0 (\red^n(M, s))$.
 
@@
 
-Given a closed SPCF term $M$, we transform it to a term $\transform{M}$ of the nondeterministic simply-typed lambda-calculus as follows.
+Given a closed PPCF term $M$, we transform it to a term $\transform{M}$ of the nondeterministic simply-typed lambda-calculus as follows.
 (We may assume that the nondeterministic calculus is generated from a base type $\iota$ with a $\iota$-type constant symbol $r$, and a function symbol $\bot : A$  for each function type $A$.)
@@
 \item replace every $\tY \, f \, x . N$ by $\lambda x . \transform{N}[\bot / f]$
-
-\item replace every $\tscore (N)$ by $\transform{N}$ (we set aside the requirement that $\tscore$ is undefined on negative numbers)
 \end{enumerate}
@@
 \transform{\tif{B}{M_1}{M_2}} &:= \big(\lambda z . (\transform{M_1} + \transform{M_2})\big) \, \transform{B} \\
-\transform{\tY \, f \, y . N} &:= \lambda y . \transform{N}[\bot / f]\\
-\transform{\tscore (N)} &:= \transform{N}
+\transform{\tY \, f \, y . N} &:= \lambda y . \transform{N}[\bot / f]
 \end{align*}
 In the above, variables $z, z_1, \cdots, z_n$ are assumed to be fresh; $f$ ranges over primitive functions and numerals.
-In transforming $\tscore (N)$, we set aside the requirement that $\tscore$ is undefined on negative numbers.
 

### 2021-01-04 LICS21/supermartingales.tex
Global change: SPCF -> PPCF.

@@
 
-Henceforth, fix the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$, and a closed SPCF term $M$.
+Henceforth, fix the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$, and a closed PPCF term $M$.
 For $n \geq 0$, define the random variables 
@@
 \label{thm:rankable and strict rankable}
-If a closed SPCF term $M$ is rankable (respectively, strictly rankable) by $f$ 
+If a closed PPCF term $M$ is rankable (respectively, strictly rankable) by $f$ 
 then $(f(M_n))_{n \geq 0}$ is a supermartingale (respectively, ranking supermartingale w.r.t.~stopping time $T_M$) adapted to $(\mathcal{F}_n)_{n \geq 0}$.
@@
 
-We say that a given SPCF term is: %\emph{type-1} (respectively 2, 3 and 4) if it has the shape $E[\tY \lambda x. N]$ (respectively $E[\tsample]$, $E[R]$ where $R$ is any other redex, and of a value).
+We say that a given PPCF term is: %\emph{type-1} (respectively 2, 3 and 4) if it has the shape $E[\tY \lambda x. N]$ (respectively $E[\tsample]$, $E[R]$ where $R$ is any other redex, and of a value).
 \emph{type 1} if it has the shape $E[\tY \lambda x.
@@
 
-In the rest of this section we show that if $M \in \Lambda^0$ is rankable, then $M$ is AST: in other words, the method of ranking function is sound for proving a.s.~termination of SPCF programs.
+In the rest of this section we show that if $M \in \Lambda^0$ is rankable, then $M$ is AST: in other words, the method of ranking function is sound for proving a.s.~termination of PPCF programs.
 
@@
 \begin{enumerate}
-\item If a closed SPCF term $M$ is rankable, then $M$ is AST (equivalently, $T^{\tY}_M < \infty$ almost surely), and $M$ is $\tY$-positively almost sure terminating ($\tY$-PAST) i.e.~$\expect{T^{\tY}_M} < \infty$.
+\item If a closed PPCF term $M$ is rankable, then $M$ is AST (equivalently, $T^{\tY}_M < \infty$ almost surely), and $M$ is $\tY$-positively almost sure terminating ($\tY$-PAST) i.e.~$\expect{T^{\tY}_M} < \infty$.
 
-\item If a closed SPCF term $M$ is strictly rankable, then $\expect{T_M} < \infty$ i.e.~$M$ is positively almost sure terminating (PAST).
+\item If a closed PPCF term $M$ is strictly rankable, then $\expect{T_M} < \infty$ i.e.~$M$ is positively almost sure terminating (PAST).
 \end{enumerate}
@@
 
-Thus the method of (strict) ranking function is sound for proving (positive) a.s.~termination of SPCF programs.
+Thus the method of (strict) ranking function is sound for proving (positive) a.s.~termination of PPCF programs.
 It is in fact also complete in a sense: if $\expect{T^\tY_N} < \infty$ for all $N \in \mathit{Rch}(M)$ then $M$ is rankable (\Cref{thm:minimal}).

### 2021-01-04 LICS21/SPCF.tex
Minor: def. of PPCF

@@
 
-The language PPCF is a simply-typed lambda calculus with sampling of real numbers from the closed interval $[0,1]$, \changed[lo]{following \cite{MakOPW21} except the score function (\Cref{rem:score})}.
+The language PPCF is a simply-typed lambda calculus with sampling of real numbers from the closed interval $[0,1]$. 
+%\changed[lo]{following \cite{MakOPW21} except the score function (\Cref{rem:score})}. 
 Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
@@
 %\paragraph{}
-To define the reduction relation, let \emph{evaluation contexts} be of the form:
+To define the call-by-value reduction relation, let \emph{evaluation contexts} be of the form:
 \begin{align*}

### 2021-01-04 LICS21/ranking.tex
Minor: def. of PPCF

@@
 The definition of rankability is also inconvenient for syntactic sugar.
-It could be useful, for example, to define $M \oplus_p N = \tif{\tsample - p < 0} M N$, where $M \oplus_p N$ reduces to $M$ or $N$, depending on the first value of $s \in \entrosp$\lo{Need a name for $s$.}, with probability $p$ resp.~$(1-p)$.
+It could be useful, for example, to define $M \oplus_p N = \tif{\tsample - p < 0} M N$, where $M \oplus_p N$ reduces to $M$ or $N$, depending on the first value of $s \in \entrosp$, with probability $p$ resp.~$(1-p)$.
 Technically though, it reduces first to $\tif{\underline r - p < 0} M N$ for all $r \in I$, so those terms all need values of the ranking function too.

### 2021-01-05 LICS21/antitone.tex
Sec. VI. Ex. VI.2

@@
 
-\changed[akr]{Note that antitone ranking functions are actually a generalisation of ranking functions, even though their name may suggest that they are a type of ranking functions.}
+\changed[lo]{Note that antitone ranking functions are actually a generalisation of ranking functions, even though the way we reference them may suggest that they are a type of ranking functions.}
 
@@
 However for the definition to make sense, $T$ must be the stopping time $T(s) := \min \set{n \mid Y_n(s) = 0}$.}
-\akr{The definition you gave for an antitone ranking function is incorrect, because the antitone ranking functions are a proper subset of ranking functions.
+%\akr{The definition you gave for an antitone ranking function is incorrect, because the antitone ranking functions are a proper subset of ranking functions.
 Although we may have been referring to antitone ranking functions as a type of ranking function, they're a generalisation, so they don't actually meet the definition of a ranking function in general.}

### 2021-01-05 LICS21/confluent.tex
Sec. VI. Ex. VI.2

@@
 M = \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})
-\qquad
+\quad
 \alpha =\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2
@@
 Two subterms $N_1$ and $N_2$ of a term $M$, corresponding to positions $\alpha_1$ and $\alpha_2$, can overlap in a few different ways. 
-If $\alpha_1$ is an initial segment (i.e.~prefix) of $\alpha_2$ (written as $\alpha_1 \leq \alpha_2$), then $N_2$ is also a subterm of $N_1$.
+If $\alpha_1$ is a prefix of $\alpha_2$ (written as $\alpha_1 \leq \alpha_2$), then $N_2$ is also a subterm of $N_1$.
 If neither $\alpha_1 \leq \alpha_2$ nor $\alpha_1 \geq \alpha_2$, the positions are said to be \emph{disjoint}.
@@
 
-We are thus led to consider the reduction relation on skeletons (and positions in a skeleton), which can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, and $\tsample$ reducing to $\skeletonPlaceholder$.
+We are thus led to consider the reduction relation on skeletons (and positions in a skeleton), which can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, and $\tsample$ reducing to $\skeletonPlaceholder$, \changed[lo]{and $(\lambda x.A) \, \skeletonPlaceholder$ reducing to $A[\skeletonPlaceholder / x]$}.
 For example, we have 
@@
 Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a $\to$-reduction sequence of skeletons starting at $\mathit{Sk}(M)$, and the second of which is a position in the final skeleton of the reduction sequence.
-As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the standard sampling semantics, there will be (with one more caveat introduced later) a sample from $I$ pre-selected for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.
+As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the standard sampling semantics, 
+%there will be (with one more caveat introduced later) a sample from $I$ pre-selected for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.
+\changed[lo]{modified traces, which are elements of $I^{L_0(M)}$ (with one more caveat introduced later)}, will be used to pre-select a sample from $I$ for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.
 
@@
 \end{align*}
-If terms rather than skeletons were used to label samples, the set of traces where $A[\tsample]$ terminates would be
+If terms rather than skeletons were used to label samples, the set of \changed[akr]{``traces''} \lo{Readers may not realise that traces here means elements of $I^{L_0(M)}$ (but with samples labelled by terms).
+I propose replacing the highlighted ``traces'' by ``modified traces'' (which is introduced two paragraphs before).
+I think this might be a little clearer, but feel free to change back.} where $A[\tsample]$ terminates would be
 \begin{align*}
@@
 \lo{We have just defined $L_0(M)$, but what is $L_s(\cdot)$ (subscript $s$)? I assume it is the version of $L_0(M)$ that labels samples by terms.} 
-
 \akr{$L_s(M)$ is the set of potential positions (quotiented by $\sim^*$) where there is a $\tsample$.
 I don't want to move this example until after $L_s(M)$ has actually been defined, but also I don't want to imply that $I^{L_0(M)}$ is the actual entropy space too explicitly, so I'll just leave the set $s$ is restricted to implicit for now.}
+\lo{Noted.}
 
+\iffalse
 \lo{The occurrence of $L_s(M)$ in the set comprehension below should be $L_0(M)$?}
@@
 \lo{I think we should write ``$[A[\tsample], A[\underline r], \ldots, B$'' in the line above as ``$A[\tsample] \to A[\underline r] \to^\ast B$''; similarly $\mathit{Sk}(A[\tsample]) \to \mathit{Sk}(A[\underline 0]) \to^\ast \mathit{Sk}(B)$ below.}
+\fi
 This is a rather unwieldy expression, but the crucial part is that $r$ occurs twice in the conditions on $s$: once as the value a sample must take, and once in the location of a sample. 
@@
 
-For this particular term, $\mathit{Sk}(A[\underline r])$ does not depend on the value of $r$, therefore the set where it terminates becomes simply
+For this particular term, $\mathit{Sk}(A[\underline r])$ does not depend on the value of $r$, therefore the set where it terminates becomes simply the following, which is measurable.
 \begin{align*}
@@
 \end{align*}
-which is measurable.
-\qed
 \end{example}
@@
 The new sample space is then defined as $I^{L_s(M)}$, with the Borel $\sigma$-algebra and product measure.
-\changed[lo]{Since $I^{L_s(M)}$ is a countable product, the measure space is well-defined \citep[Cor.~2.7.3]{AshDD00}.}
+Since $I^{L_s(M)}$ is a countable product, the measure space is well-defined \citep[Cor.~2.7.3]{AshDD00}.
 
@@
 
+\begin{definition}
 Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function, so it is denoted ``$\Rightarrow$'' instead of ``$\red$'', and it relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
@@
 \end{align*}
+\end{definition}
 

### 2021-01-05 LICS21/ranking.tex
Sec. VI. Ex. VI.2

@@
 \[
-\Xi := \tY \lambda f n. \tif{\tsample - 0.5 < 0}{n \, \underline 0}{f (\lambda x. n (n x))}
+\Xi := \tY \lambda f \, n. \tif{\tsample - 0.5 < 0}{n \, \underline 0}{f (\lambda x. n (n \, x))}
 \] 
@@
 The definition of rankability is also inconvenient for syntactic sugar.
-It could be useful, for example, to define $M \oplus_p N = \tif{\tsample - p < 0} M N$, where $M \oplus_p N$ reduces to $M$ or $N$, depending on the first value of $s \in \entrosp$, with probability $p$ resp.~$(1-p)$.
+It could be useful, for example, to define $M \oplus_p N := \tif{\tsample - p < 0} M N$, where $M \oplus_p N$ reduces to $M$ or $N$, depending on the first value of $s \in \entrosp$, with probability $p$ resp.~$(1-p)$.
 Technically though, it reduces first to $\tif{\underline r - p < 0} M N$ for all $r \in I$, so those terms all need values of the ranking function too.
@@
     \item $f$ is defined at $M$.
-    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually \akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and ast, but that doesn't actually work because of terms reachable with 0 probability.
-Of course, astness doesn't imply rankability, so any definition of sparse ranking function that includes all Y-past terms wouldn't always extend to a total ranking function.} reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
+    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually \akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and AST, but that doesn't actually work because of terms reachable with 0 probability.
+Of course, ASTness doesn't imply rankability, so any definition of sparse ranking function that includes all Y-PAST terms wouldn't always extend to a total ranking function.} reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
 \end{itemize}
@@
 \begin{align*}
-(\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\ N & \mapsto 2 \\
-\underline i \oplus_{0.5} (\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\ (\underline i + 1) & \mapsto 1 \\
+(\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\, N & \mapsto 2 \\
+\underline i \oplus_{0.5} (\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\, (\underline i + 1) & \mapsto 1 \\
 \underline i & \mapsto 0.

### 2021-01-06 LICS21/SPCF.tex
Tidying up comments

@@
 Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
-\akr{\begin{align*}
+\begin{align*}
   \text{\em types } A, B &::= \textsf{R}  \mid  A \to B \\
@@
   \text{\em terms } M, N &::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY \, M \\
-   & \qquad \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample
-\end{align*}}
-\lo{Currently $\tY$ is not a legal term.
-This is unnatural: there is no reason to so restrict $\tY$.
-Following \citep[p.~376]{DBLP:conf/fsttcs/Sieber90}, we should define $\tY$ as a value (and so a proper term); it follows that $\tY \, M$ is redundant as a clause.}
-\lo{N.B. It is also standard to regard variables as values in a CBV language.}
-
-\lo{Problem: $\tY \, \tY$ is then a ``stuck'' term: neither value nor redex.
-A possible solution is to redefine the $\tY$-redex rule as $\tY \, V \to \lambda z .
-V \, (\tY \, V) \, z$, but I expect this would affect many proofs in the confluent semantics section, and there may not be time for such a change.}
-\begin{align*}
-  \text{\em types } A, B &::= \textsf{R}  \mid  A \to B \\
-  \text{\em values } V &::= x \mid \lambda x.M  \mid  \underline{r} \mid \tY \\
-  \text{\em terms } M, N &::= V  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n) \\
    & \qquad \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample

### 2021-01-06 LICS21/ranking.tex
Tidying up comments

@@
 \fi
-Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.\lo{TODO: reduce the figure to avoid overflow.}\akr{It still overflows the gutter, but not the actual text of the other column.}\lo{OK.
-Thanks.} The meanings of the substitutions (to be applied to the contractum) and conditions that label some of the reductions in the diagram should be self-explanatory.
+Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.
+%\lo{TODO: reduce the figure to avoid overflow.}\akr{It still overflows the gutter, but not the actual text of the other column.}\lo{OK. Thanks.} 
+The meanings of the substitutions (to be applied to the contractum) and conditions that label some of the reductions in the diagram should be self-explanatory.
 

### 2021-01-06 LICS21/supermartingales.tex
Tidying up comments

@@
 \begin{remark}
-\changed[lo]{\Cref{lem:rank-PAST} is a slight extension of \citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15},
-which asserts the same result but for the specific stopping time $T : \omega \mapsto \min\set{n \mid Y_n(\omega) = 0}$.}
+\Cref{lem:rank-PAST} is a slight extension of \citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15},
+which asserts the same result but for the specific stopping time $T : \omega \mapsto \min\set{n \mid Y_n(\omega) = 0}$.
 \end{remark}

### 2021-01-07 LICS21/SPCF.tex
Resolve comments

@@
 To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples.
-Let $ I := [0,1] \subset \mathbb{R} $, and define $\entrosp := I^{\mathbb{N}}$, with the \changed[lo]{Borel $\sigma$-algebra $\Sigma_\entrosp$} and the probability measure, denoted $\mu_{\entrosp}$, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
+Let $ I := [0,1] \subset \mathbb{R} $, and define $\entrosp := I^{\mathbb{N}}$, with the Borel $\sigma$-algebra $\Sigma_\entrosp$ and the probability measure, denoted $\mu_{\entrosp}$, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
 Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and $\mu_{\entrosp} (\prod_{i=0}^\infty X_i) := \prod_{i=0}^\infty \Leb(X_i)$.
 The maps $\pi_h:\entrosp \to I, \; \pi_t:\entrosp \to \entrosp$ popping the first element are then measurable.
-\changed[lo]{Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$ the \emph{entropy space}, and call elements of $\entrosp$ \emph{traces}.}
+Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$ the \emph{entropy space}, and call elements of $\entrosp$ \emph{traces}.
 

### 2021-01-07 LICS21/confluent-proofs.tex
Resolve comments

@@
 Suppose for induction that $A \sim_\uparrow^k (D,\delta) \sim_\downarrow^* (E,\epsilon) \sim_\uparrow^* C$ for some $D, \delta, E, \epsilon$, and $0 \leq k \leq n$, where in the reduction sequence $D \to^* E$, for any pair of reductions at positions which aren't disjoint, the reduction at the innermost (greater) position occurs first.
-Reduction sequences with this property will be called ``parallel'', as it is analagous to \changed[lo]{the parallel reduction introduced by Tait and Martin-L\"of in their proof of the Church-Rosser theorem \citep{Bar84,Takahashi1995}.}
-  \akr{TODO: find this reference, for the proof of the Church-Rosser theorem by Tait and Martin-L\"of.}
-  \lo{Tait and Martin-L\"of's proof is folklore; I don't think there is a first-hand documentary record.} This induction is in reverse, with $k$ decreasing from $n$ to $0$.
+Reduction sequences with this property will be called ``parallel'', as it is analagous to the parallel reduction introduced by Tait and Martin-L\"of in their proof of the Church-Rosser theorem \citep{Bar84,Takahashi1995}.
+This induction is in reverse, with $k$ decreasing from $n$ to $0$.
 If $k = n$, simply set $(D,\delta) = B$ and $(E,\epsilon) = C$.

### 2021-01-07 LICS21/confluent.tex
Resolve comments

@@
 
-We are thus led to consider the reduction relation on skeletons (and positions in a skeleton), which can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, and $\tsample$ reducing to $\skeletonPlaceholder$, \changed[lo]{and $(\lambda x.A) \, \skeletonPlaceholder$ reducing to $A[\skeletonPlaceholder / x]$}.
+We are thus led to consider the reduction relation on skeletons (and positions in a skeleton), which can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, $\tsample$ reducing to $\skeletonPlaceholder$, and $\skeletonPlaceholder$ considered as (the skeletal equivalent of) a value, so that $(\lambda x.A) \, \skeletonPlaceholder$ reduces to $A[\skeletonPlaceholder / x]$.
 For example, we have 
@@
 Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a $\to$-reduction sequence of skeletons starting at $\mathit{Sk}(M)$, and the second of which is a position in the final skeleton of the reduction sequence.
-As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the standard sampling semantics, 
-%there will be (with one more caveat introduced later) a sample from $I$ pre-selected for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.
-\changed[lo]{modified traces, which are elements of $I^{L_0(M)}$ (with one more caveat introduced later)}, will be used to pre-select a sample from $I$ for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.
+As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the standard sampling semantics, modified traces, which are elements of $I^{L_0(M)}$ (with one more caveat introduced later), will be used to pre-select a sample from $I$ for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.
 
@@
 \end{align*}
-If terms rather than skeletons were used to label samples, the set of \changed[akr]{``traces''} \lo{Readers may not realise that traces here means elements of $I^{L_0(M)}$ (but with samples labelled by terms).
-I propose replacing the highlighted ``traces'' by ``modified traces'' (which is introduced two paragraphs before).
-I think this might be a little clearer, but feel free to change back.} where $A[\tsample]$ terminates would be
+If terms rather than skeletons were used to label samples, the set of modified traces where $A[\tsample]$ terminates would be
 \begin{align*}
@@
 \end{align*}
-
-\lo{We have just defined $L_0(M)$, but what is $L_s(\cdot)$ (subscript $s$)? I assume it is the version of $L_0(M)$ that labels samples by terms.} 
-\akr{$L_s(M)$ is the set of potential positions (quotiented by $\sim^*$) where there is a $\tsample$.
-I don't want to move this example until after $L_s(M)$ has actually been defined, but also I don't want to imply that $I^{L_0(M)}$ is the actual entropy space too explicitly, so I'll just leave the set $s$ is restricted to implicit for now.}
-\lo{Noted.}
 
@@
 Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function, so it is denoted ``$\Rightarrow$'' instead of ``$\red$'', and it relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
-\changed[lo]{We write an element of $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ as $(M', s)$ where the \emph{term} $M' \in \Lambda^0$ and $s \in  I^{L_s(M')}$.}
+We write an element of $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ as $(M', s)$ where the \emph{term} $M' \in \Lambda^0$ and $s \in  I^{L_s(M')}$.
 \begin{align*}

### 2021-01-07 LICS21/notes.tex
Added notes.tex

% !TEX root = main.tex

\begin{align*}
  \text{\em types } A, B &::= \textsf{R}  \mid  A \to B \\
  \text{\em values } V &::= x \mid \lambda x.M  \mid  \underline{r} \mid \tY \\
  \text{\em terms } M, N &::= V  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n) \\
   & \qquad \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample
\end{align*}
Currently $\tY$ is not a legal term. This is unnatural: there is no reason to so restrict $\tY$. Following \citep[p.~376]{DBLP:conf/fsttcs/Sieber90}, we should define $\tY$ as a value (and so a proper term); it follows that $\tY \, M$ is redundant as a clause.

It is also standard to regard variables as values in a CBV language.

Problem: $\tY \, \tY$ is then a ``stuck'' term: neither value nor redex. A possible solution is to redefine the $\tY$-redex rule as $\tY \, V \to \lambda z . V \, (\tY \, V) \, z$, but I expect this would affect many proofs in the confluent semantics section, and there may not be time for such a change.

### 2021-01-07 LICS21/antitone.tex
Add motivating examples for the confluent semantics.

@@
 \paragraph*{\Cref{ex:ac-ranking} revisited}
-\akr{All of these ranking functions are slightly incorrect, because $\Xi \, \underline n$ reduces to something like $\Xi \, (\underline n + 1)$ then $(\lambda z.
-\text{something}[\Xi/f]) (\underline n + 1)$, rather than reducing the argument first.
-This could be fixed by appealing to the confluent semantics version of the antitone ranking theorem, but it would be nice to have at least some examples here illustrating how antitone ranking functions work, rather than actually saving all these examples until the end.
-
-Even though the difference is quite trivial, this was actually the original motivating example for the confluent trace semantics.}
 Programs $M_1$ and $M_2$ are AST. For $M_1$, the sparse ranking function 
@@
 % \end{align*}
-For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$, each with probability $1/2$, with one $\tY$-reduction.
+For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$,\footnote{Technically this is not quite correct, because of the distinction between $\underline n + 1$ and $\underline{n + 1}$, but this ranking function can still be made to work by appealing to \Cref{thm:confluent ranking}, as can the others in this section with similar issues.} each with probability $1/2$, with one $\tY$-reduction.
 Now

### 2021-01-07 LICS21/confluent-proofs.tex
Add motivating examples for the confluent semantics.

@@
 \end{proof}
+
+\confluentNoStronger*
+\begin{proof}
+Suppose that $(M,s) \Rightarrow^*_{\cbv} (F,s')$ for some trace $s$ and term $F$ which is not AST.
+Let $P$ be the finite set of potential positions in $M$ such that the corresponding sample is used in the reductions $(M,s) \Rightarrow^*_{\cbv} (F,s')$.
+As $F$ is not AST, the set $T_0 = \{t \in I^{L_s(M) \setminus P} \mid M \text{ terminates with } (t,s|_P) \text{ and } \cbv \}$ has non-zero measure.
+
+Take some order $p_1, \dots$ on the elements of $P$, and define $T_n$ recursively as follows.
+If the sample at $p_n$ is eventually used in the sequence $(M,(t,s|_P)) \Rightarrow^*_r \dots$ for some positive measure subset of $T_{n-1}$, then let $T_n$ be the subset of $T_{n-1}$ where it is used within $m$ steps, for the minimal $m$ such that this is of positive measure.
+Otherwise, let $T_n$ be the subset of $T_{n-1}$ where the sample at $p_n$ is never used.
+Let $T$ be the final such $T_n$.
+Also, let $P_i$ be the set of $p_n$ such that the second case was taken in the definition of $T_n$ i.e.~the sample at $p_n$ is never used in the reduction sequences starting at $(M,(t,s|_P))$ with strategy $r$, and let $k$ be the maximum value of $m$.
+
+Let $T' = T \times \prod_{p \in P_i} I \times \prod{p \in P \setminus P_i} \{s(p)\}$.
+Although $T'$ itself may have a measure of 0 as a subset of $I^{L_s(M)}$, it still has a natural measure of its own.
+It also has the properties that $M$ does not terminate with $\cbv$ and any trace in $T'$ (therefore $M$ also does not terminate with $r$ and any trace in $T'$, by \Cref{thm:CbvIsTerminatingest}), and for any trace $t \in T'$, if $(M,t) \Rightarrow^k_r (M',t')$, none of the potential positions in $M'$ correspond to any of the elements of $P \setminus P_i$.
+There are finitely many skeletal reduction sequences $k$ steps long starting at $M$ and following $r$, each one corresponding to some measurable subset of $I^{L_s(M)}$, so pick one such that its intersection with $T'$ is a non-null subset of $T'$, let this subset be $Q$.
+
+Pick some $p \in Q$, and let $(M,p) \Rightarrow^k_r (N,p')$.
+The injection $i(M \to^k N)$ is the same as it would be for any other element of $Q$, and its image is disjoint from $P \setminus P_i$.
+The inverse image under composition with $i(M \to^k N)$ of $Q$ is therefore a non-null subset of $I^{L_s(N)}$, and $N$ fails to terminate with $r$ and every trace in this set, therefore $N$ is not AST with respect to $r$, therefore $N$ is not AST, but $N \in Rch_r(M)$, therefore not every element of $Rch_r(M)$ is AST.
+
+Taking the contrapositive of this, we have the desired result.
+\end{proof}

### 2021-01-07 LICS21/confluent.tex
Add motivating examples for the confluent semantics.

@@
 
-\begin{theorem}
+\begin{theorem} \label{thm:confluent ranking}
 If a closed PPCF term $M$ has a ranking function, sparse ranking function, antitone ranking function or antitone sparse ranking function with respect to a reduction strategy $r$, then $M$ is AST with respect to $r$, and AST.
 \end{theorem}
-\lo{Give an example or two of a reduction strategy (for $\Rightarrow$) that is not $\cbv$.}
+
+\paragraph{}
+If, as seems likely, \Cref{thm:antitone rankable implies termination} is complete for terms from which every reachable term is AST, then by the following theorem, \Cref{thm:confluent ranking} can't actually prove the termination of any terms not already provable by \Cref{thm:antitone rankable implies termination} and \Cref{thm:antitone partial implies rankable}.
+
+\begin{restatable}{theorem}{confluentNoStronger}
+For any closed term $M$ and reduction strategy $r$ on $M$, if every term in $Rch_r(M)$ is AST, then every term in $Rch(M)$ is AST.
+\end{restatable}
+
+However, there are some terms which terminate more simply and directly via some alternative reduction strategy.
+For these examples, a custom reduction strategy is specified for each term under consideration.
+
+\begin{example}
+The following term picks a random natural number $n$, then computes $\sum_{k=0}^n k^n$.
+\begin{align*}
+M =&\ (\lambda n. P[n]) \lfloor \tsample^{-1/2} \rfloor \\
+P[n] =&\ (\lambda p. \Xi[p] n) (\lambda x. \Theta[x] n)) \\
+\Xi[p] =&\ \tY \lambda f n. \tif{n = 0}{\underline 0}{p n + f (n-1)} \\
+\Theta[x] =&\ \tY \lambda f n. \tif{n = 0}{\underline 1}{x \times f (n-1)}.
+\end{align*}
+
+Let $r$ be the reduction strategy which reduces $M$ in the order
+\begin{align*}
+&\ M \\
+\to^* &\ P[\underline n] \\
+\to^* &\ (\lambda p.\ \Xi[p]\ \underline n)\ (\lambda x. \underbrace{x \times \dots \times x \times}_n \underline 1) \\
+\to^{\phantom *} &\ \Xi[\lambda x. \underbrace{x \times \dots \times x \times}_n \underline 1]\ \underline n \\
+\to^{\phantom *} &\ \tif{\underline n = 0}{\underline 0}{\underline{n^n} + \Xi[\lambda x. x \times \dots \underline 1]\ \underline{n-1}} \\
+\to^* &\ \underline{n^n} + \underline{(n-1)^n} + \dots + \underline 0 \\
+\to^* &\ \underline{\sum_{k=0}^n k^n}
+\end{align*}
+(this is a little underspecified, but the details are not important). A suitable sparse ranking function for it is
+\begin{align*}
+M \mapsto & \frac{\pi^2}{3} \\
+P[\underline n] \mapsto & 2n+2 \\
+(\lambda p.\ \Xi[p]\ \underline n)\ (\lambda x. \underbrace{x \times \dots \times x \times}_{n-k} (\Theta[x]\ \underline k)) \mapsto & n + k + 2 \\
+\Xi[\lambda x. \underbrace{x \times \dots \times x \times}_n \underline 1]\ \underline n \mapsto & n + 1 \\
+\underline{n^n} + \dots + \underline{(n-k+1)^n} + \Xi[\lambda x. x \times \dots \underline 1]\ \underline k \mapsto & k + 1 \\
+\underline{\sum_{k=0}^n k^n} \mapsto & 0.
+\end{align*}
+
+The first step is justified because the probability of reaching $P[\underline n]$ for any specific $n \geq 0$ is $(n+1)^{-2} - (n+2)^{-2} = \frac{2n + 3}{(n+1)^2(n+2)^2}$, therefore the expected next value of the ranking function is $\sum_{n=0}^{\infty} \frac{(2n+2)(2n + 3)}{(n+1)^2(n+2)^2} = \frac{\pi^2}{3}$.
+The other steps are deterministic, and involve 1 or 0 $\tY$-reductions each.
+Given that all of the reduction steps after the first are deterministic, and the number of $\tY$ reduction steps is not that hard to count, defining the sparse ranking function's values only at $M$ and $P[\underline n]$ would also have been reasonable, although in a similar term which had more random samples throughout, that would not have been so simple.
+
+Providing a ranking function of some sort for this term in a similar level of detail would have been rather more complicated using only the standard reduction strategy.
+The recursion in $\Theta$ would have had to be evaluated separately for every time it was used, so there would have been more (and more complex) terms to consider in the reduction sequence.
+Also, the number of $\tY$-reductions from $P[\underline n]$ would have been $n^2 + 2n + 1$ instead of $2n + 2$, therefore the expected number of $\tY$ reductions starting from $M$ would be infinite, i.e.~$M$ is not $\tY$-PAST, therefore it is not even rankable.
+It would be possible to define an antitone sparse ranking function for it, but it would have to include terms such as
+\begin{align*}
+& \underline{n^n} + \dots + \underline{(n-k_1+1)^n} + \\
+& (\underbrace{\underline{n-k_1} \times \dots \times \underline{n-k_1} \times}_{n-k_2-1} (\Theta[\underline{n-k_1}] (\underline{k_2} - 1))) + \\
+& \Xi[\lambda x. \Theta[x] \underline n] (\underline{n-k_1} - 1) \mapsto \log((n+1)k_1 + k_2)
+\end{align*}
+as only one of the several cases, which is significantly more complicated and difficult to use.
+
+It wouldn't even be possible in this case to give a sparser version of the ranking function defined only at $M$ and $P[\underline n]$, because the amount that an antitone sparse ranking function must decrease at each $\tY$-reduction step depends on the value of the ranking function at the next term where it is defined, therefore it is necessary to have these intermediate steps in order for the ranking function to be able to change sufficiently slowly.
+\end{example}
+
+Even some of the examples given earlier implicitly use slightly non-standard reduction strategies to make the definitions of their ranking functions slightly similar.
+For example, the antitone sparse ranking function in \Cref{ex:unbiased random walk} was defined at $\Theta_2\ \underline n$ (where $\Theta_2 = \tY \, \lambda f n .
+\, \tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$), but $\Theta_2\ \underline{10}$ actually reduces to $\Theta_2\ (\underline{10} - 1)$ or $\Theta_2\ (\underline{10} + 1)$, and (taking the second of these), it then reduces to $(\lambda z.
+(\lambda n.
+\, \tif{n = 0}{0}{\Theta_2 \, (n - 1) \oplus_{1/2} \Theta_2 \, (n + 1)}) z) \, (\underline{10} + 1)$, then $(\lambda z.
+(\lambda n.
+\, \tif{n = 0}{0}{\Theta_2 \, (n - 1) \oplus_{1/2} \Theta_2 \, (n + 1)}) z) \, \underline{11}$, bypassing $\Theta_2\ \underline{11}$ entirely.
+The antitone sparse ranking function in the form given can be justified anyway, by considering it to be using a reduction strategy which is the same as $\cbv$ except that in $\Theta_2\ (\underline f \underline n)$, it evaluates the redex at $@_2$ first.

### 2021-01-07 LICS21/applications.tex
Add brief discussion of application of \sim to inference

@@
 \label{sec:applications}
+Although the definition of potential positions in a term in \Cref{sec:confluent} was intended simply to define a variant of sampling semantics that has a restricted version of the Church-Rosser property, it is likely to also be applicable to probabilistic inference algorithms as well.
+
+In the Metropolis-Hastings inference algorithm, the program is executed, with a random value being chosen each time a $\tsample$ redex is evaluated.
+The trace of samples used is recorded, then for the next step, it is modified slightly before the program is run again.
+This time, when $\tsample$ redexes are evaluated, they may be taken from the pre-determined trace instead of chosen randomly.
+The change to the trace may be accepted or rejected, depending on the probability weight associated with each execution of the program (based on a conditioning language construct), and after this process is repeated sufficiently many times, the distribution of results of the program tends towards the true posterior distribution.
+
+It is essential to the efficiency of this algorithm that after modifying the trace, the weight of the resultant program execution is likely to be similar to the weight of the previous execution.
+This is satisfied by the values of the samples being similar to their original values, which (for a sufficiently well-behaved program) will tend to make the entire execution proceed similarly.
+In the simplest version of this algorithm, the samples in the trace are used one at a time, in the order they are encountered.
+However, if as a result of some random choice, more or fewer samples are taken during some stage of the program, subsequent samples in the trace will be displaced, and end up being used for different purposes than previously, thereby decreasing the correlation between the program runs.
+
+It is possible to reduce this problem by labelling the samples in the trace by something other than the order in which they are used.
+If the labels of samples in the trace correspond closely to the roles that the samples play in the program execution, this will increase the efficiency of the inference algorithm.
+In \ref{wingate2011lightweight}, for example, something rather like a stack trace is used.
+Potential positions could also be used for a similar purpose, although in practice, a more compact representation than an entire skeletal reduction sequence would be necessary.

### 2021-01-09 LICS21/applications.tex
Add an introduction, and more further directions and things.

@@
 If the labels of samples in the trace correspond closely to the roles that the samples play in the program execution, this will increase the efficiency of the inference algorithm.
-In \ref{wingate2011lightweight}, for example, something rather like a stack trace is used.
-Potential positions could also be used for a similar purpose, although in practice, a more compact representation than an entire skeletal reduction sequence would be necessary.
+In \citep{wingate2011lightweight}, for example, something rather like a stack trace is used, although this labelling scheme turns out to have various problems.\citep{kiselyov2016problems} Potential positions could also be used for a similar purpose, although in practice, a more compact representation than an entire skeletal reduction sequence would be necessary.
+Selecting an appropriate addressing scheme can also be important in the context of variational inference.\citep[\S 6.2.2]{paige2016automatic}
+
+The $\sim$ relation as it has been defined is not exactly what would be necessary for inference (for example, $\sim_c$ only becomes relevant when there are multiple different reduction strategies being considered), but something much like it is still likely to be useful.
+
+\paragraph{}
+Another possible application for the confluent trace semantics would be as a justification, within a sampling-style semantics, of some forms of equational reasoning.
+By introducing reduction steps that go backwards to terms not reachable from the original term, it may in some cases be simplified, or equivalences between terms may be proved, by appealing to confluence results such as \Cref{churchRosser} and \Cref{thm:CbvIsTerminatingest}.

### 2021-01-09 LICS21/intro.tex
Mainly LaTeX reformatting to fit 2-column style
Further comments on figure about reachable terms (geometric distribution example)
Add an introduction, and more further directions and things.

@@
+% !TEX root = main.tex
+
 \section{Introduction}
@@
 Various theorems about probabilistic programs rely on the assumption that the program terminates almost surely.
+One possible approach to prove almost sure termination is to find some variant on the program state that decreases on average sufficiently quickly that it must at some point reach 0, at which point the program terminates.
+In other words, the program's behavior is used to define an associated ranking supermartingale.
 Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \citep{DBLP:journals/pacmpl/McIverMKK18}.
 This paper's contribution is to extend this method to a higher-order setting.
+
+The language PPCF used in this paper is simply typed, with an explicit recursion primitive, $\tY$.
+As the type system already constrains the terms enough to force termination in the absence of the recursion construct, it is only the $\tY$-reduction steps that must be counted, making defining ranking functions somewhat easier.
+
+The notion of a sparse ranking function is introduced to make constructing ranking functions more convenient.
+Most of the individual execution steps of a typical program are trivial and easy to mentally skip over.
+Sparse ranking functions can be defined only for those points in the execution of a program which are semantically important, while all of the other intermediate steps can be ignored.
+This also makes the ranking function method of proving AST more compatible with syntactic sugar, because the intermediate reduction steps implicit in the simplified notation can be ignored.
+
+A ranking function (or sparse ranking function) provides a bound on the expected number of $\tY$-reduction steps before the program terminates, therefore ranking functions cannot be constructed for terms whose expected number of $\tY$-reduction steps is infinite, such as the simplest implementation of the 1D unbiased random walk.
+This restriction can be removed by generalising ranking functions to antitone ranking functions, which rather than having to decrease by a constant amount for each $\tY$-reduction step, may decrease by a variable amount, depending on the value of the ranking function.
+This method is capable of proving the almost sure termination of programs which terminate arbitrarily slowly.
+Our progress condition is equivalent to that given in \citep{DBLP:journals/pacmpl/McIverMKK18} (by applying a transformation to the ranking function), but needs to be different in order to be compatible with the presence of additional intermediate reduction steps where the ranking function is not required to decrease at all.
+
+Basic (non-random) lambda calculus has the very useful Church-Rosser property, which implies (among other things) that even if execution of a program starts in a different order, it will still reach the same normal form eventually (assuming it does reach a normal form).
+Probabilistic lambda calculus does not have this property, because random choices may be duplicated, and evaluating the same sub-term multiple times can yield different results.
+However, with a restricted set of reduction strategies, Church-Rosserness may be regained.
+A novel addressing scheme for the possible random choices in a program's execution is introduced, which ensures that the same random choices are taken at corresponding positions in alternative reduction sequences, so that the same eventual result can be reached.
+This is then used to prove yet another extension to the ranking function theorem, that ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and rankability in this sense still imples almost sure termination.
+This confluent trace semantics has other possible applications as well, for example in inference algorithms.
+
+\subsubsection*{Contributions}
+
+This is the first application of martingales to probabilistic lambda calculus, and the first version of sampling semantics that's capable of satisfying the (restricted) Church-Rosser property.

### 2021-01-09 LICS21/related.tex
Mainly LaTeX reformatting to fit 2-column style
Add an introduction, and more further directions and things.

@@
+% !TEX root = main.tex
 \section{Related work, conclusion and further directions}
+\label{sec:related}
+We have not yet been able to prove our conjecture that every term is antitone rankable if all of the terms reachable from it are AST, which would imply that \Cref{thm:antitone rankable implies termination} is almost complete.
+
+The definition \label{def:more general red} of redexes and which positions are acceptable to reduce at is sufficiently restrictive to guarantee Church-Rosserness, but is also a little more restrictive in some cases than is necessary for this purpose.
+For example, if the argument of a function is not yet a value, but its reduction to a value would be deterministic, or the function is affine, then applying the function before reducing its argument would not cause any problematic duplication of random samples.
+Similarly, if $\tsample$ occurs at position $\alpha;@_2;\beta;\lambda;\gamma$, but the function at $\alpha;@_1$ is affine, evaluatinf the $\tsample$ may in some cases also work fine.
+A more complete characterisation of which redexes can be reduced without breaking Church-Rosserness could be interesting.
+
+Although this is a broadly applicable method of proving almost sure termination of probabilistic functional programs, and \Cref{thm:antitone partial implies rankable,thm:confluent ranking} make it more convenient to use, some method of automating the construction and checking of (antitone) ranking functions, even partially, could make it considerably more practically useful, especially in cases where almost sure termination is merely a side-condition for some other theorem or algorithm to be applicable.

### 2021-01-09 LICS21/confluent-proofs.tex
Trim down non-appendix size.

@@
 % !TEX root = main.tex
+\label{apx:confluent}
 The rules defining $\sim_c$ can be more intuitively understood as diagrams.
@@
 If a reduction sequence includes one branch of one of these cases at some point in it, then the positions in the reduction sequence obtained by substituting in the other branch are all related by $\sim_c$ to the corresponding positions in the original reduction sequence.
-
-\lo{TODO. Remove the white space by repositioning the sim diagrams.}
 \include{sim-diagram}
@@
 \begin{proof}
-\akr{I'm not sure whether this proof is sufficiently detailed. It would be pretty easy to fill it in more, just very verbose.}
-
 For both $\red$ and $\Rightarrow_{\cbv}$, the probability measure on the sample space can be used to define a measure on the reduction sequences of each finite length.
@@
 For the inductive case, suppose that the distributions of reduction sequences of length $n$ are equal.
-\akr{TODO: mention the case that it has already terminated by step $n$, or is about to.
-The two versions of the semantics handle it a little differently.}
 
@@
 \end{proof}
+
+\Cref{ex:sum of powers}
+Recall that
+\begin{align*}
+M =&\ (\lambda n. P[n]) \lfloor \tsample^{-1/2} \rfloor \\
+P[n] =&\ (\lambda p. \Xi[p] n) (\lambda x. \Theta[x] n)) \\
+\Xi[p] =&\ \tY \lambda f n. \tif{n = 0}{\underline 0}{p n + f (n-1)} \\
+\Theta[x] =&\ \tY \lambda f n. \tif{n = 0}{\underline 1}{x \times f (n-1)}.
+\end{align*}
+
+Let $r$ be the reduction strategy which reduces $M$ in the order
+\begin{align*}
+&\ M \\
+\to^* &\ P[\underline n] \\
+\to^* &\ (\lambda p.\ \Xi[p]\ \underline n)\ (\lambda x. \underbrace{x \times \dots \times x \times}_n \underline 1) \\
+\to^{\phantom *} &\ \Xi[\lambda x. \underbrace{x \times \dots \times x \times}_n \underline 1]\ \underline n \\
+\to^{\phantom *} &\ \tif{\underline n = 0}{\underline 0}{\underline{n^n} + \Xi[\lambda x. x \times \dots \underline 1]\ \underline{n-1}} \\
+\to^* &\ \underline{n^n} + \underline{(n-1)^n} + \dots + \underline 0 \\
+\to^* &\ \underline{\sum_{k=0}^n k^n}
+\end{align*}
+(this is a little underspecified, but the details are not important). A suitable sparse ranking function for it is
+\begin{align*}
+M \mapsto & \frac{\pi^2}{3} \\
+P[\underline n] \mapsto & 2n+2 \\
+(\lambda p.\ \Xi[p]\ \underline n)\ (\lambda x. \underbrace{x \times \dots \times x \times}_{n-k} (\Theta[x]\ \underline k)) \mapsto & n + k + 2 \\
+\Xi[\lambda x. \underbrace{x \times \dots \times x \times}_n \underline 1]\ \underline n \mapsto & n + 1 \\
+\underline{n^n} + \dots + \underline{(n-k+1)^n} + \Xi[\lambda x. x \times \dots \underline 1]\ \underline k \mapsto & k + 1 \\
+\underline{\sum_{k=0}^n k^n} \mapsto & 0.
+\end{align*}
+
+The first step is justified because the probability of reaching $P[\underline n]$ for any specific $n \geq 0$ is $(n+1)^{-2} - (n+2)^{-2} = \frac{2n + 3}{(n+1)^2(n+2)^2}$, therefore the expected next value of the ranking function is $\sum_{n=0}^{\infty} \frac{(2n+2)(2n + 3)}{(n+1)^2(n+2)^2} = \frac{\pi^2}{3}$.
+The other steps are deterministic, and involve 1 or 0 $\tY$-reductions each.
+Given that all of the reduction steps after the first are deterministic, and the number of $\tY$ reduction steps is not that hard to count, defining the sparse ranking function's values only at $M$ and $P[\underline n]$ would also have been reasonable, although in a similar term which had more random samples throughout, that would not have been so simple.
+
+Providing a ranking function of some sort for this term in a similar level of detail would have been rather more complicated using only the standard reduction strategy.
+The recursion in $\Theta$ would have had to be evaluated separately for every time it was used, so there would have been more (and more complex) terms to consider in the reduction sequence.
+Also, the number of $\tY$-reductions from $P[\underline n]$ would have been $n^2 + 2n + 1$ instead of $2n + 2$, therefore the expected number of $\tY$ reductions starting from $M$ would be infinite, i.e.~$M$ is not $\tY$-PAST, therefore it is not even rankable.
+It would be possible to define an antitone sparse ranking function for it, but it would have to include terms such as
+\begin{align*}
+& \underline{n^n} + \dots + \underline{(n-k_1+1)^n} + \\
+& (\underbrace{\underline{n-k_1} \times \dots \times \underline{n-k_1} \times}_{n-k_2-1} (\Theta[\underline{n-k_1}] (\underline{k_2} - 1))) + \\
+& \Xi[\lambda x. \Theta[x] \underline n] (\underline{n-k_1} - 1) \mapsto \log((n+1)k_1 + k_2)
+\end{align*}
+as only one of the several cases, which is significantly more complicated and difficult to use.
+
+It wouldn't even be possible in this case to give a sparser version of the ranking function defined only at $M$ and $P[\underline n]$, because the amount that an antitone sparse ranking function must decrease at each $\tY$-reduction step depends on the value of the ranking function at the next term where it is defined, therefore it is necessary to have these intermediate steps in order for the ranking function to be able to change sufficiently slowly.

### 2021-01-09 LICS21/ranking-proofs.tex
Trim down non-appendix size.

--- 2021-01-04 LICS21/ranking.tex
@@
 % !TEX root = main.tex
-\section{Constructing ranking functions}
-\label{sec:ranking}
-
-Although rankability implies almost sure termination, the converse does not hold in general. For example,
-\begin{equation}
-\tif{{-}\tsample < 0}{\underline{0}}{(\tY \lambda x. x) \, \underline 0}
-\label{ex:0 probability reachable}
-\end{equation}
-terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x. x) \, \underline 0$ is reachable, although that has probability 0. 
-Not only is this counterexample AST, it's PAST. %positively almost surely terminating i.e.~the expected time to termination is finite.
-
-A ranking function can be constructed under the stronger assumptions that, for every $N$ reachable from $M$, the expected number of $\tY$-reduction steps from $N$ to a value is finite. 
-In particular, the expected number of $\tY$-reduction steps from each reachable term is a ranking function. 
-However, note that a finite number of expected $\tY$-reduction steps does not necessarily imply a finite number of expected total reduction steps.
-
-\begin{example}
-\label{ex:tY finite does not imply t finite}
-The term $M = \Xi \, (\lambda x. x+1)$ where
-\[
-\Xi := \tY \lambda f n. \tif{\tsample - 0.5 < 0}{n \, \underline 0}{f (\lambda x. n (n x))}
-\] 
-terminates with only 2 $\tY$-reductions on average, i.e.,~$\expect{T_M^\tY} = 2$, but applies the increment function $2^n$ times with probability $2^{-n-1}$ for $n \geq 0$, which diverges, i.e.,~$\expect{T_M} = \infty$.
-\end{example}
-
-\begin{theorem} \label{thm:minimal}
-Given a closed term $M$, the function $f:Rch(M) \to \mathbb R$ given by 
-\[
-f(N) := \mathbb E [\text{number of }\tY\text{-reduction steps from }N\text{ to a value}]
-\] 
-if it exists, is the least of all possible ranking functions of $M$.
-\end{theorem}
+\ThmMinimal*
 \begin{proof}
@@
 \end{proof}
-
-\subsection{Sparse ranking functions}
-Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost sure termination, so it isn't useful for proving it.
-
-\begin{example}[Geometric distribution]
-% Consider the term
-% \begin{align*}
-% &(\tY \lambda f, n. \\
-% &\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
-% &) \, \underline{0}
-% \end{align*}
-%\[
-%\big(\tY \lambda f, n. \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
-%\big) \, \underline{0}
-%\]
-Let
-\[
-\Theta := \lambda f, n. \tif{\tsample - 0.5 < 0}{n}{f (n+1)}. 
-\]
-The term  $(\tY \, \Theta) \, \underline 0$ generates a geometric distribution.
-\iffalse
-\lo{\Cref{fig:geometric distribution} is confusing: the arrow means either one-step reduction, or its reflexive transitive closure, or ``has the same syntactic shape''.}
-
-\akr{The arrows all mean the one-step reduction $\to$, I just didn't intend for the values of $i$ and $r$ to be consistent between the terms.
-I would much rather have the diagram be complete (with all the reductions and all the terms).}
-
-\lo{Perhaps you have in mind an informal, schematic interpretation, reading a term $C[i, r]$ as ``there exists a term $C[i, r]$, for appropriate $i, r \in \mathbb{R}$''.
-One way to make this precise is to consider the graph obtained by quotienting the one-step reduction relation by an appropriate equivalence relation over terms (that specifies this scheme reading).}
-
-\lo{\Cref{fig:geometric distribution 1} is less ambitious, but it is not confusing in that arrow does mean one-step reduction (also note $r'$), and it would be correct to say (cf.~the preceding blue-highlighted phrase): ``Despite its simplicity, $\mathit{Rch}(\tY \, \Theta \, \underline 0)$ contains all the terms in \Cref{fig:geometric distribution 1}, for each $i \geq 0$''.}
-
-\lo{The graph can be simplified somewhat if we redefine general recursion using a fixpoint operator $\tY_{f, n}\Theta$ with redex rule as per \Cref{eq:letrec redex}: see \Cref{fig:geometric distribution 2}.
-But (especially in view of the vast investment in the confluent trace semantics) it is perhaps too late to change things now?}
-\fi
-Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.\lo{TODO: reduce the figure to avoid overflow.}\akr{It still overflows the gutter, but not the actual text of the other column.}\lo{OK.
-Thanks.} The meanings of the substitutions (to be applied to the contractum) and conditions that label some of the reductions in the diagram should be self-explanatory.
 
@@
 
-\iffalse
-\begin{figure}
-\centering
-\begin{tikzpicture}[node distance=0.5em, anchor=west]
-    \node (1) at (0,-0) {$\tY\, \Theta\, \underline i$ \quad [provided $i = 0$]};
-    \node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, \underline i$};
-    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})})\, \underline i$};
-    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (5) at (0,-4) {$\tif{\underline r\, \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (6) at (0,-5) {$\tif{\underline r' < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (7) at (2,-6) [text width=] {$\underline i$};
-    \node (8) at (0,-7) {$\tY\, \Theta\, (\underline i\, \underline{+ 1})$};
-    \node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, (\underline i\, \underline{+ 1})$};
-    \node (10) at (0,-9) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, \underline{i + 1}$};
-
-    \draw [->] (1.south west) to [bend right=45] (2.north west);
-    \draw [->] (2.south west) to [bend right=45] (3.north west);
-    \draw [->] (3.south west) to [bend right=45] (4.north west);
-    \draw [->] (4.south west) to [bend right=45] (5.north west);
-    \draw [->] (5.south west) to [bend right=45] (6.north west);
-    \draw [->] (6.south west) to [bend right=45] (8.north west);
-    \draw [->] (8.south west) to [bend right=45] (9.north west);
-    \draw [->] (9.south west) to [bend right=45] (10.north west);
-    \draw [->] (6) to (7);
-\end{tikzpicture}
-\caption{Some reachable terms of $\tY \, \Theta \, \underline 0$\label{fig:geometric distribution 1}}
-\end{figure}
-
-\begin{figure}
-\centering
-\begin{tikzpicture}[node distance=0.5em, anchor=west]
-    \node (1) at (0,-0) {$\tY_{f, n}\Theta\, \underline i$};
-    %\node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY_{f, n}\Theta\, (n \underline{+ 1})}) z)\, \underline i$};
-    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY_{f, n}\Theta\, (n \underline{+ 1})})\, \underline i$};
-    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY_{f, n}\Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (5) at (0,-4) {$\tif{\underline r\, \underline{- 0.5} < 0}{\underline i}{\tY_{f, n}\Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (6) at (0,-5) {$\tif{\underline r' < 0}{\underline i}{\tY_{f, n}\Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (7) at (2,-6) [text width=] {$\underline i$};
-    \node (8) at (0,-7) {$\tY_{f, n}\Theta\, (\underline i\, \underline{+ 1})$};
-    \node (9) at (0,-8) {$\tY_{f, n}\Theta\, \underline{i + 1}$};
-    %\node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY_{f, n}\Theta\, (n \underline{+ 1})}) z)\, (\underline i\, \underline{+ 1})$};
-    %\node (10) at (0,-9) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY_{f, n}\Theta\, (n \underline{+ 1})}) z)\, \underline{i + 1}$};
-
-    \draw [->] (1.south west) to [bend right=45] (3.north west);
-    %\draw [->] (2.south west) to [bend right=45] (3.north west);
-    \draw [->] (3.south west) to [bend right=45] (4.north west);
-    \draw [->] (4.south west) to [bend right=45] (5.north west);
-    \draw [->] (5.south west) to [bend right=45] (6.north west);
-    \draw [->] (6.south west) to [bend right=45] (8.north west);
-    \draw [->] (8.south west) to [bend right=45] (9.north west);
-    %\draw [->] (9.south west) to [bend right=45] (10.north west);
-    \draw [->] (6) to (7);
-\end{tikzpicture}
-\caption{Some reachable terms of $\tY_{f, n}\Theta \, \underline 0$\label{fig:geometric distribution 2}}
-\end{figure}
-\fi
-\end{example}
-
-Even in this simple case, defining a ranking function explicitly is awkward because of the number of cases, although in most cases, because the value need only be greater than or equal to that of the next term in sequence, it suffices to take the ranking function as having the same value as the next term, so that overall it takes only 3 distinct values.
-(We will explain why later.)
-
-The definition of rankability is also inconvenient for syntactic sugar.
-It could be useful, for example, to define $M \oplus_p N = \tif{\tsample - p < 0} M N$, where $M \oplus_p N$ reduces to $M$ or $N$, depending on the first value of $s \in \entrosp$, with probability $p$ resp.~$(1-p)$.
-Technically though, it reduces first to $\tif{\underline r - p < 0} M N$ for all $r \in I$, so those terms all need values of the ranking function too.
-
-%\paragraph{}
-In both of these cases, there are only some values of the ranking function that are semantically important. 
-\begin{definition}
-Define a \emph{sparse ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \rightharpoonup \mathbb R$ such that
-\begin{itemize}
-    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
-    \item $f$ is defined at $M$.
-    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually \akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and ast, but that doesn't actually work because of terms reachable with 0 probability.
-Of course, astness doesn't imply rankability, so any definition of sparse ranking function that includes all Y-past terms wouldn't always extend to a total ranking function.} reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
-\end{itemize}
-\end{definition}
-A sparse ranking function that is total is just a ranking function.
-Providing a sparse ranking function is essentially part way between providing a ranking function and directly proving almost sure termination.
-
-\begin{theorem} \label{thm:partial implies rankable}
-Every sparse ranking function is a restriction of a ranking function.
-\end{theorem}
+\PartialImpliesRankable*
 \begin{proof}
@@
 Define $f_2(N) = \int_\entrosp f_1(\nnext(N,s)) + g(N,s) \, \mu_{\entrosp}(\mathrm d s)$.
-The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction).
-\lo{TODO. The preceding sentence deserves a proof.}
+The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction, by the same argument as \Cref{thm:minimal}).
 \end{proof}
-
-As a corollary, any term which admits a sparse ranking function terminates almost surely.
-
-\subsection{Examples}
-Let $M \oplus_p N = \tif{\tsample - p < 0} M N$, for $p \in (0,1]$, then there are the pseudo-reduction relations
-\begin{align*}
-E[M \oplus_p N] \to^3 & E[M] & \\
-E[M \oplus_p N] \to^3 & E[N] & \\
-\red^3(E[M \oplus_p N], s) = & \left\{
-    \begin{array}{ll}
-        (E[M],\pi_t(s)) & \text{if } \pi_h(s) < p \\
-        (E[N],\pi_t(s)) & \text{if } \pi_h(s) \geq p. \\
-    \end{array} \right .
-\end{align*}
-A sparse ranking function could be defined with respect to this shortcut reduction simply by replacing $\to$ and $\red$ in the definition of a sparse ranking function by a version that goes straight from $N \oplus_p O$ to $N$ or $O$.
-Such a pseudo-sparse ranking function would then be a partial function from a subset of $Rch(M)$, so it could also be considered as a partial function from all of $Rch(M)$, and it would in fact also be an actual sparse ranking function.
-It is therefore possible to prove rankability directly using the shortcutted reductions.
-
-A similar procedure would work for other forms of syntactic sugar.
-If a closed term $N$ eventually reduces to one of a set of other terms $\{N_i \mid i \in I\}$ with certain probabilities, a sparse ranking function defined with respect to a reduction sequence that skips straight from $N$ to $N_i$ is also a valid sparse ranking function for the original reduction function, and therefore its existence implies almost sure termination.
-There is a caveat, however, that $\tY$-reduction steps skipped over in the shortcut still need to be counted for the expected number of $\tY$-reduction steps.
-
-%\paragraph{}
-With this abbreviation, the geometric distribution example from earlier can be written as $(\tY \lambda f n.\ n \oplus_{0.5} f(n+1)) \, \underline 0$.
-It is then easy to see that the following is a sparse ranking function:
-\begin{align*}
-(\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\ N & \mapsto 2 \\
-\underline i \oplus_{0.5} (\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\ (\underline i + 1) & \mapsto 1 \\
-\underline i & \mapsto 0.
-\end{align*}
-
-In fact, even the partial function $\tY \Theta N \mapsto 2$ alone is a sparse ranking function for this term.

### 2021-01-09 LICS21/ranking.tex
Trim down non-appendix size.

@@
 
-\begin{theorem} \label{thm:minimal}
+\begin{restatable}{theorem}{ThmMinimal} \label{thm:minimal}
 Given a closed term $M$, the function $f:Rch(M) \to \mathbb R$ given by 
@@
 if it exists, is the least of all possible ranking functions of $M$.
-\end{theorem}
-\begin{proof}
-Let $f$ be the candidate least ranking function defined above, and suppose $g$ is another ranking function such that $f(N) > g(N)$ for some $N \in Rch(M)$.
-The restrictions of $f$ and $g$ to $Rch(N)$ have the same properties assumed of $f$ and $g$, so assume w.l.o.g.~that $N=M$.
-The difference $g - f$ is then a supermartingale (with the same setup as in the proof of %\Cref{thm:rankable gives supermartingale}
-%Theorem \ref{rankable implies ast}) 
-\Cref{thm:rankable and strict rankable});
-therefore $\mathbb E[g(M_n)] \leq \mathbb E[f(M_n)] + g(M)-f(M)$, for all $n$.
-Now $\mathbb E[f(M_n)] = \sum_{k=n}^\infty \mathbb P[M_k = E[\tY N] \text{ for some }E, N] \to 0 \text{ as } n \to \infty$; 
-therefore as $g(M) - f(M) < 0$, eventually $\mathbb E[g(M_n)] < 0$, which is impossible. 
-It follows that $g \geq f$ as required.
-
-In order for $f$ to be the least ranking function of $M$, it also has to actually be a ranking function itself.
-Each of the conditions on a ranking function is easily verified from the definition of $f$.
-\end{proof}
+\end{restatable}
 
@@
 The term  $(\tY \, \Theta) \, \underline 0$ generates a geometric distribution.
-\iffalse
-\lo{\Cref{fig:geometric distribution} is confusing: the arrow means either one-step reduction, or its reflexive transitive closure, or ``has the same syntactic shape''.}
-
-\akr{The arrows all mean the one-step reduction $\to$, I just didn't intend for the values of $i$ and $r$ to be consistent between the terms.
-I would much rather have the diagram be complete (with all the reductions and all the terms).}
-
-\lo{Perhaps you have in mind an informal, schematic interpretation, reading a term $C[i, r]$ as ``there exists a term $C[i, r]$, for appropriate $i, r \in \mathbb{R}$''.
-One way to make this precise is to consider the graph obtained by quotienting the one-step reduction relation by an appropriate equivalence relation over terms (that specifies this scheme reading).}
-
-\lo{\Cref{fig:geometric distribution 1} is less ambitious, but it is not confusing in that arrow does mean one-step reduction (also note $r'$), and it would be correct to say (cf.~the preceding blue-highlighted phrase): ``Despite its simplicity, $\mathit{Rch}(\tY \, \Theta \, \underline 0)$ contains all the terms in \Cref{fig:geometric distribution 1}, for each $i \geq 0$''.}
-
-\lo{The graph can be simplified somewhat if we redefine general recursion using a fixpoint operator $\tY_{f, n}\Theta$ with redex rule as per \Cref{eq:letrec redex}: see \Cref{fig:geometric distribution 2}.
-But (especially in view of the vast investment in the confluent trace semantics) it is perhaps too late to change things now?}
-\fi
-Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.
+Despite its simplicity, its $\mathit{Rch}$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.
 %\lo{TODO: reduce the figure to avoid overflow.}\akr{It still overflows the gutter, but not the actual text of the other column.}\lo{OK. Thanks.} 
@@
 
-\begin{figure}
-\begin{tikzpicture}[node distance=0.5em, anchor=west]
-    \node (1) at (0,-0) {$\tY\, \Theta\, \underline 0$};
-    \node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, \underline i$};
-    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})})\, \underline i$};
-    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (5) at (0,-4) {$\tif{\underline r\, \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (6) at (0,-5) {$\tif{\underline {r - 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (7) at (2,-6.3) [text width=] {$\underline i$};
-    \node (8) at (0,-7) {$\tY\, \Theta\, (\underline i\, \underline{+ 1})$};
-    \node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, (\underline i\, \underline{+ 1})$};
-
-    \draw [->] (1.south west) to [bend right=45, "$i := 0$"'] (2.north west);
-    \draw [->] (2.south west) to [bend right=45] (3.north west);
-    \draw [->] (3.south west) to [bend right=45] (4.north west);
-    \draw [->] (4.south west) to [bend right=45] (5.north west);
-    \draw [->] (5.south west) to [bend right=45] (6.north west);
-    \draw [->] (6.south west) to [bend right=45, "$r \geq 0.5$"] (8.north west);
-    \draw [->] (8.south west) to [bend right=45] (9.north west);
-
-    \draw [->] (9.west) to [bend left=30, "$i := i + 1$" {below,sloped}] (2.west);
-    \draw [->] (6) to ["$r < 0.5$"] (7);
-\end{tikzpicture}
-\caption{The reachable terms of $\tY \, \Theta \, \underline 0$\label{fig:geometric distribution}}
-\end{figure}
-
-\iffalse
-\begin{figure}
-\centering
-\begin{tikzpicture}[node distance=0.5em, anchor=west]
-    \node (1) at (0,-0) {$\tY\, \Theta\, \underline i$ \quad [provided $i = 0$]};
-    \node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, \underline i$};
-    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})})\, \underline i$};
-    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (5) at (0,-4) {$\tif{\underline r\, \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (6) at (0,-5) {$\tif{\underline r' < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (7) at (2,-6) [text width=] {$\underline i$};
-    \node (8) at (0,-7) {$\tY\, \Theta\, (\underline i\, \underline{+ 1})$};
-    \node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, (\underline i\, \underline{+ 1})$};
-    \node (10) at (0,-9) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, \underline{i + 1}$};
-
-    \draw [->] (1.south west) to [bend right=45] (2.north west);
-    \draw [->] (2.south west) to [bend right=45] (3.north west);
-    \draw [->] (3.south west) to [bend right=45] (4.north west);
-    \draw [->] (4.south west) to [bend right=45] (5.north west);
-    \draw [->] (5.south west) to [bend right=45] (6.north west);
-    \draw [->] (6.south west) to [bend right=45] (8.north west);
-    \draw [->] (8.south west) to [bend right=45] (9.north west);
-    \draw [->] (9.south west) to [bend right=45] (10.north west);
-    \draw [->] (6) to (7);
-\end{tikzpicture}
-\caption{Some reachable terms of $\tY \, \Theta \, \underline 0$\label{fig:geometric distribution 1}}
-\end{figure}
-
-\begin{figure}
-\centering
-\begin{tikzpicture}[node distance=0.5em, anchor=west]
-    \node (1) at (0,-0) {$\tY_{f, n}\Theta\, \underline i$};
-    %\node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY_{f, n}\Theta\, (n \underline{+ 1})}) z)\, \underline i$};
-    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY_{f, n}\Theta\, (n \underline{+ 1})})\, \underline i$};
-    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY_{f, n}\Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (5) at (0,-4) {$\tif{\underline r\, \underline{- 0.5} < 0}{\underline i}{\tY_{f, n}\Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (6) at (0,-5) {$\tif{\underline r' < 0}{\underline i}{\tY_{f, n}\Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (7) at (2,-6) [text width=] {$\underline i$};
-    \node (8) at (0,-7) {$\tY_{f, n}\Theta\, (\underline i\, \underline{+ 1})$};
-    \node (9) at (0,-8) {$\tY_{f, n}\Theta\, \underline{i + 1}$};
-    %\node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY_{f, n}\Theta\, (n \underline{+ 1})}) z)\, (\underline i\, \underline{+ 1})$};
-    %\node (10) at (0,-9) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY_{f, n}\Theta\, (n \underline{+ 1})}) z)\, \underline{i + 1}$};
-
-    \draw [->] (1.south west) to [bend right=45] (3.north west);
-    %\draw [->] (2.south west) to [bend right=45] (3.north west);
-    \draw [->] (3.south west) to [bend right=45] (4.north west);
-    \draw [->] (4.south west) to [bend right=45] (5.north west);
-    \draw [->] (5.south west) to [bend right=45] (6.north west);
-    \draw [->] (6.south west) to [bend right=45] (8.north west);
-    \draw [->] (8.south west) to [bend right=45] (9.north west);
-    %\draw [->] (9.south west) to [bend right=45] (10.north west);
-    \draw [->] (6) to (7);
-\end{tikzpicture}
-\caption{Some reachable terms of $\tY_{f, n}\Theta \, \underline 0$\label{fig:geometric distribution 2}}
-\end{figure}
-\fi
 \end{example}
@@
     \item $f$ is defined at $M$.
-    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually \akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and AST, but that doesn't actually work because of terms reachable with 0 probability.
-Of course, ASTness doesn't imply rankability, so any definition of sparse ranking function that includes all Y-PAST terms wouldn't always extend to a total ranking function.} reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
+    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually 
+    %\akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and AST, but that doesn't actually work because of terms reachable with 0 probability.
+Of course, ASTness doesn't imply rankability, so any definition of sparse ranking function that includes all Y-PAST terms wouldn't always extend to a total ranking function.}
+     reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
 \end{itemize}
@@
 
-\begin{theorem} \label{thm:partial implies rankable}
+\begin{restatable}{theorem}{PartialImpliesRankable} \label{thm:partial implies rankable}
 Every sparse ranking function is a restriction of a ranking function.
-\end{theorem}
-\begin{proof}
-Take a closed term $M$ and sparse ranking function $f$ on $M$. Define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
-\begin{align*}
-    f_1(N) & := f(N) \text{ whenever $f(N)$ is defined,} \\
-    f_1(V) & := 0 \text{ for values $V$ not in the domain of $f$.}
-\end{align*}
-
-Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) := \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
-The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on sparse ranking functions.
-Define $f_2(N) = \int_\entrosp f_1(\nnext(N,s)) + g(N,s) \, \mu_{\entrosp}(\mathrm d s)$.
-The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction).
-\lo{TODO. The preceding sentence deserves a proof.}
-\end{proof}
+\end{restatable}
 

### 2021-01-09 LICS21/supermartingales.tex
Trim down non-appendix size.

@@
 It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \geq 0}$ is the reduction sequence starting from $M$, then $(f(M_n))_{n \geq 0}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales.
-\lo{It is confusing (strictly speaking, incorrect) to say that $(M_n)_{n \geq 0}$ is a reduction sequence from $M$. Formally $M_n$ is the random variable $M_n: s \mapsto \pi_0 (\red^n(M, s))$.}
-\akr{I guess it's implicitly identifying a sequence of functions ($(M_n)_{n \geq 0}$), and a function producing sequences (a reduction-sequence-valued random variable), but this sort of thing seems very standard in dealing with random variables (e.g.~applying a function to some random variables, which is really a sort of concatenation).}
+%\lo{It is confusing (strictly speaking, incorrect) to say that $(M_n)_{n \geq 0}$ is a reduction sequence from $M$. Formally $M_n$ is the random variable $M_n: s \mapsto \pi_0 (\red^n(M, s))$.}
+%\akr{I guess it's implicitly identifying a sequence of functions ($(M_n)_{n \geq 0}$), and a function producing sequences (a reduction-sequence-valued random variable), but this sort of thing seems very standard in dealing with random variables (e.g.~applying a function to some random variables, which is really a sort of concatenation).}
 
@@
 Our first result is the following theorem.
-\begin{theorem}[Rankability] 
+\begin{restatable}{theorem}{rankableAndStrictRankable}[Deriving supermartingales] 
 \label{thm:rankable and strict rankable}
@@
 %where $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
-\end{theorem}
-\lo{@Andrew: A better name for this theorem: ``Ranking function'', ``Supermartingale'', ``(Ranking) supermartingale''?}
+\end{restatable}
+%\lo{@Andrew: A better name for this theorem: ``Ranking function'', ``Supermartingale'', ``(Ranking) supermartingale''?}
 
 %\subsection{Proof of \Cref{thm:rankable and strict rankable}.}
-We shall obtain the theorem as a corollary of a technical lemma (\Cref{lem:key rankable}).
-
-\iffalse
-\lo{We need to show that each $f(M_n)$ is integrable. LO claimed earlier that $0 \leq f(N) \leq f(M)$ for all $N \in \mathit{Rch}(M)$.
-This is of false: take $f$ with $v \, \tsample \mapsto 1$ and $v \, \underline{r} \mapsto 2\, r$ for some value $v$. 
-However, it is true that
-\[
-\int_{S} \mu(\dif s) \, |f(M_n)(s)| 
-=
-\int_{S} \mu(\dif s) \, f(M_n)(s)
-\leq
-\int_{S} \mu(\dif s) \, f(M_0)(s)
-=
-f(M)
-\]
-the inequality above follows from \Cref{lem:key rankable}.
-}
-\fi
-
-We say that a given PPCF term is: %\emph{type-1} (respectively 2, 3 and 4) if it has the shape $E[\tY \lambda x. N]$ (respectively $E[\tsample]$, $E[R]$ where $R$ is any other redex, and of a value).
-\emph{type 1} if it has the shape $E[\tY \lambda x.
-N]$, \emph{type 2} if it has the shape $E[\tsample]$; \emph{type 3} if it has the shape $E[R]$ where $R$ is any other redex, and \emph{type 4} if it is a value.
-Henceforth fix an $n \geq 0$, and define $\mathbf{T}_i := \set{s \mid M_n(s) \hbox{ is type $i$}}$.
-It is straightforward to see that each $\mathbf{T}_i \in \calF_n$, and $\set{\mathbf{T}_1, \mathbf{T}_2, \mathbf{T}_3, \mathbf{T}_4}$ is a partition of $\entrosp$.
-%By abuse of language, we refer to the three respective types of redexes as type-$1$, 2 and 3.
-Hence it suffices to prove the following lemma.
-
-\iffalse
-For $i \in \{1, 2, 3\}$ and $n \geq 0$, define function $f_{i, n+1}(M) : \entrosp \to \Real$ by
-\[
-f_{i, n+1}(M)(s) :=
-\begin{cases}
-f(M_{n+1}(s)) & \hbox{if }s \in \mathbf{T}_i := 
-\set{s \mid M_n(s) \textrm{ is type-$i$}}\\
-0 & \textrm{otherwise}
-\end{cases}
-\]
-\begin{lemma}
-\label{lem:inde}
-For each $i$ and $n$, $f_{i, n+1}(M)$ is %not just $\mathcal{F}_{n+1}$-measurable but also 
-$\mathcal{F}_{n}$-measurable.
-\end{lemma}
-
-\begin{proof} First fix notation
-\begin{align*}
-\sigma(M_{n+1}) &= \sigma\big(\set{M_{n+1}^{-1}(U_{\beta_k})
-\mid \beta_k\in \mathsf{Sk}_k, U_{\beta_k} \in \mathcal{B}(\Real^k), k \geq 0}\big)\\
-\sigma(M_{n}) &= \sigma\big(\set{M_{n}^{-1}(U_{\alpha_j})
-\mid \alpha_j\in \mathsf{Sk}_j, U_{\alpha_j} \in \mathcal{B}(\Real^j), j \geq 0}\big);
-\end{align*}
-and for $\beta_k \in \mathsf{Sk}_i$, we write $\beta_k[\dagger]$ to mean the instantiation of $\beta_k$ by a $k$-vector of numerals $\dagger$; $\alpha_j[\dagger']$ has the same meaning.
-Now for each $\beta_k[\dagger] = E[R']$ where $R'$ is the contractum of a type-$i$ redex $R$, there exist $\alpha_j \in \mathsf{Sk}_j$ and $j$-vector $\dagger'$ such that $\alpha_j[\dagger'] = E[R]$ of type $i$.
-Moreover, for each $U_{\beta_k}$, there exists $U_{\alpha_j}$ such that $M_{n+1}^{-1}(U_{\beta_k}) = M_{n}^{-1}(U_{\alpha_j}) \in \sigma(M_n)$.
-\end{proof}
-
-Plainly
-\(
-f(M_{n+1}) = %\sum_{i=1}^3 f_{i, n+1}(M).
-f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M),
-\)
-$\mu$-almost surely.
-%(Unless otherwise stated, all equations and inequations between random variables are assumed to hold only $\mu$-a.s.)
-Therefore 
-\[
-\expect{f(M_{n+1}) \mid \mathcal{F}_n} \nonumber \\
-= 
-\expect{f_{1, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{2, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{3, n+1}(M) \mid \mathcal{F}_n} 
-\]
-\begin{align}
-& \expect{f(M_{n+1}) \mid \mathcal{F}_n} \nonumber \\
-& = \expect{f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M)  \mid \mathcal{F}_n} 
-\nonumber \\
-& = \expect{f_{1, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{2, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{3, n+1}(M) \mid \mathcal{F}_n} 
-\label{eqn:linear cond ex} 
-& = f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M)
-\label{eqn:inde cond exp}
-& = f_{1, n+1}(M) + \int_I f() + f_{3, n+1}(M) \label{eqn:inde cond exp} \\
-\end{align}
-\Cref{eqn:linear cond ex} follows from the linearity of conditional expectation. 
-%\Cref{eqn:inde cond exp} is justified because $f_{i, n+1}(M)$ is $\mathcal{F}_n$-measurable (\Cref{lem:inde}).
-
-It remains to show: for all $A \in \calF_n$
-\begin{align}
-& \int_A \dif \mu \, \expect{f(M_{n+1}) \mid \mathcal{F}_n} \nonumber \\
-& \leq  \int_A \mu (\dif s) \, \big(f(M_{n})[s \in \mathbf{T}_1] + f(M_{n})[s \in \mathbf{T}_2] + f(M_{n})[s \in \mathbf{T}_3]\big)
-\label{eqn:ts} \\
-& = \int_A \dif \mu \, f(M_n). 
-\label{eqn:partition} 
-\end{align}
-where $[s \in \mathbf{T}_i]$ is the Iverson bracket.
-
-Here we use the notation: given function $g : \entrosp \to \Real$ and $U \subseteq \entrosp$, $g[U] : \entrosp \to \Real$ denotes the function
-\[
-g[U](x) :=
-\begin{cases}
-g(x) & \textrm{$x \in U$}\\
-0 & \textrm{otherwise}
-\end{cases}
-\]
-
-The inequality (\ref{eqn:ts}) follows immediately from the lemma below; 
-and (\ref{eqn:partition}) holds because $\{\mathbf{T}_1, \mathbf{T}_2, \mathbf{T}_3, \mathbf{T}_4\}$ is a partition of $\entrosp$.
-\fi
-
-\begin{restatable}[Technical]{lemma}{TechnicalRankingSupermartingale}
-\label{lem:key rankable}
-For all $i \in \set{1, 2, 3, 4}$ and $A \in \calF_n$
-\[
-\int_A \mu_{\entrosp}(\dif s) \, f(M_{n+1})[s \in \mathbf{T}_i] \leq \int_A \mu_{\entrosp}(\dif s) \, f(M_n)[s \in \mathbf{T}_i]
-\; \footnote{where (the Iverson bracket) $[P] := 1$ if the statement $P$ hold, and 0 otherwise.}
-\] 
-%$\int_A \mu_{\entrosp}(\dif s) \, f(M_{n+1}) \leq \int_A \mu_{\entrosp}(\dif s) \, f(M_n)$, i.e., 
-Hence $\mathbb{E}[f(M_{n+1}) \mid \calF_n] \leq f(M_n)$ a.s.
-\end{restatable}
-As an immediate corollary of \Cref{lem:key rankable}, each $f(M_n)$ is integrable
-\iffalse
-\begin{align*}
-\int_{\entrosp} \mu_{\entrosp}(\dif s) \, |f(M_n)(s)| 
-&=
-\int_{\entrosp} \mu_{\entrosp}(\dif s) \, f(M_n)(s)
-\\
-&\leq
-\int_{\entrosp} \mu_{\entrosp}(\dif s) \, f(M_0)(s)
-=
-f(M).
-\end{align*}
-
-[** To see $f_{2, n+1}(M) \leq f(M_n)[T_2]$, take $s \in T_2$. Then $f_{2, n+1}(M)(s) = f(E[\underline{a}])$ and $M_n = E[\tsample]$, for some $a \in [0, 1]$ and evaluation context $E$. Hence 
-\[
-f_{2, n+1}(M)(s) \leq \int_I f(E[\underline{r}]) \, \mu_{leb}(\textrm{d} r)
-\leq f(E[\tsample]) = f(M_n)[T_2](s).
-\]
-**]
-\fi
-This concludes the proof of \Cref{thm:rankable and strict rankable}.
+The ranking supermartingale condition is satisfied essentially by a case analysis on the type of redex (\Cref{lem:key rankable}).
 
@@
 & \qquad\qquad\qquad\qquad\qquad \hbox{or of form } E[\tY \lambda x. N] \} \nonumber \\
-X_n(s) & := f(M_n(s)) \nonumber \\
-Y_n(s) & := X_{T_n}(s) \nonumber \\
+Y_n(s) & := f(M_{T_n}(s)) \nonumber \\
 T_M^\tY(s) &:= \min \set{n \mid M_{T_n(s)}(s) \textrm{ is a value}} \label{eq:y stopping time} 
-%T_M(s) & := \min \set{n \mid X_n(s) = 0} \nonumber
 \end{align}
 
-%The random variable $T_0(s)$ is the least $k \geq 0$ such that $M_k(s)$ is either a value or a $\tY$-redex.
-
-%$T_1(s)$ is the least $k \geq 0$ such that $M_k(s)$ is either a value or the second occurrence of a $\tY$-redex.
-
-%Thus the random variable $T_n(s)$ is the least $k \geq 0$ such that $M_k(s)$ is either a value or the $(n+1)$-occurrence of a $\tY$-redex (in the reduction sequence $M_0(s), M_1(s), \ldots$).
-%Notice that each $T_n$ is a stopping time w.r.t.~filtration $(\calF_n)_{n \geq 0}$.
 We first state a useful property about r.v.~$T_0, T_1, T_2, \cdots$.

### 2021-01-09 LICS21/antitone-proofs.tex

--- 2020-12-28 relativised-completeness.tex
@@
-% !TEX root = ./probabilisticProgrammingMartingales.tex
+% !TEX root = main.tex
+\label{apx:antitone}
+\AntitonePartial*
+\begin{proof}
+  Take a closed term $M$ and an antitone sparse ranking function $f$ on $M$, with a corresponding antitone function $\epsilon$.
+Assume wlog that $\epsilon$ is convex (as if it isn't, we can just take its convex hull instead).
+As in Theorem \ref{thm:partial implies rankable}, define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
+  \begin{align*}
+    f_1(N) &= f(N) \text{ whenever $f(N)$ is defined},\\
+    f_1(V) &= 0 \text{ for values $V$ not in the domain of $f$.}
+  \end{align*}
+
+  Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) := \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
+  The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on antitone partial ranking functions.
+Define $f_2(N) := \int_\entrosp f_1(\nnext(N,s)) + \epsilon(f_1(\nnext(N,s))) g(N,s) \, \mu_{\entrosp}(\mathrm d s)$.
+For any term $N$ where $f$ is defined, $f_2(N) = f(N)$, and the value that $f_2$ would have at $N$ if $f$ were not defined at $N$ is $\leq f(N)$, by the third condition on antitone partial ranking functions.
+In order to show that $f_2$ is an antitone partial ranking function, it therefore suffices to show that the value that $f_2$ would have had at each term if $f$ were not defined at that term is at least the expectation of $f_2$ after one reduction step (plus $\epsilon(f_2(N))$ if the reduction step is a $tY$-reduction).
+For any term $N$ which is not of the form $E[R]$ for some $\tY$-redex $R$, this is trivial.
+If $R$ is a $\tY$-redex, then $\epsilon(f_2(N)) \leq \int_\entrosp \epsilon(f_1(\nnext(N,s)))\, \mu_{\entrosp}(\mathrm d s)$ by the convexity of $\epsilon$, because $n$ is bounded.
+Therefore, the (total) function $f_2$, which agrees with $f$ on $f$'s domain, is an antitone ranking function on $M$ (with the same function $\epsilon$ if it's convex).
+\end{proof}
+
+\Cref{ex:raven complex}
+Let $\Theta := \tY \, \lambda f x . \tif{x \leq 0}{0}{f(x - \tsample)}$. 
+We can construct a sparse ranking function $f$ for $\Theta \, 10$ as follows:
+\begin{align*}
+\Theta \, \underline l 
+&\mapsto 
+l + 2
+\\
+\tif{l \leq 0}{0}{\Theta \, (l - \tsample)}
+&\mapsto
+l + 1
+\\
+0 &\mapsto 0.
+\end{align*}
+
+For a more complex (not $\tY$-PAST) example, consider the following ``continuous random walk'': $\Xi \, 10$ where
+\[
+%\underbrace{\big(
+\Xi := \tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample + 1/2)} 
+%\big)}_{\Xi} \, 10
+\]
+Let $g(x) := 2 + \ln(x + 1)$, and let $\epsilon$ be the function specified by
+\[
+\epsilon(g(x-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y.
+\]
+The limit of this as $x \to \infty$ and $g(x+1/2) \to \infty$ is 0, and $\frac {\mathrm d}{\mathrm dx} \epsilon(g(x+1/2)) = \frac 1 {x+1} + \ln(1 - \frac 1 {x + 1/2}) < 0$, and $g$ is monotonic increasing;
+therefore $\epsilon$ is antitone and bounded below by 0.
+We define an antitone sparse ranking function by:
+\[
+\Xi \, l 
+\mapsto 
+g(l), \qquad 
+0 \mapsto 0
+\]
+
+% \begin{align*}
+% \Xi \, l 
+% &\mapsto 
+% g(l)
+% \\
+% 0 &\mapsto 0
+% \end{align*}
+The value of $g$ after one $\tY$-reduction step is at least $g(l-1/2)$, therefore the expectation of $\epsilon$ after one $\tY$-reduction step is at most $\epsilon(g(l-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y$.
+Thus 
+\begin{itemize}
+\item in case $l > 0$, $g$ decreases by the required amount
+\item in case $l \leq 0$, 
+\(
+g(\Xi \, l) \geq 2 + \ln(1/2) > \ln(\frac{3 \sqrt 3} 2) - 1 = \epsilon(g(0-1/2))
+\) 
+as well.
+\end{itemize}
+Hence this is a valid antitone sparse ranking function and the term is AST.
+
+\Cref{ex:Fair-in-the-limit random walk}
+\[
+\underbrace{\big
+(\tY \, f \, x . 
+\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)}_{\Xi} 
+\, 10
+\]
+To construct an antitone ranking function, we solve the recurrence relation:
+% \[
+% \left\{
+% \begin{array}{rll}
+% z_0 &=& 0\\
+% n \geq 1, \quad z_n &>& \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
+% \end{array}
+% \right.
+% \]
+\begin{align*}
+z_0 &= 0\\
+n \geq 1, \quad z_n &> \textstyle \frac{n}{2n + 1} \, z_{n-1} + \frac{n+1}{2n + 1} \, z_{n+1}.
+\end{align*}
+For $n > 2$, $z_n = \ln(n-1)$ works.
+The expected decrease is $\frac 1 2(\ln(1+\frac 1 {(n-1)^2-1}) - \frac 1 {2n+1} \ln(1 + \frac 2 {n-2}))$, and using the fact that $\frac x {x+1} \leq \ln (1+x) \leq x$ for $x > 0$, this is at least $\frac 1 2(\frac 1 {(n-1)^2} - \frac 1 {2n + 1} \frac 2 {n-2}) = \frac {n^2}{2(n-1)^2(2n+1)(n-2)}$, which (again for $n > 2$) is positive and antitone.
+For $n = 0, 1, 2$, we take $\epsilon$ to be 9/40 (the same as its value at 3), then set $z_2 = 2 \ln 2 - \ln 3 - 1, z_1 = 3 \ln 2 - 2 \ln 3 - 3, z_0 = 4 \ln 2 - 3 \ln 3 - 6$.
+Some of those values are negative, but it's still bounded below, so by adding a constant offset it can be corrected, and the term is AST.
+
+\Cref{ex:escaping spline}
+\[
+\underbrace{\big
+(\tY \, f \, x . \,
+0 \oplus_{\frac{1}{x+1}} f(x + 1) \big)}_{\Xi} 
+\, 10
+\]
+
+In this case, the fact that the ranking function must decrease at each $\tY$-step (in an antitone sparse ranking function) by the expected value of $\epsilon$ not at the current term, but at the next term where the ranking function is defined, is a little harder to deal with, because the variable $x$ can change all the way to $0$ in one step, therefore simply adding a small offset doesn't suffice to compensate for this fact.
+
+Consider the candidate ranking function $\Xi \, n \mapsto n + 1$.
+For each $n$, $\Xi \, n$ reduces to either $0$ (with probability $\frac 1 {n + 1}$) or $\Xi \, (n + 1)$, therefore the expected value of the ranking function is $\frac{n(n+2)}{n+1} = n + 1 - \frac 1 {(n+1)^2}$, and the required decrease is $\frac{\epsilon(0) + n \epsilon(n + 2)}{n + 1}$, therefore eventually, the expected decrease isn't enough, whatever the value of $\epsilon(0)$ is.
+
+If instead, we take the ranking function to be defined after the $\tY$-reduction but before the $\tsample$-reduction as well, this can be resolved.
+Letting $\Theta[n] = \underline 0 \oplus_{\frac 1 n} \Xi \, n$:
+\begin{align*}
+\Theta[n] &\mapsto n \\
+\Xi \, 10 & \mapsto 12.
+\end{align*}
+For each $n$, $\Theta[n]$ reduces to either $0$ or $\Theta[n+1]$, with a $\tY$-reduction only in the latter case, and the condition that this is an antitone sparse ranking function is that $n \geq \frac{(n-1)(n+1)}{n} + \frac{n-1}{n} \epsilon(n+1)$ and $12 \geq 11 + \epsilon(11)$.
+These are satisfied by setting $\epsilon(x) = \min(1, 1/x)$, therefore this term is AST.
+
+\medskip
+
+Many of the recent advances in the development of AST verification methods \citep{DBLP:conf/pldi/ChenH20,DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18,DBLP:conf/aplas/HuangFC18,DBLP:conf/popl/ChatterjeeNZ17,DBLP:journals/pacmpl/AgrawalC018,DBLP:conf/cav/ChatterjeeFG16,DBLP:conf/lics/OlmedoKKM16,DBLP:journals/pacmpl/Huang0CG19} are concerned with loop-based programs.
+We can view such loops as tail-recursive programs that are, in particular, \emph{affine recursive}, i.e., 
+in each evaluation (or run) of the body of the recursion, recursive calls are made from at most one call site \citep[\S 4.1]{DBLP:journals/toplas/LagoG19}.
+(Note that whether a program is affine recursive cannot be checked by just counting textual occurrences of variables.)
+Termination analysis of \emph{non-affine recursive} probabilistic programs does not seem to have received much attention.
+Methods such as those presented in \citep{DBLP:journals/toplas/LagoG19} are explicitly restricted to affine programs, and are unsound otherwise.
+By contrast, many probabilistic programming languages allow for richer recursive structures \citep{DBLP:conf/pkdd/TolpinMW15,DBLP:conf/uai/GoodmanMRBT08,DBLP:journals/corr/MansinghkaSP14}.
+
+\medskip
+
+\Cref{ex:non-affine recursion}
+Let 
+\[
+\Xi := \tY \lambda f x. x \oplus_{2/3} f (f (x + 1))
+\] 
+A suitable sparse ranking function for $\Xi \, \underline 1$ would be $\Xi^n \, \underline i \mapsto 3 \, n$ (for $n \geq 0$), because $\Xi^n \, \underline i$ reduces to either $\Xi^{n+1} \, \underline{i+1}$ or $\Xi^{n-1} \, \underline{i+1}$, with probabilities $1/3$ and $2/3$.
+The value of $i$ does not actually matter for the progress of this recursion.
+It is basically another variant of the biased random walk, except that the relevant variable is the number of copies of $\Xi$, instead of a real number in the term.
+
+There is an obvious source of deterministic (i.e.~non-probabilistic) higher-order functions defined by recursion, viz., \emph{higher-order recursion schemes} (HORS) (see e.g.~\citep{DBLP:conf/lics/Ong06,DBLP:conf/lics/Ong15}).
+(Incidentally the (sparse) ranking function method is just as applicable to the termination analysis of \emph{deterministic} PPCF programs.)
+Recently \cite{DBLP:conf/lics/KobayashiLG19} have extended HORS to \emph{probabilistic higher-order recursion schemes} (PHORS), which are HORS augmented with probabilistic (binary) branching $\oplus_p$.
+As HORS are in essence the $\lambda \tY$-calculus \citep{DBLP:conf/lics/Statman02} (i.e.~pure simply-typed lambda calculus with recursion, generated from a finite base type), PHORS are definable in PPCF:
+(order-$n$) PHORS is encodable as (order-$n$) (call-by-name) PPCF, but the former is strictly less expressive (because the underlying recursion schemes are not Turing complete). 
+%Some interesting PPCF terms such as \Cref{ex:non-affine with continuous distribution} %\refExample{complexExample} 
+%cannot be expressed as PHORS.
+%\akr{As I removed the referenced example, and I don't even know what a PHORS is, I've just removed this comment too.}
+A relevant result here is that the AST problem is decidable for order-1 PHORS (by reduction to the PSPACE-hard solvability of finite systems of polynomial equations with real coefficients; see \citep{DBLP:journals/jacm/EtessamiY09}), but undecidable for order-2 PHORS.
+
+\Cref{ex:higher-order recursion}
+Consider the higher-order function
+$\Xi : (\textsf{R} \to \textsf{R} \to \textsf{R}) \to \textsf{R} \to \textsf{R} \to \textsf{R}$
+recursively defined by
+\[
+\begin{array}{l}
+\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} . \\
+\quad\tif{n \leq 0}{s}{
+f \, n \, (\varphi \, f \, s \, (n - 1))
+\oplus_p
+f \, n \, (\varphi \, f \, s \, (n + 1))
+}
+\end{array}
+\]
+Let $F : \textsf{R} \to \textsf{R} \to \textsf{R}$ be a function such that $F\ \underline n\ \underline m$ terminates with no $\tY$ reductions for all $m$ and $n$.
+Then $\Xi\ F\ \underline x\ \underline y$ has the antitone sparse ranking function $F \underline{n_0} (F \underline{n_1} (\dots(\Xi\ F\ \underline x\ \underline n)\dots) \mapsto g_1(n)$, where $g_1$ is as in \Cref{ex:unbiased random walk}, using the same antitone function too.
+(Again, this is not actually the correct reduction order, in that $F$ should be applied to its argument $n_i$ and reduced to a value before its second argument is expanded, but this would complicate the presentation, and this version can be justified by \Cref{thm:confluent ranking} instead.)
+
+The inequalities required for this to be an antitone partial supermartingale are satisfied with just the same reasoning as in \Cref{ex:unbiased random walk}, therefore this term too is antitone rankable.
+
+\antitoneSeemsComplete*
 
@@
 \item For $x \geq 2^k$, $\epsilon(x) \leq \frac{2^k}{t_{k+1}-t_k}$. 
-\lo{This would imply $\frac{2^k}{t_{k+1}-t_k} \geq \epsilon(2^{k-1})$. Why?} \akr{Corrected: it should have been min, not max.}
 \end{itemize}
@@
     & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \sum_{m=t_k}^{n-1} \frac{2^k}{t_{k+1}-t_k}) \\
+\end{align*}
+%Allow a page break so that the sim-diagram lines up neatly.
+\begin{align*}
     & = \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \frac{2^k (n - t_k)}{t_{k+1}-t_k}) \\
@@
 \begin{align*}
-    & \expect{Y_{n+1} \mid \mathcal F_n} \\
-    = & \expect{y_{n+1} \, 1_{\set{T \geq n+1}} \mid \mathcal F_n} \\
-    = & y_{n+1} \, \expect{1_{\set{T \geq n+1}} \mid \mathcal F_n} \\ 
-    = & y_{n+1} \, 1_{\set{T \geq n}} \frac{\mathbb P[T > n+1]}{\mathbb P[T > n]} \\
-    = & (y_n - \epsilon(y_n)) \, 1_{\set{T \geq n}} \\
-    = & Y_n - \epsilon(y_n) \, 1_{\set{T \geq n}}
+    \expect{Y_{n+1} \mid \mathcal F_n} 
+    &=  \expect{y_{n+1} \, {\bf 1}_{\set{T \geq n+1}} \mid \mathcal F_n} \\
+    &=  y_{n+1} \, \expect{{\bf 1}_{\set{T \geq n+1}} \mid \mathcal F_n} \\ 
+    &=  y_{n+1} \, {\bf 1}_{\set{T \geq n}} \textstyle \frac{\mathbb P[T > n+1]}{\mathbb P[T > n]} \\
+    &=  (y_n - \epsilon(y_n)) \, {\bf 1}_{\set{T \geq n}} \\
+    &= Y_n - \epsilon(y_n) \, {\bf 1}_{\set{T \geq n}}
 \end{align*}

### 2021-01-09 LICS21/antitone.tex

@@
 \begin{definition}
-\changed[akr]{The definition of \emph{antitone ranking function} $f$ for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
+The definition of \emph{antitone ranking function} $f$ for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
 we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) function $\epsilon : \nnReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
 \[f(E[R']) \leq f(E[R]) - \epsilon(f(E[R])) \] 
-where $R \to R'$ is the $\tY$-redex rule.}
+where $R \to R'$ is the $\tY$-redex rule.
 Any closed term for which an antitone ranking function exists is called \emph{antitone rankable}.
@@
 
-\changed[lo]{Note that antitone ranking functions are actually a generalisation of ranking functions, even though the way we reference them may suggest that they are a type of ranking functions.}
+Note that antitone ranking functions are actually a generalisation of ranking functions, even though the way we reference them may suggest that they are a type of ranking functions.
 
@@
 \dfrac{\expect{Y_n} - \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
-  \step[\leq]{$Y_n \geq 0$ always}
+  \step[\leq]{}
 \dfrac{\expect{Y_n}}{\mathbb P[T > n]}
-  \step[\leq]{$(Y_n)_n$ is a supermartingale}
+  \step[\leq]{}
 \dfrac{\expect{Y_0}}{\mathbb P[T > n]}
@@
 \expect{Y_n-Y_{n+1}}
-  \step[=]{$\calF_n \subseteq \calF_{n+1}$, def.~\& linearity of cond.~expectation}
+  \step[=]{}
 \expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
-  \step[\geq]{antitone strict assumption}
+  \step[\geq]{}
 \expect{\epsilon(Y_n) \cdot {\bf 1}_{\set{T > n}}}
 %  \step[=]{definition of conditional expectation for a discrete variable}
-\step[=]{def.~expectation conditioning on an \emph{event}}
+\step[=]{}
 \mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
@@
 \begin{example}
-\iffalse
-\akr{Consider a random variable $X$, uniformly distributed in the interval $[-1,1]$, $T := \min \set{n \mid 2^{-n} < X}$, 
-\[
-Y_k := 
-\begin{cases}
-4-2^{-k} & \hbox{if $k < T$}\\
-0 & \hbox{otherwise}
-\end{cases} 
-\]
-} \lo{The following reformulation defines $T$ and $Y_k$ directly as a measurable function on a probability space (rather than via a r.v.~$X$).}
-\fi
 Take the probability space $(\Omega, \calF, \mathbb{P})$ where $\Omega$ is the closed interval of reals $[-1, 1]$, $\calF$ the Borel $\sigma$-algebra, and $\mathbb{P}$ the corresponding Lebesgue probability measure.
@@
 \end{align*}
-\lo{N.B.~$\set{T = n} = (\frac{1}{2^{n}}, \frac{1}{2^{n-1}}]$, and so, $\set{T > n} = [-1, \frac{1}{2^n}]$.}
 Plainly, $T$ is a stopping time, and $(Y_n)_{n \geq 0}$ is a supermartingale, adapted to the filtration $(\calF_n)_{n \geq 0}$ where $\calF_n = \calF$ for all $n$. 
@@
 
-\begin{theorem}  \label{thm:antitone partial implies rankable}
+\begin{restatable}{theorem}{AntitonePartial}  \label{thm:antitone partial implies rankable}
   Every antitone sparse ranking function is a restriction of an antitone ranking function.
-\end{theorem}
-\begin{proof}
-  Take a closed term $M$ and an antitone sparse ranking function $f$ on $M$, with a corresponding antitone function $\epsilon$.
-Assume wlog that $\epsilon$ is convex (as if it isn't, we can just take its convex hull instead).
-As in Theorem \ref{thm:partial implies rankable}, define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
-  \begin{align*}
-    f_1(N) &= f(N) \text{ whenever $f(N)$ is defined},\\
-    f_1(V) &= 0 \text{ for values $V$ not in the domain of $f$.}
-  \end{align*}
-
-  Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) := \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
-  The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on antitone partial ranking functions.
-Define $f_2(N) := \int_\entrosp f_1(\nnext(N,s)) + \epsilon(f_1(\nnext(N,s))) g(N,s) \, \mu_{\entrosp}(\mathrm d s)$.
-For any term $N$ where $f$ is defined, $f_2(N) = f(N)$, and the value that $f_2$ would have at $N$ if $f$ were not defined at $N$ is $\leq f(N)$, by the third condition on antitone partial ranking functions.
-In order to show that $f_2$ is an antitone partial ranking function, it therefore suffices to show that the value that $f_2$ would have had at each term if $f$ were not defined at that term is at least the expectation of $f_2$ after one reduction step (plus $\epsilon(f_2(N))$ if the reduction step is a $tY$-reduction).
-For any term $N$ which is not of the form $E[R]$ for some $\tY$-redex $R$, this is trivial.
-If $R$ is a $\tY$-redex, then $\epsilon(f_2(N)) \leq \int_\entrosp \epsilon(f_1(\nnext(N,s)))\, \mu_{\entrosp}(\mathrm d s)$ by the convexity of $\epsilon$, because $n$ is bounded.
-Therefore, the (total) function $f_2$, which agrees with $f$ on $f$'s domain, is an antitone ranking function on $M$ (with the same function $\epsilon$ if it's convex).
-\end{proof}
+\end{restatable}
 
@@
 Programs $M_1$ and $M_2$ are AST. For $M_1$, the sparse ranking function 
-\[
+\begin{align*}
 \big(\tY \, \lambda f n . \, \tif{n=0}{0}{f \, (n-1) \oplus_{3/2} f \, (n+1)}\big) \, \underline x \mapsto 3x + 1
-\] 
+\end{align*}
 suffices to prove its termination (and could equivalently be considered an antitone sparse ranking function with the constant antitone function $\epsilon_1(x) = 1$).
@@
 \label{ex:unbiased random walk}
-For $M_2$, define the function $g_1 : \nnReal \to \nnReal$ by $g_1(x) = \log(x+1) + 1$.
+For $M_2$, define the function $g_1 : \nnReal \to \nnReal$ by $g_1(x) = \ln(x+1) + 1$.
 Using shorthand 
@@
 & g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 \\
-  &=  \log \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) \\
-  &=  \frac 1 2 \log\left(\frac{(n+1)^2}{n(n+2)}\right) \\
-  &=  \frac 1 2 \log\left(1 + \frac 1 {n(n+2)}\right) \\
-  &>  \frac 1 {4n(n+2)} \\
-  &>  \frac 1 2 \frac 1 {5(n+1)^2} + \frac 1 2 \frac 1 {5(n+3)^2}
+  &=  \ln \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) \\
+  &=  \frac 1 2 \ln\left(\frac{(n+1)^2}{n(n+2)}\right) \\
+  &=  \frac 1 2 \ln\left(1 + \frac 1 {n(n+2)}\right) \\
+  &>  \frac 1 {n(n+2) + 1} \\
+  &>  \frac 1 2 \frac 1 {(n+1)^2} + \frac 1 2 \frac 1 {(n+3)^2}
 \end{align*}
-\changed[lo]{(The first inequality follows from the fact that for all $x \in (0, 1)$, $\log_2 (1 + x) > x$.)}
 Moreover $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
-Therefore by setting $\epsilon_2(x) = \frac 1 {5(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met.
+Therefore by setting $\epsilon_2(x) = \frac 1 {(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met.
 It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone sparse ranking function, and $M_2$ is AST.
 \end{example}
+
+All of the following examples also have antitone ranking functions, provided in Appendix \ref{apx:antitone}.
+
+Sampling from continuous distributions is an essential feature of statistical probabilistic programming languages.
+(See e.g.~Church \citep{DBLP:conf/uai/GoodmanMRBT08}, Stan \citep{carpenter2017stan}, Anglican \citep{DBLP:conf/pkdd/TolpinMW15}, Gen \citep{cusumano-towner2019Gen}, Pyro \citep{bingham2019Pyro}, Edward \citep{tran2016edward} and Turing \citep{ge2018Turing}.)
+Methods of proving AST of probabilistic computation have been developed for probabilistic programs with discrete distributions (see e.g.~\citep{DBLP:journals/toplas/LagoG19,DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/lics/KobayashiLG19,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05}).
+To our knowledge, the problem of proving AST of probabilistic programs with continuous distribution is new.
 
@@
 \]
-We can construct a sparse ranking function $f$ as follows:
-\begin{align*}
-\Theta \, \underline l 
-&\mapsto 
-l + 2
-\\
-\tif{l \leq 0}{0}{\Theta \, (l - \tsample)}
-&\mapsto
-l + 1
-\\
-0 &\mapsto 0.
-\end{align*}
 
@@
 \]
-Let $g(x) := 2 + \log(x + 1)$, and let $\epsilon$ be the function specified by
-\[
-\epsilon(g(x-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y.
-\]
-The limit of this as $x \to \infty$ and $g(x+1/2) \to \infty$ is 0, and $\frac {\mathrm d}{\mathrm dx} \epsilon(g(x+1/2)) = \frac 1 {x+1} + \log(1 - \frac 1 {x + 1/2}) < 0$, and $g$ is monotonic increasing;
-therefore $\epsilon$ is antitone and bounded below by 0.
-We define an antitone sparse ranking function by:
-\[
-\Xi \, l 
-\mapsto 
-g(l), \qquad 
-0 \mapsto 0
-\]
-
-% \begin{align*}
-% \Xi \, l 
-% &\mapsto 
-% g(l)
-% \\
-% 0 &\mapsto 0
-% \end{align*}
-The value of $g$ after one $\tY$-reduction step is at least $g(l-1/2)$, therefore the expectation of $\epsilon$ after one $\tY$-reduction step is at most $\epsilon(g(l-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y$.
-Thus 
-\begin{itemize}
-\item in case $l > 0$, $g$ decreases by the required amount
-\item in case $l \leq 0$, 
-\(
-g(\Xi \, l) \geq 2 + \log(1/2) > \log(\frac{3 \sqrt 3} 2) - 1 = \epsilon(g(0-1/2))
-\) 
-as well.
-\end{itemize}
-Hence this is a valid antitone sparse ranking function and the term is AST.
 \end{example}
@@
 \[
-\underbrace{\big
+\big
 (\tY \, f \, x . 
-\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)}_{\Xi} 
+\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)
 \, 10
 \]
-To construct an antitone ranking function, we solve the recurrence relation:
-% \[
-% \left\{
-% \begin{array}{rll}
-% z_0 &=& 0\\
-% n \geq 1, \quad z_n &>& \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
-% \end{array}
-% \right.
-% \]
-\begin{align*}
-z_0 &= 0\\
-n \geq 1, \quad z_n &> \textstyle \frac{n}{2n + 1} \, z_{n-1} + \frac{n+1}{2n + 1} \, z_{n+1}.
-\end{align*}
-For $n > 2$, $z_n = \log(n-1)$ works.
-The expected decrease is $\frac 1 2(\log(1+\frac 1 {(n-1)^2-1}) - \frac 1 {2n+1} \log(1 + \frac 2 {n-2}))$, and using the fact that $\frac x {x+1} \leq \log (1+x) \leq x$ for $x > 0$,\lo{The preceding inequality is incorrect: for $x \in (0, 1)$, $\log (1 + x) > x$.}
-\lo{OK.
-You mean $\ln$ rather than $\log_2$.} this is at least $\frac 1 2(\frac 1 {(n-1)^2} - \frac 1 {2n + 1} \frac 2 {n-2}) = \frac {n^2}{2(n-1)^2(2n+1)(n-2)}$, which (again for $n > 2$) is positive and antitone.
-For $n = 0, 1, 2$, we take $\epsilon$ to be 9/40 (the same as its value at 3), then set $z_2 = 2 \log 2 - \log 3 - 1, z_1 = 3 \log 2 - 2 \log 3 - 3, z_0 = 4 \log 2 - 3 \log 3 - 6$.
-Some of those values are negative, but it's still bounded below, so by adding a constant offset it can be corrected, and the term is AST.
-
 \end{example}
@@
 \]
+\end{example}
 
-In this case, the fact that the ranking function must decrease at each $\tY$-step (in an antitone sparse ranking function) by the expected value of $\epsilon$ not at the current term, but at the next term where the ranking function is defined, is a little harder to deal with, because the variable $x$ can change all the way to $0$ in one step, therefore simply adding a small offset doesn't suffice to compensate for this fact.
+\begin{example}[Non-affine recursion] \label{ex:non-affine recursion}
+\[
+(\tY \lambda f x. x \oplus_{2/3} f (f (x + 1)))\ \underline 1
+\]
+\end{example}
 
-Consider the candidate ranking function $\Xi \, n \mapsto n + 1$.
-For each $n$, $\Xi \, n$ reduces to either $0$ (with probability $\frac 1 {n + 1}$) or $\Xi \, (n + 1)$, therefore the expected value of the ranking function is $\frac{n(n+2)}{n+1} = n + 1 - \frac 1 {(n+1)^2}$, and the required decrease is $\frac{\epsilon(0) + n \epsilon(n + 2)}{n + 1}$, therefore eventually, the expected decrease isn't enough, whatever the value of $\epsilon(0)$ is.
-
-If instead, we take the ranking function to be defined after the $\tY$-reduction but before the $\tsample$-reduction as well, this can be resolved.
-Letting $\Theta[n] = \underline 0 \oplus_{\frac 1 n} \Xi \, n$:
-\begin{align*}
-\Theta[n] &\mapsto n \\
-\Xi \, 10 & \mapsto 12.
-\end{align*}
-For each $n$, $\Theta[n]$ reduces to either $0$ or $\Theta[n+1]$, with a $\tY$-reduction only in the latter case, and the condition that this is an antitone sparse ranking function is that $n \geq \frac{(n-1)(n+1)}{n} + \frac{n-1}{n} \epsilon(n+1)$ and $12 \geq 11 + \epsilon(11)$.
-These are satisfied by setting $\epsilon(x) = \min(1, 1/x)$, therefore this term is AST.
+\begin{example}[Higher-order recursion] \label{ex:higher-order recursion}
+Consider the higher-order function
+$\Xi : (\textsf{R} \to \textsf{R} \to \textsf{R}) \to \textsf{R} \to \textsf{R} \to \textsf{R}$
+recursively defined\footnote{Inspired by examples in \citep{DBLP:journals/pacmpl/BurnOR18,DBLP:conf/lics/OngW19}.} by
+\[
+\begin{array}{l}
+\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} . \\
+\quad\tif{n \leq 0}{s}{
+f \, n \, (\varphi \, f \, s \, (n - 1))
+\oplus_p
+f \, n \, (\varphi \, f \, s \, (n + 1))
+}
+\end{array}
+\]
+For any $F : \textsf{R} \to \textsf{R} \to \textsf{R}$ such that $F\ \underline n\ \underline m$ terminates with no $\tY$ reductions for all $m$ and $n$, $\Xi\ F\ \underline x\ \underline y$ is antitone rankable.
 \end{example}
@@
 \end{restatable}
-
-\medskip
-
-\lo{
-Motivation: Non-affine recursion. This could be placed in \Cref{sec:intro}.
-}
-Many of the recent advances in the development of AST verification methods \citep{DBLP:conf/pldi/ChenH20,DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18,DBLP:conf/aplas/HuangFC18,DBLP:conf/popl/ChatterjeeNZ17,DBLP:journals/pacmpl/AgrawalC018,DBLP:conf/cav/ChatterjeeFG16,DBLP:conf/lics/OlmedoKKM16,DBLP:journals/pacmpl/Huang0CG19} are concerned with loop-based programs.
-We can view such loops as tail-recursive programs that are, in particular, \emph{affine recursive}, i.e., 
-in each evaluation (or run) of the body of the recursion, recursive calls are made from at most one call site \citep[\S 4.1]{DBLP:journals/toplas/LagoG19}.
-(Note that whether a program is affine recursive cannot be checked by just counting textual occurrences of variables.)
-Termination analysis of \emph{non-affine recursive} probabilistic programs does not seem to have received much attention.
-Methods such as those presented in \citep{DBLP:journals/toplas/LagoG19} are explicitly restricted to affine programs, and are unsound otherwise.
-By contrast, many probabilistic programming languages allow for richer recursive structures \citep{DBLP:conf/pkdd/TolpinMW15,DBLP:conf/uai/GoodmanMRBT08,DBLP:journals/corr/MansinghkaSP14}.
-
-\medskip
-
-\begin{example}[Non-affine recursion]
-\changed[akr]{Let 
-\[
-\Xi := \tY \lambda f x. x \oplus_{2/3} f (f (x + 1))
-\] 
-A sparse ranking function for $\Xi \, 1$ would be $\Xi^n \, i \mapsto 3 \, n$ (for $n \geq 0$), but w.r.t.~the reduction strategy that first reduces the argument of an application to a value, then the function, then the application itself.
-
-For other values of $p > 1/2$, $n/(2p-1)$ works, and for $p = 1/2$, something much like the antitone ranking function for \Cref{ex:unbiased random walk} would work.
-
-A standard sparse ranking function (i.e.~one that doesn't depend on the confluent semantics) would have a slightly more verbose definition, as the functions that accumulate would not be $\Xi$, but what $\Xi$ reduces to after one step.}
-
-\lo{This is a genuine (and simple) \emph{non-affine recursive program}, i.e., recursive programs that can, during the evaluation of the recursive body, make multiple recursive
-calls (of a first-order function) from distinct call sites. 
-We can use this example to motivate the confluent trace semantics, though it is perhaps not the most compelling possible.
-Reason: one could use a more restrictive but equally expressive recursion \emph{operator} (as opposed to a $\tY$-\emph{combinator}, which is a constant) to implement general recursion which would make it unnecessary to consider an alternative reduction strategy when applying the sparse ranking function theorem.
-E.g.~define letrec-terms, $\mathsf{letrec} \, f^{A \to B} \, x^A . M$, as values, via the formation rule
-\[{\Gamma, f: A \to B, x:A  \vdash M : B} \over
-\Gamma \vdash \mathsf{letrec} \, f^{A \to B} \, x^A . M : A \to B
-\]
-with the redex rule 
-\begin{equation}
-(\mathsf{letrec} \, f^{A \to B} \, x^A . M^B) \, V \to M[\mathsf{letrec} \, f \, x . M / f, V / x].
-\label{eq:letrec redex}
-\end{equation}
-}
-\end{example}
-
-\lo{
-DRAFT. Motivation: Continuous distributions. This could be placed in \Cref{sec:intro}.
-}
-Sampling from continuous distributions is an essential feature of statistical probabilistic programming languages.
-(See e.g.~Church \citep{DBLP:conf/uai/GoodmanMRBT08}, Stan \citep{carpenter2017stan}, Anglican \citep{DBLP:conf/pkdd/TolpinMW15}, Gen \citep{cusumano-towner2019Gen}, Pyro \citep{bingham2019Pyro}, Edward \citep{tran2016edward} and Turing \citep{ge2018Turing}.)
-Methods of proving AST of probabilistic computation have been developed for probabilistic programs with discrete distributions (see e.g.~\citep{DBLP:journals/toplas/LagoG19,DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/lics/KobayashiLG19,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05}).
-To our knowledge, the problem of proving AST of probabilistic programs with continuous distribution is new.
-
-\lo{The following examples are for our discussion at 5:30. Details yet to be checked.}
-
-\begin{example}[Non-affine recursion with continuous distribution]
-\label{ex:non-affine with continuous distribution}
-Consider 
-\[
-\Xi := \tY \lambda f x . \big(\mathsf{let} \; e = \tsample \; \mathsf{in} \;
-\mathsf{if}\boldsymbol{(}e \leq p, x, X\boldsymbol{)}\big)
-\]
-where
-\(
-X = \big( f^3(x+1) \oplus_e f^2(x+1) \big) \oplus_{\mathit{sig}(x)} f^2(x+1) 
-\)
-and $\mathit{sig}(x)$ is the sigmoid function.
-
-Not only is this program non-affine recursive,
-note also the use of a random sample, $e$, as a first-class value, and the subsequent use as a probability in the binary choice $\oplus_e$.
-Such a computation cannot be modelled via discrete distributions.
-
-We can use the ranking function method (coupled with the solution of linear recurrence relations) to show that provided $p \geq \sqrt{7} - 2$, the program $\Xi \, \underline r$ is AST for all $r \in \Real$.
-
-\lo{This example is taken from the series of 3d-printing-company running examples (Beutner-Ong 2021).
-This paper is as yet unpublished, and effectively uncitable for the time being (because of strict double-blind submission rules).}
-\end{example}
-
-\lo{I believe the ranking function method can be used to prove soundness of the Beutner-Ong counting-based method (which reduces the AST problem of first-order (possibly non-affine) recursive PPCF programs to random walk on $\mathbb{N}$), based on Theorem 5.9 and Lemma 5.10 in \emph{op.~cit.}}
-
-\iffalse
-\[\begin{array}{l}
-\mathsf{let} \; \mathit{add} \; x \; y = x + y \\
-\mathsf{letrec} \; \mathit{iter} \; f \; s \; n = 
-\tif{n \leq 0}{s}{f \; n \; (\mathit{iter} \; f \; s \; (n - 1)}
-\end{array}
-\]
-\fi
-
-\lo{
-DRAFT. Motivation: higher-order recursion. This could be placed in \Cref{sec:intro} or \Cref{sec:related}.
-}
-There is an obvious source of deterministic (i.e.~non-probabilistic) higher-order functions defined by recursion, viz., \emph{higher-order recursion schemes} (HORS) (see e.g.~\citep{DBLP:conf/lics/Ong06,DBLP:conf/lics/Ong15}).
-(Incidentally the (sparse) ranking function method is just as applicable to the termination analysis of \emph{deterministic} PPCF programs.)
-Recently \cite{DBLP:conf/lics/KobayashiLG19} have extended HORS to \emph{probabilistic higher-order recursion schemes} (PHORS), which are HORS augmented with probabilistic (binary) branching $\oplus_p$.
-As HORS are in essence the $\lambda \tY$-calculus \citep{DBLP:conf/lics/Statman02} (i.e.~pure simply-typed lambda calculus with recursion, generated from a finite base type), PHORS are definable in PPCF:
-(order-$n$) PHORS is encodable as (order-$n$) (call-by-name) PPCF, but the former is strictly less expressive (because the underlying recursion schemes are not Turing complete). 
-Some interesting PPCF terms such as \Cref{ex:non-affine with continuous distribution} %\refExample{complexExample} 
-cannot be expressed as PHORS.
-A relevant result here is that the AST problem is decidable for order-1 PHORS (by reduction to the PSPACE-hard solvability of finite systems of polynomial equations with real coefficients; see \citep{DBLP:journals/jacm/EtessamiY09}), but undecidable for order-2 PHORS.
-
-\begin{example}[Higher-order recursion]
-Consider the higher-order function
-$\Xi : (\textsf{R} \to \textsf{R} \to \textsf{R}) \to \textsf{R} \to \textsf{R} \to \textsf{R}$
-recursively defined\footnote{Inspired by examples in \citep{DBLP:journals/pacmpl/BurnOR18,DBLP:conf/lics/OngW19}.} by
-\[
-\begin{array}{l}
-\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} . \\
-\quad\tif{n \leq 0}{s}{
-f \, n \, (\varphi \, f \, s \, (n - 1))
-\oplus_p
-f \, (n+1) \, (\varphi \, f \, s \, n)
-}
-\end{array}
-\]
-\lo{QUESTION: Let $g : \textsf{R} \to \textsf{R} \to \textsf{R}$ be a function that terminates on all arguments.
-Then $\Xi \, g$ is AST for $p \geq 1/2$.}
-\akr{In order to actually construct a ranking function, it will be necessary to assume a bound on the number of $\tY$-reduction steps in the reduction of $g \, \underline n \, \underline m$ too.}
-
-\akr{I suspect that the term you actually had in mind for the higher-order recursion example was different from what you wrote, because you said that it terminates for $p \geq 1/2$, but it actually terminates for all $p > 0$.
-Passing $n+1$ as an argument to $f$ doesn't actually make the recursion any longer.
-Did you mean
-\[
-\begin{array}{l}
-\tY \lambda \varphi f s n. \\
-\qquad \tif{n \leq 0}{s}{ f \, n \, (\varphi \, f \, s \, (n - 1)) \oplus_p f \, n \, (\varphi \, f \, s \, (n+1))}
-\end{array}
-\]}
-\end{example}

### 2021-01-09 LICS21/confluent.tex

--- 2021-01-07 LICS21/confluent.tex
@@
 To be more explicit, the modified forms of the theorems are:
-\begin{definition}\rm
-A \emph{ranking function on $M$ with respect to a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
-\begin{itemize}
-    \item $f(N) \geq 1+ f(N')$ if $N[r(N)]$ is a $\tY$-redex and $N$ reduces to $N'$ at $r(N)$
-    \item $f(N) \geq \int_I f(N[\underline{x}/r(N)]) \, \Leb(\mathrm{d}x)$ if $N[r(N)] = \tsample$
-
-    \item $f(N) \geq f(N')$ if $r(N)$ is any other redex, where $N \to N'$ at $r(N)$.
-\end{itemize}
-Any closed term for which a ranking function with respect to a reduction strategy $r$ exists is called \emph{rankable} with respect to $r$. 
-\end{definition}
-
 \begin{definition}
-A \emph{sparse ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : Rch_r(M) \rightharpoonup \mathbb R$ such that
-\begin{itemize}
-    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
-    \item $f$ is defined at $M$.
-    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ at the positions specified by $r$ will eventually reach some $O$ such that either $r(O)$ isn't defined or $f(O)$ is, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
-\end{itemize}
-\end{definition}
-
-\begin{definition}\rm
 An \emph{antitone ranking function on $M$ with respect to a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ such that
@@
 
-\begin{theorem}
+If the antitone function is just the constant function $1$ instead, then it is called a \emph{(sparse) ranking function on $M$ with respect to $r$} instead, and the term is called \emph{rankable} with respect to $r$.
+
+\begin{theorem} \label{thm:confluent ranking}
 If a closed PPCF term $M$ has a ranking function, sparse ranking function, antitone ranking function or antitone sparse ranking function with respect to a reduction strategy $r$, then $M$ is AST with respect to $r$, and AST.
 \end{theorem}
-\lo{Give an example or two of a reduction strategy (for $\Rightarrow$) that is not $\cbv$.}
+
+\paragraph{}
+If, as seems likely, \Cref{thm:antitone rankable implies termination} is complete for terms from which every reachable term is AST, then by the following theorem, \Cref{thm:confluent ranking} can't actually prove the termination of any terms not already provable by \Cref{thm:antitone rankable implies termination} and \Cref{thm:antitone partial implies rankable}.
+
+\begin{restatable}{theorem}{confluentNoStronger}
+For any closed term $M$ and reduction strategy $r$ on $M$, if every term in $Rch_r(M)$ is AST, then every term in $Rch(M)$ is AST.
+\end{restatable}
+
+However, there are some terms which terminate more simply and directly via some alternative reduction strategy.
+For these examples, a custom reduction strategy is specified for each term under consideration.
+
+\begin{example} \label{ex:sum of powers}
+The following term picks a random natural number $n$, then computes $\sum_{k=0}^n k^n$.
+\begin{align*}
+M =&\ (\lambda n. P[n]) \lfloor \tsample^{-1/2} \rfloor \\
+P[n] =&\ (\lambda p. \Xi[p] n) (\lambda x. \Theta[x] n)) \\
+\Xi[p] =&\ \tY \lambda f n. \tif{n = 0}{\underline 0}{p n + f (n-1)} \\
+\Theta[x] =&\ \tY \lambda f n. \tif{n = 0}{\underline 1}{x \times f (n-1)}.
+\end{align*}
+
+With the CBV reduction strategy, the recursion in $\Theta[x]$ (that defines the function $k \mapsto k^x$) is executed separately for every term in the sum in $\Xi$.
+Not only is this much more complicated than simplifying $\Theta[x] \underline n$ first, it requires considerably more $\tY$-reduction steps, so many so that $M$ is $\tY$-PAST with respect to a reduction strategy that simplifies $\Theta[x] \underline n$ first (call it ``$r$''), but isn't $\tY$-PAST with respect to the standard reduction strategy, so that $M$ is rankable with respect to $r$, but isn't rankable.
+It is still antitone rankable, but this is much more complicated.
+Details of the ranking function construction with and without using the confluent semantics are given in Appendix \ref{apx:confluent}.
+\end{example}
+
+Even some of the examples given earlier implicitly use slightly non-standard reduction strategies to make the definitions of their ranking functions slightly simpler.
+For example, the antitone sparse ranking function in \Cref{ex:unbiased random walk} was defined at $\Theta_2\ \underline n$ (where $\Theta_2 = \tY \, \lambda f n .
+\, \tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$), but $\Theta_2\ \underline{10}$ actually reduces to $\Theta_2\ (\underline{10} - 1)$ or $\Theta_2\ (\underline{10} + 1)$, and (taking the second of these), it then reduces to $(\lambda z.
+(\lambda n.
+\, \tif{n = 0}{0}{\Theta_2 \, (n - 1) \oplus_{1/2} \Theta_2 \, (n + 1)}) z) \, (\underline{10} + 1)$, then $(\lambda z.
+(\lambda n.
+\, \tif{n = 0}{0}{\Theta_2 \, (n - 1) \oplus_{1/2} \Theta_2 \, (n + 1)}) z) \, \underline{11}$, bypassing $\Theta_2\ \underline{11}$ entirely.
+The antitone sparse ranking function in the form given can be justified anyway, by considering it to be using a reduction strategy which is the same as $\cbv$ except that in $\Theta_2\ (\underline f \underline n)$, it evaluates the redex at $@_2$ first.
+
+\Cref{ex:higher-order recursion} also makes use of an alternative reduction strategy.
+In this case, it is to allow the antitone sparse ranking function to skip over all of those terms where the unknown function $F$ is part way through evaluation.

### 2021-01-09 LICS21/supermartingales-proofs.tex

@@
 
-\TechnicalRankingSupermartingale*
+\rankableAndStrictRankable*
+We shall obtain the theorem as a corollary of a technical lemma (\Cref{lem:key rankable}).
+
+We say that a given PPCF term is: 
+\emph{type 1} if it has the shape $E[\tY \lambda x.
+N]$, \emph{type 2} if it has the shape $E[\tsample]$; \emph{type 3} if it has the shape $E[R]$ where $R$ is any other redex, and \emph{type 4} if it is a value.
+Henceforth fix an $n \geq 0$, and define $\mathbf{T}_i := \set{s \mid M_n(s) \hbox{ is type $i$}}$.
+It is straightforward to see that each $\mathbf{T}_i \in \calF_n$, and $\set{\mathbf{T}_1, \mathbf{T}_2, \mathbf{T}_3, \mathbf{T}_4}$ is a partition of $\entrosp$.
+Hence it suffices to prove the following lemma.
+
+\begin{restatable}[Technical]{lemma}{TechnicalRankingSupermartingale}
+\label{lem:key rankable}
+For all $i \in \set{1, 2, 3, 4}$ and $A \in \calF_n$
+\[
+\int_A \mu_{\entrosp}(\dif s) \, f(M_{n+1})[s \in \mathbf{T}_i] \leq \int_A \mu_{\entrosp}(\dif s) \, f(M_n)[s \in \mathbf{T}_i]
+\; \footnote{where (the Iverson bracket) $[P] := 1$ if the statement $P$ hold, and 0 otherwise.}
+\] 
+%$\int_A \mu_{\entrosp}(\dif s) \, f(M_{n+1}) \leq \int_A \mu_{\entrosp}(\dif s) \, f(M_n)$, i.e., 
+Hence $\mathbb{E}[f(M_{n+1}) \mid \calF_n] \leq f(M_n)$ a.s.
+\end{restatable}
 
@@
 
+As an immediate corollary of \Cref{lem:key rankable}, each $f(M_n)$ is integrable.
+This concludes the proof of \Cref{thm:rankable and strict rankable}.
+
 \newcommand\transform[1]{{\ulcorner{#1}\urcorner}}
@@
 
-\lo{Note: \cite{DBLP:conf/lfcs/Groote94} considered a call-by-name version of a simply-typed nondeterministic lambda-calculus, and proved it strongly normalising, using a reducibility argument \`a la Tait.
-One should check that it extends to the call-by-value calculus.
-}
-
 \iffalse

### 2021-01-10 LICS21/SPCF.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
Changed bib style to IEEEtran; \citep -> \cite

Repositioned bibliography

@@
 \end{equation*}
-\lo{Both $\to$ and $\red$ are called reduction relation.}
+%\lo{Both $\to$ and $\red$ are called reduction relation.}
 
@@
 \label{rem:score}
-This definition of almost sure termination is equivalent to that given in \citep{MakOPW21} (although the program semantics is stated in a slightly different way), except for the presence of a $\tscore(r)$ construct for soft conditioning.
+This definition of almost sure termination is equivalent to that given in \cite{MakOPW21} (although the program semantics is stated in a slightly different way), except for the presence of a $\tscore(r)$ construct for soft conditioning.
 The $\tscore(r)$ construct is not relevant to termination except that it fails if its argument is less than or equal to zero, which allows computations to fail after finitely many steps. 

### 2021-01-10 LICS21/antitone-proofs.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
Changed bib style to IEEEtran; \citep -> \cite

Repositioned bibliography

@@
 
-Many of the recent advances in the development of AST verification methods \citep{DBLP:conf/pldi/ChenH20,DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18,DBLP:conf/aplas/HuangFC18,DBLP:conf/popl/ChatterjeeNZ17,DBLP:journals/pacmpl/AgrawalC018,DBLP:conf/cav/ChatterjeeFG16,DBLP:conf/lics/OlmedoKKM16,DBLP:journals/pacmpl/Huang0CG19} are concerned with loop-based programs.
+Many of the recent advances in the development of AST verification methods \cite{DBLP:conf/pldi/ChenH20,DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18,DBLP:conf/aplas/HuangFC18,DBLP:conf/popl/ChatterjeeNZ17,DBLP:journals/pacmpl/AgrawalC018,DBLP:conf/cav/ChatterjeeFG16,DBLP:conf/lics/OlmedoKKM16,DBLP:journals/pacmpl/Huang0CG19} are concerned with loop-based programs.
 We can view such loops as tail-recursive programs that are, in particular, \emph{affine recursive}, i.e., 
-in each evaluation (or run) of the body of the recursion, recursive calls are made from at most one call site \citep[\S 4.1]{DBLP:journals/toplas/LagoG19}.
+in each evaluation (or run) of the body of the recursion, recursive calls are made from at most one call site \cite[\S 4.1]{DBLP:journals/toplas/LagoG19}.
 (Note that whether a program is affine recursive cannot be checked by just counting textual occurrences of variables.)
 Termination analysis of \emph{non-affine recursive} probabilistic programs does not seem to have received much attention.
-Methods such as those presented in \citep{DBLP:journals/toplas/LagoG19} are explicitly restricted to affine programs, and are unsound otherwise.
-By contrast, many probabilistic programming languages allow for richer recursive structures \citep{DBLP:conf/pkdd/TolpinMW15,DBLP:conf/uai/GoodmanMRBT08,DBLP:journals/corr/MansinghkaSP14}.
+Methods such as those presented in \cite{DBLP:journals/toplas/LagoG19} are explicitly restricted to affine programs, and are unsound otherwise.
+By contrast, many probabilistic programming languages allow for richer recursive structures \cite{DBLP:conf/pkdd/TolpinMW15,DBLP:conf/uai/GoodmanMRBT08,DBLP:journals/corr/MansinghkaSP14}.
 
@@
 
-There is an obvious source of deterministic (i.e.~non-probabilistic) higher-order functions defined by recursion, viz., \emph{higher-order recursion schemes} (HORS) (see e.g.~\citep{DBLP:conf/lics/Ong06,DBLP:conf/lics/Ong15}).
+There is an obvious source of deterministic (i.e.~non-probabilistic) higher-order functions defined by recursion, viz., \emph{higher-order recursion schemes} (HORS) (see e.g.~\cite{DBLP:conf/lics/Ong06,DBLP:conf/lics/Ong15}).
 (Incidentally the (sparse) ranking function method is just as applicable to the termination analysis of \emph{deterministic} PPCF programs.)
 Recently \cite{DBLP:conf/lics/KobayashiLG19} have extended HORS to \emph{probabilistic higher-order recursion schemes} (PHORS), which are HORS augmented with probabilistic (binary) branching $\oplus_p$.
-As HORS are in essence the $\lambda \tY$-calculus \citep{DBLP:conf/lics/Statman02} (i.e.~pure simply-typed lambda calculus with recursion, generated from a finite base type), PHORS are definable in PPCF:
+As HORS are in essence the $\lambda \tY$-calculus \cite{DBLP:conf/lics/Statman02} (i.e.~pure simply-typed lambda calculus with recursion, generated from a finite base type), PHORS are definable in PPCF:
 (order-$n$) PHORS is encodable as (order-$n$) (call-by-name) PPCF, but the former is strictly less expressive (because the underlying recursion schemes are not Turing complete). 
@@
 %\akr{As I removed the referenced example, and I don't even know what a PHORS is, I've just removed this comment too.}
-A relevant result here is that the AST problem is decidable for order-1 PHORS (by reduction to the PSPACE-hard solvability of finite systems of polynomial equations with real coefficients; see \citep{DBLP:journals/jacm/EtessamiY09}), but undecidable for order-2 PHORS.
+A relevant result here is that the AST problem is decidable for order-1 PHORS (by reduction to the PSPACE-hard solvability of finite systems of polynomial equations with real coefficients; see \cite{DBLP:journals/jacm/EtessamiY09}), but undecidable for order-2 PHORS.
 

### 2021-01-10 LICS21/antitone.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
Changed bib style to IEEEtran; \citep -> \cite

Repositioned bibliography

@@
 Sampling from continuous distributions is an essential feature of statistical probabilistic programming languages.
-(See e.g.~Church \citep{DBLP:conf/uai/GoodmanMRBT08}, Stan \citep{carpenter2017stan}, Anglican \citep{DBLP:conf/pkdd/TolpinMW15}, Gen \citep{cusumano-towner2019Gen}, Pyro \citep{bingham2019Pyro}, Edward \citep{tran2016edward} and Turing \citep{ge2018Turing}.)
-Methods of proving AST of probabilistic computation have been developed for probabilistic programs with discrete distributions (see e.g.~\citep{DBLP:journals/toplas/LagoG19,DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/lics/KobayashiLG19,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05}).
+(See e.g.~Church \cite{DBLP:conf/uai/GoodmanMRBT08}, Stan \cite{carpenter2017stan}, Anglican \cite{DBLP:conf/pkdd/TolpinMW15}, Gen \cite{cusumano-towner2019Gen}, Pyro \cite{bingham2019Pyro}, Edward \cite{tran2016edward} and Turing \cite{ge2018Turing}.)
+Methods of proving AST of probabilistic computation have been developed for probabilistic programs with discrete distributions (see e.g.~\cite{DBLP:journals/toplas/LagoG19,DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/lics/KobayashiLG19,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05}).
 To our knowledge, the problem of proving AST of probabilistic programs with continuous distribution is new.
@@
 \begin{example}[Fair-in-the-limit random walk]
-\label{ex:Fair-in-the-limit random walk}\citep[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}
+\label{ex:Fair-in-the-limit random walk}\cite[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}
 \[
@@
 \begin{example}[Escaping spline]
-\label{ex:escaping spline}\citep[\S 5.4]{DBLP:journals/pacmpl/McIverMKK18}
+\label{ex:escaping spline}\cite[\S 5.4]{DBLP:journals/pacmpl/McIverMKK18}
 \[
@@
 $\Xi : (\textsf{R} \to \textsf{R} \to \textsf{R}) \to \textsf{R} \to \textsf{R} \to \textsf{R}$
-recursively defined\footnote{Inspired by examples in \citep{DBLP:journals/pacmpl/BurnOR18,DBLP:conf/lics/OngW19}.} by
+recursively defined\footnote{Inspired by examples in \cite{DBLP:journals/pacmpl/BurnOR18,DBLP:conf/lics/OngW19}.} by
 \[

### 2021-01-10 LICS21/applications.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
Changed bib style to IEEEtran; \citep -> \cite

Repositioned bibliography

@@
 If the labels of samples in the trace correspond closely to the roles that the samples play in the program execution, this will increase the efficiency of the inference algorithm.
-In \citep{wingate2011lightweight}, for example, something rather like a stack trace is used, although this labelling scheme turns out to have various problems.\citep{kiselyov2016problems} Potential positions could also be used for a similar purpose, although in practice, a more compact representation than an entire skeletal reduction sequence would be necessary.
-Selecting an appropriate addressing scheme can also be important in the context of variational inference.\citep[\S 6.2.2]{paige2016automatic}
+In \cite{wingate2011lightweight}, for example, something rather like a stack trace is used, although this labelling scheme turns out to have various problems.\cite{kiselyov2016problems} Potential positions could also be used for a similar purpose, although in practice, a more compact representation than an entire skeletal reduction sequence would be necessary.
+Selecting an appropriate addressing scheme can also be important in the context of variational inference.\cite[\S 6.2.2]{paige2016automatic}
 

### 2021-01-10 LICS21/confluent-proofs.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
Changed bib style to IEEEtran; \citep -> \cite

Repositioned bibliography

@@
 Suppose for induction that $A \sim_\uparrow^k (D,\delta) \sim_\downarrow^* (E,\epsilon) \sim_\uparrow^* C$ for some $D, \delta, E, \epsilon$, and $0 \leq k \leq n$, where in the reduction sequence $D \to^* E$, for any pair of reductions at positions which aren't disjoint, the reduction at the innermost (greater) position occurs first.
-Reduction sequences with this property will be called ``parallel'', as it is analagous to the parallel reduction introduced by Tait and Martin-L\"of in their proof of the Church-Rosser theorem \citep{Bar84,Takahashi1995}.
+Reduction sequences with this property will be called ``parallel'', as it is analagous to the parallel reduction introduced by Tait and Martin-L\"of in their proof of the Church-Rosser theorem \cite{Bar84,Takahashi1995}.
 This induction is in reverse, with $k$ decreasing from $n$ to $0$.

### 2021-01-10 LICS21/intro.tex
minor
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
Changed bib style to IEEEtran; \citep -> \cite

Repositioned bibliography

@@
 In other words, the program's behavior is used to define an associated ranking supermartingale.
-Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \citep{DBLP:journals/pacmpl/McIverMKK18}.
+Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \cite{DBLP:journals/pacmpl/McIverMKK18}.
 This paper's contribution is to extend this method to a higher-order setting.
@@
 This method is capable of proving the almost sure termination of programs which terminate arbitrarily slowly.
-Our progress condition is equivalent to that given in \citep{DBLP:journals/pacmpl/McIverMKK18} (by applying a transformation to the ranking function), but needs to be different in order to be compatible with the presence of additional intermediate reduction steps where the ranking function is not required to decrease at all.
+Our progress condition is equivalent to that given in \cite{DBLP:journals/pacmpl/McIverMKK18} (by applying a transformation to the ranking function), but needs to be different in order to be compatible with the presence of additional intermediate reduction steps where the ranking function is not required to decrease at all.
 
@@
 This is then used to prove yet another extension to the ranking function theorem, that ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and rankability in this sense still imples almost sure termination.
-This confluent trace semantics has other possible applications as well, for example in inference algorithms.
+This confluent trace semantics has other possible applications as well, for example in Bayesian inference algorithms.
 

### 2021-01-10 LICS21/supermartingales.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
Changed bib style to IEEEtran; \citep -> \cite

Repositioned bibliography

@@
 One approach to proving that a term terminates almost surely is to find some variant that is bounded below and, on average, decreases sufficiently quickly that it must eventually reach 0, similarly to the approach taken by
-%\citep{DBLP:journals/pacmpl/McIverMKK18} 
+%\cite{DBLP:journals/pacmpl/McIverMKK18} 
 \cite{DBLP:conf/popl/FioritiH15} and others for imperative programs.
@@
 \begin{remark}
-Our notion of ranking supermartingale is a slight generalisation of the original definition, due to \citep{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, which does not refer to a stopping time explicitly:
+Our notion of ranking supermartingale is a slight generalisation of the original definition, due to \cite{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, which does not refer to a stopping time explicitly:
 Fioriti and Hermanns call a supermartingale $(Y_n)_{n \geq 0}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot {\bf 1}_{\set{Y_n > 0}}$.
@@
 \begin{restatable}{lemma}{rankPast}
-%\begin{lemma}%[{\citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}}]
-% \citep[Prop 1]{DBLP:conf/popl/ChatterjeeFNH16}}
+%\begin{lemma}%[{\cite[Lemma 5.5]{DBLP:conf/popl/FioritiH15}}]
+% \cite[Prop 1]{DBLP:conf/popl/ChatterjeeFNH16}}
 \label{lem:rank-PAST}
@@
 \begin{remark}
-\Cref{lem:rank-PAST} is a slight extension of \citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15},
+\Cref{lem:rank-PAST} is a slight extension of \cite[Lemma 5.5]{DBLP:conf/popl/FioritiH15},
 which asserts the same result but for the specific stopping time $T : \omega \mapsto \min\set{n \mid Y_n(\omega) = 0}$.

### 2021-01-10 LICS21/SPCF.tex
Sec. 1: Added outline, "additional materials in appendices". Tidying things up, saving space.

@@
 %\paragraph{}
-Note that the requirement that $M$ be closed is not important for the proofs of the theorems, but is simply required to ensure that $M$ reaches a value, rather than a general normal form like $x \, y$.
+The requirement that $M$ be closed is not important for the proofs of the theorems, but is simply needed to ensure that $M$ reaches a value, rather than a general normal form like $x \, y$.
 

### 2021-01-10 LICS21/confluent.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
Changed bib style to IEEEtran; \citep -> \cite

Repositioned bibliography
Sec. 1: Added outline, "additional materials in appendices". Tidying things up, saving space.

@@
 The new sample space is then defined as $I^{L_s(M)}$, with the Borel $\sigma$-algebra and product measure.
-Since $I^{L_s(M)}$ is a countable product, the measure space is well-defined \citep[Cor.~2.7.3]{AshDD00}.
+Since $I^{L_s(M)}$ is a countable product, the measure space is well-defined \cite[Cor.~2.7.3]{AshDD00}.
 
@@
 \begin{proof}
-Suppose $M$ is AST with respect to $f$.
+Suppose $M$ is AST w.r.t.~$f$.
 Let the set of samples with which it terminates with this reduction strategy be $X$.
@@
 The domain of definition of the ranking functions also needs to be changed from the reachable terms $Rch(M)$ to the \emph{reachable terms with respect to the reduction strategy $r$},
-\begin{align*}
-& Rch_r(M) = \\ & \{N \mid \exists n, (N_i)_{0 \leq i \leq n}: N_0 = M, N_n = N, N_i \to N_{i+1} \text{ at } r(N_i)\}.
-\end{align*}
+\[
+\begin{array}{l}
+Rch_r(M) := \{N \mid \exists n, (N_i)_{0 \leq i \leq n} : {}\\
+\qquad\qquad\qquad N_0 = M, N_n = N, N_i \to N_{i+1} \text{ at } r(N_i)\}.
+\end{array}
+\]
 
-To be more explicit, the modified forms of the theorems are:
+More explicitly, the modified forms of the theorems are:
 \begin{definition}
@@
 It is still antitone rankable, but this is much more complicated.
-Details of the ranking function construction with and without using the confluent semantics are given in Appendix \ref{apx:confluent}.
+Details of the ranking function construction with and without using the confluent semantics are given in \Cref{apx:confluent}.
 \end{example}

### 2021-01-10 LICS21/intro.tex
Sec. 1: Added outline, "additional materials in appendices". Tidying things up, saving space.

@@
 Various theorems about probabilistic programs rely on the assumption that the program terminates almost surely.
-One possible approach to prove almost sure termination is to find some variant on the program state that decreases on average sufficiently quickly that it must at some point reach 0, at which point the program terminates.
-In other words, the program's behavior is used to define an associated ranking supermartingale.
-Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \cite{DBLP:journals/pacmpl/McIverMKK18}.
+One possible approach to prove almost sure termination (AST) is to find some variant on the program state that decreases on average sufficiently quickly that it must at some point reach 0, at which point the program terminates.
+In other words, the program's behavior is used to define an associated (ranking) supermartingale.
+Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \cite{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18}.
 This paper's contribution is to extend this method to a higher-order setting.
@@
 The language PPCF used in this paper is simply typed, with an explicit recursion primitive, $\tY$.
-As the type system already constrains the terms enough to force termination in the absence of the recursion construct, it is only the $\tY$-reduction steps that must be counted, making defining ranking functions somewhat easier.
+As the type system already constrains the terms enough to force termination in the absence of the recursion construct \cite{tait1967,BarendregtDS2010}, it is only the $\tY$-reduction steps that must be counted, making defining ranking functions somewhat easier.
 
@@
 This is the first application of martingales to probabilistic lambda calculus, and the first version of sampling semantics that's capable of satisfying the (restricted) Church-Rosser property.
+
+\lo{To continue.}
+
+\subsubsection*{Outline}
+
+We present the syntax and sampling (or trace) semantics of PPCF in \Cref{sec:PPCF}.
+In \Cref{sec:supermartingales}, we show that ranking functions on terms induce supermartingales, which form the basis of a sound method for proving AST.
+We introduce sparse ranking functions in \Cref{sec:ranking} and antitone ranking functions in \Cref{sec:antitone}, and illustrate how they can be used to prove AST via examples. 
+In \Cref{sec:confluent}, we present a confluent trace semantics and demonstrate its usefulness.
+We discuss further applications of the confluent semantics in \Cref{sec:applications}; and conclude with comments on related work and further directions in \Cref{sec:related}.
+
+\paragraph*{Additional materials} Further details of some of the examples and all missing proofs can be found in the appendices.

### 2021-01-10 LICS21/ranking.tex
Sec. 1: Added outline, "additional materials in appendices". Tidying things up, saving space.

@@
 Let $M \oplus_p N = \tif{\tsample - p < 0} M N$, for $p \in (0,1]$, then there are the pseudo-reduction relations
-\begin{align*}
-E[M \oplus_p N] \to^3 & E[M] & \\
-E[M \oplus_p N] \to^3 & E[N] & \\
-\red^3(E[M \oplus_p N], s) = & \left\{
+\[
+\begin{array}{c}
+E[M \oplus_p N] \to^3  E[M] \qquad
+E[M \oplus_p N] \to^3  E[N] \\[2mm]
+\red^3(E[M \oplus_p N], s) = \left\{
     \begin{array}{ll}
         (E[M],\pi_t(s)) & \text{if } \pi_h(s) < p \\
-        (E[N],\pi_t(s)) & \text{if } \pi_h(s) \geq p. \\
+        (E[N],\pi_t(s)) & \text{if } \pi_h(s) \geq p. 
     \end{array} \right .
-\end{align*}
+\end{array}
+\]
+% \begin{align*}
+% E[M \oplus_p N] \to^3 & E[M] & \\
+% E[M \oplus_p N] \to^3 & E[N] & \\
+% \red^3(E[M \oplus_p N], s) = & \left\{
+%     \begin{array}{ll}
+%         (E[M],\pi_t(s)) & \text{if } \pi_h(s) < p \\
+%         (E[N],\pi_t(s)) & \text{if } \pi_h(s) \geq p. \\
+%     \end{array} \right .
+% \end{align*}
 A sparse ranking function could be defined with respect to this shortcut reduction simply by replacing $\to$ and $\red$ in the definition of a sparse ranking function by a version that goes straight from $N \oplus_p O$ to $N$ or $O$.

### 2021-01-10 LICS21/supermartingales.tex
Sec. 1: Added outline, "additional materials in appendices". Tidying things up, saving space.

@@
     \item $f(E[\tsample]) \geq \int_I f(E[\underline{x}]) \, \Leb(\mathrm{d}x)$
-    \item $f(E[R]) \geq f(E[R'])$ for any other redex $R$, where $R \to R'$.
+    \item $f(E[R]) \geq f(E[R'])$ for any other redex $R$ with $R \to R'$
 \end{enumerate}
@@
 Any closed term for which a ranking (respectively, strict ranking) function exists is called \emph{rankable} (respectively, \emph{strictly rankable}). 
-For example, the term $(\tY \lambda x.x) \, \underline 0$ is not rankable.
+For example, $(\tY \lambda x.x) \, \underline 0$ is not rankable.
 \end{definition}
@@
 
-The following is an iterated version of Doob's well-known Optional Sampling Theorem (see, e.g., \cite[\S 6.7]{AshDD00} and \cite[Thm.~7.2]{DBLP:conf/popl/FioritiH15}).
+The following is an iterated version of Doob's well-known Optional Sampling Theorem 
+%(see, e.g., \cite[\S 6.7]{AshDD00} and \cite[Thm.~7.2]{DBLP:conf/popl/FioritiH15}).
+(see e.g.~\cite{AshDD00,DBLP:conf/popl/FioritiH15}).
 \begin{theorem}[Optional Sampling]
@@
 Our first result is the following theorem.
-\begin{restatable}{theorem}{rankableAndStrictRankable}[Deriving supermartingales] 
+\begin{restatable}[Deriving supermartingales]{theorem}{rankableAndStrictRankable}
 \label{thm:rankable and strict rankable}
@@
 
-\item If a closed PPCF term $M$ is strictly rankable, then $\expect{T_M} < \infty$ i.e.~$M$ is positively almost sure terminating (PAST).
+\item If a closed PPCF term $M$ is strictly rankable, then $\expect{T_M} < \infty$ i.e.~$M$ is positively a.s.~terminating (PAST).
 \end{enumerate}
@@
 %It follows immediately from \Cref{lem:rank-PAST} that $\expect{T_M} < \infty$.
-Statement 2 follows at once from \Cref{thm:rankable and strict rankable} and \Cref{lem:rank-PAST}.
+Statement 2 follows from \Cref{thm:rankable and strict rankable} and \Cref{lem:rank-PAST}.
 \end{proof}

### 2021-01-10 LICS21/antitone.tex
Sec. 1: Added outline, "additional materials in appendices". Tidying things up, saving space.
Abstract: minor
Reformatted def. of sparse ranking function (and two other similar def.), making it more compact.

@@
 The term $M_1$ is rankable, terminating in 31 $\tY$-steps on average, and $M_3$ only has a $1/1024$ chance of terminating, but in between, $M_2$ is AST, but isn't $\tY$-PAST, therefore it isn't rankable and \Cref{thm:rankable implies termination} is insufficient to prove its termination.
-We therefore want to find a generalised notion of ranking function so that $M_2$ becomes rankable, and then prove it sound i.e.~rankable in this generalised sense implies AST.
+Thus we seek a generalised notion of ranking function so that $M_2$ becomes rankable, and then prove it sound i.e.~rankable in this generalised sense implies AST.
 
@@
 \dfrac{\expect{Y_n}}{\mathbb P[T > n]}
-  \step[\leq]{}
-\dfrac{\expect{Y_0}}{\mathbb P[T > n]}
+%  \step[\leq]{}
+%\dfrac{\expect{Y_0}}{\mathbb P[T > n]}
+\mathbin{=} \dfrac{\expect{Y_0}}{\mathbb P[T > n]}
 \end{calculation}
@@
 \expect{Y_n-Y_{n+1}}
-  \step[=]{}
+%  \step[=]{}
+\mathbin{=}
 \expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
@@
 %  \step[=]{definition of conditional expectation for a discrete variable}
-\step[=]{}
+%\step[=]{}
+\mathbin{=}
 \mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
@@
 %\lo{$\expect{Y_0 - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})}$}
-which is a contradiction, therefore the claim must be true, therefore $P[T > n]$ tends to 0 as $n$ tends to infinity, therefore $T < \infty$ almost surely.
+which is a contradiction, therefore the claim must be true, therefore $P[T > n] \to 0$ as $n \to \infty$, therefore $T < \infty$ a.s.
 \end{proof}
@@
 \Cref{thm:rankable implies termination},
-take the probability space $(\entrosp, \Sigma_\entrosp, \mu_\entrosp)$, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$. 
+take the probability space $(\entrosp, \Sigma_\entrosp, \mu_\entrosp)$, and define the same r.v.~$T_n, X_n, Y_n, T_M^{\tY}$. 
 Thanks to \Cref{thm:rankable and strict rankable}, $(Y_n)_{n \geq 0}$ is a supermartingale; and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale w.r.t.~stopping time $T_M^{\tY}$.
@@
 
-Define an \emph{antitone sparse ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$
+Define an \emph{antitone sparse ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$:
+\begin{inparaenum}[(i)]
+\item $f(N) \geq 0$ for all $N \in \dom(f)$,
+\item $M \in \dom(f)$,
+\item for any $N \in \dom(f)$, evaluation of $N$ will eventually reach some $O$ which is either a value or in $\dom(f)$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of $\dom(f)$).
+\end{inparaenum}
+\iffalse
 \begin{itemize}
@@
 \end{itemize}
-
+\fi
 \begin{restatable}{theorem}{AntitonePartial}  \label{thm:antitone partial implies rankable}
@@
 by
-\[
-f: {\Theta_2} \, n 
-\mapsto 
-g_2(n), \qquad 
-0 \mapsto 0
-\]
+\(
+f_2: {\Theta_2} \, n \mapsto 
+g_2(n)
+\)
+and
+$0 \mapsto 0$
 for $n \in \mathbb N$.
@@
 Now
-\begin{align*}
+\begin{align*}%\textstyle
 & g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 \\
-  &=  \ln \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) \\
-  &=  \frac 1 2 \ln\left(\frac{(n+1)^2}{n(n+2)}\right) \\
-  &=  \frac 1 2 \ln\left(1 + \frac 1 {n(n+2)}\right) \\
-  &>  \frac 1 {n(n+2) + 1} \\
+  &=  \ln \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) 
+  =  \frac 1 2 \ln\left(\frac{(n+1)^2}{n(n+2)}\right) \\
+  &=  \frac 1 2 \ln\left(1 + \frac 1 {n(n+2)}\right) 
+  >  \frac 1 {n(n+2) + 1} \\
   &>  \frac 1 2 \frac 1 {(n+1)^2} + \frac 1 2 \frac 1 {(n+3)^2}
@@
 Methods of proving AST of probabilistic computation have been developed for probabilistic programs with discrete distributions (see e.g.~\cite{DBLP:journals/toplas/LagoG19,DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/lics/KobayashiLG19,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05}).
-To our knowledge, the problem of proving AST of probabilistic programs with continuous distribution is new.
+To our knowledge, the problem of proving AST of probabilistic functional programs with continuous distribution is new.
 
@@
 \[
-\underbrace{\big
-(\tY \, f \, x . \,
-0 \oplus_{\frac{1}{x+1}} f(x + 1) \big)}_{\Xi} 
+%\underbrace{
+\big(\tY \, f \, x . \,
+0 \oplus_{\frac{1}{x+1}} f(x + 1) 
+\big)
+%}_{\Xi} 
 \, 10
@@
 \Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} . \\
-\quad\tif{n \leq 0}{s}{
+\qquad\tif{n \leq 0}{s}{
 f \, n \, (\varphi \, f \, s \, (n - 1))

### 2021-01-10 LICS21/applications.tex
Abstract: minor
Reformatted def. of sparse ranking function (and two other similar def.), making it more compact.

@@
 If the labels of samples in the trace correspond closely to the roles that the samples play in the program execution, this will increase the efficiency of the inference algorithm.
-In \cite{wingate2011lightweight}, for example, something rather like a stack trace is used, although this labelling scheme turns out to have various problems.\cite{kiselyov2016problems} Potential positions could also be used for a similar purpose, although in practice, a more compact representation than an entire skeletal reduction sequence would be necessary.
-Selecting an appropriate addressing scheme can also be important in the context of variational inference.\cite[\S 6.2.2]{paige2016automatic}
+In \cite{wingate2011lightweight}, for example, something rather like a stack trace is used, although this labelling scheme turns out to have various problems \cite{kiselyov2016problems}. 
+Potential positions could also be used for a similar purpose, although in practice, a more compact representation than an entire skeletal reduction sequence would be necessary. 
+Selecting an appropriate addressing scheme can also be important in the context of variational inference \cite[\S 6.2.2]{paige2016automatic}.
 

### 2021-01-10 LICS21/confluent.tex
Reformatted def. of sparse ranking function (and two other similar def.), making it more compact.

@@
 \end{align*}
-The \emph{subterm of $M$ at a position $\alpha$}, denoted $M \mid \alpha$, is defined as
+The \emph{subterm of $M$ at $\alpha$}, denoted $M \mid \alpha$, is defined as
 \begin{align*}
@@
 \begin{example} Consider the term $M = \tY (\lambda f x.\ \tif{\tsample - \underline{0.5} < 0}{f \, x}{x}) \, \underline 0$, which reduces after a few steps to $N = \tif{\tsample - \underline{0.5}  < 0}{M}{\underline 0}$.
-If the pre-selected sample for $(N,\textsf{if}_1;\underline{-}_1)$ is less than 0.5, $N$ reduces back to $M$, then $N$ again, then the same sample is used the next time, therefore it's an infinite loop, whereas if samples are labelled by reduction sequences, the samples for $M \to \dots \to N$ are independent from the samples for $M \to \dots \to N \to M \to \dots \to N$, and so on.
+If the pre-selected sample for $(N,\textsf{if}_1;\underline{-}_1)$ is less than 0.5, $N$ reduces back to $M$, then $N$ again, then the same sample is used the next time, therefore it's an infinite loop, whereas if samples are labelled by reduction sequences, the samples for $M \to^\ast N$ are independent from the samples for $M \to^\ast N \to M \to^\ast N$, and so on.
 \end{example}
@@
 The reduction sequences of skeletons will often be discussed as though they were just skeletons, identifying them with their final skeletons.
-With this abuse of notation, a reduction sequence $N$ (actually $N_1 \to \dots \to N_n = N$) may be said to reduce to a reduction sequence $O$, where the reduction sequence implicitly associated with the final skeleton $O$ is $N_1 \to \dots \to N_n \to O$.
+With this abuse of notation, a reduction sequence $N$ (actually $N_1 \to^\ast N_n = N$) may be said to reduce to a reduction sequence $O$, where the reduction sequence implicitly associated with the final skeleton $O$ is $N_1 \to^\ast N_n \to O$.
 
@@
 An \emph{antitone sparse ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : Rch_r(M) \rightharpoonup \mathbb R$ such that there exists some antitone function $\epsilon : \nnReal \to \pReal$ such that
+\begin{inparaenum}[(i)]
+    \item $f(N) \geq 0$ for all $N \in \dom(f)$,
+    \item $M \in \dom(f)$,
+    \item for any $N \in \dom(f)$, evaluation of $N$ at the positions specified by $r$ will eventually reach some $O$ such that either $r(O)$ isn't defined or $f(O)$ is, and $f(N) \geq \mathbb E[f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of $\dom(f)$).
+\end{inparaenum}
+\iffalse
 \begin{itemize}
@@
 \end{itemize}
+\fi
 \end{definition}

### 2021-01-10 LICS21/ranking.tex
Reformatted def. of sparse ranking function (and two other similar def.), making it more compact.

@@
 Define a \emph{sparse ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \rightharpoonup \mathbb R$ such that
+\begin{inparaenum}[(i)]
+    \item $f(N) \geq 0$ for all $N \in \dom(f)$,
+    \item $M \in \dom(f)$,
+    \item for any $N \in \dom(f)$, evaluation of $N$ will eventually 
+    %\akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and AST, but that doesn't actually work because of terms reachable with 0 probability.
+Of course, ASTness doesn't imply rankability, so any definition of sparse ranking function that includes all Y-PAST terms wouldn't always extend to a total ranking function.}
+     reach some $O$ which is either a value or in $\dom(f)$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of $\dom(f)$).
+\end{inparaenum}
+
+\iffalse
 \begin{itemize}
@@
 A similar procedure would work for other forms of syntactic sugar.
-If a closed term $N$ eventually reduces to one of a set of other terms $\{N_i \mid i \in I\}$ with certain probabilities, a sparse ranking function defined with respect to a reduction sequence that skips straight from $N$ to $N_i$ is also a valid sparse ranking function for the original reduction function, and therefore its existence implies almost sure termination.
+If a closed term $N$ eventually reduces to one of a set of other terms $\{N_i \mid i \in I\}$ with certain probabilities, a sparse ranking function defined w.r.t.~a reduction sequence that skips straight from $N$ to $N_i$ is also a valid sparse ranking function for the original reduction function, and therefore its existence implies almost sure termination.
 There is a caveat, however, that $\tY$-reduction steps skipped over in the shortcut still need to be counted for the expected number of $\tY$-reduction steps.

### 2021-01-10 LICS21/related.tex
Sec. 1: Added outline, "additional materials in appendices". Tidying things up, saving space.
Reformatted def. of sparse ranking function (and two other similar def.), making it more compact.

@@
 % !TEX root = main.tex
-\section{Related work, conclusion and further directions}
+\section{Related work and further directions}
+
+\lo{TODO. Related work (.75 column). Discuss \cite{DBLP:conf/birthday/BlancLM05}.}
+
+\lo{Our results specialise to proving termination of non-random lambda calculus, which is much studied.
+Our approach is closest in spirit to Jones' work on size-change argument.
+Dal Lago's work builds on this.
+We think that martingales are a good framework for generalising / unifying with size-change work.}
+
+\lo{Other approaches, mainly based on types, are complementary to ours.}
+
+\lo{Kobayashi et al.: PHORS are strictly weaker in expressivity.}
+
 \label{sec:related}
@@
 
-The definition \label{def:more general red} of redexes and which positions are acceptable to reduce at is sufficiently restrictive to guarantee Church-Rosserness, but is also a little more restrictive in some cases than is necessary for this purpose.
+The \Cref{def:more general red} of redexes and which positions are acceptable to reduce at is sufficiently restrictive to guarantee Church-Rosserness, but is also a little more restrictive in some cases than is necessary for this purpose.
 For example, if the argument of a function is not yet a value, but its reduction to a value would be deterministic, or the function is affine, then applying the function before reducing its argument would not cause any problematic duplication of random samples.

### 2021-01-10 LICS21/applications.tex
abstract: tweak inserted version (for ease of reference, when people give comments) \paragraph{} -> \medskip Inserted todo points in Sec. I (intro) and Sec. VII (related work)

@@
 
-\paragraph{}
+\medskip
 Another possible application for the confluent trace semantics would be as a justification, within a sampling-style semantics, of some forms of equational reasoning.

### 2021-01-10 LICS21/confluent.tex
abstract: tweak inserted version (for ease of reference, when people give comments) \paragraph{} -> \medskip Inserted todo points in Sec. I (intro) and Sec. VII (related work)

@@
 
-\paragraph{}
+\medskip
 Even with this restriction, a sampling semantics in the style of the one already defined would not be entirely Church-Rosser, as, for example, $\red^3(\tsample - \tsample, (1,0,\dots))$ would be either $1$ or $-1$ depending on the order of evaluation of the $\tsample$s, as that determines which sample from the pre-selected sequence is used for each one.
@@
 
-\paragraph{}
+\medskip
 Labelling the pre-chosen samples by the positions in the term would also not work because in some cases, a $\tsample$ will be duplicated before being reduced, for example, in $(\lambda x.
@@
 
-\paragraph{}
+\medskip
 This is still not quite sufficient because sometimes the same samples must be used at corresponding positions in different reduction sequences.
@@
 
-\paragraph{}
+\medskip
 Before defining the new version of the reduction relation $\red$, the following lemma is necessary for it to be well-defined.
@@
 
-\paragraph{}
+\medskip
 This reduction relation now has all of the properties required of it.
@@
 
-\paragraph{}
+\medskip
 The reduction relation $\Rightarrow$ is nondeterministic, so it admits multiple possible reduction strategies. 
@@
 
-\paragraph{}
+\medskip
 If, as seems likely, \Cref{thm:antitone rankable implies termination} is complete for terms from which every reachable term is AST, then by the following theorem, \Cref{thm:confluent ranking} can't actually prove the termination of any terms not already provable by \Cref{thm:antitone rankable implies termination} and \Cref{thm:antitone partial implies rankable}.

### 2021-01-10 LICS21/confluent-proofs.tex
\ref -> \Cref

@@
 In the other direction, suppose $(N_1, \alpha_1) \sim_M^* (N_2, \alpha_2)$ where both $N_1$ and $N_2$ are descendants of $N$.
-By Lemma~\ref{downAcrossUp}, take $(N_1, \alpha_1) \sim_{p,M}^* (N'_1, \alpha') \sim_{c,M}^* (N'_2, \alpha') \sim_{p,M}^* (N_2, \alpha_2)$.
+By \Cref{downAcrossUp}, take $(N_1, \alpha_1) \sim_{p,M}^* (N'_1, \alpha') \sim_{c,M}^* (N'_2, \alpha') \sim_{p,M}^* (N_2, \alpha_2)$.
 The $\sim_p^*$ steps remain within $Rch(N)$, and $\sim_p$ does not depend on the history of the reduction sequences, therefore $(N_1, \alpha_1) \sim_{p,N}^* (N'_1, \alpha') \sim_{c,M}^* (N'_2, \alpha') \sim_{p,N}^* (N_2, \alpha_2)$.
 
-Let $X$ be the call-by-value reduction sequence related to both $N'_1$ and $N'_2$ by $\sim_{c,M}$ given by Lemma~\ref{canonicalCousins}. 
-As $M \to N$ is the first reduction in the sequence $N'_1$, it is the last to be affected by the $\sim_{c,M}$ sequence $N'_1 \sim_{c,M}^* X$ given by Lemma \ref{canonicalCousins} therefore it can be split into $N'_1 \sim_{c,M}^* Y_1 \sim_{c,M}^* X$ where $Y_1$ is in CBV order except possibly for its first reduction, which is still $M \to N$.
+Let $X$ be the call-by-value reduction sequence related to both $N'_1$ and $N'_2$ by $\sim_{c,M}$ given by \Cref{canonicalCousins}. 
+As $M \to N$ is the first reduction in the sequence $N'_1$, it is the last to be affected by the $\sim_{c,M}$ sequence $N'_1 \sim_{c,M}^* X$ given by \Cref{canonicalCousins} therefore it can be split into $N'_1 \sim_{c,M}^* Y_1 \sim_{c,M}^* X$ where $Y_1$ is in CBV order except possibly for its first reduction, which is still $M \to N$.
 Let the position of the reduction $M \to N$ be $\beta$.
@@
 In the case that $(M,s) \Rightarrow^* (M_1,s_1)$ or $(M,s) \Rightarrow^* (M_2,s_2)$ by multiple steps, it is possible to repeatedly replace a pair of the form $A \Leftarrow B \Rightarrow C$ by $A \Rightarrow^* D \Leftarrow^* C$ by the construction above, but it is not immediate that this process terminates, because each $\Rightarrow$ or $\Leftarrow$ may be replaced by multiple, so the sequence of $\Rightarrow$s and $\Leftarrow$s from $(M_1,s_1)$ to $(M_2,s_2)$ may get longer at some steps.
-However, termination can be proved by noting that the structure of this process is identical to the process of swapping $\sim_\uparrow$ and $\sim_\downarrow$ steps in Lemma \ref{downAcrossUp}.
+However, termination can be proved by noting that the structure of this process is identical to the process of swapping $\sim_\uparrow$ and $\sim_\downarrow$ steps in \Cref{downAcrossUp}.
 To be more precise, in this case there is a sequence of $\Leftarrow$ and $\Rightarrow$ steps, each of which has an associated reduction position and initial and final terms, and if a $\Leftarrow$ immediately precedes a $\Rightarrow$, they may be swapped to produce some number of $\Rightarrow$s, followed by some number of $\Leftarrow$s.
-In the case of Lemma \ref{downAcrossUp}, there is a sequence of $\sim_\uparrow$ and $\sim_\downarrow$ steps, each of which has an associated reduction position and initial and final skeletons, and if a $\sim_\uparrow$ immediately precedes a $\sim_\downarrow$, they may be swapped to produce some number of $\sim_\downarrow$s followed by some number of $\sim_\uparrow$s.
+In the case of \Cref{downAcrossUp}, there is a sequence of $\sim_\uparrow$ and $\sim_\downarrow$ steps, each of which has an associated reduction position and initial and final skeletons, and if a $\sim_\uparrow$ immediately precedes a $\sim_\downarrow$, they may be swapped to produce some number of $\sim_\downarrow$s followed by some number of $\sim_\uparrow$s.
 In both cases, the number and reduction positions of the resultant steps are determined by the case of $\sim_c$ that matches the way the initial reduction positions overlap, and the initial skeleton (or the skeleton of the initial term).
-The same argument that the process in Lemma \ref{downAcrossUp} terminated is therefore applicable here too.
+The same argument that the process in \Cref{downAcrossUp} terminated is therefore applicable here too.
 At every stage, the sequence of reductions that results from one of the initial $\Rightarrow$s is a parallel sequence, and swapping a parallel sequence of $\Rightarrow$s with a $\Leftarrow$ always terminates, therefore each $\Rightarrow$ in turn can be moved past all of the $\Leftarrow$s, and the process as a whole will terminate in a state where all of the $\Rightarrow$s precede all of the $\Leftarrow$s, i.e.~a pair of reduction sequences $(M_1,s_1) \Rightarrow^* (M',s') \Leftarrow^* (M_2,s_2)$.
@@
 \begin{proof}
-Suppose on the contrary that $(M,\alpha) \sim^* (A,\beta)$, where $N \to^* A$, then by Lemma \ref{simpDoesn'tMix}, $(M,\alpha) \sim_p^* (A,\beta)$.
+Suppose on the contrary that $(M,\alpha) \sim^* (A,\beta)$, where $N \to^* A$, then by \Cref{simpDoesn'tMix}, $(M,\alpha) \sim_p^* (A,\beta)$.
 $M \neq A$ therefore $(M,\alpha) \sim_p$ some position in $N$, but in all the cases of the definition of $\sim_p$, no position in the child term is related to the position of the redex.
@@
 
-Essentially what Lemma~\ref{lem:redexDestroyed} demonstrates is that the samples taken during any reduction sequence are independent of each other. This is made more precise in the following lemmas.
+Essentially what \Cref{lem:redexDestroyed} demonstrates is that the samples taken during any reduction sequence are independent of each other. This is made more precise in the following lemmas.
 
@@
 The distribution of $s(k)$ conditional on reaching $N$ is therefore the uniform distribution on $I$.
-Similarly for the other semantics, if the distribution of remaining samples after $n$ steps is independent of the term by Lemma \ref{lem:independentSamples}, therefore the distribution of $s(\mathit{Sk}(N),\cbv(N))$ conditional on $N$ is the uniform distribution on $I$.
+Similarly for the other semantics, if the distribution of remaining samples after $n$ steps is independent of the term by \Cref{lem:independentSamples}, therefore the distribution of $s(\mathit{Sk}(N),\cbv(N))$ conditional on $N$ is the uniform distribution on $I$.
 In either case, the distribution of the next term conditional on the previous term, in the case that it reduces randomly, is equal to the image under $r \mapsto E[\underline r]$ of the uniform distribution on $I$.
@@
 Let $(M,s) = (M_0,s_0) \Rightarrow_f \dots \Rightarrow_f (M_n,s_n)$, where $M_n$ is a value.
-By Lemma \ref{canonicalCousins}, the reduction sequence $(\mathit{Sk}(M_i))_i$ is related by $\sim_c^*$ to some skeletal reduction sequence $X$ which is in call-by-value order.
+By \Cref{canonicalCousins}, the reduction sequence $(\mathit{Sk}(M_i))_i$ is related by $\sim_c^*$ to some skeletal reduction sequence $X$ which is in call-by-value order.
 For each skeletal reduction sequence $X_j$ in the sequence $(\mathit{Sk}(M_i))_i \sim_c \dots \sim_c X$, it is possible to define a reduction sequence of terms such that $(M,s) = (N_{j,0},s_{j,0}) \Rightarrow \dots \Rightarrow (N_{j,n_j},s_{j,n_j})$ and $\mathit{Sk}(N_{j,k}) = X_{j,k}$, by at each step applying the reduction at the same place as in $X_j$, essentially just filling in the reals in the skeletal reduction sequence to make it a reduction sequence of terms (it will be shown shortly that the correct branch is taken in each $\textsf{if}$-reduction).
@@
 To prove that this holds true for $j+1$ as well, it suffices to show that the corresponding term and samples immediately after the reductions involved in this $\sim_c$ step match, as from that point on, the $\Rightarrow$-sequences will match, as they start from the same point and have all the same reduction positions.
-Letting the term and remaining samples immediately before the reductions involved in this $\sim_c$ step be $(N,t)$ (which is the same in both $N_j$ and $N_{j+1}$ for the same reason again, they start at the same point and the sequences of reduction positions up to this point are the same), and the terms and samples after these reductions be $(O_1,t_1)$ and $(O_2,t_2)$, so that these match the $N$, $O_1$ and $O_2$ in the definition of $\sim_c$, $t_1 = t_2$ because both of them are, by Lemma \ref{lem:sim-M-N}, equal to the restriction of $t$ to the subset of $Rch(N)$ that starts with the reduction sequences to $O_1$ and $O_2$ respectively, and the potential positions in those subsets of $Rch(N)$ are identified with each other by $\sim$ because of the same case of $\sim_c$ that means $N_j$ and $N_{j+1}$ are related.
+Letting the term and remaining samples immediately before the reductions involved in this $\sim_c$ step be $(N,t)$ (which is the same in both $N_j$ and $N_{j+1}$ for the same reason again, they start at the same point and the sequences of reduction positions up to this point are the same), and the terms and samples after these reductions be $(O_1,t_1)$ and $(O_2,t_2)$, so that these match the $N$, $O_1$ and $O_2$ in the definition of $\sim_c$, $t_1 = t_2$ because both of them are, by \Cref{lem:sim-M-N}, equal to the restriction of $t$ to the subset of $Rch(N)$ that starts with the reduction sequences to $O_1$ and $O_2$ respectively, and the potential positions in those subsets of $Rch(N)$ are identified with each other by $\sim$ because of the same case of $\sim_c$ that means $N_j$ and $N_{j+1}$ are related.
 If none of the relevant reductions are $\tsample$-reductions, $O_1 = O_2$ trivially.
 If there are samples taken (which must be the arbitrary reduction at $\beta$), it is also required that the same samples are taken in both branches.
-Similarly to Lemma \ref{churchRosser}, the positions of these $\tsample$-reductions in each branch are related by $\sim_p^*$, therefore they correspond to the same sample in $t$, therefore $O_1 = O_2$.
+Similarly to \Cref{churchRosser}, the positions of these $\tsample$-reductions in each branch are related by $\sim_p^*$, therefore they correspond to the same sample in $t$, therefore $O_1 = O_2$.
 The fact that the terms are equal also implies that the same branch of an $\textsf{if}$ statements will be taken.

### 2021-01-10 LICS21/related.tex
Introduce Conjecture environment for the completeness conjecture.

@@
 \label{sec:related}
-We have not yet been able to prove our conjecture that every term is antitone rankable if all of the terms reachable from it are AST, which would imply that \Cref{thm:antitone rankable implies termination} is almost complete.
+We have not yet been able to prove \Cref{conj:antitone} which would imply that \Cref{thm:antitone rankable implies termination} is almost complete.
+\begin{conjecture}
+\label{conj:antitone}
+Every term is antitone rankable if all of the terms reachable from it are AST.
+\end{conjecture}
 

### 2021-01-12 LICS21/SPCF.tex
Abstract: minor
Sec. VII and VIII: comments and suggestions.

@@
 
-\subsection{Sampling semantics}
+\subsection{Trace semantics}
 \label{sec:sampling semantics}
@@
 %\paragraph{}
-The requirement that $M$ be closed is not important for the proofs of the theorems, but is simply needed to ensure that $M$ reaches a value, rather than a general normal form like $x \, y$.
+%The requirement that $M$ be closed is not important for the proofs of the theorems, but is simply needed to ensure that $M$ reaches a value, rather than a general normal form like $x \, y$.
+\lo{This can be omitted.}
 
@@
 The $\tscore(r)$ construct is not relevant to termination except that it fails if its argument is less than or equal to zero, which allows computations to fail after finitely many steps. 
-If such a thing were added, it would merely require an additional side condition to our main theorems, that no failing term was reachable (or, in the case of the confluent semantics based results, no term where the reduction stragegy is not defined is reachable via the reduction strategy unless it's a value).
+If such a thing were added, it would merely require an additional side condition to our main theorems, that no failing term was reachable (or, in the case of the confluent semantics based results, no term where the reduction strategy is not defined is reachable via the reduction strategy unless it's a value).
 \end{remark}

### 2021-01-12 LICS21/applications.tex
Sec. VII and VIII: comments and suggestions.

@@
 \label{sec:applications}
-Although the definition of potential positions in a term in \Cref{sec:confluent} was intended simply to define a variant of sampling semantics that has a restricted version of the Church-Rosser property, it is likely to also be applicable to probabilistic inference algorithms as well.
+Although the definition of potential positions in a term in \Cref{sec:confluent} was intended simply to define a variant of trace semantics that has a restricted version of the Church-Rosser property, it is likely to also be applicable to \changed[lo]{inference algorithms (for probabilistic programs) as well}.
 
-In the Metropolis-Hastings inference algorithm, the program is executed, with a random value being chosen each time a $\tsample$ redex is evaluated.
+\lo{What you described in the following paragraph is not the Metropolis-Hastings algorithm (for MH algorithm, see e.g.~Algorithm 1, Slide 18 of 5-MCMC-part2.pdf \footnote{\url{https://www.cs.ox.ac.uk/teaching/materials20-21/SPP/Lectures/5-MCMC-part2.pdf}, \cite{Ong20}}), but a version of component-wise MH adapted to (models \emph{qua}) probabilistic programs (see Slide 36 onwards, 6-operational semantics.pdf \footnote{\url{https://www.cs.ox.ac.uk/teaching/materials20-21/SPP/Lectures/6-Operational-Semantics-21Nov20.pdf}, \cite{Ong20} \label{fnote:lmh}}).}
+
+%In the Metropolis-Hastings inference algorithm, 
+\changed[lo]{In lightweight Metropolis-Hastings (LMH) or single-site MH (originally suggested by \cite{wingate2011lightweight}; see also \cite{rainforth2017Automating})},
+the program is executed, with a random value being chosen each time a $\tsample$ redex is evaluated.
 The trace of samples used is recorded, then for the next step, it is modified slightly before the program is run again.
@@
 If the labels of samples in the trace correspond closely to the roles that the samples play in the program execution, this will increase the efficiency of the inference algorithm.
-In \cite{wingate2011lightweight}, for example, something rather like a stack trace is used, although this labelling scheme turns out to have various problems \cite{kiselyov2016problems}. 
+In \cite{wingate2011lightweight}, for example, something rather like a stack trace is used, although this assdressing scheme turns out to have various problems \cite{kiselyov2016problems,Hur2015a}. 
 Potential positions could also be used for a similar purpose, although in practice, a more compact representation than an entire skeletal reduction sequence would be necessary. 
+\lo{It would be very interesting (and rewarding, I believe) to work out the details.
+There is a formalisation of LMH for Idealised Anglican in \Cref{fnote:lmh}.
+SPCF is a much nicer probabilistic programming language (in purified / idealised form).}
 Selecting an appropriate addressing scheme can also be important in the context of variational inference \cite[\S 6.2.2]{paige2016automatic}.
@@
 \medskip
+
 Another possible application for the confluent trace semantics would be as a justification, within a sampling-style semantics, of some forms of equational reasoning. 
 By introducing reduction steps that go backwards to terms not reachable from the original term, it may in some cases be simplified, or equivalences between terms may be proved, by appealing to confluence results such as \Cref{churchRosser} and \Cref{thm:CbvIsTerminatingest}.
+\lo{Can you provide an example (and present it in the appendix)?}

### 2021-01-12 LICS21/confluent-proofs.tex
Sec. VII and VIII: comments and suggestions.

@@
 By \Cref{downAcrossUp}, take $(N_1, \alpha_1) \sim_{p,M}^* (N'_1, \alpha') \sim_{c,M}^* (N'_2, \alpha') \sim_{p,M}^* (N_2, \alpha_2)$.
-The $\sim_p^*$ steps remain within $Rch(N)$, and $\sim_p$ does not depend on the history of the reduction sequences, therefore $(N_1, \alpha_1) \sim_{p,N}^* (N'_1, \alpha') \sim_{c,M}^* (N'_2, \alpha') \sim_{p,N}^* (N_2, \alpha_2)$.
+The $\sim_p^*$ steps remain within $\mathit{Rch}(N)$, and $\sim_p$ does not depend on the history of the reduction sequences, therefore $(N_1, \alpha_1) \sim_{p,N}^* (N'_1, \alpha') \sim_{c,M}^* (N'_2, \alpha') \sim_{p,N}^* (N_2, \alpha_2)$.
 
@@
 
-Combining this with the $\sim_{p,N}^*$s at the beginning and end then yields the desired result that $(N_1,\alpha_1) \sim_N^* (N_2, \alpha_2)$, therefore the restriction of $\sim_M^*$ to $\sim_N$'s domain ($Rch(N)$) is a subset of $\sim_N^*$ therefore the two versions of $\sim$ match on this domain, as desired.
+Combining this with the $\sim_{p,N}^*$s at the beginning and end then yields the desired result that $(N_1,\alpha_1) \sim_N^* (N_2, \alpha_2)$, therefore the restriction of $\sim_M^*$ to $\sim_N$'s domain ($\mathit{Rch}(N)$) is a subset of $\sim_N^*$ therefore the two versions of $\sim$ match on this domain, as desired.
 \end{proof}
@@
 Let\begin{align*}
-R_s & = \{\mathit{Sk}(O) \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) = \tsample, O \to N[r]\} \\
-R_d & = \{\mathit{Sk}(O) \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) \neq \tsample, O \to N[r]\}.
+R_s & = \{\mathit{Sk}(O) \mid O \in \mathit{Rch}(M), r \in \mathbb R^k, O | f(O) = \tsample, O \to N[r]\} \\
+R_d & = \{\mathit{Sk}(O) \mid O \in \mathit{Rch}(M), r \in \mathbb R^k, O | f(O) \neq \tsample, O \to N[r]\}.
 \end{align*}
@@
 To prove that this holds true for $j+1$ as well, it suffices to show that the corresponding term and samples immediately after the reductions involved in this $\sim_c$ step match, as from that point on, the $\Rightarrow$-sequences will match, as they start from the same point and have all the same reduction positions.
-Letting the term and remaining samples immediately before the reductions involved in this $\sim_c$ step be $(N,t)$ (which is the same in both $N_j$ and $N_{j+1}$ for the same reason again, they start at the same point and the sequences of reduction positions up to this point are the same), and the terms and samples after these reductions be $(O_1,t_1)$ and $(O_2,t_2)$, so that these match the $N$, $O_1$ and $O_2$ in the definition of $\sim_c$, $t_1 = t_2$ because both of them are, by \Cref{lem:sim-M-N}, equal to the restriction of $t$ to the subset of $Rch(N)$ that starts with the reduction sequences to $O_1$ and $O_2$ respectively, and the potential positions in those subsets of $Rch(N)$ are identified with each other by $\sim$ because of the same case of $\sim_c$ that means $N_j$ and $N_{j+1}$ are related.
+Letting the term and remaining samples immediately before the reductions involved in this $\sim_c$ step be $(N,t)$ (which is the same in both $N_j$ and $N_{j+1}$ for the same reason again, they start at the same point and the sequences of reduction positions up to this point are the same), and the terms and samples after these reductions be $(O_1,t_1)$ and $(O_2,t_2)$, so that these match the $N$, $O_1$ and $O_2$ in the definition of $\sim_c$, $t_1 = t_2$ because both of them are, by \Cref{lem:sim-M-N}, equal to the restriction of $t$ to the subset of $\mathit{Rch}(N)$ that starts with the reduction sequences to $O_1$ and $O_2$ respectively, and the potential positions in those subsets of $\mathit{Rch}(N)$ are identified with each other by $\sim$ because of the same case of $\sim_c$ that means $N_j$ and $N_{j+1}$ are related.
 If none of the relevant reductions are $\tsample$-reductions, $O_1 = O_2$ trivially.
@@
 The injection $i(M \to^k N)$ is the same as it would be for any other element of $Q$, and its image is disjoint from $P \setminus P_i$.
-The inverse image under composition with $i(M \to^k N)$ of $Q$ is therefore a non-null subset of $I^{L_s(N)}$, and $N$ fails to terminate with $r$ and every trace in this set, therefore $N$ is not AST with respect to $r$, therefore $N$ is not AST, but $N \in Rch_r(M)$, therefore not every element of $Rch_r(M)$ is AST.
+The inverse image under composition with $i(M \to^k N)$ of $Q$ is therefore a non-null subset of $I^{L_s(N)}$, and $N$ fails to terminate with $r$ and every trace in this set, therefore $N$ is not AST with respect to $r$, therefore $N$ is not AST, but $N \in \mathit{Rch}_r(M)$, therefore not every element of $\mathit{Rch}_r(M)$ is AST.
 

### 2021-01-12 LICS21/confluent.tex
Sec. VII and VIII: comments and suggestions.

@@
 The reduction relation $\Rightarrow$ is nondeterministic, so it admits multiple possible reduction strategies. 
-A \emph{reduction strategy} starting from a closed term $M$ is a partial function $f$ from $Rch(M)$ to positions, such that for any reachable term $N$ where $f$ is defined, $f(N)$ is a position of a redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
+A \emph{reduction strategy} starting from a closed term $M$ is a partial function $f$ from $\mathit{Rch}(M)$ to positions, such that for any reachable term $N$ where $f$ is defined, $f(N)$ is a position of a redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
 Using a reduction strategy $f$, a subset of $\Rightarrow$ that isn't nondeterministic, $\Rightarrow_f$, can be defined by $(N,s) \Rightarrow_f (N',s')$ just if $(N,s) \Rightarrow (N',s')$ and $N$ reduces to $N'$ with the redex at $f(N)$.
@@
 
-The domain of definition of the ranking functions also needs to be changed from the reachable terms $Rch(M)$ to the \emph{reachable terms with respect to the reduction strategy $r$},
+The domain of definition of the ranking functions also needs to be changed from the reachable terms $\mathit{Rch}(M)$ to the \emph{reachable terms with respect to the reduction strategy $r$},
 \[
 \begin{array}{l}
-Rch_r(M) := \{N \mid \exists n, (N_i)_{0 \leq i \leq n} : {}\\
+\mathit{Rch}_r(M) := \{N \mid \exists n, (N_i)_{0 \leq i \leq n} : {}\\
 \qquad\qquad\qquad N_0 = M, N_n = N, N_i \to N_{i+1} \text{ at } r(N_i)\}.
@@
 \begin{definition}
-An \emph{antitone sparse ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : Rch_r(M) \rightharpoonup \mathbb R$ such that there exists some antitone function $\epsilon : \nnReal \to \pReal$ such that
+An \emph{antitone sparse ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : \mathit{Rch}_r(M) \rightharpoonup \mathbb R$ such that there exists some antitone function $\epsilon : \nnReal \to \pReal$ such that
 \begin{inparaenum}[(i)]
@@
 \begin{restatable}{theorem}{confluentNoStronger}
-For any closed term $M$ and reduction strategy $r$ on $M$, if every term in $Rch_r(M)$ is AST, then every term in $Rch(M)$ is AST.
+For any closed term $M$ and reduction strategy $r$ on $M$, if every term in $\mathit{Rch}_r(M)$ is AST, then every term in $\mathit{Rch}(M)$ is AST.
 \end{restatable}
@@
 M =&\ (\lambda n. P[n]) \lfloor \tsample^{-1/2} \rfloor \\
-P[n] =&\ (\lambda p. \Xi[p] n) (\lambda x. \Theta[x] n)) \\
-\Xi[p] =&\ \tY \lambda f n. \tif{n = 0}{\underline 0}{p n + f (n-1)} \\
+P[n] =&\ (\lambda p. \Xi[p] \, n) (\lambda x. \Theta[x] \, n)) \\
+\Xi[p] =&\ \tY \lambda f n. \tif{n = 0}{\underline 0}{p \, n + f (n-1)} \\
 \Theta[x] =&\ \tY \lambda f n. \tif{n = 0}{\underline 1}{x \times f (n-1)}.
@@
 With the CBV reduction strategy, the recursion in $\Theta[x]$ (that defines the function $k \mapsto k^x$) is executed separately for every term in the sum in $\Xi$.
-Not only is this much more complicated than simplifying $\Theta[x] \underline n$ first, it requires considerably more $\tY$-reduction steps, so many so that $M$ is $\tY$-PAST with respect to a reduction strategy that simplifies $\Theta[x] \underline n$ first (call it ``$r$''), but isn't $\tY$-PAST with respect to the standard reduction strategy, so that $M$ is rankable with respect to $r$, but isn't rankable.
+Not only is this much more complicated than simplifying $\Theta[x] \, \underline n$ first, it requires considerably more $\tY$-reduction steps, so many so that $M$ is $\tY$-PAST with respect to a reduction strategy that simplifies $\Theta[x] \, \underline n$ first (call it ``$r$''), but isn't $\tY$-PAST with respect to the standard reduction strategy, so that $M$ is rankable with respect to $r$, but isn't rankable.
 It is still antitone rankable, but this is much more complicated.

### 2021-01-12 LICS21/intro.tex
Reformatted def. of sparse ranking function (and two other similar def.), making it more compact.
\ref -> \Cref
Sec. VII and VIII: comments and suggestions.

@@
 \label{sec:intro}
+
+\lo{Proving termination of higher-order probabilistic programs with continuous distribution is important but difficult.}
 
@@
 
-The notion of a sparse ranking function is introduced to make constructing ranking functions more convenient.
+The notion of a \emph{sparse ranking function} is introduced to make constructing ranking functions more convenient.
 Most of the individual execution steps of a typical program are trivial and easy to mentally skip over.
@@
 
-\lo{To continue.}
+\lo{Ongoing.}
 

### 2021-01-12 LICS21/related.tex
Sec. VII and VIII: comments and suggestions.

@@
 For example, if the argument of a function is not yet a value, but its reduction to a value would be deterministic, or the function is affine, then applying the function before reducing its argument would not cause any problematic duplication of random samples.
-Similarly, if $\tsample$ occurs at position $\alpha;@_2;\beta;\lambda;\gamma$, but the function at $\alpha;@_1$ is affine, evaluatinf the $\tsample$ may in some cases also work fine.
+Similarly, if $\tsample$ occurs at position $\alpha;@_2;\beta;\lambda;\gamma$, but the function at $\alpha;@_1$ is affine, evaluating the $\tsample$ may in some cases also work fine.
 A more complete characterisation of which redexes can be reduced without breaking Church-Rosserness could be interesting.

### 2021-01-12 LICS21/related.tex
Added some references

@@
 \section{Related work and further directions}
+\label{sec:related}
 
@@
 \lo{Our results specialise to proving termination of non-random lambda calculus, which is much studied.
-Our approach is closest in spirit to Jones' work on size-change argument.
-Dal Lago's work builds on this.
+Our approach is closest in spirit to Jones' work on size-change principle \cite{DBLP:journals/lmcs/JonesB08,DBLP:conf/aplas/SereniJ05}.
+Dal Lago's \cite{DBLP:journals/toplas/LagoG19} work builds on this.
 We think that martingales are a good framework for generalising / unifying with size-change work.}
@@
 
-\lo{Kobayashi et al.: PHORS are strictly weaker in expressivity.}
+\lo{Kobayashi et al.: PHORS are strictly weaker in expressivity \cite{DBLP:conf/lics/KobayashiLG1}.}
 
-\label{sec:related}
 We have not yet been able to prove \Cref{conj:antitone} which would imply that \Cref{thm:antitone rankable implies termination} is almost complete.

### 2021-01-13 LICS21/applications.tex
Consistently use "trace semantics" rather than "sampling semantics"

@@
 
-Another possible application for the confluent trace semantics would be as a justification, within a sampling-style semantics, of some forms of equational reasoning. 
+Another possible application for the confluent trace semantics would be as a justification, within a trace-style semantics, of some forms of equational reasoning. 
 By introducing reduction steps that go backwards to terms not reachable from the original term, it may in some cases be simplified, or equivalences between terms may be proved, by appealing to confluence results such as \Cref{churchRosser} and \Cref{thm:CbvIsTerminatingest}.

### 2021-01-13 LICS21/confluent.tex
Consistently use "trace semantics" rather than "sampling semantics"

@@
 \medskip
-Even with this restriction, a sampling semantics in the style of the one already defined would not be entirely Church-Rosser, as, for example, $\red^3(\tsample - \tsample, (1,0,\dots))$ would be either $1$ or $-1$ depending on the order of evaluation of the $\tsample$s, as that determines which sample from the pre-selected sequence is used for each one.
+Even with this restriction, a trace semantics in the style of the one already defined would not be entirely Church-Rosser, as, for example, $\red^3(\tsample - \tsample, (1,0,\dots))$ would be either $1$ or $-1$ depending on the order of evaluation of the $\tsample$s, as that determines which sample from the pre-selected sequence is used for each one.
 To fix this, rather than pre-selecting samples according to the order they'll be drawn in, select them according to the position in the term where they'll be used instead.
@@
 Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a $\to$-reduction sequence of skeletons starting at $\mathit{Sk}(M)$, and the second of which is a position in the final skeleton of the reduction sequence.
-As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the standard sampling semantics, modified traces, which are elements of $I^{L_0(M)}$ (with one more caveat introduced later), will be used to pre-select a sample from $I$ for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.
+As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the standard trace semantics, modified traces, which are elements of $I^{L_0(M)}$ (with one more caveat introduced later), will be used to pre-select a sample from $I$ for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.
 
@@
 
-With these definitions, it is now possible to relate the confluent sampling semantics back to the standard sampling semantics.
+With these definitions, it is now possible to relate the confluent trace semantics back to the standard trace semantics.
 

### 2021-01-13 LICS21/intro.tex
Consistently use "trace semantics" rather than "sampling semantics"

@@
 
-This is the first application of martingales to probabilistic lambda calculus, and the first version of sampling semantics that's capable of satisfying the (restricted) Church-Rosser property.
+This is the first application of martingales to probabilistic lambda calculus, and the first version of trace semantics that's capable of satisfying the (restricted) Church-Rosser property.
 
@@
 
-We present the syntax and sampling (or trace) semantics of PPCF in \Cref{sec:PPCF}.
+We present the syntax and trace (or sampling) semantics of PPCF in \Cref{sec:PPCF}.
 In \Cref{sec:supermartingales}, we show that ranking functions on terms induce supermartingales, which form the basis of a sound method for proving AST.

### 2021-01-13 LICS21/notes.tex
Add removed example back into notes.tex

@@
 V \, (\tY \, V) \, z$, but I expect this would affect many proofs in the confluent semantics section, and there may not be time for such a change.
+
+\begin{example}[Non-affine recursion with continuous distribution]
+\label{ex:non-affine with continuous distribution}
+Consider 
+\[
+\Xi := \tY \lambda f x . \big(\mathsf{let} \; e = \tsample \; \mathsf{in} \;
+\mathsf{if}\boldsymbol{(}e \leq p, x, X\boldsymbol{)}\big)
+\]
+where
+\(
+X = \big( f^3(x+1) \oplus_e f^2(x+1) \big) \oplus_{\mathit{sig}(x)} f^2(x+1) 
+\)
+and $\mathit{sig}(x)$ is the sigmoid function.
+
+Not only is this program non-affine recursive,
+note also the use of a random sample, $e$, as a first-class value, and the subsequent use as a probability in the binary choice $\oplus_e$.
+Such a computation cannot be modelled via discrete distributions.
+
+We can use the ranking function method (coupled with the solution of linear recurrence relations) to show that provided $p \geq \sqrt{7} - 2$, the program $\Xi \, \underline r$ is AST for all $r \in \Real$.
+
+\lo{This example is taken from the series of 3d-printing-company running examples (Beutner-Ong 2021).
+This paper is as yet unpublished, and effectively uncitable for the time being (because of strict double-blind submission rules).}
+\end{example}

### 2021-01-13 LICS21/antitone-proofs.tex
Sec. 1: Added outline, "additional materials in appendices". Tidying things up, saving space.
Fix typo (g_1 vs g_2)

@@
 Let $F : \textsf{R} \to \textsf{R} \to \textsf{R}$ be a function such that $F\ \underline n\ \underline m$ terminates with no $\tY$ reductions for all $m$ and $n$.
-Then $\Xi\ F\ \underline x\ \underline y$ has the antitone sparse ranking function $F \underline{n_0} (F \underline{n_1} (\dots(\Xi\ F\ \underline x\ \underline n)\dots) \mapsto g_1(n)$, where $g_1$ is as in \Cref{ex:unbiased random walk}, using the same antitone function too.
+Then $\Xi\ F\ \underline x\ \underline y$ has the antitone sparse ranking function $F \underline{n_0} (F \underline{n_1} (\dots(\Xi\ F\ \underline x\ \underline n)\dots) \mapsto g_2(n)$, where $g_2$ is as in \Cref{ex:unbiased random walk}, using the same antitone function too.
 (Again, this is not actually the correct reduction order, in that $F$ should be applied to its argument $n_i$ and reduced to a value before its second argument is expanded, but this would complicate the presentation, and this version can be justified by \Cref{thm:confluent ranking} instead.)
@@
     & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \sum_{m=t_k}^{n-1} \frac{2^k}{t_{k+1}-t_k}) \\
-\end{align*}
+%\end{align*}
 %Allow a page break so that the sim-diagram lines up neatly.
-\begin{align*}
+%\begin{align*}
     & = \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \frac{2^k (n - t_k)}{t_{k+1}-t_k}) \\

### 2021-01-14 LICS21/antitone-proofs.tex
Add more explanation for the confluent semantics.

@@
 Then $\Xi\ F\ \underline x\ \underline y$ has the antitone sparse ranking function $F \underline{n_0} (F \underline{n_1} (\dots(\Xi\ F\ \underline x\ \underline n)\dots) \mapsto g_2(n)$, where $g_2$ is as in \Cref{ex:unbiased random walk}, using the same antitone function too.
-(Again, this is not actually the correct reduction order, in that $F$ should be applied to its argument $n_i$ and reduced to a value before its second argument is expanded, but this would complicate the presentation, and this version can be justified by \Cref{thm:confluent ranking} instead.)
 
 The inequalities required for this to be an antitone partial supermartingale are satisfied with just the same reasoning as in \Cref{ex:unbiased random walk}, therefore this term too is antitone rankable.
+
+Again, this is not actually the correct reduction order, in that $F$ should be applied to its argument $n_i$ and reduced to a value before its second argument is expanded, but this would complicate the presentation, and this version can be justified by \Cref{thm:confluent ranking} instead.
+The reduction strategy implied here should be clear enough (assuming that, where it isn't specified, it just matches $\cbv$), but to be more precise, let $r(F \underline{n_0} (\dots(F\ \underline {n_{k-1}}\ X)\dots)) = @_2^k;\cbv(X)$, where $X$ is not a value or of the form $F\ \underline{n_k}\ Z$ for some non-value $Z$.
+All terms match this pattern for exactly one value of $k \geq 0$, except for values.
+Within $X$, there is necessarily a redex at $\cbv(X)$.
+The only conditions that $@_2^k;\cbv(X)$ is also a position of a redex in the whole term are that if the redex is a $\tsample$, it is not inside a $\lambda$ or $\tY$ on the right of an application, which is true because $\cbv$ never selects a redex inside of a $\lambda$ or $\tY$.
+After $n$ reaches $0$, the term is $F \underline{n_0} (\dots(F\ \underline {n_k}\ \underline x)\dots)$, then the innermost $F$ is evaluated completely in $\cbv$ order, therefore (by assumption) it terminates.
+Because there are no $\tY$-reductions in the evaluation of $F$, it can't reach any other term of the form $F\ \underline{n_k'}\ X$, therefore $k$ never changes until this whole subexpression reaches a number, at which point $k$ decreases by 1 and the next $F$ is evaluated.
 

### 2021-01-14 LICS21/confluent-proofs.tex
Add more explanation for the confluent semantics.

@@
 Also, the number of $\tY$-reductions from $P[\underline n]$ would have been $n^2 + 2n + 1$ instead of $2n + 2$, therefore the expected number of $\tY$ reductions starting from $M$ would be infinite, i.e.~$M$ is not $\tY$-PAST, therefore it is not even rankable.
-It would be possible to define an antitone sparse ranking function for it, but it would have to include terms such as
+It would be possible to define an antitone sparse ranking function for it, but it is much more difficult to construct:
 \begin{align*}
-& \underline{n^n} + \dots + \underline{(n-k_1+1)^n} + \\
-& (\underbrace{\underline{n-k_1} \times \dots \times \underline{n-k_1} \times}_{n-k_2-1} (\Theta[\underline{n-k_1}] (\underline{k_2} - 1))) + \\
-& \Xi[\lambda x. \Theta[x] \underline n] (\underline{n-k_1} - 1) \mapsto \log((n+1)k_1 + k_2)
+& M \mapsto 1.47 \\
+& P[\underline n] \mapsto 1 + \ln((n+1)n + 1) \\
+& \Xi[\lambda x. \Theta[x] \underline n] \underline n \mapsto 1 + \ln((n+1)n + 1) \\
+& \underline{n^n} + \dots + \underline{k_1^n} + \\
+& \Xi[\lambda x. \Theta[x] \underline n] (\underline{k_1} - 1) \mapsto 1 + \ln((n+1)(k_1-1)+1) \\
+& \underline{n^n} + \dots + \underline{(k_1+1)^n} + \Theta[\underline{k_1}] \underline{n} + \\
+& \Xi[\lambda x. \Theta[x] \underline n] (\underline{k_1} - 1) \mapsto 1 + \ln((n+1)k_1 + 1) \\
+& \underline{n^n} + \dots + \underline{(k_1+1)^n} + \\
+& \underbrace{\underline{k_1} \times \dots \times \underline{k_1} \times}_{n-k_2+1} \Theta[\underline{k_1}] (\underline{k_2} - 1) + \\
+& \Xi[\lambda x. \Theta[x] \underline n] (\underline{k_1} - 1) \mapsto 1 + \ln((n+1)k_1 + k_2 + 1) \\
+& \underline{\sum_{k=0}^n k^n} \mapsto 0.
 \end{align*}
-as only one of the several cases, which is significantly more complicated and difficult to use.
 

### 2021-01-14 LICS21/confluent.tex
Add more explanation for the confluent semantics.

@@
 
-When proving almost sure termination in this way, it is necessary to consider which terms a given term may reduce to.
-Sometimes however, the reduction that the programmer has in mind may not be strictly the call-by-value order defined so far; or considering an alternative reduction order may be simpler or more intuitive.
+When proving almost sure termination in this way, it is necessary to consider all of the terms a given term may reduce to, and organise them into a manageable number of cases to assign ranking function values.
+Sometimes however, the reduction that the programmer has in mind may not be strictly the call-by-value order defined so far; or the number of cases necessary for defining a sparse ranking function is excessive because the reduction does not proceed in a neat and orderly manner.
+In cases such as these, it may be much more convenient to instead consider ranking functions with respect to alternative reduction strategies constructed to match the individual term in question, thereby reducing the number of cases needed and simplifying the reachable terms.
+It will be shown in \Cref{thm:confluent ranking} that ranking functions with respect to alternative reduction strategies are sound for proving AST in the usual sense (i.e.~with respect to the standard CBV reduction strategy).
+In order to obtain this result, we prove a relation between the results of using different reduction strategies, by first constructing a confluent trace semantics that is equivalent to the standard semantics as a special case.
 
@@
 \begin{example} Consider the term $M = \tY (\lambda f x.\ \tif{\tsample - \underline{0.5} < 0}{f \, x}{x}) \, \underline 0$, which reduces after a few steps to $N = \tif{\tsample - \underline{0.5}  < 0}{M}{\underline 0}$.
-If the pre-selected sample for $(N,\textsf{if}_1;\underline{-}_1)$ is less than 0.5, $N$ reduces back to $M$, then $N$ again, then the same sample is used the next time, therefore it's an infinite loop, whereas if samples are labelled by reduction sequences, the samples for $M \to^\ast N$ are independent from the samples for $M \to^\ast N \to M \to^\ast N$, and so on.
+If we label samples by just skeletons and positions, and the pre-selected sample for $(\mathit{Sk}(N),\textsf{if}_1;\underline{-}_1)$ is less than 0.5, $N$ reduces back to $M$, then $N$ again, then the same sample is used the next time, therefore it's an infinite loop, whereas if samples are labelled by reduction sequences, the samples for $M \to^\ast N$ are independent from the samples for $M \to^\ast N \to M \to^\ast N$, and so on.
 \end{example}
@@
 
+When constructing an (antitone) sparse ranking function with respect to an alternative reduction strategy, it is intuitively simplest to define the sparse ranking function and the reduction strategy together.
+For each key reachable term where the sparse ranking function is defined, simply decide which sub-term should be reduced next, and how far before the next key reachable term.
+The only restrictions to bear in mind on evaluating sub-terms are that a $\tsample$ may not be evaluated inside a $\lambda$ or $\tY$ on the right of an application, and a $\beta$ redex may not be reduced until its argument is a value.
+This can be seen as applied to \Cref{ex:sum of powers} at the end of \Cref{apx:confluent}.
+
 \medskip

### 2021-01-15 LICS21/antitone-proofs.tex
Cut another proof

@@
 \label{apx:antitone}
+\ACStrict*
+\begin{proof}
+First, as $(Y_n)$ is a supermartingale, $\expect{Y_n} \leq \expect{Y_0}$. 
+Therefore
+\begin{calculation}
+\expect{Y_n \mid T > n}
+   \step[=]{rearranging terms}
+% \dfrac{\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
+% + 
+% \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
+% - 
+% \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
+\dfrac{
+\begin{array}{l}
+\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
++ 
+\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
+\\
+{} - 
+\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}
+\end{array}
+}{\mathbb P[T > n]}
+  \step[=]{definition of conditional expectation}
+\dfrac{\expect{Y_n} - \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
+  \step[\leq]{}
+\dfrac{\expect{Y_n}}{\mathbb P[T > n]}
+%  \step[\leq]{}
+%\dfrac{\expect{Y_0}}{\mathbb P[T > n]}
+\mathbin{=} \dfrac{\expect{Y_0}}{\mathbb P[T > n]}
+\end{calculation}
+
+\emph{Claim}: For all $0 < x \leq 1$, 
+\(
+\mathbb P[T > B_x] \leq x,
+\)
+where $B_x = \left \lceil \dfrac{\expect{Y_0}+1}{x \; \epsilon(\expect{Y_0} \, x^{-1})} \right \rceil$ and $\epsilon : \nnReal \to \pReal$ is the antitone function.
+
+\smallskip
+
+As the convex hull of $\epsilon$ (the greatest convex function less than or equal to it) satisfies all the conditions assumed of $\epsilon$, in addition to being convex, assume wlog that $\epsilon$ is convex.
+
+Assume for a contradiction that $\mathbb P[T > B_x] > x$.
+Then, take $n \leq B_x$. 
+We have
+%\lo{Change of justification of the second equality below: event, not discrete r.v. Recall: expectation conditioning on r.v.~(resp.~event) is a r.v.~(resp.~number).}
+\begin{calculation}
+\expect{Y_n-Y_{n+1}}
+%  \step[=]{}
+\mathbin{=}
+\expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
+  \step[\geq]{}
+\expect{\epsilon(Y_n) \cdot {\bf 1}_{\set{T > n}}}
+%  \step[=]{definition of conditional expectation for a discrete variable}
+%\step[=]{}
+\mathbin{=}
+\mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
+  \step[\geq]{Jensen's inequality}
+\mathbb P[T > n]\ \epsilon(\expect{Y_n \mid T > n})
+  \step[\geq]{proved earlier}
+\mathbb P[T > n]\ \epsilon\Big(\dfrac{\expect{Y_0}}{\mathbb P[T > n]}\Big)
+  \step[>]{assumption, $\mathbb P[T > n] \geq \mathbb P[T > B_x] > x$}
+x \; \epsilon\big(\expect{Y_0} \, x^{-1}\big).
+\end{calculation}
+Therefore, by a telescoping sum
+\begin{align*}
+\expect{Y_{B_x}} 
+&= \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})\\
+&\leq -1 < 0 
+\end{align*}
+%\lo{$\expect{Y_0 - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})}$}
+which is a contradiction, therefore the claim must be true, therefore $P[T > n] \to 0$ as $n \to \infty$, therefore $T < \infty$ a.s.
+\end{proof}
+
 \AntitonePartial*

### 2021-01-15 LICS21/antitone.tex
Introduce Conjecture environment for the completeness conjecture.
Fix typo (g_1 vs g_2)
Cut another proof

@@
 
-\begin{theorem}
+\begin{restatable}{theorem}{ACStrict}
 \label{thm:a-c strict}
@@
 Then $T < \infty$ a.s.
-\end{theorem}
+\end{restatable}
 
-\begin{proof}
-First, as $(Y_n)$ is a supermartingale, $\expect{Y_n} \leq \expect{Y_0}$. 
-Therefore
-\begin{calculation}
-\expect{Y_n \mid T > n}
-   \step[=]{rearranging terms}
-% \dfrac{\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
-% + 
-% \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
-% - 
-% \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
-\dfrac{
-\begin{array}{l}
-\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
-+ 
-\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
-\\
-{} - 
-\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}
-\end{array}
-}{\mathbb P[T > n]}
-  \step[=]{definition of conditional expectation}
-\dfrac{\expect{Y_n} - \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
-  \step[\leq]{}
-\dfrac{\expect{Y_n}}{\mathbb P[T > n]}
-%  \step[\leq]{}
-%\dfrac{\expect{Y_0}}{\mathbb P[T > n]}
-\mathbin{=} \dfrac{\expect{Y_0}}{\mathbb P[T > n]}
-\end{calculation}
-
-\emph{Claim}: For all $0 < x \leq 1$, 
-\(
-\mathbb P[T > B_x] \leq x,
-\)
-where $B_x = \left \lceil \dfrac{\expect{Y_0}+1}{x \; \epsilon(\expect{Y_0} \, x^{-1})} \right \rceil$ and $\epsilon : \nnReal \to \pReal$ is the antitone function.
-
-\smallskip
-
-As the convex hull of $\epsilon$ (the greatest convex function less than or equal to it) satisfies all the conditions assumed of $\epsilon$, in addition to being convex, assume wlog that $\epsilon$ is convex.
-
-Assume for a contradiction that $\mathbb P[T > B_x] > x$.
-Then, take $n \leq B_x$. 
-We have
-%\lo{Change of justification of the second equality below: event, not discrete r.v. Recall: expectation conditioning on r.v.~(resp.~event) is a r.v.~(resp.~number).}
-\begin{calculation}
-\expect{Y_n-Y_{n+1}}
-%  \step[=]{}
-\mathbin{=}
-\expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
-  \step[\geq]{}
-\expect{\epsilon(Y_n) \cdot {\bf 1}_{\set{T > n}}}
-%  \step[=]{definition of conditional expectation for a discrete variable}
-%\step[=]{}
-\mathbin{=}
-\mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
-  \step[\geq]{Jensen's inequality}
-\mathbb P[T > n]\ \epsilon(\expect{Y_n \mid T > n})
-  \step[\geq]{proved earlier}
-\mathbb P[T > n]\ \epsilon\Big(\dfrac{\expect{Y_0}}{\mathbb P[T > n]}\Big)
-  \step[>]{assumption, $\mathbb P[T > n] \geq \mathbb P[T > B_x] > x$}
-x \; \epsilon\big(\expect{Y_0} \, x^{-1}\big).
-\end{calculation}
-Therefore, by a telescoping sum
-\begin{align*}
-\expect{Y_{B_x}} 
-&= \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})\\
-&\leq -1 < 0 
-\end{align*}
-%\lo{$\expect{Y_0 - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})}$}
-which is a contradiction, therefore the claim must be true, therefore $P[T > n] \to 0$ as $n \to \infty$, therefore $T < \infty$ a.s.
-\end{proof}
-
-Note that it is essential in this proof that $\epsilon$ be defined for all $\mathbb R_{\geq 0}$, not just the values that $(Y_n)_{n \geq 0}$ actually takes (or at least if $(Y_n)_{n \geq 0}$ is uniformly bounded, $\epsilon$ must be positive at the supremum as well as the actual values).
+It is essential in this theorem that $\epsilon$ be defined for all $\mathbb R_{\geq 0}$, not just the values that $(Y_n)_{n \geq 0}$ actually takes (or at least if $(Y_n)_{n \geq 0}$ is uniformly bounded, $\epsilon$ must be positive at the supremum as well as the actual values).
 
@@
 \label{ex:unbiased random walk}
-For $M_2$, define the function $g_1 : \nnReal \to \nnReal$ by $g_1(x) = \ln(x+1) + 1$.
+For $M_2$, define the function $g_2 : \nnReal \to \nnReal$ by $g_2(x) = \ln(x+1) + 1$.
 Using shorthand 

### 2021-01-15 LICS21/applications.tex
Cut another proof

--- 2021-01-10 LICS21/applications.tex
@@
 \label{sec:applications}
-Although the definition of potential positions in a term in \Cref{sec:confluent} was intended simply to define a variant of sampling semantics that has a restricted version of the Church-Rosser property, it is likely to also be applicable to probabilistic inference algorithms as well.
+Although the definition of potential positions in a term in \Cref{sec:confluent} was intended simply to define a variant of trace semantics that has a restricted version of the Church-Rosser property, it is likely to also be applicable to inference algorithms (for probabilistic programs) as well.
 
-In the Metropolis-Hastings inference algorithm, the program is executed, with a random value being chosen each time a $\tsample$ redex is evaluated.
+%\lo{What you described in the following paragraph is not the Metropolis-Hastings algorithm (for MH algorithm, see e.g.~Algorithm 1, Slide 18 of 5-MCMC-part2.pdf \footnote{\url{https://www.cs.ox.ac.uk/teaching/materials20-21/SPP/Lectures/5-MCMC-part2.pdf}, \cite{Ong20}}), but a version of component-wise MH adapted to (models \emph{qua}) probabilistic programs (see Slide 36 onwards, 6-operational semantics.pdf \footnote{\url{https://www.cs.ox.ac.uk/teaching/materials20-21/SPP/Lectures/6-Operational-Semantics-21Nov20.pdf}, \cite{Ong20} \label{fnote:lmh}}).}
+
+In lightweight Metropolis-Hastings (LMH) or single-site MH (originally suggested by \cite{wingate2011lightweight}; see also \cite{rainforth2017Automating}),
+the program is executed, with a random value being chosen each time a $\tsample$ redex is evaluated.
 The trace of samples used is recorded, then for the next step, it is modified slightly before the program is run again.
@@
 If the labels of samples in the trace correspond closely to the roles that the samples play in the program execution, this will increase the efficiency of the inference algorithm.
-In \cite{wingate2011lightweight}, for example, something rather like a stack trace is used, although this labelling scheme turns out to have various problems \cite{kiselyov2016problems}. 
+In \cite{wingate2011lightweight}, for example, something rather like a stack trace is used, although this assdressing scheme turns out to have various problems \cite{kiselyov2016problems,Hur2015a}. 
 Potential positions could also be used for a similar purpose, although in practice, a more compact representation than an entire skeletal reduction sequence would be necessary. 
+%\lo{It would be very interesting (and rewarding, I believe) to work out the details.
+There is a formalisation of LMH for Idealised Anglican in \Cref{fnote:lmh}.
+SPCF is a much nicer probabilistic programming language (in purified / idealised form).}
 Selecting an appropriate addressing scheme can also be important in the context of variational inference \cite[\S 6.2.2]{paige2016automatic}.
@@
 \medskip
-Another possible application for the confluent trace semantics would be as a justification, within a sampling-style semantics, of some forms of equational reasoning.
+
+Another possible application for the confluent trace semantics would be as a justification, within a trace-style semantics, of some forms of equational reasoning. 
 By introducing reduction steps that go backwards to terms not reachable from the original term, it may in some cases be simplified, or equivalences between terms may be proved, by appealing to confluence results such as \Cref{churchRosser} and \Cref{thm:CbvIsTerminatingest}.
+\lo{Can you provide an example (and present it in the appendix)?} \akr{I don't have any compelling example for this in mind.}

### 2021-01-15 LICS21/confluent.tex
Referenced an example.

@@
 When proving almost sure termination in this way, it is necessary to consider all of the terms a given term may reduce to, and organise them into a manageable number of cases to assign ranking function values.
-Sometimes however, the reduction that the programmer has in mind may not be strictly the call-by-value order defined so far; or the number of cases necessary for defining a sparse ranking function is excessive because the reduction does not proceed in a neat and orderly manner.
-In cases such as these, it may be much more convenient to instead consider ranking functions with respect to alternative reduction strategies constructed to match the individual term in question, thereby reducing the number of cases needed and simplifying the reachable terms.
+Sometimes however, the reduction that the programmer has in mind may not be strictly the call-by-value order defined so far; 
+or the number of cases necessary for defining a sparse ranking function is excessive because the reduction does not proceed in a neat and orderly manner. 
+In cases such as these \changed[lo]{(see e.g.~\cref{ex:sum of powers})}, it may be much more convenient to instead consider ranking functions with respect to alternative reduction strategies constructed to match the individual term in question,
+thereby reducing the number of cases needed and simplifying the reachable terms. 
 It will be shown in \Cref{thm:confluent ranking} that ranking functions with respect to alternative reduction strategies are sound for proving AST in the usual sense (i.e.~with respect to the standard CBV reduction strategy).
-In order to obtain this result, we prove a relation between the results of using different reduction strategies, by first constructing a confluent trace semantics that is equivalent to the standard semantics as a special case.
+In order to obtain this result, we prove a relation between the results of using different reduction strategies, 
+by first constructing a confluent trace semantics that is equivalent to the standard semantics as a special case.
 

### 2021-01-15 LICS21/related.tex

@@
 
-\lo{TODO. Related work (.75 column). Discuss \cite{DBLP:conf/birthday/BlancLM05}.}
+\lo{TODO.
+Related work (.75 column).
+Discuss \cite{DBLP:conf/birthday/BlancLM05}.} \akr{I think that that labelling scheme is not really that related.
+Would a note like "This is another scheme for labelling positions in reachable terms, therefore it appears superficially similar, but it's not actually that related." actually be worthwhile?
+I suppose it could at least reassure readers that we're aware of that paper and it isn't the same thing, but I don't know how likely readers are to actually draw the relation in the first place.}
 
@@
 
-\lo{Kobayashi et al.: PHORS are strictly weaker in expressivity \cite{DBLP:conf/lics/KobayashiLG1}.}
+\lo{Kobayashi et al.: PHORS are strictly weaker in expressivity \cite{DBLP:conf/lics/KobayashiLG19}.}
+
+The main theorem in \cite{DBLP:journals/pacmpl/McIverMKK18} is similar to our \ref{thm:antitone rankable implies termination}, but in an imperative language.
+While we require that the antitone ranking function decreases for every $\tY$-reduction step, they require that it decreases for every iteration of a certain while loop (the ranking function in that case is defined in the context of a particular loop), which is similar to the recursion construct $\tY$ but limited to tail recursion.
+The difference in the exact progress condition is not significant.
+Their progress condition, that there are antitone functions $p$ and $d$ such that the ranking function decreases from $x$ by at least $d(x)$ with probability at least $p(x)$, would not have worked directly because it does not cope as easily with the fact that the ranking function may change in value at the intermediate non-$\tY$ reduction steps (provided it does not increase in expectation).
+However, it is easy enough to convert a ranking function that satisfies either progress condition to the other.
+
+The confluent semantics, and the results based on it, is new in the setting of a functional language, because an imperative language does not have any equivalent of other redexes, or a similar nondeterministic structure.
+The sparse ranking function construction is also more useful in a functional language (although it would be possible to give an equivalent in an imperative language), because in an imperative language where the order of execution is more rigid, the iterations of the while loop provide a natural set of checkpoints where it is reasonable to give values of the ranking function.
 

### 2021-01-15 LICS21/confluent-proofs.tex

@@
 Let the position of the reduction $M \to N$ be $\beta$.
-\akr{TODO: fill in the details that are missing here. It shouldn't be too hard, it's just a bit sketchy at present.} 
 The rearrangement of the reduction sequence $Y_1 \sim_{c,M} X$ consists of moving the reduction at $\beta$ down past the other reductions in $Y_1$, possibly duplicating or deleting it in the process, but not affecting the positions or the correct order for any of the other reductions.

### 2021-01-16 LICS21/related.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper Related work

--- 2021-01-12 LICS21/related.tex
@@
 \section{Related work and further directions}
+\label{sec:related}
 
-\lo{TODO. Related work (.75 column). Discuss \cite{DBLP:conf/birthday/BlancLM05}.}
+\lo{TODO.
+Related work (.75 column).
+Discuss \cite{DBLP:conf/birthday/BlancLM05}.} \akr{I think that that labelling scheme is not really that related.
+Would a note like ``This is another scheme for labelling positions in reachable terms, therefore it appears superficially similar, but it's not actually that related.'' actually be worthwhile?
+I suppose it could at least reassure readers that we're aware of that paper and it isn't the same thing, but I don't know how likely readers are to actually draw the relation in the first place.}
 
-\lo{Our results specialise to proving termination of non-random lambda calculus, which is much studied.
-Our approach is closest in spirit to Jones' work on size-change argument.
-Dal Lago's work builds on this.
-We think that martingales are a good framework for generalising / unifying with size-change work.}
+Most methods for proving AST of programs do not support continuous distributions \cite{DBLP:journals/toplas/LagoG19,DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/lics/KobayashiLG19,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05}.
+The few that do \cite{DBLP:conf/popl/FioritiH15,DBLP:conf/pldi/ChenH20,DBLP:journals/toplas/ChatterjeeFNH18} are for first-order imperative programs.
 
-\lo{Other approaches, mainly based on types, are complementary to ours.}
+A fortiori, our results specialise to a method for proving termination of non-random lambda calculus. 
+Thus restricted, our approach is closest in spirit to Jones' work on size-change termination \cite{DBLP:journals/lmcs/JonesB08,DBLP:conf/aplas/SereniJ05}. 
+Dal Lago and Grellois \cite{DBLP:journals/toplas/LagoG19} extend this to programs with discrete distributions, but their method is restricted to first-order affine recursion.
+Breuvart and Dal Lago \cite{DBLP:conf/ppdp/BreuvartL18} develop systems of intersection types from which the termination probability of higher-order programs (with discrete distributions) can be inferred from (infinitely many) type derivations.
 
-\lo{Kobayashi et al.: PHORS are strictly weaker in expressivity.}
+Kobayashi et al.~\cite{DBLP:conf/lics/KobayashiLG19} show that the termination probability of CbN order-$n$ probabilistic recursion schemes ($n$-PHORS) can be obtained as the least fixpoint of suitable order-$(n-1)$ fixpoint equations, which are solvable using Kleene fixpoint iteration.
+By contrast, our approach works for programs with continuous distributions. 
+It is worth noting that (order-$n$) PHORS is strictly less expressive then (order-$n$) CbN PPCF (because the underlying recursion schemes are not Turing complete). 
+Some interesting PPCF terms such as \cref{ex:raven complex,ex:higher-order recursion}
+%\refExample{complexExample} 
+are not expressible as PHORS.
 
-\label{sec:related}
-We have not yet been able to prove \Cref{conj:antitone} which would imply that \Cref{thm:antitone rankable implies termination} is almost complete.
+The main result in the work \cite{DBLP:journals/pacmpl/McIverMKK18} by McIver et al.~is similar to our \cref{thm:antitone rankable implies termination}, but in an imperative language.
+While we require that the antitone ranking function decreases for every $\tY$-reduction step, they require that it decreases for every iteration of a certain while loop (the ranking function in that case is defined in the context of a particular loop), which is similar to $\tY$ but limited to tail recursion.
+The difference in the exact progress condition is not significant. 
+\akr{Their progress condition, that there are antitone functions $p$ and $d$ such that the ranking function decreases from $x$ by at least $d(x)$ with probability at least $p(x)$, would not have worked directly because it does not cope as easily with the fact that the ranking function may change in value at the intermediate non-$\tY$ reduction steps (provided it does not increase in expectation).
+However, it is easy enough to convert a ranking function that satisfies either progress condition to the other.}
+\lo{We may have to omit this.}
+
+The confluent semantics, and the results based on it, is new in the setting of a functional language, because an imperative language does not have any equivalent of other redexes, or a similar nondeterministic structure.
+The sparse ranking function construction is also more useful in a functional language (although it would be possible to give an equivalent in an imperative language), because in an imperative language where the order of execution is more rigid, the iterations of the while loop provide a natural set of checkpoints where it is reasonable to give values of the ranking function.
+
+\paragraph*{Further directions} We have not yet been able to prove \Cref{conj:antitone} which would imply that \Cref{thm:antitone rankable implies termination} is almost complete.
 \begin{conjecture}

### 2021-01-16 LICS21/SPCF.tex
Related work Introduction: ongoing Title Various sections: minor formatting so save space

@@
 
-The one-step reduction is given by the function $\red : \Lambda^0 \times \entrosp \to \Lambda^0 \times \entrosp$ where
+\changed[lo]{The one-step reduction of the \emph{trace-based} (or \emph{sampling-style}) \emph{operational semantics} \cite{DBLP:journals/jcss/Kozen81,DBLP:conf/icfp/BorgstromLGS16,DBLP:conf/esop/CulpepperC17}} is given by the function $\red : \Lambda^0 \times \entrosp \to \Lambda^0 \times \entrosp$ where
 \begin{equation*}
@@
 
-The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid M \text{ terminates at } s \}$ is measurable.
-We say that a term $M$ is \emph{almost surely terminating} (AST) just if $\mu_{\entrosp}(\{s \mid M \text{ terminates at } s\}) = 1$.
+The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid \hbox{$M$ terminates at $s$}\}$ is measurable.
+We say that a term $M$ is \emph{almost surely terminating} (AST) just if $\mu_{\entrosp}(\{s \mid \hbox{$M$ terminates at $s$}\}) = 1$.
 
@@
 which generates a geometric distribution, terminates on the set $\entrosp \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas 
-\[
+\(
 \tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \underline 0}
-\] 
+\), 
 which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.
@@
 \label{rem:score}
-This definition of almost sure termination is equivalent to that given in \cite{MakOPW21} (although the program semantics is stated in a slightly different way), except for the presence of a $\tscore(r)$ construct for soft conditioning.
-The $\tscore(r)$ construct is not relevant to termination except that it fails if its argument is less than or equal to zero, which allows computations to fail after finitely many steps. 
+Our definition of AST is equivalent to that given in \cite{MakOPW21} (although the program semantics is stated in a slightly different way), except for the presence of a $\tscore(r)$ construct for soft conditioning.
+The $\tscore(r)$ construct is irrelevant to termination except that it fails if its argument is less than or equal to zero, which allows computations to fail after finitely many steps. 
 If such a thing were added, it would merely require an additional side condition to our main theorems, that no failing term was reachable (or, in the case of the confluent semantics based results, no term where the reduction strategy is not defined is reachable via the reduction strategy unless it's a value).

### 2021-01-16 LICS21/antitone.tex
Related work Introduction: ongoing Title Various sections: minor formatting so save space

@@
 \paragraph*{Completeness}
-It seems very likely that this method is complete in the case that every reachable term is AST, but we have been unable to actually prove this.
+It seems very likely that this method is complete in the case that every reachable term is AST (\cref{conj:antitone}), but we have been unable to actually prove this.
 It is certainly at least capable of proving the termination of terms which terminate arbitrarily slowly (in the sense that there is no similar limitation to Theorem \ref{thm:rankable implies termination}, which can only prove termination of $\tY$-PAST terms).

### 2021-01-16 LICS21/confluent.tex
Related work Introduction: ongoing Title Various sections: minor formatting so save space

--- 2021-01-14 LICS21/confluent.tex
@@
 When proving almost sure termination in this way, it is necessary to consider all of the terms a given term may reduce to, and organise them into a manageable number of cases to assign ranking function values.
-Sometimes however, the reduction that the programmer has in mind may not be strictly the call-by-value order defined so far; or the number of cases necessary for defining a sparse ranking function is excessive because the reduction does not proceed in a neat and orderly manner.
-In cases such as these, it may be much more convenient to instead consider ranking functions with respect to alternative reduction strategies constructed to match the individual term in question, thereby reducing the number of cases needed and simplifying the reachable terms.
+Sometimes however, the reduction that the programmer has in mind may not be strictly the call-by-value order defined so far; 
+or the number of cases necessary for defining a sparse ranking function is excessive because the reduction does not proceed in a neat and orderly manner. 
+In cases such as these (e.g.~\cref{ex:sum of powers}), it may be much more convenient to instead consider ranking functions with respect to alternative reduction strategies constructed to match the individual term in question,
+thereby reducing the number of cases needed and simplifying the reachable terms. 
 It will be shown in \Cref{thm:confluent ranking} that ranking functions with respect to alternative reduction strategies are sound for proving AST in the usual sense (i.e.~with respect to the standard CBV reduction strategy).
-In order to obtain this result, we prove a relation between the results of using different reduction strategies, by first constructing a confluent trace semantics that is equivalent to the standard semantics as a special case.
+In order to obtain this result, we prove a relation between the results of using different reduction strategies, 
+by first constructing a confluent trace semantics that is equivalent to the standard semantics as a special case.
 
@@
 For example, let 
-\[
+\(
 M = \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})
-\quad
+\)
+and
+\(
 \alpha =\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2
-\]
+\)
 then 

### 2021-01-16 LICS21/intro.tex
Related work Introduction: ongoing Title Various sections: minor formatting so save space

@@
 Various theorems about probabilistic programs rely on the assumption that the program terminates almost surely.
-One possible approach to prove almost sure termination (AST) is to find some variant on the program state that decreases on average sufficiently quickly that it must at some point reach 0, at which point the program terminates.
-In other words, the program's behavior is used to define an associated (ranking) supermartingale.
+One possible approach to prove almost sure termination (AST) is to find some variant on the program state, called \emph{ranking function}, that decreases on average sufficiently quickly that it must at some point reach 0, at which point the program terminates.
+In other words, the program's behavior is used to define an associated (ranking) \emph{supermartingale}.
 Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \cite{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18}.
@@
 
-The language PPCF used in this paper is simply typed, with an explicit recursion primitive, $\tY$.
+The language PPCF used in this paper is simply typed, with sampling from a continuous distribution, and an explicit recursion primitive, $\tY$. 
 As the type system already constrains the terms enough to force termination in the absence of the recursion construct \cite{tait1967,BarendregtDS2010}, it is only the $\tY$-reduction steps that must be counted, making defining ranking functions somewhat easier.
 
-The notion of a \emph{sparse ranking function} is introduced to make constructing ranking functions more convenient.
+We introduce the notion of \emph{sparse ranking function} to make constructing ranking functions more convenient.
 Most of the individual execution steps of a typical program are trivial and easy to mentally skip over.
@@
 A ranking function (or sparse ranking function) provides a bound on the expected number of $\tY$-reduction steps before the program terminates, therefore ranking functions cannot be constructed for terms whose expected number of $\tY$-reduction steps is infinite, such as the simplest implementation of the 1D unbiased random walk.
-This restriction can be removed by generalising ranking functions to antitone ranking functions, which rather than having to decrease by a constant amount for each $\tY$-reduction step, may decrease by a variable amount, depending on the value of the ranking function.
+This restriction can be removed by generalising ranking functions to \emph{antitone ranking functions}, which rather than having to decrease by a constant amount for each $\tY$-reduction step, may decrease by a variable amount, depending on the value of the ranking function.
 This method is capable of proving the almost sure termination of programs which terminate arbitrarily slowly.
-Our progress condition is equivalent to that given in \cite{DBLP:journals/pacmpl/McIverMKK18} (by applying a transformation to the ranking function), but needs to be different in order to be compatible with the presence of additional intermediate reduction steps where the ranking function is not required to decrease at all.
+Our progress condition is equivalent to that given by McIver et al.~\cite{DBLP:journals/pacmpl/McIverMKK18} (by applying a transformation to the ranking function), but needs to be different in order to be compatible with the presence of additional intermediate reduction steps where the ranking function is not required to decrease at all.
 
 Basic (non-random) lambda calculus has the very useful Church-Rosser property, which implies (among other things) that even if execution of a program starts in a different order, it will still reach the same normal form eventually (assuming it does reach a normal form).
-Probabilistic lambda calculus does not have this property, because random choices may be duplicated, and evaluating the same sub-term multiple times can yield different results.
+Probabilistic lambda calculus does not have this property, because random choices may be duplicated, and evaluating the same subterm multiple times can yield different results. 
 However, with a restricted set of reduction strategies, Church-Rosserness may be regained. 
-A novel addressing scheme for the possible random choices in a program's execution is introduced, which ensures that the same random choices are taken at corresponding positions in alternative reduction sequences, so that the same eventual result can be reached.
+We introduce a novel addressing scheme for the possible random choices in a program's execution, which ensures that the same random choices are taken at corresponding positions in alternative reduction sequences, so that the same eventual result can be reached.
 This is then used to prove yet another extension to the ranking function theorem, that ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and rankability in this sense still imples almost sure termination.
@@
 
-We present the syntax and trace (or sampling) semantics of PPCF in \Cref{sec:PPCF}.
+We present the syntax and trace semantics of PPCF in \Cref{sec:PPCF}.
 In \Cref{sec:supermartingales}, we show that ranking functions on terms induce supermartingales, which form the basis of a sound method for proving AST.

### 2021-01-16 LICS21/ranking.tex
Related work Introduction: ongoing Title Various sections: minor formatting so save space

@@
 Although rankability implies almost sure termination, the converse does not hold in general. For example,
-\begin{equation}
+\begin{equation*}
 \tif{{-}\tsample < 0}{\underline{0}}{(\tY \lambda x. x) \, \underline 0}
 \label{ex:0 probability reachable}
-\end{equation}
+\end{equation*}
 terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x. x) \, \underline 0$ is reachable, although that has probability 0. 
@@
 In particular, the expected number of $\tY$-reduction steps from each reachable term is a ranking function. 
-However, note that a finite number of expected $\tY$-reduction steps does not necessarily imply a finite number of expected total reduction steps.
+However, a finite number of expected $\tY$-reduction steps does not necessarily imply a finite number of expected total reduction steps.
 

### 2021-01-16 LICS21/related.tex
Related work Introduction: ongoing Title Various sections: minor formatting so save space

@@
 
-Most methods for proving AST of programs do not support continuous distributions \cite{DBLP:journals/toplas/LagoG19,DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/lics/KobayashiLG19,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05}.
+Most methods for proving AST of programs in the literature do not support continuous distributions \cite{DBLP:journals/toplas/LagoG19,DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/lics/KobayashiLG19,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05}.
 The few that do \cite{DBLP:conf/popl/FioritiH15,DBLP:conf/pldi/ChenH20,DBLP:journals/toplas/ChatterjeeFNH18} are for first-order imperative programs.
@@
 Thus restricted, our approach is closest in spirit to Jones' work on size-change termination \cite{DBLP:journals/lmcs/JonesB08,DBLP:conf/aplas/SereniJ05}. 
-Dal Lago and Grellois \cite{DBLP:journals/toplas/LagoG19} extend this to programs with discrete distributions, but their method is restricted to first-order affine recursion.
+Dal Lago and Grellois \cite{DBLP:journals/toplas/LagoG19} extend this to programs with discrete distributions, but their method is limited to affine recursion of first-order functions.
 Breuvart and Dal Lago \cite{DBLP:conf/ppdp/BreuvartL18} develop systems of intersection types from which the termination probability of higher-order programs (with discrete distributions) can be inferred from (infinitely many) type derivations.
 
-Kobayashi et al.~\cite{DBLP:conf/lics/KobayashiLG19} show that the termination probability of CbN order-$n$ probabilistic recursion schemes ($n$-PHORS) can be obtained as the least fixpoint of suitable order-$(n-1)$ fixpoint equations, which are solvable using Kleene fixpoint iteration.
+Kobayashi et al.~\cite{DBLP:conf/lics/KobayashiLG19} show that the termination probability of order-$n$ probabilistic recursion schemes ($n$-PHORS) can be obtained as the least fixpoint of suitable order-$(n-1)$ fixpoint equations, which are solvable by Kleene fixpoint iteration.
 By contrast, our approach works for programs with continuous distributions. 
-It is worth noting that (order-$n$) PHORS is strictly less expressive then (order-$n$) CbN PPCF (because the underlying recursion schemes are not Turing complete). 
-Some interesting PPCF terms such as \cref{ex:raven complex,ex:higher-order recursion}
+Note that $n$-PHORS is strictly less expressive then order-$n$ CbN PPCF (because the underlying recursion schemes are not Turing complete):
+PPCF terms in \cref{ex:raven complex,ex:higher-order recursion}
 %\refExample{complexExample} 
-are not expressible as PHORS.
+are not definable as PHORS.
 
-The main result in the work \cite{DBLP:journals/pacmpl/McIverMKK18} by McIver et al.~is similar to our \cref{thm:antitone rankable implies termination}, but in an imperative language.
+The main result by McIver et al.~\cite{DBLP:journals/pacmpl/McIverMKK18} is similar to our \cref{thm:antitone rankable implies termination}, but in an imperative language.
 While we require that the antitone ranking function decreases for every $\tY$-reduction step, they require that it decreases for every iteration of a certain while loop (the ranking function in that case is defined in the context of a particular loop), which is similar to $\tY$ but limited to tail recursion.
@@
 However, it is easy enough to convert a ranking function that satisfies either progress condition to the other.}
-\lo{We may have to omit this.}
+\lo{Can you replace the highlighted by a brief comment on their discussion of completeness?}
 
@@
 \paragraph*{Further directions} We have not yet been able to prove \Cref{conj:antitone} which would imply that \Cref{thm:antitone rankable implies termination} is almost complete.
-\begin{conjecture}
+\begin{conjecture}[Completeness]
 \label{conj:antitone}
-Every term is antitone rankable if all of the terms reachable from it are AST.
+Every PPCF term is antitone rankable if all terms reachable from it are AST.
 \end{conjecture}

### 2021-01-16 LICS21/SPCF.tex
Introduction:ongoing Reformatted last antitone sparse ranking function in appendix

@@
 
-The language PPCF is a simply-typed lambda calculus with sampling of real numbers from the closed interval $[0,1]$. 
+The language PPCF is a CbV version of PCF with sampling of real numbers from the closed interval $[0,1]$ \cite{Ehrhard2018c,DBLP:journals/pacmpl/EhrhardPT18,MakOPW21}.
 %\changed[lo]{following \cite{MakOPW21} except the score function (\Cref{rem:score})}. 
@@
 We write $\to^\ast$ for the reflexive, transitive closure of $\to$.
-
-Every closed well-typed term either is a value or reduces to another term.
+Every closed term either is a value or reduces to another term.
 
@@
 Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $\mathit{Sk}(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$.
-Following \cite{DBLP:conf/icfp/BorgstromLGS16}, the $\sigma$-algebra on $\Lambda$ is defined by identifying $\Lambda$ with $\bigcup_{m\geq 0} \big(\mathsf{Sk}_m \times \Real^m \big)$, where $\mathsf{Sk}_m$ is the set of skeletons containing $m$ occurrences of $\skeletonPlaceholder$.
+Following \cite{DBLP:conf/icfp/BorgstromLGS16,DBLP:conf/lics/StatonYWHK16,DBLP:journals/pacmpl/EhrhardPT18}, the $\sigma$-algebra on $\Lambda$ is defined by identifying $\Lambda$ with $\bigcup_{m\geq 0} \big(\mathsf{Sk}_m \times \Real^m \big)$, where $\mathsf{Sk}_m$ is the set of skeletons containing $m$ occurrences of $\skeletonPlaceholder$.
 Thus identified, we give $\Lambda$ the countable disjoint union topology of the product topology of the discrete topology on $\mathsf{Sk}_m$ and the standard topology on $\Real^m$, and take the corresponding Borel $\sigma$-algebra.
@@
 \label{rem:score}
-Our definition of AST is equivalent to that given in \cite{MakOPW21} (although the program semantics is stated in a slightly different way), except for the presence of a $\tscore(r)$ construct for soft conditioning.
-The $\tscore(r)$ construct is irrelevant to termination except that it fails if its argument is less than or equal to zero, which allows computations to fail after finitely many steps. 
+Our definition of AST is equivalent to that given in \cite{MakOPW21} (although the program semantics is stated in a slightly different way), except for the presence of a $\tscore$ construct for soft conditioning.
+The $\tscore$ construct is irrelevant to termination except that it fails if its argument is negative, thus allowing computations to fail after finitely many steps. 
 If such a thing were added, it would merely require an additional side condition to our main theorems, that no failing term was reachable (or, in the case of the confluent semantics based results, no term where the reduction strategy is not defined is reachable via the reduction strategy unless it's a value).

### 2021-01-16 LICS21/antitone.tex
Introduction:ongoing Reformatted last antitone sparse ranking function in appendix

@@
 \fi
-\begin{restatable}{theorem}{AntitonePartial}  \label{thm:antitone partial implies rankable}
+\begin{restatable}[Sparse function]{theorem}{AntitonePartial}  \label{thm:antitone partial implies rankable}
   Every antitone sparse ranking function is a restriction of an antitone ranking function.

### 2021-01-16 LICS21/confluent.tex
Introduction:ongoing Reformatted last antitone sparse ranking function in appendix

@@
 If terms rather than skeletons were used to label samples, the set of modified traces where $A[\tsample]$ terminates would be
-\begin{align*}
-\bigcup_{r \in [0,1]} \Big\{s \mid\ & s(A[\tsample], \mathsf{if}_1;\underline{-}_1;@_1;\mathsf{if}_1) = r, \\
-& s(A[\tsample] \to A[\underline r] \to^* B, \mathsf{if}_1 ; \underline{-}_1) > 0.5 \Big\}.
-\end{align*}
-
+\[
+\begin{array}{l}
+\bigcup_{r \in [0,1]} \big\{s \mid\ s(A[\tsample], \mathsf{if}_1;\underline{-}_1;@_1;\mathsf{if}_1) = r, \\
+\qquad\qquad\qquad s(A[\tsample] \to A[\underline r] \to^* B, \mathsf{if}_1 ; \underline{-}_1) > 0.5 \big\}.
+\end{array}
+\]
 \iffalse
@@
 The relation $\sim$ is defined as the union of the minimal symmetric relations $\sim_p$ (``$p$'' for parent-child) and $\sim_c$ (``$c$'' for cousin) satisfying
-\begin{enumerate}
+\begin{asparaenum}[(i)]
     \item If $N$ reduces to $O$ with the redex at position $\alpha$, and $\beta$ is a position in $N$ disjoint from $\alpha$, then $(N,\beta) \sim_p (O,\beta)$.
@@
     (in which case $O_1$ and $O_2$ are equal as skeletons, but with different reduction sequences), $O_1'$ and $O_2'$ are the results of applying some reduction sequence to each of $O_1$ and $O_2$ (the same reductions in each case, which is always possible because they're equal terms), and $\delta$ is a position in $O_1'$ (or equivalently $O_2'$), then $(O_1',\delta) \sim_c (O_2',\delta)$.
-\end{enumerate}
+\end{asparaenum}
 The $\sim_c$-rules are illustrated in \Cref{sim-diagram}.
@@
 
-\begin{theorem} \label{thm:confluent ranking}
-If a closed PPCF term $M$ has a ranking function, sparse ranking function, antitone ranking function or antitone sparse ranking function with respect to a reduction strategy $r$, then $M$ is AST with respect to $r$, and AST.
+\begin{theorem}[Confluent ranking] \label{thm:confluent ranking}
+If a closed PPCF term $M$ has a ranking function, sparse ranking function, antitone ranking function or antitone sparse ranking function w.r.t.~a reduction strategy $r$, then $M$ is AST w.r.t.~$r$, and AST.
 \end{theorem}
@@
 \begin{restatable}{theorem}{confluentNoStronger}
+\label{thm:confluentNoStronger}
 For any closed term $M$ and reduction strategy $r$ on $M$, if every term in $\mathit{Rch}_r(M)$ is AST, then every term in $\mathit{Rch}(M)$ is AST.

### 2021-01-16 LICS21/intro.tex
Introduction:ongoing Reformatted last antitone sparse ranking function in appendix

@@
 
+\medskip
+
 Various theorems about probabilistic programs rely on the assumption that the program terminates almost surely. 
@@
 
-The language PPCF used in this paper is simply typed, with sampling from a continuous distribution, and an explicit recursion primitive, $\tY$. 
-As the type system already constrains the terms enough to force termination in the absence of the recursion construct \cite{tait1967,BarendregtDS2010}, it is only the $\tY$-reduction steps that must be counted, making defining ranking functions somewhat easier.
+The language PPCF used in this paper is simply typed, with random continuous sampling, and an explicit recursion primitive, $\tY$. 
+We define a \emph{ranking function} on a term $M$ to be a non-negative measurable function on the reachable terms from $M$ whose expected value decreases or is unchanged as the term is reduced.
+Because the type system already constrains the terms enough to force termination in the absence of the recursion construct \cite{tait1967,BarendregtDS2010}, it is only the $\tY$-reduction steps that must be counted, making defining ranking functions somewhat easier.
+Using supermartingales and Doob's (iterated) Optional Sampling Theorem, we show the soundness of rankability (\cref{thm:rankable implies termination}): if a PPCF term has a ranking function then it is AST.
 
-We introduce the notion of \emph{sparse ranking function} to make constructing ranking functions more convenient.
+To define a ranking function, one has to try to organise the set of reachable terms (of the program being analysed) into a manageable number of syntactic cases. 
+Unfortunately (for any fixed reduction strategy) there are programs whose reachable terms span such a large number of cases that it would be extremely difficult to analyse by hand, if at all possible.
+How can the construction of ranking functions be made possible or easier?
+
+Our first answer is the notion of \emph{sparse ranking function}. 
 Most of the individual execution steps of a typical program are trivial and easy to mentally skip over.
 Sparse ranking functions can be defined only for those points in the execution of a program which are semantically important, while all of the other intermediate steps can be ignored.
+Yet, they are no less efficacious for proving AST (\cref{thm:partial implies rankable}): every sparse ranking function is a restriction of a ranking function.
 This also makes the ranking function method of proving AST more compatible with syntactic sugar, because the intermediate reduction steps implicit in the simplified notation can be ignored.
@@
 This restriction can be removed by generalising ranking functions to \emph{antitone ranking functions}, which rather than having to decrease by a constant amount for each $\tY$-reduction step, may decrease by a variable amount, depending on the value of the ranking function.
-This method is capable of proving the almost sure termination of programs which terminate arbitrarily slowly.
-Our progress condition is equivalent to that given by McIver et al.~\cite{DBLP:journals/pacmpl/McIverMKK18} (by applying a transformation to the ranking function), but needs to be different in order to be compatible with the presence of additional intermediate reduction steps where the ranking function is not required to decrease at all.
+Thanks to this feature, the antitone ranking function method is capable of handling programs which terminate arbitrarily slowly. \lo{TODO: Reference some examples.}
+We show that this method is sound for proving AST (\cref{thm:antitone rankable implies termination}), 
+and it also enjoys a sparse function theorem (\cref{thm:antitone partial implies rankable}).
+\akr{Our progress condition is equivalent to that given by McIver et al.~\cite{DBLP:journals/pacmpl/McIverMKK18} (by applying a transformation to the ranking function), but needs to be different in order to be compatible with the presence of additional intermediate reduction steps where the ranking function is not required to decrease at all.}
+\lo{The same point is made in \cref{sec:related}, so the preceding can be omitted.}
 
@@
 We introduce a novel addressing scheme for the possible random choices in a program's execution, which ensures that the same random choices are taken at corresponding positions in alternative reduction sequences, so that the same eventual result can be reached.
-This is then used to prove yet another extension to the ranking function theorem, that ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and rankability in this sense still imples almost sure termination.
-This confluent trace semantics has other possible applications as well, for example in Bayesian inference algorithms.
+This is then used to prove yet another extension to the ranking function theorem, that (\cref{thm:confluent ranking}) ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and (\cref{thm:confluentNoStronger}) rankability in this sense still imples almost sure termination.
+We stress that the confluent trace semantics is of wider interest, and has other possible applications as well, for example in Bayesian inference algorithms.
+
+By combining these methods we can prove AST of a variety of PPCF programs, including cases (e.g.~\cref{ex:non-affine recursion,ex:higher-order recursion}) that are beyond the reach of methods in the literature.
 

### 2021-01-16 LICS21/ranking.tex
Introduction:ongoing Reformatted last antitone sparse ranking function in appendix

@@
 
-\begin{restatable}{theorem}{PartialImpliesRankable} \label{thm:partial implies rankable}
+\begin{restatable}[Sparse function]{theorem}{PartialImpliesRankable} \label{thm:partial implies rankable}
 Every sparse ranking function is a restriction of a ranking function.

### 2021-01-16 LICS21/supermartingales.tex
Introduction:ongoing Reformatted last antitone sparse ranking function in appendix

@@
 \begin{remark}
-Our notion of ranking supermartingale is a slight generalisation of the original definition, due to \cite{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, which does not refer to a stopping time explicitly:
-Fioriti and Hermanns call a supermartingale $(Y_n)_{n \geq 0}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot {\bf 1}_{\set{Y_n > 0}}$.
+Our notion of ranking supermartingale is a minor generalisation of the original definition, due to \cite{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, which does not refer to a stopping time explicitly.
+%: Fioriti and Hermanns call a supermartingale $(Y_n)_{n \geq 0}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot {\bf 1}_{\set{Y_n > 0}}$.
 \end{remark}
@@
 
-In the rest of this section we show that if $M \in \Lambda^0$ is rankable, then $M$ is AST: in other words, the method of ranking function is sound for proving a.s.~termination of PPCF programs.
+In the rest of this section we show that if a PPCF term is rankable, then it is AST. 
+%in other words, the method of ranking function is sound for proving a.s.~termination of PPCF programs.
 
-Let $f$ be a ranking function on $M$.
+Let $f$ be a ranking function on $M \in \Lambda^0$.
 Define random variables on the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$:
@@
 Note that, as the type system ensures that the reduction relation excluding $\tY$-reduction is strongly normalising, only finitely many reductions can occur in a row without one of them being a $\tY$-reduction, therefore $T_M^\tY < \infty$ a.s.~iff $M$ is AST.
-Moreover 
+Moreover: 
 \begin{restatable}{lemma}{TMtYStoppingTime}
@@
 Let $f$ be a ranking function on $M$.
-For 1, since $(T_n)_{n \geq 0}$ is an increasing sequence of stopping times, each is adapted to $(\calF_n)_{n \geq 0}$ and bounded (\Cref{lem:TnBounded}),
+For (i), since $(T_n)_{n \geq 0}$ is an increasing sequence of stopping times, each is adapted to $(\calF_n)_{n \geq 0}$ and bounded (\Cref{lem:TnBounded}),
 and $(f(M_n))_{n \geq 0}$ is a supermartingale also adapted to $(\calF_n)_{n \geq 0}$ (\Cref{thm:rankable and strict rankable}),
@@
 Therefore, by \Cref{lem:rank-PAST}, $T^{\tY}_M < \infty$ a.s.~and $\expect{T^{\tY}_M} < \infty$.
-
 %For 2, we have that $(f(M_n))_{n \geq 0}$ is a ranking supermartingale (\Cref{thm:rankable and strict rankable}) w.r.t.~stopping time $T_M$. 
 %It follows immediately from \Cref{lem:rank-PAST} that $\expect{T_M} < \infty$.
-Statement 2 follows from \Cref{thm:rankable and strict rankable} and \Cref{lem:rank-PAST}.
+Statement (ii) follows at once from \Cref{thm:rankable and strict rankable} and \Cref{lem:rank-PAST}.
 \end{proof}

### 2021-01-16 LICS21/intro.tex
minor
Reworked introduction. Added conclusion.

@@
 
-\lo{Proving termination of higher-order probabilistic programs with continuous distribution is important but difficult.}
+Probabilistic (or randomised) programs have long been recognised as central to the efficient solution of algorithmic problems \cite{Rabin1976}.
+More recently, in probabilistic programming \cite{DBLP:conf/icse/GordonHNR14,rainforth2017Automating,vandemeent2018Introduction}, probabilistic programs are used to express generative models whose posterior probability can be computed by general purpose inference engines.
+Sampling from continuous distributions is an essential construct of probabilistic programming languages, such as Church \cite{DBLP:conf/uai/GoodmanMRBT08}, Stan \cite{carpenter2017stan}, Anglican \cite{DBLP:conf/pkdd/TolpinMW15}, Gen \cite{cusumano-towner2019Gen}, Pyro \cite{bingham2019Pyro}, Edward \cite{tran2016edward} and Turing \cite{ge2018Turing}.
+Another key feature is higher order; in fact some of the most influential probabilistic languages are functional.
 
-\medskip
+In this paper we study a central property of probabilistic programs: \emph{termination}. 
+A program is \emph{almost surely terminating} (AST) if a run of it terminates with probability $1$.
+When a probabilistic program implements a solution to an algorithmic problem, one naturally requires the computation to terminate with a high probability, usually $1$.
+In probabilistic programming, guarantees of AST are equally important. 
+Indeed, it is standard for designers and implementors of probabilistic programming systems to regard non-AST programs as defining invalid models, and hence inadmissible \cite{Rainforth2017,DBLP:conf/uai/GoodmanMRBT08}.
 
-Various theorems about probabilistic programs rely on the assumption that the program terminates almost surely. 
-One possible approach to prove almost sure termination (AST) is to find some variant on the program state, called \emph{ranking function}, that decreases on average sufficiently quickly that it must at some point reach 0, at which point the program terminates.
+Moreover, various theorems about probabilistic programs rely on the assumption that the program terminates almost surely. 
+A recent result \cite{MakOPW20} proves that AST programs have density functions that are differentiable almost everywhere.
+This is significant for Bayesian machine learning, because the latter property is a precondition for the correctness of some of the most scalable inference algorithms, such as Hamiltonian Monte Carlo \cite{DBLP:conf/aistats/ZhouGKRYW19,Nishimura2020b} and reparameterised gradient variational inference \cite{DBLP:conf/nips/LeeYY18}.
+
+\subsection*{Contributions}
+
+A possible approach to prove almost sure termination (AST) is to find some variant on the program state, called \emph{ranking function}, that decreases on average sufficiently quickly that it must at some point reach 0, at which point the program terminates.
 In other words, the program's behavior is used to define an associated (ranking) \emph{supermartingale}.
@@
 This is then used to prove yet another extension to the ranking function theorem, that (\cref{thm:confluent ranking}) ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and (\cref{thm:confluentNoStronger}) rankability in this sense still imples almost sure termination.
-We stress that the confluent trace semantics is of wider interest, and has other possible applications as well, for example in Bayesian inference algorithms.
+The confluent trace semantics is of wider interest, and has other possible applications as well, for example in Bayesian inference algorithms.
 
@@
 
-\subsubsection*{Contributions}
-
-This is the first application of martingales to probabilistic lambda calculus, and the first version of trace semantics that's capable of satisfying the (restricted) Church-Rosser property.
-
-\lo{Ongoing.}
-
-\subsubsection*{Outline}
+\paragraph*{Outline}
 

### 2021-01-16 LICS21/related.tex
Introduction:ongoing Reformatted last antitone sparse ranking function in appendix
Reworked introduction. Added conclusion.

@@
 I suppose it could at least reassure readers that we're aware of that paper and it isn't the same thing, but I don't know how likely readers are to actually draw the relation in the first place.}
+\lo{On reflection, I agree it is not necessary to discuss this.}
 
@@
 Although this is a broadly applicable method of proving almost sure termination of probabilistic functional programs, and \Cref{thm:antitone partial implies rankable,thm:confluent ranking} make it more convenient to use, some method of automating the construction and checking of (antitone) ranking functions, even partially, could make it considerably more practically useful, especially in cases where almost sure termination is merely a side-condition for some other theorem or algorithm to be applicable.
+
+\paragraph*{Conclusions}
+
+We have presented the first application of martingales to probabilistic lambda calculus, and the first version of trace semantics that's capable of satisfying the (restricted) Church-Rosser property.
+Though the construction of ranking functions is inherently difficult, using sparse ranking functions, antitone ranking functions and ranking functions w.r.t.~an alternative reduction strategy, we have shown that a great variety of programs can be shown to be AST, including some that are beyond the reach of methods in the literature.

### 2021-01-16 LICS21/antitone.tex
Reworked introduction. Added conclusion.
Introduction: added complexity results

@@
 
+\iffalse
 Sampling from continuous distributions is an essential feature of statistical probabilistic programming languages.
@@
 To our knowledge, the problem of proving AST of probabilistic functional programs with continuous distribution is new.
+\fi
 
@@
 
-\begin{example}[Non-affine recursion] \label{ex:non-affine recursion}
+\begin{example}[Non-affine recursion]
+\label{ex:non-affine recursion easy}
 \[

### 2021-01-16 LICS21/intro.tex
Introduction: added complexity results

@@
 
-Probabilistic (or randomised) programs have long been recognised as central to the efficient solution of algorithmic problems \cite{Rabin1976}.
+Probabilistic (or randomised) programs have long been recognised as essential to the efficient solution of many algorithmic problems \cite{Rabin1976}.
 More recently, in probabilistic programming \cite{DBLP:conf/icse/GordonHNR14,rainforth2017Automating,vandemeent2018Introduction}, probabilistic programs are used to express generative models whose posterior probability can be computed by general purpose inference engines.
 Sampling from continuous distributions is an essential construct of probabilistic programming languages, such as Church \cite{DBLP:conf/uai/GoodmanMRBT08}, Stan \cite{carpenter2017stan}, Anglican \cite{DBLP:conf/pkdd/TolpinMW15}, Gen \cite{cusumano-towner2019Gen}, Pyro \cite{bingham2019Pyro}, Edward \cite{tran2016edward} and Turing \cite{ge2018Turing}.
-Another key feature is higher order; in fact some of the most influential probabilistic languages are functional.
+Another key feature is higher order; in fact some of the most influential probabilistic programming languages are functional.
 
 In this paper we study a central property of probabilistic programs: \emph{termination}. 
-A program is \emph{almost surely terminating} (AST) if a run of it terminates with probability $1$.
-When a probabilistic program implements a solution to an algorithmic problem, one naturally requires the computation to terminate with a high probability, usually $1$.
+A program is \emph{almost surely terminating} (AST) if a run of it terminates with probability one.
+When a probabilistic program implements a solution to an algorithmic problem, we naturally require the computation to terminate with a high probability, usually $1$.
 In probabilistic programming, guarantees of AST are equally important. 
@@
 
-Moreover, various theorems about probabilistic programs rely on the assumption that the program terminates almost surely. 
+Moreover various theorems about probabilistic programs rely on the assumption that the program terminates almost surely. 
 A recent result \cite{MakOPW20} proves that AST programs have density functions that are differentiable almost everywhere.
-This is significant for Bayesian machine learning, because the latter property is a precondition for the correctness of some of the most scalable inference algorithms, such as Hamiltonian Monte Carlo \cite{DBLP:conf/aistats/ZhouGKRYW19,Nishimura2020b} and reparameterised gradient variational inference \cite{DBLP:conf/nips/LeeYY18}.
+This is significant for Bayesian machine learning, because almost everywhere differentiability is a precondition for the correctness of some of the most scalable inference algorithms, such as Hamiltonian Monte Carlo \cite{DBLP:conf/aistats/ZhouGKRYW19,Nishimura2020b} and reparameterised gradient variational inference \cite{DBLP:conf/nips/LeeYY18}.
+
+Despite its importance, the AST problem is difficult: 
+Deciding AST of first-order imperative programs with discrete probabilities is $\Pi^0_2$-complete \cite{DBLP:conf/mfcs/KaminskiK15}.
+In recent unpublished work, the AST problem for higher-order programs with continuous distributions (and suitable primitive functions) is shown to have the same complexity.
+Though the problem of proving AST of imperative probabilistic programs is much studied \cite{DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05,DBLP:conf/popl/FioritiH15,DBLP:conf/pldi/ChenH20,DBLP:journals/toplas/ChatterjeeFNH18}, to our knowledge the problem of proving AST of higher-order programs with continuous distribution is new.
 
@@
 
-A possible approach to prove almost sure termination (AST) is to find some variant on the program state, called \emph{ranking function}, that decreases on average sufficiently quickly that it must at some point reach 0, at which point the program terminates.
+A possible approach to prove AST is to find some variant on the program state, called \emph{ranking function}, that decreases on average sufficiently quickly that it must at some point reach 0, at which point the program terminates.
 In other words, the program's behavior is used to define an associated (ranking) \emph{supermartingale}.
@@
 Because the type system already constrains the terms enough to force termination in the absence of the recursion construct \cite{tait1967,BarendregtDS2010}, it is only the $\tY$-reduction steps that must be counted, making defining ranking functions somewhat easier.
-Using supermartingales and Doob's (iterated) Optional Sampling Theorem, we show the soundness of rankability (\cref{thm:rankable implies termination}): if a PPCF term has a ranking function then it is AST.
+Using supermartingales and Doob's (iterated) Optional Sampling Theorem, we show the soundness of rankability (\cref{thm:rankable implies termination}): if a term has a ranking function, it is AST.
 
@@
 Thanks to this feature, the antitone ranking function method is capable of handling programs which terminate arbitrarily slowly. \lo{TODO: Reference some examples.}
-We show that this method is sound for proving AST (\cref{thm:antitone rankable implies termination}), 
-and it also enjoys a sparse function theorem (\cref{thm:antitone partial implies rankable}).
+We show that this method is also sound for proving AST (\cref{thm:antitone rankable implies termination}), 
+moreover it also enjoys a sparse function theorem (\cref{thm:antitone partial implies rankable}).
 \akr{Our progress condition is equivalent to that given by McIver et al.~\cite{DBLP:journals/pacmpl/McIverMKK18} (by applying a transformation to the ranking function), but needs to be different in order to be compatible with the presence of additional intermediate reduction steps where the ranking function is not required to decrease at all.}
@@
 We introduce a novel addressing scheme for the possible random choices in a program's execution, which ensures that the same random choices are taken at corresponding positions in alternative reduction sequences, so that the same eventual result can be reached.
-This is then used to prove yet another extension to the ranking function theorem, that (\cref{thm:confluent ranking}) ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and (\cref{thm:confluentNoStronger}) rankability in this sense still imples almost sure termination.
+This is then used to prove yet another extension to the ranking function theorem, that (\cref{thm:confluent ranking}) ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and (\cref{cor:Reduction strategy independence}) rankability in this sense still imples almost sure termination.
 The confluent trace semantics is of wider interest, and has other possible applications as well, for example in Bayesian inference algorithms.
 
-By combining these methods we can prove AST of a variety of PPCF programs, including cases (e.g.~\cref{ex:non-affine recursion,ex:higher-order recursion}) that are beyond the reach of methods in the literature.
+By combining these methods we can prove AST of a variety of PPCF programs that are beyond the reach of methods in the literature, including non-affine recursion\footnote{\label{fnote:non-affine} Non-affine recursive programs are recursive programs that can, during the evaluation of the recursive body, make multiple recursive calls (of a first-order function) from distinct call sites.} (\cref{ex:non-affine recursion easy}) and recursions that define high-order functions (\cref{ex:higher-order recursion}).
 

### 2021-01-17 LICS21/SPCF.tex
Various minors

@@
 
-The language PPCF is a CbV version of PCF with sampling of real numbers from the closed interval $[0,1]$ \cite{Ehrhard2018c,DBLP:journals/pacmpl/EhrhardPT18,MakOPW21}.
+The language PPCF is a call-by-value version of PCF with sampling of real numbers from the closed interval $[0,1]$ \cite{Ehrhard2018c,DBLP:journals/pacmpl/EhrhardPT18,MakOPW21}.
 %\changed[lo]{following \cite{MakOPW21} except the score function (\Cref{rem:score})}. 
@@
 Let $ I := [0,1] \subset \mathbb{R} $, and define $\entrosp := I^{\mathbb{N}}$, with the Borel $\sigma$-algebra $\Sigma_\entrosp$ and the probability measure, denoted $\mu_{\entrosp}$, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
-Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and $\mu_{\entrosp} (\prod_{i=0}^\infty X_i) := \prod_{i=0}^\infty \Leb(X_i)$.
-The maps $\pi_h:\entrosp \to I, \; \pi_t:\entrosp \to \entrosp$ popping the first element are then measurable.
-Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$ the \emph{entropy space}, and call elements of $\entrosp$ \emph{traces}.
+Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and $\mu_{\entrosp} (\prod_{i=0}^\infty X_i) := \prod_{i=0}^\infty \Leb(X_i)$, \changed[lo]{writing $\Leb$ for the Lebesgue measure.}
+\changed[lo]{The maps $\pi_h:\entrosp \to I$ (projecting to the first element) and $\pi_t:\entrosp \to \entrosp$ (popping the first element) are then measurable.}
+Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$ the \emph{entropy space}, and elements of $\entrosp$ \emph{traces}.
 
@@
 
-The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid \hbox{$M$ terminates at $s$}\}$ is measurable.
+The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid \hbox{$M$ terminates at $s$}\}$ is measurable \cite{DBLP:conf/icfp/BorgstromLGS16,MakOP20a}.
 We say that a term $M$ is \emph{almost surely terminating} (AST) just if $\mu_{\entrosp}(\{s \mid \hbox{$M$ terminates at $s$}\}) = 1$.
@@
 \(
-\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \underline 0}
+\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \, \underline 0}
 \), 

### 2021-01-17 LICS21/intro.tex
Various minors

@@
 Indeed, it is standard for designers and implementors of probabilistic programming systems to regard non-AST programs as defining invalid models, and hence inadmissible \cite{Rainforth2017,DBLP:conf/uai/GoodmanMRBT08}.
+(Yet none of these systems provides any support for the development or verification of AST programs.)
 
@@
 
-Despite its importance, the AST problem is difficult: 
+The AST problem is not just important, but also difficult: 
 Deciding AST of first-order imperative programs with discrete probabilities is $\Pi^0_2$-complete \cite{DBLP:conf/mfcs/KaminskiK15}.
 In recent unpublished work, the AST problem for higher-order programs with continuous distributions (and suitable primitive functions) is shown to have the same complexity.
-Though the problem of proving AST of imperative probabilistic programs is much studied \cite{DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05,DBLP:conf/popl/FioritiH15,DBLP:conf/pldi/ChenH20,DBLP:journals/toplas/ChatterjeeFNH18}, to our knowledge the problem of proving AST of higher-order programs with continuous distribution is new.
+Though the problem of proving AST of imperative probabilistic programs is much studied \cite{DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05,DBLP:conf/popl/FioritiH15,DBLP:conf/pldi/ChenH20,DBLP:journals/toplas/ChatterjeeFNH18}, to our knowledge the problem of proving AST of higher-order programs with continuous distributions is new.
 

### 2021-01-17 LICS21/related.tex
Introduction: added complexity results
Various minors

--- 2021-01-16 LICS21/related.tex
@@
 Breuvart and Dal Lago \cite{DBLP:conf/ppdp/BreuvartL18} develop systems of intersection types from which the termination probability of higher-order programs (with discrete distributions) can be inferred from (infinitely many) type derivations.
+\lo{Mention Dal Lago et al.
+POPL21.
+Proving PAST.}
 
@@
 By contrast, our approach works for programs with continuous distributions. 
-Note that $n$-PHORS is strictly less expressive then order-$n$ CbN PPCF (because the underlying recursion schemes are not Turing complete):
-PPCF terms in \cref{ex:raven complex,ex:higher-order recursion}
+Note that $n$-PHORS is strictly less expressive then order-$n$ CbN PPCF:
+% (because the underlying recursion schemes are not Turing complete):
+the PPCF terms in \cref{ex:raven complex,ex:higher-order recursion}
 %\refExample{complexExample} 
@@
 Although this is a broadly applicable method of proving almost sure termination of probabilistic functional programs, and \Cref{thm:antitone partial implies rankable,thm:confluent ranking} make it more convenient to use, some method of automating the construction and checking of (antitone) ranking functions, even partially, could make it considerably more practically useful, especially in cases where almost sure termination is merely a side-condition for some other theorem or algorithm to be applicable.
+
+\paragraph*{Conclusions}
+
+We have presented the first application of martingales to probabilistic lambda calculus, and the first version of trace semantics that's capable of satisfying the (restricted) Church-Rosser property.
+Though the construction of ranking functions is inherently difficult, using sparse ranking functions, antitone ranking functions and ranking functions w.r.t.~an alternative reduction strategy, we have shown that a great variety of programs can be proved to be AST, including some that are beyond the reach of methods in the literature.

### 2021-01-17 LICS21/supermartingales.tex
Various minors

@@
 
-It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \geq 0}$ is the reduction sequence starting from $M$, then $(f(M_n))_{n \geq 0}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales.
-%\lo{It is confusing (strictly speaking, incorrect) to say that $(M_n)_{n \geq 0}$ is a reduction sequence from $M$. Formally $M_n$ is the random variable $M_n: s \mapsto \pi_0 (\red^n(M, s))$.}
-%\akr{I guess it's implicitly identifying a sequence of functions ($(M_n)_{n \geq 0}$), and a function producing sequences (a reduction-sequence-valued random variable), but this sort of thing seems very standard in dealing with random variables (e.g.~applying a function to some random variables, which is really a sort of concatenation).}
-
+It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \geq 0}$ \changed[lo]{(qua stochastic process)} is the reduction sequence starting from $M$, then $(f(M_n))_{n \geq 0}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales (see e.g.~\cite{Williams1999} for details).
+\iffalse
+\lo{It is confusing (strictly speaking, incorrect) to say that $(M_n)_{n \geq 0}$ is a reduction sequence from $M$. Formally $M_n$ is the random variable $M_n: s \mapsto \pi_0 (\red^n(M, s))$.}
+\akr{I guess it's implicitly identifying a sequence of functions ($(M_n)_{n \geq 0}$), and a function producing sequences (a reduction-sequence-valued random variable), but this sort of thing seems very standard in dealing with random variables (e.g.~applying a function to some random variables, which is really a sort of concatenation).}
+\fi
 %\paragraph{}
 
-Fix a probability space $(\Omega, \calF, \mathbb{P})$ and a filtration $(\calF_n)_{n \geq 0}$. 
-(In subsequent applications to termination analysis, $\Omega = \entrosp$, and $\mathbb{P} = \mu_{\entrosp}$.)
-Let $T$ be a random variable that takes values in $\mathbb{N} \cup \set{\infty}$.
-We call $T$ a \emph{stopping time adapted to $(\calF_n)_{n \geq 0}$} just if $\set{T = n} \in \calF_n$, for all $n \geq 0$.
+Fix a probability space $(\Omega, \calF, \mathbb{P})$ and a filtration $(\calF_n)_{n \geq 0}$ (i.e.~$\calF_n \subseteq \calF$ is a $\sigma$-algebra, and $\calF_n \subseteq \calF_{n+1}$ for all $n$). 
+%(In subsequent applications to termination analysis, $\Omega = \entrosp$, and $\mathbb{P} = \mu_{\entrosp}$.)
+Let $T$ be a r.v.~that takes values in $\mathbb{N} \cup \set{\infty}$.
+We call $T$ a \emph{stopping time adapted to $(\calF_n)_{n \geq 0}$} just if $\set{T = n} \in \calF_n$, for all $n$.
 
 \begin{definition}\rm
-\begin{enumerate}
-\item A sequence of random variables $(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \emph{supermartingale} if for all $n \geq 0$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~(i.e.~for all $A \in \calF_n$, $\int_A \mathbb{P}(\dif \omega) \, Y_{n+1}(\omega) \leq \int_A \mathbb{P}(\dif \omega) \,Y_n(\omega)$).
+\begin{asparaenum}[(i)]
+\item A sequence of r.v.~$(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \emph{supermartingale} if for all $n \geq 0$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~(i.e.~for all $A \in \calF_n$, $\int_A \mathbb{P}(\dif \omega) \, Y_{n+1}(\omega) \leq \int_A \mathbb{P}(\dif \omega) \,Y_n(\omega)$).
 \item Let $\epsilon > 0$. 
@@
 A \emph{ranking supermartingale} w.r.t.~$T$ is a $\epsilon$-ranking supermartingale w.r.t.~$T$ for some $\epsilon > 0$.
-\end{enumerate}
+\end{asparaenum}
 \end{definition}
@@
 \begin{remark}
-Our notion of ranking supermartingale is a minor generalisation of the original definition, due to \cite{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, which does not refer to a stopping time explicitly.
+Our notion of ranking supermartingale is a slight generalisation of the original definition \cite{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, \changed[lo]{which does not involve an arbitrary stopping time}.
 %: Fioriti and Hermanns call a supermartingale $(Y_n)_{n \geq 0}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot {\bf 1}_{\set{Y_n > 0}}$.

### 2021-01-17 LICS21/SPCF.tex
Broken reference fixed, and other minors.

@@
 
-The language PPCF is a call-by-value version of PCF with sampling of real numbers from the closed interval $[0,1]$ \cite{Ehrhard2018c,DBLP:journals/pacmpl/EhrhardPT18,MakOPW21}.
+The language PPCF is a call-by-value (CBV) version of PCF with sampling of real numbers from the closed interval $[0,1]$ \cite{Ehrhard2018c,DBLP:journals/pacmpl/EhrhardPT18,MakOPW21}.
 %\changed[lo]{following \cite{MakOPW21} except the score function (\Cref{rem:score})}. 
@@
 %\paragraph{}
-To define the call-by-value reduction relation, let \emph{evaluation contexts} be of the form:
+To define the CBV reduction relation, let \emph{evaluation contexts} be of the form:
 \begin{align*}

### 2021-01-17 LICS21/antitone.tex
Broken reference fixed, and other minors.

@@
 \]
-For any $F : \textsf{R} \to \textsf{R} \to \textsf{R}$ such that $F\ \underline n\ \underline m$ terminates with no $\tY$ reductions for all $m$ and $n$, $\Xi\ F\ \underline x\ \underline y$ is antitone rankable.
+For any $F : \textsf{R} \to \textsf{R} \to \textsf{R}$ such that $F\, \underline n\, \underline m$ terminates with no $\tY$ reductions for all $m$ and $n$, $\Xi\, F\, \underline x\, \underline y$ is antitone rankable.
 \end{example}

### 2021-01-17 LICS21/confluent.tex
Introduction: added complexity results
Broken reference fixed, and other minors.

@@
 
-\begin{corollary}
+\begin{corollary}[Reduction strategy independence]
+\label{cor:Reduction strategy independence}
 If $M$ is AST with respect to any reduction strategy, it is AST.
@@
 
-\begin{example} \label{ex:sum of powers}
+\begin{example}[Sum of powers] \label{ex:sum of powers}
 The following term picks a random natural number $n$, then computes $\sum_{k=0}^n k^n$.

### 2021-01-17 LICS21/intro.tex
Broken reference fixed, and other minors.

@@
 
-To define a ranking function, one has to try to organise the set of reachable terms (of the program being analysed) into a manageable number of syntactic cases. 
+To define a ranking function, one would typically try to organise the set of reachable terms (of the program being analysed) into a manageable number of syntactic cases. 
 Unfortunately (for any fixed reduction strategy) there are programs whose reachable terms span such a large number of cases that it would be extremely difficult to analyse by hand, if at all possible.

### 2021-01-18 LICS21/SPCF.tex
Minor fixes

@@
 To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples.
-Let $ I := [0,1] \subset \mathbb{R} $, and define $\entrosp := I^{\mathbb{N}}$, with the Borel $\sigma$-algebra $\Sigma_\entrosp$ and the probability measure, denoted $\mu_{\entrosp}$, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element.
-Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and $\mu_{\entrosp} (\prod_{i=0}^\infty X_i) := \prod_{i=0}^\infty \Leb(X_i)$, \changed[lo]{writing $\Leb$ for the Lebesgue measure.}
+Let $ I := [0,1] \subset \mathbb{R} $, and define $\entrosp := I^{\mathbb{N}}$, with the Borel $\sigma$-algebra $\Sigma_\entrosp$.
+Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and the probability measure $\mu_{\entrosp}$ is given by $\mu_{\entrosp} (\prod_{i=0}^\infty X_i) := \prod_{i=0}^\infty \Leb(X_i)$, \changed[lo]{writing $\Leb$ for the Lebesgue measure.}
 \changed[lo]{The maps $\pi_h:\entrosp \to I$ (projecting to the first element) and $\pi_t:\entrosp \to \entrosp$ (popping the first element) are then measurable.}
@@
 \[
-(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0},
+(\tY \lambda f n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0},
 \] 

### 2021-01-18 LICS21/antitone-proofs.tex
Broken reference fixed, and other minors.
Minor fixes

@@
 
-\medskip
+%\medskip
+
+\subsubsection*{Non-affine recursion}
 
@@
 
-\Cref{ex:non-affine recursion}
+\Cref{ex:non-affine recursion easy}
 Let 
@@
 It is basically another variant of the biased random walk, except that the relevant variable is the number of copies of $\Xi$, instead of a real number in the term.
+
+\subsubsection*{Higher-order recursion}
 
@@
 Within $X$, there is necessarily a redex at $\cbv(X)$.
-The only conditions that $@_2^k;\cbv(X)$ is also a position of a redex in the whole term are that if the redex is a $\tsample$, it is not inside a $\lambda$ or $\tY$ on the right of an application, which is true because $\cbv$ never selects a redex inside of a $\lambda$ or $\tY$.
+The only conditions that $@_2^k;\cbv(X)$ is also a position of a redex in the whole term are that if the redex is a $\tsample$, it is not inside a $\lambda$ or $\tY$ on the right of an application, which is true because $\cbv$ never selects a redex inside of a $\lambda$.
 After $n$ reaches $0$, the term is $F \underline{n_0} (\dots(F\ \underline {n_k}\ \underline x)\dots)$, then the innermost $F$ is evaluated completely in $\cbv$ order, therefore (by assumption) it terminates.

### 2021-01-18 LICS21/antitone.tex
Minor fixes

@@
 
+\begin{definition}
 Define an \emph{antitone sparse ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$:
@@
 \end{inparaenum}
-\iffalse
-\begin{itemize}
-\item $f(N) \geq 0$ for all $N$ where $f$ is defined.
-\item $f$ is defined at $M$.
-\item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
-\end{itemize}
-\fi
+\end{definition}
+
 \begin{restatable}[Sparse function]{theorem}{AntitonePartial}  \label{thm:antitone partial implies rankable}
@@
 \begin{example}[Continuous random walk]\label{ex:raven complex}
-In PPCF we can construct a function whose argument changes by a random amount at each recursive call: $\Theta \, 10$ where
+In PPCF we can construct a function whose argument changes by a random amount at each recursive call: $\Theta \, \underline{10}$ where
 \[
 %\underbrace{\big
-\Theta := \tY \, \lambda f x . \tif{x \leq 0}{0}{f(x - \tsample)}
+\Theta := \tY \, \lambda f x .\ \tif{x \leq 0}{\underline 0}{f(x - \tsample)}
 %}_{\Theta} \, 10
@@
 
-For a more complex (not $\tY$-PAST) example, consider the following ``continuous random walk'': $\Xi \, 10$ where
+For a more complex (not $\tY$-PAST) example, consider the following continuous random walk: $\Xi \, \underline{10}$ where
 \[
-%\underbrace{\big(
-\Xi := \tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample + 1/2)} 
-%\big)}_{\Xi} \, 10
+\Xi := \tY \, \lambda f x .\ \tif{x \leq 0}{\underline 0}{f(x - \tsample + 1/2)} 
 \]
@@
 \big
-(\tY \, f \, x . 
-\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)
-\, 10
+(\tY\ \lambda f x .\ 
+\tif{x \leq 0}{\underline 0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)
+\, \underline{10}
 \]
@@
 %\underbrace{
-\big(\tY \, f \, x . \,
-0 \oplus_{\frac{1}{x+1}} f(x + 1) 
+\big(\tY\ \lambda f x . \,
+\underline 0 \oplus_{\frac{1}{x+1}} f(x + 1) 
 \big)
 %}_{\Xi} 
-\, 10
+\, \underline{10}
 \]

### 2021-01-18 LICS21/applications.tex
Introduction:ongoing Reformatted last antitone sparse ranking function in appendix
Minor fixes

@@
 If the labels of samples in the trace correspond closely to the roles that the samples play in the program execution, this will increase the efficiency of the inference algorithm.
-In \cite{wingate2011lightweight}, for example, something rather like a stack trace is used, although this assdressing scheme turns out to have various problems \cite{kiselyov2016problems,Hur2015a}. 
+In \cite{wingate2011lightweight}, for example, something rather like a stack trace is used, although this addressing scheme turns out to have various problems \cite{kiselyov2016problems,Hur2015a}. 
 Potential positions could also be used for a similar purpose, although in practice, a more compact representation than an entire skeletal reduction sequence would be necessary. 
@@
 SPCF is a much nicer probabilistic programming language (in purified / idealised form).}
-Selecting an appropriate addressing scheme can also be important in the context of variational inference \cite[\S 6.2.2]{paige2016automatic}.
+Selecting an appropriate addressing scheme is also important for variational inference \cite{paige2016automatic}.
+%\cite[\S 6.2.2]{paige2016automatic}.
 

### 2021-01-18 LICS21/confluent-proofs.tex
Introduction:ongoing Reformatted last antitone sparse ranking function in appendix
minor
Minor fixes

@@
 
+It is also true that the distributions of terms produced by the standard semantics and by the reduction strategy $\cbv$ are the same, for much the same reason, although that is not relevant to our main results.
+
 \CbvIsTerminatingest*
@@
 \end{proof}
+
+It is not quite true, however, that all reduction strategies produce the same values (barring some reduction strategies terminating where others don't), because it is possible for a reduction strategy to overshoot the value that the standard semantics produces by, for example, performing reductions inside of a lambda.
+Even for the same value of the samples, the final values may be different.
+A more restricted result is true, however, that if a term terminates with some reduction strategy $f$ and samples $s$, the value that $\Rightarrow_{\cbv}$ terminates with can reduce via $\Rightarrow$ to the value that $\Rightarrow_f$ terminates with (as a corollary to this proof).
 
@@
 It would be possible to define an antitone sparse ranking function for it, but it is much more difficult to construct:
+\iffalse
 \begin{align*}
@@
 \end{align*}
+\fi
+\begin{align*}
+M & \mapsto 1.47 \\
+P[\underline n] & \mapsto 1 + \ln((n+1)n + 1) \\
+\Xi[\lambda x. \Theta[x] \, \underline n] \, \underline n & \mapsto 1 + \ln((n+1)n + 1) \\
+%\underline{n^n} + \dots + \underline{k_1^n} + \\
+%\Xi[\lambda x. \Theta[x] \underline n] (\underline{k_1} - 1) & \mapsto 1 + \ln((n+1)(k_1-1)+1) \\
+\left.\begin{array}{r}
+\underline{n^n} + \dots + \underline{k_1^n} + {}\\
+\Xi[\lambda x. \Theta[x] \underline n] (\underline{k_1} - 1) 
+\end{array}\right\} 
+& \mapsto 1 + \ln((n+1)(k_1-1)+1) \\
+%\underline{n^n} + \dots + \underline{(k_1+1)^n} + \Theta[\underline{k_1}] \underline{n} + \\
+%\Xi[\lambda x. \Theta[x] \underline n] (\underline{k_1} - 1) & \mapsto 1 + \ln((n+1)k_1 + 1) \\
+\left.\begin{array}{r}
+\underline{n^n} + \dots + \underline{(k_1+1)^n} +  {}\\
+\Theta[\underline{k_1}] \, \underline{n} + \Xi[\lambda x. \Theta[x] \, \underline n] (\underline{k_1} - 1) 
+\end{array}\right\}
+& \mapsto 1 + \ln((n+1)k_1 + 1) \\
+%\underline{n^n} + \dots + \underline{(k_1+1)^n} + \\
+%\underbrace{\underline{k_1} \times \dots \times \underline{k_1} \times}_{n-k_2+1} \Theta[\underline{k_1}] (\underline{k_2} - 1) + \\
+%\Xi[\lambda x. \Theta[x] \underline n] (\underline{k_1} - 1) & \mapsto 1 + \ln((n+1)k_1 + k_2 + 1) \\
+\left.\begin{array}{r}\underline{n^n} + \dots + \underline{(k_1+1)^n} + {} \\
+\underbrace{\underline{k_1} \times \dots \times \underline{k_1} \times}_{n-k_2+1} \Theta[\underline{k_1}] (\underline{k_2} - 1) + {} \\
+\Xi[\lambda x. \Theta[x] \, \underline n] (\underline{k_1} - 1)
+\end{array}\right\} & \mapsto 1 + \ln((n+1)k_1 + k_2 + 1) \\
+\underline{\textstyle\sum_{k=0}^n k^n} & \mapsto 0.
+\end{align*}
 

### 2021-01-18 LICS21/confluent.tex
Minor fixes

@@
 \begin{align*}
-A[x] &= \tif{\tif{x>0}{I}{I} \, \tsample - 0.5 > 0}{0}{\Omega}\\
-B &= \tif{\tsample - 0.5 > 0}{0}{\Omega}
+A[x] &= \tif{\tif{x>0}{I}{I} (\lambda y.\tsample)\ \underline 0 - \underline{0.5} > 0}{\underline 0}{\Omega}\\
+B &= \tif{\tsample - \underline{0.5} > 0}{\underline 0}{\Omega}
 \end{align*}
@@
 \begin{array}{l}
-\bigcup_{r \in [0,1]} \big\{s \mid\ s(A[\tsample], \mathsf{if}_1;\underline{-}_1;@_1;\mathsf{if}_1) = r, \\
+\bigcup_{r \in [0,1]} \big\{s \mid\ s(A[\tsample], \mathsf{if}_1;\underline{-}_1;@_1;@_1;\mathsf{if}_1) = r, \\
 \qquad\qquad\qquad s(A[\tsample] \to A[\underline r] \to^* B, \mathsf{if}_1 ; \underline{-}_1) > 0.5 \big\}.
@@
     \end{enumerate}
-    (in which case $O_1$ and $O_2$ are equal as skeletons, but with different reduction sequences), $O_1'$ and $O_2'$ are the results of applying some reduction sequence to each of $O_1$ and $O_2$ (the same reductions in each case, which is always possible because they're equal terms), and $\delta$ is a position in $O_1'$ (or equivalently $O_2'$), then $(O_1',\delta) \sim_c (O_2',\delta)$.
+    (in which case $O_1$ and $O_2$ are equal as skeletons, but with different reduction sequences), $O_1'$ and $O_2'$ are the results of applying some reduction sequence to each of $O_1$ and $O_2$ (the same reductions in each case, which is always possible because they're equal skeletons), and $\delta$ is a position in $O_1'$ (or equivalently $O_2'$), then $(O_1',\delta) \sim_c (O_2',\delta)$.
 \end{asparaenum}
@@
 
-It is also true that the distributions of terms produced by the standard semantics and by the reduction strategy $\cbv$ are the same, for much the same reason, although that is not relevant to our main results.
-
 \begin{restatable}{theorem}{CbvIsTerminatingest} \label{thm:CbvIsTerminatingest}
@@
 \end{restatable}
-
-It is not quite true, however, that all reduction strategies produce the same values (barring some reduction strategies terminating where others don't), because it is possible for a reduction strategy to overshoot the value that the standard semantics produces by, for example, performing reductions inside of a lambda.
-Even for the same value of the samples, the final values may be different.
-A more restricted result is true, however, that if a term terminates with some reduction strategy $f$ and samples $s$, the value that $\Rightarrow_{\cbv}$ terminates with can reduce via $\Rightarrow$ to the value that $\Rightarrow_f$ terminates with (as a corollary to the proof of \Cref{thm:CbvIsTerminatingest}).
 

### 2021-01-18 LICS21/intro.tex
Minor fixes

@@
 In this paper we study a central property of probabilistic programs: \emph{termination}.
-A program is \emph{almost surely terminating} (AST) if a run of it terminates with probability one.
-When a probabilistic program implements a solution to an algorithmic problem, we naturally require the computation to terminate with a high probability, usually $1$.
-In probabilistic programming, guarantees of AST are equally important. 
+When a probabilistic program implements a solution to an algorithmic problem, we naturally require the computation to terminate with probability $1$, in which case the program is called \emph{almost surely terminating} (AST).
 Indeed, it is standard for designers and implementors of probabilistic programming systems to regard non-AST programs as defining invalid models, and hence inadmissible \cite{Rainforth2017,DBLP:conf/uai/GoodmanMRBT08}.

### 2021-01-18 LICS21/supermartingales.tex
Minor fixes

@@
 
-It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \geq 0}$ \changed[lo]{(qua stochastic process)} is the reduction sequence starting from $M$, then $(f(M_n))_{n \geq 0}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales (see e.g.~\cite{Williams1999} for details).
+It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \geq 0}$ is the reduction sequence starting from $M$ (considered as a stochastic process), then $(f(M_n))_{n \geq 0}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales (see e.g.~\cite{Williams1999} for details).
 \iffalse
@@
 \begin{asparaenum}[(i)]
-\item A sequence of r.v.~$(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \emph{supermartingale} if for all $n \geq 0$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~(i.e.~for all $A \in \calF_n$, $\int_A \mathbb{P}(\dif \omega) \, Y_{n+1}(\omega) \leq \int_A \mathbb{P}(\dif \omega) \,Y_n(\omega)$).
+\item A sequence of r.v.s $(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \emph{supermartingale} if for all $n \geq 0$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~(i.e.~for all $A \in \calF_n$, $\int_A \mathbb{P}(\dif \omega) \, Y_{n+1}(\omega) \leq \int_A \mathbb{P}(\dif \omega) \,Y_n(\omega)$).
 \item Let $\epsilon > 0$. 
@@
 
-We first state a useful property about r.v.~$T_0, T_1, T_2, \cdots$.
+We first state a useful property about r.v.s $T_0, T_1, T_2, \dots$.
 \begin{restatable}{lemma}{TnBounded}

### 2021-01-18 LICS21/notes.tex
Added remark after assumptions

@@
 This is unnatural: there is no reason to so restrict $\tY$.
-Following \citep[p.~376]{DBLP:conf/fsttcs/Sieber90}, we should define $\tY$ as a value (and so a proper term); it follows that $\tY \, M$ is redundant as a clause.
+Following \cite[p.~376]{DBLP:conf/fsttcs/Sieber90}, we should define $\tY$ as a value (and so a proper term); it follows that $\tY \, M$ is redundant as a clause.
 
@@
 \end{example}
+
+AKR: {I'd suggest, as a simplified version of the non-affine continuous example, $(\tY \lambda f x.
+(\mathsf{let} \; e = \tsample \; \mathsf{in} \; \tif{e \leq p}{x}{f^2(x+1) \oplus_{sig(x)} f^3(x+1)}) \underline 0$ (or possibly that, but with some simpler algebraic function instead of the sigmoid).
+This removes the extra $\oplus$ and makes $x$ more relevant, but I think it preserves all the important features.}
+
+AKR: {Oh, I forgot a bit. $e$ is meant to be used again too. That can just be involved in the probability of the $\oplus$ or something, instead of having a separate one.}

### 2021-01-18 LICS21/notes.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper Sec. I: tweak

@@
 AKR: {I'd suggest, as a simplified version of the non-affine continuous example, $(\tY \lambda f x.
-(\mathsf{let} \; e = \tsample \; \mathsf{in} \; \tif{e \leq p}{x}{f^2(x+1) \oplus_{sig(x)} f^3(x+1)}) \underline 0$ (or possibly that, but with some simpler algebraic function instead of the sigmoid).
+(\mathsf{let} \; e = \tsample \; \mathsf{in} \; \tif{e \leq p}{x}{f^2(x+1) \oplus_{sig(x)} f^3(x+1)}) \, \underline 0$ (or possibly that, but with some simpler algebraic function instead of the sigmoid).
 This removes the extra $\oplus$ and makes $x$ more relevant, but I think it preserves all the important features.}

### 2021-01-21 LICS21/related.tex
Minor fixes
CBN -> call-by-name

@@
 By contrast, our approach works for programs with continuous distributions. 
-Note that $n$-PHORS is strictly less expressive then order-$n$ CbN PPCF:
+Note that $n$-PHORS is strictly less expressive then order-$n$ call-by-name PPCF:
 % (because the underlying recursion schemes are not Turing complete):
@@
 \label{conj:antitone}
-Every PPCF term is antitone rankable if all terms reachable from it are AST.
+If every term reachable from a certain PPCF term is AST, then that term is antitone rankable.
 \end{conjecture}

### 2021-01-21 LICS21/antitone-proofs.tex
Various minor fixes in the appendices

@@
 \dfrac{\expect{Y_n} - \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
-  \step[\leq]{}
+  \step[\leq]{$Y_n \geq 0$ always}
 \dfrac{\expect{Y_n}}{\mathbb P[T > n]}
-%  \step[\leq]{}
-%\dfrac{\expect{Y_0}}{\mathbb P[T > n]}
-\mathbin{=} \dfrac{\expect{Y_0}}{\mathbb P[T > n]}
+  \step[\leq]{$(Y_n)_n$ is a supermartingale}
+\dfrac{\expect{Y_0}}{\mathbb P[T > n]}
 \end{calculation}
@@
 We have
-%\lo{Change of justification of the second equality below: event, not discrete r.v. Recall: expectation conditioning on r.v.~(resp.~event) is a r.v.~(resp.~number).}
 \begin{calculation}
 \expect{Y_n-Y_{n+1}}
-%  \step[=]{}
-\mathbin{=}
+  \step[=]{$\calF_n \subseteq \calF_{n+1}$, def.~\& linearity of cond.~expectation}
 \expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
-  \step[\geq]{}
+  \step[\geq]{antitone strict assumption}
 \expect{\epsilon(Y_n) \cdot {\bf 1}_{\set{T > n}}}
-%  \step[=]{definition of conditional expectation for a discrete variable}
-%\step[=]{}
-\mathbin{=}
+  \step[=]{definition of expectation conditioning on an event}
 \mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
@@
 
-  Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) := \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
+  Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) := \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY\, \lambda x.
+N'],s') \} \right |$.
   The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on antitone partial ranking functions.
@@
 Let $\Theta := \tY \, \lambda f x . \tif{x \leq 0}{0}{f(x - \tsample)}$. 
-We can construct a sparse ranking function $f$ for $\Theta \, 10$ as follows:
+We can construct a sparse ranking function $f$ for $\Theta \, \underline{10}$ as follows:
 \begin{align*}
@@
 \\
-\tif{l \leq 0}{0}{\Theta \, (l - \tsample)}
+\tif{\underline l \leq 0}{0}{\Theta \, (\underline l - \tsample)}
 &\mapsto
@@
 \\
-0 &\mapsto 0.
+\underline 0 &\mapsto 0.
 \end{align*}
 
-For a more complex (not $\tY$-PAST) example, consider the following ``continuous random walk'': $\Xi \, 10$ where
+For a more complex (not $\tY$-PAST) example, consider the following ``continuous random walk'': $\Xi \, \underline{10}$ where
 \[
 %\underbrace{\big(
-\Xi := \tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample + 1/2)} 
+\Xi := \tY \, \lambda f x . \tif{x \leq 0}{0}{f(x - \tsample + 1/2)} 
 %\big)}_{\Xi} \, 10
@@
 \[
-\Xi \, l 
+\Xi \, \underline l 
 \mapsto 
 g(l), \qquad 
-0 \mapsto 0
+\underline 0 \mapsto 0
 \]
@@
 \underbrace{\big
-(\tY \, f \, x . 
+(\tY \, \lambda f x .\ 
 \tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)}_{\Xi} 
-\, 10
+\, \underline{10}
 \]
@@
 \underbrace{\big
-(\tY \, f \, x . \,
+(\tY \, \lambda f x . \,
 0 \oplus_{\frac{1}{x+1}} f(x + 1) \big)}_{\Xi} 
-\, 10
+\, \underline{10}
 \]
@@
 
-Consider the candidate ranking function $\Xi \, n \mapsto n + 1$.
-For each $n$, $\Xi \, n$ reduces to either $0$ (with probability $\frac 1 {n + 1}$) or $\Xi \, (n + 1)$, therefore the expected value of the ranking function is $\frac{n(n+2)}{n+1} = n + 1 - \frac 1 {(n+1)^2}$, and the required decrease is $\frac{\epsilon(0) + n \epsilon(n + 2)}{n + 1}$, therefore eventually, the expected decrease isn't enough, whatever the value of $\epsilon(0)$ is.
+Consider the candidate ranking function $\Xi \, \underline n \mapsto n + 1$.
+For each $n$, $\Xi \, \underline n$ reduces to either $\underline 0$ (with probability $\frac 1 {n + 1}$) or $\Xi \, \underline{n + 1}$, therefore the expected value of the ranking function is $\frac{n(n+2)}{n+1} = n + 1 - \frac 1 {(n+1)^2}$, and the required decrease is $\frac{\epsilon(0) + n \epsilon(n + 2)}{n + 1}$, therefore eventually, the expected decrease isn't enough, whatever the value of $\epsilon(0)$ is.
 
@@
 \begin{align*}
-\Theta[n] &\mapsto n \\
-\Xi \, 10 & \mapsto 12.
+\Theta[\underline n] &\mapsto n \\
+\Xi \, \underline{10} & \mapsto 12.
 \end{align*}
@@
 Within $X$, there is necessarily a redex at $\cbv(X)$.
-The only conditions that $@_2^k;\cbv(X)$ is also a position of a redex in the whole term are that if the redex is a $\tsample$, it is not inside a $\lambda$ or $\tY$ on the right of an application, which is true because $\cbv$ never selects a redex inside of a $\lambda$.
+The only conditions that $@_2^k;\cbv(X)$ is also a position of a redex in the whole term are that if the redex is a $\tsample$, it is not inside a $\lambda$ that's inside a $\tY$ or on the right of an application, which is true because $\cbv$ never selects a redex inside of a $\lambda$.
 After $n$ reaches $0$, the term is $F \underline{n_0} (\dots(F\ \underline {n_k}\ \underline x)\dots)$, then the innermost $F$ is evaluated completely in $\cbv$ order, therefore (by assumption) it terminates.

### 2021-01-21 LICS21/confluent-proofs.tex
Avoid formula exceeding column width
Various minor fixes in the appendices

@@
 There are only multiple $\sim_\downarrow$ steps in the result if the position of the reduction $B \to C$ is inside the position of the reduction $B \to A$, and similarly, there are only multiple $\sim_\uparrow$ steps in the result if the position of the reduction $B \to A$ is inside the position of the reduction $B \to C$.
-If there are multiple $\sim_\downarrow$ steps, they are not necessarily parallel, but in the only case where they aren't (case f of $sim_c$), the subsequence of $\sim_\downarrow$ steps with the reductions at $\alpha;\lambda;@_1;\beta$ then $\alpha;\lambda;@_1;\gamma_i;\tY;\lambda;\beta$ for each $i$ where $\gamma_i \geq \beta$ (in the notation of the definition of $\sim_c$) can be replaced by $\sim_\downarrow$s with the reductions at $\alpha;\lambda;@_1;\gamma'_i;\tY;\lambda;\beta$ then $\alpha;\lambda;@_1;\beta$, then another $\sim_c$, where $(\gamma'_i)$ is the list of positions in $A$ (\emph{before} the reduction at $\beta$ to $A'$, unlike $(\gamma_i)$) where $A | \gamma'_i = x$ and $\gamma'_i > \beta$.
+If there are multiple $\sim_\downarrow$ steps, they are not necessarily parallel, but in the only case where they aren't (case f of $\sim_c$), the subsequence of $\sim_\downarrow$ steps with the reductions at $\alpha;\lambda;@_1;\beta$ then $\alpha;\lambda;@_1;\gamma_i;\tY;\lambda;\beta$ for each $i$ where $\gamma_i \geq \beta$ (in the notation of the definition of $\sim_c$) can be replaced by $\sim_\downarrow$s with the reductions at $\alpha;\lambda;@_1;\gamma'_i;\tY;\lambda;\beta$ then $\alpha;\lambda;@_1;\beta$, then another $\sim_c$, where $(\gamma'_i)$ is the list of positions in $A$ (\emph{before} the reduction at $\beta$ to $A'$, unlike $(\gamma_i)$) where $A | \gamma'_i = x$ and $\gamma'_i > \beta$.
 
@@
 \begin{lemma} \label{canonicalCousins}
-For any reduction sequence $X$, there is a unique reduction sequence $X_{cbv}$ that is related to $X$ by $\sim_c$ and is in call-by-value order, and if $X \sim_c^* Y$, then $X_{cbv} = Y_{cbv}$.
+For any reduction sequence $X$, there is a unique reduction sequence $X_{cbv}$ that is related to $X$ by $\sim_c^*$ and is in call-by-value order, and if $X \sim_c^* Y$, then $X_{cbv} = Y_{cbv}$.
 \end{lemma}
@@
 Now consider the specific sequence of positions of the reductions in $X$ involved in either of these steps.
-Let the last of these positions be $\alpha$, the first $\gamma$, and the second-last, third-last and so on $\beta_1, \beta_2, \dots$ respectively, and let the positions of the redexes in $Y$ involved in $Y \sim_c X$ be $\beta$ then $(\gamma_i)$ in order..
+Let the last of these positions be $\alpha$, the first $\gamma$, and the second-last, third-last and so on $\beta_1, \beta_2, \dots$ respectively, and let the positions of the redexes in $Y$ involved in $Y \sim_c X$ be $\beta$ then $(\gamma_i)$ in order.
 Then $X \sim_c X_1$ swaps $\beta_1$ and $\alpha$, and $\alpha$ comes before $\beta_1$ in CBV order (i.e.~$\alpha < \beta_1$ or $\alpha$ is left of $\beta_1$), and the step $Y \sim_c X$ either proceeds forwards towards CBV order, swapping $\beta$ below $\gamma_1 = \gamma$ to form $\gamma$ then $(\beta_i)$, or proceeds backwards, taking $Y$ even farther from CBV order, swapping the $(\gamma_i)$ to above $\beta$, forming $\gamma$ then $\beta_1 = \beta$.
@@
 They do not depend on $\alpha$ or $\gamma$, so that if both initial positions were moved somewhere else the relative positions of the images of $\gamma$ would be unaffected, and similarly if some position within $\gamma$ were swapped with $\alpha$, the relative positions of its images would be the same (except to the extent that, in case f, the positions of the relevant variable are taken after the inner reduction takes place, so that some of those may still change).
-Because there are so many of them and they don't actually affect the multiset of positions where reductions take place, I will be ignoring swaps by case a in the proof of this case, and just assuming that all the final positions end up in the correct order.
+Because there are so many of them and they don't actually affect the multiset of positions where reductions take place, we will be ignoring swaps by case a in the proof of this case, and just assuming that all the final positions end up in the correct order.
 
-  The final set of reduction positions that I will be proving both $(X_i)_i$ and $(Y_j)_j$ reach is $\alpha, (\alpha;\delta_i;\gamma', (\alpha;\delta_i;\gamma';\epsilon_j;\beta')_j)_i$, where as usual the sequences are expanded out to the full list.
+  The final set of reduction positions that we will be proving both $(X_i)_i$ and $(Y_j)_j$ reach is $\alpha, (\alpha;\delta_i;\gamma', (\alpha;\delta_i;\gamma';\epsilon_j;\beta')_j)_i$, where as usual the sequences are expanded out to the full list.
 For the sequence $(X_i)_i$, first the images of $\beta$ in $X_0$ are $(\gamma,\epsilon_i,\beta')$.
@@
 First $\gamma$ swaps with $\alpha$, yielding $\alpha$ and $(\alpha;\delta_i;\gamma')_i$, then $\beta$ swaps with $\alpha$, yielding one image of $\beta$ for each $\delta_i$, except that, in case f, those images at positions inside $\alpha;\lambda;@_1;\gamma'$ may differ because they are earlier in the reduction sequence than any of the images of $\gamma$.
-Each of the images of $\beta$ in turn it moved to the corresponding image of $\gamma$, except that those overlapping with $\alpha;\lambda;@_1;\gamma'$ may be duplicated or deleted on the way.
+Each of the images of $\beta$ in turn is moved to the corresponding image of $\gamma$, except that those overlapping with $\alpha;\lambda;@_1;\gamma'$ may be duplicated or deleted on the way.
 The images of $\beta$ after just this process (which does not actually correspond to any term in $(Y_j)_j$, but it's sufficiently independent that the order doesn't matter that much) are $(\alpha;\delta_i;\gamma';\beta'')$ (where $\gamma;\beta'' = \beta$).
@@
 The term $M \mid \alpha_1$ can't be $\tsample$, because it has strict subterms, but the case that $M \mid \alpha_2 = \tsample$ is still somewhat more complicated. 
-$\alpha_2$ can't be within $\alpha_1 ; @_2$ or $\alpha_1 ; Y$ (cases e and f of $\sim_c$) because those are not valid positions to reduce a $\tsample$, 
-and the case that $\alpha_2$ is within the branch of an if statement that is deleted by the reduction at $\alpha_1$ (case b) doesn't actually present a problem because there is no corresponding reduction to $M \to M_2$ in the other branch,
+$\alpha_2$ can't be within $\alpha_1 ; @_2$ or $\alpha_1 ; \tY ; \lambda$ (cases e and f of $\sim_c$) because if $\alpha_2 > \alpha_1 ; @_2$, $M \mid \alpha_1;@_2$ must be a value, therefore $\alpha_2 > \alpha_1;@_2;\lambda$ and those are not valid positions to reduce a $\tsample$.
+The case that $\alpha_2$ is within the branch of an if statement that is deleted by the reduction at $\alpha_1$ (case b) doesn't actually present a problem because there is no corresponding reduction to $M \to M_2$ in the other branch,
 which leaves cases c and d, that $\alpha_2$ is in the $\tif{\cdot}{\cdot}{\cdot}$ branch that isn't deleted, and that $\alpha_2 < \alpha_1 ; @_1 ; \lambda$.  
@@
 If $\kappa$ is disjoint from $\beta$, then $(N,\zeta) \sim_p (J,\zeta) \sim_p (O_2,\alpha;\kappa') = (O_2;\theta)$ by cases 1 and 2 of $\sim_p$.
-In the other case, that $\kappa$ isn't disjoint from $\beta$, $\kappa > \beta$ because none of the positions $<= \alpha;@_1;\lambda;\beta$ in $N$ are related to any position in $I$ by $\sim_p$.
+In the other case, that $\kappa$ isn't disjoint from $\beta$, $\kappa > \beta$ because none of the positions $\leq \alpha;@_1;\lambda;\beta$ in $N$ are related to any position in $I$ by $\sim_p$.
 As $(I,\alpha;\kappa) \sim_p (O_1,\alpha;\kappa')$ (with the redex at $\alpha;\beta$), for exactly the same reason $(N,\alpha;@_1;\lambda;\kappa) \sim_p (J,\alpha;@_1;\lambda;\kappa')$ (with the redex at $\alpha;@_1;\lambda;\beta$).
@@
 \begin{proof}
-If the result holds for all sets $S$ of the form $\{s \in I^{L_s(N)} \mid \forall j : s(j) \in x_i\}$, where $(x_j)_{j \in J}$ is a family of measurable subsets of $I$ indexed by some finite set $J \subset L_s(N)$ of positions, it also holds for all other $S$ by taking limits and disjoint unions.
+If the result holds for all sets $S$ of the form $\{s \in I^{L_s(N)} \mid \forall j : s(j) \in x_i\}$, where $(x_j)_{j \in J}$ is a family of measurable subsets of $I$ indexed by some finite set $J \subset L_s(N)$ of potential positions, it also holds for all other $S$ by taking limits and disjoint unions.
 
-Take such a $(x_j)_{j \in J}$, where $x_j = x_k$ and define $K = i(M \to N)[J]$.
+Take such a $(x_j)_{j \in J}$, and define $K = i(M \to N)[J]$.
 Because $i(M \to N)$ is injective, it defines a bijection between $J$ and $K$.
@@
 Let\begin{align*}
-R_s & = \{\mathit{Sk}(O) \mid O \in \mathit{Rch}(M), r \in \mathbb R^k, O | f(O) = \tsample, O \to N[r]\} \\
-R_d & = \{\mathit{Sk}(O) \mid O \in \mathit{Rch}(M), r \in \mathbb R^k, O | f(O) \neq \tsample, O \to N[r]\}.
+R_s & = \{\mathit{Sk}(O) \mid O \in \mathit{Rch}(M), r \in \mathbb R^k, O | f(O[r]) = \tsample, O \to N[r]\} \\
+R_d & = \{\mathit{Sk}(O) \mid O \in \mathit{Rch}(M), r \in \mathbb R^k, O | f(O[r]) \neq \tsample, O \to N[r]\}.
 \end{align*}
@@
 $g_O(r) = r'$ if $O[r] \to N[r']$ for $O \in R_d$, and $g_O(s, r) = r'$ if $O[r] \to N[r']$ with $s$ as the value the sample takes in the reduction.
-All of these functions are measurable, as they are just rearrangements of vector components, or in the case that $O|f(O) = \underline x(\skeletonPlaceholder, \dots, \skeletonPlaceholder)$, a combination fo the measurable function $x$ with a rearrangement of vector components.
+All of these functions are measurable, as they are just rearrangements of vector components, or in the case that $O|f(O[r]) = \underline x(\skeletonPlaceholder, \dots, \skeletonPlaceholder)$, a combination of the measurable function $x$ with a rearrangement of vector components.
 In the case that $O \in R_s$, the function $g_O$ simply inserts the sample into the vector of reals at a certain index, call it $h_O$.
@@
 = & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}:\\
-    & g_O(r,s'(O,f(O))) \in T, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
+    & g_O(r,s'(O,f(O[r]))) \in T, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O[r])))]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
      + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}:\\
@@
 = & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)}:\\
-    & s'(O,f(O)) \in T_{h(O)}, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
+    & s'(O,f(O[r])) \in T_{h(O)}, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O[r])))]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
     + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}:\\
@@
 = & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
-    & \mu(\{s' \in I^{L_s(O)} \mid s'(O,f(O)) \in T_{h(O)}, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O)))]) \in S\})\\
+    & \mu(\{s' \in I^{L_s(O)} \mid s'(O,f(O[r])) \in T_{h(O)}, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O[r])))]) \in S\})\\
     + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
@@
     + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : s'(O,f(O)) \in T_{h(O)}, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S)\\
+= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : s'(O,f(O[r])) \in T_{h(O)}, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S)\\
     + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
-= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \mathbb R^{k-1}, s' \in I^{L_s(O)} : g_O(r,s'(O,f(O))) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S)\\
+= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \mathbb R^{k-1}, s' \in I^{L_s(O)} : g_O(r,s'(O,f(O[r]))) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S)\\
     + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
@@
 
-Take some order $p_1, \dots$ on the elements of $P$, and define $T_n$ recursively as follows.
-If the sample at $p_n$ is eventually used in the sequence $(M,(t,s|_P)) \Rightarrow^*_r \dots$ for some positive measure subset of $T_{n-1}$, then let $T_n$ be the subset of $T_{n-1}$ where it is used within $m$ steps, for the minimal $m$ such that this is of positive measure.
+Take some order $p_1, \dots$ on the elements of $P$, and define $T_n \subset I^{L_s(M)}$ recursively as follows.
+If the sample at $p_n$ is eventually used in the sequence $(M,(t,s|_P)) \Rightarrow^*_r \dots$ for some positive measure subset of the traces $t$ in $T_{n-1}$, then let $T_n$ be the subset of $T_{n-1}$ where this sample is used within $m$ steps, for the minimal $m$ such that this is of positive measure.
 Otherwise, let $T_n$ be the subset of $T_{n-1}$ where the sample at $p_n$ is never used.
@@
 
-Let $T' = T \times \prod_{p \in P_i} I \times \prod{p \in P \setminus P_i} \{s(p)\}$.
-Although $T'$ itself may have a measure of 0 as a subset of $I^{L_s(M)}$, it still has a natural measure of its own.
+Let $T' = T \times \prod_{p \in P_i} I \times \prod_{p \in P \setminus P_i} \{s(p)\}$.
+Although $T'$ itself may have a measure of 0 as a subset of $I^{L_s(M)}$, it still has a natural measure-space structure of its own.
 It also has the properties that $M$ does not terminate with $\cbv$ and any trace in $T'$ (therefore $M$ also does not terminate with $r$ and any trace in $T'$, by \Cref{thm:CbvIsTerminatingest}), and for any trace $t \in T'$, if $(M,t) \Rightarrow^k_r (M',t')$, none of the potential positions in $M'$ correspond to any of the elements of $P \setminus P_i$.
@@
 \to^{\phantom *} &\ \Xi[\lambda x. \underbrace{x \times \dots \times x \times}_n \underline 1]\ \underline n \\
-\to^{\phantom *} &\ \tif{\underline n = 0}{\underline 0}{\underline{n^n} + \Xi[\lambda x. x \times \dots \underline 1]\ \underline{n-1}} \\
 \to^* &\ \underline{n^n} + \underline{(n-1)^n} + \dots + \underline 0 \\
@@
 Providing a ranking function of some sort for this term in a similar level of detail would have been rather more complicated using only the standard reduction strategy.
-The recursion in $\Theta$ would have had to be evaluated separately for every time it was used, so there would have been more (and more complex) terms to consider in the reduction sequence.
+The recursion in $\Theta$ would have to be evaluated separately for every time it was used, so there would have been more (and more complex) terms to consider in the reduction sequence.
 Also, the number of $\tY$-reductions from $P[\underline n]$ would have been $n^2 + 2n + 1$ instead of $2n + 2$, therefore the expected number of $\tY$ reductions starting from $M$ would be infinite, i.e.~$M$ is not $\tY$-PAST, therefore it is not even rankable.
 It would be possible to define an antitone sparse ranking function for it, but it is much more difficult to construct:
-\iffalse
-\begin{align*}
-& M \mapsto 1.47 \\
-& P[\underline n] \mapsto 1 + \ln((n+1)n + 1) \\
-& \Xi[\lambda x. \Theta[x] \underline n] \underline n \mapsto 1 + \ln((n+1)n + 1) \\
-& \underline{n^n} + \dots + \underline{k_1^n} + \\
-& \Xi[\lambda x. \Theta[x] \underline n] (\underline{k_1} - 1) \mapsto 1 + \ln((n+1)(k_1-1)+1) \\
-& \underline{n^n} + \dots + \underline{(k_1+1)^n} + \Theta[\underline{k_1}] \underline{n} + \\
-& \Xi[\lambda x. \Theta[x] \underline n] (\underline{k_1} - 1) \mapsto 1 + \ln((n+1)k_1 + 1) \\
-& \underline{n^n} + \dots + \underline{(k_1+1)^n} + \\
-& \underbrace{\underline{k_1} \times \dots \times \underline{k_1} \times}_{n-k_2+1} \Theta[\underline{k_1}] (\underline{k_2} - 1) + \\
-& \Xi[\lambda x. \Theta[x] \underline n] (\underline{k_1} - 1) \mapsto 1 + \ln((n+1)k_1 + k_2 + 1) \\
-& \underline{\sum_{k=0}^n k^n} \mapsto 0.
-\end{align*}
-\fi
 \begin{align*}
@@
 \Xi[\lambda x. \Theta[x] \, \underline n] \, \underline n & \mapsto 1 + \ln((n+1)n + 1) \\
-%\underline{n^n} + \dots + \underline{k_1^n} + \\
-%\Xi[\lambda x. \Theta[x] \underline n] (\underline{k_1} - 1) & \mapsto 1 + \ln((n+1)(k_1-1)+1) \\
 \left.\begin{array}{r}
@@
 & \mapsto 1 + \ln((n+1)(k_1-1)+1) \\
-%\underline{n^n} + \dots + \underline{(k_1+1)^n} + \Theta[\underline{k_1}] \underline{n} + \\
-%\Xi[\lambda x. \Theta[x] \underline n] (\underline{k_1} - 1) & \mapsto 1 + \ln((n+1)k_1 + 1) \\
 \left.\begin{array}{r}
 \underline{n^n} + \dots + \underline{(k_1+1)^n} +  {}\\
-\Theta[\underline{k_1}] \, \underline{n} + \Xi[\lambda x. \Theta[x] \, \underline n] (\underline{k_1} - 1) 
+\Theta[\underline{k_1}] \, \underline{n} +  {}\\
+\Xi[\lambda x. \Theta[x] \, \underline n] (\underline{k_1} - 1) 
 \end{array}\right\}
 & \mapsto 1 + \ln((n+1)k_1 + 1) \\
-%\underline{n^n} + \dots + \underline{(k_1+1)^n} + \\
-%\underbrace{\underline{k_1} \times \dots \times \underline{k_1} \times}_{n-k_2+1} \Theta[\underline{k_1}] (\underline{k_2} - 1) + \\
-%\Xi[\lambda x. \Theta[x] \underline n] (\underline{k_1} - 1) & \mapsto 1 + \ln((n+1)k_1 + k_2 + 1) \\
 \left.\begin{array}{r}\underline{n^n} + \dots + \underline{(k_1+1)^n} + {} \\
-\underbrace{\underline{k_1} \times \dots \times \underline{k_1} \times}_{n-k_2+1} \Theta[\underline{k_1}] (\underline{k_2} - 1) + {} \\
+\underbrace{\underline{k_1} \times \dots \times \underline{k_1} \times}_{n-k_2+1} {} \\
+\Theta[\underline{k_1}] (\underline{k_2} - 1) + {} \\
 \Xi[\lambda x. \Theta[x] \, \underline n] (\underline{k_1} - 1)

### 2021-01-21 LICS21/confluent.tex
Various minor fixes in the appendices

@@
 The reduction relation $\Rightarrow$ is nondeterministic, so it admits multiple possible reduction strategies. 
-A \emph{reduction strategy} starting from a closed term $M$ is a partial function $f$ from $\mathit{Rch}(M)$ to positions, such that for any reachable term $N$ where $f$ is defined, $f(N)$ is a position of a redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
+A \emph{reduction strategy} starting from a closed term $M$ is a measurable partial function $f$ from $\mathit{Rch}(M)$ to positions, such that for any reachable term $N$ where $f$ is defined, $f(N)$ is a position of a redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
 Using a reduction strategy $f$, a subset of $\Rightarrow$ that isn't nondeterministic, $\Rightarrow_f$, can be defined by $(N,s) \Rightarrow_f (N',s')$ just if $(N,s) \Rightarrow (N',s')$ and $N$ reduces to $N'$ with the redex at $f(N)$.
@@
 For each key reachable term where the sparse ranking function is defined, simply decide which sub-term should be reduced next, and how far before the next key reachable term.
-The only restrictions to bear in mind on evaluating sub-terms are that a $\tsample$ may not be evaluated inside a $\lambda$ or $\tY$ on the right of an application, and a $\beta$ redex may not be reduced until its argument is a value.
+The only restrictions to bear in mind on evaluating sub-terms are that a $\tsample$ may not be evaluated inside a $\lambda$ that's inside a $\tY$ or on the right of an application, and a $\beta$ redex may not be reduced until its argument is a value.
 This can be seen as applied to \Cref{ex:sum of powers} at the end of \Cref{apx:confluent}.

### 2021-01-21 LICS21/ranking-proofs.tex
Various minor fixes in the appendices

@@
 The restrictions of $f$ and $g$ to $Rch(N)$ have the same properties assumed of $f$ and $g$, so assume w.l.o.g.~that $N=M$.
-The difference $g - f$ is then a supermartingale (with the same setup as in the proof of %\Cref{thm:rankable gives supermartingale}
-%Theorem \ref{rankable implies ast}) 
+The difference $g - f$ is then a supermartingale (with the same setup as in
 \Cref{thm:rankable and strict rankable});
@@
 
-Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) := \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$.
+Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) := \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY\, \lambda x.
+N'],s') \} \right |$.
 The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on sparse ranking functions.

### 2021-01-21 LICS21/supermartingales-proofs.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
Changed bib style to IEEEtran; \citep -> \cite

Repositioned bibliography
Various minor fixes in the appendices

@@
 %and the filtration $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
-The $\calF_n$-measurability of $M_n$ (and hence of $\#\mathrm{draw}_n$) follows from \citep{DBLP:conf/icfp/BorgstromLGS16}.
+The $\calF_n$-measurability of $M_n$ (and hence of $\#\mathrm{draw}_n$) follows from \cite{DBLP:conf/icfp/BorgstromLGS16}.
 %$\#\mathrm{draw}_n$ is a stopping time (bounded by $n$). 
@@
 f(M_{n+1})[s \in \mathbf{T}_2] = \sum_{i \in \calI} 
-f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])[s \in U_i]
+f(E_i[\sigma_i(s)][\underline{\rho(s)}])[s \in U_i]
 \label{eqn:f 2 n+1}
@@
 \item $\set{U_i}_{i \in \calI}$ is a partition of $\mathbf{T}_2$, 
-where each $U_i$ is determined by a skeletal term $E_i[\cdot][\tsample]$ and a number (of draws) $l_i \leq n$, satisfying
-\(
-s \in U_i
-\iff
-(E_i[\underline{r}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s),
-\;
-\hbox{for some }r \in \Real^{j_i}.
-\)
+where each $U_i$ is determined by a skeletal environment $E_i$ and a number (of draws) $l_i \leq n$, so that $U_i$ is the set of traces where after $n$ reduction steps, $l_i$ samples have been used, and the term has reached $E_i[r][\tsample]$ for some $r \in \Real^{j_i}$.
+%\(
+%s \in U_i
+%\iff
+%(E_i[\underline{r}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s),
+%\;
+%\hbox{for some }r \in \Real^{j_i}.
+%\)
 (We use the (measurable) function $\sigma_i : \entrosp \to \Real^{j_i}$ to skolemise the existentially quantified $r$.)
@@
 \[
-U_i := \#\mathrm{draw}_n^{-1}(l_i) \cap M_n^{-1}(E_i[\cdot][\tsample], \Real^{j_i}) \in \calF_n.
+U_i := \#\mathrm{draw}_n^{-1}[l_i] \cap M_n^{-1}[\{E_i[r][\tsample] \mid r \in \Real^{j_i}\}] \in \calF_n.
 \]
@@
 \[%\textstyle
-\int_I \Leb(\dif r) \, f(E_i[\underline{\widehat{\sigma_i}(u)}][\underline{r}])) \leq f(E_i[\underline{\widehat{\sigma_i}(u)}][\tsample])).
+\int_I \Leb(\dif r) \, f(E_i[\widehat{\sigma_i}(u)][\underline{r}])) \leq f(E_i[\widehat{\sigma_i}(u)][\tsample])).
 \]
@@
 \begin{align*}
-& \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, \int_I \textrm{Leb}(\dif r) \, f(E_i[\underline{\widehat{\sigma_i}(u)}][\underline{r}])) \\
-& \leq \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, f(E_i[\underline{\widehat{\sigma_i}(u)}][\tsample])).
+& \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, \int_I \textrm{Leb}(\dif r) \, f(E_i[\widehat{\sigma_i}(u)][\underline{r}])) \\
+& \leq \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, f(E_i[\widehat{\sigma_i}(u)][\tsample])).
 \end{align*}
@@
 \begin{align*}
-& \int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\underline{\widehat{\sigma_i}(u')}][\underline{\widehat{\rho}(u')}])) \\
+& \int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\widehat{\sigma_i}(u')][\underline{\widehat{\rho}(u')}])) \\
 & \leq 
-\int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\underline{\widehat{\sigma_i}(u')}][\tsample]))
+\int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u') \, f(E_i[\widehat{\sigma_i}(u')][\tsample]))
 \end{align*}
@@
 \begin{align}
-& \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])) \nonumber \\
+& \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\sigma_i(s)][\underline{\rho(s)}])) \nonumber \\
 & \leq 
-\int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\tsample])).
+\int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\sigma_i(s)][\tsample])).
 \label{eqn:a u ui}
@@
 \begin{align*}
-& \int_A \mu_{\entrosp}(\dif s) \, f_{n+1}(M)[s \in \mathbf{T}_i] \\
+& \int_A \mu_{\entrosp}(\dif s) \, f(M_{n+1})[s \in \mathbf{T}_2] \\
 &= 
-\int_A \mu_{\entrosp}(\dif s) \, \sum_{i \in \calI} f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])[s \in U_i] \\
+\int_A \mu_{\entrosp}(\dif s) \, \sum_{i \in \calI} f(E_i[\sigma_i(s)][\underline{\rho(s)}])[s \in U_i] \\
 &= 
-\sum_{i \in \calI} \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])\\
+\sum_{i \in \calI} \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\sigma_i(s)][\underline{\rho(s)}])\\
 &\leq 
-\sum_{i \in \calI} \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\tsample])
+\sum_{i \in \calI} \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\sigma_i(s)][\tsample])
 \quad \hbox{$\because$ (\ref{eqn:a u ui})}\\
 &= 
-\int_{A} \mu_{\entrosp}(\dif s) \, \sum_{i \in \calI} \, f(E_i[\underline{\sigma_i(s)}][\tsample])[s \in U_i]\\
+\int_{A} \mu_{\entrosp}(\dif s) \, \sum_{i \in \calI} \, f(E_i[\sigma_i(s)][\tsample])[s \in U_i]\\
 &= \int_{A} \mu_{\entrosp}(\dif s) f_{n}(M)[s \in \mathbf{T}_2]
@@
 \transform{\tif{B}{M_1}{M_2}} &:= \big(\lambda z . (\transform{M_1} + \transform{M_2})\big) \, \transform{B} \\
-\transform{\tY \, f \, y . N} &:= \lambda y . \transform{N}[\bot / f]
+\transform{\tY N} &:= (\lambda y . \bot) \transform{N}
 \end{align*}
@@
 
-It is straightforward to see that for every trace $s \in \entrosp$, there is a reduction sequence of $\transform{M}$ that simulates (an initial subsequene of) the reduction of $M$ under $s$;
-and the simulation can be made lockstep.
+It is straightforward to see that for every trace $s \in \entrosp$, there is a reduction sequence of $\transform{M}$ that simulates (an initial subsequene of) the reduction of $M$ under $s$.
 
@@
 To see $\set{T_M^\tY = n} \in \calF_{T_n}$ for a given $n$, we need to show that $\set{T_M^\tY = n} \cap \set{T_n \leq i} \in \calF_i$, for all $i \in \omega$.
-Let $\cal V$ be the measurable subset of $\Lambda^0$ given by the countable union of $\set{\alpha_j} \times \mathbb{R}^j$ where $\alpha_j$ ranges over skeletons which are values with $j$ place-holders, and $j$ ranges over $\omega$.
-Let $\cal W$ be the measurable subset of $\Lambda^0$ given by the countable union of $\set{\beta_j} \times \mathbb{R}^j$ where $\beta_j$ ranges over skeletons which are not values, and with $j$ place-holders, and $j$ ranges over $\omega$.
-%Then $\set{T_M^\tY = n} \cap \set{T_n \leq i}$ is given by
+Let $\cal V$ be the set of values in $\Lambda^0$ (which is measurable), and let $\cal W$ be the non-values.
 Then, it suffices to observe that $\set{T_M^\tY = n} \cap \set{T_n \leq i}
 = \bigcup_{l=n}^i 
-\big( M_l^{-1} ({\cal V}) \cap  
-M_{l-1}^{-1} ({\cal W}) \cap 
+\big( M_l^{-1} [{\cal V}] \cap  
+M_{l-1}^{-1} [{\cal W}] \cap 
 \set{T_n = l} \big)
-$.
+$, where each of $M_l^{-1} [{\cal V}]$, $M_{l-1}^{-1} [{\cal W}]$, and $\set{T_n = l}$ is in $\calF_i$.
 \end{proof}

### 2021-01-21 LICS21/antitone-proofs.tex
Add a simplified non-affine and continuous example.

@@
 
-\Cref{ex:escaping spline}
+\begin{example}[Escaping spline]
+\label{ex:escaping spline}\cite[\S 5.4]{DBLP:journals/pacmpl/McIverMKK18}
 \[
@@
 These are satisfied by setting $\epsilon(x) = \min(1, 1/x)$, therefore this term is AST.
+\end{example}
 
@@
 It is basically another variant of the biased random walk, except that the relevant variable is the number of copies of $\Xi$, instead of a real number in the term.
+
+\Cref{ex:non-affine continuous}
+Let
+\begin{align*}
+\Xi & = \tY\, \lambda f x.\, (\lambda e. X[e]) \tsample \\
+X[e] & = \tif{e \leq p - x^{-2}}{x+1}{f^2(x+1) \oplus_e f(x+1)}
+\end{align*}
+
+Not only is this program non-affine recursive,
+note also the use of a random sample, $e$, as a first-class value, and the subsequent use as a probability in the binary choice $\oplus_e$.
+Such a computation cannot be modelled via discrete distributions.
+
+We can use the ranking function method (coupled with the solution of linear recurrence relations) to show that provided $p \geq \frac{5 - \sqrt{21}}{2}$, the program $\Xi \, \underline 3$ is AST.
+Consider the edge case, that $p = \frac{5 - \sqrt{21}}{2}$ exactly.
+
+The term $\Xi^n\ \underline x$ (for $n > 0$, $x^{-2} < p$) reduces to either $\Xi^{n-1}\ \underline{x+1}$, $\Xi^n\ \underline{x+1}$ or $\Xi^{n+1}\ \underline{x+1}$, with probabilities $p-x^{-2}$, $(1-p+x^{-2})\frac{p-x^{-2}}{2}$ and $(1-p+x^{-2})(1-\frac{p-x^{-2}}{2})$ respectively.
+Let $a(\Xi^n\ \underline x) = n + 2x^{-1}$.
+This is a supermartingale (because the decrease in $2x^{-1}$ as $x$ increases is enough to offset the average increase in $n$), but it does not satisfy the antitone-strict progress condition.
+It does however have a bounded-below variance, so the usual method of using $\ln a$ instead of just $a$ works.
+Calculating the exact amount that $\ln a$ decreases is not necessary, because it can be bounded as follows: $a$ changes by at least $1/2$ with probability at least $1/11$ (assuming $x \geq 3$), therefore $\ln(a)$ decreases in expectation by at least $\frac{1}{11}(\ln(a+\frac{1}{2}) - \ln(a) -\frac 1 {2a})$, using the fact that a linear approximation to $\ln$, applied to $a$, at least doesn't increase, then adding the deviation of $\ln$ from its linear approximation $\ln(a) + \frac{x-a}{a}$ (and using the fact that the linear approximation is everywhere an overestimate), we obtain a sufficiently strong bound on the decrease of $\ln(a)$ that it's an antitone sparse ranking function.
+(It can obviously be extended to $\Xi\ \underline 1$ too, which reduces to $\Xi^n\ \underline 3$ or a value after a bounded number of steps, but that complicates the analysis a little.)
 

### 2021-01-21 LICS21/confluent-proofs.tex
Align things to reduce the amount of empty page.

@@
 If a reduction sequence includes one branch of one of these cases at some point in it, then the positions in the reduction sequence obtained by substituting in the other branch are all related by $\sim_c$ to the corresponding positions in the original reduction sequence.
-\include{sim-diagram}
-%The diagram is not actually in the figure environment because that doesn't work properly with it extending over several pages. The empty figure is just needed to get the label and caption.
-\begin{figure}[htb]
-\caption{Illustration of the $\sim_c$ rules. \label{sim-diagram}
-}
-\end{figure}
 
@@
 Case a is also similar, except that the relevant positions are all disjoint rather than contained in one another, and either branch could be considered the right branch.
+
+\include{sim-diagram}
+%The diagram is not actually in the figure environment because that doesn't work properly with it extending over several pages. The empty figure is just needed to get the label and caption.
+\begin{figure}[htb]
+\caption{Illustration of the $\sim_c$ rules. \label{sim-diagram}
+}
+\end{figure}
 
@@
 In the case that $O \in R_s$, the function $g_O$ simply inserts the sample into the vector of reals at a certain index, call it $h_O$.
-\clearpage
+
 \begin{align*}

### 2021-01-21 LICS21/intro.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper Sec. I: tweak
Merge branches 'master' and 'master' of https://github.com/4Denthusiast/prob-programming-termpaper Added reference to more complex non-affine recursion example.

@@
 
-Moreover various theorems about probabilistic programs rely on the assumption that the program terminates almost surely. 
+Moreover various theorems about probabilistic programs rely on the assumption that the program terminates almost surely (see e.g.~\cite{DBLP:conf/uai/GoodmanMRBT08}). 
 A recent result \cite{MakOPW20} proves that AST programs have density functions that are differentiable almost everywhere.
@@
 
-By combining these methods we can prove AST of a variety of PPCF programs that are beyond the reach of methods in the literature, including non-affine recursion\footnote{\label{fnote:non-affine} Non-affine recursive programs are recursive programs that can, during the evaluation of the recursive body, make multiple recursive calls (of a first-order function) from distinct call sites.} (\cref{ex:non-affine recursion easy}) and recursions that define high-order functions (\cref{ex:higher-order recursion}).
+By combining these methods we can prove AST of a variety of PPCF programs that are beyond the reach of methods in the literature, including non-affine recursion\footnote{\label{fnote:non-affine} Non-affine recursive programs are recursive programs that can, during the evaluation of the recursive body, make multiple recursive calls (of a first-order function) from distinct call sites.} (\cref{ex:non-affine recursion easy,ex:non-affine continuous}) and recursions that define high-order functions (\cref{ex:higher-order recursion}).
 

### 2021-01-21 LICS21/SPCF.tex
Minor; letter case

@@
 
-The language PPCF is a call-by-value (CBV) version of PCF with sampling of real numbers from the closed interval $[0,1]$ \cite{Ehrhard2018c,DBLP:journals/pacmpl/EhrhardPT18,MakOPW21}.
+The language PPCF is a call-by-value (CBV) version of PCF \cite{} with sampling of real numbers from the closed interval $[0,1]$ \cite{Ehrhard2018c,DBLP:journals/pacmpl/EhrhardPT18,MakOPW21}.
 %\changed[lo]{following \cite{MakOPW21} except the score function (\Cref{rem:score})}. 

### 2021-01-21 LICS21/antitone.tex
Add a simplified non-affine and continuous example.
Minor; letter case

@@
 
-\begin{example}[Escaping spline]
-\label{ex:escaping spline}\cite[\S 5.4]{DBLP:journals/pacmpl/McIverMKK18}
-\[
-%\underbrace{
-\big(\tY\ \lambda f x . \,
-\underline 0 \oplus_{\frac{1}{x+1}} f(x + 1) 
-\big)
-%}_{\Xi} 
-\, \underline{10}
-\]
-\end{example}
-
 \begin{example}[Non-affine recursion]
@@
 \]
+\end{example}
+
+\begin{example}[More complex non-affine recursion]
+\label{ex:non-affine continuous}
+\begin{align*}
+& (\tY\, \lambda f x.\, (\lambda e. X[e]) \tsample)\ \underline 0 \\
+& X[e] = \tif{e \leq p - x^{-2}}{x}{f^2(x+1) \oplus_e f(x+1)}
+\end{align*}
 \end{example}
@@
 It seems very likely that this method is complete in the case that every reachable term is AST (\cref{conj:antitone}), but we have been unable to actually prove this.
-It is certainly at least capable of proving the termination of terms which terminate arbitrarily slowly (in the sense that there is no similar limitation to Theorem \ref{thm:rankable implies termination}, which can only prove termination of $\tY$-PAST terms).
+It is certainly at least capable of proving the termination of terms which terminate arbitrarily slowly (in the sense that there is no similar limitation to \cref{thm:rankable implies termination}, which can only prove termination of $\tY$-PAST terms).
 

### 2021-01-21 LICS21/confluent.tex
Minor; letter case

@@
 \, \tif{n = 0}{0}{\Theta_2 \, (n - 1) \oplus_{1/2} \Theta_2 \, (n + 1)}) z) \, \underline{11}$, bypassing $\Theta_2\ \underline{11}$ entirely.
-The antitone sparse ranking function in the form given can be justified anyway, by considering it to be using a reduction strategy which is the same as $\cbv$ except that in $\Theta_2\ (\underline f \underline n)$, it evaluates the redex at $@_2$ first.
+The antitone sparse ranking function in the form given can be justified anyway, by considering it to be using a reduction strategy which is the same as $\cbv$ except that in $\Theta_2\ (\underline f \, \underline n)$, it evaluates the redex at $@_2$ first.
 

### 2021-01-21 LICS21/related.tex
Add a simplified non-affine and continuous example.
Related work: added reference to more complex non-affine recursion example.

@@
 % (because the underlying recursion schemes are not Turing complete):
-the PPCF terms in \cref{ex:raven complex,ex:higher-order recursion}
+the PPCF terms in \cref{ex:raven complex,ex:non-affine continuous,ex:higher-order recursion}
 %\refExample{complexExample} 
@@
 While we require that the antitone ranking function decreases for every $\tY$-reduction step, they require that it decreases for every iteration of a certain while loop (the ranking function in that case is defined in the context of a particular loop), which is similar to $\tY$ but limited to tail recursion.
-The difference in the exact progress condition is not significant. 
+The difference in the exact progress condition is insignificant. A ranking function satisfying either progress condition can easily be converted to satisfy the other.
 \akr{Their progress condition, that there are antitone functions $p$ and $d$ such that the ranking function decreases from $x$ by at least $d(x)$ with probability at least $p(x)$, would not have worked directly because it does not cope as easily with the fact that the ranking function may change in value at the intermediate non-$\tY$ reduction steps (provided it does not increase in expectation).
@@
 \lo{Can you replace the highlighted by a brief comment on their discussion of completeness?}
+\akr{I would, but there's no space. Their completeness results seem pretty weak anyway.}
 

### 2021-01-21 LICS21/SPCF.tex
Remove duplicate references

@@
 
-The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid \hbox{$M$ terminates at $s$}\}$ is measurable \cite{DBLP:conf/icfp/BorgstromLGS16,MakOP20a}.
+The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid \hbox{$M$ terminates at $s$}\}$ is measurable \cite{DBLP:conf/icfp/BorgstromLGS16,MakOPW21}.
 We say that a term $M$ is \emph{almost surely terminating} (AST) just if $\mu_{\entrosp}(\{s \mid \hbox{$M$ terminates at $s$}\}) = 1$.

### 2021-01-22 LICS21/antitone-proofs.tex
Various minors

@@
 
-The term $\Xi^n\ \underline x$ (for $n > 0$, $x^{-2} < p$) reduces to either $\Xi^{n-1}\ \underline{x+1}$, $\Xi^n\ \underline{x+1}$ or $\Xi^{n+1}\ \underline{x+1}$, with probabilities $p-x^{-2}$, $(1-p+x^{-2})\frac{p-x^{-2}}{2}$ and $(1-p+x^{-2})(1-\frac{p-x^{-2}}{2})$ respectively.
-Let $a(\Xi^n\ \underline x) = n + 2x^{-1}$.
-This is a supermartingale (because the decrease in $2x^{-1}$ as $x$ increases is enough to offset the average increase in $n$), but it does not satisfy the antitone-strict progress condition.
+The term $\Xi^n\, \underline x$ (for $n > 0$, $x^{-2} < p$) reduces to either $\Xi^{n-1}\, \underline{x+1}$, $\Xi^n\ \underline{x+1}$ or $\Xi^{n+1}\, \underline{x+1}$, 
+with probabilities $p-x^{-2}$, $(1-p+x^{-2})\frac{p-x^{-2}}{2}$ and $(1-p+x^{-2})(1-\frac{p-x^{-2}}{2})$ respectively. 
+Let $a(\Xi^n\, \underline x) = n + 2 \, x^{-1}$. 
+This is a supermartingale (because the decrease in $2 \, x^{-1}$ as $x$ increases is enough to offset the average increase in $n$), but it does not satisfy the antitone-strict progress condition. 
 It does however have a bounded-below variance, so the usual method of using $\ln a$ instead of just $a$ works. 
-Calculating the exact amount that $\ln a$ decreases is not necessary, because it can be bounded as follows: $a$ changes by at least $1/2$ with probability at least $1/11$ (assuming $x \geq 3$), therefore $\ln(a)$ decreases in expectation by at least $\frac{1}{11}(\ln(a+\frac{1}{2}) - \ln(a) -\frac 1 {2a})$, using the fact that a linear approximation to $\ln$, applied to $a$, at least doesn't increase, then adding the deviation of $\ln$ from its linear approximation $\ln(a) + \frac{x-a}{a}$ (and using the fact that the linear approximation is everywhere an overestimate), we obtain a sufficiently strong bound on the decrease of $\ln(a)$ that it's an antitone sparse ranking function.
-(It can obviously be extended to $\Xi\ \underline 1$ too, which reduces to $\Xi^n\ \underline 3$ or a value after a bounded number of steps, but that complicates the analysis a little.)
+Calculating the exact amount that $\ln a$ decreases is not necessary, because it can be bounded as follows: 
+$a$ changes by at least $1/2$ with probability at least $1/11$ (assuming $x \geq 3$), 
+therefore $\ln(a)$ decreases in expectation by at least $\frac{1}{11}(\ln(a+\frac{1}{2}) - \ln(a) -\frac 1 {2a})$, 
+using the fact that a linear approximation to $\ln$, applied to $a$, at least doesn't increase, 
+then adding the deviation of $\ln$ from its linear approximation $\ln(a) + \frac{x-a}{a}$ 
+(and using the fact that the linear approximation is everywhere an overestimate), 
+we obtain a sufficiently strong bound on the decrease of $\ln(a)$ that it's an antitone sparse ranking function. 
+(It can obviously be extended to $\Xi\, \underline 1$ too, which reduces to $\Xi^n\, \underline 3$ or a value after a bounded number of steps, but that complicates the analysis a little.)
 

### 2021-01-22 LICS21/intro.tex
Remove duplicate references
Introduction: added "(under double-blind review elsewhere)"

@@
 Moreover various theorems about probabilistic programs rely on the assumption that the program terminates almost surely (see e.g.~\cite{DBLP:conf/uai/GoodmanMRBT08}). 
-A recent result \cite{MakOPW20} proves that AST programs have density functions that are differentiable almost everywhere.
+A recent result \cite{MakOPW21} proves that AST programs have density functions that are differentiable almost everywhere.
 This is significant for Bayesian machine learning, because almost everywhere differentiability is a precondition for the correctness of some of the most scalable inference algorithms, such as Hamiltonian Monte Carlo \cite{DBLP:conf/aistats/ZhouGKRYW19,Nishimura2020b} and reparameterised gradient variational inference \cite{DBLP:conf/nips/LeeYY18}.
@@
 Deciding AST of first-order imperative programs with discrete probabilities is $\Pi^0_2$-complete \cite{DBLP:conf/mfcs/KaminskiK15}.
-In recent unpublished work, the AST problem for higher-order programs with continuous distributions (and suitable primitive functions) is shown to have the same complexity.
+In recent unpublished work (under double-blind review elsewhere), the AST problem for higher-order programs with continuous distributions (and suitable primitive functions) is shown to have the same complexity.
 Though the problem of proving AST of imperative probabilistic programs is much studied \cite{DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05,DBLP:conf/popl/FioritiH15,DBLP:conf/pldi/ChenH20,DBLP:journals/toplas/ChatterjeeFNH18}, to our knowledge the problem of proving AST of higher-order programs with continuous distributions is new.

### 2021-01-22 LICS21/related.tex
Related work: added "We think that the supermartingales of our paper are a good basis for giving a unifying account."

@@
 Dal Lago and Grellois \cite{DBLP:journals/toplas/LagoG19} extend this to programs with discrete distributions, but their method is limited to affine recursion of first-order functions.
-Breuvart and Dal Lago \cite{DBLP:conf/ppdp/BreuvartL18} develop systems of intersection types from which the termination probability of higher-order programs (with discrete distributions) can be inferred from (infinitely many) type derivations.
+We think that the supermartingales of our paper are a good basis for giving a unifying account.
+Breuvart and Dal Lago \cite{DBLP:conf/ppdp/BreuvartL18} develop systems of intersection types from which the termination probability of programs (with discrete distributions) can be inferred from (infinitely many) type derivations.
 \lo{Mention Dal Lago et al.

### 2021-01-22 LICS21/intro.tex
approach to prove -> approach to proving

@@
 
-A possible approach to prove AST is to find some variant on the program state, called \emph{ranking function}, that decreases on average sufficiently quickly that it must at some point reach 0, at which point the program terminates.
+A possible approach to proving AST is to find some variant on the program state, called \emph{ranking function}, that decreases on average sufficiently quickly that it must at some point reach 0, at which point the program terminates.
 In other words, the program's behavior is used to define an associated (ranking) \emph{supermartingale}.

### 2021-01-23 LICS21/SPCF.tex
Minor reformatting in appendix; added reference

@@
 \fi
-For example, the term 
+\begin{example} The term 
 \[
@@
 \] 
-which generates a geometric distribution, terminates on the set $\entrosp \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas 
+which generates a geometric distribution, terminates on the set $\entrosp \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely; whereas 
 \(
@@
 which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.
-
+\end{example}
 %\paragraph{}

### 2021-01-23 LICS21/antitone-proofs.tex
Minor reformatting in appendix; added reference

@@
 
-\Cref{ex:raven complex}
+\subsection*{\Cref{ex:raven complex}}
 Let $\Theta := \tY \, \lambda f x . \tif{x \leq 0}{0}{f(x - \tsample)}$. 
@@
 
-\Cref{ex:Fair-in-the-limit random walk}
+\subsection*{\Cref{ex:Fair-in-the-limit random walk}}
 \[
@@
 
-\subsubsection*{Non-affine recursion}
+\subsection*{Non-affine recursion}
 
@@
 
-\medskip
-
-\Cref{ex:non-affine recursion easy}
+\subsection*{\Cref{ex:non-affine recursion easy}}
 Let 
@@
 
-\subsubsection*{Higher-order recursion}
+\subsection*{Higher-order recursion}
 
@@
 
-\Cref{ex:higher-order recursion}
+\subsection*{\Cref{ex:higher-order recursion}}
 Consider the higher-order function

### 2021-01-23 LICS21/intro.tex
Minor reformatting in appendix; added reference

@@
 Deciding AST of first-order imperative programs with discrete probabilities is $\Pi^0_2$-complete \cite{DBLP:conf/mfcs/KaminskiK15}.
-In recent unpublished work (under double-blind review elsewhere), the AST problem for higher-order programs with continuous distributions (and suitable primitive functions) is shown to have the same complexity.
+In recent work (under double-blind review elsewhere), the AST problem for higher-order programs with continuous distributions (and suitable primitive functions) is shown to have the same complexity.
 Though the problem of proving AST of imperative probabilistic programs is much studied \cite{DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05,DBLP:conf/popl/FioritiH15,DBLP:conf/pldi/ChenH20,DBLP:journals/toplas/ChatterjeeFNH18}, to our knowledge the problem of proving AST of higher-order programs with continuous distributions is new.

### 2021-01-23 LICS21/supermartingales.tex
Minor reformatting in appendix; added reference

@@
 
-It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \geq 0}$ is the reduction sequence starting from $M$ (considered as a stochastic process), then $(f(M_n))_{n \geq 0}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales (see e.g.~\cite{Williams1999} for details).
+It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \geq 0}$ is the reduction sequence starting from $M$ (considered as a stochastic process), then $(f(M_n))_{n \geq 0}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales (see e.g.~\cite{Williams1999,Durrett2019} for details).
 \iffalse

### 2021-01-23 LICS21/antitone-proofs.tex
Appendix C: minor reformatting / editing

@@
 
-\subsection*{\Cref{ex:raven complex}}
+\subsection*{\Cref{ex:raven complex} (Continuous random walk)}
 Let $\Theta := \tY \, \lambda f x . \tif{x \leq 0}{0}{f(x - \tsample)}$. 
@@
 
-\subsection*{\Cref{ex:Fair-in-the-limit random walk}}
+\subsection*{\Cref{ex:Fair-in-the-limit random walk} (Fair-in-the-limit random walk)}
 \[
@@
 
-\subsection*{\Cref{ex:non-affine recursion easy}}
+\subsection*{\Cref{ex:non-affine recursion easy} (Non-affine recursion)}
 Let 
@@
 
-\Cref{ex:non-affine continuous}
+\subsection*{\Cref{ex:non-affine continuous} (More complex non-affine recursion)}
 Let
@@
 
-There is an obvious source of deterministic (i.e.~non-probabilistic) higher-order functions defined by recursion, viz., \emph{higher-order recursion schemes} (HORS) (see e.g.~\cite{DBLP:conf/lics/Ong06,DBLP:conf/lics/Ong15}).
-(Incidentally the (sparse) ranking function method is just as applicable to the termination analysis of \emph{deterministic} PPCF programs.)
+There is an obvious source of deterministic (i.e.~non-random) higher-order functions defined by recursion, viz., \emph{higher-order recursion schemes} (HORS) (see e.g.~\cite{DBLP:conf/lics/Ong06,DBLP:conf/lics/Ong15}).
+%(Incidentally the (sparse) ranking function method is just as applicable to the termination analysis of \emph{deterministic} PPCF programs.)
 Recently \cite{DBLP:conf/lics/KobayashiLG19} have extended HORS to \emph{probabilistic higher-order recursion schemes} (PHORS), which are HORS augmented with probabilistic (binary) branching $\oplus_p$.
@@
 
-\subsection*{\Cref{ex:higher-order recursion}}
+\subsection*{\Cref{ex:higher-order recursion} (Higher-order recursion)}
 Consider the higher-order function

### 2021-01-23 LICS21/antitone-proofs.tex
Appendix C: add names to examples and various minors

@@
 \begin{align*}
-\Xi & = \tY\, \lambda f x.\, (\lambda e. X[e]) \tsample \\
+\Xi & = \tY\, \lambda f x.\, (\lambda e. X[e]) \, \tsample \\
 X[e] & = \tif{e \leq p - x^{-2}}{x+1}{f^2(x+1) \oplus_e f(x+1)}
@@
 
-There is an obvious source of deterministic (i.e.~non-random) higher-order functions defined by recursion, viz., \emph{higher-order recursion schemes} (HORS) (see e.g.~\cite{DBLP:conf/lics/Ong06,DBLP:conf/lics/Ong15}).
+There is an important formalism for constructing (non-random) higher-order recursive functions, viz., \emph{higher-order recursion schemes} (HORS) (see e.g.~\cite{DBLP:conf/lics/Ong06,DBLP:conf/lics/Ong15}).
 %(Incidentally the (sparse) ranking function method is just as applicable to the termination analysis of \emph{deterministic} PPCF programs.)
@@
 As HORS are in essence the $\lambda \tY$-calculus \cite{DBLP:conf/lics/Statman02} (i.e.~pure simply-typed lambda calculus with recursion, generated from a finite base type), PHORS are definable in PPCF:
-(order-$n$) PHORS is encodable as (order-$n$) (call-by-name) PPCF, but the former is strictly less expressive (because the underlying recursion schemes are not Turing complete). 
+(order-$n$) PHORS is encodable as (order-$n$) call-by-name PPCF, but the former is strictly less expressive (because the underlying HORS is not Turing complete). 
+For example, the PPCF term in \Cref{ex:higher-order recursion} is not definable in PHORS.
 %Some interesting PPCF terms such as \Cref{ex:non-affine with continuous distribution} %\refExample{complexExample} 
@@
 %\akr{As I removed the referenced example, and I don't even know what a PHORS is, I've just removed this comment too.}
-A relevant result here is that the AST problem is decidable for order-1 PHORS (by reduction to the PSPACE-hard solvability of finite systems of polynomial equations with real coefficients; see \cite{DBLP:journals/jacm/EtessamiY09}), but undecidable for order-2 PHORS.
+
+A relevant result here is that the AST problem is decidable for order-1 PHORS (by reduction to the PSPACE-hard solvability of finite systems of polynomial equations with real coefficients; see \cite{DBLP:journals/jacm/EtessamiY09}), but undecidable for PHORS with order greater than 1 \cite{DBLP:conf/lics/KobayashiLG19}.
 

### 2021-01-24 LICS21/SPCF.tex
Use \defn to make key definienda stand out - easier for reviewers to find definitions

@@
 \changed[lo]{The maps $\pi_h:\entrosp \to I$ (projecting to the first element) and $\pi_t:\entrosp \to \entrosp$ (popping the first element) are then measurable.}
-Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$ the \emph{entropy space}, and elements of $\entrosp$ \emph{traces}.
+Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$ the \emph{entropy space}, and elements of $\entrosp$ \defn{traces}.
 
-Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $\mathit{Sk}(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$.
+Define a \defn{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $\mathit{Sk}(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$.
 Following \cite{DBLP:conf/icfp/BorgstromLGS16,DBLP:conf/lics/StatonYWHK16,DBLP:journals/pacmpl/EhrhardPT18}, the $\sigma$-algebra on $\Lambda$ is defined by identifying $\Lambda$ with $\bigcup_{m\geq 0} \big(\mathsf{Sk}_m \times \Real^m \big)$, where $\mathsf{Sk}_m$ is the set of skeletons containing $m$ occurrences of $\skeletonPlaceholder$.
@@
 The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid \hbox{$M$ terminates at $s$}\}$ is measurable \cite{DBLP:conf/icfp/BorgstromLGS16,MakOPW21}.
-We say that a term $M$ is \emph{almost surely terminating} (AST) just if $\mu_{\entrosp}(\{s \mid \hbox{$M$ terminates at $s$}\}) = 1$.
+We say that a term $M$ is \defn{almost surely terminating} (\defn{AST}) just if $\mu_{\entrosp}(\{s \mid \hbox{$M$ terminates at $s$}\}) = 1$.
 
@@
 \fi
-\begin{example} The term 
+\begin{example}[Geometric distribution] The term 
 \[

### 2021-01-24 LICS21/antitone.tex
Various minors
Use \defn to make key definienda stand out - easier for reviewers to find definitions

@@
 \begin{definition}
-The definition of \emph{antitone ranking function} $f$ for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
+The definition of \defn{antitone ranking function} $f$ for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
 we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) function $\epsilon : \nnReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
@@
 where $R \to R'$ is the $\tY$-redex rule.
-Any closed term for which an antitone ranking function exists is called \emph{antitone rankable}.
+Any closed term for which an antitone ranking function exists is called \defn{antitone rankable}.
 \iffalse
@@
 Given a probability space $(\Omega, \calF, \mathbb{P})$, and a supermartingale $(Y_n)_{n \geq 0}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \geq 0}$,
-we say that $(Y_n)_{n \geq 0}$ is an \emph{antitone strict supermartingale w.r.t.~$T$} if for all $n \geq 0$, we have $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
+we say that $(Y_n)_{n \geq 0}$ is an \defn{antitone strict supermartingale w.r.t.~$T$} if for all $n \geq 0$, we have $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
 \(
@@
 \begin{definition}
-Define an \emph{antitone sparse ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$:
+Define an \defn{antitone sparse ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$:
 \begin{inparaenum}[(i)]
@@
 \[
-(\tY \lambda f x. x \oplus_{2/3} f (f (x + 1)))\ \underline 1
+(\tY \lambda f x. x \oplus_{2/3} f (f (x + 1)))\, \underline 1
 \]
@@
 \begin{align*}
-& (\tY\, \lambda f x.\, (\lambda e. X[e]) \tsample)\ \underline 0 \\
-& X[e] = \tif{e \leq p - x^{-2}}{x}{f^2(x+1) \oplus_e f(x+1)}
+& (\tY\, \lambda f x.\, (\lambda e. X[e]) \, \tsample)\, \underline 0 \\
+& X[e] := \tif{e \leq p - x^{-2}}{x}{f^2(x+1) \oplus_e f(x+1)}
 \end{align*}

### 2021-01-24 LICS21/confluent.tex
Use \defn to make key definienda stand out - easier for reviewers to find definitions

@@
 
-A \emph{position} is a finite sequence of steps into a term, defined inductively as
+A \defn{position} is a finite sequence of steps into a term, defined inductively as
 \begin{align*}
@@
 \end{align*}
-The \emph{subterm of $M$ at $\alpha$}, denoted $M \mid \alpha$, is defined as
+The \defn{subterm of $M$ at $\alpha$}, denoted $M \mid \alpha$, is defined as
 \begin{align*}
@@
 \end{align*}
-In each of these cases, $M \mid \alpha$ is the \emph{redex}, and the reduction takes place at $\alpha$.
+In each of these cases, $M \mid \alpha$ is the \defn{redex}, and the reduction takes place at $\alpha$.
 \end{definition}
@@
 
-The reflexive transitive closure $\sim^*$ of this relation is used to define the set of \emph{potential positions} $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable skeletons.
+The reflexive transitive closure $\sim^*$ of this relation is used to define the set of \defn{potential positions} $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable skeletons.
 If $(N,\alpha) \sim^* (O,\beta)$, then $N \mid \alpha$ and $O \mid \beta$ both have the same shape (i.e.~they're either both the placeholder $\skeletonPlaceholder$, both variables, both applications, both $\tsample$s etc.), therefore it's well-defined to talk of the set of potential positions where there is a $\tsample$, $L_s(M)$.
@@
 The reduction relation $\Rightarrow$ is nondeterministic, so it admits multiple possible reduction strategies. 
-A \emph{reduction strategy} starting from a closed term $M$ is a measurable partial function $f$ from $\mathit{Rch}(M)$ to positions, such that for any reachable term $N$ where $f$ is defined, $f(N)$ is a position of a redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
+%\begin{definition}
+A \defn{reduction strategy} starting from a closed term $M$ is a measurable partial function $f$ from $\mathit{Rch}(M)$ to positions, such that for any reachable term $N$ where $f$ is defined, $f(N)$ is a position of a redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
 Using a reduction strategy $f$, a subset of $\Rightarrow$ that isn't nondeterministic, $\Rightarrow_f$, can be defined by $(N,s) \Rightarrow_f (N',s')$ just if $(N,s) \Rightarrow (N',s')$ and $N$ reduces to $N'$ with the redex at $f(N)$.
+%\end{definition}
 
@@
 \begin{definition}
-An \emph{antitone ranking function on $M$ with respect to a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ such that
+An \defn{antitone ranking function on $M$ with respect to a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ such that
 \begin{itemize}
@@
 \end{itemize}
-Any closed term for which an antitone ranking function with respect to a reduction strategy $r$ exists is called \emph{antitone rankable} with respect to $r$. 
+Any closed term for which an antitone ranking function with respect to a reduction strategy $r$ exists is called \defn{antitone rankable} with respect to $r$. 
 \end{definition}
@@
 \begin{definition}
-An \emph{antitone sparse ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : \mathit{Rch}_r(M) \rightharpoonup \mathbb R$ such that there exists some antitone function $\epsilon : \nnReal \to \pReal$ such that
+An \defn{antitone sparse ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : \mathit{Rch}_r(M) \rightharpoonup \mathbb R$ such that there exists some antitone function $\epsilon : \nnReal \to \pReal$ such that
 \begin{inparaenum}[(i)]
@@
 
-If the antitone function is just the constant function $1$ instead, then it is called a \emph{(sparse) ranking function on $M$ with respect to $r$} instead, and the term is called \emph{rankable} with respect to $r$.
+If the antitone function is just the constant function $1$ instead, then it is called a \defn{(sparse) ranking function on $M$ with respect to $r$} instead, and the term is called \defn{rankable} with respect to $r$.
 

### 2021-01-24 LICS21/supermartingales.tex
Use \defn to make key definienda stand out - easier for reviewers to find definitions

@@
 \label{def:ranking function}
-A \emph{ranking function on $M \in \Lambda^0$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$\footnote{The fact that the ranking function can be 0 in some cases before it reaches a value is necessary to get \Cref{thm:minimal} to work neatly.}, and
+A \defn{ranking function on $M \in \Lambda^0$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$\footnote{The fact that the ranking function can be 0 in some cases before it reaches a value is necessary to get \Cref{thm:minimal} to work neatly.}, and
 \begin{enumerate}
@@
 \fi
-We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.
+We say that the ranking function $f$ is \defn{strict} if there exists $\epsilon > 0$ such that for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.
 
-Any closed term for which a ranking (respectively, strict ranking) function exists is called \emph{rankable} (respectively, \emph{strictly rankable}). 
+Any closed term for which a ranking (respectively, strict ranking) function exists is called \defn{rankable} (respectively, \defn{strictly rankable}). 
 For example, $(\tY \lambda x.x) \, \underline 0$ is not rankable.
@@
 Let $T$ be a r.v.~that takes values in $\mathbb{N} \cup \set{\infty}$.
-We call $T$ a \emph{stopping time adapted to $(\calF_n)_{n \geq 0}$} just if $\set{T = n} \in \calF_n$, for all $n$.
+We call $T$ a \defn{stopping time} \emph{adapted to $(\calF_n)_{n \geq 0}$} just if $\set{T = n} \in \calF_n$, for all $n$.
 
@@
 \begin{asparaenum}[(i)]
-\item A sequence of r.v.s $(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \emph{supermartingale} if for all $n \geq 0$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~(i.e.~for all $A \in \calF_n$, $\int_A \mathbb{P}(\dif \omega) \, Y_{n+1}(\omega) \leq \int_A \mathbb{P}(\dif \omega) \,Y_n(\omega)$).
+\item A sequence of r.v.s $(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \defn{supermartingale} if for all $n \geq 0$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~(i.e.~for all $A \in \calF_n$, $\int_A \mathbb{P}(\dif \omega) \, Y_{n+1}(\omega) \leq \int_A \mathbb{P}(\dif \omega) \,Y_n(\omega)$).
 \item Let $\epsilon > 0$. 
 Given a stopping time $T$ and a supermartingale $(Y_n)_{n \geq 0}$, both adapted to filtration $(\calF_n)_{n \geq 0}$, 
-we say that $(Y_n)_{n \geq 0}$ is a \emph{$\epsilon$-ranking supermartingale w.r.t.~$T$} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.
+we say that $(Y_n)_{n \geq 0}$ is a $\epsilon$-\defn{ranking supermartingale w.r.t.~$T$} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.
 \footnote{For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable defined by ${\bf 1}_A(\omega) := 1$ if $\omega \in A$, and $0$ otherwise.} %\lo{\LaTeX\ does not allow the preceding bracketed sentence to be typeset as a footnote.
@@
 Stranger.}
-A \emph{ranking supermartingale} w.r.t.~$T$ is a $\epsilon$-ranking supermartingale w.r.t.~$T$ for some $\epsilon > 0$.
+A \defn{ranking supermartingale} w.r.t.~$T$ is a $\epsilon$-ranking supermartingale w.r.t.~$T$ for some $\epsilon > 0$.
 \end{asparaenum}
@@
 and the filtration $(\mathcal{F}_n)_{n \geq 0}$ where $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
-Thus $T_M$ is the \emph{runtime} of $M$ (and $M$ is AST iff $\mu_{\entrosp}(T_M < \infty) = 1$).
+Thus $T_M$ is the \defn{runtime} of $M$ (and $M$ is AST iff $\mu_{\entrosp}(T_M < \infty) = 1$).
 

### 2021-01-24 LICS21/confluent.tex
A few more \defn

@@
 \end{align*}
-In each of these cases, $M \mid \alpha$ is the \defn{redex}, and the reduction takes place at $\alpha$.
+In each of these cases, $M \mid \alpha$ is the \defn{redex}, and the \defn{reduction takes place at $\alpha$}.
 \end{definition}
@@
 
-The domain of definition of the ranking functions also needs to be changed from the reachable terms $\mathit{Rch}(M)$ to the \emph{reachable terms with respect to the reduction strategy $r$},
+The domain of definition of the ranking functions also needs to be changed from the reachable terms $\mathit{Rch}(M)$ to the \defn{reachable terms with respect to the reduction strategy $r$},
 \[

### 2021-01-25 LICS21/ranking.tex
Tweak

@@
 terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x. x) \, \underline 0$ is reachable, although that has probability 0. 
-Not only is this counterexample AST, it's PAST. %positively almost surely terminating i.e.~the expected time to termination is finite.
+Not only is this counterexample AST, it's postively almost sure terminating (PAST). %positively almost surely terminating i.e.~the expected time to termination is finite.
 
@@
 \begin{definition}
-Define a \emph{sparse ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \rightharpoonup \mathbb R$ such that
+Define a \defn{sparse ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \rightharpoonup \mathbb R$ such that
 \begin{inparaenum}[(i)]

### 2021-01-25 LICS21/supermartingales.tex
A few more \defn
Tweak

@@
 Stranger.}
-A \defn{ranking supermartingale} w.r.t.~$T$ is a $\epsilon$-ranking supermartingale w.r.t.~$T$ for some $\epsilon > 0$.
+A \defn{ranking supermartingale w.r.t.~$T$} is a $\epsilon$-ranking supermartingale w.r.t.~$T$ for some $\epsilon > 0$.
 \end{asparaenum}
@@
 \begin{remark}
-\Cref{lem:rank-PAST} is a slight extension of \cite[Lemma 5.5]{DBLP:conf/popl/FioritiH15},
+\Cref{lem:rank-PAST} is a slight extension of \cite[Lem.~5.5]{DBLP:conf/popl/FioritiH15},
 which asserts the same result but for the specific stopping time $T : \omega \mapsto \min\set{n \mid Y_n(\omega) = 0}$.

### 2021-01-25 LICS21/antitone.tex
Explicitly define PAST and Y-PAST. Cleaner and shorter restatement of Thm III.10. Insert forward reference to explanation (made a remark) in footnote 4. \defn in Sec. VI

@@
 % \end{align*}
-For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$,\footnote{Technically this is not quite correct, because of the distinction between $\underline n + 1$ and $\underline{n + 1}$, but this ranking function can still be made to work by appealing to \Cref{thm:confluent ranking}, as can the others in this section with similar issues.} each with probability $1/2$, with one $\tY$-reduction.
+For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$,\footnote{Technically this is not quite correct, because of the distinction between $\underline n + 1$ and $\underline{n + 1}$, but this ranking function can still be made to work by appealing to the Confluent Ranking \Cref{thm:confluent ranking}, as can the others in this section with similar issues.
+See \cref{rem:ex corrected using confluent semantics} for a fuller account.} each with probability $1/2$, with one $\tY$-reduction.
 Now

### 2021-01-25 LICS21/confluent.tex
Explicitly define PAST and Y-PAST. Cleaner and shorter restatement of Thm III.10. Insert forward reference to explanation (made a remark) in footnote 4. \defn in Sec. VI

--- 2021-01-24 LICS21/confluent.tex
@@
 
-We are thus led to consider the reduction relation on skeletons (and positions in a skeleton), which can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, $\tsample$ reducing to $\skeletonPlaceholder$, and $\skeletonPlaceholder$ considered as (the skeletal equivalent of) a value, so that $(\lambda x.A) \, \skeletonPlaceholder$ reduces to $A[\skeletonPlaceholder / x]$.
+\paragraph*{Reduction relation on skeletons} We are thus led to consider the reduction relation on skeletons (and positions in a skeleton), which can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, $\tsample$ reducing to $\skeletonPlaceholder$, and $\skeletonPlaceholder$ considered as (the skeletal equivalent of) a value, so that $(\lambda x.A) \, \skeletonPlaceholder$ reduces to $A[\skeletonPlaceholder / x]$.
 For example, we have 
@@
 \begin{align*}
-A[x] &= \tif{\tif{x>0}{I}{I} (\lambda y.\tsample)\ \underline 0 - \underline{0.5} > 0}{\underline 0}{\Omega}\\
+A[x] &= \tif{\tif{x>0}{I}{I} (\lambda y.\tsample)\, \underline 0 - \underline{0.5} > 0}{\underline 0}{\Omega}\\
 B &= \tif{\tsample - \underline{0.5} > 0}{\underline 0}{\Omega}
@@
 This is a rather unwieldy expression, but the crucial part is that $r$ occurs twice in the conditions on $s$: once as the value a sample must take, and once in the location of a sample. 
-This set is unmeasurable, therefore the termination probability would not even be well-defined. 
+As this set is unmeasurable, the termination probability would not even be well-defined. 
 Labelling samples by skeletons instead, this problem does not occur because there are only countably many skeleton, and at each step in a reduction sequence, only finitely many could have occurred yet.
@@
 
-\begin{example} Consider the term $M = \tY (\lambda f x.\ \tif{\tsample - \underline{0.5} < 0}{f \, x}{x}) \, \underline 0$, which reduces after a few steps to $N = \tif{\tsample - \underline{0.5}  < 0}{M}{\underline 0}$.
+\begin{example} Consider the term $M = \tY (\lambda f x.\, \tif{\tsample - \underline{0.5} < 0}{f \, x}{x}) \, \underline 0$, which reduces after a few steps to $N = \tif{\tsample - \underline{0.5}  < 0}{M}{\underline 0}$.
 If we label samples by just skeletons and positions, and the pre-selected sample for $(\mathit{Sk}(N),\textsf{if}_1;\underline{-}_1)$ is less than 0.5, $N$ reduces back to $M$, then $N$ again, then the same sample is used the next time, therefore it's an infinite loop, whereas if samples are labelled by reduction sequences, the samples for $M \to^\ast N$ are independent from the samples for $M \to^\ast N \to M \to^\ast N$, and so on.
@@
 
-\medskip
+%\medskip
 This is still not quite sufficient because sometimes the same samples must be used at corresponding positions in different reduction sequences.
@@
     \item If $N$, $O_1$ and $O_2$ match any of the following cases:
-    \begin{enumerate}
+    \begin{compactenum}
         \item $N$ contains redexes at disjoint positions $\alpha_1$ and $\alpha_2$, $O_1$ is $N$ reduced first at $\alpha_1$ then $\alpha_2$ and $O_2$ is $N$ reduced first at $\alpha_2$ then at $\alpha_1$.
@@
 A)$, $A$ reduced at $\beta$ is $A'$, $(\gamma_i)_i$ is a list of all the positions where $A' \mid \gamma  = x$, ordered from left to right, $O_1$ is $N$ reduced first at $\alpha;\tY;\lambda;\beta$ then at $\alpha$, and $O_2$ is $N$ reduced first at $\alpha$ then at $\alpha;\lambda;@_1;\gamma_i;Y;\lambda;\beta$ for each $i$ in order where $\gamma_i$ is left of $\beta$ then at $\alpha;\lambda;@_1;\beta$ then at $\alpha;\lambda;@_1;\gamma_i;Y;\lambda;\beta$ for the remaining values of $i$.
-    \end{enumerate}
+    \end{compactenum}
     (in which case $O_1$ and $O_2$ are equal as skeletons, but with different reduction sequences), $O_1'$ and $O_2'$ are the results of applying some reduction sequence to each of $O_1$ and $O_2$ (the same reductions in each case, which is always possible because they're equal skeletons), and $\delta$ is a position in $O_1'$ (or equivalently $O_2'$), then $(O_1',\delta) \sim_c (O_2',\delta)$.
@@
 
-The reflexive transitive closure $\sim^*$ of this relation is used to define the set of \defn{potential positions} $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable skeletons.
-If $(N,\alpha) \sim^* (O,\beta)$, then $N \mid \alpha$ and $O \mid \beta$ both have the same shape (i.e.~they're either both the placeholder $\skeletonPlaceholder$, both variables, both applications, both $\tsample$s etc.), therefore it's well-defined to talk of the set of potential positions where there is a $\tsample$, $L_s(M)$.
-Formally $L_s(M) := \{[(X, \alpha)]_{\sim^\ast_M} : X \mid \alpha = \tsample\}$.
+The reflexive transitive closure $\sim^*$ of this relation is used to define the set of \emph{potential positions} $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable skeletons.
+If $(N,\alpha) \sim^* (O,\beta)$, then $N \mid \alpha$ and $O \mid \beta$ both have the same shape (i.e.~they're either both the placeholder $\skeletonPlaceholder$, both variables, both applications, both $\tsample$s etc.), therefore it's well-defined to talk of the set of \defn{potential positions where there is a $\tsample$}, $L_s(M)$.
+Formally 
+\[
+L_s(M) := \{[(X, \alpha)]_{\sim^\ast_M} : X \mid \alpha = \tsample\}.
+\]
 The new sample space is then defined as $I^{L_s(M)}$, with the Borel $\sigma$-algebra and product measure.
@@
 
-\begin{definition}
+\begin{definition}[$\Rightarrow$-reduction]
 Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function, so it is denoted ``$\Rightarrow$'' instead of ``$\red$'', and it relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
@@
 \begin{definition}
-An \defn{antitone ranking function on $M$ with respect to a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ such that
+An \defn{antitone ranking function on $M$ w.r.t.~a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ such that
 \begin{itemize}
@@
 \end{itemize}
-Any closed term for which an antitone ranking function with respect to a reduction strategy $r$ exists is called \defn{antitone rankable} with respect to $r$. 
+Any closed term for which an antitone ranking function with respect to a reduction strategy $r$ exists is called \defn{antitone rankable w.r.t.~$r$}. 
 \end{definition}
@@
 \begin{definition}
-An \defn{antitone sparse ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : \mathit{Rch}_r(M) \rightharpoonup \mathbb R$ such that there exists some antitone function $\epsilon : \nnReal \to \pReal$ such that
+An \defn{antitone sparse ranking function on $M$ w.r.t.~a reduction strategy $r$} is a partial function $f : \mathit{Rch}_r(M) \rightharpoonup \mathbb R$ such that there exists some antitone function $\epsilon : \nnReal \to \pReal$ such that
 \begin{inparaenum}[(i)]
@@
 
-If the antitone function is just the constant function $1$ instead, then it is called a \defn{(sparse) ranking function on $M$ with respect to $r$} instead, and the term is called \defn{rankable} with respect to $r$.
+If the antitone function is just the constant function $1$ instead, then it is called a \defn{(sparse) ranking function on $M$ w.r.t.~$r$} instead, and the term is called \defn{rankable w.r.t.~$r$}.
 
@@
 \begin{align*}
-M =&\ (\lambda n. P[n]) \lfloor \tsample^{-1/2} \rfloor \\
-P[n] =&\ (\lambda p. \Xi[p] \, n) (\lambda x. \Theta[x] \, n)) \\
-\Xi[p] =&\ \tY \lambda f n. \tif{n = 0}{\underline 0}{p \, n + f (n-1)} \\
-\Theta[x] =&\ \tY \lambda f n. \tif{n = 0}{\underline 1}{x \times f (n-1)}.
+M  &= (\lambda n. P[n]) \lfloor \tsample^{-1/2} \rfloor \\
+P[n]  &= (\lambda p. \Xi[p] \, n) (\lambda x. \Theta[x] \, n)) \\
+\Xi[p]  &= \tY \lambda f n. \tif{n = 0}{\underline 0}{p \, n + f (n-1)} \\
+\Theta[x]  &= \tY \lambda f n. \tif{n = 0}{\underline 1}{x \times f (n-1)}.
 \end{align*}
@@
 
+\begin{remark}
+\label{rem:ex corrected using confluent semantics}
 Even some of the examples given earlier implicitly use slightly non-standard reduction strategies to make the definitions of their ranking functions slightly simpler.
-For example, the antitone sparse ranking function in \Cref{ex:unbiased random walk} was defined at $\Theta_2\ \underline n$ (where $\Theta_2 = \tY \, \lambda f n .
-\, \tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$), but $\Theta_2\ \underline{10}$ actually reduces to $\Theta_2\ (\underline{10} - 1)$ or $\Theta_2\ (\underline{10} + 1)$, and (taking the second of these), it then reduces to $(\lambda z.
+For example, the antitone sparse ranking function in \Cref{ex:unbiased random walk} was defined at $\Theta_2\, \underline n$ (where $\Theta_2 = \tY \, \lambda f n .
+\, \tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$), but $\Theta_2\, \underline{10}$ actually reduces to $\Theta_2\, (\underline{10} - 1)$ or $\Theta_2\, (\underline{10} + 1)$, and (taking the second of these), it then reduces to $(\lambda z.
 (\lambda n.
@@
 (\lambda n.
-\, \tif{n = 0}{0}{\Theta_2 \, (n - 1) \oplus_{1/2} \Theta_2 \, (n + 1)}) z) \, \underline{11}$, bypassing $\Theta_2\ \underline{11}$ entirely.
-The antitone sparse ranking function in the form given can be justified anyway, by considering it to be using a reduction strategy which is the same as $\cbv$ except that in $\Theta_2\ (\underline f \, \underline n)$, it evaluates the redex at $@_2$ first.
+\, \tif{n = 0}{0}{\Theta_2 \, (n - 1) \oplus_{1/2} \Theta_2 \, (n + 1)}) z) \, \underline{11}$, bypassing $\Theta_2\, \underline{11}$ entirely.
+The antitone sparse ranking function in the form given can be justified anyway, by considering it to be using a reduction strategy which is the same as $\cbv$ except that in $\Theta_2\, (\underline f \, \underline n)$, it evaluates the redex at $@_2$ first.
 
@@
 In this case, it is to allow the antitone sparse ranking function to skip over all of those terms where the unknown function $F$ is part way through evaluation.
+\end{remark}

### 2021-01-25 LICS21/confluent.tex
minor
More signposting for Sec. VI

@@
 
-\begin{definition}[$\Rightarrow$-reduction]
+\begin{definition}[$\Rightarrow$ reduction]
 Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function, so it is denoted ``$\Rightarrow$'' instead of ``$\red$'', and it relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
@@
 
-\medskip
+\paragraph*{Reduction strategy}
 The reduction relation $\Rightarrow$ is nondeterministic, so it admits multiple possible reduction strategies. 
@@
 A closed term $M$ terminates with a given reduction strategy $f$ and samples $s$ if there is some natural number $n$ such that $(M,s) \Rightarrow_f^n (N,s')$ where $f$ gives no reduction at $N$. 
-The term terminates almost surely with respect to $f$ if it terminates with $f$ for almost all $s$.
+The term \emph{terminates almost surely w.r.t.~$f$} if it terminates with $f$ for almost all $s$.
 
@@
 
-However, there are some terms which terminate more simply and directly via some alternative reduction strategy.
+\paragraph*{Custom reduction strategy for AST analysis} However, there are some terms which terminate more simply and directly via some alternative reduction strategy.
 For these examples, a custom reduction strategy is specified for each term under consideration.

### 2021-01-25 LICS21/confluent.tex
minor

@@
 
-\paragraph*{Reduction strategy}
+\paragraph*{Reduction strategy determinising $\Rightarrow$}
 The reduction relation $\Rightarrow$ is nondeterministic, so it admits multiple possible reduction strategies. 

### 2021-01-25 LICS21/SPCF.tex
Tweak

@@
 
-The language PPCF is a call-by-value (CBV) version of PCF \cite{} with sampling of real numbers from the closed interval $[0,1]$ \cite{Ehrhard2018c,DBLP:journals/pacmpl/EhrhardPT18,MakOPW21}.
+The language PPCF is a call-by-value (CBV) version of PCF with sampling of real numbers from the closed interval $[0,1]$ \cite{Ehrhard2018c,DBLP:journals/pacmpl/EhrhardPT18,MakOPW21}.
 %\changed[lo]{following \cite{MakOPW21} except the score function (\Cref{rem:score})}. 

### 2021-03-12 LICS21/review response.txt
Add a draft review response.

(@Luke: This is a bit over the word limit, but some of that is comments to you. If it's still too long after you've responded, we can cut some of the more minor ones as you suggested.)

Reviewer 1 asked what "\mathsf R", "r.v." and "a.s." are. "\mathsf R" refers to the base type for programs which reduce to real-number values. "r.v." means random variable, and "a.s." means almost surely.

Reviewer 2 said that there were no examples of reduction strategies beyond CBV. As the other reduction strategies mainly considered in this work (i.e. used with Theorem V1.16) are intended to be specifically designed for particular programs, the other examples are given along with particular example programs, specifically the extended discussions of examples V.14 and VI.17 in the appendices.

CBN is not considered a valid reduction strategy because it violates the condition that the argument of a beta redex must be a value.

(@Luke: I don't understand why they're objecting to "Y \lambda f n. M". Their objection isn't the colon (while the syntax for lambda terms was defined with a dot) because they also object to it again later where we did use a dot. The abbreviation "\lambda f n. M" for "\lambda f. \lambda n. M" seems sufficiently standard that I don't expect that this is their objection either. Maybe they're expecting brackets ("Y (\lambda f n. M)")? I don't want to reply that this objection is incorrect without understanding what the objection is, but I think it is incorrect.)

Example IV.1 actually is rankable, as rankability corresponds (approximately) to Y-PAST.

(@Luke: I remember looking up what "contractum" meant, but I can't find it now.)

At the end of section IV, the next reduction in i ⊕_0.5 (Y \lambda fn. n ⊕_0.5 f(n+1)) (i+1) is the ⊕_0.5 (or, to be more precise, the sample implicit in that notation), not the Y, because ⊕ is syntactic sugar for a sort of if, and the branches of an if are not evaluated before one of them is selected.

Example V.1 is simply meant to illustrate that there are some terms which terminate too slowly to be rankable.

The relation defined in VI.1 is non-deterministic, and alpha can take any value that locates one of these types of redexes.

Labelling the samples with elements of a set A simply refers to picking samples s from I^A, then using s(a) for the sample reduction whenever the relevant position/number of samples previoustly taken/reduction sequence and position (depending on the labelling) is a.

In example VI.2, it's not clear whether you're referring to the skeleton placeholder number, sans-serif X, or simply that x should be uppercase because it's a term (which would be correct). "I" here is meant as shorthand for the identity function.

Regarding the general comments on section 6.
The restrictions in reduction strategies are the extra conditions in Def VI.1 on what reductions are valid (specifically, that the argument of a beta redex must be a value, and the restrictions on positions where sample reductions can occur).
The body of a Y subterm is effectively both the function and the argument in an application. If it contians multiple occurrences of the relevant variable, it will be duplicated by the Y reduction. More specifically, consider (Y \lambda f n. if n = 0 then (\lambda x. sample) else \lambda x.(f 0 0 + f 0 0)) 1 0. This reduces to something like (lambda x. sample) 0 + (\lambda x. sample) 0. If the sample had been allowed to be reduced first, the samples would have ended up correlated. The fact that the position has to have a Y and a lambda in it to be excluded is simply to accommodate things like Y if sample > 0.5 then M else N, where there actually should be reductions inside the Y.
Yes, (lambda x. x 0 + x 0) (lambda y. sample) illustrates why the reduction strategies must be restricted. The correct distribution of results is triangular, with support [0,2] and peak at 1. The (invalid) reduction strategy that reduces the sample first would instead produce a uniform distribution on [0,2], because it would be taking one sample and scaling it up rather than taking two independent samples. The restriction that sample reductions can't occur at positions in which lambda occurs after @_2 is indeed to avoid problems with terms like this.
In Definition VI.8, s is a function L_s(M) to I, and i(M -> N) is a function L_s(N) to L_s(M), therefore the composition i(M -> N) then s is valid.

Regarding reviewer 3's questions about the completeness conjecture:
I am not aware of any particularly relevant previous literature. I did spend a while trying to solve the problem, and do intend to try some more in future, but didn't have time to continue. Part of what makes this question so difficult is that you have to construct an antitone function that represents the rate of convergence of all of the different states the program could be in, and assign all the program states ranking function values, despite it be not obvious how to compare, for example, one state which would either terminate in a few steps or extremely slowly, and another which would terminate after a long time, but more deterministically. The fact that, for constructing a ranking function, even branches with 0 probability must be considered is an additional cause of difficulty, as they can terminate arbitrarily much slower than the program itself.

### 2021-03-12 LICS21/antitone-proofs.tex

@@
 
-\subsection*{\Cref{ex:Fair-in-the-limit random walk} (Fair-in-the-limit random walk)}
+\subsection*{\Cref{ex:Fair-in-the-limit random walk} (Fair-in-the-limit random walk) \cite[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}}
 \[
@@
 Let $F : \textsf{R} \to \textsf{R} \to \textsf{R}$ be a function such that $F\ \underline n\ \underline m$ terminates with no $\tY$ reductions for all $m$ and $n$.
-Then $\Xi\ F\ \underline x\ \underline y$ has the antitone sparse ranking function $F \underline{n_0} (F \underline{n_1} (\dots(\Xi\ F\ \underline x\ \underline n)\dots) \mapsto g_2(n)$, where $g_2$ is as in \Cref{ex:unbiased random walk}, using the same antitone function too.
+Then $\Xi\ F\ \underline x\ \underline y$ has the antitone sparse ranking function
+\[
+F \, \underline{n_0} \, (F \, \underline{n_1} \, (\dots(\Xi\ F\, \underline x\, \underline n)\dots) \mapsto g_2(n),
+\] 
+where $g_2$ is as in \Cref{ex:unbiased random walk}, using the same antitone function too.
 
@@
 
-Again, this is not actually the correct reduction order, in that $F$ should be applied to its argument $n_i$ and reduced to a value before its second argument is expanded, but this would complicate the presentation, and this version can be justified by \Cref{thm:confluent ranking} instead.
-The reduction strategy implied here should be clear enough (assuming that, where it isn't specified, it just matches $\cbv$), but to be more precise, let $r(F \underline{n_0} (\dots(F\ \underline {n_{k-1}}\ X)\dots)) = @_2^k;\cbv(X)$, where $X$ is not a value or of the form $F\ \underline{n_k}\ Z$ for some non-value $Z$.
-All terms match this pattern for exactly one value of $k \geq 0$, except for values.
+Again, this is not actually the correct reduction order, in that $F$ should be applied to its argument $n_i$ and reduced to a value before its second argument is expanded, 
+but this would complicate the presentation, and this version can be justified by \Cref{thm:confluent ranking} instead. 
+The reduction strategy implied here should be clear enough (assuming that, where it isn't specified, it just matches $\cbv$), but to be more precise, 
+let 
+\[
+r \, (F \, \underline{n_0} \, (\dots(F\, \underline {n_{k-1}}\, X)\dots)) = @_2^k;\cbv(X),
+\] 
+where $X$ is not a value or of the form $F\ \underline{n_k}\ Z$ for some non-value $Z$. All terms match this pattern for exactly one value of $k \geq 0$, except for values. 
 Within $X$, there is necessarily a redex at $\cbv(X)$. 
-The only conditions that $@_2^k;\cbv(X)$ is also a position of a redex in the whole term are that if the redex is a $\tsample$, it is not inside a $\lambda$ that's inside a $\tY$ or on the right of an application, which is true because $\cbv$ never selects a redex inside of a $\lambda$.
-After $n$ reaches $0$, the term is $F \underline{n_0} (\dots(F\ \underline {n_k}\ \underline x)\dots)$, then the innermost $F$ is evaluated completely in $\cbv$ order, therefore (by assumption) it terminates.
-Because there are no $\tY$-reductions in the evaluation of $F$, it can't reach any other term of the form $F\ \underline{n_k'}\ X$, therefore $k$ never changes until this whole subexpression reaches a number, at which point $k$ decreases by 1 and the next $F$ is evaluated.
+The only conditions that $@_2^k;\cbv(X)$ is also a position of a redex in the whole term are that if the redex is a $\tsample$, 
+it is not inside a $\lambda$ that's inside a $\tY$ or on the right of an application, which is true because $\cbv$ never selects a redex inside of a $\lambda$. 
+After $n$ reaches $0$, the term is $F \, \underline{n_0} \, (\dots(F\, \underline {n_k} \, \underline x)\dots)$, then the innermost $F$ is evaluated completely in $\cbv$ order, therefore (by assumption) it terminates.
+Because there are no $\tY$-reductions in the evaluation of $F$, it can't reach any other term of the form $F\, \underline{n_k'}\, X$, 
+therefore $k$ never changes until this whole subexpression reaches a number, at which point $k$ decreases by 1 and the next $F$ is evaluated.
 

### 2021-03-12 LICS21/antitone.tex
Tweak

@@
 \label{ex:ac-ranking}
-\begin{enumerate}
-\item 1D biased (towards 0) random walk
+\begin{compactenum}[(i)]
+\item One-dimensional (1D) biased (towards 0) random walk
 \[
@@
 \]
+%While {x > 0} do {x := x-1 \oplus_2/3 x:= x+1}
+%M2 = While {x > 0} do {x := x-1 \oplus_2/3 x:= x+2}
+\end{compactenum}
+\end{example}
 
-%While {x > 0} do {x := x-1 \oplus_2/3 x:= x+1}
-
-%M2 = While {x > 0} do {x := x-1 \oplus_2/3 x:= x+2}
-\end{enumerate}
-\end{example}
+\medskip
 
@@
 
-All of the following examples also have antitone ranking functions, provided in Appendix \ref{apx:antitone}.
+\paragraph*{More challenging examples} All of the following examples also have antitone ranking functions, provided in Appendix \ref{apx:antitone}.
 
@@
 \begin{align*}
-& (\tY\, \lambda f x.\, (\lambda e. X[e]) \, \tsample)\, \underline 0 \\
+& (\tY\, \lambda f x.\, (\lambda e. X[e]) \, \tsample)\, \underline 0 \quad \hbox{where}\\
 & X[e] := \tif{e \leq p - x^{-2}}{x}{f^2(x+1) \oplus_e f(x+1)}

### 2021-03-12 LICS21/applications.tex

@@
 
-In lightweight Metropolis-Hastings (LMH) or single-site MH (originally suggested by \cite{wingate2011lightweight}; see also \cite{rainforth2017Automating}),
+In Lightweight Metropolis-Hastings (LMH) or Single-site MH (originally suggested by \cite{wingate2011lightweight}; see also \cite{rainforth2017Automating}),
 the program is executed, with a random value being chosen each time a $\tsample$ redex is evaluated.

### 2021-03-12 LICS21/confluent.tex

@@
 
-With this notation, a more general reduction relation $\to$ can be defined.
+Thus we can define a more general reduction relation\footnote{In this section $\to$ always means the relation defined in \cref{def:more general red}.} $\to$.
 \begin{definition}
@@
 
-\medskip
+%\medskip
 Labelling the pre-chosen samples by the positions in the term would also not work because in some cases, a $\tsample$ will be duplicated before being reduced, for example, in $(\lambda x.
@@
 
-\begin{example}[Labelling samples by terms]
+\begin{example}%[Labelling samples by terms]
 Consider the terms

### 2021-03-12 LICS21/supermartingales-proofs.tex
Minor reformatting in appendix; added reference

@@
 
+\medskip
+
 \newcommand\transform[1]{{\ulcorner{#1}\urcorner}}
@@
 Finally thanks to the following
-\begin{theorem}[de Groote]
+\begin{theorem}[de Groote \cite{DBLP:conf/lfcs/Groote94}]
 \label{thm:de groote}
-The simply-typed nondeterministic lambda-calculus is strongly normalising. \qed
+The simply-typed nondeterministic lambda calculus is strongly normalising. \qed
 \end{theorem}
@@
 \end{enumerate}
+This concludes the proof of \cref{lem:TnBounded}.
 
@@
 
+\medskip
+
 \TMtYStoppingTime*

### 2021-03-12 LICS21/supermartingales.tex

@@
 and the filtration $(\mathcal{F}_n)_{n \geq 0}$ where $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
-Thus $T_M$ is the \defn{runtime} of $M$ (and $M$ is AST iff $\mu_{\entrosp}(T_M < \infty) = 1$).
+Thus $T_M$ is the \defn{runtime} of $M$ (and $M$ is AST iff $\mu_{\entrosp}(T_M < \infty) = 1$); we say that $M$ is \defn{positively almost surely terminating} (\defn{PAST}) if $\expect{T_M} < \infty$.
 
@@
 
-The random variable $T_M^\tY$, which we call the \emph{$\tY$-runtime of $M$}, can equivalently be defined as the number of $\tY$-reduction steps in the reduction sequence of $M$. 
+The random variable $T_M^\tY$, which we call the $\tY$-\defn{runtime} of $M$, can equivalently be defined as the number of $\tY$-reduction steps in the reduction sequence of $M$. 
 Note that, as the type system ensures that the reduction relation excluding $\tY$-reduction is strongly normalising, only finitely many reductions can occur in a row without one of them being a $\tY$-reduction, therefore $T_M^\tY < \infty$ a.s.~iff $M$ is AST.
-Moreover: 
+We say that $M$ is $\tY$-\defn{PAST} if $\expect{T^{\tY}_M} < \infty$.
+
 \begin{restatable}{lemma}{TMtYStoppingTime}
@@
 \begin{enumerate}
-\item If a closed PPCF term $M$ is rankable, then $M$ is AST (equivalently, $T^{\tY}_M < \infty$ almost surely), and $M$ is $\tY$-positively almost sure terminating ($\tY$-PAST) i.e.~$\expect{T^{\tY}_M} < \infty$.
+\item If a closed PPCF term $M$ is rankable, then $M$ is AST and $\tY$-PAST.
+%AST (equivalently, $T^{\tY}_M < \infty$ almost surely), and $M$ is $\tY$-positively almost sure terminating ($\tY$-PAST) i.e.~$\expect{T^{\tY}_M} < \infty$.
 
-\item If a closed PPCF term $M$ is strictly rankable, then $\expect{T_M} < \infty$ i.e.~$M$ is positively a.s.~terminating (PAST).
+\item If a closed PPCF term $M$ is strictly rankable, then $M$ is PAST.
+%$\expect{T_M} < \infty$ i.e.~$M$ is PAST.
 \end{enumerate}

### 2021-03-17 thesis proposal/thesis proposal.tex
Add thesis proposal

\documentclass[titlepage]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\newcommand{\tY}{\mathsf{Y}}
\newcommand{\tsample}{\mathsf{sample}}

\title{DPhil Thesis Proposal: Semantics of Probabilistic Functional Languages}
\author{Andrew Kenyon-Roberts}

\begin{document}
\maketitle
\section{Introduction}

%Existing approaches:
    %brief summary, axes of difference: operational, denotational, distribution-based, trace-based
    %Operational Semantics
    %Denotational Semantics
        %Denotational Semantics in the non-probabilistic case, types and inconsistencies
        %QBSes
    %Distribution-based Semantics
    %Trace-based Semantics

\section{My Previous Research}
My thesis will be building on the paper I recently submitted to LICS (in collaboration with Luke Ong), Supermartingales, Ranking Functions and Probabilistic Lambda Calculus. In it, we extend existing work on proving almost-sure termination using ranking functions (functions from the program state to non-negative real numbers which at each step have a non-positive expected change, and which satisfy some further progress condition) to the context of a functional language with continuous distributions. Because the language used is simply-typed, the only way that non-termination is possible is using the explicit recursion contruct, $\tY$, therefore adding a condition that the ranking function must decrease by 1 for every $\tY$-reduction step provides a bound on the expected number of $\tY$-reduction steps, which implies almost sure termination. This result is then extended in three ways (which can all be combined).

The first extension is sparse ranking functions. A sparse ranking function may be defined at a subset of the reachable states of the program, with the condition instead being that the expected value of the ranking function at the next step where it is defined is not greater than the current value (and, if there are intermediate $\tY$-reduction steps, there is a corresponding strict decrease). Although this does not increase the strength of the technique overall, it makes it easier to apply, because the ranking function only needs to be defined at those points in the program's execution which are actually interesting. This extension is new, having not been applicable in the context of first-order languages because the completed body of a while loop provides a natural set of checkpoints.

The second extension is antitone ranking functions. This idea is taken from \cite{some citation needed}, just adapted to the new functional context. A function $\epsilon : \mathbb R_{\geq 0} \to \mathbb R_{> 0}$, which is required to be monotoically decreasing, is added to the definition, and rather than the ranking function having to decrease by 1 at every $\tY$-reduction step, it only has to decrease by $\epsilon$ of the current value of the ranking function. This allows the termination of programs which terminate much more slowly (such as the unbiased random walk) to be proven, whereas the basic result was only applicable to terms which terminated with a finite expected number of $\tY$-reduction steps.

The third extension, and the most interesting and relevant to my intended thesis topic, is ranking functions with respect to alternative reduction strategies. This result is proven using a variant of trace semantics which is able to satisfy a restricted confluence property. Probabilistic functional languages in general do not have the same sort of confluence properties as pure lambda calculus. For example, assuming that samples are drawn from the uniform distribution on $[0,1]$, the term $(\lambda x. x + x) \tsample$ evaluates either to the uniform distribution on $[0,2]$ or to the triangle-shaped distribution with the same support, with its peak at 1, depending on whether the sample is evaluated first (in which case, one value is doubled) or the $\beta$-redex is evaluated first (in which case, two independent samples are taken and added). If, however, certain reductions are excluded (in the results I proved in this paper, it's $\beta$-reductions whose arguments aren't values, and $\tsample$-reductions inside $\lambda$s either inside $\tY$s or in the argument of applications, although other criteria would also be possible), confluence can be re-gained.

This is enough to ensure that the distribution of results is (barring cases where some reduction strategies may terminate while others don't) the same, but that would be difficult to prove directly, and the result doesn't extend to the same traces giving the same results. For example, if the trace is simply a sequence of samples which are used in the order in which the $\tsample$-reductions occur, then the term $\tsample - \tsample$ reduces with the trace $0, 1, ...$ to either $1$ or $-1$, depending on which $\tsample$ is reduced first. This example can be fixed by labelling the samples in the trace by positions in the term rather than the order in which they're used, but in other cases, that would still not be sufficient. Consider, for example, $(\lambda f. f 0 + f 0) (\lambda x. \tsample)$. This reduces (by some reduction strategies) to $\tsample + \tsample$, but the $\tsample$s here don't correspond in a unique way to the $\tsample$s in the initial term, therefore labelling samples by positions in the initial term isn't sufficient. Perhaps it would be possible to in some way label those $\tsample$s by a combination of the position of the $\tsample$ they're derived from and the occurrences of the variable $f$ that they were substituted in to, but the approach I ended up taking was a somewhat more brute-force solution, in that if any labelling scheme would have worked, so would this one, by design. Samples are labelled by a combination of a term reachable from the initial term (technically a reachable skeleton instead, with all the real numbers removed, for measure-theoretic reasons) and a position within that term, with an equivalence relation defined on these (reachable term, position) pairs that is the minimum required to ensure confluence.

This definition is sufficient to prove the AST result that is nominally the main focus of that paper, but also seems like something that has a lot of potential to be used for other purposes. The limits of confluence for probabilistic functional languages is not something I have seen discussed extensively elsewhere, except the fact that some reduction strategies (specifically call-by-name and call-by-value) give different results. The way that confluence is obtained with a trace-based semantics also seems likely to extend to a denotational trace semantics.

\section{Planned research}
\subsection{Using Labelling to Combine Denotational and Trace Semantics}
Trace-based semantics is, in a sense, finer than distribution-based semantics. Given a measurable function $f : \Lambda \times \mathbb S \to A$ that describes some property of a term $\in \Lambda$ by using a trace from some sampling space $\mathbb S$, which has some fixed probability measure on it $\mu_{\mathbb S}$, this can be converted to a kernel from $\Lambda$ to $A$ using the push-forward measure of $\mu_{\mathbb S}$, but some information is lost in the process: which specific trace lead to which outcome. There is no corresponding simple way to convert a distribution-based semantics to a trace-based semantics that actually results in all the correct structure expected of a trace-based semantics. In trace-based semantics, it is possible to define things like correlations between the random variables it defines, because they are all defined in the same probability space, whereas this is not possible given merely a distribution of outputs.

There are also some advantages of denotational over operational semantics. It provides meanings to sub-programs that can be reasoned about independently of their context. It can be more suitable for proving things like the effectiveness of a type system for proving termination, as the structure of the semantics more closely matches the derivations of typing judgements.%This paragraph could really do with some references. They don't even have to be probabilistic, just demonstrations of how useful denotational semantics in general can be.

Existing approaches are either denotational and distribution-based, operational and distribution-based, or operational and trace-based. The combination of denotational and trace-based is more difficult to define, because some way of distinguishing multiple copies of the same sample statement is needed, so that they can derive their randomness from different parts of the trace. In operational semantics, this is easy enough, because the argument of a beta redex does not need to be evaluated at all before it is duplicated by the reduction, and its copies can simply be dealt with one at a time, like any other independent subterms. In denotational semantics, it is not the terms themselves which are combined in an application, but their interpretations, therefore the interpretation of the argument needs some way to retain the possibility of having multiple different results from the random process, even though there is only one trace used overall.

An extension of the labelling scheme used in the confluent trace semantics of my previous paper could provide just such a possibility. The exact labelling scheme used there depends explicitly on the reduction sequences (and is also rather convoluted), so it is not suitable to use directly in defining a denotational semantics, but it seems highly likely that some labelling scheme that's equivalent but differently defined would be. If all of the subterms of the original term are given unique labels, and the labels form some sort of magma, so that when a reduction occurs, the labels of the argument's subterms and the function's subterms can be combined in some way to produce the labels of the reduct's subterms, then a trace that contains a value for every possible combination of labels could be used. The random samples taken in the evaluation of each subterm then depend on the transformation of the labels that occurs, which is distinct for each copy of the subterm.

\subsection{Using labelling to perform efficient inference}
Another possible application of a labelling scheme for sample is in improving the efficiency of certain inference algorithms. In lightweight metropolis-hastings, the program is first executed, picking random values for all the samples taken, with the trace of samples being remembered. Then, random variations on the trace are tested, re-running the program with the modified trace and accepting or rejecting the modification based on the difference in the likelihood of the runs. It is essential to the efficiency of this algorithm that the score of a modified run is highly correlated to the score of the original run, as otherwise the acceptance ratio becomes low. At the same time, it is necessary that the proposed changes actually change the trace by not too small an amount, so that the distribution of traces converges to the stationary distribution quickly.

The correlation between runs can be impeded if the number of samples taken varies depending on the results of earlier random choices. In the basic version of the algorithm, the trace is simply a list of the samples in the order in which they are taken. If some sample is never used, however, the rest of the samples in the list become misaligned, and end up taking different roles in the resulting execution, completely changing the score with only a small change to the trace.

This, then, is another possible application of a labelling scheme similar to the one previously mentioned. By better approximating the roles that samples play in the program execution by the labels, the problem of trace misalignment can be reduced. In \cite{lee2019towards}, a similar approach is taken for variational inference with a first-order language, but in that case, they require that the labels be explicitly provided in the program (potentially being computed using the values of other variables). There is also a similar proposal in \cite{pmlr-v15-wingate11a}.

The requirements of a labelling scheme for this purpose are somewhat different from the previous cases. In the other cases, the labels needed to match across program executions that differ nondeterministically in their reduction order, but for the purposes of inference, the relevant different executions differ by the randomly selected samples. A labelling scheme adapted to the problem of inference could probably be simplified relative to the confluent trace semantics defined in my previous paper, or the more algebraic variant I intend to develop for the denotational trace semantics mentioned earlier. I am not yet sure how similar it would end up, after this simplification, to the much simpler labelling scheme described in \cite{pmlr-v15-wingate11a}.

\subsection{Completeness of Antitone Rankability}
A term is rankable iff every term reachable from it terminates with finitely many $\tY$-reduction steps in expectation. This essentially puts a limit on how slowly a term can terminate while still being rankable, so any term which is not $\tY$-positive almost surely terminating cannot be proven AST by assigning it a ranking function. With antitone ranking functions, there is no analogous restriction: there are arbitrarily slowly terminating terms (i.e. for any probability distribution on $\mathbb N$, there is a term whose distribution of number of $\tY$-reduction steps before termination is the given distribution) which are antitone rankable. It therefore seems rather likely that any term such that every term reachable from is is AST is antitone rankable. I intend to try to prove this result, even though it does not relate very closely to my overall thesis topic, because it would make my existing work more complete.
\end{document}

### 2021-03-18 thesis proposal/thesis proposal.tex
Add introduction and start prior work section
I know I need to add references. I'm planning to do that after most
of the writing.

@@
 \newcommand{\tsample}{\mathsf{sample}}
+\newcommand{\tscore}{\mathsf{score}}
 
@@
 \section{Introduction}
+The augumentation of lambda calculus with random sampling and conditioning, the fundamental constructs of probabilistic programming, can be interpreted formally in many different ways, and provides certain unique challenges in defining the language's semantics.
+The standard approach to defining random things, where all the random values are considered as functions from a probability space, has the problem that the random samples, which should form the basis of this probability space, are not specified in advance.
+There are various possible schemes of organising the probability space such that this data is available, known as trace or entropy space semantics.
+The conditioning construct, which modifies the posterior probability of an event, fits somewhat unnaturally in this framework, and has to be specified as an extra step that modifies the probabilities at the end.
 
-%Existing approaches:
-    %brief summary, axes of difference: operational, denotational, distribution-based, trace-based
-    %Operational Semantics
-    %Denotational Semantics
+Alternatively, the distribution of results can be specified more directly as a distribution.
+This omits any information on joint distributions from the semantics, but this itself can be in some ways an advantage, as a subterm can be duplicated without being correlated to itself.
+It would be natural to attempt to use the category of measurable spaces, however, that turns out to not be cartesian closed.
+This motivates the introduction of quasi-Borel spaces as an alternative way in which a distribution can be defined, which has a more restrictive set of morphisms and does not have this issue, but is nevertheless still able to represent all the functions definable as programs.
+
+For my thesis, I plan to extend the existing literature and my own previous work in defining the semantics of probabilistic functional languages.
+
+\section{Existing approaches}
+The various existing approaches to defining the semantics of probabilistic functional languages can generally be categorised as operational or denotational (depending on whether it works by defining the state of the program at each step, or by defining the meanings of subprograms independently and combining them together) (this distinction applies to functional languages more generally, rather than being specific to probabilistic languages), and distribution-based or trace-based (depending on whether only the distribution of outputs is defined, or the output that occurs is defined as a function of a random input).
+
+As a running example, I will be using a simply-typed call by value language with a random sampling construct in the form of a probabilistic choice $M \oplus N$, a conditioning construct $\tscore(r)$ for non-negative real constants $r$, and countably many primitive values of some base type $\mathsf T$.
+Although this language is rather deficient (there is no way to use continuous distributions, for example, and no recursion), it will be sufficient to illustrate the different styles of semantics.
+
+\subsection{Operational Semantics}
+In operational semantics, the meaning of the program is defined by repeatedly transforming a term into other terms, until it reaches a value, a term which is considered finished, which is the output.
+A small-step trace-based operational semantics of the example language can be defined as follows.
+
+First, an \emph{environment} is defined as
+\begin{equation*}
+E := \cdot \mid E M \mid V E,
+\end{equation*}
+so that an environment is a term with a single hole, where $M$ is a term and $V$ is a value.
+Substituting terms into environments is defined in the usual way.
+The single-step reduction relation can then be defined as
+\begin{align*}
+(E[(\lambda x. M) V], t, s) & \to (M[V/x], t, s) \\
+(E[M \oplus N], t, s) & \to (E[M], 0:t, s) \\
+(E[M \oplus N], t, s) & \to (E[N], 1:t, s) \\
+(E[\tscore(r)], t, s) & \to (E[(\lambda x. x)], t, rs),
+\end{align*}
+where $t$ is a list of booleans, $:$ is the cons operator, and $s$ is a real number. The probability of $M$ reaching a value $V$ can then be defined as
+\[
+\sum_{(M,[],1) \to^* (V,t,s)} 2^{-\mathrm{length}(t)} s.
+\]
+
+Alternatively, this could be defined in a different but closely related version of operational semantics, the big-step semantics:
+\begin{align*}
+\frac{(M, t_0, s_0) \Downarrow (\lambda x. M', t_1, s_1) \; (N, t_1, s_1) \Downarrow (N', t_2, s_2) \; (M'[N'/x], t_2, s_2) \Downarrow (O, t_3, s_3)}{(M N, t_0, s_0) \Downarrow (O, t_3, s_3)} \quad
+\frac{(M, 0:t_0, s_0) \Downarrow (M', t_1, s_1)}{(M \oplus N, t_0, s_0) \Downarrow (M', t_1, s_1)} \quad
+\frac{(N, 1:t_0, s_0) \Downarrow (N', t_1, s_1)}{(M \oplus N, t_0, s_0) \Downarrow (N', t_1, s_1)} \quad
+\frac{}{(\tscore(r), t, s) \Downarrow (\lambda x. x, t, rs)}.
+\end{align*}
+The notation in this case could be simplified somewhat, but that would obscure the connection to the small-step semantics.
+If $V$ is a value, then $(M, t_0, s_0) \to^* (V, t_1, s_1)$ iff $(M, t_0, s_0) \Downarrow (V, t_1, s_1)$.
+
+\subsection{Denotational Semantics}
         %Denotational Semantics in the non-probabilistic case, types and inconsistencies

### 2021-03-23 thesis proposal/thesis proposal.tex
Write more about existing approaches

@@
 \usepackage{amsfonts}
+\usepackage{stmaryrd}
 \newcommand{\tY}{\mathsf{Y}}
@@
 \newcommand{\tscore}{\mathsf{score}}
+\newcommand{\denotation}[1]{\llbracket #1 \rrbracket}
 
@@
 
-As a running example, I will be using a simply-typed call by value language with a random sampling construct in the form of a probabilistic choice $M \oplus N$, a conditioning construct $\tscore(r)$ for non-negative real constants $r$, and countably many primitive values of some base type $\mathsf T$.
+As a running example, I will be using a simply-typed call by value language with a random sampling construct in the form of a probabilistic choice $M \oplus N$, a conditioning construct $\tscore(r)$ for non-negative real constants $r$, and a countable set $T$ of primitive values of some base type $\mathsf T$.
 Although this language is rather deficient (there is no way to use continuous distributions, for example, and no recursion), it will be sufficient to illustrate the different styles of semantics.
 
-\subsection{Operational Semantics}
+\subsection{Operational Semantics} \label{operational semantics}
 In operational semantics, the meaning of the program is defined by repeatedly transforming a term into other terms, until it reaches a value, a term which is considered finished, which is the output.
@@
 (E[(\lambda x. M) V], t, s) & \to (M[V/x], t, s) \\
-(E[M \oplus N], t, s) & \to (E[M], 0:t, s) \\
-(E[M \oplus N], t, s) & \to (E[N], 1:t, s) \\
+(E[M \oplus N], 0:t, s) & \to (E[M], t, s) \\
+(E[M \oplus N], 1:t, s) & \to (E[N], t, s) \\
 (E[\tscore(r)], t, s) & \to (E[(\lambda x. x)], t, rs),
@@
 \[
-\sum_{(M,[],1) \to^* (V,t,s)} 2^{-\mathrm{length}(t)} s.
+K \sum_{(M,t,1) \to^* (V,[],s)} 2^{-\mathrm{length}(t)} s,
 \]
+where $K$ is a normalising constant to make the probabilities all add up to 1.
 
 Alternatively, this could be defined in a different but closely related version of operational semantics, the big-step semantics:
-\begin{align*}
-\frac{(M, t_0, s_0) \Downarrow (\lambda x. M', t_1, s_1) \; (N, t_1, s_1) \Downarrow (N', t_2, s_2) \; (M'[N'/x], t_2, s_2) \Downarrow (O, t_3, s_3)}{(M N, t_0, s_0) \Downarrow (O, t_3, s_3)} \quad
+\begin{multline*}
+\frac{(M, t_0, s_0) \Downarrow (\lambda x. M', t_1, s_1) \; (N, t_1, s_1) \Downarrow (N', t_2, s_2) \; (M'[N'/x], t_2, s_2) \Downarrow (O, t_3, s_3)}{(M N, t_0, s_0) \Downarrow (O, t_3, s_3)} \\
 \frac{(M, 0:t_0, s_0) \Downarrow (M', t_1, s_1)}{(M \oplus N, t_0, s_0) \Downarrow (M', t_1, s_1)} \quad
-\frac{(N, 1:t_0, s_0) \Downarrow (N', t_1, s_1)}{(M \oplus N, t_0, s_0) \Downarrow (N', t_1, s_1)} \quad
+\frac{(N, 1:t_0, s_0) \Downarrow (N', t_1, s_1)}{(M \oplus N, t_0, s_0) \Downarrow (N', t_1, s_1)} \\
 \frac{}{(\tscore(r), t, s) \Downarrow (\lambda x. x, t, rs)}.
-\end{align*}
+\end{multline*}
 The notation in this case could be simplified somewhat, but that would obscure the connection to the small-step semantics.
@@
 \subsection{Denotational Semantics}
-        %Denotational Semantics in the non-probabilistic case, types and inconsistencies
-        %QBSes
-    %Distribution-based Semantics
-    %Trace-based Semantics
+The general format of denotational semantics is a function $\denotation M_\gamma$, from a term $M$ to its value, given the values of all the free variables in the context $\gamma$.
+The type of output depends on the type of the term, given by $\denotation A$ which is defined recursively.
+For a non-probabilistic language, it would be something like $\denotation{\mathsf T} = T$, $\denotation{A \to B} = \denotation B^{\denotation A}$.
+If $M$ satisfies the typing judgement $\Gamma \vdash M : A$, and $\gamma \in \denotation \Gamma$ (i.e.~$\gamma$ is a function from the variables in $\Gamma$ to values such that if $x:B$ is in $\Gamma$, $\gamma (x) \in \denotation B$), then $\denotation M_\gamma \in \denotation A$.
+
+For the example language but without the probabilistic constructs (i.e.~just simply-typed lambda calculus with constants), a denotational semantics could be defined as
+\begin{align*}
+\denotation t_\gamma & = t\,(\text{for }t \in T) \\
+\denotation x_\gamma & = \gamma(x) \\
+\denotation{M N}_\gamma & = \denotation M_\gamma(\denotation N_\gamma) \\
+\denotation{\lambda x. M }_\gamma(X) & = \denotation M_{\gamma ; x \mapsto X}.
+\end{align*}
+In order to define a denotational semantics for a probabilistic language, the result would have to be not just a value, but some sort of distribution of values.
+The rule $\denotation{A \to B} = \denotation B^{\denotation A}$ also has to be changed to take into account that a function may produce outputs that differ randomly each time it is called, so something like the set of kernels from $\denotation A$ to $\denotation B$ (a trace-based denotational semantics would also probably be possible, but much more complicated, but as far as I am aware, such a thing has never been defined).
+This would require a measure structure on $\denotation A$ and $\denotation B$ to be defined, but in some cases it is not possible to define a measurable structure on the set of kernels between two measurable spaces such that the application map taking a kernel $A \rightsquigarrow B$ and an element of A is itself measurable, therefore the denotations of higher-order types could be undefined.
+This motivates the introduction of quasi-Borel spaces.
+
+\subsubsection{Quasi-Borel Spaces}
+A quasi-Borel space is a pair of a base space $X$ with a set of random variables $M_X \subset X^{\mathbb R}$ that satisfies certain axioms.
+In many ways, quasi-Borel spaces are similar to measurable spaces.
+A $\sigma$-algebra on $X$ can be converted to a quasi-Borel space structure by taking all the measurable functions as random variables, and a quasi-Borel space can be converted to a measurable space by taking as measurable sets those whose inverse images under all random variables are measurable.
+There is a natural notion of QBS morphisms (functions whose compositions with random variables are random variables) and measures defined by combining measures on $\mathbb R$ with random variables which can be used to define integrals.
+The crucial difference to measurable spaces is that the set of morphisms betweem two QBSes is itself a QBS, so that the category of QBSes is cartesian closed.
+
+Let $G(X)$ be the set of measures on a QBS $X$, equipped with its own QBS structure, with the dirac delta $\delta : X \to G(X)$ and $\eta : (X \to G(X)) \to (G(X) \to G(X))$ as the monad operations, and let $QBS(X, Y)$ be the set of morphisms between two $QBSes$.
+Let $T$ be equipped with the discrete QBS structure.
+A denotational semantics for the example language, such that $\denotation M_\gamma \in G(\denotation A)$ if $\Gamma \vdash M : A$ and $\gamma \in \denotation \Gamma$ can then be given as follows.
+
+\begin{align*}
+\denotation{\mathsf T} & = T \\
+\denotation{A \to B} & = \denotation A \to G(\denotation B) \\
+\denotation{t}_\gamma & = \delta(t) \\
+\denotation{x}_\gamma & = \delta(\gamma(x)) \\
+\denotation{M N}_\gamma & = \eta(f \mapsto \eta(f)(\denotation N_\gamma))(\denotation M_\gamma) \\
+\denotation{\lambda x. M}_\gamma(X) & = \denotation M_{\gamma ; x \mapsto X} \\
+\denotation{M \oplus N}_\gamma & = \frac 1 2 \denotation M_\gamma + \frac 1 2 \denotation N \gamma \\
+\denotation{\tscore(r)}_\gamma & = r \denotation{\lambda x. x}_\gamma = r \delta(x \mapsto \delta(x)).
+\end{align*}
+
+Although the type of the result is different, this denotational semantics can still be shown to be equivalent to the operational semantics previously presented.
+For programs of type $\mathsf T$, both versions of the semantics produce a distribution on $T$ (one a QBS-style distribution and the other a measure-space-style distribution, but these can be converted), which are equal.
+The statement for function types is a bit more complicated and less direct, because the denotational semantics produces a distribution on (QBS) kernels, while the operational semantics produces a distribution on lambda terms.
+
+Having a denotational semantics available makes certain sorts of proofs more convenient.
+For example, if there's a program rearrangement that can be expressed as replacing $A[M,N]$ with $B[M,N]$ (for arbitrary $M, N$), then the proof of equivalence can be quite direct, showing that the way $\denotation M$ and $\denotation N$ are combined is equivalent (possibly subject to some conditions on certain variables not being free in $M$ and $N$, which implies that $\denotation M_\gamma$ and $\denotation N_\gamma$ are independent of the values of those variables in $\gamma$), whereas with operational semantics, it is much more complicated to keep track of the way the arbitrary subterms are duplicated, transformed and combined as execution progresses.
+Similarly, the notion of contextual equivalence is not necessary in denotational semantics (being trivially equivalent to equality of values).
+
+Another useful property of denotation semantics is that termination properties are much more directly visible.
+If the type system is strong enough to imply termination, this manifests directly as every term being given a value.
+If non-terminating terms are possible, it will be necessary to introduce, as part of the semantics, special cases to deal with that, otherwise the definition will simply not be consistent.
+
+\subsection{Trace-based semantics}
+A trace-based semantics for the example language was already described in Section \ref{operational semantics}.
+In trace-based semantics, the results are defined not just as distributions, but as random variables.
+Everything can be seen as taking place in a probability space (neglecting $\tscore$ which, at least if it is allowed to take values greater than 1, does not entirely fit), with the term after $n$ steps, the values of certain samples, the running time, and the final value all being random variables in the same space.
+
+There are a few variations on how the traces work.
+In some versions (as is the case in Section \ref{operational semantics}), each run of the program is assigned a trace of finite length, containing exactly those samples actually used in that run.
+If the set of results for an individual sampling operation is $S$ (in this case it's the set of 2 elements, but in a language with continuous distributions, it could be $\mathbb R$, or an interval, or something else), the set of traces is $\bigcup_{n \in \mathbb N} S^n$, and the measure defined on the set of traces is not actually a probability measure, as each set $S^n$ has measure 1.
+This still acts somewhat like a probability measure though, as traces of different lengths are mutually exclusive.
+If a trace $a$ is a prefix of a trace $b$ (with $b = a + c$), then it is not possible that $(M, a, 1) \to^* (V, [], s)$ and also $(M, b, 1) \to^* (V', [], s')$, because the evaluation with any given trace is deterministic, and $(M, b, 1) \to^* (V, c, s)$, at which point it could not proceed any further, having already reached a value.
+For any term, the set of traces for which it terminates therefore has measure at most 1.
+
+Alternatively, the trace could contain infinitely many values, with any finite execution of a program leaving some of the samples unused.
+This is in some ways a little simpler, because the set of traces can simply have a probability distribution, and the trace does not need to depend in any way on the program (by having its length matched).
+An additional variant of this style of trace is to split the trace in three parts, all infinitely long, for the application case of the big-step semantics, rather than having the each step (evaluating the function, evaluating the argument, then evaluating the result of the $\beta$ reduction) use however many samples it needs from the start of the trace in turn.
+
+\subsection{Distribution-Based Semantics}
+Rather than defining the value of a program as a random variable and then defining the distribution based on that, it is possible to define the distribution of results directly.
+The example given above of operational semantics was distribution based (in the QBS sense of ``distrbution''), but to more clearly illustrate the difference, I will give a distrbution-based operational semantics for the example language, using the same definition of environments as in the trace-based semantics.
+
+\begin{align*}
+s(E[(\lambda x. M) V]) & = \delta(M[V/x]) \\
+s(E[M \oplus N]) & = \frac 1 2 \delta(E[M]) + \frac 1 2 \delta(E[N]) \\
+s(E[\tscore(r)]) & = r \delta(E[(\lambda x. x)]) \\
+s(V) & = \delta(V).
+\end{align*}
+
+The multi-step function $s^n$ can then be defined as $s^0(M) = M$, $s^{n+1}(M) = \int s^n(x)\,\mathrm d s(M)(x)$, then $s^\infty(M)$ can be defined as $\lim_{n \to \infty} s^n(M)$, which (after re-normalising) gives the distribution of final values.
+In this case, everything terminates because of the type system, therefore the limit is always well-defined.
+For an untyped language, or one with an unrestricted recursion operator, the definition of the final result would be a little more complicated, but the general idea is the same.
 

### 2021-03-23 LICS21/review response.txt
Some comments

@@
 As the other reduction strategies mainly considered in this work (i.e.
-used with Theorem V1.16) are intended to be specifically designed for particular programs, the other examples are given along with particular example programs, specifically the extended discussions of examples V.14 and VI.17 in the appendices.
+used with Theorem V1.16) are intended to be specifically designed for particular programs, the other examples are given along with particular example programs, specifically the extended discussions of Examples V.14 and VI.17 in the appendices.
 
-CBN is not considered a valid reduction strategy because it violates the condition that the argument of a beta redex must be a value.
+CBN is not considered a valid reduction strategy because it violates the condition that the argument of a beta redex must be a value, which follows from the definition of reduction strategy (p.
+10) and of redex (Def VI.1).
 
@@
 
-(@Luke: I remember looking up what "contractum" meant, but I can't find it now.)
+(@Luke: I remember looking up what "contractum" meant, but I can't find it now.) [See my answer on Slack.]
 
-At the end of section IV, the next reduction in i ⊕_0.5 (Y \lambda fn.
-n ⊕_0.5 f(n+1)) (i+1) is the ⊕_0.5 (or, to be more precise, the sample implicit in that notation), not the Y, because ⊕ is syntactic sugar for a sort of if, and the branches of an if are not evaluated before one of them is selected.
+At the end of Section IV, the next reduction in 
+   i ⊕_0.5 (Y \lambda fn. n ⊕_0.5 f(n+1)) (i+1) 
+is the ⊕_0.5 (or, to be more precise, the sample implicit in that notation), not the Y, because ⊕ is syntactic sugar for a sort of if, and the branches of an if are not evaluated before one of them is selected.
 
@@
 
-The relation defined in VI.1 is non-deterministic, and alpha can take any value that locates one of these types of redexes.
+The relation defined in VI.1 is nondeterministic, and alpha can take any value that locates one of these types of redexes.
 
-Labelling the samples with elements of a set A simply refers to picking samples s from I^A, then using s(a) for the sample reduction whenever the relevant position/number of samples previoustly taken/reduction sequence and position (depending on the labelling) is a.
+Labelling the samples with elements of a set A simply refers to picking samples s from I^A, then using s(a) for the sample reduction whenever the relevant position/number of samples previously taken/reduction sequence and position (depending on the labelling) is a.
 
-In example VI.2, it's not clear whether you're referring to the skeleton placeholder number, sans-serif X, or simply that x should be uppercase because it's a term (which would be correct).
+In Example VI.2, it's not clear whether you're referring to the skeleton placeholder number, sans-serif X, or simply that x should be uppercase because it's a term (which would be correct).
 "I" here is meant as shorthand for the identity function.
@@
 Regarding the general comments on section 6.
-The restrictions in reduction strategies are the extra conditions in Def VI.1 on what reductions are valid (specifically, that the argument of a beta redex must be a value, and the restrictions on positions where sample reductions can occur).
+The restrictions in reduction strategies are the extra conditions in Definition VI.1 on what reductions are valid (specifically, that the argument of a beta redex must be a value, and the restrictions on positions where sample reductions can occur).
 The body of a Y subterm is effectively both the function and the argument in an application.
-If it contians multiple occurrences of the relevant variable, it will be duplicated by the Y reduction.
-More specifically, consider (Y \lambda f n.
-if n = 0 then (\lambda x.
-sample) else \lambda x.(f 0 0 + f 0 0)) 1 0.
-This reduces to something like (lambda x.
-sample) 0 + (\lambda x.
-sample) 0.
+If it contains multiple occurrences of the relevant variable, it will be duplicated by the Y reduction.
+More specifically, consider
+   (Y \lambda f n. if n = 0 then (\lambda x. sample) else \lambda x.(f 0 0 + f 0 0)) 1 0. 
+This reduces to something like 
+   (lambda x. sample) 0 + (\lambda x. sample) 0. 
 If the sample had been allowed to be reduced first, the samples would have ended up correlated.
-The fact that the position has to have a Y and a lambda in it to be excluded is simply to accommodate things like Y if sample > 0.5 then M else N, where there actually should be reductions inside the Y.
+The fact that the position has to have a Y and a lambda in it to be excluded is simply to accommodate things like
+   Y if sample > 0.5 then M else N
+where there actually should be reductions inside the Y.
+
 Yes, (lambda x.
@@
 Regarding reviewer 3's questions about the completeness conjecture:
-I am not aware of any particularly relevant previous literature.
-I did spend a while trying to solve the problem, and do intend to try some more in future, but didn't have time to continue.
+We are not aware of any particularly relevant previous literature.
+We did spend a while trying to solve the problem, and do intend to try some more in future, but didn't have time to continue.
 Part of what makes this question so difficult is that you have to construct an antitone function that represents the rate of convergence of all of the different states the program could be in, and assign all the program states ranking function values, despite it be not obvious how to compare, for example, one state which would either terminate in a few steps or extremely slowly, and another which would terminate after a long time, but more deterministically.

### 2021-03-24 thesis proposal/thesis proposal.tex

@@
 \documentclass[titlepage]{article}
+
+\newif\ifdraft
+\drafttrue %% To hide all comments and highlighting, just comment this line out 
+\input{../Style/comment}
 \usepackage{amsmath}
@@
 It provides meanings to sub-programs that can be reasoned about independently of their context.
-It can be more suitable for proving things like the effectiveness of a type system for proving termination, as the structure of the semantics more closely matches the derivations of typing judgements.%This paragraph could really do with some references.
+It can be more suitable for proving things like the effectiveness of a type system for proving termination \lo{Yes, provided the denotational semantics is \emph{adequate} (which in the non-random case means: $M$ converges iff the denotation of $M$ is not $\bot$), which is a non-trivial ask in probabilistic computation -- witness \cite{DBLP:journals/pacmpl/VakarKS19}.}, as the structure of the semantics more closely matches the derivations of typing judgements.%This paragraph could really do with some references.
 They don't even have to be probabilistic, just demonstrations of how useful denotational semantics in general can be.
+\lo{These are the advantages of a \emph{compositional} semantics, and compositionality is the defining feature of denotational semantics.}
 
@@
 
+\lo{A denotational trace-based semantics for SPCF does sound like an interesting problem.
+What might one be able to do with this model, which one can't with the standard trace-based operational semantics?}
+
+\lo{To envisage what such a model might look like, I find it helpful to consider the stochastic symbolic execution semantics of SPCF \cite{MakOP20b}.} \lo{To be continued.}
+
 An extension of the labelling scheme used in the confluent trace semantics of my previous paper could provide just such a possibility.
@@
 The random samples taken in the evaluation of each subterm then depend on the transformation of the labels that occurs, which is distinct for each copy of the subterm.
+
+\lo{To give a denotational semantics of the simply-typed lambda calculus is to construct a cartesian closed category.
+One would expect a denotational trace-based semantics to have a strong operational flavour.
+This brings to mind game semantics, which is an unusual denotational semantics in exactly this sense that it is operational in  character.
+As you may know, it gave rise to the first syntax-independent fully abstract model for PCF.}
 
@@
 I intend to try to prove this result, even though it does not relate very closely to my overall thesis topic, because it would make my existing work more complete.
+
+\bibliographystyle{apalike}
+\bibliography{../Bib/references-Beutner,../Bib/references}
+
 \end{document}

### 2021-03-30 thesis proposal/thesis proposal.tex
Add references

--- 2021-03-23 thesis proposal/thesis proposal.tex
@@
 \usepackage{amsfonts}
+\usepackage{natbib}
+\setcitestyle{numbers, open={[}, close={]}}
+\bibliographystyle{plainnat}
 \usepackage{stmaryrd}
@@
 \author{Andrew Kenyon-Roberts}
+
+\usepackage{hyperref}
 
@@
 This omits any information on joint distributions from the semantics, but this itself can be in some ways an advantage, as a subterm can be duplicated without being correlated to itself.
-It would be natural to attempt to use the category of measurable spaces, however, that turns out to not be cartesian closed.
-This motivates the introduction of quasi-Borel spaces as an alternative way in which a distribution can be defined, which has a more restrictive set of morphisms and does not have this issue, but is nevertheless still able to represent all the functions definable as programs.
+To define a denotational semantics for a higher-order probabilistic language, it would be natural to attempt to use the category of measurable spaces, however, that turns out to not be cartesian closed.
+This motivates the introduction of quasi-Borel spaces \cite{8005137} as an alternative way in which a distribution can be defined, which has a more restrictive set of morphisms and does not have this issue, but is nevertheless still able to represent all the functions definable as programs.
 
@@
 In operational semantics, the meaning of the program is defined by repeatedly transforming a term into other terms, until it reaches a value, a term which is considered finished, which is the output.
+Several variants of operational semantics can be seen, for example, in \cite{borgstrom2016lambda}.
 A small-step trace-based operational semantics of the example language can be defined as follows.
@@
 The type of output depends on the type of the term, given by $\denotation A$ which is defined recursively.
-For a non-probabilistic language, it would be something like $\denotation{\mathsf T} = T$, $\denotation{A \to B} = \denotation B^{\denotation A}$.
-If $M$ satisfies the typing judgement $\Gamma \vdash M : A$, and $\gamma \in \denotation \Gamma$ (i.e.~$\gamma$ is a function from the variables in $\Gamma$ to values such that if $x:B$ is in $\Gamma$, $\gamma (x) \in \denotation B$), then $\denotation M_\gamma \in \denotation A$.
+For a non-probabilistic language, it would be something like $\denotation{\mathsf T} = T$, $\denotation{A \to B} = \denotation B^{\denotation A}$.\cite{barendregt2013lambda} If $M$ satisfies the typing judgement $\Gamma \vdash M : A$, and $\gamma \in \denotation \Gamma$ (i.e.~$\gamma$ is a function from the variables in $\Gamma$ to values such that if $x:B$ is in $\Gamma$, $\gamma (x) \in \denotation B$), then $\denotation M_\gamma \in \denotation A$.
+
+A key property that makes it denotational semantics (the result of the operational semantics given above could be reformatted into something like this, for example) is that the meaning of each term is defined in terms of the meanings of its subterms.
+\cite[\S 3.1]{MOSSES1990575}
 
@@
 This would require a measure structure on $\denotation A$ and $\denotation B$ to be defined, but in some cases it is not possible to define a measurable structure on the set of kernels between two measurable spaces such that the application map taking a kernel $A \rightsquigarrow B$ and an element of A is itself measurable, therefore the denotations of higher-order types could be undefined.
-This motivates the introduction of quasi-Borel spaces.
+This motivates the introduction of quasi-Borel spaces.\cite{8005137}
 
@@
 Having a denotational semantics available makes certain sorts of proofs more convenient.
-For example, if there's a program rearrangement that can be expressed as replacing $A[M,N]$ with $B[M,N]$ (for arbitrary $M, N$), then the proof of equivalence can be quite direct, showing that the way $\denotation M$ and $\denotation N$ are combined is equivalent (possibly subject to some conditions on certain variables not being free in $M$ and $N$, which implies that $\denotation M_\gamma$ and $\denotation N_\gamma$ are independent of the values of those variables in $\gamma$), whereas with operational semantics, it is much more complicated to keep track of the way the arbitrary subterms are duplicated, transformed and combined as execution progresses.
-Similarly, the notion of contextual equivalence is not necessary in denotational semantics (being trivially equivalent to equality of values).
+Suppose there's a program rearrangement that can be expressed as replacing $A[M,N]$ with $B[M,N]$ (for arbitrary $M, N$), then the proof of equivalence can be quite direct, showing that the ways $\denotation M$ and $\denotation N$ are combined by $A$ and $B$ are equivalent (possibly subject to some conditions on certain variables not being free in $M$ and $N$, which implies that $\denotation M_\gamma$ and $\denotation N_\gamma$ are independent of the values of those variables in $\gamma$), whereas with operational semantics, it is much more complicated to keep track of the way the arbitrary subterms are duplicated, transformed and combined as execution progresses.
+For example, in \cite{staton2017commutative}, the equivalence of $\mathsf{let}\ x = A\ \mathsf{in\ let}\ y = B\ \mathsf{in}\ C$ and $\mathsf{let}\ y=B\ \mathsf{in\ let}\ x=A\ \mathsf{in}\ C$ was proven using a denotational semantics (subject to $x$ and $y$ not being free in $A$ and $B$).
+Similarly, the notion of contextual equivalence is not generally necessary in denotational semantics, being trivially equivalent to equality of values (though there are some cases where it could still apply, if some class of terms do not count as valid complete programs).
 
-Another useful property of denotation semantics is that termination properties are much more directly visible.
+Another useful property of denotation semantics is that termination properties, to the extent that they exist, are much more directly visible.
 If the type system is strong enough to imply termination, this manifests directly as every term being given a value.
-If non-terminating terms are possible, it will be necessary to introduce, as part of the semantics, special cases to deal with that, otherwise the definition will simply not be consistent.
+If non-terminating terms are possible, it will be necessary to introduce, as part of the semantics, special cases to deal with that,\cite[\S 5.2]{MOSSES1990575} otherwise the definition will simply not be consistent.
 
@@
 There are a few variations on how the traces work.
-In some versions (as is the case in Section \ref{operational semantics}), each run of the program is assigned a trace of finite length, containing exactly those samples actually used in that run.
+In some versions (as is the case in Section \ref{operational semantics}, and \cite{borgstrom2016lambda}), each run of the program is assigned a trace of finite length, containing exactly those samples actually used in that run.
 If the set of results for an individual sampling operation is $S$ (in this case it's the set of 2 elements, but in a language with continuous distributions, it could be $\mathbb R$, or an interval, or something else), the set of traces is $\bigcup_{n \in \mathbb N} S^n$, and the measure defined on the set of traces is not actually a probability measure, as each set $S^n$ has measure 1.
@@
 This is in some ways a little simpler, because the set of traces can simply have a probability distribution, and the trace does not need to depend in any way on the program (by having its length matched).
-An additional variant of this style of trace is to split the trace in three parts, all infinitely long, for the application case of the big-step semantics, rather than having the each step (evaluating the function, evaluating the argument, then evaluating the result of the $\beta$ reduction) use however many samples it needs from the start of the trace in turn.
+An additional variant of this style of trace, used for example in \cite{culpepper17}, is to split the trace in three parts, all infinitely long, for the application case of the big-step semantics, rather than having the each step (evaluating the function, evaluating the argument, then evaluating the result of the $\beta$ reduction) use however many samples it needs from the start of the trace in turn.
 
@@
 
+The distinction between distribution and trace-based semantics is discussed in \cite{borgstrom2016lambda}, which gives both for the same language, and proves their equivalence (in the sense that the distributions derived from the trace-based semantics are the same as the distributions that the distribution-based semantics produces directly).
+
 \section{My Previous Research}
@@
 The second extension is antitone ranking functions.
-This idea is taken from \cite{some citation needed}, just adapted to the new functional context.
+This is closely related to the progress condition given in \cite{mciver2017new}, just adapted to the new functional context.
 A function $\epsilon : \mathbb R_{\geq 0} \to \mathbb R_{> 0}$, which is required to be monotoically decreasing, is added to the definition, and rather than the ranking function having to decrease by 1 at every $\tY$-reduction step, it only has to decrease by $\epsilon$ of the current value of the ranking function.
@@
 I intend to try to prove this result, even though it does not relate very closely to my overall thesis topic, because it would make my existing work more complete.
+
+\bibliography{references}
 \end{document}

### 2021-04-26 LICS21/SPCF.tex
Respond to reviews, "\defn" -> "\emph", "Church-Rosser" -> "confluent"

--- 2021-01-23 LICS21/SPCF.tex
@@
   E[\tY \lambda x. M] & \to E[\lambda z. M[(\tY \lambda x. M)/x] z] \\
-  & \qquad \qquad \text{ where $z$ is not free in $s$}\\
+  & \qquad \qquad \text{ where $z$ is not free in $M$}\\
   E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_1] \text{ where }r < 0 \\
@@
 \label{sec:sampling semantics}
-This version of the reduction relation allows $\tsample$ to reduce to any number in $[0,1]$. 
-To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples.
+The relation $\to$ allows $\tsample$ to reduce to any number in $[0,1]$. 
+This defines which results are possible, but to specify the exact probabilities, an additional argument is needed to determine the outcome of random samples.
 Let $ I := [0,1] \subset \mathbb{R} $, and define $\entrosp := I^{\mathbb{N}}$, with the Borel $\sigma$-algebra $\Sigma_\entrosp$.
@@
 The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid \hbox{$M$ terminates at $s$}\}$ is measurable \cite{DBLP:conf/icfp/BorgstromLGS16,MakOPW21}.
-We say that a term $M$ is \emph{almost surely terminating} (AST) just if $\mu_{\entrosp}(\{s \mid \hbox{$M$ terminates at $s$}\}) = 1$.
+We say that a term $M$ is \emph{almost surely terminating} (\emph{AST}) just if $\mu_{\entrosp}(\{s \mid \hbox{$M$ terminates at $s$}\}) = 1$.
 
@@
 \fi
-\begin{example} The term 
+\begin{example}[Geometric distribution] The term 
 \[
-(\tY \lambda f n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0},
+(\tY \lambda f n. \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0},
 \] 

### 2021-04-26 LICS21/antitone.tex
Respond to reviews, "\defn" -> "\emph", "Church-Rosser" -> "confluent"

@@
 \big(\tY \, \lambda f n . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{3/2} f \, (n + 1)}\big) \, 10
+\tif{n = 0}{0}{f \, (n - 1) \oplus_{2/3} f \, (n + 1)}\big) \, 10
 \]
@@
 \begin{definition}
-The definition of \defn{antitone ranking function} $f$ for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
+The definition of \emph{antitone ranking function} $f$ for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
 we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) function $\epsilon : \nnReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
@@
 where $R \to R'$ is the $\tY$-redex rule.
-Any closed term for which an antitone ranking function exists is called \defn{antitone rankable}.
+Any closed term for which an antitone ranking function exists is called \emph{antitone rankable}.
 \iffalse
@@
 Given a probability space $(\Omega, \calF, \mathbb{P})$, and a supermartingale $(Y_n)_{n \geq 0}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \geq 0}$,
-we say that $(Y_n)_{n \geq 0}$ is an \defn{antitone strict supermartingale w.r.t.~$T$} if for all $n \geq 0$, we have $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
+we say that $(Y_n)_{n \geq 0}$ is an \emph{antitone strict supermartingale w.r.t.~$T$} if for all $n \geq 0$, we have $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
 \(
@@
 
-It is essential in this theorem that $\epsilon$ be defined for all $\mathbb R_{\geq 0}$, not just the values that $(Y_n)_{n \geq 0}$ actually takes (or at least if $(Y_n)_{n \geq 0}$ is uniformly bounded, $\epsilon$ must be positive at the supremum as well as the actual values).
+It is essential in this theorem that $\epsilon$ be defined for all $\mathbb R_{\geq 0}$, not just the values that $(Y_n)_{n \geq 0}$ actually takes (or at least if $(Y_n)_{n \geq 0}$ is uniformly bounded, $\epsilon$ must be positive at the supremum as well as the realized values).
 
@@
 Take the probability space $(\Omega, \calF, \mathbb{P})$ where $\Omega$ is the closed interval of reals $[-1, 1]$, $\calF$ the Borel $\sigma$-algebra, and $\mathbb{P}$ the corresponding Lebesgue probability measure.
-Let $\omega \in [-1, 1]$. Define random variables $T$ and $(Y_k)_{k \in \omega}$:
+Let $\omega \in [-1, 1]$. Define random variables $T$ and $(Y_k)_{k \geq 0}$:
 \begin{align*}
@@
 \begin{definition}
-Define an \defn{antitone sparse ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$:
+Define an \emph{antitone sparse ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$:
 \begin{inparaenum}[(i)]

### 2021-04-26 LICS21/applications.tex
Respond to reviews, "\defn" -> "\emph", "Church-Rosser" -> "confluent"

@@
 \label{sec:applications}
-Although the definition of potential positions in a term in \Cref{sec:confluent} was intended simply to define a variant of trace semantics that has a restricted version of the Church-Rosser property, it is likely to also be applicable to inference algorithms (for probabilistic programs) as well.
+Although the definition of potential positions in a term in \Cref{sec:confluent} was intended simply to define a variant of trace semantics that has a restricted version of confluence, it is likely to also be applicable to inference algorithms (for probabilistic programs) as well.
 

### 2021-04-26 LICS21/confluent.tex
Respond to reviews, "\defn" -> "\emph", "Church-Rosser" -> "confluent"

@@
 thereby reducing the number of cases needed and simplifying the reachable terms.
-It will be shown in \Cref{thm:confluent ranking} that ranking functions with respect to alternative reduction strategies are sound for proving AST in the usual sense (i.e.~with respect to the standard CBV reduction strategy).
-In order to obtain this result, we prove a relation between the results of using different reduction strategies, 
-by first constructing a confluent trace semantics that is equivalent to the standard semantics as a special case.
 
-Non-probabilistic lambda calculi generally have the Church-Rosser property, that if a term $A$ reduces to both $B_1$ and $B_2$, there is some $C$ 
-%with reduction sequences $B_1 \to^* C$ and $B_2 \to^* C$, 
+In order to be able to use alternative reduction strategies to prove properties of the program defined using the usual reduction strategy, we first prove the relation between these different strategies, providing a confluent version of the semantics.
+It is therefore necessary to have some way of relating the samples that will be used in runs of the program using different reduction strategies.
+Positions are defined as a way of addressing sample statements within a program independently of the reduction order.
+Next, a version of the reduction relation is presented that is nondeterministic, so that it allows a choice of what order to perform reductions in.
+The notion of positions is extended to potential positions, for samples which may appear later in the reduction sequence but not necessarily in the initial term.
+In order to allow potential positions in different reduction sequences to be considered equivalent, a relation $\sim^*$ is defined, and finally, all of these are used to construct a confluent version of the trace semantics, $\Rightarrow$, that is still nondeterministic in reduction order, but does specify the outcome of random choices.
+The desired properties of this semantics (confluence and equivalence to the standard semantics) are proved (\Cref{churchRosser,thm:AstEquivalence}), then it is used to give extended versions of ranking functions and the related AST results, \Cref{thm:confluent ranking}.
+
+Non-probabilistic lambda calculi are generally confluent, i.e.~if a term $A$ reduces to both $B_1$ and $B_2$, there is some $C$ 
 to which both $B_1$ and $B_2$ reduce,
@@
 \medskip
-Even with this restriction, a trace semantics in the style of the one already defined would not be entirely Church-Rosser, as, for example, $\red^3(\tsample - \tsample, (1,0,\dots))$ would be either $1$ or $-1$ depending on the order of evaluation of the $\tsample$s, as that determines which sample from the pre-selected sequence is used for each one.
+Even with such a restriction, a trace semantics in the style of the one already defined would not be entirely confluent, as, for example, $\red^3(\tsample - \tsample, (1,0,\dots))$ would be either $1$ or $-1$ depending on the order of evaluation of the $\tsample$s, as that determines which sample from the pre-selected sequence is used for each one.
 To fix this, rather than pre-selecting samples according to the order they'll be drawn in, select them according to the position in the term where they'll be used instead.
 
-A \defn{position} is a finite sequence of steps into a term, defined inductively as
+A \emph{position} is a finite sequence of steps into a term, defined inductively as
 \begin{align*}
@@
 \end{align*}
-The \defn{subterm of $M$ at $\alpha$}, denoted $M \mid \alpha$, is defined as
+The \emph{subterm of $M$ at $\alpha$}, denoted $M \mid \alpha$, is defined as
 \begin{align*}
@@
 \lambda x. M \mid \lambda ; \alpha & = M \mid \alpha \\
-M_1 M_2 \mid @_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2 \\
+M_1 \, M_2 \mid @_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2 \\
 \underline f(M_1,\dots,M_n) \mid \underline f_i ; \alpha & = M_i \mid \alpha \quad \text{for }i \leq n \\
@@
 \end{align*}
-In each of these cases, $M \mid \alpha$ is the \defn{redex}, and the \defn{reduction takes place at $\alpha$}.
+In each of these cases, $M \mid \alpha$ is the \emph{redex}, and the reduction \emph{takes place at $\alpha$}.
+Each subterm can be a redex in at most one way, but there can be multiple redexes at different positions.
 \end{definition}
 
+The argument of a $\beta$ redex and the body of a $\tY$ redex may be duplicated by those reductions.
+It is therefore these cases that need to be handled carefully to avoid duplicating samples at the wrong time.
+In both cases, the potentially duplicated part must already be a value, which excludes terms like $\tsample$ or $\tsample + \underline 1$, which should be evaluated before being duplicated.
+In the other direction, if a $\tsample$ occurs inside of a $\lambda$, it may need to be duplicated before being evaluated, which is why a $\tsample$ reduction isn't allowed inside a $\lambda$ inside a $\tY$ or the right side of an application.
+These restrictions are in some cases unnecessarily strict, for example, in $(\lambda x.
+x) ((\lambda y.
+\tsample) \underline 0)$, it would be fine to evaluate the sample first, but they are at least sufficient to ensure confluence in terms of the distribution of results.
+Getting individual traces to behave correctly will take more work though.
+
 %\medskip
-Labelling the pre-chosen samples by the positions in the term would also not work because in some cases, a $\tsample$ will be duplicated before being reduced, for example, in $(\lambda x.
+Labelling the pre-chosen samples by the positions in the term by using $I^{\{\alpha \mid \, (M \mid \alpha) = \tsample\}}$ as the trace space would not be sufficient to solve the issue of different samples being used in corresponding locations in different reduction sequences because in some cases, a $\tsample$ will be duplicated before being reduced, for example, in $(\lambda x.
 x \, {\underline 0} \, \mathbin{\underline{+}} \, {x \, \underline 0}) (\lambda y.
@@
 Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a $\to$-reduction sequence of skeletons starting at $\mathit{Sk}(M)$, and the second of which is a position in the final skeleton of the reduction sequence.
-As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the standard trace semantics, modified traces, which are elements of $I^{L_0(M)}$ (with one more caveat introduced later), will be used to pre-select a sample from $I$ for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.
+As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the standard trace semantics, modified traces, which are elements of $I^{L_0(M)}$ (with one more caveat introduced after Def.
+\ref{defn:sim}), will be used to pre-select a sample from $I$ for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.
 
@@
 \begin{align*}
-A[x] &= \tif{\tif{x>0}{I}{I} (\lambda y.\tsample)\, \underline 0 - \underline{0.5} > 0}{\underline 0}{\Omega}\\
+A[M] &= \tif{\tif{M>0}{I}{I} (\lambda y.\tsample)\, \underline 0 - \underline{0.5} > 0}{\underline 0}{\Omega}\\
 B &= \tif{\tsample - \underline{0.5} > 0}{\underline 0}{\Omega}
@@
 %\medskip
-This is still not quite sufficient because sometimes the same samples must be used at corresponding positions in different reduction sequences.
+This is still not quite sufficient to attain confluence because sometimes the same samples must be used at corresponding positions in different reduction sequences.
 
@@
 In the reduction $M \to N_1$, the sample labelled $(M, \underline{+}_1)$ is used, and in the reduction $N_2 \to O$, the sample labelled $(M \to N_2, \underline{+}_1)$ is used. 
-Each of these samples becomes the value of the first numeral in $O$ in their respective reduction sequences, therefore in order for Church-Rosserness to be attained, they must be the same.
+Each of these samples becomes the value of the first numeral in $O$ in their respective reduction sequences, therefore in order for confluence to be attained, they must be the same.
 Which elements of $L_0(M)$ must match can be described by the relation $\sim^*$:
@@
 
-\begin{definition}
+\begin{definition} \label{defn:sim}
 The relation $\sim$ is defined as the union of the minimal symmetric relations $\sim_p$ (``$p$'' for parent-child) and $\sim_c$ (``$c$'' for cousin) satisfying
@@
 The reflexive transitive closure $\sim^*$ of this relation is used to define the set of \emph{potential positions} $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable skeletons.
-If $(N,\alpha) \sim^* (O,\beta)$, then $N \mid \alpha$ and $O \mid \beta$ both have the same shape (i.e.~they're either both the placeholder $\skeletonPlaceholder$, both variables, both applications, both $\tsample$s etc.), therefore it's well-defined to talk of the set of \defn{potential positions where there is a $\tsample$}, $L_s(M)$.
-Formally 
-\[
-L_s(M) := \{[(X, \alpha)]_{\sim^\ast_M} : X \mid \alpha = \tsample\}.
-\]
+If $(N,\alpha) \sim^* (O,\beta)$, then $N \mid \alpha$ and $O \mid \beta$ both have the same shape (i.e.~they're either both the placeholder $\skeletonPlaceholder$, both variables, both applications, both $\tsample$s etc.), therefore it's well-defined to talk of the set of \emph{potential positions where there is a $\tsample$}, $L_s(M)$.
 The new sample space is then defined as $I^{L_s(M)}$, with the Borel $\sigma$-algebra and product measure.
@@
 \begin{restatable}{lemma}{churchRosser} \label{churchRosser}
-The relation $\Rightarrow$ is Church-Rosser.
+The relation $\Rightarrow$ is confluent.
 \end{restatable}
@@
 %\begin{definition}
-A \defn{reduction strategy} starting from a closed term $M$ is a measurable partial function $f$ from $\mathit{Rch}(M)$ to positions, such that for any reachable term $N$ where $f$ is defined, $f(N)$ is a position of a redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
+A \emph{reduction strategy} starting from a closed term $M$ is a measurable partial function $f$ from $\mathit{Rch}(M)$ to positions, such that for any reachable term $N$ where $f$ is defined, $f(N)$ is a position of a redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
 Using a reduction strategy $f$, a subset of $\Rightarrow$ that isn't nondeterministic, $\Rightarrow_f$, can be defined by $(N,s) \Rightarrow_f (N',s')$ just if $(N,s) \Rightarrow (N',s')$ and $N$ reduces to $N'$ with the redex at $f(N)$.
@@
 \begin{definition}
-An \defn{antitone ranking function on $M$ w.r.t.~a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ such that
+An \emph{antitone ranking function on $M$ w.r.t.~a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ such that
 \begin{itemize}
@@
 \end{itemize}
-Any closed term for which an antitone ranking function with respect to a reduction strategy $r$ exists is called \defn{antitone rankable w.r.t.~$r$}. 
+Any closed term for which an antitone ranking function with respect to a reduction strategy $r$ exists is called \emph{antitone rankable w.r.t.~$r$}. 
 \end{definition}
@@
 \begin{definition}
-An \defn{antitone sparse ranking function on $M$ w.r.t.~a reduction strategy $r$} is a partial function $f : \mathit{Rch}_r(M) \rightharpoonup \mathbb R$ such that there exists some antitone function $\epsilon : \nnReal \to \pReal$ such that
+An \emph{antitone sparse ranking function on $M$ w.r.t.~a reduction strategy $r$} is a partial function $f : \mathit{Rch}_r(M) \rightharpoonup \mathbb R$ such that there exists some antitone function $\epsilon : \nnReal \to \pReal$ such that
 \begin{inparaenum}[(i)]
@@
 
-If the antitone function is just the constant function $1$ instead, then it is called a \defn{(sparse) ranking function on $M$ w.r.t.~$r$} instead, and the term is called \defn{rankable w.r.t.~$r$}.
+If the antitone function is just the constant function $1$ instead, then it is called a \emph{(sparse) ranking function on $M$ w.r.t.~$r$} instead, and the term is called \emph{rankable w.r.t.~$r$}.
 

### 2021-04-26 LICS21/intro.tex
Tweaks
Respond to reviews, "\defn" -> "\emph", "Church-Rosser" -> "confluent"

@@
 
-Basic (non-random) lambda calculus has the very useful Church-Rosser property, which implies (among other things) that even if execution of a program starts in a different order, it will still reach the same normal form eventually (assuming it does reach a normal form).
+Basic (non-random) lambda calculus has the very useful property of confluence, which implies (among other things) that even if execution of a program starts in a different order, it will still reach the same normal form eventually (assuming it does reach a normal form).
 Probabilistic lambda calculus does not have this property, because random choices may be duplicated, and evaluating the same subterm multiple times can yield different results. 
-However, with a restricted set of reduction strategies, Church-Rosserness may be regained. 
+However, with a restricted set of reduction strategies, confluence may be regained. 
 We introduce a novel addressing scheme for the possible random choices in a program's execution, which ensures that the same random choices are taken at corresponding positions in alternative reduction sequences, so that the same eventual result can be reached.
@@
 
-By combining these methods we can prove AST of a variety of PPCF programs that are beyond the reach of methods in the literature, including non-affine recursion\footnote{\label{fnote:non-affine} Non-affine recursive programs are recursive programs that can, during the evaluation of the recursive body, make multiple recursive calls (of a first-order function) from distinct call sites.} (\cref{ex:non-affine recursion easy,ex:non-affine continuous}) and recursions that define high-order functions (\cref{ex:higher-order recursion}).
+By combining these methods we can prove AST of a variety of PPCF programs that are beyond the reach of methods in the literature, including non-affine recursion\footnote{\label{fnote:non-affine} Non-affine recursive programs are recursive programs that can, during the evaluation of the recursive body, make multiple recursive calls (of a first-order function) from distinct call sites.} (\cref{ex:non-affine recursion easy,ex:non-affine continuous}) and recursions that define higher-order functions (\cref{ex:higher-order recursion}).
 
@@
 
-\paragraph*{Additional materials} Further details of some of the examples and all missing proofs can be found in the appendices.
+\paragraph*{Additional materials} Further details of some of the examples and all missing proofs can be found in the appendix.

### 2021-04-26 LICS21/ranking.tex
Respond to reviews, "\defn" -> "\emph", "Church-Rosser" -> "confluent"

--- 2021-01-16 LICS21/ranking.tex
@@
 \] 
-terminates with only 2 $\tY$-reductions on average, i.e.,~$\expect{T_M^\tY} = 2$, but applies the increment function $2^n$ times with probability $2^{-n-1}$ for $n \geq 0$, which diverges, i.e.,~$\expect{T_M} = \infty$.
+is $\tY$-PAST and rankable, terminating with only 2 $\tY$-reductions on average, i.e.,~$\expect{T_M^\tY} = 2$, but applies the increment function $2^n$ times with probability $2^{-n-1}$ for $n \geq 0$, which diverges, i.e.,~$\expect{T_M} = \infty$.
 \end{example}
@@
 %\lo{TODO: reduce the figure to avoid overflow.}\akr{It still overflows the gutter, but not the actual text of the other column.}\lo{OK. Thanks.}
-The meanings of the substitutions (to be applied to the contractum) and conditions that label some of the reductions in the diagram should be self-explanatory.
 
@@
 The definition of rankability is also inconvenient for syntactic sugar.
-It could be useful, for example, to define $M \oplus_p N := \tif{\tsample - p < 0} M N$, where $M \oplus_p N$ reduces to $M$ or $N$, depending on the first value of $s \in \entrosp$, with probability $p$ resp.~$(1-p)$.
+It could be useful, for example, to define \mbox{$M \oplus_p N := \tif{\tsample - p < 0} M N$}, where $M \oplus_p N$ reduces to $M$ or $N$, depending on the first value of $s \in \entrosp$, with probability $p$ resp.~$(1-p)$.
 Technically though, it reduces first to $\tif{\underline r - p < 0} M N$ for all $r \in I$, so those terms all need values of the ranking function too.

### 2021-04-26 LICS21/related.tex
Respond to reviews, "\defn" -> "\emph", "Church-Rosser" -> "confluent"

@@
 % (because the underlying recursion schemes are not Turing complete):
-the PPCF terms in \cref{ex:raven complex,ex:non-affine continuous,ex:higher-order recursion}
+for example, the PPCF terms in \cref{ex:raven complex,ex:non-affine continuous,ex:higher-order recursion}
 %\refExample{complexExample} 
@@
 
-The \Cref{def:more general red} of redexes and which positions are acceptable to reduce at is sufficiently restrictive to guarantee Church-Rosserness, but is also a little more restrictive in some cases than is necessary for this purpose.
+The \Cref{def:more general red} of redexes and which positions are acceptable to reduce at is sufficiently restrictive to guarantee confluence, but is also a little more restrictive in some cases than is necessary for this purpose.
 For example, if the argument of a function is not yet a value, but its reduction to a value would be deterministic, or the function is affine, then applying the function before reducing its argument would not cause any problematic duplication of random samples.
 Similarly, if $\tsample$ occurs at position $\alpha;@_2;\beta;\lambda;\gamma$, but the function at $\alpha;@_1$ is affine, evaluating the $\tsample$ may in some cases also work fine.
-A more complete characterisation of which redexes can be reduced without breaking Church-Rosserness could be interesting.
+A more complete characterisation of which redexes can be reduced without breaking confluence could be interesting.
 
@@
 
-We have presented the first application of martingales to probabilistic lambda calculus, and the first version of trace semantics that's capable of satisfying the (restricted) Church-Rosser property.
+We have presented the first application of martingales to probabilistic lambda calculus, and the first version of trace semantics that's capable of satisfying a restricted form of confluence.
 Though the construction of ranking functions is inherently difficult, using sparse ranking functions, antitone ranking functions and ranking functions w.r.t.~an alternative reduction strategy, we have shown that a great variety of programs can be proved to be AST, including some that are beyond the reach of methods in the literature.

### 2021-04-26 LICS21/supermartingales-proofs.tex
Respond to reviews, "\defn" -> "\emph", "Church-Rosser" -> "confluent"

@@
 
-Given an $\omega$-sequence (e.g.~$s \in \entrosp$) and $m \geq 0$, we write $s_{\leq m} \in I^m$ to mean the prefix of $s$ of length $m$.
+Given an $\mathbb N$-indexed sequence (e.g.~$s \in \entrosp$) and $m \geq 0$, we write $s_{\leq m} \in I^m$ to mean the prefix of $s$ of length $m$.
 For $n \geq 0$, define
@@
 For any $s'\in \entrosp$, if $s_{\leq l} = s_{\leq l}'$ then $s' \in A$.
-It follows that ${\set{s_{\leq l}} \cdot I^\omega} \subseteq A$.
+It follows that ${\set{s_{\leq l}} \cdot I^{\mathbb N}} \subseteq A$.
 \lo{NOTE. To fix notation:
@@
 
-Observe that if $s \in U_i$ then $\set{s_{\leq l_i}} \cdot I^\omega \subseteq U_i$;
-in fact $(U_i)_{\leq l_i} \cdot I^\omega = U_i$.
-%Moreover, for any $A \in \calF_n$, if $s \in A$ then $\set{s_{\leq l}} \cdot I^\omega \subseteq A$.
+Observe that if $s \in U_i$ then $\set{s_{\leq l_i}} \cdot I^{\mathbb N} \subseteq U_i$;
+in fact $(U_i)_{\leq l_i} \cdot I^{\mathbb N} = U_i$.
+%Moreover, for any $A \in \calF_n$, if $s \in A$ then $\set{s_{\leq l}} \cdot I^{\mathbb N} \subseteq A$.
 This means that for any measurable $g : \entrosp \to \Real_{\geq 0}$, if $g(s)$ only depends on the prefix of $s$ of length $(l_i+1)$, then, writing $\widehat{g} : I^{l_i+1} \to \Real_{\geq 0}$ where $g(s) = \widehat{g}(s_{\leq l_i+1})$, we have: for any $A \in \calF_n$
@@
 \end{align*}
-Since $\Leb_{l+1}$ is the (unique) product measure satisfying $\Leb_{l+1}(V \times B) = \Leb_l(V) \cdot \Leb(B)$, and $(U_i)_{\leq l_i} \cdot I^\omega = U_i$, we have
+Since $\Leb_{l+1}$ is the (unique) product measure satisfying $\Leb_{l+1}(V \times B) = \Leb_l(V) \cdot \Leb(B)$, and $(U_i)_{\leq l_i} \cdot I^{\mathbb N} = U_i$, we have
 %it follows from the preceding observation that
@@
 
-We first show that $T_1$ is bounded i.e.~$T_1 \leq n$, for some $n \in \omega$.
+We first show that $T_1$ is bounded i.e.~$T_1 \leq n$, for some $n \in \mathbb N$.
 
@@
 %For each $n \in \omega$, $\set{T_M^\tY = n} \in \calF_{T_n}$.
-To see $\set{T_M^\tY = n} \in \calF_{T_n}$ for a given $n$, we need to show that $\set{T_M^\tY = n} \cap \set{T_n \leq i} \in \calF_i$, for all $i \in \omega$.
+To see $\set{T_M^\tY = n} \in \calF_{T_n}$ for a given $n$, we need to show that $\set{T_M^\tY = n} \cap \set{T_n \leq i} \in \calF_i$, for all $i \in \mathbb N$.
 Let $\cal V$ be the set of values in $\Lambda^0$ (which is measurable), and let $\cal W$ be the non-values.

### 2021-04-26 LICS21/supermartingales.tex
Respond to reviews, "\defn" -> "\emph", "Church-Rosser" -> "confluent"

@@
 \label{def:ranking function}
-A \defn{ranking function on $M \in \Lambda^0$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$\footnote{The fact that the ranking function can be 0 in some cases before it reaches a value is necessary to get \Cref{thm:minimal} to work neatly.}, and
+A \emph{ranking function on $M \in \Lambda^0$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$,\footnote{The fact that the ranking function can be 0 in some cases before it reaches a value is necessary to get \Cref{thm:minimal} to work neatly.} and
 \begin{enumerate}
@@
 \fi
-We say that the ranking function $f$ is \defn{strict} if there exists $\epsilon > 0$ such that for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.
+We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.
 
-Any closed term for which a ranking (respectively, strict ranking) function exists is called \defn{rankable} (respectively, \defn{strictly rankable}). 
+Any closed term for which a ranking (respectively, strict ranking) function exists is called \emph{rankable} (respectively, \emph{strictly rankable}). 
 For example, $(\tY \lambda x.x) \, \underline 0$ is not rankable.
@@
 Let $T$ be a r.v.~that takes values in $\mathbb{N} \cup \set{\infty}$.
-We call $T$ a \defn{stopping time} \emph{adapted to $(\calF_n)_{n \geq 0}$} just if $\set{T = n} \in \calF_n$, for all $n$.
+We call $T$ a \emph{stopping time adapted to $(\calF_n)_{n \geq 0}$} just if $\set{T = n} \in \calF_n$, for all $n$.
 
@@
 \begin{asparaenum}[(i)]
-\item A sequence of r.v.s $(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \defn{supermartingale} if for all $n \geq 0$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~(i.e.~for all $A \in \calF_n$, $\int_A \mathbb{P}(\dif \omega) \, Y_{n+1}(\omega) \leq \int_A \mathbb{P}(\dif \omega) \,Y_n(\omega)$).
+\item A sequence of r.v.s $(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \emph{supermartingale} if for all $n \geq 0$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~(i.e.~for all $A \in \calF_n$, $\int_A Y_{n+1}(\omega) \, \mathbb{P}(\dif \omega) \leq \int_A Y_n(\omega) \, \mathbb{P}(\dif \omega)$).
 \item Let $\epsilon > 0$. 
 Given a stopping time $T$ and a supermartingale $(Y_n)_{n \geq 0}$, both adapted to filtration $(\calF_n)_{n \geq 0}$, 
-we say that $(Y_n)_{n \geq 0}$ is a $\epsilon$-\defn{ranking supermartingale w.r.t.~$T$} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.
-\footnote{For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable defined by ${\bf 1}_A(\omega) := 1$ if $\omega \in A$, and $0$ otherwise.} %\lo{\LaTeX\ does not allow the preceding bracketed sentence to be typeset as a footnote.
+we say that $(Y_n)_{n \geq 0}$ is a $\epsilon$-\emph{ranking supermartingale w.r.t.~$T$} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.\footnote{For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable defined by ${\bf 1}_A(\omega) := 1$ if $\omega \in A$, and $0$ otherwise.} %\lo{\LaTeX\ does not allow the preceding bracketed sentence to be typeset as a footnote.
 Strange!}\lo{Now it does.
 Stranger.}
-A \defn{ranking supermartingale w.r.t.~$T$} is a $\epsilon$-ranking supermartingale w.r.t.~$T$ for some $\epsilon > 0$.
+A \emph{ranking supermartingale w.r.t.~$T$} is a $\epsilon$-ranking supermartingale w.r.t.~$T$ for some $\epsilon > 0$.
 \end{asparaenum}
@@
 and the filtration $(\mathcal{F}_n)_{n \geq 0}$ where $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
-Thus $T_M$ is the \defn{runtime} of $M$ (and $M$ is AST iff $\mu_{\entrosp}(T_M < \infty) = 1$); we say that $M$ is \defn{positively almost surely terminating} (\defn{PAST}) if $\expect{T_M} < \infty$.
+Thus $T_M$ is the \emph{runtime} of $M$ (and $M$ is AST iff $\mu_{\entrosp}(T_M < \infty) = 1$); we say that $M$ is \emph{positively almost surely terminating} (\emph{PAST}) if $\expect{T_M} < \infty$.
 
@@
 & \qquad\qquad\qquad\qquad\qquad \hbox{or of form } E[\tY \lambda x. N] \} \nonumber \\
-Y_n(s) & := f(M_{T_n}(s)) \nonumber \\
+Y_n(s) & := f(M_{T_n(s)}(s)) \nonumber \\
 T_M^\tY(s) &:= \min \set{n \mid M_{T_n(s)}(s) \textrm{ is a value}} \label{eq:y stopping time} 
@@
 
-The random variable $T_M^\tY$, which we call the $\tY$-\defn{runtime} of $M$, can equivalently be defined as the number of $\tY$-reduction steps in the reduction sequence of $M$. 
+The random variable $T_M^\tY$, which we call the $\tY$-\emph{runtime} of $M$, can equivalently be defined as the number of $\tY$-reduction steps in the reduction sequence of $M$. 
 Note that, as the type system ensures that the reduction relation excluding $\tY$-reduction is strongly normalising, only finitely many reductions can occur in a row without one of them being a $\tY$-reduction, therefore $T_M^\tY < \infty$ a.s.~iff $M$ is AST.
-We say that $M$ is $\tY$-\defn{PAST} if $\expect{T^{\tY}_M} < \infty$.
+We say that $M$ is $\tY$-\emph{PAST} if $\expect{T^{\tY}_M} < \infty$.
 

### 2021-04-26 thesis proposal/thesis proposal.tex

@@
 
+\iffalse
+\bibliographystyle{apalike}
+\bibliography{../Bib/references-Beutner,../Bib/references}
+\fi
+
 \bibliography{references}

### 2021-04-26 LICS21/SPCF.tex
Edit to focus on confluence result for IWC.
Remove everything except the sections on the confluent trace semantics
and syntax. Add a new introduction. Remove references to other parts
of the paper. Generally edit so that it makes sense on its own.

--- 2021-01-25 LICS21/SPCF.tex
@@
 % !TEX root = main.tex
-\section{Probabilistic PCF}
+\section{Syntax of Probabilistic PCF}
 \label{sec:PPCF}
 
-\subsection{Syntax of PPCF}
-
-The language PPCF is a call-by-value (CBV) version of PCF with sampling of real numbers from the closed interval $[0,1]$ \cite{Ehrhard2018c,DBLP:journals/pacmpl/EhrhardPT18,MakOPW21}.
+The language PPCF is a call-by-value version of PCF with sampling of real numbers from the closed interval $[0,1]$ \cite{Ehrhard2018c,DBLP:journals/pacmpl/EhrhardPT18,MakOPW21}.
 %\changed[lo]{following \cite{MakOPW21} except the score function (\Cref{rem:score})}. 
@@
 The typing rules are standard (see \Cref{fig:typing rules}).
+The restriction to well-typed terms is only necessary here in order to avoid reaching terms which contain nonsense such as applying a number as though it were a function, so a more liberal type system would work just as well.
+Simple types are just used for simplicity.
 Terms are identified up to $\alpha$-equivalence, as usual. 
@@
 
-%\paragraph{}
-To define the CBV reduction relation, let \emph{evaluation contexts} be of the form:
+\begin{figure*}[htb]
 \begin{align*}
-  E ::= & \, [\,] \mid E \, M \mid V \, E \mid \underline{f}(\underline r_1,\dots ,\underline r_{k-1}, E, M_{k+1}, \dots, M_n) \\ & \mid \tY E \mid \tif{E<0}{M_1}{M_2}
+  \frac{}{\Gamma ; x:A \vdash x:A} 
+  \qquad
+  \frac{\Gamma ; x:A \vdash M : B}{\Gamma \vdash \lambda x.M : A \to B} 
+  \qquad
+  \frac{\Gamma \vdash M:A \to B \quad \Gamma \vdash N : A}{\Gamma \vdash M \, N : B} 
+  \\ \\
+  \qquad
+  \frac{\Gamma \vdash M : (A \to B) \to (A \to B)}{\Gamma \vdash \tY M : (A \to B)}
+  \qquad
+  \frac{\Gamma \vdash M : \textsf{R} \quad \Gamma \vdash N_1 : A \quad \Gamma \vdash N_2 : A}{\Gamma \vdash \tif{M < 0}{N_1}{N_2} : A}    
+  \\ \\
+  \qquad
+  \frac{}{\underline{r} : \textsf{R}} 
+  \qquad
+  \frac{}{\Gamma \vdash \tsample : \textsf{R}}
+  \qquad
+  \frac{\Gamma \vdash M_1:\textsf{R} \quad \dots \quad \Gamma \vdash M_n:\textsf{R}}{\Gamma \vdash \underline{f}(M_1,\dots,M_n) : \textsf{R}} \ (f : \mathbb{R}^n \to \mathbb{R})
 \end{align*}
-then a term reduces if it is formed by substituting a redex in a context i.e.
-\begin{align*}
-  E[(\lambda x.M) V] & \to E[M[V/x]] \\
-  E[\underline f (\underline r_1, \dots , \underline r_n)] & \to E[\underline{f(r_1,\dots,r_n)}] \\
-  E[\tY \lambda x. M] & \to E[\lambda z. M[(\tY \lambda x. M)/x] z] \\
-  & \qquad \qquad \text{ where $z$ is not free in $s$}\\
-  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_1] \text{ where }r < 0 \\
-  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_2] \text{ where }r \geq 0 \\
-  E[\tsample] & \to E[\underline r] \text{ where } r \in [0,1].
-\end{align*}
-We write $\to^\ast$ for the reflexive, transitive closure of $\to$.
-Every closed term either is a value or reduces to another term.
-
-\subsection{Trace semantics}
-\label{sec:sampling semantics}
-This version of the reduction relation allows $\tsample$ to reduce to any number in $[0,1]$. 
-To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples.
-Let $ I := [0,1] \subset \mathbb{R} $, and define $\entrosp := I^{\mathbb{N}}$, with the Borel $\sigma$-algebra $\Sigma_\entrosp$.
-Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and the probability measure $\mu_{\entrosp}$ is given by $\mu_{\entrosp} (\prod_{i=0}^\infty X_i) := \prod_{i=0}^\infty \Leb(X_i)$, \changed[lo]{writing $\Leb$ for the Lebesgue measure.}
-\changed[lo]{The maps $\pi_h:\entrosp \to I$ (projecting to the first element) and $\pi_t:\entrosp \to \entrosp$ (popping the first element) are then measurable.}
-Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$ the \emph{entropy space}, and elements of $\entrosp$ \defn{traces}.
-
-Define a \defn{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $\mathit{Sk}(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$.
-Following \cite{DBLP:conf/icfp/BorgstromLGS16,DBLP:conf/lics/StatonYWHK16,DBLP:journals/pacmpl/EhrhardPT18}, the $\sigma$-algebra on $\Lambda$ is defined by identifying $\Lambda$ with $\bigcup_{m\geq 0} \big(\mathsf{Sk}_m \times \Real^m \big)$, where $\mathsf{Sk}_m$ is the set of skeletons containing $m$ occurrences of $\skeletonPlaceholder$.
-Thus identified, we give $\Lambda$ the countable disjoint union topology of the product topology of the discrete topology on $\mathsf{Sk}_m$ and the standard topology on $\Real^m$, and take the corresponding Borel $\sigma$-algebra.
-Note that the connected components of $\Lambda$ have the form $\{S\} \times \Real^m$, with $S$ ranging over $\mathsf{Sk}_m$, and $m$ over $\mathbb{N}$.
-
-\changed[lo]{The one-step reduction of the \emph{trace-based} (or \emph{sampling-style}) \emph{operational semantics} \cite{DBLP:journals/jcss/Kozen81,DBLP:conf/icfp/BorgstromLGS16,DBLP:conf/esop/CulpepperC17}} is given by the function $\red : \Lambda^0 \times \entrosp \to \Lambda^0 \times \entrosp$ where
-\begin{equation*}
-\red(M,s) := \left\{
-    \begin{array}{ll}
-        (E[N],s) & \text{if } M = E[R], R \to N\\
-        & \qquad \text{ and } R \neq \tsample \\
-        (E[\underline{\pi_h(s)}],\pi_t(s)) & \text{if } M = E[\tsample] \\
-        (M,s) & \text{if } M \text{ is a value}
-    \end{array} \right .
-\end{equation*}
-%\lo{Both $\to$ and $\red$ are called reduction relation.}
-
-The result after $n$ steps is then simply 
-\[
-\red^n(M,s) = \underbrace{\red(\ldots\red(}_n M,s)\ldots)
-\] 
-and the limit $\red^\infty$ can then be defined as a partial function as $\lim_{n \to \infty} \red^n(M,s)$ whenever that sequence becomes constant by reaching a value.
-A term $M$ terminates for a sample sequence $s$ if the limit $\red^\infty(M,s)$ is defined.
-
-The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid \hbox{$M$ terminates at $s$}\}$ is measurable \cite{DBLP:conf/icfp/BorgstromLGS16,MakOPW21}.
-We say that a term $M$ is \defn{almost surely terminating} (\defn{AST}) just if $\mu_{\entrosp}(\{s \mid \hbox{$M$ terminates at $s$}\}) = 1$.
-
-\iffalse
-\lo{Alternatively, define the \emph{runtime of $M$} to be the random variable 
-\[
-T_M(s) := 
-\begin{cases}
-\min \set{n \mid \pi_0(\red^n(M, s)) \textrm{ is a value}} & \hbox{if $\red^\infty(M,s)$ is defined}\\
-\infty & \hbox{otherwise}
-\end{cases}
-\]
-Equivalently, we say that $M$ is \emph{almost surely terminating} (AST) if $T_M < \infty$ a.s.; 
-and $M$ is \emph{positively almost surely terminating} (PAST) if $\expect{T_M} < \infty$.}
-\fi
-\begin{example}[Geometric distribution] The term 
-\[
-(\tY \lambda f n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0},
-\] 
-which generates a geometric distribution, terminates on the set $\entrosp \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely; whereas 
-\(
-\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \, \underline 0}
-\), 
-which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.
-\end{example}
-%\paragraph{}
-%The requirement that $M$ be closed is not important for the proofs of the theorems, but is simply needed to ensure that $M$ reaches a value, rather than a general normal form like $x \, y$.
-\lo{This can be omitted.}
-
-%\paragraph{}
-\begin{remark}
-\label{rem:score}
-Our definition of AST is equivalent to that given in \cite{MakOPW21} (although the program semantics is stated in a slightly different way), except for the presence of a $\tscore$ construct for soft conditioning.
-The $\tscore$ construct is irrelevant to termination except that it fails if its argument is negative, thus allowing computations to fail after finitely many steps. 
-If such a thing were added, it would merely require an additional side condition to our main theorems, that no failing term was reachable (or, in the case of the confluent semantics based results, no term where the reduction strategy is not defined is reachable via the reduction strategy unless it's a value).
-\end{remark}
-
-\iffalse
-\lo{Your operational semantics does not maintain a record of the current weight of the reduction.
-A.s.~termination does depend on $\tscore$: see \cite[\S 4.3]{MakOPW21}\footnote{\url{https://arxiv.org/abs/2004.03924}}.
-I think it important to take the behaviour of $\tscore$ into account;
-you should do it as a future task.}
-\fi
+\caption{Typing rules of PPCF \label{fig:typing rules}}
+\end{figure*}

### 2021-04-26 LICS21/confluent.tex
Edit to focus on confluence result for IWC.
Remove everything except the sections on the confluent trace semantics
and syntax. Add a new introduction. Remove references to other parts
of the paper. Generally edit so that it makes sense on its own.

@@
 % !TEX root = main.tex
-\section{Confluent trace semantics}
-\label{sec:confluent}
-
-When proving almost sure termination in this way, it is necessary to consider all of the terms a given term may reduce to, and organise them into a manageable number of cases to assign ranking function values.
-Sometimes however, the reduction that the programmer has in mind may not be strictly the call-by-value order defined so far; 
-or the number of cases necessary for defining a sparse ranking function is excessive because the reduction does not proceed in a neat and orderly manner. 
-In cases such as these (e.g.~\cref{ex:sum of powers}), it may be much more convenient to instead consider ranking functions with respect to alternative reduction strategies constructed to match the individual term in question,
-thereby reducing the number of cases needed and simplifying the reachable terms.
-
-In order to be able to use alternative reduction strategies to prove properties of the program defined using the usual reduction strategy, we first prove the relation between these different strategies, providing a confluent version of the semantics.
-It is therefore necessary to have some way of relating the samples that will be used in runs of the program using different reduction strategies.
-Positions are defined as a way of addressing sample statements within a program independently of the reduction order.
-Next, a version of the reduction relation is presented that is nondeterministic, so that it allows a choice of what order to perform reductions in.
-The notion of positions is extended to potential positions, for samples which may appear later in the reduction sequence but not necessarily in the initial term.
-In order to allow potential positions in different reduction sequences to be considered equivalent, a relation $\sim^*$ is defined, and finally, all of these are used to construct a confluent version of the trace semantics, $\Rightarrow$, that is still nondeterministic in reduction order, but does specify the outcome of random choices.
-The desired properties of this semantics (confluence and equivalence to the standard semantics) are proved (\Cref{churchRosser,thm:AstEquivalence}), then it is used to give extended versions of ranking functions and the related AST results, \Cref{thm:confluent ranking}.
-
-Non-probabilistic lambda calculi are generally confluent, i.e.~if a term $A$ reduces to both $B_1$ and $B_2$, there is some $C$ 
-to which both $B_1$ and $B_2$ reduce,
-so the reduction order mostly doesn't matter. 
-In the probabilistic case, this may not be true, because $\beta$-reduction can duplicate $\tsample$s, so the outputs of the copies of the sample may be identical or independent, depending on whether the sample is taken before or after $\beta$-reduction.
-There are, however, some restricted variations on the reduction order that do not have this problem.
-
-\medskip
-Even with such a restriction, a trace semantics in the style of the one already defined would not be entirely confluent, as, for example, $\red^3(\tsample - \tsample, (1,0,\dots))$ would be either $1$ or $-1$ depending on the order of evaluation of the $\tsample$s, as that determines which sample from the pre-selected sequence is used for each one.
-To fix this, rather than pre-selecting samples according to the order they'll be drawn in, select them according to the position in the term where they'll be used instead.
+\section{Positions}
 
@@
 
-Thus we can define a more general reduction relation\footnote{In this section $\to$ always means the relation defined in \cref{def:more general red}.} $\to$.
+Thus we can define a nondeterministic reduction relation $\to$.
 \begin{definition}
@@
 
-%\medskip
+\section{Skeletal Reduction Sequences}
 Labelling the pre-chosen samples by the positions in the term by using $I^{\{\alpha \mid \, (M \mid \alpha) = \tsample\}}$ as the trace space would not be sufficient to solve the issue of different samples being used in corresponding locations in different reduction sequences because in some cases, a $\tsample$ will be duplicated before being reduced, for example, in $(\lambda x.
@@
 
-\paragraph*{Reduction relation on skeletons} We are thus led to consider the reduction relation on skeletons (and positions in a skeleton), which can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, $\tsample$ reducing to $\skeletonPlaceholder$, and $\skeletonPlaceholder$ considered as (the skeletal equivalent of) a value, so that $(\lambda x.A) \, \skeletonPlaceholder$ reduces to $A[\skeletonPlaceholder / x]$.
+We are thus led to consider the reduction relation on skeletons.
+Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $\mathit{Sk}(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$.
+Positions in a skeleton and the reduction relation $\to$ on skeletons can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, $\tsample$ reducing to $\skeletonPlaceholder$, and $\skeletonPlaceholder$ considered as (the skeletal equivalent of) a value, so that $(\lambda x.A) \, \skeletonPlaceholder$ reduces to $A[\skeletonPlaceholder / x]$.
 For example, we have 
@@
 
-Reduction sequences are used rather than reachable skeletons because if the same skeleton is reached twice, different samples may be needed.
+Reduction sequences are used rather than reachable skeletons because if the same skeleton is reached twice, different samples may be needed:
 
@@
 
-%\medskip
+\section{Potential Positions}
 This is still not quite sufficient to attain confluence because sometimes the same samples must be used at corresponding positions in different reduction sequences.
@@
 \end{asparaenum}
-The $\sim_c$-rules are illustrated in \Cref{sim-diagram}.
 \end{definition}
@@
 
-\medskip
-Before defining the new version of the reduction relation $\red$, the following lemma is necessary for it to be well-defined.
+\section{The Confluent Trace Semantics}
+Before defining the new version of the reduction relation, the following lemma is necessary for it to be well-defined.
 
@@
 \begin{definition}[$\Rightarrow$ reduction]
-Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function, so it is denoted ``$\Rightarrow$'' instead of ``$\red$'', and it relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
-We write an element of $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ as $(M', s)$ where the \emph{term} $M' \in \Lambda^0$ and $s \in  I^{L_s(M')}$.
+This version of the reduction relation now specifies the results of $\tsample$ reductions, but is still nondeterministic with respect to the order of reduction.
+It relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
+We write an element of $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ as $(M', s)$ where the term $M' \in \Lambda^0$ and $s \in  I^{L_s(M')}$.
 \begin{align*}
@@
 
-\paragraph*{Reduction strategy determinising $\Rightarrow$}
+In order to show that $\Rightarrow$ behaves as expected, the following lemma is also necessary, in order to show that a sample is never used multiple times in the same reduction sequence:
+\begin{lemma} \label{lem:redexDestroyed}
+If $M \to N$, with the redex at position $\alpha$, then no position in any term reachable from $N$ is related by $\sim^*$ to $(M,\alpha)$.
+\end{lemma}
+
 The reduction relation $\Rightarrow$ is nondeterministic, so it admits multiple possible reduction strategies. 
@@
 
-With these definitions, it is now possible to relate the confluent trace semantics back to the standard trace semantics.
+This reduction strategy allows the confluent trace semantics to be related to the standard version of the trace semantics with a fixed reduction order and linear traces.
+In \cite{arxiv}, which gives the full definition of the standard trace semantics, this is used to prove the following theorems that allow termination results to be transferred from the confluent trace semantics to the standard trace semantics.
 
@@
 \end{proof}
-
-All of the theorems on the termination of rankable terms therefore extend to other reduction strategies too.
-The proofs of \Cref{thm:rankable implies termination,thm:partial implies rankable,thm:antitone rankable implies termination,thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies.
-They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because \Cref{thm:de groote} applies equally to any reduction strategy).
-For a reduction strategy $r$ and term $M$, just substitute $N[r(N)]$ being a $\tY$-redex for $N$ being of the form $E[\tY \lambda x.
-O]$, and $r(N)$ being undefined for $N$ being a value.
-
-The domain of definition of the ranking functions also needs to be changed from the reachable terms $\mathit{Rch}(M)$ to the \emph{reachable terms with respect to the reduction strategy $r$},
-\[
-\begin{array}{l}
-\mathit{Rch}_r(M) := \{N \mid \exists n, (N_i)_{0 \leq i \leq n} : {}\\
-\qquad\qquad\qquad N_0 = M, N_n = N, N_i \to N_{i+1} \text{ at } r(N_i)\}.
-\end{array}
-\]
-
-More explicitly, the modified forms of the theorems are:
-\begin{definition}
-An \emph{antitone ranking function on $M$ w.r.t.~a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ such that
-\begin{itemize}
-    \item $f(N) \geq \epsilon(f(N)) + f(N')$ if $N[r(N)]$ is a $\tY$-redex and $N$ reduces to $N'$ at $r(N)$
-    \item $f(N) \geq \int_I f(N[\underline{x}/r(N)]) \, \Leb(\mathrm{d}x)$ if $N[r(N)] = \tsample$
-
-    \item $f(N) \geq f(N')$ if $r(N)$ is any other redex, where $N \to N'$ at $r(N)$.
-\end{itemize}
-Any closed term for which an antitone ranking function with respect to a reduction strategy $r$ exists is called \emph{antitone rankable w.r.t.~$r$}. 
-\end{definition}
-
-\begin{definition}
-An \emph{antitone sparse ranking function on $M$ w.r.t.~a reduction strategy $r$} is a partial function $f : \mathit{Rch}_r(M) \rightharpoonup \mathbb R$ such that there exists some antitone function $\epsilon : \nnReal \to \pReal$ such that
-\begin{inparaenum}[(i)]
-    \item $f(N) \geq 0$ for all $N \in \dom(f)$,
-    \item $M \in \dom(f)$,
-    \item for any $N \in \dom(f)$, evaluation of $N$ at the positions specified by $r$ will eventually reach some $O$ such that either $r(O)$ isn't defined or $f(O)$ is, and $f(N) \geq \mathbb E[f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of $\dom(f)$).
-\end{inparaenum}
-\iffalse
-\begin{itemize}
-    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
-    \item $f$ is defined at $M$.
-    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ at the positions specified by $r$ will eventually reach some $O$ such that either $r(O)$ isn't defined or $f(O)$ is, and $f(N) \geq \mathbb E[f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
-\end{itemize}
-\fi
-\end{definition}
-
-If the antitone function is just the constant function $1$ instead, then it is called a \emph{(sparse) ranking function on $M$ w.r.t.~$r$} instead, and the term is called \emph{rankable w.r.t.~$r$}.
-
-\begin{theorem}[Confluent ranking] \label{thm:confluent ranking}
-If a closed PPCF term $M$ has a ranking function, sparse ranking function, antitone ranking function or antitone sparse ranking function w.r.t.~a reduction strategy $r$, then $M$ is AST w.r.t.~$r$, and AST.
-\end{theorem}
-
-When constructing an (antitone) sparse ranking function with respect to an alternative reduction strategy, it is intuitively simplest to define the sparse ranking function and the reduction strategy together.
-For each key reachable term where the sparse ranking function is defined, simply decide which sub-term should be reduced next, and how far before the next key reachable term.
-The only restrictions to bear in mind on evaluating sub-terms are that a $\tsample$ may not be evaluated inside a $\lambda$ that's inside a $\tY$ or on the right of an application, and a $\beta$ redex may not be reduced until its argument is a value.
-This can be seen as applied to \Cref{ex:sum of powers} at the end of \Cref{apx:confluent}.
-
-\medskip
-If, as seems likely, \Cref{thm:antitone rankable implies termination} is complete for terms from which every reachable term is AST, then by the following theorem, \Cref{thm:confluent ranking} can't actually prove the termination of any terms not already provable by \Cref{thm:antitone rankable implies termination} and \Cref{thm:antitone partial implies rankable}.
-
-\begin{restatable}{theorem}{confluentNoStronger}
-\label{thm:confluentNoStronger}
-For any closed term $M$ and reduction strategy $r$ on $M$, if every term in $\mathit{Rch}_r(M)$ is AST, then every term in $\mathit{Rch}(M)$ is AST.
-\end{restatable}
-
-\paragraph*{Custom reduction strategy for AST analysis} However, there are some terms which terminate more simply and directly via some alternative reduction strategy.
-For these examples, a custom reduction strategy is specified for each term under consideration.
-
-\begin{example}[Sum of powers] \label{ex:sum of powers}
-The following term picks a random natural number $n$, then computes $\sum_{k=0}^n k^n$.
-\begin{align*}
-M  &= (\lambda n. P[n]) \lfloor \tsample^{-1/2} \rfloor \\
-P[n]  &= (\lambda p. \Xi[p] \, n) (\lambda x. \Theta[x] \, n)) \\
-\Xi[p]  &= \tY \lambda f n. \tif{n = 0}{\underline 0}{p \, n + f (n-1)} \\
-\Theta[x]  &= \tY \lambda f n. \tif{n = 0}{\underline 1}{x \times f (n-1)}.
-\end{align*}
-
-With the CBV reduction strategy, the recursion in $\Theta[x]$ (that defines the function $k \mapsto k^x$) is executed separately for every term in the sum in $\Xi$.
-Not only is this much more complicated than simplifying $\Theta[x] \, \underline n$ first, it requires considerably more $\tY$-reduction steps, so many so that $M$ is $\tY$-PAST with respect to a reduction strategy that simplifies $\Theta[x] \, \underline n$ first (call it ``$r$''), but isn't $\tY$-PAST with respect to the standard reduction strategy, so that $M$ is rankable with respect to $r$, but isn't rankable.
-It is still antitone rankable, but this is much more complicated.
-Details of the ranking function construction with and without using the confluent semantics are given in \Cref{apx:confluent}.
-\end{example}
-
-\begin{remark}
-\label{rem:ex corrected using confluent semantics}
-Even some of the examples given earlier implicitly use slightly non-standard reduction strategies to make the definitions of their ranking functions slightly simpler.
-For example, the antitone sparse ranking function in \Cref{ex:unbiased random walk} was defined at $\Theta_2\, \underline n$ (where $\Theta_2 = \tY \, \lambda f n .
-\, \tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$), but $\Theta_2\, \underline{10}$ actually reduces to $\Theta_2\, (\underline{10} - 1)$ or $\Theta_2\, (\underline{10} + 1)$, and (taking the second of these), it then reduces to $(\lambda z.
-(\lambda n.
-\, \tif{n = 0}{0}{\Theta_2 \, (n - 1) \oplus_{1/2} \Theta_2 \, (n + 1)}) z) \, (\underline{10} + 1)$, then $(\lambda z.
-(\lambda n.
-\, \tif{n = 0}{0}{\Theta_2 \, (n - 1) \oplus_{1/2} \Theta_2 \, (n + 1)}) z) \, \underline{11}$, bypassing $\Theta_2\, \underline{11}$ entirely.
-The antitone sparse ranking function in the form given can be justified anyway, by considering it to be using a reduction strategy which is the same as $\cbv$ except that in $\Theta_2\, (\underline f \, \underline n)$, it evaluates the redex at $@_2$ first.
-
-\Cref{ex:higher-order recursion} also makes use of an alternative reduction strategy.
-In this case, it is to allow the antitone sparse ranking function to skip over all of those terms where the unknown function $F$ is part way through evaluation.
-\end{remark}

### 2021-04-26 LICS21/intro.tex
Edit to focus on confluence result for IWC.
Remove everything except the sections on the confluent trace semantics
and syntax. Add a new introduction. Remove references to other parts
of the paper. Generally edit so that it makes sense on its own.

@@
 
-Probabilistic (or randomised) programs have long been recognised as essential to the efficient solution of many algorithmic problems \cite{Rabin1976}.
-More recently, in probabilistic programming \cite{DBLP:conf/icse/GordonHNR14,rainforth2017Automating,vandemeent2018Introduction}, probabilistic programs are used to express generative models whose posterior probability can be computed by general purpose inference engines.
-Sampling from continuous distributions is an essential construct of probabilistic programming languages, such as Church \cite{DBLP:conf/uai/GoodmanMRBT08}, Stan \cite{carpenter2017stan}, Anglican \cite{DBLP:conf/pkdd/TolpinMW15}, Gen \cite{cusumano-towner2019Gen}, Pyro \cite{bingham2019Pyro}, Edward \cite{tran2016edward} and Turing \cite{ge2018Turing}.
-Another key feature is higher order; in fact some of the most influential probabilistic programming languages are functional.
+Non-probabilistic lambda calculi are generally confluent, i.e.~if a term $A$ reduces to both $B_1$ and $B_2$, there is some $C$ 
+to which both $B_1$ and $B_2$ reduce,
+so the reduction order mostly doesn't matter. 
+In the probabilistic case, this may not be true, because $\beta$-reduction can duplicate $\tsample$s, so the outputs of the copies of the $\tsample$ may be identical or independent, depending on whether the sample is taken before or after $\beta$-reduction.
+Consider for example the term $(\lambda x.
+x + x)\, \tsample$, where $\tsample$ reduces to a number chosen uniformly at random from the interval $I = [0,1]$.
+If it is reduced in call-by-value order, first the $\tsample$ reduces to some number $r$, then the $\beta$ redex is evaluated, then $r$ is added to itself, yielding $\underline{2r}$.
+If it is reduced in call-by-name order instead, first the $\beta$ redex is reduced, yielding $\tsample + \tsample$, then the $\tsample$s are evaluated independently and added, yielding $\underline{r + r'}$.
+As $r$ and $r'$ are independent, the distribution of results is triangular, with support $[0,2]$ and peak at $1$, which is different from the uniform distribution of results in the CbV case.
 
-In this paper we study a central property of probabilistic programs: \emph{termination}.
-When a probabilistic program implements a solution to an algorithmic problem, we naturally require the computation to terminate with probability $1$, in which case the program is called \emph{almost surely terminating} (AST).
-Indeed, it is standard for designers and implementors of probabilistic programming systems to regard non-AST programs as defining invalid models, and hence inadmissible \cite{Rainforth2017,DBLP:conf/uai/GoodmanMRBT08}.
-(Yet none of these systems provides any support for the development or verification of AST programs.)
+The results obtained by CbV and CbN evaluation differ in a significant way, however, there are some cases where the order of evaluation doesn't matter.
+For example, in $\tsample + \tsample$, the order in which the $\tsample$s are evaluated doesn't affect the final result, and in $(\lambda x.
+\tsample)\, \underline 0$, the $\beta$ redex and the $\tsample$ can be evaluated in either order.
+In order to obtain the desired confluence result, we restrict our attention to a class of reduction strategies that are equivalent to CbV, as the CbN semantics is less expressive.
 
-Moreover various theorems about probabilistic programs rely on the assumption that the program terminates almost surely (see e.g.~\cite{DBLP:conf/uai/GoodmanMRBT08}). 
-A recent result \cite{MakOPW21} proves that AST programs have density functions that are differentiable almost everywhere.
-This is significant for Bayesian machine learning, because almost everywhere differentiability is a precondition for the correctness of some of the most scalable inference algorithms, such as Hamiltonian Monte Carlo \cite{DBLP:conf/aistats/ZhouGKRYW19,Nishimura2020b} and reparameterised gradient variational inference \cite{DBLP:conf/nips/LeeYY18}.
+Even with such a restriction, a trace semantics in the usual style would not be entirely confluent.
+In the normal sort of trace semantics\cite{borgstrom2016lambda}, there is a sequence of samples, the \emph{trace}, selected at random from a trace space such as $I^{\mathbb N}$, then for every $\tsample$ statement reduction, the next sample from the trace is used in order, so that the samples in the trace are effectively each labelled by a number corresponding to the execution order of the $\tsample$ statements.
+Consider the evaluation of the term $\tsample - \tsample$ using one of these simple linear traces, $(1, 0, \dots)$.
+It would reduce to either $1$ or $-1$ depending on the order of evaluation of the $\tsample$s, as that determines which sample from the pre-selected sequence is used for each one.
+To fix this, rather than pre-selecting samples according to the order they'll be drawn in, they can be labelled according to the position in the term where they'll be used instead.
 
-The AST problem is not just important, but also difficult: 
-Deciding AST of first-order imperative programs with discrete probabilities is $\Pi^0_2$-complete \cite{DBLP:conf/mfcs/KaminskiK15}.
-In recent work (under double-blind review elsewhere), the AST problem for higher-order programs with continuous distributions (and suitable primitive functions) is shown to have the same complexity.
-Though the problem of proving AST of imperative probabilistic programs is much studied \cite{DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05,DBLP:conf/popl/FioritiH15,DBLP:conf/pldi/ChenH20,DBLP:journals/toplas/ChatterjeeFNH18}, to our knowledge the problem of proving AST of higher-order programs with continuous distributions is new.
+Further details, including all of the missing proofs, can be found in \cite[\S IV, \S D]{arxiv}.
 
-\subsection*{Contributions}
-
-A possible approach to proving AST is to find some variant on the program state, called \emph{ranking function}, that decreases on average sufficiently quickly that it must at some point reach 0, at which point the program terminates.
-In other words, the program's behavior is used to define an associated (ranking) \emph{supermartingale}.
-Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \cite{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18}.
-This paper's contribution is to extend this method to a higher-order setting.
-
-The language PPCF used in this paper is simply typed, with random continuous sampling, and an explicit recursion primitive, $\tY$. 
-We define a \emph{ranking function} on a term $M$ to be a non-negative measurable function on the reachable terms from $M$ whose expected value decreases or is unchanged as the term is reduced.
-Because the type system already constrains the terms enough to force termination in the absence of the recursion construct \cite{tait1967,BarendregtDS2010}, it is only the $\tY$-reduction steps that must be counted, making defining ranking functions somewhat easier.
-Using supermartingales and Doob's (iterated) Optional Sampling Theorem, we show the soundness of rankability (\cref{thm:rankable implies termination}): if a term has a ranking function, it is AST.
-
-To define a ranking function, one would typically try to organise the set of reachable terms (of the program being analysed) into a manageable number of syntactic cases. 
-Unfortunately (for any fixed reduction strategy) there are programs whose reachable terms span such a large number of cases that it would be extremely difficult to analyse by hand, if at all possible.
-How can the construction of ranking functions be made possible or easier?
-
-Our first answer is the notion of \emph{sparse ranking function}. 
-Most of the individual execution steps of a typical program are trivial and easy to mentally skip over.
-Sparse ranking functions can be defined only for those points in the execution of a program which are semantically important, while all of the other intermediate steps can be ignored.
-Yet, they are no less efficacious for proving AST (\cref{thm:partial implies rankable}): every sparse ranking function is a restriction of a ranking function.
-This also makes the ranking function method of proving AST more compatible with syntactic sugar, because the intermediate reduction steps implicit in the simplified notation can be ignored.
-
-A ranking function (or sparse ranking function) provides a bound on the expected number of $\tY$-reduction steps before the program terminates, therefore ranking functions cannot be constructed for terms whose expected number of $\tY$-reduction steps is infinite, such as the simplest implementation of the 1D unbiased random walk.
-This restriction can be removed by generalising ranking functions to \emph{antitone ranking functions}, which rather than having to decrease by a constant amount for each $\tY$-reduction step, may decrease by a variable amount, depending on the value of the ranking function.
-Thanks to this feature, the antitone ranking function method is capable of handling programs which terminate arbitrarily slowly. \lo{TODO: Reference some examples.}
-We show that this method is also sound for proving AST (\cref{thm:antitone rankable implies termination}), 
-moreover it also enjoys a sparse function theorem (\cref{thm:antitone partial implies rankable}).
-\akr{Our progress condition is equivalent to that given by McIver et al.~\cite{DBLP:journals/pacmpl/McIverMKK18} (by applying a transformation to the ranking function), but needs to be different in order to be compatible with the presence of additional intermediate reduction steps where the ranking function is not required to decrease at all.}
-\lo{The same point is made in \cref{sec:related}, so the preceding can be omitted.}
-
-Basic (non-random) lambda calculus has the very useful property of confluence, which implies (among other things) that even if execution of a program starts in a different order, it will still reach the same normal form eventually (assuming it does reach a normal form).
-Probabilistic lambda calculus does not have this property, because random choices may be duplicated, and evaluating the same subterm multiple times can yield different results. 
-However, with a restricted set of reduction strategies, confluence may be regained. 
-We introduce a novel addressing scheme for the possible random choices in a program's execution, which ensures that the same random choices are taken at corresponding positions in alternative reduction sequences, so that the same eventual result can be reached.
-This is then used to prove yet another extension to the ranking function theorem, that (\cref{thm:confluent ranking}) ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and (\cref{cor:Reduction strategy independence}) rankability in this sense still imples almost sure termination.
-The confluent trace semantics is of wider interest, and has other possible applications as well, for example in Bayesian inference algorithms.
-
-By combining these methods we can prove AST of a variety of PPCF programs that are beyond the reach of methods in the literature, including non-affine recursion\footnote{\label{fnote:non-affine} Non-affine recursive programs are recursive programs that can, during the evaluation of the recursive body, make multiple recursive calls (of a first-order function) from distinct call sites.} (\cref{ex:non-affine recursion easy,ex:non-affine continuous}) and recursions that define higher-order functions (\cref{ex:higher-order recursion}).
-
-\paragraph*{Outline}
-
-We present the syntax and trace semantics of PPCF in \Cref{sec:PPCF}.
-In \Cref{sec:supermartingales}, we show that ranking functions on terms induce supermartingales, which form the basis of a sound method for proving AST.
-We introduce sparse ranking functions in \Cref{sec:ranking} and antitone ranking functions in \Cref{sec:antitone}, and illustrate how they can be used to prove AST via examples. 
-In \Cref{sec:confluent}, we present a confluent trace semantics and demonstrate its usefulness.
-We discuss further applications of the confluent semantics in \Cref{sec:applications}; and conclude with comments on related work and further directions in \Cref{sec:related}.
-
-\paragraph*{Additional materials} Further details of some of the examples and all missing proofs can be found in the appendix.
+\subsection{Outline}
+First, the syntax of the language PPCF is introduced.
+Positions are defined as a way of addressing sample statements within a program independently of the reduction order.
+Next, a version of the reduction relation is presented that is nondeterministic, so that it allows a choice of what order to perform reductions in.
+The notion of positions is extended to potential positions, for samples which may appear later in the reduction sequence but not necessarily in the initial term.
+In order to allow potential positions in different reduction sequences to be considered equivalent, a relation $\sim^*$ is defined, and finally, all of these are used to construct a confluent version of the trace semantics, $\Rightarrow$, that is still nondeterministic in reduction order, but does specify the outcome of random choices.

### 2021-04-28 LICS21/SPCF.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
# Conflicts:
#	thesis proposal/thesis proposal.tex

Minors; references

--- 2021-04-26 LICS21/SPCF.tex
@@
 
-The language PPCF is a call-by-value (CBV) version of PCF with sampling of real numbers from the closed interval $[0,1]$ \cite{Ehrhard2018c,DBLP:journals/pacmpl/EhrhardPT18,MakOPW21}.
-%\changed[lo]{following \cite{MakOPW21} except the score function (\Cref{rem:score})}. 
+The language PPCF is a call-by-value (CBV) version of PCF with sampling of real numbers from the closed interval $[0,1]$ \cite{Ehrhard2018c,DBLP:journals/pacmpl/EhrhardPT18,DBLP:conf/esop/MakOPW21}.
+%\changed[lo]{following \cite{DBLP:conf/esop/MakOPW21} except the score function (\Cref{rem:score})}. 
 Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
@@
 
-The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid \hbox{$M$ terminates at $s$}\}$ is measurable \cite{DBLP:conf/icfp/BorgstromLGS16,MakOPW21}.
+The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid \hbox{$M$ terminates at $s$}\}$ is measurable \cite{DBLP:conf/icfp/BorgstromLGS16,DBLP:conf/esop/MakOPW21}.
 We say that a term $M$ is \emph{almost surely terminating} (\emph{AST}) just if $\mu_{\entrosp}(\{s \mid \hbox{$M$ terminates at $s$}\}) = 1$.
@@
 \label{rem:score}
-Our definition of AST is equivalent to that given in \cite{MakOPW21} (although the program semantics is stated in a slightly different way), except for the presence of a $\tscore$ construct for soft conditioning.
+Our definition of AST is equivalent to that given in \cite{DBLP:conf/esop/MakOPW21} (although the program semantics is stated in a slightly different way), except for the presence of a $\tscore$ construct for soft conditioning.
 The $\tscore$ construct is irrelevant to termination except that it fails if its argument is negative, thus allowing computations to fail after finitely many steps. 
@@
 \lo{Your operational semantics does not maintain a record of the current weight of the reduction.
-A.s.~termination does depend on $\tscore$: see \cite[\S 4.3]{MakOPW21}\footnote{\url{https://arxiv.org/abs/2004.03924}}.
+A.s.~termination does depend on $\tscore$: see \cite[\S 4.3]{DBLP:conf/esop/MakOPW21}\footnote{\url{https://arxiv.org/abs/2004.03924}}.
 I think it important to take the behaviour of $\tscore$ into account;

### 2021-04-28 LICS21/intro.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
# Conflicts:
#	thesis proposal/thesis proposal.tex

Minors; references

--- 2021-04-26 LICS21/intro.tex
@@
 Moreover various theorems about probabilistic programs rely on the assumption that the program terminates almost surely (see e.g.~\cite{DBLP:conf/uai/GoodmanMRBT08}). 
-A recent result \cite{MakOPW21} proves that AST programs have density functions that are differentiable almost everywhere.
+A recent result \cite{DBLP:conf/esop/MakOPW21} proves that AST programs have density functions that are differentiable almost everywhere.
 This is significant for Bayesian machine learning, because almost everywhere differentiability is a precondition for the correctness of some of the most scalable inference algorithms, such as Hamiltonian Monte Carlo \cite{DBLP:conf/aistats/ZhouGKRYW19,Nishimura2020b} and reparameterised gradient variational inference \cite{DBLP:conf/nips/LeeYY18}.
@@
 Deciding AST of first-order imperative programs with discrete probabilities is $\Pi^0_2$-complete \cite{DBLP:conf/mfcs/KaminskiK15}.
-In recent work (under double-blind review elsewhere), the AST problem for higher-order programs with continuous distributions (and suitable primitive functions) is shown to have the same complexity.
+In recent work \cite{BeutnerO21} %(under double-blind review elsewhere), 
+the AST problem for higher-order programs with continuous distributions (and suitable primitive functions) is shown to have the same complexity.
 Though the problem of proving AST of imperative probabilistic programs is much studied \cite{DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05,DBLP:conf/popl/FioritiH15,DBLP:conf/pldi/ChenH20,DBLP:journals/toplas/ChatterjeeFNH18}, to our knowledge the problem of proving AST of higher-order programs with continuous distributions is new.

### 2021-04-28 probabilisticProgrammingMartingales.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
# Conflicts:
#	thesis proposal/thesis proposal.tex

Minors; references

@@
 
-The language SPCF is a simply-typed lambda calculus with sampling of real numbers from $[0,1]$ and unbounded scoring, following \cite{MakOPW21}.
+The language SPCF is a simply-typed lambda calculus with sampling of real numbers from $[0,1]$ and unbounded scoring, following \cite{DBLP:conf/esop/MakOPW21}.
 Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
@@
 \paragraph{}
-This definition of almost sure termination is equivalent to that given in \citep{MakOPW21}, although the program semantics is stated in a slightly different way.
+This definition of almost sure termination is equivalent to that given in \citep{DBLP:conf/esop/MakOPW21}, although the program semantics is stated in a slightly different way.
 In particular, the argument to $\tscore$ is not relevant to termination (except for the possibility that its argument's evaluation wouldn't terminate).

### 2021-04-29 LICS21/intro.tex
Fix references
It should not be possible to actually run bibtex without errors.

@@
 When a probabilistic program implements a solution to an algorithmic problem, we naturally require the computation to terminate with probability $1$, in which case the program is called \emph{almost surely terminating} (AST).
-Indeed, it is standard for designers and implementors of probabilistic programming systems to regard non-AST programs as defining invalid models, and hence inadmissible \cite{Rainforth2017,DBLP:conf/uai/GoodmanMRBT08}.
+Indeed, it is standard for designers and implementors of probabilistic programming systems to regard non-AST programs as defining invalid models, and hence inadmissible \cite{rainforth2017Automating,DBLP:conf/uai/GoodmanMRBT08}.
 (Yet none of these systems provides any support for the development or verification of AST programs.)

### 2021-04-29 LICS21/confluent.tex
Remove appendices

--- 2021-04-26 LICS21/confluent.tex
@@
 
-\paragraph*{Reduction strategy determinising $\Rightarrow$}
 The reduction relation $\Rightarrow$ is nondeterministic, so it admits multiple possible reduction strategies. 
@@
 The proofs of \Cref{thm:rankable implies termination,thm:partial implies rankable,thm:antitone rankable implies termination,thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies.
-They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because \Cref{thm:de groote} applies equally to any reduction strategy).
+They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because the termination property implied by the type system applies equally to any reduction strategy).
 For a reduction strategy $r$ and term $M$, just substitute $N[r(N)]$ being a $\tY$-redex for $N$ being of the form $E[\tY \lambda x.
@@
 The only restrictions to bear in mind on evaluating sub-terms are that a $\tsample$ may not be evaluated inside a $\lambda$ that's inside a $\tY$ or on the right of an application, and a $\beta$ redex may not be reduced until its argument is a value.
-This can be seen as applied to \Cref{ex:sum of powers} at the end of \Cref{apx:confluent}.
 
@@
 
-\paragraph*{Custom reduction strategy for AST analysis} However, there are some terms which terminate more simply and directly via some alternative reduction strategy.
+However, there are some terms which terminate more simply and directly via some alternative reduction strategy.
 For these examples, a custom reduction strategy is specified for each term under consideration.
@@
 It is still antitone rankable, but this is much more complicated.
-Details of the ranking function construction with and without using the confluent semantics are given in \Cref{apx:confluent}.
 \end{example}

### 2021-04-29 LICS21/antitone.tex
Remove appendices
Cut down the length some more

@@
 
-\begin{example}[Random walk]
+\begin{example}[Random walks]
 \label{ex:ac-ranking}
-\begin{compactenum}[(i)]
-\item One-dimensional (1D) biased (towards 0) random walk
-\[
+\begin{align*}
 M_1 = 
 \big(\tY \, \lambda f n . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{2/3} f \, (n + 1)}\big) \, 10
-\]
-
-\item 1D unbiased random walk
-\[
+\tif{n = 0}{0}{f \, (n - 1) \oplus_{2/3} f \, (n + 1)}\big) \, 10 \\
 M_2 = 
 \big(\tY \, \lambda f n . \, 
-\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}\big) \, 10
-\]
-
-\item 1D biased (away from 0) random walk
-\[
+\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}\big) \, 10 \\
 M_3 = 
@@
 \tif{n = 0}{0}{f \, (n - 1) \oplus_{1/3} f \, (n + 2)}\big) \, 10
-\]
-%While {x > 0} do {x := x-1 \oplus_2/3 x:= x+1}
-%M2 = While {x > 0} do {x := x-1 \oplus_2/3 x:= x+2}
-\end{compactenum}
+\end{align*}
 \end{example}
-
-\medskip
 
@@
 Now
-\begin{align*}%\textstyle
-& g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 \\
-  &=  \ln \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) 
-  =  \frac 1 2 \ln\left(\frac{(n+1)^2}{n(n+2)}\right) \\
-  &=  \frac 1 2 \ln\left(1 + \frac 1 {n(n+2)}\right) 
-  >  \frac 1 {n(n+2) + 1} \\
-  &>  \frac 1 2 \frac 1 {(n+1)^2} + \frac 1 2 \frac 1 {(n+3)^2}
-\end{align*}
+\[
+g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2  >  \frac 1 2 \frac 1 {(n+1)^2} + \frac 1 2 \frac 1 {(n+3)^2}
+\]
 Moreover $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
@@
 
-\paragraph*{More challenging examples} All of the following examples also have antitone ranking functions, provided in Appendix \ref{apx:antitone}.
+\paragraph*{More challenging examples} All of the following examples also have antitone ranking functions, provided in \cite{arxiv}.
 

### 2021-04-29 LICS21/confluent.tex
Cut down the length some more

@@
 
-In order to be able to use alternative reduction strategies to prove properties of the program defined using the usual reduction strategy, we first prove the relation between these different strategies, providing a confluent version of the semantics.
+In order to be able to use alternative reduction strategies to prove properties of the program defined using the usual strategy, we prove the relation between these different strategies, providing a confluent version of the semantics.
 It is therefore necessary to have some way of relating the samples that will be used in runs of the program using different reduction strategies.
-Positions are defined as a way of addressing sample statements within a program independently of the reduction order.
-Next, a version of the reduction relation is presented that is nondeterministic, so that it allows a choice of what order to perform reductions in.
+Positions are defined as a way of addressing subterms independently of the reduction order.
+Next, we present a nondeterministic version of the reduction relation, that allows a choice of what order to perform reductions in.
 The notion of positions is extended to potential positions, for samples which may appear later in the reduction sequence but not necessarily in the initial term.
@@
 
-Non-probabilistic lambda calculi are generally confluent, i.e.~if a term $A$ reduces to both $B_1$ and $B_2$, there is some $C$ 
-to which both $B_1$ and $B_2$ reduce,
+Non-probabilistic lambda calculi are generally confluent, i.e.~if $A \to B_1, B_2$, then $\exists C.\ B_1, B_2 \to^* C$,
 so the reduction order mostly doesn't matter. 
@@
   \text{if } M \mid \alpha = \underline f (\underline r_1, \dots , \underline r_n),\ & M \to M[\underline{f(r_1,\dots,r_n)}/\alpha] \\
-  \text{if } M \mid \alpha = \tY \lambda x. N,\ & M \to M[\lambda z. N[(\tY \lambda x. N)/x] z/\alpha]\\ \text{where }z&\text{ is not free in $N$}\\
+  \text{if } M \mid \alpha = \tY \lambda x. N,\ & M \to M[\lambda z. N[(\tY \lambda x. N)/x] z/\alpha]\\ \text{where }z&\text{ is not free in $N$} \\
   \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_1/\alpha] \text{ where }r < 0 \\
@@
 \end{definition}
-
-The argument of a $\beta$ redex and the body of a $\tY$ redex may be duplicated by those reductions.
-It is therefore these cases that need to be handled carefully to avoid duplicating samples at the wrong time.
-In both cases, the potentially duplicated part must already be a value, which excludes terms like $\tsample$ or $\tsample + \underline 1$, which should be evaluated before being duplicated.
-In the other direction, if a $\tsample$ occurs inside of a $\lambda$, it may need to be duplicated before being evaluated, which is why a $\tsample$ reduction isn't allowed inside a $\lambda$ inside a $\tY$ or the right side of an application.
-These restrictions are in some cases unnecessarily strict, for example, in $(\lambda x.
-x) ((\lambda y.
-\tsample) \underline 0)$, it would be fine to evaluate the sample first, but they are at least sufficient to ensure confluence in terms of the distribution of results.
-Getting individual traces to behave correctly will take more work though.
 

### 2021-04-29 LICS21/intro.tex
Remove appendices
Cut down the length some more

@@
 Indeed, it is standard for designers and implementors of probabilistic programming systems to regard non-AST programs as defining invalid models, and hence inadmissible \cite{rainforth2017Automating,DBLP:conf/uai/GoodmanMRBT08}.
-(Yet none of these systems provides any support for the development or verification of AST programs.)
-
-Moreover various theorems about probabilistic programs rely on the assumption that the program terminates almost surely (see e.g.~\cite{DBLP:conf/uai/GoodmanMRBT08}). 
-A recent result \cite{DBLP:conf/esop/MakOPW21} proves that AST programs have density functions that are differentiable almost everywhere.
-This is significant for Bayesian machine learning, because almost everywhere differentiability is a precondition for the correctness of some of the most scalable inference algorithms, such as Hamiltonian Monte Carlo \cite{DBLP:conf/aistats/ZhouGKRYW19,Nishimura2020b} and reparameterised gradient variational inference \cite{DBLP:conf/nips/LeeYY18}.
+(Yet none of these systems provides any support for the development or verification of AST programs.) Moreover various theorems about probabilistic programs rely on the assumption that the program terminates almost surely (see e.g.~\cite{DBLP:conf/uai/GoodmanMRBT08, DBLP:conf/esop/MakOPW21}).
 
@@
 
-A possible approach to proving AST is to find some variant on the program state, called \emph{ranking function}, that decreases on average sufficiently quickly that it must at some point reach 0, at which point the program terminates.
-In other words, the program's behavior is used to define an associated (ranking) \emph{supermartingale}.
+Our approach to proving AST is to find some variant on the program state, called \emph{ranking function}, that decreases on average sufficiently quickly that it must at some point reach 0, at which point the program terminates.
+Equivalently, the program's behavior is used to define an associated (ranking) \emph{supermartingale}.
 Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \cite{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18}.
@@
 Our first answer is the notion of \emph{sparse ranking function}. 
-Most of the individual execution steps of a typical program are trivial and easy to mentally skip over.
+Most of the individual execution steps of a typical program are trivial and easy to ignore.
 Sparse ranking functions can be defined only for those points in the execution of a program which are semantically important, while all of the other intermediate steps can be ignored.
-Yet, they are no less efficacious for proving AST (\cref{thm:partial implies rankable}): every sparse ranking function is a restriction of a ranking function.
-This also makes the ranking function method of proving AST more compatible with syntactic sugar, because the intermediate reduction steps implicit in the simplified notation can be ignored.
+Yet, the existence of a sparce ranking function still implies AST (\cref{thm:partial implies rankable}).
+This also makes this method of proving AST more compatible with syntactic sugar, because the intermediate reduction steps implicit in the simplified notation can be ignored.
 
-A ranking function (or sparse ranking function) provides a bound on the expected number of $\tY$-reduction steps before the program terminates, therefore ranking functions cannot be constructed for terms whose expected number of $\tY$-reduction steps is infinite, such as the simplest implementation of the 1D unbiased random walk.
+A ranking function (or sparse ranking function) provides a bound on the expected number of $\tY$-reduction steps before the program terminates, therefore ranking functions cannot be constructed for terms whose expected number of $\tY$-reduction steps is infinite, such as the 1D unbiased random walk.
 This restriction can be removed by generalising ranking functions to \emph{antitone ranking functions}, which rather than having to decrease by a constant amount for each $\tY$-reduction step, may decrease by a variable amount, depending on the value of the ranking function.
@@
 We introduce a novel addressing scheme for the possible random choices in a program's execution, which ensures that the same random choices are taken at corresponding positions in alternative reduction sequences, so that the same eventual result can be reached.
-This is then used to prove yet another extension to the ranking function theorem, that (\cref{thm:confluent ranking}) ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and (\cref{cor:Reduction strategy independence}) rankability in this sense still imples almost sure termination.
+This is then used to prove yet another extension to the ranking function theorem (\cref{thm:confluent ranking}), that ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and rankability in this sense still imples almost sure termination.
 The confluent trace semantics is of wider interest, and has other possible applications as well, for example in Bayesian inference algorithms.
@@
 
-\paragraph*{Additional materials} Further details of some of the examples and all missing proofs can be found in the appendix.
+\paragraph*{Additional materials} Further details of some of the examples and all missing proofs can be found in the complete version of this paper on arXiv.\cite{arxiv}

### 2021-04-29 LICS21/related.tex
Cut down the length some more

@@
 
-The \Cref{def:more general red} of redexes and which positions are acceptable to reduce at is sufficiently restrictive to guarantee confluence, but is also a little more restrictive in some cases than is necessary for this purpose.
-For example, if the argument of a function is not yet a value, but its reduction to a value would be deterministic, or the function is affine, then applying the function before reducing its argument would not cause any problematic duplication of random samples.
-Similarly, if $\tsample$ occurs at position $\alpha;@_2;\beta;\lambda;\gamma$, but the function at $\alpha;@_1$ is affine, evaluating the $\tsample$ may in some cases also work fine.
+The \Cref{def:more general red} of redexes and which positions are acceptable to reduce at is sufficiently restrictive to guarantee confluence, but there are several cases where it is a more restrictive than is necessary for this purpose.
+For example, if the argument of a function is not yet a value, but its reduction to a value would be deterministic, applying the function before reducing its argument would not cause any problematic duplication of random samples.
 A more complete characterisation of which redexes can be reduced without breaking confluence could be interesting.

### 2021-04-30 LICS21/antitone-proofs.tex
Move figures back to main text, and make alignment tweaks.

@@
 
-Not only is this program non-affine recursive,
-note also the use of a random sample, $e$, as a first-class value, and the subsequent use as a probability in the binary choice $\oplus_e$.
-Such a computation cannot be modelled via discrete distributions.
+This example is a little more complex because of the use of a random sample, $e$, as a first class value, which cannot be modelled via discrete distributions.
 
@@
 \] 
-where $X$ is not a value or of the form $F\ \underline{n_k}\ Z$ for some non-value $Z$. All terms match this pattern for exactly one value of $k \geq 0$, except for values. 
+where $X$ is not a value or of the form $F\ \underline{n_k}\ Z$ for some non-value $Z$. 
 Within $X$, there is necessarily a redex at $\cbv(X)$. 
@@
 By the induction hypothesis, $y_{t_k} \geq 2^k$.
+
 \begin{align*}
@@
 
-Although the stronger induction hypothesis was necessary for the induction to work, the only reason this is needed is to prove that $y_n > 0$ for all $n$, which implies that $Y_n \geq 0$. 
+The fact $y_n \geq 2^k$ for some $k$ implies that $y_n > 0$ for all $n$, which implies that $Y_n \geq 0$. 
 The other condition for $(Y_n)$ to be an antitone-strict supermartingale is

### 2021-04-30 LICS21/confluent.tex
Move figures back to main text, and make alignment tweaks.

--- 2021-04-29 LICS21/confluent.tex
@@
 Thus we can define a more general reduction relation\footnote{In this section $\to$ always means the relation defined in \cref{def:more general red}.} $\to$.
+
 \begin{definition}
@@
 \end{asparaenum}
+The $\sim_c$-rules are illustrated in \Cref{sim-diagram}.
 \end{definition}
@@
 
+\paragraph*{Reduction strategy determinising $\Rightarrow$}
 The reduction relation $\Rightarrow$ is nondeterministic, so it admits multiple possible reduction strategies. 
@@
 The proofs of \Cref{thm:rankable implies termination,thm:partial implies rankable,thm:antitone rankable implies termination,thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies.
-They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because the termination property implied by the type system applies equally to any reduction strategy).
+They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because \Cref{thm:de groote} applies equally to any reduction strategy).
 For a reduction strategy $r$ and term $M$, just substitute $N[r(N)]$ being a $\tY$-redex for $N$ being of the form $E[\tY \lambda x.
@@
 The only restrictions to bear in mind on evaluating sub-terms are that a $\tsample$ may not be evaluated inside a $\lambda$ that's inside a $\tY$ or on the right of an application, and a $\beta$ redex may not be reduced until its argument is a value.
+This can be seen as applied to \Cref{ex:sum of powers} at the end of \Cref{apx:confluent}.
 
@@
 It is still antitone rankable, but this is much more complicated.
+Details of the ranking function construction with and without using the confluent semantics are given in \Cref{apx:confluent}.
 \end{example}

### 2021-04-30 LICS21/ranking-proofs.tex
Move figures back to main text, and make alignment tweaks.

@@
 
-\begin{figure}
-\begin{tikzpicture}[node distance=0.5em, anchor=west]
-    \node (1) at (0,-0) {$\tY\, \Theta\, \underline 0$};
-    \node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, \underline i$};
-    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})})\, \underline i$};
-    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (5) at (0,-4) {$\tif{\underline r\, \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (6) at (0,-5) {$\tif{\underline {r - 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
-    \node (7) at (2,-6.3) [text width=] {$\underline i$};
-    \node (8) at (0,-7) {$\tY\, \Theta\, (\underline i\, \underline{+ 1})$};
-    \node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, (\underline i\, \underline{+ 1})$};
-
-    \draw [->] (1.south west) to [bend right=45, "$i := 0$"'] (2.north west);
-    \draw [->] (2.south west) to [bend right=45] (3.north west);
-    \draw [->] (3.south west) to [bend right=45] (4.north west);
-    \draw [->] (4.south west) to [bend right=45] (5.north west);
-    \draw [->] (5.south west) to [bend right=45] (6.north west);
-    \draw [->] (6.south west) to [bend right=45, "$r \geq 0.5$"] (8.north west);
-    \draw [->] (8.south west) to [bend right=45] (9.north west);
-
-    \draw [->] (9.west) to [bend left=30, "$i := i + 1$" {below,sloped}] (2.west);
-    \draw [->] (6) to ["$r < 0.5$"] (7);
-\end{tikzpicture}
-\caption{The reachable terms of $\tY \, \Theta \, \underline 0$\label{fig:geometric distribution}}
-\end{figure}
-
 \PartialImpliesRankable*

### 2021-04-30 LICS21/ranking.tex
Remove appendices
Move figures back to main text, and make alignment tweaks.

@@
 Despite its simplicity, its $\mathit{Rch}$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.
-%\lo{TODO: reduce the figure to avoid overflow.}\akr{It still overflows the gutter, but not the actual text of the other column.}\lo{OK. Thanks.}
+
+\begin{figure}[h]
+\begin{tikzpicture}[node distance=0.5em, anchor=west]
+    \node (1) at (0,-0) {$\tY\, \Theta\, \underline 0$};
+    \node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, \underline i$};
+    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})})\, \underline i$};
+    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
+    \node (5) at (0,-4) {$\tif{\underline r\, \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
+    \node (6) at (0,-5) {$\tif{\underline {r - 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
+    \node (7) at (2,-6.3) [text width=] {$\underline i$};
+    \node (8) at (0,-7) {$\tY\, \Theta\, (\underline i\, \underline{+ 1})$};
+    \node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, (\underline i\, \underline{+ 1})$};
+
+    \draw [->] (1.south west) to [bend right=45, "$i := 0$"'] (2.north west);
+    \draw [->] (2.south west) to [bend right=45] (3.north west);
+    \draw [->] (3.south west) to [bend right=45] (4.north west);
+    \draw [->] (4.south west) to [bend right=45] (5.north west);
+    \draw [->] (5.south west) to [bend right=45] (6.north west);
+    \draw [->] (6.south west) to [bend right=45, "$r \geq 0.5$"] (8.north west);
+    \draw [->] (8.south west) to [bend right=45] (9.north west);
+
+    \draw [->] (9.west) to [bend left=30, "$i := i + 1$" {below,sloped}] (2.west);
+    \draw [->] (6) to ["$r < 0.5$"] (7);
+\end{tikzpicture}
+\caption{The reachable terms of $\tY \, \Theta \, \underline 0$\label{fig:geometric distribution}}
+\end{figure}
 

### 2021-09-19 LICS21/intro.tex
Style/commenting.sty: draftnote -> draftnotee (to avoid clash with msc.sty) added: /userpackage{savesym} to get around clash with "comment"

--- 2021-04-29 LICS21/intro.tex
@@
 More recently, in probabilistic programming \cite{DBLP:conf/icse/GordonHNR14,rainforth2017Automating,vandemeent2018Introduction}, probabilistic programs are used to express generative models whose posterior probability can be computed by general purpose inference engines.
-Sampling from continuous distributions is an essential construct of probabilistic programming languages, such as Church \cite{DBLP:conf/uai/GoodmanMRBT08}, Stan \cite{carpenter2017stan}, Anglican \cite{DBLP:conf/pkdd/TolpinMW15}, Gen \cite{cusumano-towner2019Gen}, Pyro \cite{bingham2019Pyro}, Edward \cite{tran2016edward} and Turing \cite{ge2018Turing}.
+Sampling from continuous distributions is an essential construct of probabilistic programming languages, such as Church \citep{DBLP:conf/uai/GoodmanMRBT08}, Stan \cite{carpenter2017stan}, Anglican \cite{DBLP:conf/pkdd/TolpinMW15}, Gen \cite{cusumano-towner2019Gen}, Pyro \cite{bingham2019Pyro}, Edward \cite{tran2016edward} and Turing \cite{ge2018Turing}.
 Another key feature is higher order; in fact some of the most influential probabilistic programming languages are functional.

### 2021-09-19 LICS21/intro.tex
Change some \cite to \citep - ongoing

--- 2021-04-29 LICS21/intro.tex
@@
 % !TEX root = main.tex
+
+\lo{CONVENTION
+
+- Use {\tt citep} to render ``(\emph{author}, \emph{year})'' as opposed to ``\emph{author} (\emph{year})''}
 
@@
 Probabilistic (or randomised) programs have long been recognised as essential to the efficient solution of many algorithmic problems \cite{Rabin1976}.
-More recently, in probabilistic programming \cite{DBLP:conf/icse/GordonHNR14,rainforth2017Automating,vandemeent2018Introduction}, probabilistic programs are used to express generative models whose posterior probability can be computed by general purpose inference engines.
-Sampling from continuous distributions is an essential construct of probabilistic programming languages, such as Church \cite{DBLP:conf/uai/GoodmanMRBT08}, Stan \cite{carpenter2017stan}, Anglican \cite{DBLP:conf/pkdd/TolpinMW15}, Gen \cite{cusumano-towner2019Gen}, Pyro \cite{bingham2019Pyro}, Edward \cite{tran2016edward} and Turing \cite{ge2018Turing}.
+More recently, in probabilistic programming \citep{DBLP:conf/icse/GordonHNR14,rainforth2017Automating,vandemeent2018Introduction}, probabilistic programs are used to express generative models whose posterior probability can be computed by general purpose inference engines.
+Sampling from continuous distributions is an essential construct of probabilistic programming languages, such as Church \citep{DBLP:conf/uai/GoodmanMRBT08}, Stan \citep{carpenter2017stan}, Anglican \cite{DBLP:conf/pkdd/TolpinMW15}, Gen \cite{cusumano-towner2019Gen}, Pyro \cite{bingham2019Pyro}, Edward \cite{tran2016edward} and Turing \cite{ge2018Turing}.
 Another key feature is higher order; in fact some of the most influential probabilistic programming languages are functional.
@@
 the AST problem for higher-order programs with continuous distributions (and suitable primitive functions) is shown to have the same complexity.
-Though the problem of proving AST of imperative probabilistic programs is much studied \cite{DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05,DBLP:conf/popl/FioritiH15,DBLP:conf/pldi/ChenH20,DBLP:journals/toplas/ChatterjeeFNH18}, to our knowledge the problem of proving AST of higher-order programs with continuous distributions is new.
+Though the problem of proving AST of imperative probabilistic programs is much studied \citet{DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05,DBLP:conf/popl/FioritiH15,DBLP:conf/pldi/ChenH20,DBLP:journals/toplas/ChatterjeeFNH18}, to our knowledge the problem of proving AST of higher-order programs with continuous distributions is new.
 

### 2021-09-19 LICS21/intro.tex
Removed \onecolumn and \twocolumn - seems to compile now

@@
 
-- Use {\tt citep} to render ``(\emph{author}, \emph{year})'' as opposed to ``\emph{author} (\emph{year})''}
+- Use {\tt citep} to render ``(\emph{author}, \emph{year})''; default {\tt cite} is ``\emph{author} (\emph{year})''}
 
@@
 We introduce a novel addressing scheme for the possible random choices in a program's execution, which ensures that the same random choices are taken at corresponding positions in alternative reduction sequences, so that the same eventual result can be reached.
-This is then used to prove yet another extension to the ranking function theorem, that (\cref{thm:confluent ranking}) ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and (\cref{cor:Reduction strategy independence}) rankability in this sense still imples almost sure termination.
+This is then used to prove yet another extension to the ranking function theorem, that (\cref{thm:confluent ranking}) \lo{Preceding reference generated is wrong; it should be a theorem} ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and (\cref{cor:Reduction strategy independence}) rankability in this sense still imples almost sure termination.
 The confluent trace semantics is of wider interest, and has other possible applications as well, for example in Bayesian inference algorithms.

### 2021-09-22 LICS21/SPCF.tex
Remove appendices
Move figures back to main text, and make alignment tweaks.
Moving some proofs back to main text: ongoing

@@
 The set of all terms is denoted $\Lambda$, and the set of closed terms is denoted $\Lambda^0$.
+
+\input{apx-typing-rules}
 
@@
 
-\changed[lo]{The one-step reduction of the \emph{trace-based} (or \emph{sampling-style}) \emph{operational semantics} \cite{DBLP:journals/jcss/Kozen81,DBLP:conf/icfp/BorgstromLGS16,DBLP:conf/esop/CulpepperC17}} is given by the function $\red : \Lambda^0 \times \entrosp \to \Lambda^0 \times \entrosp$ where
+\changed[lo]{The one-step reduction of the \emph{trace-based} (or \emph{sampling-style}) \emph{operational semantics} \citep{DBLP:journals/jcss/Kozen81,DBLP:conf/icfp/BorgstromLGS16,DBLP:conf/esop/CulpepperC17}} is given by the function $\red : \Lambda^0 \times \entrosp \to \Lambda^0 \times \entrosp$ where
 \begin{equation*}

### 2021-09-22 LICS21/antitone.tex
Moving some proofs back to main text: ongoing

--- 2021-04-26 LICS21/antitone.tex
@@
 Moreover $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
-Therefore by setting $\epsilon_2(x) = \frac 1 {(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met.
+Therefore by setting $\epsilon_2(x) = 1 / {(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met.
 It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone sparse ranking function, and $M_2$ is AST.
@@
 The following theorem does not prove completeness, but is suggestive in that direction:
-\begin{restatable}{theorem}{antitoneSeemsComplete}
+%\begin{restatable}{theorem}{antitoneSeemsComplete}
+\begin{theorem}
+\label{thm:towards completeness}
 For any stopping time $T$ which is almost surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
-\end{restatable}
+\end{theorem}
+%\end{restatable}
+
+\begin{proof}
+
+If $T$ is bounded by $b$ a.s.~(for $n$ constant), the statement is trivially true by taking $\epsilon$ to be constantly 1, and $Y_n = b - n$.
+Otherwise, let $(t_n)_{n \geq 0}$ be defined recursively such that 
+\begin{align*}
+t_0 &= 0\\
+t_{n+1} &> t_n\\
+\mathbb P[T > t_{n+1}] &\leq \frac 1 4 \mathbb P[T > t_n].
+\end{align*}
+We will then define a supermartingale $(Y_n)_{n \geq 0}$ and nonrandom sequence $(y_n)_{n \geq 0}$ such that $Y_n = 0$ iff $T < n$, $Y_n = y_n$ iff $T \geq n$, and $y_n \geq 2^k$ for $n \geq t_k$.
+
+The antitone function $\epsilon$ is defined piecewise and recursively in such a way as to force all the necessary constraints to hold:
+\begin{align*}
+    \text{for }x \in [0,1),\ &\epsilon(x) = 1 \\
+    \text{for }x \in [2^k,2^{k+1}),\ &\epsilon(x) = \min \left (\epsilon(2^{k-1}), \frac{2^k}{t_{k+1}-t_k} \right ).
+\end{align*}
+This ensures that
+\begin{itemize}
+\item $\epsilon$ is weakly decreasing. 
+\item $\epsilon$ is bounded below by 0.
+\item For $x \geq 2^k$, $\epsilon(x) \leq \frac{2^k}{t_{k+1}-t_k}$. 
+\end{itemize}
+
+The sequence $(y_n)_{n \geq 0}$ is then defined recursively by
+\begin{align*}
+    y_0 = 2, \quad
+    y_{n+1} = (y_n - \epsilon(y_n)) \frac{\mathbb P[T > n]}{\mathbb P[T > n+1]}.
+\end{align*}
+
+The fact $y_{t_k} \geq 2^{k+1}$ is proven by induction on $k$.
+For base case, $2 \geq 2$, as required.
+For the inductive case $k+1$, we first do another induction to prove that $y_n \geq 2^k$ for all $t_k \leq n \leq t_{k+1}$.
+The base case of this inner induction follows from the outer induction hypothesis a fortiori.
+Take the greatest $k$ such that $n > t_k$.
+By the induction hypothesis, $y_{t_k} \geq 2^k$.
+
+\begin{align*}
+    y_n & = y_{t_k} \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} - \sum_{m=t_k}^{n-1} \epsilon(y_m) \frac{\mathbb P[T > m+1]}{\mathbb P[T > n]} \\
+    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \sum_{m=t_k}^{n-1} \epsilon(y_m)) \\
+    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \sum_{m=t_k}^{n-1} \frac{2^k}{t_{k+1}-t_k}) \\
+%\end{align*}
+%Allow a page break so that the sim-diagram lines up neatly.
+%\begin{align*}
+    & = \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \frac{2^k (n - t_k)}{t_{k+1}-t_k}) \\
+    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - 2^k) \\
+    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (2^{k+1} - 2^k) \\
+    & = \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} 2^k \\
+    & \geq 2^k
+\end{align*}
+as required. Substituting $n = t_{k+1}$ and using the fact that $\frac{\mathbb P[T > t_k]}{\mathbb P[T > t_{k+1}]} > 4$, the same reasoning gives $y_{t_{k+1}} \geq 2^{k+2}$, as required.
+
+The fact $y_n \geq 2^k$ for some $k$ implies that $y_n > 0$ for all $n$, which implies that $Y_n \geq 0$. 
+The other condition for $(Y_n)$ to be an antitone-strict supermartingale is
+\begin{align*}
+    \expect{Y_{n+1} \mid \mathcal F_n} 
+    &=  \expect{y_{n+1} \, {\bf 1}_{\set{T \geq n+1}} \mid \mathcal F_n} \\
+    &=  y_{n+1} \, \expect{{\bf 1}_{\set{T \geq n+1}} \mid \mathcal F_n} \\ 
+    &=  y_{n+1} \, {\bf 1}_{\set{T \geq n}} \textstyle \frac{\mathbb P[T > n+1]}{\mathbb P[T > n]} \\
+    &=  (y_n - \epsilon(y_n)) \, {\bf 1}_{\set{T \geq n}} \\
+    &= Y_n - \epsilon(y_n) \, {\bf 1}_{\set{T \geq n}}
+\end{align*}
+therefore $(Y_n)_n$ is an antitone strict supermartingale with respect to $(T,\epsilon)$.
+\end{proof}
+
+The assumption that $(\mathcal F_n)_{n \geq 0}$ is the coarsest filtration to which $T$ is adapted is used in the equality between the third and fourth lines here.
+This condition is the main reason that this proof does not extend directly to a completeness result for antitone ranking functions.

### 2021-09-22 LICS21/supermartingales.tex
Remove appendices
Moving some proofs back to main text: ongoing

@@
 \begin{remark}
-Our notion of ranking supermartingale is a slight generalisation of the original definition \cite{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, \changed[lo]{which does not involve an arbitrary stopping time}.
+Our notion of ranking supermartingale is a slight generalisation of the original definition \citep{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, which does not involve an arbitrary stopping time.
 %: Fioriti and Hermanns call a supermartingale $(Y_n)_{n \geq 0}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot {\bf 1}_{\set{Y_n > 0}}$.
@@
 \label{lem:rank-PAST}
-Let $(Y_n)_{n \geq 0}$ be a $\epsilon$-ranking supermartingale w.r.t.~the stopping time $T$, then $T < \infty$ a.s., and $\expect{T} \leq \frac{\expect{Y_0}}{\epsilon}$.
+Let $(Y_n)_{n \geq 0}$ be a $\epsilon$-ranking supermartingale w.r.t.~the stopping time $T$, then $T < \infty$ a.s., and %$\expect{T} \leq \frac{\expect{Y_0}}{\epsilon}$.
+$\expect{T} \leq {\expect{Y_0}}/{\epsilon}$.
 %\end{lemma}
@@
 \begin{remark}
-\Cref{lem:rank-PAST} is a slight extension of \cite[Lem.~5.5]{DBLP:conf/popl/FioritiH15},
+Lem.~\ref{lem:rank-PAST} is a slight extension of \cite[Lem.~5.5]{DBLP:conf/popl/FioritiH15},
 which asserts the same result but for the specific stopping time $T : \omega \mapsto \min\set{n \mid Y_n(\omega) = 0}$.
@@
 %(see, e.g., \cite[\S 6.7]{AshDD00} and \cite[Thm.~7.2]{DBLP:conf/popl/FioritiH15}).
-(see e.g.~\cite{AshDD00,DBLP:conf/popl/FioritiH15}).
+(see e.g.~\citep{AshDD00,DBLP:conf/popl/FioritiH15}).
 \begin{theorem}[Optional Sampling]
@@
 
+%\cref{thm:rankable implies termination}
+
 \begin{proof}

### 2021-09-27 LICS21/SPCF.tex
Fix citations and references

@@
 
-The language PPCF is a call-by-value (CBV) version of PCF with sampling of real numbers from the closed interval $[0,1]$ \cite{Ehrhard2018c,DBLP:journals/pacmpl/EhrhardPT18,DBLP:conf/esop/MakOPW21}.
+The language PPCF is a call-by-value (CBV) version of PCF with sampling of real numbers from the closed interval $[0,1]$ \citep{Ehrhard2018c,DBLP:journals/pacmpl/EhrhardPT18,DBLP:conf/esop/MakOPW21}.
 %\changed[lo]{following \cite{DBLP:conf/esop/MakOPW21} except the score function (\Cref{rem:score})}. 
@@
 
-The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid \hbox{$M$ terminates at $s$}\}$ is measurable \cite{DBLP:conf/icfp/BorgstromLGS16,DBLP:conf/esop/MakOPW21}.
+The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid \hbox{$M$ terminates at $s$}\}$ is measurable \citep{DBLP:conf/icfp/BorgstromLGS16,DBLP:conf/esop/MakOPW21}.
 We say that a term $M$ is \emph{almost surely terminating} (\emph{AST}) just if $\mu_{\entrosp}(\{s \mid \hbox{$M$ terminates at $s$}\}) = 1$.

### 2021-09-27 LICS21/antitone-proofs.tex
Moving some proofs back to main text: ongoing
Fix citations and references

@@
 
-\subsection*{\Cref{ex:Fair-in-the-limit random walk} (Fair-in-the-limit random walk) \cite[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}}
+\subsection*{\Cref{ex:Fair-in-the-limit random walk} (Fair-in-the-limit random walk) \citep[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}}
 \[
@@
 
-Many of the recent advances in the development of AST verification methods \cite{DBLP:conf/pldi/ChenH20,DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18,DBLP:conf/aplas/HuangFC18,DBLP:conf/popl/ChatterjeeNZ17,DBLP:journals/pacmpl/AgrawalC018,DBLP:conf/cav/ChatterjeeFG16,DBLP:conf/lics/OlmedoKKM16,DBLP:journals/pacmpl/Huang0CG19} are concerned with loop-based programs.
+Many of the recent advances in the development of AST verification methods \citep{DBLP:conf/pldi/ChenH20,DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18,DBLP:conf/aplas/HuangFC18,DBLP:conf/popl/ChatterjeeNZ17,DBLP:journals/pacmpl/AgrawalC018,DBLP:conf/cav/ChatterjeeFG16,DBLP:conf/lics/OlmedoKKM16,DBLP:journals/pacmpl/Huang0CG19} are concerned with loop-based programs.
 We can view such loops as tail-recursive programs that are, in particular, \emph{affine recursive}, i.e., 
@@
 Methods such as those presented in \cite{DBLP:journals/toplas/LagoG19} are explicitly restricted to affine programs, and are unsound otherwise.
-By contrast, many probabilistic programming languages allow for richer recursive structures \cite{DBLP:conf/pkdd/TolpinMW15,DBLP:conf/uai/GoodmanMRBT08,DBLP:journals/corr/MansinghkaSP14}.
+By contrast, many probabilistic programming languages allow for richer recursive structures \citep{DBLP:conf/pkdd/TolpinMW15,DBLP:conf/uai/GoodmanMRBT08,DBLP:journals/corr/MansinghkaSP14}.
 
@@
 Recently \cite{DBLP:conf/lics/KobayashiLG19} have extended HORS to \emph{probabilistic higher-order recursion schemes} (PHORS), which are HORS augmented with probabilistic (binary) branching $\oplus_p$.
-As HORS are in essence the $\lambda \tY$-calculus \cite{DBLP:conf/lics/Statman02} (i.e.~pure simply-typed lambda calculus with recursion, generated from a finite base type), PHORS are definable in PPCF:
+As HORS are in essence the $\lambda \tY$-calculus \citep{DBLP:conf/lics/Statman02} (i.e.~pure simply-typed lambda calculus with recursion, generated from a finite base type), PHORS are definable in PPCF:
 (order-$n$) PHORS is encodable as (order-$n$) call-by-name PPCF, but the former is strictly less expressive (because the underlying HORS is not Turing complete). 
@@
 
-A relevant result here is that the AST problem is decidable for order-1 PHORS (by reduction to the PSPACE-hard solvability of finite systems of polynomial equations with real coefficients; see \cite{DBLP:journals/jacm/EtessamiY09}), but undecidable for PHORS with order greater than 1 \cite{DBLP:conf/lics/KobayashiLG19}.
+A relevant result here is that the AST problem is decidable for order-1 PHORS (by reduction to the PSPACE-hard solvability of finite systems of polynomial equations with real coefficients; see \cite{DBLP:journals/jacm/EtessamiY09}), but undecidable for PHORS with order greater than 1 \citep{DBLP:conf/lics/KobayashiLG19}.
 
@@
 
+\iffalse
 \antitoneSeemsComplete*
@@
 This condition is the main reason that this proof does not extend directly to a completeness result for antitone ranking functions.
+\fi

### 2021-09-27 LICS21/antitone.tex
Fix citations and references

@@
 Sampling from continuous distributions is an essential feature of statistical probabilistic programming languages.
-(See e.g.~Church \cite{DBLP:conf/uai/GoodmanMRBT08}, Stan \cite{carpenter2017stan}, Anglican \cite{DBLP:conf/pkdd/TolpinMW15}, Gen \cite{cusumano-towner2019Gen}, Pyro \cite{bingham2019Pyro}, Edward \cite{tran2016edward} and Turing \cite{ge2018Turing}.)
+(See e.g.~Church \citep{DBLP:conf/uai/GoodmanMRBT08}, Stan \citep{carpenter2017stan}, Anglican \citep{DBLP:conf/pkdd/TolpinMW15}, Gen \citep{cusumano-towner2019Gen}, Pyro \citep{bingham2019Pyro}, Edward \citep{tran2016edward} and Turing \citep{ge2018Turing}.)
 Methods of proving AST of probabilistic computation have been developed for probabilistic programs with discrete distributions (see e.g.~\cite{DBLP:journals/toplas/LagoG19,DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/lics/KobayashiLG19,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05}).

### 2021-09-27 LICS21/applications.tex
Fix citations and references

@@
 If the labels of samples in the trace correspond closely to the roles that the samples play in the program execution, this will increase the efficiency of the inference algorithm.
-In \cite{wingate2011lightweight}, for example, something rather like a stack trace is used, although this addressing scheme turns out to have various problems \cite{kiselyov2016problems,Hur2015a}. 
+In \cite{wingate2011lightweight}, for example, something rather like a stack trace is used, although this addressing scheme turns out to have various problems \citep{kiselyov2016problems,Hur2015a}. 
 Potential positions could also be used for a similar purpose, although in practice, a more compact representation than an entire skeletal reduction sequence would be necessary. 
@@
 SPCF is a much nicer probabilistic programming language (in purified / idealised form).}
-Selecting an appropriate addressing scheme is also important for variational inference \cite{paige2016automatic}.
+Selecting an appropriate addressing scheme is also important for variational inference \citep{paige2016automatic}.
 %\cite[\S 6.2.2]{paige2016automatic}.

### 2021-09-27 LICS21/confluent-proofs.tex
Move figures back to main text, and make alignment tweaks.
Removed \onecolumn and \twocolumn - seems to compile now
Fix citations and references

@@
 If a reduction sequence includes one branch of one of these cases at some point in it, then the positions in the reduction sequence obtained by substituting in the other branch are all related by $\sim_c$ to the corresponding positions in the original reduction sequence.
+
+\include{sim-diagram}
 
@@
 Case a is also similar, except that the relevant positions are all disjoint rather than contained in one another, and either branch could be considered the right branch.
-
-\include{sim-diagram}
-%The diagram is not actually in the figure environment because that doesn't work properly with it extending over several pages. The empty figure is just needed to get the label and caption.
-\begin{figure}[htb]
-\caption{Illustration of the $\sim_c$ rules. \label{sim-diagram}
-}
-\end{figure}
 
@@
 Suppose for induction that $A \sim_\uparrow^k (D,\delta) \sim_\downarrow^* (E,\epsilon) \sim_\uparrow^* C$ for some $D, \delta, E, \epsilon$, and $0 \leq k \leq n$, where in the reduction sequence $D \to^* E$, for any pair of reductions at positions which aren't disjoint, the reduction at the innermost (greater) position occurs first.
-Reduction sequences with this property will be called ``parallel'', as it is analagous to the parallel reduction introduced by Tait and Martin-L\"of in their proof of the Church-Rosser theorem \cite{Bar84,Takahashi1995}.
+Reduction sequences with this property will be called ``parallel'', as it is analagous to the parallel reduction introduced by Tait and Martin-L\"of in their proof of the Church-Rosser theorem \citep{Bar84,Takahashi1995}.
 This induction is in reverse, with $k$ decreasing from $n$ to $0$.
@@
 \end{proof}
-\clearpage
 

### 2021-09-27 LICS21/intro.tex
Fix citations and references

@@
 
-Probabilistic (or randomised) programs have long been recognised as essential to the efficient solution of many algorithmic problems \cite{Rabin1976}.
+Probabilistic (or randomised) programs have long been recognised as essential to the efficient solution of many algorithmic problems \citep{Rabin1976}.
 More recently, in probabilistic programming \citep{DBLP:conf/icse/GordonHNR14,rainforth2017Automating,vandemeent2018Introduction}, probabilistic programs are used to express generative models whose posterior probability can be computed by general purpose inference engines.
-Sampling from continuous distributions is an essential construct of probabilistic programming languages, such as Church \citep{DBLP:conf/uai/GoodmanMRBT08}, Stan \citep{carpenter2017stan}, Anglican \cite{DBLP:conf/pkdd/TolpinMW15}, Gen \cite{cusumano-towner2019Gen}, Pyro \cite{bingham2019Pyro}, Edward \cite{tran2016edward} and Turing \cite{ge2018Turing}.
+Sampling from continuous distributions is an essential construct of probabilistic programming languages, such as Church \citep{DBLP:conf/uai/GoodmanMRBT08}, Stan \citep{carpenter2017stan}, Anglican \citep{DBLP:conf/pkdd/TolpinMW15}, Gen \citep{cusumano-towner2019Gen}, Pyro \citep{bingham2019Pyro}, Edward \citep{tran2016edward} and Turing \citep{ge2018Turing}.
 Another key feature is higher order; in fact some of the most influential probabilistic programming languages are functional.
@@
 When a probabilistic program implements a solution to an algorithmic problem, we naturally require the computation to terminate with probability $1$, in which case the program is called \emph{almost surely terminating} (AST).
-Indeed, it is standard for designers and implementors of probabilistic programming systems to regard non-AST programs as defining invalid models, and hence inadmissible \cite{rainforth2017Automating,DBLP:conf/uai/GoodmanMRBT08}.
+Indeed, it is standard for designers and implementors of probabilistic programming systems to regard non-AST programs as defining invalid models, and hence inadmissible \citep{rainforth2017Automating,DBLP:conf/uai/GoodmanMRBT08}.
 (Yet none of these systems provides any support for the development or verification of AST programs.)
@@
 Moreover various theorems about probabilistic programs rely on the assumption that the program terminates almost surely (see e.g.~\cite{DBLP:conf/uai/GoodmanMRBT08}). 
-A recent result \cite{DBLP:conf/esop/MakOPW21} proves that AST programs have density functions that are differentiable almost everywhere.
-This is significant for Bayesian machine learning, because almost everywhere differentiability is a precondition for the correctness of some of the most scalable inference algorithms, such as Hamiltonian Monte Carlo \cite{DBLP:conf/aistats/ZhouGKRYW19,Nishimura2020b} and reparameterised gradient variational inference \cite{DBLP:conf/nips/LeeYY18}.
+A recent result \citep{DBLP:conf/esop/MakOPW21} proves that AST programs have density functions that are differentiable almost everywhere.
+This is significant for Bayesian machine learning, because almost everywhere differentiability is a precondition for the correctness of some of the most scalable inference algorithms, such as Hamiltonian Monte Carlo \citep{DBLP:conf/aistats/ZhouGKRYW19,Nishimura2020b} and reparameterised gradient variational inference \citep{DBLP:conf/nips/LeeYY18}.
 
 The AST problem is not just important, but also difficult: 
-Deciding AST of first-order imperative programs with discrete probabilities is $\Pi^0_2$-complete \cite{DBLP:conf/mfcs/KaminskiK15}.
-In recent work \cite{BeutnerO21} %(under double-blind review elsewhere), 
+Deciding AST of first-order imperative programs with discrete probabilities is $\Pi^0_2$-complete \citep{DBLP:conf/mfcs/KaminskiK15}.
+In recent work \citep{BeutnerO21} %(under double-blind review elsewhere), 
 the AST problem for higher-order programs with continuous distributions (and suitable primitive functions) is shown to have the same complexity.
@@
 In other words, the program's behavior is used to define an associated (ranking) \emph{supermartingale}.
-Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \cite{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18}.
+Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \citep{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18}.
 This paper's contribution is to extend this method to a higher-order setting.
@@
 We define a \emph{ranking function} on a term $M$ to be a non-negative measurable function on the reachable terms from $M$ whose expected value decreases or is unchanged as the term is reduced.
-Because the type system already constrains the terms enough to force termination in the absence of the recursion construct \cite{tait1967,BarendregtDS2010}, it is only the $\tY$-reduction steps that must be counted, making defining ranking functions somewhat easier.
+Because the type system already constrains the terms enough to force termination in the absence of the recursion construct \citep{tait1967,BarendregtDS2010}, it is only the $\tY$-reduction steps that must be counted, making defining ranking functions somewhat easier.
 Using supermartingales and Doob's (iterated) Optional Sampling Theorem, we show the soundness of rankability (\cref{thm:rankable implies termination}): if a term has a ranking function, it is AST.
@@
 moreover it also enjoys a sparse function theorem (\cref{thm:antitone partial implies rankable}).
-\akr{Our progress condition is equivalent to that given by McIver et al.~\cite{DBLP:journals/pacmpl/McIverMKK18} (by applying a transformation to the ranking function), but needs to be different in order to be compatible with the presence of additional intermediate reduction steps where the ranking function is not required to decrease at all.}
+\akr{Our progress condition is equivalent to that given by \cite{DBLP:journals/pacmpl/McIverMKK18} (by applying a transformation to the ranking function), but needs to be different in order to be compatible with the presence of additional intermediate reduction steps where the ranking function is not required to decrease at all.}
 \lo{The same point is made in \cref{sec:related}, so the preceding can be omitted.}

### 2021-09-27 LICS21/related.tex
Fix citations and references

--- 2021-04-26 LICS21/related.tex
@@
 
-Most methods for proving AST of programs in the literature do not support continuous distributions \cite{DBLP:journals/toplas/LagoG19,DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/lics/KobayashiLG19,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05}.
-The few that do \cite{DBLP:conf/popl/FioritiH15,DBLP:conf/pldi/ChenH20,DBLP:journals/toplas/ChatterjeeFNH18} are for first-order imperative programs.
+Most methods for proving AST of programs in the literature do not support continuous distributions \citep{DBLP:journals/toplas/LagoG19,DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/lics/KobayashiLG19,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05}.
+The few that do \citep{DBLP:conf/popl/FioritiH15,DBLP:conf/pldi/ChenH20,DBLP:journals/toplas/ChatterjeeFNH18} are for first-order imperative programs.
 
 A fortiori, our results specialise to a method for proving termination of non-random lambda calculus. 
-Thus restricted, our approach is closest in spirit to Jones' work on size-change termination \cite{DBLP:journals/lmcs/JonesB08,DBLP:conf/aplas/SereniJ05}. 
-Dal Lago and Grellois \cite{DBLP:journals/toplas/LagoG19} extend this to programs with discrete distributions, but their method is limited to affine recursion of first-order functions.
+Thus restricted, our approach is closest in spirit to Jones' work on size-change termination \citep{DBLP:journals/lmcs/JonesB08,DBLP:conf/aplas/SereniJ05}. 
+\cite{DBLP:journals/toplas/LagoG19} extends this to programs with discrete distributions, but their method is limited to affine recursion of first-order functions.
 We think that the supermartingales of our paper are a good basis for giving a unifying account.
-Breuvart and Dal Lago \cite{DBLP:conf/ppdp/BreuvartL18} develop systems of intersection types from which the termination probability of programs (with discrete distributions) can be inferred from (infinitely many) type derivations.
+\cite{DBLP:conf/ppdp/BreuvartL18} develops systems of intersection types from which the termination probability of programs (with discrete distributions) can be inferred from (infinitely many) type derivations.
 \lo{Mention Dal Lago et al.
@@
 
-Kobayashi et al.~\cite{DBLP:conf/lics/KobayashiLG19} show that the termination probability of order-$n$ probabilistic recursion schemes ($n$-PHORS) can be obtained as the least fixpoint of suitable order-$(n-1)$ fixpoint equations, which are solvable by Kleene fixpoint iteration.
+\cite{DBLP:conf/lics/KobayashiLG19} shows that the termination probability of order-$n$ probabilistic recursion schemes ($n$-PHORS) can be obtained as the least fixpoint of suitable order-$(n-1)$ fixpoint equations, which are solvable by Kleene fixpoint iteration.
 By contrast, our approach works for programs with continuous distributions. 
@@
 
-The main result by McIver et al.~\cite{DBLP:journals/pacmpl/McIverMKK18} is similar to our \cref{thm:antitone rankable implies termination}, but in an imperative language.
+The main result of \cite{DBLP:journals/pacmpl/McIverMKK18} is similar to our \cref{thm:antitone rankable implies termination}, but in an imperative language.
 While we require that the antitone ranking function decreases for every $\tY$-reduction step, they require that it decreases for every iteration of a certain while loop (the ranking function in that case is defined in the context of a particular loop), which is similar to $\tY$ but limited to tail recursion.

### 2021-10-01 LICS21/antitone-proofs.tex
Use \ref instead of \Cref.

@@
 
-\subsection*{\Cref{ex:raven complex} (Continuous random walk)}
+\subsection*{Example \ref{ex:raven complex} (Continuous random walk)}
 Let $\Theta := \tY \, \lambda f x . \tif{x \leq 0}{0}{f(x - \tsample)}$. 
@@
 
-\subsection*{\Cref{ex:Fair-in-the-limit random walk} (Fair-in-the-limit random walk) \citep[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}}
+\subsection*{Example \ref{ex:Fair-in-the-limit random walk} (Fair-in-the-limit random walk) \citep[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}}
 \[
@@
 
-\subsection*{\Cref{ex:non-affine recursion easy} (Non-affine recursion)}
+\subsection*{Example \ref{ex:non-affine recursion easy} (Non-affine recursion)}
 Let 
@@
 
-\subsection*{\Cref{ex:non-affine continuous} (More complex non-affine recursion)}
+\subsection*{Example \ref{ex:non-affine continuous} (More complex non-affine recursion)}
 Let
@@
 (order-$n$) PHORS is encodable as (order-$n$) call-by-name PPCF, but the former is strictly less expressive (because the underlying HORS is not Turing complete). 
-For example, the PPCF term in \Cref{ex:higher-order recursion} is not definable in PHORS.
+For example, the PPCF term in Example \ref{ex:higher-order recursion} is not definable in PHORS.
 %Some interesting PPCF terms such as \Cref{ex:non-affine with continuous distribution} %\refExample{complexExample} 
@@
 
-\subsection*{\Cref{ex:higher-order recursion} (Higher-order recursion)}
+\subsection*{Example \ref{ex:higher-order recursion} (Higher-order recursion)}
 Consider the higher-order function
@@
 \] 
-where $g_2$ is as in \Cref{ex:unbiased random walk}, using the same antitone function too.
+where $g_2$ is as in Example \ref{ex:unbiased random walk}, using the same antitone function too.
 
-The inequalities required for this to be an antitone partial supermartingale are satisfied with just the same reasoning as in \Cref{ex:unbiased random walk}, therefore this term too is antitone rankable.
+The inequalities required for this to be an antitone partial supermartingale are satisfied with just the same reasoning as in Example \ref{ex:unbiased random walk}, therefore this term too is antitone rankable.
 
 Again, this is not actually the correct reduction order, in that $F$ should be applied to its argument $n_i$ and reduced to a value before its second argument is expanded, 
-but this would complicate the presentation, and this version can be justified by \Cref{thm:confluent ranking} instead. 
+but this would complicate the presentation, and this version can be justified by Theorem \ref{thm:confluent ranking} instead. 
 The reduction strategy implied here should be clear enough (assuming that, where it isn't specified, it just matches $\cbv$), but to be more precise, 

### 2021-10-01 LICS21/antitone.tex
Use \ref instead of \Cref.

@@
 
-The term $M_1$ is rankable, terminating in 31 $\tY$-steps on average, and $M_3$ only has a $1/1024$ chance of terminating, but in between, $M_2$ is AST, but isn't $\tY$-PAST, therefore it isn't rankable and \Cref{thm:rankable implies termination} is insufficient to prove its termination.
+The term $M_1$ is rankable, terminating in 31 $\tY$-steps on average, and $M_3$ only has a $1/1024$ chance of terminating, but in between, $M_2$ is AST, but isn't $\tY$-PAST, therefore it isn't rankable and Theorem \ref{thm:rankable implies termination} is insufficient to prove its termination.
 Thus we seek a generalised notion of ranking function so that $M_2$ becomes rankable, and then prove it sound i.e.~rankable in this generalised sense implies AST.
@@
 As in %Theorem \ref{thm:basicRankingSoundness}, 
-\Cref{thm:rankable implies termination},
+Theorem \ref{thm:rankable implies termination},
 take the probability space $(\entrosp, \Sigma_\entrosp, \mu_\entrosp)$, and define the same r.v.~$T_n, X_n, Y_n, T_M^{\tY}$. 
-Thanks to \Cref{thm:rankable and strict rankable}, $(Y_n)_{n \geq 0}$ is a supermartingale; and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale w.r.t.~stopping time $T_M^{\tY}$.
-Thus, by \Cref{thm:a-c strict}, $T_M^{\tY} < \infty$ a.s.
+Thanks to Theorem \ref{thm:rankable and strict rankable}, $(Y_n)_{n \geq 0}$ is a supermartingale; and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale w.r.t.~stopping time $T_M^{\tY}$.
+Thus, by Theorem \ref{thm:a-c strict}, $T_M^{\tY} < \infty$ a.s.
 \end{proof}
@@
 
-\paragraph*{\Cref{ex:ac-ranking} revisited}
+\paragraph*{Example \ref{ex:ac-ranking} revisited}
 Programs $M_1$ and $M_2$ are AST. For $M_1$, the sparse ranking function 
@@
 % \end{align*}
-For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$,\footnote{Technically this is not quite correct, because of the distinction between $\underline n + 1$ and $\underline{n + 1}$, but this ranking function can still be made to work by appealing to the Confluent Ranking \Cref{thm:confluent ranking}, as can the others in this section with similar issues.
-See \cref{rem:ex corrected using confluent semantics} for a fuller account.} each with probability $1/2$, with one $\tY$-reduction.
+For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$,\footnote{Technically this is not quite correct, because of the distinction between $\underline n + 1$ and $\underline{n + 1}$, but this ranking function can still be made to work by appealing to the Confluent Ranking Theorem \ref{thm:confluent ranking}, as can the others in this section with similar issues.
+See Remark \ref{rem:ex corrected using confluent semantics} for a fuller account.} each with probability $1/2$, with one $\tY$-reduction.
 Now
@@
 \paragraph*{Completeness}
-It seems very likely that this method is complete in the case that every reachable term is AST (\cref{conj:antitone}), but we have been unable to actually prove this.
-It is certainly at least capable of proving the termination of terms which terminate arbitrarily slowly (in the sense that there is no similar limitation to \cref{thm:rankable implies termination}, which can only prove termination of $\tY$-PAST terms).
+It seems very likely that this method is complete in the case that every reachable term is AST (Conjecture \ref{conj:antitone}), but we have been unable to actually prove this.
+It is certainly at least capable of proving the termination of terms which terminate arbitrarily slowly (in the sense that there is no similar limitation to Theorem \ref{thm:rankable implies termination}, which can only prove termination of $\tY$-PAST terms).
 

### 2021-10-01 LICS21/applications.tex
Use \ref instead of \Cref.

@@
 \label{sec:applications}
-Although the definition of potential positions in a term in \Cref{sec:confluent} was intended simply to define a variant of trace semantics that has a restricted version of confluence, it is likely to also be applicable to inference algorithms (for probabilistic programs) as well.
+Although the definition of potential positions in a term in Section \ref{sec:confluent} was intended simply to define a variant of trace semantics that has a restricted version of confluence, it is likely to also be applicable to inference algorithms (for probabilistic programs) as well.
 
@@
 Another possible application for the confluent trace semantics would be as a justification, within a trace-style semantics, of some forms of equational reasoning. 
-By introducing reduction steps that go backwards to terms not reachable from the original term, it may in some cases be simplified, or equivalences between terms may be proved, by appealing to confluence results such as \Cref{churchRosser} and \Cref{thm:CbvIsTerminatingest}.
+By introducing reduction steps that go backwards to terms not reachable from the original term, it may in some cases be simplified, or equivalences between terms may be proved, by appealing to confluence results such as Lemma \ref{churchRosser} and Theorem \ref{thm:CbvIsTerminatingest}.
 \lo{Can you provide an example (and present it in the appendix)?} \akr{I don't have any compelling example for this in mind.}

### 2021-10-01 LICS21/confluent-proofs.tex
Use \ref instead of \Cref.

@@
 They are adjacent in the sequence, as the CBV order is a total order.
-Let the positions of these reductions be $\alpha$ and $\beta$ respecively.
+Let the positions of these reductions be $\alpha$ and $\beta$ respectively.
 If they are disjoint, they can be considered as one of the branches of $\sim_c$ case a.
@@
 \paragraph{}
-If $Y \sim_c X \sim_c^* X_{cbv}$, and the reductions involved in $Y \sim_c X$ are not the last pair of reductions in $Y$ that aren't in CBV order, the order of the $\sim_c$s can be rearranged so that the sequence $Y \sim_c^* X_{cbv}$ is the canonical sequence given above, therefore $Y_{cbv} = X_{cbv}$.
+If $Y \sim_c X \sim_c^* X_{cbv}$, and the reductions involved in $Y \sim_c X$ are not the last pair of reductions in $Y$ that aren't in CBV order, it will be shown that the order of the $\sim_c$s can be rearranged so that the sequence $Y \sim_c^* X_{cbv}$ is the canonical sequence given above, therefore $Y_{cbv} = X_{cbv}$.
 To be more precise about this rearrangement, consider the sequence of $\sim_c$ steps from $Y$ to $X_{cbv}$.
@@
 In the other direction, suppose $(N_1, \alpha_1) \sim_M^* (N_2, \alpha_2)$ where both $N_1$ and $N_2$ are descendants of $N$.
-By \Cref{downAcrossUp}, take $(N_1, \alpha_1) \sim_{p,M}^* (N'_1, \alpha') \sim_{c,M}^* (N'_2, \alpha') \sim_{p,M}^* (N_2, \alpha_2)$.
+By Lemma \ref{downAcrossUp}, take $(N_1, \alpha_1) \sim_{p,M}^* (N'_1, \alpha') \sim_{c,M}^* (N'_2, \alpha') \sim_{p,M}^* (N_2, \alpha_2)$.
 The $\sim_p^*$ steps remain within $\mathit{Rch}(N)$, and $\sim_p$ does not depend on the history of the reduction sequences, therefore $(N_1, \alpha_1) \sim_{p,N}^* (N'_1, \alpha') \sim_{c,M}^* (N'_2, \alpha') \sim_{p,N}^* (N_2, \alpha_2)$.
 
-Let $X$ be the call-by-value reduction sequence related to both $N'_1$ and $N'_2$ by $\sim_{c,M}$ given by \Cref{canonicalCousins}. 
-As $M \to N$ is the first reduction in the sequence $N'_1$, it is the last to be affected by the $\sim_{c,M}$ sequence $N'_1 \sim_{c,M}^* X$ given by \Cref{canonicalCousins} therefore it can be split into $N'_1 \sim_{c,M}^* Y_1 \sim_{c,M}^* X$ where $Y_1$ is in CBV order except possibly for its first reduction, which is still $M \to N$.
+Let $X$ be the call-by-value reduction sequence related to both $N'_1$ and $N'_2$ by $\sim_{c,M}$ given by Lemma \ref{canonicalCousins}. 
+As $M \to N$ is the first reduction in the sequence $N'_1$, it is the last to be affected by the $\sim_{c,M}$ sequence $N'_1 \sim_{c,M}^* X$ given by Lemma \ref{canonicalCousins} therefore it can be split into $N'_1 \sim_{c,M}^* Y_1 \sim_{c,M}^* X$ where $Y_1$ is in CBV order except possibly for its first reduction, which is still $M \to N$.
 Let the position of the reduction $M \to N$ be $\beta$.
@@
 In the case that $(M,s) \Rightarrow^* (M_1,s_1)$ or $(M,s) \Rightarrow^* (M_2,s_2)$ by multiple steps, it is possible to repeatedly replace a pair of the form $A \Leftarrow B \Rightarrow C$ by $A \Rightarrow^* D \Leftarrow^* C$ by the construction above, but it is not immediate that this process terminates, because each $\Rightarrow$ or $\Leftarrow$ may be replaced by multiple, so the sequence of $\Rightarrow$s and $\Leftarrow$s from $(M_1,s_1)$ to $(M_2,s_2)$ may get longer at some steps.
-However, termination can be proved by noting that the structure of this process is identical to the process of swapping $\sim_\uparrow$ and $\sim_\downarrow$ steps in \Cref{downAcrossUp}.
+However, termination can be proved by noting that the structure of this process is identical to the process of swapping $\sim_\uparrow$ and $\sim_\downarrow$ steps in Lemma \ref{downAcrossUp}.
 To be more precise, in this case there is a sequence of $\Leftarrow$ and $\Rightarrow$ steps, each of which has an associated reduction position and initial and final terms, and if a $\Leftarrow$ immediately precedes a $\Rightarrow$, they may be swapped to produce some number of $\Rightarrow$s, followed by some number of $\Leftarrow$s.
-In the case of \Cref{downAcrossUp}, there is a sequence of $\sim_\uparrow$ and $\sim_\downarrow$ steps, each of which has an associated reduction position and initial and final skeletons, and if a $\sim_\uparrow$ immediately precedes a $\sim_\downarrow$, they may be swapped to produce some number of $\sim_\downarrow$s followed by some number of $\sim_\uparrow$s.
+In the case of Lemma \ref{downAcrossUp}, there is a sequence of $\sim_\uparrow$ and $\sim_\downarrow$ steps, each of which has an associated reduction position and initial and final skeletons, and if a $\sim_\uparrow$ immediately precedes a $\sim_\downarrow$, they may be swapped to produce some number of $\sim_\downarrow$s followed by some number of $\sim_\uparrow$s.
 In both cases, the number and reduction positions of the resultant steps are determined by the case of $\sim_c$ that matches the way the initial reduction positions overlap, and the initial skeleton (or the skeleton of the initial term).
-The same argument that the process in \Cref{downAcrossUp} terminated is therefore applicable here too.
+The same argument that the process in Lemma \ref{downAcrossUp} terminated is therefore applicable here too.
 At every stage, the sequence of reductions that results from one of the initial $\Rightarrow$s is a parallel sequence, and swapping a parallel sequence of $\Rightarrow$s with a $\Leftarrow$ always terminates, therefore each $\Rightarrow$ in turn can be moved past all of the $\Leftarrow$s, and the process as a whole will terminate in a state where all of the $\Rightarrow$s precede all of the $\Leftarrow$s, i.e.~a pair of reduction sequences $(M_1,s_1) \Rightarrow^* (M',s') \Leftarrow^* (M_2,s_2)$.
@@
 \begin{proof}
-Suppose on the contrary that $(M,\alpha) \sim^* (A,\beta)$, where $N \to^* A$, then by \Cref{simpDoesn'tMix}, $(M,\alpha) \sim_p^* (A,\beta)$.
+Suppose on the contrary that $(M,\alpha) \sim^* (A,\beta)$, where $N \to^* A$, then by Lemma \ref{simpDoesn'tMix}, $(M,\alpha) \sim_p^* (A,\beta)$.
 $M \neq A$ therefore $(M,\alpha) \sim_p$ some position in $N$, but in all the cases of the definition of $\sim_p$, no position in the child term is related to the position of the redex.
@@
 
-Essentially what \Cref{lem:redexDestroyed} demonstrates is that the samples taken during any reduction sequence are independent of each other. This is made more precise in the following lemmas.
+Essentially what Lemma \ref{lem:redexDestroyed} demonstrates is that the samples taken during any reduction sequence are independent of each other. This is made more precise in the following lemmas.
 
@@
 The distribution of $s(k)$ conditional on reaching $N$ is therefore the uniform distribution on $I$.
-Similarly for the other semantics, if the distribution of remaining samples after $n$ steps is independent of the term by \Cref{lem:independentSamples}, therefore the distribution of $s(\mathit{Sk}(N),\cbv(N))$ conditional on $N$ is the uniform distribution on $I$.
+Similarly for the other semantics, if the distribution of remaining samples after $n$ steps is independent of the term by Lemma \ref{lem:independentSamples}, therefore the distribution of $s(\mathit{Sk}(N),\cbv(N))$ conditional on $N$ is the uniform distribution on $I$.
 In either case, the distribution of the next term conditional on the previous term, in the case that it reduces randomly, is equal to the image under $r \mapsto E[\underline r]$ of the uniform distribution on $I$.
@@
 Let $(M,s) = (M_0,s_0) \Rightarrow_f \dots \Rightarrow_f (M_n,s_n)$, where $M_n$ is a value.
-By \Cref{canonicalCousins}, the reduction sequence $(\mathit{Sk}(M_i))_i$ is related by $\sim_c^*$ to some skeletal reduction sequence $X$ which is in call-by-value order.
+By Lemma \ref{canonicalCousins}, the reduction sequence $(\mathit{Sk}(M_i))_i$ is related by $\sim_c^*$ to some skeletal reduction sequence $X$ which is in call-by-value order.
 For each skeletal reduction sequence $X_j$ in the sequence $(\mathit{Sk}(M_i))_i \sim_c \dots \sim_c X$, it is possible to define a reduction sequence of terms such that $(M,s) = (N_{j,0},s_{j,0}) \Rightarrow \dots \Rightarrow (N_{j,n_j},s_{j,n_j})$ and $\mathit{Sk}(N_{j,k}) = X_{j,k}$, by at each step applying the reduction at the same place as in $X_j$, essentially just filling in the reals in the skeletal reduction sequence to make it a reduction sequence of terms (it will be shown shortly that the correct branch is taken in each $\textsf{if}$-reduction).
@@
 To prove that this holds true for $j+1$ as well, it suffices to show that the corresponding term and samples immediately after the reductions involved in this $\sim_c$ step match, as from that point on, the $\Rightarrow$-sequences will match, as they start from the same point and have all the same reduction positions.
-Letting the term and remaining samples immediately before the reductions involved in this $\sim_c$ step be $(N,t)$ (which is the same in both $N_j$ and $N_{j+1}$ for the same reason again, they start at the same point and the sequences of reduction positions up to this point are the same), and the terms and samples after these reductions be $(O_1,t_1)$ and $(O_2,t_2)$, so that these match the $N$, $O_1$ and $O_2$ in the definition of $\sim_c$, $t_1 = t_2$ because both of them are, by \Cref{lem:sim-M-N}, equal to the restriction of $t$ to the subset of $\mathit{Rch}(N)$ that starts with the reduction sequences to $O_1$ and $O_2$ respectively, and the potential positions in those subsets of $\mathit{Rch}(N)$ are identified with each other by $\sim$ because of the same case of $\sim_c$ that means $N_j$ and $N_{j+1}$ are related.
+Letting the term and remaining samples immediately before the reductions involved in this $\sim_c$ step be $(N,t)$ (which is the same in both $N_j$ and $N_{j+1}$ for the same reason again, they start at the same point and the sequences of reduction positions up to this point are the same), and the terms and samples after these reductions be $(O_1,t_1)$ and $(O_2,t_2)$, so that these match the $N$, $O_1$ and $O_2$ in the definition of $\sim_c$, $t_1 = t_2$ because both of them are, by Lemma \ref{lem:sim-M-N}, equal to the restriction of $t$ to the subset of $\mathit{Rch}(N)$ that starts with the reduction sequences to $O_1$ and $O_2$ respectively, and the potential positions in those subsets of $\mathit{Rch}(N)$ are identified with each other by $\sim$ because of the same case of $\sim_c$ that means $N_j$ and $N_{j+1}$ are related.
 If none of the relevant reductions are $\tsample$-reductions, $O_1 = O_2$ trivially.
 If there are samples taken (which must be the arbitrary reduction at $\beta$), it is also required that the same samples are taken in both branches.
-Similarly to \Cref{churchRosser}, the positions of these $\tsample$-reductions in each branch are related by $\sim_p^*$, therefore they correspond to the same sample in $t$, therefore $O_1 = O_2$.
+Similarly to Lemma \ref{churchRosser}, the positions of these $\tsample$-reductions in each branch are related by $\sim_p^*$, therefore they correspond to the same sample in $t$, therefore $O_1 = O_2$.
 The fact that the terms are equal also implies that the same branch of an $\textsf{if}$ statements will be taken.
@@
 Although $T'$ itself may have a measure of 0 as a subset of $I^{L_s(M)}$, it still has a natural measure-space structure of its own.
-It also has the properties that $M$ does not terminate with $\cbv$ and any trace in $T'$ (therefore $M$ also does not terminate with $r$ and any trace in $T'$, by \Cref{thm:CbvIsTerminatingest}), and for any trace $t \in T'$, if $(M,t) \Rightarrow^k_r (M',t')$, none of the potential positions in $M'$ correspond to any of the elements of $P \setminus P_i$.
+It also has the properties that $M$ does not terminate with $\cbv$ and any trace in $T'$ (therefore $M$ also does not terminate with $r$ and any trace in $T'$, by Theorem \ref{thm:CbvIsTerminatingest}), and for any trace $t \in T'$, if $(M,t) \Rightarrow^k_r (M',t')$, none of the potential positions in $M'$ correspond to any of the elements of $P \setminus P_i$.
 There are finitely many skeletal reduction sequences $k$ steps long starting at $M$ and following $r$, each one corresponding to some measurable subset of $I^{L_s(M)}$, so pick one such that its intersection with $T'$ is a non-null subset of $T'$, let this subset be $Q$.
@@
 
-\Cref{ex:sum of powers}
+Example \ref{ex:sum of powers}
 Recall that

### 2021-10-01 LICS21/confluent.tex
Use \ref instead of \Cref.

@@
 or the number of cases necessary for defining a sparse ranking function is excessive because the reduction does not proceed in a neat and orderly manner. 
-In cases such as these (e.g.~\cref{ex:sum of powers}), it may be much more convenient to instead consider ranking functions with respect to alternative reduction strategies constructed to match the individual term in question,
+In cases such as these (e.g.
+Example \ref{ex:sum of powers}), it may be much more convenient to instead consider ranking functions with respect to alternative reduction strategies constructed to match the individual term in question,
 thereby reducing the number of cases needed and simplifying the reachable terms.
@@
 In order to allow potential positions in different reduction sequences to be considered equivalent, a relation $\sim^*$ is defined, and finally, all of these are used to construct a confluent version of the trace semantics, $\Rightarrow$, that is still nondeterministic in reduction order, but does specify the outcome of random choices.
-The desired properties of this semantics (confluence and equivalence to the standard semantics) are proved (\Cref{churchRosser,thm:AstEquivalence}), then it is used to give extended versions of ranking functions and the related AST results, \Cref{thm:confluent ranking}.
+The desired properties of this semantics (confluence and equivalence to the standard semantics) are proved (Lemma \ref{churchRosser} and Theorem \ref{thm:AstEquivalence}), then it is used to give extended versions of ranking functions and the related AST results, Theorem \ref{thm:confluent ranking}.
 
@@
 
-Thus we can define a more general reduction relation\footnote{In this section $\to$ always means the relation defined in \cref{def:more general red}.} $\to$.
+Thus we can define a more general reduction relation\footnote{In this section $\to$ always means the relation defined in Definition \ref{def:more general red}.} $\to$.
 
@@
 \end{asparaenum}
-The $\sim_c$-rules are illustrated in \Cref{sim-diagram}.
+The $\sim_c$-rules are illustrated in Figure \ref{sim-diagram}.
 \end{definition}
@@
 \begin{example}
-In \Cref{ex:cousin sim}, $(M,\underline{+}_1) \sim_p (M \to N_2,\underline{+}_1)$ by case i of $\sim_p$ (because the reduction $M \to N_2$ occurs at $\underline{+}_2$, which is disjoint from $\underline{+}_1$), and similarly, $(M,\underline{+}_2) \sim_p (M \to N_1,\underline{+}_2)$.
+In Example \ref{ex:cousin sim}, $(M,\underline{+}_1) \sim_p (M \to N_2,\underline{+}_1)$ by case i of $\sim_p$ (because the reduction $M \to N_2$ occurs at $\underline{+}_2$, which is disjoint from $\underline{+}_1$), and similarly, $(M,\underline{+}_2) \sim_p (M \to N_1,\underline{+}_2)$.
 
@@
 There are then reduction sequences $M_{sss} \to M_{Xss} \to M_{XXs} \to M_{XXX} \to \skeletonPlaceholder$ and $M_{sss} \to M_{sXs} \to M_{XXs} \to M_{XXX} \to \skeletonPlaceholder$.
-For the first two reductions, these reduction sequences take the same samples by $\sim_p$, case i, as in \Cref{ex:cousin sim}~.
+For the first two reductions, these reduction sequences take the same samples by $\sim_p$, case i, as in Example \ref{ex:cousin sim}~.
 The next reduction uses the samples labelled by $(M_{sss} \to M_{Xss} \to M_{XXs}, \underline{+}_3)$ and $(M_{sss} \to M_{sXs} \to M_{XXs}, \underline{+}_3)$, which are related by $\sim_c$, case a, therefore when these reduction sequences reach $M_{XXX}$, they still contain all the same numbers, as desired.
@@
 At each reduction step $M \to N$, the sample space must be restricted from $I^{L_s(M)}$ to $I^{L_s(N)}$. 
-The injection $L_0(N) \to L_0(M)$ is trivial to define by appending $\mathit{Sk}(M) \to \mathit{Sk}(N)$ to each path, and using \Cref{lem:sim-M-N}, this induces a corresponding injection on the quotient, $L(N) \to L(M)$.
+The injection $L_0(N) \to L_0(M)$ is trivial to define by appending $\mathit{Sk}(M) \to \mathit{Sk}(N)$ to each path, and using Lemma \ref{lem:sim-M-N}, this induces a corresponding injection on the quotient, $L(N) \to L(M)$.
 The corresponding map $L_s(N) \to L_s(M)$ is then denoted $i(M \to N)$.
@@
 This reduction relation now has all of the properties required of it.
-In particular, it can be considered an extension of the standard trace semantics (as will be seen later in \Cref{thm:AstEquivalence}), and also:
+In particular, it can be considered an extension of the standard trace semantics (as will be seen later in Theorem \ref{thm:AstEquivalence}), and also:
 \begin{restatable}{lemma}{churchRosser} \label{churchRosser}
@@
 Let the set of samples with which it terminates with this reduction strategy be $X$.
-By \Cref{thm:CbvIsTerminatingest}, $M$ also terminates with $\cbv$ and every element of $X$, and $X$ has measure 1, by assumption, therefore $M$ is AST with respect to $\cbv$ therefore by \Cref{thm:AstEquivalence} it is AST.
+By Theorem \ref{thm:CbvIsTerminatingest}, $M$ also terminates with $\cbv$ and every element of $X$, and $X$ has measure 1, by assumption, therefore $M$ is AST with respect to $\cbv$ therefore by Theorem \ref{thm:AstEquivalence} it is AST.
 \end{proof}
@@
 All of the theorems on the termination of rankable terms therefore extend to other reduction strategies too.
-The proofs of \Cref{thm:rankable implies termination,thm:partial implies rankable,thm:antitone rankable implies termination,thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies.
-They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because \Cref{thm:de groote} applies equally to any reduction strategy).
+The proofs of Theorem \ref{thm:rankable implies termination,thm:partial implies rankable,thm:antitone rankable implies termination,thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies.
+They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because Theorem \ref{thm:de groote} applies equally to any reduction strategy).
 For a reduction strategy $r$ and term $M$, just substitute $N[r(N)]$ being a $\tY$-redex for $N$ being of the form $E[\tY \lambda x.
@@
 The only restrictions to bear in mind on evaluating sub-terms are that a $\tsample$ may not be evaluated inside a $\lambda$ that's inside a $\tY$ or on the right of an application, and a $\beta$ redex may not be reduced until its argument is a value.
-This can be seen as applied to \Cref{ex:sum of powers} at the end of \Cref{apx:confluent}.
+This can be seen as applied to Example \ref{ex:sum of powers} at the end of Appendix \ref{apx:confluent}.
 
 \medskip
-If, as seems likely, \Cref{thm:antitone rankable implies termination} is complete for terms from which every reachable term is AST, then by the following theorem, \Cref{thm:confluent ranking} can't actually prove the termination of any terms not already provable by \Cref{thm:antitone rankable implies termination} and \Cref{thm:antitone partial implies rankable}.
+If, as seems likely, Theorem \ref{thm:antitone rankable implies termination} is complete for terms from which every reachable term is AST, then by the following theorem, Theorem \ref{thm:confluent ranking} can't actually prove the termination of any terms not already provable by Theorem \ref{thm:antitone rankable implies termination} and Theorem \ref{thm:antitone partial implies rankable}.
 
@@
 It is still antitone rankable, but this is much more complicated.
-Details of the ranking function construction with and without using the confluent semantics are given in \Cref{apx:confluent}.
+Details of the ranking function construction with and without using the confluent semantics are given in Appendix \ref{apx:confluent}.
 \end{example}
@@
 Even some of the examples given earlier implicitly use slightly non-standard reduction strategies to make the definitions of their ranking functions slightly simpler.
-For example, the antitone sparse ranking function in \Cref{ex:unbiased random walk} was defined at $\Theta_2\, \underline n$ (where $\Theta_2 = \tY \, \lambda f n .
+For example, the antitone sparse ranking function in Example \ref{ex:unbiased random walk} was defined at $\Theta_2\, \underline n$ (where $\Theta_2 = \tY \, \lambda f n .
 \, \tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$), but $\Theta_2\, \underline{10}$ actually reduces to $\Theta_2\, (\underline{10} - 1)$ or $\Theta_2\, (\underline{10} + 1)$, and (taking the second of these), it then reduces to $(\lambda z.
@@
 
-\Cref{ex:higher-order recursion} also makes use of an alternative reduction strategy.
+Example \ref{ex:higher-order recursion} also makes use of an alternative reduction strategy.
 In this case, it is to allow the antitone sparse ranking function to skip over all of those terms where the unknown function $F$ is part way through evaluation.

### 2021-10-01 LICS21/intro.tex
Use \ref instead of \Cref.

@@
 Because the type system already constrains the terms enough to force termination in the absence of the recursion construct \citep{tait1967,BarendregtDS2010}, it is only the $\tY$-reduction steps that must be counted, making defining ranking functions somewhat easier.
-Using supermartingales and Doob's (iterated) Optional Sampling Theorem, we show the soundness of rankability (\cref{thm:rankable implies termination}): if a term has a ranking function, it is AST.
+Using supermartingales and Doob's (iterated) Optional Sampling Theorem, we show the soundness of rankability (Theorem \ref{thm:rankable implies termination}): if a term has a ranking function, it is AST.
 
@@
 Sparse ranking functions can be defined only for those points in the execution of a program which are semantically important, while all of the other intermediate steps can be ignored.
-Yet, they are no less efficacious for proving AST (\cref{thm:partial implies rankable}): every sparse ranking function is a restriction of a ranking function.
+Yet, they are no less efficacious for proving AST (Theorem \ref{thm:partial implies rankable}): every sparse ranking function is a restriction of a ranking function.
 This also makes the ranking function method of proving AST more compatible with syntactic sugar, because the intermediate reduction steps implicit in the simplified notation can be ignored.
@@
 Thanks to this feature, the antitone ranking function method is capable of handling programs which terminate arbitrarily slowly. \lo{TODO: Reference some examples.}
-We show that this method is also sound for proving AST (\cref{thm:antitone rankable implies termination}), 
-moreover it also enjoys a sparse function theorem (\cref{thm:antitone partial implies rankable}).
+We show that this method is also sound for proving AST (Theorem \ref{thm:antitone rankable implies termination}), 
+moreover it also enjoys a sparse function theorem (Theorem \ref{thm:antitone partial implies rankable}).
 \akr{Our progress condition is equivalent to that given by \cite{DBLP:journals/pacmpl/McIverMKK18} (by applying a transformation to the ranking function), but needs to be different in order to be compatible with the presence of additional intermediate reduction steps where the ranking function is not required to decrease at all.}
-\lo{The same point is made in \cref{sec:related}, so the preceding can be omitted.}
+\lo{The same point is made in Section \ref{sec:related}, so the preceding can be omitted.}
 
@@
 We introduce a novel addressing scheme for the possible random choices in a program's execution, which ensures that the same random choices are taken at corresponding positions in alternative reduction sequences, so that the same eventual result can be reached.
-This is then used to prove yet another extension to the ranking function theorem, that (\cref{thm:confluent ranking}) \lo{Preceding reference generated is wrong; it should be a theorem} ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and (\cref{cor:Reduction strategy independence}) rankability in this sense still imples almost sure termination.
+This is then used to prove yet another extension to the ranking function theorem, that (Theorem \ref{thm:confluent ranking}) \lo{Preceding reference generated is wrong; it should be a theorem} ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and (Corollary \ref{cor:Reduction strategy independence}) rankability in this sense still imples almost sure termination.
 The confluent trace semantics is of wider interest, and has other possible applications as well, for example in Bayesian inference algorithms.
 
-By combining these methods we can prove AST of a variety of PPCF programs that are beyond the reach of methods in the literature, including non-affine recursion\footnote{\label{fnote:non-affine} Non-affine recursive programs are recursive programs that can, during the evaluation of the recursive body, make multiple recursive calls (of a first-order function) from distinct call sites.} (\cref{ex:non-affine recursion easy,ex:non-affine continuous}) and recursions that define higher-order functions (\cref{ex:higher-order recursion}).
+By combining these methods we can prove AST of a variety of PPCF programs that are beyond the reach of methods in the literature, including non-affine recursion\footnote{\label{fnote:non-affine} Non-affine recursive programs are recursive programs that can, during the evaluation of the recursive body, make multiple recursive calls (of a first-order function) from distinct call sites.} (Examples \ref{ex:non-affine recursion easy,ex:non-affine continuous}) and recursions that define higher-order functions (Example \ref{ex:higher-order recursion}).
 
@@
 
-We present the syntax and trace semantics of PPCF in \Cref{sec:PPCF}.
-In \Cref{sec:supermartingales}, we show that ranking functions on terms induce supermartingales, which form the basis of a sound method for proving AST.
-We introduce sparse ranking functions in \Cref{sec:ranking} and antitone ranking functions in \Cref{sec:antitone}, and illustrate how they can be used to prove AST via examples. 
-In \Cref{sec:confluent}, we present a confluent trace semantics and demonstrate its usefulness.
-We discuss further applications of the confluent semantics in \Cref{sec:applications}; and conclude with comments on related work and further directions in \Cref{sec:related}.
+We present the syntax and trace semantics of PPCF in Section \ref{sec:PPCF}.
+In Section \ref{sec:supermartingales}, we show that ranking functions on terms induce supermartingales, which form the basis of a sound method for proving AST.
+We introduce sparse ranking functions in Section \ref{sec:ranking} and antitone ranking functions in Section \ref{sec:antitone}, and illustrate how they can be used to prove AST via examples. 
+In Section \ref{sec:confluent}, we present a confluent trace semantics and demonstrate its usefulness.
+We discuss further applications of the confluent semantics in Section \ref{sec:applications}; and conclude with comments on related work and further directions in Section \ref{sec:related}.
 

### 2021-10-01 LICS21/ranking-proofs.tex
Use \ref instead of \Cref.

@@
 The difference $g - f$ is then a supermartingale (with the same setup as in
-\Cref{thm:rankable and strict rankable});
+Theorem \ref{thm:rankable and strict rankable});
 therefore $\mathbb E[g(M_n)] \leq \mathbb E[f(M_n)] + g(M)-f(M)$, for all $n$.
@@
 Define $f_2(N) = \int_\entrosp f_1(\nnext(N,s)) + g(N,s) \, \mu_{\entrosp}(\mathrm d s)$.
-The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction, by the same argument as \Cref{thm:minimal}).
+The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction, by the same argument as Theorem \ref{thm:minimal}).
 \end{proof}

### 2021-10-01 LICS21/ranking.tex
Use \ref instead of \Cref.

@@
 \subsection{Sparse ranking functions}
-Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost sure termination, so it isn't useful for proving it.
+Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and Theorem \ref{thm:minimal} makes even stronger assumptions than almost sure termination, so it isn't useful for proving it.
 
@@
 The term  $(\tY \, \Theta) \, \underline 0$ generates a geometric distribution.
-Despite its simplicity, its $\mathit{Rch}$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.
+Despite its simplicity, its $\mathit{Rch}$ contains all the terms in Figure \ref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.
 

### 2021-10-01 LICS21/related.tex
Use \ref instead of \Cref.

@@
 % (because the underlying recursion schemes are not Turing complete):
-for example, the PPCF terms in \cref{ex:raven complex,ex:non-affine continuous,ex:higher-order recursion}
+for example, the PPCF terms in Examples \ref{ex:raven complex,ex:non-affine continuous,ex:higher-order recursion}
 %\refExample{complexExample} 
@@
 
-The main result of \cite{DBLP:journals/pacmpl/McIverMKK18} is similar to our \cref{thm:antitone rankable implies termination}, but in an imperative language.
+The main result of \cite{DBLP:journals/pacmpl/McIverMKK18} is similar to our Theorem \ref{thm:antitone rankable implies termination}, but in an imperative language.
 While we require that the antitone ranking function decreases for every $\tY$-reduction step, they require that it decreases for every iteration of a certain while loop (the ranking function in that case is defined in the context of a particular loop), which is similar to $\tY$ but limited to tail recursion.
@@
 
-\paragraph*{Further directions} We have not yet been able to prove \Cref{conj:antitone} which would imply that \Cref{thm:antitone rankable implies termination} is almost complete.
+\paragraph*{Further directions} We have not yet been able to prove Conjecture\ref{conj:antitone} which would imply that Theorem \ref{thm:antitone rankable implies termination} is almost complete.
 \begin{conjecture}[Completeness]
@@
 
-The \Cref{def:more general red} of redexes and which positions are acceptable to reduce at is sufficiently restrictive to guarantee confluence, but is also a little more restrictive in some cases than is necessary for this purpose.
+Definition \ref{def:more general red}, defining redexes and which positions are acceptable to reduce at is sufficiently restrictive to guarantee confluence, but is also a little more restrictive in some cases than is necessary for this purpose.
 For example, if the argument of a function is not yet a value, but its reduction to a value would be deterministic, or the function is affine, then applying the function before reducing its argument would not cause any problematic duplication of random samples.
@@
 
-Although this is a broadly applicable method of proving almost sure termination of probabilistic functional programs, and \Cref{thm:antitone partial implies rankable,thm:confluent ranking} make it more convenient to use, some method of automating the construction and checking of (antitone) ranking functions, even partially, could make it considerably more practically useful, especially in cases where almost sure termination is merely a side-condition for some other theorem or algorithm to be applicable.
+Although this is a broadly applicable method of proving almost sure termination of probabilistic functional programs, and Theorems \ref{thm:antitone partial implies rankable,thm:confluent ranking} make it more convenient to use, some method of automating the construction and checking of (antitone) ranking functions, even partially, could make it considerably more practically useful, especially in cases where almost sure termination is merely a side-condition for some other theorem or algorithm to be applicable.
 

### 2021-10-01 LICS21/supermartingales-proofs.tex
Fix compile errors
Fix citations and references
Use \ref instead of \Cref.

@@
 \rankableAndStrictRankable*
-We shall obtain the theorem as a corollary of a technical lemma (\Cref{lem:key rankable}).
+We shall obtain the theorem as a corollary of a technical lemma (Lemma \ref{lem:key rankable}).
 
@@
 
-As an immediate corollary of \Cref{lem:key rankable}, each $f(M_n)$ is integrable.
-This concludes the proof of \Cref{thm:rankable and strict rankable}.
+As an immediate corollary of Lemma \ref{lem:key rankable}, each $f(M_n)$ is integrable.
+This concludes the proof of Theorem \ref{thm:rankable and strict rankable}.
 
@@
 Finally thanks to the following
-\begin{theorem}[de Groote \cite{DBLP:conf/lfcs/Groote94}]
+\begin{theorem}[\cite{DBLP:conf/lfcs/Groote94}]
 \label{thm:de groote}
@@
 \end{enumerate}
-This concludes the proof of \cref{lem:TnBounded}.
+This concludes the proof of Lemma \ref{lem:TnBounded}.
 
@@
 To see $\set{T_M^\tY = n} \in \calF_{T_n}$ for a given $n$, we need to show that $\set{T_M^\tY = n} \cap \set{T_n \leq i} \in \calF_i$, for all $i \in \mathbb N$.
-Let $\cal V$ be the set of values in $\Lambda^0$ (which is measurable), and let $\cal W$ be the non-values.
+Let $\mathcal V$ be the set of values in $\Lambda^0$ (which is measurable), and let $\mathcal W$ be the non-values.
 Then, it suffices to observe that $\set{T_M^\tY = n} \cap \set{T_n \leq i}
 = \bigcup_{l=n}^i 
-\big( M_l^{-1} [{\cal V}] \cap  
-M_{l-1}^{-1} [{\cal W}] \cap 
+\big( M_l^{-1} [{\mathcal V}] \cap  
+M_{l-1}^{-1} [{\mathcal W}] \cap 
 \set{T_n = l} \big)
-$, where each of $M_l^{-1} [{\cal V}]$, $M_{l-1}^{-1} [{\cal W}]$, and $\set{T_n = l}$ is in $\calF_i$.
+$, where each of $M_l^{-1} [{\mathcal V}]$, $M_{l-1}^{-1} [{\mathcal W}]$, and $\set{T_n = l}$ is in $\calF_i$.
 \end{proof}

### 2021-10-01 LICS21/supermartingales.tex
Use \ref instead of \Cref.

@@
 \label{def:ranking function}
-A \emph{ranking function on $M \in \Lambda^0$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$,\footnote{The fact that the ranking function can be 0 in some cases before it reaches a value is necessary to get \Cref{thm:minimal} to work neatly.} and
+A \emph{ranking function on $M \in \Lambda^0$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$,\footnote{The fact that the ranking function can be 0 in some cases before it reaches a value is necessary to get Theorem \ref{thm:minimal} to work neatly.} and
 \begin{enumerate}
@@
 It seems cleaner to include this in the definition.}
-\akr{The fact that the ranking function can be 0 in some cases before it reaches a value is necessary to get \Cref{thm:minimal} to work neatly.}
-\lo{OK. Let's leave \Cref{def:ranking function} as it is.}
+\akr{The fact that the ranking function can be 0 in some cases before it reaches a value is necessary to get Theorem \ref{thm:minimal} to work neatly.}
+\lo{OK. Let's leave Definition \ref{def:ranking function} as it is.}
 \fi
@@
 \begin{remark}
-Lem.~\ref{lem:rank-PAST} is a slight extension of \cite[Lem.~5.5]{DBLP:conf/popl/FioritiH15},
+Lemma \ref{lem:rank-PAST} is a slight extension of \cite[Lem.~5.5]{DBLP:conf/popl/FioritiH15},
 which asserts the same result but for the specific stopping time $T : \omega \mapsto \min\set{n \mid Y_n(\omega) = 0}$.
@@
 %\subsection{Proof of \Cref{thm:rankable and strict rankable}.}
-The ranking supermartingale condition is satisfied essentially by a case analysis on the type of redex (\Cref{lem:key rankable}).
+The ranking supermartingale condition is satisfied essentially by a case analysis on the type of redex (Lemma \ref{lem:key rankable}).
 
@@
 Let $f$ be a ranking function on $M$.
-For (i), since $(T_n)_{n \geq 0}$ is an increasing sequence of stopping times, each is adapted to $(\calF_n)_{n \geq 0}$ and bounded (\Cref{lem:TnBounded}),
-and $(f(M_n))_{n \geq 0}$ is a supermartingale also adapted to $(\calF_n)_{n \geq 0}$ (\Cref{thm:rankable and strict rankable}),
-it follows from the Optional Sampling \Cref{thm:optional sampling} that $(Y_n)_{n \geq 0}$ %where $Y_n := f(M_{T_n})$ 
+For (i), since $(T_n)_{n \geq 0}$ is an increasing sequence of stopping times, each is adapted to $(\calF_n)_{n \geq 0}$ and bounded (Lemma \ref{lem:TnBounded}),
+and $(f(M_n))_{n \geq 0}$ is a supermartingale also adapted to $(\calF_n)_{n \geq 0}$ (Theorem \ref{thm:rankable and strict rankable}),
+it follows from the Optional Sampling Theorem \ref{thm:optional sampling} that $(Y_n)_{n \geq 0}$ %where $Y_n := f(M_{T_n})$ 
 is a supermartingale adapted to $(\calF_{T_n})_{n \geq 0}$.
-Notice the stopping time $T_M^\tY$ is also adapted to $(\calF_{T_n})_{n \geq 0}$ (\Cref{lem:TMtY is a stopping time}); and we have that $(Y_n)_{n \geq 0}$ is a $1$-ranking supermartingale.
-Therefore, by \Cref{lem:rank-PAST}, $T^{\tY}_M < \infty$ a.s.~and $\expect{T^{\tY}_M} < \infty$.
+Notice the stopping time $T_M^\tY$ is also adapted to $(\calF_{T_n})_{n \geq 0}$ (Lemma \ref{lem:TMtY is a stopping time}); and we have that $(Y_n)_{n \geq 0}$ is a $1$-ranking supermartingale.
+Therefore, by Lemma \ref{lem:rank-PAST}, $T^{\tY}_M < \infty$ a.s.~and $\expect{T^{\tY}_M} < \infty$.
 %For 2, we have that $(f(M_n))_{n \geq 0}$ is a ranking supermartingale (\Cref{thm:rankable and strict rankable}) w.r.t.~stopping time $T_M$. 
 %It follows immediately from \Cref{lem:rank-PAST} that $\expect{T_M} < \infty$.
-Statement (ii) follows at once from \Cref{thm:rankable and strict rankable} and \Cref{lem:rank-PAST}.
+Statement (ii) follows at once from Theorem \ref{thm:rankable and strict rankable} and Lemma \ref{lem:rank-PAST}.
 \end{proof}
@@
 Thus the method of (strict) ranking function is sound for proving (positive) a.s.~termination of PPCF programs.
-It is in fact also complete in a sense: if $\expect{T^\tY_N} < \infty$ for all $N \in \mathit{Rch}(M)$ then $M$ is rankable (\Cref{thm:minimal}).
+It is in fact also complete in a sense: if $\expect{T^\tY_N} < \infty$ for all $N \in \mathit{Rch}(M)$ then $M$ is rankable (Theorem \ref{thm:minimal}).

### 2021-10-01 LICS21/antitone.tex
Replace "theorem" environment with "therm", to get numbering back.

@@
 
-\begin{restatable}{theorem}{ACStrict}
+\begin{restatable}{therm}{ACStrict}
 \label{thm:a-c strict}
@@
 
-\begin{theorem}[Antitone ranking function soundness] \label{thm:antitone rankable implies termination}
+\begin{therm}[Antitone ranking function soundness] \label{thm:antitone rankable implies termination}
 If a closed PPCF term $M$ is antitone rankable, then $T_M^{\tY} < \infty$ a.s.~(equivalently, $M$ is AST).
-\end{theorem}
+\end{therm}
 \begin{proof}
@@
 
-\begin{restatable}[Sparse function]{theorem}{AntitonePartial}  \label{thm:antitone partial implies rankable}
+\begin{restatable}[Sparse function]{therm}{AntitonePartial}  \label{thm:antitone partial implies rankable}
   Every antitone sparse ranking function is a restriction of an antitone ranking function.
@@
 The following theorem does not prove completeness, but is suggestive in that direction:
-%\begin{restatable}{theorem}{antitoneSeemsComplete}
-\begin{theorem}
+%\begin{restatable}{therm}{antitoneSeemsComplete}
+\begin{therm}
 \label{thm:towards completeness}
 For any stopping time $T$ which is almost surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
-\end{theorem}
+\end{therm}
 %\end{restatable}

### 2021-10-01 LICS21/confluent.tex
Replace "theorem" environment with "therm", to get numbering back.

@@
 
-\begin{restatable}{theorem}{AstEquivalence} \label{thm:AstEquivalence}
+\begin{restatable}{therm}{AstEquivalence} \label{thm:AstEquivalence}
 A closed term $M$ is AST with respect to $\cbv$ iff it is AST.
@@
 
-\begin{restatable}{theorem}{CbvIsTerminatingest} \label{thm:CbvIsTerminatingest}
+\begin{restatable}{therm}{CbvIsTerminatingest} \label{thm:CbvIsTerminatingest}
 If $M$ terminates with some reduction strategy $f$ and trace $s$, it terminates with $\cbv$ and $s$.
@@
 All of the theorems on the termination of rankable terms therefore extend to other reduction strategies too.
-The proofs of Theorem \ref{thm:rankable implies termination,thm:partial implies rankable,thm:antitone rankable implies termination,thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies.
+The proofs of Theorems \ref{thm:rankable implies termination}, \ref{thm:partial implies rankable}, \ref{thm:antitone rankable implies termination} and \ref{thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies.
 They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because Theorem \ref{thm:de groote} applies equally to any reduction strategy).
@@
 
-\begin{theorem}[Confluent ranking] \label{thm:confluent ranking}
+\begin{therm}[Confluent ranking] \label{thm:confluent ranking}
 If a closed PPCF term $M$ has a ranking function, sparse ranking function, antitone ranking function or antitone sparse ranking function w.r.t.~a reduction strategy $r$, then $M$ is AST w.r.t.~$r$, and AST.
-\end{theorem}
+\end{therm}
 
@@
 
-\begin{restatable}{theorem}{confluentNoStronger}
+\begin{restatable}{therm}{confluentNoStronger}
 \label{thm:confluentNoStronger}

### 2021-10-01 LICS21/intro.tex
Replace "theorem" environment with "therm", to get numbering back.

@@
 
-By combining these methods we can prove AST of a variety of PPCF programs that are beyond the reach of methods in the literature, including non-affine recursion\footnote{\label{fnote:non-affine} Non-affine recursive programs are recursive programs that can, during the evaluation of the recursive body, make multiple recursive calls (of a first-order function) from distinct call sites.} (Examples \ref{ex:non-affine recursion easy,ex:non-affine continuous}) and recursions that define higher-order functions (Example \ref{ex:higher-order recursion}).
+By combining these methods we can prove AST of a variety of PPCF programs that are beyond the reach of methods in the literature, including non-affine recursion\footnote{\label{fnote:non-affine} Non-affine recursive programs are recursive programs that can, during the evaluation of the recursive body, make multiple recursive calls (of a first-order function) from distinct call sites.} (Examples \ref{ex:non-affine recursion easy} and \ref{ex:non-affine continuous}) and recursions that define higher-order functions (Example \ref{ex:higher-order recursion}).
 

### 2021-10-01 LICS21/ranking.tex
Replace "theorem" environment with "therm", to get numbering back.

@@
 
-\begin{restatable}{theorem}{ThmMinimal} \label{thm:minimal}
+\begin{restatable}{therm}{ThmMinimal} \label{thm:minimal}
 Given a closed term $M$, the function $f:Rch(M) \to \mathbb R$ given by 
@@
 
-\begin{restatable}[Sparse function]{theorem}{PartialImpliesRankable} \label{thm:partial implies rankable}
+\begin{restatable}[Sparse function]{therm}{PartialImpliesRankable} \label{thm:partial implies rankable}
 Every sparse ranking function is a restriction of a ranking function.

### 2021-10-01 LICS21/related.tex
Replace "theorem" environment with "therm", to get numbering back.

@@
 % (because the underlying recursion schemes are not Turing complete):
-for example, the PPCF terms in Examples \ref{ex:raven complex,ex:non-affine continuous,ex:higher-order recursion}
+for example, the PPCF terms in Examples \ref{ex:raven complex}, \ref{ex:non-affine continuous} and \ref{ex:higher-order recursion}
 %\refExample{complexExample} 
@@
 
-Although this is a broadly applicable method of proving almost sure termination of probabilistic functional programs, and Theorems \ref{thm:antitone partial implies rankable,thm:confluent ranking} make it more convenient to use, some method of automating the construction and checking of (antitone) ranking functions, even partially, could make it considerably more practically useful, especially in cases where almost sure termination is merely a side-condition for some other theorem or algorithm to be applicable.
+Although this is a broadly applicable method of proving almost sure termination of probabilistic functional programs, and Theorems \ref{thm:antitone partial implies rankable} and \ref{thm:confluent ranking} make it more convenient to use, some method of automating the construction and checking of (antitone) ranking functions, even partially, could make it considerably more practically useful, especially in cases where almost sure termination is merely a side-condition for some other theorem or algorithm to be applicable.
 

### 2021-10-01 LICS21/LICS 207 reviews and rebuttal.txt
" \ref" -> "~\ref"

Review 1

Overall evaluation:	2: (accept)

The paper develops a method of establishing that a higher-order probabilistic program with continuous distributions almost surely terminates in the context of lambda calculus. There are a number of new ideas in the paper, including the use of ranking supermartingales, sparse tracking functions, antitone ranking functions, and restricted sets of reduction strategies that ensure the Church-Rosser property. Although the method doesn't look like a completed project, the advances are impressive (at least, to a non-specialist).

For a non-expert in the area, the paper is a challenging read, with tough notation, but it contains a lot of examples (also further developed in the appendix). All in all, I think this is a very solid piece of work and recommend acceptance.

- What is \mathsf R in the definition of types?
- What is r.v. before Definition III.2?
- What is a.s. in Definition III.2?

===

Review 2

Overall evaluation:	1: (weak accept)

# Summary

This paper considers how to prove almost-sure termination (AST) for
probabilistic higher-order programs. AST is the natural generalization of
termination to probabilistic programs, and there has been plenty of work in
recent years on verifying this property. Existing work largely considers
first-order, imperative programs, developing the method of ranking
supermartingales (and generalizations, such as antitone ranking
supermartingales) to verify AST. This paper adapts these ideas to higher-order
probabilistic PCF, and also with continuous sampling. The authors introduce
sparse ranking supermartingales to make it possible to avoid defining the
ranking function for "uninteresting" steps, and antitone ranking
supermartingales to prove AST when the expected termination time is not finite.
The last contribution in the paper considers when AST under the "standard
reduction" (CBV) can be shown by proving AST under an alternative reduction
strategy, which may be significantly easier for some programs.

# General comments

This paper tackles an important and interesting problem. To date most work on
termination of probabilistic programs has considered first-order programs
(though there are certainly exceptions, as discussed in related work), and this
work takes many ideas from the first-order literature and transports it to the
higher-order setting. The idea of a sparse ranking supermartingale is nice, and
addresses a particular issue in the higher-order setting---it may be
inconvenient to define the ranking function for all reduction steps, and we may
want to focus on just specific, "interesting" reduction steps. Likewise,
antitone ranking supermartingales have been considered in the imperative
setting, but this paper shows that they can also prove AST for higher-order
programs.

My main criticism of the paper is in the last contribution, which has the
potential to be the most interesting---a choice of reduction strategies is
something that is inherently a feature of higher-order programs, and indeed it
would be interesting to consider adapting the reduction strategy in order to
simply reasoning about AST. The high level goal is clear to me. However, I found
the section where this contribution is described (Section 6) to be totally
impenetrable. The section starts by defining positions, subterms, alternative
reduction relation, reduction relation on skeletons, and goes on, but it was
never clear to me why the authors are doing all of this. Even a small roadmap
would really help me understand why we are covering these things and why we are
defining these concepts, and how they fit in to the overall goal. Currently,
Section 6 is a series of highly intricate technical definitions with somewhat
cryptic examples, but there is little intuition and no sense of how the pieces
fit together, and what the goal is.

It seems to me that Corollary VI.12 is the main result of the section. But I was
also very confused here, and before in the definition of reduction strategy. The
example of reduction strategy shows CBV, but gives no further examples. In
particular, is CBN a valid reduction strategy under this definition? If not, why
not? And if so, VI.12 seems strange---if M is AST wrt to CBN, is it also AST wrt
to CBV? That seems highly suspicious...

# Comment after PC discussion

Thanks for the clarifications. The PC members agree on the importance of the paper. However, there is also some concern on the accessibility of the current version. We encourage the authors to work in order to make Section VI more accessible. Moreover, we suggest incorporating the responses given in the rebuttal into the paper. In particular, it would be valuable to expand accordingly the discussion concerning Theorem V.15.

# Specific comments

* p2: Church-Rosserness -> the Church-Rosser property
* p2: "We introduce a novel addressing scheme for the possible random choices in
a program’s execution, which ensures that the same random choices are taken at
corresponding positions in alternative reduction sequences, so that the same
eventual result can be reached." :: this description sounds like a
probabilistic coupling is being constructed (?) is there any connection?
* p2: free in s -> free in M (?)
* p3: "Y lambda f n :" this syntax doesn't match the grammar (throughout)
* p3: "The score construct is irrelevant to termination except that it fails if
its argument is negative, thus allowing computations to fail after finitely
many steps." :: arguably, it is not clear whether termination is an
interesting/meaningful concept in the presence of score/conditioning
* p3: for all N2. -> for all N.2 (all footnotes)
* p3: "Let T be a r.v." :: maybe give the type of T?
* p3: int P(d omega) Yn(w) :: maybe swap for consistency? int Yn(w) P(d omega)
* p4: extra space before footnote
* p4: "we have that (Yn)n≥0 is a 1-ranking supermartingale" :: this needs more
justification. for instance, why is not an eps-ranking supermartingale, where
does the 1 come from?
* p4: Equation (1): M_{T_n}(s) -> M_{T_n(s)}(s)?
* p5: Section 4 said that “A finite number of expected Y-reduction steps does
not necessarily imply a finite number of expected total reduction steps.” But
under lemma III.8, it is said that T_M^Y < ∞ a.s. iff M is AST. More
discussions about the connection between Y-PAST and AST/PAST would be helpful
and why Example IV.1 is not rankable (by mapping to Y reductions needed) will
be appreciated.
* p5: contractum :: what is this?
* p5: "to define M ⊕p N" :: bad linebreak here
* p5: "Y lambda f n." grammar
* p5: On the bottom of page 5, why the map described is a sparse ranking
function? (I thought the expected Y reduction from i (+)_0.5 (Y \lambda fn.
n (+)_0.5 f(n+1)) (i +1) is 0.5 * 1 + 0.5 * 0.5 * 2 + ... + 0.5^c * c + ... >
1)
* p6: maybe cite prior work on antitone ranking functions?
* p6: what are these examples supposed to show?
* p6: Example V.1, M1, 3/2 -> 2/3 ?
* p6: actual values -> realized values
* p6: Example V.5 is a nice example, but it seems like a curiosity rather than a
central point. the space might be better used for something else, e.g., to
explain the next section.
* p6: "(Yk) k \in omega" :: if omega is supposed to be the natural numbers, then
this clashes with the previous omega on this line.
* p6: Plainly -> It can be seen that
* p7: "Therefore by setting epsilon2(x) = ..." :: where does e come from? this
needs more explanation.
* p7: "Even with this restriction" :: what restriction?
* p7: a roadmap here would be nice. you define positions, then subterms, then a
more general reduction relation, then a reduction relation on skeletons, ...
but where are we going and why are we doing all of this?
* p8: missing space: M1 M2
* p8: Definition VI.1 to be clear, is this relation is non-deterministic? that
is, is it for any position alpha? also, this definition is hard to
parse---consider putting "then" after the commas.
* p8: "Labelling the pre-chosen samples by the positions" :: where is this
labelling? example?
* p8: footnote 6: why not just use a different arrow?
* p8: "with one more caveat introduced later" :: hard to follow. where will this
caveat be introduced?
* p8: Example VI.2: should A[x] be A[X], since X is a hole?
* I see that this defines the semantics of A[r], where A is a skeleton, and r is
a vector of real numbers in Sec 2. But here, I'm not sure if A is a skeleton,
or a program, and what does A[x] and A[X] mean?
* Is "I" in A[x] a function? You used I for intervals too.
* "r occurs twice in the conditions on s: once as the value a sample must take,
and once in the location of a sample". I have difficulties to see where r
appears as the location of a sample.
* p8: "this set is unmeasurable" :: is this obvious? why?
* p9: "This is still not sufficient" :: for what?
* p9: "in which case" :: what case is this referring to?
* p9: I could not understand the purpose of Definition VI.5.
* p10: is CBN a reduction strategy, under your definition? if not, why not? if
so, VI.12 looks suspicious. Another example of a reduction strategy would help.
* p11: Applications: this section is highly speculative, since it does not
appear that any of the claims have been worked out in detail. maybe this is
better under future work, or discussion?

# More on Section 6:
* At the bottom of page 7, what kind of restrictions are needed to exclude the problem caused by the reduction strategy making copies of samples to be identical v.s.. independent?
* In Definition VI.I, could you give more intuition why λ should not occur after Y in α?
Is the example (lambda x. x 0 + x 0) (lambda y. sample) the kind of programs where making copies identical vs. independent causes problem?
Is that example meant to illustrate why λ should not occur after @2 in α?
Also, it will be helpful to highlight restriction here has to do with restriction on reduction strategies that satisfy confluence.
* Probably some high-level intuition for complicated cases in Definition VI.5, besides the discussions in examples VI.4 and example VI.6
* In Definition VI.8, what is s \circ i(M→N) ? Is circ composition? The type doesn’t seem to match.
* Theorem VI.11 was confusing to me at first. Discussing how all the previous definitions imply the restrictions on reduction strategies would help. (Maybe moving some lines from the paragraph under Theorem VI.15)

===

Review 3

Overall evaluation:	1: (weak accept)

The paper develops new techniques for proving almost-sure termination of probabilistic programs written in a prototypical higher-order typed functional language. This adds to a large body of work on formal methods for probabilistic programming language. The paper improves on previous results by dealing with higher-order programs and by considering a random choice operator (called sample) with a continuous distribution. The method underlying termination proofs is based on (sparse) antitone ranking functions.

I am not really an expert in the field but it does not seem that these ingredients are very original. However they are nicely combined, the paper is very well written, with a good set of illustrative example programs. The technical work is presented in a clear way, with the result that it does not seem overly difficult.

It is regrettable that the author(s) do not comment more on their ability to prove completeness of the method [for imperative programs and ranking functions in a well-founded partial ordering, completeness is easy to show]. Is this just lack of time or is there a well-identified stumbling work that could be mentioned? If a program is AST every reachable term has a stopping time and these stopping times can be compared when we're talking about terms reachable from one another. Now we have to turn this into a ranking function. Is the problem caused by the existence of loops between reachable terms (even when guaranteed to terminate almost surely)? Is the problem about showing that a complex set of recursive inequations indeed admit solutions?

===

Review 4
Overall evaluation:	2: (accept)

This paper extends expectation-based reasoning for (positive) almost-sure termination from first-order imperative programs a la [18,25] to higher-order functional programs, i.e. to a probabilistic lambda calculus. One could argue that this extension is rather straight-forward. I would instead say that it is rather _clean_ and hence very valuable. Furthermore, "executing" lambda-terms is much less straight-forward than executing first-order programs due to (a) potentially many intermediate reduction steps until reaching another reduction of the Y-combinator (analogous to reaching the loop-head in a first-order program) and (b) potentially ambiguous reduction orders. Both (a) and (b) pose technical difficulties in lifting the (P)AST proof rules from first-order to higher order. The solution proposed for (a) - sparse ranking functions - is again a very clean and easy to comprehend solution. The solution for (b), I had difficulties understanding, which is partly because I am not an expert on lambda calculi.

Overall, I would - perhaps weakly - support acceptance of this paper. Why weakly? I found the paper rather dense, Section VI (Confluent Trace Semantics) in particular. I believe, it is quite difficult to read this paper, if one does not speak (simply typed) lambda fluently and I also believe that this aspect could well be improved significantly in order to make the paper more accessible to a wider audience.

Questions for Rebuttal:

===================

Regarding Theorem V.15: Could you elaborate on why exactly this theorem does NOT solve the open problem of completeness of antitone ranking supermartingales? Where/what exactly is the gap? What's missing to prove completeness? Furthermore, what _does_ Theorem V.15 telle us, intuitively?

More Detailed Comments:

======================

p2: high-order -> higher-order

p2: The role of the environment Gamma is not at all clear here. Only in Figure 1 it is needed and without explanation, Figure 1 is very difficult to understand for someone who is not fluent in lambda calculi. Please explain the role of Gamma and provide some intuition on the Y-combinator in order to make the paper more accessible to non-lambda-experts.

p2, Section II.B begins with "This version...". Which version is meant here? Furthermore, the second sentence is entirely unclear to me. And the first two paragraphs of this subsection I find rather dense and lacking intuition. This could be improved quite easily.

p3, Definition III.1 Please put a space after the superscript 2 in N². I first thought the authors mean a pair NxN instead of a footnote.

===

REBUTTAL

Reviewer 1: "\mathsf R" refers to the base type for programs which reduce to real-number values. "r.v." means random variable, and “a.s.” means almost surely.

Reviewer 2 said that there were no examples of reduction strategies beyond CBV. As the other reduction strategies we’re mainly considering (i.e. those used with Theorem VI.16) are intended to be designed for particular programs, the other examples are given along with particular example programs, specifically the extended discussions of Examples V.14 and VI.17 in the appendices.

CBN is not considered a valid reduction strategy because it violates the condition that the argument of a beta redex must be a value, which follows from the definition of reduction strategy (p. 10) and redex (Definition VI.1).

Example V.1 is simply meant to illustrate that there are some terms which terminate too slowly to be rankable.

Labelling the samples with elements of a set A simply refers to picking traces s from I^A, then using s(a) for the sample reduction whenever the relevant position (or other label) is a.

Section 6:

The restrictions in reduction strategies are the extra conditions in Definition VI.1 on what reductions are valid (as described in the paragraph after Theorem VI.15).

The body of a Y subterm is effectively both the function and the argument in an application. If it contains multiple occurrences of the relevant variable, it will be duplicated by the Y reduction, potentially causing a similar issue to the case of (\lambda f. f 0 + f 0) (\lambda x. sample) if the sample were evaluated first.

Reading guide for Section 6: In order to be able to use alternative reduction strategies to prove properties of the program defined using the usual reduction strategy, it is necessary first to prove the relation between these different strategies, i.e. to provide a confluent version of the semantics. It is therefore necessary to have some way of relating the samples that will be used in runs of the program using different reduction strategies. Positions are defined as a way of addressing sample statements within a program independently of the reduction order. Next, a version of the reduction relation is presented that is nondeterministic, so that it allows a choice of what order to perform reductions in. The notion of positions is extended to potential positions, for samples which may appear later in the reduction sequence but not necessarily in the initial term. In order to allow potential positions in different reduction sequences to be considered equivalent, a relation ~* is defined, and finally, all of these are used to construct a confluent version of the trace semantics, ⇒, that is still nondeterministic in reduction order, but does specify the outcome of random choices. The desired properties of this semantics (confluence and equivalence to the standard semantics) are proved, then it is used to give extended versions of ranking functions and the related AST results.

Reviewer 3:

Regarding the conjecture, we are unaware of any particularly relevant previous literature. We did spend a while trying to solve the problem, and will try some more in future. Part of what makes this question so difficult is that you have to construct an antitone function that represents the rate of convergence of all of the different states the program could be in, and simultaneously assign all the program states ranking function values. The fact that, for constructing a ranking function, even branches with 0 probability must be considered is an additional cause of difficulty, as they can terminate arbitrarily much slower than the program itself.

===
Rebuttal for Reviewer 4:

** Regarding Theorem V.15

The key weakness of Theorem V.15 is the condition “if (F_n)_n is the coarsest filtration to which T is adapted”. If we set the stopping time T to the termination time of a program, and (F_n)_n to the filtration defined by the revelation of samples as the program progresses, then an antitone supermartingale wrt T would (subject to the supermartingale and the program aligning properly, which would probably not be too hard to arrange) provide an antitone ranking function for the term. However, a supermartingale wrt the filtration that only observes whether the term has terminated yet may not be a supermartingale wrt the filtration that has all the information on the term’s progress, so V.15 does not quite match up to what is needed for the completeness Conjecture.

What it does imply is that there is no restriction on how slowly an antitone rankable term can terminate, analogous to the fact that if a term is rankable, it is Y-PAST. For example, it proves that there are terms such that the expectation of the square-root of their stopping time, or the logarithm of their stopping time, is unbounded, but which are still antitone rankable (because for any stopping time, it is possible to construct a term whose Y running time matches that stopping time, and where the constraint on the filtrations is satisfied).

** Regarding the rather dense Section VI

Although Section VI is rather dense, that complexity is necessary because of the complexity of the results presented. It takes several steps to reach the key definition VI.8, and it is not even possible to state the main results of that section until that definition. We have attempted to explain briefly the reasons why each of these steps is necessary (see paragraph “Reading guide for Section VI” of the rebuttal), although these are necessarily a little abstract until the reader has seen the whole picture. Admittedly this could be improved somewhat by summarising all of the steps of the definition at the start.

** Responses to other comments

The type of a term (and whether it’s well-typed at all) can depend on the type of its free variables, therefore the typing judgement Γ ⊢ M : A includes both the type A of the term M and a map Γ of free variables to types. The only typing rules involving this in a non-trivial way in this case are lambda abstraction (because the lambda binds a free variable) and the variable rule (because the type of a variable is just whatever is specified for it in the context). The typing system is not really a key component of this paper, which is why little emphasis is put on it. Any type system which prevents ill-formed terms like 2 (λx.x) + 1 and which is strong enough to prove that a term that does not undergo any Y reductions terminates would have worked equally well.

The phrase “This version” in section II.B refers to the distinction between \to and red, which is in a sense a refinement of \to.

### 2021-10-01 LICS21/rebuttal-long.txt
" \ref" -> "~\ref"

Reviewer 1:  "\mathsf R" refers to the base type for programs which reduce to real-number values. "r.v." means random variable, and “a.s.” means almost surely.

Reviewer 2 said that there were no examples of reduction strategies beyond CBV. As the other reduction strategies mainly considered in this work (i.e. used with Theorem VI.16) are intended to be specifically designed for particular programs, the other examples are given along with particular example programs, specifically the extended discussions of Examples V.14 and VI.17 in the appendices.

Regarding Corollary VI.12, CBN is not considered a valid reduction strategy because it violates the condition that the argument of a beta redex must be a value, which follows from the definition of reduction strategy (p. 10) and redex (Definition VI.1).

At the end of Section IV, the next reduction in 
   i ⊕_0.5 (Y \lambda fn. n ⊕_0.5 f(n+1)) (i+1) 
is the ⊕_0.5 (or, to be more precise, the sample implicit in that notation), not the Y.

Example V.1 is simply meant to illustrate that there are some terms which terminate too slowly to be rankable.

Labelling the samples with elements of a set A simply refers to picking samples s from I^A, then using s(a) for the sample reduction whenever the relevant position/number of samples previously taken/reduction sequence and position (depending on the labelling) is a.

Section 6:

The restrictions in reduction strategies are the extra conditions in Definition VI.1 on what reductions are valid (specifically, that the argument of a beta redex must be a value, and the restrictions on positions where sample reductions can occur).

The body of a Y subterm is effectively both the function and the argument in an application. If it contains multiple occurrences of the relevant variable, it will be duplicated by the Y reduction. More specifically, consider 
   (Y \lambda f n. if n = 0 then (\lambda x. sample) else \lambda x.(f 0 0 + f 0 0)) 1 0. 
This reduces to something like 
   (\lambda x. sample) 0 + (\lambda x. sample) 0. 
If the sample had been allowed to be reduced first, the samples would have ended up correlated. The fact that the position has to have a Y and a lambda in it to be excluded is simply to accommodate things like 
   Y if sample > 0.5 then M else N
where there actually should be reductions inside the Y.

In order to be able to use alternative reduction strategies to prove properties of the program defined using the usual reduction strategy, it is necessary first to prove the relation between these different strategies, i.e. to provide a confluent version of the semantics. It is therefore necessary to have some way of relating the samples that will be used in runs of the program using different reduction strategies. Positions are defined as a way of addressing sample statements within a program independently of the reduction order. Next, a version of the reduction relation is presented that is nondeterministic, so that it allows a choice of what order to perform reductions in. The notion of positions is extended to potential positions, for samples which may appear later in the reduction sequence but not necessarily in the initial term. In order to allow potential positions in different reduction sequences to be considered equivalent, a relation ~* is defined, and finally, all of these are used to construct a confluent version of the trace semantics, ⇒, that is still nondeterministic in reduction order, but that does specify the outcome of random choices. The desired properties of this semantics (confluence and equivalence to the standard semantics) are proved, then it is used to give extended versions of ranking functions and the related AST results.

Reviewer 3:

Regarding the conjecture, we are unaware of any particularly relevant previous literature. We did spend a while trying to solve the problem, and will try some more in future. Part of what makes this question so difficult is that you have to construct an antitone function that represents the rate of convergence of all of the different states the program could be in, and assign all the program states ranking function values, despite it not being obvious how to compare, for example, one state which would either terminate in a few steps or extremely slowly, and another which would terminate after a long time, but more deterministically. The fact that, for constructing a ranking function, even branches with 0 probability must be considered is an additional cause of difficulty, as they can terminate arbitrarily much slower than the program itself.

### 2021-10-01 LICS21/antitone copy.tex
" \ref" -> "~\ref"

% !TEX root = main.tex
\section{Antitone ranking functions}
\label{sec:antitone}

\begin{example}[Random walk]
\label{ex:ac-ranking}
\begin{enumerate}
\item 1D biased (towards 0) random walk
\[
M_1 = 
\big(\tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{3/2} f \, (n + 1)}\big) \, 10
\]

\item 1D unbiased random walk
\[
M_2 = 
\big(\tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}\big) \, 10
\]

\item 1D biased (away from 0) random walk
\[
M_3 = 
\big(\tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/3} f \, (n + 2)}\big) \, 10
\]

%While {x > 0} do {x := x-1 \oplus_2/3 x:= x+1}

%M2 = While {x > 0} do {x := x-1 \oplus_2/3 x:= x+2}
\end{enumerate}
\end{example}

The term $M_1$ is rankable, terminating in 31 $\tY$-steps on average, and $M_3$ only has a $1/1024$ chance of terminating, but in between, $M_2$ is AST, but isn't $\tY$-PAST, therefore it isn't rankable and \Cref{thm:rankable implies termination} is insufficient to prove its termination. We therefore want to find a generalised notion of ranking function so that $M_2$ becomes rankable, and then prove it sound i.e.~rankable in this generalised sense implies AST.

%\paragraph*{Idea}

\begin{definition}
\changed[akr]{The definition of \emph{antitone ranking function} $f$ for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) function $\epsilon : \nnReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
\[f(E[R']) \leq f(E[R]) - \epsilon(f(E[R])) \] 
where $R \to R'$ is the $\tY$-redex rule.}
Any closed term for which an antitone ranking function exists is called \emph{antitone rankable}.
\iffalse
Given a ranking function $f$ on $M \in \Lambda^0$, we say that $f$ is \emph{antitone} if there exists an antitone function\footnote{i.e.~$r < r'$ implies $\epsilon(r) \geq \epsilon(r')$} $\epsilon : \nnReal \to \pReal$ such that %the ranking function $f$ satisfies
\[
f(E[R']) \leq f(E[R]) - \epsilon(f(E[R])) 
\]
for all $E[R'] \in \mathit{Rch}(M)$ where $R \to R'$ is the $\tY$-redex rule.
\fi
\end{definition}

\changed[lo]{Note that antitone ranking functions are actually a generalisation of ranking functions, even though the way we reference them may suggest that they are a type of ranking functions.}

\iffalse
\lo{An aside: This CBV $\tY$-rule seems cleaner:
\[
\big(\tY f^{A \to B} \, x^A \, . \, \theta^B \big) \, v \to \theta[(\tY f \, x \, . \, \theta) / f, v / x].
\]
We assume $\tY f^{A \to B} \, x^A \, . \, \theta^B$ is a value.}
\fi

\begin{definition}
Given a probability space $(\Omega, \calF, \mathbb{P})$, and a supermartingale $(Y_n)_{n \geq 0}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \geq 0}$,
we say that $(Y_n)_{n \geq 0}$ is an \emph{antitone strict supermartingale w.r.t.~$T$} if for all $n \geq 0$, we have $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
\(
\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot {\bf 1}_{\set{T > n}}
\).
%Any closed term for which an antitone ranking function exists is called \emph{antitone rankable}.
\end{definition}

%\lo{The preceding definition does not place any constraint on the stopping time $T$ (other than that it is w.r.t.~the filtration $(\calF_n)_{n \geq 0}$). However for the definition to make sense, $T$ must be the stopping time $T(s) := \min \set{n \mid Y_n(s) = 0}$.}
%\akr{The definition you gave for an antitone ranking function is incorrect, because the antitone ranking functions are a proper subset of ranking functions. Although we may have been referring to antitone ranking functions as a type of ranking function, they're a generalisation, so they don't actually meet the definition of a ranking function in general.}

\begin{theorem}
\label{thm:a-c strict}
Let $(Y_n)_{n \geq 0}$ be an antitone strict supermartingale w.r.t.~stopping time $T$. 
Then $T < \infty$ a.s.
\end{theorem}

\begin{proof}
First, as $(Y_n)$ is a supermartingale, $\expect{Y_n} \leq \expect{Y_0}$. 
Therefore
\begin{calculation}
\expect{Y_n \mid T > n}
   \step[=]{rearranging terms}
% \dfrac{\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
% + 
% \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
% - 
% \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
\dfrac{
\begin{array}{l}
\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
+ 
\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
\\
{} - 
\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}
\end{array}
}{\mathbb P[T > n]}
  \step[=]{definition of conditional expectation}
\dfrac{\expect{Y_n} - \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
  \step[\leq]{$Y_n \geq 0$ always}
\dfrac{\expect{Y_n}}{\mathbb P[T > n]}
  \step[\leq]{$(Y_n)_n$ is a supermartingale}
\dfrac{\expect{Y_0}}{\mathbb P[T > n]}
\end{calculation}

\emph{Claim}: For all $0 < x \leq 1$, 
\(
\mathbb P[T > B_x] \leq x,
\)
where $B_x = \left \lceil \dfrac{\expect{Y_0}+1}{x \; \epsilon(\expect{Y_0} \, x^{-1})} \right \rceil$ and $\epsilon : \nnReal \to \pReal$ is the antitone function.

\smallskip

As the convex hull of $\epsilon$ (the greatest convex function less than or equal to it) satisfies all the conditions assumed of $\epsilon$, in addition to being convex, assume wlog that $\epsilon$ is convex.

Assume for a contradiction that $\mathbb P[T > B_x] > x$.
Then, take $n \leq B_x$. 
We have
%\lo{Change of justification of the second equality below: event, not discrete r.v. Recall: expectation conditioning on r.v.~(resp.~event) is a r.v.~(resp.~number).}
\begin{calculation}
\expect{Y_n-Y_{n+1}}
  \step[=]{$\calF_n \subseteq \calF_{n+1}$, def.~\& linearity of cond.~expectation}
\expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
  \step[\geq]{antitone strict assumption}
\expect{\epsilon(Y_n) \cdot {\bf 1}_{\set{T > n}}}
%  \step[=]{definition of conditional expectation for a discrete variable}
\step[=]{def.~expectation conditioning on an \emph{event}}
\mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
  \step[\geq]{Jensen's inequality}
\mathbb P[T > n]\ \epsilon(\expect{Y_n \mid T > n})
  \step[\geq]{proved earlier}
\mathbb P[T > n]\ \epsilon\Big(\dfrac{\expect{Y_0}}{\mathbb P[T > n]}\Big)
  \step[>]{assumption, $\mathbb P[T > n] \geq \mathbb P[T > B_x] > x$}
x \; \epsilon\big(\expect{Y_0} \, x^{-1}\big).
\end{calculation}
Therefore, by a telescoping sum
\begin{align*}
\expect{Y_{B_x}} 
&= \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})\\
&\leq -1 < 0 
\end{align*}
%\lo{$\expect{Y_0 - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})}$}
which is a contradiction, therefore the claim must be true, therefore $P[T > n]$ tends to 0 as $n$ tends to infinity, therefore $T < \infty$ almost surely.
\end{proof}

Note that it is essential in this proof that $\epsilon$ be defined for all $\mathbb R_{\geq 0}$, not just the values that $(Y_n)_{n \geq 0}$ actually takes (or at least if $(Y_n)_{n \geq 0}$ is uniformly bounded, $\epsilon$ must be positive at the supremum as well as the actual values).

\begin{example}
\iffalse
\akr{Consider a random variable $X$, uniformly distributed in the interval $[-1,1]$, $T := \min \set{n \mid 2^{-n} < X}$, 
\[
Y_k := 
\begin{cases}
4-2^{-k} & \hbox{if $k < T$}\\
0 & \hbox{otherwise}
\end{cases} 
\]
} \lo{The following reformulation defines $T$ and $Y_k$ directly as a measurable function on a probability space (rather than via a r.v.~$X$).}
\fi
Take the probability space $(\Omega, \calF, \mathbb{P})$ where $\Omega$ is the closed interval of reals $[-1, 1]$, $\calF$ the Borel $\sigma$-algebra, and $\mathbb{P}$ the corresponding Lebesgue probability measure.
Let $\omega \in [-1, 1]$. Define random variables $T$ and $(Y_k)_{k \in \omega}$:
\begin{align*}
T(\omega) & := 
\begin{cases}
\min\set{n \in \mathbb{N} \mid \omega > 2^{-n}} & \hbox{if $\omega \in (0, 1]$}\\
\infty & \hbox{otherwise}
\end{cases} 
\\
Y_k(\omega) & := 
\begin{cases}
4-2^{-k} & \hbox{if $\omega \in [-1, 2^{-k}]$ (equivalently $k < T(\omega)$)}\\
0 & \hbox{otherwise}
\end{cases} 
\end{align*}
\lo{N.B.~$\set{T = n} = (\frac{1}{2^{n}}, \frac{1}{2^{n-1}}]$, and so, $\set{T > n} = [-1, \frac{1}{2^n}]$.}
Plainly, $T$ is a stopping time, and $(Y_n)_{n \geq 0}$ is a supermartingale, adapted to the filtration $(\calF_n)_{n \geq 0}$ where $\calF_n = \calF$ for all $n$. 
In this case, $(Y_n)_{n \geq 0}$ either tends to $4$ as $n \to \infty$, or drops to 0 at some point, with an exponentially decreasing probability. 
With respect to the antitone function $\epsilon(x) = \frac{4-x}{4}$ and stopping time $T$, we have that $(Y_n)_{n \geq 0}$ is an antitone strict supermartingale, except that $\epsilon(4) = 0$, and $T = \infty$ with probability $\frac 1 2$, even though $4$ is larger than any value $(Y_n)_{n \geq 0}$ can actually take, and $\epsilon(Y_n)$ never actually reaches $0$.
%\lo{@Andrew: I refer to the preceding highlighted claim. Did you mean: for all $n, s$, $\epsilon(Y_n(s)) > Y_n(s)$?} \akr{No. This phrasing should be clearer.}
\end{example}

%\changed[lo]{Let $f$ be a ranking function on a closed PPCF term $M$. We say that $M$ is \emph{antitone rankable} by $f$ if the supermartingale $(f(M_n))_{n \geq 0}$ and the stopping time $T_M^{\tY}$, both adapted to the fitration $(\calF_n)_{n \geq 0}$ where $\calF_n := \sigma(M_1, \ldots, M_n)$, are antitone strict.}

\begin{theorem}[Antitone ranking function soundness] \label{thm:antitone rankable implies termination}
If a closed PPCF term $M$ is antitone rankable, then $T_M^{\tY} < \infty$ a.s.~(equivalently, $M$ is AST).
\end{theorem}
\begin{proof}
As in %Theorem~\ref{thm:basicRankingSoundness}, 
\Cref{thm:rankable implies termination},
take the probability space $(\entrosp, \Sigma_\entrosp, \mu_\entrosp)$, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$. 
Thanks to \Cref{thm:rankable and strict rankable}, $(Y_n)_{n \geq 0}$ is a supermartingale; and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale w.r.t.~stopping time $T_M^{\tY}$. 
Thus, by \Cref{thm:a-c strict}, $T_M^{\tY} < \infty$ a.s.
\end{proof}

%\paragraph{}
As before, constructing antitone ranking fuctions completely is not necessary, and there is a corresponding notion of an antitone sparse ranking function.

Define an \emph{antitone sparse ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$
\begin{itemize}
\item $f(N) \geq 0$ for all $N$ where $f$ is defined.
\item $f$ is defined at $M$.
\item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}

\begin{theorem}  \label{thm:antitone partial implies rankable}
  Every antitone sparse ranking function is a restriction of an antitone ranking function.
\end{theorem}
\begin{proof}
  Take a closed term $M$ and an antitone sparse ranking function $f$ on $M$, with a corresponding antitone function $\epsilon$. Assume wlog that $\epsilon$ is convex (as if it isn't, we can just take its convex hull instead). As in Theorem~\ref{thm:partial implies rankable}, define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
  \begin{align*}
    f_1(N) &= f(N) \text{ whenever $f(N)$ is defined},\\
    f_1(V) &= 0 \text{ for values $V$ not in the domain of $f$.}
  \end{align*}

  Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) := \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$. 
  The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on antitone partial ranking functions. Define $f_2(N) := \int_\entrosp f_1(\nnext(N,s)) + \epsilon(f_1(\nnext(N,s))) g(N,s) \, \mu_{\entrosp}(\mathrm d s)$. For any term $N$ where $f$ is defined, $f_2(N) = f(N)$, and the value that $f_2$ would have at $N$ if $f$ were not defined at $N$ is $\leq f(N)$, by the third condition on antitone partial ranking functions. In order to show that $f_2$ is an antitone partial ranking function, it therefore suffices to show that the value that $f_2$ would have had at each term if $f$ were not defined at that term is at least the expectation of $f_2$ after one reduction step (plus $\epsilon(f_2(N))$ if the reduction step is a $tY$-reduction). For any term $N$ which is not of the form $E[R]$ for some $\tY$-redex $R$, this is trivial. If $R$ is a $\tY$-redex, then $\epsilon(f_2(N)) \leq \int_\entrosp \epsilon(f_1(\nnext(N,s)))\, \mu_{\entrosp}(\mathrm d s)$ by the convexity of $\epsilon$, because $n$ is bounded. Therefore, the (total) function $f_2$, which agrees with $f$ on $f$'s domain, is an antitone ranking function on $M$ (with the same function $\epsilon$ if it's convex).
\end{proof}

As a corollary, any term which admits an antitone sparse ranking function terminates almost surely.

\paragraph*{\Cref{ex:ac-ranking} revisited}
Programs $M_1$ and $M_2$ are AST. For $M_1$, the sparse ranking function 
\[
\big(\tY \, \lambda f n . \, \tif{n=0}{0}{f \, (n-1) \oplus_{3/2} f \, (n+1)}\big) \, \underline x \mapsto 3x + 1
\] 
suffices to prove its termination (and could equivalently be considered an antitone sparse ranking function with the constant antitone function $\epsilon_1(x) = 1$).

\begin{example}[Unbiased random walk]
\label{ex:unbiased random walk}
For $M_2$, define the function $g_1 : \nnReal \to \nnReal$ by $g_1(x) = \log(x+1) + 1$.
Using shorthand 
$\Theta_2 = \tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
we can define an antitone sparse ranking function $f_2 : \mathit{Rch}(M_2) \rightharpoonup \nnReal$ 
%with antitone function $\epsilon_2$ defined as follows (for $n \in \mathbb N$):
by
\[
f: {\Theta_2} \, n 
\mapsto 
g_2(n), \qquad 
0 \mapsto 0
\]
for $n \in \mathbb N$.
% \begin{align*}
% {\Theta_2} \, n 
% &\mapsto 
% g_2(n)
% \\
% 0 &\mapsto 0.
% \end{align*}
For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$,\footnote{Technically this is not quite correct, because of the distinction between $\underline n + 1$ and $\underline{n + 1}$, but this ranking function can still be made to work by appealing to \Cref{thm:confluent ranking}, as can the others in this section with similar issues.} each with probability $1/2$, with one $\tY$-reduction.
Now
\begin{align*}
& g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 \\
  &=  \log \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) \\
  &=  \frac 1 2 \log\left(\frac{(n+1)^2}{n(n+2)}\right) \\
  &=  \frac 1 2 \log\left(1 + \frac 1 {n(n+2)}\right) \\
  &>  \frac 1 {4n(n+2)} \\
  &>  \frac 1 2 \frac 1 {5(n+1)^2} + \frac 1 2 \frac 1 {5(n+3)^2}
\end{align*}
\changed[lo]{(The first inequality follows from the fact that for all $x \in (0, 1)$, $\log_2 (1 + x) > x$.)}
Moreover $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
Therefore by setting $\epsilon_2(x) = \frac 1 {5(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met. It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone sparse ranking function, and $M_2$ is AST.
\end{example}

\begin{example}[Continuous random walk]\label{ex:raven complex}
In PPCF we can construct a function whose argument changes by a random amount at each recursive call: $\Theta \, 10$ where
\[
%\underbrace{\big
\Theta := \tY \, \lambda f x . \tif{x \leq 0}{0}{f(x - \tsample)}
%}_{\Theta} \, 10
\]
We can construct a sparse ranking function $f$ as follows:
\begin{align*}
\Theta \, \underline l 
&\mapsto 
l + 2
\\
\tif{l \leq 0}{0}{\Theta \, (l - \tsample)}
&\mapsto
l + 1
\\
0 &\mapsto 0.
\end{align*}

For a more complex (not $\tY$-PAST) example, consider the following ``continuous random walk'': $\Xi \, 10$ where
\[
%\underbrace{\big(
\Xi := \tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample + 1/2)} 
%\big)}_{\Xi} \, 10
\]
Let $g(x) := 2 + \log(x + 1)$, and let $\epsilon$ be the function specified by
\[
\epsilon(g(x-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y.
\]
The limit of this as $x \to \infty$ and $g(x+1/2) \to \infty$ is 0, and $\frac {\mathrm d}{\mathrm dx} \epsilon(g(x+1/2)) = \frac 1 {x+1} + \log(1 - \frac 1 {x + 1/2}) < 0$, and $g$ is monotonic increasing; 
therefore $\epsilon$ is antitone and bounded below by 0.
We define an antitone sparse ranking function by:
\[
\Xi \, l 
\mapsto 
g(l), \qquad 
0 \mapsto 0
\]

% \begin{align*}
% \Xi \, l 
% &\mapsto 
% g(l)
% \\
% 0 &\mapsto 0
% \end{align*}
The value of $g$ after one $\tY$-reduction step is at least $g(l-1/2)$, therefore the expectation of $\epsilon$ after one $\tY$-reduction step is at most $\epsilon(g(l-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y$. 
Thus 
\begin{itemize}
\item in case $l > 0$, $g$ decreases by the required amount
\item in case $l \leq 0$, 
\(
g(\Xi \, l) \geq 2 + \log(1/2) > \log(\frac{3 \sqrt 3} 2) - 1 = \epsilon(g(0-1/2))
\) 
as well.
\end{itemize}
Hence this is a valid antitone sparse ranking function and the term is AST.
\end{example}

\begin{example}[Fair-in-the-limit random walk]
\label{ex:Fair-in-the-limit random walk}\citep[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}
\[
\underbrace{\big
(\tY \, f \, x . 
\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)}_{\Xi} 
\, 10
\]
To construct an antitone ranking function, we solve the recurrence relation:
% \[
% \left\{
% \begin{array}{rll}
% z_0 &=& 0\\
% n \geq 1, \quad z_n &>& \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
% \end{array}
% \right.
% \]
\begin{align*}
z_0 &= 0\\
n \geq 1, \quad z_n &> \textstyle \frac{n}{2n + 1} \, z_{n-1} + \frac{n+1}{2n + 1} \, z_{n+1}.
\end{align*}
For $n > 2$, $z_n = \log(n-1)$ works. The expected decrease is $\frac 1 2(\log(1+\frac 1 {(n-1)^2-1}) - \frac 1 {2n+1} \log(1 + \frac 2 {n-2}))$, and using the fact that $\frac x {x+1} \leq \log (1+x) \leq x$ for $x > 0$,\lo{The preceding inequality is incorrect: for $x \in (0, 1)$, $\log (1 + x) > x$.}
\lo{OK. You mean $\ln$ rather than $\log_2$.} this is at least $\frac 1 2(\frac 1 {(n-1)^2} - \frac 1 {2n + 1} \frac 2 {n-2}) = \frac {n^2}{2(n-1)^2(2n+1)(n-2)}$, which (again for $n > 2$) is positive and antitone. For $n = 0, 1, 2$, we take $\epsilon$ to be 9/40 (the same as its value at 3), then set $z_2 = 2 \log 2 - \log 3 - 1, z_1 = 3 \log 2 - 2 \log 3 - 3, z_0 = 4 \log 2 - 3 \log 3 - 6$. Some of those values are negative, but it's still bounded below, so by adding a constant offset it can be corrected, and the term is AST.

\end{example}

\begin{example}[Escaping spline]
\label{ex:escaping spline}\citep[\S 5.4]{DBLP:journals/pacmpl/McIverMKK18}
\[
\underbrace{\big
(\tY \, f \, x . \,
0 \oplus_{\frac{1}{x+1}} f(x + 1) \big)}_{\Xi} 
\, 10
\]

In this case, the fact that the ranking function must decrease at each $\tY$-step (in an antitone sparse ranking function) by the expected value of $\epsilon$ not at the current term, but at the next term where the ranking function is defined, is a little harder to deal with, because the variable $x$ can change all the way to $0$ in one step, therefore simply adding a small offset doesn't suffice to compensate for this fact.

Consider the candidate ranking function $\Xi \, n \mapsto n + 1$. For each $n$, $\Xi \, n$ reduces to either $0$ (with probability $\frac 1 {n + 1}$) or $\Xi \, (n + 1)$, therefore the expected value of the ranking function is $\frac{n(n+2)}{n+1} = n + 1 - \frac 1 {(n+1)^2}$, and the required decrease is $\frac{\epsilon(0) + n \epsilon(n + 2)}{n + 1}$, therefore eventually, the expected decrease isn't enough, whatever the value of $\epsilon(0)$ is.

If instead, we take the ranking function to be defined after the $\tY$-reduction but before the $\tsample$-reduction as well, this can be resolved. Letting $\Theta[n] = \underline 0 \oplus_{\frac 1 n} \Xi \, n$:
\begin{align*}
\Theta[n] &\mapsto n \\
\Xi \, 10 & \mapsto 12.
\end{align*}
For each $n$, $\Theta[n]$ reduces to either $0$ or $\Theta[n+1]$, with a $\tY$-reduction only in the latter case, and the condition that this is an antitone sparse ranking function is that $n \geq \frac{(n-1)(n+1)}{n} + \frac{n-1}{n} \epsilon(n+1)$ and $12 \geq 11 + \epsilon(11)$. These are satisfied by setting $\epsilon(x) = \min(1, 1/x)$, therefore this term is AST.
\end{example}

\paragraph*{Completeness}
It seems very likely that this method is complete in the case that every reachable term is AST, but we have been unable to actually prove this. It is certainly at least capable of proving the termination of terms which terminate arbitrarily slowly (in the sense that there is no similar limitation to Theorem~\ref{thm:rankable implies termination}, which can only prove termination of $\tY$-PAST terms).

The following theorem does not prove completeness, but is suggestive in that direction:
\begin{restatable}{theorem}{antitoneSeemsComplete}
For any stopping time $T$ which is almost surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
\end{restatable}

\medskip

\lo{
Motivation: Non-affine recursion. This could be placed in \Cref{sec:intro}.
}
Many of the recent advances in the development of AST verification methods \citep{DBLP:conf/pldi/ChenH20,DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18,DBLP:conf/aplas/HuangFC18,DBLP:conf/popl/ChatterjeeNZ17,DBLP:journals/pacmpl/AgrawalC018,DBLP:conf/cav/ChatterjeeFG16,DBLP:conf/lics/OlmedoKKM16,DBLP:journals/pacmpl/Huang0CG19} are concerned with loop-based programs.
We can view such loops as tail-recursive programs that are, in particular, \emph{affine recursive}, i.e., 
in each evaluation (or run) of the body of the recursion, recursive calls are made from at most one call site \citep[\S 4.1]{DBLP:journals/toplas/LagoG19}.
(Note that whether a program is affine recursive cannot be checked by just counting textual occurrences of variables.)
Termination analysis of \emph{non-affine recursive} probabilistic programs does not seem to have received much attention.
Methods such as those presented in \citep{DBLP:journals/toplas/LagoG19} are explicitly restricted to affine programs, and are unsound otherwise.
By contrast, many probabilistic programming languages allow for richer recursive structures \citep{DBLP:conf/pkdd/TolpinMW15,DBLP:conf/uai/GoodmanMRBT08,DBLP:journals/corr/MansinghkaSP14}.

\medskip

\begin{example}[Non-affine recursion]
\changed[akr]{Let 
\[
\Xi := \tY \lambda f x. x \oplus_{2/3} f (f (x + 1))
\] 
A sparse ranking function for $\Xi \, 1$ would be $\Xi^n \, i \mapsto 3 \, n$ (for $n \geq 0$), but w.r.t.~the reduction strategy that first reduces the argument of an application to a value, then the function, then the application itself.

For other values of $p > 1/2$, $n/(2p-1)$ works, and for $p = 1/2$, something much like the antitone ranking function for \Cref{ex:unbiased random walk} would work.

A standard sparse ranking function (i.e.~one that doesn't depend on the confluent semantics) would have a slightly more verbose definition, as the functions that accumulate would not be $\Xi$, but what $\Xi$ reduces to after one step.}

\lo{This is a genuine (and simple) \emph{non-affine recursive program}, i.e., recursive programs that can, during the evaluation of the recursive body, make multiple recursive
calls (of a first-order function) from distinct call sites. 
We can use this example to motivate the confluent trace semantics, though it is perhaps not the most compelling possible.
Reason: one could use a more restrictive but equally expressive recursion \emph{operator} (as opposed to a $\tY$-\emph{combinator}, which is a constant) to implement general recursion which would make it unnecessary to consider an alternative reduction strategy when applying the sparse ranking function theorem. 
E.g.~define letrec-terms, $\mathsf{letrec} \, f^{A \to B} \, x^A . M$, as values, via the formation rule
\[{\Gamma, f: A \to B, x:A  \vdash M : B} \over
\Gamma \vdash \mathsf{letrec} \, f^{A \to B} \, x^A . M : A \to B
\]
with the redex rule 
\begin{equation}
(\mathsf{letrec} \, f^{A \to B} \, x^A . M^B) \, V \to M[\mathsf{letrec} \, f \, x . M / f, V / x].
\label{eq:letrec redex}
\end{equation}
}
\end{example}

\lo{
DRAFT. Motivation: Continuous distributions. This could be placed in \Cref{sec:intro}.
}
Sampling from continuous distributions is an essential feature of statistical probabilistic programming languages. (See e.g.~Church \citep{DBLP:conf/uai/GoodmanMRBT08}, Stan \citep{carpenter2017stan}, Anglican \citep{DBLP:conf/pkdd/TolpinMW15}, Gen \citep{cusumano-towner2019Gen}, Pyro \citep{bingham2019Pyro}, Edward \citep{tran2016edward} and Turing \citep{ge2018Turing}.)
Methods of proving AST of probabilistic computation have been developed for probabilistic programs with discrete distributions (see e.g.~\citep{DBLP:journals/toplas/LagoG19,DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/lics/KobayashiLG19,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05}).
To our knowledge, the problem of proving AST of probabilistic programs with continuous distribution is new.

\lo{The following examples are for our discussion at 5:30. Details yet to be checked.}

\begin{example}[Non-affine recursion with continuous distribution]
\label{ex:non-affine with continuous distribution}
Consider 
\[
\Xi := \tY \lambda f x . \big(\mathsf{let} \; e = \tsample \; \mathsf{in} \;
\mathsf{if}\boldsymbol{(}e \leq p, x, X\boldsymbol{)}\big)
\]
where
\(
X = \big( f^3(x+1) \oplus_e f^2(x+1) \big) \oplus_{\mathit{sig}(x)} f^2(x+1) 
\)
and $\mathit{sig}(x)$ is the sigmoid function.

Not only is this program non-affine recursive,
note also the use of a random sample, $e$, as a first-class value, and the subsequent use as a probability in the binary choice $\oplus_e$.
Such a computation cannot be modelled via discrete distributions.

We can use the ranking function method (coupled with the solution of linear recurrence relations) to show that provided $p \geq \sqrt{7} - 2$, the program $\Xi \, \underline r$ is AST for all $r \in \Real$.

\lo{This example is taken from the series of 3d-printing-company running examples (Beutner-Ong 2021).
This paper is as yet unpublished, and effectively uncitable for the time being (because of strict double-blind submission rules).}
\end{example}

\lo{I believe the ranking function method can be used to prove soundness of the Beutner-Ong counting-based method (which reduces the AST problem of first-order (possibly non-affine) recursive PPCF programs to random walk on $\mathbb{N}$), based on Theorem 5.9 and Lemma 5.10 in \emph{op.~cit.}}

\iffalse
\[\begin{array}{l}
\mathsf{let} \; \mathit{add} \; x \; y = x + y \\
\mathsf{letrec} \; \mathit{iter} \; f \; s \; n = 
\tif{n \leq 0}{s}{f \; n \; (\mathit{iter} \; f \; s \; (n - 1)}
\end{array}
\]
\fi

\lo{
DRAFT. Motivation: higher-order recursion. This could be placed in \Cref{sec:intro} or \Cref{sec:related}.
}
There is an obvious source of deterministic (i.e.~non-probabilistic) higher-order functions defined by recursion, viz., \emph{higher-order recursion schemes} (HORS) (see e.g.~\citep{DBLP:conf/lics/Ong06,DBLP:conf/lics/Ong15}).
(Incidentally the (sparse) ranking function method is just as applicable to the termination analysis of \emph{deterministic} PPCF programs.)
Recently \cite{DBLP:conf/lics/KobayashiLG19} have extended HORS to \emph{probabilistic higher-order recursion schemes} (PHORS), which are HORS augmented with probabilistic (binary) branching $\oplus_p$.
As HORS are in essence the $\lambda \tY$-calculus \citep{DBLP:conf/lics/Statman02} (i.e.~pure simply-typed lambda calculus with recursion, generated from a finite base type), PHORS are definable in PPCF:
(order-$n$) PHORS is encodable as (order-$n$) (call-by-name) PPCF, but the former is strictly less expressive (because the underlying recursion schemes are not Turing complete). 
Some interesting PPCF terms such as \Cref{ex:non-affine with continuous distribution} %\refExample{complexExample} 
cannot be expressed as PHORS.
A relevant result here is that the AST problem is decidable for order-1 PHORS (by reduction to the PSPACE-hard solvability of finite systems of polynomial equations with real coefficients; see \citep{DBLP:journals/jacm/EtessamiY09}), but undecidable for order-2 PHORS.

\begin{example}[Higher-order recursion]
Consider the higher-order function
$\Xi : (\textsf{R} \to \textsf{R} \to \textsf{R}) \to \textsf{R} \to \textsf{R} \to \textsf{R}$
recursively defined\footnote{Inspired by examples in \citep{DBLP:journals/pacmpl/BurnOR18,DBLP:conf/lics/OngW19}.} by
\[
\begin{array}{l}
\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} . \\
\quad\tif{n \leq 0}{s}{
f \, n \, (\varphi \, f \, s \, (n - 1))
\oplus_p
f \, (n+1) \, (\varphi \, f \, s \, n)
}
\end{array}
\]
\lo{QUESTION: Let $g : \textsf{R} \to \textsf{R} \to \textsf{R}$ be a function that terminates on all arguments.
Then $\Xi \, g$ is AST for $p \geq 1/2$.}
\akr{In order to actually construct a ranking function, it will be necessary to assume a bound on the number of $\tY$-reduction steps in the reduction of $g \, \underline n \, \underline m$ too.}

\akr{I suspect that the term you actually had in mind for the higher-order recursion example was different from what you wrote, because you said that it terminates for $p \geq 1/2$, but it actually terminates for all $p > 0$. Passing $n+1$ as an argument to $f$ doesn't actually make the recursion any longer. Did you mean 
\[
\begin{array}{l}
\tY \lambda \varphi f s n. \\
\qquad \tif{n \leq 0}{s}{ f \, n \, (\varphi \, f \, s \, (n - 1)) \oplus_p f \, n \, (\varphi \, f \, s \, (n+1))}
\end{array}
\]}
\end{example}

### 2021-10-01 LICS21/antitone-proofs.tex
" \ref" -> "~\ref"

--- 2021-09-27 LICS21/antitone-proofs.tex
@@
 Assume wlog that $\epsilon$ is convex (as if it isn't, we can just take its convex hull instead).
-As in Theorem \ref{thm:partial implies rankable}, define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
+As in Theorem~\ref{thm:partial implies rankable}, define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
   \begin{align*}
@@
 
-\subsection*{\Cref{ex:raven complex} (Continuous random walk)}
+\subsection*{Example~\ref{ex:raven complex} (Continuous random walk)}
 Let $\Theta := \tY \, \lambda f x . \tif{x \leq 0}{0}{f(x - \tsample)}$. 
@@
 
-\subsection*{\Cref{ex:Fair-in-the-limit random walk} (Fair-in-the-limit random walk) \citep[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}}
+\subsection*{Example~\ref{ex:Fair-in-the-limit random walk} (Fair-in-the-limit random walk) \citep[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}}
 \[
@@
 
-\subsection*{\Cref{ex:non-affine recursion easy} (Non-affine recursion)}
+\subsection*{Example~\ref{ex:non-affine recursion easy} (Non-affine recursion)}
 Let 
@@
 
-\subsection*{\Cref{ex:non-affine continuous} (More complex non-affine recursion)}
+\subsection*{Example~\ref{ex:non-affine continuous} (More complex non-affine recursion)}
 Let
@@
 (order-$n$) PHORS is encodable as (order-$n$) call-by-name PPCF, but the former is strictly less expressive (because the underlying HORS is not Turing complete). 
-For example, the PPCF term in \Cref{ex:higher-order recursion} is not definable in PHORS.
+For example, the PPCF term in Example~\ref{ex:higher-order recursion} is not definable in PHORS.
 %Some interesting PPCF terms such as \Cref{ex:non-affine with continuous distribution} %\refExample{complexExample} 
@@
 
-\subsection*{\Cref{ex:higher-order recursion} (Higher-order recursion)}
+\subsection*{Example~\ref{ex:higher-order recursion} (Higher-order recursion)}
 Consider the higher-order function
@@
 \] 
-where $g_2$ is as in \Cref{ex:unbiased random walk}, using the same antitone function too.
+where $g_2$ is as in Example~\ref{ex:unbiased random walk}, using the same antitone function too.
 
-The inequalities required for this to be an antitone partial supermartingale are satisfied with just the same reasoning as in \Cref{ex:unbiased random walk}, therefore this term too is antitone rankable.
+The inequalities required for this to be an antitone partial supermartingale are satisfied with just the same reasoning as in Example~\ref{ex:unbiased random walk}, therefore this term too is antitone rankable.
 
 Again, this is not actually the correct reduction order, in that $F$ should be applied to its argument $n_i$ and reduced to a value before its second argument is expanded, 
-but this would complicate the presentation, and this version can be justified by \Cref{thm:confluent ranking} instead. 
+but this would complicate the presentation, and this version can be justified by Theorem~\ref{thm:confluent ranking} instead. 
 The reduction strategy implied here should be clear enough (assuming that, where it isn't specified, it just matches $\cbv$), but to be more precise, 

### 2021-10-01 LICS21/antitone.tex
" \ref" -> "~\ref"

--- 2021-09-27 LICS21/antitone.tex
@@
 
-The term $M_1$ is rankable, terminating in 31 $\tY$-steps on average, and $M_3$ only has a $1/1024$ chance of terminating, but in between, $M_2$ is AST, but isn't $\tY$-PAST, therefore it isn't rankable and \Cref{thm:rankable implies termination} is insufficient to prove its termination.
+The term $M_1$ is rankable, terminating in 31 $\tY$-steps on average, and $M_3$ only has a $1/1024$ chance of terminating, but in between, $M_2$ is AST, but isn't $\tY$-PAST, therefore it isn't rankable and Theorem~\ref{thm:rankable implies termination} is insufficient to prove its termination.
 Thus we seek a generalised notion of ranking function so that $M_2$ becomes rankable, and then prove it sound i.e.~rankable in this generalised sense implies AST.
@@
 \begin{proof}
-As in %Theorem \ref{thm:basicRankingSoundness}, 
-\Cref{thm:rankable implies termination},
+As in %Theorem~\ref{thm:basicRankingSoundness}, 
+Theorem~\ref{thm:rankable implies termination},
 take the probability space $(\entrosp, \Sigma_\entrosp, \mu_\entrosp)$, and define the same r.v.~$T_n, X_n, Y_n, T_M^{\tY}$. 
-Thanks to \Cref{thm:rankable and strict rankable}, $(Y_n)_{n \geq 0}$ is a supermartingale; and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale w.r.t.~stopping time $T_M^{\tY}$.
-Thus, by \Cref{thm:a-c strict}, $T_M^{\tY} < \infty$ a.s.
+Thanks to Theorem~\ref{thm:rankable and strict rankable}, $(Y_n)_{n \geq 0}$ is a supermartingale; and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale w.r.t.~stopping time $T_M^{\tY}$.
+Thus, by Theorem~\ref{thm:a-c strict}, $T_M^{\tY} < \infty$ a.s.
 \end{proof}
@@
 
-\paragraph*{\Cref{ex:ac-ranking} revisited}
+\paragraph*{Example~\ref{ex:ac-ranking} revisited}
 Programs $M_1$ and $M_2$ are AST. For $M_1$, the sparse ranking function 
@@
 % \end{align*}
-For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$,\footnote{Technically this is not quite correct, because of the distinction between $\underline n + 1$ and $\underline{n + 1}$, but this ranking function can still be made to work by appealing to the Confluent Ranking \Cref{thm:confluent ranking}, as can the others in this section with similar issues.
-See \cref{rem:ex corrected using confluent semantics} for a fuller account.} each with probability $1/2$, with one $\tY$-reduction.
+For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$,\footnote{Technically this is not quite correct, because of the distinction between $\underline n + 1$ and $\underline{n + 1}$, but this ranking function can still be made to work by appealing to the Confluent Ranking Theorem~\ref{thm:confluent ranking}, as can the others in this section with similar issues.
+See Remark~\ref{rem:ex corrected using confluent semantics} for a fuller account.} each with probability $1/2$, with one $\tY$-reduction.
 Now
@@
 
-\paragraph*{More challenging examples} All of the following examples also have antitone ranking functions, provided in Appendix \ref{apx:antitone}.
+\paragraph*{More challenging examples} All of the following examples also have antitone ranking functions, provided in Appendix~\ref{apx:antitone}.
 
@@
 \paragraph*{Completeness}
-It seems very likely that this method is complete in the case that every reachable term is AST (\cref{conj:antitone}), but we have been unable to actually prove this.
-It is certainly at least capable of proving the termination of terms which terminate arbitrarily slowly (in the sense that there is no similar limitation to \cref{thm:rankable implies termination}, which can only prove termination of $\tY$-PAST terms).
+It seems very likely that this method is complete in the case that every reachable term is AST (Conjecture~\ref{conj:antitone}), but we have been unable to actually prove this.
+It is certainly at least capable of proving the termination of terms which terminate arbitrarily slowly (in the sense that there is no similar limitation to Theorem~\ref{thm:rankable implies termination}, which can only prove termination of $\tY$-PAST terms).
 

### 2021-10-01 LICS21/applications.tex
" \ref" -> "~\ref"

--- 2021-09-27 LICS21/applications.tex
@@
 \label{sec:applications}
-Although the definition of potential positions in a term in \Cref{sec:confluent} was intended simply to define a variant of trace semantics that has a restricted version of confluence, it is likely to also be applicable to inference algorithms (for probabilistic programs) as well.
+Although the definition of potential positions in a term in Section~\ref{sec:confluent} was intended simply to define a variant of trace semantics that has a restricted version of confluence, it is likely to also be applicable to inference algorithms (for probabilistic programs) as well.
 
@@
 Another possible application for the confluent trace semantics would be as a justification, within a trace-style semantics, of some forms of equational reasoning. 
-By introducing reduction steps that go backwards to terms not reachable from the original term, it may in some cases be simplified, or equivalences between terms may be proved, by appealing to confluence results such as \Cref{churchRosser} and \Cref{thm:CbvIsTerminatingest}.
+By introducing reduction steps that go backwards to terms not reachable from the original term, it may in some cases be simplified, or equivalences between terms may be proved, by appealing to confluence results such as Lemma~\ref{churchRosser} and Theorem~\ref{thm:CbvIsTerminatingest}.
 \lo{Can you provide an example (and present it in the appendix)?} \akr{I don't have any compelling example for this in mind.}

### 2021-10-01 LICS21/confluent-proofs.tex
" \ref" -> "~\ref"

@@
 In the other direction, suppose $(N_1, \alpha_1) \sim_M^* (N_2, \alpha_2)$ where both $N_1$ and $N_2$ are descendants of $N$.
-By Lemma \ref{downAcrossUp}, take $(N_1, \alpha_1) \sim_{p,M}^* (N'_1, \alpha') \sim_{c,M}^* (N'_2, \alpha') \sim_{p,M}^* (N_2, \alpha_2)$.
+By Lemma~\ref{downAcrossUp}, take $(N_1, \alpha_1) \sim_{p,M}^* (N'_1, \alpha') \sim_{c,M}^* (N'_2, \alpha') \sim_{p,M}^* (N_2, \alpha_2)$.
 The $\sim_p^*$ steps remain within $\mathit{Rch}(N)$, and $\sim_p$ does not depend on the history of the reduction sequences, therefore $(N_1, \alpha_1) \sim_{p,N}^* (N'_1, \alpha') \sim_{c,M}^* (N'_2, \alpha') \sim_{p,N}^* (N_2, \alpha_2)$.
 
-Let $X$ be the call-by-value reduction sequence related to both $N'_1$ and $N'_2$ by $\sim_{c,M}$ given by Lemma \ref{canonicalCousins}. 
-As $M \to N$ is the first reduction in the sequence $N'_1$, it is the last to be affected by the $\sim_{c,M}$ sequence $N'_1 \sim_{c,M}^* X$ given by Lemma \ref{canonicalCousins} therefore it can be split into $N'_1 \sim_{c,M}^* Y_1 \sim_{c,M}^* X$ where $Y_1$ is in CBV order except possibly for its first reduction, which is still $M \to N$.
+Let $X$ be the call-by-value reduction sequence related to both $N'_1$ and $N'_2$ by $\sim_{c,M}$ given by Lemma~\ref{canonicalCousins}. 
+As $M \to N$ is the first reduction in the sequence $N'_1$, it is the last to be affected by the $\sim_{c,M}$ sequence $N'_1 \sim_{c,M}^* X$ given by Lemma~\ref{canonicalCousins} therefore it can be split into $N'_1 \sim_{c,M}^* Y_1 \sim_{c,M}^* X$ where $Y_1$ is in CBV order except possibly for its first reduction, which is still $M \to N$.
 Let the position of the reduction $M \to N$ be $\beta$.
@@
 In the case that $(M,s) \Rightarrow^* (M_1,s_1)$ or $(M,s) \Rightarrow^* (M_2,s_2)$ by multiple steps, it is possible to repeatedly replace a pair of the form $A \Leftarrow B \Rightarrow C$ by $A \Rightarrow^* D \Leftarrow^* C$ by the construction above, but it is not immediate that this process terminates, because each $\Rightarrow$ or $\Leftarrow$ may be replaced by multiple, so the sequence of $\Rightarrow$s and $\Leftarrow$s from $(M_1,s_1)$ to $(M_2,s_2)$ may get longer at some steps.
-However, termination can be proved by noting that the structure of this process is identical to the process of swapping $\sim_\uparrow$ and $\sim_\downarrow$ steps in Lemma \ref{downAcrossUp}.
+However, termination can be proved by noting that the structure of this process is identical to the process of swapping $\sim_\uparrow$ and $\sim_\downarrow$ steps in Lemma~\ref{downAcrossUp}.
 To be more precise, in this case there is a sequence of $\Leftarrow$ and $\Rightarrow$ steps, each of which has an associated reduction position and initial and final terms, and if a $\Leftarrow$ immediately precedes a $\Rightarrow$, they may be swapped to produce some number of $\Rightarrow$s, followed by some number of $\Leftarrow$s.
-In the case of Lemma \ref{downAcrossUp}, there is a sequence of $\sim_\uparrow$ and $\sim_\downarrow$ steps, each of which has an associated reduction position and initial and final skeletons, and if a $\sim_\uparrow$ immediately precedes a $\sim_\downarrow$, they may be swapped to produce some number of $\sim_\downarrow$s followed by some number of $\sim_\uparrow$s.
+In the case of Lemma~\ref{downAcrossUp}, there is a sequence of $\sim_\uparrow$ and $\sim_\downarrow$ steps, each of which has an associated reduction position and initial and final skeletons, and if a $\sim_\uparrow$ immediately precedes a $\sim_\downarrow$, they may be swapped to produce some number of $\sim_\downarrow$s followed by some number of $\sim_\uparrow$s.
 In both cases, the number and reduction positions of the resultant steps are determined by the case of $\sim_c$ that matches the way the initial reduction positions overlap, and the initial skeleton (or the skeleton of the initial term).
-The same argument that the process in Lemma \ref{downAcrossUp} terminated is therefore applicable here too.
+The same argument that the process in Lemma~\ref{downAcrossUp} terminated is therefore applicable here too.
 At every stage, the sequence of reductions that results from one of the initial $\Rightarrow$s is a parallel sequence, and swapping a parallel sequence of $\Rightarrow$s with a $\Leftarrow$ always terminates, therefore each $\Rightarrow$ in turn can be moved past all of the $\Leftarrow$s, and the process as a whole will terminate in a state where all of the $\Rightarrow$s precede all of the $\Leftarrow$s, i.e.~a pair of reduction sequences $(M_1,s_1) \Rightarrow^* (M',s') \Leftarrow^* (M_2,s_2)$.
@@
 \begin{proof}
-Suppose on the contrary that $(M,\alpha) \sim^* (A,\beta)$, where $N \to^* A$, then by Lemma \ref{simpDoesn'tMix}, $(M,\alpha) \sim_p^* (A,\beta)$.
+Suppose on the contrary that $(M,\alpha) \sim^* (A,\beta)$, where $N \to^* A$, then by Lemma~\ref{simpDoesn'tMix}, $(M,\alpha) \sim_p^* (A,\beta)$.
 $M \neq A$ therefore $(M,\alpha) \sim_p$ some position in $N$, but in all the cases of the definition of $\sim_p$, no position in the child term is related to the position of the redex.
@@
 
-Essentially what Lemma \ref{lem:redexDestroyed} demonstrates is that the samples taken during any reduction sequence are independent of each other. This is made more precise in the following lemmas.
+Essentially what Lemma~\ref{lem:redexDestroyed} demonstrates is that the samples taken during any reduction sequence are independent of each other. This is made more precise in the following lemmas.
 
@@
 The distribution of $s(k)$ conditional on reaching $N$ is therefore the uniform distribution on $I$.
-Similarly for the other semantics, if the distribution of remaining samples after $n$ steps is independent of the term by Lemma \ref{lem:independentSamples}, therefore the distribution of $s(\mathit{Sk}(N),\cbv(N))$ conditional on $N$ is the uniform distribution on $I$.
+Similarly for the other semantics, if the distribution of remaining samples after $n$ steps is independent of the term by Lemma~\ref{lem:independentSamples}, therefore the distribution of $s(\mathit{Sk}(N),\cbv(N))$ conditional on $N$ is the uniform distribution on $I$.
 In either case, the distribution of the next term conditional on the previous term, in the case that it reduces randomly, is equal to the image under $r \mapsto E[\underline r]$ of the uniform distribution on $I$.
@@
 Let $(M,s) = (M_0,s_0) \Rightarrow_f \dots \Rightarrow_f (M_n,s_n)$, where $M_n$ is a value.
-By Lemma \ref{canonicalCousins}, the reduction sequence $(\mathit{Sk}(M_i))_i$ is related by $\sim_c^*$ to some skeletal reduction sequence $X$ which is in call-by-value order.
+By Lemma~\ref{canonicalCousins}, the reduction sequence $(\mathit{Sk}(M_i))_i$ is related by $\sim_c^*$ to some skeletal reduction sequence $X$ which is in call-by-value order.
 For each skeletal reduction sequence $X_j$ in the sequence $(\mathit{Sk}(M_i))_i \sim_c \dots \sim_c X$, it is possible to define a reduction sequence of terms such that $(M,s) = (N_{j,0},s_{j,0}) \Rightarrow \dots \Rightarrow (N_{j,n_j},s_{j,n_j})$ and $\mathit{Sk}(N_{j,k}) = X_{j,k}$, by at each step applying the reduction at the same place as in $X_j$, essentially just filling in the reals in the skeletal reduction sequence to make it a reduction sequence of terms (it will be shown shortly that the correct branch is taken in each $\textsf{if}$-reduction).
@@
 To prove that this holds true for $j+1$ as well, it suffices to show that the corresponding term and samples immediately after the reductions involved in this $\sim_c$ step match, as from that point on, the $\Rightarrow$-sequences will match, as they start from the same point and have all the same reduction positions.
-Letting the term and remaining samples immediately before the reductions involved in this $\sim_c$ step be $(N,t)$ (which is the same in both $N_j$ and $N_{j+1}$ for the same reason again, they start at the same point and the sequences of reduction positions up to this point are the same), and the terms and samples after these reductions be $(O_1,t_1)$ and $(O_2,t_2)$, so that these match the $N$, $O_1$ and $O_2$ in the definition of $\sim_c$, $t_1 = t_2$ because both of them are, by Lemma \ref{lem:sim-M-N}, equal to the restriction of $t$ to the subset of $\mathit{Rch}(N)$ that starts with the reduction sequences to $O_1$ and $O_2$ respectively, and the potential positions in those subsets of $\mathit{Rch}(N)$ are identified with each other by $\sim$ because of the same case of $\sim_c$ that means $N_j$ and $N_{j+1}$ are related.
+Letting the term and remaining samples immediately before the reductions involved in this $\sim_c$ step be $(N,t)$ (which is the same in both $N_j$ and $N_{j+1}$ for the same reason again, they start at the same point and the sequences of reduction positions up to this point are the same), and the terms and samples after these reductions be $(O_1,t_1)$ and $(O_2,t_2)$, so that these match the $N$, $O_1$ and $O_2$ in the definition of $\sim_c$, $t_1 = t_2$ because both of them are, by Lemma~\ref{lem:sim-M-N}, equal to the restriction of $t$ to the subset of $\mathit{Rch}(N)$ that starts with the reduction sequences to $O_1$ and $O_2$ respectively, and the potential positions in those subsets of $\mathit{Rch}(N)$ are identified with each other by $\sim$ because of the same case of $\sim_c$ that means $N_j$ and $N_{j+1}$ are related.
 If none of the relevant reductions are $\tsample$-reductions, $O_1 = O_2$ trivially.
 If there are samples taken (which must be the arbitrary reduction at $\beta$), it is also required that the same samples are taken in both branches.
-Similarly to Lemma \ref{churchRosser}, the positions of these $\tsample$-reductions in each branch are related by $\sim_p^*$, therefore they correspond to the same sample in $t$, therefore $O_1 = O_2$.
+Similarly to Lemma~\ref{churchRosser}, the positions of these $\tsample$-reductions in each branch are related by $\sim_p^*$, therefore they correspond to the same sample in $t$, therefore $O_1 = O_2$.
 The fact that the terms are equal also implies that the same branch of an $\textsf{if}$ statements will be taken.
@@
 Although $T'$ itself may have a measure of 0 as a subset of $I^{L_s(M)}$, it still has a natural measure-space structure of its own.
-It also has the properties that $M$ does not terminate with $\cbv$ and any trace in $T'$ (therefore $M$ also does not terminate with $r$ and any trace in $T'$, by Theorem \ref{thm:CbvIsTerminatingest}), and for any trace $t \in T'$, if $(M,t) \Rightarrow^k_r (M',t')$, none of the potential positions in $M'$ correspond to any of the elements of $P \setminus P_i$.
+It also has the properties that $M$ does not terminate with $\cbv$ and any trace in $T'$ (therefore $M$ also does not terminate with $r$ and any trace in $T'$, by Theorem~\ref{thm:CbvIsTerminatingest}), and for any trace $t \in T'$, if $(M,t) \Rightarrow^k_r (M',t')$, none of the potential positions in $M'$ correspond to any of the elements of $P \setminus P_i$.
 There are finitely many skeletal reduction sequences $k$ steps long starting at $M$ and following $r$, each one corresponding to some measurable subset of $I^{L_s(M)}$, so pick one such that its intersection with $T'$ is a non-null subset of $T'$, let this subset be $Q$.
@@
 
-Example \ref{ex:sum of powers}
+Example~\ref{ex:sum of powers}
 Recall that

### 2021-10-01 LICS21/confluent.tex
" \ref" -> "~\ref"

--- 2021-04-30 LICS21/confluent.tex
@@
 or the number of cases necessary for defining a sparse ranking function is excessive because the reduction does not proceed in a neat and orderly manner. 
-In cases such as these (e.g.~\cref{ex:sum of powers}), it may be much more convenient to instead consider ranking functions with respect to alternative reduction strategies constructed to match the individual term in question,
+In cases such as these (e.g.
+Example~\ref{ex:sum of powers}), it may be much more convenient to instead consider ranking functions with respect to alternative reduction strategies constructed to match the individual term in question,
 thereby reducing the number of cases needed and simplifying the reachable terms.
@@
 In order to allow potential positions in different reduction sequences to be considered equivalent, a relation $\sim^*$ is defined, and finally, all of these are used to construct a confluent version of the trace semantics, $\Rightarrow$, that is still nondeterministic in reduction order, but does specify the outcome of random choices.
-The desired properties of this semantics (confluence and equivalence to the standard semantics) are proved (\Cref{churchRosser,thm:AstEquivalence}), then it is used to give extended versions of ranking functions and the related AST results, \Cref{thm:confluent ranking}.
+The desired properties of this semantics (confluence and equivalence to the standard semantics) are proved (Lemma~\ref{churchRosser} and Theorem~\ref{thm:AstEquivalence}), then it is used to give extended versions of ranking functions and the related AST results, Theorem~\ref{thm:confluent ranking}.
 
@@
 
-Thus we can define a more general reduction relation\footnote{In this section $\to$ always means the relation defined in \cref{def:more general red}.} $\to$.
+Thus we can define a more general reduction relation\footnote{In this section $\to$ always means the relation defined in Definition~\ref{def:more general red}.} $\to$.
 
@@
 Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a $\to$-reduction sequence of skeletons starting at $\mathit{Sk}(M)$, and the second of which is a position in the final skeleton of the reduction sequence.
-As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the standard trace semantics, modified traces, which are elements of $I^{L_0(M)}$ (with one more caveat introduced after Def.
-\ref{defn:sim}), will be used to pre-select a sample from $I$ for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.
+As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the standard trace semantics, modified traces, which are elements of $I^{L_0(M)}$ (with one more caveat introduced after Def.~\ref{defn:sim}), will be used to pre-select a sample from $I$ for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.
 
@@
 \end{asparaenum}
-The $\sim_c$-rules are illustrated in \Cref{sim-diagram}.
+The $\sim_c$-rules are illustrated in Figure~\ref{sim-diagram}.
 \end{definition}
@@
 \begin{example}
-In \Cref{ex:cousin sim}, $(M,\underline{+}_1) \sim_p (M \to N_2,\underline{+}_1)$ by case i of $\sim_p$ (because the reduction $M \to N_2$ occurs at $\underline{+}_2$, which is disjoint from $\underline{+}_1$), and similarly, $(M,\underline{+}_2) \sim_p (M \to N_1,\underline{+}_2)$.
+In Example~\ref{ex:cousin sim}, $(M,\underline{+}_1) \sim_p (M \to N_2,\underline{+}_1)$ by case i of $\sim_p$ (because the reduction $M \to N_2$ occurs at $\underline{+}_2$, which is disjoint from $\underline{+}_1$), and similarly, $(M,\underline{+}_2) \sim_p (M \to N_1,\underline{+}_2)$.
 
@@
 There are then reduction sequences $M_{sss} \to M_{Xss} \to M_{XXs} \to M_{XXX} \to \skeletonPlaceholder$ and $M_{sss} \to M_{sXs} \to M_{XXs} \to M_{XXX} \to \skeletonPlaceholder$.
-For the first two reductions, these reduction sequences take the same samples by $\sim_p$, case i, as in \Cref{ex:cousin sim}~.
+For the first two reductions, these reduction sequences take the same samples by $\sim_p$, case i, as in Example~\ref{ex:cousin sim}~.
 The next reduction uses the samples labelled by $(M_{sss} \to M_{Xss} \to M_{XXs}, \underline{+}_3)$ and $(M_{sss} \to M_{sXs} \to M_{XXs}, \underline{+}_3)$, which are related by $\sim_c$, case a, therefore when these reduction sequences reach $M_{XXX}$, they still contain all the same numbers, as desired.
@@
 At each reduction step $M \to N$, the sample space must be restricted from $I^{L_s(M)}$ to $I^{L_s(N)}$. 
-The injection $L_0(N) \to L_0(M)$ is trivial to define by appending $\mathit{Sk}(M) \to \mathit{Sk}(N)$ to each path, and using \Cref{lem:sim-M-N}, this induces a corresponding injection on the quotient, $L(N) \to L(M)$.
+The injection $L_0(N) \to L_0(M)$ is trivial to define by appending $\mathit{Sk}(M) \to \mathit{Sk}(N)$ to each path, and using Lemma~\ref{lem:sim-M-N}, this induces a corresponding injection on the quotient, $L(N) \to L(M)$.
 The corresponding map $L_s(N) \to L_s(M)$ is then denoted $i(M \to N)$.
@@
 This reduction relation now has all of the properties required of it.
-In particular, it can be considered an extension of the standard trace semantics (as will be seen later in \Cref{thm:AstEquivalence}), and also:
+In particular, it can be considered an extension of the standard trace semantics (as will be seen later in Theorem~\ref{thm:AstEquivalence}), and also:
 \begin{restatable}{lemma}{churchRosser} \label{churchRosser}
@@
 Let the set of samples with which it terminates with this reduction strategy be $X$.
-By \Cref{thm:CbvIsTerminatingest}, $M$ also terminates with $\cbv$ and every element of $X$, and $X$ has measure 1, by assumption, therefore $M$ is AST with respect to $\cbv$ therefore by \Cref{thm:AstEquivalence} it is AST.
+By Theorem~\ref{thm:CbvIsTerminatingest}, $M$ also terminates with $\cbv$ and every element of $X$, and $X$ has measure 1, by assumption, therefore $M$ is AST with respect to $\cbv$ therefore by Theorem~\ref{thm:AstEquivalence} it is AST.
 \end{proof}
@@
 All of the theorems on the termination of rankable terms therefore extend to other reduction strategies too.
-The proofs of \Cref{thm:rankable implies termination,thm:partial implies rankable,thm:antitone rankable implies termination,thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies.
-They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because \Cref{thm:de groote} applies equally to any reduction strategy).
+The proofs of Theorem~\ref{thm:rankable implies termination,thm:partial implies rankable,thm:antitone rankable implies termination,thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies.
+They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because Theorem~\ref{thm:de groote} applies equally to any reduction strategy).
 For a reduction strategy $r$ and term $M$, just substitute $N[r(N)]$ being a $\tY$-redex for $N$ being of the form $E[\tY \lambda x.
@@
 The only restrictions to bear in mind on evaluating sub-terms are that a $\tsample$ may not be evaluated inside a $\lambda$ that's inside a $\tY$ or on the right of an application, and a $\beta$ redex may not be reduced until its argument is a value.
-This can be seen as applied to \Cref{ex:sum of powers} at the end of \Cref{apx:confluent}.
+This can be seen as applied to Example~\ref{ex:sum of powers} at the end of Appendix~\ref{apx:confluent}.
 
 \medskip
-If, as seems likely, \Cref{thm:antitone rankable implies termination} is complete for terms from which every reachable term is AST, then by the following theorem, \Cref{thm:confluent ranking} can't actually prove the termination of any terms not already provable by \Cref{thm:antitone rankable implies termination} and \Cref{thm:antitone partial implies rankable}.
+If, as seems likely, Theorem~\ref{thm:antitone rankable implies termination} is complete for terms from which every reachable term is AST, then by the following theorem, Theorem~\ref{thm:confluent ranking} can't actually prove the termination of any terms not already provable by Theorem~\ref{thm:antitone rankable implies termination} and Theorem~\ref{thm:antitone partial implies rankable}.
 
@@
 It is still antitone rankable, but this is much more complicated.
-Details of the ranking function construction with and without using the confluent semantics are given in \Cref{apx:confluent}.
+Details of the ranking function construction with and without using the confluent semantics are given in Appendix~\ref{apx:confluent}.
 \end{example}
@@
 Even some of the examples given earlier implicitly use slightly non-standard reduction strategies to make the definitions of their ranking functions slightly simpler.
-For example, the antitone sparse ranking function in \Cref{ex:unbiased random walk} was defined at $\Theta_2\, \underline n$ (where $\Theta_2 = \tY \, \lambda f n .
+For example, the antitone sparse ranking function in Example~\ref{ex:unbiased random walk} was defined at $\Theta_2\, \underline n$ (where $\Theta_2 = \tY \, \lambda f n .
 \, \tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$), but $\Theta_2\, \underline{10}$ actually reduces to $\Theta_2\, (\underline{10} - 1)$ or $\Theta_2\, (\underline{10} + 1)$, and (taking the second of these), it then reduces to $(\lambda z.
@@
 
-\Cref{ex:higher-order recursion} also makes use of an alternative reduction strategy.
+Example~\ref{ex:higher-order recursion} also makes use of an alternative reduction strategy.
 In this case, it is to allow the antitone sparse ranking function to skip over all of those terms where the unknown function $F$ is part way through evaluation.

### 2021-10-01 LICS21/grid-game.tex
" \ref" -> "~\ref"

\documentclass{article}
\usepackage{amsmath,amssymb,latexsym}

\begin{document}
The problem I discussed during our meeting today would, more precisely, be something like this.

Given a function $F$ from $\mathbb N^d \to X$, define the $n$-neighbourhood of a point $x \in \mathbb N^d$ to be $f_{x, n} : [-n,n]^d \to X$, $f_{x,n}(y) = f(x+y)$ (where the interval is a subset of $\mathbb N$.

Given a prior probability distribution on $X$ and a weighting function $w : \mathbb N^d \times (X^{[-n,n]^d}) \to \mathbb R$, define the associated probability distribution on $X^{\mathbb N^d}$ by setting its marginal distribution on any finite set of points $P$ equal to the limit as the finite region $R \subset \mathbb N^d$ goes to $\mathbb N^d$ of the marginal distribution on $P$ of the distribution obtained by multiplying the prior (given by the product distribution of $|R|$ copies of $X$) by the posterior (given by the product of $w(x, \hbox{the $n$-neighbourhood of $x$})$ for every $x$ such that the neighbourhood is contained in $R$) and re-normalising, assuming the limit exists (which it sometimes doesn't).

The problem, then, is to provide an algorithm that, given $n$, the prior distribution on $X$, the weighting function $w$, a tolerance value $\epsilon > 0$, a point $x$ and a trace/source of randomness, will output an element of $X$ in such a way that the joint distribution obtained by evaluating the algorithm with the same trace for all points in a region $P$ has an error of at most $\epsilon |P|$ (in the sense that the information of the algorithm's output distribution relative to the correct distribution on $P$ is at most $\epsilon |P|$).

There are a lot of omitted details in terms of how everything is encoded, but I don't expect that to matter all that much. Also, I'm reasonably confident that this problem doesn't have a full solution (there's probably some sort of way of encoding halting problems in this form), but it might well have a partial solution given some sort of restriction on $w$.

The way of defining the global distribution in terms of local weights that I used here is standard in statistical mechanics. A solution for $n=1$ would imply a solution for arbitrary $n$.

\paragraph{Example}

One application of such an algorithm would be sampling a state of the Ising model. In that case, set $d=2, n=1$, $X=\{\mathit{up}, \mathit{down}\}$, the prior distribution on $X$ giving each of the 2 states $1/2$ chance, and 
\[
w(f) = \exp\left(\frac{(f(-1,0) = f(0,0) \mathbin{?} 1 : -1) + (f(0,-1) = f(0,0) \mathbin{?} 1 : -1)}{T}\right)
\] 
where $T$ is the temperature parameter. 
This means that fields where the state at adjacent points is equal are more likely. 
For temperatures above some critical temperature (I can't remember the value, but it doesn't really matter), the true distribution of results has patches of each state, with the patches getting larger as $T$ decreases.

There is therefore a correlation between the states at different points, that falls off exponentially with distance. 
(The behaviour below the critical temperature is more strange and discontinuous, and that case should possibly be excluded from consideration). 
The algorithm has to approximate this distribution, but it can't generate the entire infinite grid, so it has to generate finite portions at a time. When generating (separately) the values for two different points, it is given the coordinates of each point, but it's given the same source of randomness for both of them. 
The correct distribution for each point individually is just $1/2$ chance of each state, but it has to (for sufficiently low $\epsilon$) use the shared randomness to generate correctly correlated results.

\end{document}

### 2021-10-01 LICS21/intro.tex
" \ref" -> "~\ref"

--- 2021-09-27 LICS21/intro.tex
@@
 Because the type system already constrains the terms enough to force termination in the absence of the recursion construct \citep{tait1967,BarendregtDS2010}, it is only the $\tY$-reduction steps that must be counted, making defining ranking functions somewhat easier.
-Using supermartingales and Doob's (iterated) Optional Sampling Theorem, we show the soundness of rankability (\cref{thm:rankable implies termination}): if a term has a ranking function, it is AST.
+Using supermartingales and Doob's (iterated) Optional Sampling Theorem, we show the soundness of rankability (Theorem~\ref{thm:rankable implies termination}): if a term has a ranking function, it is AST.
 
@@
 Sparse ranking functions can be defined only for those points in the execution of a program which are semantically important, while all of the other intermediate steps can be ignored.
-Yet, they are no less efficacious for proving AST (\cref{thm:partial implies rankable}): every sparse ranking function is a restriction of a ranking function.
+Yet, they are no less efficacious for proving AST (Theorem~\ref{thm:partial implies rankable}): every sparse ranking function is a restriction of a ranking function.
 This also makes the ranking function method of proving AST more compatible with syntactic sugar, because the intermediate reduction steps implicit in the simplified notation can be ignored.
@@
 Thanks to this feature, the antitone ranking function method is capable of handling programs which terminate arbitrarily slowly. \lo{TODO: Reference some examples.}
-We show that this method is also sound for proving AST (\cref{thm:antitone rankable implies termination}), 
-moreover it also enjoys a sparse function theorem (\cref{thm:antitone partial implies rankable}).
+We show that this method is also sound for proving AST (Theorem~\ref{thm:antitone rankable implies termination}), 
+moreover it also enjoys a sparse function theorem (Theorem~\ref{thm:antitone partial implies rankable}).
 \akr{Our progress condition is equivalent to that given by \cite{DBLP:journals/pacmpl/McIverMKK18} (by applying a transformation to the ranking function), but needs to be different in order to be compatible with the presence of additional intermediate reduction steps where the ranking function is not required to decrease at all.}
-\lo{The same point is made in \cref{sec:related}, so the preceding can be omitted.}
+\lo{The same point is made in Section~\ref{sec:related}, so the preceding can be omitted.}
 
@@
 We introduce a novel addressing scheme for the possible random choices in a program's execution, which ensures that the same random choices are taken at corresponding positions in alternative reduction sequences, so that the same eventual result can be reached.
-This is then used to prove yet another extension to the ranking function theorem, that (\cref{thm:confluent ranking}) \lo{Preceding reference generated is wrong; it should be a theorem} ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and (\cref{cor:Reduction strategy independence}) rankability in this sense still imples almost sure termination.
+This is then used to prove yet another extension to the ranking function theorem, that (Theorem~\ref{thm:confluent ranking}) \lo{Preceding reference generated is wrong; it should be a theorem} ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and (Corollary~\ref{cor:Reduction strategy independence}) rankability in this sense still imples almost sure termination.
 The confluent trace semantics is of wider interest, and has other possible applications as well, for example in Bayesian inference algorithms.
 
-By combining these methods we can prove AST of a variety of PPCF programs that are beyond the reach of methods in the literature, including non-affine recursion\footnote{\label{fnote:non-affine} Non-affine recursive programs are recursive programs that can, during the evaluation of the recursive body, make multiple recursive calls (of a first-order function) from distinct call sites.} (\cref{ex:non-affine recursion easy,ex:non-affine continuous}) and recursions that define higher-order functions (\cref{ex:higher-order recursion}).
+By combining these methods we can prove AST of a variety of PPCF programs that are beyond the reach of methods in the literature, including non-affine recursion\footnote{\label{fnote:non-affine} Non-affine recursive programs are recursive programs that can, during the evaluation of the recursive body, make multiple recursive calls (of a first-order function) from distinct call sites.} (Examples~\ref{ex:non-affine recursion easy,ex:non-affine continuous}) and recursions that define higher-order functions (Example~\ref{ex:higher-order recursion}).
 
@@
 
-We present the syntax and trace semantics of PPCF in \Cref{sec:PPCF}.
-In \Cref{sec:supermartingales}, we show that ranking functions on terms induce supermartingales, which form the basis of a sound method for proving AST.
-We introduce sparse ranking functions in \Cref{sec:ranking} and antitone ranking functions in \Cref{sec:antitone}, and illustrate how they can be used to prove AST via examples. 
-In \Cref{sec:confluent}, we present a confluent trace semantics and demonstrate its usefulness.
-We discuss further applications of the confluent semantics in \Cref{sec:applications}; and conclude with comments on related work and further directions in \Cref{sec:related}.
+We present the syntax and trace semantics of PPCF in Section~\ref{sec:PPCF}.
+In Section~\ref{sec:supermartingales}, we show that ranking functions on terms induce supermartingales, which form the basis of a sound method for proving AST.
+We introduce sparse ranking functions in Section~\ref{sec:ranking} and antitone ranking functions in Section~\ref{sec:antitone}, and illustrate how they can be used to prove AST via examples. 
+In Section~\ref{sec:confluent}, we present a confluent trace semantics and demonstrate its usefulness.
+We discuss further applications of the confluent semantics in Section~\ref{sec:applications}; and conclude with comments on related work and further directions in Section~\ref{sec:related}.
 

### 2021-10-01 LICS21/ranking-proofs.tex
" \ref" -> "~\ref"

--- 2021-04-30 LICS21/ranking-proofs.tex
@@
 The difference $g - f$ is then a supermartingale (with the same setup as in
-\Cref{thm:rankable and strict rankable});
+Theorem~\ref{thm:rankable and strict rankable});
 therefore $\mathbb E[g(M_n)] \leq \mathbb E[f(M_n)] + g(M)-f(M)$, for all $n$.
@@
 Define $f_2(N) = \int_\entrosp f_1(\nnext(N,s)) + g(N,s) \, \mu_{\entrosp}(\mathrm d s)$.
-The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction, by the same argument as \Cref{thm:minimal}).
+The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction, by the same argument as Theorem~\ref{thm:minimal}).
 \end{proof}

### 2021-10-01 LICS21/ranking.tex
" \ref" -> "~\ref"

--- 2021-04-30 LICS21/ranking.tex
@@
 \subsection{Sparse ranking functions}
-Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost sure termination, so it isn't useful for proving it.
+Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and Theorem~\ref{thm:minimal} makes even stronger assumptions than almost sure termination, so it isn't useful for proving it.
 
@@
 The term  $(\tY \, \Theta) \, \underline 0$ generates a geometric distribution.
-Despite its simplicity, its $\mathit{Rch}$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.
+Despite its simplicity, its $\mathit{Rch}$ contains all the terms in Figure~\ref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.
 

### 2021-10-01 LICS21/related.tex
" \ref" -> "~\ref"

--- 2021-09-27 LICS21/related.tex
@@
 % (because the underlying recursion schemes are not Turing complete):
-for example, the PPCF terms in \cref{ex:raven complex,ex:non-affine continuous,ex:higher-order recursion}
+for example, the PPCF terms in Examples~\ref{ex:raven complex,ex:non-affine continuous,ex:higher-order recursion}
 %\refExample{complexExample} 
@@
 
-The main result of \cite{DBLP:journals/pacmpl/McIverMKK18} is similar to our \cref{thm:antitone rankable implies termination}, but in an imperative language.
+The main result of \cite{DBLP:journals/pacmpl/McIverMKK18} is similar to our Theorem~\ref{thm:antitone rankable implies termination}, but in an imperative language.
 While we require that the antitone ranking function decreases for every $\tY$-reduction step, they require that it decreases for every iteration of a certain while loop (the ranking function in that case is defined in the context of a particular loop), which is similar to $\tY$ but limited to tail recursion.
@@
 
-\paragraph*{Further directions} We have not yet been able to prove \Cref{conj:antitone} which would imply that \Cref{thm:antitone rankable implies termination} is almost complete.
+\paragraph*{Further directions} We have not yet been able to prove Conjecture~\ref{conj:antitone} which would imply that Theorem~\ref{thm:antitone rankable implies termination} is almost complete.
 \begin{conjecture}[Completeness]
@@
 
-The \Cref{def:more general red} of redexes and which positions are acceptable to reduce at is sufficiently restrictive to guarantee confluence, but is also a little more restrictive in some cases than is necessary for this purpose.
+Definition~\ref{def:more general red}, defining redexes and which positions are acceptable to reduce at is sufficiently restrictive to guarantee confluence, but is also a little more restrictive in some cases than is necessary for this purpose.
 For example, if the argument of a function is not yet a value, but its reduction to a value would be deterministic, or the function is affine, then applying the function before reducing its argument would not cause any problematic duplication of random samples.
@@
 
-Although this is a broadly applicable method of proving almost sure termination of probabilistic functional programs, and \Cref{thm:antitone partial implies rankable,thm:confluent ranking} make it more convenient to use, some method of automating the construction and checking of (antitone) ranking functions, even partially, could make it considerably more practically useful, especially in cases where almost sure termination is merely a side-condition for some other theorem or algorithm to be applicable.
+Although this is a broadly applicable method of proving almost sure termination of probabilistic functional programs, and Theorems~\ref{thm:antitone partial implies rankable,thm:confluent ranking} make it more convenient to use, some method of automating the construction and checking of (antitone) ranking functions, even partially, could make it considerably more practically useful, especially in cases where almost sure termination is merely a side-condition for some other theorem or algorithm to be applicable.
 

### 2021-10-01 LICS21/supermartingales-proofs.tex
Replace "theorem" environment with "therm", to get numbering back.
" \ref" -> "~\ref"

@@
 \rankableAndStrictRankable*
-We shall obtain the theorem as a corollary of a technical lemma (Lemma \ref{lem:key rankable}).
+We shall obtain the theorem as a corollary of a technical lemma (Lemma~\ref{lem:key rankable}).
 
@@
 
-As an immediate corollary of Lemma \ref{lem:key rankable}, each $f(M_n)$ is integrable.
-This concludes the proof of Theorem \ref{thm:rankable and strict rankable}.
+As an immediate corollary of Lemma~\ref{lem:key rankable}, each $f(M_n)$ is integrable.
+This concludes the proof of Theorem~\ref{thm:rankable and strict rankable}.
 
@@
 \end{enumerate}
-This concludes the proof of Lemma \ref{lem:TnBounded}.
+This concludes the proof of Lemma~\ref{lem:TnBounded}.
 

### 2021-10-01 LICS21/supermartingales.tex
Replace "theorem" environment with "therm", to get numbering back.
" \ref" -> "~\ref"

--- 2021-09-22 LICS21/supermartingales.tex
@@
 \label{def:ranking function}
-A \emph{ranking function on $M \in \Lambda^0$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$,\footnote{The fact that the ranking function can be 0 in some cases before it reaches a value is necessary to get \Cref{thm:minimal} to work neatly.} and
+A \emph{ranking function on $M \in \Lambda^0$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$,\footnote{The fact that the ranking function can be 0 in some cases before it reaches a value is necessary to get Theorem~\ref{thm:minimal} to work neatly.} and
 \begin{enumerate}
@@
 It seems cleaner to include this in the definition.}
-\akr{The fact that the ranking function can be 0 in some cases before it reaches a value is necessary to get \Cref{thm:minimal} to work neatly.}
-\lo{OK. Let's leave \Cref{def:ranking function} as it is.}
+\akr{The fact that the ranking function can be 0 in some cases before it reaches a value is necessary to get Theorem~\ref{thm:minimal} to work neatly.}
+\lo{OK. Let's leave Definition~\ref{def:ranking function} as it is.}
 \fi
@@
 \begin{remark}
-Lem.~\ref{lem:rank-PAST} is a slight extension of \cite[Lem.~5.5]{DBLP:conf/popl/FioritiH15},
+Lemma~\ref{lem:rank-PAST} is a slight extension of \cite[Lem.~5.5]{DBLP:conf/popl/FioritiH15},
 which asserts the same result but for the specific stopping time $T : \omega \mapsto \min\set{n \mid Y_n(\omega) = 0}$.
@@
 %\subsection{Proof of \Cref{thm:rankable and strict rankable}.}
-The ranking supermartingale condition is satisfied essentially by a case analysis on the type of redex (\Cref{lem:key rankable}).
+The ranking supermartingale condition is satisfied essentially by a case analysis on the type of redex (Lemma~\ref{lem:key rankable}).
 
@@
 Let $f$ be a ranking function on $M$.
-For (i), since $(T_n)_{n \geq 0}$ is an increasing sequence of stopping times, each is adapted to $(\calF_n)_{n \geq 0}$ and bounded (\Cref{lem:TnBounded}),
-and $(f(M_n))_{n \geq 0}$ is a supermartingale also adapted to $(\calF_n)_{n \geq 0}$ (\Cref{thm:rankable and strict rankable}),
-it follows from the Optional Sampling \Cref{thm:optional sampling} that $(Y_n)_{n \geq 0}$ %where $Y_n := f(M_{T_n})$ 
+For (i), since $(T_n)_{n \geq 0}$ is an increasing sequence of stopping times, each is adapted to $(\calF_n)_{n \geq 0}$ and bounded (Lemma~\ref{lem:TnBounded}),
+and $(f(M_n))_{n \geq 0}$ is a supermartingale also adapted to $(\calF_n)_{n \geq 0}$ (Theorem~\ref{thm:rankable and strict rankable}),
+it follows from the Optional Sampling Theorem~\ref{thm:optional sampling} that $(Y_n)_{n \geq 0}$ %where $Y_n := f(M_{T_n})$ 
 is a supermartingale adapted to $(\calF_{T_n})_{n \geq 0}$.
-Notice the stopping time $T_M^\tY$ is also adapted to $(\calF_{T_n})_{n \geq 0}$ (\Cref{lem:TMtY is a stopping time}); and we have that $(Y_n)_{n \geq 0}$ is a $1$-ranking supermartingale.
-Therefore, by \Cref{lem:rank-PAST}, $T^{\tY}_M < \infty$ a.s.~and $\expect{T^{\tY}_M} < \infty$.
+Notice the stopping time $T_M^\tY$ is also adapted to $(\calF_{T_n})_{n \geq 0}$ (Lemma~\ref{lem:TMtY is a stopping time}); and we have that $(Y_n)_{n \geq 0}$ is a $1$-ranking supermartingale.
+Therefore, by Lemma~\ref{lem:rank-PAST}, $T^{\tY}_M < \infty$ a.s.~and $\expect{T^{\tY}_M} < \infty$.
 %For 2, we have that $(f(M_n))_{n \geq 0}$ is a ranking supermartingale (\Cref{thm:rankable and strict rankable}) w.r.t.~stopping time $T_M$. 
 %It follows immediately from \Cref{lem:rank-PAST} that $\expect{T_M} < \infty$.
-Statement (ii) follows at once from \Cref{thm:rankable and strict rankable} and \Cref{lem:rank-PAST}.
+Statement (ii) follows at once from Theorem~\ref{thm:rankable and strict rankable} and Lemma~\ref{lem:rank-PAST}.
 \end{proof}
@@
 Thus the method of (strict) ranking function is sound for proving (positive) a.s.~termination of PPCF programs.
-It is in fact also complete in a sense: if $\expect{T^\tY_N} < \infty$ for all $N \in \mathit{Rch}(M)$ then $M$ is rankable (\Cref{thm:minimal}).
+It is in fact also complete in a sense: if $\expect{T^\tY_N} < \infty$ for all $N \in \mathit{Rch}(M)$ then $M$ is rankable (Theorem~\ref{thm:minimal}).

### 2021-10-01 probabilisticProgrammingMartingales.tex
" \ref" -> "~\ref"

@@
 The difference $g - f$ is then a supermartingale (with the same setup as in the proof of %\Cref{thm:rankable gives supermartingale}
-%Theorem \ref{rankable implies ast}) 
+%Theorem~\ref{rankable implies ast}) 
 \Cref{thm:rankable and strict rankable});
@@
 \begin{proof}
-As in %Theorem \ref{thm:basicRankingSoundness}, 
+As in %Theorem~\ref{thm:basicRankingSoundness}, 
 \Cref{thm:rankable implies termination},
@@
 It seems very likely that this method is complete in the case that every reachable term is AST, but we have been unable to actually prove this.
-It is certainly at least capable of proving the termination of terms which terminate arbitrarily slowly (in the sense that there is no similar limitation to Theorem \ref{thm:rankable implies termination}, which can only prove termination of $\tY$-PAST terms).
+It is certainly at least capable of proving the termination of terms which terminate arbitrarily slowly (in the sense that there is no similar limitation to Theorem~\ref{thm:rankable implies termination}, which can only prove termination of $\tY$-PAST terms).
 
@@
 Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $Sk(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$.
-\lo{Do we actually use the notation $S[r]$ in what follows?} \akr{Yes, in Lemma \ref{lem:independentSamples} for example.}
+\lo{Do we actually use the notation $S[r]$ in what follows?} \akr{Yes, in Lemma~\ref{lem:independentSamples} for example.}
 \lo{Skeleton is defined in Sec.~\ref{sec:sampling semantics}.}
@@
 Let $X$ be the call-by-value reduction sequence related to both $N'_1$ and $N'_2$ by $\sim_{c,M}$ given by Lemma~\ref{canonicalCousins}. 
-As $M \to N$ is the first reduction in the sequence $N'_1$, it is the last to be affected by the $\sim_{c,M}$ sequence $N'_1 \sim_{c,M}^* X$ given by Lemma \ref{canonicalCousins} therefore it can be split into $N'_1 \sim_{c,M}^* Y_1 \sim_{c,M}^* X$ where $Y_1$ is in CBV order except possibly for its first reduction, which is still $M \to N$.
+As $M \to N$ is the first reduction in the sequence $N'_1$, it is the last to be affected by the $\sim_{c,M}$ sequence $N'_1 \sim_{c,M}^* X$ given by Lemma~\ref{canonicalCousins} therefore it can be split into $N'_1 \sim_{c,M}^* Y_1 \sim_{c,M}^* X$ where $Y_1$ is in CBV order except possibly for its first reduction, which is still $M \to N$.
 Let the position of the reduction $M \to N$ be $\beta$.
@@
 The distribution of $s(k)$ conditional on reaching $N$ is therefore the uniform distribution on $I$.
-Similarly for the other semantics, if the distribution of remaining samples after $n$ steps is independent of the term by Lemma \ref{lem:independentSamples}, therefore the distribution of $s(Sk(N),\cbv(N))$ conditional on $N$ is the uniform distribution on $I$.
+Similarly for the other semantics, if the distribution of remaining samples after $n$ steps is independent of the term by Lemma~\ref{lem:independentSamples}, therefore the distribution of $s(Sk(N),\cbv(N))$ conditional on $N$ is the uniform distribution on $I$.
 In either case, the distribution of the next term conditional on the previous term, in the case that it reduces randomly, is equal to the image under $r \mapsto E[\underline r]$ of the uniform distribution on $I$.
@@
 Let $(M,s) = (M_0,s_0) \Rightarrow_f \dots \Rightarrow_f (M_n,s_n)$, where $M_n$ is a value.
-By Lemma \ref{canonicalCousins}, the reduction sequence $(Sk(M_i))_i$ is related by $\sim_c^*$ to some skeletal reduction sequence $X$ which is in call-by-value order.
+By Lemma~\ref{canonicalCousins}, the reduction sequence $(Sk(M_i))_i$ is related by $\sim_c^*$ to some skeletal reduction sequence $X$ which is in call-by-value order.
 For each skeletal reduction sequence $X_j$ in the sequence $(Sk(M_i))_i \sim_c \dots \sim_c X$, it is possible to define a reduction sequence of terms such that $(M,s) = (N_{j,0},s_{j,0}) \Rightarrow \dots \Rightarrow (N_{j,n_j},s_{j,n_j})$ and $Sk(N_{j,k}) = X_{j,k}$, by at each step applying the reduction at the same place as in $X_j$, essentially just filling in the reals in the skeletal reduction sequence to make it a reduction sequence of terms (it will be shown shortly that the correct branch is taken in each $\textsf{if}$-reduction).
@@
 To prove that this holds true for $j+1$ as well, it suffices to show that the corresponding term and samples immediately after the reductions involved in this $\sim_c$ step match, as from that point on, the $\Rightarrow$-sequences will match, as they start from the same point and have all the same reduction positions.
-Letting the term and remaining samples immediately before the reductions involved in this $\sim_c$ step be $(N,t)$ (which is the same in both $N_j$ and $N_{j+1}$ for the same reason again, they start at the same point and the sequences of reduction positions up to this point are the same), and the terms and samples after these reductions be $(O_1,t_1)$ and $(O_2,t_2)$, so that these match the $N$, $O_1$ and $O_2$ in the definition of $\sim_c$, $t_1 = t_2$ because both of them are, by Lemma \ref{lem:sim-M-N}, equal to the restriction of $t$ to the subset of $Rch(N)$ that starts with the reduction sequences to $O_1$ and $O_2$ respectively, and the potential positions in those subsets of $Rch(N)$ are identified with each other by $\sim$ because of the same case of $\sim_c$ that means $N_j$ and $N_{j+1}$ are related.
+Letting the term and remaining samples immediately before the reductions involved in this $\sim_c$ step be $(N,t)$ (which is the same in both $N_j$ and $N_{j+1}$ for the same reason again, they start at the same point and the sequences of reduction positions up to this point are the same), and the terms and samples after these reductions be $(O_1,t_1)$ and $(O_2,t_2)$, so that these match the $N$, $O_1$ and $O_2$ in the definition of $\sim_c$, $t_1 = t_2$ because both of them are, by Lemma~\ref{lem:sim-M-N}, equal to the restriction of $t$ to the subset of $Rch(N)$ that starts with the reduction sequences to $O_1$ and $O_2$ respectively, and the potential positions in those subsets of $Rch(N)$ are identified with each other by $\sim$ because of the same case of $\sim_c$ that means $N_j$ and $N_{j+1}$ are related.
 If none of the relevant reductions are $\tsample$-reductions, $O_1 = O_2$ trivially.
 If there are samples taken (which must be the arbitrary reduction at $\beta$), it is also required that the same samples are taken in both branches.
-Similarly to Lemma \ref{churchRosser}, the positions of these $\tsample$-reductions in each branch are related by $\sim_p^*$, therefore they correspond to the same sample in $t$, therefore $O_1 = O_2$.
+Similarly to Lemma~\ref{churchRosser}, the positions of these $\tsample$-reductions in each branch are related by $\sim_p^*$, therefore they correspond to the same sample in $t$, therefore $O_1 = O_2$.
 The fact that the terms are equal also implies that the same branch of an $\textsf{if}$ statements will be taken.
@@
 All of the theorems on the termination of rankable terms therefore extend to other reduction strategies too.
-The proofs of Theorems \ref{thm:rankable implies termination}, \ref{thm:partial implies rankable}, \ref{thm:antitone rankable implies termination} and \ref{thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies.
-They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because Theorem \ref{thm:de groote} applies equally to any reduction strategy).
+The proofs of Theorems~\ref{thm:rankable implies termination},~\ref{thm:partial implies rankable},~\ref{thm:antitone rankable implies termination} and~\ref{thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies.
+They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because Theorem~\ref{thm:de groote} applies equally to any reduction strategy).
 For a reduction strategy $r$ and term $M$, just substitute $N[r(N)]$ being a $\tY$-redex for $N$ being of the form $E[\tY \lambda x.

### 2021-10-01 ranking-supermartingales-8May20.tex
" \ref" -> "~\ref"

--- 2021-01-02 ranking-supermartingales.tex
@@
 %!TEX root = probabilisticProgrammingMartingales.tex
-\paragraph{}
+\subsection*{Reorganised proof}
 
@@
 \begin{definition}\rm
-\begin{enumerate}
-\item A sequence of random variables $(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \emph{supermartingale} if for all $n \in \omega$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~\changed[lo]{(i.e.~$\forall A \in \calF_n$, $\int_A \dif \mu \, Y_{n+1} \leq \int_A \dif \mu \,Y_n$).}
-\item Let $\epsilon > 0$.
-A stopping time $T$ and a supermartingale $(Y_n)_{n \geq 0}$, both adapted to filtration $(\calF_n)_{n \geq 0}$, are a \emph{$\epsilon$-ranking supermartingale} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.\footnote{For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable $s \mapsto \begin{cases}
-1 & \hbox{if $s \in A$}\\
-0 & \hbox{otherwise}
-\end{cases}$.}
-A \emph{ranking supermartingale} is a $\epsilon$-ranking supermartingale for some $\epsilon > 0$.
-\end{enumerate}
+A sequence of random variables $(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \emph{supermartingale} if for all $n \in \omega$, $\expect{|Y_n|} < \infty$, and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$.
+\lo{The preceding means: $\forall A \in \calF_n$, $\int_A \dif \mu \, Y_{n+1} \leq \int_A \dif \mu \,Y_n$.}
+For $\epsilon > 0$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \, \mathbf{1}_{\set{Y_n > 0}}$.
+\citep{DBLP:conf/popl/FioritiH15,DBLP:conf/popl/ChatterjeeFNH16}
 \end{definition}
 
-\begin{remark}
-\changed[lo]{The original definition of ranking supermartingale, due to \citep{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, does not refer to a stopping time explicitly:
-Fioriti and Hermanns call a supermartingale $(Y_n)_{n \geq 0}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot {\bf 1}_{\set{Y_n > 0}}$.}
-\end{remark}
+Intuitively $Y_n$ is the rank of the program after $n$ steps of computation.
+(Assume that the rank of a term is 0 iff it is a value.) In a $\epsilon$-ranking supermartingale, each computation step causes a strict decrease in rank, provided the term being reduced is not a value.
 
-Intuitively $Y_n$ gives the rank of the program after $n$ steps of computation, and $T$ is the time at which it reaches a value (which may be infinite if it fails to terminate).
-In a $\epsilon$-ranking supermartingale, each computation step causes a strict decrease in rank of at least $\epsilon$, provided the term in question is not a value.
-
-Let $T$ and $T'$ be stopping times adapted to $(\calF_n)_{n \geq 0}$.
-Recall the $\sigma$-algebra (consisting of measurable subsets ``prior to $T$'')
+Let $S$ and $T$ be stopping times adapted to $(\calF_n)_{n \geq 0}$.
+Recall the $\sigma$-algebra (consisting of measurable subsets ``priori to $T$'')
 \[
@@
 \]
-and if $T \leq T'$, then $\calF_{T} \subseteq \calF_{T'}$.
+and if $S \leq T$, then $\calF_S \subseteq \calF_{T}$.
 
-The following is an iterated version of Doob's well-known Optional Sampling Theorem (see, e.g., \cite[\S 6.7]{AshDD00} and \cite[Theorem 7.2]{DBLP:conf/popl/FioritiH15}).
+The following is an iterated version of Doob's well-known Optional Sampling Theorem (see, e.g., \cite[\S 6.7]{AshDD00}).
 \begin{theorem}[Optional Sampling]
 \label{thm:optional sampling}
-Let $(X_n)_{n \geq 0}$ be a supermartingale, and $(T_n)_{n \geq 0}$ a sequence of increasing stopping times, both adapted to filtration $(\calF_n)_{n \geq 0}$, then $(X_{T_n})_{n \geq 0}$ is a supermartingale adapted to $(\calF_{T_n})_{n \geq 0}$ if one of the following conditions holds:
+Let $(X_n)_{n \geq 0}$ be a supermartingale, and $(T_n)_{n \geq 0}$ a sequence of increasing stopping times, then $(X_{T_n})_{n \geq 0}$ is a supermartingale adapted to $(\calF_{T_n})_{n \geq 0}$ if one of the following conditions holds:
 \begin{enumerate}
@@
 
-\begin{lemma}%[{\citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}}]
-% \citep[Prop 1]{DBLP:conf/popl/ChatterjeeFNH16}}
-\label{lem:rank-PAST}
-Let $(Y_n)_{n \geq 0}$ be a $\epsilon$-ranking supermartingale with respect to the stopping time $T$, then $T < \infty$ almost surely, and $\expect{T} \leq \dfrac{\expect{Y_0}}{\epsilon}$.
-\end{lemma}
+\begin{theorem}[{\citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}, \citep[Prop 1]{DBLP:conf/popl/ChatterjeeFNH16}}]
+\label{thm:rank-PAST}
+Let $(Y_n)_{n \geq 0}$ be a $\epsilon$-ranking supermartingale, and set $T(s) := \min \set{n \mid Y_n(s) = 0}$.
 
-\begin{proof}
-We first prove by induction on $n$: 
-\[
-\forall n \in \omega \, . \, \expect{Y_n} \leq \expect{Y_0} - \epsilon \cdot \Big(\sum_{i=0}^{n-1} \mathbb{P}[T > i]\Big).
-\]
-It follows that $\sum_{i=0}^{\infty} \mathbb{P}[T > i]$ converges; hence $\lim_{i \to \infty} \mathbb{P}[T > i] = 0$ and so $\mathbb{P}[T < \infty] = 1$.
-It then remains to observe: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=0} \mathbb{P}[T > i]$.
-\end{proof}
-
-\changed[lo]{\Cref{lem:rank-PAST} is a slight extension of \citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}. 
-In Lemma 5.5 of op.~cit., the stopping time $T$ is assumed to be the random variable $s \mapsto \min\set{n \mid Y_n(s) = 0}$.}
-
-\medskip
-
-Our first result is the following theorem.
-\begin{theorem}[Rankability] 
-\label{thm:rankable and strict rankable}
-If a closed SPCF term is rankable (respectively, strictly rankable) by $f$, then $(f(M_n))_{n \geq 0}$ is a 
-supermartingale (respectively, ranking supermartingale) adapted to the filtration $(\mathcal{F}_n)_{n \geq 0}$ where $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
+Then $T < \infty$ almost surely, and $\expect{T} \leq \dfrac{\expect{Y_0}}{\epsilon}$.
 \end{theorem}
 
-%\subsection{Proof of \Cref{thm:rankable and strict rankable}.}
-\emph{Notation}. Given an $\omega$-sequence (e.g.~$s \in S$) and $m \in \omega$, we write $s_{\leq m} \in I^m$ to mean the prefix of $s$ of length $m$.
+\paragraph{Proof of \Cref{thm:rankable gives supermartingale}.}
+\emph{Notation}. Given an $\omega$-sequence (e.g.~$s \in S$) and $m \in \omega$, we write $s_{\leq m} \in I^m$ to mean the $m$-long prefix of $s$.
 
@@
 M_n(s) &:= \pi_0 (\red^n(M, s))\\
-\ndraw{n}{s} &:= |\{k < n \mid \exists E: M_k(s) = E[\tsample]\}|,
+\ndraw{n}{s} &:= l \textrm{ where $\pi_1 (\red^n(M, s)) = \underbrace{\pi_t( \cdots (\pi_t}_l(s))$}
 \end{align*}
- so that $\pi_1 (\red^n(M, s)) = \overbrace{\pi_t( \cdots (\pi_t}^{\ndraw n s}(s))$, and the filtration $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
-The $\calF_n$-measurability of $M_n$ (and hence of $\#\mathrm{draw}_n$) follows from \citep{DBLP:conf/icfp/BorgstromLGS16}.
-%$\#\mathrm{draw}_n$ is a stopping time (bounded by $n$). 
-\iffalse
-\akr{$\#\mathrm{draw}_n$ is not a stopping time.
-It doesn't look like you actually use this claim anyway, so it should just be fine to remove, but did you mean something different?} \lo{I agree, and I don't actually use this claim.}
-\fi
+and the filtration $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
+The $\calF_n$-measurability of $M_n$ (and hence $\#\mathrm{draw}_n$) follow from \citep{DBLP:conf/icfp/BorgstromLGS16};
+$\#\mathrm{draw}_n$ is a stopping time (bounded by $n$).
 Take $s \in A \in \calF_n$ with $\ndraw{n}{s} = l$.
@@
 
-\iffalse
-\lo{We need to show that each $f(M_n)$ is integrable. LO claimed earlier that $0 \leq f(N) \leq f(M)$ for all $N \in \mathit{Rch}(M)$.
-This is of false: take $f$ with $v \, \tsample \mapsto 1$ and $v \, \underline{r} \mapsto 2\, r$ for some value $v$. 
-However, it is true that
-\[
-\int_{S} \mu(\dif s) \, |f(M_n)(s)| 
-=
-\int_{S} \mu(\dif s) \, f(M_n)(s)
-\leq
-\int_{S} \mu(\dif s) \, f(M_0)(s)
-=
-f(M)
-\]
-the inequality above follows from \Cref{lem:key rankable}.
-}
-\fi
-
-We say that a given SPCF term is: %\emph{type-1} (respectively 2, 3 and 4) if it has the shape $E[\tY \lambda x. N]$ (respectively $E[\tsample]$, $E[R]$ where $R$ is any other redex, and of a value).
-\emph{type-1} if it has the shape $E[\tY \lambda x.
-N]$, \emph{type-2} if it has the shape $E[\tsample]$; \emph{type-3} if it has the shape $E[R]$ where $R$ is any other redex, and \emph{type-4} if it is a value.
-Henceforth fix an $n \geq 0$, and define $\mathbf{T}_i := \set{s \mid M_n(s) \hbox{ is type-$i$}}$.
+We say that a given SPCF term is \emph{type-1} (respectively 2, 3 and 4) if it has the shape $E[\tY (\lambda x. N)]$ (respectively $E[\tsample]$, $E[R]$ where $R$ is any other redex, and of a value).
+Define $\mathbf{T}_i := 
+\set{s \mid M_n(s) \hbox{ is type-$i$}}$.
 It is straightforward to see that each $\mathbf{T}_i \in \calF_n$, and $\set{\mathbf{T}_1, \mathbf{T}_2, \mathbf{T}_3, \mathbf{T}_4}$ is a partition of $S$.
 %By abuse of language, we refer to the three respective types of redexes as type-$1$, 2 and 3.
-Hence it suffices to prove the following lemma.
 
-\iffalse
+%Henceforth fix $M \in \Lambda^0$.
 For $i \in \{1, 2, 3\}$ and $n \in \omega$, define function $f_{i, n+1}(M) : S \to \Real$ by
@@
 and (\ref{eqn:partition}) holds because $\{\mathbf{T}_1, \mathbf{T}_2, \mathbf{T}_3, \mathbf{T}_4\}$ is a partition of $S$.
-\fi
 
 \begin{lemma}
-\label{lem:key rankable}
-For all $i \in \set{1, 2, 3, 4}$ and $A \in \calF_n$
+For all $i \in \set{1, 2, 3}$ and $A \in \calF_n$
 \[
-\int_A \mu(\dif s) \, f(M_{n+1})[s \in \mathbf{T}_i] \leq \int_A \mu(\dif s) \, f(M_n)[s \in \mathbf{T}_i]
-\; \footnote{where (the Iverson bracket) $[P] := 1$ if the statement $P$ hold, and 0 otherwise.}
+\int_A \dif \mu \, f_{i, n+1}(M) \leq \int_A \mu(\dif s) \, f(M_n)[s \in \mathbf{T}_i].
 \] 
@@
 \begin{equation}
-f(M_{n+1})[s \in \mathbf{T}_2] = \sum_{i \in \calI} 
-f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])[s \in U_i]
+f_{2, n+1}(M) = \sum_{i \in \calI} 
+\bblambda s . f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])[s \in U_i]
 \label{eqn:f 2 n+1}
@@
 \item $\calI$ is a countable indexing set
-\item $E_i[\cdot][\tsample] \in \mathsf{Sk}_{j_i}$, and $\sigma_i : S \to \Real^{j_i}$, and $\rho(s) := \pi_h(\pi_1(\red^n(M, s))) = \pi_h(\pi_t^{l_i}(s)) \in \Real$ 
-\item $\set{U_i}_{i \in \calI}$ is a partition of $\mathbf{T}_2$, 
-where each $U_i$ is determined by a skeletal term $E_i[\cdot][\tsample]$ and a number (of draws) $l_i \leq n$, satisfying
-\[
-s \in U_i
-\iff
-(E_i[\underline{r}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s),
-\;
-\hbox{for some }r \in \Real^{j_i}
-\]
-(We use the (measurable) function $\sigma_i : S \to \Real^{j_i}$ to skolemise the existentially quantified $r$.)
-
-Equivalently
-\(
-U_i := \#\mathrm{draw}_n^{-1}(l_i) \cap M_n^{-1}(E_i[\cdot][\tsample], \Real^{j_i}) \in \calF_n.
-\)
-%\lo{I.e.~$(E_i[\underline{\sigma_i(s)}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s)$ iff $s \in U_i$.}
+\item $E_i \in \bigcup_{j \in \omega}\mathsf{Sk}_j$, and $\sigma_i : S \to \bigcup_{j \in \omega}\Real^j$ is a random variable
+\item $\rho(s) = \pi_h(\pi_1(\red^n(M, s))) \in \Real$ 
+\item $\set{U_i}_{i \in \calI}$ is a partition of $\mathbf{T}_2$ 
+\item Each $U_i$ is determined by a skeletal term, namely, $E_i$, and the number of draws (i.e.~$\ndraw{n}{s}$) consumed in $n$ steps of computation of $M$, in the sense that for all $s, s' \in U_i$, 
+%$\pi_0(\red^n(M, s))$ and $\pi_0(\red^n(M, s))$ 
+$M_n(s)$ and $M_n(s')$ have the same underlying skeleton $E_i$ and $\ndraw{n}{s} = \ndraw{n}{s'}$.
+Notice that $E_i[\underline{\sigma_i(s)}][{\tsample}] = M_{n}(s)$ for $s \in U_i$.
 \end{itemize}
 
-Observe that if $s \in U_i$ then $\set{s_{\leq l_i}} \cdot I^\omega \subseteq U_i$;
-in fact $(U_i)_{\leq l_i} \cdot I^\omega = U_i$.
-%Moreover, for any $A \in \calF_n$, if $s \in A$ then $\set{s_{\leq l}} \cdot I^\omega \subseteq A$.
-This means that for any measurable $g : S \to \Real_{\geq 0}$, if $g(s)$ only depends on the prefix of $s$ of length $(l_i+1)$, then, writing $\widehat{g} : I^{l_i+1} \to \Real_{\geq 0}$ where $g(s) = \widehat{g}(s_{\leq l_i+1})$, we have: for any $A \in \calF_n$
+Observe that if $s \in U_i$ and $\ndraw{n}{s} = l$ then $s_{\leq l} = s_{\leq l}'$ implies $M_n(s) = M_n(s')$.
+It follows that $\set{s_{\leq l}} \cdot I^\omega \subseteq U_i$ (where $\cdot$ means concatenation of sequences).
+Moreover, for any $A \in \calF_n$, if $s \in A$ then $\set{s_{\leq l}} \cdot I^\omega \subseteq A$.
+This means that for any measurable $g$, if $g(s)$ only depends on the $(l+1)$-long prefix of $s$ (i.e.~$g(s) = g'(s_{\leq l+1})$), then 
 \begin{equation}
 \int_{A \cap U_i}  \mu(\dif s) \, g(s) = 
-\int_{(A \cap U_i)_{\leq l_i+1}} \Leb_{l_i+1}(\dif t) \, \widehat{g}(t)
+\int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif t) \, g'(t)
 \label{eqn:truncate}
 \end{equation}
-%(For simplicity, we will write $g' = g$ in the following.)
-%where $A_{\leq m} := \set{s_{\leq m} \mid s \in A}$.
+where $A_{\leq m} := \set{s_{\leq m} \mid s \in A}$, and $\Leb_n$ is the Lebesgue measure on $I^n$.
 
-Take $s \in U_i$, and set $l = l_i$.
-Plainly $\sigma_i(s)$ depends on $s_{\leq l}$, and $\rho(s)$ depends on $s_{\leq l +1}$.
-Take $u \in (U_i)_{\leq l}$.
-It then follows from the definition of ranking function that
+Take $s \in U_i$ with $\ndraw{n}{s} = l$.
+Now $\sigma_i(s)$ depends on $s_{\leq l}$, and $\rho(s)$ depends on $s_{\leq l+1}$.
+Let $u$ range over $(U_i)_{\leq l}$.
+Then, it follows from the definition of $f$ that
 \[
-\int_I \Leb(\dif r) \, f(E_i[\underline{\widehat{\sigma_i}(u)}][\underline{r}])) \leq f(E_i[\underline{\widehat{\sigma_i}(u)}][\tsample])).
+\int_I \Leb(\dif r) \, f(E_i[\underline{\sigma_i(u)}][\underline{r}])) \leq f(E_i[\underline{\sigma_i(u)}][\tsample])).
 \]
-Take $A \in \calF_n$, and integrating both sides, we get
+Integrating both sides, we get
 \begin{align*}
-& \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, \int_I \textrm{Leb}(\dif r) \, f(E_i[\underline{\widehat{\sigma_i}(u)}][\underline{r}])) \\
-& \leq \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, f(E_i[\underline{\widehat{\sigma_i}(u)}][\tsample])).
+& \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, \int_I \textrm{Leb}(\dif r) \, f(E_i[\underline{\sigma_i(u)}][\underline{r}])) \\
+\leq & \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, f(E_i[\underline{\sigma_i(u)}][\tsample]))
 \end{align*}
-Since $\Leb_{l+1}$ is the (unique) product measure satisfying $\Leb_{l+1}(V \times B) = \Leb_l(V) \cdot \Leb(B)$, and $(U_i)_{\leq l_i} \cdot I^\omega = U_i$, we have
-%it follows from the preceding observation that
+It follows from the preceding observation that
 \begin{align*}
-& \int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\underline{\widehat{\sigma_i}(u')}][\underline{\widehat{\rho}(u')}])) \\
+& \int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\underline{\sigma_i(u')}][\underline{\rho(u')}])) \\
 & \leq 
-\int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\underline{\widehat{\sigma_i}(u')}][\tsample]))
+\int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\underline{\sigma_i(u')}][\tsample]))
 \end{align*}
-and so, by (\ref{eqn:truncate})
+and so we have
 \begin{equation}
@@
 
-Now, integrating both sides of (\ref{eqn:f 2 n+1}), we have
+Finally, integrating both sides of (\ref{eqn:f 2 n+1}), we have
 \begin{align*}
-\int_A \mu(\dif s) \, f_{n+1}(M)[s \in \mathbf{T}_i] 
+\int_A \mu(\dif s) \, f_{2, n+1}(M)(s) 
 &= 
@@
 
-As an immediate corollary of \Cref{lem:key rankable}, each $f(M_n)$ is integrable:
-\[
-\int_{S} \mu(\dif s) \, |f(M_n)(s)| 
-=
-\int_{S} \mu(\dif s) \, f(M_n)(s)
-\leq
-\int_{S} \mu(\dif s) \, f(M_0)(s)
-=
-f(M).
-\]
-
 \iffalse
@@
 \fi
-This concludes the proof of %\Cref{thm:rankable gives supermartingale}. 
-\Cref{thm:rankable and strict rankable}. \hfill \qed
+This concludes the proof of \Cref{thm:rankable gives supermartingale}. \hfill \qed
 
 \medskip
-As before, fix a closed SPCF term $M \in \Lambda^0$.
-Define random variables on the probability space $(S, \calF, \mu)$:
-\begin{align}
-T_{-1}(s) & := -1 \nonumber \\
-T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value or of form } E[\tY \lambda x. N] \} \nonumber \\
-X_n(s) & := f(M_n(s)) \nonumber \\
-Y_n(s) & := X_{T_n}(s) \nonumber \\
-T_M^\tY(s) &:= \min \set{n \mid M_{T_n(s)}(s) \textrm{ is a value}} \label{eq:y stopping time}
-\end{align}
-The random variable $T_M^\tY$, which we call the \emph{$\tY$-runtime of $M$}, can equivalently be defined as the number of $\tY$-reduction steps in the reduction sequence of $M$. 
-Note that, as the type system ensures that the reduction relation excluding $\tY$-reduction is strongly normalising, only finitely many reductions can occur in a row without one of them being a $\tY$-reduction, therefore $T_M^\tY < \infty$ a.s.~iff $M$ is AST.
+Define random variables on $(S, \calF, \mu)$.
+\begin{align*}
+T_0(s) & := 0 \\
+T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value or of form } E[\tY (\lambda x. N)] \}\\
+X_n(s) & := f(M_n(s)) \\
+Y_n(s) & := X_{T_n}(s)
+\end{align*}
 
-The r.v.~$T_M^\tY$ is a stopping time adapted to the filtration $(\calF_{T_n})_{n \geq 0}$.
+Observe that $T_0, T_1, T_2, \cdots$ are an increasing sequence of stopping times adapted to $(\calF_n)_{n \geq 0}$, and each $T_i$ is bounded.
+As $(X_n)_{n \geq 0}$ is a supermartingale (\Cref{thm:rankable gives supermartingale}), it follows from \Cref{thm:optional sampling} that $(Y_n)_{n \geq 0}$ is a supermartingale (in fact, a $1$-ranking one).
+Therefore, by \Cref{thm:rank-PAST}, $M$ is PAST.
 
-\begin{theorem}[Soundness] \label{thm:rankable implies termination}
-\begin{enumerate}
-\item If a closed SPCF term $M$ is rankable, then $M$ is AST (equivalently, $T^{\tY}_M < \infty$ almost surely), and $M$ is $\tY$-positively almost sure terminating ($\tY$-PAST) i.e.~$\expect{T^{\tY}_M} < \infty$.
+Thus, to summarise
 
-\item If a closed SPCF term $M$ is strictly rankable, then $\expect{T_M} < \infty$ i.e.~$M$ is positively almost sure terminating (PAST).
-\end{enumerate}
-\end{theorem}
-
-\begin{proof}
-Let $f$ be a ranking function on $M$.
-For 1, observe that $T_0, T_1, T_2, \cdots$ are an increasing sequence of stopping times, each is bounded \lo{Prove it.} and adapted to $(\calF_n)_{n \geq 0}$.
-Then, since $(f(M_n))_{n \geq 0}$ is a supermartingale also adapted to $(\calF_n)_{n \geq 0}$ (\Cref{thm:rankable and strict rankable}),
-it follows from the Optional Sampling \Cref{thm:optional sampling} that $(Y_n)_{n \geq 0}$ where $Y_n := f(M_{T_n})$ is a supermartingale adapted to the filtration $(\calF_{T_n})_{n \geq 0}$.
-Since $T_M^\tY$ is a stopping time also adapted to $(\calF_{T_n})_{n \geq 0}$, we have that $(Y_n)_{n \geq 0}$ is a $1$-ranking supermartingale.
-Therefore, by \Cref{lem:rank-PAST}, $T^{\tY}_M < \infty$ a.s.~and $\expect{T^{\tY}_M} < \infty$.
-
-Statement 2 follows immediately from \Cref{lem:rank-PAST}.
-\end{proof}
-
-Thus the method of (strict) ranking function is sound for proving (positive) a.s.~termination of SPCF programs.
-It is in fact also complete in a sense: if $\expect{T^\tY_N} < \infty$ for all $N \in \mathit{Rch}(M)$ then $M$ is rankable (\Cref{thm:minimal}).
+\begin{corollary}
+If a closed SPCF term is rankable, then it is positively almost surely terminating.
+\end{corollary}

### 2021-10-01 thesis proposal/thesis proposal copy.tex
" \ref" -> "~\ref"

--- 2021-03-24 thesis proposal/thesis proposal.tex
@@
 \usepackage{amsfonts}
-\usepackage{stmaryrd}
 \newcommand{\tY}{\mathsf{Y}}
@@
 \newcommand{\tscore}{\mathsf{score}}
-\newcommand{\denotation}[1]{\llbracket #1 \rrbracket}
 
@@
 
-As a running example, I will be using a simply-typed call by value language with a random sampling construct in the form of a probabilistic choice $M \oplus N$, a conditioning construct $\tscore(r)$ for non-negative real constants $r$, and a countable set $T$ of primitive values of some base type $\mathsf T$.
+As a running example, I will be using a simply-typed call by value language with a random sampling construct in the form of a probabilistic choice $M \oplus N$, a conditioning construct $\tscore(r)$ for non-negative real constants $r$, and countably many primitive values of some base type $\mathsf T$.
 Although this language is rather deficient (there is no way to use continuous distributions, for example, and no recursion), it will be sufficient to illustrate the different styles of semantics.
 
-\subsection{Operational Semantics} \label{operational semantics}
+\subsection{Operational Semantics}
 In operational semantics, the meaning of the program is defined by repeatedly transforming a term into other terms, until it reaches a value, a term which is considered finished, which is the output.
@@
 (E[(\lambda x. M) V], t, s) & \to (M[V/x], t, s) \\
-(E[M \oplus N], 0:t, s) & \to (E[M], t, s) \\
-(E[M \oplus N], 1:t, s) & \to (E[N], t, s) \\
+(E[M \oplus N], t, s) & \to (E[M], 0:t, s) \\
+(E[M \oplus N], t, s) & \to (E[N], 1:t, s) \\
 (E[\tscore(r)], t, s) & \to (E[(\lambda x. x)], t, rs),
@@
 \[
-K \sum_{(M,t,1) \to^* (V,[],s)} 2^{-\mathrm{length}(t)} s,
+\sum_{(M,[],1) \to^* (V,t,s)} 2^{-\mathrm{length}(t)} s.
 \]
-where $K$ is a normalising constant to make the probabilities all add up to 1.
 
 Alternatively, this could be defined in a different but closely related version of operational semantics, the big-step semantics:
-\begin{multline*}
-\frac{(M, t_0, s_0) \Downarrow (\lambda x. M', t_1, s_1) \; (N, t_1, s_1) \Downarrow (N', t_2, s_2) \; (M'[N'/x], t_2, s_2) \Downarrow (O, t_3, s_3)}{(M N, t_0, s_0) \Downarrow (O, t_3, s_3)} \\
+\begin{align*}
+\frac{(M, t_0, s_0) \Downarrow (\lambda x. M', t_1, s_1) \; (N, t_1, s_1) \Downarrow (N', t_2, s_2) \; (M'[N'/x], t_2, s_2) \Downarrow (O, t_3, s_3)}{(M N, t_0, s_0) \Downarrow (O, t_3, s_3)} \quad
 \frac{(M, 0:t_0, s_0) \Downarrow (M', t_1, s_1)}{(M \oplus N, t_0, s_0) \Downarrow (M', t_1, s_1)} \quad
-\frac{(N, 1:t_0, s_0) \Downarrow (N', t_1, s_1)}{(M \oplus N, t_0, s_0) \Downarrow (N', t_1, s_1)} \\
+\frac{(N, 1:t_0, s_0) \Downarrow (N', t_1, s_1)}{(M \oplus N, t_0, s_0) \Downarrow (N', t_1, s_1)} \quad
 \frac{}{(\tscore(r), t, s) \Downarrow (\lambda x. x, t, rs)}.
-\end{multline*}
+\end{align*}
 The notation in this case could be simplified somewhat, but that would obscure the connection to the small-step semantics.
@@
 \subsection{Denotational Semantics}
-The general format of denotational semantics is a function $\denotation M_\gamma$, from a term $M$ to its value, given the values of all the free variables in the context $\gamma$.
-The type of output depends on the type of the term, given by $\denotation A$ which is defined recursively.
-For a non-probabilistic language, it would be something like $\denotation{\mathsf T} = T$, $\denotation{A \to B} = \denotation B^{\denotation A}$.
-If $M$ satisfies the typing judgement $\Gamma \vdash M : A$, and $\gamma \in \denotation \Gamma$ (i.e.~$\gamma$ is a function from the variables in $\Gamma$ to values such that if $x:B$ is in $\Gamma$, $\gamma (x) \in \denotation B$), then $\denotation M_\gamma \in \denotation A$.
-
-For the example language but without the probabilistic constructs (i.e.~just simply-typed lambda calculus with constants), a denotational semantics could be defined as
-\begin{align*}
-\denotation t_\gamma & = t\,(\text{for }t \in T) \\
-\denotation x_\gamma & = \gamma(x) \\
-\denotation{M N}_\gamma & = \denotation M_\gamma(\denotation N_\gamma) \\
-\denotation{\lambda x. M }_\gamma(X) & = \denotation M_{\gamma ; x \mapsto X}.
-\end{align*}
-In order to define a denotational semantics for a probabilistic language, the result would have to be not just a value, but some sort of distribution of values.
-The rule $\denotation{A \to B} = \denotation B^{\denotation A}$ also has to be changed to take into account that a function may produce outputs that differ randomly each time it is called, so something like the set of kernels from $\denotation A$ to $\denotation B$ (a trace-based denotational semantics would also probably be possible, but much more complicated, but as far as I am aware, such a thing has never been defined).
-This would require a measure structure on $\denotation A$ and $\denotation B$ to be defined, but in some cases it is not possible to define a measurable structure on the set of kernels between two measurable spaces such that the application map taking a kernel $A \rightsquigarrow B$ and an element of A is itself measurable, therefore the denotations of higher-order types could be undefined.
-This motivates the introduction of quasi-Borel spaces.
-
-\subsubsection{Quasi-Borel Spaces}
-A quasi-Borel space is a pair of a base space $X$ with a set of random variables $M_X \subset X^{\mathbb R}$ that satisfies certain axioms.
-In many ways, quasi-Borel spaces are similar to measurable spaces.
-A $\sigma$-algebra on $X$ can be converted to a quasi-Borel space structure by taking all the measurable functions as random variables, and a quasi-Borel space can be converted to a measurable space by taking as measurable sets those whose inverse images under all random variables are measurable.
-There is a natural notion of QBS morphisms (functions whose compositions with random variables are random variables) and measures defined by combining measures on $\mathbb R$ with random variables which can be used to define integrals.
-The crucial difference to measurable spaces is that the set of morphisms betweem two QBSes is itself a QBS, so that the category of QBSes is cartesian closed.
-
-Let $G(X)$ be the set of measures on a QBS $X$, equipped with its own QBS structure, with the dirac delta $\delta : X \to G(X)$ and $\eta : (X \to G(X)) \to (G(X) \to G(X))$ as the monad operations, and let $QBS(X, Y)$ be the set of morphisms between two $QBSes$.
-Let $T$ be equipped with the discrete QBS structure.
-A denotational semantics for the example language, such that $\denotation M_\gamma \in G(\denotation A)$ if $\Gamma \vdash M : A$ and $\gamma \in \denotation \Gamma$ can then be given as follows.
-
-\begin{align*}
-\denotation{\mathsf T} & = T \\
-\denotation{A \to B} & = \denotation A \to G(\denotation B) \\
-\denotation{t}_\gamma & = \delta(t) \\
-\denotation{x}_\gamma & = \delta(\gamma(x)) \\
-\denotation{M N}_\gamma & = \eta(f \mapsto \eta(f)(\denotation N_\gamma))(\denotation M_\gamma) \\
-\denotation{\lambda x. M}_\gamma(X) & = \denotation M_{\gamma ; x \mapsto X} \\
-\denotation{M \oplus N}_\gamma & = \frac 1 2 \denotation M_\gamma + \frac 1 2 \denotation N \gamma \\
-\denotation{\tscore(r)}_\gamma & = r \denotation{\lambda x. x}_\gamma = r \delta(x \mapsto \delta(x)).
-\end{align*}
-
-Although the type of the result is different, this denotational semantics can still be shown to be equivalent to the operational semantics previously presented.
-For programs of type $\mathsf T$, both versions of the semantics produce a distribution on $T$ (one a QBS-style distribution and the other a measure-space-style distribution, but these can be converted), which are equal.
-The statement for function types is a bit more complicated and less direct, because the denotational semantics produces a distribution on (QBS) kernels, while the operational semantics produces a distribution on lambda terms.
-
-Having a denotational semantics available makes certain sorts of proofs more convenient.
-For example, if there's a program rearrangement that can be expressed as replacing $A[M,N]$ with $B[M,N]$ (for arbitrary $M, N$), then the proof of equivalence can be quite direct, showing that the way $\denotation M$ and $\denotation N$ are combined is equivalent (possibly subject to some conditions on certain variables not being free in $M$ and $N$, which implies that $\denotation M_\gamma$ and $\denotation N_\gamma$ are independent of the values of those variables in $\gamma$), whereas with operational semantics, it is much more complicated to keep track of the way the arbitrary subterms are duplicated, transformed and combined as execution progresses.
-Similarly, the notion of contextual equivalence is not necessary in denotational semantics (being trivially equivalent to equality of values).
-
-Another useful property of denotation semantics is that termination properties are much more directly visible.
-If the type system is strong enough to imply termination, this manifests directly as every term being given a value.
-If non-terminating terms are possible, it will be necessary to introduce, as part of the semantics, special cases to deal with that, otherwise the definition will simply not be consistent.
-
-\subsection{Trace-based semantics}
-A trace-based semantics for the example language was already described in Section \ref{operational semantics}.
-In trace-based semantics, the results are defined not just as distributions, but as random variables.
-Everything can be seen as taking place in a probability space (neglecting $\tscore$ which, at least if it is allowed to take values greater than 1, does not entirely fit), with the term after $n$ steps, the values of certain samples, the running time, and the final value all being random variables in the same space.
-
-There are a few variations on how the traces work.
-In some versions (as is the case in Section \ref{operational semantics}), each run of the program is assigned a trace of finite length, containing exactly those samples actually used in that run.
-If the set of results for an individual sampling operation is $S$ (in this case it's the set of 2 elements, but in a language with continuous distributions, it could be $\mathbb R$, or an interval, or something else), the set of traces is $\bigcup_{n \in \mathbb N} S^n$, and the measure defined on the set of traces is not actually a probability measure, as each set $S^n$ has measure 1.
-This still acts somewhat like a probability measure though, as traces of different lengths are mutually exclusive.
-If a trace $a$ is a prefix of a trace $b$ (with $b = a + c$), then it is not possible that $(M, a, 1) \to^* (V, [], s)$ and also $(M, b, 1) \to^* (V', [], s')$, because the evaluation with any given trace is deterministic, and $(M, b, 1) \to^* (V, c, s)$, at which point it could not proceed any further, having already reached a value.
-For any term, the set of traces for which it terminates therefore has measure at most 1.
-
-Alternatively, the trace could contain infinitely many values, with any finite execution of a program leaving some of the samples unused.
-This is in some ways a little simpler, because the set of traces can simply have a probability distribution, and the trace does not need to depend in any way on the program (by having its length matched).
-An additional variant of this style of trace is to split the trace in three parts, all infinitely long, for the application case of the big-step semantics, rather than having the each step (evaluating the function, evaluating the argument, then evaluating the result of the $\beta$ reduction) use however many samples it needs from the start of the trace in turn.
-
-\subsection{Distribution-Based Semantics}
-Rather than defining the value of a program as a random variable and then defining the distribution based on that, it is possible to define the distribution of results directly.
-The example given above of operational semantics was distribution based (in the QBS sense of ``distrbution''), but to more clearly illustrate the difference, I will give a distrbution-based operational semantics for the example language, using the same definition of environments as in the trace-based semantics.
-
-\begin{align*}
-s(E[(\lambda x. M) V]) & = \delta(M[V/x]) \\
-s(E[M \oplus N]) & = \frac 1 2 \delta(E[M]) + \frac 1 2 \delta(E[N]) \\
-s(E[\tscore(r)]) & = r \delta(E[(\lambda x. x)]) \\
-s(V) & = \delta(V).
-\end{align*}
-
-The multi-step function $s^n$ can then be defined as $s^0(M) = M$, $s^{n+1}(M) = \int s^n(x)\,\mathrm d s(M)(x)$, then $s^\infty(M)$ can be defined as $\lim_{n \to \infty} s^n(M)$, which (after re-normalising) gives the distribution of final values.
-In this case, everything terminates because of the type system, therefore the limit is always well-defined.
-For an untyped language, or one with an unrestricted recursion operator, the definition of the final result would be a little more complicated, but the general idea is the same.
+        %Denotational Semantics in the non-probabilistic case, types and inconsistencies
+        %QBSes
+    %Distribution-based Semantics
+    %Trace-based Semantics
 

### 2021-10-01 LICS21/confluent.tex
Dealt with all unresolved references.

@@
 All of the theorems on the termination of rankable terms therefore extend to other reduction strategies too.
-The proofs of Theorem~\ref{thm:rankable implies termination,thm:partial implies rankable,thm:antitone rankable implies termination,thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies.
+The proofs of Theorems~\ref{thm:rankable implies termination}, \ref{thm:partial implies rankable},
+\ref{thm:antitone rankable implies termination} and~\ref{thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies.
 They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because Theorem~\ref{thm:de groote} applies equally to any reduction strategy).
@@
 \[
-\begin{array}{l}
-\mathit{Rch}_r(M) := \{N \mid \exists n, (N_i)_{0 \leq i \leq n} : {}\\
-\qquad\qquad\qquad N_0 = M, N_n = N, N_i \to N_{i+1} \text{ at } r(N_i)\}.
-\end{array}
+%\begin{array}{l}
+\mathit{Rch}_r(M) := \{N \mid \exists n, (N_i)_{0 \leq i \leq n} : 
+N_0 = M, N_n = N, N_i \to N_{i+1} \text{ at } r(N_i)\}.
+%\end{array}
 \]

### 2021-10-01 LICS21/intro.tex
Dealt with all unresolved references.

@@
 
-By combining these methods we can prove AST of a variety of PPCF programs that are beyond the reach of methods in the literature, including non-affine recursion\footnote{\label{fnote:non-affine} Non-affine recursive programs are recursive programs that can, during the evaluation of the recursive body, make multiple recursive calls (of a first-order function) from distinct call sites.} (Examples~\ref{ex:non-affine recursion easy,ex:non-affine continuous}) and recursions that define higher-order functions (Example~\ref{ex:higher-order recursion}).
+By combining these methods we can prove AST of a variety of PPCF programs that are beyond the reach of methods in the literature, including non-affine recursion\footnote{\label{fnote:non-affine} Non-affine recursive programs are recursive programs that can, during the evaluation of the recursive body, make multiple recursive calls (of a first-order function) from distinct call sites.} (Examples~\ref{ex:non-affine recursion easy} and~\ref{ex:non-affine continuous}) and recursions that define higher-order functions (Example~\ref{ex:higher-order recursion}).
 

### 2021-10-01 LICS21/related.tex
Dealt with all unresolved references.

@@
 % (because the underlying recursion schemes are not Turing complete):
-for example, the PPCF terms in Examples~\ref{ex:raven complex,ex:non-affine continuous,ex:higher-order recursion}
+for example, the PPCF terms in Examples~\ref{ex:raven complex},
+\ref{ex:non-affine continuous} and
+\ref{ex:higher-order recursion}
 %\refExample{complexExample} 
@@
 
-Although this is a broadly applicable method of proving almost sure termination of probabilistic functional programs, and Theorems~\ref{thm:antitone partial implies rankable,thm:confluent ranking} make it more convenient to use, some method of automating the construction and checking of (antitone) ranking functions, even partially, could make it considerably more practically useful, especially in cases where almost sure termination is merely a side-condition for some other theorem or algorithm to be applicable.
+Although this is a broadly applicable method of proving almost sure termination of probabilistic functional programs, and Theorems~\ref{thm:antitone partial implies rankable} and \ref{thm:confluent ranking} make it more convenient to use, some method of automating the construction and checking of (antitone) ranking functions, even partially, could make it considerably more practically useful, especially in cases where almost sure termination is merely a side-condition for some other theorem or algorithm to be applicable.
 

### 2021-10-01 LICS21/SPCF.tex
Use \ref instead of \Cref.
" \ref" -> "~\ref"
Various minors (formatting)

@@
   \text{\em values } V &::= \lambda x.M  \mid  \underline{r} \\
-  \text{\em terms } M, N &::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY \, M \\
-   & \qquad \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample
+  \text{\em terms } M, N &::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY \, M \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample
 \end{align*}
-The typing rules are standard (see \Cref{fig:typing rules}).
+The typing rules are standard (see Figure~\ref{fig:typing rules}).
 Terms are identified up to $\alpha$-equivalence, as usual. 
@@
 To define the CBV reduction relation, let \emph{evaluation contexts} be of the form:
-\begin{align*}
-  E ::= & \, [\,] \mid E \, M \mid V \, E \mid \underline{f}(\underline r_1,\dots ,\underline r_{k-1}, E, M_{k+1}, \dots, M_n) \\ & \mid \tY E \mid \tif{E<0}{M_1}{M_2}
-\end{align*}
+\[
+  E ::= [\,] \mid E \, M \mid V \, E \mid \underline{f}(\underline r_1,\dots ,\underline r_{k-1}, E, M_{k+1}, \dots, M_n) \mid \tY E \mid \tif{E<0}{M_1}{M_2}
+\]
 then a term reduces if it is formed by substituting a redex in a context i.e.
@@
   E[\underline f (\underline r_1, \dots , \underline r_n)] & \to E[\underline{f(r_1,\dots,r_n)}] \\
-  E[\tY \lambda x. M] & \to E[\lambda z. M[(\tY \lambda x. M)/x] z] \\
-  & \qquad \qquad \text{ where $z$ is not free in $M$}\\
+  E[\tY \lambda x. M] & \to E[\lambda z. M[(\tY \lambda x. M)/x] z]\quad \text{ where $z$ is not free in $M$}\\
   E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_1] \text{ where }r < 0 \\

### 2021-10-01 LICS21/antitone.tex
Various minors (formatting)

@@
 \begin{align*}%\textstyle
-& g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 \\
+g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 
   &=  \ln \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) 
@@
 \label{ex:non-affine continuous}
+\iffalse
 \begin{align*}
@@
 \end{align*}
+\fi
+\[
+(\tY\, \lambda f x.\, (\lambda e. X[e]) \, \tsample)\, \underline 0 \quad \hbox{where} \quad
+X[e] := \tif{e \leq p - x^{-2}}{x}{f^2(x+1) \oplus_e f(x+1)}
+\]
 \end{example}
@@
 recursively defined\footnote{Inspired by examples in \cite{DBLP:journals/pacmpl/BurnOR18,DBLP:conf/lics/OngW19}.} by
+\[
+\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} \,.\, \tif{n \leq 0}{s}{
+f \, n \, (\varphi \, f \, s \, (n - 1))
+\oplus_p
+f \, n \, (\varphi \, f \, s \, (n + 1))
+}
+\]
+\iffalse
 \[
@@
 \]
+\fi
 For any $F : \textsf{R} \to \textsf{R} \to \textsf{R}$ such that $F\, \underline n\, \underline m$ terminates with no $\tY$ reductions for all $m$ and $n$, $\Xi\, F\, \underline x\, \underline y$ is antitone rankable.

### 2021-10-01 LICS21/SPCF.tex
Latex minors

@@
     \begin{array}{ll}
-        (E[N],s) & \text{if } M = E[R], R \to N\\
-        & \qquad \text{ and } R \neq \tsample \\
+        (E[N],s) & \text{if } M = E[R], R \to N \text{ and } R \neq \tsample \\
         (E[\underline{\pi_h(s)}],\pi_t(s)) & \text{if } M = E[\tsample] \\

### 2021-10-16 LICS21/antitone.tex
Moved all appendices (except those concerning Sec. 6) to main body. Moved references to the end of document.

@@
 
-\begin{restatable}{therm}{ACStrict}
+\begin{therm}
+%\begin{restatable}{therm}{ACStrict}
 \label{thm:a-c strict}
@@
 Then $T < \infty$ a.s.
-\end{restatable}
+%\end{restatable}
+\end{therm}
 
@@
 
+\begin{proof}
+First, as $(Y_n)$ is a supermartingale, $\expect{Y_n} \leq \expect{Y_0}$. 
+Therefore
+\begin{calculation}
+\expect{Y_n \mid T > n}
+   \step[=]{rearranging terms}
+% \dfrac{\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
+% + 
+% \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
+% - 
+% \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
+\dfrac{
+\begin{array}{l}
+\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
++ 
+\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
+\\
+{} - 
+\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}
+\end{array}
+}{\mathbb P[T > n]}
+  \step[=]{definition of conditional expectation}
+\dfrac{\expect{Y_n} - \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
+  \step[\leq]{$Y_n \geq 0$ always}
+\dfrac{\expect{Y_n}}{\mathbb P[T > n]}
+  \step[\leq]{$(Y_n)_n$ is a supermartingale}
+\dfrac{\expect{Y_0}}{\mathbb P[T > n]}
+\end{calculation}
+
+\emph{Claim}: For all $0 < x \leq 1$, 
+\(
+\mathbb P[T > B_x] \leq x,
+\)
+where $B_x = \left \lceil \dfrac{\expect{Y_0}+1}{x \; \epsilon(\expect{Y_0} \, x^{-1})} \right \rceil$ and $\epsilon : \nnReal \to \pReal$ is the antitone function.
+
+\smallskip
+
+As the convex hull of $\epsilon$ (the greatest convex function less than or equal to it) satisfies all the conditions assumed of $\epsilon$, in addition to being convex, assume wlog that $\epsilon$ is convex.
+
+Assume for a contradiction that $\mathbb P[T > B_x] > x$.
+Then, take $n \leq B_x$. 
+We have
+\begin{calculation}
+\expect{Y_n-Y_{n+1}}
+  \step[=]{$\calF_n \subseteq \calF_{n+1}$, def.~\& linearity of cond.~expectation}
+\expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
+  \step[\geq]{antitone strict assumption}
+\expect{\epsilon(Y_n) \cdot {\bf 1}_{\set{T > n}}}
+  \step[=]{definition of expectation conditioning on an event}
+\mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
+  \step[\geq]{Jensen's inequality}
+\mathbb P[T > n]\ \epsilon(\expect{Y_n \mid T > n})
+  \step[\geq]{proved earlier}
+\mathbb P[T > n]\ \epsilon\Big(\dfrac{\expect{Y_0}}{\mathbb P[T > n]}\Big)
+  \step[>]{assumption, $\mathbb P[T > n] \geq \mathbb P[T > B_x] > x$}
+x \; \epsilon\big(\expect{Y_0} \, x^{-1}\big).
+\end{calculation}
+Therefore, by a telescoping sum
+\begin{align*}
+\expect{Y_{B_x}} 
+&= \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})\\
+&\leq -1 < 0 
+\end{align*}
+%\lo{$\expect{Y_0 - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})}$}
+which is a contradiction, therefore the claim must be true, therefore $P[T > n] \to 0$ as $n \to \infty$, therefore $T < \infty$ a.s.
+\end{proof}
 \begin{example}
@@
 
-\begin{restatable}[Sparse function]{therm}{AntitonePartial}  \label{thm:antitone partial implies rankable}
+\begin{therm}
+%\begin{restatable}[Sparse function]{therm}{AntitonePartial}  
+\label{thm:antitone partial implies rankable}
   Every antitone sparse ranking function is a restriction of an antitone ranking function.
-\end{restatable}
+%\end{restatable}
+\end{therm}
 
+\begin{proof}
+  Take a closed term $M$ and an antitone sparse ranking function $f$ on $M$, with a corresponding antitone function $\epsilon$.
+Assume wlog that $\epsilon$ is convex (as if it isn't, we can just take its convex hull instead).
+As in Theorem~\ref{thm:partial implies rankable}, define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
+  \begin{align*}
+    f_1(N) &= f(N) \text{ whenever $f(N)$ is defined},\\
+    f_1(V) &= 0 \text{ for values $V$ not in the domain of $f$.}
+  \end{align*}
+
+  Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) := \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY\, \lambda x.
+N'],s') \} \right |$.
+  The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on antitone partial ranking functions.
+Define $f_2(N) := \int_\entrosp f_1(\nnext(N,s)) + \epsilon(f_1(\nnext(N,s))) g(N,s) \, \mu_{\entrosp}(\mathrm d s)$.
+For any term $N$ where $f$ is defined, $f_2(N) = f(N)$, and the value that $f_2$ would have at $N$ if $f$ were not defined at $N$ is $\leq f(N)$, by the third condition on antitone partial ranking functions.
+In order to show that $f_2$ is an antitone partial ranking function, it therefore suffices to show that the value that $f_2$ would have had at each term if $f$ were not defined at that term is at least the expectation of $f_2$ after one reduction step (plus $\epsilon(f_2(N))$ if the reduction step is a $tY$-reduction).
+For any term $N$ which is not of the form $E[R]$ for some $\tY$-redex $R$, this is trivial.
+If $R$ is a $\tY$-redex, then $\epsilon(f_2(N)) \leq \int_\entrosp \epsilon(f_1(\nnext(N,s)))\, \mu_{\entrosp}(\mathrm d s)$ by the convexity of $\epsilon$, because $n$ is bounded.
+Therefore, the (total) function $f_2$, which agrees with $f$ on $f$'s domain, is an antitone ranking function on $M$ (with the same function $\epsilon$ if it's convex).
+\end{proof}
 As a corollary, any term which admits an antitone sparse ranking function terminates almost surely.
 
-\paragraph*{Example~\ref{ex:ac-ranking} revisited}
+\begin{example}[Example~\ref{ex:ac-ranking} revisited]
 Programs $M_1$ and $M_2$ are AST. For $M_1$, the sparse ranking function 
@@
 suffices to prove its termination (and could equivalently be considered an antitone sparse ranking function with the constant antitone function $\epsilon_1(x) = 1$).
+\end{example}
 
@@
 g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 
-  &=  \ln \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) 
+  &=  \ln \left(\frac{n+1}{\sqrt{n(n+2)}}\right) 
   =  \frac 1 2 \ln\left(\frac{(n+1)^2}{n(n+2)}\right) \\
@@
 
-\paragraph*{More challenging examples} All of the following examples also have antitone ranking functions, provided in Appendix~\ref{apx:antitone}.
+\subsection{More challenging examples}
 
-\iffalse
+%\iffalse
 Sampling from continuous distributions is an essential feature of statistical probabilistic programming languages.
@@
 To our knowledge, the problem of proving AST of probabilistic functional programs with continuous distribution is new.
-\fi
+%\fi
 
+All of the following examples also have antitone ranking functions.
+%, provided in Appendix~\ref{apx:antitone}.
+
+\iffalse
 \begin{example}[Continuous random walk]\label{ex:raven complex}
@@
 \end{example}
+\fi
 
-\paragraph*{Completeness}
+\begin{example}[Continuous random walk]
+\label{ex:raven complex}
+Let $\Theta := \tY \, \lambda f x . \tif{x \leq 0}{0}{f(x - \tsample)}$. 
+We can construct a sparse ranking function $f$ for $\Theta \, \underline{10}$ as follows:
+\begin{align*}
+\Theta \, \underline l 
+&\mapsto 
+l + 2
+\\
+\tif{\underline l \leq 0}{0}{\Theta \, (\underline l - \tsample)}
+&\mapsto
+l + 1
+\\
+\underline 0 &\mapsto 0.
+\end{align*}
+
+For a more complex (not $\tY$-PAST) example, consider the following ``continuous random walk'': $\Xi \, \underline{10}$ where
+\[
+%\underbrace{\big(
+\Xi := \tY \, \lambda f x . \tif{x \leq 0}{0}{f(x - \tsample + 1/2)} 
+%\big)}_{\Xi} \, 10
+\]
+Let $g(x) := 2 + \ln(x + 1)$, and let $\epsilon$ be the function specified by
+\[
+\epsilon(g(x-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y.
+\]
+The limit of this as $x \to \infty$ and $g(x+1/2) \to \infty$ is 0, and $\frac {\mathrm d}{\mathrm dx} \epsilon(g(x+1/2)) = \frac 1 {x+1} + \ln(1 - \frac 1 {x + 1/2}) < 0$, and $g$ is monotonic increasing;
+therefore $\epsilon$ is antitone and bounded below by 0.
+We define an antitone sparse ranking function by:
+\[
+\Xi \, \underline l 
+\mapsto 
+g(l), \qquad 
+\underline 0 \mapsto 0
+\]
+
+% \begin{align*}
+% \Xi \, l 
+% &\mapsto 
+% g(l)
+% \\
+% 0 &\mapsto 0
+% \end{align*}
+The value of $g$ after one $\tY$-reduction step is at least $g(l-1/2)$, therefore the expectation of $\epsilon$ after one $\tY$-reduction step is at most $\epsilon(g(l-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y$.
+Thus 
+\begin{itemize}
+\item in case $l > 0$, $g$ decreases by the required amount
+\item in case $l \leq 0$, 
+\(
+g(\Xi \, l) \geq 2 + \ln(1/2) > \ln(\frac{3 \sqrt 3} 2) - 1 = \epsilon(g(0-1/2))
+\) 
+as well.
+\end{itemize}
+Hence this is a valid antitone sparse ranking function and the term is AST.
+\end{example}
+
+\begin{example}[Fair-in-the-limit random walk] 
+\label{ex:Fair-in-the-limit random walk}
+\citep[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}
+\[
+\underbrace{\big
+(\tY \, \lambda f x .\ 
+\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)}_{\Xi} 
+\, \underline{10}
+\]
+To construct an antitone ranking function, we solve the recurrence relation:
+% \[
+% \left\{
+% \begin{array}{rll}
+% z_0 &=& 0\\
+% n \geq 1, \quad z_n &>& \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
+% \end{array}
+% \right.
+% \]
+\begin{align*}
+z_0 &= 0\\
+n \geq 1, \quad z_n &> \textstyle \frac{n}{2n + 1} \, z_{n-1} + \frac{n+1}{2n + 1} \, z_{n+1}.
+\end{align*}
+For $n > 2$, $z_n = \ln(n-1)$ works.
+The expected decrease is $\frac 1 2(\ln(1+\frac 1 {(n-1)^2-1}) - \frac 1 {2n+1} \ln(1 + \frac 2 {n-2}))$, and using the fact that $\frac x {x+1} \leq \ln (1+x) \leq x$ for $x > 0$, this is at least $\frac 1 2(\frac 1 {(n-1)^2} - \frac 1 {2n + 1} \frac 2 {n-2}) = \frac {n^2}{2(n-1)^2(2n+1)(n-2)}$, which (again for $n > 2$) is positive and antitone.
+For $n = 0, 1, 2$, we take $\epsilon$ to be 9/40 (the same as its value at 3), then set $z_2 = 2 \ln 2 - \ln 3 - 1, z_1 = 3 \ln 2 - 2 \ln 3 - 3, z_0 = 4 \ln 2 - 3 \ln 3 - 6$.
+Some of those values are negative, but it's still bounded below, so by adding a constant offset it can be corrected, and the term is AST.
+\end{example}
+
+\begin{example}[Escaping spline]
+\label{ex:escaping spline}\cite[\S 5.4]{DBLP:journals/pacmpl/McIverMKK18}
+\[
+\underbrace{\big
+(\tY \, \lambda f x . \,
+0 \oplus_{\frac{1}{x+1}} f(x + 1) \big)}_{\Xi} 
+\, \underline{10}
+\]
+
+In this case, the fact that the ranking function must decrease at each $\tY$-step (in an antitone sparse ranking function) by the expected value of $\epsilon$ not at the current term, but at the next term where the ranking function is defined, is a little harder to deal with, because the variable $x$ can change all the way to $0$ in one step, therefore simply adding a small offset doesn't suffice to compensate for this fact.
+
+Consider the candidate ranking function $\Xi \, \underline n \mapsto n + 1$.
+For each $n$, $\Xi \, \underline n$ reduces to either $\underline 0$ (with probability $\frac 1 {n + 1}$) or $\Xi \, \underline{n + 1}$, therefore the expected value of the ranking function is $\frac{n(n+2)}{n+1} = n + 1 - \frac 1 {(n+1)^2}$, and the required decrease is $\frac{\epsilon(0) + n \epsilon(n + 2)}{n + 1}$, therefore eventually, the expected decrease isn't enough, whatever the value of $\epsilon(0)$ is.
+
+If instead, we take the ranking function to be defined after the $\tY$-reduction but before the $\tsample$-reduction as well, this can be resolved.
+Letting $\Theta[n] = \underline 0 \oplus_{\frac 1 n} \Xi \, n$:
+\begin{align*}
+\Theta[\underline n] &\mapsto n \\
+\Xi \, \underline{10} & \mapsto 12.
+\end{align*}
+For each $n$, $\Theta[n]$ reduces to either $0$ or $\Theta[n+1]$, with a $\tY$-reduction only in the latter case, and the condition that this is an antitone sparse ranking function is that $n \geq \frac{(n-1)(n+1)}{n} + \frac{n-1}{n} \epsilon(n+1)$ and $12 \geq 11 + \epsilon(11)$.
+These are satisfied by setting $\epsilon(x) = \min(1, 1/x)$, therefore this term is AST.
+\end{example}
+
+%\medskip
+
+\subsection*{Non-affine recursion}
+
+Many of the recent advances in the development of AST verification methods \citep{DBLP:conf/pldi/ChenH20,DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18,DBLP:conf/aplas/HuangFC18,DBLP:conf/popl/ChatterjeeNZ17,DBLP:journals/pacmpl/AgrawalC018,DBLP:conf/cav/ChatterjeeFG16,DBLP:conf/lics/OlmedoKKM16,DBLP:journals/pacmpl/Huang0CG19} are concerned with loop-based programs.
+We can view such loops as tail-recursive programs that are, in particular, \emph{affine recursive}, i.e., 
+in each evaluation (or run) of the body of the recursion, recursive calls are made from at most one call site \cite[\S 4.1]{DBLP:journals/toplas/LagoG19}.
+(Note that whether a program is affine recursive cannot be checked by just counting textual occurrences of variables.)
+Termination analysis of \emph{non-affine recursive} probabilistic programs does not seem to have received much attention.
+Methods such as those presented in \cite{DBLP:journals/toplas/LagoG19} are explicitly restricted to affine programs, and are unsound otherwise.
+By contrast, many probabilistic programming languages allow for richer recursive structures \citep{DBLP:conf/pkdd/TolpinMW15,DBLP:conf/uai/GoodmanMRBT08,DBLP:journals/corr/MansinghkaSP14}.
+
+\begin{example}[Non-affine recursion]
+\label{ex:non-affine recursion easy}
+Let 
+\[
+\Xi := \tY \lambda f x. x \oplus_{2/3} f (f (x + 1))
+\] 
+A suitable sparse ranking function for $\Xi \, \underline 1$ would be $\Xi^n \, \underline i \mapsto 3 \, n$ (for $n \geq 0$), because $\Xi^n \, \underline i$ reduces to either $\Xi^{n+1} \, \underline{i+1}$ or $\Xi^{n-1} \, \underline{i+1}$, with probabilities $1/3$ and $2/3$.
+The value of $i$ does not actually matter for the progress of this recursion.
+It is basically another variant of the biased random walk, except that the relevant variable is the number of copies of $\Xi$, instead of a real number in the term.
+\end{example}
+
+\begin{example}[More complex non-affine recursion]
+\label{ex:non-affine continuous}
+Let
+\begin{align*}
+\Xi & = \tY\, \lambda f x.\, (\lambda e. X[e]) \, \tsample \\
+X[e] & = \tif{e \leq p - x^{-2}}{x+1}{f^2(x+1) \oplus_e f(x+1)}
+\end{align*}
+
+This example is a little more complex because of the use of a random sample, $e$, as a first class value, which cannot be modelled via discrete distributions.
+
+We can use the ranking function method (coupled with the solution of linear recurrence relations) to show that provided $p \geq \frac{5 - \sqrt{21}}{2}$, the program $\Xi \, \underline 3$ is AST. 
+Consider the edge case, that $p = \frac{5 - \sqrt{21}}{2}$ exactly.
+
+The term $\Xi^n\, \underline x$ (for $n > 0$, $x^{-2} < p$) reduces to either $\Xi^{n-1}\, \underline{x+1}$, $\Xi^n\ \underline{x+1}$ or $\Xi^{n+1}\, \underline{x+1}$, 
+with probabilities $p-x^{-2}$, $(1-p+x^{-2})\frac{p-x^{-2}}{2}$ and $(1-p+x^{-2})(1-\frac{p-x^{-2}}{2})$ respectively. 
+Let $a(\Xi^n\, \underline x) = n + 2 \, x^{-1}$. 
+This is a supermartingale (because the decrease in $2 \, x^{-1}$ as $x$ increases is enough to offset the average increase in $n$), but it does not satisfy the antitone-strict progress condition. 
+It does however have a bounded-below variance, so the usual method of using $\ln a$ instead of just $a$ works. 
+Calculating the exact amount that $\ln a$ decreases is not necessary, because it can be bounded as follows: 
+$a$ changes by at least $1/2$ with probability at least $1/11$ (assuming $x \geq 3$), 
+therefore $\ln(a)$ decreases in expectation by at least $\frac{1}{11}(\ln(a+\frac{1}{2}) - \ln(a) -\frac 1 {2a})$, 
+using the fact that a linear approximation to $\ln$, applied to $a$, at least doesn't increase, 
+then adding the deviation of $\ln$ from its linear approximation $\ln(a) + \frac{x-a}{a}$ 
+(and using the fact that the linear approximation is everywhere an overestimate), 
+we obtain a sufficiently strong bound on the decrease of $\ln(a)$ that it's an antitone sparse ranking function. 
+(It can obviously be extended to $\Xi\, \underline 1$ too, which reduces to $\Xi^n\, \underline 3$ or a value after a bounded number of steps, but that complicates the analysis a little.)
+\end{example}
+
+\subsection*{Higher-order recursion}
+
+There is an important formalism for constructing (non-random) higher-order recursive functions, viz., \emph{higher-order recursion schemes} (HORS) (see e.g.~\cite{DBLP:conf/lics/Ong06,DBLP:conf/lics/Ong15}).
+%(Incidentally the (sparse) ranking function method is just as applicable to the termination analysis of \emph{deterministic} PPCF programs.)
+Recently \cite{DBLP:conf/lics/KobayashiLG19} have extended HORS to \emph{probabilistic higher-order recursion schemes} (PHORS), which are HORS augmented with probabilistic (binary) branching $\oplus_p$.
+As HORS are in essence the $\lambda \tY$-calculus \citep{DBLP:conf/lics/Statman02} (i.e.~pure simply-typed lambda calculus with recursion, generated from a finite base type), PHORS are definable in PPCF:
+(order-$n$) PHORS is encodable as (order-$n$) call-by-name PPCF, but the former is strictly less expressive (because the underlying HORS is not Turing complete). 
+For example, the PPCF term in Example~\ref{ex:higher-order recursion} is not definable in PHORS.
+%Some interesting PPCF terms such as \Cref{ex:non-affine with continuous distribution} %\refExample{complexExample} 
+%cannot be expressed as PHORS.
+%\akr{As I removed the referenced example, and I don't even know what a PHORS is, I've just removed this comment too.}
+
+A relevant result here is that the AST problem is decidable for order-1 PHORS (by reduction to the PSPACE-hard solvability of finite systems of polynomial equations with real coefficients; see \cite{DBLP:journals/jacm/EtessamiY09}), but undecidable for PHORS with order greater than 1 \citep{DBLP:conf/lics/KobayashiLG19}.
+
+\begin{example}[Higher-order recursion]
+\label{ex:higher-order recursion}
+Consider the higher-order function
+$\Xi : (\textsf{R} \to \textsf{R} \to \textsf{R}) \to \textsf{R} \to \textsf{R} \to \textsf{R}$
+recursively defined by
+\[
+\begin{array}{l}
+\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} . \\
+\quad\tif{n \leq 0}{s}{
+f \, n \, (\varphi \, f \, s \, (n - 1))
+\oplus_p
+f \, n \, (\varphi \, f \, s \, (n + 1))
+}
+\end{array}
+\]
+Let $F : \textsf{R} \to \textsf{R} \to \textsf{R}$ be a function such that $F\ \underline n\ \underline m$ terminates with no $\tY$ reductions for all $m$ and $n$.
+Then $\Xi\ F\ \underline x\ \underline y$ has the antitone sparse ranking function
+\[
+F \, \underline{n_0} \, (F \, \underline{n_1} \, (\dots(\Xi\ F\, \underline x\, \underline n)\dots) \mapsto g_2(n),
+\] 
+where $g_2$ is as in Example~\ref{ex:unbiased random walk}, using the same antitone function too.
+
+The inequalities required for this to be an antitone partial supermartingale are satisfied with just the same reasoning as in Example~\ref{ex:unbiased random walk}, therefore this term too is antitone rankable.
+
+Again, this is not actually the correct reduction order, in that $F$ should be applied to its argument $n_i$ and reduced to a value before its second argument is expanded, 
+but this would complicate the presentation, and this version can be justified by Theorem~\ref{thm:confluent ranking} instead. 
+The reduction strategy implied here should be clear enough (assuming that, where it isn't specified, it just matches $\cbv$), but to be more precise, 
+let 
+\[
+r \, (F \, \underline{n_0} \, (\dots(F\, \underline {n_{k-1}}\, X)\dots)) = @_2^k;\cbv(X),
+\] 
+where $X$ is not a value or of the form $F\ \underline{n_k}\ Z$ for some non-value $Z$. 
+Within $X$, there is necessarily a redex at $\cbv(X)$. 
+The only conditions that $@_2^k;\cbv(X)$ is also a position of a redex in the whole term are that if the redex is a $\tsample$, 
+it is not inside a $\lambda$ that's inside a $\tY$ or on the right of an application, which is true because $\cbv$ never selects a redex inside of a $\lambda$. 
+After $n$ reaches $0$, the term is $F \, \underline{n_0} \, (\dots(F\, \underline {n_k} \, \underline x)\dots)$, then the innermost $F$ is evaluated completely in $\cbv$ order, therefore (by assumption) it terminates.
+Because there are no $\tY$-reductions in the evaluation of $F$, it can't reach any other term of the form $F\, \underline{n_k'}\, X$, 
+therefore $k$ never changes until this whole subexpression reaches a number, at which point $k$ decreases by 1 and the next $F$ is evaluated.
+\end{example}
+
+\iffalse
+\antitoneSeemsComplete*
+
+\begin{proof}
+If $T$ is bounded by $b$ a.s.~(for $n$ constant), the statement is trivially true by taking $\epsilon$ to be constantly 1, and $Y_n = b - n$.
+Otherwise, let $(t_n)_{n \geq 0}$ be defined recursively such that $t_0 = 0$, $t_{n+1} > t_n$ and $\mathbb P[T > t_{n+1}] \leq \frac 1 4 \mathbb P[T > t_n]$.
+We will then define a supermartingale $(Y_n)_{n \geq 0}$ and nonrandom sequence $(y_n)_{n \geq 0}$ such that $Y_n = 0$ iff $T < n$, $Y_n = y_n$ iff $T \geq n$, and $y_n \geq 2^k$ for $n \geq t_k$.
+
+The antitone function $\epsilon$ is defined piecewise and recursively in such a way as to force all the necessary constraints to hold:
+\begin{align*}
+    \text{for }x \in [0,1),\ &\epsilon(x) = 1 \\
+    \text{for }x \in [2^k,2^{k+1}),\ &\epsilon(x) = \min \left (\epsilon(2^{k-1}), \frac{2^k}{t_{k+1}-t_k} \right ).
+\end{align*}
+This ensures that
+\begin{itemize}
+\item $\epsilon$ is weakly decreasing. 
+\item $\epsilon$ is bounded below by 0.
+\item For $x \geq 2^k$, $\epsilon(x) \leq \frac{2^k}{t_{k+1}-t_k}$. 
+\end{itemize}
+
+The sequence $(y_n)_{n \geq 0}$ is then defined recursively by
+\begin{align*}
+    y_0 = 2, \quad
+    y_{n+1} = (y_n - \epsilon(y_n)) \frac{\mathbb P[T > n]}{\mathbb P[T > n+1]}.
+\end{align*}
+
+The fact $y_{t_k} \geq 2^{k+1}$ is proven by induction on $k$.
+For base case, $2 \geq 2$, as required.
+For the inductive case $k+1$, we first do another induction to prove that $y_n \geq 2^k$ for all $t_k \leq n \leq t_{k+1}$.
+The base case of this inner induction follows from the outer induction hypothesis a fortiori.
+Take the greatest $k$ such that $n > t_k$.
+By the induction hypothesis, $y_{t_k} \geq 2^k$.
+
+\begin{align*}
+    y_n & = y_{t_k} \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} - \sum_{m=t_k}^{n-1} \epsilon(y_m) \frac{\mathbb P[T > m+1]}{\mathbb P[T > n]} \\
+    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \sum_{m=t_k}^{n-1} \epsilon(y_m)) \\
+    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \sum_{m=t_k}^{n-1} \frac{2^k}{t_{k+1}-t_k}) \\
+%\end{align*}
+%Allow a page break so that the sim-diagram lines up neatly.
+%\begin{align*}
+    & = \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \frac{2^k (n - t_k)}{t_{k+1}-t_k}) \\
+    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - 2^k) \\
+    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (2^{k+1} - 2^k) \\
+    & = \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} 2^k \\
+    & \geq 2^k
+\end{align*}
+as required. Substituting $n = t_{k+1}$ and using the fact that $\frac{\mathbb P[T > t_k]}{\mathbb P[T > t_{k+1}]} > 4$, the same reasoning gives $y_{t_{k+1}} \geq 2^{k+2}$, as required.
+
+The fact $y_n \geq 2^k$ for some $k$ implies that $y_n > 0$ for all $n$, which implies that $Y_n \geq 0$. 
+The other condition for $(Y_n)$ to be an antitone-strict supermartingale is
+\begin{align*}
+    \expect{Y_{n+1} \mid \mathcal F_n} 
+    &=  \expect{y_{n+1} \, {\bf 1}_{\set{T \geq n+1}} \mid \mathcal F_n} \\
+    &=  y_{n+1} \, \expect{{\bf 1}_{\set{T \geq n+1}} \mid \mathcal F_n} \\ 
+    &=  y_{n+1} \, {\bf 1}_{\set{T \geq n}} \textstyle \frac{\mathbb P[T > n+1]}{\mathbb P[T > n]} \\
+    &=  (y_n - \epsilon(y_n)) \, {\bf 1}_{\set{T \geq n}} \\
+    &= Y_n - \epsilon(y_n) \, {\bf 1}_{\set{T \geq n}}
+\end{align*}
+therefore $(Y_n)_n$ is an antitone strict supermartingale with respect to $(T,\epsilon)$.
+\end{proof}
+
+The assumption that $(\mathcal F_n)_{n \geq 0}$ is the coarsest filtration to which $T$ is adapted is used in the equality between the third and fourth lines here.
+This condition is the main reason that this proof does not extend directly to a completeness result for antitone ranking functions.
+\fi
+
+\subsection{Completeness}
 It seems very likely that this method is complete in the case that every reachable term is AST (Conjecture~\ref{conj:antitone}), but we have been unable to actually prove this.

### 2021-10-16 LICS21/ranking.tex
Various minors (formatting)
Moved all appendices (except those concerning Sec. 6) to main body. Moved references to the end of document.

@@
 
-\begin{restatable}{therm}{ThmMinimal} \label{thm:minimal}
+\begin{therm}
+%\begin{restatable}{therm}{ThmMinimal} 
+\label{thm:minimal}
 Given a closed term $M$, the function $f:Rch(M) \to \mathbb R$ given by 
@@
 if it exists, is the least of all possible ranking functions of $M$.
-\end{restatable}
+%\end{restatable}
+\end{therm}
+
+\begin{proof}
+Let $f$ be the candidate least ranking function defined above, and suppose $g$ is another ranking function such that $f(N) > g(N)$ for some $N \in Rch(M)$.
+The restrictions of $f$ and $g$ to $Rch(N)$ have the same properties assumed of $f$ and $g$, so assume w.l.o.g.~that $N=M$.
+The difference $g - f$ is then a supermartingale (with the same setup as in
+Theorem~\ref{thm:rankable and strict rankable});
+therefore $\mathbb E[g(M_n)] \leq \mathbb E[f(M_n)] + g(M)-f(M)$, for all $n$.
+Now $\mathbb E[f(M_n)] = \sum_{k=n}^\infty \mathbb P[M_k = E[\tY N] \text{ for some }E, N] \to 0 \text{ as } n \to \infty$; 
+therefore as $g(M) - f(M) < 0$, eventually $\mathbb E[g(M_n)] < 0$, which is impossible. 
+It follows that $g \geq f$ as required.
+
+In order for $f$ to be the least ranking function of $M$, it also has to actually be a ranking function itself.
+Each of the conditions on a ranking function is easily verified from the definition of $f$.
+\end{proof}
 
@@
 \begin{figure}[h]
+\begin{center}
 \begin{tikzpicture}[node distance=0.5em, anchor=west]
@@
 \end{tikzpicture}
+\end{center}
 \caption{The reachable terms of $\tY \, \Theta \, \underline 0$\label{fig:geometric distribution}}
@@
 
-\begin{restatable}[Sparse function]{therm}{PartialImpliesRankable} \label{thm:partial implies rankable}
+\begin{therm}
+%\begin{restatable}[Sparse function]{therm}{PartialImpliesRankable} 
+\label{thm:partial implies rankable}
 Every sparse ranking function is a restriction of a ranking function.
-\end{restatable}
+%\end{restatable}
+\end{therm}
 
+%\PartialImpliesRankable*
+\begin{proof}
+Take a closed term $M$ and sparse ranking function $f$ on $M$. Define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
+\begin{align*}
+    f_1(N) & := f(N) \text{ whenever $f(N)$ is defined,} \\
+    f_1(V) & := 0 \text{ for values $V$ not in the domain of $f$.}
+\end{align*}
+
+Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) := \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY\, \lambda x.
+N'],s') \} \right |$.
+The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on sparse ranking functions.
+Define $f_2(N) = \int_\entrosp f_1(\nnext(N,s)) + g(N,s) \, \mu_{\entrosp}(\mathrm d s)$.
+The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction, by the same argument as Theorem~\ref{thm:minimal}).
+\end{proof}
 As a corollary, any term which admits a sparse ranking function terminates almost surely.

### 2021-10-16 LICS21/supermartingales.tex
Latex minors
Moved all appendices (except those concerning Sec. 6) to main body. Moved references to the end of document.

@@
 
-\begin{restatable}{lemma}{rankPast}
-%\begin{lemma}%[{\cite[Lemma 5.5]{DBLP:conf/popl/FioritiH15}}]
-% \cite[Prop 1]{DBLP:conf/popl/ChatterjeeFNH16}}
-\label{lem:rank-PAST}
+%\begin{restatable}{lemma}{rankPast}
+\begin{lemma}\label{lem:rank-PAST}
 Let $(Y_n)_{n \geq 0}$ be a $\epsilon$-ranking supermartingale w.r.t.~the stopping time $T$, then $T < \infty$ a.s., and %$\expect{T} \leq \frac{\expect{Y_0}}{\epsilon}$.
 $\expect{T} \leq {\expect{Y_0}}/{\epsilon}$.
-%\end{lemma}
-\end{restatable}
+\end{lemma}
+%\end{restatable}
+
+\begin{proof}%\rankPast
+We first prove by induction: for all $n \geq 0$
+\[
+%\forall n \geq 0 \, . \, 
+\expect{Y_n} \leq \expect{Y_0} - \epsilon \cdot \big(\textstyle\sum_{i=0}^{n-1} \mathbb{P}[T > i]\big).
+\]
+It follows that $\sum_{i=0}^{\infty} \mathbb{P}[T > i]$ converges; hence we have $\lim_{i \to \infty} \mathbb{P}[T > i] = 0$ and so $\mathbb{P}[T < \infty] = 1$.
+It then remains to observe: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=0} \mathbb{P}[T > i]$.
+\end{proof}
 
@@
 (see e.g.~\citep{AshDD00,DBLP:conf/popl/FioritiH15}).
-\begin{theorem}[Optional Sampling]
+\begin{therm}[Optional Sampling]
 \label{thm:optional sampling}
@@
 \end{enumerate}
-\end{theorem}
+\end{therm}
 
@@
 Our first result is the following theorem.
-\begin{restatable}[Deriving supermartingales]{theorem}{rankableAndStrictRankable}
+\begin{restatable}[Deriving supermartingales]{therm}{rankableAndStrictRankable}
 \label{thm:rankable and strict rankable}
@@
 %\subsection{Proof of \Cref{thm:rankable and strict rankable}.}
-The ranking supermartingale condition is satisfied essentially by a case analysis on the type of redex (Lemma~\ref{lem:key rankable}).
+
+%The ranking supermartingale condition is satisfied essentially by a case analysis on the type of redex (Lemma~\ref{lem:key rankable}).
+
+We shall obtain the theorem as a corollary of a technical lemma (Lemma~\ref{lem:key rankable}).
+
+We say that a given PPCF term is: 
+\emph{type 1} if it has the shape $E[\tY \lambda x.
+N]$, \emph{type 2} if it has the shape $E[\tsample]$; \emph{type 3} if it has the shape $E[R]$ where $R$ is any other redex, and \emph{type 4} if it is a value.
+Henceforth fix an $n \geq 0$, and define $\mathbf{T}_i := \set{s \mid M_n(s) \hbox{ is type $i$}}$.
+It is straightforward to see that each $\mathbf{T}_i \in \calF_n$, and $\set{\mathbf{T}_1, \mathbf{T}_2, \mathbf{T}_3, \mathbf{T}_4}$ is a partition of $\entrosp$.
+Hence it suffices to prove the following lemma.
+
+\begin{restatable}[Technical]{lemma}{TechnicalRankingSupermartingale}
+\label{lem:key rankable}
+For all $i \in \set{1, 2, 3, 4}$ and $A \in \calF_n$
+\[
+\int_A \mu_{\entrosp}(\dif s) \, f(M_{n+1})[s \in \mathbf{T}_i] \leq \int_A \mu_{\entrosp}(\dif s) \, f(M_n)[s \in \mathbf{T}_i]
+\; \footnote{where (the Iverson bracket) $[P] := 1$ if the statement $P$ hold, and 0 otherwise.}
+\] 
+%$\int_A \mu_{\entrosp}(\dif s) \, f(M_{n+1}) \leq \int_A \mu_{\entrosp}(\dif s) \, f(M_n)$, i.e., 
+Hence $\mathbb{E}[f(M_{n+1}) \mid \calF_n] \leq f(M_n)$ a.s.
+\end{restatable}
+
+First, some notation.
+
+Given an $\mathbb N$-indexed sequence (e.g.~$s \in \entrosp$) and $m \geq 0$, we write $s_{\leq m} \in I^m$ to mean the prefix of $s$ of length $m$.
+For $n \geq 0$, define
+\begin{align*}
+%M_n(s) &:= \pi_0 (\red^n(M, s))\\
+\ndraw{n}{s} &:= |\{k < n \mid \exists E: M_k(s) = E[\tsample]\}|
+\end{align*}
+so that $\pi_1 (\red^n(M, s)) = %\overbrace{\pi_t( \cdots (\pi_t}^{\ndraw n s}(s))$. 
+\pi_t( \cdots (\pi_t (s))$, with $\pi_t$ applied ${\ndraw n s}$ times.
+%and the filtration $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
+The $\calF_n$-measurability of $M_n$ (and hence of $\#\mathrm{draw}_n$) follows from \cite{DBLP:conf/icfp/BorgstromLGS16}.
+%$\#\mathrm{draw}_n$ is a stopping time (bounded by $n$). 
+\iffalse
+\akr{$\#\mathrm{draw}_n$ is not a stopping time.
+It doesn't look like you actually use this claim anyway, so it should just be fine to remove, but did you mean something different?} \lo{I agree, and I don't actually use this claim.}
+\fi
+Take $s \in A \in \calF_n$ with $\ndraw{n}{s} = l$.
+For any $s'\in \entrosp$, if $s_{\leq l} = s_{\leq l}'$ then $s' \in A$.
+It follows that ${\set{s_{\leq l}} \cdot I^{\mathbb N}} \subseteq A$.
+\lo{NOTE. To fix notation:
+\(
+\sigma(M_{n}) = \sigma\big(\set{M_{n}^{-1}(\alpha_j, U_{\alpha_j})
+\mid \alpha_j\in \mathsf{Sk}_j, U_{\alpha_j} \in \mathcal{B}(\Real^j), j \geq 0}\big)
+\)
+}
+%Take $M \in \Lambda^0$ (closed PPCF terms). Define, for each $n \in \omega$, the random variable $M_n : \entrosp \to \Lambda^0$ by $M_n(s) := \pi_0 (\red^n(M, s))$.
+
+\begin{proof}
+We show the non-trivial case of $i = 2$.
+First we express 
+\begin{equation}
+f(M_{n+1})[s \in \mathbf{T}_2] = \sum_{i \in \calI} 
+f(E_i[\sigma_i(s)][\underline{\rho(s)}])[s \in U_i]
+\label{eqn:f 2 n+1}
+\end{equation}
+where 
+\begin{itemize}
+\item $\calI$ is a countable indexing set
+\item $E_i[\cdot][\tsample] \in \mathsf{Sk}_{j_i}$, and $\sigma_i : \entrosp \to \Real^{j_i}$, and $\rho(s) := \pi_h(\pi_1(\red^n(M, s))) = \pi_h(\pi_t^{l_i}(s)) \in \Real$ 
+\item $\set{U_i}_{i \in \calI}$ is a partition of $\mathbf{T}_2$, 
+where each $U_i$ is determined by a skeletal environment $E_i$ and a number (of draws) $l_i \leq n$, so that $U_i$ is the set of traces where after $n$ reduction steps, $l_i$ samples have been used, and the term has reached $E_i[r][\tsample]$ for some $r \in \Real^{j_i}$.
+%\(
+%s \in U_i
+%\iff
+%(E_i[\underline{r}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s),
+%\;
+%\hbox{for some }r \in \Real^{j_i}.
+%\)
+(We use the (measurable) function $\sigma_i : \entrosp \to \Real^{j_i}$ to skolemise the existentially quantified $r$.)
+Equivalently
+\[
+U_i := \#\mathrm{draw}_n^{-1}[l_i] \cap M_n^{-1}[\{E_i[r][\tsample] \mid r \in \Real^{j_i}\}] \in \calF_n.
+\]
+%\lo{I.e.~$(E_i[\underline{\sigma_i(s)}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s)$ iff $s \in U_i$.}
+\end{itemize}
+
+Observe that if $s \in U_i$ then $\set{s_{\leq l_i}} \cdot I^{\mathbb N} \subseteq U_i$;
+in fact $(U_i)_{\leq l_i} \cdot I^{\mathbb N} = U_i$.
+%Moreover, for any $A \in \calF_n$, if $s \in A$ then $\set{s_{\leq l}} \cdot I^{\mathbb N} \subseteq A$.
+This means that for any measurable $g : \entrosp \to \Real_{\geq 0}$, if $g(s)$ only depends on the prefix of $s$ of length $(l_i+1)$, then, writing $\widehat{g} : I^{l_i+1} \to \Real_{\geq 0}$ where $g(s) = \widehat{g}(s_{\leq l_i+1})$, we have: for any $A \in \calF_n$
+\begin{equation}
+\int_{A \cap U_i}  \mu_{\entrosp}(\dif s) \, g(s) = 
+\int_{(A \cap U_i)_{\leq l_i+1}} \Leb_{l_i+1}(\dif t) \, \widehat{g}(t)
+\label{eqn:truncate}
+\end{equation}
+%(For simplicity, we will write $g' = g$ in the following.)
+%where $A_{\leq m} := \set{s_{\leq m} \mid s \in A}$.
+
+Take $s \in U_i$, and set $l = l_i$.
+Plainly $\sigma_i(s)$ depends on $s_{\leq l}$, and $\rho(s)$ depends on $s_{\leq l +1}$.
+Take $u \in (U_i)_{\leq l}$.
+It then follows from the definition of ranking function that
+\[%\textstyle
+\int_I \Leb(\dif r) \, f(E_i[\widehat{\sigma_i}(u)][\underline{r}])) \leq f(E_i[\widehat{\sigma_i}(u)][\tsample])).
+\]
+Take $A \in \calF_n$, and integrating both sides, we get
+\begin{align*}
+& \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, \int_I \textrm{Leb}(\dif r) \, f(E_i[\widehat{\sigma_i}(u)][\underline{r}])) \\
+& \leq \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, f(E_i[\widehat{\sigma_i}(u)][\tsample])).
+\end{align*}
+Since $\Leb_{l+1}$ is the (unique) product measure satisfying $\Leb_{l+1}(V \times B) = \Leb_l(V) \cdot \Leb(B)$, and $(U_i)_{\leq l_i} \cdot I^{\mathbb N} = U_i$, we have
+%it follows from the preceding observation that
+\begin{align*}
+& \int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\widehat{\sigma_i}(u')][\underline{\widehat{\rho}(u')}])) \\
+& \leq 
+\int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u') \, f(E_i[\widehat{\sigma_i}(u')][\tsample]))
+\end{align*}
+and so, by (\ref{eqn:truncate})
+\begin{align}
+& \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\sigma_i(s)][\underline{\rho(s)}])) \nonumber \\
+& \leq 
+\int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\sigma_i(s)][\tsample])).
+\label{eqn:a u ui}
+\end{align}
+
+Now, integrating both sides of (\ref{eqn:f 2 n+1}), we have
+\begin{align*}
+& \int_A \mu_{\entrosp}(\dif s) \, f(M_{n+1})[s \in \mathbf{T}_2] \\
+&= 
+\int_A \mu_{\entrosp}(\dif s) \, \sum_{i \in \calI} f(E_i[\sigma_i(s)][\underline{\rho(s)}])[s \in U_i] \\
+&= 
+\sum_{i \in \calI} \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\sigma_i(s)][\underline{\rho(s)}])\\
+&\leq 
+\sum_{i \in \calI} \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\sigma_i(s)][\tsample])
+\quad \hbox{$\because$ (\ref{eqn:a u ui})}\\
+&= 
+\int_{A} \mu_{\entrosp}(\dif s) \, \sum_{i \in \calI} \, f(E_i[\sigma_i(s)][\tsample])[s \in U_i]\\
+&= \int_{A} \mu_{\entrosp}(\dif s) f_{n}(M)[s \in \mathbf{T}_2]
+%\quad \hbox{$\because$ (\ref{eqn:f 2 n+1})}
+\end{align*}
+
+\end{proof}
+
+As an immediate corollary of Lemma~\ref{lem:key rankable}, each $f(M_n)$ is integrable.
+This concludes the proof of Theorem~\ref{thm:rankable and strict rankable}.
 
@@
 T_{-1}(s) & := -1 \nonumber \\
-T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value} \nonumber\\
-& \qquad\qquad\qquad\qquad\qquad \hbox{or of form } E[\tY \lambda x. N] \} \nonumber \\
+T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value, or of form } E[\tY \lambda x. N] \} \nonumber \\
 Y_n(s) & := f(M_{T_n(s)}(s)) \nonumber \\
@@
 
+\newcommand\transform[1]{{\ulcorner{#1}\urcorner}}
+
+%\section{$\tY$ stopping times are finite}
+\iffalse
+As before, fix a $M \in \Lambda^0$.
+Recall the random variables on $(S, \calF, \mu)$:
+\begin{align*}
+T_0(s) & := 0 \\
+T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value or of form } E[\tY \lambda x. N] \}
+\end{align*}
+\fi
+
+We first show that $T_1$ is bounded i.e.~$T_1 \leq n$, for some $n \in \mathbb N$.
+
+A first proof idea is to construct a reduction argument to the strong normalisation of the simply-typed lambda-calculus.
+ There is a classical transform of conditionals into the pure lambda-calculus.
+However, each evaluation of $\tsample$ (almost surely) returns a different number, which complicates such a transform.
+
+Given a closed PPCF term $M$, we transform it to a term $\transform{M}$ of the nondeterministic simply-typed lambda-calculus as follows.
+(We may assume that the nondeterministic calculus is generated from a base type $\iota$ with a $\iota$-type constant symbol $r$, and a function symbol $\bot : A$  for each function type $A$.)
+\iffalse
+\begin{enumerate}
+\item replace every $\tsample$ by $r$
+
+\item replace each base-type subterm $f \, M_1 \cdots M_n$ by $(\lambda x_1 \cdots x_n . r) \, \transform{M_1} \cdots \transform{M_n}$ where $f$ is a primitive function
+
+\item replace every subterm $\tif{B}{M_1}{M_2}$ by
+$(\lambda x . \transform{M_1} + \transform{M_2}) \, \transform{B}$, for a fresh $x$
+
+\item replace every $\tY \, f \, x . N$ by $\lambda x . \transform{N}[\bot / f]$
+\end{enumerate}
+\lo{The above rules can be defined formally.}
+\fi
+\begin{align*}
+\transform{\tsample} &:= r\\
+\transform{y} &:= y \\
+\transform{\lambda y . N} &:= \lambda y . \transform{N}\\
+\transform{M_1 \, M_2} &:= \transform{M_1} \, \transform{M_2} \\
+n \geq 0, \; \transform{f \, M_1 \cdots M_n} &:= (\lambda z_1 \cdots z_n . r) \, \transform{M_1} \cdots \transform{M_n} \\
+\transform{\tif{B}{M_1}{M_2}} &:= \big(\lambda z . (\transform{M_1} + \transform{M_2})\big) \, \transform{B} \\
+\transform{\tY N} &:= (\lambda y . \bot) \transform{N}
+\end{align*}
+In the above, variables $z, z_1, \cdots, z_n$ are assumed to be fresh; $f$ ranges over primitive functions and numerals.
+
+The idea is that $\transform{M}$ captures the initial, non-recursive operational behaviour of $M$, so that $\transform{M}$ simulates the reduction of $M$ until the latter reaches a value, or a term of the form $E[\tY \lambda x.
+N]$.
+
+It is straightforward to see that for every trace $s \in \entrosp$, there is a reduction sequence of $\transform{M}$ that simulates (an initial subsequene of) the reduction of $M$ under $s$.
+
+Finally thanks to the following
+\begin{therm}[\cite{DBLP:conf/lfcs/Groote94}]
+\label{thm:de groote}
+The simply-typed nondeterministic lambda calculus is strongly normalising. \qed
+\end{therm}
+we have:
+
+\begin{enumerate}
+\item Every $\transform{M}$-reduction terminates on reaching $r$, or a term of the shape $E[\bot N_1 \cdots N_n]$.
+
+\item Further there is a finite bound (say $l$) on the length of such $\transform{M}$-reduction sequences; and $l$ bounds the stopping time $T_1$.
+\end{enumerate}
+This concludes the proof of Lemma~\ref{lem:TnBounded}.
 %The random variable $T_M$ is a stopping time w.r.t.~filtration $(\calF_n)_{n \geq 0}$. 
@@
 
-\begin{restatable}{lemma}{TMtYStoppingTime}
+\begin{lemma}
+%\begin{restatable}{lemma}{TMtYStoppingTime}
 \label{lem:TMtY is a stopping time}
 $T_M^\tY$ is a stopping time adapted to $(\calF_{T_n})_{n \geq 0}$.
-\end{restatable}
+%\end{restatable}
+\end{lemma}
 
-\begin{theorem}[Soundness of rankability] \label{thm:rankable implies termination}
+\begin{proof}
+%For each $n \in \omega$, $\set{T_M^\tY = n} \in \calF_{T_n}$.
+To see $\set{T_M^\tY = n} \in \calF_{T_n}$ for a given $n$, we need to show that $\set{T_M^\tY = n} \cap \set{T_n \leq i} \in \calF_i$, for all $i \in \mathbb N$.
+Let $\mathcal V$ be the set of values in $\Lambda^0$ (which is measurable), and let $\mathcal W$ be the non-values.
+Then, it suffices to observe that $\set{T_M^\tY = n} \cap \set{T_n \leq i}
+= \bigcup_{l=n}^i 
+\big( M_l^{-1} [{\mathcal V}] \cap  
+M_{l-1}^{-1} [{\mathcal W}] \cap 
+\set{T_n = l} \big)
+$, where each of $M_l^{-1} [{\mathcal V}]$, $M_{l-1}^{-1} [{\mathcal W}]$, and $\set{T_n = l}$ is in $\calF_i$.
+\end{proof}
+
+\begin{therm}[Soundness of rankability] \label{thm:rankable implies termination}
 \begin{enumerate}

### 2021-10-16 LICS21/SPCF.tex
for important technical words, changed \emph to \defn moved discussion of sum of powers from appendix to main body other minors

@@
 \changed[lo]{The maps $\pi_h:\entrosp \to I$ (projecting to the first element) and $\pi_t:\entrosp \to \entrosp$ (popping the first element) are then measurable.}
-Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$ the \emph{entropy space}, and elements of $\entrosp$ \emph{traces}.
+Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$ the \defn{entropy space}, and elements of $\entrosp$ \defn{traces}.
 
-Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $\mathit{Sk}(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$.
+Define a \defn{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $\mathit{Sk}(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$.
 Following \cite{DBLP:conf/icfp/BorgstromLGS16,DBLP:conf/lics/StatonYWHK16,DBLP:journals/pacmpl/EhrhardPT18}, the $\sigma$-algebra on $\Lambda$ is defined by identifying $\Lambda$ with $\bigcup_{m\geq 0} \big(\mathsf{Sk}_m \times \Real^m \big)$, where $\mathsf{Sk}_m$ is the set of skeletons containing $m$ occurrences of $\skeletonPlaceholder$.
@@
 The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid \hbox{$M$ terminates at $s$}\}$ is measurable \citep{DBLP:conf/icfp/BorgstromLGS16,DBLP:conf/esop/MakOPW21}.
-We say that a term $M$ is \emph{almost surely terminating} (\emph{AST}) just if $\mu_{\entrosp}(\{s \mid \hbox{$M$ terminates at $s$}\}) = 1$.
+We say that a term $M$ is \defn{almost surely terminating} (\emph{AST}) just if $\mu_{\entrosp}(\{s \mid \hbox{$M$ terminates at $s$}\}) = 1$.
 

### 2021-10-16 LICS21/antitone.tex
for important technical words, changed \emph to \defn moved discussion of sum of powers from appendix to main body other minors

@@
 
-\begin{definition}
-The definition of \emph{antitone ranking function} $f$ for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
+\begin{definition}\rm
+The definition of \defn{antitone ranking function} $f$ for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
 we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) function $\epsilon : \nnReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
@@
 
-\begin{definition}
+\begin{definition}\rm
 Given a probability space $(\Omega, \calF, \mathbb{P})$, and a supermartingale $(Y_n)_{n \geq 0}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \geq 0}$,
-we say that $(Y_n)_{n \geq 0}$ is an \emph{antitone strict supermartingale w.r.t.~$T$} if for all $n \geq 0$, we have $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
+we say that $(Y_n)_{n \geq 0}$ is an \defn{antitone strict supermartingale w.r.t.~$T$} if for all $n \geq 0$, we have $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
 \(
@@
 
-\begin{definition}
-Define an \emph{antitone sparse ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$:
-\begin{inparaenum}[(i)]
+\begin{definition}\rm
+Define an \defn{antitone sparse ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$:
+\begin{asparaenum}[(i)]
 \item $f(N) \geq 0$ for all $N \in \dom(f)$,
@@
 \item for any $N \in \dom(f)$, evaluation of $N$ will eventually reach some $O$ which is either a value or in $\dom(f)$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of $\dom(f)$).
-\end{inparaenum}
+\end{asparaenum}
 \end{definition}
 
-\begin{therm}
+\begin{therm}[Sparse function]
 %\begin{restatable}[Sparse function]{therm}{AntitonePartial}  
@@
 (Note that whether a program is affine recursive cannot be checked by just counting textual occurrences of variables.)
-Termination analysis of \emph{non-affine recursive} probabilistic programs does not seem to have received much attention.
+Termination analysis of \emph{non-affine recursive} probabilistic programs has not received much attention (with the exception of \citep{BeutnerO21}).
 Methods such as those presented in \cite{DBLP:journals/toplas/LagoG19} are explicitly restricted to affine programs, and are unsound otherwise.
@@
 \[
-\begin{array}{l}
-\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} . \\
-\quad\tif{n \leq 0}{s}{
+%\begin{array}{l}
+\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} \, . \, 
+\tif{n \leq 0}{s}{
 f \, n \, (\varphi \, f \, s \, (n - 1))
@@
 }
-\end{array}
+%\end{array}
 \]

### 2021-10-16 LICS21/confluent-proofs.tex
for important technical words, changed \emph to \defn moved discussion of sum of powers from appendix to main body other minors

@@
 
-\paragraph{}
+%\paragraph{}
 Although the six cases in the definition of $\sim_c$ are defined separately, there are some commonalities worth noting.
@@
 
-\paragraph{}
+%\paragraph{}
 If $Y \sim_c X \sim_c^* X_{cbv}$, and the reductions involved in $Y \sim_c X$ are not the last pair of reductions in $Y$ that aren't in CBV order, it will be shown that the order of the $\sim_c$s can be rearranged so that the sequence $Y \sim_c^* X_{cbv}$ is the canonical sequence given above, therefore $Y_{cbv} = X_{cbv}$.
@@
 
-\paragraph{}
+%\paragraph{}
 If $Y$ is in CBV order and $Y \sim_c^* X$, then $Y_{cbv} = X_{cbv}$ but also $Y_{cbv} = Y$ because the sequence $(Y_i)_i$ terminates immediately therefore $X_{cbv}$ is the unique reduction sequence in call-by-value order that's related to $X$ by $\sim_c^*$.
@@
 
-\paragraph{}
+%\paragraph{}
 In the case that $(M,s) \Rightarrow^* (M_1,s_1)$ or $(M,s) \Rightarrow^* (M_2,s_2)$ by multiple steps, it is possible to repeatedly replace a pair of the form $A \Leftarrow B \Rightarrow C$ by $A \Rightarrow^* D \Leftarrow^* C$ by the construction above, but it is not immediate that this process terminates, because each $\Rightarrow$ or $\Leftarrow$ may be replaced by multiple, so the sequence of $\Rightarrow$s and $\Leftarrow$s from $(M_1,s_1)$ to $(M_2,s_2)$ may get longer at some steps.
@@
 
-  \paragraph{}
+  %\paragraph{}
   There is therefore a reduction sequence $(M,s) = (N_0,s_0') \Rightarrow \dots \Rightarrow (N_m, s_m')$ which ends in a value ($N_m = M_n$) and is in CBV order.
@@
 
+\iffalse
 Example~\ref{ex:sum of powers}
@@
 It wouldn't even be possible in this case to give a sparser version of the ranking function defined only at $M$ and $P[\underline n]$, because the amount that an antitone sparse ranking function must decrease at each $\tY$-reduction step depends on the value of the ranking function at the next term where it is defined, therefore it is necessary to have these intermediate steps in order for the ranking function to be able to change sufficiently slowly.
+\fi

### 2021-10-16 LICS21/confluent.tex
Various minors (formatting)
for important technical words, changed \emph to \defn moved discussion of sum of powers from appendix to main body other minors

@@
 
-A \emph{position} is a finite sequence of steps into a term, defined inductively as
+A \defn{position} is a finite sequence of steps into a term, defined inductively as
+\iffalse
 \begin{align*}
@@
 \end{align*}
-The \emph{subterm of $M$ at $\alpha$}, denoted $M \mid \alpha$, is defined as
+\fi
+\[
+\alpha ::= {\cdot} \mid \lambda ; \alpha \mid @_1 ; \alpha \mid @_2 ; \alpha \mid \underline f_i ; \alpha \mid \tY ; \alpha \mid \textsf{if}_1 ; \alpha \mid \textsf{if}_2 ; \alpha \mid \textsf{if}_3 ; \alpha.
+\]
+The \defn{subterm of $M$ at $\alpha$}, denoted $M \mid \alpha$, is defined as
 \begin{align*}
@@
 so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda$ is undefined). 
-A position such that $M\mid \alpha$ does exist is said to \emph{occur} in $M$. 
+A position such that $M\mid \alpha$ does exist is said to \defn{occur} in $M$. 
 %\changed[lo]{
-\emph{Substitution} of $N$ at position $\alpha$ in $M$, written $M[N/\alpha]$, is defined similarly.
+\defn{Substitution} of $N$ at position $\alpha$ in $M$, written $M[N/\alpha]$, is defined similarly.
 For example, let 
@@
 If $\alpha_1$ is a prefix of $\alpha_2$ (written as $\alpha_1 \leq \alpha_2$), then $N_2$ is also a subterm of $N_1$.
-If neither $\alpha_1 \leq \alpha_2$ nor $\alpha_1 \geq \alpha_2$, the positions are said to be \emph{disjoint}.
+If neither $\alpha_1 \leq \alpha_2$ nor $\alpha_1 \geq \alpha_2$, the positions are said to be \defn{disjoint}.
 The notion of disjointness is mostly relevant in that if $\alpha_1$ and $\alpha_2$ are disjoint, performing a substitution at $\alpha_1$ will leave the subterm at $\alpha_2$ unaffected.
@@
 
-\begin{definition}
+\begin{definition}\rm
 \label{def:more general red}
@@
 \end{align*}
-In each of these cases, $M \mid \alpha$ is the \emph{redex}, and the reduction \emph{takes place at $\alpha$}.
+In each of these cases, $M \mid \alpha$ is the \defn{redex}, and the reduction \emph{takes place at $\alpha$}.
 Each subterm can be a redex in at most one way, but there can be multiple redexes at different positions.
@@
 
-\paragraph*{Reduction relation on skeletons} We are thus led to consider the reduction relation on skeletons (and positions in a skeleton), which can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, $\tsample$ reducing to $\skeletonPlaceholder$, and $\skeletonPlaceholder$ considered as (the skeletal equivalent of) a value, so that $(\lambda x.A) \, \skeletonPlaceholder$ reduces to $A[\skeletonPlaceholder / x]$.
+\subsection{Reduction relation on skeletons} We are thus led to consider the reduction relation on skeletons (and positions in a skeleton), which can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, $\tsample$ reducing to $\skeletonPlaceholder$, and $\skeletonPlaceholder$ considered as (the skeletal equivalent of) a value, so that $(\lambda x.A) \, \skeletonPlaceholder$ reduces to $A[\skeletonPlaceholder / x]$.
 For example, we have 
@@
 
-\begin{definition} \label{defn:sim}
+\begin{definition}\rm \label{defn:sim}
 The relation $\sim$ is defined as the union of the minimal symmetric relations $\sim_p$ (``$p$'' for parent-child) and $\sim_c$ (``$c$'' for cousin) satisfying
@@
 
-\begin{definition}[$\Rightarrow$ reduction]
+\begin{definition}[$\Rightarrow$ reduction]\rm
 Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function, so it is denoted ``$\Rightarrow$'' instead of ``$\red$'', and it relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
@@
 
-\paragraph*{Reduction strategy determinising $\Rightarrow$}
+\subsection{Reduction strategy determinising $\Rightarrow$}
 The reduction relation $\Rightarrow$ is nondeterministic, so it admits multiple possible reduction strategies. 
 %\begin{definition}
-A \emph{reduction strategy} starting from a closed term $M$ is a measurable partial function $f$ from $\mathit{Rch}(M)$ to positions, such that for any reachable term $N$ where $f$ is defined, $f(N)$ is a position of a redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
+A \defn{reduction strategy} starting from a closed term $M$ is a measurable partial function $f$ from $\mathit{Rch}(M)$ to positions, such that for any reachable term $N$ where $f$ is defined, $f(N)$ is a position of a redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
 Using a reduction strategy $f$, a subset of $\Rightarrow$ that isn't nondeterministic, $\Rightarrow_f$, can be defined by $(N,s) \Rightarrow_f (N',s')$ just if $(N,s) \Rightarrow (N',s')$ and $N$ reduces to $N'$ with the redex at $f(N)$.
@@
 A closed term $M$ terminates with a given reduction strategy $f$ and samples $s$ if there is some natural number $n$ such that $(M,s) \Rightarrow_f^n (N,s')$ where $f$ gives no reduction at $N$. 
-The term \emph{terminates almost surely w.r.t.~$f$} if it terminates with $f$ for almost all $s$.
+The term \defn{terminates almost surely w.r.t.~$f$} if it terminates with $f$ for almost all $s$.
 
@@
 More explicitly, the modified forms of the theorems are:
-\begin{definition}
-An \emph{antitone ranking function on $M$ w.r.t.~a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ such that
+\begin{definition}\rm
+An \defn{antitone ranking function on $M$ w.r.t.~a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ such that
 \begin{itemize}
@@
 \end{itemize}
-Any closed term for which an antitone ranking function with respect to a reduction strategy $r$ exists is called \emph{antitone rankable w.r.t.~$r$}. 
+Any closed term for which an antitone ranking function with respect to a reduction strategy $r$ exists is called \defn{antitone rankable w.r.t.~$r$}. 
 \end{definition}
 
-\begin{definition}
-An \emph{antitone sparse ranking function on $M$ w.r.t.~a reduction strategy $r$} is a partial function $f : \mathit{Rch}_r(M) \rightharpoonup \mathbb R$ such that there exists some antitone function $\epsilon : \nnReal \to \pReal$ such that
+\begin{definition}\rm
+An \defn{antitone sparse ranking function on $M$ w.r.t.~a reduction strategy $r$} is a partial function $f : \mathit{Rch}_r(M) \rightharpoonup \mathbb R$ such that there exists some antitone function $\epsilon : \nnReal \to \pReal$ such that
 \begin{inparaenum}[(i)]
@@
 
-If the antitone function is just the constant function $1$ instead, then it is called a \emph{(sparse) ranking function on $M$ w.r.t.~$r$} instead, and the term is called \emph{rankable w.r.t.~$r$}.
+If the antitone function is just the constant function $1$ instead, then it is called a \defn{(sparse) ranking function on $M$ w.r.t.~$r$} instead, and the term is called \defn{rankable w.r.t.~$r$}.
 
@@
 
-\paragraph*{Custom reduction strategy for AST analysis} However, there are some terms which terminate more simply and directly via some alternative reduction strategy.
+\subsection{Custom reduction strategy for AST analysis} However, there are some terms which terminate more simply and directly via some alternative reduction strategy.
 For these examples, a custom reduction strategy is specified for each term under consideration.
@@
 
+Let $r$ be the reduction strategy which reduces $M$ in the order
+\begin{align*}
+&\ M \\
+\to^* &\ P[\underline n] \\
+\to^* &\ (\lambda p.\ \Xi[p]\ \underline n)\ (\lambda x. \underbrace{x \times \dots \times x \times}_n \underline 1) \\
+\to^{\phantom *} &\ \Xi[\lambda x. \underbrace{x \times \dots \times x \times}_n \underline 1]\ \underline n \\
+\to^* &\ \underline{n^n} + \underline{(n-1)^n} + \dots + \underline 0 \\
+\to^* &\ \underline{\sum_{k=0}^n k^n}
+\end{align*}
+(this is a little underspecified, but the details are not important). A suitable sparse ranking function for it is
+\begin{align*}
+M \mapsto & \frac{\pi^2}{3} \\
+P[\underline n] \mapsto & 2n+2 \\
+(\lambda p.\ \Xi[p]\ \underline n)\ (\lambda x. \underbrace{x \times \dots \times x \times}_{n-k} (\Theta[x]\ \underline k)) \mapsto & n + k + 2 \\
+\Xi[\lambda x. \underbrace{x \times \dots \times x \times}_n \underline 1]\ \underline n \mapsto & n + 1 \\
+\underline{n^n} + \dots + \underline{(n-k+1)^n} + \Xi[\lambda x. x \times \dots \underline 1]\ \underline k \mapsto & k + 1 \\
+\underline{\sum_{k=0}^n k^n} \mapsto & 0.
+\end{align*}
+
+The first step is justified because the probability of reaching $P[\underline n]$ for any specific $n \geq 0$ is $(n+1)^{-2} - (n+2)^{-2} = \frac{2n + 3}{(n+1)^2(n+2)^2}$, therefore the expected next value of the ranking function is $\sum_{n=0}^{\infty} \frac{(2n+2)(2n + 3)}{(n+1)^2(n+2)^2} = \frac{\pi^2}{3}$.
+The other steps are deterministic, and involve 1 or 0 $\tY$-reductions each.
+Given that all of the reduction steps after the first are deterministic, and the number of $\tY$ reduction steps is not that hard to count, defining the sparse ranking function's values only at $M$ and $P[\underline n]$ would also have been reasonable, although in a similar term which had more random samples throughout, that would not have been so simple.
+
+Providing a ranking function of some sort for this term in a similar level of detail would have been rather more complicated using only the standard reduction strategy.
+The recursion in $\Theta$ would have to be evaluated separately for every time it was used, so there would have been more (and more complex) terms to consider in the reduction sequence.
+Also, the number of $\tY$-reductions from $P[\underline n]$ would have been $n^2 + 2n + 1$ instead of $2n + 2$, therefore the expected number of $\tY$ reductions starting from $M$ would be infinite, i.e.~$M$ is not $\tY$-PAST, therefore it is not even rankable.
+It would be possible to define an antitone sparse ranking function for it, but it is much more difficult to construct:
+\begin{align*}
+M & \mapsto 1.47 \\
+P[\underline n] & \mapsto 1 + \ln((n+1)n + 1) \\
+\Xi[\lambda x. \Theta[x] \, \underline n] \, \underline n & \mapsto 1 + \ln((n+1)n + 1) \\
+\left.\begin{array}{r}
+\underline{n^n} + \dots + \underline{k_1^n} + {}\\
+\Xi[\lambda x. \Theta[x] \underline n] (\underline{k_1} - 1) 
+\end{array}\right\} 
+& \mapsto 1 + \ln((n+1)(k_1-1)+1) \\
+\left.\begin{array}{r}
+\underline{n^n} + \dots + \underline{(k_1+1)^n} +  {}\\
+\Theta[\underline{k_1}] \, \underline{n} +  {}\\
+\Xi[\lambda x. \Theta[x] \, \underline n] (\underline{k_1} - 1) 
+\end{array}\right\}
+& \mapsto 1 + \ln((n+1)k_1 + 1) \\
+\left.\begin{array}{r}\underline{n^n} + \dots + \underline{(k_1+1)^n} + {} \\
+\underbrace{\underline{k_1} \times \dots \times \underline{k_1} \times}_{n-k_2+1} {} \\
+\Theta[\underline{k_1}] (\underline{k_2} - 1) + {} \\
+\Xi[\lambda x. \Theta[x] \, \underline n] (\underline{k_1} - 1)
+\end{array}\right\} & \mapsto 1 + \ln((n+1)k_1 + k_2 + 1) \\
+\underline{\textstyle\sum_{k=0}^n k^n} & \mapsto 0.
+\end{align*}
+
+It wouldn't even be possible in this case to give a sparser version of the ranking function defined only at $M$ and $P[\underline n]$, because the amount that an antitone sparse ranking function must decrease at each $\tY$-reduction step depends on the value of the ranking function at the next term where it is defined, therefore it is necessary to have these intermediate steps in order for the ranking function to be able to change sufficiently slowly.
 With the CBV reduction strategy, the recursion in $\Theta[x]$ (that defines the function $k \mapsto k^x$) is executed separately for every term in the sum in $\Xi$.
@@
 It is still antitone rankable, but this is much more complicated.
-Details of the ranking function construction with and without using the confluent semantics are given in Appendix~\ref{apx:confluent}.
+%Details of the ranking function construction with and without using the confluent semantics are given in Appendix~\ref{apx:confluent}.
 \end{example}

### 2021-10-16 LICS21/intro.tex
for important technical words, changed \emph to \defn moved discussion of sum of powers from appendix to main body other minors

@@
 Deciding AST of first-order imperative programs with discrete probabilities is $\Pi^0_2$-complete \citep{DBLP:conf/mfcs/KaminskiK15}.
-In recent work \citep{BeutnerO21} %(under double-blind review elsewhere), 
+In recent work by \cite{BeutnerO21} %(under double-blind review elsewhere), 
 the AST problem for higher-order programs with continuous distributions (and suitable primitive functions) is shown to have the same complexity.
-Though the problem of proving AST of imperative probabilistic programs is much studied \citet{DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05,DBLP:conf/popl/FioritiH15,DBLP:conf/pldi/ChenH20,DBLP:journals/toplas/ChatterjeeFNH18}, to our knowledge the problem of proving AST of higher-order programs with continuous distributions is new.
+Though the problem of proving AST of imperative probabilistic programs is much studied \citep{DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05,DBLP:conf/popl/FioritiH15,DBLP:conf/pldi/ChenH20,DBLP:journals/toplas/ChatterjeeFNH18}, to our knowledge the problem of proving AST of higher-order programs with continuous distributions is new.
 
@@
 In other words, the program's behavior is used to define an associated (ranking) \emph{supermartingale}.
-Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \citep{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18}.
+Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs: see, for example, \citep{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18}.
 This paper's contribution is to extend this method to a higher-order setting.
@@
 Sparse ranking functions can be defined only for those points in the execution of a program which are semantically important, while all of the other intermediate steps can be ignored. 
-Yet, they are no less efficacious for proving AST (Theorem~\ref{thm:partial implies rankable}): every sparse ranking function is a restriction of a ranking function.
+Yet, they are no less efficacious for proving AST: we prove a \emph{sparse function theorem} (Theorem~\ref{thm:partial implies rankable}): 
+every sparse ranking function is a restriction of a ranking function. 
 This also makes the ranking function method of proving AST more compatible with syntactic sugar, because the intermediate reduction steps implicit in the simplified notation can be ignored.
@@
 
-\paragraph*{Outline}
+\subsection*{Outline}
 
@@
 
-\paragraph*{Additional materials} Further details of some of the examples and all missing proofs can be found in the appendix.
+\paragraph*{Appendix} All missing proofs of theorems in Section~\ref{sec:confluent} and further details can be found in the appendix.

### 2021-10-16 LICS21/ranking.tex
for important technical words, changed \emph to \defn moved discussion of sum of powers from appendix to main body other minors

@@
 In both of these cases, there are only some values of the ranking function that are semantically important. 
-\begin{definition}
-Define a \emph{sparse ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \rightharpoonup \mathbb R$ such that
+\begin{definition}\rm
+Define a \defn{sparse ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \rightharpoonup \mathbb R$ such that
 \begin{inparaenum}[(i)]
@@
 
-\begin{therm}
+\begin{therm}[Sparse function]
 %\begin{restatable}[Sparse function]{therm}{PartialImpliesRankable} 

### 2021-10-16 LICS21/related.tex
for important technical words, changed \emph to \defn moved discussion of sum of powers from appendix to main body other minors

@@
 
-\paragraph*{Further directions} We have not yet been able to prove Conjecture~\ref{conj:antitone} which would imply that Theorem~\ref{thm:antitone rankable implies termination} is almost complete.
+\subsection*{Further directions} We have not yet been able to prove Conjecture~\ref{conj:antitone} which would imply that Theorem~\ref{thm:antitone rankable implies termination} is almost complete.
 \begin{conjecture}[Completeness]
@@
 
-\paragraph*{Conclusions}
+\subsection*{Conclusions}
 

### 2021-10-16 LICS21/supermartingales.tex
for important technical words, changed \emph to \defn moved discussion of sum of powers from appendix to main body other minors

@@
 These variants are defined as functions from reachable terms (i.e.~possible states of the program's execution) to real numbers. 
-Specifically, let the set of reachable terms from a given closed starting term $M$ be $Rch(M) := \{N \in \Lambda \mid M \to^* N \}$, with the $\sigma$-algebra induced as a subset of $\Lambda$.
+Specifically, let the set of reachable terms from a given closed starting term $M$ be 
+\[
+\mathit{Rch}(M) := \{N \in \Lambda \mid M \to^* N \}
+\] 
+with the $\sigma$-algebra induced as a subset of $\Lambda$.
 
@@
 \label{def:ranking function}
-A \emph{ranking function on $M \in \Lambda^0$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$,\footnote{The fact that the ranking function can be 0 in some cases before it reaches a value is necessary to get Theorem~\ref{thm:minimal} to work neatly.} and
+A \defn{ranking function on $M \in \Lambda^0$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$,\footnote{The fact that the ranking function can be 0 in some cases before it reaches a value is necessary to get Theorem~\ref{thm:minimal} to work neatly.} and
 \begin{enumerate}
@@
 Let $T$ be a r.v.~that takes values in $\mathbb{N} \cup \set{\infty}$.
-We call $T$ a \emph{stopping time adapted to $(\calF_n)_{n \geq 0}$} just if $\set{T = n} \in \calF_n$, for all $n$.
+We call $T$ a \defn{stopping time adapted to $(\calF_n)_{n \geq 0}$} just if $\set{T = n} \in \calF_n$, for all $n$.
 
@@
 \begin{asparaenum}[(i)]
-\item A sequence of r.v.s $(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \emph{supermartingale} if for all $n \geq 0$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~(i.e.~for all $A \in \calF_n$, $\int_A Y_{n+1}(\omega) \, \mathbb{P}(\dif \omega) \leq \int_A Y_n(\omega) \, \mathbb{P}(\dif \omega)$).
+\item A sequence of r.v.s $(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \defn{supermartingale} if for all $n \geq 0$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~(i.e.~for all $A \in \calF_n$, $\int_A Y_{n+1}(\omega) \, \mathbb{P}(\dif \omega) \leq \int_A Y_n(\omega) \, \mathbb{P}(\dif \omega)$).
 \item Let $\epsilon > 0$. 
 Given a stopping time $T$ and a supermartingale $(Y_n)_{n \geq 0}$, both adapted to filtration $(\calF_n)_{n \geq 0}$, 
-we say that $(Y_n)_{n \geq 0}$ is a $\epsilon$-\emph{ranking supermartingale w.r.t.~$T$} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.\footnote{For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable defined by ${\bf 1}_A(\omega) := 1$ if $\omega \in A$, and $0$ otherwise.} %\lo{\LaTeX\ does not allow the preceding bracketed sentence to be typeset as a footnote.
+we say that $(Y_n)_{n \geq 0}$ is a $\epsilon$-\defn{ranking supermartingale w.r.t.~$T$} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.\footnote{For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable defined by ${\bf 1}_A(\omega) := 1$ if $\omega \in A$, and $0$ otherwise.} %\lo{\LaTeX\ does not allow the preceding bracketed sentence to be typeset as a footnote.
 Strange!}\lo{Now it does.
 Stranger.}
-A \emph{ranking supermartingale w.r.t.~$T$} is a $\epsilon$-ranking supermartingale w.r.t.~$T$ for some $\epsilon > 0$.
+A \defn{ranking supermartingale w.r.t.~$T$} is a $\epsilon$-ranking supermartingale w.r.t.~$T$ for some $\epsilon > 0$.
 \end{asparaenum}
@@
 
-\begin{remark}
+%\begin{remark}
 Our notion of ranking supermartingale is a slight generalisation of the original definition \citep{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, which does not involve an arbitrary stopping time.
 %: Fioriti and Hermanns call a supermartingale $(Y_n)_{n \geq 0}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot {\bf 1}_{\set{Y_n > 0}}$.
-\end{remark}
+%\end{remark}
 
@@
 
-\begin{remark}
+%\begin{remark}
 Lemma~\ref{lem:rank-PAST} is a slight extension of \cite[Lem.~5.5]{DBLP:conf/popl/FioritiH15},
 which asserts the same result but for the specific stopping time $T : \omega \mapsto \min\set{n \mid Y_n(\omega) = 0}$.
-\end{remark}
+%\end{remark}
 
@@
 and the filtration $(\mathcal{F}_n)_{n \geq 0}$ where $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
-Thus $T_M$ is the \emph{runtime} of $M$ (and $M$ is AST iff $\mu_{\entrosp}(T_M < \infty) = 1$); we say that $M$ is \emph{positively almost surely terminating} (\emph{PAST}) if $\expect{T_M} < \infty$.
+Thus $T_M$ is the \defn{runtime} of $M$ (and $M$ is AST iff $\mu_{\entrosp}(T_M < \infty) = 1$); we say that $M$ is \defn{positively almost surely terminating} (\defn{PAST}) if $\expect{T_M} < \infty$.
 
@@
 %and the filtration $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
-The $\calF_n$-measurability of $M_n$ (and hence of $\#\mathrm{draw}_n$) follows from \cite{DBLP:conf/icfp/BorgstromLGS16}.
+The $\calF_n$-measurability of $M_n$ (and hence of $\#\mathrm{draw}_n$) follows from \citep{DBLP:conf/icfp/BorgstromLGS16}.
 %$\#\mathrm{draw}_n$ is a stopping time (bounded by $n$). 
 \iffalse
-\akr{$\#\mathrm{draw}_n$ is not a stopping time.
-It doesn't look like you actually use this claim anyway, so it should just be fine to remove, but did you mean something different?} \lo{I agree, and I don't actually use this claim.}
+\akr{$\#\mathrm{draw}_n$ is not a stopping time. It doesn't look like you actually use this claim anyway, so it should just be fine to remove, but did you mean something different?} 
+\lo{I agree, and I don't actually use this claim.}
 \fi
@@
 Take $A \in \calF_n$, and integrating both sides, we get
-\begin{align*}
-& \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, \int_I \textrm{Leb}(\dif r) \, f(E_i[\widehat{\sigma_i}(u)][\underline{r}])) \\
-& \leq \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, f(E_i[\widehat{\sigma_i}(u)][\tsample])).
-\end{align*}
+\[
+%\begin{align*}
+\int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, \int_I \textrm{Leb}(\dif r) \, f(E_i[\widehat{\sigma_i}(u)][\underline{r}])) 
+\leq \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, f(E_i[\widehat{\sigma_i}(u)][\tsample])).
+%\end{align*}
+\]
 Since $\Leb_{l+1}$ is the (unique) product measure satisfying $\Leb_{l+1}(V \times B) = \Leb_l(V) \cdot \Leb(B)$, and $(U_i)_{\leq l_i} \cdot I^{\mathbb N} = U_i$, we have
 %it follows from the preceding observation that
-\begin{align*}
-& \int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\widehat{\sigma_i}(u')][\underline{\widehat{\rho}(u')}])) \\
-& \leq 
+\[
+%\begin{align*}
+\int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\widehat{\sigma_i}(u')][\underline{\widehat{\rho}(u')}])) 
+\leq 
 \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u') \, f(E_i[\widehat{\sigma_i}(u')][\tsample]))
-\end{align*}
+%\end{align*}
+\]
 and so, by (\ref{eqn:truncate})
-\begin{align}
-& \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\sigma_i(s)][\underline{\rho(s)}])) \nonumber \\
-& \leq 
+\[
+%\begin{align}
+\int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\sigma_i(s)][\underline{\rho(s)}])) \nonumber \leq 
 \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\sigma_i(s)][\tsample])).
 \label{eqn:a u ui}
-\end{align}
-
+%\end{align}
+\]
 Now, integrating both sides of (\ref{eqn:f 2 n+1}), we have
 \begin{align*}
-& \int_A \mu_{\entrosp}(\dif s) \, f(M_{n+1})[s \in \mathbf{T}_2] \\
+\int_A \mu_{\entrosp}(\dif s) \, f(M_{n+1})[s \in \mathbf{T}_2] 
 &= 
@@
 We first state a useful property about r.v.s $T_0, T_1, T_2, \dots$.
-\begin{restatable}{lemma}{TnBounded}
+\begin{lemma}
+%\begin{restatable}{lemma}{TnBounded}
 \label{lem:TnBounded}
 $(T_n)_{n \geq 0}$ is an increasing sequence of stopping times adapted to $(\calF_n)_{n \geq 0}$, and each $T_i$ is bounded.
-\end{restatable}
+%\end{restatable}
+\end{lemma}
 
@@
 
+\begin{proof}
 We first show that $T_1$ is bounded i.e.~$T_1 \leq n$, for some $n \in \mathbb N$.
@@
 
-Finally thanks to the following
+Finally since the simply-typed nondeterministic lambda calculus is strongly normalising \citep{DBLP:conf/lfcs/Groote94}
+\iffalse
+thanks to the following
 \begin{therm}[\cite{DBLP:conf/lfcs/Groote94}]
@@
 \end{therm}
-we have:
-
+\fi
+it remains to observe:
 \begin{enumerate}
-\item Every $\transform{M}$-reduction terminates on reaching $r$, or a term of the shape $E[\bot N_1 \cdots N_n]$.
+\item Every $\transform{M}$-reduction terminates on reaching $r$, or a term of the shape $E[\bot \, N_1 \cdots N_n]$.
 
@@
 \end{enumerate}
-This concludes the proof of Lemma~\ref{lem:TnBounded}.
+%This concludes the proof of Lemma~\ref{lem:TnBounded}.
 %The random variable $T_M$ is a stopping time w.r.t.~filtration $(\calF_n)_{n \geq 0}$. 
 %Since $f(M_n(s)) = 0$ iff $M_n(s)$ is a value, $T_M$ is the \emph{runtime of $M$}.
+\end{proof}
 
@@
 Note that, as the type system ensures that the reduction relation excluding $\tY$-reduction is strongly normalising, only finitely many reductions can occur in a row without one of them being a $\tY$-reduction, therefore $T_M^\tY < \infty$ a.s.~iff $M$ is AST.
-We say that $M$ is $\tY$-\emph{PAST} if $\expect{T^{\tY}_M} < \infty$.
+We say that $M$ is $\tY$-\defn{PAST} if $\expect{T^{\tY}_M} < \infty$.
 
@@
 %It follows immediately from \Cref{lem:rank-PAST} that $\expect{T_M} < \infty$.
+
 Statement (ii) follows at once from Theorem~\ref{thm:rankable and strict rankable} and Lemma~\ref{lem:rank-PAST}.
@@
 
-Thus the method of (strict) ranking function is sound for proving (positive) a.s.~termination of PPCF programs.
+Thus the method of (strict) ranking function is sound for proving (positive) almost sure termination of PPCF programs.
 It is in fact also complete in a sense: if $\expect{T^\tY_N} < \infty$ for all $N \in \mathit{Rch}(M)$ then $M$ is rankable (Theorem~\ref{thm:minimal}).

### 2021-10-16 LICS21/antitone.tex
new subsections and subsection titles

@@
 
+\subsection{Antitone strict supermartingales}
+
 \begin{definition}\rm
@@
 
-\emph{Claim}: For all $0 < x \leq 1$, 
+\noindent\textbf{Claim}:\emph{For all $0 < x \leq 1$, 
 \(
@@
 \)
-where $B_x = \left \lceil \dfrac{\expect{Y_0}+1}{x \; \epsilon(\expect{Y_0} \, x^{-1})} \right \rceil$ and $\epsilon : \nnReal \to \pReal$ is the antitone function.
+where $B_x = \left \lceil \dfrac{\expect{Y_0}+1}{x \; \epsilon(\expect{Y_0} \, x^{-1})} \right \rceil$ and $\epsilon : \nnReal \to \pReal$ is the antitone function.}
 
@@
 Therefore, by a telescoping sum
-\begin{align*}
+\[
 \expect{Y_{B_x}} 
-&= \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})\\
-&\leq -1 < 0 
-\end{align*}
+= \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})
+\leq -1 < 0 
+\]
 %\lo{$\expect{Y_0 - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})}$}
@@
 \begin{align*}
-\Theta[\underline n] &\mapsto n \\
+\Theta[\underline n] &\mapsto n &
 \Xi \, \underline{10} & \mapsto 12.
@@
 
-\subsection*{Non-affine recursion}
+\subsubsection*{Non-affine recursion}
 
@@
 
-\subsection*{Higher-order recursion}
+\subsubsection*{Higher-order recursion}
 
@@
 
-\subsection{Completeness}
+\subsection{Towards completeness}
 It seems very likely that this method is complete in the case that every reachable term is AST (Conjecture~\ref{conj:antitone}), but we have been unable to actually prove this.

### 2021-10-16 LICS21/supermartingales.tex
new subsections and subsection titles

@@
 \label{sec:supermartingales}
+
+\subsection{Ranking functions and supermartingales}
 
@@
 
-\subsection{Ranking functions and supermartingales}
+\subsection{Deriving supermartingales}
 
@@
 Given a closed PPCF term $M$, we transform it to a term $\transform{M}$ of the nondeterministic simply-typed lambda-calculus as follows.
-(We may assume that the nondeterministic calculus is generated from a base type $\iota$ with a $\iota$-type constant symbol $r$, and a function symbol $\bot : A$  for each function type $A$.)
+(We assume that the nondeterministic calculus is generated from a base type $\iota$ with a $\iota$-type constant symbol $r$, and a function symbol $\bot : A$  for each function type $A$.)
 \iffalse
@@
 \transform{\tif{B}{M_1}{M_2}} &:= \big(\lambda z . (\transform{M_1} + \transform{M_2})\big) \, \transform{B} \\
-\transform{\tY N} &:= (\lambda y . \bot) \transform{N}
+\transform{\tY N} &:= (\lambda y . \bot) \, \transform{N}
 \end{align*}

### 2021-10-16 LICS21/confluent-proofs.tex
LaTeX: Removed some white space in the Appendix

@@
 \begin{lemma} \label{lem:independentSamples}
-For any initial term $M$, reduction strategy $f$ on $M$, natural number $n$, skeleton $N$ with $k$ holes, measurable set $T \subset \mathbb R ^ k$ and measurable set $S$ of samples in $I^{L_s(N)}$, $\mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')\}) = \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in I^{L_s(N)} : (M,s) \Rightarrow_f^n (N[r], s')\}) \mu(S)$
+For any initial term $M$, reduction strategy $f$ on $M$, natural number $n$, skeleton $N$ with $k$ holes, measurable set $T \subset \mathbb R ^ k$ and measurable set $S$ of samples in $I^{L_s(N)}$
+\begin{align}
+& \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')\}) \nonumber \\
+& = \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in I^{L_s(N)} : (M,s) \Rightarrow_f^n (N[r], s')\}) \mu(S)
+\label{eq:inde samples}
+\end{align}
 \end{lemma}
@@
 
+We show Equation~\ref{eq:inde samples} in Figure~\ref{fig:inde samples}.
+
+\begin{figure}[h!]
+\small
 \begin{align*}
@@
 \end{align*}
+
+\medskip
+
+\caption{Proof of Equation~\ref{eq:inde samples} 
+\label{fig:inde samples}}
+\end{figure}
 \end{proof}

### 2021-10-16 LICS21/intro.tex
LaTeX: Removed some white space in the Appendix

@@
 
-The language PPCF used in this paper is simply typed, with random continuous sampling, and an explicit recursion primitive, $\tY$. 
+The language Probabilistic PCF (PPCF) used in this paper is simply typed, with random continuous sampling, and an explicit recursion primitive, $\tY$. 
 We define a \emph{ranking function} on a term $M$ to be a non-negative measurable function on the reachable terms from $M$ whose expected value decreases or is unchanged as the term is reduced.
@@
 In Section~\ref{sec:supermartingales}, we show that ranking functions on terms induce supermartingales, which form the basis of a sound method for proving AST.
-We introduce sparse ranking functions in Section~\ref{sec:ranking} and antitone ranking functions in Section~\ref{sec:antitone}, and illustrate how they can be used to prove AST via examples. 
+We introduce sparse ranking functions in Section~\ref{sec:ranking} and antitone ranking functions in Section~\ref{sec:antitone}, and prove sparse function theorems, illustrating how (sparse) (antitone) ranking functions can be used to prove AST via numerous examples.
 In Section~\ref{sec:confluent}, we present a confluent trace semantics and demonstrate its usefulness.

### 2021-10-16 LICS21/ranking.tex
new subsections and subsection titles
LaTeX: Removed some white space in the Appendix

@@
 
-\subsection{Sparse ranking functions}
+\subsection{Sparse ranking functions and sparse function theorem}
 Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and Theorem~\ref{thm:minimal} makes even stronger assumptions than almost sure termination, so it isn't useful for proving it.

### 2021-10-17 LICS21/antitone.tex
various minors

@@
 
+\subsection{Antitone ranking functions and antitone strict supermartingales}
+
 \begin{definition}\rm
@@
 \fi
-
-\subsection{Antitone strict supermartingales}
 
@@
 
-A relevant result here is that the AST problem is decidable for order-1 PHORS (by reduction to the PSPACE-hard solvability of finite systems of polynomial equations with real coefficients; see \cite{DBLP:journals/jacm/EtessamiY09}), but undecidable for PHORS with order greater than 1 \citep{DBLP:conf/lics/KobayashiLG19}.
+A relevant result here is that the AST problem is decidable for order-1 PHORS (by reduction to the PSPACE-hard solvability of finite systems of polynomial equations with real coefficients; see \citep{DBLP:journals/jacm/EtessamiY09}), but undecidable for PHORS with order greater than 1 \citep{DBLP:conf/lics/KobayashiLG19}.
 

### 2021-10-17 LICS21/supermartingales.tex
various minors

@@
 %(In subsequent applications to termination analysis, $\Omega = \entrosp$, and $\mathbb{P} = \mu_{\entrosp}$.)
-Let $T$ be a r.v.~that takes values in $\mathbb{N} \cup \set{\infty}$.
+Let $T$ be a random variable~that takes values in $\mathbb{N} \cup \set{\infty}$.
 We call $T$ a \defn{stopping time adapted to $(\calF_n)_{n \geq 0}$} just if $\set{T = n} \in \calF_n$, for all $n$.
@@
 \begin{asparaenum}[(i)]
-\item A sequence of r.v.s $(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \defn{supermartingale} if for all $n \geq 0$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~(i.e.~for all $A \in \calF_n$, $\int_A Y_{n+1}(\omega) \, \mathbb{P}(\dif \omega) \leq \int_A Y_n(\omega) \, \mathbb{P}(\dif \omega)$).
+\item A sequence of random variables $(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \defn{supermartingale} if for all $n \geq 0$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~(i.e.~for all $A \in \calF_n$, $\int_A Y_{n+1}(\omega) \, \mathbb{P}(\dif \omega) \leq \int_A Y_n(\omega) \, \mathbb{P}(\dif \omega)$).
 \item Let $\epsilon > 0$. 
@@
 %\begin{remark}
-Lemma~\ref{lem:rank-PAST} is a slight extension of \cite[Lem.~5.5]{DBLP:conf/popl/FioritiH15},
+Lemma~\ref{lem:rank-PAST} is a slight extension of \cite[Lemma 5.5]{DBLP:conf/popl/FioritiH15},
 which asserts the same result but for the specific stopping time $T : \omega \mapsto \min\set{n \mid Y_n(\omega) = 0}$.
@@
 and so, by (\ref{eqn:truncate})
-\[
+\begin{equation}
 %\begin{align}
@@
 %\end{align}
-\]
+\end{equation}
 Now, integrating both sides of (\ref{eqn:f 2 n+1}), we have
@@
 \sum_{i \in \calI} \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\sigma_i(s)][\tsample])
-\quad \hbox{$\because$ (\ref{eqn:a u ui})}\\
+\quad \hbox{(Inequality~(\ref{eqn:a u ui}))}\\
 &= 
@@
 
-We first state a useful property about r.v.s $T_0, T_1, T_2, \dots$.
+We first state a useful property of the random variables $T_0, T_1, T_2, \dots$.
 \begin{lemma}
@@
 \transform{\tif{B}{M_1}{M_2}} &:= \big(\lambda z . (\transform{M_1} + \transform{M_2})\big) \, \transform{B} \\
-\transform{\tY N} &:= (\lambda y . \bot) \, \transform{N}
+\transform{\tY \, N} &:= (\lambda y . \bot) \, \transform{N}
 \end{align*}
@@
 
-Finally since the simply-typed nondeterministic lambda calculus is strongly normalising \citep{DBLP:conf/lfcs/Groote94}
+Finally, since the simply-typed nondeterministic lambda calculus is strongly normalising \citep{DBLP:conf/lfcs/Groote94},
 \iffalse

### 2021-10-18 LICS21/antitone.tex
Start adding the new stronger (yet still partial) completeness result

@@
 The following theorem does not prove completeness, but is suggestive in that direction:
-%\begin{restatable}{therm}{antitoneSeemsComplete}
+\begin{therm}
+\label{thm:towards completeness supermartingales}
+For any stopping time $T$ which is almost surely finite, adapted to a filtration $(\calF_n)_n$, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
+\end{therm}
+
+\begin{proof}
+First, take the case where $T$ is bounded by some integer $b$. Setting $\epsilon(x) = 1$ and $Y_n = \max(b-n, 0)$ trivially gives an antitone ranking supermartingale for $T$.
+
+Assume, for the other case, that $T$ is unbounded.
+Let $f(n) = \mathbb P[T < n]$, let $g : [0,1) \to \mathbb N$ be a non-decreasing function such that $g(f(n)) \geq n$ \akr{This was originally $f(g(x)) \geq x$.
+Maybe that was right?
+Maybe I need both.
+I think that would be fine.
+I'll see how the rest of the proof goes.} (essentially just taking the inverse of $f$ and filling in the gaps) and let $\epsilon : \mathbb R_{\geq 0} \to \mathbb R_{>0}$ be a convex antitone function less than or equal to $(g(1-2^{-x}))^{-1}$.
+Then we can define the supermartingale $(Y_n)_n$ by $Y_n = \mathbb E[-\log_2(1-f(T-n)) \mid \calF_n]$.
+Because $T$ is unbounded, $f$ never reaches 1 therefore the logarithm is finite.
+
+In order to show that the expectation $\mathbb E[-\log_2(1-f(T-n))]$ converges, consider its definition as a Lebesgue integral (using the fact that the argument of the expectation here is non-negative):
+\begin{align*}
+& \mathbb E[-\log_2(1-f(T-n))] \\
+= & \int_0^\infty \mathbb P[-\log_2(1-f(T-n)) > x] \, \mathrm d x \\
+= & \int_0^\infty \mathbb P[f(T-n) > 1-2^{-x}] \, \mathrm d x \\
+\leq & \int_0^\infty \mathbb P[f(T) > 1-2^{-x}] \, \mathrm d x \\
+= & \int_0^\infty \mu(\{a \mid \mu(\{b \mid T(b) < T(a)\}) > 1-2^{-x}\}) \, \mathrm d x \\
+\end{align*}
+Call the set $\{a \mid \mu(\{b \mid T(b) < T(a)\}) > 1-y\}$ $A_y$.
+Because the predicate $\mu(\{b \mid T(b) < T(a)\}) > 1-y$ is monotonic in $T(a)$, all elements of $A_y$ have higher values of $T$ than all non-elements of $A_y$.
+Let $a$ be the element of $A_y$ with the minimum value of $T$.
+Then $1-y < \mu(\{b \mid T(b) < T(a)\}) = \mu(A_y^\complement) = 1 - \mu(A_y)$ therefore $\mu(A_y) < y$.
+The integral is therefore less than
+\begin{align*}
+\int_0^\infty 2^{-x} \, \mathrm d x < \infty
+\end{align*}
+therefore $(Y_n)_n$ is well defined.
+
+The sequence is clearly non-negative and adapted to the filtration, so the only condition left to check is the progress condition.
+\begin{align*}
+& \epsilon(Y_n) \\
+= & \epsilon(\mathbb E[-\log_2(1-f(T-n)) \mid \calF_n]) \\
+\leq & \mathbb E[\epsilon(-\log_2(1-f(T-n))) \mid \calF_n] \\
+\leq & \mathbb E[(g(1-2^{\log_2(1-f(T-n))}))^{-1} \mid \calF_n] \\
+= & \mathbb E[(g(f(T-n)))^{-1} \mid \calF_n] \\
+\end{align*}
+
+\akr{I haven't finished writing the proof.}
+
+\end{proof}
+
+This construction can be applied to produce antitone ranking functions in most but not all cases.
+More than simply every reachable term being AST, they must all terminate uniformly quickly.
+The restriction, more specifically, is as follows:
+
 \begin{therm}
 \label{thm:towards completeness}
-For any stopping time $T$ which is almost surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
+For any closed term $M$, if there is some function $t : \mathbb N \to [0,1]$ that tends to 1 such that for any $N \in \mathit{Rch}(M)$, the probability that $N$ terminates within $n$ steps is at least $t(n)$, then $M$ is antitone rankable.
 \end{therm}
-%\end{restatable}
 
 \begin{proof}
-
-If $T$ is bounded by $b$ a.s.~(for $n$ constant), the statement is trivially true by taking $\epsilon$ to be constantly 1, and $Y_n = b - n$.
-Otherwise, let $(t_n)_{n \geq 0}$ be defined recursively such that 
-\begin{align*}
-t_0 &= 0\\
-t_{n+1} &> t_n\\
-\mathbb P[T > t_{n+1}] &\leq \frac 1 4 \mathbb P[T > t_n].
-\end{align*}
-We will then define a supermartingale $(Y_n)_{n \geq 0}$ and nonrandom sequence $(y_n)_{n \geq 0}$ such that $Y_n = 0$ iff $T < n$, $Y_n = y_n$ iff $T \geq n$, and $y_n \geq 2^k$ for $n \geq t_k$.
-
-The antitone function $\epsilon$ is defined piecewise and recursively in such a way as to force all the necessary constraints to hold:
-\begin{align*}
-    \text{for }x \in [0,1),\ &\epsilon(x) = 1 \\
-    \text{for }x \in [2^k,2^{k+1}),\ &\epsilon(x) = \min \left (\epsilon(2^{k-1}), \frac{2^k}{t_{k+1}-t_k} \right ).
-\end{align*}
-This ensures that
-\begin{itemize}
-\item $\epsilon$ is weakly decreasing. 
-\item $\epsilon$ is bounded below by 0.
-\item For $x \geq 2^k$, $\epsilon(x) \leq \frac{2^k}{t_{k+1}-t_k}$. 
-\end{itemize}
-
-The sequence $(y_n)_{n \geq 0}$ is then defined recursively by
-\begin{align*}
-    y_0 = 2, \quad
-    y_{n+1} = (y_n - \epsilon(y_n)) \frac{\mathbb P[T > n]}{\mathbb P[T > n+1]}.
-\end{align*}
-
-The fact $y_{t_k} \geq 2^{k+1}$ is proven by induction on $k$.
-For base case, $2 \geq 2$, as required.
-For the inductive case $k+1$, we first do another induction to prove that $y_n \geq 2^k$ for all $t_k \leq n \leq t_{k+1}$.
-The base case of this inner induction follows from the outer induction hypothesis a fortiori.
-Take the greatest $k$ such that $n > t_k$.
-By the induction hypothesis, $y_{t_k} \geq 2^k$.
-
-\begin{align*}
-    y_n & = y_{t_k} \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} - \sum_{m=t_k}^{n-1} \epsilon(y_m) \frac{\mathbb P[T > m+1]}{\mathbb P[T > n]} \\
-    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \sum_{m=t_k}^{n-1} \epsilon(y_m)) \\
-    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \sum_{m=t_k}^{n-1} \frac{2^k}{t_{k+1}-t_k}) \\
-%\end{align*}
-%Allow a page break so that the sim-diagram lines up neatly.
-%\begin{align*}
-    & = \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \frac{2^k (n - t_k)}{t_{k+1}-t_k}) \\
-    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - 2^k) \\
-    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (2^{k+1} - 2^k) \\
-    & = \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} 2^k \\
-    & \geq 2^k
-\end{align*}
-as required. Substituting $n = t_{k+1}$ and using the fact that $\frac{\mathbb P[T > t_k]}{\mathbb P[T > t_{k+1}]} > 4$, the same reasoning gives $y_{t_{k+1}} \geq 2^{k+2}$, as required.
-
-The fact $y_n \geq 2^k$ for some $k$ implies that $y_n > 0$ for all $n$, which implies that $Y_n \geq 0$. 
-The other condition for $(Y_n)$ to be an antitone-strict supermartingale is
-\begin{align*}
-    \expect{Y_{n+1} \mid \mathcal F_n} 
-    &=  \expect{y_{n+1} \, {\bf 1}_{\set{T \geq n+1}} \mid \mathcal F_n} \\
-    &=  y_{n+1} \, \expect{{\bf 1}_{\set{T \geq n+1}} \mid \mathcal F_n} \\ 
-    &=  y_{n+1} \, {\bf 1}_{\set{T \geq n}} \textstyle \frac{\mathbb P[T > n+1]}{\mathbb P[T > n]} \\
-    &=  (y_n - \epsilon(y_n)) \, {\bf 1}_{\set{T \geq n}} \\
-    &= Y_n - \epsilon(y_n) \, {\bf 1}_{\set{T \geq n}}
-\end{align*}
-therefore $(Y_n)_n$ is an antitone strict supermartingale with respect to $(T,\epsilon)$.
 \end{proof}
 
-The assumption that $(\mathcal F_n)_{n \geq 0}$ is the coarsest filtration to which $T$ is adapted is used in the equality between the third and fourth lines here.
-This condition is the main reason that this proof does not extend directly to a completeness result for antitone ranking functions.
+This theorem is not able to account for all antitone rankable terms.
+
+\begin{example}
+Let $M = (\tY \lambda f s.
+\tif{-\tsample < 0}{(\tY \lambda r n.
+\tif{n < 0}{\underline 0}{r (n - 1)}) s \lfloor \frac 1 \tsample \rfloor}{f (\lambda n.
+2^{s n})}) (\lambda n.
+n)$.
+It first draws samples repeatedly until one of them is not exactly 0, then choses a random number, applies a function to it, and then loops that many times.
+Every time the sample drawn is exactly 0, the function gets exponentially faster growing.
+It therefore doesn't satisfy the conditions of Theorem \ref{thm:towards completeness} because the term reached after $n$ draws of 0 terminates slower as $n$ increases.
+However, it is antitone rankable nevertheless.
+\end{example}

### 2021-11-02 LICS21/antitone.tex
minor
Finish antitone completeness proof for abstract stopping times.

@@
 Assume, for the other case, that $T$ is unbounded.
-Let $f(n) = \mathbb P[T < n]$, let $g : [0,1) \to \mathbb N$ be a non-decreasing function such that $g(f(n)) \geq n$ \akr{This was originally $f(g(x)) \geq x$.
-Maybe that was right?
-Maybe I need both.
-I think that would be fine.
-I'll see how the rest of the proof goes.} (essentially just taking the inverse of $f$ and filling in the gaps) and let $\epsilon : \mathbb R_{\geq 0} \to \mathbb R_{>0}$ be a convex antitone function less than or equal to $(g(1-2^{-x}))^{-1}$.
-Then we can define the supermartingale $(Y_n)_n$ by $Y_n = \mathbb E[-\log_2(1-f(T-n)) \mid \calF_n]$.
+Let $f(n) \leq \mathbb P[T < n]$ where $f$ tends to 1 and is concave for arguments $\geq 0$, let $g : [0,1) \to \mathbb N$ be a non-decreasing function such that $g(f(n)) \geq n$ (essentially just taking the inverse of $f$ and filling in the gaps) and let $\epsilon : \mathbb R_{\geq 0} \to \mathbb R_{>0}$ be a convex antitone function $\leq \Delta[f](g(1-e^{-x})-1)$ where $\Delta[f](n) = f(n+1) - f(n)$.
+Then we can define the supermartingale $(Y_n)_n$ by $Y_n = \mathbb E[-\log(1-f(T-n)) \mid \calF_n]$.
 Because $T$ is unbounded, $f$ never reaches 1 therefore the logarithm is finite.
 
-In order to show that the expectation $\mathbb E[-\log_2(1-f(T-n))]$ converges, consider its definition as a Lebesgue integral (using the fact that the argument of the expectation here is non-negative):
+In order to show that the expectation $\mathbb E[-\log(1-f(T-n))]$ converges, consider its definition as a Lebesgue integral (using the fact that the argument of the expectation here is non-negative):
 \begin{align*}
-& \mathbb E[-\log_2(1-f(T-n))] \\
-= & \int_0^\infty \mathbb P[-\log_2(1-f(T-n)) > x] \, \mathrm d x \\
-= & \int_0^\infty \mathbb P[f(T-n) > 1-2^{-x}] \, \mathrm d x \\
-\leq & \int_0^\infty \mathbb P[f(T) > 1-2^{-x}] \, \mathrm d x \\
-= & \int_0^\infty \mu(\{a \mid \mu(\{b \mid T(b) < T(a)\}) > 1-2^{-x}\}) \, \mathrm d x \\
+& \mathbb E[-\log(1-f(T-n))] \\
+= & \int_0^\infty \mathbb P[-\log(1-f(T-n)) > x] \, \mathrm d x \\
+= & \int_0^\infty \mathbb P[f(T-n) > 1-e^{-x}] \, \mathrm d x \\
+\leq & \int_0^\infty \mathbb P[f(T) > 1-e^{-x}] \, \mathrm d x \\
+\leq & \int_0^\infty \mu(\{a \mid \mu(\{b \mid T(b) < T(a)\}) > 1-e^{-x}\}) \, \mathrm d x \\
 \end{align*}
-Call the set $\{a \mid \mu(\{b \mid T(b) < T(a)\}) > 1-y\}$ $A_y$.
+Let $A_y$ be the set $\{a \mid \mu(\{b \mid T(b) < T(a)\}) > 1-y\}$.
 Because the predicate $\mu(\{b \mid T(b) < T(a)\}) > 1-y$ is monotonic in $T(a)$, all elements of $A_y$ have higher values of $T$ than all non-elements of $A_y$.
@@
 \begin{align*}
-\int_0^\infty 2^{-x} \, \mathrm d x < \infty
+\int_0^\infty e^{-x} \, \mathrm d x < \infty
 \end{align*}
@@
 \begin{align*}
-& \epsilon(Y_n) \\
-= & \epsilon(\mathbb E[-\log_2(1-f(T-n)) \mid \calF_n]) \\
-\leq & \mathbb E[\epsilon(-\log_2(1-f(T-n))) \mid \calF_n] \\
-\leq & \mathbb E[(g(1-2^{\log_2(1-f(T-n))}))^{-1} \mid \calF_n] \\
-= & \mathbb E[(g(f(T-n)))^{-1} \mid \calF_n] \\
+& \epsilon(Y_n) \mathbf 1_{T > n} \\
+= & \epsilon(\mathbb E[-\log(1-f(T-n)) \mid \calF_n]) \mathbf 1_{T > n} \\
+\leq & \mathbb E[\epsilon(-\log(1-f(T-n))) \mathbf 1_{T > n} \mid \calF_n] \\
+\leq & \mathbb E[\Delta[f](g(1-e^{\log(1-f(T-n))})-1) \mathbf 1_{T > n} \mid \calF_n] \\
+= & \mathbb E[\Delta[f](g(f(T-n))-1) \mathbf 1_{T > n} \mid \calF_n] \\
+\leq & \mathbb E[\Delta[f](T-n-1) \mid \calF_n] \\
+= & \mathbb E[f(T-n)-f(T-n-1) \mid \calF_n] \\
+\leq & \mathbb E[\frac{f(T-n)-f(T-n-1)}{1-f(T-n-1)} \mid \calF_n] \\
+\leq & \mathbb E[\log(1-f(T-n-1)) - \log(1-f(T-n)) \mid \calF_n] \\
+= & Y_n - \mathbb E[Y_{n+1} \mid \calF_n]
 \end{align*}
-
-\akr{I haven't finished writing the proof.}
+This relies on the facts that $\epsilon$ is convex, $g(f(n)) \geq n$, and $\Delta[f]$ is antitone for arguments $\geq 0$.
 

### 2021-11-08 LICS21/antitone.tex
Add more proof in the direction of completeness, and a partial couterexample

@@
 \begin{align*}
-\Xi & = \tY\, \lambda f x.\, (\lambda e. X[e]) \, \tsample \\
-X[e] & = \tif{e \leq p - x^{-2}}{x+1}{f^2(x+1) \oplus_e f(x+1)}
+\Xi & := \tY\, \lambda f x.\, (\lambda e. X[e]) \, \tsample \\
+X[e] & := \tif{e \leq p - x^{-2}}{x+1}{f^2(x+1) \oplus_e f(x+1)}
 \end{align*}
@@
 \label{thm:towards completeness}
-For any closed term $M$, if there is some function $t : \mathbb N \to [0,1]$ that tends to 1 such that for any $N \in \mathit{Rch}(M)$, the probability that $N$ terminates within $n$ steps is at least $t(n)$, then $M$ is antitone rankable.
+For any closed term $M$, if there is some function $t : \mathbb N \to [0,1]$ that tends to 1 such that for any $N \in \mathit{Rch}(M)$, there exists some $m$ such that the probability that $N$ terminates within $n$ steps is at least $\frac{t(n+m)-t(m)}{1-t(m)}$, then $M$ is antitone rankable.
 \end{therm}
@@
 \begin{proof}
+The same construction as in the previous lemma applies, with the stopping time $T$ being the $\tY$-runtime $T_M^\tY$, but with $f$ being a concave function $\leq t(n)$ instead of $\leq \mathbb P[T < n]$.
+The definition of $\epsilon$ is identical, and the definition of the ranking function is similar, but taking the $\tY$-runtime starting at $N$ rather than taking the conditional expectation, so it's $\mathbb E[-\log(1-f(T_N^\tY))]$.
+While the conditional expectation is only defined up to a set of measure 0, this definition is largely equivalent but works for all $N$.
+\akr{I feel like it's neater to just leave it at that, and imply the proof that the expectation is bounded is the same as in the more abstract case, but the details are a little different.}
+
+Despite the differences between the constructions, the proofs that the ranking function is finite and satisfies the progress condition are generally similar to the proofs in the case of the abstract stopping time.
 \end{proof}
 
-This theorem is not able to account for all antitone rankable terms.
+This theorem is not able to account for all terms whose antitone rankability is allowed by the theorems proven so far, as it requires not only that all reachable terms terminate, but that they terminate uniformly quickly.
+In practice though this is likely to include all reasonable programs.
 
 \begin{example}
-Let $M = (\tY \lambda f s.
-\tif{-\tsample < 0}{(\tY \lambda r n.
-\tif{n < 0}{\underline 0}{r (n - 1)}) s \lfloor \frac 1 \tsample \rfloor}{f (\lambda n.
-2^{s n})}) (\lambda n.
-n)$.
-It first draws samples repeatedly until one of them is not exactly 0, then choses a random number, applies a function to it, and then loops that many times.
-Every time the sample drawn is exactly 0, the function gets exponentially faster growing.
-It therefore doesn't satisfy the conditions of Theorem \ref{thm:towards completeness} because the term reached after $n$ draws of 0 terminates slower as $n$ increases.
-However, it is antitone rankable nevertheless.
+The number of functions $\mathbb N \to \mathbb N$ is equal to $|\mathbb R|$, therefore by assigning some code and uncurrying, it is possible to define a function $Z : \mathbb R \times \mathbb N \to \mathbb N$ such that $Z(r, \cdot)$ can be any function $\mathbb N \to \mathbb N$ for some $r \in [0,1]$, and it is even possible to do this in such a way that $Z$ is measurable.
+As PPCF contains every measurable real function as a primitive, it is therefore possible for a PPCF program to randomly pick a function $\mathbb N \to \mathbb N$ so that every such function is possible.
+
+Let $M = (\lambda r m.
+(\tY \lambda w n.
+\tif{n < 0}{\underline 0}{w (n-1)}) \underline Z(r, m)) \tsample \lfloor \frac 1 \tsample \rfloor$.
+This program first picks a random function, then picks an argument for that function from $\mathbb N$, then counts down from the result to 0.
+Assuming for convenience that $\frac 1 0$ evaluates to $0$, this clearly terminates almost surely, as do all reachable terms, but for any candidate $t$, there is some $r$ such that $Z(r, \cdot)$ is much faster growing than $t^{-1}$, therefore the condition for Theorem \ref{thm:towards completeness} is not satisfied for $M$.
+\akr{I don't actually know whether $M$ is antitone rankable.
+Now I've thought of the example, my guess is that it isn't, which would imply that the completeness conjecture isn't true.
+If I manage to work it out either way, I'll want to change some of what I've written to account for that, but I'm working for now on the assumption that I won't work it out before publishing.} It is not clear whether this term is antitone rankable at all.
+However, the construction relies crucially on the peculiar property of PPCF that programs can sample an infinite amount of random data at once (in the form of unlimited precision real numbers).
+The peculiar function $Z$ also plays a crucial role in the construction, but it would be possible to implement it entirely in terms of basic arithmetic operations, so the fact that it is permitted as a primitive should not be taken as a sign that this is an issue unique to an abstract uncomputable language like PPCF.
 \end{example}

### 2021-11-09 LICS21/antitone.tex
minor

--- 2021-11-02 LICS21/antitone.tex
@@
 \begin{example}
-Let $M = (\tY \lambda f s.
-\tif{-\tsample < 0}{(\tY \lambda r n.
-\tif{n < 0}{\underline 0}{r (n - 1)}) s \lfloor \frac 1 \tsample \rfloor}{f (\lambda n.
-2^{s n})}) (\lambda n.
-n)$.
+Take the term 
+\[
+M = (\tY \lambda f s. \tif{-\tsample < 0}{(\tY \lambda r n. \tif{n < 0}{\underline 0}{r (n - 1)}) s \lfloor \frac 1 \tsample \rfloor}{f (\lambda n. 2^{s n})}) (\lambda n. n).
+\] 
 It first draws samples repeatedly until one of them is not exactly 0, then choses a random number, applies a function to it, and then loops that many times.

### 2021-11-26 LICS21/related.tex
Update because the completeness conjecture is false

@@
 
-\subsection*{Further directions} We have not yet been able to prove Conjecture~\ref{conj:antitone} which would imply that Theorem~\ref{thm:antitone rankable implies termination} is almost complete.
-\begin{conjecture}[Completeness]
-\label{conj:antitone}
-If every term reachable from a certain PPCF term is AST, then that term is antitone rankable.
-\end{conjecture}
+\subsection*{Further directions} All of the variants of the method we have presented for proving almost sure termination are incomplete, being unable to deal with terms like Example \ref{ex:antitone isn't complete}, so extending it to be applicable to more terms would likely be possible.
 

### 2021-11-27 LICS21/antitone.tex

--- 2021-11-08 LICS21/antitone.tex
@@
 \subsection{Towards completeness}
-It seems very likely that this method is complete in the case that every reachable term is AST (Conjecture~\ref{conj:antitone}), but we have been unable to actually prove this.
-It is certainly at least capable of proving the termination of terms which terminate arbitrarily slowly (in the sense that there is no similar limitation to Theorem~\ref{thm:rankable implies termination}, which can only prove termination of $\tY$-PAST terms).
+Unlike Theorem~\ref{thm:rankable implies termination}, which can only prove termination of $\tY$-PAST terms, this method is capable of proving termination of terms which terminate arbitrarily slowly.
 
-The following theorem does not prove completeness, but is suggestive in that direction:
+More precisely:
 \begin{therm}
@@
 This construction can be applied to produce antitone ranking functions in most but not all cases.
-More than simply every reachable term being AST, they must all terminate uniformly quickly.
-The restriction, more specifically, is as follows:
+As with the simple version of ranking functions, an antitone ranking function can only exist if every reachable term is AST, because every reachable term must itself be assigned a value and is therefore antitone rankable.
+An additional restriction is necessary to apply this argument to constructing antitone ranking functions though, to ensure that all the reachable subterms (particularly those only reachable with probability 0, which are less constrained) are antitone rankable with respect to the same antitone function $\epsilon$.
 
@@
 This theorem is not able to account for all terms whose antitone rankability is allowed by the theorems proven so far, as it requires not only that all reachable terms terminate, but that they terminate uniformly quickly.
-In practice though this is likely to include all reasonable programs.
+In practice this seems likely to include most reasonable programs, but there are still some cases where it doesn't work.
 
-\begin{example}
+\begin{example} \label{ex:antitone isn't complete}
 The number of functions $\mathbb N \to \mathbb N$ is equal to $|\mathbb R|$, therefore by assigning some code and uncurrying, it is possible to define a function $Z : \mathbb R \times \mathbb N \to \mathbb N$ such that $Z(r, \cdot)$ can be any function $\mathbb N \to \mathbb N$ for some $r \in [0,1]$, and it is even possible to do this in such a way that $Z$ is measurable.
@@
 This program first picks a random function, then picks an argument for that function from $\mathbb N$, then counts down from the result to 0.
-Assuming for convenience that $\frac 1 0$ evaluates to $0$, this clearly terminates almost surely, as do all reachable terms, but for any candidate $t$, there is some $r$ such that $Z(r, \cdot)$ is much faster growing than $t^{-1}$, therefore the condition for Theorem \ref{thm:towards completeness} is not satisfied for $M$.
-\akr{I don't actually know whether $M$ is antitone rankable.
-Now I've thought of the example, my guess is that it isn't, which would imply that the completeness conjecture isn't true.
-If I manage to work it out either way, I'll want to change some of what I've written to account for that, but I'm working for now on the assumption that I won't work it out before publishing.} It is not clear whether this term is antitone rankable at all.
-However, the construction relies crucially on the peculiar property of PPCF that programs can sample an infinite amount of random data at once (in the form of unlimited precision real numbers).
+Assuming for convenience that $\frac 1 0$ evaluates to $0$, this clearly terminates almost surely, as do all terms reachable from it, but for any candidate $t$, there is some $r$ such that $Z(r, \cdot)$ is much faster growing than $t^{-1}$, therefore the condition for Theorem \ref{thm:towards completeness} is not satisfied for $M$.
+It turns out this term is not antitone rankable at all.
+For any $\epsilon$, there is some $f: \mathbb N \to \mathbb N$ sufficiently fast growing that the ranking function value required for each $(\lambda m.
+(\tY \lambda w n.
+\tif{n < 0}{\underline 0}{w (n-1)}) \underline f(m)) \underline u$ is at least inversely proportional to the chance of $\lfloor \frac 1 \tsample \rfloor$ evaluating to $\underline u$, therefore $(\lambda m.
+(\tY \lambda w n.
+\tif{n < 0}{\underline 0}{w (n-1)}) \underline f(m)) \tsample \lfloor \frac 1 \tsample \rfloor$ as a whole is not rankable w.r.t.
+that $\epsilon$, and $f$ is $Z(r, \cdot)$ for some $r$, therefore $M$ isn't antitone rankable at all.
+
+This term is therefore a counterexample to the completeness conjecture in a previous version of this paper\cite{thisLics}, but the construction relies crucially on the peculiar property of PPCF that programs can sample an infinite amount of random data at once (in the form of unlimited precision real numbers).
 The peculiar function $Z$ also plays a crucial role in the construction, but it would be possible to implement it entirely in terms of basic arithmetic operations, so the fact that it is permitted as a primitive should not be taken as a sign that this is an issue unique to an abstract uncomputable language like PPCF.

### 2021-11-29 LICS21/antitone.tex
Minor question about Example 35

@@
 \begin{definition}\rm
-Given a probability space $(\Omega, \calF, \mathbb{P})$, and a supermartingale $(Y_n)_{n \geq 0}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \geq 0}$,
+Given a probability space $(\Omega, \calF, \mathbb{P})$, a supermartingale $(Y_n)_{n \geq 0}$, and a stopping time $T$ adapted to filtration $(\calF_n)_{n \geq 0}$,
 we say that $(Y_n)_{n \geq 0}$ is an \defn{antitone strict supermartingale w.r.t.~$T$} if for all $n \geq 0$, we have $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
@@
 \label{thm:towards completeness supermartingales}
-For any stopping time $T$ which is almost surely finite, adapted to a filtration $(\calF_n)_n$, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
+For any stopping time $T$ which is almost surely finite, adapted to a filtration $(\calF_n)_n$, 
+there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
 \end{therm}
@@
 
-Let $M = (\lambda r m.
-(\tY \lambda w n.
-\tif{n < 0}{\underline 0}{w (n-1)}) \underline Z(r, m)) \tsample \lfloor \frac 1 \tsample \rfloor$.
+Let 
+\[
+M = \Big(\lambda r \; m. 
+\big(\tY \lambda w \; n. \tif{n < 0}{\underline 0}{w (n-1)}
+\big) \; \underline Z(r, m)
+\Big) \; \tsample \; 
+\left\lfloor \tfrac 1 \tsample 
+\right\rfloor.
+\] 
 This program first picks a random function, then picks an argument for that function from $\mathbb N$, then counts down from the result to 0. 
-Assuming for convenience that $\frac 1 0$ evaluates to $0$, this clearly terminates almost surely, as do all terms reachable from it, but for any candidate $t$, there is some $r$ such that $Z(r, \cdot)$ is much faster growing than $t^{-1}$, therefore the condition for Theorem \ref{thm:towards completeness} is not satisfied for $M$.
-It turns out this term is not antitone rankable at all.
-For any $\epsilon$, there is some $f: \mathbb N \to \mathbb N$ sufficiently fast growing that the ranking function value required for each $(\lambda m.
-(\tY \lambda w n.
-\tif{n < 0}{\underline 0}{w (n-1)}) \underline f(m)) \underline u$ is at least inversely proportional to the chance of $\lfloor \frac 1 \tsample \rfloor$ evaluating to $\underline u$, therefore $(\lambda m.
-(\tY \lambda w n.
-\tif{n < 0}{\underline 0}{w (n-1)}) \underline f(m)) \tsample \lfloor \frac 1 \tsample \rfloor$ as a whole is not rankable w.r.t.
-that $\epsilon$, and $f$ is $Z(r, \cdot)$ for some $r$, therefore $M$ isn't antitone rankable at all.
+Assuming for convenience that $\frac 1 0$ evaluates to $0$, this clearly terminates almost surely, as do all terms reachable from it.
 
-This term is therefore a counterexample to the completeness conjecture in a previous version of this paper\cite{thisLics}, but the construction relies crucially on the peculiar property of PPCF that programs can sample an infinite amount of random data at once (in the form of unlimited precision real numbers).
-The peculiar function $Z$ also plays a crucial role in the construction, but it would be possible to implement it entirely in terms of basic arithmetic operations, so the fact that it is permitted as a primitive should not be taken as a sign that this is an issue unique to an abstract uncomputable language like PPCF.
+However for any candidate $t$ \lo{of type $\mathbb N \to \mathbb N$? or $\mathbb N \to [0, 1]$ as in Theorem~\ref{thm:towards completeness}?}, 
+there is some $r$ such that $Z(r, \cdot)$ is much faster growing than $t^{-1}$, therefore the condition for Theorem \ref{thm:towards completeness} is not satisfied for $M$.
+
+It turns out that $M$ is not antitone rankable at all. 
+For any antitone $\epsilon : \nnReal \to \pReal$,
+there is some $f: \mathbb N \to \mathbb N$ sufficiently fast growing that the ranking function value required for each 
+\[
+\Big(
+\lambda m. 
+\big(
+\tY \lambda w n. \tif{n < 0}{\underline 0}{w (n-1)}
+\big)
+\; \underline f(m)
+\Big) 
+\; \underline u
+\] 
+is at least inversely proportional to the chance of $\lfloor \frac 1 \tsample \rfloor$ evaluating to $\underline u$, therefore 
+\changed[lo]{ 
+\[
+\Big(\lambda m. 
+\big(
+\tY \lambda w \; n. \tif{n < 0}{\underline 0}{w (n-1)}
+\big)
+\; \underline f(m)
+\Big) 
+%\; \tsample 
+\; \left\lfloor \tfrac 1 \tsample \right\rfloor
+\]}
+as a whole is not rankable w.r.t. that $\epsilon$, and $f$ is $Z(r, \cdot)$ for some $r$.
+Therefore $M$ isn't antitone rankable at all.
+
+\medskip
+
+This term is therefore a counterexample to the completeness conjecture in the conference (LICS 2021) version of this paper \citep{DBLP:conf/lics/Kenyon-RobertsO21}, but the construction relies crucially on the peculiar property of PPCF that programs can sample an infinite amount of random data at once (in the form of unlimited precision real numbers).
+The peculiar function $Z$ also plays a crucial role in the construction, 
+but it would be possible to implement it entirely in terms of basic arithmetic operations, 
+so the fact that it is permitted as a primitive should not be taken as a sign that this is an issue unique to an abstract uncomputable language like PPCF.
 \end{example}

### 2021-12-17 LICS21/antitone.tex
Mostly minor fixes
Things caught in proofreading. The one major error I found was in
the definition of CBV order of reductions in the appendix.

@@
 \dfrac{
-\begin{array}{l}
-\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
-+ 
-\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
-\\
-{} - 
-\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}
-\end{array}
+\mathbb P[T > n] \, \expect{Y_n \mid T > n} + \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} - \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}
 }{\mathbb P[T > n]}
@@
 \begin{example}
-Take the probability space $(\Omega, \calF, \mathbb{P})$ where $\Omega$ is the closed interval of reals $[-1, 1]$, $\calF$ the Borel $\sigma$-algebra, and $\mathbb{P}$ the corresponding Lebesgue probability measure.
+To demonstrate why it is necessary that $\epsilon(x) > 0$ for all $x$, not merely for those values the ranking function actually takes, take the probability space $(\Omega, \calF, \mathbb{P})$ where $\Omega$ is the closed interval of reals $[-1, 1]$, $\calF$ the Borel $\sigma$-algebra, and $\mathbb{P}$ the corresponding Lebesgue probability measure.
 Let $\omega \in [-1, 1]$. Define random variables $T$ and $(Y_k)_{k \geq 0}$:
@@
 
-\subsection{Towards completeness}
+\subsection{Generality and incompleteness}
 Unlike Theorem~\ref{thm:rankable implies termination}, which can only prove termination of $\tY$-PAST terms, this method is capable of proving termination of terms which terminate arbitrarily slowly.
@@
 This construction can be applied to produce antitone ranking functions in most but not all cases.
-As with the simple version of ranking functions, an antitone ranking function can only exist if every reachable term is AST, because every reachable term must itself be assigned a value and is therefore antitone rankable.
+If cases with probability 0 could be entirely ignored, as they are in the context of supermartingales, this would be enough to prove that antitone ranking functions were a complete method of proving AST.
+However, as with the simple version of ranking functions, an antitone ranking function can only exist if every reachable term is AST, because every reachable term must itself be assigned a value and is therefore antitone rankable.
 An additional restriction is necessary to apply this argument to constructing antitone ranking functions though, to ensure that all the reachable subterms (particularly those only reachable with probability 0, which are less constrained) are antitone rankable with respect to the same antitone function $\epsilon$.
@@
 \begin{example} \label{ex:antitone isn't complete}
-The number of functions $\mathbb N \to \mathbb N$ is equal to $|\mathbb R|$, therefore by assigning some code and uncurrying, it is possible to define a function $Z : \mathbb R \times \mathbb N \to \mathbb N$ such that $Z(r, \cdot)$ can be any function $\mathbb N \to \mathbb N$ for some $r \in [0,1]$, and it is even possible to do this in such a way that $Z$ is measurable.
+The number of functions $\mathbb N \to \mathbb N$ is equal to $|\mathbb R|$, therefore by assigning some code and uncurrying, it is possible to define a function $Z : \mathbb R \times \mathbb N \to \mathbb N$ such that $Z(r, \cdot)$ can be any function $\mathbb N \to \mathbb N$ given the right value of $r \in [0,1]$, and it is even possible to do this in such a way that $Z$ is measurable.
 As PPCF contains every measurable real function as a primitive, it is therefore possible for a PPCF program to randomly pick a function $\mathbb N \to \mathbb N$ so that every such function is possible.
@@
 
-However for any candidate $t$ \lo{of type $\mathbb N \to \mathbb N$? or $\mathbb N \to [0, 1]$ as in Theorem~\ref{thm:towards completeness}?}, 
-there is some $r$ such that $Z(r, \cdot)$ is much faster growing than $t^{-1}$, therefore the condition for Theorem \ref{thm:towards completeness} is not satisfied for $M$.
+However for any candidate $t$, there is some $r$ such that $Z(r, \cdot)$ is much faster growing than $n \mapsto t^{-1}(\frac 1 n)$, therefore the condition for Theorem \ref{thm:towards completeness} is not satisfied for $M$.
 
 It turns out that $M$ is not antitone rankable at all. 
-For any antitone $\epsilon : \nnReal \to \pReal$,
-there is some $f: \mathbb N \to \mathbb N$ sufficiently fast growing that the ranking function value required for each 
+Every antitone $\epsilon : \nnReal \to \pReal$ defines some correspondence between expected $\tY$ runtimes of terms and how high the ranking function value must be,
+therefore there is some $f: \mathbb N \to \mathbb N$ sufficiently fast growing that the ranking function value required for each 
 \[
@@
 is at least inversely proportional to the chance of $\lfloor \frac 1 \tsample \rfloor$ evaluating to $\underline u$, therefore 
-\changed[lo]{ 
 \[
@@
 \; \left\lfloor \tfrac 1 \tsample \right\rfloor
-\]}
+\]
 as a whole is not rankable w.r.t. that $\epsilon$, and $f$ is $Z(r, \cdot)$ for some $r$.
@@
 This term is therefore a counterexample to the completeness conjecture in the conference (LICS 2021) version of this paper \citep{DBLP:conf/lics/Kenyon-RobertsO21}, but the construction relies crucially on the peculiar property of PPCF that programs can sample an infinite amount of random data at once (in the form of unlimited precision real numbers).
-The peculiar function $Z$ also plays a crucial role in the construction, 
+The unusual function $Z$ also plays a crucial role in the construction, 
 but it would be possible to implement it entirely in terms of basic arithmetic operations, 

### 2021-12-17 LICS21/confluent-proofs.tex
various minors
Mostly minor fixes
Things caught in proofreading. The one major error I found was in
the definition of CBV order of reductions in the appendix.

@@
 For this, the member of the class that's in call-by-value order is used.
-A reduction sequence is in call-by-value order if for every pair of reductions in the sequence at positions $\alpha$ and $\beta$, where the reduction at $\alpha$ happens first, either $\alpha \leq \beta$ or $\alpha$ and $\beta$ are disjoint, with $\alpha$ to the left of $\beta$ (i.e.~the first elements of the sequences $\alpha, \beta$ where they differ are $@_1$ and $@_2$, $\underline f_i$ and $\underline f_j$ for $i < j$, or $\textsf{if}_i$ and $\textsf{if}_j$ for $i < j$, in that order).
+A reduction sequence is in call-by-value order if for every pair of reductions in the sequence at positions $\alpha$ and $\beta$, where the reduction at $\alpha$ happens first, either $\alpha \leq \beta$, the redex at $\beta$ doesn't form until after the reduction at $\alpha$, or $\alpha$ and $\beta$ are disjoint, with $\alpha$ to the left of $\beta$ (i.e.~the first elements of the sequences $\alpha, \beta$ where they differ are $@_1$ and $@_2$, $\underline f_i$ and $\underline f_j$ for $i < j$, or $\textsf{if}_i$ and $\textsf{if}_j$ for $i < j$, in that order).
 Note that although the reductions that occur in the sequence are in call-by-value order, it may not be the actual call-by-value reduction sequence starting from that point because some redexes may remain un-reduced even when other reductions that should happen afterwards occur.
@@
 
-We show Equation~\ref{eq:inde samples} in Figure~\ref{fig:inde samples}.
+We show Equation~(\ref{eq:inde samples}) in Figure~\ref{fig:inde samples}.
 
@@
 The distribution of $s(k)$ conditional on reaching $N$ is therefore the uniform distribution on $I$.
-Similarly for the other semantics, if the distribution of remaining samples after $n$ steps is independent of the term by Lemma~\ref{lem:independentSamples}, therefore the distribution of $s(\mathit{Sk}(N),\cbv(N))$ conditional on $N$ is the uniform distribution on $I$.
+Similarly for the other semantics, the distribution of remaining samples after $n$ steps is independent of the term by Lemma~\ref{lem:independentSamples}, therefore the distribution of $s(\mathit{Sk}(N),\cbv(N))$ conditional on $N$ is the uniform distribution on $I$.
 In either case, the distribution of the next term conditional on the previous term, in the case that it reduces randomly, is equal to the image under $r \mapsto E[\underline r]$ of the uniform distribution on $I$.
 
-In any case, the distribution of reduction sequences after $n+1$ steps, conditional on the reduction sequence after $n$ steps, as defined by $\Rightarrow_{\cbv}$ and by $\red$, is equal, any by the inductive hypothesis the distributions after $n$ steps are equal, therefore by integrating over the reduction sequence up to step $n$, the distributions on reduction sequences up to step $n+1$ are equal.
+In any case, the distribution of reduction sequences after $n+1$ steps, conditional on the reduction sequence after $n$ steps, as defined by $\Rightarrow_{\cbv}$ and by $\red$, is equal, and by the inductive hypothesis the distributions after $n$ steps are equal, therefore by integrating over the reduction sequence up to step $n$, the distributions on reduction sequences up to step $n+1$ are equal.
 

### 2021-12-17 LICS21/confluent.tex
Mostly minor fixes
Things caught in proofreading. The one major error I found was in
the definition of CBV order of reductions in the appendix.

@@
 In order to be able to use alternative reduction strategies to prove properties of the program defined using the usual reduction strategy, we first prove the relation between these different strategies, providing a confluent version of the semantics.
-It is therefore necessary to have some way of relating the samples that will be used in runs of the program using different reduction strategies.
+It is therefore necessary to have some way of relating the samples that will be used in runs of the program using different reduction strategies, to ensure that the same numbers are used regardless of what order they're used in.
 Positions are defined as a way of addressing sample statements within a program independently of the reduction order.
 Next, a version of the reduction relation is presented that is nondeterministic, so that it allows a choice of what order to perform reductions in.
-The notion of positions is extended to potential positions, for samples which may appear later in the reduction sequence but not necessarily in the initial term.
+The notion of positions is extended to potential positions, so as to be able to also label samples which may appear later in the reduction sequence but not necessarily in the initial term.
 In order to allow potential positions in different reduction sequences to be considered equivalent, a relation $\sim^*$ is defined, and finally, all of these are used to construct a confluent version of the trace semantics, $\Rightarrow$, that is still nondeterministic in reduction order, but does specify the outcome of random choices.
@@
 
+\medskip
 Non-probabilistic lambda calculi are generally confluent, i.e.~if a term $A$ reduces to both $B_1$ and $B_2$, there is some $C$ 
@@
 
-\medskip
-Even with such a restriction, a trace semantics in the style of the one already defined would not be entirely confluent, as, for example, $\red^3(\tsample - \tsample, (1,0,\dots))$ would be either $1$ or $-1$ depending on the order of evaluation of the $\tsample$s, as that determines which sample from the pre-selected sequence is used for each one.
+Even with a suitable restriction, a trace semantics in the style of the one already defined would not be entirely confluent, as, for example, $\red^3(\tsample - \tsample, (1,0,\dots))$ would be either $1$ or $-1$ depending on the order of evaluation of the $\tsample$s, as that determines which sample from the pre-selected sequence is used for each one.
 To fix this, rather than pre-selecting samples according to the order they'll be drawn in, select them according to the position in the term where they'll be used instead.
+
+\pagebreak %Shift how things align to avoid a page break at a really bad location in defn:sim.
 
@@
 \end{align*}
-In each of these cases, $M \mid \alpha$ is the \defn{redex}, and the reduction \emph{takes place at $\alpha$}.
+In each of these cases, $M \mid \alpha$ is the \defn{redex}, and the reduction \defn{takes place at $\alpha$}.
 Each subterm can be a redex in at most one way, but there can be multiple redexes at different positions.
@@
 
-%\medskip
+\medskip %Shift how things align to avoid a page break at a really bad location in defn:sim.
 This is still not quite sufficient to attain confluence because sometimes the same samples must be used at corresponding positions in different reduction sequences.
@@
 
+\medskip %Shift how things align to avoid a page break at a really bad location in defn:sim.
 \begin{definition}\rm \label{defn:sim}
 The relation $\sim$ is defined as the union of the minimal symmetric relations $\sim_p$ (``$p$'' for parent-child) and $\sim_c$ (``$c$'' for cousin) satisfying
-\begin{asparaenum}[(i)]
+\begin{compactenum}[(i)]
     \item If $N$ reduces to $O$ with the redex at position $\alpha$, and $\beta$ is a position in $N$ disjoint from $\alpha$, then $(N,\beta) \sim_p (O,\beta)$.
@@
     (in which case $O_1$ and $O_2$ are equal as skeletons, but with different reduction sequences), $O_1'$ and $O_2'$ are the results of applying some reduction sequence to each of $O_1$ and $O_2$ (the same reductions in each case, which is always possible because they're equal skeletons), and $\delta$ is a position in $O_1'$ (or equivalently $O_2'$), then $(O_1',\delta) \sim_c (O_2',\delta)$.
-\end{asparaenum}
+\end{compactenum}
+
 The $\sim_c$-rules are illustrated in Figure~\ref{sim-diagram}.
@@
 \ref{thm:antitone rankable implies termination} and~\ref{thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies.
-They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because Theorem~\ref{thm:de groote} applies equally to any reduction strategy).
+They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because Lemma~\ref{lem:TnBounded} applies equally to any reduction strategy).
 For a reduction strategy $r$ and term $M$, just substitute $N[r(N)]$ being a $\tY$-redex for $N$ being of the form $E[\tY \lambda x.
@@
 \medskip
-If, as seems likely, Theorem~\ref{thm:antitone rankable implies termination} is complete for terms from which every reachable term is AST, then by the following theorem, Theorem~\ref{thm:confluent ranking} can't actually prove the termination of any terms not already provable by Theorem~\ref{thm:antitone rankable implies termination} and Theorem~\ref{thm:antitone partial implies rankable}.
+It's not clear that Theorem~\ref{thm:confluent ranking} is actually any stronger than Theorems~\ref{thm:antitone rankable implies termination} and~\ref{thm:antitone partial implies rankable} in the sense of being able to prove termination of an more terms.
+In particular, it cannot be used to avoid the restriction to terms where all reachable terms are also AST.
 
@@
 M  &= (\lambda n. P[n]) \lfloor \tsample^{-1/2} \rfloor \\
-P[n]  &= (\lambda p. \Xi[p] \, n) (\lambda x. \Theta[x] \, n)) \\
+P[n]  &= (\lambda p. \Xi[p] \, n) (\lambda x. \Theta[x] \, n) \\
 \Xi[p]  &= \tY \lambda f n. \tif{n = 0}{\underline 0}{p \, n + f (n-1)} \\

### 2021-12-17 LICS21/supermartingales.tex
Mostly minor fixes
Things caught in proofreading. The one major error I found was in
the definition of CBV order of reductions in the appendix.

@@
 and the filtration $(\mathcal{F}_n)_{n \geq 0}$ where $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
-Thus $T_M$ is the \defn{runtime} of $M$ (and $M$ is AST iff $\mu_{\entrosp}(T_M < \infty) = 1$); we say that $M$ is \defn{positively almost surely terminating} (\defn{PAST}) if $\expect{T_M} < \infty$.
+Thus $T_M$ is the \defn{runtime} of $M$ (and $M$ is AST iff $\mu_{\entrosp}(\{T_M < \infty\}) = 1$); we say that $M$ is \defn{positively almost surely terminating} (\defn{PAST}) if $\expect{T_M} < \infty$.
 
@@
 \begin{enumerate}
-\item Every $\transform{M}$-reduction terminates on reaching $r$, or a term of the shape $E[\bot \, N_1 \cdots N_n]$.
+\item Every $\transform{M}$-reduction terminates on reaching a value, or a term of the shape $E[\bot \, N_1 \cdots N_n]$.
 
@@
 \end{enumerate}
-%This concludes the proof of Lemma~\ref{lem:TnBounded}.
-%The random variable $T_M$ is a stopping time w.r.t.~filtration $(\calF_n)_{n \geq 0}$. 
-%Since $f(M_n(s)) = 0$ iff $M_n(s)$ is a value, $T_M$ is the \emph{runtime of $M$}.
+From there, all the other stopping times $T_n$ can be shown to be bounded using a similar argument by induction.
+There are finitely many possible values of $\transform{M_{T_n+1}}$, and for each of these possibilities, there is a finite bound on $T_{n+1}$ by the same argument as before, therefore $T_{n+1}$ is bounded by the maximum of these.
+
+The event $\{T_n = k\}$ is equal to the event \{exactly $n$ of the terms $M_0, \dots, M_{k-1}$, and $M_k$ itself, are values or of the form $E[\tY \lambda x.N]$\}, which depends only on $M_0, \dots, M_k$ and is therefore in $\mathcal F_k$, therefore $T_n$ is a stopping time adapted to $(\mathcal F_n)_{n \geq 0}$.
 \end{proof}
